### 2020-06-01 docs/article_bigsimr.tex
created article skeleton

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={GIVE ME A TITLE},
  pdfauthor={A.G. Schissler; A.B. PERSON},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage[]{natbib}
\bibliographystyle{apalike}

\title{GIVE ME A TITLE}
\author{A.G. Schissler \and A.B. PERSON}
\date{Monday, June 01, 2020}

\begin{document}
\maketitle

{
\setcounter{tocdepth}{3}
\tableofcontents
}
\hypertarget{prerequisites}{%
\section{Prerequisites}\label{prerequisites}}

This is a \emph{sample} book written in \textbf{Markdown}. You can use anything that Pandoc's Markdown supports, e.g., a math equation \(a^2 + b^2 = c^2\).

The \textbf{bookdown} package can be installed from CRAN or Github:

Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading \texttt{\#}.

To compile this example to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): \url{https://yihui.name/tinytex/}.

  \bibliography{bibliography.bib,packages.bib}

\end{document}

### 2020-06-01 index.Rmd
created article skeleton

--- 
title: "GIVE ME A TITLE"
author: [A.G. Schissler, A.B. PERSON]
date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: article
bibliography: [bibliography.bib, packages.bib]
biblio-style: apalike
link-citations: yes
---

# Prerequisites

This is a _sample_ book written in **Markdown**. You can use anything that Pandoc's Markdown supports, e.g., a math equation $a^2 + b^2 = c^2$.

The **bookdown** package can be installed from CRAN or Github:

```{r eval=FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")
```

Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`.

To compile this example to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): <https://yihui.name/tinytex/>.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

### 2020-06-01 docs/article_bigsimr.tex
added references section

@@
 \setcounter{secnumdepth}{5}
+\usepackage{booktabs}
+\usepackage{amsthm}
+\makeatletter
+\def\thm@space@setup{%
+  \thm@preskip=8pt plus 2pt minus 4pt
+  \thm@postskip=\thm@preskip
+}
+\makeatother
 \usepackage[]{natbib}
@@
 \maketitle
+\begin{abstract}
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Vestibulum laoreet orci non elit fermentum mollis vel sit amet dolor.
+Proin ullamcorper, diam ac consectetur molestie, urna urna aliquet augue, sit amet vulputate felis massa ut libero.
+Integer mattis eget lorem vel viverra.
+Maecenas consequat feugiat mauris vitae porta.
+Mauris nec vulputate libero.
+Sed a turpis eget nisl tempus ultricies.
+Sed accumsan at augue a bibendum.
+Integer blandit, sem non imperdiet pharetra, tellus augue lobortis est, eget vestibulum massa ex eget purus.
+Suspendisse eleifend ullamcorper elit vitae pulvinar.
+Praesent vel tristique diam, vitae tempus leo.
+Nam placerat, massa a sodales porttitor, mauris felis ullamcorper eros, ut mattis lectus nisi vitae leo.
+Proin justo ante, fringilla quis consectetur at, euismod eget augue.
+In in nisi quis nunc fermentum semper ut a velit.
+Quisque suscipit dui in auctor tempor.
+Integer eu dui dictum, laoreet urna nec, tristique risus.
+Ut non tellus in metus.
+\end{abstract}
 
@@
 
+This is an example of a reference without braces \citet{xie2015} and a reference with braces \citep{xie2015}
+
   \bibliography{bibliography.bib,packages.bib}

### 2020-06-01 index.Rmd
added references section

@@
 link-citations: yes
+abstract: |
+  Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Vestibulum laoreet orci non elit fermentum mollis vel sit amet dolor.
+Proin ullamcorper, diam ac consectetur molestie, urna urna aliquet augue, sit amet vulputate felis massa ut libero.
+Integer mattis eget lorem vel viverra.
+Maecenas consequat feugiat mauris vitae porta.
+Mauris nec vulputate libero.
+Sed a turpis eget nisl tempus ultricies.
+Sed accumsan at augue a bibendum.
+Integer blandit, sem non imperdiet pharetra, tellus augue lobortis est, eget vestibulum massa ex eget purus.
+Suspendisse eleifend ullamcorper elit vitae pulvinar.
+Praesent vel tristique diam, vitae tempus leo.
+Nam placerat, massa a sodales porttitor, mauris felis ullamcorper eros, ut mattis lectus nisi vitae leo.
+Proin justo ante, fringilla quis consectetur at, euismod eget augue.
+In in nisi quis nunc fermentum semper ut a velit.
+Quisque suscipit dui in auctor tempor.
+Integer eu dui dictum, laoreet urna nec, tristique risus.
+Ut non tellus in metus.
 ---
@@
 
-Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`.
+Remember each Rmd file contains one and only one chapter/section, and a chapter is defined by the first-level heading `#`.
 
 To compile this example to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): <https://yihui.name/tinytex/>.
+
+This is an example of a reference without braces @xie2015 and a reference with braces [@xie2015]
 

### 2020-06-02 index.Rmd
insert abstract

--- 2020-06-01 index.Rmd
@@
 link-citations: yes
+abstract: |
+  In this era of Big Data, it is critical to realistically simulate data to conduct informative Monte Carlo studies.
+This is problematic when data are inherently multivariate while at the same time are (ultra-) high dimensional.
+This situation appears frequently in observational data found on online and in high-throughput biomedical experiments (e.g., RNA-sequencing).
+Due to the difficulty in simulating realistic correlated data points, researchers often resort to simulation designs that posit independence --- greatly diminishing the insight into the empirical operating characteristics of any proposed methodology.
+Major challenges lie in the computational complexity involved in simulating these massive random vectors.
+We propose a fairly general, scalable procedure to simulate high-dimensional multivariate distributions with pre-specified marginal characteristics and dependency characteristics.
+As a motivating example, we use our methodology to study large-scale statistical inferential procedures applied to cancer-related RNA-sequencing data sets.
+The proposed algorithm is implemented as the `bigsimr` R package.
 ---

### 2020-06-02 index.Rmd
insert working title

@@
 --- 
-title: "GIVE ME A TITLE"
-author: [A.G. Schissler, A.B. PERSON]
+title: "Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package"
+author: [A.G. Schissler, A. Knudson]
 date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"
@@
 ```{r eval=FALSE}
-install.packages("bookdown")
+## install.packages("bookdown")
 # or the development version
@@
 ```{r include=FALSE}
+## use the pandoc bundled with R Studio
+## rmarkdown::find_pandoc()
 # automatically create a bib database for R packages

### 2020-06-02 index.Rmd
organize refs II

@@
 documentclass: article
-bibliography: [bibliography.bib, packages.bib]
+bibliography: [bigsimr.bib, packages.bib]
 biblio-style: apalike
@@
 
-```{r eval=FALSE}
-## install.packages("bookdown")
+```{r eval=FALSE, echo=FALSE}
+install.packages("bookdown")
 # or the development version
@@
 ```
-
-Remember each Rmd file contains one and only one chapter/section, and a chapter is defined by the first-level heading `#`.
-
-To compile this example to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): <https://yihui.name/tinytex/>.
-
-This is an example of a reference without braces @xie2015 and a reference with braces [@xie2015]
 
@@
 ```
+
+```{bash copyBib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
+cp /Users/alfred/Dropbox/bib/bigsimr.bib ./.
+```
+
+Kendall's tau is great [@K38].

### 2020-06-02 index.Rmd
insert outline from esco2020_talk

@@
 
-# Prerequisites
+```{r LoadLib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
+library(ggplot2)
+library(tidyverse)
+library(knitr)
+library(dplyr)
+```
 
-This is a _sample_ book written in **Markdown**. You can use anything that Pandoc's Markdown supports, e.g., a math equation $a^2 + b^2 = c^2$.
-
-The **bookdown** package can be installed from CRAN or Github:
-
-```{r eval=FALSE, echo=FALSE}
-install.packages("bookdown")
-# or the development version
-# devtools::install_github("rstudio/bookdown")
+```{bash copyBib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
+cp /Users/alfred/Dropbox/bib/bigsimr.bib ./.
 ```
@@
 ```{r include=FALSE}
-## use the pandoc bundled with R Studio
-## rmarkdown::find_pandoc()
 # automatically create a bib database for R packages
@@
 
-```{bash copyBib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
-cp /Users/alfred/Dropbox/bib/bigsimr.bib ./.
+# Introduction
+
+Kendall's tau is great [@K38].
+
+Intergene correlation [@Schissler2019].
+
+# Multivariate correlation bounds
+
+# Simulation methodology
+
+# The `bigsimr` R package
+
+Short description and key features.
+
+```{r setup, eval=TRUE}
+## Using conda
+reticulate::use_condaenv("py37")
+library(bigsimr)
 ```
 
-Kendall's tau is great [@K38].
+## Basic use
+
+short description
+
+----
+### Specifying marginals
+
+As stated earlier, to generate multivariate data, we need a list of marginals (and their parameters), and a correlation structure (matrix).
+The marginal distributions can be built up as a list of lists, where each sublist contains the information for the target distribution.
+
+```{r}
+margins = list(
+  list("norm", mean = 3.14, sd = 0.1),
+  list("beta", shape1 = 1, shape2 = 4),
+  list("nbinom", size = 10, prob = 0.75)
+)
+```
+
+The things to point out here are that in each sublist (marginal), the first item is an unnamed character string with the R name of the distribution *without a letter prefix*.
+E.g.
+instead of `rnorm`, we pass in just `"norm"`.
+The second thing to note is that the remaining items are *named* arguments that go along with the distribution.
+A full list of built-in distributions is found in the appendix.
+
+----
+### Specify correlation
+
+The next step is to define a correlation structure for the multivariate distribution.
+This correlation matrix can either come from observed data, or we can set it ourselves, or we can generate a random correlation matrix via `bigsimr::rcor`.
+Let's create a simple correlation matrix where all off-diagonal elements are 0.5.
+Since we have 3 marginals, we need a $3\times 3$ matrix.
+
+```{r}
+rho <- matrix(0.5, nrow = 3, ncol = 3)
+diag(rho) <- 1.0
+rho
+```
+
+Finally we can generate a random vector with our specified marginals and correlation structure.
+The last argument, `type`, is looking to know what kind of correlation matrix it is receiving.
+Right now it can handle Pearson, Spearman, or Kendal.
+
+----
+### Small sample
+
+```{r}
+x = rvec(10, rho = rho, params = margins, type = "pearson")
+```
+
+On my machine, there is no dedicated GPU, so I would see the following warning message once per session.
+
+```{r, echo=FALSE}
+warning("warning.warn('No GPU/TPU found, falling back to CPU.')")
+```
+
+Taking a look at our random vector, we see that it hs 10 rows and 3 columns, one column for each marginal.
+
+```{r}
+x
+```
+
+We can simulate many more samples and then check the histogram of each margin, as well as the estimated correlation between the columns.
+
+----
+### Scaling up N
+
+```{r, fig.width=7}
+x = rvec(10000, rho = rho, params = margins, type = "pearson")
+
+par(mfrow=c(1,3))
+hist(x[,1], breaks = 30, xlab = "", main = "Normal")
+hist(x[,2], breaks = 30, xlab = "", main = "Beta")
+hist(x[,3], breaks = 30, xlab = "", main = "Negative Binomial")
+```
+
+----
+### Evaluation
+
+```{r}
+cor(x)
+```
+
+We can see that even wih 10,000 samples, the estimated correlation of the simulated data is not exactly the same as the target correlation.
+This can be explained by the fact that some correlations are simply not possible due to the discrete nature of certain distributions.
+Another possibility is that the copula algorith is biased and needs correction.
+
+----
+### List of supported distributions
+
+```{r, eval=FALSE}
+all_dists <- list(
+  list(dist = "beta", shape1, shape2),
+  list(dist = "binom", size, prob),
+  list(dist = "cauchy", location, scale),
+  list(dist = "chisq", df),
+  list(dist = "exp", rate),
+  list(dist = "f", df1, df2),
+  list(dist = "gamma", shape, rate),
+  list(dist = "geom", prob),
+  list(dist = "hyper", m, n, k),
+  list(dist = "logis", location, scale),
+  list(dist = "lnorm", meanlog, sdlog),
+  list(dist = "nbinom", size, prob),
+  list(dist = "norm", mean, sd),
+  list(dist = "pois", lambda),
+  list(dist = "t", df),
+  list(dist = "unif", min, max),
+  list(dist = "weibull", shape, scale),
+  list(dist = "wilcox", m, n),
+  list(dist = "signrank", n)
+)
+```
+
+----
+## Comparsions to other software
+
+----
+### Speed
+
+- `copula`
+- `Genord`
+- `Multiord`
+- other?
+- CPU benchmarking
+- GPU benchmarking
+
+----
+### Accuracy
+
+blah
+
+----
+### Other comparisons
+
+blah
+
+# Some applications
+
+## Simulation-based correlation checking
+
+## Simulating RNA-seq data
+
+## Simulation-based correlation testing
+
+## Simulation-based probability estimation
+
+# Conclusions

### 2020-06-02 index.Rmd
draft introduction

@@
 library(dplyr)
+set.seed(06082020)
 ```
@@
 
-Kendall's tau is great [@K38].
+Massive high-dimensional data sets are commonplace in many areas of scientific inquiry.
+As new methods are developed for these data, a fundamental challenge lies in designing and conducting simulation studies to assess operating characteristics of proposed methodology, such as false positive rates, statistical power, and robustness --- often in comparison to existing methods.
+Another application lies in the discovering the sampling distribution of test statistic under hypothesized null models in complex scenarios.
+Such Monte Carlo studies in the high-dimensional setting often become difficult to conduct with existing algorithms and tools, or require expertise in computing on clusters.
+This is particularly true when simulating massive multivariate, non-normal distributions.
+
+Along those lines, simulating high-dimension dependent negative binomial data motivates this work --- in pursuit of modeling RNA-sequencing data [@Wang2009b; @Conesa2016b] derived from breast cancer patients.
+This labroratory procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions.
+Often researchers posit a negative binomial model (refs) as counts are often over (or under) expressed that a Poisson model would suggest.
+Yet due to inherent biological processes, gene expression data exhibits correlation (coexpression) across genes [@BE07; @Schissler2018].
+RNA-sequencing analysis has garnered much interest in the statistical literature (refs).
+Often researchers simulate independently or from their proposed model in order condunct Monte Carlo studies (refs).
+An alternavitive strategy is to think *generatively* about the data collection process and scientific domain knowledge and design simulations with probabilitistic models that reflect those processes.
+In our motivating example, we'll study the consequences correlation in the simulation study replicating test statistics from an illustrative, classical large-scale hypothesis testing study [see @Efron2004].
+
+With our application in mind and seeking a general-purpose algorithm with broad applications, our purposed methodology should possess the following properities (adapted from criteria from @Nik13a):
+
+* P1: Wide range of dependence, allowing both positive and negative dependence
+* P2: Flexible dependence, meaning that the number of bivariate marginals is (approximately) equal to the number of dependence parameters.
+* P3: Flexible marginal modeling, generating hetereogenous data --- possibly from differing probability families.
+* P4: Computationally feasible cumulative distribution function (cdf) for likelihood estimation.
+
+Moreover, in simulation method must scale to high dimensions:
+
+* S1: Procedure must scale to high dimensions, computable in a reasonable amount time.
+* S2: Procedure must scale to high dimensions while maintaining accuracy.
+
+As others @MB13 have noted, however, simulating dependent data can be challenging.
+A central issue lies characterizing dependency between components in the high-dimensional random vector.
+The choice of correlation in typical practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g.
+non-normal, discrete, finite support, etc).
+Here we propose a scheme that ignores the analytic goal --- and, instead, aims to match the empirically-derived estimated correlation data and marginal characteristics.
+For normal data, the Pearson product-moment correlation describes the dependency completely.
+As we will see, however, simulating abrirtary random vectors with a given Pearson correlation matrix is computationally intense [@Chen2001, @Xia17], violating property **S1**.
+On the other hand, an analyst may consider the use of non-parametric correlation measures, such as Spearman's $\rho$ and Kendall's $\tau$, particularly when modeling non-normal data as in our application.
+Adding to the complexity is the well-known, marginal-dependent constraints on the possible bivariate correlation (a consequence of the bounds of bivariate distribution functions --- the *Frechet-Hoeffding bounds* [@Nelsen2007]).
+Denote the pointwise upper bound as $M(y_i, y_{i^\prime})$ These bounds give strict restraints on the possible bounds and cannot be overcome through algorithm design.
+Yet not all multivariate simulation approaches obtain these bounds [for example REF].
+Computationally, algorithms that rely on serial computation scale poorly to high dimensions (refs - hierarchy) and fail to make use of modern high-performance computing including parallelization and graphical processing unit acceleration [@Li2019gpu].
+
+To meet the desired properties in the face of the challenging setting, we present a general-purpose, scalable multivariate simulation algorithm.
+The crux of the method lies in the construction of a Gaussian copula [refs].
+The idea that many multivariate and marginally heterogenous distribution can be constructed in such a manner is well known [refs].
+This article's contriubition lies in it's application to high-dimenisonal data through a high-performance implementation (`bigsimr`) and speed-focused algorithm design.
+The algorithm design relies on useful properties of non-standard correlation measures, namely Kendall's $\tau$ and Spearman's $\rho$.
+We'll show that quality of simulation does not require the use of the standard Pearson correlation matrix, even for normally distributed data.
+The purpose of our method is to generate random vectors that match exactly specified any possible Kendall's $\tau$ or Spearman's $\rho$, and approximately Pearson product-moment correlation.
+
+The study proceeds with more details on our motivating example in RNA-sequencing breast cancer data.
+Then we describe and justify our simulation methodology, followed by extensive Monte Carlo studies under various distributional assumputions --- emphasizing normal and non-normal scenarios.
+In these studies, we evaluate the accuracy and speed of our approach in comparison to other readily availble software.
+After evaluating *in silico*, we revisit our motivating
+example by employing our method and compare our simulations to empirically-derived statistics. Finally, we'll make concluding remarks regarding the method's utility and future directions.
 

### 2020-06-02 index.Rmd
removed horizontal rules from talk
insert first equation
outline and add lesser sections

@@
 
+## General algorithm
+
+## Rank-based dependence measures
+
+## Pearson correlation
+
+Difficulties
+
 # The `bigsimr` R package
 
-Short description and key features.
+Short description and key features, including GPU acceleration.
 
@@
 
-----
 ### Specifying marginals
@@
 
-----
 ### Specify correlation
@@
 
-----
 ### Small sample
@@
 
-----
 ### Scaling up N
@@
 
-----
 ### Evaluation
@@
 
-----
 ### List of supported distributions
@@
 
-----
 ## Comparsions to other software
 
-----
 ### Speed
@@
 
-----
 ### Accuracy
@@
 
-----
 ### Other comparisons
@@
 
+A multivariate approach?
+
+Or all the pairs?
+
+Generate, sort, and correlate algorithm @Demirtas2011
+
+```{r, computeFrechetBounds, echo = TRUE, eval=TRUE}
+
+## simulation based
+## devtools::install_github("adknudson/bigsimr", ref="develop")
+library(bigsimr)
+## ?rvec
+
+d <- 5
+rho <- rcor(d)
+margins <- list(
+    list("nbinom", size = 5, prob = 0.3),
+    list("exp", rate = 4),
+    list("binom", size = 5, prob = 0.7),
+    list("norm", mean = 10, sd = 3),
+    list("pois", lambda = 10)
+)
+
+margins2 <- list(
+    list("nbinom", 5, 0.3),
+    list("exp", 4),
+    list("binom", 5, 0.7),
+    list("norm", 10, 3),
+    list("pois", 10)
+)
+
+params = margins
+rho = rho
+## cores = parallel::detectCores() - 1
+cores = 2
+reps = 1e3
+type = "spearman"
+
+## help(package="bigsimr")
+rho_bounds <- computeCorBounds(params = margins, cores = cores, type = "spearman", reps = reps)
+rho_bounds
+```
+
 ## Simulating RNA-seq data
+
+We apply our methodology to simulate RNA-sequencing data sets based on samples derived from breast cancer patients' tumors (the BRCA data set in TCGA).
+The data are freely available as part of the BRCA data set within the The Cancer Genome Atlas data warehouse and were downloaded on BLAH using the Harvard's Broad Institute's Firehouse interface.
+For each of the 1093 patients, the gene expression (abundance of messenger-RNA) from 20501 genes were counted using RNA-sequencing.
+The resultant data set can be represented as a matrix of size 1093 x 20501 with discrete values aligned to HUGO gene symbols for each patient.
+In this illustrative case study, we model the genewise marginal distributions as hetereogeneous negative binomial with pmf given by Equation \@ref(eq:nb) with two parameters, $p$ and $r$ (OR SWITCH TO mu and delta).
+Modeling RNA-seq counts as negative binomially distributed random variables is commonplace (e.g., @Zhao2018) since overdispersion is often observed for most expressed genes.
+We begin by estimating the marginal negative binomial parameters for each group using a simple method of moments approach by matching the negative binomial parameters with the sample mean and variance.
+Table/Figure XXX summarizes the estimated means, variances, and NB parameters as described in Equation \@ref(eq:nb).
+Notably some genes show underdisperation compared to a Poisson random varible.
+For simplicity we exclude these cases, but nothing in the later described simulation method prevents using a different marginal distribution in those cases (such as a Maxwell distribution REF Sellers).
+
+\begin{equation} 
+  g(n) = \mathbb P(Y=n) = \frac{\Gamma(n+r)}{\Gamma(r)n!} p^r(1-p)^n, \,\,\, n\in
+\mathbb N_0.
+  (\#eq:nb)
+\end{equation}
 
@@
 
-## Simulation-based probability estimation
+## Simulation-based probability calculations
+
+# Conclusion/Discussion
+
+- Although we recommend to use the rank-based measures
+- discuss the omission (or inclusion) of high dimensional covariance estimators?
+- We've shown that the multivariate NB simulations outperform the Independent sims.
+But the agreement is still poor.
+This could motivate other more appropriate models for high-expressing RNA-seq data than the negative binomial distribution.
+This algorithm lends itself to exploration of flexible probability models to inspire model selection in the developemnt novel RNA-seq analytic methodology.
+
+We've introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a high-performance implementation called `bigsimr` (github url).
+The parallelized (multi-core) algorithm is simple and easy to understand as an application of Gaussian copulas.
+This method is largely inspired by @MB13, but with contributions with regard to scaleability, implementation, and application in high-throughput biomedical data (RNA-sequencing).
+We advocate the use of Kendall's $\tau$ as it better captures correlation among components of non-normal, as well as non-linear patterns of association.
+Moreover, the matching Kendall's $\tau$ is nearly trival due to the invariant of monotone transformations, after an adjustment to the input correlation matrix.
+Interestingly, our simulation studies show the use of Kendall's $\tau$ to works as well as using Pearson correlation coefficient when simulating from multivariate normal distributions.
+
+We also show utility in our methodology through an application to differential gene expression analysis from RNA-sequencing data.
+The application results show that correlations indeed matter in the large-scale hypothesis testing, as many others have noted (for example, see @BE07, @Wu2012b).
+We also hope that the application provides an example workflow --- and other strategy to use in simulation design.
+Even the best-performing simulations we provide show a gap from the empirical distribution.
+To keep the demonstration straightforward, we did not attempt to match the empirical distribution of test statistics as precisely as possible.
+Yet our methodology could be more creatively applied to meet that goal.
+One could consider marginal distributions from different families, such as Poisson for a subset of genes.
+One could imagine finding a best fitting probability distribution among a class of distributions for each gene and this could perhaps a better fit.
+One could imagine additonal structures/features in the data that an analyset could model to improve the correspondence with the emprical values, for example row correlations and high-dimensional covariance estimators (see @Won2013g).
+
+There are of course limitations to the methodology and implementation.
+The algorithm requires invertible marginal cdfs.
+This leaves some restriction to the available marginal probability distributions (DOES IT?
+FOR EXAMPLE?) An additional source of error is after converting dependencies (step i; ${\bf R_{Kendall}})$) the resultant matrix may not be positive definite (PD).
+Then the nearest PD matrix is used.
+From a practical computing standpoint, the user's computing resource provides the ultimate limit on how large a dimension can be simulated using the `bigsimr` R package.
+A user must consider carefully the available memory and cores when conducting a Monte Carlo experiment.
+The algorithm does amends itself well to parallelization and graphical processing unit acceleration [@Li2019gpu] and advanced users may take advantage of those techniques.
+
+Future work includes developing scalable algorithms to match the Pearson correlation matrix exactly.
+Further, more sophicated approaches could involve simulation algorithms that are "estimation awarene" and so could explore the role of covariance estimation within the simulation.
+Lastly, these may reveal new approaches to estimating and evaluating high dimensional regression and Bayesian models.
+On the implementation side, employing graphical process unit acceleration could produce substaintial speedups over our multi-core approach.
 
 # Conclusions
+
+# Supplementary Materials
+
+We provide an open-source implementation of our methology as the `bigsimr` R package, hosted on github.
+
+# Acknowledgement(s)
+
+The authors gratefully acknowledge the helpful discussions with Professor Walter W. Piegorsch and Professor Edward J. Bedrick during this project's conception.
+
+# Disclosure statement
+
+The authors report no conflict of interest. DO WE?
+
+# Funding
+
+FUNDING FROM EVERYONE HERE. CP3 grant.
+
+# Nomenclature/Notation
+
+# Notes
+
+Mention Python?

### 2020-06-02 index.Rmd
format pdf in tf article

@@
 site: bookdown::bookdown_site
-output: bookdown::gitbook
+output: 
+  bookdown::gitbook:
+    base_format: rticles::tf_article
 documentclass: article

### 2020-06-04 notebooks/bivariateGamma.rmd
add notebooks for temporary investigations

---
title: Bivariate Gamma Correlation Bounds Comparison
author: |-
    A. Grant Schissler\
    Department of Mathematics & Statistics\
    The University of Nevada, Reno. \
    aschissler@unr.edu
date: '`r Sys.Date()`'
output:
    html_document:
      toc: yes
      toc_depth: 2
bibliography: /Users/alfred/Dropbox/bib/library.bib
abstract:
    Quick study checking the `bigsimr` Pearson matching scheme while checking whether the methods obtain pairwise Frechet bounds.
---

```{r LoadLib, echo = TRUE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
library(ggplot2)
library(tidyverse)
library(knitr)
library(dplyr)
knitr::opts_chunk$set(cache=TRUE, autodep=TRUE, cache.comments=TRUE)
set.seed(06032020)
library(bigsimr)
CORES <- parallel::detectCores() - 1
## CORES <- 1
## devtools::install_github( "superdesolator/NORTARA" )
library(nortaRA)
```

# Configure global parameters

```{r, globalParams}
## bivariate gamma
d <- 2
shape1 <- 5
shape2 <- 5
margins <- list(
    list("gamma", shape = shape1),
    list("gamma", shape = shape2)
)

## for evaluation of obtaining the bounds
gridSteps <- 1e2
N <- 1e5

## storing the results
boundsDat <- NULL
simRes <- NULL
```

# Compute bounds under each type of dependency

## Pearson correlation bounds

```{r, boundsPearson}
type <- 'pearson'
## Using GSC method of Demirtas2011
(tmpBounds <- computeCorBounds( params = margins, cores = CORES, type = type, reps = N ))
tmpBoundsDat <- data.frame( shape1 = shape1, shape2 = shape2, type = type, upper = tmpBounds$upper[ 1, 2 ], lower = tmpBounds$lower[ 1, 2 ] )
boundsDat <- rbind( boundsDat, tmpBoundsDat )
```

## Spearman correlation bounds

```{r, boundsSpearman}
type <- 'spearman'
## Using GSC method of Demirtas2011
(tmpBounds <- computeCorBounds( params = margins, cores = CORES, type = type, reps = N ))
tmpBoundsDat <- data.frame( shape1 = shape1, shape2 = shape2, type = type, upper = tmpBounds$upper[ 1, 2 ], lower = tmpBounds$lower[ 1, 2 ] )
boundsDat <- rbind( boundsDat, tmpBoundsDat )
```

## Kendall correlation bounds

```{r, boundsKendall, eval=FALSE}
type <- 'kendall'
## Using GSC method of Demirtas2011
(tmpBounds <- computeCorBounds( params = margins, cores = CORES, type = type, reps = N ))
tmpBoundsDat <- data.frame( shape1 = shape1, shape2 = shape2, type = type, upper = tmpBounds$upper[ 1, 2 ], lower = tmpBounds$lower[ 1, 2 ] )
boundsDat <- rbind( boundsDat, tmpBoundsDat )
```

# Simulating and checking the empirical bounds obtainment

## `bigsimr` simulations

### Pearson

```{r bigsimrPearson, eval = TRUE, results = 'hide'}
type <- 'pearson'
## now show that our simulation methods works
tmpLower <- boundsDat$lower[ boundsDat$type == type ]
tmpUpper <- boundsDat$upper[ boundsDat$type == type ]
rVec <- seq( tmpLower, tmpUpper, length.out = gridSteps)  ## loose approximation for now

## loop for simplicity see if fast enough
for (r in rVec) {
    print( paste0( 'Working on true rho of ', r ) )
    rho <- matrix(r, nrow = d, ncol = d)
    diag(rho) <- 1.0
    simData  <- as.data.frame( rvec(n = N, rho = rho, params = margins, type = type, cores = CORES, adjustForDiscrete = FALSE) )
    tmpSimData <- data.frame( shape1 = shape1, shape2 = shape2, trueRho = r, estRho = cor(simData, method = type)[ 1, 2], method = 'bigsimr' , type = type)
    simRes <- rbind( simRes, tmpSimData )
}
```

### Spearman

```{r bigsimrSpearman, eval = TRUE, results = 'hide'}
type <- 'spearman'
## now show that our simulation methods works
tmpLower <- boundsDat$lower[ boundsDat$type == type ]
tmpUpper <- boundsDat$upper[ boundsDat$type == type ]
rVec <- seq( tmpLower, tmpUpper, length.out = gridSteps)  ## loose approximation for now

## loop for simplicity see if fast enough
for (r in rVec) {
    print( paste0( 'Working on true rho of ', r ) )
    rho <- matrix(r, nrow = d, ncol = d)
    diag(rho) <- 1.0
    simData  <- as.data.frame( rvec(n = N, rho = rho, params = margins, type = type, cores = CORES, adjustForDiscrete = FALSE) )
    tmpSimData <- data.frame( shape1 = shape1, shape2 = shape2, trueRho = r, estRho = cor(simData, method = type)[ 1, 2], method = 'bigsimr' , type = type)
    simRes <- rbind( simRes, tmpSimData )
}
```

### Kendall

```{r bigsimrKendall, eval = FALSE}
type <- 'kendall'
## now show that our simulation methods works
tmpLower <- boundsDat$lower[ boundsDat$type == type ]
tmpUpper <- boundsDat$upper[ boundsDat$type == type ]
rVec <- seq( tmpLower, tmpUpper, length.out = gridSteps )

## loop for simplicity see if fast enough
for (r in rVec) {
    print( paste0( 'Working on true rho of ', r ) )
    rho <- matrix(r, nrow = d, ncol = d)
    diag(rho) <- 1.0
    simData  <- as.data.frame( rvec(n = N, rho = rho, params = margins, type = type, cores = CORES, adjustForDiscrete = FALSE) )
    tmpSimData <- data.frame( shape1 = shape1, shape2 = shape2, trueRho = r, estRho = cor(simData, method = type)[ 1, 2], method = 'bigsimr' , type = type)
    simRes <- rbind( simRes, tmpSimData )
}
```

## `NORTARA` simulations

```{r nortaraSetup}
## slightly different input format than bigsimr
invcdfnames <- c("qgamma","qgamma")
paramslists <- list(
    m1 = list(shape = shape1 ),
    m2 = list(shape = shape2 )
)
```

### Pearson

```{r nortaraPearson, eval = TRUE, warning=FALSE, results='hide'}
type <- 'pearson'
## now show that our simulation methods works
tmpLower <- boundsDat$lower[ boundsDat$type == type ]
tmpUpper <- boundsDat$upper[ boundsDat$type == type ]
esp <- 1e-3
rVec <- seq( tmpLower+esp, tmpUpper-esp, length.out = gridSteps ) 
## drop N for this one
tmpN <- 1e4
## loop for simplicity see if fast enough
for (r in rVec) {
    print( paste0( 'Working on true rho of ', r ) )
    rho <- matrix(r, nrow = d, ncol = d)
    diag(rho) <- 1.0
    simData  <- nortaRA::gennortaRA(n = tmpN, cor_matrix = rho, invcdfnames, paramslists)
    tmpSimData <- data.frame( shape1 = shape1, shape2 = shape2, trueRho = r, estRho = cor(simData, method = type)[ 1, 2], method = 'nortara' , type = type)
    simRes <- rbind( simRes, tmpSimData )
}
```

# Check the performance around the bounds

```{r obtainment}
## numerically compare to the bounds

```

# Visualize the results

```{r empVspecPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.path='fig/plot-', out.width= '75%', fig.align='center', fig.ext='pdf', fig.cap = 'Plot of the specified (true) versus empirical (computed) correlations for bivariate lognormal data with $p_1 = p_2 = 0.2$ generated by the standard approach, bigsimr, and T-Poisson. Empirical correlations are the estimated correlations from the simulated samples.'}
p0 <- ggplot(data = simRes, aes(x = trueRho, y = estRho, color = factor(method)) )
p0 + geom_point() + facet_grid(. ~ type ) + geom_abline(slope = 1, intercept = 0 )
```

### 2020-06-04 index.Rmd
Finish adapting old content and improve outline

@@
 
+## Ultra High Dimensional multivariate modeling and simulation desired properties
+
 With our application in mind and seeking a general-purpose algorithm with broad applications, our purposed methodology should possess the following properities (adapted from criteria from @Nik13a):
@@
 * P3: Flexible marginal modeling, generating hetereogenous data --- possibly from differing probability families.
-* P4: Computationally feasible cumulative distribution function (cdf) for likelihood estimation.
 
-Moreover, in simulation method must scale to high dimensions:
+Moreover, the simulation method must scale to high dimensions:
 
@@
 
-Intergene correlation [@Schissler2019].
+# Background and notation
 
-# Multivariate correlation bounds
+## Gaussian copulas
 
-# Simulation methodology
+We present an general-purpose, scalable multivaritae simulation algorithm.
+The crux of the method is construction of a Gaussian copula.
+This idea is well known [refs].
+A copula is a distribution function on $[0,1]^d$ describing a random vector with standard uniform marginals.
+Moreover, for any random vector ${\bf X}=(X_1, \ldots, X_d)$ with cumulative distribution function (CDF) $F$ and marginal CDFs $F_i$ there is a copula function
+$C(u_1, \ldots, u_d)$ so that
 
-## General algorithm
+\[
+F(x_1, \ldots,x_d) = \mathbb P(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), \,\,\, x_i\in \mathbb R, i=1,\ldots,d. 
+\]
 
-## Rank-based dependence measures
+A Gaussian copula is the case where all marginal CDFs $F_i$ are the standard normal cdf, $\Phi$.
+The Gaussian copula is the one that corresponds to a multivariate normal distribution with standard normal marginal distributions and covariance matrix ${\bf R}$.
+(Since the marginals are standard normal, this ${\bf R}$ is also the correlation matrix).
+If $F_{{\bf R}}$ is the CDF of such multivariate normal distribution, then the corresponding Gaussian copula $C_{{\bf R}}$ is defined through
 
-## Pearson correlation
+\begin{equation}
+(\#eq:gauss)
+F_{{\bf R}}(x_1, \ldots, x_d) = C_{{\bf R}}(\Phi(x_1), \ldots, \Phi(x_d)),
+\end{equation}
 
-Difficulties
+where $\Phi(\cdot)$ is the standard normal CDF.
+Note that the copula $C_{{\bf R}}$ is simply the CDF of the random vector $(\Phi(X_1), \ldots, \Phi(X_d))$, where  $(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R})$.
+
+Sklar's Theorem (ref) guarantee's that any random vector with computational feasible inverse CDFs (P4 above) and obtainable correlation matrix (within the Frechet bounds) can be obtained via transformations involving copula functions.
+Namely, to simulate a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$, we can construct a Gaussian copula ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$.
+When an $Y_i$ is discrete, some care must be taken to define $F_i$.
+Letting
+
+\begin{equation}
+F_{i}^{-1} = inf\{y:F_{i}(y) \geq u \}
+(\#eq:inverseCDF)
+\end{equation}
+
+\noindent ensures that $Y_i \sim F_i$.
+
+Simulation of a general multivariate random vector ${\bf T}$ with margins $F_i$ based on this copula is quite simple.
+Ensuring that one obtains a certian dependence among the marginal distributions, however, proves challenging in our proposed Gaussian-copula-based scheme.
+Two core issues: 1) marginal characteristics induce bounds on the possible bivariate correlations and 2) montone transformations deform the Pearson correlation which no closed form expression exists in general (ref?
+Chen2001).
+The Frechet bounds are well-known.
+
+## Measures of dependency
+
+The population correlation coefficient, commonly called the Pearson (product-moment) correlation coefficient, describes the linear association between two random variables $X$ and $Y$ and is given by
+
+\begin{equation}
+(\#eq:pearson)
+\rho(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ var(X)var(Y)\right]^{1/2}} 
+\end{equation}
+
+As @MB13 and @MK01 discuss, for a bivariate normal $(X,Y)$ random vector the Pearson correlation adequately describes the dependency between the components.
+$\rho$'s utility in non-normal or non-linear associations is lacking (@MK01).
+Rank-based (ordinal) approaches performance better in these settings, such as Spearman's (denoted $\rho_s$) and Kendall's $\tau$.
+Define
+
+\begin{equation}
+(\#eq:spearman)
+\rho_s(X,Y) = 3 \left[ P\left[ (X_1 - X_2)(Y_1-Y_3) > 0 \right] - P\left[ (X_1 - X_2)(Y_1-Y_3) < 0 \right] \right]
+\end{equation}
+
+\noindent where $(X_1, Y_1) \overset{d}{=} (X,Y), X_2 \overset{d}{=} X, Y_3 \overset{d}{=} Y$ with $X_2$ and $Y_3$ are independent of one other and of $(X_1, Y_1)$.
+For continuous marginals, the measure works well due to zero probability of ties.
+For discrete marginals, however, one could perform a rescaled version of $\rho_s$ for random variables $X,Y$ with pmfs (or pdfs) $p(x)$ and $q(y)$, respectively.
+
+\begin{equation}
+(\#eq:spearmanRescaled)
+\rho_{RS}(X,Y) = \frac{\rho_s(X,Y)}{ \left[ \left[ 1 - \sum_x p(x)^3 \right] \left[ 1 - \sum_y q(y)^3 \right] \right]^{1/2}}
+\end{equation}
+
+\noindent Kendall's $\tau$ is the probability of concordant pairs minus the probability of discordant pairs and is given compactly as
+
+\begin{equation}
+  (\#eq:tau)
+  \tau(X,Y) = \frac{\rho_s(X,Y)}{ \left[ \left[ 1 - \sum_x p(x)^3 \right] \left[ 1 - \sum_y q(y)^3 \right] \right]^{1/2}}
+\end{equation}
+
+## Marginal-dependent bivariate correlation bounds
+
+The pairwise correlation between two correlated random variables cannot in general obtain the full range of possibile values, $[-1,-1]$.
+The range, called the Frechet(-Hoeffding) bounds, is a well-known function of the marginal distributions and are given by [@BF17]:
+
+\begin{equation}
+(\#eq:frechet)
+\rho^{max} = \rho \left( F^{-1}_1 (U), F^{-1}_2 (U) \right), \quad \rho^{min} = \rho \left( F^{-1}_1 (U), F^{-1}_2 (1 - U) \right)
+\end{equation}
+
+\noindent where $U$ is a uniform random variable in $(0,1)$, and $F^{-1}_1, F^{-1}_2$ are the inverse cdf of random variables $X_1$ and $X_2$, respectively.
+For discrete random variables, define $F^{-1}$ as in Equation \@ref(eq:inverseCDF).
+
+# Simulation algorithm
+
+## Algorithm description
+
+This section describes the method for simulating a random vector $\bf Y$ with $Y_i$ components for $i=1,2,\ldots,d$. Each $Y_i$ has a specified marginal
+distribution function $F_i$ and its inverse.
+To characterize dependency, every pair $(Y_i, Y_j)$ has either a specified Pearson correlation \@ref(eq:pearson), (rescaled) Spearman correlation \@ref(eq:spearmanRescaled) , or Kendall's $\tau$ (Equation \@ref(eq:tau) ).
+The method only approximately matches the Pearson correlation in general, whereas the rank-based methods are exact.
+
+The method is best understand as a **parallelized Gaussian copula** (see Equation \@ref(eq:gauss).
+We shall see that contructing continuous joint distributions that match a target Spearman or Kendall's correlations computes easily when employing Gaussian copulas, since this measures are invariant under the monotone transformations involved [refs].
+To do this, we take advantage of a closed form relationship [ref?] between Kendall's $\tau$ and Pearson's correlation coefficient for bivariate normal random variables:
+
+\begin{equation}
+(\#eq:convertKendall)
+r_{Pearson} = sin \left( \tau_{Kendall} \times \frac{\pi}{2} \right), 
+\end{equation}
+
+\noindent and similarly for Spearman's $\rho$ [@K58],
+
+\begin{equation}
+(\#eq:convertSpearman)
+\rho_{Pearson} = 2 \times sin \left( \rho_{Spearman} \times \frac{\pi}{6} \right).
+\end{equation}
+
+For discrete marginals, achieving a target Spearman correlation under this scheme is possible by using components rom Equation \@ref(eq:spearmanRescaled) to further adjust the input correlation matrix.
+Let the unscaled Spearman correlation coefficients be $\rho_{s} \left(Y_{i}, Y_{i^\prime}\right)$ for two marginal distributions and divide the target correlation by the product in the denominator of Equation \@ref(eq:spearmanRescaled).
+Let these adjustment factors be denoted as $a_i = \left[ 1 - \sum_y p_i(y)^3 \right]^{1/2}$ and specifically rescale the target Spearman correlation matrix by
+
+\begin{equation}
+(\#eq:convertSpearmanDiscrete)
+\rho_{rs} \left(Y_{i}, Y_{i^\prime}\right) = \frac{\rho_{s} \left(Y_{i}, Y_{i^\prime}\right)}{a_i \times a_{i^\prime}}.
+\end{equation}
+
+In a similiar fashion, we rescale Kendall's $\tau$ to adjust the input correlation matrix. The conversion formula is given by
+
+\begin{equation}
+(\#eq:convertKendallDiscrete)
+\rho_{rs} \left(Y_{i}, Y_{i^\prime}\right) = \frac{\rho_{s} \left(Y_{i}, Y_{i^\prime}\right)}{a_i \times a_{i^\prime}}.
+\end{equation}
+
+In contrast the rank-based correlations, matching specified Pearson correlation coefficients exactly is computational intense in this scheme.
+In general, there is no closed form correspondence and involving computing or approximating $\binom{d}{2}$ integrals of the form $EY_iY_j = \int \int y_i y_j f_{X|r}(F_i^{-1}(\Phi(z_i)), F_j^{-1}(\Phi(z_j))dy_idy_j$, for $i,j=1,2,\ldots,d$.
+For accurate numeric approximation of these integrals, the functions must be evaluated hundreds of times.
+Others have used efficient Monte Carlo integration schemes (see @Chen2001), but scale poorly to large dimension in reasonable times (property **S2**).
+Despite all this, if one does desire to characterize dependency using Pearson correlations, we often see in practice --- and it is theoretically justified under certain conditions (@Song00) --- that simply using the target Pearson correlation matrix as the initial conditions to our proposed algorithm will lead to approximate matching in the resultant distribution.
+
+## Simulation Algorithm
+
+Putting the together the facts provided in the equations above, we come the following proposed simulation algorithm to produce a random vector ${\bf Y}$ with specified Spearman's correlation and marginal distributions.
+Note that all computational steps can be parallelized as each operation can be done or either the $d$ marginals or the $\binom{d}{2}$ pairs for the correlation values.
+Even the generation of multivariate normal random vectors is parallelized through optimized matrix multiplication/decomposition routines.
+
+### Inputs
+(1) Marginal characteristics including the same of distributional family and parameter values, denotes as $F_i$ for marginal component random variable $Y_i$ for $i=1,\ldots,d$.
+(2) A target (specified) Spearman's correlation matrix ${\bf R_{Spearman}}$ with
+each $\rho_{Spearman}$ within the Frechet limits. Alternatively, the rescaled
+Spearman's correlation (see Equation \@ref(eq:convertKendallDiscrete) ) can be provided.
+(3) An error tolerance $\epsilon$ for finding the nearest positive definite matrix a transformed correlation matrix (see step iii below). Alternatively, a maximum number of iterations can be supplied.
+
+### Algorithm
+(i) Compute Spearman's $\rho_{rs}$ (see Equation
+\@ref(eq:convertKendallDiscrete)) from the specified $\rho_{spearman}$ for each
+pair of marginal random variables $(Y_i,Y_{i^\prime})$ with $i \neq i^\prime$
+when either marginal is discrete. In such large scale computations, it may be
+that numerically $\rho_{rs}$ is be larger/smaller than the Frechet bounds (or
+even $\pm 1$). To guard aganist this, set $\rho_{rs} = min( \rho_{rs}, M)$,
+where $M$ is the upper Frechet bound (see [Section Introduction]) and similarly set $\rho_{rs} = max( \rho_{rs}, W)$, where $W$ is the lower Frechet bound for this pair of distributions.
+Gather these $\rho_{rs} \lq s$ into a new input correlation matrix ${\bf R_{rs}}$.
+(ii) Convert ${\bf R_{Spearman}}$ into ${\bf R_{Pearson}}$ via $\rho_{Pearson} = 2 \times sin \left( \rho_{rs} \times \frac{\pi}{6} \right)$.   
+(iii) Finally, ensure that ${\bf R_{Pearson}}$ is positive definite, by finding the nearest positive definite correlation matrix ${\bf R}$ --- using the `R` routine `nearPD` in the `Matrix` package --- within an error tolerance of $\epsilon$.
+(iv) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R})$;  
+(v) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$;  
+(vi) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;
 
@@
 
+The most computationally extensive lies in the second step of our algorithm. We
+use a optimized and parallelized multivariate normal simulator within the `R`
+package `mvnfast`. The rest of the code is also parallelized, but these steps
+run rapidly in serial computation except for extremely large dimension.
+
 ```{r setup, eval=TRUE}
@@
 library(bigsimr)
+## help(package = 'bigsimr')
 ```
@@
 
-### List of supported distributions
+# Monte Carlo evaluation and Comparsions to other software
 
-```{r, eval=FALSE}
-all_dists <- list(
-  list(dist = "beta", shape1, shape2),
-  list(dist = "binom", size, prob),
-  list(dist = "cauchy", location, scale),
-  list(dist = "chisq", df),
-  list(dist = "exp", rate),
-  list(dist = "f", df1, df2),
-  list(dist = "gamma", shape, rate),
-  list(dist = "geom", prob),
-  list(dist = "hyper", m, n, k),
-  list(dist = "logis", location, scale),
-  list(dist = "lnorm", meanlog, sdlog),
-  list(dist = "nbinom", size, prob),
-  list(dist = "norm", mean, sd),
-  list(dist = "pois", lambda),
-  list(dist = "t", df),
-  list(dist = "unif", min, max),
-  list(dist = "weibull", shape, scale),
-  list(dist = "wilcox", m, n),
-  list(dist = "signrank", n)
-)
-```
+Before applying our methodology to real data simulation, we conduct several Monte Carlo studies to investigate method performance in comparison to other existing implementations.
+We focus the numerical experiments on assessing how well the procedure scales to high dimension with respect to reasonable computation times (property S1 above) and accurately matching marginal and dependency paramaters.
+The simulations will proceed in increasing complexity --- leading up to the setting in our motivating example.
+We begin by exploring simple exchangeable (constant) correlation structures under a large number of simulation replicates while applying the algorithm above to first continous then discrete marginal distributions.
+We conclude the simulation studies with $10^5$ replicated data sets with sample sizes and distributional characteristics corresponding the estimates from the motivating example data.
 
-## Comparsions to other software
+The CPU-based computation times we report were from runs using eight dual-threaded 3.9GHz
+Xeon Gold processors on a linux workstation, allowing for parallel instructions
+over 16 cores. We choos 15 cores for our experiments.
 
-### Speed
+## Other multivariate simulation software
+
+Describe the competing software and algorithms briefly.
 
@@
 - `Multiord`
-- other?
-- CPU benchmarking
-- GPU benchmarking
+- `nortaRA` only does Pearson matching.
 
-### Accuracy
+## Simulation I: Bivariate simulation using rank-based correlation
 
-blah
+### Continuous example: Bivariate Exponential
 
-### Other comparisons
+Both Spearman's and Kendall's measures of dependency are invariant under
+strictly monotone transformations (ref). This in turn provides exact simulation
+of continuos marginals $F_i$ $i=1,\ldots,d$ since their corresponding quantile
+function are increasing montonically over their support. The results below
+provide computations times for $N=10^5$ under increasingly large dimension $d$.
+For simplicity we assume multivariate Normal with varying parameters and to
+additionally demonstrate the utility of Kendall's $\tau$ even in the Gaussian
+setting (as opposed to the nature choice of Pearson's correlation). To layer on complexity, we now assume a multivariate gamma distribution with fixed marginal parameters.
 
-blah
+Compare speed and accuracy.
+
+```{r tabGamma, echo = TRUE}
+## computation times normal
+
+```
+### Discrete example: Bivariate Geometric
+
+Compare speed and accuracy.
+
+## Simulation II: Ultra HD multivariate simulation with identical margins and exchangeable rank-based correlation.
+
+### UHD Continuous example: Multivariate Gamma
+
+Compare speed and accuracy.
+
+### UHD Discrete example: Multivariate Negative Binomial
+
+Compare speed and accuracy.
+
+## Simulation III: Ultra HD multivariate simulatino with heterogenous margins and exchangeable rank-based correlation.
+
+### UHD Heteregeous continuous example: Multivariate Gamma
+
+Compare speed and accuracy.
+
+### UHD Heteregeousdiscrete example: Multivariate Negative Binomial
+
+Compare speed and accuracy.
 
@@
 
-## Simulation-based correlation checking
+## Simulating Ultra High Dimensional RNA-seq data
 
-A multivariate approach?
+We apply our methodology to simulate RNA-sequencing data sets based on samples derived from breast cancer patients' tumors (the BRCA data set in TCGA).
+The data are freely available as part of the BRCA data set within the The Cancer Genome Atlas data warehouse and were downloaded on BLAH using the Harvard's Broad Institute's Firehouse interface.
+For each of the 1093 patients, the gene expression (abundance of messenger-RNA) from 20501 genes were counted using RNA-sequencing.
+The resultant data set can be represented as a matrix of size 1093 x 20501 with discrete values aligned to HUGO gene symbols for each patient.
+In this illustrative case study, we model the genewise marginal distributions as hetereogeneous negative binomial with pmf given by Equation \@ref(eq:nb) with two parameters, $p$ and $r$ (OR SWITCH TO mu and delta).
+Modeling RNA-seq counts as negative binomially distributed random variables is commonplace (e.g., @Zhao2018) since overdispersion is often observed for most expressed genes.
+We begin by estimating the marginal negative binomial parameters for each group using a simple method of moments approach by matching the negative binomial parameters with the sample mean and variance.
+Table/Figure XXX summarizes the estimated means, variances, and NB parameters as described in Equation \@ref(eq:nb).
+Notably some genes show underdisperation compared to a Poisson random varible.
+For simplicity we exclude these cases, but nothing in the later described simulation method prevents using a different marginal distribution in those cases (such as a Maxwell distribution REF Sellers).
 
-Or all the pairs?
+Intergene correlation [@Schissler2019].
 
-Generate, sort, and correlate algorithm @Demirtas2011
+\begin{equation} 
+  g(n) = \mathbb P(Y=n) = \frac{\Gamma(n+r)}{\Gamma(r)n!} p^r(1-p)^n, \,\,\, n\in
+\mathbb N_0.
+  (\#eq:nb)
+\end{equation}
 
-```{r, computeFrechetBounds, echo = TRUE, eval=TRUE}
+```{r brca, echo=TRUE, eval=FALSE}
+## full workflow simulating RNA-seq data
+
+```
+
+The data are overdispersed:
+
+```{r figNBratio, echo = FALSE, eval=FALSE, message=FALSE, warning=FALSE, fig.path='fig/plot-', out.width= '75%', fig.align='center', fig.ext='pdf', fig.cap="The genes display extreme overdispersion.
+Using group-specific sample means and variances for each of the d=4777 genes, we display the distributions oflog of variance to mean ratio (log(sigma2/mu))."}
+## http://www.sthda.com/english/wiki/ggplot2-histogram-easy-histogram-graph-with-ggplot2-r-package
+## devtools::install_github("kassambara/easyGgplot2")
+library(easyGgplot2)
+## p0 <- ggplot(data = nb_dat, aes(x = log(nb_var / nb_mean), y = stat(density), fill=group))
+## p0 + geom_histogram(bins = 30) + labs(x="log of variance to mean ratio") + theme_bw()
+nb_dat$logRatio <- log(nb_dat$nb_var / nb_dat$nb_mean)
+ggplot2.histogram(data=nb_dat, xName='logRatio',
+    groupName='group', legendPosition="right",
+    alpha=0.5, addDensity=TRUE,
+    addMeanLine=TRUE, meanLineColor="white", meanLineSize=1.5)
+```
+
+We see that the all genes studies display overdispersion (variance greater than mean) compared to a Poisson model for the gene counts.
+The genes are highly hetereogeneous even when dealing with highly expressed genes after filtering.
+The facts taken together motivate the negative binomial model for the marginal distributions.
+The genes are slightly more variable within the deceased group (expected with smaller sample size and potentially due to biomedical considerations).
+
+Next we perform some essential pre-processing of the data.
+Since our goal is to simulate meaningful multivariate constructions, we must first find a subset of 20501 genes that are expressed.
+In order words, we'll filter out the low expressing genes to allow greater range of possible $d$-variate correlations (see @NK10 for details OR SHOULD I EXPLAIN THIS MORE CAREFULLY?).
+Notably, many RNA-seq analytic workflows filter out low expressing genes (see \cite{Conesa2016b}) and this, admittedly, is usually done in an \emph {ad hoc} fashion.
+In this study, we prefer an inclusion criterion that is theoretically motivated.
+Based on the results in Section BLAH, we filter genes with a sample mean less than 12.
+In this way, we remove 4762 (23.2\%) of the genes, retaining 15,739 genes for further analysis.
+The filtered low-expressing genes will be simulated as independent negative binomial random variables to form a complete simulated transcriptome.
+
+RNA-sequencing data derived from breast cancer patients (n=1093), with 20501 genes measured (from TCGA).
+We filter to genes with average expression greater than 10000 counts (d = 4777) across both groups.
+Consider basic differentially expressed genes (DEG) analysis between surviving (n0 = 941) and deceased (n1 = 152) patients.
+This setting can be described as large-scale simultaneous hypothesis testing, under correlation.
+We should evaluate existing and new methodology using simulations that reflect dependency among variables.
+But multivariate simulation tools often scale Tpoorly to high dimension or do not model the full range of dependency.
+Perform a two-sample $t$ test for each 4777 genes on the observed RNA-seq counts ($empirical$).
+Call a "Z score" for the $i^{th}$ gene $Z_{i} = \Phi^{-1}(t_{\approx 272}(t_i))$.
+
+Our method requires some measure of that the covariance structure is pre-specified.
+In practice, however, high-dimensional covariance estimation is not that easy.
+Indeed there is much recent interest in this area (cite a few recent high visibility articles --- see the GPU-NORTA pub).
+Here we do not provide a comprehensive review on the topic.
+Instead we only seek to use covariance estimator that only guarantees postive-semidefiniteness while maintaining adequate estimation properties.
+To this end we chose to use the condition-number-regularized (\textsc{condreg}) approach of \cite{Won2013g}.
+This covariance estimation procedure restricts the ratio of the largest eigenvalue to the smallest eigenvalue to improve numerical results.
+But the method employs a penalized Gaussian likelihood.
+And so, rather than estimating the covariance on the discrete counts directly, we first perform a $log_{2}(x +1)$ transformation to better suit this modeling assumption.
+\textsc{condreg} requires a parameter, $\kappa_{max}$, that controls the largest condition number of the resultant estimated covariance matrix.
+Here we aim to allow a large number of nonzero pairwise correlations and so arbitrary set $\kappa_{max}=10^{4}$.
+The routine ran without issue on a MacBook Pro carrying BLAH in XXX units of time.
+
+Most pairwise correlations are nearly zero with XXX \% less than |0.1|.
+Yet there are dense subsets of correlated genes.
+And even small correlations among many variables can disrupt the operating characteristics of commonly used statistical inferential procedures \cite{Schissler2018}.
+
+In this case study in the large-scale hypothesis testing in pursuit of detecting differentially expressed genes, we aim to simulate RNA-seq counts using a multivariate negative binomial.
+To evaluate the simulation's utility, we systematically compare the simulated results to the empirical results.
+In short we computed t-statistics and p-values for each of $d=4777$ genes between two groups of breast cancer patients (1=deceased, 0=alive; $n_0 = 152, n_1=941$).
+Further details on the empirical results are provided in Section BLAH.
+We compare two aspects in the simulation design 1) Choice of correlation measure while modeling marginal distributions consistently and 2) how well does the best-performing simulation agree with the empirical results.
+Informed by the simulation studies in Section BLAH, we anticipate a far amount of variation at such small sample sizes and large $d$ and so replicate the entire study 100 times.
+
+To recap from the discussion above (Section BLAH), we estimated negative binomial parameters for each of the $d=4777$ genes and also their Pearson, Spearman, and Kendall's correlation coefficients from n=1093 breast cancer patients, within each vital status group (0=deceased, 1=alive).
+We generated 100 sythentic data sets for each group, with the number of $d$-dimensional random vectors produced equal to the corresponding sample sizes ($n_0 = 152, n_1=941$).
+These 100 samples are used to understand uncertainty in this setting and we will use the mean simulated values to compare to the empirical (and also explore the worst-case scenarios).
+
+Model genes as marginally negative binomial with heterogenous genewise parameters.
+Compute sample correlations (using desired dependency measure) for the 11,407,476 genes pairs, for each group.
+To simulate gene expression counts, our goal is to produce a random vector ${\bf Y}=(Y_1, \ldots, Y_d)$ with **correlated** NB components.
+To do that, we start with a sequence of **independent** Poisson processes $N_i(t)$, $i=1,\ldots,d$, where the rate of the process $N_i(t)$ is $\lambda_i=(1-p_i)/p_i>0$ (so that $p_i=1/(1+\lambda_i)$).
+Now, we let ${\bf T} = (T_1, \ldots T_d)$ have a multivariate distribution on $\mathbb R_+^d$ with the PDF $f_{{\bf T}}({\bf t})$.
+Then, we define
+
+We rescaled the target Spearman correlation to account for the probability of ties via Equation \@ref(eq:convertSpearmanDiscrete).
+The serial computation took approximately 10 minutes on XXX.
+Notably, the difference between target and adjusted matrices was very small (mean relative difference: 0.708) in this particular configuration of negative binomial parameters.
+For problems with smaller counts and higher probabilities of ties, the adjustment can be substantial (as observed in the simulation studies above in Section BLAH).
+
+- Kendall's $\tau$ was estimated efficently using the `pcaPP::cor.fk` function in `R`.
+
+## Ultra High-Dimensional simulation-based correlation hypothesis testing
+
+```{r brcaSigmaTesting, echo=TRUE, eval=FALSE}
+## full workflow simulating RNA-seq data
+
+```
+
+## Simulation-based computation of correlation bounds
+
+Using the Generate, sort, and correlate algorithm @Demirtas2011
+
+```{r, computeBounds, echo = TRUE, eval=FALSE}
 
@@
 
-margins2 <- list(
-    list("nbinom", 5, 0.3),
-    list("exp", 4),
-    list("binom", 5, 0.7),
-    list("norm", 10, 3),
-    list("pois", 10)
-)
-
 params = margins
@@
 ## help(package="bigsimr")
-rho_bounds <- computeCorBounds(params = margins, cores = cores, type = "spearman", reps = reps)
+rho_bounds <- computeCorBounds(params = margins, cores = cores, type = type, reps = reps)
 rho_bounds
@@
 
-## Simulating RNA-seq data
+## Simulation-based joint probability calculations
 
-We apply our methodology to simulate RNA-sequencing data sets based on samples derived from breast cancer patients' tumors (the BRCA data set in TCGA).
-The data are freely available as part of the BRCA data set within the The Cancer Genome Atlas data warehouse and were downloaded on BLAH using the Harvard's Broad Institute's Firehouse interface.
-For each of the 1093 patients, the gene expression (abundance of messenger-RNA) from 20501 genes were counted using RNA-sequencing.
-The resultant data set can be represented as a matrix of size 1093 x 20501 with discrete values aligned to HUGO gene symbols for each patient.
-In this illustrative case study, we model the genewise marginal distributions as hetereogeneous negative binomial with pmf given by Equation \@ref(eq:nb) with two parameters, $p$ and $r$ (OR SWITCH TO mu and delta).
-Modeling RNA-seq counts as negative binomially distributed random variables is commonplace (e.g., @Zhao2018) since overdispersion is often observed for most expressed genes.
-We begin by estimating the marginal negative binomial parameters for each group using a simple method of moments approach by matching the negative binomial parameters with the sample mean and variance.
-Table/Figure XXX summarizes the estimated means, variances, and NB parameters as described in Equation \@ref(eq:nb).
-Notably some genes show underdisperation compared to a Poisson random varible.
-For simplicity we exclude these cases, but nothing in the later described simulation method prevents using a different marginal distribution in those cases (such as a Maxwell distribution REF Sellers).
+```{r densityEvaluation, echo=TRUE, eval=FALSE}
+## full workflow simulating RNA-seq data
 
-\begin{equation} 
-  g(n) = \mathbb P(Y=n) = \frac{\Gamma(n+r)}{\Gamma(r)n!} p^r(1-p)^n, \,\,\, n\in
-\mathbb N_0.
-  (\#eq:nb)
-\end{equation}
-
-## Simulation-based correlation testing
-
-## Simulation-based probability calculations
+```
 
@@
 
-# Conclusions
-
 # Supplementary Materials
@@
 
+### List of supported distributions
+
+```{r, eval=FALSE}
+all_dists <- list(
+  list(dist = "beta", shape1, shape2),
+  list(dist = "binom", size, prob),
+  list(dist = "cauchy", location, scale),
+  list(dist = "chisq", df),
+  list(dist = "exp", rate),
+  list(dist = "f", df1, df2),
+  list(dist = "gamma", shape, rate),
+  list(dist = "geom", prob),
+  list(dist = "hyper", m, n, k),
+  list(dist = "logis", location, scale),
+  list(dist = "lnorm", meanlog, sdlog),
+  list(dist = "nbinom", size, prob),
+  list(dist = "norm", mean, sd),
+  list(dist = "pois", lambda),
+  list(dist = "t", df),
+  list(dist = "unif", min, max),
+  list(dist = "weibull", shape, scale),
+  list(dist = "wilcox", m, n),
+  list(dist = "signrank", n)
+)
+```
+
 Mention Python?

### 2020-06-04 010-background.Rmd
split up document into sections. did spelling check

--- 2020-06-02 index.Rmd
@@
---- 
-title: "Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package"
-author: [A.G. Schissler, A. Knudson]
-date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"
-site: bookdown::bookdown_site
-output: bookdown::gitbook
-documentclass: article
-bibliography: [bigsimr.bib, packages.bib]
-biblio-style: apalike
-link-citations: yes
-abstract: |
-  In this era of Big Data, it is critical to realistically simulate data to conduct informative Monte Carlo studies.
-This is problematic when data are inherently multivariate while at the same time are (ultra-) high dimensional.
-This situation appears frequently in observational data found on online and in high-throughput biomedical experiments (e.g., RNA-sequencing).
-Due to the difficulty in simulating realistic correlated data points, researchers often resort to simulation designs that posit independence --- greatly diminishing the insight into the empirical operating characteristics of any proposed methodology.
-Major challenges lie in the computational complexity involved in simulating these massive random vectors.
-We propose a fairly general, scalable procedure to simulate high-dimensional multivariate distributions with pre-specified marginal characteristics and dependency characteristics.
-As a motivating example, we use our methodology to study large-scale statistical inferential procedures applied to cancer-related RNA-sequencing data sets.
-The proposed algorithm is implemented as the `bigsimr` R package.
----
+# Background and notation
 
-```{r LoadLib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
-library(ggplot2)
-library(tidyverse)
-library(knitr)
-library(dplyr)
-set.seed(06082020)
-```
+## Gaussian copulas
 
-```{bash copyBib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
-cp /Users/alfred/Dropbox/bib/bigsimr.bib ./.
-```
+We present an general-purpose, scalable multivariate simulation algorithm.
+The crux of the method is construction of a Gaussian copula.
+This idea is well known [refs].
+A copula is a distribution function on $[0,1]^d$ describing a random vector with standard uniform marginals.
+Moreover, for any random vector ${\bf X}=(X_1, \ldots, X_d)$ with cumulative distribution function (CDF) $F$ and marginal CDFs $F_i$ there is a copula function
+$C(u_1, \ldots, u_d)$ so that
 
-```{r include=FALSE}
-# automatically create a bib database for R packages
-knitr::write_bib(c(
-  .packages(), 'bookdown', 'knitr', 'rmarkdown'
-), 'packages.bib')
-```
+\[
+F(x_1, \ldots,x_d) = \mathbb P(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), \,\,\, x_i\in \mathbb R, i=1,\ldots,d. 
+\]
 
-# Introduction
-
-Massive high-dimensional data sets are commonplace in many areas of scientific inquiry.
-As new methods are developed for these data, a fundamental challenge lies in designing and conducting simulation studies to assess operating characteristics of proposed methodology, such as false positive rates, statistical power, and robustness --- often in comparison to existing methods.
-Another application lies in the discovering the sampling distribution of test statistic under hypothesized null models in complex scenarios.
-Such Monte Carlo studies in the high-dimensional setting often become difficult to conduct with existing algorithms and tools, or require expertise in computing on clusters.
-This is particularly true when simulating massive multivariate, non-normal distributions.
-
-Along those lines, simulating high-dimension dependent negative binomial data motivates this work --- in pursuit of modeling RNA-sequencing data [@Wang2009b; @Conesa2016b] derived from breast cancer patients.
-This labroratory procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions.
-Often researchers posit a negative binomial model (refs) as counts are often over (or under) expressed that a Poisson model would suggest.
-Yet due to inherent biological processes, gene expression data exhibits correlation (coexpression) across genes [@BE07; @Schissler2018].
-RNA-sequencing analysis has garnered much interest in the statistical literature (refs).
-Often researchers simulate independently or from their proposed model in order condunct Monte Carlo studies (refs).
-An alternavitive strategy is to think *generatively* about the data collection process and scientific domain knowledge and design simulations with probabilitistic models that reflect those processes.
-In our motivating example, we'll study the consequences correlation in the simulation study replicating test statistics from an illustrative, classical large-scale hypothesis testing study [see @Efron2004].
-
-With our application in mind and seeking a general-purpose algorithm with broad applications, our purposed methodology should possess the following properities (adapted from criteria from @Nik13a):
-
-* P1: Wide range of dependence, allowing both positive and negative dependence
-* P2: Flexible dependence, meaning that the number of bivariate marginals is (approximately) equal to the number of dependence parameters.
-* P3: Flexible marginal modeling, generating hetereogenous data --- possibly from differing probability families.
-* P4: Computationally feasible cumulative distribution function (cdf) for likelihood estimation.
-
-Moreover, in simulation method must scale to high dimensions:
-
-* S1: Procedure must scale to high dimensions, computable in a reasonable amount time.
-* S2: Procedure must scale to high dimensions while maintaining accuracy.
-
-As others @MB13 have noted, however, simulating dependent data can be challenging.
-A central issue lies characterizing dependency between components in the high-dimensional random vector.
-The choice of correlation in typical practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g.
-non-normal, discrete, finite support, etc).
-Here we propose a scheme that ignores the analytic goal --- and, instead, aims to match the empirically-derived estimated correlation data and marginal characteristics.
-For normal data, the Pearson product-moment correlation describes the dependency completely.
-As we will see, however, simulating abrirtary random vectors with a given Pearson correlation matrix is computationally intense [@Chen2001, @Xia17], violating property **S1**.
-On the other hand, an analyst may consider the use of non-parametric correlation measures, such as Spearman's $\rho$ and Kendall's $\tau$, particularly when modeling non-normal data as in our application.
-Adding to the complexity is the well-known, marginal-dependent constraints on the possible bivariate correlation (a consequence of the bounds of bivariate distribution functions --- the *Frechet-Hoeffding bounds* [@Nelsen2007]).
-Denote the pointwise upper bound as $M(y_i, y_{i^\prime})$ These bounds give strict restraints on the possible bounds and cannot be overcome through algorithm design.
-Yet not all multivariate simulation approaches obtain these bounds [for example REF].
-Computationally, algorithms that rely on serial computation scale poorly to high dimensions (refs - hierarchy) and fail to make use of modern high-performance computing including parallelization and graphical processing unit acceleration [@Li2019gpu].
-
-To meet the desired properties in the face of the challenging setting, we present a general-purpose, scalable multivariate simulation algorithm.
-The crux of the method lies in the construction of a Gaussian copula [refs].
-The idea that many multivariate and marginally heterogenous distribution can be constructed in such a manner is well known [refs].
-This article's contriubition lies in it's application to high-dimenisonal data through a high-performance implementation (`bigsimr`) and speed-focused algorithm design.
-The algorithm design relies on useful properties of non-standard correlation measures, namely Kendall's $\tau$ and Spearman's $\rho$.
-We'll show that quality of simulation does not require the use of the standard Pearson correlation matrix, even for normally distributed data.
-The purpose of our method is to generate random vectors that match exactly specified any possible Kendall's $\tau$ or Spearman's $\rho$, and approximately Pearson product-moment correlation.
-
-The study proceeds with more details on our motivating example in RNA-sequencing breast cancer data.
-Then we describe and justify our simulation methodology, followed by extensive Monte Carlo studies under various distributional assumputions --- emphasizing normal and non-normal scenarios.
-In these studies, we evaluate the accuracy and speed of our approach in comparison to other readily availble software.
-After evaluating *in silico*, we revisit our motivating
-example by employing our method and compare our simulations to empirically-derived statistics. Finally, we'll make concluding remarks regarding the method's utility and future directions.
-
-Intergene correlation [@Schissler2019].
-
-# Multivariate correlation bounds
-
-# Simulation methodology
-
-## General algorithm
-
-## Rank-based dependence measures
-
-## Pearson correlation
-
-Difficulties
-
-# The `bigsimr` R package
-
-Short description and key features, including GPU acceleration.
-
-```{r setup, eval=TRUE}
-## Using conda
-reticulate::use_condaenv("py37")
-library(bigsimr)
-```
-
-## Basic use
-
-short description
-
-### Specifying marginals
-
-As stated earlier, to generate multivariate data, we need a list of marginals (and their parameters), and a correlation structure (matrix).
-The marginal distributions can be built up as a list of lists, where each sublist contains the information for the target distribution.
-
-```{r}
-margins = list(
-  list("norm", mean = 3.14, sd = 0.1),
-  list("beta", shape1 = 1, shape2 = 4),
-  list("nbinom", size = 10, prob = 0.75)
-)
-```
-
-The things to point out here are that in each sublist (marginal), the first item is an unnamed character string with the R name of the distribution *without a letter prefix*.
-E.g.
-instead of `rnorm`, we pass in just `"norm"`.
-The second thing to note is that the remaining items are *named* arguments that go along with the distribution.
-A full list of built-in distributions is found in the appendix.
-
-### Specify correlation
-
-The next step is to define a correlation structure for the multivariate distribution.
-This correlation matrix can either come from observed data, or we can set it ourselves, or we can generate a random correlation matrix via `bigsimr::rcor`.
-Let's create a simple correlation matrix where all off-diagonal elements are 0.5.
-Since we have 3 marginals, we need a $3\times 3$ matrix.
-
-```{r}
-rho <- matrix(0.5, nrow = 3, ncol = 3)
-diag(rho) <- 1.0
-rho
-```
-
-Finally we can generate a random vector with our specified marginals and correlation structure.
-The last argument, `type`, is looking to know what kind of correlation matrix it is receiving.
-Right now it can handle Pearson, Spearman, or Kendal.
-
-### Small sample
-
-```{r}
-x = rvec(10, rho = rho, params = margins, type = "pearson")
-```
-
-On my machine, there is no dedicated GPU, so I would see the following warning message once per session.
-
-```{r, echo=FALSE}
-warning("warning.warn('No GPU/TPU found, falling back to CPU.')")
-```
-
-Taking a look at our random vector, we see that it hs 10 rows and 3 columns, one column for each marginal.
-
-```{r}
-x
-```
-
-We can simulate many more samples and then check the histogram of each margin, as well as the estimated correlation between the columns.
-
-### Scaling up N
-
-```{r, fig.width=7}
-x = rvec(10000, rho = rho, params = margins, type = "pearson")
-
-par(mfrow=c(1,3))
-hist(x[,1], breaks = 30, xlab = "", main = "Normal")
-hist(x[,2], breaks = 30, xlab = "", main = "Beta")
-hist(x[,3], breaks = 30, xlab = "", main = "Negative Binomial")
-```
-
-### Evaluation
-
-```{r}
-cor(x)
-```
-
-We can see that even wih 10,000 samples, the estimated correlation of the simulated data is not exactly the same as the target correlation.
-This can be explained by the fact that some correlations are simply not possible due to the discrete nature of certain distributions.
-Another possibility is that the copula algorith is biased and needs correction.
-
-### List of supported distributions
-
-```{r, eval=FALSE}
-all_dists <- list(
-  list(dist = "beta", shape1, shape2),
-  list(dist = "binom", size, prob),
-  list(dist = "cauchy", location, scale),
-  list(dist = "chisq", df),
-  list(dist = "exp", rate),
-  list(dist = "f", df1, df2),
-  list(dist = "gamma", shape, rate),
-  list(dist = "geom", prob),
-  list(dist = "hyper", m, n, k),
-  list(dist = "logis", location, scale),
-  list(dist = "lnorm", meanlog, sdlog),
-  list(dist = "nbinom", size, prob),
-  list(dist = "norm", mean, sd),
-  list(dist = "pois", lambda),
-  list(dist = "t", df),
-  list(dist = "unif", min, max),
-  list(dist = "weibull", shape, scale),
-  list(dist = "wilcox", m, n),
-  list(dist = "signrank", n)
-)
-```
-
-## Comparsions to other software
-
-### Speed
-
-- `copula`
-- `Genord`
-- `Multiord`
-- other?
-- CPU benchmarking
-- GPU benchmarking
-
-### Accuracy
-
-blah
-
-### Other comparisons
-
-blah
-
-# Some applications
-
-## Simulation-based correlation checking
-
-A multivariate approach?
-
-Or all the pairs?
-
-Generate, sort, and correlate algorithm @Demirtas2011
-
-```{r, computeFrechetBounds, echo = TRUE, eval=TRUE}
-
-## simulation based
-## devtools::install_github("adknudson/bigsimr", ref="develop")
-library(bigsimr)
-## ?rvec
-
-d <- 5
-rho <- rcor(d)
-margins <- list(
-    list("nbinom", size = 5, prob = 0.3),
-    list("exp", rate = 4),
-    list("binom", size = 5, prob = 0.7),
-    list("norm", mean = 10, sd = 3),
-    list("pois", lambda = 10)
-)
-
-margins2 <- list(
-    list("nbinom", 5, 0.3),
-    list("exp", 4),
-    list("binom", 5, 0.7),
-    list("norm", 10, 3),
-    list("pois", 10)
-)
-
-params = margins
-rho = rho
-## cores = parallel::detectCores() - 1
-cores = 2
-reps = 1e3
-type = "spearman"
-
-## help(package="bigsimr")
-rho_bounds <- computeCorBounds(params = margins, cores = cores, type = "spearman", reps = reps)
-rho_bounds
-```
-
-## Simulating RNA-seq data
-
-We apply our methodology to simulate RNA-sequencing data sets based on samples derived from breast cancer patients' tumors (the BRCA data set in TCGA).
-The data are freely available as part of the BRCA data set within the The Cancer Genome Atlas data warehouse and were downloaded on BLAH using the Harvard's Broad Institute's Firehouse interface.
-For each of the 1093 patients, the gene expression (abundance of messenger-RNA) from 20501 genes were counted using RNA-sequencing.
-The resultant data set can be represented as a matrix of size 1093 x 20501 with discrete values aligned to HUGO gene symbols for each patient.
-In this illustrative case study, we model the genewise marginal distributions as hetereogeneous negative binomial with pmf given by Equation \@ref(eq:nb) with two parameters, $p$ and $r$ (OR SWITCH TO mu and delta).
-Modeling RNA-seq counts as negative binomially distributed random variables is commonplace (e.g., @Zhao2018) since overdispersion is often observed for most expressed genes.
-We begin by estimating the marginal negative binomial parameters for each group using a simple method of moments approach by matching the negative binomial parameters with the sample mean and variance.
-Table/Figure XXX summarizes the estimated means, variances, and NB parameters as described in Equation \@ref(eq:nb).
-Notably some genes show underdisperation compared to a Poisson random varible.
-For simplicity we exclude these cases, but nothing in the later described simulation method prevents using a different marginal distribution in those cases (such as a Maxwell distribution REF Sellers).
+A Gaussian copula is the case where all marginal CDFs $F_i$ are the standard normal cdf, $\Phi$.
+The Gaussian copula is the one that corresponds to a multivariate normal distribution with standard normal marginal distributions and covariance matrix ${\bf R}$.
+(Since the marginals are standard normal, this ${\bf R}$ is also the correlation matrix).
+If $F_{{\bf R}}$ is the CDF of such multivariate normal distribution, then the corresponding Gaussian copula $C_{{\bf R}}$ is defined through
 
 \begin{equation}
-  g(n) = \mathbb P(Y=n) = \frac{\Gamma(n+r)}{\Gamma(r)n!} p^r(1-p)^n, \,\,\, n\in
-\mathbb N_0.
-  (\#eq:nb)
+(\#eq:gauss)
+F_{{\bf R}}(x_1, \ldots, x_d) = C_{{\bf R}}(\Phi(x_1), \ldots, \Phi(x_d)),
 \end{equation}
 
-## Simulation-based correlation testing
+where $\Phi(\cdot)$ is the standard normal CDF.
+Note that the copula $C_{{\bf R}}$ is simply the CDF of the random vector $(\Phi(X_1), \ldots, \Phi(X_d))$, where  $(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R})$.
 
-## Simulation-based probability calculations
+Sklar's Theorem (ref) guarantee's that any random vector with computational feasible inverse CDFs (P4 above) and obtainable correlation matrix (within the Frechet bounds) can be obtained via transformations involving copula functions.
+Namely, to simulate a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$, we can construct a Gaussian copula ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$.
+When an $Y_i$ is discrete, some care must be taken to define $F_i$.
+Letting
 
-# Conclusion/Discussion
+\begin{equation}
+F_{i}^{-1} = inf\{y:F_{i}(y) \geq u \}
+(\#eq:inverseCDF)
+\end{equation}
 
-- Although we recommend to use the rank-based measures
-- discuss the omission (or inclusion) of high dimensional covariance estimators?
-- We've shown that the multivariate NB simulations outperform the Independent sims.
-But the agreement is still poor.
-This could motivate other more appropriate models for high-expressing RNA-seq data than the negative binomial distribution.
-This algorithm lends itself to exploration of flexible probability models to inspire model selection in the developemnt novel RNA-seq analytic methodology.
+\noindent ensures that $Y_i \sim F_i$.
 
-We've introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a high-performance implementation called `bigsimr` (github url).
-The parallelized (multi-core) algorithm is simple and easy to understand as an application of Gaussian copulas.
-This method is largely inspired by @MB13, but with contributions with regard to scaleability, implementation, and application in high-throughput biomedical data (RNA-sequencing).
-We advocate the use of Kendall's $\tau$ as it better captures correlation among components of non-normal, as well as non-linear patterns of association.
-Moreover, the matching Kendall's $\tau$ is nearly trival due to the invariant of monotone transformations, after an adjustment to the input correlation matrix.
-Interestingly, our simulation studies show the use of Kendall's $\tau$ to works as well as using Pearson correlation coefficient when simulating from multivariate normal distributions.
+Simulation of a general multivariate random vector ${\bf T}$ with margins $F_i$ based on this copula is quite simple.
+Ensuring that one obtains a certain dependence among the marginal distributions, however, proves challenging in our proposed Gaussian-copula-based scheme.
+Two core issues: 1) marginal characteristics induce bounds on the possible bivariate correlations and 2) monotone transformations deform the Pearson correlation which no closed form expression exists in general (ref?
+Chen2001).
+The Frechet bounds are well-known.
 
-We also show utility in our methodology through an application to differential gene expression analysis from RNA-sequencing data.
-The application results show that correlations indeed matter in the large-scale hypothesis testing, as many others have noted (for example, see @BE07, @Wu2012b).
-We also hope that the application provides an example workflow --- and other strategy to use in simulation design.
-Even the best-performing simulations we provide show a gap from the empirical distribution.
-To keep the demonstration straightforward, we did not attempt to match the empirical distribution of test statistics as precisely as possible.
-Yet our methodology could be more creatively applied to meet that goal.
-One could consider marginal distributions from different families, such as Poisson for a subset of genes.
-One could imagine finding a best fitting probability distribution among a class of distributions for each gene and this could perhaps a better fit.
-One could imagine additonal structures/features in the data that an analyset could model to improve the correspondence with the emprical values, for example row correlations and high-dimensional covariance estimators (see @Won2013g).
+## Measures of dependency
 
-There are of course limitations to the methodology and implementation.
-The algorithm requires invertible marginal cdfs.
-This leaves some restriction to the available marginal probability distributions (DOES IT?
-FOR EXAMPLE?) An additional source of error is after converting dependencies (step i; ${\bf R_{Kendall}})$) the resultant matrix may not be positive definite (PD).
-Then the nearest PD matrix is used.
-From a practical computing standpoint, the user's computing resource provides the ultimate limit on how large a dimension can be simulated using the `bigsimr` R package.
-A user must consider carefully the available memory and cores when conducting a Monte Carlo experiment.
-The algorithm does amends itself well to parallelization and graphical processing unit acceleration [@Li2019gpu] and advanced users may take advantage of those techniques.
+The population correlation coefficient, commonly called the Pearson (product-moment) correlation coefficient, describes the linear association between two random variables $X$ and $Y$ and is given by
 
-Future work includes developing scalable algorithms to match the Pearson correlation matrix exactly.
-Further, more sophicated approaches could involve simulation algorithms that are "estimation awarene" and so could explore the role of covariance estimation within the simulation.
-Lastly, these may reveal new approaches to estimating and evaluating high dimensional regression and Bayesian models.
-On the implementation side, employing graphical process unit acceleration could produce substaintial speedups over our multi-core approach.
+\begin{equation}
+(\#eq:pearson)
+\rho(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ var(X)var(Y)\right]^{1/2}} 
+\end{equation}
 
-# Conclusions
+As @MB13 and @MK01 discuss, for a bivariate normal $(X,Y)$ random vector the Pearson correlation adequately describes the dependency between the components.
+$\rho$'s utility in non-normal or non-linear associations is lacking (@MK01).
+Rank-based (ordinal) approaches performance better in these settings, such as Spearman's (denoted $\rho_s$) and Kendall's $\tau$.
+Define
 
-# Supplementary Materials
+\begin{equation}
+(\#eq:spearman)
+\rho_s(X,Y) = 3 \left[ P\left[ (X_1 - X_2)(Y_1-Y_3) > 0 \right] - P\left[ (X_1 - X_2)(Y_1-Y_3) < 0 \right] \right]
+\end{equation}
 
-We provide an open-source implementation of our methology as the `bigsimr` R package, hosted on github.
+\noindent where $(X_1, Y_1) \overset{d}{=} (X,Y), X_2 \overset{d}{=} X, Y_3 \overset{d}{=} Y$ with $X_2$ and $Y_3$ are independent of one other and of $(X_1, Y_1)$.
+For continuous marginals, the measure works well due to zero probability of ties.
+For discrete marginals, however, one could perform a rescaled version of $\rho_s$ for random variables $X,Y$ with pmfs (or pdfs) $p(x)$ and $q(y)$, respectively.
 
-# Acknowledgement(s)
+\begin{equation}
+(\#eq:spearmanRescaled)
+\rho_{RS}(X,Y) = \frac{\rho_s(X,Y)}{ \left[ \left[ 1 - \sum_x p(x)^3 \right] \left[ 1 - \sum_y q(y)^3 \right] \right]^{1/2}}
+\end{equation}
 
-The authors gratefully acknowledge the helpful discussions with Professor Walter W. Piegorsch and Professor Edward J. Bedrick during this project's conception.
+\noindent Kendall's $\tau$ is the probability of concordant pairs minus the probability of discordant pairs and is given compactly as
 
-# Disclosure statement
+\begin{equation}
+  (\#eq:tau)
+  \tau(X,Y) = \frac{\rho_s(X,Y)}{ \left[ \left[ 1 - \sum_x p(x)^3 \right] \left[ 1 - \sum_y q(y)^3 \right] \right]^{1/2}}
+\end{equation}
 
-The authors report no conflict of interest. DO WE?
+## Marginal-dependent bivariate correlation bounds
 
-# Funding
+The pairwise correlation between two correlated random variables cannot in general obtain the full range of possible values, $[-1,-1]$.
+The range, called the Frechet(-Hoeffding) bounds, is a well-known function of the marginal distributions and are given by [@BF17]:
 
-FUNDING FROM EVERYONE HERE. CP3 grant.
+\begin{equation}
+(\#eq:frechet)
+\rho^{max} = \rho \left( F^{-1}_1 (U), F^{-1}_2 (U) \right), \quad \rho^{min} = \rho \left( F^{-1}_1 (U), F^{-1}_2 (1 - U) \right)
+\end{equation}
 
-# Nomenclature/Notation
-
-# Notes
-
-Mention Python?
+\noindent where $U$ is a uniform random variable in $(0,1)$, and $F^{-1}_1, F^{-1}_2$ are the inverse cdf of random variables $X_1$ and $X_2$, respectively.
+For discrete random variables, define $F^{-1}$ as in Equation \@ref(eq:inverseCDF).

### 2020-06-04 020-simulation-algorithm.Rmd
split up document into sections. did spelling check

--- 2020-06-02 index.Rmd
@@
---- 
-title: "Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package"
-author: [A.G. Schissler, A. Knudson]
-date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"
-site: bookdown::bookdown_site
-output: bookdown::gitbook
-documentclass: article
-bibliography: [bigsimr.bib, packages.bib]
-biblio-style: apalike
-link-citations: yes
-abstract: |
-  In this era of Big Data, it is critical to realistically simulate data to conduct informative Monte Carlo studies.
-This is problematic when data are inherently multivariate while at the same time are (ultra-) high dimensional.
-This situation appears frequently in observational data found on online and in high-throughput biomedical experiments (e.g., RNA-sequencing).
-Due to the difficulty in simulating realistic correlated data points, researchers often resort to simulation designs that posit independence --- greatly diminishing the insight into the empirical operating characteristics of any proposed methodology.
-Major challenges lie in the computational complexity involved in simulating these massive random vectors.
-We propose a fairly general, scalable procedure to simulate high-dimensional multivariate distributions with pre-specified marginal characteristics and dependency characteristics.
-As a motivating example, we use our methodology to study large-scale statistical inferential procedures applied to cancer-related RNA-sequencing data sets.
-The proposed algorithm is implemented as the `bigsimr` R package.
----
+# Simulation algorithm
 
-```{r LoadLib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
-library(ggplot2)
-library(tidyverse)
-library(knitr)
-library(dplyr)
-set.seed(06082020)
-```
+## Algorithm description
 
-```{bash copyBib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
-cp /Users/alfred/Dropbox/bib/bigsimr.bib ./.
-```
+This section describes the method for simulating a random vector $\bf Y$ with $Y_i$ components for $i=1,2,\ldots,d$. Each $Y_i$ has a specified marginal
+distribution function $F_i$ and its inverse.
+To characterize dependency, every pair $(Y_i, Y_j)$ has either a specified Pearson correlation \@ref(eq:pearson), (rescaled) Spearman correlation \@ref(eq:spearmanRescaled) , or Kendall's $\tau$ (Equation \@ref(eq:tau) ).
+The method only approximately matches the Pearson correlation in general, whereas the rank-based methods are exact.
 
-```{r include=FALSE}
-# automatically create a bib database for R packages
-knitr::write_bib(c(
-  .packages(), 'bookdown', 'knitr', 'rmarkdown'
-), 'packages.bib')
-```
-
-# Introduction
-
-Massive high-dimensional data sets are commonplace in many areas of scientific inquiry.
-As new methods are developed for these data, a fundamental challenge lies in designing and conducting simulation studies to assess operating characteristics of proposed methodology, such as false positive rates, statistical power, and robustness --- often in comparison to existing methods.
-Another application lies in the discovering the sampling distribution of test statistic under hypothesized null models in complex scenarios.
-Such Monte Carlo studies in the high-dimensional setting often become difficult to conduct with existing algorithms and tools, or require expertise in computing on clusters.
-This is particularly true when simulating massive multivariate, non-normal distributions.
-
-Along those lines, simulating high-dimension dependent negative binomial data motivates this work --- in pursuit of modeling RNA-sequencing data [@Wang2009b; @Conesa2016b] derived from breast cancer patients.
-This labroratory procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions.
-Often researchers posit a negative binomial model (refs) as counts are often over (or under) expressed that a Poisson model would suggest.
-Yet due to inherent biological processes, gene expression data exhibits correlation (coexpression) across genes [@BE07; @Schissler2018].
-RNA-sequencing analysis has garnered much interest in the statistical literature (refs).
-Often researchers simulate independently or from their proposed model in order condunct Monte Carlo studies (refs).
-An alternavitive strategy is to think *generatively* about the data collection process and scientific domain knowledge and design simulations with probabilitistic models that reflect those processes.
-In our motivating example, we'll study the consequences correlation in the simulation study replicating test statistics from an illustrative, classical large-scale hypothesis testing study [see @Efron2004].
-
-With our application in mind and seeking a general-purpose algorithm with broad applications, our purposed methodology should possess the following properities (adapted from criteria from @Nik13a):
-
-* P1: Wide range of dependence, allowing both positive and negative dependence
-* P2: Flexible dependence, meaning that the number of bivariate marginals is (approximately) equal to the number of dependence parameters.
-* P3: Flexible marginal modeling, generating hetereogenous data --- possibly from differing probability families.
-* P4: Computationally feasible cumulative distribution function (cdf) for likelihood estimation.
-
-Moreover, in simulation method must scale to high dimensions:
-
-* S1: Procedure must scale to high dimensions, computable in a reasonable amount time.
-* S2: Procedure must scale to high dimensions while maintaining accuracy.
-
-As others @MB13 have noted, however, simulating dependent data can be challenging.
-A central issue lies characterizing dependency between components in the high-dimensional random vector.
-The choice of correlation in typical practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g.
-non-normal, discrete, finite support, etc).
-Here we propose a scheme that ignores the analytic goal --- and, instead, aims to match the empirically-derived estimated correlation data and marginal characteristics.
-For normal data, the Pearson product-moment correlation describes the dependency completely.
-As we will see, however, simulating abrirtary random vectors with a given Pearson correlation matrix is computationally intense [@Chen2001, @Xia17], violating property **S1**.
-On the other hand, an analyst may consider the use of non-parametric correlation measures, such as Spearman's $\rho$ and Kendall's $\tau$, particularly when modeling non-normal data as in our application.
-Adding to the complexity is the well-known, marginal-dependent constraints on the possible bivariate correlation (a consequence of the bounds of bivariate distribution functions --- the *Frechet-Hoeffding bounds* [@Nelsen2007]).
-Denote the pointwise upper bound as $M(y_i, y_{i^\prime})$ These bounds give strict restraints on the possible bounds and cannot be overcome through algorithm design.
-Yet not all multivariate simulation approaches obtain these bounds [for example REF].
-Computationally, algorithms that rely on serial computation scale poorly to high dimensions (refs - hierarchy) and fail to make use of modern high-performance computing including parallelization and graphical processing unit acceleration [@Li2019gpu].
-
-To meet the desired properties in the face of the challenging setting, we present a general-purpose, scalable multivariate simulation algorithm.
-The crux of the method lies in the construction of a Gaussian copula [refs].
-The idea that many multivariate and marginally heterogenous distribution can be constructed in such a manner is well known [refs].
-This article's contriubition lies in it's application to high-dimenisonal data through a high-performance implementation (`bigsimr`) and speed-focused algorithm design.
-The algorithm design relies on useful properties of non-standard correlation measures, namely Kendall's $\tau$ and Spearman's $\rho$.
-We'll show that quality of simulation does not require the use of the standard Pearson correlation matrix, even for normally distributed data.
-The purpose of our method is to generate random vectors that match exactly specified any possible Kendall's $\tau$ or Spearman's $\rho$, and approximately Pearson product-moment correlation.
-
-The study proceeds with more details on our motivating example in RNA-sequencing breast cancer data.
-Then we describe and justify our simulation methodology, followed by extensive Monte Carlo studies under various distributional assumputions --- emphasizing normal and non-normal scenarios.
-In these studies, we evaluate the accuracy and speed of our approach in comparison to other readily availble software.
-After evaluating *in silico*, we revisit our motivating
-example by employing our method and compare our simulations to empirically-derived statistics. Finally, we'll make concluding remarks regarding the method's utility and future directions.
-
-Intergene correlation [@Schissler2019].
-
-# Multivariate correlation bounds
-
-# Simulation methodology
-
-## General algorithm
-
-## Rank-based dependence measures
-
-## Pearson correlation
-
-Difficulties
-
-# The `bigsimr` R package
-
-Short description and key features, including GPU acceleration.
-
-```{r setup, eval=TRUE}
-## Using conda
-reticulate::use_condaenv("py37")
-library(bigsimr)
-```
-
-## Basic use
-
-short description
-
-### Specifying marginals
-
-As stated earlier, to generate multivariate data, we need a list of marginals (and their parameters), and a correlation structure (matrix).
-The marginal distributions can be built up as a list of lists, where each sublist contains the information for the target distribution.
-
-```{r}
-margins = list(
-  list("norm", mean = 3.14, sd = 0.1),
-  list("beta", shape1 = 1, shape2 = 4),
-  list("nbinom", size = 10, prob = 0.75)
-)
-```
-
-The things to point out here are that in each sublist (marginal), the first item is an unnamed character string with the R name of the distribution *without a letter prefix*.
-E.g.
-instead of `rnorm`, we pass in just `"norm"`.
-The second thing to note is that the remaining items are *named* arguments that go along with the distribution.
-A full list of built-in distributions is found in the appendix.
-
-### Specify correlation
-
-The next step is to define a correlation structure for the multivariate distribution.
-This correlation matrix can either come from observed data, or we can set it ourselves, or we can generate a random correlation matrix via `bigsimr::rcor`.
-Let's create a simple correlation matrix where all off-diagonal elements are 0.5.
-Since we have 3 marginals, we need a $3\times 3$ matrix.
-
-```{r}
-rho <- matrix(0.5, nrow = 3, ncol = 3)
-diag(rho) <- 1.0
-rho
-```
-
-Finally we can generate a random vector with our specified marginals and correlation structure.
-The last argument, `type`, is looking to know what kind of correlation matrix it is receiving.
-Right now it can handle Pearson, Spearman, or Kendal.
-
-### Small sample
-
-```{r}
-x = rvec(10, rho = rho, params = margins, type = "pearson")
-```
-
-On my machine, there is no dedicated GPU, so I would see the following warning message once per session.
-
-```{r, echo=FALSE}
-warning("warning.warn('No GPU/TPU found, falling back to CPU.')")
-```
-
-Taking a look at our random vector, we see that it hs 10 rows and 3 columns, one column for each marginal.
-
-```{r}
-x
-```
-
-We can simulate many more samples and then check the histogram of each margin, as well as the estimated correlation between the columns.
-
-### Scaling up N
-
-```{r, fig.width=7}
-x = rvec(10000, rho = rho, params = margins, type = "pearson")
-
-par(mfrow=c(1,3))
-hist(x[,1], breaks = 30, xlab = "", main = "Normal")
-hist(x[,2], breaks = 30, xlab = "", main = "Beta")
-hist(x[,3], breaks = 30, xlab = "", main = "Negative Binomial")
-```
-
-### Evaluation
-
-```{r}
-cor(x)
-```
-
-We can see that even wih 10,000 samples, the estimated correlation of the simulated data is not exactly the same as the target correlation.
-This can be explained by the fact that some correlations are simply not possible due to the discrete nature of certain distributions.
-Another possibility is that the copula algorith is biased and needs correction.
-
-### List of supported distributions
-
-```{r, eval=FALSE}
-all_dists <- list(
-  list(dist = "beta", shape1, shape2),
-  list(dist = "binom", size, prob),
-  list(dist = "cauchy", location, scale),
-  list(dist = "chisq", df),
-  list(dist = "exp", rate),
-  list(dist = "f", df1, df2),
-  list(dist = "gamma", shape, rate),
-  list(dist = "geom", prob),
-  list(dist = "hyper", m, n, k),
-  list(dist = "logis", location, scale),
-  list(dist = "lnorm", meanlog, sdlog),
-  list(dist = "nbinom", size, prob),
-  list(dist = "norm", mean, sd),
-  list(dist = "pois", lambda),
-  list(dist = "t", df),
-  list(dist = "unif", min, max),
-  list(dist = "weibull", shape, scale),
-  list(dist = "wilcox", m, n),
-  list(dist = "signrank", n)
-)
-```
-
-## Comparsions to other software
-
-### Speed
-
-- `copula`
-- `Genord`
-- `Multiord`
-- other?
-- CPU benchmarking
-- GPU benchmarking
-
-### Accuracy
-
-blah
-
-### Other comparisons
-
-blah
-
-# Some applications
-
-## Simulation-based correlation checking
-
-A multivariate approach?
-
-Or all the pairs?
-
-Generate, sort, and correlate algorithm @Demirtas2011
-
-```{r, computeFrechetBounds, echo = TRUE, eval=TRUE}
-
-## simulation based
-## devtools::install_github("adknudson/bigsimr", ref="develop")
-library(bigsimr)
-## ?rvec
-
-d <- 5
-rho <- rcor(d)
-margins <- list(
-    list("nbinom", size = 5, prob = 0.3),
-    list("exp", rate = 4),
-    list("binom", size = 5, prob = 0.7),
-    list("norm", mean = 10, sd = 3),
-    list("pois", lambda = 10)
-)
-
-margins2 <- list(
-    list("nbinom", 5, 0.3),
-    list("exp", 4),
-    list("binom", 5, 0.7),
-    list("norm", 10, 3),
-    list("pois", 10)
-)
-
-params = margins
-rho = rho
-## cores = parallel::detectCores() - 1
-cores = 2
-reps = 1e3
-type = "spearman"
-
-## help(package="bigsimr")
-rho_bounds <- computeCorBounds(params = margins, cores = cores, type = "spearman", reps = reps)
-rho_bounds
-```
-
-## Simulating RNA-seq data
-
-We apply our methodology to simulate RNA-sequencing data sets based on samples derived from breast cancer patients' tumors (the BRCA data set in TCGA).
-The data are freely available as part of the BRCA data set within the The Cancer Genome Atlas data warehouse and were downloaded on BLAH using the Harvard's Broad Institute's Firehouse interface.
-For each of the 1093 patients, the gene expression (abundance of messenger-RNA) from 20501 genes were counted using RNA-sequencing.
-The resultant data set can be represented as a matrix of size 1093 x 20501 with discrete values aligned to HUGO gene symbols for each patient.
-In this illustrative case study, we model the genewise marginal distributions as hetereogeneous negative binomial with pmf given by Equation \@ref(eq:nb) with two parameters, $p$ and $r$ (OR SWITCH TO mu and delta).
-Modeling RNA-seq counts as negative binomially distributed random variables is commonplace (e.g., @Zhao2018) since overdispersion is often observed for most expressed genes.
-We begin by estimating the marginal negative binomial parameters for each group using a simple method of moments approach by matching the negative binomial parameters with the sample mean and variance.
-Table/Figure XXX summarizes the estimated means, variances, and NB parameters as described in Equation \@ref(eq:nb).
-Notably some genes show underdisperation compared to a Poisson random varible.
-For simplicity we exclude these cases, but nothing in the later described simulation method prevents using a different marginal distribution in those cases (such as a Maxwell distribution REF Sellers).
+The method is best understand as a **parallelized Gaussian copula** (see Equation \@ref(eq:gauss).
+We shall see that constructing continuous joint distributions that match a target Spearman or Kendall's correlations computes easily when employing Gaussian copulas, since this measures are invariant under the monotone transformations involved [refs].
+To do this, we take advantage of a closed form relationship [ref?] between Kendall's $\tau$ and Pearson's correlation coefficient for bivariate normal random variables:
 
 \begin{equation}
-  g(n) = \mathbb P(Y=n) = \frac{\Gamma(n+r)}{\Gamma(r)n!} p^r(1-p)^n, \,\,\, n\in
-\mathbb N_0.
-  (\#eq:nb)
+(\#eq:convertKendall)
+r_{Pearson} = sin \left( \tau_{Kendall} \times \frac{\pi}{2} \right), 
 \end{equation}
 
-## Simulation-based correlation testing
+\noindent and similarly for Spearman's $\rho$ [@K58],
 
-## Simulation-based probability calculations
+\begin{equation}
+(\#eq:convertSpearman)
+\rho_{Pearson} = 2 \times sin \left( \rho_{Spearman} \times \frac{\pi}{6} \right).
+\end{equation}
 
-# Conclusion/Discussion
+For discrete marginals, achieving a target Spearman correlation under this scheme is possible by using components from Equation \@ref(eq:spearmanRescaled) to further adjust the input correlation matrix.
+Let the unscaled Spearman correlation coefficients be $\rho_{s} \left(Y_{i}, Y_{i^\prime}\right)$ for two marginal distributions and divide the target correlation by the product in the denominator of Equation \@ref(eq:spearmanRescaled).
+Let these adjustment factors be denoted as $a_i = \left[ 1 - \sum_y p_i(y)^3 \right]^{1/2}$ and specifically rescale the target Spearman correlation matrix by
 
-- Although we recommend to use the rank-based measures
-- discuss the omission (or inclusion) of high dimensional covariance estimators?
-- We've shown that the multivariate NB simulations outperform the Independent sims.
-But the agreement is still poor.
-This could motivate other more appropriate models for high-expressing RNA-seq data than the negative binomial distribution.
-This algorithm lends itself to exploration of flexible probability models to inspire model selection in the developemnt novel RNA-seq analytic methodology.
+\begin{equation}
+(\#eq:convertSpearmanDiscrete)
+\rho_{rs} \left(Y_{i}, Y_{i^\prime}\right) = \frac{\rho_{s} \left(Y_{i}, Y_{i^\prime}\right)}{a_i \times a_{i^\prime}}.
+\end{equation}
 
-We've introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a high-performance implementation called `bigsimr` (github url).
-The parallelized (multi-core) algorithm is simple and easy to understand as an application of Gaussian copulas.
-This method is largely inspired by @MB13, but with contributions with regard to scaleability, implementation, and application in high-throughput biomedical data (RNA-sequencing).
-We advocate the use of Kendall's $\tau$ as it better captures correlation among components of non-normal, as well as non-linear patterns of association.
-Moreover, the matching Kendall's $\tau$ is nearly trival due to the invariant of monotone transformations, after an adjustment to the input correlation matrix.
-Interestingly, our simulation studies show the use of Kendall's $\tau$ to works as well as using Pearson correlation coefficient when simulating from multivariate normal distributions.
+In a similar fashion, we rescale Kendall's $\tau$ to adjust the input correlation matrix. The conversion formula is given by
 
-We also show utility in our methodology through an application to differential gene expression analysis from RNA-sequencing data.
-The application results show that correlations indeed matter in the large-scale hypothesis testing, as many others have noted (for example, see @BE07, @Wu2012b).
-We also hope that the application provides an example workflow --- and other strategy to use in simulation design.
-Even the best-performing simulations we provide show a gap from the empirical distribution.
-To keep the demonstration straightforward, we did not attempt to match the empirical distribution of test statistics as precisely as possible.
-Yet our methodology could be more creatively applied to meet that goal.
-One could consider marginal distributions from different families, such as Poisson for a subset of genes.
-One could imagine finding a best fitting probability distribution among a class of distributions for each gene and this could perhaps a better fit.
-One could imagine additonal structures/features in the data that an analyset could model to improve the correspondence with the emprical values, for example row correlations and high-dimensional covariance estimators (see @Won2013g).
+\begin{equation}
+(\#eq:convertKendallDiscrete)
+\rho_{rs} \left(Y_{i}, Y_{i^\prime}\right) = \frac{\rho_{s} \left(Y_{i}, Y_{i^\prime}\right)}{a_i \times a_{i^\prime}}.
+\end{equation}
 
-There are of course limitations to the methodology and implementation.
-The algorithm requires invertible marginal cdfs.
-This leaves some restriction to the available marginal probability distributions (DOES IT?
-FOR EXAMPLE?) An additional source of error is after converting dependencies (step i; ${\bf R_{Kendall}})$) the resultant matrix may not be positive definite (PD).
-Then the nearest PD matrix is used.
-From a practical computing standpoint, the user's computing resource provides the ultimate limit on how large a dimension can be simulated using the `bigsimr` R package.
-A user must consider carefully the available memory and cores when conducting a Monte Carlo experiment.
-The algorithm does amends itself well to parallelization and graphical processing unit acceleration [@Li2019gpu] and advanced users may take advantage of those techniques.
+In contrast the rank-based correlations, matching specified Pearson correlation coefficients exactly is computational intense in this scheme.
+In general, there is no closed form correspondence and involving computing or approximating $\binom{d}{2}$ integrals of the form $EY_iY_j = \int \int y_i y_j f_{X|r}(F_i^{-1}(\Phi(z_i)), F_j^{-1}(\Phi(z_j))dy_idy_j$, for $i,j=1,2,\ldots,d$.
+For accurate numeric approximation of these integrals, the functions must be evaluated hundreds of times.
+Others have used efficient Monte Carlo integration schemes (see @Chen2001), but scale poorly to large dimension in reasonable times (property **S2**).
+Despite all this, if one does desire to characterize dependency using Pearson correlations, we often see in practice --- and it is theoretically justified under certain conditions (@Song00) --- that simply using the target Pearson correlation matrix as the initial conditions to our proposed algorithm will lead to approximate matching in the resultant distribution.
 
-Future work includes developing scalable algorithms to match the Pearson correlation matrix exactly.
-Further, more sophicated approaches could involve simulation algorithms that are "estimation awarene" and so could explore the role of covariance estimation within the simulation.
-Lastly, these may reveal new approaches to estimating and evaluating high dimensional regression and Bayesian models.
-On the implementation side, employing graphical process unit acceleration could produce substaintial speedups over our multi-core approach.
+## Simulation Algorithm
 
-# Conclusions
+Putting the together the facts provided in the equations above, we come the following proposed simulation algorithm to produce a random vector ${\bf Y}$ with specified Spearman's correlation and marginal distributions.
+Note that all computational steps can be parallelized as each operation can be done or either the $d$ marginals or the $\binom{d}{2}$ pairs for the correlation values.
+Even the generation of multivariate normal random vectors is parallelized through optimized matrix multiplication/decomposition routines.
 
-# Supplementary Materials
+### Inputs
+(1) Marginal characteristics including the same of distributional family and parameter values, denotes as $F_i$ for marginal component random variable $Y_i$ for $i=1,\ldots,d$.
+(2) A target (specified) Spearman's correlation matrix ${\bf R_{Spearman}}$ with
+each $\rho_{Spearman}$ within the Frechet limits. Alternatively, the rescaled
+Spearman's correlation (see Equation \@ref(eq:convertKendallDiscrete) ) can be provided.
+(3) An error tolerance $\epsilon$ for finding the nearest positive definite matrix a transformed correlation matrix (see step iii below). Alternatively, a maximum number of iterations can be supplied.
 
-We provide an open-source implementation of our methology as the `bigsimr` R package, hosted on github.
-
-# Acknowledgement(s)
-
-The authors gratefully acknowledge the helpful discussions with Professor Walter W. Piegorsch and Professor Edward J. Bedrick during this project's conception.
-
-# Disclosure statement
-
-The authors report no conflict of interest. DO WE?
-
-# Funding
-
-FUNDING FROM EVERYONE HERE. CP3 grant.
-
-# Nomenclature/Notation
-
-# Notes
-
-Mention Python?
+### Algorithm
+(i) Compute Spearman's $\rho_{rs}$ (see Equation
+\@ref(eq:convertKendallDiscrete)) from the specified $\rho_{spearman}$ for each
+pair of marginal random variables $(Y_i,Y_{i^\prime})$ with $i \neq i^\prime$
+when either marginal is discrete. In such large scale computations, it may be
+that numerically $\rho_{rs}$ is be larger/smaller than the Frechet bounds (or
+even $\pm 1$). To guard against this, set $\rho_{rs} = min( \rho_{rs}, M)$,
+where $M$ is the upper Frechet bound (see [Section Introduction]) and similarly set $\rho_{rs} = max( \rho_{rs}, W)$, where $W$ is the lower Frechet bound for this pair of distributions.
+Gather these $\rho_{rs} \lq s$ into a new input correlation matrix ${\bf R_{rs}}$.
+(ii) Convert ${\bf R_{Spearman}}$ into ${\bf R_{Pearson}}$ via $\rho_{Pearson} = 2 \times sin \left( \rho_{rs} \times \frac{\pi}{6} \right)$.   
+(iii) Finally, ensure that ${\bf R_{Pearson}}$ is positive definite, by finding the nearest positive definite correlation matrix ${\bf R}$ --- using the `R` routine `nearPD` in the `Matrix` package --- within an error tolerance of $\epsilon$.
+(iv) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R})$;  
+(v) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$;  
+(vi) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;

### 2020-06-04 030-bigsimr-package.Rmd
split up document into sections. did spelling check

--- 2020-06-02 index.Rmd
@@
---- 
-title: "Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package"
-author: [A.G. Schissler, A. Knudson]
-date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"
-site: bookdown::bookdown_site
-output: bookdown::gitbook
-documentclass: article
-bibliography: [bigsimr.bib, packages.bib]
-biblio-style: apalike
-link-citations: yes
-abstract: |
-  In this era of Big Data, it is critical to realistically simulate data to conduct informative Monte Carlo studies.
-This is problematic when data are inherently multivariate while at the same time are (ultra-) high dimensional.
-This situation appears frequently in observational data found on online and in high-throughput biomedical experiments (e.g., RNA-sequencing).
-Due to the difficulty in simulating realistic correlated data points, researchers often resort to simulation designs that posit independence --- greatly diminishing the insight into the empirical operating characteristics of any proposed methodology.
-Major challenges lie in the computational complexity involved in simulating these massive random vectors.
-We propose a fairly general, scalable procedure to simulate high-dimensional multivariate distributions with pre-specified marginal characteristics and dependency characteristics.
-As a motivating example, we use our methodology to study large-scale statistical inferential procedures applied to cancer-related RNA-sequencing data sets.
-The proposed algorithm is implemented as the `bigsimr` R package.
----
-
-```{r LoadLib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
-library(ggplot2)
-library(tidyverse)
-library(knitr)
-library(dplyr)
-set.seed(06082020)
-```
-
-```{bash copyBib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
-cp /Users/alfred/Dropbox/bib/bigsimr.bib ./.
-```
-
-```{r include=FALSE}
-# automatically create a bib database for R packages
-knitr::write_bib(c(
-  .packages(), 'bookdown', 'knitr', 'rmarkdown'
-), 'packages.bib')
-```
-
-# Introduction
-
-Massive high-dimensional data sets are commonplace in many areas of scientific inquiry.
-As new methods are developed for these data, a fundamental challenge lies in designing and conducting simulation studies to assess operating characteristics of proposed methodology, such as false positive rates, statistical power, and robustness --- often in comparison to existing methods.
-Another application lies in the discovering the sampling distribution of test statistic under hypothesized null models in complex scenarios.
-Such Monte Carlo studies in the high-dimensional setting often become difficult to conduct with existing algorithms and tools, or require expertise in computing on clusters.
-This is particularly true when simulating massive multivariate, non-normal distributions.
-
-Along those lines, simulating high-dimension dependent negative binomial data motivates this work --- in pursuit of modeling RNA-sequencing data [@Wang2009b; @Conesa2016b] derived from breast cancer patients.
-This labroratory procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions.
-Often researchers posit a negative binomial model (refs) as counts are often over (or under) expressed that a Poisson model would suggest.
-Yet due to inherent biological processes, gene expression data exhibits correlation (coexpression) across genes [@BE07; @Schissler2018].
-RNA-sequencing analysis has garnered much interest in the statistical literature (refs).
-Often researchers simulate independently or from their proposed model in order condunct Monte Carlo studies (refs).
-An alternavitive strategy is to think *generatively* about the data collection process and scientific domain knowledge and design simulations with probabilitistic models that reflect those processes.
-In our motivating example, we'll study the consequences correlation in the simulation study replicating test statistics from an illustrative, classical large-scale hypothesis testing study [see @Efron2004].
-
-With our application in mind and seeking a general-purpose algorithm with broad applications, our purposed methodology should possess the following properities (adapted from criteria from @Nik13a):
-
-* P1: Wide range of dependence, allowing both positive and negative dependence
-* P2: Flexible dependence, meaning that the number of bivariate marginals is (approximately) equal to the number of dependence parameters.
-* P3: Flexible marginal modeling, generating hetereogenous data --- possibly from differing probability families.
-* P4: Computationally feasible cumulative distribution function (cdf) for likelihood estimation.
-
-Moreover, in simulation method must scale to high dimensions:
-
-* S1: Procedure must scale to high dimensions, computable in a reasonable amount time.
-* S2: Procedure must scale to high dimensions while maintaining accuracy.
-
-As others @MB13 have noted, however, simulating dependent data can be challenging.
-A central issue lies characterizing dependency between components in the high-dimensional random vector.
-The choice of correlation in typical practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g.
-non-normal, discrete, finite support, etc).
-Here we propose a scheme that ignores the analytic goal --- and, instead, aims to match the empirically-derived estimated correlation data and marginal characteristics.
-For normal data, the Pearson product-moment correlation describes the dependency completely.
-As we will see, however, simulating abrirtary random vectors with a given Pearson correlation matrix is computationally intense [@Chen2001, @Xia17], violating property **S1**.
-On the other hand, an analyst may consider the use of non-parametric correlation measures, such as Spearman's $\rho$ and Kendall's $\tau$, particularly when modeling non-normal data as in our application.
-Adding to the complexity is the well-known, marginal-dependent constraints on the possible bivariate correlation (a consequence of the bounds of bivariate distribution functions --- the *Frechet-Hoeffding bounds* [@Nelsen2007]).
-Denote the pointwise upper bound as $M(y_i, y_{i^\prime})$ These bounds give strict restraints on the possible bounds and cannot be overcome through algorithm design.
-Yet not all multivariate simulation approaches obtain these bounds [for example REF].
-Computationally, algorithms that rely on serial computation scale poorly to high dimensions (refs - hierarchy) and fail to make use of modern high-performance computing including parallelization and graphical processing unit acceleration [@Li2019gpu].
-
-To meet the desired properties in the face of the challenging setting, we present a general-purpose, scalable multivariate simulation algorithm.
-The crux of the method lies in the construction of a Gaussian copula [refs].
-The idea that many multivariate and marginally heterogenous distribution can be constructed in such a manner is well known [refs].
-This article's contriubition lies in it's application to high-dimenisonal data through a high-performance implementation (`bigsimr`) and speed-focused algorithm design.
-The algorithm design relies on useful properties of non-standard correlation measures, namely Kendall's $\tau$ and Spearman's $\rho$.
-We'll show that quality of simulation does not require the use of the standard Pearson correlation matrix, even for normally distributed data.
-The purpose of our method is to generate random vectors that match exactly specified any possible Kendall's $\tau$ or Spearman's $\rho$, and approximately Pearson product-moment correlation.
-
-The study proceeds with more details on our motivating example in RNA-sequencing breast cancer data.
-Then we describe and justify our simulation methodology, followed by extensive Monte Carlo studies under various distributional assumputions --- emphasizing normal and non-normal scenarios.
-In these studies, we evaluate the accuracy and speed of our approach in comparison to other readily availble software.
-After evaluating *in silico*, we revisit our motivating
-example by employing our method and compare our simulations to empirically-derived statistics. Finally, we'll make concluding remarks regarding the method's utility and future directions.
-
-Intergene correlation [@Schissler2019].
-
-# Multivariate correlation bounds
-
-# Simulation methodology
-
-## General algorithm
-
-## Rank-based dependence measures
-
-## Pearson correlation
-
-Difficulties
-
 # The `bigsimr` R package
@@
 Short description and key features, including GPU acceleration.
+
+The most computationally extensive lies in the second step of our algorithm. We
+use a optimized and parallelized multivariate normal simulator within the `R`
+package `mvnfast`. The rest of the code is also parallelized, but these steps
+run rapidly in serial computation except for extremely large dimension.
 
@@
 The last argument, `type`, is looking to know what kind of correlation matrix it is receiving.
-Right now it can handle Pearson, Spearman, or Kendal.
+Right now it can handle Pearson, Spearman, or Kendall.
 
@@
 
-Taking a look at our random vector, we see that it hs 10 rows and 3 columns, one column for each marginal.
+Taking a look at our random vector, we see that it is 10 rows and 3 columns, one column for each marginal.
 
@@
 
-We can see that even wih 10,000 samples, the estimated correlation of the simulated data is not exactly the same as the target correlation.
+We can see that even with 10,000 samples, the estimated correlation of the simulated data is not exactly the same as the target correlation.
 This can be explained by the fact that some correlations are simply not possible due to the discrete nature of certain distributions.
-Another possibility is that the copula algorith is biased and needs correction.
-
-### List of supported distributions
-
-```{r, eval=FALSE}
-all_dists <- list(
-  list(dist = "beta", shape1, shape2),
-  list(dist = "binom", size, prob),
-  list(dist = "cauchy", location, scale),
-  list(dist = "chisq", df),
-  list(dist = "exp", rate),
-  list(dist = "f", df1, df2),
-  list(dist = "gamma", shape, rate),
-  list(dist = "geom", prob),
-  list(dist = "hyper", m, n, k),
-  list(dist = "logis", location, scale),
-  list(dist = "lnorm", meanlog, sdlog),
-  list(dist = "nbinom", size, prob),
-  list(dist = "norm", mean, sd),
-  list(dist = "pois", lambda),
-  list(dist = "t", df),
-  list(dist = "unif", min, max),
-  list(dist = "weibull", shape, scale),
-  list(dist = "wilcox", m, n),
-  list(dist = "signrank", n)
-)
-```
-
-## Comparsions to other software
-
-### Speed
-
-- `copula`
-- `Genord`
-- `Multiord`
-- other?
-- CPU benchmarking
-- GPU benchmarking
-
-### Accuracy
-
-blah
-
-### Other comparisons
-
-blah
-
-# Some applications
-
-## Simulation-based correlation checking
-
-A multivariate approach?
-
-Or all the pairs?
-
-Generate, sort, and correlate algorithm @Demirtas2011
-
-```{r, computeFrechetBounds, echo = TRUE, eval=TRUE}
-
-## simulation based
-## devtools::install_github("adknudson/bigsimr", ref="develop")
-library(bigsimr)
-## ?rvec
-
-d <- 5
-rho <- rcor(d)
-margins <- list(
-    list("nbinom", size = 5, prob = 0.3),
-    list("exp", rate = 4),
-    list("binom", size = 5, prob = 0.7),
-    list("norm", mean = 10, sd = 3),
-    list("pois", lambda = 10)
-)
-
-margins2 <- list(
-    list("nbinom", 5, 0.3),
-    list("exp", 4),
-    list("binom", 5, 0.7),
-    list("norm", 10, 3),
-    list("pois", 10)
-)
-
-params = margins
-rho = rho
-## cores = parallel::detectCores() - 1
-cores = 2
-reps = 1e3
-type = "spearman"
-
-## help(package="bigsimr")
-rho_bounds <- computeCorBounds(params = margins, cores = cores, type = "spearman", reps = reps)
-rho_bounds
-```
-
-## Simulating RNA-seq data
-
-We apply our methodology to simulate RNA-sequencing data sets based on samples derived from breast cancer patients' tumors (the BRCA data set in TCGA).
-The data are freely available as part of the BRCA data set within the The Cancer Genome Atlas data warehouse and were downloaded on BLAH using the Harvard's Broad Institute's Firehouse interface.
-For each of the 1093 patients, the gene expression (abundance of messenger-RNA) from 20501 genes were counted using RNA-sequencing.
-The resultant data set can be represented as a matrix of size 1093 x 20501 with discrete values aligned to HUGO gene symbols for each patient.
-In this illustrative case study, we model the genewise marginal distributions as hetereogeneous negative binomial with pmf given by Equation \@ref(eq:nb) with two parameters, $p$ and $r$ (OR SWITCH TO mu and delta).
-Modeling RNA-seq counts as negative binomially distributed random variables is commonplace (e.g., @Zhao2018) since overdispersion is often observed for most expressed genes.
-We begin by estimating the marginal negative binomial parameters for each group using a simple method of moments approach by matching the negative binomial parameters with the sample mean and variance.
-Table/Figure XXX summarizes the estimated means, variances, and NB parameters as described in Equation \@ref(eq:nb).
-Notably some genes show underdisperation compared to a Poisson random varible.
-For simplicity we exclude these cases, but nothing in the later described simulation method prevents using a different marginal distribution in those cases (such as a Maxwell distribution REF Sellers).
-
-\begin{equation} 
-  g(n) = \mathbb P(Y=n) = \frac{\Gamma(n+r)}{\Gamma(r)n!} p^r(1-p)^n, \,\,\, n\in
-\mathbb N_0.
-  (\#eq:nb)
-\end{equation}
-
-## Simulation-based correlation testing
-
-## Simulation-based probability calculations
-
-# Conclusion/Discussion
-
-- Although we recommend to use the rank-based measures
-- discuss the omission (or inclusion) of high dimensional covariance estimators?
-- We've shown that the multivariate NB simulations outperform the Independent sims.
-But the agreement is still poor.
-This could motivate other more appropriate models for high-expressing RNA-seq data than the negative binomial distribution.
-This algorithm lends itself to exploration of flexible probability models to inspire model selection in the developemnt novel RNA-seq analytic methodology.
-
-We've introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a high-performance implementation called `bigsimr` (github url).
-The parallelized (multi-core) algorithm is simple and easy to understand as an application of Gaussian copulas.
-This method is largely inspired by @MB13, but with contributions with regard to scaleability, implementation, and application in high-throughput biomedical data (RNA-sequencing).
-We advocate the use of Kendall's $\tau$ as it better captures correlation among components of non-normal, as well as non-linear patterns of association.
-Moreover, the matching Kendall's $\tau$ is nearly trival due to the invariant of monotone transformations, after an adjustment to the input correlation matrix.
-Interestingly, our simulation studies show the use of Kendall's $\tau$ to works as well as using Pearson correlation coefficient when simulating from multivariate normal distributions.
-
-We also show utility in our methodology through an application to differential gene expression analysis from RNA-sequencing data.
-The application results show that correlations indeed matter in the large-scale hypothesis testing, as many others have noted (for example, see @BE07, @Wu2012b).
-We also hope that the application provides an example workflow --- and other strategy to use in simulation design.
-Even the best-performing simulations we provide show a gap from the empirical distribution.
-To keep the demonstration straightforward, we did not attempt to match the empirical distribution of test statistics as precisely as possible.
-Yet our methodology could be more creatively applied to meet that goal.
-One could consider marginal distributions from different families, such as Poisson for a subset of genes.
-One could imagine finding a best fitting probability distribution among a class of distributions for each gene and this could perhaps a better fit.
-One could imagine additonal structures/features in the data that an analyset could model to improve the correspondence with the emprical values, for example row correlations and high-dimensional covariance estimators (see @Won2013g).
-
-There are of course limitations to the methodology and implementation.
-The algorithm requires invertible marginal cdfs.
-This leaves some restriction to the available marginal probability distributions (DOES IT?
-FOR EXAMPLE?) An additional source of error is after converting dependencies (step i; ${\bf R_{Kendall}})$) the resultant matrix may not be positive definite (PD).
-Then the nearest PD matrix is used.
-From a practical computing standpoint, the user's computing resource provides the ultimate limit on how large a dimension can be simulated using the `bigsimr` R package.
-A user must consider carefully the available memory and cores when conducting a Monte Carlo experiment.
-The algorithm does amends itself well to parallelization and graphical processing unit acceleration [@Li2019gpu] and advanced users may take advantage of those techniques.
-
-Future work includes developing scalable algorithms to match the Pearson correlation matrix exactly.
-Further, more sophicated approaches could involve simulation algorithms that are "estimation awarene" and so could explore the role of covariance estimation within the simulation.
-Lastly, these may reveal new approaches to estimating and evaluating high dimensional regression and Bayesian models.
-On the implementation side, employing graphical process unit acceleration could produce substaintial speedups over our multi-core approach.
-
-# Conclusions
-
-# Supplementary Materials
-
-We provide an open-source implementation of our methology as the `bigsimr` R package, hosted on github.
-
-# Acknowledgement(s)
-
-The authors gratefully acknowledge the helpful discussions with Professor Walter W. Piegorsch and Professor Edward J. Bedrick during this project's conception.
-
-# Disclosure statement
-
-The authors report no conflict of interest. DO WE?
-
-# Funding
-
-FUNDING FROM EVERYONE HERE. CP3 grant.
-
-# Nomenclature/Notation
-
-# Notes
-
-Mention Python?
+Another possibility is that the copula algorithm is biased and needs correction.

### 2020-06-04 050-some-applications.Rmd
split up document into sections. did spelling check

--- 2020-06-02 index.Rmd
@@
---- 
-title: "Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package"
-author: [A.G. Schissler, A. Knudson]
-date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"
-site: bookdown::bookdown_site
-output: bookdown::gitbook
-documentclass: article
-bibliography: [bigsimr.bib, packages.bib]
-biblio-style: apalike
-link-citations: yes
-abstract: |
-  In this era of Big Data, it is critical to realistically simulate data to conduct informative Monte Carlo studies.
-This is problematic when data are inherently multivariate while at the same time are (ultra-) high dimensional.
-This situation appears frequently in observational data found on online and in high-throughput biomedical experiments (e.g., RNA-sequencing).
-Due to the difficulty in simulating realistic correlated data points, researchers often resort to simulation designs that posit independence --- greatly diminishing the insight into the empirical operating characteristics of any proposed methodology.
-Major challenges lie in the computational complexity involved in simulating these massive random vectors.
-We propose a fairly general, scalable procedure to simulate high-dimensional multivariate distributions with pre-specified marginal characteristics and dependency characteristics.
-As a motivating example, we use our methodology to study large-scale statistical inferential procedures applied to cancer-related RNA-sequencing data sets.
-The proposed algorithm is implemented as the `bigsimr` R package.
----
+# Some applications
 
-```{r LoadLib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
-library(ggplot2)
-library(tidyverse)
-library(knitr)
-library(dplyr)
-set.seed(06082020)
-```
+## Simulating Ultra High Dimensional RNA-seq data
 
-```{bash copyBib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
-cp /Users/alfred/Dropbox/bib/bigsimr.bib ./.
-```
-
-```{r include=FALSE}
-# automatically create a bib database for R packages
-knitr::write_bib(c(
-  .packages(), 'bookdown', 'knitr', 'rmarkdown'
-), 'packages.bib')
-```
-
-# Introduction
-
-Massive high-dimensional data sets are commonplace in many areas of scientific inquiry.
-As new methods are developed for these data, a fundamental challenge lies in designing and conducting simulation studies to assess operating characteristics of proposed methodology, such as false positive rates, statistical power, and robustness --- often in comparison to existing methods.
-Another application lies in the discovering the sampling distribution of test statistic under hypothesized null models in complex scenarios.
-Such Monte Carlo studies in the high-dimensional setting often become difficult to conduct with existing algorithms and tools, or require expertise in computing on clusters.
-This is particularly true when simulating massive multivariate, non-normal distributions.
-
-Along those lines, simulating high-dimension dependent negative binomial data motivates this work --- in pursuit of modeling RNA-sequencing data [@Wang2009b; @Conesa2016b] derived from breast cancer patients.
-This labroratory procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions.
-Often researchers posit a negative binomial model (refs) as counts are often over (or under) expressed that a Poisson model would suggest.
-Yet due to inherent biological processes, gene expression data exhibits correlation (coexpression) across genes [@BE07; @Schissler2018].
-RNA-sequencing analysis has garnered much interest in the statistical literature (refs).
-Often researchers simulate independently or from their proposed model in order condunct Monte Carlo studies (refs).
-An alternavitive strategy is to think *generatively* about the data collection process and scientific domain knowledge and design simulations with probabilitistic models that reflect those processes.
-In our motivating example, we'll study the consequences correlation in the simulation study replicating test statistics from an illustrative, classical large-scale hypothesis testing study [see @Efron2004].
-
-With our application in mind and seeking a general-purpose algorithm with broad applications, our purposed methodology should possess the following properities (adapted from criteria from @Nik13a):
-
-* P1: Wide range of dependence, allowing both positive and negative dependence
-* P2: Flexible dependence, meaning that the number of bivariate marginals is (approximately) equal to the number of dependence parameters.
-* P3: Flexible marginal modeling, generating hetereogenous data --- possibly from differing probability families.
-* P4: Computationally feasible cumulative distribution function (cdf) for likelihood estimation.
-
-Moreover, in simulation method must scale to high dimensions:
-
-* S1: Procedure must scale to high dimensions, computable in a reasonable amount time.
-* S2: Procedure must scale to high dimensions while maintaining accuracy.
-
-As others @MB13 have noted, however, simulating dependent data can be challenging.
-A central issue lies characterizing dependency between components in the high-dimensional random vector.
-The choice of correlation in typical practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g.
-non-normal, discrete, finite support, etc).
-Here we propose a scheme that ignores the analytic goal --- and, instead, aims to match the empirically-derived estimated correlation data and marginal characteristics.
-For normal data, the Pearson product-moment correlation describes the dependency completely.
-As we will see, however, simulating abrirtary random vectors with a given Pearson correlation matrix is computationally intense [@Chen2001, @Xia17], violating property **S1**.
-On the other hand, an analyst may consider the use of non-parametric correlation measures, such as Spearman's $\rho$ and Kendall's $\tau$, particularly when modeling non-normal data as in our application.
-Adding to the complexity is the well-known, marginal-dependent constraints on the possible bivariate correlation (a consequence of the bounds of bivariate distribution functions --- the *Frechet-Hoeffding bounds* [@Nelsen2007]).
-Denote the pointwise upper bound as $M(y_i, y_{i^\prime})$ These bounds give strict restraints on the possible bounds and cannot be overcome through algorithm design.
-Yet not all multivariate simulation approaches obtain these bounds [for example REF].
-Computationally, algorithms that rely on serial computation scale poorly to high dimensions (refs - hierarchy) and fail to make use of modern high-performance computing including parallelization and graphical processing unit acceleration [@Li2019gpu].
-
-To meet the desired properties in the face of the challenging setting, we present a general-purpose, scalable multivariate simulation algorithm.
-The crux of the method lies in the construction of a Gaussian copula [refs].
-The idea that many multivariate and marginally heterogenous distribution can be constructed in such a manner is well known [refs].
-This article's contriubition lies in it's application to high-dimenisonal data through a high-performance implementation (`bigsimr`) and speed-focused algorithm design.
-The algorithm design relies on useful properties of non-standard correlation measures, namely Kendall's $\tau$ and Spearman's $\rho$.
-We'll show that quality of simulation does not require the use of the standard Pearson correlation matrix, even for normally distributed data.
-The purpose of our method is to generate random vectors that match exactly specified any possible Kendall's $\tau$ or Spearman's $\rho$, and approximately Pearson product-moment correlation.
-
-The study proceeds with more details on our motivating example in RNA-sequencing breast cancer data.
-Then we describe and justify our simulation methodology, followed by extensive Monte Carlo studies under various distributional assumputions --- emphasizing normal and non-normal scenarios.
-In these studies, we evaluate the accuracy and speed of our approach in comparison to other readily availble software.
-After evaluating *in silico*, we revisit our motivating
-example by employing our method and compare our simulations to empirically-derived statistics. Finally, we'll make concluding remarks regarding the method's utility and future directions.
+We apply our methodology to simulate RNA-sequencing data sets based on samples derived from breast cancer patients' tumors (the BRCA data set in TCGA).
+The data are freely available as part of the BRCA data set within the The Cancer Genome Atlas data warehouse and were downloaded on BLAH using the Harvard's Broad Institute's Firehouse interface.
+For each of the 1093 patients, the gene expression (abundance of messenger-RNA) from 20501 genes were counted using RNA-sequencing.
+The resultant data set can be represented as a matrix of size 1093 x 20501 with discrete values aligned to HUGO gene symbols for each patient.
+In this illustrative case study, we model the gene-wise marginal distributions as heterogeneous negative binomial with pmf given by Equation \@ref(eq:nb) with two parameters, $p$ and $r$ (OR SWITCH TO mu and delta).
+Modeling RNA-seq counts as negative binomially distributed random variables is commonplace (e.g., @Zhao2018) since overdispersion is often observed for most expressed genes.
+We begin by estimating the marginal negative binomial parameters for each group using a simple method of moments approach by matching the negative binomial parameters with the sample mean and variance.
+Table/Figure XXX summarizes the estimated means, variances, and NB parameters as described in Equation \@ref(eq:nb).
+Notably some genes show underdispersion compared to a Poisson random variable.
+For simplicity we exclude these cases, but nothing in the later described simulation method prevents using a different marginal distribution in those cases (such as a Maxwell distribution REF Sellers).
 
@@
 
-# Multivariate correlation bounds
+\begin{equation} 
+  g(n) = \mathbb P(Y=n) = \frac{\Gamma(n+r)}{\Gamma(r)n!} p^r(1-p)^n, \,\,\, n\in
+\mathbb N_0.
+  (\#eq:nb)
+\end{equation}
 
-# Simulation methodology
+```{r brca, echo=TRUE, eval=FALSE}
+## full workflow simulating RNA-seq data
 
-## General algorithm
-
-## Rank-based dependence measures
-
-## Pearson correlation
-
-Difficulties
-
-# The `bigsimr` R package
-
-Short description and key features, including GPU acceleration.
-
-```{r setup, eval=TRUE}
-## Using conda
-reticulate::use_condaenv("py37")
-library(bigsimr)
 ```
 
-## Basic use
+The data are overdispersed:
 
-short description
-
-### Specifying marginals
-
-As stated earlier, to generate multivariate data, we need a list of marginals (and their parameters), and a correlation structure (matrix).
-The marginal distributions can be built up as a list of lists, where each sublist contains the information for the target distribution.
-
-```{r}
-margins = list(
-  list("norm", mean = 3.14, sd = 0.1),
-  list("beta", shape1 = 1, shape2 = 4),
-  list("nbinom", size = 10, prob = 0.75)
-)
+```{r figNBratio, echo = FALSE, eval=FALSE, message=FALSE, warning=FALSE, fig.path='fig/plot-', out.width= '75%', fig.align='center', fig.ext='pdf', fig.cap="The genes display extreme overdispersion.
+Using group-specific sample means and variances for each of the d=4777 genes, we display the distributions oflog of variance to mean ratio (log(sigma2/mu))."}
+## http://www.sthda.com/english/wiki/ggplot2-histogram-easy-histogram-graph-with-ggplot2-r-package
+## devtools::install_github("kassambara/easyGgplot2")
+library(easyGgplot2)
+## p0 <- ggplot(data = nb_dat, aes(x = log(nb_var / nb_mean), y = stat(density), fill=group))
+## p0 + geom_histogram(bins = 30) + labs(x="log of variance to mean ratio") + theme_bw()
+nb_dat$logRatio <- log(nb_dat$nb_var / nb_dat$nb_mean)
+ggplot2.histogram(data=nb_dat, xName='logRatio',
+    groupName='group', legendPosition="right",
+    alpha=0.5, addDensity=TRUE,
+    addMeanLine=TRUE, meanLineColor="white", meanLineSize=1.5)
 ```
 
-The things to point out here are that in each sublist (marginal), the first item is an unnamed character string with the R name of the distribution *without a letter prefix*.
-E.g.
-instead of `rnorm`, we pass in just `"norm"`.
-The second thing to note is that the remaining items are *named* arguments that go along with the distribution.
-A full list of built-in distributions is found in the appendix.
+We see that the all genes studies display overdispersion (variance greater than mean) compared to a Poisson model for the gene counts.
+The genes are highly heterogeneous even when dealing with highly expressed genes after filtering.
+The facts taken together motivate the negative binomial model for the marginal distributions.
+The genes are slightly more variable within the deceased group (expected with smaller sample size and potentially due to biomedical considerations).
 
-### Specify correlation
+Next we perform some essential pre-processing of the data.
+Since our goal is to simulate meaningful multivariate constructions, we must first find a subset of 20501 genes that are expressed.
+In order words, we'll filter out the low expressing genes to allow greater range of possible $d$-variate correlations (see @NK10 for details OR SHOULD I EXPLAIN THIS MORE CAREFULLY?).
+Notably, many RNA-seq analytic workflows filter out low expressing genes (see \cite{Conesa2016b}) and this, admittedly, is usually done in an \emph {ad hoc} fashion.
+In this study, we prefer an inclusion criterion that is theoretically motivated.
+Based on the results in Section BLAH, we filter genes with a sample mean less than 12.
+In this way, we remove 4762 (23.2\%) of the genes, retaining 15,739 genes for further analysis.
+The filtered low-expressing genes will be simulated as independent negative binomial random variables to form a complete simulated transcriptome.
 
-The next step is to define a correlation structure for the multivariate distribution.
-This correlation matrix can either come from observed data, or we can set it ourselves, or we can generate a random correlation matrix via `bigsimr::rcor`.
-Let's create a simple correlation matrix where all off-diagonal elements are 0.5.
-Since we have 3 marginals, we need a $3\times 3$ matrix.
+RNA-sequencing data derived from breast cancer patients (n=1093), with 20501 genes measured (from TCGA).
+We filter to genes with average expression greater than 10000 counts (d = 4777) across both groups.
+Consider basic differentially expressed genes (DEG) analysis between surviving (n0 = 941) and deceased (n1 = 152) patients.
+This setting can be described as large-scale simultaneous hypothesis testing, under correlation.
+We should evaluate existing and new methodology using simulations that reflect dependency among variables.
+But multivariate simulation tools often scale poorly to high dimension or do not model the full range of dependency.
+Perform a two-sample $t$ test for each 4777 genes on the observed RNA-seq counts ($empirical$).
+Call a "Z score" for the $i^{th}$ gene $Z_{i} = \Phi^{-1}(t_{\approx 272}(t_i))$.
 
-```{r}
-rho <- matrix(0.5, nrow = 3, ncol = 3)
-diag(rho) <- 1.0
-rho
+Our method requires some measure of that the covariance structure is pre-specified.
+In practice, however, high-dimensional covariance estimation is not that easy.
+Indeed there is much recent interest in this area (cite a few recent high visibility articles --- see the GPU-NORTA pub).
+Here we do not provide a comprehensive review on the topic.
+Instead we only seek to use covariance estimator that only guarantees positive-semidefiniteness while maintaining adequate estimation properties.
+To this end we chose to use the condition-number-regularized (\textsc{condreg}) approach of \cite{Won2013g}.
+This covariance estimation procedure restricts the ratio of the largest eigenvalue to the smallest eigenvalue to improve numerical results.
+But the method employs a penalized Gaussian likelihood.
+And so, rather than estimating the covariance on the discrete counts directly, we first perform a $log_{2}(x +1)$ transformation to better suit this modeling assumption.
+\textsc{condreg} requires a parameter, $\kappa_{max}$, that controls the largest condition number of the resultant estimated covariance matrix.
+Here we aim to allow a large number of nonzero pairwise correlations and so arbitrary set $\kappa_{max}=10^{4}$.
+The routine ran without issue on a MacBook Pro carrying BLAH in XXX units of time.
+
+Most pairwise correlations are nearly zero with XXX \% less than |0.1|.
+Yet there are dense subsets of correlated genes.
+And even small correlations among many variables can disrupt the operating characteristics of commonly used statistical inferential procedures \cite{Schissler2018}.
+
+In this case study in the large-scale hypothesis testing in pursuit of detecting differentially expressed genes, we aim to simulate RNA-seq counts using a multivariate negative binomial.
+To evaluate the simulation's utility, we systematically compare the simulated results to the empirical results.
+In short we computed t-statistics and p-values for each of $d=4777$ genes between two groups of breast cancer patients (1=deceased, 0=alive; $n_0 = 152, n_1=941$).
+Further details on the empirical results are provided in Section BLAH.
+We compare two aspects in the simulation design 1) Choice of correlation measure while modeling marginal distributions consistently and 2) how well does the best-performing simulation agree with the empirical results.
+Informed by the simulation studies in Section BLAH, we anticipate a far amount of variation at such small sample sizes and large $d$ and so replicate the entire study 100 times.
+
+To recap from the discussion above (Section BLAH), we estimated negative binomial parameters for each of the $d=4777$ genes and also their Pearson, Spearman, and Kendall's correlation coefficients from n=1093 breast cancer patients, within each vital status group (0=deceased, 1=alive).
+We generated 100 synthetic data sets for each group, with the number of $d$-dimensional random vectors produced equal to the corresponding sample sizes ($n_0 = 152, n_1=941$).
+These 100 samples are used to understand uncertainty in this setting and we will use the mean simulated values to compare to the empirical (and also explore the worst-case scenarios).
+
+Model genes as marginally negative binomial with heterogeneous gene-wise parameters.
+Compute sample correlations (using desired dependency measure) for the 11,407,476 genes pairs, for each group.
+To simulate gene expression counts, our goal is to produce a random vector ${\bf Y}=(Y_1, \ldots, Y_d)$ with **correlated** NB components.
+To do that, we start with a sequence of **independent** Poisson processes $N_i(t)$, $i=1,\ldots,d$, where the rate of the process $N_i(t)$ is $\lambda_i=(1-p_i)/p_i>0$ (so that $p_i=1/(1+\lambda_i)$).
+Now, we let ${\bf T} = (T_1, \ldots T_d)$ have a multivariate distribution on $\mathbb R_+^d$ with the PDF $f_{{\bf T}}({\bf t})$.
+Then, we define
+
+We rescaled the target Spearman correlation to account for the probability of ties via Equation \@ref(eq:convertSpearmanDiscrete).
+The serial computation took approximately 10 minutes on XXX.
+Notably, the difference between target and adjusted matrices was very small (mean relative difference: 0.708) in this particular configuration of negative binomial parameters.
+For problems with smaller counts and higher probabilities of ties, the adjustment can be substantial (as observed in the simulation studies above in Section BLAH).
+
+- Kendall's $\tau$ was estimated efficiently using the `pcaPP::cor.fk` function in `R`.
+
+## Ultra High-Dimensional simulation-based correlation hypothesis testing
+
+```{r brcaSigmaTesting, echo=TRUE, eval=FALSE}
+## full workflow simulating RNA-seq data
+
 ```
 
-Finally we can generate a random vector with our specified marginals and correlation structure.
-The last argument, `type`, is looking to know what kind of correlation matrix it is receiving.
-Right now it can handle Pearson, Spearman, or Kendal.
+## Simulation-based computation of correlation bounds
 
-### Small sample
+Using the Generate, sort, and correlate algorithm @Demirtas2011
 
-```{r}
-x = rvec(10, rho = rho, params = margins, type = "pearson")
-```
-
-On my machine, there is no dedicated GPU, so I would see the following warning message once per session.
-
-```{r, echo=FALSE}
-warning("warning.warn('No GPU/TPU found, falling back to CPU.')")
-```
-
-Taking a look at our random vector, we see that it hs 10 rows and 3 columns, one column for each marginal.
-
-```{r}
-x
-```
-
-We can simulate many more samples and then check the histogram of each margin, as well as the estimated correlation between the columns.
-
-### Scaling up N
-
-```{r, fig.width=7}
-x = rvec(10000, rho = rho, params = margins, type = "pearson")
-
-par(mfrow=c(1,3))
-hist(x[,1], breaks = 30, xlab = "", main = "Normal")
-hist(x[,2], breaks = 30, xlab = "", main = "Beta")
-hist(x[,3], breaks = 30, xlab = "", main = "Negative Binomial")
-```
-
-### Evaluation
-
-```{r}
-cor(x)
-```
-
-We can see that even wih 10,000 samples, the estimated correlation of the simulated data is not exactly the same as the target correlation.
-This can be explained by the fact that some correlations are simply not possible due to the discrete nature of certain distributions.
-Another possibility is that the copula algorith is biased and needs correction.
-
-### List of supported distributions
-
-```{r, eval=FALSE}
-all_dists <- list(
-  list(dist = "beta", shape1, shape2),
-  list(dist = "binom", size, prob),
-  list(dist = "cauchy", location, scale),
-  list(dist = "chisq", df),
-  list(dist = "exp", rate),
-  list(dist = "f", df1, df2),
-  list(dist = "gamma", shape, rate),
-  list(dist = "geom", prob),
-  list(dist = "hyper", m, n, k),
-  list(dist = "logis", location, scale),
-  list(dist = "lnorm", meanlog, sdlog),
-  list(dist = "nbinom", size, prob),
-  list(dist = "norm", mean, sd),
-  list(dist = "pois", lambda),
-  list(dist = "t", df),
-  list(dist = "unif", min, max),
-  list(dist = "weibull", shape, scale),
-  list(dist = "wilcox", m, n),
-  list(dist = "signrank", n)
-)
-```
-
-## Comparsions to other software
-
-### Speed
-
-- `copula`
-- `Genord`
-- `Multiord`
-- other?
-- CPU benchmarking
-- GPU benchmarking
-
-### Accuracy
-
-blah
-
-### Other comparisons
-
-blah
-
-# Some applications
-
-## Simulation-based correlation checking
-
-A multivariate approach?
-
-Or all the pairs?
-
-Generate, sort, and correlate algorithm @Demirtas2011
-
-```{r, computeFrechetBounds, echo = TRUE, eval=TRUE}
+```{r, computeBounds, echo = TRUE, eval=FALSE}
 
@@
 
-## Simulating RNA-seq data
+## Simulation-based joint probability calculations
 
-We apply our methodology to simulate RNA-sequencing data sets based on samples derived from breast cancer patients' tumors (the BRCA data set in TCGA).
-The data are freely available as part of the BRCA data set within the The Cancer Genome Atlas data warehouse and were downloaded on BLAH using the Harvard's Broad Institute's Firehouse interface.
-For each of the 1093 patients, the gene expression (abundance of messenger-RNA) from 20501 genes were counted using RNA-sequencing.
-The resultant data set can be represented as a matrix of size 1093 x 20501 with discrete values aligned to HUGO gene symbols for each patient.
-In this illustrative case study, we model the genewise marginal distributions as hetereogeneous negative binomial with pmf given by Equation \@ref(eq:nb) with two parameters, $p$ and $r$ (OR SWITCH TO mu and delta).
-Modeling RNA-seq counts as negative binomially distributed random variables is commonplace (e.g., @Zhao2018) since overdispersion is often observed for most expressed genes.
-We begin by estimating the marginal negative binomial parameters for each group using a simple method of moments approach by matching the negative binomial parameters with the sample mean and variance.
-Table/Figure XXX summarizes the estimated means, variances, and NB parameters as described in Equation \@ref(eq:nb).
-Notably some genes show underdisperation compared to a Poisson random varible.
-For simplicity we exclude these cases, but nothing in the later described simulation method prevents using a different marginal distribution in those cases (such as a Maxwell distribution REF Sellers).
+```{r densityEvaluation, echo=TRUE, eval=FALSE}
+## full workflow simulating RNA-seq data
 
-\begin{equation} 
-  g(n) = \mathbb P(Y=n) = \frac{\Gamma(n+r)}{\Gamma(r)n!} p^r(1-p)^n, \,\,\, n\in
-\mathbb N_0.
-  (\#eq:nb)
-\end{equation}
-
-## Simulation-based correlation testing
-
-## Simulation-based probability calculations
-
-# Conclusion/Discussion
-
-- Although we recommend to use the rank-based measures
-- discuss the omission (or inclusion) of high dimensional covariance estimators?
-- We've shown that the multivariate NB simulations outperform the Independent sims.
-But the agreement is still poor.
-This could motivate other more appropriate models for high-expressing RNA-seq data than the negative binomial distribution.
-This algorithm lends itself to exploration of flexible probability models to inspire model selection in the developemnt novel RNA-seq analytic methodology.
-
-We've introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a high-performance implementation called `bigsimr` (github url).
-The parallelized (multi-core) algorithm is simple and easy to understand as an application of Gaussian copulas.
-This method is largely inspired by @MB13, but with contributions with regard to scaleability, implementation, and application in high-throughput biomedical data (RNA-sequencing).
-We advocate the use of Kendall's $\tau$ as it better captures correlation among components of non-normal, as well as non-linear patterns of association.
-Moreover, the matching Kendall's $\tau$ is nearly trival due to the invariant of monotone transformations, after an adjustment to the input correlation matrix.
-Interestingly, our simulation studies show the use of Kendall's $\tau$ to works as well as using Pearson correlation coefficient when simulating from multivariate normal distributions.
-
-We also show utility in our methodology through an application to differential gene expression analysis from RNA-sequencing data.
-The application results show that correlations indeed matter in the large-scale hypothesis testing, as many others have noted (for example, see @BE07, @Wu2012b).
-We also hope that the application provides an example workflow --- and other strategy to use in simulation design.
-Even the best-performing simulations we provide show a gap from the empirical distribution.
-To keep the demonstration straightforward, we did not attempt to match the empirical distribution of test statistics as precisely as possible.
-Yet our methodology could be more creatively applied to meet that goal.
-One could consider marginal distributions from different families, such as Poisson for a subset of genes.
-One could imagine finding a best fitting probability distribution among a class of distributions for each gene and this could perhaps a better fit.
-One could imagine additonal structures/features in the data that an analyset could model to improve the correspondence with the emprical values, for example row correlations and high-dimensional covariance estimators (see @Won2013g).
-
-There are of course limitations to the methodology and implementation.
-The algorithm requires invertible marginal cdfs.
-This leaves some restriction to the available marginal probability distributions (DOES IT?
-FOR EXAMPLE?) An additional source of error is after converting dependencies (step i; ${\bf R_{Kendall}})$) the resultant matrix may not be positive definite (PD).
-Then the nearest PD matrix is used.
-From a practical computing standpoint, the user's computing resource provides the ultimate limit on how large a dimension can be simulated using the `bigsimr` R package.
-A user must consider carefully the available memory and cores when conducting a Monte Carlo experiment.
-The algorithm does amends itself well to parallelization and graphical processing unit acceleration [@Li2019gpu] and advanced users may take advantage of those techniques.
-
-Future work includes developing scalable algorithms to match the Pearson correlation matrix exactly.
-Further, more sophicated approaches could involve simulation algorithms that are "estimation awarene" and so could explore the role of covariance estimation within the simulation.
-Lastly, these may reveal new approaches to estimating and evaluating high dimensional regression and Bayesian models.
-On the implementation side, employing graphical process unit acceleration could produce substaintial speedups over our multi-core approach.
-
-# Conclusions
-
-# Supplementary Materials
-
-We provide an open-source implementation of our methology as the `bigsimr` R package, hosted on github.
-
-# Acknowledgement(s)
-
-The authors gratefully acknowledge the helpful discussions with Professor Walter W. Piegorsch and Professor Edward J. Bedrick during this project's conception.
-
-# Disclosure statement
-
-The authors report no conflict of interest. DO WE?
-
-# Funding
-
-FUNDING FROM EVERYONE HERE. CP3 grant.
-
-# Nomenclature/Notation
-
-# Notes
-
-Mention Python?
+```

### 2020-06-04 060-conclusion-discussion.Rmd
split up document into sections. did spelling check

--- 2020-06-02 index.Rmd
@@
---- 
-title: "Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package"
-author: [A.G. Schissler, A. Knudson]
-date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"
-site: bookdown::bookdown_site
-output: bookdown::gitbook
-documentclass: article
-bibliography: [bigsimr.bib, packages.bib]
-biblio-style: apalike
-link-citations: yes
-abstract: |
-  In this era of Big Data, it is critical to realistically simulate data to conduct informative Monte Carlo studies.
-This is problematic when data are inherently multivariate while at the same time are (ultra-) high dimensional.
-This situation appears frequently in observational data found on online and in high-throughput biomedical experiments (e.g., RNA-sequencing).
-Due to the difficulty in simulating realistic correlated data points, researchers often resort to simulation designs that posit independence --- greatly diminishing the insight into the empirical operating characteristics of any proposed methodology.
-Major challenges lie in the computational complexity involved in simulating these massive random vectors.
-We propose a fairly general, scalable procedure to simulate high-dimensional multivariate distributions with pre-specified marginal characteristics and dependency characteristics.
-As a motivating example, we use our methodology to study large-scale statistical inferential procedures applied to cancer-related RNA-sequencing data sets.
-The proposed algorithm is implemented as the `bigsimr` R package.
----
-
-```{r LoadLib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
-library(ggplot2)
-library(tidyverse)
-library(knitr)
-library(dplyr)
-set.seed(06082020)
-```
-
-```{bash copyBib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
-cp /Users/alfred/Dropbox/bib/bigsimr.bib ./.
-```
-
-```{r include=FALSE}
-# automatically create a bib database for R packages
-knitr::write_bib(c(
-  .packages(), 'bookdown', 'knitr', 'rmarkdown'
-), 'packages.bib')
-```
-
-# Introduction
-
-Massive high-dimensional data sets are commonplace in many areas of scientific inquiry.
-As new methods are developed for these data, a fundamental challenge lies in designing and conducting simulation studies to assess operating characteristics of proposed methodology, such as false positive rates, statistical power, and robustness --- often in comparison to existing methods.
-Another application lies in the discovering the sampling distribution of test statistic under hypothesized null models in complex scenarios.
-Such Monte Carlo studies in the high-dimensional setting often become difficult to conduct with existing algorithms and tools, or require expertise in computing on clusters.
-This is particularly true when simulating massive multivariate, non-normal distributions.
-
-Along those lines, simulating high-dimension dependent negative binomial data motivates this work --- in pursuit of modeling RNA-sequencing data [@Wang2009b; @Conesa2016b] derived from breast cancer patients.
-This labroratory procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions.
-Often researchers posit a negative binomial model (refs) as counts are often over (or under) expressed that a Poisson model would suggest.
-Yet due to inherent biological processes, gene expression data exhibits correlation (coexpression) across genes [@BE07; @Schissler2018].
-RNA-sequencing analysis has garnered much interest in the statistical literature (refs).
-Often researchers simulate independently or from their proposed model in order condunct Monte Carlo studies (refs).
-An alternavitive strategy is to think *generatively* about the data collection process and scientific domain knowledge and design simulations with probabilitistic models that reflect those processes.
-In our motivating example, we'll study the consequences correlation in the simulation study replicating test statistics from an illustrative, classical large-scale hypothesis testing study [see @Efron2004].
-
-With our application in mind and seeking a general-purpose algorithm with broad applications, our purposed methodology should possess the following properities (adapted from criteria from @Nik13a):
-
-* P1: Wide range of dependence, allowing both positive and negative dependence
-* P2: Flexible dependence, meaning that the number of bivariate marginals is (approximately) equal to the number of dependence parameters.
-* P3: Flexible marginal modeling, generating hetereogenous data --- possibly from differing probability families.
-* P4: Computationally feasible cumulative distribution function (cdf) for likelihood estimation.
-
-Moreover, in simulation method must scale to high dimensions:
-
-* S1: Procedure must scale to high dimensions, computable in a reasonable amount time.
-* S2: Procedure must scale to high dimensions while maintaining accuracy.
-
-As others @MB13 have noted, however, simulating dependent data can be challenging.
-A central issue lies characterizing dependency between components in the high-dimensional random vector.
-The choice of correlation in typical practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g.
-non-normal, discrete, finite support, etc).
-Here we propose a scheme that ignores the analytic goal --- and, instead, aims to match the empirically-derived estimated correlation data and marginal characteristics.
-For normal data, the Pearson product-moment correlation describes the dependency completely.
-As we will see, however, simulating abrirtary random vectors with a given Pearson correlation matrix is computationally intense [@Chen2001, @Xia17], violating property **S1**.
-On the other hand, an analyst may consider the use of non-parametric correlation measures, such as Spearman's $\rho$ and Kendall's $\tau$, particularly when modeling non-normal data as in our application.
-Adding to the complexity is the well-known, marginal-dependent constraints on the possible bivariate correlation (a consequence of the bounds of bivariate distribution functions --- the *Frechet-Hoeffding bounds* [@Nelsen2007]).
-Denote the pointwise upper bound as $M(y_i, y_{i^\prime})$ These bounds give strict restraints on the possible bounds and cannot be overcome through algorithm design.
-Yet not all multivariate simulation approaches obtain these bounds [for example REF].
-Computationally, algorithms that rely on serial computation scale poorly to high dimensions (refs - hierarchy) and fail to make use of modern high-performance computing including parallelization and graphical processing unit acceleration [@Li2019gpu].
-
-To meet the desired properties in the face of the challenging setting, we present a general-purpose, scalable multivariate simulation algorithm.
-The crux of the method lies in the construction of a Gaussian copula [refs].
-The idea that many multivariate and marginally heterogenous distribution can be constructed in such a manner is well known [refs].
-This article's contriubition lies in it's application to high-dimenisonal data through a high-performance implementation (`bigsimr`) and speed-focused algorithm design.
-The algorithm design relies on useful properties of non-standard correlation measures, namely Kendall's $\tau$ and Spearman's $\rho$.
-We'll show that quality of simulation does not require the use of the standard Pearson correlation matrix, even for normally distributed data.
-The purpose of our method is to generate random vectors that match exactly specified any possible Kendall's $\tau$ or Spearman's $\rho$, and approximately Pearson product-moment correlation.
-
-The study proceeds with more details on our motivating example in RNA-sequencing breast cancer data.
-Then we describe and justify our simulation methodology, followed by extensive Monte Carlo studies under various distributional assumputions --- emphasizing normal and non-normal scenarios.
-In these studies, we evaluate the accuracy and speed of our approach in comparison to other readily availble software.
-After evaluating *in silico*, we revisit our motivating
-example by employing our method and compare our simulations to empirically-derived statistics. Finally, we'll make concluding remarks regarding the method's utility and future directions.
-
-Intergene correlation [@Schissler2019].
-
-# Multivariate correlation bounds
-
-# Simulation methodology
-
-## General algorithm
-
-## Rank-based dependence measures
-
-## Pearson correlation
-
-Difficulties
-
-# The `bigsimr` R package
-
-Short description and key features, including GPU acceleration.
-
-```{r setup, eval=TRUE}
-## Using conda
-reticulate::use_condaenv("py37")
-library(bigsimr)
-```
-
-## Basic use
-
-short description
-
-### Specifying marginals
-
-As stated earlier, to generate multivariate data, we need a list of marginals (and their parameters), and a correlation structure (matrix).
-The marginal distributions can be built up as a list of lists, where each sublist contains the information for the target distribution.
-
-```{r}
-margins = list(
-  list("norm", mean = 3.14, sd = 0.1),
-  list("beta", shape1 = 1, shape2 = 4),
-  list("nbinom", size = 10, prob = 0.75)
-)
-```
-
-The things to point out here are that in each sublist (marginal), the first item is an unnamed character string with the R name of the distribution *without a letter prefix*.
-E.g.
-instead of `rnorm`, we pass in just `"norm"`.
-The second thing to note is that the remaining items are *named* arguments that go along with the distribution.
-A full list of built-in distributions is found in the appendix.
-
-### Specify correlation
-
-The next step is to define a correlation structure for the multivariate distribution.
-This correlation matrix can either come from observed data, or we can set it ourselves, or we can generate a random correlation matrix via `bigsimr::rcor`.
-Let's create a simple correlation matrix where all off-diagonal elements are 0.5.
-Since we have 3 marginals, we need a $3\times 3$ matrix.
-
-```{r}
-rho <- matrix(0.5, nrow = 3, ncol = 3)
-diag(rho) <- 1.0
-rho
-```
-
-Finally we can generate a random vector with our specified marginals and correlation structure.
-The last argument, `type`, is looking to know what kind of correlation matrix it is receiving.
-Right now it can handle Pearson, Spearman, or Kendal.
-
-### Small sample
-
-```{r}
-x = rvec(10, rho = rho, params = margins, type = "pearson")
-```
-
-On my machine, there is no dedicated GPU, so I would see the following warning message once per session.
-
-```{r, echo=FALSE}
-warning("warning.warn('No GPU/TPU found, falling back to CPU.')")
-```
-
-Taking a look at our random vector, we see that it hs 10 rows and 3 columns, one column for each marginal.
-
-```{r}
-x
-```
-
-We can simulate many more samples and then check the histogram of each margin, as well as the estimated correlation between the columns.
-
-### Scaling up N
-
-```{r, fig.width=7}
-x = rvec(10000, rho = rho, params = margins, type = "pearson")
-
-par(mfrow=c(1,3))
-hist(x[,1], breaks = 30, xlab = "", main = "Normal")
-hist(x[,2], breaks = 30, xlab = "", main = "Beta")
-hist(x[,3], breaks = 30, xlab = "", main = "Negative Binomial")
-```
-
-### Evaluation
-
-```{r}
-cor(x)
-```
-
-We can see that even wih 10,000 samples, the estimated correlation of the simulated data is not exactly the same as the target correlation.
-This can be explained by the fact that some correlations are simply not possible due to the discrete nature of certain distributions.
-Another possibility is that the copula algorith is biased and needs correction.
-
-### List of supported distributions
-
-```{r, eval=FALSE}
-all_dists <- list(
-  list(dist = "beta", shape1, shape2),
-  list(dist = "binom", size, prob),
-  list(dist = "cauchy", location, scale),
-  list(dist = "chisq", df),
-  list(dist = "exp", rate),
-  list(dist = "f", df1, df2),
-  list(dist = "gamma", shape, rate),
-  list(dist = "geom", prob),
-  list(dist = "hyper", m, n, k),
-  list(dist = "logis", location, scale),
-  list(dist = "lnorm", meanlog, sdlog),
-  list(dist = "nbinom", size, prob),
-  list(dist = "norm", mean, sd),
-  list(dist = "pois", lambda),
-  list(dist = "t", df),
-  list(dist = "unif", min, max),
-  list(dist = "weibull", shape, scale),
-  list(dist = "wilcox", m, n),
-  list(dist = "signrank", n)
-)
-```
-
-## Comparsions to other software
-
-### Speed
-
-- `copula`
-- `Genord`
-- `Multiord`
-- other?
-- CPU benchmarking
-- GPU benchmarking
-
-### Accuracy
-
-blah
-
-### Other comparisons
-
-blah
-
-# Some applications
-
-## Simulation-based correlation checking
-
-A multivariate approach?
-
-Or all the pairs?
-
-Generate, sort, and correlate algorithm @Demirtas2011
-
-```{r, computeFrechetBounds, echo = TRUE, eval=TRUE}
-
-## simulation based
-## devtools::install_github("adknudson/bigsimr", ref="develop")
-library(bigsimr)
-## ?rvec
-
-d <- 5
-rho <- rcor(d)
-margins <- list(
-    list("nbinom", size = 5, prob = 0.3),
-    list("exp", rate = 4),
-    list("binom", size = 5, prob = 0.7),
-    list("norm", mean = 10, sd = 3),
-    list("pois", lambda = 10)
-)
-
-margins2 <- list(
-    list("nbinom", 5, 0.3),
-    list("exp", 4),
-    list("binom", 5, 0.7),
-    list("norm", 10, 3),
-    list("pois", 10)
-)
-
-params = margins
-rho = rho
-## cores = parallel::detectCores() - 1
-cores = 2
-reps = 1e3
-type = "spearman"
-
-## help(package="bigsimr")
-rho_bounds <- computeCorBounds(params = margins, cores = cores, type = "spearman", reps = reps)
-rho_bounds
-```
-
-## Simulating RNA-seq data
-
-We apply our methodology to simulate RNA-sequencing data sets based on samples derived from breast cancer patients' tumors (the BRCA data set in TCGA).
-The data are freely available as part of the BRCA data set within the The Cancer Genome Atlas data warehouse and were downloaded on BLAH using the Harvard's Broad Institute's Firehouse interface.
-For each of the 1093 patients, the gene expression (abundance of messenger-RNA) from 20501 genes were counted using RNA-sequencing.
-The resultant data set can be represented as a matrix of size 1093 x 20501 with discrete values aligned to HUGO gene symbols for each patient.
-In this illustrative case study, we model the genewise marginal distributions as hetereogeneous negative binomial with pmf given by Equation \@ref(eq:nb) with two parameters, $p$ and $r$ (OR SWITCH TO mu and delta).
-Modeling RNA-seq counts as negative binomially distributed random variables is commonplace (e.g., @Zhao2018) since overdispersion is often observed for most expressed genes.
-We begin by estimating the marginal negative binomial parameters for each group using a simple method of moments approach by matching the negative binomial parameters with the sample mean and variance.
-Table/Figure XXX summarizes the estimated means, variances, and NB parameters as described in Equation \@ref(eq:nb).
-Notably some genes show underdisperation compared to a Poisson random varible.
-For simplicity we exclude these cases, but nothing in the later described simulation method prevents using a different marginal distribution in those cases (such as a Maxwell distribution REF Sellers).
-
-\begin{equation} 
-  g(n) = \mathbb P(Y=n) = \frac{\Gamma(n+r)}{\Gamma(r)n!} p^r(1-p)^n, \,\,\, n\in
-\mathbb N_0.
-  (\#eq:nb)
-\end{equation}
-
-## Simulation-based correlation testing
-
-## Simulation-based probability calculations
-
 # Conclusion/Discussion
@@
 This could motivate other more appropriate models for high-expressing RNA-seq data than the negative binomial distribution.
-This algorithm lends itself to exploration of flexible probability models to inspire model selection in the developemnt novel RNA-seq analytic methodology.
+This algorithm lends itself to exploration of flexible probability models to inspire model selection in the development novel RNA-seq analytic methodology.
 
@@
 The parallelized (multi-core) algorithm is simple and easy to understand as an application of Gaussian copulas.
-This method is largely inspired by @MB13, but with contributions with regard to scaleability, implementation, and application in high-throughput biomedical data (RNA-sequencing).
+This method is largely inspired by @MB13, but with contributions with regard to scalability, implementation, and application in high-throughput biomedical data (RNA-sequencing).
 We advocate the use of Kendall's $\tau$ as it better captures correlation among components of non-normal, as well as non-linear patterns of association.
-Moreover, the matching Kendall's $\tau$ is nearly trival due to the invariant of monotone transformations, after an adjustment to the input correlation matrix.
+Moreover, the matching Kendall's $\tau$ is nearly trivial due to the invariant of monotone transformations, after an adjustment to the input correlation matrix.
 Interestingly, our simulation studies show the use of Kendall's $\tau$ to works as well as using Pearson correlation coefficient when simulating from multivariate normal distributions.
@@
 One could imagine finding a best fitting probability distribution among a class of distributions for each gene and this could perhaps a better fit.
-One could imagine additonal structures/features in the data that an analyset could model to improve the correspondence with the emprical values, for example row correlations and high-dimensional covariance estimators (see @Won2013g).
+One could imagine additional structures/features in the data that an analyst could model to improve the correspondence with the empirical values, for example row correlations and high-dimensional covariance estimators (see @Won2013g).
 
@@
 Future work includes developing scalable algorithms to match the Pearson correlation matrix exactly.
-Further, more sophicated approaches could involve simulation algorithms that are "estimation awarene" and so could explore the role of covariance estimation within the simulation.
+Further, more sophisticated approaches could involve simulation algorithms that are "estimation aware" and so could explore the role of covariance estimation within the simulation.
 Lastly, these may reveal new approaches to estimating and evaluating high dimensional regression and Bayesian models.
-On the implementation side, employing graphical process unit acceleration could produce substaintial speedups over our multi-core approach.
-
-# Conclusions
-
-# Supplementary Materials
-
-We provide an open-source implementation of our methology as the `bigsimr` R package, hosted on github.
-
-# Acknowledgement(s)
-
-The authors gratefully acknowledge the helpful discussions with Professor Walter W. Piegorsch and Professor Edward J. Bedrick during this project's conception.
-
-# Disclosure statement
-
-The authors report no conflict of interest. DO WE?
-
-# Funding
-
-FUNDING FROM EVERYONE HERE. CP3 grant.
-
-# Nomenclature/Notation
-
-# Notes
-
-Mention Python?
+On the implementation side, employing graphical process unit acceleration could produce substantial speedups over our multi-core approach.

### 2020-06-04 docs/article_bigsimr.tex
split up document into sections. did spelling check

--- 2020-06-01 docs/article_bigsimr.tex
@@
 \hypersetup{
-  pdftitle={GIVE ME A TITLE},
-  pdfauthor={A.G. Schissler; A.B. PERSON},
+  pdftitle={Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package},
+  pdfauthor={A.G. Schissler; A. Knudson},
   hidelinks,
@@
 \urlstyle{same} % disable monospaced font for URLs
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{248,248,248}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 \usepackage{longtable,booktabs}
@@
 
-\title{GIVE ME A TITLE}
-\author{A.G. Schissler \and A.B. PERSON}
-\date{Monday, June 01, 2020}
+\title{Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package}
+\author{A.G. Schissler \and A. Knudson}
+\date{Thursday, June 04, 2020}
 
@@
 \maketitle
+\begin{abstract}
+In this era of Big Data, it is critical to realistically simulate data to conduct informative Monte Carlo studies.
+This is problematic when data are inherently multivariate while at the same time are (ultra-) high dimensional.
+This situation appears frequently in observational data found on online and in high-throughput biomedical experiments (e.g., RNA-sequencing).
+Due to the difficulty in simulating realistic correlated data points, researchers often resort to simulation designs that posit independence --- greatly diminishing the insight into the empirical operating characteristics of any proposed methodology.
+Major challenges lie in the computational complexity involved in simulating these massive random vectors.
+We propose a fairly general, scalable procedure to simulate high-dimensional multivariate distributions with pre-specified marginal characteristics and dependency characteristics.
+As a motivating example, we use our methodology to study large-scale statistical inferential procedures applied to cancer-related RNA-sequencing data sets.
+The proposed algorithm is implemented as the \texttt{bigsimr} R package.
+\end{abstract}
 
@@
 }
-\hypertarget{prerequisites}{%
-\section{Prerequisites}\label{prerequisites}}
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
-This is a \emph{sample} book written in \textbf{Markdown}. You can use anything that Pandoc's Markdown supports, e.g., a math equation \(a^2 + b^2 = c^2\).
+Massive high-dimensional data sets are commonplace in many areas of scientific inquiry.
+As new methods are developed for these data, a fundamental challenge lies in designing and conducting simulation studies to assess operating characteristics of proposed methodology, such as false positive rates, statistical power, and robustness --- often in comparison to existing methods.
+Another application lies in the discovering the sampling distribution of test statistic under hypothesized null models in complex scenarios.
+Such Monte Carlo studies in the high-dimensional setting often become difficult to conduct with existing algorithms and tools, or require expertise in computing on clusters.
+This is particularly true when simulating massive multivariate, non-normal distributions.
 
-The \textbf{bookdown} package can be installed from CRAN or Github:
+Along those lines, simulating high-dimension dependent negative binomial data motivates this work --- in pursuit of modeling RNA-sequencing data \citep{Wang2009b, Conesa2016b} derived from breast cancer patients.
+This laboratory procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions.
+Often researchers posit a negative binomial model (refs) as counts are often over (or under) expressed that a Poisson model would suggest.
+Yet due to inherent biological processes, gene expression data exhibits correlation (coexpression) across genes \citep{BE07, Schissler2018}.
+RNA-sequencing analysis has garnered much interest in the statistical literature (refs).
+Often researchers simulate independently or from their proposed model in order to conduct Monte Carlo studies (refs).
+An alternative strategy is to think \emph{generatively} about the data collection process and scientific domain knowledge and design simulations with probabilistic models that reflect those processes.
+In our motivating example, we'll study the consequences correlation in the simulation study replicating test statistics from an illustrative, classical large-scale hypothesis testing study \citep[see][]{Efron2004}.
 
-Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading \texttt{\#}.
+\hypertarget{ultra-high-dimensional-multivariate-modeling-and-simulation-desired-properties}{%
+\subsection{Ultra High Dimensional multivariate modeling and simulation desired properties}\label{ultra-high-dimensional-multivariate-modeling-and-simulation-desired-properties}}
 
-To compile this example to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): \url{https://yihui.name/tinytex/}.
+With our application in mind and seeking a general-purpose algorithm with broad applications, our purposed methodology should possess the following properties (adapted from criteria from \citet{Nik13a}):
 
-  \bibliography{bibliography.bib,packages.bib}
+\begin{itemize}
+\tightlist
+\item
+  P1: Wide range of dependence, allowing both positive and negative dependence
+\item
+  P2: Flexible dependence, meaning that the number of bivariate marginals is (approximately) equal to the number of dependence parameters.
+\item
+  P3: Flexible marginal modeling, generating heterogeneous data --- possibly from differing probability families.
+\end{itemize}
+
+Moreover, the simulation method must scale to high dimensions:
+
+\begin{itemize}
+\tightlist
+\item
+  S1: Procedure must scale to high dimensions, computable in a reasonable amount time.
+\item
+  S2: Procedure must scale to high dimensions while maintaining accuracy.
+\end{itemize}
+
+As others \citet{MB13} have noted, however, simulating dependent data can be challenging.
+A central issue lies characterizing dependency between components in the high-dimensional random vector.
+The choice of correlation in typical practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g.~non-normal, discrete, finite support, etc).
+Here we propose a scheme that ignores the analytic goal --- and, instead, aims to match the empirically-derived estimated correlation data and marginal characteristics.
+For normal data, the Pearson product-moment correlation describes the dependency completely.
+As we will see, however, simulating arbitrary random vectors with a given Pearson correlation matrix is computationally intense \citep[\citet{Xia17}]{Chen2001}, violating property \textbf{S1}.
+On the other hand, an analyst may consider the use of non-parametric correlation measures, such as Spearman's \(\rho\) and Kendall's \(\tau\), particularly when modeling non-normal data as in our application.
+Adding to the complexity is the well-known, marginal-dependent constraints on the possible bivariate correlation (a consequence of the bounds of bivariate distribution functions --- the \emph{Frechet-Hoeffding bounds} \citep{Nelsen2007}).
+Denote the pointwise upper bound as \(M(y_i, y_{i^\prime})\) These bounds give strict restraints on the possible bounds and cannot be overcome through algorithm design.
+Yet not all multivariate simulation approaches obtain these bounds {[}for example REF{]}.
+Computationally, algorithms that rely on serial computation scale poorly to high dimensions (refs - hierarchy) and fail to make use of modern high-performance computing including parallelization and graphical processing unit acceleration \citep{Li2019gpu}.
+
+To meet the desired properties in the face of the challenging setting, we present a general-purpose, scalable multivariate simulation algorithm.
+The crux of the method lies in the construction of a Gaussian copula {[}refs{]}.
+The idea that many multivariate and marginally heterogeneous distribution can be constructed in such a manner is well known {[}refs{]}.
+This article's contribution lies in it's application to high-dimensional data through a high-performance implementation (\texttt{bigsimr}) and speed-focused algorithm design.
+The algorithm design relies on useful properties of non-standard correlation measures, namely Kendall's \(\tau\) and Spearman's \(\rho\).
+We'll show that quality of simulation does not require the use of the standard Pearson correlation matrix, even for normally distributed data.
+The purpose of our method is to generate random vectors that match exactly specified any possible Kendall's \(\tau\) or Spearman's \(\rho\), and approximately Pearson product-moment correlation.
+
+The study proceeds with more details on our motivating example in RNA-sequencing breast cancer data.
+Then we describe and justify our simulation methodology, followed by extensive Monte Carlo studies under various distributional assumptions --- emphasizing normal and non-normal scenarios.
+In these studies, we evaluate the accuracy and speed of our approach in comparison to other readily available software.
+After evaluating \emph{in silico}, we revisit our motivating
+example by employing our method and compare our simulations to empirically-derived statistics. Finally, we'll make concluding remarks regarding the method's utility and future directions.
+
+\hypertarget{background-and-notation}{%
+\section{Background and notation}\label{background-and-notation}}
+
+\hypertarget{gaussian-copulas}{%
+\subsection{Gaussian copulas}\label{gaussian-copulas}}
+
+We present an general-purpose, scalable multivariate simulation algorithm.
+The crux of the method is construction of a Gaussian copula.
+This idea is well known {[}refs{]}.
+A copula is a distribution function on \([0,1]^d\) describing a random vector with standard uniform marginals.
+Moreover, for any random vector \({\bf X}=(X_1, \ldots, X_d)\) with cumulative distribution function (CDF) \(F\) and marginal CDFs \(F_i\) there is a copula function
+\(C(u_1, \ldots, u_d)\) so that
+
+\[
+F(x_1, \ldots,x_d) = \mathbb P(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), \,\,\, x_i\in \mathbb R, i=1,\ldots,d. 
+\]
+
+A Gaussian copula is the case where all marginal CDFs \(F_i\) are the standard normal cdf, \(\Phi\).
+The Gaussian copula is the one that corresponds to a multivariate normal distribution with standard normal marginal distributions and covariance matrix \({\bf R}\).
+(Since the marginals are standard normal, this \({\bf R}\) is also the correlation matrix).
+If \(F_{{\bf R}}\) is the CDF of such multivariate normal distribution, then the corresponding Gaussian copula \(C_{{\bf R}}\) is defined through
+
+\begin{equation}
+\label{eq:gauss}
+F_{{\bf R}}(x_1, \ldots, x_d) = C_{{\bf R}}(\Phi(x_1), \ldots, \Phi(x_d)),
+\end{equation}
+
+where \(\Phi(\cdot)\) is the standard normal CDF.
+Note that the copula \(C_{{\bf R}}\) is simply the CDF of the random vector \((\Phi(X_1), \ldots, \Phi(X_d))\), where \((X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R})\).
+
+Sklar's Theorem (ref) guarantee's that any random vector with computational feasible inverse CDFs (P4 above) and obtainable correlation matrix (within the Frechet bounds) can be obtained via transformations involving copula functions.
+Namely, to simulate a random vector \({\bf Y} = (Y_1, \ldots, Y_d)\), where \(Y_i=F_i^{-1}(U_i)\), \(i=1, \ldots, d\), we can construct a Gaussian copula \({\bf U} = (U_1, \ldots, U_d)\) viz \(U_i=\Phi(X_i)\), \(i=1, \ldots, d\).
+When an \(Y_i\) is discrete, some care must be taken to define \(F_i\).
+Letting
+
+\begin{equation}
+F_{i}^{-1} = inf\{y:F_{i}(y) \geq u \}
+\label{eq:inverseCDF}
+\end{equation}
+
+\noindent ensures that \(Y_i \sim F_i\).
+
+Simulation of a general multivariate random vector \({\bf T}\) with margins \(F_i\) based on this copula is quite simple.
+Ensuring that one obtains a certain dependence among the marginal distributions, however, proves challenging in our proposed Gaussian-copula-based scheme.
+Two core issues: 1) marginal characteristics induce bounds on the possible bivariate correlations and 2) monotone transformations deform the Pearson correlation which no closed form expression exists in general (ref?
+Chen2001).
+The Frechet bounds are well-known.
+
+\hypertarget{measures-of-dependency}{%
+\subsection{Measures of dependency}\label{measures-of-dependency}}
+
+The population correlation coefficient, commonly called the Pearson (product-moment) correlation coefficient, describes the linear association between two random variables \(X\) and \(Y\) and is given by
+
+\begin{equation}
+\label{eq:pearson}
+\rho(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ var(X)var(Y)\right]^{1/2}} 
+\end{equation}
+
+As \citet{MB13} and \citet{MK01} discuss, for a bivariate normal \((X,Y)\) random vector the Pearson correlation adequately describes the dependency between the components.
+\(\rho\)'s utility in non-normal or non-linear associations is lacking (\citet{MK01}).
+Rank-based (ordinal) approaches performance better in these settings, such as Spearman's (denoted \(\rho_s\)) and Kendall's \(\tau\).
+Define
+
+\begin{equation}
+\label{eq:spearman}
+\rho_s(X,Y) = 3 \left[ P\left[ (X_1 - X_2)(Y_1-Y_3) > 0 \right] - P\left[ (X_1 - X_2)(Y_1-Y_3) < 0 \right] \right]
+\end{equation}
+
+\noindent where \((X_1, Y_1) \overset{d}{=} (X,Y), X_2 \overset{d}{=} X, Y_3 \overset{d}{=} Y\) with \(X_2\) and \(Y_3\) are independent of one other and of \((X_1, Y_1)\).
+For continuous marginals, the measure works well due to zero probability of ties.
+For discrete marginals, however, one could perform a rescaled version of \(\rho_s\) for random variables \(X,Y\) with pmfs (or pdfs) \(p(x)\) and \(q(y)\), respectively.
+
+\begin{equation}
+\label{eq:spearmanRescaled}
+\rho_{RS}(X,Y) = \frac{\rho_s(X,Y)}{ \left[ \left[ 1 - \sum_x p(x)^3 \right] \left[ 1 - \sum_y q(y)^3 \right] \right]^{1/2}}
+\end{equation}
+
+\noindent Kendall's \(\tau\) is the probability of concordant pairs minus the probability of discordant pairs and is given compactly as
+
+\begin{equation}
+  \label{eq:tau}
+  \tau(X,Y) = \frac{\rho_s(X,Y)}{ \left[ \left[ 1 - \sum_x p(x)^3 \right] \left[ 1 - \sum_y q(y)^3 \right] \right]^{1/2}}
+\end{equation}
+
+\hypertarget{marginal-dependent-bivariate-correlation-bounds}{%
+\subsection{Marginal-dependent bivariate correlation bounds}\label{marginal-dependent-bivariate-correlation-bounds}}
+
+The pairwise correlation between two correlated random variables cannot in general obtain the full range of possible values, \([-1,-1]\).
+The range, called the Frechet(-Hoeffding) bounds, is a well-known function of the marginal distributions and are given by \citep{BF17}:
+
+\begin{equation}
+\label{eq:frechet}
+\rho^{max} = \rho \left( F^{-1}_1 (U), F^{-1}_2 (U) \right), \quad \rho^{min} = \rho \left( F^{-1}_1 (U), F^{-1}_2 (1 - U) \right)
+\end{equation}
+
+\noindent where \(U\) is a uniform random variable in \((0,1)\), and \(F^{-1}_1, F^{-1}_2\) are the inverse cdf of random variables \(X_1\) and \(X_2\), respectively.
+For discrete random variables, define \(F^{-1}\) as in Equation \eqref{eq:inverseCDF}.
+
+\hypertarget{simulation-algorithm}{%
+\section{Simulation algorithm}\label{simulation-algorithm}}
+
+\hypertarget{algorithm-description}{%
+\subsection{Algorithm description}\label{algorithm-description}}
+
+This section describes the method for simulating a random vector \(\bf Y\) with \(Y_i\) components for \(i=1,2,\ldots,d\). Each \(Y_i\) has a specified marginal
+distribution function \(F_i\) and its inverse.
+To characterize dependency, every pair \((Y_i, Y_j)\) has either a specified Pearson correlation \eqref{eq:pearson}, (rescaled) Spearman correlation \eqref{eq:spearmanRescaled} , or Kendall's \(\tau\) (Equation \eqref{eq:tau} ).
+The method only approximately matches the Pearson correlation in general, whereas the rank-based methods are exact.
+
+The method is best understand as a \textbf{parallelized Gaussian copula} (see Equation \eqref{eq:gauss}.
+We shall see that constructing continuous joint distributions that match a target Spearman or Kendall's correlations computes easily when employing Gaussian copulas, since this measures are invariant under the monotone transformations involved {[}refs{]}.
+To do this, we take advantage of a closed form relationship {[}ref?{]} between Kendall's \(\tau\) and Pearson's correlation coefficient for bivariate normal random variables:
+
+\begin{equation}
+\label{eq:convertKendall}
+r_{Pearson} = sin \left( \tau_{Kendall} \times \frac{\pi}{2} \right), 
+\end{equation}
+
+\noindent and similarly for Spearman's \(\rho\) \citep{K58},
+
+\begin{equation}
+\label{eq:convertSpearman}
+\rho_{Pearson} = 2 \times sin \left( \rho_{Spearman} \times \frac{\pi}{6} \right).
+\end{equation}
+
+For discrete marginals, achieving a target Spearman correlation under this scheme is possible by using components from Equation \eqref{eq:spearmanRescaled} to further adjust the input correlation matrix.
+Let the unscaled Spearman correlation coefficients be \(\rho_{s} \left(Y_{i}, Y_{i^\prime}\right)\) for two marginal distributions and divide the target correlation by the product in the denominator of Equation \eqref{eq:spearmanRescaled}.
+Let these adjustment factors be denoted as \(a_i = \left[ 1 - \sum_y p_i(y)^3 \right]^{1/2}\) and specifically rescale the target Spearman correlation matrix by
+
+\begin{equation}
+\label{eq:convertSpearmanDiscrete}
+\rho_{rs} \left(Y_{i}, Y_{i^\prime}\right) = \frac{\rho_{s} \left(Y_{i}, Y_{i^\prime}\right)}{a_i \times a_{i^\prime}}.
+\end{equation}
+
+In a similar fashion, we rescale Kendall's \(\tau\) to adjust the input correlation matrix. The conversion formula is given by
+
+\begin{equation}
+\label{eq:convertKendallDiscrete}
+\rho_{rs} \left(Y_{i}, Y_{i^\prime}\right) = \frac{\rho_{s} \left(Y_{i}, Y_{i^\prime}\right)}{a_i \times a_{i^\prime}}.
+\end{equation}
+
+In contrast the rank-based correlations, matching specified Pearson correlation coefficients exactly is computational intense in this scheme.
+In general, there is no closed form correspondence and involving computing or approximating \(\binom{d}{2}\) integrals of the form \(EY_iY_j = \int \int y_i y_j f_{X|r}(F_i^{-1}(\Phi(z_i)), F_j^{-1}(\Phi(z_j))dy_idy_j\), for \(i,j=1,2,\ldots,d\).
+For accurate numeric approximation of these integrals, the functions must be evaluated hundreds of times.
+Others have used efficient Monte Carlo integration schemes (see \citet{Chen2001}), but scale poorly to large dimension in reasonable times (property \textbf{S2}).
+Despite all this, if one does desire to characterize dependency using Pearson correlations, we often see in practice --- and it is theoretically justified under certain conditions (\citet{Song00}) --- that simply using the target Pearson correlation matrix as the initial conditions to our proposed algorithm will lead to approximate matching in the resultant distribution.
+
+\hypertarget{simulation-algorithm-1}{%
+\subsection{Simulation Algorithm}\label{simulation-algorithm-1}}
+
+Putting the together the facts provided in the equations above, we come the following proposed simulation algorithm to produce a random vector \({\bf Y}\) with specified Spearman's correlation and marginal distributions.
+Note that all computational steps can be parallelized as each operation can be done or either the \(d\) marginals or the \(\binom{d}{2}\) pairs for the correlation values.
+Even the generation of multivariate normal random vectors is parallelized through optimized matrix multiplication/decomposition routines.
+
+\hypertarget{inputs}{%
+\subsubsection{Inputs}\label{inputs}}
+
+\begin{enumerate}
+\def\labelenumi{(\arabic{enumi})}
+\tightlist
+\item
+  Marginal characteristics including the same of distributional family and parameter values, denotes as \(F_i\) for marginal component random variable \(Y_i\) for \(i=1,\ldots,d\).
+\item
+  A target (specified) Spearman's correlation matrix \({\bf R_{Spearman}}\) with
+  each \(\rho_{Spearman}\) within the Frechet limits. Alternatively, the rescaled
+  Spearman's correlation (see Equation \eqref{eq:convertKendallDiscrete} ) can be provided.
+\item
+  An error tolerance \(\epsilon\) for finding the nearest positive definite matrix a transformed correlation matrix (see step iii below). Alternatively, a maximum number of iterations can be supplied.
+\end{enumerate}
+
+\hypertarget{algorithm}{%
+\subsubsection{Algorithm}\label{algorithm}}
+
+\begin{enumerate}
+\def\labelenumi{(\roman{enumi})}
+\tightlist
+\item
+  Compute Spearman's \(\rho_{rs}\) (see Equation
+  \eqref{eq:convertKendallDiscrete}) from the specified \(\rho_{spearman}\) for each
+  pair of marginal random variables \((Y_i,Y_{i^\prime})\) with \(i \neq i^\prime\)
+  when either marginal is discrete. In such large scale computations, it may be
+  that numerically \(\rho_{rs}\) is be larger/smaller than the Frechet bounds (or
+  even \(\pm 1\)). To guard against this, set \(\rho_{rs} = min( \rho_{rs}, M)\),
+  where \(M\) is the upper Frechet bound (see {[}Section Introduction{]}) and similarly set \(\rho_{rs} = max( \rho_{rs}, W)\), where \(W\) is the lower Frechet bound for this pair of distributions.
+Gather these \(\rho_{rs} \lq s\) into a new input correlation matrix \({\bf R_{rs}}\).\\
+\item
+  Convert \({\bf R_{Spearman}}\) into \({\bf R_{Pearson}}\) via \(\rho_{Pearson} = 2 \times sin \left( \rho_{rs} \times \frac{\pi}{6} \right)\).\\
+\item
+  Finally, ensure that \({\bf R_{Pearson}}\) is positive definite, by finding the nearest positive definite correlation matrix \({\bf R}\) --- using the \texttt{R} routine \texttt{nearPD} in the \texttt{Matrix} package --- within an error tolerance of \(\epsilon\).\\
+\item
+  Generate \({\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R})\);\\
+\item
+  Transform \({\bf X}\) to \({\bf U} = (U_1, \ldots, U_d)\) viz \(U_i=\Phi(X_i)\), \(i=1, \ldots, d\);\\
+\item
+  Return \({\bf Y} = (Y_1, \ldots, Y_d)\), where \(Y_i=F_i^{-1}(U_i)\), \(i=1, \ldots, d\);
+\end{enumerate}
+
+\hypertarget{the-bigsimr-r-package}{%
+\section{\texorpdfstring{The \texttt{bigsimr} R package}{The bigsimr R package}}\label{the-bigsimr-r-package}}
+
+Short description and key features, including GPU acceleration.
+
+The most computationally extensive lies in the second step of our algorithm. We
+use a optimized and parallelized multivariate normal simulator within the \texttt{R}
+package \texttt{mvnfast}. The rest of the code is also parallelized, but these steps
+run rapidly in serial computation except for extremely large dimension.
+
+\hypertarget{basic-use}{%
+\subsection{Basic use}\label{basic-use}}
+
+short description
+
+\hypertarget{specifying-marginals}{%
+\subsubsection{Specifying marginals}\label{specifying-marginals}}
+
+As stated earlier, to generate multivariate data, we need a list of marginals (and their parameters), and a correlation structure (matrix).
+The marginal distributions can be built up as a list of lists, where each sublist contains the information for the target distribution.
+
+The things to point out here are that in each sublist (marginal), the first item is an unnamed character string with the R name of the distribution \emph{without a letter prefix}.
+E.g.
+instead of \texttt{rnorm}, we pass in just \texttt{"norm"}.
+The second thing to note is that the remaining items are \emph{named} arguments that go along with the distribution.
+A full list of built-in distributions is found in the appendix.
+
+\hypertarget{specify-correlation}{%
+\subsubsection{Specify correlation}\label{specify-correlation}}
+
+The next step is to define a correlation structure for the multivariate distribution.
+This correlation matrix can either come from observed data, or we can set it ourselves, or we can generate a random correlation matrix via \texttt{bigsimr::rcor}.
+Let's create a simple correlation matrix where all off-diagonal elements are 0.5.
+Since we have 3 marginals, we need a \(3\times 3\) matrix.
+
+\begin{verbatim}
+     [,1] [,2] [,3]
+[1,]  1.0  0.5  0.5
+[2,]  0.5  1.0  0.5
+[3,]  0.5  0.5  1.0
+\end{verbatim}
+
+Finally we can generate a random vector with our specified marginals and correlation structure.
+The last argument, \texttt{type}, is looking to know what kind of correlation matrix it is receiving.
+Right now it can handle Pearson, Spearman, or Kendall.
+
+\hypertarget{small-sample}{%
+\subsubsection{Small sample}\label{small-sample}}
+
+On my machine, there is no dedicated GPU, so I would see the following warning message once per session.
+
+Taking a look at our random vector, we see that it is 10 rows and 3 columns, one column for each marginal.
+
+\begin{verbatim}
+       [,1]     [,2] [,3]
+ [1,] 3.046 0.003646    2
+ [2,] 3.180 0.091149    5
+ [3,] 3.070 0.014344    2
+ [4,] 3.157 0.164819    2
+ [5,] 3.222 0.529193    5
+ [6,] 3.118 0.341439    5
+ [7,] 3.056 0.028928    2
+ [8,] 3.187 0.119534    4
+ [9,] 3.118 0.413526    3
+[10,] 3.225 0.431735    3
+\end{verbatim}
+
+We can simulate many more samples and then check the histogram of each margin, as well as the estimated correlation between the columns.
+
+\hypertarget{scaling-up-n}{%
+\subsubsection{Scaling up N}\label{scaling-up-n}}
+
+\begin{center}\includegraphics[width=0.85\linewidth]{article_bigsimr_files/figure-latex/unnamed-chunk-14-1} \end{center}
+
+\hypertarget{evaluation}{%
+\subsubsection{Evaluation}\label{evaluation}}
+
+\begin{verbatim}
+       [,1]   [,2]   [,3]
+[1,] 1.0000 0.4886 0.4945
+[2,] 0.4886 1.0000 0.4882
+[3,] 0.4945 0.4882 1.0000
+\end{verbatim}
+
+We can see that even with 10,000 samples, the estimated correlation of the simulated data is not exactly the same as the target correlation.
+This can be explained by the fact that some correlations are simply not possible due to the discrete nature of certain distributions.
+Another possibility is that the copula algorithm is biased and needs correction.
+
+\hypertarget{monte-carlo-evaluation-and-comparsions-to-other-software}{%
+\section{Monte Carlo evaluation and Comparsions to other software}\label{monte-carlo-evaluation-and-comparsions-to-other-software}}
+
+Before applying our methodology to real data simulation, we conduct several Monte Carlo studies to investigate method performance in comparison to other existing implementations.
+We focus the numerical experiments on assessing how well the procedure scales to high dimension with respect to reasonable computation times (property S1 above) and accurately matching marginal and dependency parameters.
+The simulations will proceed in increasing complexity --- leading up to the setting in our motivating example.
+We begin by exploring simple exchangeable (constant) correlation structures under a large number of simulation replicates while applying the algorithm above to first continuous then discrete marginal distributions.
+We conclude the simulation studies with \(10^5\) replicated data sets with sample sizes and distributional characteristics corresponding the estimates from the motivating example data.
+
+The CPU-based computation times we report were from runs using eight dual-threaded 3.9GHz
+Xeon Gold processors on a linux workstation, allowing for parallel instructions
+over 16 cores. We chose 15 cores for our experiments.
+
+\hypertarget{other-multivariate-simulation-software}{%
+\subsection{Other multivariate simulation software}\label{other-multivariate-simulation-software}}
+
+Describe the competing software and algorithms briefly.
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{copula}
+\item
+  \texttt{Genord}
+\item
+  \texttt{Multiord}
+\item
+  \texttt{nortaRA} only does Pearson matching.
+\end{itemize}
+
+\hypertarget{simulation-i-bivariate-simulation-using-rank-based-correlation}{%
+\subsection{Simulation I: Bivariate simulation using rank-based correlation}\label{simulation-i-bivariate-simulation-using-rank-based-correlation}}
+
+\hypertarget{continuous-example-bivariate-exponential}{%
+\subsubsection{Continuous example: Bivariate Exponential}\label{continuous-example-bivariate-exponential}}
+
+Both Spearman's and Kendall's measures of dependency are invariant under
+strictly monotone transformations (ref). This in turn provides exact simulation
+of continuous marginals \(F_i\) \(i=1,\ldots,d\) since their corresponding quantile
+function are increasing monotonically over their support. The results below
+provide computations times for \(N=10^5\) under increasingly large dimension \(d\).
+For simplicity we assume multivariate Normal with varying parameters and to
+additionally demonstrate the utility of Kendall's \(\tau\) even in the Gaussian
+setting (as opposed to the nature choice of Pearson's correlation). To layer on complexity, we now assume a multivariate gamma distribution with fixed marginal parameters.
+
+Compare speed and accuracy.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## computation times normal}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{discrete-example-bivariate-geometric}{%
+\subsubsection{Discrete example: Bivariate Geometric}\label{discrete-example-bivariate-geometric}}
+
+Compare speed and accuracy.
+
+\hypertarget{simulation-ii-ultra-hd-multivariate-simulation-with-identical-margins-and-exchangeable-rank-based-correlation.}{%
+\subsection{Simulation II: Ultra HD multivariate simulation with identical margins and exchangeable rank-based correlation.}\label{simulation-ii-ultra-hd-multivariate-simulation-with-identical-margins-and-exchangeable-rank-based-correlation.}}
+
+\hypertarget{uhd-continuous-example-multivariate-gamma}{%
+\subsubsection{UHD Continuous example: Multivariate Gamma}\label{uhd-continuous-example-multivariate-gamma}}
+
+Compare speed and accuracy.
+
+\hypertarget{uhd-discrete-example-multivariate-negative-binomial}{%
+\subsubsection{UHD Discrete example: Multivariate Negative Binomial}\label{uhd-discrete-example-multivariate-negative-binomial}}
+
+Compare speed and accuracy.
+
+\hypertarget{simulation-iii-ultra-hd-multivariate-simulatino-with-heterogenous-margins-and-exchangeable-rank-based-correlation.}{%
+\subsection{Simulation III: Ultra HD multivariate simulatino with heterogenous margins and exchangeable rank-based correlation.}\label{simulation-iii-ultra-hd-multivariate-simulatino-with-heterogenous-margins-and-exchangeable-rank-based-correlation.}}
+
+\hypertarget{uhd-heteregeous-continuous-example-multivariate-gamma}{%
+\subsubsection{UHD Heteregeous continuous example: Multivariate Gamma}\label{uhd-heteregeous-continuous-example-multivariate-gamma}}
+
+Compare speed and accuracy.
+
+\hypertarget{uhd-heteregeousdiscrete-example-multivariate-negative-binomial}{%
+\subsubsection{UHD Heteregeousdiscrete example: Multivariate Negative Binomial}\label{uhd-heteregeousdiscrete-example-multivariate-negative-binomial}}
+
+Compare speed and accuracy.
+
+\hypertarget{some-applications}{%
+\section{Some applications}\label{some-applications}}
+
+\hypertarget{simulating-ultra-high-dimensional-rna-seq-data}{%
+\subsection{Simulating Ultra High Dimensional RNA-seq data}\label{simulating-ultra-high-dimensional-rna-seq-data}}
+
+We apply our methodology to simulate RNA-sequencing data sets based on samples derived from breast cancer patients' tumors (the BRCA data set in TCGA).
+The data are freely available as part of the BRCA data set within the The Cancer Genome Atlas data warehouse and were downloaded on BLAH using the Harvard's Broad Institute's Firehouse interface.
+For each of the 1093 patients, the gene expression (abundance of messenger-RNA) from 20501 genes were counted using RNA-sequencing.
+The resultant data set can be represented as a matrix of size 1093 x 20501 with discrete values aligned to HUGO gene symbols for each patient.
+In this illustrative case study, we model the gene-wise marginal distributions as heterogeneous negative binomial with pmf given by Equation \eqref{eq:nb} with two parameters, \(p\) and \(r\) (OR SWITCH TO mu and delta).
+Modeling RNA-seq counts as negative binomially distributed random variables is commonplace (e.g., \citet{Zhao2018}) since overdispersion is often observed for most expressed genes.
+We begin by estimating the marginal negative binomial parameters for each group using a simple method of moments approach by matching the negative binomial parameters with the sample mean and variance.
+Table/Figure XXX summarizes the estimated means, variances, and NB parameters as described in Equation \eqref{eq:nb}.
+Notably some genes show underdispersion compared to a Poisson random variable.
+For simplicity we exclude these cases, but nothing in the later described simulation method prevents using a different marginal distribution in those cases (such as a Maxwell distribution REF Sellers).
+
+Intergene correlation \citep{Schissler2019}.
+
+\begin{equation} 
+  g(n) = \mathbb P(Y=n) = \frac{\Gamma(n+r)}{\Gamma(r)n!} p^r(1-p)^n, \,\,\, n\in
+\mathbb N_0.
+  \label{eq:nb}
+\end{equation}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## full workflow simulating RNA-seq data}
+\end{Highlighting}
+\end{Shaded}
+
+The data are overdispersed:
+
+We see that the all genes studies display overdispersion (variance greater than mean) compared to a Poisson model for the gene counts.
+The genes are highly heterogeneous even when dealing with highly expressed genes after filtering.
+The facts taken together motivate the negative binomial model for the marginal distributions.
+The genes are slightly more variable within the deceased group (expected with smaller sample size and potentially due to biomedical considerations).
+
+Next we perform some essential pre-processing of the data.
+Since our goal is to simulate meaningful multivariate constructions, we must first find a subset of 20501 genes that are expressed.
+In order words, we'll filter out the low expressing genes to allow greater range of possible \(d\)-variate correlations (see \citet{NK10} for details OR SHOULD I EXPLAIN THIS MORE CAREFULLY?).
+Notably, many RNA-seq analytic workflows filter out low expressing genes (see \cite{Conesa2016b}) and this, admittedly, is usually done in an \emph {ad hoc} fashion.
+In this study, we prefer an inclusion criterion that is theoretically motivated.
+Based on the results in Section BLAH, we filter genes with a sample mean less than 12.
+In this way, we remove 4762 (23.2\%) of the genes, retaining 15,739 genes for further analysis.
+The filtered low-expressing genes will be simulated as independent negative binomial random variables to form a complete simulated transcriptome.
+
+RNA-sequencing data derived from breast cancer patients (n=1093), with 20501 genes measured (from TCGA).
+We filter to genes with average expression greater than 10000 counts (d = 4777) across both groups.
+Consider basic differentially expressed genes (DEG) analysis between surviving (n0 = 941) and deceased (n1 = 152) patients.
+This setting can be described as large-scale simultaneous hypothesis testing, under correlation.
+We should evaluate existing and new methodology using simulations that reflect dependency among variables.
+But multivariate simulation tools often scale poorly to high dimension or do not model the full range of dependency.
+Perform a two-sample \(t\) test for each 4777 genes on the observed RNA-seq counts (\(empirical\)).
+Call a ``Z score'' for the \(i^{th}\) gene \(Z_{i} = \Phi^{-1}(t_{\approx 272}(t_i))\).
+
+Our method requires some measure of that the covariance structure is pre-specified.
+In practice, however, high-dimensional covariance estimation is not that easy.
+Indeed there is much recent interest in this area (cite a few recent high visibility articles --- see the GPU-NORTA pub).
+Here we do not provide a comprehensive review on the topic.
+Instead we only seek to use covariance estimator that only guarantees positive-semidefiniteness while maintaining adequate estimation properties.
+To this end we chose to use the condition-number-regularized (\textsc{condreg}) approach of \cite{Won2013g}.
+This covariance estimation procedure restricts the ratio of the largest eigenvalue to the smallest eigenvalue to improve numerical results.
+But the method employs a penalized Gaussian likelihood.
+And so, rather than estimating the covariance on the discrete counts directly, we first perform a \(log_{2}(x +1)\) transformation to better suit this modeling assumption.
+\textsc{condreg} requires a parameter, \(\kappa_{max}\), that controls the largest condition number of the resultant estimated covariance matrix.
+Here we aim to allow a large number of nonzero pairwise correlations and so arbitrary set \(\kappa_{max}=10^{4}\).
+The routine ran without issue on a MacBook Pro carrying BLAH in XXX units of time.
+
+Most pairwise correlations are nearly zero with XXX \% less than \textbar0.1\textbar.
+Yet there are dense subsets of correlated genes.
+And even small correlations among many variables can disrupt the operating characteristics of commonly used statistical inferential procedures \cite{Schissler2018}.
+
+In this case study in the large-scale hypothesis testing in pursuit of detecting differentially expressed genes, we aim to simulate RNA-seq counts using a multivariate negative binomial.
+To evaluate the simulation's utility, we systematically compare the simulated results to the empirical results.
+In short we computed t-statistics and p-values for each of \(d=4777\) genes between two groups of breast cancer patients (1=deceased, 0=alive; \(n_0 = 152, n_1=941\)).
+Further details on the empirical results are provided in Section BLAH.
+We compare two aspects in the simulation design 1) Choice of correlation measure while modeling marginal distributions consistently and 2) how well does the best-performing simulation agree with the empirical results.
+Informed by the simulation studies in Section BLAH, we anticipate a far amount of variation at such small sample sizes and large \(d\) and so replicate the entire study 100 times.
+
+To recap from the discussion above (Section BLAH), we estimated negative binomial parameters for each of the \(d=4777\) genes and also their Pearson, Spearman, and Kendall's correlation coefficients from n=1093 breast cancer patients, within each vital status group (0=deceased, 1=alive).
+We generated 100 synthetic data sets for each group, with the number of \(d\)-dimensional random vectors produced equal to the corresponding sample sizes (\(n_0 = 152, n_1=941\)).
+These 100 samples are used to understand uncertainty in this setting and we will use the mean simulated values to compare to the empirical (and also explore the worst-case scenarios).
+
+Model genes as marginally negative binomial with heterogeneous gene-wise parameters.
+Compute sample correlations (using desired dependency measure) for the 11,407,476 genes pairs, for each group.
+To simulate gene expression counts, our goal is to produce a random vector \({\bf Y}=(Y_1, \ldots, Y_d)\) with \textbf{correlated} NB components.
+To do that, we start with a sequence of \textbf{independent} Poisson processes \(N_i(t)\), \(i=1,\ldots,d\), where the rate of the process \(N_i(t)\) is \(\lambda_i=(1-p_i)/p_i>0\) (so that \(p_i=1/(1+\lambda_i)\)).
+Now, we let \({\bf T} = (T_1, \ldots T_d)\) have a multivariate distribution on \(\mathbb R_+^d\) with the PDF \(f_{{\bf T}}({\bf t})\).
+Then, we define
+
+We rescaled the target Spearman correlation to account for the probability of ties via Equation \eqref{eq:convertSpearmanDiscrete}.
+The serial computation took approximately 10 minutes on XXX.
+Notably, the difference between target and adjusted matrices was very small (mean relative difference: 0.708) in this particular configuration of negative binomial parameters.
+For problems with smaller counts and higher probabilities of ties, the adjustment can be substantial (as observed in the simulation studies above in Section BLAH).
+
+\begin{itemize}
+\tightlist
+\item
+  Kendall's \(\tau\) was estimated efficiently using the \texttt{pcaPP::cor.fk} function in \texttt{R}.
+\end{itemize}
+
+\hypertarget{ultra-high-dimensional-simulation-based-correlation-hypothesis-testing}{%
+\subsection{Ultra High-Dimensional simulation-based correlation hypothesis testing}\label{ultra-high-dimensional-simulation-based-correlation-hypothesis-testing}}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## full workflow simulating RNA-seq data}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{simulation-based-computation-of-correlation-bounds}{%
+\subsection{Simulation-based computation of correlation bounds}\label{simulation-based-computation-of-correlation-bounds}}
+
+Using the Generate, sort, and correlate algorithm \citet{Demirtas2011}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+
+\CommentTok{## simulation based}
+\CommentTok{## devtools::install_github("adknudson/bigsimr", ref="develop")}
+\KeywordTok{library}\NormalTok{(bigsimr)}
+\CommentTok{## ?rvec}
+
+\NormalTok{d <-}\StringTok{ }\DecValTok{5}
+\NormalTok{rho <-}\StringTok{ }\KeywordTok{rcor}\NormalTok{(d)}
+\NormalTok{margins <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}
+    \KeywordTok{list}\NormalTok{(}\StringTok{"nbinom"}\NormalTok{, }\DataTypeTok{size =} \DecValTok{5}\NormalTok{, }\DataTypeTok{prob =} \FloatTok{0.3}\NormalTok{),}
+    \KeywordTok{list}\NormalTok{(}\StringTok{"exp"}\NormalTok{, }\DataTypeTok{rate =} \DecValTok{4}\NormalTok{),}
+    \KeywordTok{list}\NormalTok{(}\StringTok{"binom"}\NormalTok{, }\DataTypeTok{size =} \DecValTok{5}\NormalTok{, }\DataTypeTok{prob =} \FloatTok{0.7}\NormalTok{),}
+    \KeywordTok{list}\NormalTok{(}\StringTok{"norm"}\NormalTok{, }\DataTypeTok{mean =} \DecValTok{10}\NormalTok{, }\DataTypeTok{sd =} \DecValTok{3}\NormalTok{),}
+    \KeywordTok{list}\NormalTok{(}\StringTok{"pois"}\NormalTok{, }\DataTypeTok{lambda =} \DecValTok{10}\NormalTok{)}
+\NormalTok{)}
+
+\NormalTok{params =}\StringTok{ }\NormalTok{margins}
+\NormalTok{rho =}\StringTok{ }\NormalTok{rho}
+\CommentTok{## cores = parallel::detectCores() - 1}
+\NormalTok{cores =}\StringTok{ }\DecValTok{2}
+\NormalTok{reps =}\StringTok{ }\FloatTok{1e3}
+\NormalTok{type =}\StringTok{ "spearman"}
+
+\CommentTok{## help(package="bigsimr")}
+\NormalTok{rho_bounds <-}\StringTok{ }\KeywordTok{computeCorBounds}\NormalTok{(}\DataTypeTok{params =}\NormalTok{ margins, }\DataTypeTok{cores =}\NormalTok{ cores, }\DataTypeTok{type =}\NormalTok{ type, }\DataTypeTok{reps =}\NormalTok{ reps)}
+\NormalTok{rho_bounds}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{simulation-based-joint-probability-calculations}{%
+\subsection{Simulation-based joint probability calculations}\label{simulation-based-joint-probability-calculations}}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## full workflow simulating RNA-seq data}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{conclusiondiscussion}{%
+\section{Conclusion/Discussion}\label{conclusiondiscussion}}
+
+\begin{itemize}
+\tightlist
+\item
+  Although we recommend to use the rank-based measures
+\item
+  discuss the omission (or inclusion) of high dimensional covariance estimators?
+\item
+  We've shown that the multivariate NB simulations outperform the Independent sims.
+But the agreement is still poor.
+This could motivate other more appropriate models for high-expressing RNA-seq data than the negative binomial distribution.
+This algorithm lends itself to exploration of flexible probability models to inspire model selection in the development novel RNA-seq analytic methodology.
+\end{itemize}
+
+We've introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a high-performance implementation called \texttt{bigsimr} (github url).
+The parallelized (multi-core) algorithm is simple and easy to understand as an application of Gaussian copulas.
+This method is largely inspired by \citet{MB13}, but with contributions with regard to scalability, implementation, and application in high-throughput biomedical data (RNA-sequencing).
+We advocate the use of Kendall's \(\tau\) as it better captures correlation among components of non-normal, as well as non-linear patterns of association.
+Moreover, the matching Kendall's \(\tau\) is nearly trivial due to the invariant of monotone transformations, after an adjustment to the input correlation matrix.
+Interestingly, our simulation studies show the use of Kendall's \(\tau\) to works as well as using Pearson correlation coefficient when simulating from multivariate normal distributions.
+
+We also show utility in our methodology through an application to differential gene expression analysis from RNA-sequencing data.
+The application results show that correlations indeed matter in the large-scale hypothesis testing, as many others have noted (for example, see \citet{BE07}, \citet{Wu2012b}).
+We also hope that the application provides an example workflow --- and other strategy to use in simulation design.
+Even the best-performing simulations we provide show a gap from the empirical distribution.
+To keep the demonstration straightforward, we did not attempt to match the empirical distribution of test statistics as precisely as possible.
+Yet our methodology could be more creatively applied to meet that goal.
+One could consider marginal distributions from different families, such as Poisson for a subset of genes.
+One could imagine finding a best fitting probability distribution among a class of distributions for each gene and this could perhaps a better fit.
+One could imagine additional structures/features in the data that an analyst could model to improve the correspondence with the empirical values, for example row correlations and high-dimensional covariance estimators (see \citet{Won2013g}).
+
+There are of course limitations to the methodology and implementation.
+The algorithm requires invertible marginal cdfs.
+This leaves some restriction to the available marginal probability distributions (DOES IT?
+FOR EXAMPLE?) An additional source of error is after converting dependencies (step i; \({\bf R_{Kendall}})\)) the resultant matrix may not be positive definite (PD).
+Then the nearest PD matrix is used.
+From a practical computing standpoint, the user's computing resource provides the ultimate limit on how large a dimension can be simulated using the \texttt{bigsimr} R package.
+A user must consider carefully the available memory and cores when conducting a Monte Carlo experiment.
+The algorithm does amends itself well to parallelization and graphical processing unit acceleration \citep{Li2019gpu} and advanced users may take advantage of those techniques.
+
+Future work includes developing scalable algorithms to match the Pearson correlation matrix exactly.
+Further, more sophisticated approaches could involve simulation algorithms that are ``estimation aware'' and so could explore the role of covariance estimation within the simulation.
+Lastly, these may reveal new approaches to estimating and evaluating high dimensional regression and Bayesian models.
+On the implementation side, employing graphical process unit acceleration could produce substantial speedups over our multi-core approach.
+
+\hypertarget{supplementary-materials}{%
+\section{Supplementary Materials}\label{supplementary-materials}}
+
+We provide an open-source implementation of our methology as the \texttt{bigsimr} R package, hosted on github.
+
+\hypertarget{acknowledgements}{%
+\section{Acknowledgement(s)}\label{acknowledgements}}
+
+The authors gratefully acknowledge the helpful discussions with Professor Walter W. Piegorsch and Professor Edward J. Bedrick during this project's conception.
+
+\hypertarget{disclosure-statement}{%
+\section{Disclosure statement}\label{disclosure-statement}}
+
+The authors report no conflict of interest. DO WE?
+
+\hypertarget{funding}{%
+\section{Funding}\label{funding}}
+
+FUNDING FROM EVERYONE HERE. CP3 grant.
+
+\hypertarget{nomenclaturenotation}{%
+\section{Nomenclature/Notation}\label{nomenclaturenotation}}
+
+\hypertarget{notes}{%
+\section{Notes}\label{notes}}
+
+\hypertarget{list-of-supported-distributions}{%
+\subsubsection{List of supported distributions}\label{list-of-supported-distributions}}
+
+Mention Python?
+
+  \bibliography{bigsimr.bib,packages.bib}
 

### 2020-06-04 index.Rmd
split up document into sections. did spelling check

--- 2020-06-02 index.Rmd
@@
 
-```{bash copyBib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
+```{bash copyBib, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, results = "hide"}
 cp /Users/alfred/Dropbox/bib/bigsimr.bib ./.
@@
 Along those lines, simulating high-dimension dependent negative binomial data motivates this work --- in pursuit of modeling RNA-sequencing data [@Wang2009b; @Conesa2016b] derived from breast cancer patients.
-This labroratory procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions.
+This laboratory procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions.
 Often researchers posit a negative binomial model (refs) as counts are often over (or under) expressed that a Poisson model would suggest.
@@
 RNA-sequencing analysis has garnered much interest in the statistical literature (refs).
-Often researchers simulate independently or from their proposed model in order condunct Monte Carlo studies (refs).
-An alternavitive strategy is to think *generatively* about the data collection process and scientific domain knowledge and design simulations with probabilitistic models that reflect those processes.
+Often researchers simulate independently or from their proposed model in order to conduct Monte Carlo studies (refs).
+An alternative strategy is to think *generatively* about the data collection process and scientific domain knowledge and design simulations with probabilistic models that reflect those processes.
 In our motivating example, we'll study the consequences correlation in the simulation study replicating test statistics from an illustrative, classical large-scale hypothesis testing study [see @Efron2004].
 
-With our application in mind and seeking a general-purpose algorithm with broad applications, our purposed methodology should possess the following properities (adapted from criteria from @Nik13a):
+## Ultra High Dimensional multivariate modeling and simulation desired properties
+
+With our application in mind and seeking a general-purpose algorithm with broad applications, our purposed methodology should possess the following properties (adapted from criteria from @Nik13a):
 
@@
 * P2: Flexible dependence, meaning that the number of bivariate marginals is (approximately) equal to the number of dependence parameters.
-* P3: Flexible marginal modeling, generating hetereogenous data --- possibly from differing probability families.
-* P4: Computationally feasible cumulative distribution function (cdf) for likelihood estimation.
+* P3: Flexible marginal modeling, generating heterogeneous data --- possibly from differing probability families.
 
-Moreover, in simulation method must scale to high dimensions:
+Moreover, the simulation method must scale to high dimensions:
 
@@
 For normal data, the Pearson product-moment correlation describes the dependency completely.
-As we will see, however, simulating abrirtary random vectors with a given Pearson correlation matrix is computationally intense [@Chen2001, @Xia17], violating property **S1**.
+As we will see, however, simulating arbitrary random vectors with a given Pearson correlation matrix is computationally intense [@Chen2001, @Xia17], violating property **S1**.
 On the other hand, an analyst may consider the use of non-parametric correlation measures, such as Spearman's $\rho$ and Kendall's $\tau$, particularly when modeling non-normal data as in our application.
@@
 The crux of the method lies in the construction of a Gaussian copula [refs].
-The idea that many multivariate and marginally heterogenous distribution can be constructed in such a manner is well known [refs].
-This article's contriubition lies in it's application to high-dimenisonal data through a high-performance implementation (`bigsimr`) and speed-focused algorithm design.
+The idea that many multivariate and marginally heterogeneous distribution can be constructed in such a manner is well known [refs].
+This article's contribution lies in it's application to high-dimensional data through a high-performance implementation (`bigsimr`) and speed-focused algorithm design.
 The algorithm design relies on useful properties of non-standard correlation measures, namely Kendall's $\tau$ and Spearman's $\rho$.
@@
 The study proceeds with more details on our motivating example in RNA-sequencing breast cancer data.
-Then we describe and justify our simulation methodology, followed by extensive Monte Carlo studies under various distributional assumputions --- emphasizing normal and non-normal scenarios.
-In these studies, we evaluate the accuracy and speed of our approach in comparison to other readily availble software.
+Then we describe and justify our simulation methodology, followed by extensive Monte Carlo studies under various distributional assumptions --- emphasizing normal and non-normal scenarios.
+In these studies, we evaluate the accuracy and speed of our approach in comparison to other readily available software.
 After evaluating *in silico*, we revisit our motivating
 example by employing our method and compare our simulations to empirically-derived statistics. Finally, we'll make concluding remarks regarding the method's utility and future directions.
-
-Intergene correlation [@Schissler2019].
-
-# Multivariate correlation bounds
-
-# Simulation methodology
-
-## General algorithm
-
-## Rank-based dependence measures
-
-## Pearson correlation
-
-Difficulties
-
-# The `bigsimr` R package
-
-Short description and key features, including GPU acceleration.
-
-```{r setup, eval=TRUE}
-## Using conda
-reticulate::use_condaenv("py37")
-library(bigsimr)
-```
-
-## Basic use
-
-short description
-
-### Specifying marginals
-
-As stated earlier, to generate multivariate data, we need a list of marginals (and their parameters), and a correlation structure (matrix).
-The marginal distributions can be built up as a list of lists, where each sublist contains the information for the target distribution.
-
-```{r}
-margins = list(
-  list("norm", mean = 3.14, sd = 0.1),
-  list("beta", shape1 = 1, shape2 = 4),
-  list("nbinom", size = 10, prob = 0.75)
-)
-```
-
-The things to point out here are that in each sublist (marginal), the first item is an unnamed character string with the R name of the distribution *without a letter prefix*.
-E.g.
-instead of `rnorm`, we pass in just `"norm"`.
-The second thing to note is that the remaining items are *named* arguments that go along with the distribution.
-A full list of built-in distributions is found in the appendix.
-
-### Specify correlation
-
-The next step is to define a correlation structure for the multivariate distribution.
-This correlation matrix can either come from observed data, or we can set it ourselves, or we can generate a random correlation matrix via `bigsimr::rcor`.
-Let's create a simple correlation matrix where all off-diagonal elements are 0.5.
-Since we have 3 marginals, we need a $3\times 3$ matrix.
-
-```{r}
-rho <- matrix(0.5, nrow = 3, ncol = 3)
-diag(rho) <- 1.0
-rho
-```
-
-Finally we can generate a random vector with our specified marginals and correlation structure.
-The last argument, `type`, is looking to know what kind of correlation matrix it is receiving.
-Right now it can handle Pearson, Spearman, or Kendal.
-
-### Small sample
-
-```{r}
-x = rvec(10, rho = rho, params = margins, type = "pearson")
-```
-
-On my machine, there is no dedicated GPU, so I would see the following warning message once per session.
-
-```{r, echo=FALSE}
-warning("warning.warn('No GPU/TPU found, falling back to CPU.')")
-```
-
-Taking a look at our random vector, we see that it hs 10 rows and 3 columns, one column for each marginal.
-
-```{r}
-x
-```
-
-We can simulate many more samples and then check the histogram of each margin, as well as the estimated correlation between the columns.
-
-### Scaling up N
-
-```{r, fig.width=7}
-x = rvec(10000, rho = rho, params = margins, type = "pearson")
-
-par(mfrow=c(1,3))
-hist(x[,1], breaks = 30, xlab = "", main = "Normal")
-hist(x[,2], breaks = 30, xlab = "", main = "Beta")
-hist(x[,3], breaks = 30, xlab = "", main = "Negative Binomial")
-```
-
-### Evaluation
-
-```{r}
-cor(x)
-```
-
-We can see that even wih 10,000 samples, the estimated correlation of the simulated data is not exactly the same as the target correlation.
-This can be explained by the fact that some correlations are simply not possible due to the discrete nature of certain distributions.
-Another possibility is that the copula algorith is biased and needs correction.
-
-### List of supported distributions
-
-```{r, eval=FALSE}
-all_dists <- list(
-  list(dist = "beta", shape1, shape2),
-  list(dist = "binom", size, prob),
-  list(dist = "cauchy", location, scale),
-  list(dist = "chisq", df),
-  list(dist = "exp", rate),
-  list(dist = "f", df1, df2),
-  list(dist = "gamma", shape, rate),
-  list(dist = "geom", prob),
-  list(dist = "hyper", m, n, k),
-  list(dist = "logis", location, scale),
-  list(dist = "lnorm", meanlog, sdlog),
-  list(dist = "nbinom", size, prob),
-  list(dist = "norm", mean, sd),
-  list(dist = "pois", lambda),
-  list(dist = "t", df),
-  list(dist = "unif", min, max),
-  list(dist = "weibull", shape, scale),
-  list(dist = "wilcox", m, n),
-  list(dist = "signrank", n)
-)
-```
-
-## Comparsions to other software
-
-### Speed
-
-- `copula`
-- `Genord`
-- `Multiord`
-- other?
-- CPU benchmarking
-- GPU benchmarking
-
-### Accuracy
-
-blah
-
-### Other comparisons
-
-blah
-
-# Some applications
-
-## Simulation-based correlation checking
-
-A multivariate approach?
-
-Or all the pairs?
-
-Generate, sort, and correlate algorithm @Demirtas2011
-
-```{r, computeFrechetBounds, echo = TRUE, eval=TRUE}
-
-## simulation based
-## devtools::install_github("adknudson/bigsimr", ref="develop")
-library(bigsimr)
-## ?rvec
-
-d <- 5
-rho <- rcor(d)
-margins <- list(
-    list("nbinom", size = 5, prob = 0.3),
-    list("exp", rate = 4),
-    list("binom", size = 5, prob = 0.7),
-    list("norm", mean = 10, sd = 3),
-    list("pois", lambda = 10)
-)
-
-margins2 <- list(
-    list("nbinom", 5, 0.3),
-    list("exp", 4),
-    list("binom", 5, 0.7),
-    list("norm", 10, 3),
-    list("pois", 10)
-)
-
-params = margins
-rho = rho
-## cores = parallel::detectCores() - 1
-cores = 2
-reps = 1e3
-type = "spearman"
-
-## help(package="bigsimr")
-rho_bounds <- computeCorBounds(params = margins, cores = cores, type = "spearman", reps = reps)
-rho_bounds
-```
-
-## Simulating RNA-seq data
-
-We apply our methodology to simulate RNA-sequencing data sets based on samples derived from breast cancer patients' tumors (the BRCA data set in TCGA).
-The data are freely available as part of the BRCA data set within the The Cancer Genome Atlas data warehouse and were downloaded on BLAH using the Harvard's Broad Institute's Firehouse interface.
-For each of the 1093 patients, the gene expression (abundance of messenger-RNA) from 20501 genes were counted using RNA-sequencing.
-The resultant data set can be represented as a matrix of size 1093 x 20501 with discrete values aligned to HUGO gene symbols for each patient.
-In this illustrative case study, we model the genewise marginal distributions as hetereogeneous negative binomial with pmf given by Equation \@ref(eq:nb) with two parameters, $p$ and $r$ (OR SWITCH TO mu and delta).
-Modeling RNA-seq counts as negative binomially distributed random variables is commonplace (e.g., @Zhao2018) since overdispersion is often observed for most expressed genes.
-We begin by estimating the marginal negative binomial parameters for each group using a simple method of moments approach by matching the negative binomial parameters with the sample mean and variance.
-Table/Figure XXX summarizes the estimated means, variances, and NB parameters as described in Equation \@ref(eq:nb).
-Notably some genes show underdisperation compared to a Poisson random varible.
-For simplicity we exclude these cases, but nothing in the later described simulation method prevents using a different marginal distribution in those cases (such as a Maxwell distribution REF Sellers).
-
-\begin{equation} 
-  g(n) = \mathbb P(Y=n) = \frac{\Gamma(n+r)}{\Gamma(r)n!} p^r(1-p)^n, \,\,\, n\in
-\mathbb N_0.
-  (\#eq:nb)
-\end{equation}
-
-## Simulation-based correlation testing
-
-## Simulation-based probability calculations
-
-# Conclusion/Discussion
-
-- Although we recommend to use the rank-based measures
-- discuss the omission (or inclusion) of high dimensional covariance estimators?
-- We've shown that the multivariate NB simulations outperform the Independent sims.
-But the agreement is still poor.
-This could motivate other more appropriate models for high-expressing RNA-seq data than the negative binomial distribution.
-This algorithm lends itself to exploration of flexible probability models to inspire model selection in the developemnt novel RNA-seq analytic methodology.
-
-We've introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a high-performance implementation called `bigsimr` (github url).
-The parallelized (multi-core) algorithm is simple and easy to understand as an application of Gaussian copulas.
-This method is largely inspired by @MB13, but with contributions with regard to scaleability, implementation, and application in high-throughput biomedical data (RNA-sequencing).
-We advocate the use of Kendall's $\tau$ as it better captures correlation among components of non-normal, as well as non-linear patterns of association.
-Moreover, the matching Kendall's $\tau$ is nearly trival due to the invariant of monotone transformations, after an adjustment to the input correlation matrix.
-Interestingly, our simulation studies show the use of Kendall's $\tau$ to works as well as using Pearson correlation coefficient when simulating from multivariate normal distributions.
-
-We also show utility in our methodology through an application to differential gene expression analysis from RNA-sequencing data.
-The application results show that correlations indeed matter in the large-scale hypothesis testing, as many others have noted (for example, see @BE07, @Wu2012b).
-We also hope that the application provides an example workflow --- and other strategy to use in simulation design.
-Even the best-performing simulations we provide show a gap from the empirical distribution.
-To keep the demonstration straightforward, we did not attempt to match the empirical distribution of test statistics as precisely as possible.
-Yet our methodology could be more creatively applied to meet that goal.
-One could consider marginal distributions from different families, such as Poisson for a subset of genes.
-One could imagine finding a best fitting probability distribution among a class of distributions for each gene and this could perhaps a better fit.
-One could imagine additonal structures/features in the data that an analyset could model to improve the correspondence with the emprical values, for example row correlations and high-dimensional covariance estimators (see @Won2013g).
-
-There are of course limitations to the methodology and implementation.
-The algorithm requires invertible marginal cdfs.
-This leaves some restriction to the available marginal probability distributions (DOES IT?
-FOR EXAMPLE?) An additional source of error is after converting dependencies (step i; ${\bf R_{Kendall}})$) the resultant matrix may not be positive definite (PD).
-Then the nearest PD matrix is used.
-From a practical computing standpoint, the user's computing resource provides the ultimate limit on how large a dimension can be simulated using the `bigsimr` R package.
-A user must consider carefully the available memory and cores when conducting a Monte Carlo experiment.
-The algorithm does amends itself well to parallelization and graphical processing unit acceleration [@Li2019gpu] and advanced users may take advantage of those techniques.
-
-Future work includes developing scalable algorithms to match the Pearson correlation matrix exactly.
-Further, more sophicated approaches could involve simulation algorithms that are "estimation awarene" and so could explore the role of covariance estimation within the simulation.
-Lastly, these may reveal new approaches to estimating and evaluating high dimensional regression and Bayesian models.
-On the implementation side, employing graphical process unit acceleration could produce substaintial speedups over our multi-core approach.
-
-# Conclusions
-
-# Supplementary Materials
-
-We provide an open-source implementation of our methology as the `bigsimr` R package, hosted on github.
-
-# Acknowledgement(s)
-
-The authors gratefully acknowledge the helpful discussions with Professor Walter W. Piegorsch and Professor Edward J. Bedrick during this project's conception.
-
-# Disclosure statement
-
-The authors report no conflict of interest. DO WE?
-
-# Funding
-
-FUNDING FROM EVERYONE HERE. CP3 grant.
-
-# Nomenclature/Notation
-
-# Notes
-
-Mention Python?

### 2020-06-04 040-monte-carlo-evaluation.Rmd

--- 2020-06-01 docs/article_bigsimr.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-]{article}
-\usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={GIVE ME A TITLE},
-  pdfauthor={A.G. Schissler; A.B. PERSON},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage{longtable,booktabs}
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-\usepackage{graphicx,grffile}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
-\usepackage[]{natbib}
-\bibliographystyle{apalike}
+# Monte Carlo evaluation and Comparsions to other software
 
-\title{GIVE ME A TITLE}
-\author{A.G. Schissler \and A.B. PERSON}
-\date{Monday, June 01, 2020}
+Before applying our methodology to real data simulation, we conduct several Monte Carlo studies to investigate method performance in comparison to other existing implementations.
+We focus the numerical experiments on assessing how well the procedure scales to high dimension with respect to reasonable computation times (property S1 above) and accurately matching marginal and dependency parameters.
+The simulations will proceed in increasing complexity --- leading up to the setting in our motivating example.
+We begin by exploring simple exchangeable (constant) correlation structures under a large number of simulation replicates while applying the algorithm above to first continuous then discrete marginal distributions.
+We conclude the simulation studies with $10^5$ replicated data sets with sample sizes and distributional characteristics corresponding the estimates from the motivating example data.
 
-\begin{document}
-\maketitle
+The CPU-based computation times we report were from runs using eight dual-threaded 3.9GHz
+Xeon Gold processors on a linux workstation, allowing for parallel instructions
+over 16 cores. We chose 15 cores for our experiments.
 
-{
-\setcounter{tocdepth}{3}
-\tableofcontents
-}
-\hypertarget{prerequisites}{%
-\section{Prerequisites}\label{prerequisites}}
+## Other multivariate simulation software
 
-This is a \emph{sample} book written in \textbf{Markdown}. You can use anything that Pandoc's Markdown supports, e.g., a math equation \(a^2 + b^2 = c^2\).
+Describe the competing software and algorithms briefly.
 
-The \textbf{bookdown} package can be installed from CRAN or Github:
+- `copula`
+- `Genord`
+- `Multiord`
+- `nortaRA` only does Pearson matching.
 
-Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading \texttt{\#}.
+## Simulation I: Bivariate simulation using rank-based correlation
 
-To compile this example to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): \url{https://yihui.name/tinytex/}.
+### Continuous example: Bivariate Exponential
 
-  \bibliography{bibliography.bib,packages.bib}
+Both Spearman's and Kendall's measures of dependency are invariant under
+strictly monotone transformations (ref). This in turn provides exact simulation
+of continuous marginals $F_i$ $i=1,\ldots,d$ since their corresponding quantile
+function are increasing monotonically over their support. The results below
+provide computations times for $N=10^5$ under increasingly large dimension $d$.
+For simplicity we assume multivariate Normal with varying parameters and to
+additionally demonstrate the utility of Kendall's $\tau$ even in the Gaussian
+setting (as opposed to the nature choice of Pearson's correlation). To layer on complexity, we now assume a multivariate gamma distribution with fixed marginal parameters.
 
-\end{document}
+Compare speed and accuracy.
+
+```{r tabGamma, echo = TRUE}
+## computation times normal
+
+```
+### Discrete example: Bivariate Geometric
+
+Compare speed and accuracy.
+
+## Simulation II: Ultra HD multivariate simulation with identical margins and exchangeable rank-based correlation.
+
+### UHD Continuous example: Multivariate Gamma
+
+Compare speed and accuracy.
+
+### UHD Discrete example: Multivariate Negative Binomial
+
+Compare speed and accuracy.
+
+## Simulation III: Ultra HD multivariate simulatino with heterogenous margins and exchangeable rank-based correlation.
+
+### UHD Heteregeous continuous example: Multivariate Gamma
+
+Compare speed and accuracy.
+
+### UHD Heteregeousdiscrete example: Multivariate Negative Binomial
+
+Compare speed and accuracy.

### 2020-06-04 070-supplementary-materials.Rmd

# Supplementary Materials

We provide an open-source implementation of our methology as the `bigsimr` R package, hosted on github.

# Acknowledgement(s)

The authors gratefully acknowledge the helpful discussions with Professor Walter W. Piegorsch and Professor Edward J. Bedrick during this project's conception.

# Disclosure statement

The authors report no conflict of interest. DO WE?

# Funding

FUNDING FROM EVERYONE HERE. CP3 grant.

# Nomenclature/Notation

# Notes

### List of supported distributions

```{r, eval=FALSE}
all_dists <- list(
  list(dist = "beta", shape1, shape2),
  list(dist = "binom", size, prob),
  list(dist = "cauchy", location, scale),
  list(dist = "chisq", df),
  list(dist = "exp", rate),
  list(dist = "f", df1, df2),
  list(dist = "gamma", shape, rate),
  list(dist = "geom", prob),
  list(dist = "hyper", m, n, k),
  list(dist = "logis", location, scale),
  list(dist = "lnorm", meanlog, sdlog),
  list(dist = "nbinom", size, prob),
  list(dist = "norm", mean, sd),
  list(dist = "pois", lambda),
  list(dist = "t", df),
  list(dist = "unif", min, max),
  list(dist = "weibull", shape, scale),
  list(dist = "wilcox", m, n),
  list(dist = "signrank", n)
)
```

Mention Python?

### 2020-06-15 020-simulation-algorithm.Rmd
migrate to html based preview

@@
 
-### Inputs
-(1) Marginal characteristics including the same of distributional family and parameter values, denotes as $F_i$ for marginal component random variable $Y_i$ for $i=1,\ldots,d$.
-(2) A target (specified) Spearman's correlation matrix ${\bf R_{Spearman}}$ with
-each $\rho_{Spearman}$ within the Frechet limits. Alternatively, the rescaled
-Spearman's correlation (see Equation \@ref(eq:convertKendallDiscrete) ) can be provided.
-(3) An error tolerance $\epsilon$ for finding the nearest positive definite matrix a transformed correlation matrix (see step iii below). Alternatively, a maximum number of iterations can be supplied.
+### Algorithm 1: Matching Pearson approximately
 
-### Algorithm
-(i) Compute Spearman's $\rho_{rs}$ (see Equation
-\@ref(eq:convertKendallDiscrete)) from the specified $\rho_{spearman}$ for each
-pair of marginal random variables $(Y_i,Y_{i^\prime})$ with $i \neq i^\prime$
-when either marginal is discrete. In such large scale computations, it may be
-that numerically $\rho_{rs}$ is be larger/smaller than the Frechet bounds (or
-even $\pm 1$). To guard against this, set $\rho_{rs} = min( \rho_{rs}, M)$,
-where $M$ is the upper Frechet bound (see [Section Introduction]) and similarly set $\rho_{rs} = max( \rho_{rs}, W)$, where $W$ is the lower Frechet bound for this pair of distributions.
-Gather these $\rho_{rs} \lq s$ into a new input correlation matrix ${\bf R_{rs}}$.
-(ii) Convert ${\bf R_{Spearman}}$ into ${\bf R_{Pearson}}$ via $\rho_{Pearson} = 2 \times sin \left( \rho_{rs} \times \frac{\pi}{6} \right)$.   
-(iii) Finally, ensure that ${\bf R_{Pearson}}$ is positive definite, by finding the nearest positive definite correlation matrix ${\bf R}$ --- using the `R` routine `nearPD` in the `Matrix` package --- within an error tolerance of $\epsilon$.
-(iv) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R})$;  
-(v) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$;  
-(vi) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;
+We aim to simulate multivariate data with a Pearson correlation
+matrix ${\bf R}$.
+
+(i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R})$;  
+(ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$;  
+(iii) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;
+
+* Where $F_i^{-1}$ is the inverse CDF for the $i^{th}$ gene's marginal
+distribution.  
+* And ${\bf R}$ is the target Pearson correlation matrix.  
+* Pearson's correlation is **not** invariant under the montonically increasing
+  transformations in steps 3, 4.  
+* So the resulting simulated random vectors will have the **exact** margins but **approximate** ${\bf R}$.
+
+### Algorithm 2: Matching Spearman's exactly
+
+(i) Convert ${\bf R}_{Spearman}$ into ${\bf R}_{Pearson}$ via $\rho_{Pearson} =
+  2 \times sin \left( \rho_{Spearman} \times \frac{\pi}{6} \right)$;  
+(ii) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf} R_{Pearson})$;  
+(iii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$;  
+(iv) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;
+
+- Spearman's $\rho$ relates to the difference in the ranks of pairs of values.
+- We take advantage of a closed form correspondence between Pearson and Spearman
+  for *normal random variables* in step 1 [@BF17].
+- Since Spearman's $\rho$ is invariant under the monotonically increasing
+  transformations in steps 3 and 4, the resulting simulated random vectors will
+  have the **exact** margins and **exact** Spearman correlation (up to Monte Carlo/numeric error).
+
+### Algorithm 3: Matching Kendall's exactly
+
+(i) Convert ${\bf R}_{Kendall}$ into ${\bf R}_{Pearson}$ via $\rho_{Pearson} = sin \left( \tau \times \frac{\pi}{2} \right)$;  
+(ii) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R}_{Pearson})$;  
+(iii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$;  
+(iv) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;
+
+- Kendall's $\tau$ is the difference in the probabilities of concordant and discordant pairs.
+- We use the 1-1 relationship between Pearson and Kendall's $\tau$ for *normal
+  rvs* in step 1.
+- Since Kendall's $\rho$ is invariant under the montonically increasing
+  transformations in steps 3 and 4,  
+ the resulting simulated random vectors will have the **exact** margins and
+ **exact** Kendall's $\tau$  (up to Monte Carlo/numeric error).
+
+## Algorithm notes
+
+- Discrete margins pose some difficulties and matching *unadjusted* correlation
+  measures is not exact (empirically approximate).
+- Exact Pearson solutions are possible but require computing $d \choose 2$
+   double integrals $EY_iY_j = \int \int y_i y_j f_{X|r}(F_i^{-1}(\Phi(z_i)),
+   F_j^{-1}(\Phi(z_j))dy_idy_j$. We have yet to accelerate this process.
+- We GPU accelerate whenever possible during our proposed High-Performance NORTA
+  algorithm.
+- One thorny problem when dealing with high-dimensional correlation matrices is
+they can become non-positive definite through either during estimation or
+bivariate transformations.
+- When converting nonparametric correlations to Pearson (first step above) the
+  resultant correlation may not be positive definite (PD).
+- In that case, we replace ${\bf R}_{Pearson}$ with a "close" PD matrix ${\bf \tilde{R}}_{Pearson}$.
+- In practice, this loss in accuracy typically has little impact on performance,
+  but the algorithm needs acceleration (`Matrix::nearPD()` violates property S1).
+
+test workflow again

### 2020-06-15 030-bigsimr-package.Rmd
migrate to html based preview

@@
 
-```{r setup, eval=TRUE}
-## Using conda
-reticulate::use_condaenv("py37")
+```{r setup, eval = TRUE}
 library(bigsimr)
-## help(package = 'bigsimr')
+# Reticulate needs to be able to find the python binary with `jax` installed
+reticulate::use_condaenv("bigsimr-cpu")
 ```
@@
 ```{r}
-margins = list(
-  list("norm", mean = 3.14, sd = 0.1),
-  list("beta", shape1 = 1, shape2 = 4),
-  list("nbinom", size = 10, prob = 0.75)
+margins = alist(
+  qnorm(mean = 3.14, sd = 0.1),
+  qbeta(shape1 = 1, shape2 = 4),
+  qnbinom(size = 10, prob = 0.75)
 )
@@
 ```{r}
-x = rvec(10, rho = rho, params = margins, type = "pearson")
+x <- rvec(10, rho = rho, margins = margins, type = "spearman")
 ```
-
-On my machine, there is no dedicated GPU, so I would see the following warning message once per session.
 
@@
 ```{r, fig.width=7}
-x = rvec(10000, rho = rho, params = margins, type = "pearson")
+x <- rvec(100000, rho = rho, margins = margins, type = "spearman")
 
@@
 ```{r}
-cor(x)
+cor(x, method = 'spearman')
 ```
 
-We can see that even with 10,000 samples, the estimated correlation of the simulated data is not exactly the same as the target correlation.
+We can see that even with 100,000 samples, the estimated correlation of the simulated data is not exactly the same as the target correlation.
 This can be explained by the fact that some correlations are simply not possible due to the discrete nature of certain distributions.

### 2020-06-15 050-some-applications.Rmd
migrate to html based preview

@@
 # Some applications
+
+```{r LoadLib050, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
+library(ggplot2)
+library(tidyverse)
+library(knitr)
+library(dplyr)
+## devtools::install_github( "adknudson/bigsimr", ref = 'develop')
+library(bigsimr)
+reticulate::use_condaenv("py37")
+CORES <- parallel::detectCores() - 1
+set.seed(06082020)
+## devtools::install_github("ggobi/GGally")
+## devtools::install_github( "adknudson/bigsimr")
+## A conda environment is also activitied (see README)
+library(bigsimr)
+library(GGally)
+```
 
@@
 
-```{r brca, echo=TRUE, eval=FALSE}
+```{r processBRCA, echo=TRUE, eval=TRUE, cache=TRUE}
 ## full workflow simulating RNA-seq data
-
+allDat <- readRDS( file = "~/Downloads/complete_processed_tcga2stat_RNASeq2_with_clinical.rds" )
+lastClinical <- which( names(allDat) == 'tumorsize' )
+brca <- allDat[ allDat$disease == "BRCA", (lastClinical+1):ncol(allDat) ]
+## remove naively the RSEM adjustment
+brca <- round(brca, 0)
+ncol(brca) ## num of genes 20501
+## compute the median expression! avoid the 0s
+brcaMedian <- apply(brca, 2, median)
+## retain top (1-probs)*100% highest expressing genes for illustration
+myProb <- 0.95 
+cutPoint <- quantile( x = brcaMedian, probs = myProb )
+## genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
+genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
+brca <- brca[ , genesToKeep ]
+## ncol(brca) / 20501
+(d <- length(genesToKeep))
 ```
@@
 
-```{r figNBratio, echo = FALSE, eval=FALSE, message=FALSE, warning=FALSE, fig.path='fig/plot-', out.width= '75%', fig.align='center', fig.ext='pdf', fig.cap="The genes display extreme overdispersion.
-Using group-specific sample means and variances for each of the d=4777 genes, we display the distributions oflog of variance to mean ratio (log(sigma2/mu))."}
-## http://www.sthda.com/english/wiki/ggplot2-histogram-easy-histogram-graph-with-ggplot2-r-package
-## devtools::install_github("kassambara/easyGgplot2")
-library(easyGgplot2)
-## p0 <- ggplot(data = nb_dat, aes(x = log(nb_var / nb_mean), y = stat(density), fill=group))
-## p0 + geom_histogram(bins = 30) + labs(x="log of variance to mean ratio") + theme_bw()
-nb_dat$logRatio <- log(nb_dat$nb_var / nb_dat$nb_mean)
-ggplot2.histogram(data=nb_dat, xName='logRatio',
-    groupName='group', legendPosition="right",
-    alpha=0.5, addDensity=TRUE,
-    addMeanLine=TRUE, meanLineColor="white", meanLineSize=1.5)
+```{r figNBratio, echo = TRUE, eval=TRUE, message=FALSE, warning=FALSE, out.width= '75%', fig.align='center', fig.cap='The genes display extreme overdispersion.
+Using group-specific sample means and variances for each of the d=`r d` genes, we display the distributions oflog of variance to mean ratio.'}
+
+logVarOverMean <- function( x ) {
+    log ( var(x) / mean (x) )
+}
+## check whether NB makes sense
+
+summaryBRCA <- brca %>% 
+    summarize_all(  list(~ logVarOverMean( . ) ) )
+summaryBRCA <- as.data.frame( t(as.data.frame( summaryBRCA )) )
+names(summaryBRCA) <- "logVarOverMean"
+## sum(summaryBRCA$logVarOverMean < 0)  ## no underdispersed genes
+ggplot(data = summaryBRCA, mapping = aes(x = logVarOverMean)) +
+  geom_histogram(color = "white", bins = 20)
 ```
@@
 
-- Kendall's $\tau$ was estimated efficiently using the `pcaPP::cor.fk` function in `R`.
+```{r estRhoBRCA, echo=TRUE, eval=TRUE}
+## full workflow simulating RNA-seq data
 
-## Ultra High-Dimensional simulation-based correlation hypothesis testing
+## 1. Estimate Spearman's correlation on the count data
+corType <- 'spearman'
+system.time( rho <- bigsimr::fastCor( brca, method = corType ) )
+## Describe the correlations
+```
+
+```{r estMargins, echo = TRUE, eval = TRUE, cache=TRUE}
+## 2. Estimate NegBin parameters using Method of Moments
+estimateNegBinMoM <- function(tmpGene, minP = 1e-5) {
+    tmpMean <- mean(tmpGene)
+    tmpVar <- var(tmpGene)
+    ## relate to nbinom parameters
+    ## See ?rbinom for details.
+    p <- tmpMean / tmpVar
+    if (p < minP) {p <- minP } ## maybe add noise here
+    n <- ( tmpMean * p) / ( 1  - p )  
+    ## format for bigsimr margins
+    return( list("nbinom", size = n, prob = p) )
+}
+brcaMargins <- apply( unname(as.matrix(brca)), 2, estimateNegBinMoM )
+## describe the margins
+## check for too small prob
+## head( sort( unlist( lapply( brcaMargins, function(x) {x$prob} ) ) ) )
+## head( sort( unlist( lapply( brcaMargins, function(x) {x$size} ) ) ) )
+```
+
+```{r simBRCA, echo=TRUE, eval = FALSE, cache=TRUE}
+## 3. Generate the simulated samples
+## N <- 10
+## system.time( simBRCA <- rvec(N, rho = rho[1:2, 1:2], params = brcaMargins[1:2], type = corType, adjustForDiscrete = FALSE) )
+N <- nrow(brca)
+system.time( simBRCA <- rvec(N, rho = rho, params = brcaMargins, cores = CORES, type = corType, adjustForDiscrete = FALSE) )
+simBRCA[1:2, 1:2 ]
+
+```
+
+```{r evalBRCA, echo=TRUE, eval = FALSE, cache=TRUE}
+
+## Describe and plot real data
+GGally::ggpairs(data = as.data.frame(brca[ ,1:3] ) )
+
+## Describe and plot simulations
+GGally::ggpairs(data = as.data.frame(simBRCA[ ,1:3] ) )
+
+## check 1st moments
+trueMu <- colMeans(brca)
+simMu <- colMeans(simBRCA)
+qplot( x = trueMu, y = simMu ) + geom_abline(slope = 1, intercept = 0)
+
+## check 2nd moment
+trueVar <- apply( brca, 2, var)
+simVar <- apply( simBRCA, 2, var)
+qplot( x = trueVar, y = simVar ) + geom_abline(slope = 1, intercept = 0)
+
+## check correlation
+system.time( simRho <- bigsimr::fastCor( simBRCA, method = corType ) )
+rho[1:5, 1:5]
+simRho[1:5, 1:5]
+trueRho <- rho[lower.tri(rho)]
+simRho <- simRho[lower.tri(simRho)]
+qplot( x = trueRho, y = simRho ) + geom_abline(slope = 1, intercept = 0)
+```
+
+## Ultra High-Dimensional simulation-based correlation hypothesis testing under
+
+Since many of the correlations are near zero, one could ask whether it is
+possible that
 
 ```{r brcaSigmaTesting, echo=TRUE, eval=FALSE}
-## full workflow simulating RNA-seq data
 

### 2020-06-15 docs/article_bigsimr.tex
migrate to html based preview

@@
 \urlstyle{same} % disable monospaced font for URLs
+\usepackage[margin=1in]{geometry}
 \usepackage{color}
@@
 \makesavenoteenv{longtable}
-\usepackage{graphicx,grffile}
+\usepackage{graphicx}
 \makeatletter
@@
 \author{A.G. Schissler \and A. Knudson}
-\date{Thursday, June 04, 2020}
+\date{Monday, June 15, 2020}
 
@@
         [,1]     [,2] [,3]
- [1,] 3.046 0.003646    2
- [2,] 3.180 0.091149    5
- [3,] 3.070 0.014344    2
- [4,] 3.157 0.164819    2
- [5,] 3.222 0.529193    5
- [6,] 3.118 0.341439    5
- [7,] 3.056 0.028928    2
- [8,] 3.187 0.119534    4
- [9,] 3.118 0.413526    3
-[10,] 3.225 0.431735    3
+  [1,] 3.044 0.095446    3
+  [2,] 3.266 0.149347    4
+  [3,] 3.210 0.368758    7
+  [4,] 3.096 0.273881    1
+  [5,] 3.262 0.167132    8
+  [6,] 3.186 0.297752    4
+  [7,] 3.244 0.378209    2
+  [8,] 3.092 0.061396    2
+  [9,] 3.153 0.388307    5
+ [10,] 3.191 0.017892    2
+ [11,] 3.021 0.143984    1
+ [12,] 2.977 0.306241    2
+ [13,] 3.327 0.311243    6
+ [14,] 3.087 0.166009    3
+ [15,] 3.146 0.462549    2
+ [16,] 3.130 0.045254    3
+ [17,] 3.215 0.341481    3
+ [18,] 3.153 0.481414    9
+ [19,] 3.311 0.536290    6
+ [20,] 3.054 0.183528    4
+ [21,] 3.017 0.011316    1
+ [22,] 3.195 0.293363    1
+ [23,] 3.017 0.077518    1
+ [24,] 3.024 0.041316    1
+ [25,] 3.272 0.330607    6
+ [26,] 3.218 0.206071    5
+ [27,] 2.984 0.001637    2
+ [28,] 3.156 0.114774    3
+ [29,] 3.269 0.089939    3
+ [30,] 3.103 0.177539    3
+ [31,] 3.147 0.138347    6
+ [32,] 3.157 0.093564    4
+ [33,] 3.318 0.356667    2
+ [34,] 3.111 0.231386    4
+ [35,] 3.174 0.659264    2
+ [36,] 3.218 0.293820    4
+ [37,] 3.264 0.240420    6
+ [38,] 3.072 0.255088    3
+ [39,] 3.131 0.168695    0
+ [40,] 3.280 0.491158    6
+ [41,] 3.189 0.238835    5
+ [42,] 3.017 0.011214    1
+ [43,] 3.145 0.022872    1
+ [44,] 3.111 0.144358    1
+ [45,] 3.236 0.085572    3
+ [46,] 3.221 0.309900    7
+ [47,] 2.955 0.014524    3
+ [48,] 3.069 0.011569    2
+ [49,] 2.970 0.075424    2
+ [50,] 3.265 0.233258    5
+ [51,] 3.125 0.223936    2
+ [52,] 3.118 0.100138    3
+ [53,] 3.042 0.113373    2
+ [54,] 2.993 0.335292    3
+ [55,] 3.073 0.020944    1
+ [56,] 3.178 0.256723    4
+ [57,] 3.242 0.679650    7
+ [58,] 3.204 0.228560    5
+ [59,] 3.325 0.254273    9
+ [60,] 2.918 0.109870    2
+ [61,] 3.106 0.134519    5
+ [62,] 3.035 0.022232    1
+ [63,] 3.238 0.029679    3
+ [64,] 3.163 0.159671    3
+ [65,] 3.250 0.265043    5
+ [66,] 3.139 0.109471    6
+ [67,] 2.952 0.069780    3
+ [68,] 3.092 0.103947    6
+ [69,] 3.163 0.196133    6
+ [70,] 2.993 0.079026    1
+ [71,] 2.990 0.043229    2
+ [72,] 3.057 0.018904    2
+ [73,] 2.985 0.009260    1
+ [74,] 3.224 0.778705    6
+ [75,] 3.231 0.238377    4
+ [76,] 3.083 0.014756    2
+ [77,] 3.175 0.262330    1
+ [78,] 3.141 0.106700    3
+ [79,] 3.047 0.007255    2
+ [80,] 3.182 0.334552    2
+ [81,] 3.184 0.303953    5
+ [82,] 3.033 0.182557    2
+ [83,] 3.114 0.227299    4
+ [84,] 3.066 0.057184    5
+ [85,] 3.077 0.069102    4
+ [86,] 2.866 0.004398    1
+ [87,] 2.920 0.100838    2
+ [88,] 3.219 0.175613    6
+ [89,] 3.080 0.094636    1
+ [90,] 3.145 0.076677    4
+ [91,] 3.176 0.243460    4
+ [92,] 3.005 0.230508    2
+ [93,] 3.056 0.315420    5
+ [94,] 3.162 0.524840    5
+ [95,] 3.332 0.685218    7
+ [96,] 3.120 0.209939    0
+ [97,] 3.184 0.174008    2
+ [98,] 3.037 0.037463    2
+ [99,] 3.064 0.013679    1
+[100,] 3.098 0.101754    3
 \end{verbatim}
@@
        [,1]   [,2]   [,3]
-[1,] 1.0000 0.4886 0.4945
-[2,] 0.4886 1.0000 0.4882
-[3,] 0.4945 0.4882 1.0000
+[1,] 1.0000 0.4989 0.4963
+[2,] 0.4989 1.0000 0.4929
+[3,] 0.4963 0.4929 1.0000
 \end{verbatim}
 
-We can see that even with 10,000 samples, the estimated correlation of the simulated data is not exactly the same as the target correlation.
+We can see that even with 100,000 samples, the estimated correlation of the simulated data is not exactly the same as the target correlation.
 This can be explained by the fact that some correlations are simply not possible due to the discrete nature of certain distributions.
@@
 \begin{Highlighting}[]
-\CommentTok{## computation times normal}
+\CommentTok{\#\# computation times normal}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{## full workflow simulating RNA-seq data}
+\CommentTok{\#\# full workflow simulating RNA{-}seq data}
+\NormalTok{allDat \textless{}{-}}\StringTok{ }\KeywordTok{readRDS}\NormalTok{( }\DataTypeTok{file =} \StringTok{"\textasciitilde{}/Downloads/complete\_processed\_tcga2stat\_RNASeq2\_with\_clinical.rds"}\NormalTok{ )}
+\NormalTok{lastClinical \textless{}{-}}\StringTok{ }\KeywordTok{which}\NormalTok{( }\KeywordTok{names}\NormalTok{(allDat) }\OperatorTok{==}\StringTok{ \textquotesingle{}tumorsize\textquotesingle{}}\NormalTok{ )}
+\NormalTok{brca \textless{}{-}}\StringTok{ }\NormalTok{allDat[ allDat}\OperatorTok{$}\NormalTok{disease }\OperatorTok{==}\StringTok{ "BRCA"}\NormalTok{, (lastClinical}\OperatorTok{+}\DecValTok{1}\NormalTok{)}\OperatorTok{:}\KeywordTok{ncol}\NormalTok{(allDat) ]}
+\CommentTok{\#\# remove naively the RSEM adjustment}
+\NormalTok{brca \textless{}{-}}\StringTok{ }\KeywordTok{round}\NormalTok{(brca, }\DecValTok{0}\NormalTok{)}
+\KeywordTok{ncol}\NormalTok{(brca) }\CommentTok{\#\# num of genes 20501}
+\NormalTok{[}\DecValTok{1}\NormalTok{] }\DecValTok{20501}
+\CommentTok{\#\# compute the median expression! avoid the 0s}
+\NormalTok{brcaMedian \textless{}{-}}\StringTok{ }\KeywordTok{apply}\NormalTok{(brca, }\DecValTok{2}\NormalTok{, median)}
+\CommentTok{\#\# retain top (1{-}probs)*100\% highest expressing genes for illustration}
+\NormalTok{myProb \textless{}{-}}\StringTok{ }\FloatTok{0.95} 
+\NormalTok{cutPoint \textless{}{-}}\StringTok{ }\KeywordTok{quantile}\NormalTok{( }\DataTypeTok{x =}\NormalTok{ brcaMedian, }\DataTypeTok{probs =}\NormalTok{ myProb )}
+\CommentTok{\#\# genesToKeep \textless{}{-} names( brcaMedian ) [ which(brcaMedian \textgreater{}= cutPoint) ]}
+\NormalTok{genesToKeep \textless{}{-}}\StringTok{ }\KeywordTok{names}\NormalTok{( brcaMedian ) [ }\KeywordTok{which}\NormalTok{(brcaMedian }\OperatorTok{\textgreater{}=}\StringTok{ }\NormalTok{cutPoint) ]}
+\NormalTok{brca \textless{}{-}}\StringTok{ }\NormalTok{brca[ , genesToKeep ]}
+\CommentTok{\#\# ncol(brca) / 20501}
+\NormalTok{(d \textless{}{-}}\StringTok{ }\KeywordTok{length}\NormalTok{(genesToKeep))}
+\NormalTok{[}\DecValTok{1}\NormalTok{] }\DecValTok{1026}
 \end{Highlighting}
@@
 The data are overdispersed:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+
+\NormalTok{logVarOverMean \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{( x ) \{}
+    \KeywordTok{log}\NormalTok{ ( }\KeywordTok{var}\NormalTok{(x) }\OperatorTok{/}\StringTok{ }\KeywordTok{mean}\NormalTok{ (x) )}
+\NormalTok{\}}
+\CommentTok{\#\# check whether NB makes sense}
+
+\NormalTok{summaryBRCA \textless{}{-}}\StringTok{ }\NormalTok{brca }\OperatorTok{\%\textgreater{}\%}\StringTok{ }
+\StringTok{    }\KeywordTok{summarize\_all}\NormalTok{(  }\KeywordTok{list}\NormalTok{(}\OperatorTok{\textasciitilde{}}\StringTok{ }\KeywordTok{logVarOverMean}\NormalTok{( . ) ) )}
+\NormalTok{summaryBRCA \textless{}{-}}\StringTok{ }\KeywordTok{as.data.frame}\NormalTok{( }\KeywordTok{t}\NormalTok{(}\KeywordTok{as.data.frame}\NormalTok{( summaryBRCA )) )}
+\KeywordTok{names}\NormalTok{(summaryBRCA) \textless{}{-}}\StringTok{ "logVarOverMean"}
+\CommentTok{\#\# sum(summaryBRCA$logVarOverMean \textless{} 0)  \#\# no underdispersed genes}
+\KeywordTok{ggplot}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ summaryBRCA, }\DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ logVarOverMean)) }\OperatorTok{+}
+\StringTok{  }\KeywordTok{geom\_histogram}\NormalTok{(}\DataTypeTok{color =} \StringTok{"white"}\NormalTok{, }\DataTypeTok{bins =} \DecValTok{20}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{article_bigsimr_files/figure-latex/figNBratio-1}
+
+}
+
+\caption{The genes display extreme overdispersion.
+Using group-specific sample means and variances for each of the d=`r d` genes, we display the distributions oflog of variance to mean ratio.}\label{fig:figNBratio}
+\end{figure}
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  Kendall's \(\tau\) was estimated efficiently using the \texttt{pcaPP::cor.fk} function in \texttt{R}.
-\end{itemize}
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# full workflow simulating RNA{-}seq data}
 
-\hypertarget{ultra-high-dimensional-simulation-based-correlation-hypothesis-testing}{%
-\subsection{Ultra High-Dimensional simulation-based correlation hypothesis testing}\label{ultra-high-dimensional-simulation-based-correlation-hypothesis-testing}}
+\CommentTok{\#\# 1. Estimate Spearman\textquotesingle{}s correlation on the count data}
+\NormalTok{corType \textless{}{-}}\StringTok{ \textquotesingle{}spearman\textquotesingle{}}
+\KeywordTok{system.time}\NormalTok{( rho \textless{}{-}}\StringTok{ }\NormalTok{bigsimr}\OperatorTok{::}\KeywordTok{fastCor}\NormalTok{( brca, }\DataTypeTok{method =}\NormalTok{ corType ) )}
+\NormalTok{   user  system elapsed }
+  \FloatTok{0.654}   \FloatTok{0.013}   \FloatTok{0.673} 
+\CommentTok{\#\# Describe the correlations}
+\end{Highlighting}
+\end{Shaded}
 
@@
 \begin{Highlighting}[]
-\CommentTok{## full workflow simulating RNA-seq data}
+\CommentTok{\#\# 2. Estimate NegBin parameters using Method of Moments}
+\NormalTok{estimateNegBinMoM \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(tmpGene, }\DataTypeTok{minP =} \FloatTok{1e{-}5}\NormalTok{) \{}
+\NormalTok{    tmpMean \textless{}{-}}\StringTok{ }\KeywordTok{mean}\NormalTok{(tmpGene)}
+\NormalTok{    tmpVar \textless{}{-}}\StringTok{ }\KeywordTok{var}\NormalTok{(tmpGene)}
+    \CommentTok{\#\# relate to nbinom parameters}
+    \CommentTok{\#\# See ?rbinom for details.}
+\NormalTok{    p \textless{}{-}}\StringTok{ }\NormalTok{tmpMean }\OperatorTok{/}\StringTok{ }\NormalTok{tmpVar}
+    \ControlFlowTok{if}\NormalTok{ (p }\OperatorTok{\textless{}}\StringTok{ }\NormalTok{minP) \{p \textless{}{-}}\StringTok{ }\NormalTok{minP \} }\CommentTok{\#\# maybe add noise here}
+\NormalTok{    n \textless{}{-}}\StringTok{ }\NormalTok{( tmpMean }\OperatorTok{*}\StringTok{ }\NormalTok{p) }\OperatorTok{/}\StringTok{ }\NormalTok{( }\DecValTok{1}  \OperatorTok{{-}}\StringTok{ }\NormalTok{p )  }
+    \CommentTok{\#\# format for bigsimr margins}
+    \KeywordTok{return}\NormalTok{( }\KeywordTok{list}\NormalTok{(}\StringTok{"nbinom"}\NormalTok{, }\DataTypeTok{size =}\NormalTok{ n, }\DataTypeTok{prob =}\NormalTok{ p) )}
+\NormalTok{\}}
+\NormalTok{brcaMargins \textless{}{-}}\StringTok{ }\KeywordTok{apply}\NormalTok{( }\KeywordTok{unname}\NormalTok{(}\KeywordTok{as.matrix}\NormalTok{(brca)), }\DecValTok{2}\NormalTok{, estimateNegBinMoM )}
+\CommentTok{\#\# describe the margins}
+\CommentTok{\#\# check for too small prob}
+\CommentTok{\#\# head( sort( unlist( lapply( brcaMargins, function(x) \{x$prob\} ) ) ) )}
+\CommentTok{\#\# head( sort( unlist( lapply( brcaMargins, function(x) \{x$size\} ) ) ) )}
 \end{Highlighting}
 \end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# 3. Generate the simulated samples}
+\CommentTok{\#\# N \textless{}{-} 10}
+\CommentTok{\#\# system.time( simBRCA \textless{}{-} rvec(N, rho = rho[1:2, 1:2], params = brcaMargins[1:2], type = corType, adjustForDiscrete = FALSE) )}
+\NormalTok{N \textless{}{-}}\StringTok{ }\KeywordTok{nrow}\NormalTok{(brca)}
+\KeywordTok{system.time}\NormalTok{( simBRCA \textless{}{-}}\StringTok{ }\KeywordTok{rvec}\NormalTok{(N, }\DataTypeTok{rho =}\NormalTok{ rho, }\DataTypeTok{params =}\NormalTok{ brcaMargins, }\DataTypeTok{cores =}\NormalTok{ CORES, }\DataTypeTok{type =}\NormalTok{ corType, }\DataTypeTok{adjustForDiscrete =} \OtherTok{FALSE}\NormalTok{) )}
+\NormalTok{simBRCA[}\DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{, }\DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{ ]}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+
+\CommentTok{\#\# Describe and plot real data}
+\NormalTok{GGally}\OperatorTok{::}\KeywordTok{ggpairs}\NormalTok{(}\DataTypeTok{data =} \KeywordTok{as.data.frame}\NormalTok{(brca[ ,}\DecValTok{1}\OperatorTok{:}\DecValTok{3}\NormalTok{] ) )}
+
+\CommentTok{\#\# Describe and plot simulations}
+\NormalTok{GGally}\OperatorTok{::}\KeywordTok{ggpairs}\NormalTok{(}\DataTypeTok{data =} \KeywordTok{as.data.frame}\NormalTok{(simBRCA[ ,}\DecValTok{1}\OperatorTok{:}\DecValTok{3}\NormalTok{] ) )}
+
+\CommentTok{\#\# check 1st moments}
+\NormalTok{trueMu \textless{}{-}}\StringTok{ }\KeywordTok{colMeans}\NormalTok{(brca)}
+\NormalTok{simMu \textless{}{-}}\StringTok{ }\KeywordTok{colMeans}\NormalTok{(simBRCA)}
+\KeywordTok{qplot}\NormalTok{( }\DataTypeTok{x =}\NormalTok{ trueMu, }\DataTypeTok{y =}\NormalTok{ simMu ) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom\_abline}\NormalTok{(}\DataTypeTok{slope =} \DecValTok{1}\NormalTok{, }\DataTypeTok{intercept =} \DecValTok{0}\NormalTok{)}
+
+\CommentTok{\#\# check 2nd moment}
+\NormalTok{trueVar \textless{}{-}}\StringTok{ }\KeywordTok{apply}\NormalTok{( brca, }\DecValTok{2}\NormalTok{, var)}
+\NormalTok{simVar \textless{}{-}}\StringTok{ }\KeywordTok{apply}\NormalTok{( simBRCA, }\DecValTok{2}\NormalTok{, var)}
+\KeywordTok{qplot}\NormalTok{( }\DataTypeTok{x =}\NormalTok{ trueVar, }\DataTypeTok{y =}\NormalTok{ simVar ) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom\_abline}\NormalTok{(}\DataTypeTok{slope =} \DecValTok{1}\NormalTok{, }\DataTypeTok{intercept =} \DecValTok{0}\NormalTok{)}
+
+\CommentTok{\#\# check correlation}
+\KeywordTok{system.time}\NormalTok{( simRho \textless{}{-}}\StringTok{ }\NormalTok{bigsimr}\OperatorTok{::}\KeywordTok{fastCor}\NormalTok{( simBRCA, }\DataTypeTok{method =}\NormalTok{ corType ) )}
+\NormalTok{rho[}\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{, }\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{]}
+\NormalTok{simRho[}\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{, }\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{]}
+\NormalTok{trueRho \textless{}{-}}\StringTok{ }\NormalTok{rho[}\KeywordTok{lower.tri}\NormalTok{(rho)]}
+\NormalTok{simRho \textless{}{-}}\StringTok{ }\NormalTok{simRho[}\KeywordTok{lower.tri}\NormalTok{(simRho)]}
+\KeywordTok{qplot}\NormalTok{( }\DataTypeTok{x =}\NormalTok{ trueRho, }\DataTypeTok{y =}\NormalTok{ simRho ) }\OperatorTok{+}\StringTok{ }\KeywordTok{geom\_abline}\NormalTok{(}\DataTypeTok{slope =} \DecValTok{1}\NormalTok{, }\DataTypeTok{intercept =} \DecValTok{0}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{ultra-high-dimensional-simulation-based-correlation-hypothesis-testing-under}{%
+\subsection{Ultra High-Dimensional simulation-based correlation hypothesis testing under}\label{ultra-high-dimensional-simulation-based-correlation-hypothesis-testing-under}}
+
+Since many of the correlations are near zero, one could ask whether it is
+possible that
 
@@
 
-\CommentTok{## simulation based}
-\CommentTok{## devtools::install_github("adknudson/bigsimr", ref="develop")}
+\CommentTok{\#\# simulation based}
+\CommentTok{\#\# devtools::install\_github("adknudson/bigsimr", ref="develop")}
 \KeywordTok{library}\NormalTok{(bigsimr)}
-\CommentTok{## ?rvec}
+\CommentTok{\#\# ?rvec}
 
-\NormalTok{d <-}\StringTok{ }\DecValTok{5}
-\NormalTok{rho <-}\StringTok{ }\KeywordTok{rcor}\NormalTok{(d)}
-\NormalTok{margins <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}
+\NormalTok{d \textless{}{-}}\StringTok{ }\DecValTok{5}
+\NormalTok{rho \textless{}{-}}\StringTok{ }\KeywordTok{rcor}\NormalTok{(d)}
+\NormalTok{margins \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}
     \KeywordTok{list}\NormalTok{(}\StringTok{"nbinom"}\NormalTok{, }\DataTypeTok{size =} \DecValTok{5}\NormalTok{, }\DataTypeTok{prob =} \FloatTok{0.3}\NormalTok{),}
@@
 \NormalTok{rho =}\StringTok{ }\NormalTok{rho}
-\CommentTok{## cores = parallel::detectCores() - 1}
+\CommentTok{\#\# cores = parallel::detectCores() {-} 1}
 \NormalTok{cores =}\StringTok{ }\DecValTok{2}
@@
 
-\CommentTok{## help(package="bigsimr")}
-\NormalTok{rho_bounds <-}\StringTok{ }\KeywordTok{computeCorBounds}\NormalTok{(}\DataTypeTok{params =}\NormalTok{ margins, }\DataTypeTok{cores =}\NormalTok{ cores, }\DataTypeTok{type =}\NormalTok{ type, }\DataTypeTok{reps =}\NormalTok{ reps)}
-\NormalTok{rho_bounds}
+\CommentTok{\#\# help(package="bigsimr")}
+\NormalTok{rho\_bounds \textless{}{-}}\StringTok{ }\KeywordTok{computeCorBounds}\NormalTok{(}\DataTypeTok{params =}\NormalTok{ margins, }\DataTypeTok{cores =}\NormalTok{ cores, }\DataTypeTok{type =}\NormalTok{ type, }\DataTypeTok{reps =}\NormalTok{ reps)}
+\NormalTok{rho\_bounds}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{## full workflow simulating RNA-seq data}
+\CommentTok{\#\# full workflow simulating RNA{-}seq data}
 \end{Highlighting}
@@
 One could imagine additional structures/features in the data that an analyst could model to improve the correspondence with the empirical values, for example row correlations and high-dimensional covariance estimators (see \citet{Won2013g}).
+
+Mention \texttt{rslurm}
 

### 2020-06-15 index.Rmd
migrate to html based preview

@@
 site: bookdown::bookdown_site
-output: 
-  bookdown::gitbook:
-    base_format: rticles::tf_article
-documentclass: article
 bibliography: [bigsimr.bib, packages.bib]
@@
 library(dplyr)
+## devtools::install_github( "adknudson/bigsimr", ref = 'develop')
 set.seed(06082020)

### 2020-06-16 050-some-applications.Rmd
update case study with new syntax

@@
 ## retain top (1-probs)*100% highest expressing genes for illustration
-myProb <- 0.95 
+myProb <- 0.99
 cutPoint <- quantile( x = brcaMedian, probs = myProb )
@@
 ## 2. Estimate NegBin parameters using Method of Moments
-estimateNegBinMoM <- function(tmpGene, minP = 1e-5) {
+estimateNegBinMoM <- function(tmpGene) {
     tmpMean <- mean(tmpGene)
@@
     p <- tmpMean / tmpVar
-    if (p < minP) {p <- minP } ## maybe add noise here
     n <- ( tmpMean * p) / ( 1  - p )  
     ## format for bigsimr margins
-    return( list("nbinom", size = n, prob = p) )
+    get( 'qnbinom')
+    return( str2lang( paste0( "qnbinom(size = ", n, ", prob = ", p, ")" ) ) )
 }
 brcaMargins <- apply( unname(as.matrix(brca)), 2, estimateNegBinMoM )
+## brcaMargins <- apply( unname(as.matrix(brca))[ , 1:3], 2, estimateNegBinMoM )
 ## describe the margins
@@
 
-```{r simBRCA, echo=TRUE, eval = FALSE, cache=TRUE}
+```{r simBRCA, echo=TRUE, eval = TRUE, cache=TRUE}
 ## 3. Generate the simulated samples
-## N <- 10
-## system.time( simBRCA <- rvec(N, rho = rho[1:2, 1:2], params = brcaMargins[1:2], type = corType, adjustForDiscrete = FALSE) )
 N <- nrow(brca)
-system.time( simBRCA <- rvec(N, rho = rho, params = brcaMargins, cores = CORES, type = corType, adjustForDiscrete = FALSE) )
+system.time( simBRCA <- rvec(n = N, rho = rho, margins = brcaMargins, cores = CORES, type = corType) )
 simBRCA[1:2, 1:2 ]

### 2020-06-16 index.Rmd
update case study with new syntax

@@
 title: "Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package"
-author: [A.G. Schissler, A. Knudson]
+author: [A.G. Schissler, A. Knudson, T. Kozubowski, A. Panorksa]
 date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"

### 2020-06-19 index.Rmd
add serve_book() convenience comment

@@
 ## devtools::install_github( "adknudson/bigsimr", ref = 'develop')
+## bookdown::serve_book()
 set.seed(06082020)

### 2020-06-20 030-bigsimr-package.Rmd
update case study with new syntax
draft rslurm section

@@
 ```{r setup, eval = TRUE}
+## bookdown::preview_chapter('030-bigsimr-package.Rmd')
 library(bigsimr)
-# Reticulate needs to be able to find the python binary with `jax` installed
-reticulate::use_condaenv("bigsimr-cpu")
 ```
 
-## Basic use
-
-short description
-
-### Specifying marginals
+## Specifying marginals
 
@@
 
-### Specify correlation
+## Specify correlation
 
@@
 
-### Small sample
+## Small sample
 
@@
 
-### Scaling up N
+## Scaling up N
 
 ```{r, fig.width=7}
-x <- rvec(100000, rho = rho, margins = margins, type = "spearman")
+x <- rvec(1e6, rho = rho, margins = margins, type = "spearman")
 
@@
 
-### Evaluation
+## Evaluation
 
@@
 Another possibility is that the copula algorithm is biased and needs correction.
+
+## Easy cluster implementation through `rslurm`
+
+Though `bigsimr` runs quickly, at large $d$ users may want to run jobs on a shared computing server.
+The R package `rslurm` makes it esay to run embarrasingly large parallel `rvec` calls.
+This example assumes that `bigsimr` is installed on a system
+with a slurm scheduler installed.
+
+```{r exampleRslurm, echo=FALSE, eval=FALSE, include=FALSE}
+## run this code once and use the generated rds later.
+## cyberhelp.sesync.org/rslurm/articles/rslurm.html
+test_func <- function(par_mu, par_sd) {
+    samp <- rnorm(10^6, par_mu, par_sd)
+    c(s_mu = mean(samp), s_sd = sd(samp))
+}
+pars <- data.frame(par_mu = 1:10,
+                   par_sd = seq(0.1, 1, length.out = 10))
+library(rslurm)
+sjob <- slurm_apply(test_func, pars, jobname = 'test_apply',
+                    nodes = 2, cpus_per_node = 2, submit = TRUE)
+list.files('_rslurm_test_apply', 'results')
+res <- get_slurm_out(sjob, outtype = 'table')
+head(res, 3)
+cleanup_files(sjob) ## careful this deletes output also
+
+## single function evaluation
+sjob <- slurm_call(test_func, jobname = 'test_call',
+                   list(par_mu = 5, par_sd = 1), submit = FALSE)
+
+## try rvec
+sjob <- slurm_call(rvec, jobname = 'rvec',
+                   list(par_mu = 5, par_sd = 1), submit = FALSE)
+```
+
+```{r rvecRslurm, echo=TRUE, eval=FALSE}
+library(rslurm)
+## a single call of rvec
+## had to edit rslurm/templates/submit_single_sh.txt
+## source ~/.bashrc
+## conda activate bigsimr
+## conda env config vars list
+sjob <- slurm_call(rvec, jobname = 'rvec',
+                   list(n=1e6,
+                        rho = rho,
+                        margins = margins,
+                        type = "spearman"),
+                   submit = TRUE)
+
+res <- readRDS(file.path('_rslurm_rvec', list.files('_rslurm_rvec', 'results')))
+head(res, 3)
+cleanup_files(sjob) ## careful this deletes output also
+```
+
+```{r rvecRslurmApply, echo=TRUE, eval=TRUE}
+library(rslurm)
+## a single call of rvec
+## had to edit rslurm/templates/submit_single_sh.txt
+## source ~/.bashrc
+## conda activate bigsimr
+## conda env config vars list
+sjob <- slurm_call(rvec, jobname = 'rvec',
+                   list(n=1e6,
+                        rho = rho,
+                        margins = margins,
+                        type = "spearman"),
+                   submit = TRUE)
+
+res <- readRDS(file.path('_rslurm_rvec', list.files('_rslurm_rvec', 'results')))
+head(res, 3)
+cleanup_files(sjob) ## careful this deletes output also
+```

### 2020-06-20 index.Rmd
draft rslurm section

--- 2020-06-16 index.Rmd
@@
 library(dplyr)
-## devtools::install_github( "adknudson/bigsimr", ref = 'develop')
+## rmote::start_rmote()
+## library(ggplot2); qplot(mpg, wt, data=mtcars, colour=cyl)
+## bookdown::serve_book()
 set.seed(06082020)

### 2020-06-20 030-bigsimr-package.Rmd
dont evaluate rslurm jobs

@@
 ## bookdown::preview_chapter('030-bigsimr-package.Rmd')
+## bookdown::render_book('index.html')
 library(bigsimr)
@@
                    submit = TRUE)
+```
 
+```{r rvecRslurmRes, echo=TRUE, eval=TRUE}
 res <- readRDS(file.path('_rslurm_rvec', list.files('_rslurm_rvec', 'results')))
 head(res, 3)
-cleanup_files(sjob) ## careful this deletes output also
+## cleanup_files(sjob) ## careful this deletes output also
 ```
-
-```{r rvecRslurmApply, echo=TRUE, eval=TRUE}
-library(rslurm)
-## a single call of rvec
-## had to edit rslurm/templates/submit_single_sh.txt
-## source ~/.bashrc
-## conda activate bigsimr
-## conda env config vars list
-sjob <- slurm_call(rvec, jobname = 'rvec',
-                   list(n=1e6,
-                        rho = rho,
-                        margins = margins,
-                        type = "spearman"),
-                   submit = TRUE)
-
-res <- readRDS(file.path('_rslurm_rvec', list.files('_rslurm_rvec', 'results')))
-head(res, 3)
-cleanup_files(sjob) ## careful this deletes output also
-```

### 2020-06-20 030-bigsimr-package.Rmd
move rslurm example to advanced use, make corr test simpler

@@
 
+## Basic use
+
 ```{r setup, eval = TRUE}
-## bookdown::preview_chapter('030-bigsimr-package.Rmd')
+## setwd('~/Research/article_bigsimr/'); bookdown::preview_chapter('030-bigsimr-package.Rmd')
 ## bookdown::render_book('index.html')
 library(bigsimr)
+## reticulate::use_condaenv( 'bigsimr-cpu' )
 ```
 
-## Specifying marginals
+### Specifying marginals
 
@@
 
-## Specify correlation
+### Specify correlation
 
@@
 
-## Small sample
+### Generating a few random vectors
 
@@
 
-## Scaling up N
+### Scaling up N to 1,000,000
 
@@
 
-## Evaluation
+### Check the performance with fast correlation estimation
 
 ```{r}
-cor(x, method = 'spearman')
+bigsimr::fastCor(x, method = 'spearman')
 ```
@@
 
-## Easy cluster implementation through `rslurm`
+## Advanced use
+
+### Using multicore
+
+Using multicore is easy simply specify the number of cores.
+
+### Simulation-based computation of correlation bounds
+
+Using the Generate, sort, and correlate algorithm @Demirtas2011  and TAS issue
+
+- use a large number of reps
+
+```{r, computeBounds, echo = TRUE, eval=FALSE}
+## Possibly bug? or configuration
+## Error: segfault from C stack overflow on my Okapi bigsimr env 
+rho_bounds <- bigsimr::computeCorBounds(margins = margins, type = 'spearman', cores = 1, reps = 1e5)
+rho_bounds
+```
+
+### Using `bigsimr` on a computing cluster via `rslurm`
 
@@
 The R package `rslurm` makes it esay to run embarrasingly large parallel `rvec` calls.
-This example assumes that `bigsimr` is installed on a system
-with a slurm scheduler installed.
+This example assumes that `bigsimr` is installed on a system with a slurm scheduler installed.
 
@@
 ```
+
+Now, let's show off the real power of combining `bigsimr` and `rslurm` by simulating many correlation structures for these three marginals.
+The `rslurm::slumr_map` syntax mirrors the familiar `base::lapply` and `purrr::map` functions.
+
+```{r rvecRslurmMap, echo=TRUE, eval=FALSE}
+library(rslurm)
+## a single call of rvec
+## had to edit rslurm/templates/submit_sh.txt
+## source ~/.bashrc
+## conda activate bigsimr
+## conda env config vars list
+set.seed(06202020)
+simReps <- 100
+rhoList <- replicate( n = simReps, bigsimr::rcor(d = length(margins), constant_rho = TRUE), simplify=FALSE )
+sjob <- slurm_map(x = rhoList,
+                  f = rvec,
+                  jobname = 'rvecMap',
+                  n=1e6,
+                  margins = margins,
+                  type = "spearman",
+                  nodes = 4,
+                  cpus_per_node = 4,
+                  submit = TRUE)
+
+```
+
+On a cluster carrying 24 nodes with 48 threads, these 100 jobs completed in about a minute. Let's evaluate the quality of simulation performance.
+
+```{r rvecRslurmMapRes, echo=TRUE, eval=FALSE}
+list.files('_rslurm_rvecMap', 'results')
+res <- rslurm::get_slurm_out(sjob, outtype = 'raw')
+rhoHats <- lapply( res, bigsimr::fastCor, method = 'spearman' )
+
+resData <- NULL
+simreps <- length(rhoHats)
+for (i in 1:simreps){
+    tmpRhoHat <- rhoHats[[i]]
+    ## average together
+    margins
+    avgRhoHat <- tmpRhoHat[lower.tri(tmpRhoHat)]
+    tmpPair <- paste0( "pair", apply( combn(x = length(margins), 2), 2, paste, collapse = "_" ) )
+    tmpResData <- data.frame( simNum=i, rho=rhoList[[i]][1,2], rhoHat=tmpRhoHat[lower.tri(tmpRhoHat)], pair=tmpPair )
+    resData <- rbind(resData, tmpResData)
+}
+
+## "wrangle" into a tibble
+library(tidyverse)
+res <- as_tibble(resData)
+## save for viz
+saveRDS( res, file = 'data/rvecRslurmMapRes.rds' )
+
+## cleanup careful this deletes output also
+## cleanup_files(sjob) 
+```
+
+```{r rvecRslurmMapFig, echo = TRUE, eval=TRUE, message=FALSE, warning=FALSE, out.width= '75%', fig.align='center', fig.cap='The continous marginal pairs exact reproduce the specified correlation whereas the discrete pairs (green nad blue) show a downward bias.
+Future work with employ a discrete adjusted procedures using *rescaling* on the traditional correlation coefficient.'}
+library(tidyverse)
+
+res <- readRDS( 'data/rvecRslurmMapRes.rds' )
+res
+
+res %>%
+    ggplot( aes( x = rho, y = rhoHat, color = pair ) )  +
+    geom_point() +
+    geom_abline( slope = 1, intercept = 0 )
+```

### 2020-06-20 050-some-applications.Rmd
draft rslurm section
move rslurm example to advanced use, make corr test simpler

@@
 Since many of the correlations are near zero, one could ask whether it is
-possible that
+possible that test the hypothesis that $H_0: R = R_0$.
 
-```{r brcaSigmaTesting, echo=TRUE, eval=FALSE}
+```{r corrTest, echo=FALSE, eval=FALSE, include=FALSE}
 
-```
-
-## Simulation-based computation of correlation bounds
-
-Using the Generate, sort, and correlate algorithm @Demirtas2011
-
-```{r, computeBounds, echo = TRUE, eval=FALSE}
-
-## simulation based
-## devtools::install_github("adknudson/bigsimr", ref="develop")
-library(bigsimr)
-## ?rvec
-
-d <- 5
-rho <- rcor(d)
-margins <- list(
-    list("nbinom", size = 5, prob = 0.3),
-    list("exp", rate = 4),
-    list("binom", size = 5, prob = 0.7),
-    list("norm", mean = 10, sd = 3),
-    list("pois", lambda = 10)
-)
-
-params = margins
-rho = rho
-## cores = parallel::detectCores() - 1
-cores = 2
-reps = 1e3
-type = "spearman"
-
-## help(package="bigsimr")
-rho_bounds <- computeCorBounds(params = margins, cores = cores, type = type, reps = reps)
-rho_bounds
 ```

### 2020-09-18 020-simulation-algorithm.Rmd
update case study with new syntax
get started after a long break and note the current limitations

@@
 
+- discuss the consequences of the transformations to obtain the corresponding Pearson matrix and describe the quadratic algorithm implemented to address this (Alex).
 - Discrete margins pose some difficulties and matching *unadjusted* correlation
@@
    F_j^{-1}(\Phi(z_j))dy_idy_j$. We have yet to accelerate this process.
-- We GPU accelerate whenever possible during our proposed High-Performance NORTA
+- We GPU accelerate certain steps (Alex which steps?) possible during our proposed High-Performance NORTA
   algorithm.
@@
 test workflow again
+test after a long while

### 2020-09-18 030-bigsimr-package.Rmd
get started after a long break and note the current limitations

@@
 
-```{r setup, eval = TRUE}
+```{r setup3, eval = TRUE}
 ## setwd('~/Research/article_bigsimr/'); bookdown::preview_chapter('030-bigsimr-package.Rmd')
@@
 library(bigsimr)
-## reticulate::use_condaenv( 'bigsimr-cpu' )
+library(reticulate)
+reticulate::use_condaenv("bigsimr-cpu")
 ```
@@
 
-```{r}
+```{r setMargins}
 margins = alist(
@@
 
-```{r}
+```{r setCor}
 rho <- matrix(0.5, nrow = 3, ncol = 3)
@@
 
-```{r}
-x <- rvec(10, rho = rho, margins = margins, type = "spearman")
+```{r rvec10}
+x <- bigsimr::rvec(10, rho = rho, margins = margins, type = "spearman")
 ```
 
-```{r, echo=FALSE}
+```{r gpuWarn, echo=FALSE}
 warning("warning.warn('No GPU/TPU found, falling back to CPU.')")
@@
 
-```{r}
+```{r printX}
 x
@@
 
-```{r, fig.width=7}
+```{r rvec1e6, fig.width=7}
 x <- rvec(1e6, rho = rho, margins = margins, type = "spearman")
@@
 
-```{r}
+```{r fastCor}
 bigsimr::fastCor(x, method = 'spearman')
@@
 
-```{r, computeBounds, echo = TRUE, eval=FALSE}
+```{r computeBounds, echo = TRUE, eval=FALSE}
 ## Possibly bug? or configuration
@@
 This example assumes that `bigsimr` is installed on a system with a slurm scheduler installed.
-
-```{r exampleRslurm, echo=FALSE, eval=FALSE, include=FALSE}
-## run this code once and use the generated rds later.
-## cyberhelp.sesync.org/rslurm/articles/rslurm.html
-test_func <- function(par_mu, par_sd) {
-    samp <- rnorm(10^6, par_mu, par_sd)
-    c(s_mu = mean(samp), s_sd = sd(samp))
-}
-pars <- data.frame(par_mu = 1:10,
-                   par_sd = seq(0.1, 1, length.out = 10))
-library(rslurm)
-sjob <- slurm_apply(test_func, pars, jobname = 'test_apply',
-                    nodes = 2, cpus_per_node = 2, submit = TRUE)
-list.files('_rslurm_test_apply', 'results')
-res <- get_slurm_out(sjob, outtype = 'table')
-head(res, 3)
-cleanup_files(sjob) ## careful this deletes output also
-
-## single function evaluation
-sjob <- slurm_call(test_func, jobname = 'test_call',
-                   list(par_mu = 5, par_sd = 1), submit = FALSE)
-
-## try rvec
-sjob <- slurm_call(rvec, jobname = 'rvec',
-                   list(par_mu = 5, par_sd = 1), submit = FALSE)
-```
 
@@
 
-```{r rvecRslurmRes, echo=TRUE, eval=TRUE}
+```{r rvecRslurmRes, echo=TRUE, eval=FALSE}
 res <- readRDS(file.path('_rslurm_rvec', list.files('_rslurm_rvec', 'results')))

### 2020-09-18 050-some-applications.Rmd
get started after a long break and note the current limitations

@@
 library(bigsimr)
-reticulate::use_condaenv("py37")
+reticulate::use_condaenv("bigsimr")
 CORES <- parallel::detectCores() - 1

### 2020-09-18 060-conclusion-discussion.Rmd
migrate to html based preview
get started after a long break and note the current limitations

@@
 
+Mention `rslurm`
+
 There are of course limitations to the methodology and implementation.
-The algorithm requires invertible marginal cdfs.
-This leaves some restriction to the available marginal probability distributions (DOES IT?
-FOR EXAMPLE?) An additional source of error is after converting dependencies (step i; ${\bf R_{Kendall}})$) the resultant matrix may not be positive definite (PD).
-Then the nearest PD matrix is used.
+The most obvious missing feature the proposed methodology is the inability to match a Pearson correlation matrix exactly.
+As discussed above, this is a computational intense procedure and not a natural choice for the posed problem (seeking to simulate non-normal margins in a scaleable fashion).
+Yet it is an important aspect that paralleziable, approximation based apporaches are promising [ref Hermite polynomials].
+Further, while we provide the ability of the user to specify discrete marginals, exact matching a desired dependency measure is not yet obtained.
+One potential solution for this is *rescale* to account for ties [ref].
+Finally we note, that the algorithm requires invertible marginal cdfs.
+This gives some restriction to the available marginal probability distributions (DOES IT?
+FOR EXAMPLE?).
+
 From a practical computing standpoint, the user's computing resource provides the ultimate limit on how large a dimension can be simulated using the `bigsimr` R package.
@@
 
-Future work includes developing scalable algorithms to match the Pearson correlation matrix exactly.
+Future work includes developing scalable algorithms to match the Pearson correlation matrix exactly and more methods developed specifically for discrete distributions.
 Further, more sophisticated approaches could involve simulation algorithms that are "estimation aware" and so could explore the role of covariance estimation within the simulation.

### 2020-09-18 index.Rmd
get started after a long break and note the current limitations

--- 2020-06-16 index.Rmd
@@
 title: "Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package"
-author: [A.G. Schissler, A. Knudson, T. Kozubowski, A. Panorksa]
+author: [A.G. Schissler, R.D.L. Foote, A.D. Knudson, T.J. Kozubowski, A.K. Panorska]
 date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"
@@
 
-```{r LoadLib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
+```{r indexSetup, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
 library(ggplot2)
@@
 library(dplyr)
-## devtools::install_github( "adknudson/bigsimr", ref = 'develop')
 set.seed(06082020)
@@
 
-```{bash copyBib, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, results = "hide"}
-cp /Users/alfred/Dropbox/bib/bigsimr.bib ./.
+```{r copyBib, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, results = "hide"}
+system( 'cp ~/bib/bigsimr.bib ./.' )
 ```
@@
 ), 'packages.bib')
+```
+
+```{r serveHelper, include=FALSE, eval=FALSE}
+## rmote::start_rmote()
+## library(ggplot2); qplot(mpg, wt, data=mtcars, colour=cyl)
+bookdown::serve_book()
+```
+
+```{r countWords, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, results = "hide"}
+rmdFiles <- dir(pattern = 'Rmd$', ignore.case= T)
+## rmdFiles <- rmdFiles[-grep('^_', rmdFiles)]
+rmdFiles <- rmdFiles[-grep('references', rmdFiles)]
+## rmdFiles <- rmdFiles[-grep('README', rmdFiles)]
+## rmdFiles <- rmdFiles[-grep('article', rmdFiles)] ## after generating a single document
+## wordcountaddin::word_count( rmdFiles[11] )
+(counts <- sapply( rmdFiles, wordcountaddin::word_count, simplify = T ))
+sum(counts)
 ```
@@
 
-Along those lines, simulating high-dimension dependent negative binomial data motivates this work --- in pursuit of modeling RNA-sequencing data [@Wang2009b; @Conesa2016b] derived from breast cancer patients.
+Along those lines, simulating high-dimensional non-normal data motivates this work --- in pursuit of modeling RNA-sequencing data [@Wang2009b; @Conesa2016b] derived from breast cancer patients.
 This laboratory procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions.
@@
 
-With our application in mind and seeking a general-purpose algorithm with broad applications, our purposed methodology should possess the following properties (adapted from criteria from @Nik13a):
+With our application in mind and seeking a fairly general-purpose algorithm with broad applications, our purposed methodology should possess the following properties (adapted from criteria from @Nik13a):
 

### 2020-09-21 010-background.Rmd
clarify vision and organize

@@
 This idea is well known [refs].
+We chose a Gaussian copula among the many available copula functions primarily for two reasons: 1) the computationally convient closed expression to convert among the most common dependency measures when the margins are Gaussian and 2) the availability of high-performance multivariate normal simulators in R ('mvnfast' ref).
+
 A copula is a distribution function on $[0,1]^d$ describing a random vector with standard uniform marginals.

### 2020-09-21 030-bigsimr-package.Rmd
clarify vision and organize

@@
 
-```{r setup3, eval = TRUE}
+```{r setup3, eval = FALSE}
 ## setwd('~/Research/article_bigsimr/'); bookdown::preview_chapter('030-bigsimr-package.Rmd')
@@
 
-```{r setMargins}
+```{r setMargins, eval = FALSE}
 margins = alist(
@@
 
-```{r setCor}
+```{r setCor, eval = FALSE}
 rho <- matrix(0.5, nrow = 3, ncol = 3)
@@
 
-```{r rvec10}
+```{r rvec10, eval = FALSE}
 x <- bigsimr::rvec(10, rho = rho, margins = margins, type = "spearman")
@@
 
-```{r printX}
+```{r printX, eval = FALSE}
 x
@@
 
-```{r rvec1e6, fig.width=7}
+```{r rvec1e6, fig.width=7, eval = FALSE}
 x <- rvec(1e6, rho = rho, margins = margins, type = "spearman")
@@
 
-```{r fastCor}
-bigsimr::fastCor(x, method = 'spearman')
+```{r fastCor, eval = FALSE}
+## help( package = 'bigsimr')
+bigsimr::cor_fast(x, method = 'spearman')
 ```
@@
 
-```{r computeBounds, echo = TRUE, eval=FALSE}
+```{r computeBounds, echo = FALSE, eval=FALSE}
 ## Possibly bug? or configuration
@@
 
-```{r rvecRslurm, echo=TRUE, eval=FALSE}
+```{r rvecRslurm, echo=FALSE, eval=FALSE}
 library(rslurm)
@@
 
-```{r rvecRslurmRes, echo=TRUE, eval=FALSE}
+```{r rvecRslurmRes, echo=FALSE, eval=FALSE}
 res <- readRDS(file.path('_rslurm_rvec', list.files('_rslurm_rvec', 'results')))
@@
 
-```{r rvecRslurmMap, echo=TRUE, eval=FALSE}
+```{r rvecRslurmMap, echo=FALSE, eval=FALSE}
 library(rslurm)
@@
 
-```{r rvecRslurmMapRes, echo=TRUE, eval=FALSE}
+```{r rvecRslurmMapRes, echo=FALSE, eval=FALSE}
 list.files('_rslurm_rvecMap', 'results')
@@
 
-```{r rvecRslurmMapFig, echo = TRUE, eval=TRUE, message=FALSE, warning=FALSE, out.width= '75%', fig.align='center', fig.cap='The continous marginal pairs exact reproduce the specified correlation whereas the discrete pairs (green nad blue) show a downward bias.
+```{r rvecRslurmMapFig, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, out.width= '75%', fig.align='center', fig.cap='The continous marginal pairs exact reproduce the specified correlation whereas the discrete pairs (green nad blue) show a downward bias.
 Future work with employ a discrete adjusted procedures using *rescaling* on the traditional correlation coefficient.'}

### 2020-09-21 050-some-applications.Rmd
clarify vision and organize

--- 2020-06-20 050-some-applications.Rmd
@@
 
-```{r LoadLib050, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
+```{r LoadLib050, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, results = "hide"}
 library(ggplot2)
@@
 
-```{r processBRCA, echo=TRUE, eval=TRUE, cache=TRUE}
+```{r processBRCA, echo=FALSE, eval=FALSE, cache=TRUE}
 ## full workflow simulating RNA-seq data
@@
 
-```{r figNBratio, echo = TRUE, eval=TRUE, message=FALSE, warning=FALSE, out.width= '75%', fig.align='center', fig.cap='The genes display extreme overdispersion.
+```{r figNBratio, echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE, out.width= '75%', fig.align='center', fig.cap='The genes display extreme overdispersion.
 Using group-specific sample means and variances for each of the d=`r d` genes, we display the distributions oflog of variance to mean ratio.'}
@@
 
-```{r estRhoBRCA, echo=TRUE, eval=TRUE}
+```{r estRhoBRCA, echo=FALSE, eval=FALSE}
 ## full workflow simulating RNA-seq data
@@
 
-```{r estMargins, echo = TRUE, eval = TRUE, cache=TRUE}
+```{r estMargins, echo=FALSE, eval=FALSE, cache=TRUE}
 ## 2. Estimate NegBin parameters using Method of Moments
@@
 
-```{r simBRCA, echo=TRUE, eval = TRUE, cache=TRUE}
+```{r simBRCA, echo=FALSE, eval=FALSE, cache=TRUE}
 ## 3. Generate the simulated samples
@@
 
-```{r evalBRCA, echo=TRUE, eval = FALSE, cache=TRUE}
+```{r evalBRCA, echo=FALSE, eval=FALSE, cache=TRUE}
 
@@
 
-```{r densityEvaluation, echo=TRUE, eval=FALSE}
+```{r densityEvaluation, echo=FALSE, eval=FALSE}
 ## full workflow simulating RNA-seq data

### 2020-09-21 index.Rmd
clarify vision and organize

@@
 --- 
-title: "Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package"
-author: [A.G. Schissler, R.D.L. Foote, A.D. Knudson, T.J. Kozubowski, A.K. Panorska]
+title: Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package
+author:
+  - name: Alice Anonymous
+    email: alice@example.com
+    affiliation: Some Institute of Technology
+    footnote: 1
+  - name: Bob Security
+    email: bob@example.com
+    affiliation: Another University
+  - name: Cat Memes
+    email: cat@example.com
+    affiliation: Another University
+    footnote: 2
+  - name: Derek Zoolander
+    email: derek@example.com
+    affiliation: Some Institute of Technology
+    footnote: 2
+address:
+  - code: Some Institute of Technology
+    address: Department, Street, City, State, Zip
+  - code: Another University
+    address: Department, Street, City, State, Zip
+footnote:
+  - code: 1
+    text: "Corresponding Author"
+  - code: 2
+    text: "Equal contribution"
 date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"
+bibliography:
+  - bigsimr.bib
+  - packages.bib
 site: bookdown::bookdown_site
-bibliography: [bigsimr.bib, packages.bib]
-biblio-style: apalike
-link-citations: yes
+journal: "Journal of Computational And Applied Mathematics (JCAM)"
+#linenumbers: true
+#numbersections: true
 abstract: |
@@
 ```{r indexSetup, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
-library(ggplot2)
 library(tidyverse)
 library(knitr)
-library(dplyr)
-set.seed(06082020)
+library(rmarkdown)
+library(bookdown)
+## setwd('~/article_bigsimr')
+## bookdown::render_book("index.Rmd", "bookdown::pdf_book")
+## bookdown::serve_book()
+set.seed(09212020)
 ```
@@
 
-Massive high-dimensional data sets are commonplace in many areas of scientific inquiry.
+Massive high-dimensional data sets are now commonplace in many areas of scientific inquiry.
 As new methods are developed for these data, a fundamental challenge lies in designing and conducting simulation studies to assess operating characteristics of proposed methodology, such as false positive rates, statistical power, and robustness --- often in comparison to existing methods.

### 2020-09-21 notebooks/esco2020_talk.rmd
clarify vision and organize

---
title: "Simulating Ultra High-Dimensional Multivariate Data"
subtitle: "Using the `bigsimr` R package"
author: "A. Grant Schissler, Alexander Knudson, Tomasz J. Kozubowski, Anna K. Panorska"
institute: |
    "Department of Mathematics & Statistics" |
    "University of Nevada, Reno"
date: "8 Jun 2020 (updated: `r Sys.Date()`)"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
    incremental: true
    logo: ../man/figures/logo.png
bibliography: /Users/alfred/Dropbox/bib/bigsimr.bib
abstract: "In this era of Big Data, it is critical to realistically simulate data to conduct informative Monte Carlo studies. This is problematic when data are inherently multivariate while at the same time are (ultra-) high dimensional. This situation appears frequently in observational data found on online and in high-throughput biomedical experiments (e.g., RNA-sequencing). Due to the difficulty in simulating realistic correlated data points, researchers often resort to simulation designs that posit independence --- greatly diminishing the insight into the empirical operating characteristics of any proposed methodology.  Major challenges lie in the computational complexity involved in simulating these massive random vectors. We propose a fairly general, scalable procedure to simulate high-dimensional multivariate distributions with pre-specified marginal characteristics and dependency characteristics. As a motivating example, we use our methodology to study large-scale statistical inferential procedures applied to cancer-related RNA-sequencing data sets. The proposed algorithm is implemented as the bigsimr R package."
---

```{r LoadLib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
library(ggplot2)
library(tidyverse)
library(knitr)
knitr::opts_chunk$set(cache=TRUE, autodep=TRUE, cache.comments=TRUE)
library(dplyr)
CORES <- parallel::detectCores() - 1
set.seed(06082020)
## devtools::install_github("ggobi/GGally")
## for ggpairs
library(GGally)
```

# 1 – Introduction

## Problem we're addressing

- In the era of Big Data, it is critical to realistically simulate data.
- This is computationally challenging when data are correlated and (ultra-) high dimensional.
- Many domains are producing these type of data.
- A prototypical example is RNA-sequencing data derived from high-throughput biomedical experiments.

## Our goal

- Our goal is to simulate $N$ random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with **correlated** components.
- The simulation replicates $N$ can be large (~100,000).
- And the dimension $d$ can be very large (~20,000).
- We propose a scaleable NORmal To Anything approach to this task .
- We provide a high-performance and GPU-accelerated implementation in the `R` package `bigsimr`.

## Desired high-dimensional multivariate simulation properties

Our purposed methodology should possess the following properties  
(adapted from [@Nik13a]):

* P1: Wide range of dependence, allowing both positive and negative dependence
* P2: Flexible dependence, meaning that the number of bivariate marginals is (approximately) equal to the number of dependence parameters.
* P3: Flexible marginal modeling, generating heterogeneous data --- possibly from differing probability families.

Moreover, the simulation method must scale to high dimensions:

* S1: Procedure must scale to high dimensions, computable in a reasonable amount time.
* S2: Procedure must scale to high dimensions while maintaining accuracy.

## NORmal To Anything algorithm

To simulate a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$,
there is the well-known NORTA algorithm [@Cario1997]. It follows like this:

1. Simulate a random vector $\bf Z$ with $d$ **independent** and **identical** standard normal components.
2. Determine the input matrix $\Sigma_{\bf Z}$ to corresponds with the
   specified output $\Sigma_{\bf Y}$ [@Chen2001; @Xia17]
3. Produce the Cholesky factor $M$ of $\Sigma_{\bf Z}$ so that $M M^{\prime}=\Sigma_{\bf Z}$.
4. Set $X$ by $X \gets MZ$.
5. $\text{Return} \; Y \; \text{where} \; Y_i \gets F_{Y_i}^{-1}[\Phi(X_i)], \; i=1,2,...,d$.

- We proposed a GPU-accerelated NORTA algorithm for ultra-high dimensional
multivariate data [@Li2019gpu], presented at ESC0 2018.  
- NORTA is an example of a **copula**-based algorithm. It uses the Gaussian copula [@Nelsen2007].
-  Simulating correlated discrete and count data takes more care [@Xia17],
   including slightly redefining the inverse.

## Other `R` packages for simulating random vectors

There are a few R packages for multivariate simulation.

* `copula` Highly flexible copula specification [@Yan2007]
* `nortaRA` Implements exact Pearson matching (step 2 in NORTA) [@Chen2001] 
* `Genord` Simulated correlated discrete variables [@BF17]
* `mvnfast` High-performance multivariate normal simulator [@Fasiolo2016]

# 2 Simulation algorithms

## Algorithm 1: Matching Pearson approximately

We aim to simulate multivariate data with a Pearson correlation
matrix ${\bf R}$.

(i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R})$;  
(ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$;  
(iii) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;

* Where $F_i^{-1}$ is the inverse CDF for the $i^{th}$ gene's marginal
distribution.  
* And ${\bf R}$ is the target Pearson correlation matrix.  
* Pearson's correlation is **not** invariant under the montonically increasing
  transformations in steps 3, 4.  
* So the resulting simulated random vectors will have the **exact** margins but **approximate** ${\bf R}$.

## Algorithm 2: Matching Spearman's exactly

(i) Convert ${\bf R}_{Spearman}$ into ${\bf R}_{Pearson}$ via $\rho_{Pearson} =
  2 \times sin \left( \rho_{Spearman} \times \frac{\pi}{6} \right)$;  
(ii) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf} R_{Pearson})$;  
(iii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$;  
(iv) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;

- Spearman's $\rho$ relates to the difference in the ranks of pairs of values.
- We take advantage of a closed form correspondence between Pearson and Spearman
  for *normal random variables* in step 1 [@BF17].
- Since Spearman's $\rho$ is invariant under the monotonically increasing
  transformations in steps 3 and 4, the resulting simulated random vectors will
  have the **exact** margins and **exact** Spearman correlation (up to Monte Carlo/numeric error).

## Algorithm 3: Matching Kendall's exactly

(i) Convert ${\bf R}_{Kendall}$ into ${\bf R}_{Pearson}$ via $\rho_{Pearson} = sin \left( \tau \times \frac{\pi}{2} \right)$;  
(ii) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R}_{Pearson})$;  
(iii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$;  
(iv) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;

- Kendall's $\tau$ is the difference in the probabilities of concordant and discordant pairs.
- We use the 1-1 relationship between Pearson and Kendall's $\tau$ for *normal
  rvs* in step 1.
- Since Kendall's $\rho$ is invariant under the montonically increasing
  transformations in steps 3 and 4,  
 the resulting simulated random vectors will have the **exact** margins and
 **exact** Kendall's $\tau$  (up to Monte Carlo/numeric error).

## Algorithm notes

- Discrete margins pose some difficulties and matching *unadjusted* correlation
  measures is not exact (empirically approximate).
- Exact Pearson solutions are possible but require computing $d \choose 2$
   double integrals $EY_iY_j = \int \int y_i y_j f_{X|r}(F_i^{-1}(\Phi(z_i)),
   F_j^{-1}(\Phi(z_j))dy_idy_j$. We have yet to accelerate this process.
- We GPU accelerate whenever possible during our proposed High-Performance NORTA
  algorithm.
- One thorny problem when dealing with high-dimensional correlation matrices is
they can become non-positive definite through either during estimation or
bivariate transformations.
- When converting nonparametric correlations to Pearson (first step above) the
  resultant correlation may not be positive definite (PD).
- In that case, we replace ${\bf R}_{Pearson}$ with a "close" PD matrix ${\bf \tilde{R}}_{Pearson}$.
- In practice, this loss in accuracy typically has little impact on performance,
  but the algorithm needs acceleration (`Matrix::nearPD()` violates property S1).

# 3 The `bigsimr` R package

## Installation and setup

- `bigsimr` is on `Github`: [https://github.com/adknudson/bigsimr](https://github.com/adknudson/bigsimr).
- Use `devtools` to install.
- `bigsimr` uses `JAX` python libraries for GPU acceleration.

```{r setup, eval=TRUE}
## devtools::install_github( "adknudson/bigsimr")
## A conda environment is also activitied (see README)
library(bigsimr)
```

## Simulating random vectors through `bigsimr::rvec()`

- `bigsimr`'s main function is `rvec()`. It generates random vectors with
  correlated components.
- Let's walkthrough the basic usage in a simple example.

### 1. Specify the marginals

- Create a list of lists to specify the marginal distributions
- For example, the code below specifies a 3-component mixed type distribution.

```{r}
margins = list(
  list("norm", mean = 3.14, sd = 0.1),
  list("beta", shape1 = 1, shape2 = 4),
  list("nbinom", size = 10, prob = 0.75)
)
```

- Note: we use the standard `stats` package parameterizations and labels.

---

### 2. Specify correlation structure

* Next specify the correlation structure for the desired multivariate
distribution. 
* Provide a valid correlation matrix (positive definite, one's on the diag) .

```{r}
rho <- matrix(0.5, nrow = 3, ncol = 3)
diag(rho) <- 1.0
rho
```

* Complete the correlation struction by providing the type

```{r}
corType <- "spearman"
```

---

### 3. Generate N random vectors

Let's simulate a 10 random vectors for illustration.

```{r}
x <- rvec(10, rho = rho, params = margins, type = corType)
```

If your setup has no dedicated GPU, you receive the following warning message once per session.

```{r, echo=FALSE}
warning("warning.warn('No GPU/TPU found, falling back to CPU.')")
```

----

Taking a look at our random vector, we see that it has 10 rows and 3 columns, one column for each marginal.

```{r}
x
```

---

### Scaling N to 10,000

- For accuracy long-run frequency assessment, we generate more random vectors.
- Let's simulate $N=10,000$ and check the visual appearance of the simulated values.

```{r}
x <- rvec(n = 10000, rho = rho, params = margins, type = corType)
```

```{r, echo = FALSE, fig.align='center', fig.width=4, fig.height = 4 }
GGally::ggpairs(data = as.data.frame(x) )
## par(mfrow=c(1,3))
## hist(x[,1], breaks = 30, xlab = "", main = "Normal")
## hist(x[,2], breaks = 30, xlab = "", main = "Beta")
## hist(x[,3], breaks = 30, xlab = "", main = "Negative Binomial")
```

---

```{r}
## true rho
rho
## estimated rho
bigsimr::fastCor(x, method = corType)
```

# Monte carlo experiments

## Bivariate Normal

* Let's simulate a bivariate normal and check our correlation matching
  performance as N increases.
* Here we have BVN( $\mu_1 = \mu_2 = 10, \rho_{type}$ )
* We vary $\rho$ across the entire possible range of correlations for each
  correlation type.

```{r biNormal, echo = FALSE, eval = FALSE}
mom_norm <- function(x) {
  m <- mean(x)
  s <- sd(x)
  list(mean = m, sd = s)
}

margins <- list(
   list("norm", mean = mu, sd = sigma),
   list("norm", mean = mu, sd = sigma)
)

mu <- 10
sigma <- 1
margins <- list(
    list("norm", mean = mu, sd = sigma),
    list("norm", mean = mu, sd = sigma)
)

type <- c("pearson", "spearman", "kendall")
cores <- c(1)
n <- c(1e3, 1e4, 1e5)
adjustForDiscrete <- c(FALSE)

eps <- 1e-2
grid_steps <- 100
sim_pars <- expand.grid(type = type, cores = cores, n = n,
                        stringsAsFactors = FALSE,
                        adjustForDiscrete = adjustForDiscrete)

res <- data.frame()
for (i in 1:nrow(sim_pars)) {

  type <- sim_pars$type[i]
  cores <- sim_pars$cores[i]
  n <- sim_pars$n[i]
  adjustForDiscrete <- sim_pars$adjustForDiscrete[i]

  tmp_bounds <- computeCorBounds(margins, type = type)
  cor_lo <- tmp_bounds$lower[1,2] + eps
  cor_hi <- tmp_bounds$upper[1,2] - eps
  cor_seq <- seq(cor_lo, cor_hi, length.out = grid_steps)

  for (rho in cor_seq) {
    Rho <- matrix(rho, 2, 2)
    diag(Rho) <- 1.0

    Rho <- convertCor( rho = Rho, from = type, to = 'pearson' )
    time_data <- system.time({
      x <- rmvn(n = n, mu = rep( mu, 2 ), sigma = Rho)
    })

    # Save the sims in case
    id <- paste0(
      "d", 2,
      "-N", n,
      "-c", cores,
      "-r", rho,
      "-Cor", type,
      "-adj", as.character(adjustForDiscrete),
      "-dev", "1CORE",
      "-lib", "bigsimr"
    )
    saveRDS(x, file = paste0("sims_norm/", id, ".rds"))

    # Estimate statistics
    Rho_hat <- fastCor(x, method = type)
    rho_hat <- Rho_hat[1, 2]
    norm_args_hat <- mom_norm(x[,1])
    mu_hat <- norm_args_hat$mean
    sigma_hat <- norm_args_hat$sd

    # Save the results
    res <- rbind(res, data.frame(
      method = "bigsimr",
      device = "1CORE",
      type = type,
      cores = cores,
      margins = "norm",
      adjustForDiscrete = adjustForDiscrete,
      d = 2,
      N = n,
      rho = rho,
      rho_hat = rho_hat,
      mean = mu,
      sd = sigma,
      mean_hat = mu_hat,
      sd_hat = sigma_hat,
      sim_time = unname(time_data["elapsed"])
    ))
  }
}
res$type <- factor(res$type, levels = c("pearson",  "spearman", "kendall" ) )
saveRDS(object = res, "norm_sims_ags.rds")
```

----

```{r biNormPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.height = 5, fig.width= 8, fig.align='center', fig.cap = "`bigsimr` recovers the Pearson specified correlations for MVN."}
dat <- readRDS("norm_sims_ags.rds")
dat %>%
    filter(cores == 1) %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    ## ggplot(aes(rho, rho_hat)) +
    geom_point() +
    geom_abline(slope = 1) +
    ## facet_wrap(~ + N) + theme_bw()
    facet_wrap(~ type + N) + theme_bw()

## ggsave('fig/plot-biNormPlot.pdf')
```

## Bivariate Gamma

* Similarly, let's check the performance for a non-symmetric continuous distribution: a
  standard (rate =1) bivariate gamma.
* Here we have a Bivariate Gamma with $shape_1 = shape_2 = 10, \rho_{type}$.
* We vary $\rho$ across the entire possible range of correlations for each
  correlation type.

```{r biGamma, echo = FALSE, eval = FALSE}
mom_gamma <- function(x) {
  m <- mean(x)
  s <- sd(x)
  list(shape = m^2 / s^2, rate = m / s^2)
}

shape <- 10
rate <- 1
margins <- list(
  list("gamma", shape = shape, rate = rate),
  list("gamma", shape = shape, rate = rate)
)

type <- c("pearson", "spearman", "kendall")
cores <- c(1)
n <- c(1e3, 1e4, 1e5)
adjustForDiscrete <- c(FALSE)

eps <- 1e-2
grid_steps <- 100
sim_pars <- expand.grid(type = type, cores = cores, n = n,
                        stringsAsFactors = FALSE,
                        adjustForDiscrete = adjustForDiscrete)

res <- data.frame()
for (i in 1:nrow(sim_pars)) {

  type <- sim_pars$type[i]
  cores <- sim_pars$cores[i]
  n <- sim_pars$n[i]
  adjustForDiscrete <- sim_pars$adjustForDiscrete[i]

  tmp_bounds <- computeCorBounds(margins, type = type)
  cor_lo <- tmp_bounds$lower[1,2] + eps
  cor_hi <- tmp_bounds$upper[1,2] - eps
  cor_seq <- seq(cor_lo, cor_hi, length.out = grid_steps)

  for (rho in cor_seq) {
    Rho <- matrix(rho, 2, 2)
    diag(Rho) <- 1.0

    time_data <- system.time({
      x <- rvec(n = n,
                rho = Rho,
                params = margins,
                cores = cores,
                type = type,
                adjustForDiscrete = adjustForDiscrete)
    })

    # Save the sims in case
    id <- paste0(
      "d", 2,
      "-N", n,
      "-c", cores,
      "-r", rho,
      "-Cor", type,
      "-adj", as.character(adjustForDiscrete),
      "-dev", "1CORE",
      "-lib", "bigsimr"
    )
    saveRDS(x, file = paste0("sims_gamma/", id, ".rds"))

    # Estimate statistics
    Rho_hat <- fastCor(x, method = type)
    rho_hat <- Rho_hat[1, 2]
    gamma_args_hat <- mom_gamma(x[,1])
    shape_hat <- gamma_args_hat$shape
    rate_hat <- gamma_args_hat$rate

    # Save the results
    res <- rbind(res, data.frame(
      method = "bigsimr",
      device = "1CORE",
      type = type,
      cores = cores,
      margins = "gamma",
      adjustForDiscrete = adjustForDiscrete,
      d = 2,
      N = n,
      rho = rho,
      rho_hat = rho_hat,
      shape = shape,
      rate = rate,
      shape_hat = shape_hat,
      rate_hat = rate_hat,
      sim_time = unname(time_data["elapsed"])
    ))
  }
}
res$type <- factor(res$type, levels = c("pearson",  "spearman", "kendall" ) )
saveRDS(object = res, "gamma_sims_ags.rds")
```

---

```{r biGammaPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.width= 8, fig.height = 5,fig.align='center', fig.cap = "`bigsimr` recovers correlations for bivariate gamma only approximately for Pearson but exactly for the rank-based correlations."}
dat <- readRDS("gamma_sims_ags.rds")
dat %>%
    filter(cores == 1, type == 'pearson') %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    geom_point() +
    geom_abline(slope = 1) +
    ## facet_wrap(~ type + N)
    facet_wrap(~ type + N)
```

---

```{r biGammaPlot2, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.width= 8, fig.height = 5,fig.align='center', fig.cap = "`bigsimr` recovers the correlations for bivariate gamma only approximately for Pearson but exactly for the rank-based correlations."}
dat <- readRDS("gamma_sims_ags.rds")
dat %>%
    filter(cores == 1, type == 'spearman') %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    geom_point() +
    geom_abline(slope = 1) +
    ## facet_wrap(~ type + N)
    facet_wrap(~ type + N)
```

---

```{r biGammaPlot3, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.width= 8, fig.height = 5,fig.align='center', fig.cap = "`bigsimr` recovers the correlations for bivariate gamma only approximately for Pearson but exactly for the rank-based correlations."}
dat <- readRDS("gamma_sims_ags.rds")
dat %>%
    filter(cores == 1, type == 'kendall') %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    geom_point() +
    geom_abline(slope = 1) +
    ## facet_wrap(~ type + N)
    facet_wrap(~ type + N)
```

## Bivariate Negative Binomial

* Let's check the performance for a discrete distribution: a bivariate negative binomial
* Here we have Bivariate Negative Binomial ( $prob_1 = prob_2 = 0.5, size_1 = size_2 = 4,\rho_{type}$ )
* We vary $\rho$ across the entire possible range of correlations for each
  correlation type.

```{r biNegBin, echo = FALSE, eval = FALSE}
mom_nbinom <- function(x) {
  m <- mean(x)
  s <- sd(x)
  list(size = m^2 / (s^2 - m), prob = m / s^2)
}

size <- 4
prob <- 0.5
margins <- list(
  list("nbinom", size = size, prob = prob),
  list("nbinom", size = size, prob = prob)
)

type <- c("pearson", "spearman", "kendall")
cores <- c(1)
n <- c(1e3, 1e4, 1e5)
adjustForDiscrete <- c(FALSE)

eps <- 1e-2
grid_steps <- 100
sim_pars <- expand.grid(type = type, cores = cores, n = n,
                        stringsAsFactors = FALSE,
                        adjustForDiscrete = adjustForDiscrete)

res <- data.frame()
for (i in 1:nrow(sim_pars)) {

  type <- sim_pars$type[i]
  cores <- sim_pars$cores[i]
  n <- sim_pars$n[i]
  adjustForDiscrete <- sim_pars$adjustForDiscrete[i]

  tmp_bounds <- computeCorBounds(margins, type = type)
  cor_lo <- tmp_bounds$lower[1,2] + eps
  cor_hi <- tmp_bounds$upper[1,2] - eps
  cor_seq <- seq(cor_lo, cor_hi, length.out = grid_steps)

  for (rho in cor_seq) {
    Rho <- matrix(rho, 2, 2)
    diag(Rho) <- 1.0

    time_data <- system.time({
      x <- rvec(n = n,
                rho = Rho,
                params = margins,
                cores = cores,
                type = type,
                adjustForDiscrete = adjustForDiscrete)
    })

    # Save the sims in case
    id <- paste0(
      "d", 2,
      "-N", n,
      "-c", cores,
      "-r", rho,
      "-Cor", type,
      "-adj", as.character(adjustForDiscrete),
      "-dev", "1CORE",
      "-lib", "bigsimr"
    )
    saveRDS(x, file = paste0("sims_nb/", id, ".rds"))

    # Estimate statistics
    Rho_hat <- fastCor(x, method = type)
    rho_hat <- Rho_hat[1, 2]
    nbinom_args_hat <- mom_nbinom(x[,1])
    size_hat <- nbinom_args_hat$size
    prob_hat <- nbinom_args_hat$prob

    # Save the results
    res <- rbind(res, data.frame(
      method = "bigsimr",
      device = "1CORE",
      type = type,
      cores = cores,
      margins = "nbinom",
      adjustForDiscrete = adjustForDiscrete,
      d = 2,
      N = n,
      rho = rho,
      rho_hat = rho_hat,
      size = size,
      prob = prob,
      size_hat = size_hat,
      prob_hat = prob_hat,
      sim_time = unname(time_data["elapsed"])
    ))
  }
}
res$type <- factor(res$type, levels = c("pearson",  "spearman", "kendall" ) )
saveRDS(object = res, "nbinom_sims_ags.rds")
```

---

```{r  biNegBinPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.align='center', fig.width = 8, fig.height=5,fig.cap = "`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
dat <- readRDS("nbinom_sims_ags.rds")
dat %>%
    filter(cores == 1, type == 'pearson') %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    geom_point() +
    geom_abline(slope = 1) +
    ## facet_wrap(~ type + N)
    facet_wrap(~ type + N)
```

---

```{r  biNegBinPlot2, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.align='center', fig.width = 8, fig.height=5,fig.cap = "`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
dat <- readRDS("nbinom_sims_ags.rds")
dat %>%
    filter(cores == 1, type == 'spearman') %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    geom_point() +
    geom_abline(slope = 1) +
    ## facet_wrap(~ type + N)
    facet_wrap(~ type + N)
```

---

```{r  biNegBinPlot3, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.align='center', fig.width = 8, fig.height=5,fig.cap = "`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
dat <- readRDS("nbinom_sims_ags.rds")
dat %>%
    filter(cores == 1, type == 'kendall') %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    geom_point() +
    geom_abline(slope = 1) +
    ## facet_wrap(~ type + N)
    facet_wrap(~ type + N)
```

# Scale up to Ultra-High dimensions in our motivating application

## Simulating RNA-seq data

- We have RNA-sequencing data for N=1212 breast cancer patients (TCGA BRCA).  
- There are 20,501 genes that have count data quantifying the expression of the
  each gene's mRNA.  
- We simulate N random vectors to simulate a *fake dataset* with heterogeneous
  negative binomial margins and correlations estimated from real data.

## Scale up to Ultra-High Dimensions

```{r, echo=F, out.width='80%', fig.align='center', fig.cap='Computation times as d increases. We filter to the top 1, 5, 10, 15, 20, 25% expressing genes (in terms of median expression'}
knitr::include_graphics("fig/cpu-gpu-times.png")
```

## Simulate RNA-seq data example with d=206

```{r processBRCA, echo=TRUE, eval=TRUE}
## full workflow simulating RNA-seq data
allDat <- readRDS( file = "~/Downloads/complete_processed_tcga2stat_RNASeq2_with_clinical.rds" )
lastClinical <- which( names(allDat) == 'tumorsize' )
brca <- allDat[ allDat$disease == "BRCA", (lastClinical+1):ncol(allDat) ]
## remove naively the RSEM adjustment
brca <- round(brca, 0)
dim(brca) ## num of genes 20501
## compute the median expression! avoid the 0s
brcaMedian <- apply(brca, 2, median)
## retain top (1-probs)*100% highest expressing genes for illustration
myProb <- 0.99
cutPoint <- quantile( x = brcaMedian, probs = myProb )
## genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
brca <- brca[ , genesToKeep ]
## ncol(brca) / 20501
print(d <- length(genesToKeep))
```

---

```{r figNBratio, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='The data are overdispersed (all values greater than 0)'}

logVarOverMean <- function( x ) {
    log ( var(x) / mean (x) )
}
## check whether NB makes sense

summaryBRCA <- brca %>% 
    summarize_all(  list(~ logVarOverMean( . ) ) )
summaryBRCA <- as.data.frame( t(as.data.frame( summaryBRCA )) )
names(summaryBRCA) <- "logVarOverMean"
## sum(summaryBRCA$logVarOverMean < 0)  ## no underdispersed genes
ggplot(data = summaryBRCA, mapping = aes(x = logVarOverMean)) +
  geom_histogram(color = "white", bins = 20)
```

---

### Let's walk through a workflow simulating RNA-seq data

```{r estRhoBRCA, echo=TRUE, eval=TRUE}
## 1. Estimate Spearman's correlation on the count data
corType <- 'spearman'
system.time( rho <- bigsimr::fastCor( brca, method = corType ) )
```

---

```{r estMargins, echo = TRUE, eval = TRUE}
## 2. Estimate NegBin parameters using Method of Moments
estimateNegBinMoM <- function(tmpGene, minP = 1e-7) {
    tmpMean <- mean(tmpGene)
    tmpVar <- var(tmpGene)
    ## relate to nbinom parameters
    ## See ?rbinom for details.
    p <- tmpMean / tmpVar
    if (p < minP) {p <- minP } ## maybe add noise here
    n <- ( tmpMean * p) / ( 1  - p )  
    ## format for bigsimr margins
    return( list("nbinom", size = n, prob = p) )
}
brcaMargins <- apply( unname(as.matrix(brca)), 2, estimateNegBinMoM )
## describe the margins
## check for too small prob
## head( sort( unlist( lapply( brcaMargins, function(x) {x$prob} ) ) ) )
## head( sort( unlist( lapply( brcaMargins, function(x) {x$size} ) ) ) )
```

----

```{r simBRCA, echo=TRUE, eval=TRUE}
## 3. Generate the simulated samples
N <- nrow(brca)
system.time( simBRCA <- rvec(N, rho = rho, params = brcaMargins, cores = CORES, type = corType) )
simBRCA[1:2, 1:2 ]
```

---

```{r evalBRCA2, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Real data for the 1st 3rd genes'}
## Describe and plot real data
GGally::ggpairs(data = as.data.frame(brca[ ,1:3] ) )
```

---

```{r evalBRCA3, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Simulated data for the 1st 3rd genes'}
## Describe and plot simulations
GGally::ggpairs(data = as.data.frame(simBRCA[ ,1:3] ) )
```

---

```{r evalBRCA4, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Marginal mean matching'}
## check 1st moments
trueMu <- colMeans(brca)
simMu <- colMeans(simBRCA)
qplot( x = trueMu, y = simMu ) + geom_abline(slope = 1, intercept = 0)

## check 2nd moment
## trueVar <- apply( brca, 2, var)
## simVar <- apply( simBRCA, 2, var)
## qplot( x = trueVar, y = simVar ) + geom_abline(slope = 1, intercept = 0)
```

---

```{r evalBRCA5, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width=8, fig.align='center', fig.cap="Spearman's correlation matching in RNA-seq simulations"}
## check correlation
simRho <- bigsimr::fastCor( simBRCA, method = corType )
## rho[1:5, 1:5]
## simRho[1:5, 1:5]
trueRho <- rho[lower.tri(rho)]
simRho <- simRho[lower.tri(simRho)]
qplot( x = trueRho, y = simRho ) + geom_abline(slope = 1, intercept = 0)
```

# Conclusions

## Concluding remarks

* Our scaleable NORmal To Anything provides a flexible way to simulate
  ultra high-dimensional *fake datasets*.

### Pros
* An easy-to-use tool to produce multivariate data of dimension larger than any
  other existing tool.
* Flexible modeling providing highly hetereogenous specification of joint probability models.
* The R `bigsimr`package is a high-performance and GPU-accelerated implementation
  of algorithm and other fast correlation utilities.

### Cons
* Our method requires a well-defined inverse CDF.
* Discrete margins can slightly disrupt the matching algorithm (due the probability of ties). 
* Not all algorithms are currently optimized.

## Future Work

* Investigate *rescaling* the correlations to adjust for discrete margins.
* GPU accelerate the `normal2marginal()`.
* Better handling of non-positive definiteness (fast checks for PD and faster
  implementation/algorithms of nearest PD).
* High-performance implementations for exact Pearson matching.
* Match *partial correlations*.
* Implement user-defined inverse CDFs.
* Implement high-dimensional covariance estimation algorithms.
* Take advantage of sparsity.
* Explore different simulation algorithms.

## List of supported distributions

```{r, echo=TRUE, eval=FALSE}
all_dists <- list(
  list(dist = "beta", shape1, shape2),
  list(dist = "binom", size, prob),
  list(dist = "cauchy", location, scale),
  list(dist = "chisq", df),
  list(dist = "exp", rate),
  list(dist = "f", df1, df2),
  list(dist = "gamma", shape, rate),
  list(dist = "geom", prob),
  list(dist = "hyper", m, n, k),
  list(dist = "logis", location, scale),
  list(dist = "lnorm", meanlog, sdlog),
  list(dist = "nbinom", size, prob),
  list(dist = "norm", mean, sd),
  list(dist = "pois", lambda),
  list(dist = "t", df),
  list(dist = "unif", min, max),
  list(dist = "weibull", shape, scale),
  list(dist = "wilcox", m, n),
  list(dist = "signrank", n)
)
```

<style>
slides > slide { overflow: scroll; }
slides > slide:not(.nobackground):before {
  background: none;
}
</style>

## References

### 2020-10-15 050-some-applications.Rmd
Revise rna-seq example to be less complicated

@@
 
-```{r LoadLib050, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, results = "hide"}
-library(ggplot2)
+```{r LoadLib050, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
+## devtools::install_github( "adknudson/bigsimr", ref = 'develop')
+## knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
+reticulate::use_condaenv("bigsimr")
+library(bigsimr)
 library(tidyverse)
-library(knitr)
-library(dplyr)
-## devtools::install_github( "adknudson/bigsimr", ref = 'develop')
-library(bigsimr)
-reticulate::use_condaenv("bigsimr")
-CORES <- parallel::detectCores() - 1
-set.seed(06082020)
-## devtools::install_github("ggobi/GGally")
-## devtools::install_github( "adknudson/bigsimr")
-## A conda environment is also activitied (see README)
-library(bigsimr)
-library(GGally)
+CORES <- as.integer( parallel::detectCores() - 1 )
+set.seed(10142020)
 ```
 
-## Simulating Ultra High Dimensional RNA-seq data
+## Simulating High-Dimensional RNA-seq data
 
-We apply our methodology to simulate RNA-sequencing data sets based on samples derived from breast cancer patients' tumors (the BRCA data set in TCGA).
-The data are freely available as part of the BRCA data set within the The Cancer Genome Atlas data warehouse and were downloaded on BLAH using the Harvard's Broad Institute's Firehouse interface.
-For each of the 1093 patients, the gene expression (abundance of messenger-RNA) from 20501 genes were counted using RNA-sequencing.
-The resultant data set can be represented as a matrix of size 1093 x 20501 with discrete values aligned to HUGO gene symbols for each patient.
-In this illustrative case study, we model the gene-wise marginal distributions as heterogeneous negative binomial with pmf given by Equation \@ref(eq:nb) with two parameters, $p$ and $r$ (OR SWITCH TO mu and delta).
-Modeling RNA-seq counts as negative binomially distributed random variables is commonplace (e.g., @Zhao2018) since overdispersion is often observed for most expressed genes.
-We begin by estimating the marginal negative binomial parameters for each group using a simple method of moments approach by matching the negative binomial parameters with the sample mean and variance.
-Table/Figure XXX summarizes the estimated means, variances, and NB parameters as described in Equation \@ref(eq:nb).
-Notably some genes show underdispersion compared to a Poisson random variable.
-For simplicity we exclude these cases, but nothing in the later described simulation method prevents using a different marginal distribution in those cases (such as a Maxwell distribution REF Sellers).
+This section demonstrates how to simulate high-dimensional multivariate data using `bigsimr`, aiming to replicate the structure of high dimensional dependent count data.
+In fact, simulating RNA-sequencing (RNA-seq) data is a one of the primary motivating applications of the proposed methodology, seeking scaleable Monte Carlo methods for realistic multivariate simulation (for example, see Schissler, Piegorsch, and Lussier 2018).
 
-Intergene correlation [@Schissler2019].
+The RNA-seq data generating process involves counting how often a particular messenger RNA (mRNA) is expressed in a biological sample.
+Since this is a counting processing with no upper bound, many modeling approaches discrete random variables with infinite support.
+Often the counts exhibit over-dispersion and so the negative binomial arises as a sensible model for the expression levels (\emph{gene counts}).
+Moreover, the counts are correlated (co-expressed) and cannot be assumed to behave independently.
+RNA-seq platforms quantify the entire transcriptome in one experimental run, resulting in high dimensional data.
+In humans, this results in count data corresponding to over 20,000 genes (coding genomic regions) or even over 77,000 isoforms when alternating spliced mRNA are counted.
+	This suggests simulating high-dimensional multivariate NB with heterogeneous marginals would be useful tool in the development and evaluation of RNA-seq analytics.
 
-\begin{equation} 
-  g(n) = \mathbb P(Y=n) = \frac{\Gamma(n+r)}{\Gamma(r)n!} p^r(1-p)^n, \,\,\, n\in
-\mathbb N_0.
-  (\#eq:nb)
-\end{equation}
+In an illustration of our proposed methodology applied to real data, we seek to simulate RNA-sequencing data by producing simulated random vectors with assumed marginal distributions with estimated parameters.
+Our goal is to replicate the structure of a breast cancer data set (BRCA data set from The Cancer Genome Atlas).
+For simplicity of illustration, we begin by filtering to retain the top 5\% highest expressing genes of the 20,501 gene measurements from $N=1212$ patients' tumor samples, resulting in $d=1026$ genes.
 
-```{r processBRCA, echo=FALSE, eval=FALSE, cache=TRUE}
+<!--
+All these genes exhibit over-dispersion and, so, we proceed to estimate the NB parameters $(r_i, p_i), i=1,\ldots,d$ to determine the target marginal PMFs $g_i(y_i)$ (via method of moments).
+Notably, the $\hat{p}_i's$ are small --- ranging in $[3.934 \times 10^{-6} , 1.217 \times 10^{-2}]$.
+To complete the simulation algorithm inputs, we estimate the Pearson correlation matrix $\bf{R_Y}$ and set that as the target correlation.
+ -->
+
+<!--
+With the simulation targets specified, we proceed to simulate $N=10,000$ random vectors $\bf{Y}$ $=( Y_1,\ldots,Y_d)$ with target Pearson correlation $\bf{R_Y}$ and marginal PMFs $g_i(y_i)$ using a $\bf{T}$-Poisson hierarchy of Kind II.
+Specifically, we first employ the \emph{direct Gaussian copula} approach to generate $N$ random vectors following a standard multivariate Gamma distribution $\bf{T}$ with shape parameters $r_i$ equal to the target $n_i$ and Pearson correlation matrix $\bf{R_T}$.
+Care must be taken when setting the specifying $\bf{R}$ (refer to Equation \ref{gay.cop.pdf}) --- we employ Equation \ref{mix.poi.corr} to compute the scaling factors $c_{i,j}$ and adjust the underlying correlations to ultimately match the target $\bf{R_Y}$.
+Notably, of the $525,825$ pairwise correlations from the $1026$ genes, no scale factor was less than $0.9907$, indicating the model can produce essentially the entire range of possible correlations.
+Here we are satisfied with approximate matching of the specified Gamma correlation and set $\bf{R}$ = $\bf{R_T}$ in our Gaussian copula scheme ($\bf{R}$ indicating the specified multivariate Gaussian correlation matrix).
+Finally, we generate the desired random vector $Y_i=N_i(t_i)$ by simulating Poisson counts with expected value $\mu_i=\lambda_i \times T_i$, for $i=1,\ldots,d$ (with $\lambda_i=\frac{(1-p_i)}{p_i}$) and repeat $N=10,000$ times.
+-->
+
+```{r processBRCA, echo=FALSE, eval=FALSE, results='none'}
 ## full workflow simulating RNA-seq data
@@
 brca <- allDat[ allDat$disease == "BRCA", (lastClinical+1):ncol(allDat) ]
-## remove naively the RSEM adjustment
-brca <- round(brca, 0)
-ncol(brca) ## num of genes 20501
+dim(brca) ## num of genes 20501
 ## compute the median expression! avoid the 0s
@@
 ## retain top (1-probs)*100% highest expressing genes for illustration
-myProb <- 0.99
+myProb <- 0.75
 cutPoint <- quantile( x = brcaMedian, probs = myProb )
@@
 ## ncol(brca) / 20501
-(d <- length(genesToKeep))
+print(d <- length(genesToKeep))
+## save data as rds
+saveRDS(brca, 'data/brca.rds')
 ```
 
-The data are overdispersed:
-
-```{r figNBratio, echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE, out.width= '75%', fig.align='center', fig.cap='The genes display extreme overdispersion.
-Using group-specific sample means and variances for each of the d=`r d` genes, we display the distributions oflog of variance to mean ratio.'}
-
-logVarOverMean <- function( x ) {
-    log ( var(x) / mean (x) )
-}
-## check whether NB makes sense
-
-summaryBRCA <- brca %>% 
-    summarize_all(  list(~ logVarOverMean( . ) ) )
-summaryBRCA <- as.data.frame( t(as.data.frame( summaryBRCA )) )
-names(summaryBRCA) <- "logVarOverMean"
-## sum(summaryBRCA$logVarOverMean < 0)  ## no underdispersed genes
-ggplot(data = summaryBRCA, mapping = aes(x = logVarOverMean)) +
-  geom_histogram(color = "white", bins = 20)
+```{r readBRCA, echo=TRUE, eval=TRUE}
+## full workflow simulating RNA-seq data
+brca <- readRDS(file = 'data/brca.rds') 
 ```
 
-We see that the all genes studies display overdispersion (variance greater than mean) compared to a Poisson model for the gene counts.
-The genes are highly heterogeneous even when dealing with highly expressed genes after filtering.
-The facts taken together motivate the negative binomial model for the marginal distributions.
-The genes are slightly more variable within the deceased group (expected with smaller sample size and potentially due to biomedical considerations).
+Let's walk through a workflow simulating RNA-seq data
 
-Next we perform some essential pre-processing of the data.
-Since our goal is to simulate meaningful multivariate constructions, we must first find a subset of 20501 genes that are expressed.
-In order words, we'll filter out the low expressing genes to allow greater range of possible $d$-variate correlations (see @NK10 for details OR SHOULD I EXPLAIN THIS MORE CAREFULLY?).
-Notably, many RNA-seq analytic workflows filter out low expressing genes (see \cite{Conesa2016b}) and this, admittedly, is usually done in an \emph {ad hoc} fashion.
-In this study, we prefer an inclusion criterion that is theoretically motivated.
-Based on the results in Section BLAH, we filter genes with a sample mean less than 12.
-In this way, we remove 4762 (23.2\%) of the genes, retaining 15,739 genes for further analysis.
-The filtered low-expressing genes will be simulated as independent negative binomial random variables to form a complete simulated transcriptome.
-
-RNA-sequencing data derived from breast cancer patients (n=1093), with 20501 genes measured (from TCGA).
-We filter to genes with average expression greater than 10000 counts (d = 4777) across both groups.
-Consider basic differentially expressed genes (DEG) analysis between surviving (n0 = 941) and deceased (n1 = 152) patients.
-This setting can be described as large-scale simultaneous hypothesis testing, under correlation.
-We should evaluate existing and new methodology using simulations that reflect dependency among variables.
-But multivariate simulation tools often scale poorly to high dimension or do not model the full range of dependency.
-Perform a two-sample $t$ test for each 4777 genes on the observed RNA-seq counts ($empirical$).
-Call a "Z score" for the $i^{th}$ gene $Z_{i} = \Phi^{-1}(t_{\approx 272}(t_i))$.
-
-Our method requires some measure of that the covariance structure is pre-specified.
-In practice, however, high-dimensional covariance estimation is not that easy.
-Indeed there is much recent interest in this area (cite a few recent high visibility articles --- see the GPU-NORTA pub).
-Here we do not provide a comprehensive review on the topic.
-Instead we only seek to use covariance estimator that only guarantees positive-semidefiniteness while maintaining adequate estimation properties.
-To this end we chose to use the condition-number-regularized (\textsc{condreg}) approach of \cite{Won2013g}.
-This covariance estimation procedure restricts the ratio of the largest eigenvalue to the smallest eigenvalue to improve numerical results.
-But the method employs a penalized Gaussian likelihood.
-And so, rather than estimating the covariance on the discrete counts directly, we first perform a $log_{2}(x +1)$ transformation to better suit this modeling assumption.
-\textsc{condreg} requires a parameter, $\kappa_{max}$, that controls the largest condition number of the resultant estimated covariance matrix.
-Here we aim to allow a large number of nonzero pairwise correlations and so arbitrary set $\kappa_{max}=10^{4}$.
-The routine ran without issue on a MacBook Pro carrying BLAH in XXX units of time.
-
-Most pairwise correlations are nearly zero with XXX \% less than |0.1|.
-Yet there are dense subsets of correlated genes.
-And even small correlations among many variables can disrupt the operating characteristics of commonly used statistical inferential procedures \cite{Schissler2018}.
-
-In this case study in the large-scale hypothesis testing in pursuit of detecting differentially expressed genes, we aim to simulate RNA-seq counts using a multivariate negative binomial.
-To evaluate the simulation's utility, we systematically compare the simulated results to the empirical results.
-In short we computed t-statistics and p-values for each of $d=4777$ genes between two groups of breast cancer patients (1=deceased, 0=alive; $n_0 = 152, n_1=941$).
-Further details on the empirical results are provided in Section BLAH.
-We compare two aspects in the simulation design 1) Choice of correlation measure while modeling marginal distributions consistently and 2) how well does the best-performing simulation agree with the empirical results.
-Informed by the simulation studies in Section BLAH, we anticipate a far amount of variation at such small sample sizes and large $d$ and so replicate the entire study 100 times.
-
-To recap from the discussion above (Section BLAH), we estimated negative binomial parameters for each of the $d=4777$ genes and also their Pearson, Spearman, and Kendall's correlation coefficients from n=1093 breast cancer patients, within each vital status group (0=deceased, 1=alive).
-We generated 100 synthetic data sets for each group, with the number of $d$-dimensional random vectors produced equal to the corresponding sample sizes ($n_0 = 152, n_1=941$).
-These 100 samples are used to understand uncertainty in this setting and we will use the mean simulated values to compare to the empirical (and also explore the worst-case scenarios).
-
-Model genes as marginally negative binomial with heterogeneous gene-wise parameters.
-Compute sample correlations (using desired dependency measure) for the 11,407,476 genes pairs, for each group.
-To simulate gene expression counts, our goal is to produce a random vector ${\bf Y}=(Y_1, \ldots, Y_d)$ with **correlated** NB components.
-To do that, we start with a sequence of **independent** Poisson processes $N_i(t)$, $i=1,\ldots,d$, where the rate of the process $N_i(t)$ is $\lambda_i=(1-p_i)/p_i>0$ (so that $p_i=1/(1+\lambda_i)$).
-Now, we let ${\bf T} = (T_1, \ldots T_d)$ have a multivariate distribution on $\mathbb R_+^d$ with the PDF $f_{{\bf T}}({\bf t})$.
-Then, we define
-
-We rescaled the target Spearman correlation to account for the probability of ties via Equation \@ref(eq:convertSpearmanDiscrete).
-The serial computation took approximately 10 minutes on XXX.
-Notably, the difference between target and adjusted matrices was very small (mean relative difference: 0.708) in this particular configuration of negative binomial parameters.
-For problems with smaller counts and higher probabilities of ties, the adjustment can be substantial (as observed in the simulation studies above in Section BLAH).
-
-```{r estRhoBRCA, echo=FALSE, eval=FALSE}
-## full workflow simulating RNA-seq data
-
+```{r estRhoBRCA, echo=TRUE, eval=TRUE}
 ## 1. Estimate Spearman's correlation on the count data
-corType <- 'spearman'
-system.time( rho <- bigsimr::fastCor( brca, method = corType ) )
-## Describe the correlations
+corType <- 'pearson'
+system.time( nb_Rho <- bigsimr::cor_fast( brca, method = corType ) )
 ```
 
-```{r estMargins, echo=FALSE, eval=FALSE, cache=TRUE}
+---
+
+```{r estMargins, echo = FALSE, eval = TRUE}
 ## 2. Estimate NegBin parameters using Method of Moments
-estimateNegBinMoM <- function(tmpGene) {
+estimateNegBinMoM <- function(tmpGene, minP = 1e-7) {
     tmpMean <- mean(tmpGene)
@@
     p <- tmpMean / tmpVar
+    if (p < minP) {p <- minP } ## maybe add noise here
     n <- ( tmpMean * p) / ( 1  - p )  
     ## format for bigsimr margins
-    get( 'qnbinom')
-    return( str2lang( paste0( "qnbinom(size = ", n, ", prob = ", p, ")" ) ) )
+    return( list("nbinom", size = n, prob = p) )
 }
 brcaMargins <- apply( unname(as.matrix(brca)), 2, estimateNegBinMoM )
-## brcaMargins <- apply( unname(as.matrix(brca))[ , 1:3], 2, estimateNegBinMoM )
 ## describe the margins
-## check for too small prob
-## head( sort( unlist( lapply( brcaMargins, function(x) {x$prob} ) ) ) )
+probs <- unlist( lapply( brcaMargins, function(x) {x$prob} ) )
+## summary(probs)
+sizes <- unlist( lapply( brcaMargins, function(x) {x$size} ) )
+## summary(sizes)
 ## head( sort( unlist( lapply( brcaMargins, function(x) {x$size} ) ) ) )
@@
 
-```{r simBRCA, echo=FALSE, eval=FALSE, cache=TRUE}
-## 3. Generate the simulated samples
-N <- nrow(brca)
-system.time( simBRCA <- rvec(n = N, rho = rho, margins = brcaMargins, cores = CORES, type = corType) )
-simBRCA[1:2, 1:2 ]
-
+```{r nbHelpers}
+make_nbinom_alist <- function(sizes, probs) {
+  lapply(1:length(sizes), function(i) {
+    substitute(qnbinom(size = s, prob = p), 
+               list(s = sizes[i], p = probs[i]))
+  })
+}
+## make_nbinom_alist(c(20, 21, 22), c(0.3, 0.4, 0.5))
+nbinom_mom <- function(x) {
+  m <- mean(x)
+  s <- sd(x)
+  s2 <- s^2
+  p <- m/s2
+  r <- m^2 / (s2 - m)
+  c(r, p)
+}
 ```
 
-```{r evalBRCA, echo=FALSE, eval=FALSE, cache=TRUE}
+Now specify the desired multivariate negative binomial distribution.
 
-## Describe and plot real data
-GGally::ggpairs(data = as.data.frame(brca[ ,1:3] ) )
+```{r runBRCA, results='none'}
+n <- 20000
+## d <- 100
+## n <- 100
+## d <- 5
+nb_margins <- make_nbinom_alist(sizes, probs)
+## compute theoretical bounds to make sure input is okay
+## cor_bounds( nb_margins, type = type )
+sim_nbinom <- rvec(n, nb_Rho, nb_margins, type = corType, ensure_PSD = TRUE, cores = CORES)
+## sim_nbinom <- rvec(n, nb_Rho, nb_margins, type = "pearson", ensure_PSD = TRUE, cores = 4L)
+## pairs(sim_nbinom)
+nb_Rho_hat <- cor_fast(sim_nbinom, method = corType)
+cat('The quadratic loss from the true correlation matrix is ') 
+norm(nb_Rho - nb_Rho_hat, type = "2")
+```
 
-## Describe and plot simulations
-GGally::ggpairs(data = as.data.frame(simBRCA[ ,1:3] ) )
+Figure X shows the results of our simulation by comparing the specified target parameter (horizontal axes) with the corresponding quantities estimated from the simulated data (vertical axes).
+The evaluation shows that the simulated counts approximately match the target parameters and exhibit the full range of estimated correlation from the data.
+Utilizing 15 CPU threads in a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, the simulation completed in less than 30 seconds.
 
-## check 1st moments
-trueMu <- colMeans(brca)
-simMu <- colMeans(simBRCA)
-qplot( x = trueMu, y = simMu ) + geom_abline(slope = 1, intercept = 0)
-
-## check 2nd moment
-trueVar <- apply( brca, 2, var)
-simVar <- apply( simBRCA, 2, var)
-qplot( x = trueVar, y = simVar ) + geom_abline(slope = 1, intercept = 0)
-
-## check correlation
-system.time( simRho <- bigsimr::fastCor( simBRCA, method = corType ) )
-rho[1:5, 1:5]
-simRho[1:5, 1:5]
-trueRho <- rho[lower.tri(rho)]
-simRho <- simRho[lower.tri(simRho)]
-qplot( x = trueRho, y = simRho ) + geom_abline(slope = 1, intercept = 0)
+```{r fig:BRCA, echo=F, eval=T, caption='The T-Poisson strategy produces simulated random vectors from a multivariate negative binomial (NB) that replicate the estimated structure from an RNA-seq data set.
+The dashed red lines indicated equality between estimated parameters (vertical axes; derived from the simulated data) and the specified target parameters (horizontal axes).'}
+par(mfrow=c(1,3))
+plot(nb_Rho[lower.tri(nb_Rho)], nb_Rho_hat[lower.tri(nb_Rho_hat)],
+     xlim = c(-1, 1), ylim = c(-1, 1),
+     pch = 4, cex = 1, col = rgb(0, 0, 1, 0.05),
+     xlab = "Target Correlation", ylab = "Estimated Correlation",
+     main = "bigsimr Simulation")
+abline(a = 0, b = 1, col="red", lty="dashed")
+nbinom_size_prob_hat <- apply(sim_nbinom, 2, nbinom_mom)
+plot(sizes, nbinom_size_prob_hat[1,],
+     pch = 4, cex = 1, col = rgb(0, 0, 1, 1),
+     xlab = "Target Size", ylab = "Estimated Size",
+     main = "Direct NB Simulation")
+abline(a=0, b=1, col="red", lty="dashed")
+plot(log(probs), log(nbinom_size_prob_hat[2,]),
+     pch = 4, cex = 1, col = rgb(0, 0, 1, 1),
+     xlab = "Target Log Probability", ylab = "Estimated Log Probability",
+     main = "Direct NB Simulation")
+abline(a=0, b=1, col="red", lty="dashed")
+par(mfrow=c(1,1))
 ```

### 2020-10-16 010-background.Rmd
migrating ESCO 2020 talk into sections, revising everything

@@
 For continuous marginals, the measure works well due to zero probability of ties.
+
+<!-- 
 For discrete marginals, however, one could perform a rescaled version of $\rho_s$ for random variables $X,Y$ with pmfs (or pdfs) $p(x)$ and $q(y)$, respectively.
@@
 
+-->
+
 ## Marginal-dependent bivariate correlation bounds
@@
 For discrete random variables, define $F^{-1}$ as in Equation \@ref(eq:inverseCDF).
+
+## Other `R` packages for simulating random vectors
+
+There are a few R packages for multivariate simulation.
+
+* `copula` Highly flexible copula specification [@Yan2007]
+* `nortaRA` Implements exact Pearson matching (step 2 in NORTA) [@Chen2001] 
+* `Genord` Simulated correlated discrete variables [@BF17]
+* `mvnfast` High-performance multivariate normal simulator [@Fasiolo2016]

### 2020-10-16 020-simulation-algorithm.Rmd
migrating ESCO 2020 talk into sections, revising everything

--- 2020-06-15 020-simulation-algorithm.Rmd
@@
 # Simulation algorithm
+
+```{r previewChapter020, echo = FALSE, eval = FALSE}
+bookdown::preview_chapter('020-simulation-algorithm.Rmd')
+```
 
@@
 
-The method is best understand as a **parallelized Gaussian copula** (see Equation \@ref(eq:gauss).
+The method is best understand as a **parallelized Gaussian copula** (see Equation \@ref(eq:gauss) to provide a high-performance NORTA (NORmal To Anything) algorithm.
+
+To simulate a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$,
+there is the well-known NORTA algorithm [@Cario1997]. It follows like this:
+
+1. Simulate a random vector $\bf Z$ with $d$ **independent** and **identical** standard normal components.
+2. Determine the input matrix $\Sigma_{\bf Z}$ to corresponds with the
+   specified output $\Sigma_{\bf Y}$ [@Chen2001; @Xia17]
+3. Produce the Cholesky factor $M$ of $\Sigma_{\bf Z}$ so that $M M^{\prime}=\Sigma_{\bf Z}$.
+4. Set $X$ by $X \gets MZ$.
+5. $\text{Return} \; Y \; \text{where} \; Y_i \gets F_{Y_i}^{-1}[\Phi(X_i)], \; i=1,2,...,d$.
+
 We shall see that constructing continuous joint distributions that match a target Spearman or Kendall's correlations computes easily when employing Gaussian copulas, since this measures are invariant under the monotone transformations involved [refs].
@@
 
+<!-- 
 For discrete marginals, achieving a target Spearman correlation under this scheme is possible by using components from Equation \@ref(eq:spearmanRescaled) to further adjust the input correlation matrix.
@@
 
+your comment -->
+
 In contrast the rank-based correlations, matching specified Pearson correlation coefficients exactly is computational intense in this scheme.
-In general, there is no closed form correspondence and involving computing or approximating $\binom{d}{2}$ integrals of the form $EY_iY_j = \int \int y_i y_j f_{X|r}(F_i^{-1}(\Phi(z_i)), F_j^{-1}(\Phi(z_j))dy_idy_j$, for $i,j=1,2,\ldots,d$.
+In general, there is no closed form correspondence and involving computing or approximating $\binom{d}{2}$ integrals of the form $EY_iY_j = \int \int y_i y_j f_{X|r}(F_i^{-1}(\Phi(z_i)), F_j^{-1}(\Phi(z_j))dy_idy_j$, for $i,j=1,2,\ldots,d$ (see Xia17 and MB13).
 For accurate numeric approximation of these integrals, the functions must be evaluated hundreds of times. 
 Others have used efficient Monte Carlo integration schemes (see @Chen2001), but scale poorly to large dimension in reasonable times (property **S2**). 
-Despite all this, if one does desire to characterize dependency using Pearson correlations, we often see in practice --- and it is theoretically justified under certain conditions (@Song00) --- that simply using the target Pearson correlation matrix as the initial conditions to our proposed algorithm will lead to approximate matching in the resultant distribution.
+Despite all this, if one does desire to characterize dependency using Pearson correlations, we often see in practice --- and it is theoretically justified under certain conditions (@Song00) --- 
+that simply using the target Pearson correlation matrix as the initial conditions to our proposed algorithm will lead to approximate matching in the resultant distribution.
+We'll study the robustness of our method to this limitation later in the application section (Section BLAH).
 
@@
 
-## Algorithm notes
-
-- Discrete margins pose some difficulties and matching *unadjusted* correlation
-  measures is not exact (empirically approximate).
-- Exact Pearson solutions are possible but require computing $d \choose 2$
-   double integrals $EY_iY_j = \int \int y_i y_j f_{X|r}(F_i^{-1}(\Phi(z_i)),
-   F_j^{-1}(\Phi(z_j))dy_idy_j$. We have yet to accelerate this process.
-- We GPU accelerate whenever possible during our proposed High-Performance NORTA
-  algorithm.
-- One thorny problem when dealing with high-dimensional correlation matrices is
-they can become non-positive definite through either during estimation or
+One thorny problem when dealing with high-dimensional correlation matrices is
+they can become non-positive definite through either during estimation (if done pairwise) or through the algorithm's 
 bivariate transformations.
-- When converting nonparametric correlations to Pearson (first step above) the
-  resultant correlation may not be positive definite (PD).
-- In that case, we replace ${\bf R}_{Pearson}$ with a "close" PD matrix ${\bf \tilde{R}}_{Pearson}$.
-- In practice, this loss in accuracy typically has little impact on performance,
-  but the algorithm needs acceleration (`Matrix::nearPD()` violates property S1).
-
-test workflow again

### 2020-10-16 040-monte-carlo-evaluation.Rmd
migrating ESCO 2020 talk into sections, revising everything

@@
-# Monte Carlo evaluation and Comparsions to other software
+# Monte Carlo evaluations
+
+```{r LoadLib040, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
+## devtools::install_github( "adknudson/bigsimr", ref = 'develop')
+## knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
+## reticulate::use_condaenv("bigsimr")
+library(bigsimr)
+library(tidyverse)
+## CORES <- as.integer( parallel::detectCores() - 1 )
+set.seed(10162020)
+```
+
+```{r previewChapter040, echo = FALSE, eval = FALSE}
+bookdown::preview_chapter('040-monte-carlo-evaluation.Rmd')
+```
 
@@
 The simulations will proceed in increasing complexity --- leading up to the setting in our motivating example. 
-We begin by exploring simple exchangeable (constant) correlation structures under a large number of simulation replicates while applying the algorithm above to first continuous then discrete marginal distributions.
-We conclude the simulation studies with $10^5$ replicated data sets with sample sizes and distributional characteristics corresponding the estimates from the motivating example data.
-
-The CPU-based computation times we report were from runs using eight dual-threaded 3.9GHz
-Xeon Gold processors on a linux workstation, allowing for parallel instructions
-over 16 cores. We chose 15 cores for our experiments.
-
-## Other multivariate simulation software
-
-Describe the competing software and algorithms briefly.
-
-- `copula`
-- `Genord`
-- `Multiord`
-- `nortaRA` only does Pearson matching.
+We begin by exploring simple exchangeable (constant) correlation structures under a large number of simulation replicates while applying the algorithm above to first continuous then discrete marginal distributions to test the robustness of our algorithm since exact matching is not guaranteed.
 
@@
 
-### Continuous example: Bivariate Exponential
+### Bivariate Normal
 
-Both Spearman's and Kendall's measures of dependency are invariant under
-strictly monotone transformations (ref). This in turn provides exact simulation
-of continuous marginals $F_i$ $i=1,\ldots,d$ since their corresponding quantile
-function are increasing monotonically over their support. The results below
-provide computations times for $N=10^5$ under increasingly large dimension $d$.
-For simplicity we assume multivariate Normal with varying parameters and to
-additionally demonstrate the utility of Kendall's $\tau$ even in the Gaussian
-setting (as opposed to the nature choice of Pearson's correlation). To layer on complexity, we now assume a multivariate gamma distribution with fixed marginal parameters.
+* Let's simulate a bivariate normal and check our correlation matching
+  performance as N increases.
+* Here we have BVN( $\mu_1 = \mu_2 = 10, \rho_{type}$ )
+* We vary $\rho$ across the entire possible range of correlations for each
+  correlation type.
 
-Compare speed and accuracy.
+```{r biNormal, echo = FALSE, eval = TRUE}
+mom_norm <- function(x) {
+  m <- mean(x)
+  s <- sd(x)
+  list(mean = m, sd = s)
+}
 
-```{r tabGamma, echo = TRUE}
-## computation times normal
+mu <- 10
+sigma <- 1
+margins <- alist(
+    qnorm(mean = mu, sd = sigma),
+    qnorm(mean = mu, sd = sigma)
+)
 
+type <- c("pearson", "spearman", "kendall")
+Cores <- as.integer( parallel::detectCores() - 1 )
+n <- c(1e3, 1e4, 1e5)
+adjustForDiscrete <- c(FALSE)
+
+eps <- 1e-2
+grid_steps <- 100
+sim_pars <- expand.grid(type = type, cores = cores, n = n,
+                        stringsAsFactors = FALSE,
+                        adjustForDiscrete = adjustForDiscrete)
+
+res <- data.frame()
+
+for (i in 1:nrow(sim_pars)) {
+    type <- sim_pars$type[i]
+    cores <- sim_pars$cores[i]
+    n <- sim_pars$n[i]
+    adjustForDiscrete <- sim_pars$adjustForDiscrete[i]
+    tmp_bounds <- cor_bounds(margins, type = type)
+    cor_lo <- tmp_bounds$lower[1,2] + eps
+    cor_hi <- tmp_bounds$upper[1,2] - eps
+    cor_seq <- seq(cor_lo, cor_hi, length.out = grid_steps)
+    ## rho = cor_seq[1]
+    for (rho in cor_seq) {
+        Rho <- matrix(rho, 2, 2)
+        diag(Rho) <- 1.0
+        ## Rho <- convertCor( rho = Rho, from = type, to = 'pearson' )
+        time_data <- system.time({
+            ## x <- rmvn(n = n, mu = rep( mu, 2 ), sigma = Rho)
+            ## use rvec to test our software
+            x <- rvec(n = n,
+                      rho = Rho,
+                      margins = margins,
+                      type = type,
+                      cores = cores)
+        })
+        ## Save the sims in case
+        id <- paste0(
+            "d", 2,
+            "-N", n,
+            "-c", cores,
+            "-r", rho,
+            "-Cor", type,
+            "-adj", as.character(adjustForDiscrete),
+            "-dev", "1CORE",
+            "-lib", "bigsimr"
+        )
+        ## save sims in their own subdirectory
+        dir.create("./results/sims_norm", showWarnings = FALSE)
+        saveRDS(x, file = paste0("./results/sims_norm/", id, ".rds"))
+
+        ## Estimate statistics
+        Rho_hat <- cor_fast(x, method = type)
+        rho_hat <- Rho_hat[1, 2]
+        norm_args_hat <- mom_norm(x[,1])
+        mu_hat <- norm_args_hat$mean
+        sigma_hat <- norm_args_hat$sd
+
+        ## Save the results
+        res <- rbind(res, data.frame(
+                              method = "bigsimr",
+                              device = "CPU",
+                              type = type,
+                              cores = cores,
+                              margins = "norm",
+                              adjustForDiscrete = adjustForDiscrete,
+                              d = 2,
+                              N = n,
+                              rho = rho,
+                              rho_hat = rho_hat,
+                              mean = mu,
+                              sd = sigma,
+                              mean_hat = mu_hat,
+                              sd_hat = sigma_hat,
+                              sim_time = unname(time_data["elapsed"])
+                          ))
+    }
+}
+res$type <- factor(res$type, levels = c("pearson",  "spearman", "kendall" ) )
+saveRDS(object = res, "./results/norm_sims.rds")
 ```
-### Discrete example: Bivariate Geometric
 
-Compare speed and accuracy.
+----
 
-## Simulation II: Ultra HD multivariate simulation with identical margins and exchangeable rank-based correlation.
+```{r biNormPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.height = 5, fig.width= 8, fig.align='center', fig.cap = "`bigsimr` recovers the Pearson specified correlations for MVN."}
+dat <- readRDS("./results/norm_sims.rds")
+dat %>%
+    filter(cores == 1) %>%
+    ggplot(aes(rho, rho_hat, color = type)) +
+    ## ggplot(aes(rho, rho_hat)) +
+    geom_point() +
+    geom_abline(slope = 1) +
+    ## facet_wrap(~ + N) + theme_bw()
+    facet_wrap(~ type + N) + theme_bw()
 
-### UHD Continuous example: Multivariate Gamma
+## ggsave('fig/plot-biNormPlot.pdf')
+```
 
-Compare speed and accuracy.
+## Bivariate Gamma
 
-### UHD Discrete example: Multivariate Negative Binomial
+* Similarly, let's check the performance for a non-symmetric continuous distribution: a
+  standard (rate =1) bivariate gamma.
+* Here we have a Bivariate Gamma with $shape_1 = shape_2 = 10, \rho_{type}$.
+* We vary $\rho$ across the entire possible range of correlations for each
+  correlation type.
 
-Compare speed and accuracy.
+```{r biGamma, echo = FALSE, eval = FALSE}
+mom_gamma <- function(x) {
+  m <- mean(x)
+  s <- sd(x)
+  list(shape = m^2 / s^2, rate = m / s^2)
+}
 
-## Simulation III: Ultra HD multivariate simulatino with heterogenous margins and exchangeable rank-based correlation.
+shape <- 10
+rate <- 1
+margins <- list(
+  list("gamma", shape = shape, rate = rate),
+  list("gamma", shape = shape, rate = rate)
+)
 
-### UHD Heteregeous continuous example: Multivariate Gamma
+type <- c("pearson", "spearman", "kendall")
+cores <- c(1L)
+n <- c(1e3, 1e4, 1e5)
+adjustForDiscrete <- c(FALSE)
 
-Compare speed and accuracy.
+eps <- 1e-2
+grid_steps <- 100
+sim_pars <- expand.grid(type = type, cores = cores, n = n,
+                        stringsAsFactors = FALSE,
+                        adjustForDiscrete = adjustForDiscrete)
 
-### UHD Heteregeousdiscrete example: Multivariate Negative Binomial
+res <- data.frame()
+for (i in 1:nrow(sim_pars)) {
 
-Compare speed and accuracy.
+  type <- sim_pars$type[i]
+  cores <- sim_pars$cores[i]
+  n <- sim_pars$n[i]
+  adjustForDiscrete <- sim_pars$adjustForDiscrete[i]
+
+  tmp_bounds <- computeCorBounds(margins, type = type)
+  cor_lo <- tmp_bounds$lower[1,2] + eps
+  cor_hi <- tmp_bounds$upper[1,2] - eps
+  cor_seq <- seq(cor_lo, cor_hi, length.out = grid_steps)
+
+  for (rho in cor_seq) {
+    Rho <- matrix(rho, 2, 2)
+    diag(Rho) <- 1.0
+
+    time_data <- system.time({
+      x <- rvec(n = n,
+                rho = Rho,
+                params = margins,
+                cores = cores,
+                type = type,
+                adjustForDiscrete = adjustForDiscrete)
+    })
+
+    # Save the sims in case
+    id <- paste0(
+      "d", 2,
+      "-N", n,
+      "-c", cores,
+      "-r", rho,
+      "-Cor", type,
+      "-adj", as.character(adjustForDiscrete),
+      "-dev", "1CORE",
+      "-lib", "bigsimr"
+    )
+    saveRDS(x, file = paste0("sims_gamma/", id, ".rds"))
+
+    # Estimate statistics
+    Rho_hat <- fastCor(x, method = type)
+    rho_hat <- Rho_hat[1, 2]
+    gamma_args_hat <- mom_gamma(x[,1])
+    shape_hat <- gamma_args_hat$shape
+    rate_hat <- gamma_args_hat$rate
+
+    # Save the results
+    res <- rbind(res, data.frame(
+      method = "bigsimr",
+      device = "1CORE",
+      type = type,
+      cores = cores,
+      margins = "gamma",
+      adjustForDiscrete = adjustForDiscrete,
+      d = 2,
+      N = n,
+      rho = rho,
+      rho_hat = rho_hat,
+      shape = shape,
+      rate = rate,
+      shape_hat = shape_hat,
+      rate_hat = rate_hat,
+      sim_time = unname(time_data["elapsed"])
+    ))
+  }
+}
+res$type <- factor(res$type, levels = c("pearson",  "spearman", "kendall" ) )
+saveRDS(object = res, "gamma_sims_ags.rds")
+```
+
+---
+
+```{r biGammaPlot, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.width= 8, fig.height = 5,fig.align='center', fig.cap = "`bigsimr` recovers correlations for bivariate gamma only approximately for Pearson but exactly for the rank-based correlations."}
+dat <- readRDS("gamma_sims_ags.rds")
+dat %>%
+    filter(cores == 1, type == 'pearson') %>%
+    ggplot(aes(rho, rho_hat, color = type)) +
+    geom_point() +
+    geom_abline(slope = 1) +
+    ## facet_wrap(~ type + N)
+    facet_wrap(~ type + N)
+```
+
+---
+
+```{r biGammaPlot2, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.width= 8, fig.height = 5,fig.align='center', fig.cap = "`bigsimr` recovers the correlations for bivariate gamma only approximately for Pearson but exactly for the rank-based correlations."}
+dat <- readRDS("gamma_sims_ags.rds")
+dat %>%
+    filter(cores == 1, type == 'spearman') %>%
+    ggplot(aes(rho, rho_hat, color = type)) +
+    geom_point() +
+    geom_abline(slope = 1) +
+    ## facet_wrap(~ type + N)
+    facet_wrap(~ type + N)
+```
+
+---
+
+```{r biGammaPlot3, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.width= 8, fig.height = 5,fig.align='center', fig.cap = "`bigsimr` recovers the correlations for bivariate gamma only approximately for Pearson but exactly for the rank-based correlations."}
+dat <- readRDS("gamma_sims_ags.rds")
+dat %>%
+    filter(cores == 1, type == 'kendall') %>%
+    ggplot(aes(rho, rho_hat, color = type)) +
+    geom_point() +
+    geom_abline(slope = 1) +
+    ## facet_wrap(~ type + N)
+    facet_wrap(~ type + N)
+```
+
+## Bivariate Negative Binomial
+
+* Let's check the performance for a discrete distribution: a bivariate negative binomial
+* Here we have Bivariate Negative Binomial ( $prob_1 = prob_2 = 0.5, size_1 = size_2 = 4,\rho_{type}$ )
+* We vary $\rho$ across the entire possible range of correlations for each
+  correlation type.
+
+```{r biNegBin, echo = FALSE, eval = FALSE}
+mom_nbinom <- function(x) {
+  m <- mean(x)
+  s <- sd(x)
+  list(size = m^2 / (s^2 - m), prob = m / s^2)
+}
+
+size <- 4
+prob <- 0.5
+margins <- list(
+  list("nbinom", size = size, prob = prob),
+  list("nbinom", size = size, prob = prob)
+)
+
+type <- c("pearson", "spearman", "kendall")
+cores <- c(1)
+n <- c(1e3, 1e4, 1e5)
+adjustForDiscrete <- c(FALSE)
+
+eps <- 1e-2
+grid_steps <- 100
+sim_pars <- expand.grid(type = type, cores = cores, n = n,
+                        stringsAsFactors = FALSE,
+                        adjustForDiscrete = adjustForDiscrete)
+
+res <- data.frame()
+for (i in 1:nrow(sim_pars)) {
+
+  type <- sim_pars$type[i]
+  cores <- sim_pars$cores[i]
+  n <- sim_pars$n[i]
+  adjustForDiscrete <- sim_pars$adjustForDiscrete[i]
+
+  tmp_bounds <- computeCorBounds(margins, type = type)
+  cor_lo <- tmp_bounds$lower[1,2] + eps
+  cor_hi <- tmp_bounds$upper[1,2] - eps
+  cor_seq <- seq(cor_lo, cor_hi, length.out = grid_steps)
+
+  for (rho in cor_seq) {
+    Rho <- matrix(rho, 2, 2)
+    diag(Rho) <- 1.0
+
+    time_data <- system.time({
+      x <- rvec(n = n,
+                rho = Rho,
+                params = margins,
+                cores = cores,
+                type = type,
+                adjustForDiscrete = adjustForDiscrete)
+    })
+
+    # Save the sims in case
+    id <- paste0(
+      "d", 2,
+      "-N", n,
+      "-c", cores,
+      "-r", rho,
+      "-Cor", type,
+      "-adj", as.character(adjustForDiscrete),
+      "-dev", "1CORE",
+      "-lib", "bigsimr"
+    )
+    saveRDS(x, file = paste0("sims_nb/", id, ".rds"))
+
+    # Estimate statistics
+    Rho_hat <- fastCor(x, method = type)
+    rho_hat <- Rho_hat[1, 2]
+    nbinom_args_hat <- mom_nbinom(x[,1])
+    size_hat <- nbinom_args_hat$size
+    prob_hat <- nbinom_args_hat$prob
+
+    # Save the results
+    res <- rbind(res, data.frame(
+      method = "bigsimr",
+      device = "1CORE",
+      type = type,
+      cores = cores,
+      margins = "nbinom",
+      adjustForDiscrete = adjustForDiscrete,
+      d = 2,
+      N = n,
+      rho = rho,
+      rho_hat = rho_hat,
+      size = size,
+      prob = prob,
+      size_hat = size_hat,
+      prob_hat = prob_hat,
+      sim_time = unname(time_data["elapsed"])
+    ))
+  }
+}
+res$type <- factor(res$type, levels = c("pearson",  "spearman", "kendall" ) )
+saveRDS(object = res, "nbinom_sims_ags.rds")
+```
+
+---
+
+```{r  biNegBinPlot, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.align='center', fig.width = 8, fig.height=5,fig.cap = "`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
+dat <- readRDS("nbinom_sims_ags.rds")
+dat %>%
+    filter(cores == 1, type == 'pearson') %>%
+    ggplot(aes(rho, rho_hat, color = type)) +
+    geom_point() +
+    geom_abline(slope = 1) +
+    ## facet_wrap(~ type + N)
+    facet_wrap(~ type + N)
+```
+
+---
+
+```{r  biNegBinPlot2, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.align='center', fig.width = 8, fig.height=5,fig.cap = "`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
+dat <- readRDS("nbinom_sims_ags.rds")
+dat %>%
+    filter(cores == 1, type == 'spearman') %>%
+    ggplot(aes(rho, rho_hat, color = type)) +
+    geom_point() +
+    geom_abline(slope = 1) +
+    ## facet_wrap(~ type + N)
+    facet_wrap(~ type + N)
+```
+
+---
+
+```{r  biNegBinPlot3, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.align='center', fig.width = 8, fig.height=5,fig.cap = "`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
+dat <- readRDS("nbinom_sims_ags.rds")
+dat %>%
+    filter(cores == 1, type == 'kendall') %>%
+    ggplot(aes(rho, rho_hat, color = type)) +
+    geom_point() +
+    geom_abline(slope = 1) +
+    ## facet_wrap(~ type + N)
+    facet_wrap(~ type + N)
+```

### 2020-10-16 050-motivating-application-and-examples.Rmd
migrating ESCO 2020 talk into sections, revising everything

--- 2020-10-15 050-some-applications.Rmd
@@
-# Some applications
+# Motivating application and examples
 
@@
 ## devtools::install_github( "adknudson/bigsimr", ref = 'develop')
-## knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
+knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
 reticulate::use_condaenv("bigsimr")
@@
 set.seed(10142020)
+```
+
+```{r previewChapter050, echo = FALSE, eval = FALSE}
+bookdown::preview_chapter('050-some-applications.Rmd')
 ```
@@
 
-```{r processBRCA, echo=FALSE, eval=FALSE, results='none'}
+```{r processBRCA, echo=FALSE, eval=TRUE, results='none'}
 ## full workflow simulating RNA-seq data
@@
 ## retain top (1-probs)*100% highest expressing genes for illustration
-myProb <- 0.75
+myProb <- 0.99
 cutPoint <- quantile( x = brcaMedian, probs = myProb )
@@
 ## save data as rds
-saveRDS(brca, 'data/brca.rds')
+## saveRDS(brca, './data/brca99.rds')
 ```
 
-```{r readBRCA, echo=TRUE, eval=TRUE}
+```{r readBRCA, echo=FALSE, eval=FALSE}
 ## full workflow simulating RNA-seq data
-brca <- readRDS(file = 'data/brca.rds') 
+brca <- readRDS(file = './data/brca99.rds') 
 ```
@@
 ## 1. Estimate Spearman's correlation on the count data
-corType <- 'pearson'
+## corType <- 'pearson'
+corType <- 'spearman'
 system.time( nb_Rho <- bigsimr::cor_fast( brca, method = corType ) )
 ```
-
----
 
@@
 
-```{r runBRCA, results='none'}
-n <- 20000
-## d <- 100
-## n <- 100
-## d <- 5
+```{r runBRCA, echo=TRUE, results=TRUE, cache = TRUE}
+## Set the number of random vectors
+n <- 10000
+## construct margins
 nb_margins <- make_nbinom_alist(sizes, probs)
-## compute theoretical bounds to make sure input is okay
-## cor_bounds( nb_margins, type = type )
-sim_nbinom <- rvec(n, nb_Rho, nb_margins, type = corType, ensure_PSD = TRUE, cores = CORES)
-## sim_nbinom <- rvec(n, nb_Rho, nb_margins, type = "pearson", ensure_PSD = TRUE, cores = 4L)
-## pairs(sim_nbinom)
-nb_Rho_hat <- cor_fast(sim_nbinom, method = corType)
-cat('The quadratic loss from the true correlation matrix is ') 
-norm(nb_Rho - nb_Rho_hat, type = "2")
+## run sims
+system.time( sim_nbinom <- rvec(n, nb_Rho, nb_margins, type = corType,
+                                ensure_PSD = TRUE, cores = CORES) )
 ```
@@
 
-```{r fig:BRCA, echo=F, eval=T, caption='The T-Poisson strategy produces simulated random vectors from a multivariate negative binomial (NB) that replicate the estimated structure from an RNA-seq data set.
-The dashed red lines indicated equality between estimated parameters (vertical axes; derived from the simulated data) and the specified target parameters (horizontal axes).'}
+```{r figBRCA, echo=F, eval=T, fig.cap="bigsimr produces simulated random vectors from a multivariate negative binomial that replicate the estimated structure from an RNA-seq data set.
+The dashed red lines indicated equality between estimated parameters (vertical axes; derived from the simulated data) and the specified target parameters (horizontal axes)."}
 par(mfrow=c(1,3))
+## estimate rho from sims and plot
+nb_Rho_hat <- cor_fast(sim_nbinom, method = corType)
 plot(nb_Rho[lower.tri(nb_Rho)], nb_Rho_hat[lower.tri(nb_Rho_hat)],
@@
      xlab = "Target Correlation", ylab = "Estimated Correlation",
-     main = "bigsimr Simulation")
+     main = "Correlation Matching")
 abline(a = 0, b = 1, col="red", lty="dashed")
+## estimate prob,size from sims and plot
 nbinom_size_prob_hat <- apply(sim_nbinom, 2, nbinom_mom)
@@
      xlab = "Target Size", ylab = "Estimated Size",
-     main = "Direct NB Simulation")
+     main = "Marginal NB Size Matching")
 abline(a=0, b=1, col="red", lty="dashed")
@@
      xlab = "Target Log Probability", ylab = "Estimated Log Probability",
-     main = "Direct NB Simulation")
+     main = "Marginal NB Probability Matching")
 abline(a=0, b=1, col="red", lty="dashed")

### 2020-10-16 index.Rmd
migrating ESCO 2020 talk into sections, revising everything

@@
 As a motivating example, we use our methodology to study large-scale statistical inferential procedures applied to cancer-related RNA-sequencing data sets.
-The proposed algorithm is implemented as the `bigsimr` R package.
+The proposed algorithm is implemented as the bigsimr R package.
 ---
@@
 An alternative strategy is to think *generatively* about the data collection process and scientific domain knowledge and design simulations with probabilistic models that reflect those processes.
-In our motivating example, we'll study the consequences correlation in the simulation study replicating test statistics from an illustrative, classical large-scale hypothesis testing study [see @Efron2004].
+Our methodology provides another toolkit to generate data by considering how the data arise in the experimental setting.
 
@@
 We'll show that quality of simulation does not require the use of the standard Pearson correlation matrix, even for normally distributed data.
-The purpose of our method is to generate random vectors that match exactly specified any possible Kendall's $\tau$ or Spearman's $\rho$, and approximately Pearson product-moment correlation.
+The purpose of our method is to generate random vectors that match exactly specified any possible Kendall's $\tau$ or Spearman's $\rho$, and approximately Pearson product-moment correlation.More formally, our goal is to simulate $N$ random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with **correlated** components.
+The algorithm must scale to a large number of simulation replicates,$N$ for high-dimensional multivariate constructions (dimension $d$ in the tens or hundred thousands.
 

### 2020-10-17 040-monte-carlo-evaluation.Rmd
cache computations and scale down until content finalized

@@
 ## devtools::install_github( "adknudson/bigsimr", ref = 'develop')
-## knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
+knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
 ## reticulate::use_condaenv("bigsimr")
@@
 library(tidyverse)
-## CORES <- as.integer( parallel::detectCores() - 1 )
+cores <- as.integer( parallel::detectCores() - 1 )
 set.seed(10162020)
@@
 type <- c("pearson", "spearman", "kendall")
-Cores <- as.integer( parallel::detectCores() - 1 )
-n <- c(1e3, 1e4, 1e5)
+cores <- as.integer( parallel::detectCores() - 1 )
+## n <- c(1e3, 1e4, 1e5)
+n <- c(1e1, 1e2, 1e3)
 adjustForDiscrete <- c(FALSE)
@@
 eps <- 1e-2
-grid_steps <- 100
+## grid_steps <- 100
+grid_steps <- 10
 sim_pars <- expand.grid(type = type, cores = cores, n = n,
@@
 dat %>%
-    filter(cores == 1) %>%
+    filter(cores == cores) %>%
     ggplot(aes(rho, rho_hat, color = type)) +

### 2020-10-17 050-motivating-application-and-examples.Rmd
cache computations and scale down until content finalized

@@
 library(tidyverse)
-CORES <- as.integer( parallel::detectCores() - 1 )
+cores <- as.integer( parallel::detectCores() - 1 )
 set.seed(10142020)
@@
 
-```{r processBRCA, echo=FALSE, eval=TRUE, results='none'}
+```{r processBRCA, echo=FALSE, eval=FALSE, results='none'}
 ## full workflow simulating RNA-seq data
@@
 ## save data as rds
-## saveRDS(brca, './data/brca99.rds')
+saveRDS(brca, './data/brca99.rds')
 ```
 
-```{r readBRCA, echo=FALSE, eval=FALSE}
+```{r readBRCA, echo=FALSE, eval=TRUE}
 ## full workflow simulating RNA-seq data
@@
 system.time( sim_nbinom <- rvec(n, nb_Rho, nb_margins, type = corType,
-                                ensure_PSD = TRUE, cores = CORES) )
+                                ensure_PSD = TRUE, cores = cores) )
 ```

### 2020-10-17 index.Rmd
cache computations and scale down until content finalized

@@
 ## setwd('~/article_bigsimr')
-## bookdown::render_book("index.Rmd", "bookdown::pdf_book")
-## bookdown::serve_book()
 set.seed(09212020)
@@
 ## library(ggplot2); qplot(mpg, wt, data=mtcars, colour=cyl)
+## bookdown::render_book("index.Rmd", "bookdown::pdf_book")
 bookdown::serve_book()

### 2020-10-17 040-monte-carlo-evaluation.Rmd
add content to 040-montecarlo

--- 2020-06-04 040-monte-carlo-evaluation.Rmd
@@
-# Monte Carlo evaluation and Comparsions to other software
+# Monte Carlo evaluations
+
+```{r LoadLib040, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
+## devtools::install_github( "adknudson/bigsimr", ref = 'develop')
+knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
+## reticulate::use_condaenv("bigsimr")
+library(bigsimr)
+library(tidyverse)
+cores <- as.integer( parallel::detectCores() - 1 )
+## cores <- as.integer( 1 )
+set.seed(10162020)
+```
+
+```{r previewChapter040, echo = FALSE, eval = FALSE}
+bookdown::preview_chapter('040-monte-carlo-evaluation.Rmd')
+```
 
@@
 
-### Continuous example: Bivariate Exponential
+### Bivariate Normal
 
-Both Spearman's and Kendall's measures of dependency are invariant under
-strictly monotone transformations (ref). This in turn provides exact simulation
-of continuous marginals $F_i$ $i=1,\ldots,d$ since their corresponding quantile
-function are increasing monotonically over their support. The results below
-provide computations times for $N=10^5$ under increasingly large dimension $d$.
-For simplicity we assume multivariate Normal with varying parameters and to
-additionally demonstrate the utility of Kendall's $\tau$ even in the Gaussian
-setting (as opposed to the nature choice of Pearson's correlation). To layer on complexity, we now assume a multivariate gamma distribution with fixed marginal parameters.
+* Let's simulate a bivariate normal and check our correlation matching
+  performance as N increases.
+* Here we have BVN( $\mu_1 = \mu_2 = 10, \rho_{type}$ )
+* We vary $\rho$ across the entire possible range of correlations for each
+  correlation type.
 
-Compare speed and accuracy.
+```{r biNormal, echo = FALSE, eval = TRUE}
+mom_norm <- function(x) {
+  m <- mean(x)
+  s <- sd(x)
+  list(mean = m, sd = s)
+}
 
-```{r tabGamma, echo = TRUE}
-## computation times normal
+mu <- 10
+sigma <- 1
+margins <- alist(
+    qnorm(mean = mu, sd = sigma),
+    qnorm(mean = mu, sd = sigma)
+)
 
+type <- c("pearson", "spearman", "kendall")
+## cores <- as.integer( parallel::detectCores() - 1 )
+## n <- c(1e3, 1e4, 1e5)
+n <- c(1e1, 1e2, 1e3)
+adjustForDiscrete <- c(FALSE)
+
+eps <- 1e-2
+## grid_steps <- 100
+grid_steps <- 10
+sim_pars <- expand.grid(type = type, cores = cores, n = n,
+                        stringsAsFactors = FALSE,
+                        adjustForDiscrete = adjustForDiscrete)
+
+res <- data.frame()
+
+for (i in 1:nrow(sim_pars)) {
+    type <- sim_pars$type[i]
+    cores <- sim_pars$cores[i]
+    n <- sim_pars$n[i]
+    adjustForDiscrete <- sim_pars$adjustForDiscrete[i]
+    tmp_bounds <- cor_bounds(margins, type = type)
+    cor_lo <- tmp_bounds$lower[1,2] + eps
+    cor_hi <- tmp_bounds$upper[1,2] - eps
+    cor_seq <- seq(cor_lo, cor_hi, length.out = grid_steps)
+    ## rho = cor_seq[1]
+    for (rho in cor_seq) {
+        Rho <- matrix(rho, 2, 2)
+        diag(Rho) <- 1.0
+        ## Rho <- convertCor( rho = Rho, from = type, to = 'pearson' )
+        time_data <- system.time({
+            ## x <- rmvn(n = n, mu = rep( mu, 2 ), sigma = Rho)
+            ## use rvec to test our software
+            x <- rvec(n = n,
+                      rho = Rho,
+                      margins = margins,
+                      type = type,
+                      cores = cores)
+        })
+        ## Save the sims in case
+        id <- paste0(
+            "d", 2,
+            "-N", n,
+            "-c", cores,
+            "-r", rho,
+            "-Cor", type,
+            "-adj", as.character(adjustForDiscrete),
+            "-dev", "1CORE",
+            "-lib", "bigsimr"
+        )
+        ## save sims in their own subdirectory
+        dir.create("./results/sims_norm", showWarnings = FALSE)
+        saveRDS(x, file = paste0("./results/sims_norm/", id, ".rds"))
+
+        ## Estimate statistics
+        Rho_hat <- cor_fast(x, method = type)
+        rho_hat <- Rho_hat[1, 2]
+        norm_args_hat <- mom_norm(x[,1])
+        mu_hat <- norm_args_hat$mean
+        sigma_hat <- norm_args_hat$sd
+
+        ## Save the results
+        res <- rbind(res, data.frame(
+                              method = "bigsimr",
+                              device = "CPU",
+                              type = type,
+                              cores = cores,
+                              margins = "norm",
+                              adjustForDiscrete = adjustForDiscrete,
+                              d = 2,
+                              N = n,
+                              rho = rho,
+                              rho_hat = rho_hat,
+                              mean = mu,
+                              sd = sigma,
+                              mean_hat = mu_hat,
+                              sd_hat = sigma_hat,
+                              sim_time = unname(time_data["elapsed"])
+                          ))
+    }
+}
+res$type <- factor(res$type, levels = c("pearson",  "spearman", "kendall" ) )
+saveRDS(object = res, "./results/norm_sims.rds")
 ```
-### Discrete example: Bivariate Geometric
 
-Compare speed and accuracy.
+```{r biNormPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.height = 5, fig.width= 8, fig.align='center', fig.cap = "`bigsimr` recovers the Pearson specified correlations for MVN."}
+dat <- readRDS("./results/norm_sims.rds")
+dat %>%
+    filter(cores == cores) %>%
+    ggplot(aes(rho, rho_hat, color = type)) +
+    ## ggplot(aes(rho, rho_hat)) +
+    geom_point() +
+    geom_abline(slope = 1) +
+    ## facet_wrap(~ + N) + theme_bw()
+    facet_wrap(~ type + N) + theme_bw()
 
-## Simulation II: Ultra HD multivariate simulation with identical margins and exchangeable rank-based correlation.
+## ggsave('fig/plot-biNormPlot.pdf')
+```
 
-### UHD Continuous example: Multivariate Gamma
+## Bivariate Gamma
 
-Compare speed and accuracy.
+* Similarly, let's check the performance for a non-symmetric continuous distribution: a
+  standard (rate =1) bivariate gamma.
+* Here we have a Bivariate Gamma with $shape_1 = shape_2 = 10, \rho_{type}$.
+* We vary $\rho$ across the entire possible range of correlations for each
+  correlation type.
 
-### UHD Discrete example: Multivariate Negative Binomial
+```{r biGamma, echo = FALSE, eval = TRUE}
+mom_gamma <- function(x) {
+  m <- mean(x)
+  s <- sd(x)
+  list(shape = m^2 / s^2, rate = m / s^2)
+}
 
-Compare speed and accuracy.
+shape <- 10
+rate <- 1
+margins <- alist(
+    qgamma(shape = shape, rate = rate),
+    qgamma(shape = shape, rate = rate)
+)
 
-## Simulation III: Ultra HD multivariate simulatino with heterogenous margins and exchangeable rank-based correlation.
+type <- c("pearson", "spearman", "kendall")
+## cores <- c(1L)
+## n <- c(1e3, 1e4, 1e5)
+n <- c(1e1, 1e2, 1e3)
+adjustForDiscrete <- c(FALSE)
 
-### UHD Heteregeous continuous example: Multivariate Gamma
+eps <- 1e-2
+## grid_steps <- 100
+grid_steps <- 10
+sim_pars <- expand.grid(type = type, cores = cores, n = n,
+                        stringsAsFactors = FALSE,
+                        adjustForDiscrete = adjustForDiscrete)
 
-Compare speed and accuracy.
+res <- data.frame()
+for (i in 1:nrow(sim_pars)) {
 
-### UHD Heteregeousdiscrete example: Multivariate Negative Binomial
+    type <- sim_pars$type[i]
+    cores <- sim_pars$cores[i]
+    n <- sim_pars$n[i]
+    adjustForDiscrete <- sim_pars$adjustForDiscrete[i]
 
-Compare speed and accuracy.
+    tmp_bounds <- cor_bounds(margins, type = type)
+    cor_lo <- tmp_bounds$lower[1,2] + eps
+    cor_hi <- tmp_bounds$upper[1,2] - eps
+    cor_seq <- seq(cor_lo, cor_hi, length.out = grid_steps)
+
+    ## rho = cor_seq[1]
+    for (rho in cor_seq) {
+        Rho <- matrix(rho, 2, 2)
+        diag(Rho) <- 1.0
+
+        time_data <- system.time({
+            x <- rvec(n = n,
+                      rho = Rho,
+                      margins = margins,
+                      cores = cores,
+                      type = type,
+                      ensure_PSD = TRUE)
+        })
+
+        ## Save the sims in case
+        id <- paste0(
+            "d", 2,
+            "-N", n,
+            "-c", cores,
+            "-r", rho,
+            "-Cor", type,
+            "-adj", as.character(adjustForDiscrete),
+            "-dev", "CPU",
+            "-lib", "bigsimr"
+        )
+        dir.create("./results/sims_gamma", showWarnings = FALSE)
+        saveRDS(x, file = paste0("./results/sims_gamma/", id, ".rds"))
+
+        ## Estimate statistics
+        Rho_hat <- cor_fast(x, method = type)
+        rho_hat <- Rho_hat[1, 2]
+        gamma_args_hat <- mom_gamma(x[,1])
+        shape_hat <- gamma_args_hat$shape
+        rate_hat <- gamma_args_hat$rate
+
+        ## Save the results
+        res <- rbind(res, data.frame(
+                              method = "bigsimr",
+                              device = "CPU",
+                              type = type,
+                              cores = cores,
+                              margins = "gamma",
+                              adjustForDiscrete = adjustForDiscrete,
+                              d = 2,
+                              N = n,
+                              rho = rho,
+                              rho_hat = rho_hat,
+                              shape = shape,
+                              rate = rate,
+                              shape_hat = shape_hat,
+                              rate_hat = rate_hat,
+                              sim_time = unname(time_data["elapsed"])
+                          ))
+    }
+}
+res$type <- factor(res$type, levels = c("pearson",  "spearman", "kendall" ) )
+saveRDS(object = res, "./results/sims_gamma.rds")
+```
+
+```{r biGammaPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.height = 5, fig.width= 8, fig.align='center', fig.cap = "`bigsimr` recovers the Pearson specified correlations for Bivariate Gamma."}
+dat <- readRDS("./results/sims_gamma.rds")
+dat %>%
+    filter(cores == cores) %>%
+    ggplot(aes(rho, rho_hat, color = type)) +
+    ## ggplot(aes(rho, rho_hat)) +
+    geom_point() +
+    geom_abline(slope = 1) +
+    ## facet_wrap(~ + N) + theme_bw()
+    facet_wrap(~ type + N) + theme_bw()
+
+## ggsave('fig/plot-biGammaPlot.pdf')
+```
+
+## Bivariate Negative Binomial
+
+* Let's check the performance for a discrete distribution: a bivariate negative binomial
+* Here we have Bivariate Negative Binomial ( $prob_1 = prob_2 = 0.5, size_1 = size_2 = 4,\rho_{type}$ )
+* We vary $\rho$ across the entire possible range of correlations for each
+  correlation type.
+
+```{r biNegBin, echo = FALSE, eval = TRUE}
+mom_nbinom <- function(x) {
+  m <- mean(x)
+  s <- sd(x)
+  list(size = m^2 / (s^2 - m), prob = m / s^2)
+}
+
+size <- 10
+prob <- 0.1
+margins <- alist(
+    qnbinom(size = size, prob = prob),
+    qnbinom(size = size, prob = prob)
+)
+
+type <- c("pearson", "spearman", "kendall")
+## cores <- cores
+## n <- c(1e3, 1e4, 1e5)
+n <- c(1e1, 1e2, 1e3)
+adjustForDiscrete <- c(FALSE)
+
+eps <- 1e-2
+## grid_steps <- 100
+grid_steps <- 10
+sim_pars <- expand.grid(type = type, cores = cores, n = n,
+                        stringsAsFactors = FALSE,
+                        adjustForDiscrete = adjustForDiscrete)
+
+res <- data.frame()
+for (i in 1:nrow(sim_pars)) {
+    type <- sim_pars$type[i]
+    cores <- sim_pars$cores[i]
+    n <- sim_pars$n[i]
+    adjustForDiscrete <- sim_pars$adjustForDiscrete[i]
+    ## AGS. 17 Oct 2020. cor_bounds failed.
+    ## I'll create a minimal working example and open an issue
+    ## tmp_bounds <- cor_bounds(margins, type = type)
+    ## cor_lo <- tmp_bounds$lower[1,2] + eps
+    ## cor_hi <- tmp_bounds$upper[1,2] - eps
+    cor_lo <- -0.75
+    cor_hi <- 0.95
+    cor_seq <- seq(cor_lo, cor_hi, length.out = grid_steps)
+
+    for (rho in cor_seq) {
+        Rho <- matrix(rho, 2, 2)
+        diag(Rho) <- 1.0
+
+        time_data <- system.time({
+            x <- rvec(n = n,
+                      rho = Rho,
+                      margins = margins,
+                      cores = cores,
+                      type = type,
+                      ensure_PSD = TRUE)
+        })
+
+        ## Save the sims in case
+        id <- paste0(
+            "d", 2,
+            "-N", n,
+            "-c", cores,
+            "-r", rho,
+            "-Cor", type,
+            "-adj", as.character(adjustForDiscrete),
+            "-dev", "CPU",
+            "-lib", "bigsimr"
+        )
+        dir.create("./results/nbinom_sims", showWarnings = FALSE)
+        saveRDS(x, file = paste0("./results/nbinom_sims/", id, ".rds"))
+
+        ## Estimate statistics
+        Rho_hat <- cor_fast(x, method = type)
+        rho_hat <- Rho_hat[1, 2]
+        nbinom_args_hat <- mom_nbinom(x[,1])
+        size_hat <- nbinom_args_hat$size
+        prob_hat <- nbinom_args_hat$prob
+
+        ## Save the results
+        res <- rbind(res, data.frame(
+                              method = "bigsimr",
+                              device = "CPU",
+                              type = type,
+                              cores = cores,
+                              margins = "nbinom",
+                              adjustForDiscrete = adjustForDiscrete,
+                              d = 2,
+                              N = n,
+                              rho = rho,
+                              rho_hat = rho_hat,
+                              size = size,
+                              prob = prob,
+                              size_hat = size_hat,
+                              prob_hat = prob_hat,
+                              sim_time = unname(time_data["elapsed"])
+                          ))
+    }
+}
+res$type <- factor(res$type, levels = c("pearson",  "spearman", "kendall" ) )
+saveRDS(object = res, "./results/nbinom_sims.rds")
+```
+
+```{r  biNegBinPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.align='center', fig.width = 8, fig.height=5,fig.cap = "`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
+dat <- readRDS("./results/nbinom_sims.rds")
+dat %>%
+    filter(cores == cores) %>%
+    ggplot(aes(rho, rho_hat, color = type)) +
+    ## ggplot(aes(rho, rho_hat)) +
+    geom_point() +
+    geom_abline(slope = 1) +
+    ## facet_wrap(~ + N) + theme_bw()
+    facet_wrap(~ type + N) + theme_bw()
+## ggsave('fig/plot-biNegBinPlot.pdf')
+```
+
+## Scale up to Ultra-High Dimensions
+
+```{r gpuVScpuFig, echo=F, out.width='80%', fig.align='center', fig.cap="Computation times as d increases.
+We filter to the top 1, 5, 10, 15, 20, 25\\% expressing genes (in terms of median expression.)"}
+
+knitr::include_graphics("fig/cpu-gpu-times.png")
+```

### 2020-10-18 020-simulation-algorithm.Rmd
add cmstats talk notebook and simplify algorithm description

@@
-# Simulation algorithm
+# Algorithms
 
@@
 ```
-
-## Algorithm description
 
@@
 
-## Simulation Algorithm
-
 Putting the together the facts provided in the equations above, we come the following proposed simulation algorithm to produce a random vector ${\bf Y}$ with specified Spearman's correlation and marginal distributions.
@@
 
-### Algorithm 1: Matching Pearson approximately
+## Random vector generation `bigsimr::rvec` algorithm
 
-We aim to simulate multivariate data with a Pearson correlation
-matrix ${\bf R}$.
+<mark>make this algorithm pretty later </mark>
 
-(i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R})$;  
-(ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$;  
-(iii) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;
+1. Preprocessing for nonparameteric dependency matching.  
+	(i) Convert from either ${\bf R_{Spearman}}$ or ${\bf R_{Kendall}}$ into the corresponding MVN input correlation ${\bf R_{Pearson}}$ via \@ref(eq:convertSpearman) or \@ref(eq:convertKendall), respectivity.
+	(ii) Check that ${\bf R_{Pearson}}$ is semi-positive definite.  
+	(ii) If not find a close semi-positive ${\bf \widetilde{R}_{Pearson}}$ via `cor_nearPSD()`.  
+2. NORTA transformation  
+	(1) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_{Pearson}})$;  
+	(2) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$;  
+3. Return step  
+	(3) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;
 
-* Where $F_i^{-1}$ is the inverse CDF for the $i^{th}$ gene's marginal
-distribution.  
-* And ${\bf R}$ is the target Pearson correlation matrix.  
-* Pearson's correlation is **not** invariant under the montonically increasing
-  transformations in steps 3, 4.  
-* So the resulting simulated random vectors will have the **exact** margins but **approximate** ${\bf R}$.
+## Other high-performance `bigsimr` algorithms
 
-### Algorithm 2: Matching Spearman's exactly
+<mark> ALEX, briefly discuss other algorithms here </mark>
 
-(i) Convert ${\bf R}_{Spearman}$ into ${\bf R}_{Pearson}$ via $\rho_{Pearson} =
-  2 \times sin \left( \rho_{Spearman} \times \frac{\pi}{6} \right)$;  
-(ii) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf} R_{Pearson})$;  
-(iii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$;  
-(iv) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;
-
-- Spearman's $\rho$ relates to the difference in the ranks of pairs of values.
-- We take advantage of a closed form correspondence between Pearson and Spearman
-  for *normal random variables* in step 1 [@BF17].
-- Since Spearman's $\rho$ is invariant under the monotonically increasing
-  transformations in steps 3 and 4, the resulting simulated random vectors will
-  have the **exact** margins and **exact** Spearman correlation (up to Monte Carlo/numeric error).
-
-### Algorithm 3: Matching Kendall's exactly
-
-(i) Convert ${\bf R}_{Kendall}$ into ${\bf R}_{Pearson}$ via $\rho_{Pearson} = sin \left( \tau \times \frac{\pi}{2} \right)$;  
-(ii) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R}_{Pearson})$;  
-(iii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$;  
-(iv) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;
-
-- Kendall's $\tau$ is the difference in the probabilities of concordant and discordant pairs.
-- We use the 1-1 relationship between Pearson and Kendall's $\tau$ for *normal
-  rvs* in step 1.
-- Since Kendall's $\rho$ is invariant under the montonically increasing
-  transformations in steps 3 and 4,  
- the resulting simulated random vectors will have the **exact** margins and
- **exact** Kendall's $\tau$  (up to Monte Carlo/numeric error).
-
-One thorny problem when dealing with high-dimensional correlation matrices is
-they can become non-positive definite through either during estimation (if done pairwise) or through the algorithm's 
-bivariate transformations.
+- `cor_bounds()`  
+- `cor_covert()` make sure that you are clear this is for only MVN.  
+- `cor_fast()`  
+- `cor_nearPSD()`

### 2020-10-19 020-simulation-algorithm.Rmd
add more on the motivating data and some figs

@@
 distribution function $F_i$ and its inverse.
-To characterize dependency, every pair $(Y_i, Y_j)$ has either a specified Pearson correlation \@ref(eq:pearson), (rescaled) Spearman correlation \@ref(eq:spearmanRescaled) , or Kendall's $\tau$ (Equation \@ref(eq:tau) ).
+To characterize dependency, every pair $(Y_i, Y_j)$ has either a specified Pearson correlation \@ref(eq:pearson), (rescaled) Spearman correlation \, or Kendall's $\tau$ ).
 The method only approximately matches the Pearson correlation in general, whereas the rank-based methods are exact.

### 2020-10-19 050-motivating-application-and-examples.Rmd
add more on the motivating data and some figs

@@
-# Motivating application and examples
+# Replicating motivating data and example applications
 
@@
                                 ensure_PSD = TRUE, cores = cores) )
+colnames(sim_nbinom) <- names(brca)
+```
+
+Figure X.
+
+```{r simDataFig, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Simulated data for 3 randomly selected high-expressing genes, replicating the real data structure describing in the Introduction.'}
+## retrieve the examples from an earlier chapter
+## there is probably a better way to do this
+exampleGenes <- readRDS(file = './data/exampleGenes.rds') 
+GGally::ggpairs(data = as.data.frame(sim_nbinom[ ,exampleGenes] ))
 ```

### 2020-10-19 index.Rmd
add more on the motivating data and some figs

@@
 ## bookdown::render_book("index.Rmd", "bookdown::pdf_book")
+## bookdown::preview_chapter("index.Rmd")
 bookdown::serve_book()
@@
 example by employing our method and compare our simulations to empirically-derived statistics. Finally, we'll make concluding remarks regarding the method's utility and future directions.
+
+## Motivating example
+
+- RNA-seq
+- Provide a table of values
+- check the talk history
+
+```{r realDataTabl, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
+## Describe and plot real data
+brca <- readRDS(file = './data/brca99.rds')
+set.seed(10192020)
+exampleGenes <- colnames(brca)[sort(sample(ncol(brca), 3))]
+saveRDS(exampleGenes, file = './data/exampleGenes.rds')
+kable(round(brca[1:4, exampleGenes]))
+## GGally::ggpairs(data = as.data.frame(brca[ ,exampleGenes] ))
+```
+
+Figure X.
+
+```{r realDataFig, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Real data for the 3 selected high-expressing genes'}
+## Describe and plot real data
+GGally::ggpairs(data = as.data.frame(brca[ ,exampleGenes] ))
+```

### 2020-10-20 050-examples.Rmd
decide on scope of RNA examples, revise introduction

--- 2020-10-19 050-motivating-application-and-examples.Rmd
@@
-# Replicating motivating data and example applications
+# Example applications for our motivating data
 
@@
 ```{r previewChapter050, echo = FALSE, eval = FALSE}
-bookdown::preview_chapter('050-some-applications.Rmd')
+bookdown::preview_chapter('050-examples.Rmd')
 ```
-
-## Simulating High-Dimensional RNA-seq data
 
@@
 	This suggests simulating high-dimensional multivariate NB with heterogeneous marginals would be useful tool in the development and evaluation of RNA-seq analytics.
+
+## Simulating High-Dimensional RNA-seq data
 
@@
 
-## Ultra High-Dimensional simulation-based correlation hypothesis testing under
+## Simulation-based UHD joint probability calculations
+
+To conduct statistical inference, via maximum likelihood for example, a critical task is to evaluate the joint probability mass (or density) function:
+
+$$
+P( \bf{Y} = \bf{y} ), y_i \in \chi_i.
+$$
+
+where $\chi_i$ is the sample space for the $i^{th}$ component of the random vector $\bf{Y}$.
+Compact representations with convenient computational forms are rare for high-dimensional constructions, especially with hetereogeneous, correlated marginal distributions (or margins of mixed data types).
+Given a large number simulated vectors as produced above, estimated probabilities are readily given by counting the proportion of simulated vectors meeting the desired condition.
+In our motivating application, one may ask what is the probability that all genes expressed greater than a certain threshold value $\bf{y_0}$.
+
+Then we estimate
+
+$$
+\hat{P}( \bf{Y} >= \bf{ y_0 } ) = \sum_{b=1}^B I( \bf{Y^{(b)}} > \bf{ y_0 } ) / B
+$$
+
+where $\bf{Y^{(b)}}$ is the $b^{th}$ simulated vector in a total of $B$ simulation replicates and $I(\dot)$ is the indicator function.
+For example, we can estimate from our $B=10,000$ simulated vectors the probability that all genes express more than $\bf{y_0} = 1000$.
+
+```{r densityEvaluation, echo=TRUE, eval=TRUE}
+d <- ncol(sim_nbinom)
+B <- nrow(sim_nbinom)
+threshold <- rep( 1000, d)
+mean(apply( sim_nbinom,  1,
+           function(X, y0=threshold) {
+               all( X > y0) }
+           ))
+```
+
+## UHD Parameteric bootstrap
 
@@
 ```
-
-## Simulation-based joint probability calculations
-
-```{r densityEvaluation, echo=FALSE, eval=FALSE}
-## full workflow simulating RNA-seq data
-
-```

### 2020-10-20 index.Rmd
decide on scope of RNA examples, revise introduction

@@
 As new methods are developed for these data, a fundamental challenge lies in designing and conducting simulation studies to assess operating characteristics of proposed methodology, such as false positive rates, statistical power, and robustness --- often in comparison to existing methods.
-Another application lies in the discovering the sampling distribution of test statistic under hypothesized null models in complex scenarios.
-Such Monte Carlo studies in the high-dimensional setting often become difficult to conduct with existing algorithms and tools, or require expertise in computing on clusters. 
-This is particularly true when simulating massive multivariate, non-normal distributions.
+Further efficient simulation empowers computational strategies (such as a parameteric bootstrap --- simulating from a hypothesized null model) to provide inference in analytically challenging settings.
+Such Monte Carlo techniques become difficult for high-dimensional data with the current existing algorithms and tools.
+This is particularly true when simulating massive multivariate, non-normal distributions, arising naturally in many fields of study in this era of big data.
 

### 2020-10-20 040-monte-carlo-evaluation.Rmd
dir create moved outside of loop and set recursive
some directories did not exist, and so building the book would fail

@@
 ```{r LoadLib040, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
-## devtools::install_github( "adknudson/bigsimr", ref = 'develop')
+## devtools::install_github("SchisslerGroup/bigsimr", ref = 'develop')
 knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
-## reticulate::use_condaenv("bigsimr")
+# reticulate::use_condaenv("bigsimr")
 library(bigsimr)
@@
 res <- data.frame()
-
+dir.create("./results/sims_norm", showWarnings = TRUE, recursive = TRUE)
 for (i in 1:nrow(sim_pars)) {
@@
         ## save sims in their own subdirectory
-        dir.create("./results/sims_norm", showWarnings = FALSE)
         saveRDS(x, file = paste0("./results/sims_norm/", id, ".rds"))
@@
 
+dir.create("./results/sims_gamma", showWarnings = TRUE, recursive = TRUE)
 res <- data.frame()
@@
         )
-        dir.create("./results/sims_gamma", showWarnings = FALSE)
         saveRDS(x, file = paste0("./results/sims_gamma/", id, ".rds"))
@@
 
+dir.create("./results/nbinom_sims", showWarnings = TRUE, recursive = TRUE)
 res <- data.frame()
@@
         )
-        dir.create("./results/nbinom_sims", showWarnings = FALSE)
         saveRDS(x, file = paste0("./results/nbinom_sims/", id, ".rds"))

### 2020-10-20 050-examples.Rmd
dir create moved outside of loop and set recursive
some directories did not exist, and so building the book would fail

@@
 ```{r LoadLib050, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
-## devtools::install_github( "adknudson/bigsimr", ref = 'develop')
 knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
-reticulate::use_condaenv("bigsimr")
+# reticulate::use_condaenv("bigsimr")
 library(bigsimr)
@@
 ## save data as rds
-saveRDS(brca, './data/brca99.rds')
+saveRDS(brca, 'data/brca99.rds')
 ```
@@
 ## full workflow simulating RNA-seq data
-brca <- readRDS(file = './data/brca99.rds') 
+brca <- readRDS(file = 'data/brca99.rds') 
 ```

### 2020-10-21 index.Rmd
clean up auto genereated files

@@
 Massive high-dimensional data sets are now commonplace in many areas of scientific inquiry.
-As new methods are developed for these data, a fundamental challenge lies in designing and conducting simulation studies to assess operating characteristics of proposed methodology, such as false positive rates, statistical power, and robustness --- often in comparison to existing methods.
+As new methods are developed for these data, a fundamental challenge lies in designing and conducting simulation studies to assess operating characteristics of proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness --- often in comparison to existing methods.
 Further efficient simulation empowers computational strategies (such as a parameteric bootstrap --- simulating from a hypothesized null model) to provide inference in analytically challenging settings.

### 2020-10-21 050-examples.Rmd
move brca99.rds creation to index.Rmd

@@
 -->
-
-```{r processBRCA, echo=FALSE, eval=FALSE, results='none'}
-## full workflow simulating RNA-seq data
-allDat <- readRDS( file = "~/Downloads/complete_processed_tcga2stat_RNASeq2_with_clinical.rds" )
-lastClinical <- which( names(allDat) == 'tumorsize' )
-brca <- allDat[ allDat$disease == "BRCA", (lastClinical+1):ncol(allDat) ]
-dim(brca) ## num of genes 20501
-## compute the median expression! avoid the 0s
-brcaMedian <- apply(brca, 2, median)
-## retain top (1-probs)*100% highest expressing genes for illustration
-myProb <- 0.99
-cutPoint <- quantile( x = brcaMedian, probs = myProb )
-## genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
-genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
-brca <- brca[ , genesToKeep ]
-## ncol(brca) / 20501
-print(d <- length(genesToKeep))
-## save data as rds
-saveRDS(brca, 'data/brca99.rds')
-```
 

### 2020-10-21 index.Rmd
move brca99.rds creation to index.Rmd

@@
 
+```{r processBRCA, echo=FALSE, eval=TRUE, results='none'}
+if ( !file.exists( 'data/brca99.rds' ) )  {
+    ## full workflow simulating RNA-seq data
+    allDat <- readRDS( file = "~/Downloads/complete_processed_tcga2stat_RNASeq2_with_clinical.rds" )
+    lastClinical <- which( names(allDat) == 'tumorsize' )
+    brca <- allDat[ allDat$disease == "BRCA", (lastClinical+1):ncol(allDat) ]
+    dim(brca) ## num of genes 20501
+    ## compute the median expression! avoid the 0s
+    brcaMedian <- apply(brca, 2, median)
+    ## retain top (1-probs)*100% highest expressing genes for illustration
+    myProb <- 0.99
+    cutPoint <- quantile( x = brcaMedian, probs = myProb )
+    ## genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
+    genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
+    brca <- brca[ , genesToKeep ]
+    ## ncol(brca) / 20501
+    print(d <- length(genesToKeep))
+    ## save data as rds
+    saveRDS(brca, 'data/brca99.rds')
+}
+```
+
 ```{r realDataTabl, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}

### 2020-10-21 070-supplementary-materials.Rmd
push small change after meeting
added chapter prefix

@@
 
-# Notes
-
-### List of supported distributions
-
-```{r, eval=FALSE}
+```{r ch070-all-dists, eval=FALSE}
 all_dists <- list(
@@
 ```
-
-Mention Python?

### 2020-10-21 index.Rmd
revise abstract

@@
 --- 
-title: Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package
+title: Simulating High-Dimensional Multivariate Data using the bigsimr R Package
 author:
@@
 abstract: |
-  In this era of Big Data, it is critical to realistically simulate data to conduct informative Monte Carlo studies.
-This is problematic when data are inherently multivariate while at the same time are (ultra-) high dimensional.
-This situation appears frequently in observational data found on online and in high-throughput biomedical experiments (e.g., RNA-sequencing).
-Due to the difficulty in simulating realistic correlated data points, researchers often resort to simulation designs that posit independence --- greatly diminishing the insight into the empirical operating characteristics of any proposed methodology.
-Major challenges lie in the computational complexity involved in simulating these massive random vectors.
-We propose a fairly general, scalable procedure to simulate high-dimensional multivariate distributions with pre-specified marginal characteristics and dependency characteristics.
-As a motivating example, we use our methodology to study large-scale statistical inferential procedures applied to cancer-related RNA-sequencing data sets.
-The proposed algorithm is implemented as the bigsimr R package.
+  It is critical to realistically simulate data when conducting Monte Carlo studies and methods.
+But measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments.
+Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers often resort to simulation designs that posit independence.
+This greatly diminishes insights into the empirical operating characteristics of any proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness.
+This talk introduces the bigsimr R package that provides a general, scaleable procedure to simulate high-dimensional random vectors with given marginal characteristics and dependency measures.
+We'll describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation matrices, and find close positive semi-definite matrices.
+Finally, we showcase the power of bigsimr by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d>1000$.
 ---

### 2020-10-21 030-bigsimr-package.Rmd

--- 2020-09-18 030-bigsimr-package.Rmd
@@
 
-```{r setup3, eval = TRUE}
+```{r ch030-setup, eval = FALSE}
 ## setwd('~/Research/article_bigsimr/'); bookdown::preview_chapter('030-bigsimr-package.Rmd')
@@
 
-```{r setMargins}
+```{r ch030-setMargins, eval = FALSE}
 margins = alist(
@@
 
-```{r setCor}
+```{r ch030-setCor, eval = FALSE}
 rho <- matrix(0.5, nrow = 3, ncol = 3)
@@
 
-```{r rvec10}
+```{r ch030-rvec10, eval = FALSE}
 x <- bigsimr::rvec(10, rho = rho, margins = margins, type = "spearman")
@@
 
-```{r gpuWarn, echo=FALSE}
+```{r ch030-gpuWarn, echo=FALSE}
 warning("warning.warn('No GPU/TPU found, falling back to CPU.')")
@@
 
-```{r printX}
+```{r ch030-printX, eval = FALSE}
 x
@@
 
-```{r rvec1e6, fig.width=7}
+```{r ch030-rvec1e6, fig.width=7, eval = FALSE}
 x <- rvec(1e6, rho = rho, margins = margins, type = "spearman")
@@
 
-```{r fastCor}
-bigsimr::fastCor(x, method = 'spearman')
+```{r ch030-fastCor, eval = FALSE}
+## help( package = 'bigsimr')
+bigsimr::cor_fast(x, method = 'spearman')
 ```
@@
 
-```{r computeBounds, echo = TRUE, eval=FALSE}
+```{r ch030-computeBounds, echo = FALSE, eval=FALSE}
 ## Possibly bug? or configuration
@@
 
-```{r rvecRslurm, echo=TRUE, eval=FALSE}
+```{r ch030-rvecRslurm, echo=FALSE, eval=FALSE}
 library(rslurm)
@@
 
-```{r rvecRslurmRes, echo=TRUE, eval=FALSE}
+```{r ch030-rvecRslurmRes, echo=FALSE, eval=FALSE}
 res <- readRDS(file.path('_rslurm_rvec', list.files('_rslurm_rvec', 'results')))
@@
 
-```{r rvecRslurmMap, echo=TRUE, eval=FALSE}
+```{r ch030-rvecRslurmMap, echo=FALSE, eval=FALSE}
 library(rslurm)
@@
 
-```{r rvecRslurmMapRes, echo=TRUE, eval=FALSE}
+```{r ch030-rvecRslurmMapRes, echo=FALSE, eval=FALSE}
 list.files('_rslurm_rvecMap', 'results')
@@
 
-```{r rvecRslurmMapFig, echo = TRUE, eval=TRUE, message=FALSE, warning=FALSE, out.width= '75%', fig.align='center', fig.cap='The continous marginal pairs exact reproduce the specified correlation whereas the discrete pairs (green nad blue) show a downward bias.
+```{r ch030-rvecRslurmMapFig, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, out.width= '75%', fig.align='center', fig.cap='The continous marginal pairs exact reproduce the specified correlation whereas the discrete pairs (green nad blue) show a downward bias.
 Future work with employ a discrete adjusted procedures using *rescaling* on the traditional correlation coefficient.'}

### 2020-10-21 040-monte-carlo-evaluation.Rmd
clean up auto genereated files

@@
 
-```{r LoadLib040, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
+```{r ch040-LoadLib040, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
 ## devtools::install_github("SchisslerGroup/bigsimr", ref = 'develop')
@@
 
-```{r previewChapter040, echo = FALSE, eval = FALSE}
+```{r ch040-preview, echo = FALSE, eval = FALSE}
 bookdown::preview_chapter('040-monte-carlo-evaluation.Rmd')
@@
 
-```{r biNormal, echo = FALSE, eval = TRUE}
+```{r ch040-biNormal, echo = FALSE, eval = TRUE}
 mom_norm <- function(x) {
@@
 
-```{r biNormPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.height = 5, fig.width= 8, fig.align='center', fig.cap = "`bigsimr` recovers the Pearson specified correlations for MVN."}
+```{r ch040-biNormPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.height = 5, fig.width= 8, fig.align='center', fig.cap = "`bigsimr` recovers the Pearson specified correlations for MVN."}
 dat <- readRDS("./results/norm_sims.rds")
@@
 
-```{r biGamma, echo = FALSE, eval = TRUE}
+```{r ch040-biGamma, echo = FALSE, eval = TRUE}
 mom_gamma <- function(x) {
@@
 
-```{r biGammaPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.height = 5, fig.width= 8, fig.align='center', fig.cap = "`bigsimr` recovers the Pearson specified correlations for Bivariate Gamma."}
+```{r ch040-biGammaPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.height = 5, fig.width= 8, fig.align='center', fig.cap = "`bigsimr` recovers the Pearson specified correlations for Bivariate Gamma."}
 dat <- readRDS("./results/sims_gamma.rds")
@@
 
-```{r biNegBin, echo = FALSE, eval = TRUE}
+```{r ch040-biNegBin, echo = FALSE, eval = TRUE}
 mom_nbinom <- function(x) {
@@
 
-```{r  biNegBinPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.align='center', fig.width = 8, fig.height=5,fig.cap = "`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
+```{r ch040-biNegBinPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.align='center', fig.width = 8, fig.height=5,fig.cap = "`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
 dat <- readRDS("./results/nbinom_sims.rds")
@@
 
-```{r gpuVScpuFig, echo=F, out.width='80%', fig.align='center', fig.cap="Computation times as d increases.
+```{r ch040-gpuVScpuFig, echo=F, out.width='80%', fig.align='center', fig.cap="Computation times as d increases.
 We filter to the top 1, 5, 10, 15, 20, 25\\% expressing genes (in terms of median expression.)"}
-
 knitr::include_graphics("fig/cpu-gpu-times.png")

### 2020-10-21 050-examples.Rmd

@@
 
-```{r LoadLib050, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
+```{r ch050-LoadLib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
 knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
@@
 
-```{r previewChapter050, echo = FALSE, eval = FALSE}
+```{r ch050-preview, echo = FALSE, eval = FALSE}
 bookdown::preview_chapter('050-examples.Rmd')
@@
 
-```{r readBRCA, echo=FALSE, eval=TRUE}
+```{r ch050-readBRCA, echo=FALSE, eval=TRUE}
 ## full workflow simulating RNA-seq data
@@
 
-```{r estRhoBRCA, echo=TRUE, eval=TRUE}
+```{r ch050-estRhoBRCA, echo=TRUE, eval=TRUE}
 ## 1. Estimate Spearman's correlation on the count data
@@
 
-```{r estMargins, echo = FALSE, eval = TRUE}
+```{r ch050-estMargins, echo = FALSE, eval = TRUE}
 ## 2. Estimate NegBin parameters using Method of Moments
@@
 
-```{r nbHelpers}
+```{r ch050-nbHelpers}
 make_nbinom_alist <- function(sizes, probs) {
@@
 
-```{r runBRCA, echo=TRUE, results=TRUE, cache = TRUE}
+```{r ch050-runBRCA, echo=TRUE, results=TRUE, cache = TRUE}
 ## Set the number of random vectors
@@
 
-```{r simDataFig, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Simulated data for 3 randomly selected high-expressing genes, replicating the real data structure describing in the Introduction.'}
+```{r ch050-simDataFig, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Simulated data for 3 randomly selected high-expressing genes, replicating the real data structure describing in the Introduction.'}
 ## retrieve the examples from an earlier chapter
@@
 
-```{r figBRCA, echo=F, eval=T, fig.cap="bigsimr produces simulated random vectors from a multivariate negative binomial that replicate the estimated structure from an RNA-seq data set.
+```{r ch050-figBRCA, echo=F, eval=T, fig.cap="bigsimr produces simulated random vectors from a multivariate negative binomial that replicate the estimated structure from an RNA-seq data set.
 The dashed red lines indicated equality between estimated parameters (vertical axes; derived from the simulated data) and the specified target parameters (horizontal axes)."}
@@
 
-```{r densityEvaluation, echo=TRUE, eval=TRUE}
+```{r ch050-densityEvaluation, echo=TRUE, eval=TRUE}
 d <- ncol(sim_nbinom)
@@
 
-```{r corrTest, echo=FALSE, eval=FALSE, include=FALSE}
+```{r ch050-corrTest, echo=FALSE, eval=FALSE, include=FALSE}
 

### 2020-10-22 050-examples.Rmd

@@
-# Example applications for our motivating data
+# Example applications for our motivating data {#examples}
 
@@
 brca <- readRDS(file = 'data/brca99.rds')
+## smaller set for prototyping
+## d=5
+## brca <- brca[ , 1:d]
 ```
@@
 ## 2. Estimate NegBin parameters using Method of Moments
-estimateNegBinMoM <- function(tmpGene, minP = 1e-7) {
+## estimateNegBinMoM <- function(tmpGene, minP = 1e-7) {
+estimateNegBinMoM <- function(tmpGene) {
     tmpMean <- mean(tmpGene)
@@
     p <- tmpMean / tmpVar
-    if (p < minP) {p <- minP } ## maybe add noise here
+    ## if (p < minP) {p <- minP } ## maybe add noise here
     n <- ( tmpMean * p) / ( 1  - p )  
@@
 ```
+
+Here we use the basic method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model.
+We model marginals distributions as coming from the same probability family (NB) yet are hetereogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.
+
+TYPESTE MoM estimators here.
 
@@
 
-## UHD Parameteric bootstrap
+## Monte Carlo estimation of MSE
 
-Since many of the correlations are near zero, one could ask whether it is
-possible that test the hypothesis that $H_0: R = R_0$.
+Monte Carlo (MC) methods offer flexible and straightfoward statistical inferenec in this typically difficult analytic setting.
+MC methods are routinely used many statistical inferential tasks including estimation, hypothesis testing, error rates, and empirical interval coverage rates.
+For an concise introduction to these Methos, see, for example @Rizzo2007, Ch. 6.
+As a simple example, let's now employ `bigsimr` to estimate the mean squared error (MSE) of the method of moments estimators we used in section X.
 
-```{r ch050-corrTest, echo=FALSE, eval=FALSE, include=FALSE}
+Specifically, 
+TYPESET THE MATH HERE. REF to MoM above.
 
+```{r ch050-mom-mse, echo=TRUE, eval=TRUE, cache=FALSE}
+## system( "rm results/mseSizes.rds")
+if ( !file.exists( 'results/mseSizes.rds') )  {
+    n <- nrow(brca)
+    m <- 100
+    momSizes <- replicate(n = m, expr =
+                                     { tmpSim <- rvec(n, nb_Rho, nb_margins, type = corType, ensure_PSD = TRUE, cores = cores)
+                                         tmpMom <- apply( unname(as.matrix(tmpSim)), 2, estimateNegBinMoM )
+                                         tmpSizes <- unlist( lapply( X = tmpMom, function(x) x$size ) )
+                                     }
+                          )
+    sqDiffSizes <- apply( momSizes, 2, function(x) { (x  - sizes)^2 } )
+    mseSizes <- rowMeans( sqDiffSizes )
+    saveRDS(mseSizes, 'results/mseSizes.rds')
+}
 ```
+
+Figure X.
+
+```{r ch050-fig-mseSizes, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Real data for the 3 selected high-expressing genes'}
+mseSizes <- readRDS(file = 'results/mseSizes.rds')
+qplot(mseSizes, bins = 20) + theme_bw()
+```
+
+One could then compare the MSE of the baisc MoM estimation scheme to other strategies ( such as MLE, trimmed approaches, etc.) and decide whether a more computational costly approach is warrented or if the MoM will suffice.

### 2020-10-22 030-bigsimr-package.Rmd

@@
 
-Short description and key features, including GPU acceleration.
+`bigsimr` is an R package for simulating high-dimensional multivariate data with arbitrary marginal distributions.
+The efficiency behind generating multivariate samples comes from the ability to utilize a GPU during the generation of multivariate normal samples and the subsequent transformation to uniform samples (see section \@ref(rand-vec-gen)).
+A GPU is not necessary in order to use `bigsimr`, and the speed benefits generally only come when simulating very high dimensional data ($d > 10000$).
 
-The most computationally extensive lies in the second step of our algorithm. We
-use a optimized and parallelized multivariate normal simulator within the `R`
-package `mvnfast`. The rest of the code is also parallelized, but these steps
-run rapidly in serial computation except for extremely large dimension.
+The next computational step after obtaining the correlated uniform marginals is the inverse transform into the target marginal distributions.
+Since this consists of $d$ independent transformations, we utilize parallelization to achieve higher throughput.
+As with many parallel algorithms, there is overhead associated with forking the task to utilize multiple cores, however cost is negligible compared to the total run-time.
 
@@
 
-```{r ch030-setup, eval = FALSE}
-## setwd('~/Research/article_bigsimr/'); bookdown::preview_chapter('030-bigsimr-package.Rmd')
-## bookdown::render_book('index.html')
+We're going to show the basic use and syntax of `bigsimr` by using the New York air quality data set (`airquality`) included in the R `datasets` package.
+We will focus specifically on the temperature (degrees Fahrenheit) and ozone level (parts per billion).
+
+```{r ch030-basic-example-setup, echo=TRUE, message=FALSE}
 library(bigsimr)
-library(reticulate)
-reticulate::use_condaenv("bigsimr-cpu")
+library(tidyverse)
+library(patchwork)
 ```
 
-### Specifying marginals
+```{r ch030-air-quality-data, echo=TRUE}
+df <- airquality %>%
+  select(Temp, Ozone) %>%
+  drop_na()
+```
 
-As stated earlier, to generate multivariate data, we need a list of marginals (and their parameters), and a correlation structure (matrix).
-The marginal distributions can be built up as a list of lists, where each sublist contains the information for the target distribution.
+```{r ch030-aq-glimpse}
+glimpse(df)
+```
 
-```{r ch030-setMargins, eval = FALSE}
-margins = alist(
-  qnorm(mean = 3.14, sd = 0.1),
-  qbeta(shape1 = 1, shape2 = 4),
-  qnbinom(size = 10, prob = 0.75)
+Let's look at the joint distribution of the Ozone and Temperature
+
+```{r ch030-aq-joint-dist}
+p0 <- ggplot(df, aes(Temp, Ozone)) +
+  geom_point(size = 1) +
+  theme(legend.position = "none") +
+  labs(x = "Temperature")
+
+pTemp <- ggplot(df, aes(Temp)) + 
+  geom_density() +
+  theme(axis.title.x = element_blank(),
+        axis.title.y = element_blank(),
+        axis.text = element_blank(),
+        axis.ticks = element_blank())
+
+pOzone <- ggplot(df, aes(Ozone)) + 
+  geom_density() + 
+  theme(axis.title.y = element_blank(),
+        axis.title.x = element_blank(),
+        axis.ticks = element_blank(),
+        axis.text = element_blank()) +
+  coord_flip()
+
+pTemp + plot_spacer() + p0 + pOzone + 
+  plot_layout(widths = c(3,1), heights = c(1, 3))
+```
+
+We can see that not all margins are normally distributed - the ozone level is highly skewed.
+Though we don't know the true distribution of ozone levels, we can go forward assuming that it is log-normally distributed.
+
+To simulate observations from this joint distribution, we need to estimate the correlation and the marginal parameters.
+
+### Estimating Correlation
+
+To estimate the correlation, we can use `R`'s function `cor` or we can use the convenience function `bigsimr::cor_fast` which can estimate Pearson, Spearman, or Kendall correlation using the fastest methods available from other packages.
+
+```{r ch030-aq-cor, echo=TRUE}
+(rho <- cor_fast(df, method = "pearson"))
+```
+
+### Defining Marginal Distributions
+
+Next we can estimate the marginal parameters. Assuming that `Temperature` is normally distributed, it has parameters
+
+```{r ch030-aq-temp-pars, echo=TRUE}
+df %>% 
+  select(Temp) %>% 
+  summarise_all(.funs = c(mean = mean, sd = sd))
+```
+
+and assuming that `Ozone` is log-normally distributed, it has parameters
+
+```{r ch030-aq-ozone-pars, echo=TRUE}
+mle_mean <- function(x) mean(log(x))
+mle_sd <- function(x) mean( (log(x) - mean(log(x)))^2 )
+
+df %>% 
+  select(Ozone) %>% 
+  summarise_all(.funs = c(meanlog = mle_mean, sdlog = mle_sd))
+```
+
+We need a list of marginals (and their parameters), and a correlation structure (matrix).
+The marginal distributions can be built up using R's special `alist` function.
+This allows one to enter the distributions without evaluating anything (yet).
+
+```{r ch030-margins-alist, echo=TRUE}
+margins <- alist(
+  qnorm(mean = 77.9, sd = 9.49),
+  qlnorm(meanlog = 3.42, sdlog = 0.743)
 )
@@
 
-The things to point out here are that in each sublist (marginal), the first item is an unnamed character string with the R name of the distribution *without a letter prefix*.
-E.g.
-instead of `rnorm`, we pass in just `"norm"`.
-The second thing to note is that the remaining items are *named* arguments that go along with the distribution.
-A full list of built-in distributions is found in the appendix.
+Notice that we use the *quantile* function for the marginals, as that is the workhorse of the `rvec` function.
+By using an `alist`, users can also specify their own custom distributions (see the section on [creating custom margins](#custom-margins)).
 
-### Specify correlation
+It is a bit inconvenient to have to fill in the parameter values manually each time, so we provide a convenience function called `mlist` which behaves similarly to `alist`, except that it will evaluate the right hand side of argument values within the list.
 
-The next step is to define a correlation structure for the multivariate distribution.
-This correlation matrix can either come from observed data, or we can set it ourselves, or we can generate a random correlation matrix via `bigsimr::rcor`.
-Let's create a simple correlation matrix where all off-diagonal elements are 0.5.
-Since we have 3 marginals, we need a $3\times 3$ matrix.
+```{r ch030-margins-mlist, echo=TRUE}
+margins <- mlist(
+  qnorm(mean = mean(df$Temp), sd = sd(df$Temp)),
+  qlnorm(meanlog = mle_mean(df$Ozone), sdlog = mle_sd(df$Ozone))
+)
 
-```{r ch030-setCor, eval = FALSE}
-rho <- matrix(0.5, nrow = 3, ncol = 3)
-diag(rho) <- 1.0
-rho
+margins
 ```
 
-Finally we can generate a random vector with our specified marginals and correlation structure.
-The last argument, `type`, is looking to know what kind of correlation matrix it is receiving.
-Right now it can handle Pearson, Spearman, or Kendall.
+### Correlation Bounds
 
-### Generating a few random vectors
+Given a list of margins, the theoretical lower and upper correlation coefficients can be estimated.
 
-```{r ch030-rvec10, eval = FALSE}
-x <- bigsimr::rvec(10, rho = rho, margins = margins, type = "spearman")
+```{r ch030-cor-bounds, echo=TRUE}
+cor_bounds(margins = margins, type = "pearson")
 ```
 
-```{r ch030-gpuWarn, echo=FALSE}
-warning("warning.warn('No GPU/TPU found, falling back to CPU.')")
+Since our estimated correlation is within the theoretical bounds (assuming the distributions are correct), we should be able to achieve a similar correlation in the simulated data.
+
+### Simulating Multivariate Distributions
+
+Let's now simulate 10,000 observations from the joint distribution using `bigsimr::rvec`
+
+```{r ch030-sim-margins, echo=TRUE}
+x <- rvec(10000, rho, margins, "pearson")
+
+df_sim <- as.data.frame(x)
+colnames(df_sim) <- colnames(df)
 ```
 
-Taking a look at our random vector, we see that it is 10 rows and 3 columns, one column for each marginal.
+and visualize the joint distribution
 
-```{r ch030-printX, eval = FALSE}
-x
+```{r ch030-plot-sim}
+p1 <- df_sim %>%
+  ggplot(aes(Temp, Ozone)) +
+  geom_density_2d_filled() +
+  theme(legend.position = "none") +
+  labs(x = "Temperature", y = "Ozone") +
+  scale_y_continuous(limits = c(0, max(df$Ozone))) +
+  scale_x_continuous(limits = range(df$Temp))
+
+p1Temp <- ggplot(df_sim, aes(Temp)) + 
+  geom_density(alpha = 0.5, fill = "lightseagreen") +
+  theme(axis.title.x = element_blank(),
+        axis.title.y = element_blank(),
+        axis.text = element_blank(),
+        axis.ticks = element_blank())
+
+p1Ozone <- ggplot(df_sim, aes(Ozone)) + 
+  geom_density(alpha = 0.5, fill = "lightseagreen") + 
+  theme(axis.title.y = element_blank(),
+        axis.title.x = element_blank(),
+        axis.ticks = element_blank(),
+        axis.text = element_blank()) +
+  coord_flip()
+
+p1Temp + plot_spacer() + p1 + p1Ozone + 
+  plot_layout(widths = c(3,1), heights = c(1, 3))
 ```
 
-We can simulate many more samples and then check the histogram of each margin, as well as the estimated correlation between the columns.
+### Comparison to Uncorrelated Samples
 
-### Scaling up N to 1,000,000
+We can compare the bivariate distribution above to one where no correlation is taken into account.
 
-```{r ch030-rvec1e6, fig.width=7, eval = FALSE}
-x <- rvec(1e6, rho = rho, margins = margins, type = "spearman")
+```{r ch030-compare-uncorrelated}
+df_sim2 <- data.frame(
+  Temp = rnorm(10000, mean(df$Temp), sd(df$Temp)),
+  Ozone = rlnorm(10000, mle_mean(df$Ozone), mle_sd(df$Ozone))
+)
 
-par(mfrow=c(1,3))
-hist(x[,1], breaks = 30, xlab = "", main = "Normal")
-hist(x[,2], breaks = 30, xlab = "", main = "Beta")
-hist(x[,3], breaks = 30, xlab = "", main = "Negative Binomial")
+p2 <- df_sim2 %>%
+  ggplot(aes(Temp, Ozone)) +
+  geom_density_2d_filled() +
+  theme(legend.position = "none") +
+  labs(x = "Temperature", y = "Ozone") +
+  scale_y_continuous(limits = c(0, max(df$Ozone))) +
+  scale_x_continuous(limits = range(df$Temp))
+
+p2Temp <- ggplot(df_sim2, aes(Temp)) + 
+  geom_density(alpha = 0.5, fill = "lightseagreen") +
+  theme(axis.title.x = element_blank(),
+        axis.title.y = element_blank(),
+        axis.text = element_blank(),
+        axis.ticks = element_blank())
+
+p2Ozone <- ggplot(df_sim2, aes(Ozone)) + 
+  geom_density(alpha = 0.5, fill = "lightseagreen") + 
+  theme(axis.title.y = element_blank(),
+        axis.title.x = element_blank(),
+        axis.ticks = element_blank(),
+        axis.text = element_blank()) +
+  coord_flip()
+
+p2Temp + plot_spacer() + p2 + p2Ozone + 
+  plot_layout(widths = c(3,1), heights = c(1, 3))
 ```
 
-### Check the performance with fast correlation estimation
-
-```{r ch030-fastCor, eval = FALSE}
-## help( package = 'bigsimr')
-bigsimr::cor_fast(x, method = 'spearman')
-```
-
-We can see that even with 100,000 samples, the estimated correlation of the simulated data is not exactly the same as the target correlation.
-This can be explained by the fact that some correlations are simply not possible due to the discrete nature of certain distributions.
-Another possibility is that the copula algorithm is biased and needs correction.
+Notice that this uncorrelated bivariate distribution has the same marginal distributions as the correlated example above.
 
@@
 
+### Creating a custom marginal distribution {#custom-margins}
+
+Because `bigsimr` uses an `alist` to store the margins, any kind of probability distribution can be used including custom marginal distributions not provided in base R.
+This accommodation is possible because the main function `rvec` uses the inverse CDF of a margin to transform from correlated uniform margins to the target distribution.
+
+It is highly recommended that users follow R's naming convention for probability distributions - I.e.
+prefixing distributions with `r` and `q` for *random* and *quantile* respectively.
+Only the quantile function is necessary to generate multivariate data, but if one wishes to compute the theoretical correlation bounds, then one should also specify the random number generator (RNG) function for their custom distribution.
+
+For an example, R does not have any functions for the *Pareto* distribution. We can write a quantile function and RNG function and use it in `rvec`.
+
+The Pareto CDF is defined as
+
+$$
+F(x) = 1 - \left(\frac{x_m}{x}\right)^\alpha
+$$
+
+for scale $x_m > 0$ and shape $\alpha > 0$, with support $x \in \left[x_m, \infty\right)$. From the CDF, we compute the inverse CDF
+
+$$
+F^{-1}(p) = \frac{x_m}{\left(1 - p\right)^{1/\alpha}}
+$$
+
+Next we define the Pareto quantile function in R. Note that extra checks should be taken to ensure that the arguments are valid. We omit such checks for clarity.
+
+```{r ch030-define-qpareto, echo=TRUE}
+qpareto <- function(p, scale, shape) {
+  scale / (1 - p)^(1/shape)
+}
+```
+
+Writing a RNG function for our new distribution can be accomplished by calling the quantile function on the output of a random unit uniform variable.
+
+```{r ch030-define-rpareto, echo=TRUE}
+rpareto <- function(n, scale, shape) {
+  qpareto(runif(n), scale, shape)
+}
+```
+
+Now with the quantile and RNG Pareto functions, we can use the distribution in `bigsimr` just like the other built-in distributions.
+
+```{r ch030-cust-dist-example, echo=TRUE, eval=FALSE}
+margins <- alist(
+  qnorm(mean = 3.14, sd = 0.1),
+  qbeta(shape1 = 1, shape2 = 4),
+  qnbinom(size = 10, prob = 0.75),
+  qpareto(scale = 1.11, shape = 5.55)
+)
+cor_bounds(margins, "pearson")
+rho <- cor_randPD(4)
+x <- rvec(10, rho, margins)
+```
+
+### Calculating the nearest positive (semi)definite correlation matrix
+
 ### Using multicore
 
-Using multicore is easy simply specify the number of cores.
+### Utilizing the GPU
+
+The `bigsimr` packages uses Google's `jax` python library to interface with an Nvidia CUDA enabled GPU.
 
 ### Simulation-based computation of correlation bounds
-
-Using the Generate, sort, and correlate algorithm @Demirtas2011  and TAS issue
-
-- use a large number of reps
-
-```{r ch030-computeBounds, echo = FALSE, eval=FALSE}
-## Possibly bug? or configuration
-## Error: segfault from C stack overflow on my Okapi bigsimr env 
-rho_bounds <- bigsimr::computeCorBounds(margins = margins, type = 'spearman', cores = 1, reps = 1e5)
-rho_bounds
-```
 
@@
 Though `bigsimr` runs quickly, at large $d$ users may want to run jobs on a shared computing server.
-The R package `rslurm` makes it esay to run embarrasingly large parallel `rvec` calls.
+The R package `rslurm` makes it easy to run embarrassingly large parallel `rvec` calls.
 This example assumes that `bigsimr` is installed on a system with a slurm scheduler installed.

### 2020-10-23 010-background.Rmd
decide on scope of RNA examples, revise introduction
revise introduction AGS

--- 2020-09-21 010-background.Rmd
@@
 # Background and notation
+
+Formally, this is our target:
+
+We'll show that quality of simulation does not require the use of the standard Pearson correlation matrix, even for normally distributed data.
+The purpose of our method is to generate random vectors that match exactly specified any possible Kendall's $\tau$ or Spearman's $\rho$, and approximately Pearson product-moment correlation.More formally, our goal is to simulate $N$ random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with **correlated** components.
+The algorithm must scale to a large number of simulation replicates,$N$ for high-dimensional multivariate constructions (dimension $d$ in the tens or hundred thousands.
+
+Ultra High Dimensional multivariate modeling and simulation desired properties, both basic properties (BP) and scaleability properties (SP). 
+With our application in mind and seeking a fairly general-purpose algorithm with broad applications, our purposed methodology should possess the following properties (adapted from criteria from @Nik13a):
+
+* BP1: Wide range of dependence, allowing both positive and negative dependence
+* BP2: Flexible dependence, meaning that the number of bivariate marginals is (approximately) equal to the number of dependence parameters.
+* BP3: Flexible marginal modeling, generating heterogeneous data --- possibly from differing probability families.
+
+Moreover, the simulation method must scale to high dimensions:
+
+* SP1: Procedure must scale to high dimensions, computable in a reasonable amount time.
+* SP2: Procedure must scale to high dimensions while maintaining accuracy.
+
+## Motivating example: RNA-seq data from breast cancer tumor samples
+
+Simulating high-dimensional non-normal data motivates this work --- in pursuit of modeling RNA-sequencing data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
+This laboratory procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions. 
+Often researchers posit a negative binomial model (refs) as counts are often over (or under) expressed that a Poisson model would suggest.
+Yet due to inherent biological processes, gene expression data exhibits correlation (coexpression) across genes [@BE07; @Schissler2018].
+RNA-sequencing analysis has garnered much interest in the statistical literature (refs). 
+Often researchers simulate independently or from their proposed model in order to conduct Monte Carlo studies (refs). 
+An alternative strategy is to think *generatively* about the data collection process and scientific domain knowledge and design simulations with probabilistic models that reflect those processes.
+Our methodology provides another toolkit to generate data by considering how the data arise in the experimental setting.
+
+```{r ch010-processBRCA, echo=FALSE, eval=TRUE, results='none'}
+if ( !file.exists( 'data/brca99.rds' ) )  {
+    ## full workflow simulating RNA-seq data
+    allDat <- readRDS( file = "~/Downloads/complete_processed_tcga2stat_RNASeq2_with_clinical.rds" )
+    lastClinical <- which( names(allDat) == 'tumorsize' )
+    brca <- allDat[ allDat$disease == "BRCA", (lastClinical+1):ncol(allDat) ]
+    dim(brca) ## num of genes 20501
+    ## compute the median expression! avoid the 0s
+    brcaMedian <- apply(brca, 2, median)
+    ## retain top (1-probs)*100% highest expressing genes for illustration
+    myProb <- 0.99
+    cutPoint <- quantile( x = brcaMedian, probs = myProb )
+    ## genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
+    genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
+    brca <- brca[ , genesToKeep ]
+    ## ncol(brca) / 20501
+    print(d <- length(genesToKeep))
+    ## save data as rds
+    saveRDS(brca, 'data/brca99.rds')
+}
+```
+
+```{r ch010-realDataTabl, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
+## Describe and plot real data
+brca <- readRDS(file = './data/brca99.rds')
+set.seed(10192020)
+exampleGenes <- colnames(brca)[sort(sample(ncol(brca), 3))]
+saveRDS(exampleGenes, file = './data/exampleGenes.rds')
+kable(round(brca[1:4, exampleGenes]))
+## GGally::ggpairs(data = as.data.frame(brca[ ,exampleGenes] ))
+```
+
+Figure X.
+
+```{r ch010-realDataFig, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Real data for the 3 selected high-expressing genes'}
+## Describe and plot real data
+GGally::ggpairs(data = as.data.frame(brca[ ,exampleGenes] ))
+```
+
+## Measures of dependency
+
+The population correlation coefficient, commonly called the Pearson (product-moment) correlation coefficient, describes the linear association between two random variables $X$ and $Y$ and is given by
+
+\begin{equation}
+(\#eq:pearson)
+\rho(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ var(X)var(Y)\right]^{1/2}} 
+\end{equation}
+
+As @MB13 and @MK01 discuss, for a bivariate normal $(X,Y)$ random vector the Pearson correlation adequately describes the dependency between the components.
+$\rho$'s utility in non-normal or non-linear associations is lacking (@MK01).
+Rank-based (ordinal) approaches performance better in these settings, such as Spearman's (denoted $\rho_s$) and Kendall's $\tau$.
+Define
+
+\begin{equation}
+(\#eq:spearman)
+\rho_s(X,Y) = 3 \left[ P\left[ (X_1 - X_2)(Y_1-Y_3) > 0 \right] - P\left[ (X_1 - X_2)(Y_1-Y_3) < 0 \right] \right]
+\end{equation}
+
+\noindent where $(X_1, Y_1) \overset{d}{=} (X,Y), X_2 \overset{d}{=} X, Y_3 \overset{d}{=} Y$ with $X_2$ and $Y_3$ are independent of one other and of $(X_1, Y_1)$.
+For continuous marginals, the measure works well due to zero probability of ties.
 
@@
 
-## Measures of dependency
+## Matching Pearson correlations
 
-The population correlation coefficient, commonly called the Pearson (product-moment) correlation coefficient, describes the linear association between two random variables $X$ and $Y$ and is given by
+In the NORTA, it is well known that blah
 
-\begin{equation}
-(\#eq:pearson)
-\rho(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ var(X)var(Y)\right]^{1/2}} 
-\end{equation}
+## Discrete marginal considerations
 
-As @MB13 and @MK01 discuss, for a bivariate normal $(X,Y)$ random vector the Pearson correlation adequately describes the dependency between the components.
-$\rho$'s utility in non-normal or non-linear associations is lacking (@MK01).
-Rank-based (ordinal) approaches performance better in these settings, such as Spearman's (denoted $\rho_s$) and Kendall's $\tau$.
-Define
-
-\begin{equation}
-(\#eq:spearman)
-\rho_s(X,Y) = 3 \left[ P\left[ (X_1 - X_2)(Y_1-Y_3) > 0 \right] - P\left[ (X_1 - X_2)(Y_1-Y_3) < 0 \right] \right]
-\end{equation}
-
-\noindent where $(X_1, Y_1) \overset{d}{=} (X,Y), X_2 \overset{d}{=} X, Y_3 \overset{d}{=} Y$ with $X_2$ and $Y_3$ are independent of one other and of $(X_1, Y_1)$.
-For continuous marginals, the measure works well due to zero probability of ties.
+<!-- 
 For discrete marginals, however, one could perform a rescaled version of $\rho_s$ for random variables $X,Y$ with pmfs (or pdfs) $p(x)$ and $q(y)$, respectively.
@@
 
+-->
+
 ## Marginal-dependent bivariate correlation bounds
+
+Adding to the complexity is the well-known, marginal-dependent constraints on the possible bivariate correlation.
+This is a consequence of the bounds of bivariate distribution functions --- the *Frechet-Hoeffding bounds* [@Nelsen2007]). 
+Denote the pointwise upper bound as $M(y_i, y_{i^\prime})$ These bounds give strict restraints on the possible bounds and cannot be overcome through algorithm design. 
+Yet not all multivariate simulation approaches obtain these bounds [for example REF].
+Computationally, algorithms that rely on serial computation scale poorly to high dimensions (refs - hierarchy) and fail to make use of modern high-performance computing including parallelization and graphical processing unit acceleration [@Li2019gpu].
 

### 2020-10-23 index.Rmd
add some commented keywords
revise introduction AGS

@@
 --- 
-title: Simulating High-Dimensional Multivariate Data using the bigsimr R Package
+title: "Simulating High-Dimensional Multivariate Data"
+subtitle: "using the bigsimr R Package"
 author:
-  - name: Alice Anonymous
-    email: alice@example.com
-    affiliation: Some Institute of Technology
-    footnote: 1
-  - name: Bob Security
-    email: bob@example.com
-    affiliation: Another University
-  - name: Cat Memes
-    email: cat@example.com
-    affiliation: Another University
-    footnote: 2
-  - name: Derek Zoolander
-    email: derek@example.com
-    affiliation: Some Institute of Technology
-    footnote: 2
-address:
-  - code: Some Institute of Technology
-    address: Department, Street, City, State, Zip
-  - code: Another University
-    address: Department, Street, City, State, Zip
-footnote:
-  - code: 1
-    text: "Corresponding Author"
-  - code: 2
-    text: "Equal contribution"
-date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"
+  - name: Alexander D. Knudson, Tomasz J. Kozubowski, Anna K. Panorska, Juli Petereit, and Alfred G. Schissler
 bibliography:
@@
 site: bookdown::bookdown_site
-journal: "Journal of Computational And Applied Mathematics (JCAM)"
-#linenumbers: true
-#numbersections: true
+#output: distill::distill_article
 abstract: |
@@
 ---
+
+<!-- keywords simulation multivariate models high-dimensional data R packages -->
 
@@
 ## library(ggplot2); qplot(mpg, wt, data=mtcars, colour=cyl)
-## bookdown::render_book("index.Rmd", "bookdown::pdf_book")
+## bookdown::render_book("index.Rmd", "bookdown::pdf_book", output_dir = "pdf")
 ## bookdown::preview_chapter("index.Rmd")
@@
 Massive high-dimensional data sets are now commonplace in many areas of scientific inquiry.
-As new methods are developed for these data, a fundamental challenge lies in designing and conducting simulation studies to assess operating characteristics of proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness --- often in comparison to existing methods.
-Further efficient simulation empowers computational strategies (such as a parameteric bootstrap --- simulating from a hypothesized null model) to provide inference in analytically challenging settings.
+This is particularly true when simulating massive *multivariate*, *non-normal* distributions, arising naturally in many fields of study in this era of big data.
+As new methods are developed for these data, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness --- often in comparison to existing methods.
+Further, efficient simulation empowers statistical computing strategies, such as the *parameteric bootstrap* [@Rizzo2007] to simulate from a hypothesized null model, providing inference in analytically challenging settings.
 Such Monte Carlo techniques become difficult for high-dimensional data with the current existing algorithms and tools.
-This is particularly true when simulating massive multivariate, non-normal distributions, arising naturally in many fields of study in this era of big data.
 
-Along those lines, simulating high-dimensional non-normal data motivates this work --- in pursuit of modeling RNA-sequencing data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
-This laboratory procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions. 
-Often researchers posit a negative binomial model (refs) as counts are often over (or under) expressed that a Poisson model would suggest.
-Yet due to inherent biological processes, gene expression data exhibits correlation (coexpression) across genes [@BE07; @Schissler2018].
-RNA-sequencing analysis has garnered much interest in the statistical literature (refs). 
-Often researchers simulate independently or from their proposed model in order to conduct Monte Carlo studies (refs). 
-An alternative strategy is to think *generatively* about the data collection process and scientific domain knowledge and design simulations with probabilistic models that reflect those processes.
-Our methodology provides another toolkit to generate data by considering how the data arise in the experimental setting.
+As others have noted, it can be vexing to simulate dependent, non-normal/discrete data ---  even for low dimensional settings [@MB13; @XZ19].
+For continous non--normal data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula [@Nelsen2007] approaches are well-studied with flexible, robust software available [@Yan2007; @Chen2001].
+Yet these approaches do not scale in a timely fashion to high-dimensional problems (see @Li2019gpu). 
+For discrete data, multiple techniques have been proposed.
+But some schemes have major flaws such as failing to obtain the full range of possible dependencies (for example, feasible for only positive correlations @Park1996). 
+While more recent approaches [@MB13; @Xia17; @BF17] have largely rememdied this issue for low dimesional problems, the existing tools are not designed to scale to high dimensions.
 
-## Ultra High Dimensional multivariate modeling and simulation desired properties
+Another central issue lies characterizing dependency between components in the high-dimensional random vector.
+The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g. non-normal, discrete, infinite support, etc). 
+<!-- Here we propose a scheme that ignores the analytic goal --- and, instead, aims to match the empirically-derived estimated correlation data and marginal characteristics.  -->
+For normal data, the Pearson product-moment correlation describes the dependency completely.
+As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix exactly is computationally intense [@Chen2001; @Xia17]. 
+On the other hand, an analyst may consider the use of non-parametric correlation measures to better characterize monotone dependency, such as Spearman's $\rho$ and Kendall's $\tau$.
+Throughout, we'll emphasize matching these nonparametric dependency measures as our aim lies in modeling non-normal data.
 
-With our application in mind and seeking a fairly general-purpose algorithm with broad applications, our purposed methodology should possess the following properties (adapted from criteria from @Nik13a):
+In the study, we present a scalable multivariate simulation algorithms and software. 
+The crux of the method lies in the construction of a Gaussian copula, in the spirit of the NORTA procedure.
+Further, we introduce the `bigsimr` R package that provides parallelized, high-performance software.
+The algorithm design relies on useful properties of non-parameteric correlation measures, namely invariance under monotone transformation and well-known closed form relationships between dependency measures for the multivariate normal (MVN) distribution.
 
-* P1: Wide range of dependence, allowing both positive and negative dependence
-* P2: Flexible dependence, meaning that the number of bivariate marginals is (approximately) equal to the number of dependence parameters.
-* P3: Flexible marginal modeling, generating heterogeneous data --- possibly from differing probability families.
-
-Moreover, the simulation method must scale to high dimensions:
-
-* S1: Procedure must scale to high dimensions, computable in a reasonable amount time.
-* S2: Procedure must scale to high dimensions while maintaining accuracy.
-
-As others @MB13 have noted, however, simulating dependent data can be challenging.
-A central issue lies characterizing dependency between components in the high-dimensional random vector.
-The choice of correlation in typical practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g.
-non-normal, discrete, finite support, etc).
-Here we propose a scheme that ignores the analytic goal --- and, instead, aims to match the empirically-derived estimated correlation data and marginal characteristics.
-For normal data, the Pearson product-moment correlation describes the dependency completely.
-As we will see, however, simulating arbitrary random vectors with a given Pearson correlation matrix is computationally intense [@Chen2001, @Xia17], violating property **S1**.
-On the other hand, an analyst may consider the use of non-parametric correlation measures, such as Spearman's $\rho$ and Kendall's $\tau$, particularly when modeling non-normal data as in our application.
-Adding to the complexity is the well-known, marginal-dependent constraints on the possible bivariate correlation (a consequence of the bounds of bivariate distribution functions --- the *Frechet-Hoeffding bounds* [@Nelsen2007]).
-Denote the pointwise upper bound as $M(y_i, y_{i^\prime})$ These bounds give strict restraints on the possible bounds and cannot be overcome through algorithm design.
-Yet not all multivariate simulation approaches obtain these bounds [for example REF].
-Computationally, algorithms that rely on serial computation scale poorly to high dimensions (refs - hierarchy) and fail to make use of modern high-performance computing including parallelization and graphical processing unit acceleration [@Li2019gpu].
-
-To meet the desired properties in the face of the challenging setting, we present a general-purpose, scalable multivariate simulation algorithm.
-The crux of the method lies in the construction of a Gaussian copula [refs].
-The idea that many multivariate and marginally heterogeneous distribution can be constructed in such a manner is well known [refs].
-This article's contribution lies in it's application to high-dimensional data through a high-performance implementation (`bigsimr`) and speed-focused algorithm design.
-The algorithm design relies on useful properties of non-standard correlation measures, namely Kendall's $\tau$ and Spearman's $\rho$.
-We'll show that quality of simulation does not require the use of the standard Pearson correlation matrix, even for normally distributed data.
-The purpose of our method is to generate random vectors that match exactly specified any possible Kendall's $\tau$ or Spearman's $\rho$, and approximately Pearson product-moment correlation.More formally, our goal is to simulate $N$ random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with **correlated** components.
-The algorithm must scale to a large number of simulation replicates,$N$ for high-dimensional multivariate constructions (dimension $d$ in the tens or hundred thousands.
-
-The study proceeds with more details on our motivating example in RNA-sequencing breast cancer data.
-Then we describe and justify our simulation methodology, followed by extensive Monte Carlo studies under various distributional assumptions --- emphasizing normal and non-normal scenarios.
-In these studies, we evaluate the accuracy and speed of our approach in comparison to other readily available software.
-After evaluating *in silico*, we revisit our motivating
-example by employing our method and compare our simulations to empirically-derived statistics. Finally, we'll make concluding remarks regarding the method's utility and future directions.
-
-## Motivating example
-
-- RNA-seq
-- Provide a table of values
-- check the talk history
-
-```{r processBRCA, echo=FALSE, eval=TRUE, results='none'}
-if ( !file.exists( 'data/brca99.rds' ) )  {
-    ## full workflow simulating RNA-seq data
-    allDat <- readRDS( file = "~/Downloads/complete_processed_tcga2stat_RNASeq2_with_clinical.rds" )
-    lastClinical <- which( names(allDat) == 'tumorsize' )
-    brca <- allDat[ allDat$disease == "BRCA", (lastClinical+1):ncol(allDat) ]
-    dim(brca) ## num of genes 20501
-    ## compute the median expression! avoid the 0s
-    brcaMedian <- apply(brca, 2, median)
-    ## retain top (1-probs)*100% highest expressing genes for illustration
-    myProb <- 0.99
-    cutPoint <- quantile( x = brcaMedian, probs = myProb )
-    ## genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
-    genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
-    brca <- brca[ , genesToKeep ]
-    ## ncol(brca) / 20501
-    print(d <- length(genesToKeep))
-    ## save data as rds
-    saveRDS(brca, 'data/brca99.rds')
-}
-```
-
-```{r realDataTabl, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
-## Describe and plot real data
-brca <- readRDS(file = './data/brca99.rds')
-set.seed(10192020)
-exampleGenes <- colnames(brca)[sort(sample(ncol(brca), 3))]
-saveRDS(exampleGenes, file = './data/exampleGenes.rds')
-kable(round(brca[1:4, exampleGenes]))
-## GGally::ggpairs(data = as.data.frame(brca[ ,exampleGenes] ))
-```
-
-Figure X.
-
-```{r realDataFig, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Real data for the 3 selected high-expressing genes'}
-## Describe and plot real data
-GGally::ggpairs(data = as.data.frame(brca[ ,exampleGenes] ))
-```
+The study proceeds with background and notation, including a description of our motivating example in RNA-sequencing (RNA-seq) breast cancer data.
+Then we describe and justify our simulation methodology and related algorithms.
+Next, a detailed illustrative low-dimensional example of basic and advanced use of the `bigsimr` R package is provided.
+Then we proceed with extensive Monte Carlo studies under various distributional assumptions.
+After evaluating *in silico*, we revisit our high-dimensional motivating RNA-seq example and employ our methods to commonplace statistical computing tasks.
+Finally, we'll make some concluding remarks regarding the method's utility/limitations and describe potential future directions.

### 2020-10-24 index.Rmd
lightly revise the abstract/introduction

@@
 But measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments.
-Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers often resort to simulation designs that posit independence.
+Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers may resort to simulation designs that posit independence.
 This greatly diminishes insights into the empirical operating characteristics of any proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness.
-This talk introduces the bigsimr R package that provides a general, scaleable procedure to simulate high-dimensional random vectors with given marginal characteristics and dependency measures.
+This article introduces the `bigsimr` R package that provides a flexible, scaleable procedure to simulate high-dimensional random vectors with given marginal characteristics and dependency measures.
 We'll describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation matrices, and find close positive semi-definite matrices.
-Finally, we showcase the power of bigsimr by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d>1000$.
+Finally, we demonstrate the power of `bigsimr` by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d>1000$.
 ---
@@
 As new methods are developed for these data, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness --- often in comparison to existing methods.
-Further, efficient simulation empowers statistical computing strategies, such as the *parameteric bootstrap* [@Rizzo2007] to simulate from a hypothesized null model, providing inference in analytically challenging settings.
+Further, efficient simulation empowers statistical computing strategies, such as the parameteric bootstrap [@Rizzo2007] to simulate from a hypothesized null model, providing inference in analytically challenging settings.
 Such Monte Carlo techniques become difficult for high-dimensional data with the current existing algorithms and tools.
@@
 As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix exactly is computationally intense [@Chen2001; @Xia17]. 
-On the other hand, an analyst may consider the use of non-parametric correlation measures to better characterize monotone dependency, such as Spearman's $\rho$ and Kendall's $\tau$.
+On the other hand, an analyst may consider the use of nonparametric correlation measures to better characterize monotone dependency, such as Spearman's $\rho$ and Kendall's $\tau$.
 Throughout, we'll emphasize matching these nonparametric dependency measures as our aim lies in modeling non-normal data.
 
-In the study, we present a scalable multivariate simulation algorithms and software. 
+In the study, we present scalable, flexible multivariate simulation algorithms. 
 The crux of the method lies in the construction of a Gaussian copula, in the spirit of the NORTA procedure.
 Further, we introduce the `bigsimr` R package that provides parallelized, high-performance software.
-The algorithm design relies on useful properties of non-parameteric correlation measures, namely invariance under monotone transformation and well-known closed form relationships between dependency measures for the multivariate normal (MVN) distribution.
+The algorithm design relies on useful properties of nonparameteric correlation measures, namely invariance under monotone transformation and well-known closed form relationships between dependency measures for the multivariate normal (MVN) distribution.
 
@@
 After evaluating *in silico*, we revisit our high-dimensional motivating RNA-seq example and employ our methods to commonplace statistical computing tasks.
-Finally, we'll make some concluding remarks regarding the method's utility/limitations and describe potential future directions.
+Finally, we'll discuss the method's utility, limitations, and future directions.

### 2020-10-24 010-background.Rmd
Revising background

@@
-# Background and notation
+# Background and notation {#background}
 
-Formally, this is our target:
+The article presents several algorithms to work with high-dimesional multivariate data, but all `bigsimr` algorithms were originally designed to support a single task:
+to generate random vectors drawn from given marginal distributions and component-wise dependency metrics.
+Specifically, our goal is to efficiently simulate a large number, $B$, of random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with **correlated** components and hetereogeneous marginal distributions, where $d$ can be very large (possibly millions of variables).
 
-We'll show that quality of simulation does not require the use of the standard Pearson correlation matrix, even for normally distributed data.
-The purpose of our method is to generate random vectors that match exactly specified any possible Kendall's $\tau$ or Spearman's $\rho$, and approximately Pearson product-moment correlation.More formally, our goal is to simulate $N$ random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with **correlated** components.
-The algorithm must scale to a large number of simulation replicates,$N$ for high-dimensional multivariate constructions (dimension $d$ in the tens or hundred thousands.
+When designing this methodology, we developed the following properties to guide our effort.
+We divide the properties into two categories: (1) basic properties (**BP**) and "scaleability" properties (**SP**). 
+The BPs are adapted from an existing criteria due to @Nik13a.
+Our simulation strategy should allow:
 
-Ultra High Dimensional multivariate modeling and simulation desired properties, both basic properties (BP) and scaleability properties (SP). 
-With our application in mind and seeking a fairly general-purpose algorithm with broad applications, our purposed methodology should possess the following properties (adapted from criteria from @Nik13a):
-
-* BP1: Wide range of dependence, allowing both positive and negative dependence
+* BP1: Wide range of dependences, allowing both positive and negative values, and, ideally, admitting the full range of possible values.
 * BP2: Flexible dependence, meaning that the number of bivariate marginals is (approximately) equal to the number of dependence parameters.
@@
 
-Moreover, the simulation method must scale to high dimensions:
+Moreover, the simulation method must **scale** to high dimensions:
 
@@
 * SP2: Procedure must scale to high dimensions while maintaining accuracy.
+
+To fix ideas and provide example applications enabled via `bigsimr`, we the next section describes our motivating data that originally inspired the authors` interest in creating this methodology.
 
@@
 This laboratory procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions. 
-Often researchers posit a negative binomial model (refs) as counts are often over (or under) expressed that a Poisson model would suggest.
-Yet due to inherent biological processes, gene expression data exhibits correlation (coexpression) across genes [@BE07; @Schissler2018].
-RNA-sequencing analysis has garnered much interest in the statistical literature (refs). 
-Often researchers simulate independently or from their proposed model in order to conduct Monte Carlo studies (refs). 
-An alternative strategy is to think *generatively* about the data collection process and scientific domain knowledge and design simulations with probabilistic models that reflect those processes.
-Our methodology provides another toolkit to generate data by considering how the data arise in the experimental setting.
+Often researchers posit a negative-binomial (NB) model as counts are often over-dispersed that a Poisson model would suggest.
+Yet due to inherent biological processes, gene expression data exhibits correlation (co-expression) across genes [@BE07; @Schissler2018].
 
@@
 
+To do this, we take advantage of a closed form relationship [ref?] between Kendall's $\tau$ and Pearson's correlation coefficient for bivariate normal random variables:
+
+\begin{equation}
+(\#eq:convertKendall)
+r_{Pearson} = sin \left( \tau_{Kendall} \times \frac{\pi}{2} \right), 
+\end{equation}
+
+\noindent and similarly for Spearman's $\rho$ [@K58],
+
+\begin{equation}
+(\#eq:convertSpearman)
+\rho_{Pearson} = 2 \times sin \left( \rho_{Spearman} \times \frac{\pi}{6} \right).
+\end{equation}
+
+*Marginal-dependent bivariate correlation bounds*. 
+Adding to the complexity is the well-known, marginal-dependent constraints on the possible bivariate correlation.
+This is a consequence of the bounds of bivariate distribution functions --- the *Frechet-Hoeffding bounds* [@Nelsen2007]). 
+Denote the pointwise upper bound as $M(y_i, y_{i^\prime})$ These bounds give strict restraints on the possible bounds and cannot be overcome through algorithm design. 
+Yet not all multivariate simulation approaches obtain these bounds [for example REF].
+Computationally, algorithms that rely on serial computation scale poorly to high dimensions (refs - hierarchy) and fail to make use of modern high-performance computing including parallelization and graphical processing unit acceleration [@Li2019gpu].
+
+The pairwise correlation between two correlated random variables cannot in general obtain the full range of possible values, $[-1,-1]$.
+The range, called the Frechet(-Hoeffding) bounds, is a well-known function of the marginal distributions and are given by [@BF17]:
+
+\begin{equation}
+(\#eq:frechet)
+\rho^{max} = \rho \left( F^{-1}_1 (U), F^{-1}_2 (U) \right), \quad \rho^{min} = \rho \left( F^{-1}_1 (U), F^{-1}_2 (1 - U) \right)
+\end{equation}
+
+\noindent where $U$ is a uniform random variable in $(0,1)$, and $F^{-1}_1, F^{-1}_2$ are the inverse cdf of random variables $X_1$ and $X_2$, respectively.
+For discrete random variables, define $F^{-1}$ as in Equation \@ref(eq:inverseCDF).
+
+For discrete marginals, however, one could perform a rescaled version of $\rho_s$ for random variables $X,Y$ with pmfs (or pdfs) $p(x)$ and $q(y)$, respectively.
+
+\begin{equation}
+(\#eq:spearmanRescaled)
+\rho_{RS}(X,Y) = \frac{\rho_s(X,Y)}{ \left[ \left[ 1 - \sum_x p(x)^3 \right] \left[ 1 - \sum_y q(y)^3 \right] \right]^{1/2}}
+\end{equation}
+
 ## Gaussian copulas
 
-We present an general-purpose, scalable multivariate simulation algorithm.
 The crux of the method is construction of a Gaussian copula.
@@
 The Frechet bounds are well-known.
-
-## Matching Pearson correlations
-
-In the NORTA, it is well known that blah
-
-## Discrete marginal considerations
-
-<!-- 
-For discrete marginals, however, one could perform a rescaled version of $\rho_s$ for random variables $X,Y$ with pmfs (or pdfs) $p(x)$ and $q(y)$, respectively.
-
-\begin{equation}
-(\#eq:spearmanRescaled)
-\rho_{RS}(X,Y) = \frac{\rho_s(X,Y)}{ \left[ \left[ 1 - \sum_x p(x)^3 \right] \left[ 1 - \sum_y q(y)^3 \right] \right]^{1/2}}
-\end{equation}
-
-\noindent Kendall's $\tau$ is the probability of concordant pairs minus the probability of discordant pairs and is given compactly as
-
-\begin{equation}
-  (\#eq:tau)
-  \tau(X,Y) = \frac{\rho_s(X,Y)}{ \left[ \left[ 1 - \sum_x p(x)^3 \right] \left[ 1 - \sum_y q(y)^3 \right] \right]^{1/2}}
-\end{equation}
-
--->
-
-## Marginal-dependent bivariate correlation bounds
-
-Adding to the complexity is the well-known, marginal-dependent constraints on the possible bivariate correlation.
-This is a consequence of the bounds of bivariate distribution functions --- the *Frechet-Hoeffding bounds* [@Nelsen2007]). 
-Denote the pointwise upper bound as $M(y_i, y_{i^\prime})$ These bounds give strict restraints on the possible bounds and cannot be overcome through algorithm design. 
-Yet not all multivariate simulation approaches obtain these bounds [for example REF].
-Computationally, algorithms that rely on serial computation scale poorly to high dimensions (refs - hierarchy) and fail to make use of modern high-performance computing including parallelization and graphical processing unit acceleration [@Li2019gpu].
-
-The pairwise correlation between two correlated random variables cannot in general obtain the full range of possible values, $[-1,-1]$.
-The range, called the Frechet(-Hoeffding) bounds, is a well-known function of the marginal distributions and are given by [@BF17]:
-
-\begin{equation}
-(\#eq:frechet)
-\rho^{max} = \rho \left( F^{-1}_1 (U), F^{-1}_2 (U) \right), \quad \rho^{min} = \rho \left( F^{-1}_1 (U), F^{-1}_2 (1 - U) \right)
-\end{equation}
-
-\noindent where $U$ is a uniform random variable in $(0,1)$, and $F^{-1}_1, F^{-1}_2$ are the inverse cdf of random variables $X_1$ and $X_2$, respectively.
-For discrete random variables, define $F^{-1}$ as in Equation \@ref(eq:inverseCDF).

### 2020-10-24 020-simulation-algorithm.Rmd
revise introduction AGS
Revising background

@@
-# Algorithms
+# Algorithms {#algorithms}
 
-```{r previewChapter020, echo = FALSE, eval = FALSE}
+```{r ch020-preview, echo = FALSE, eval = FALSE}
 bookdown::preview_chapter('020-simulation-algorithm.Rmd')
@@
 The method is best understand as a **parallelized Gaussian copula** (see Equation \@ref(eq:gauss) to provide a high-performance NORTA (NORmal To Anything) algorithm.
+
+`mvnfast` High-performance multivariate normal simulator [@Fasiolo2016]
+
+`nortaRA` Implements exact Pearson matching (step 2 in NORTA) [@Chen2001]
 
@@
 We shall see that constructing continuous joint distributions that match a target Spearman or Kendall's correlations computes easily when employing Gaussian copulas, since this measures are invariant under the monotone transformations involved [refs].
-To do this, we take advantage of a closed form relationship [ref?] between Kendall's $\tau$ and Pearson's correlation coefficient for bivariate normal random variables:
-
-\begin{equation}
-(\#eq:convertKendall)
-r_{Pearson} = sin \left( \tau_{Kendall} \times \frac{\pi}{2} \right), 
-\end{equation}
-
-\noindent and similarly for Spearman's $\rho$ [@K58],
-
-\begin{equation}
-(\#eq:convertSpearman)
-\rho_{Pearson} = 2 \times sin \left( \rho_{Spearman} \times \frac{\pi}{6} \right).
-\end{equation}
 
@@
 
-## Random vector generation `bigsimr::rvec` algorithm
+## Random vector generation `bigsimr::rvec` algorithm {#rand-vec-gen}
 

### 2020-10-24 040-monte-carlo-evaluation.Rmd
Revising background

@@
-# Monte Carlo evaluations
+# Monte Carlo evaluations {#simulations}
 
@@
 
-## Simulation I: Bivariate simulation using rank-based correlation
+## Bivariate experiments
 
-### Bivariate Normal
-
-* Let's simulate a bivariate normal and check our correlation matching
-  performance as N increases.
-* Here we have BVN( $\mu_1 = \mu_2 = 10, \rho_{type}$ )
-* We vary $\rho$ across the entire possible range of correlations for each
-  correlation type.
+*Bivariate Normal*.
+Let's simulate a bivariate normal and check our correlation matching performance as N increases.
+Here we have BVN( $\mu_1 = \mu_2 = 10, \rho_{type}$ ).
+We vary $\rho$ across the entire possible range of correlations for each correlation type.
 
@@
 
-## Bivariate Gamma
-
-* Similarly, let's check the performance for a non-symmetric continuous distribution: a
-  standard (rate =1) bivariate gamma.
-* Here we have a Bivariate Gamma with $shape_1 = shape_2 = 10, \rho_{type}$.
-* We vary $\rho$ across the entire possible range of correlations for each
-  correlation type.
+*Bivariate Gamma*. 
+Similarly, let's check the performance for a non-symmetric continuous distribution: a standard (rate =1) bivariate gamma.
+Here we have a Bivariate Gamma with $shape_1 = shape_2 = 10, \rho_{type}$.
+We vary $\rho$ across the entire possible range of correlations for each correlation type.
 
@@
 
-## Bivariate Negative Binomial
-
-* Let's check the performance for a discrete distribution: a bivariate negative binomial
-* Here we have Bivariate Negative Binomial ( $prob_1 = prob_2 = 0.5, size_1 = size_2 = 4,\rho_{type}$ )
-* We vary $\rho$ across the entire possible range of correlations for each
-  correlation type.
+*Bivariate Negative Binomial*. 
+Let's check the performance for a discrete distribution: a bivariate negative binomial.
+Here we have Bivariate Negative Binomial ( $prob_1 = prob_2 = 0.5, size_1 = size_2 = 4,\rho_{type}$ ).
+We vary $\rho$ across the entire possible range of correlations for each correlation type.
 

### 2020-10-25 010-background.Rmd
Revise motivating example

@@
 
-## Motivating example: RNA-seq data from breast cancer tumor samples
+## Motivating example: RNA-seq data
 
-Simulating high-dimensional non-normal data motivates this work --- in pursuit of modeling RNA-sequencing data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
-This laboratory procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions. 
-Often researchers posit a negative-binomial (NB) model as counts are often over-dispersed that a Poisson model would suggest.
+Simulating high-dimensional, non-normal data motivates this work --- in pursuit of modeling RNA-sequencing data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
+The RNA-seq data-generating process involves counting how often a particular messenger RNA (mRNA) is expressed in a biological sample.
+RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data.
+In humans, this results in count data corresponding to over 20,000 genes (coding genomic regions) or even over 77,000 isoforms when alternating spliced mRNA are counted.
 Yet due to inherent biological processes, gene expression data exhibits correlation (co-expression) across genes [@BE07; @Schissler2018].
+
+Specifically, we'll illustrate our methodology using the Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas; TCGA; see Acknowledgements).
+For simplicity, we begin by filtering to retain the top `r (1-myProb)*100`% highest expressing genes of the 20,501 gene measurements from $N=`r nrow(brca)`$ patients' tumor samples, resulting in $d=`r d`$ genes.
+Further, to illustrate our methodology's flexible/robust simulation scheme and better align with the actual data-generating process, we remove the statistical adjustment for read alignment ambiguity (RNA-seq by Expectation Maximization; RSEM; @Li2011c) by simple rounding to integer values.
+Table \@ref(tab:ch010-realDataTab) displays counts for three selected genes for the first 5 patients' breast tumor samples.
 
@@
 
-```{r ch010-realDataTabl, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
+```{r ch010-realDataTab, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
 ## Describe and plot real data
@@
 saveRDS(exampleGenes, file = './data/exampleGenes.rds')
-kable(round(brca[1:4, exampleGenes]))
+kable(round(head( brca[, exampleGenes], n = 5)),
+      booktabs = TRUE,
+      caption = 'Counts of 3 high-expressing genes for the first 5 observations of the BRCA dataset.'
+)
 ## GGally::ggpairs(data = as.data.frame(brca[ ,exampleGenes] ))
@@
 
-Figure X.
+Figure \@ref(fig:ch010-realDataFig) plots the marginal distributions and estimated Spearman's correlations (see Equation \@ref(eq:spearman) below).
 
-```{r ch010-realDataFig, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Real data for the 3 selected high-expressing genes'}
+```{r ch010-realDataFig, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's $rho$ for three example genes.
+The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations.
+Modeling these data motivate our simulation methodology."}
 ## Describe and plot real data
@@
 
-Sklar's Theorem (ref) guarantee's that any random vector with computational feasible inverse CDFs (P4 above) and obtainable correlation matrix (within the Frechet bounds) can be obtained via transformations involving copula functions.
+Sklar's Theorem (ref) guarantees that given inverse CDFs (quantile) and obtainable correlation matrix (within the Frechet bounds) can be obtained via transformations involving copula functions. 
 Namely, to simulate a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$, we can construct a Gaussian copula ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$.
-When an $Y_i$ is discrete, some care must be taken to define $F_i$.
-Letting
+When an $Y_i$ is discrete, some care must be taken to define $F_i$. Letting
 
@@
 
-Simulation of a general multivariate random vector ${\bf T}$ with margins $F_i$ based on this copula is quite simple.
-Ensuring that one obtains a certain dependence among the marginal distributions, however, proves challenging in our proposed Gaussian-copula-based scheme.
-Two core issues: 1) marginal characteristics induce bounds on the possible bivariate correlations and 2) monotone transformations deform the Pearson correlation which no closed form expression exists in general (ref?
-Chen2001).
-The Frechet bounds are well-known.
+Simulation of a multivariate random vector ${\bf T}$ with margins $F_i$ based on this copula

### 2020-10-25 050-examples.Rmd
Revise motivating example

@@
 
-The RNA-seq data generating process involves counting how often a particular messenger RNA (mRNA) is expressed in a biological sample.
-Since this is a counting processing with no upper bound, many modeling approaches discrete random variables with infinite support.
-Often the counts exhibit over-dispersion and so the negative binomial arises as a sensible model for the expression levels (\emph{gene counts}).
-Moreover, the counts are correlated (co-expressed) and cannot be assumed to behave independently.
-RNA-seq platforms quantify the entire transcriptome in one experimental run, resulting in high dimensional data.
-In humans, this results in count data corresponding to over 20,000 genes (coding genomic regions) or even over 77,000 isoforms when alternating spliced mRNA are counted.
-	This suggests simulating high-dimensional multivariate NB with heterogeneous marginals would be useful tool in the development and evaluation of RNA-seq analytics.
-
 ## Simulating High-Dimensional RNA-seq data
@@
 Our goal is to replicate the structure of a breast cancer data set (BRCA data set from The Cancer Genome Atlas).
-For simplicity of illustration, we begin by filtering to retain the top 5\% highest expressing genes of the 20,501 gene measurements from $N=1212$ patients' tumor samples, resulting in $d=1026$ genes.
+All these genes exhibit over-dispersion and, so, we proceed to estimate the NB parameters $(r_i, p_i), i=1,\ldots,d$ to determine the target marginal PMFs $g_i(y_i)$ (via method of moments).
+Often researchers posit a negative binomial (NB) model as counts are often over-dispersed that a Poisson model would suggest.
+This suggests simulating high-dimensional multivariate NB (MVB) with heterogeneous marginals would be useful tool in the development and evaluation of RNA-seq analytics.
+This procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions. 
+Notably, the $\hat{p}_i's$ are small --- ranging in $[3.934 \times 10^{-6} , 1.217 \times 10^{-2}]$.
+To complete the simulation algorithm inputs, we estimate the Pearson correlation matrix $\bf{R_Y}$ and set that as the target correlation.
 

### 2020-10-26 010-background.Rmd
draft background

@@
-# Background and notation {#background}
+# Background {#background}
 
@@
 
-Specifically, we'll illustrate our methodology using the Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas; TCGA; see Acknowledgements).
-For simplicity, we begin by filtering to retain the top `r (1-myProb)*100`% highest expressing genes of the 20,501 gene measurements from $N=`r nrow(brca)`$ patients' tumor samples, resulting in $d=`r d`$ genes.
-Further, to illustrate our methodology's flexible/robust simulation scheme and better align with the actual data-generating process, we remove the statistical adjustment for read alignment ambiguity (RNA-seq by Expectation Maximization; RSEM; @Li2011c) by simple rounding to integer values.
-Table \@ref(tab:ch010-realDataTab) displays counts for three selected genes for the first 5 patients' breast tumor samples.
-
 ```{r ch010-processBRCA, echo=FALSE, eval=TRUE, results='none'}
@@
 ```
+
+Specifically, we'll illustrate our methodology using the Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas; TCGA; see Acknowledgements).
+For simplicity, we begin by filtering to retain the top `r (1-myProb)*100`% highest expressing genes of the 20,501 gene measurements from $N=`r nrow(brca)`$ patients' tumor samples, resulting in $d=`r d`$ genes.
+This results in `r choose(d,2)` pairwise dependencies to characterize and simulate.
+Further, to illustrate our methodology's flexible/robust simulation scheme and better align with the actual data-generating process, we remove the statistical adjustment for read alignment ambiguity (RNA-seq by Expectation Maximization; RSEM; @Li2011c) by simple rounding to integer values.
+Table \@ref(tab:ch010-realDataTab) displays counts for three selected genes for the first 5 patients' breast tumor samples.
 
@@
 
-The population correlation coefficient, commonly called the Pearson (product-moment) correlation coefficient, describes the linear association between two random variables $X$ and $Y$ and is given by
+A central issue when describing multivariate distributinos is the choose of quantatitive description.
+The most widely-known is the Pearson (product-moment) correlation coefficient that describes the linear association between two random variables $X$ and $Y$ and is given by
 
@@
 (\#eq:pearson)
-\rho(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ var(X)var(Y)\right]^{1/2}} 
+\rho_P(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ var(X)var(Y)\right]^{1/2}}.
 \end{equation}
 
-As @MB13 and @MK01 discuss, for a bivariate normal $(X,Y)$ random vector the Pearson correlation adequately describes the dependency between the components.
-$\rho$'s utility in non-normal or non-linear associations is lacking (@MK01).
-Rank-based (ordinal) approaches performance better in these settings, such as Spearman's (denoted $\rho_s$) and Kendall's $\tau$.
-Define
+As @MB13 and @MK01 discuss, for a bivariate normal $(X,Y)$ random vector the Pearson correlation completely describes the dependency between the components. 
+For non-normal marginals and monotone (possibly non-linear) correlation patterns, $\rho_P$ suffers some drawbacks and may mislead or fail to capture important relationships (@MK01).
+Alternatively, in these settings, analysts often prefer rank-based correlatino measures to describe the degree of monotonic association.
+
+Two nonparameteric, rank-based measures common in practice are Spearman's correlation (denoted $\rho_S$) and Kendall's $\tau$.
+
+Formally, define
 
@@
 (\#eq:spearman)
-\rho_s(X,Y) = 3 \left[ P\left[ (X_1 - X_2)(Y_1-Y_3) > 0 \right] - P\left[ (X_1 - X_2)(Y_1-Y_3) < 0 \right] \right]
+\rho_S(X,Y) = 3 \left[ P\left[ (X_1 - X_2)(Y_1-Y_3) > 0 \right] - P\left[ (X_1 - X_2)(Y_1-Y_3) < 0 \right] \right]
 \end{equation}
@@
 \noindent where $(X_1, Y_1) \overset{d}{=} (X,Y), X_2 \overset{d}{=} X, Y_3 \overset{d}{=} Y$ with $X_2$ and $Y_3$ are independent of one other and of $(X_1, Y_1)$. 
-For continuous marginals, the measure works well due to zero probability of ties.
+Spearman's $\rho$ has an appealing correspondence as the Pearson's correlaiton coefficient on *ranks* of the values, thereby captures nonlinear yet monotone relationships.
 
-To do this, we take advantage of a closed form relationship [ref?] between Kendall's $\tau$ and Pearson's correlation coefficient for bivariate normal random variables:
+Kendall's $\tau$, on the other hand, is the difference in probabilities of concordant and discordant pairs of observations $(X_i, Y_i)$ and $(X_j, Y_j)$.
+By concordance we mean that orderings have the same direction if $X_i < X_j$, then $Y_i < Y_j$, and is determinded by the ranks of the values, not the values themselves (as in $\rho_P$).
+
+Both $\tau$ and $\rho_S$ are **invariant to under monotone transformations** of the underlying random variates.
+As we will describe more fully in the [Algorithms](#algorithms) section, this property is essential to our scheme to satistify properties SP1 and SP2, above.
+
+This is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficent, as the marginal distributions $F_i$ are fundamental to their caclutain.
+But for bivariate normal vectors, however, the correspondence is well-known:
 
@@
 (\#eq:convertKendall)
-r_{Pearson} = sin \left( \tau_{Kendall} \times \frac{\pi}{2} \right), 
+\rho_{P} = sin \left( \tau \times \frac{\pi}{2} \right), 
 \end{equation}
@@
 (\#eq:convertSpearman)
-\rho_{Pearson} = 2 \times sin \left( \rho_{Spearman} \times \frac{\pi}{6} \right).
+\rho_P = 2 \times sin \left( \rho_S \times \frac{\pi}{6} \right).
+\end{equation}
+
+These facts are also critical in our simulation algorithm to broaden the dependency measures supported by `bigsimr`.
+
+*Discrete marginal considerations*.
+Spearman's correlation for discrete marginal suffers some issues due to the nonzero probability of ties (for example, the Spearman's correlation of a discrete-valued random variable $X$ could be less than 1; @MB13).
+One remedy for this issue is to rescale Equation \@ref(eq:spearman). 
+For two random variables $X,Y$ with pmfs (or pdfs) $p(x)$ and $q(y)$, respectively, define the rescaled Spearman's correlation as
+
+\begin{equation}
+(\#eq:spearmanRescaled)
+\rho_{RS}(X,Y) = \frac{\rho_s(X,Y)}{ \left[ \left[ 1 - \sum_x p(x)^3 \right] \left[ 1 - \sum_y q(y)^3 \right] \right]^{1/2}}.
+\end{equation}
+
+Noting that with continuous marginals the rescaling returns $\rho_S$.
+For discrete marginals with large or infinite support, computing adjustment factors $\sum_x p(x)^3$, $\sum_y p(y)^3$ over all large number of pairs becomes expensive (often violating desired property *SP1*).
+And further the infinite sums must be approximately for count-valued data and potentially violating desired property *SP2*.
+
+Also we'll note that for a discrete random variable $Y_i$, some care must be taken to define the quantile function $F_{i}^{-1}$.
+Let
+
+\begin{equation}
+F_{i}^{-1} = inf\{y:F_{i}(y) \geq u \}.
+(\#eq:inverseCDF)
 \end{equation}
@@
 *Marginal-dependent bivariate correlation bounds*. 
-Adding to the complexity is the well-known, marginal-dependent constraints on the possible bivariate correlation.
-This is a consequence of the bounds of bivariate distribution functions --- the *Frechet-Hoeffding bounds* [@Nelsen2007]). 
-Denote the pointwise upper bound as $M(y_i, y_{i^\prime})$ These bounds give strict restraints on the possible bounds and cannot be overcome through algorithm design. 
-Yet not all multivariate simulation approaches obtain these bounds [for example REF].
-Computationally, algorithms that rely on serial computation scale poorly to high dimensions (refs - hierarchy) and fail to make use of modern high-performance computing including parallelization and graphical processing unit acceleration [@Li2019gpu].
-
-The pairwise correlation between two correlated random variables cannot in general obtain the full range of possible values, $[-1,-1]$.
-The range, called the Frechet(-Hoeffding) bounds, is a well-known function of the marginal distributions and are given by [@BF17]:
+Given two marginal distributions, the $\rho_P$ is not free to vary the entire range of possible correlations, $[-1,1]$.
+These so-called *Frechet-Hoeffding bounds* are well-studied (for example, see @Nelsen2007 and @BF17).
+This situatino gives strict restraints on the possible bounds and cannot be overcome through algorithm design.
+In general the bounds are given by
 
@@
 (\#eq:frechet)
-\rho^{max} = \rho \left( F^{-1}_1 (U), F^{-1}_2 (U) \right), \quad \rho^{min} = \rho \left( F^{-1}_1 (U), F^{-1}_2 (1 - U) \right)
+\rho_P^{max} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (U) \right), \quad \rho_P^{min} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (1 - U) \right)
 \end{equation}
@@
 
-For discrete marginals, however, one could perform a rescaled version of $\rho_s$ for random variables $X,Y$ with pmfs (or pdfs) $p(x)$ and $q(y)$, respectively.
-
-\begin{equation}
-(\#eq:spearmanRescaled)
-\rho_{RS}(X,Y) = \frac{\rho_s(X,Y)}{ \left[ \left[ 1 - \sum_x p(x)^3 \right] \left[ 1 - \sum_y q(y)^3 \right] \right]^{1/2}}
-\end{equation}
-
 ## Gaussian copulas
 
-The crux of the method is construction of a Gaussian copula.
-This idea is well known [refs].
-We chose a Gaussian copula among the many available copula functions primarily for two reasons: 1) the computationally convient closed expression to convert among the most common dependency measures when the margins are Gaussian and 2) the availability of high-performance multivariate normal simulators in R ('mvnfast' ref).
+There is a strong connection of our simulatino strategy to Gaussian *copulas* (for a technical introduction see @Nelsen2007).
+A copula is a distribution function on $[0,1]^d$ describing a random vector with standard uniform marginals.
+This definition provides a powerful, natural way characterize joint probability structure and the study of copulas is important in statistical theory and practice.
 
-A copula is a distribution function on $[0,1]^d$ describing a random vector with standard uniform marginals.
-Moreover, for any random vector ${\bf X}=(X_1, \ldots, X_d)$ with cumulative distribution function (CDF) $F$ and marginal CDFs $F_i$ there is a copula function
+For  any random vector ${\bf X}=(X_1, \ldots, X_d)$ with cumulative distribution function (CDF) $F$ and marginal CDFs $F_i$ there is a copula function 
 $C(u_1, \ldots, u_d)$ so that
@@
 A Gaussian copula is the case where all marginal CDFs $F_i$ are the standard normal cdf, $\Phi$. 
-The Gaussian copula is the one that corresponds to a multivariate normal distribution with standard normal marginal distributions and covariance matrix ${\bf R}$.
-(Since the marginals are standard normal, this ${\bf R}$ is also the correlation matrix).
+This representation corresponds to a multivariate normal distribution with standard normal marginal distributions and covariance matrix ${\bf R}$.
+But since the marginals are standardized to have unit variance, this ${\bf R}$ is a Pearson correlation matrix. 
 If $F_{{\bf R}}$ is the CDF of such multivariate normal distribution, then the corresponding Gaussian copula $C_{{\bf R}}$ is defined through
@@
 
-Sklar's Theorem (ref) guarantees that given inverse CDFs (quantile) and obtainable correlation matrix (within the Frechet bounds) can be obtained via transformations involving copula functions. 
-Namely, to simulate a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$, we can construct a Gaussian copula ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$.
-When an $Y_i$ is discrete, some care must be taken to define $F_i$. Letting
-
-\begin{equation}
-F_{i}^{-1} = inf\{y:F_{i}(y) \geq u \}
-(\#eq:inverseCDF)
-\end{equation}
-
-\noindent ensures that $Y_i \sim F_i$.
-
-Simulation of a multivariate random vector ${\bf T}$ with margins $F_i$ based on this copula
+Sklar's Theorem [@Sklar1959; @Ubeda-Flores2017] guarantees that given inverse CDFs $F_i^{-1}$s and a valid correlation matrix (within the Frechet bounds) *any* random vector can be obtained via transformations involving copula functions.
+For example, using Gaussian copulas, we can construct a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)$ with $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$, via ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$ povides $Y_i \sim F_i, \, \forall \, i$ .

### 2020-10-27 010-background.Rmd
revise Background

@@
 The article presents several algorithms to work with high-dimesional multivariate data, but all `bigsimr` algorithms were originally designed to support a single task:
-to generate random vectors drawn from given marginal distributions and component-wise dependency metrics.
-Specifically, our goal is to efficiently simulate a large number, $B$, of random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with **correlated** components and hetereogeneous marginal distributions, where $d$ can be very large (possibly millions of variables).
+to generate random vectors drawn from multivariate probability distribution with given marginal distributions and component-wise dependency metrics.
+Specifically, our goal is to efficiently simulate a large number, $B$, of random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with **correlated** components and hetereogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i$, and $d$ can be very large.
 
@@
 
-* BP1: Wide range of dependences, allowing both positive and negative values, and, ideally, admitting the full range of possible values.
-* BP2: Flexible dependence, meaning that the number of bivariate marginals is (approximately) equal to the number of dependence parameters.
+* BP1: A wide range of dependences, allowing both positive and negative values, and, ideally, admitting the full range of possible values.
+* BP2: Flexible dependence, meaning that the number of bivariate marginals can be equal to the number of dependence parameters.
 * BP3: Flexible marginal modeling, generating heterogeneous data --- possibly from differing probability families.
@@
 
-To fix ideas and provide example applications enabled via `bigsimr`, we the next section describes our motivating data that originally inspired the authors` interest in creating this methodology.
+To fix ideas and provide example applications enabled via `bigsimr`, the next section describes a motivating data set that originally inspired the authors` interest in developing this methodology.
 
@@
 
-Simulating high-dimensional, non-normal data motivates this work --- in pursuit of modeling RNA-sequencing data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
+Simulating high-dimensional, non-normal, correleated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
 The RNA-seq data-generating process involves counting how often a particular messenger RNA (mRNA) is expressed in a biological sample.
 RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data.
-In humans, this results in count data corresponding to over 20,000 genes (coding genomic regions) or even over 77,000 isoforms when alternating spliced mRNA are counted.
+This results in count data corresponding to over 20,000 genes (coding genomic regions) or even over 77,000 isoforms when alternating spliced mRNA are counted (for human derived samples).
 Yet due to inherent biological processes, gene expression data exhibits correlation (co-expression) across genes [@BE07; @Schissler2018].
@@
 
-Specifically, we'll illustrate our methodology using the Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas; TCGA; see Acknowledgements).
-For simplicity, we begin by filtering to retain the top `r (1-myProb)*100`% highest expressing genes of the 20,501 gene measurements from $N=`r nrow(brca)`$ patients' tumor samples, resulting in $d=`r d`$ genes.
-This results in `r choose(d,2)` pairwise dependencies to characterize and simulate.
-Further, to illustrate our methodology's flexible/robust simulation scheme and better align with the actual data-generating process, we remove the statistical adjustment for read alignment ambiguity (RNA-seq by Expectation Maximization; RSEM; @Li2011c) by simple rounding to integer values.
-Table \@ref(tab:ch010-realDataTab) displays counts for three selected genes for the first 5 patients' breast tumor samples.
+Specifically, we'll illustrate our methodology using the Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgements).
+For simplicity, we begin by filtering to retain the top `r (1-myProb)*100`% highest expressing (in terms of median expression) genes of the 20,501 gene measurements from $N=`r nrow(brca)`$ patients' tumor samples, resulting in $d=`r d`$ genes.
+This results in `r choose(d,2)` pairwise dependencies among the bivariate marginals.
+To illustrate our methodology's flexible/robust simulation scheme and better align with the actual data-generating process, we remove the statistical adjustment for read alignment ambiguity (RNA-seq by Expectation Maximization; RSEM; @Li2011c) by simple rounding to integer values.
+Table \@ref(tab:ch010-realDataTab) displays counts for three selected genes for the first five patients' breast tumor samples.
 
@@
       booktabs = TRUE,
-      caption = 'Counts of 3 high-expressing genes for the first 5 observations of the BRCA dataset.'
+      caption = 'mRNA counts for three selected high-expressing genes from the first five observations of the BRCA data set.'
 )
@@
 ## Describe and plot real data
+## NEED TO CHANGE TO SPEARMANS
 GGally::ggpairs(data = as.data.frame(brca[ ,exampleGenes] ))
@@
 
-A central issue when describing multivariate distributinos is the choose of quantatitive description.
-The most widely-known is the Pearson (product-moment) correlation coefficient that describes the linear association between two random variables $X$ and $Y$ and is given by
+In multivariate analysis, an analyst must select a metric to quantify dependency.
+The most widely-known is the Pearson (product-moment) correlation coefficient that describes the linear association between two random variables $X$ and $Y$, and, it is given by
 
@@
 
-As @MB13 and @MK01 discuss, for a bivariate normal $(X,Y)$ random vector the Pearson correlation completely describes the dependency between the components. 
-For non-normal marginals and monotone (possibly non-linear) correlation patterns, $\rho_P$ suffers some drawbacks and may mislead or fail to capture important relationships (@MK01).
-Alternatively, in these settings, analysts often prefer rank-based correlatino measures to describe the degree of monotonic association.
+As @MB13 and @MK01 discuss, for a bivariate normal $(X,Y)$ random vector, the Pearson correlation completely describes the dependency between the components. 
+For non-normal marginals with monotone correlation patterns, $\rho_P$ suffers some drawbacks and may mislead or fail to capture important relationships (@MK01).
+Alternatively, in these settings, analysts often prefer rank-based correlation measures to describe the degree of monotonic association.
 
 Two nonparameteric, rank-based measures common in practice are Spearman's correlation (denoted $\rho_S$) and Kendall's $\tau$. 
-
 Formally, define
@@
 \noindent where $(X_1, Y_1) \overset{d}{=} (X,Y), X_2 \overset{d}{=} X, Y_3 \overset{d}{=} Y$ with $X_2$ and $Y_3$ are independent of one other and of $(X_1, Y_1)$. 
-Spearman's $\rho$ has an appealing correspondence as the Pearson's correlaiton coefficient on *ranks* of the values, thereby captures nonlinear yet monotone relationships.
+Spearman's $\rho_S$ has an appealing correspondence as the Pearson's correlaiton coefficient on *ranks* of the values, thereby captures nonlinear yet monotone relationships.
 
 Kendall's $\tau$, on the other hand, is the difference in probabilities of concordant and discordant pairs of observations $(X_i, Y_i)$ and $(X_j, Y_j)$.
-By concordance we mean that orderings have the same direction if $X_i < X_j$, then $Y_i < Y_j$, and is determinded by the ranks of the values, not the values themselves (as in $\rho_P$).
+By concordance we mean that orderings have the same direction (e.g., if $X_i < X_j$, then $Y_i < Y_j$) and is determinded by the ranks of the values, not the values themselves.
 
 Both $\tau$ and $\rho_S$ are **invariant to under monotone transformations** of the underlying random variates.
-As we will describe more fully in the [Algorithms](#algorithms) section, this property is essential to our scheme to satistify properties SP1 and SP2, above.
+As we will describe more fully in the [Algorithms](#algorithms) section, this property is preferred in our scheme to satistify properties SP1 and SP2, above.
 
-This is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficent, as the marginal distributions $F_i$ are fundamental to their caclutain.
-But for bivariate normal vectors, however, the correspondence is well-known:
+*Correspondence among Pearson, Spearman, $\tau$ correlations*.
+This is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficent, as the marginal distributions $F_i$ are intrinstic in their calculation.
+But for **bivariate normal vectors**, however, the correspondence is well-known:
 
@@
 
-These facts are also critical in our simulation algorithm to broaden the dependency measures supported by `bigsimr`.
+These facts are also critical in our simulation algorithm to broaden the dependency measures supported by `bigsimr`, in a computationally effective manner.
 
 *Discrete marginal considerations*.
-Spearman's correlation for discrete marginal suffers some issues due to the nonzero probability of ties (for example, the Spearman's correlation of a discrete-valued random variable $X$ could be less than 1; @MB13).
+Spearman's correlation for discrete marginal suffers some issues due to the nonzero probability of ties (for example, the Spearman's correlation of a discrete-valued random variable $X$ with itself could be less than 1; @MB13).
 One remedy for this issue is to rescale Equation \@ref(eq:spearman). 
-For two random variables $X,Y$ with pmfs (or pdfs) $p(x)$ and $q(y)$, respectively, define the rescaled Spearman's correlation as
+For two random variables $X,Y$ with probability mass functions (PMFs) or probability densities functions (PDFs) $p(x)$ and $q(y)$, respectively, define the rescaled Spearman's correlation as
 
@@
 
-Noting that with continuous marginals the rescaling returns $\rho_S$.
-For discrete marginals with large or infinite support, computing adjustment factors $\sum_x p(x)^3$, $\sum_y p(y)^3$ over all large number of pairs becomes expensive (often violating desired property *SP1*).
-And further the infinite sums must be approximately for count-valued data and potentially violating desired property *SP2*.
+Note that with continuous marginals the rescaling returns $\rho_S$.
+For discrete marginals with large or infinite support, computing the adjustment factors $\sum_x p(x)^3$, $\sum_y p(y)^3$ over all large number of pairs becomes expensive (often violating desired property SP1).
+And further the infinite sums must be approximately for count-valued data and potentially violating desired property SP2.
 
@@
 *Marginal-dependent bivariate correlation bounds*. 
-Given two marginal distributions, the $\rho_P$ is not free to vary the entire range of possible correlations, $[-1,1]$.
-These so-called *Frechet-Hoeffding bounds* are well-studied (for example, see @Nelsen2007 and @BF17).
-This situatino gives strict restraints on the possible bounds and cannot be overcome through algorithm design.
+Given two marginal distributions, $\rho_P$ is not free to vary the entire range of possible correlations $[-1,1]$.
+The so-called *Frechet-Hoeffding bounds* are well-studied (for example, see @Nelsen2007 and @BF17).
+This situation gives strict restraints on the possible bounds and cannot be overcome through algorithm design.
 In general the bounds are given by
@@
 
-There is a strong connection of our simulatino strategy to Gaussian *copulas* (for a technical introduction see @Nelsen2007).
-A copula is a distribution function on $[0,1]^d$ describing a random vector with standard uniform marginals.
-This definition provides a powerful, natural way characterize joint probability structure and the study of copulas is important in statistical theory and practice.
+There is a strong connection of our simulation strategy to Gaussian **copulas** (for a technical introduction see @Nelsen2007).
+A copula is a distribution function on $[0,1]^d$ that describes a multivariate probability distribution with standard uniform marginals.
+This definition provides a powerful, natural way characterize joint probability structure.
+Consequently, the study of copulas is an important and active area of statistical theory and practice.
 
-For  any random vector ${\bf X}=(X_1, \ldots, X_d)$ with cumulative distribution function (CDF) $F$ and marginal CDFs $F_i$ there is a copula function 
+For any random vector ${\bf X}=(X_1, \ldots, X_d)$ with CDF $F$ and marginal CDFs $F_i$ there is a copula function 
 $C(u_1, \ldots, u_d)$ so that
@@
 A Gaussian copula is the case where all marginal CDFs $F_i$ are the standard normal cdf, $\Phi$. 
-This representation corresponds to a multivariate normal distribution with standard normal marginal distributions and covariance matrix ${\bf R}$.
-But since the marginals are standardized to have unit variance, this ${\bf R}$ is a Pearson correlation matrix. 
-If $F_{{\bf R}}$ is the CDF of such multivariate normal distribution, then the corresponding Gaussian copula $C_{{\bf R}}$ is defined through
+This representation corresponds to a multivariate normal distribution with standard normal marginal distributions and covariance matrix ${\bf R_P}$.
+But since the marginals are standardized to have unit variance, this ${\bf R_P}$ is a Pearson correlation matrix. 
+If $F_{{\bf R}}$ is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula $C_{{\bf R}}$ is defined through
 
@@
 where $\Phi(\cdot)$ is the standard normal CDF.
-Note that the copula $C_{{\bf R}}$ is simply the CDF of the random vector $(\Phi(X_1), \ldots, \Phi(X_d))$, where  $(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R})$.
+Note that the copula $C_{{\bf R}}$ is the familar multivariate normal CDF of the random vector $(\Phi(X_1), \ldots, \Phi(X_d))$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})$.
 

### 2020-10-27 020-simulation-algorithm.Rmd
draft background
revise algorithms

@@
 
-This section describes the method for simulating a random vector $\bf Y$ with $Y_i$ components for $i=1,2,\ldots,d$. Each $Y_i$ has a specified marginal
-distribution function $F_i$ and its inverse.
-To characterize dependency, every pair $(Y_i, Y_j)$ has either a specified Pearson correlation \@ref(eq:pearson), (rescaled) Spearman correlation \, or Kendall's $\tau$ ).
-The method only approximately matches the Pearson correlation in general, whereas the rank-based methods are exact.
+This section describes our methods involved in simulating a random vector $\bf Y$ with $Y_i$ components for $i=1,2,\ldots,d$.
+Each $Y_i$ has a specified marginal CDF $F_i$ and its inverse $F^{-1}_i$.
+To characterize dependency, every pair $(Y_i, Y_j)$ has a given Pearson correlation $\rho_P$, (rescaled) Spearman correlation $\rho_S$, and/or Kendall's $\tau$.
+The method is best understand as a **high-performance Gaussian copula** (Equation \@ref(eq:gauss)) providing a high-dimensianl NORTA-inspired algorithm.
 
-The method is best understand as a **parallelized Gaussian copula** (see Equation \@ref(eq:gauss) to provide a high-performance NORTA (NORmal To Anything) algorithm.
+## NORmal To Anything (NORTA)
 
-`mvnfast` High-performance multivariate normal simulator [@Fasiolo2016]
-
-`nortaRA` Implements exact Pearson matching (step 2 in NORTA) [@Chen2001]
-
-To simulate a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$,
-there is the well-known NORTA algorithm [@Cario1997]. It follows like this:
+The well-known NORTA algorithm [@Cario1997] can be used simulate a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$.
+Specifically, the NORTA algorithm follows like this:
 
 1. Simulate a random vector $\bf Z$ with $d$ **independent** and **identical** standard normal components.
-2. Determine the input matrix $\Sigma_{\bf Z}$ to corresponds with the
-   specified output $\Sigma_{\bf Y}$ [@Chen2001; @Xia17]
+2. Determine the input matrix $\Sigma_{\bf Z}$ that corresponds with the specified output $\Sigma_{\bf Y}$.
 3. Produce the Cholesky factor $M$ of $\Sigma_{\bf Z}$ so that $M M^{\prime}=\Sigma_{\bf Z}$.
@@
 
-We shall see that constructing continuous joint distributions that match a target Spearman or Kendall's correlations computes easily when employing Gaussian copulas, since this measures are invariant under the monotone transformations involved [refs].
+With modern parallelized computing, steps 1, 3, 4, 5 are readily implemented as high-performance (multicore and/or graphical-processing-unit (GPU) acceleration) algorithms, providing the greatest scaleability in a practical manner.
+
+Matching specified Pearson correlation coefficients exactly (step 2 above), however, is problematic.
+In general, there is no closed form correspondence between the components of the input $\Sigma_{\bf Z}$ and target $\Sigma_{\bf Y}$.
+Matching the correlations involves evaluating or approximating $\binom{d}{2}$ integrals of the form $EY_iY_j = \int \int y_i y_j f_{X|r}(F_i^{-1}(\Phi(z_i)), F_j^{-1}(\Phi(z_j))dy_idy_j$, for $i,j=1,2,\ldots,d$.
+For high-dimensional data, these evaluations are often too costly to enable feasible comprehensive MC studies.
+For low-dimensional problems, methods and tools exist to match exactly (see @Chen2001; @Xia17; @MB13), including the publicly available `nortaRA` R package.
+
+Our solution is to essentially avoid this complication in Pearson matching.
+Since our goal is to simulate non-normal marginals, we greatly prefer the use of rank-based measures $\rho_S$ and $\tau$ from a modeling standpoint.
+Further, $\rho_S$ and $\tau$ invariance under monotone transformation (see [Background](#background)), perserves the correlation coefficients through steps 3, 4, and 5 in the NORTA algorithm above.
+This eliminates the need for computing the $\binom{d}{2}$ integrals to match exactly (nothing is for free, however, as discussed in below in Section \@ref(rand-vec-gen)).
+
+Despite all this, if one does desire to characterize dependency using Pearson correlations, simply using the target Pearson correlation matrix as the initial conditions to our proposed algorithm will lead to approximate matching in the resultant distribution (@Song00).
+The quality of this approximation depends on the modeling goal, but in practice, for high-dimensional count data we find the accuracy to be adequate.
+Later, we'll study the robustness of our method to this limitation in selected [Monte Carlo evalutations]({#simulations}).
+
+## Random vector generation via `bigsimr::rvec()` {#rand-vec-gen}
+
+Now we describe `bigsimr::rvec()`, our algorithm to generate random vectors.
+It mirrors the classical NORTA algorithm above with some modifications for rank-based dependency matching:
+
+1. Pre-processing for nonparameteric dependency matching.  
+	(i) Convert from either ${\bf R_{Spearman}}$ or ${\bf R_{Kendall}}$ into the corresponding MVN input correlation ${\bf R_{Pearson}}$.
+	(ii) Check that ${\bf R_{Pearson}}$ is semi-positive definite.  
+	(iii) If not compute a close semi-positive ${\bf \widetilde{R}_{Pearson}}$.  
+2. Gaussian copula construction.
+	(i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_{Pearson}})$ ;  
+	(ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$;  
+3. Quantile evaluations.
+	Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;
+
+The pre-processing (Step1) takes advantage of the closed form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertSpearman) or \@ref(eq:convertKendall), respectively (implemented as `bigsimr::cor_covert()`).
+
+A complication often arises at this stage: the parallelized pairwise conversions may become non-positive definite, especially in high dimensions.
+Researchers working in multivariate computation frequently encounter such difficulties and need to find a closest positive definiteness matrix.
+The most widely used routine for this task in R is called `matrix::nearPD()` and is not suitable for high dimensions.
+To overcome this issue, we've implemented `bigsimr::nearPSD()`, a quadractically convergent Netwon method for finding the nearest correlation matrix introduced by @QS2006.
+We hope that this routine could be useful in many applications aside from our primary goal of random vector generation.
+
+Once the target margins and algorithm inputs are determined, steps 2 and 3 are essentially a NORTA algorithm with modern high-performance computing implementations.
+Specifically, step 2i uses either an efficient multicore multivariate normal simulator `mvnfast`[@Fasiolo2016] or a using Google's JAX python library for graphical-processing-unit (GPU) acceleration of the Cholesky factorization and matrix multiplication (steps 3, 4 in the NORTA algorithm in the preceeding section).
 
 <!-- 
+## Other high-performance `bigsimr` algorithms
+
+<mark> ALEX, briefly discuss other algorithms here </mark>
+- `cor_bounds()`  
+- `cor_fast()`  
+- `cor_nearPSD()`
+
 For discrete marginals, achieving a target Spearman correlation under this scheme is possible by using components from Equation \@ref(eq:spearmanRescaled) to further adjust the input correlation matrix.
@@
 your comment -->
-
-In contrast the rank-based correlations, matching specified Pearson correlation coefficients exactly is computational intense in this scheme.
-In general, there is no closed form correspondence and involving computing or approximating $\binom{d}{2}$ integrals of the form $EY_iY_j = \int \int y_i y_j f_{X|r}(F_i^{-1}(\Phi(z_i)), F_j^{-1}(\Phi(z_j))dy_idy_j$, for $i,j=1,2,\ldots,d$ (see Xia17 and MB13).
-For accurate numeric approximation of these integrals, the functions must be evaluated hundreds of times. 
-Others have used efficient Monte Carlo integration schemes (see @Chen2001), but scale poorly to large dimension in reasonable times (property **S2**). 
-Despite all this, if one does desire to characterize dependency using Pearson correlations, we often see in practice --- and it is theoretically justified under certain conditions (@Song00) --- 
-that simply using the target Pearson correlation matrix as the initial conditions to our proposed algorithm will lead to approximate matching in the resultant distribution.
-We'll study the robustness of our method to this limitation later in the application section (Section BLAH).
-
-Putting the together the facts provided in the equations above, we come the following proposed simulation algorithm to produce a random vector ${\bf Y}$ with specified Spearman's correlation and marginal distributions.
-Note that all computational steps can be parallelized as each operation can be done or either the $d$ marginals or the $\binom{d}{2}$ pairs for the correlation values.
-Even the generation of multivariate normal random vectors is parallelized through optimized matrix multiplication/decomposition routines.
-
-## Random vector generation `bigsimr::rvec` algorithm {#rand-vec-gen}
-
-<mark>make this algorithm pretty later </mark>
-
-1. Preprocessing for nonparameteric dependency matching.  
-	(i) Convert from either ${\bf R_{Spearman}}$ or ${\bf R_{Kendall}}$ into the corresponding MVN input correlation ${\bf R_{Pearson}}$ via \@ref(eq:convertSpearman) or \@ref(eq:convertKendall), respectivity.
-	(ii) Check that ${\bf R_{Pearson}}$ is semi-positive definite.  
-	(ii) If not find a close semi-positive ${\bf \widetilde{R}_{Pearson}}$ via `cor_nearPSD()`.  
-2. NORTA transformation  
-	(1) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_{Pearson}})$;  
-	(2) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$;  
-3. Return step  
-	(3) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;
-
-## Other high-performance `bigsimr` algorithms
-
-<mark> ALEX, briefly discuss other algorithms here </mark>
-
-- `cor_bounds()`  
-- `cor_covert()` make sure that you are clear this is for only MVN.  
-- `cor_fast()`  
-- `cor_nearPSD()`

### 2020-10-28 020-simulation-algorithm.Rmd
revise algorithms again

@@
 
-A complication often arises at this stage: the parallelized pairwise conversions may become non-positive definite, especially in high dimensions.
+A complication often arises at this stage: the parallelized pairwise conversions may create a (spuriously) non-positive definite correlation matrix, especially in high dimensions.
 Researchers working in multivariate computation frequently encounter such difficulties and need to find a closest positive definiteness matrix.
 The most widely used routine for this task in R is called `matrix::nearPD()` and is not suitable for high dimensions.
-To overcome this issue, we've implemented `bigsimr::nearPSD()`, a quadractically convergent Netwon method for finding the nearest correlation matrix introduced by @QS2006.
+To overcome this issue, we've implemented `bigsimr::nearPSD()`, a quadractically convergent Netwon method for finding the nearest correlation matrix, developed by @QS2006.
 We hope that this routine could be useful in many applications aside from our primary goal of random vector generation.
@@
 Once the target margins and algorithm inputs are determined, steps 2 and 3 are essentially a NORTA algorithm with modern high-performance computing implementations.
-Specifically, step 2i uses either an efficient multicore multivariate normal simulator `mvnfast`[@Fasiolo2016] or a using Google's JAX python library for graphical-processing-unit (GPU) acceleration of the Cholesky factorization and matrix multiplication (steps 3, 4 in the NORTA algorithm in the preceeding section).
+Specifically, step 2i uses either an efficient multicore multivariate normal simulator `mvnfast`[@Fasiolo2016] or a using Google's JAX python library NumPy for graphical-processing-unit (GPU) acceleration of the Cholesky factorization and matrix multiplication (steps 3, 4 in the NORTA algorithm in the preceeding section).
+(JAX is a not acronym and can be thought of a high-performance NumPy).
 

### 2020-10-28 030-bigsimr-package.Rmd
outline distribution assumptions
bigsimr R package rev1

@@
 
+The `bigsimr` package is a high-performance implementation of the proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms)).
+When designing `bigsimr`, we aimed to conveniently provide parallelized computation, with multicore and GPU acceleration.
+
+This section describes basic of `bigsimr`, by stepping through a low-dimensional (2D) simulation workflow for the often-used example data set `airquality`.
+This workflow proceeds from data, to estimation, simulation configuration, and random vector generation.
+For this low-dimensional setting, we compute using a single central processing unit (CPU) as the overhead in forking the tasks to multiple cores outways the computation gains. 
+Then we transition to advanced use where we briefly describe some of the high-performance features and syntax.
+The section concludes with a description of how to use `bigsimr` on a computer clusters through slurm scheduling via the `rslurm` package.
+
+<!-- 
 `bigsimr` is an R package for simulating high-dimensional multivariate data with arbitrary marginal distributions.
@@
 As with many parallel algorithms, there is overhead associated with forking the task to utilize multiple cores, however cost is negligible compared to the total run-time.
+-->
 
-## Basic use
+## Basic use workflow
 
 We're going to show the basic use and syntax of `bigsimr` by using the New York air quality data set (`airquality`) included in the R `datasets` package.
-We will focus specifically on the temperature (degrees Fahrenheit) and ozone level (parts per billion).
+First we load the `bigsimr` library and a few other conveninent data science packages, including the syntaxically-pleasing `tidyverse` suite of R packages.
 
@@
 ```
+
+For simplicity and to provide a minimal worknig example, we'll consider bivariate simulation of temperature, in degrees Fahrenheit, and ozone level, in parts per billion.
 
@@
 
-Let's look at the joint distribution of the Ozone and Temperature
+Figure \@ref(fig:ch030-aq-joint-dist) visualizes the bivariate relationship between Ozone and Temperature.
+We aim to simulate random two-component vectors mimicking this structure. 
+The margins are not normally distributed, particulary the ozone level exhibits a strong left skew.
 
-```{r ch030-aq-joint-dist}
+```{r ch030-aq-joint-dist, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Bivariate scatterplot of Ozone vs.
+Temperature with estimated marginal densities.
+The data are left skewed tails and appear to be correlation."}
+
 p0 <- ggplot(df, aes(Temp, Ozone)) +
@@
 
-We can see that not all margins are normally distributed - the ozone level is highly skewed.
-Though we don't know the true distribution of ozone levels, we can go forward assuming that it is log-normally distributed.
+Next we must specify the marginal distributions and correlation coefficient (type and magnitude).
+Here the analyst is free to be creative and we will not take up goodness-of-fit considerations.
+But it seems sensible without domain knowledge to estimate these quantative from the data and we have a suite of functions designed for this task.
 
-To simulate observations from this joint distribution, we need to estimate the correlation and the marginal parameters.
+*Specifying marginal distributions*
 
-### Estimating Correlation
+Though we don't know the true distribution of ozone levels or air temperatures, we'll assume that both are log-normally distributed since the values are positive and skewed.
+We'll use the well-known maximum likelihood estimators for the lognormal parameters:
 
-To estimate the correlation, we can use `R`'s function `cor` or we can use the convenience function `bigsimr::cor_fast` which can estimate Pearson, Spearman, or Kendall correlation using the fastest methods available from other packages.
-
-```{r ch030-aq-cor, echo=TRUE}
-(rho <- cor_fast(df, method = "pearson"))
-```
-
-### Defining Marginal Distributions
-
-Next we can estimate the marginal parameters. Assuming that `Temperature` is normally distributed, it has parameters
-
-```{r ch030-aq-temp-pars, echo=TRUE}
+```{r ch030-aq-temp-pars, echo=FALSE}
 df %>% 
@@
 ```
-
-and assuming that `Ozone` is log-normally distributed, it has parameters
 
@@
 df %>% 
-  select(Ozone) %>% 
+  select(Temp, Ozone) %>% 
   summarise_all(.funs = c(meanlog = mle_mean, sdlog = mle_sd))
@@
 
-We need a list of marginals (and their parameters), and a correlation structure (matrix).
-The marginal distributions can be built up using R's special `alist` function.
+Next, we'll configure the input marginals later input into `bigsimr::rvec`.
+The marginal distributions are specifying using `R`'s special `alist` function.
 This allows one to enter the distributions without evaluating anything (yet).
@@
 margins <- alist(
-  qnorm(mean = 77.9, sd = 9.49),
+  qlnorm(meanlog = 0.74, sdlog = 0.016),
   qlnorm(meanlog = 3.42, sdlog = 0.743)
@@
 
-Notice that we use the *quantile* function for the marginals, as that is the workhorse of the `rvec` function.
-By using an `alist`, users can also specify their own custom distributions (see the section on [creating custom margins](#custom-margins)).
+Notice that we use the *quantile* function for the marginals, as that is how the marginal distributions $F_i$ enter into the `bigsimr::rvec` algorthm.
+This implementation strategy supports all `R` `base` probability distributions.
+And allows flexible extensions using other `R` packages that adhere to conventions, such as `extraDistr`
+Further, by using `alist`, users can specify their own custom distributions (see below in [creating custom margins](#custom-margins)).
 
 It is a bit inconvenient to have to fill in the parameter values manually each time, so we provide a convenience function called `mlist` which behaves similarly to `alist`, except that it will evaluate the right hand side of argument values within the list.
+This is intended to help when scaling up your code to high dimensions.
 
@@
 margins <- mlist(
-  qnorm(mean = mean(df$Temp), sd = sd(df$Temp)),
+  qlnorm(meanlog = mle_mean(df$Temp), sdlog = mle_sd(df$Temp)),
   qlnorm(meanlog = mle_mean(df$Ozone), sdlog = mle_sd(df$Ozone))
 )
-
 margins
@@
 
-### Correlation Bounds
+*Specfying correlation*.
+Again the user must decide how to describe correlation, based on the particulars of the problem.
+For non-normal data and for improved simulation accuracy in our scheme, we advocate the use of rank-based correlations Spearman's $\rho_S$ and Kendall's $\tau$.
+But we also support approximate Pearson correlation coefficient matching, while cautioning the user to check the performance for their parameteric multivariate model (see [Monte Carlo evaluations](#simulations) for evaluation strategies).
+To aid in correlation specification, and estimation in general, we provide a high-performance function `bigsimr::cor_fast` which can estimate Pearson, Spearman, or Kendall correlation using the fastest methods available.
+(Anyone who has tried estimating Kendall's $\tau$ using `stats::cor` can attest that the routine does not scale to even moderate dimensions).
+Notably, these estimation methods are the standard approaches, not designed specifically designed for high-dimensional correlation estimation (see [Conclusion and Discussion]({#discussion) for more on this).
 
-Given a list of margins, the theoretical lower and upper correlation coefficients can be estimated.
+```{r ch030-aq-cor, echo=TRUE}
+type <- 'spearman'
+(rho <- cor_fast(df, method = "spearman"))
+```
+
+*Checking the theoretical correlation bounds*
+As discussed in Section \@ref(background), given a pair of marignal distributions the possible correlations are not free to vary between $[-1, 1]$. 
+To ensure that the simulation is not configured to impossible settings, we provide the `bigsimr::cor_bounds` fuction provides Monte Carlo estimated theoretical lower and upper bounds (using the Generate, Sort, and Correlate algorithm of @DH2011).
 
 ```{r ch030-cor-bounds, echo=TRUE}
-cor_bounds(margins = margins, type = "pearson")
+cor_bounds(margins = margins, type = type)
 ```
 
-Since our estimated correlation is within the theoretical bounds (assuming the distributions are correct), we should be able to achieve a similar correlation in the simulated data.
+Since our estimated correlation is within the theoretical bounds, the correlation is valid as input to `bigsimr:rvec`.
+By comparison, the Pearson correlation is restricted for these margins.
+
+```{r ch030-cor-bounds-pearson, echo=TRUE}
+cor_bounds(margins = margins, type = 'pearson')
+```
 

### 2020-10-29 030-bigsimr-package.Rmd
bigsimr R package rev2

@@
 
-The `bigsimr` package is a high-performance implementation of the proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms)).
-When designing `bigsimr`, we aimed to conveniently provide parallelized computation, with multicore and GPU acceleration.
+The `bigsimr` package is a high-performance implementation of the proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms) for details).
+When designing `bigsimr`, we aimed to conveniently provide parallelized computation, through multi-core and GPU acceleration.
 
-This section describes basic of `bigsimr`, by stepping through a low-dimensional (2D) simulation workflow for the often-used example data set `airquality`.
-This workflow proceeds from data, to estimation, simulation configuration, and random vector generation.
-For this low-dimensional setting, we compute using a single central processing unit (CPU) as the overhead in forking the tasks to multiple cores outways the computation gains. 
+This section describes basic use of `bigsimr`, by stepping through a low-dimensional (2D) simulation workflow for the often-used example data set `airquality`.
+This workflow proceeds from data, to estimation, simulation configuration, random vector generation, and result vizualization.
+For this low-dimensional setting, we compute using a single central processing unit (CPU) as the overhead in forking the tasks to multiple cores outways the computational gains. 
 Then we transition to advanced use where we briefly describe some of the high-performance features and syntax.
-The section concludes with a description of how to use `bigsimr` on a computer clusters through slurm scheduling via the `rslurm` package.
+The section concludes with a short description of how to use `bigsimr` on a computer clusters through slurm scheduling via the `rslurm` package.
 
@@
 
-## Basic use workflow
+## Basic use illustrated through a minimal example
 
-We're going to show the basic use and syntax of `bigsimr` by using the New York air quality data set (`airquality`) included in the R `datasets` package.
-First we load the `bigsimr` library and a few other conveninent data science packages, including the syntaxically-pleasing `tidyverse` suite of R packages.
+We'll demonstrate the basic use and syntax of `bigsimr` through an example workflow applied to the New York air quality data set (`airquality`) included in the R `datasets` package.
+First, we load the `bigsimr` library and a few other convenient data science packages, including the syntacically-pleasing `tidyverse` suite of `R` packages.
 
@@
 
-For simplicity and to provide a minimal worknig example, we'll consider bivariate simulation of temperature, in degrees Fahrenheit, and ozone level, in parts per billion.
+For simplicity and to provide a minimal working example, we'll consider bivariate simulation of temperature, in degrees Fahrenheit, and ozone level, in parts per billion.
 
@@
 Next we must specify the marginal distributions and correlation coefficient (type and magnitude).
-Here the analyst is free to be creative and we will not take up goodness-of-fit considerations.
-But it seems sensible without domain knowledge to estimate these quantative from the data and we have a suite of functions designed for this task.
+Here the analyst is free to be creative.
+For this example, we will not take up goodness-of-fit considerations.
+But it seems sensible without domain knowledge to estimate these quantities from the data and we have a suite of functions designed for this task.
 
-*Specifying marginal distributions*
-
+*Specifying marginal distributions*.
 Though we don't know the true distribution of ozone levels or air temperatures, we'll assume that both are log-normally distributed since the values are positive and skewed.
@@
 
-Since our estimated correlation is within the theoretical bounds, the correlation is valid as input to `bigsimr:rvec`.
-By comparison, the Pearson correlation is restricted for these margins.
+Since our estimated Spearm correlation $\hat{ \rho_S }$ is within the theoretical bounds, the correlation is valid as input to `bigsimr:rvec`.
+By comparison, the Pearson correlation $\rho_P$ is restricted to [-0.8648, 0.8727] for these margins (see `bigsimr::cor_bounds`output below).
 
@@
 
-### Simulating Multivariate Distributions
-
-Let's now simulate 10,000 observations from the joint distribution using `bigsimr::rvec`
+*Simulating random vectors*.
+Finally, we arrive at the main function of `bigsimr`, `rvec`.
+Let's now simulate $B=10,000$ observations from the assumed joint distribution of Ozone levels and Temp.
 
 ```{r ch030-sim-margins, echo=TRUE}
-x <- rvec(10000, rho, margins, "pearson")
-
+x <- rvec(10000, rho, margins, type)
 df_sim <- as.data.frame(x)
@@
 
-and visualize the joint distribution
+Figure \@ref(fig:ch030-plot-sim) plots the 10,000 simulated points.
 
-```{r ch030-plot-sim}
+```{r ch030-plot-sim, fig.cap="Contour plot and marginal densities for the simulated bivariate distribution of Air Quality Temperatures and Ozone levels.
+The simulated points mimick the observed data with respect to both the marginal characteristics and bivariate association."}
 p1 <- df_sim %>%
@@
 
-### Comparison to Uncorrelated Samples
-
+<!-- 
+*Comparison to Uncorrelated Samples*.
 We can compare the bivariate distribution above to one where no correlation is taken into account.
 
-```{r ch030-compare-uncorrelated}
+```{r ch030-compare-uncorrelated, eval=FALSE}
 df_sim2 <- data.frame(
@@
 
+your comment -->
+
 ## Advanced use
 
-### Creating a custom marginal distribution {#custom-margins}
+*Creating a custom marginal distribution*.
+Because `bigsimr` uses an `alist` to store the margins, any kind of probability distribution can be used including custom marginal distributions not provided in base `R`. 
+This accommodation is possible because the main function `rvec` uses the inverse CDF of a margin to transform from correlated uniform margins to the target distribution.
+Therefore, to specify a marginal distribution absent from an existing packge, the user simply needs to provide a closed form expression from the corresponding quantile function.
 
-Because `bigsimr` uses an `alist` to store the margins, any kind of probability distribution can be used including custom marginal distributions not provided in base R.
-This accommodation is possible because the main function `rvec` uses the inverse CDF of a margin to transform from correlated uniform margins to the target distribution.
+It is important to follow `R`'s naming convention for probability distributions.
+Users should prefix distributions with `r` and `q` for *random* and *quantile* respectively.
+Only the quantile function is necessary to simulate random vectors, but to compute the theoretical correlation bounds, it is required to supply univariate random number generator as well.
 
-It is highly recommended that users follow R's naming convention for probability distributions - I.e.
-prefixing distributions with `r` and `q` for *random* and *quantile* respectively.
-Only the quantile function is necessary to generate multivariate data, but if one wishes to compute the theoretical correlation bounds, then one should also specify the random number generator (RNG) function for their custom distribution.
-
-For an example, R does not have any functions for the *Pareto* distribution. We can write a quantile function and RNG function and use it in `rvec`.
-
-The Pareto CDF is defined as
+For an example, let's provide a custom *Pareto* distribution for use with `bigsimr`.
+The Pareto CDF is
 
@@
 
-for scale $x_m > 0$ and shape $\alpha > 0$, with support $x \in \left[x_m, \infty\right)$. From the CDF, we compute the inverse CDF
+for scale $x_m > 0$ and shape $\alpha > 0$, with support $x \in \left[x_m, \infty\right)$.
+From the CDF, we compute the inverse CDF
 
@@
 
-Next we define the Pareto quantile function in R. Note that extra checks should be taken to ensure that the arguments are valid. We omit such checks for clarity.
+Next we define the Pareto quantile function in R.
 
@@
 
-Writing a RNG function for our new distribution can be accomplished by calling the quantile function on the output of a random unit uniform variable.
+For example, writing random number generating function for our target marginal can be accomplished by calling the quantile function on a uniformly distributed random variable (the inverse transform method @Rizzo2007).
 
@@
 
-### Calculating the nearest positive (semi)definite correlation matrix
-
-### Using multicore
-
-### Utilizing the GPU
-
-The `bigsimr` packages uses Google's `jax` python library to interface with an Nvidia CUDA enabled GPU.
-
-### Simulation-based computation of correlation bounds
-
-### Using `bigsimr` on a computing cluster via `rslurm`
-
+*Using `bigsimr` on a computing cluster via `rslurm`*.
 Though `bigsimr` runs quickly, at large $d$ users may want to run jobs on a shared computing server.
@@
 
-```{r ch030-rvecRslurm, echo=FALSE, eval=FALSE}
+```{r ch030-rvecRslurm-with-comments, echo=FALSE, eval=FALSE}
 library(rslurm)
@@
 
+```{r ch030-rvecRslurm, echo=TRUE, eval=FALSE}
+library(rslurm)
+sjob <- slurm_call(rvec, jobname = 'rvec',
+                   alist(n=1e6,
+                         rho = rho,
+                        margins = margins,
+                        type = "spearman"),
+                   submit = TRUE)
+```
+
+<!-- 
+Once your jobs complete, `rslurm` conveniently stores simulation results in a standardized format.
+Retriving the data is accomplished by
+-->
+
 ```{r ch030-rvecRslurmRes, echo=FALSE, eval=FALSE}
@@
 head(res, 3)
-## cleanup_files(sjob) ## careful this deletes output also
 ```
 
-Now, let's show off the real power of combining `bigsimr` and `rslurm` by simulating many correlation structures for these three marginals.
+Now, let's show off the real power of combining `bigsimr` and `rslurm` by simulating many correlation structures.
 The `rslurm::slumr_map` syntax mirrors the familiar `base::lapply` and `purrr::map` functions.
 
-```{r ch030-rvecRslurmMap, echo=FALSE, eval=FALSE}
+```{r ch030-rvecRslurmMap-comments, echo=FALSE, eval=FALSE}
 library(rslurm)
@@
 simReps <- 100
-rhoList <- replicate( n = simReps, bigsimr::rcor(d = length(margins), constant_rho = TRUE), simplify=FALSE )
+rhoList <- replicate( n = simReps, bigsimr::cor_randPSD(d = length(margins)),
+                     simplify=FALSE )
 sjob <- slurm_map(x = rhoList,
@@
 
-On a cluster carrying 24 nodes with 48 threads, these 100 jobs completed in about a minute. Let's evaluate the quality of simulation performance.
+```{r ch030-rvecRslurmMap, echo=TRUE, eval=FALSE}
+library(rslurm)
+simReps <- 100
+rhoList <- replicate( n = simReps, bigsimr::cor_randPSD(d = length(margins)), simplify=FALSE )
+sjob <- slurm_map(x = rhoList,
+                  f = rvec,
+                  jobname = 'rvecMap',
+                  n=1e6,
+                  margins = margins,
+                  type = "spearman",
+                  nodes = 4,
+                  cpus_per_node = 4,
+                  cores = 4,
+                  submit = TRUE)
+```
+
+On a cluster carrying 24 nodes with 48 threads, these 100 jobs completed in about a minute.
+
+<!-- 
+Let's evaluate the quality of simulation performance.
+
+your comment -->
 
@@
 rhoHats <- lapply( res, bigsimr::fastCor, method = 'spearman' )
-
 resData <- NULL
@@
 }
-
 ## "wrangle" into a tibble
@@
 saveRDS( res, file = 'data/rvecRslurmMapRes.rds' )
-
 ## cleanup careful this deletes output also
@@
 
-```{r ch030-rvecRslurmMapFig, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, out.width= '75%', fig.align='center', fig.cap='The continous marginal pairs exact reproduce the specified correlation whereas the discrete pairs (green nad blue) show a downward bias.
+```{r ch030-rvecRslurmMapFig, echo = FALSE, eval=FALSE, message=FALSE, warning=FALSE, out.width= '75%', fig.align='center', fig.cap='The continous marginal pairs exact reproduce the specified correlation whereas the discrete pairs (green nad blue) show a downward bias.
 Future work with employ a discrete adjusted procedures using *rescaling* on the traditional correlation coefficient.'}
-library(tidyverse)
-
 res <- readRDS( 'data/rvecRslurmMapRes.rds' )
 res
-
 res %>%

### 2020-10-29 060-conclusion-discussion.Rmd
migrating ESCO 2020 talk into sections, revising everything
Revising background
bigsimr R package rev1
bigsimr R package rev2

@@
-# Conclusion/Discussion
+# Conclusion and discussion {#discussion}
 
@@
 This algorithm lends itself to exploration of flexible probability models to inspire model selection in the development novel RNA-seq analytic methodology.
+
+-  Simulating correlated discrete and count data takes more care [@Xia17], including slightly redefining the inverse. 
+- a user can always supply their own input Pearson after using the existing matching schemes [@XZ19].
 
@@
 
-Mention `rslurm`
+- mention that `bigsimr`
+
+Nvidia-based(copyright symbol) graphical-processing-unit (GPU), OpenCL future direction.
 

### 2020-10-29 070-supplementary-materials.Rmd
Revising background
bigsimr R package rev2

@@
-# Supplementary Materials
+# Supplementary Materials {#misc}
 
-We provide an open-source implementation of our methology as the `bigsimr` R package, hosted on github.
+We provide an open-source implementation of our methology as the `bigsimr` R package, hosted on github, https://schisslergroup.github.io/bigsimr/.
 
@@
 
-The authors gratefully acknowledge the helpful discussions with Professor Walter W. Piegorsch and Professor Edward J. Bedrick during this project's conception.
+The authors gratefully acknowledge the helpful discussions with University of Arizona's Professor Walter W. Piegorsch and Professor Edward J. Bedrick during this project's conception.
+We also gratefully acknowledge Heather Knudson's graphic design for the `bigsimr` R Package.
+The results published here are in whole or part based upon data generated by the TCGA Research Network: https://www.cancer.gov/tcga.
 
@@
 
-The authors report no conflict of interest. DO WE?
+The authors report no conflict of interest.
 
@@
 
-FUNDING FROM EVERYONE HERE. CP3 grant.
+Research reported in this publication was supported by MW-CTR-IN of the National Institutes of Health under award number NIH 1U54GM104944.
 
-# Nomenclature/Notation
+<!--
+
+Nomenclature/Notation
 
@@
 ```
+
+your comment -->

### 2020-10-29 010-background.Rmd
bigsimr R package rev3

@@
     brcaMedian <- apply(brca, 2, median)
-    ## retain top (1-probs)*100% highest expressing genes for illustration
+    ## retain top (1-myProb)*100% highest expressing genes for illustration
     myProb <- 0.99
@@
 }
+myProb <- 0.99
+brca <- readRDS(file = './data/brca99.rds')
+d <- ncol(brca)
 ```
@@
 ## Describe and plot real data
-brca <- readRDS(file = './data/brca99.rds')
 set.seed(10192020)

### 2020-10-29 030-bigsimr-package.Rmd
bigsimr R package rev3

@@
 *Specifying marginal distributions*.
-Though we don't know the true distribution of ozone levels or air temperatures, we'll assume that both are log-normally distributed since the values are positive and skewed.
-We'll use the well-known maximum likelihood estimators for the lognormal parameters:
+Of course, we don't know the true distribution of ozone levels or air temperatures.
+Based on the estimated densities in Figure \@ref(fig:ch030-aq-joint-dist), we'll assume `Temp` is normally ditsributed and `Ozone` is log-normally distributed since its values are positive and skewed.
+We'll use the well-known unbiased estimators for normal distribution and maximum likelihood estimators for the lognormal parameters:
 
-```{r ch030-aq-temp-pars, echo=FALSE}
+```{r ch030-aq-temp-pars, echo=TRUE, eval=TRUE}
 df %>% 
@@
 df %>% 
-  select(Temp, Ozone) %>% 
+  select(Ozone) %>% 
   summarise_all(.funs = c(meanlog = mle_mean, sdlog = mle_sd))
@@
 margins <- alist(
-  qlnorm(meanlog = 0.74, sdlog = 0.016),
-  qlnorm(meanlog = 3.42, sdlog = 0.743)
+    qnorm(mean = 77.871, sd = 9.4855),
+    qlnorm(meanlog = 3.419, sdlog = 0.7426),
 )
@@
 margins <- mlist(
-  qlnorm(meanlog = mle_mean(df$Temp), sdlog = mle_sd(df$Temp)),
+  qnorm(mean = mean(df$Temp), sd = sd(df$Temp)),
   qlnorm(meanlog = mle_mean(df$Ozone), sdlog = mle_sd(df$Ozone))
@@
 
-Since our estimated Spearm correlation $\hat{ \rho_S }$ is within the theoretical bounds, the correlation is valid as input to `bigsimr:rvec`.
+Since our estimated Spearm correlation $\hat{ \rho}_S$ is within the theoretical bounds, the correlation is valid as input to `bigsimr:rvec`.
 By comparison, the Pearson correlation $\rho_P$ is restricted to [-0.8648, 0.8727] for these margins (see `bigsimr::cor_bounds`output below).
@@
   theme(legend.position = "none") +
-  labs(x = "Temperature", y = "Ozone") +
+  labs(x = "Simulated Temperature", y = "Simulated Ozone") +
   scale_y_continuous(limits = c(0, max(df$Ozone))) +

### 2020-10-30 040-monte-carlo-evaluation.Rmd
bigsimr MC studies draft

@@
 
-Before applying our methodology to real data simulation, we conduct several Monte Carlo studies to investigate method performance in comparison to other existing implementations. 
-We focus the numerical experiments on assessing how well the procedure scales to high dimension with respect to reasonable computation times (property S1 above) and accurately matching marginal and dependency parameters.
-The simulations will proceed in increasing complexity --- leading up to the setting in our motivating example. 
-We begin by exploring simple exchangeable (constant) correlation structures under a large number of simulation replicates while applying the algorithm above to first continuous then discrete marginal distributions to test the robustness of our algorithm since exact matching is not guaranteed.
+Before applying our methodology to real data simulation, we conduct several Monte Carlo studies to investigate method performance.
+Since marginal parameter matching in our scheme is essentially a sequence of univariate inverse probability transforms, the challenging aspects are the accuracy of dependency matching and computational efficiency.
+And, so, we'll focus the numerical experiments on assessing how well the procedure scales to high dimension with respect to reasonable computation times (property S1 above) and accurately matching marginal and dependency parameters.
+
+Further, we select simulaiton configurations to ultimately simulate discrete-valued RNA-seq data and the simulations will proceed in increasing complexity, leading to our motivating application in Section \@ref(examples).
+We begin wiith empirically evaluating the dependency matching across all three supported correlations --- Pearson's, Spearman's, and Kendall's ---- in identical, bivariate marginal configurations.
+The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (mimicking RNA-seq counts).
 
@@
 ```
+
+Figure \@ref(fig:ch040-biNormPlot) `bigsimr` recovers the Pearson specified correlations for MVN.
 
@@
 ```
+
+Figure \@ref(fig:ch040-biGammaPlot) `bigsimr` recovers the Pearson specified correlations for bivariate gamma.
 
@@
 
+Figure \@ref(fig:ch040-biNegBinPlot) `bigsimr` recovers the Pearson specified correlations for bivariate negative binomial.
+
 ```{r ch040-biNegBinPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.align='center', fig.width = 8, fig.height=5,fig.cap = "`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
@@
 
+Taken together, the studies show our methology are accuracy across the entire range of possible correlation values for the rank-based dependency measures in continuous marginals.
+For all selected bivariate marginals, the Pearson correlation matching is approximate.
+For discrete margins, matching the dependency measures was somewhat less accurate, even for the rank-based metrics, especially near the correlation bounds.
+Yet generally the accuracy seems adequate for most applications, at least in these limited simulation settings.
+
+Of course, one could choose marginals that lead to poor matching by construction.
+In partice, we recommend users to always evaluate the accuracy for their application, using methods similar to those presented above.
+See [Discusion](discussion) for future directions for fast Pearson matching and discrete-specific modifications.
+
 ## Scale up to Ultra-High Dimensions
+
+The bivariate simulations provide a low-dimensional perspectitive.
+But, of course, this is not `bigsimr`'s primary purpose.
+We'd like the random vector methodolgy to scale to high-dimensional problems.
+
+DESCRIBE WHAT WE DID. Do we need to redo this?
 

### 2020-10-30 060-conclusion-discussion.Rmd
bigsimr MC studies draft

@@
 - mention that `bigsimr`   
-
+- comparison to other existing implementations see @Li2019gpu. copula
 Nvidia-based(copyright symbol) graphical-processing-unit (GPU), OpenCL future direction.
+- discrete matching in RNA-seq simulation may be due to large counts and ties could greatly affect
 

### 2020-10-30 040-monte-carlo-evaluation.Rmd
run bivariate sims rev2

@@
 
-*Bivariate Normal*.
-Let's simulate a bivariate normal and check our correlation matching performance as N increases.
-Here we have BVN( $\mu_1 = \mu_2 = 10, \rho_{type}$ ).
-We vary $\rho$ across the entire possible range of correlations for each correlation type.
-
 ```{r ch040-biNormal, echo = FALSE, eval = TRUE}
+mu <- 0
+sigma <- 1
+if ( !file.exists( 'results/biNorm_sims.rds' ) )  {
     mom_norm <- function(x) {
@@
     }
-
-mu <- 10
+    mu <- 0
     sigma <- 1
@@
     )
-
     type <- c("pearson", "spearman", "kendall")
-## cores <- as.integer( parallel::detectCores() - 1 )
-## n <- c(1e3, 1e4, 1e5)
-n <- c(1e1, 1e2, 1e3)
+    cores <- as.integer( parallel::detectCores() - 1 )
+    n <- c(1e3, 1e4, 1e5)
+    ## n <- c(1e1, 1e2, 1e3)
     adjustForDiscrete <- c(FALSE)
-
     eps <- 1e-2
-## grid_steps <- 100
-grid_steps <- 10
+    grid_steps <- 100
+    ## grid_steps <- 10
     sim_pars <- expand.grid(type = type, cores = cores, n = n,
@@
                             adjustForDiscrete = adjustForDiscrete)
-
     res <- data.frame()
-dir.create("./results/sims_norm", showWarnings = TRUE, recursive = TRUE)
+    dir.create("./results/biNorm_sims", showWarnings = TRUE, recursive = TRUE)
     for (i in 1:nrow(sim_pars)) {
@@
             ## save sims in their own subdirectory
-        saveRDS(x, file = paste0("./results/sims_norm/", id, ".rds"))
-
+            saveRDS(x, file = paste0("./results/biNorm_sims/", id, ".rds"))
             ## Estimate statistics
@@
             sigma_hat <- norm_args_hat$sd
-
             ## Save the results
@@
     res$type <- factor(res$type, levels = c("pearson",  "spearman", "kendall" ) )
-saveRDS(object = res, "./results/norm_sims.rds")
+    saveRDS(object = res, "./results/biNorm_sims.rds")
+}
 ```
 
+*Bivariate Normal*.
+Let's simulate a bivariate normal and check our correlation matching performance as N increases.
+Standard bivariate normal 
+Here we have BVN( $\mu_1 = \mu_2 = 0, \rho_{type}$ ).
+We vary $\rho$ across the entire possible range of correlations for each correlation type.
 Figure \@ref(fig:ch040-biNormPlot) `bigsimr` recovers the Pearson specified correlations for MVN.
@@
 ```{r ch040-biNormPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.height = 5, fig.width= 8, fig.align='center', fig.cap = "`bigsimr` recovers the Pearson specified correlations for MVN."}
-dat <- readRDS("./results/norm_sims.rds")
+dat <- readRDS("./results/biNorm_sims.rds")
 dat %>%
@@
     facet_wrap(~ type + N) + theme_bw()
-
 ## ggsave('fig/plot-biNormPlot.pdf')
@@
 
-*Bivariate Gamma*. 
-Similarly, let's check the performance for a non-symmetric continuous distribution: a standard (rate =1) bivariate gamma.
-Here we have a Bivariate Gamma with $shape_1 = shape_2 = 10, \rho_{type}$.
-We vary $\rho$ across the entire possible range of correlations for each correlation type.
-
 ```{r ch040-biGamma, echo = FALSE, eval = TRUE}
+shape <- 10
+rate <- 1
+if ( !file.exists( 'results/biGamma_sims.rds' ) )  {
     mom_gamma <- function(x) {
@@
     }
-
-shape <- 10
-rate <- 1
     margins <- alist(
@@
     ## cores <- c(1L)
-## n <- c(1e3, 1e4, 1e5)
-n <- c(1e1, 1e2, 1e3)
+    cores <- as.integer( parallel::detectCores() - 1 )
+    n <- c(1e3, 1e4, 1e5)
+    ## n <- c(1e1, 1e2, 1e3)
     adjustForDiscrete <- c(FALSE)
@@
     eps <- 1e-2
-## grid_steps <- 100
-grid_steps <- 10
+    grid_steps <- 100
+    ## grid_steps <- 10
     sim_pars <- expand.grid(type = type, cores = cores, n = n,
@@
 
-dir.create("./results/sims_gamma", showWarnings = TRUE, recursive = TRUE)
+    dir.create("./results/biGamma_sims", showWarnings = TRUE, recursive = TRUE)
     res <- data.frame()
@@
             )
-        saveRDS(x, file = paste0("./results/sims_gamma/", id, ".rds"))
+            saveRDS(x, file = paste0("./results/biGamma_sims/", id, ".rds"))
 
@@
     res$type <- factor(res$type, levels = c("pearson",  "spearman", "kendall" ) )
-saveRDS(object = res, "./results/sims_gamma.rds")
+    saveRDS(object = res, "./results/biGamma_sims.rds")
+}
 ```
 
+*Bivariate Gamma*. 
+Similarly, let's check the performance for a non-symmetric continuous distribution: a standard (rate =1) bivariate gamma.
+Here we have a Bivariate Gamma with $shape_1 = shape_2 = 10, \rho_{type}$.
+We vary $\rho$ across the entire possible range of correlations for each correlation type.
 Figure \@ref(fig:ch040-biGammaPlot) `bigsimr` recovers the Pearson specified correlations for bivariate gamma.
@@
 ```{r ch040-biGammaPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.height = 5, fig.width= 8, fig.align='center', fig.cap = "`bigsimr` recovers the Pearson specified correlations for Bivariate Gamma."}
-dat <- readRDS("./results/sims_gamma.rds")
+dat <- readRDS("./results/biGamma_sims.rds")
 dat %>%
@@
     facet_wrap(~ type + N) + theme_bw()
-
 ## ggsave('fig/plot-biGammaPlot.pdf')
@@
 
-*Bivariate Negative Binomial*. 
-Let's check the performance for a discrete distribution: a bivariate negative binomial.
-Here we have Bivariate Negative Binomial ( $prob_1 = prob_2 = 0.5, size_1 = size_2 = 4,\rho_{type}$ ).
-We vary $\rho$ across the entire possible range of correlations for each correlation type.
-
 ```{r ch040-biNegBin, echo = FALSE, eval = TRUE}
+size <- 4
+prob <- 3e-04
+if ( !file.exists( 'results/biNB_sims.rds' ) )  {
     mom_nbinom <- function(x) {
@@
 
-size <- 10
-prob <- 0.1
     margins <- alist(
@@
     type <- c("pearson", "spearman", "kendall")
+    cores <- as.integer( parallel::detectCores() - 1 )
     ## cores <- cores
-## n <- c(1e3, 1e4, 1e5)
-n <- c(1e1, 1e2, 1e3)
+    n <- c(1e3, 1e4, 1e5)
+    ## n <- c(1e1, 1e2, 1e3)
     adjustForDiscrete <- c(FALSE)
@@
 
-dir.create("./results/nbinom_sims", showWarnings = TRUE, recursive = TRUE)
+    dir.create("./results/biNB_sims", showWarnings = TRUE, recursive = TRUE)
     res <- data.frame()
@@
             )
-        saveRDS(x, file = paste0("./results/nbinom_sims/", id, ".rds"))
+            saveRDS(x, file = paste0("./results/biNB_sims/", id, ".rds"))
 
@@
     res$type <- factor(res$type, levels = c("pearson",  "spearman", "kendall" ) )
-saveRDS(object = res, "./results/nbinom_sims.rds")
+    saveRDS(object = res, "./results/biNB_sims.rds")
+}
 ```
 
+*Bivariate Negative Binomial*. 
+Let's check the performance for a discrete distribution: a bivariate negative binomial.
+Here we have Bivariate Negative Binomial ( $prob_1 = prob_2 = 0.5, size_1 = size_2 = 4,\rho_{type}$ ).
+We vary $\rho$ across the entire possible range of correlations for each correlation type.
 Figure \@ref(fig:ch040-biNegBinPlot) `bigsimr` recovers the Pearson specified correlations for bivariate negative binomial.
@@
 ```{r ch040-biNegBinPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.align='center', fig.width = 8, fig.height=5,fig.cap = "`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
-dat <- readRDS("./results/nbinom_sims.rds")
+dat <- readRDS("./results/biNB_sims.rds")
 dat %>%

### 2020-10-31 040-monte-carlo-evaluation.Rmd
revise MC studies rev3; example rev1

@@
 Let's check the performance for a discrete distribution: a bivariate negative binomial.
-Here we have Bivariate Negative Binomial ( $prob_1 = prob_2 = 0.5, size_1 = size_2 = 4,\rho_{type}$ ).
+We use values that are motivated by our RNA-seq data, namely the mean probability and sizes estimated from the data ( see [Example applications for our motivating data](examples) for estimation details): Bivariate Negative Binomial ( $prob_1 = prob_2 = `r prob`, size_1 = size_2 = 4,\rho_{type}$ ).
 We vary $\rho$ across the entire possible range of correlations for each correlation type.
@@
 Taken together, the studies show our methology are accuracy across the entire range of possible correlation values for the rank-based dependency measures in continuous marginals.
-For all selected bivariate marginals, the Pearson correlation matching is approximate.
+For the two non-normal bivariate marginals, the Pearson correlation matching is approximate.
+Table XX contains the average absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.
 For discrete margins, matching the dependency measures was somewhat less accurate, even for the rank-based metrics, especially near the correlation bounds.
-Yet generally the accuracy seems adequate for most applications, at least in these limited simulation settings.
+Since the bivariate normal exactly matches all dependency measures in our scheme, we also report relative MSE of the non-normal cases to the normal simulations.
 
+```{r ch040-biMSEtable}
+## compare relative differences in MSE for each dependency, compared to Pearson
+allDat <- NULL
+dat <- readRDS("./results/biNorm_sims.rds") %>% select( margins, type, N, rho, rho_hat  ) 
+allDat <- rbind( allDat, dat )
+dat <- readRDS("./results/biGamma_sims.rds") %>% select( margins, type, N, rho, rho_hat  )
+allDat <- rbind( allDat, dat )
+dat <- readRDS("./results/biNB_sims.rds") %>% select( margins, type, N, rho, rho_hat  )
+allDat <- rbind( allDat, dat )
+allDat$margins <- factor(allDat$margins, levels = c( 'norm', 'gamma', 'nbinom') )
+allDat$type <- factor(allDat$type, levels = c( 'kendall', 'spearman', 'pearson' ) )
+allDat$N <- factor(allDat$N, levels = c( 1000, 10000, 100000) )
+## tabMSE  <- allDat %>%
+##    group_by( N, margins ) %>%
+##    summarize( mse = mean( ( rho - rho_hat )^2 ) )
+tabMSE  <- allDat %>%
+    ## filter( N == 100000) %>%
+    group_by( N, type, margins ) %>%
+##     summarize( mse = mean( ( rho - rho_hat )^2 ) ) %>%
+    summarize( mae = mean( abs( rho - rho_hat )  ) )
+## normMSE <- as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ] ) 
+## ( tabMSE$mse - normMSE ) / normMSE
+## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) )
+## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) ) / tabMSE$mse
+knitr::kable(
+  tabMSE, booktabs = TRUE,
+  caption = 'Average abolute error in matching the target dependency and  '
+)
+```
+
+Generally, the accuracy seems adequate for most applications, at least in these limited simulation settings.
 Of course, one could choose marginals that lead to poor matching by construction.
@@
 
-The bivariate simulations provide a low-dimensional perspectitive.
-But, of course, this is not `bigsimr`'s primary purpose.
-We'd like the random vector methodolgy to scale to high-dimensional problems.
-
-DESCRIBE WHAT WE DID. Do we need to redo this?
+With some assurance of the method's accuracy from the low-dimensional perspective in the bivariate setting, we now turn to assessing whether the `bigsimr` can scale to large dimesional problems. 
+Now we seek evidence to determine whether the method can scale to higher dimensions in a practical time.
+Using our motivating RNA-seq data, described in [Background](background), we filtered the orignal 20,501 genes to the high-expressing genes at increasing percentiles, $1, 5, 10, 15, 20, 25\%$, to obtain $d=\{206, 1026, 2051, 3076, 4101, 5127\}$ marginals and $\binom{d}{2}$ pairwise correlations at each setting.
+For example, for $d=5127$ there are $13,140,501$ correlation coefficients.
+We estimated the marginal negative binomial parameters and the correlation coefficients from the RNA-seq data to seed our simulations.
+( See [Example applications for our motivating data](examples) for a detailed descrition of estimation).
+Figure \@ref(fig:ch040-gpuVScpuFig) displays computation times using various high-performance settings to produce $B=10,000$ random vectors.
+The Pearson simulations are much faster since the correlation conversion steps are avoided (pre-processing step; see [Algorithms](algorithms)), but as we known from above the accuracy will suffer slightly, especially near the negative boundary of the possible correlations.
+Matching Spearman's correlation at larger $d$ gets costly if one wants to produce $B=10,000$ random vectors at many different simulation settings.
+We recommend computing this type of simulations on a computing cluster (possibly using `rslurm`; see the [Advance Use](advanced-use)).
 

### 2020-10-31 index.Rmd
revise MC studies rev3; example rev1

@@
 But measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments.
-Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers may resort to simulation designs that posit independence.
+Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers may resort to simulation designs that posit independence or perform arbitrary data transformations.
 This greatly diminishes insights into the empirical operating characteristics of any proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness.

### 2020-11-01 040-monte-carlo-evaluation.Rmd
revise discussion, end of 1st draft

@@
 Since marginal parameter matching in our scheme is essentially a sequence of univariate inverse probability transforms, the challenging aspects are the accuracy of dependency matching and computational efficiency.
-And, so, we'll focus the numerical experiments on assessing how well the procedure scales to high dimension with respect to reasonable computation times (property S1 above) and accurately matching marginal and dependency parameters.
+And, so, we'll focus the numerical experiments on assessing how well the procedure scales to high dimension with respect to reasonable computation times (property S1 above) and accurately m
+atching marginal and dependency parameters.
 

### 2020-11-01 060-conclusion-discussion.Rmd
revise discussion, end of 1st draft

--- 2020-09-18 060-conclusion-discussion.Rmd
@@
-# Conclusion/Discussion
+# Conclusion and discussion {#discussion}
 
-- Although we recommend to use the rank-based measures
-- discuss the omission (or inclusion) of high dimensional covariance estimators?
-- We've shown that the multivariate NB simulations outperform the Independent sims.
-But the agreement is still poor.
-This could motivate other more appropriate models for high-expressing RNA-seq data than the negative binomial distribution.
-This algorithm lends itself to exploration of flexible probability models to inspire model selection in the development novel RNA-seq analytic methodology.
+We've introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a high-performance implementation called [`bigsimr`](https://schisslergroup.github.io/bigsimr/). 
+The high-performance --- efficient, multi-core and GPU enabled --- algorithms and software provide 
+The random vector generation method is largely inspired by NORTA (@Cario1997) and Gaussian copula-based approaches (@MB13, @BF17, @Xia17).
+The major contributions of our work presented here involve high-dimensional scalability, model flexibility, and high-performance implementation with broad potential data analytic applications for modern big data challenges.
 
-We've introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a high-performance implementation called `bigsimr` (github url).
-The parallelized (multi-core) algorithm is simple and easy to understand as an application of Gaussian copulas.
-This method is largely inspired by @MB13, but with contributions with regard to scalability, implementation, and application in high-throughput biomedical data (RNA-sequencing).
+It is usually customary to compare new tools and algorthms directly to existing competing methods.
+In our MC studies, however, we only employ our proposed methodology, since our previous work has show that existing tools are simply not designed or feasible to meet our HD goal (see @Li2019gpu for evalutions of the `R` `copula`package and others).
+For the bivariate simulations, existing packages such as [`nortaRA`](https://github.com/cran/NORTARA/blob/master/inst/doc/NORTARA.R) work well to match Pearson correlations exactly.
+
+<!-- 
 We advocate the use of Kendall's $\tau$ as it better captures correlation among components of non-normal, as well as non-linear patterns of association.
 Moreover, the matching Kendall's $\tau$ is nearly trivial due to the invariant of monotone transformations, after an adjustment to the input correlation matrix.
-Interestingly, our simulation studies show the use of Kendall's $\tau$ to works as well as using Pearson correlation coefficient when simulating from multivariate normal distributions.
+-->
 
+<!-- 
 We also show utility in our methodology through an application to differential gene expression analysis from RNA-sequencing data.
@@
 One could imagine additional structures/features in the data that an analyst could model to improve the correspondence with the empirical values, for example row correlations and high-dimensional covariance estimators (see @Won2013g).
+-->
 
-Mention `rslurm`
+There are limitations to the methodology and implementation.
+The most obvious missing feature the proposed methodology is the inability to match a Pearson correlation matrix exactly.
+As discussed in [Algorithms](algorithms) and extensively by @XZ19, this is a computational intense procedure and not a natural choice for characterizing dependency for non-normal marginals, especially in the high-dimensional setting.
+While we do not provide an implementation directly supporting Pearson matching, user can always supply their own input Pearson after using a supplementary matching scheme [@Cario1997; @XZ19].
 
-There are of course limitations to the methodology and implementation.
-The most obvious missing feature the proposed methodology is the inability to match a Pearson correlation matrix exactly.
-As discussed above, this is a computational intense procedure and not a natural choice for the posed problem (seeking to simulate non-normal margins in a scaleable fashion).
-Yet it is an important aspect that paralleziable, approximation based apporaches are promising [ref Hermite polynomials].
+<!-- 
 Further, while we provide the ability of the user to specify discrete marginals, exact matching a desired dependency measure is not yet obtained.
@@
 FOR EXAMPLE?).
-
 From a practical computing standpoint, the user's computing resource provides the ultimate limit on how large a dimension can be simulated using the `bigsimr` R package.
 A user must consider carefully the available memory and cores when conducting a Monte Carlo experiment.
-The algorithm does amends itself well to parallelization and graphical processing unit acceleration [@Li2019gpu] and advanced users may take advantage of those techniques.
+-->
 
-Future work includes developing scalable algorithms to match the Pearson correlation matrix exactly and more methods developed specifically for discrete distributions.
-Further, more sophisticated approaches could involve simulation algorithms that are "estimation aware" and so could explore the role of covariance estimation within the simulation.
-Lastly, these may reveal new approaches to estimating and evaluating high dimensional regression and Bayesian models.
-On the implementation side, employing graphical process unit acceleration could produce substantial speedups over our multi-core approach.
+Future work includes developing scalable algorithms to match the Pearson correlation matrix more precisely, discrete-margin specific modifications including fast Spearman's correlation rescaling (see Equation \@ref(eq:spearmanRescaled)), and *bona fide* high-dimensional correlation estimation (for example, see @Won2013g).
+From an implementation standpoint, `bigsimr` only supports Nvidia GPUs and refactoring the code using OpenCL would broaden the user base.
+As problems grow even larger, multi-GPU support is a logical progression in this highly parallelization algorthmic scheme and warrant further development.

### 2020-11-04 010-background.Rmd
edit overall. Cut RNA-seq MSE sizes. Revise RNA-seq example heavily.

--- 2020-10-27 010-background.Rmd
@@
 
-```{r ch010-processBRCA, echo=FALSE, eval=TRUE, results='none'}
-if ( !file.exists( 'data/brca99.rds' ) )  {
+```{r ch010-myProb, include=FALSE, cache = TRUE}
+myProb <- 0.99
+```
+
+```{r ch010-processBRCA, echo=FALSE, eval=TRUE, results='none', cache=F}
+brcaRDS <- paste0( "data/brca", round(myProb*100), ".rds" )
+if ( !file.exists( brcaRDS ) )  {
     ## full workflow simulating RNA-seq data
@@
     brcaMedian <- apply(brca, 2, median)
-    ## retain top (1-probs)*100% highest expressing genes for illustration
-    myProb <- 0.99
+    ## retain top (1-myProb)*100% highest expressing genes for illustration
     cutPoint <- quantile( x = brcaMedian, probs = myProb )
@@
     ## ncol(brca) / 20501
-    print(d <- length(genesToKeep))
+    ## print(d <- length(genesToKeep))
     ## save data as rds
-    saveRDS(brca, 'data/brca99.rds')
+    saveRDS(brca, brcaRDS)
 }
+brca <- readRDS(file = brcaRDS)
+d <- ncol(brca)
 ```
@@
 For simplicity, we begin by filtering to retain the top `r (1-myProb)*100`% highest expressing (in terms of median expression) genes of the 20,501 gene measurements from $N=`r nrow(brca)`$ patients' tumor samples, resulting in $d=`r d`$ genes.
-This results in `r choose(d,2)` pairwise dependencies among the bivariate marginals.
+This results in $`r choose(d,2)`$ pairwise dependencies among the bivariate marginals.
 To illustrate our methodology's flexible/robust simulation scheme and better align with the actual data-generating process, we remove the statistical adjustment for read alignment ambiguity (RNA-seq by Expectation Maximization; RSEM; @Li2011c) by simple rounding to integer values.
@@
 ## Describe and plot real data
-brca <- readRDS(file = './data/brca99.rds')
+if ( !file.exists( './data/exampleGenes.rds' ) )  {
     set.seed(10192020)
@@
     saveRDS(exampleGenes, file = './data/exampleGenes.rds')
-kable(round(head( brca[, exampleGenes], n = 5)),
+}
+exampleGenes <- readRDS('./data/exampleGenes.rds')
+knitr::kable(round(head( brca[, exampleGenes], n = 5)),
       booktabs = TRUE,

### 2020-11-04 050-examples.Rmd
edit overall. Cut RNA-seq MSE sizes. Revise RNA-seq example heavily.

@@
-# Example applications for our motivating data {#examples}
+# Example applications for RNA-seq data {#examples}
 
-```{r ch050-LoadLib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
-knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
+```{r ch050-LoadLib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide", cache=F}
+knitr::opts_chunk$set(echo = FALSE, cache = FALSE)
 # reticulate::use_condaenv("bigsimr")
@@
 
-This section demonstrates how to simulate high-dimensional multivariate data using `bigsimr`, aiming to replicate the structure of high dimensional dependent count data.
-In fact, simulating RNA-sequencing (RNA-seq) data is a one of the primary motivating applications of the proposed methodology, seeking scaleable Monte Carlo methods for realistic multivariate simulation (for example, see Schissler, Piegorsch, and Lussier 2018).
+This section demonstrates how to simulate multivariate data using `bigsimr`, aiming to replicate the structure of high-dimensional dependent count data.
+Simulating RNA-sequencing (RNA-seq) data is a primary motivating application of the proposed methodology, seeking scaleable Monte Carlo (MC) methods for realistic multivariate simulation for these data.
+Intergene correlation is an inherent part of biological processes.
+Yet many models do not account for this, leading to major distruptions to the operating characteristics of statistical estimation, testing, and prediction.
+See @Efron2012 for a detailed discussion with related methods and see @Wu2012b, @Schissler2018, @Schissler2019 for applied examples.
+The following subsections apply `bigsimr`'s methods to real RNA-seq data, including replicated an estimated parameteric structure, MC probability estimation, and MC correlation estimation efficiency assessment.
 
@@
 
+```{r ch050-readBRCA, echo=FALSE, eval=TRUE}
+## full workflow simulating RNA-seq data
+brca <- readRDS(file = brcaRDS)
+## smaller set for prototyping
+## d=5
+## brca <- brca[ , 1:d]
+```
+
 In an illustration of our proposed methodology applied to real data, we seek to simulate RNA-sequencing data by producing simulated random vectors with assumed marginal distributions with estimated parameters.
 Our goal is to replicate the structure of a breast cancer data set (BRCA data set from The Cancer Genome Atlas).
+Specifically, we will simulate $B=10,000$ random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$
 All these genes exhibit over-dispersion and, so, we proceed to estimate the NB parameters $(r_i, p_i), i=1,\ldots,d$ to determine the target marginal PMFs $g_i(y_i)$ (via method of moments).
-Often researchers posit a negative binomial (NB) model as counts are often over-dispersed that a Poisson model would suggest.
+Often researchers posit a negative binomial (NB) model as RNA-seq counts are often over-dispersed that a Poisson model would suggest.
 This suggests simulating high-dimensional multivariate NB (MVB) with heterogeneous marginals would be useful tool in the development and evaluation of RNA-seq analytics.
 This procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions. 
-Notably, the $\hat{p}_i's$ are small --- ranging in $[3.934 \times 10^{-6} , 1.217 \times 10^{-2}]$.
-To complete the simulation algorithm inputs, we estimate the Pearson correlation matrix $\bf{R_Y}$ and set that as the target correlation.
+Notably, the marginal NB probabilities $\hat{p}_i's$ are small --- ranging in $[`r min(probs)` , `r max(probs)`]$.
+This gives rise to highly variable counts and, typically, less restriction on potential pairwise correlation pairs.
+To complete the simulation algorithm inputs, we estimate the Spearman correlation matrix $\bf{R^{Spearman}_Y}$ and set that as the target correlation.
 
@@
 
-```{r ch050-readBRCA, echo=FALSE, eval=TRUE}
-## full workflow simulating RNA-seq data
-brca <- readRDS(file = 'data/brca99.rds')
-## smaller set for prototyping
-## d=5
-## brca <- brca[ , 1:d]
-```
-
-Let's walk through a workflow simulating RNA-seq data
+With our goal in mind, we first estimate the desired correlation matrix using the fast implementation provided by `bigsimr`:
 
@@
 
-```{r ch050-estMargins, echo = FALSE, eval = TRUE}
-## 2. Estimate NegBin parameters using Method of Moments
-## estimateNegBinMoM <- function(tmpGene, minP = 1e-7) {
-estimateNegBinMoM <- function(tmpGene) {
-    tmpMean <- mean(tmpGene)
-    tmpVar <- var(tmpGene)
-    ## relate to nbinom parameters
-    ## See ?rbinom for details.
-    p <- tmpMean / tmpVar
-    ## if (p < minP) {p <- minP } ## maybe add noise here
-    n <- ( tmpMean * p) / ( 1  - p )  
-    ## format for bigsimr margins
-    return( list("nbinom", size = n, prob = p) )
-}
-brcaMargins <- apply( unname(as.matrix(brca)), 2, estimateNegBinMoM )
-## describe the margins
-probs <- unlist( lapply( brcaMargins, function(x) {x$prob} ) )
-## summary(probs)
-sizes <- unlist( lapply( brcaMargins, function(x) {x$size} ) )
-## summary(sizes)
-## head( sort( unlist( lapply( brcaMargins, function(x) {x$size} ) ) ) )
-```
+Next we estimate the marginal parameters. 
+Here we use the basic method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model.
+We model the marginals distributions as coming from the same probability family (NB) yet are hetereogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.
+The code below features some advanced `R` utilities for concise, rapid, and generalizable target marginal specification:
 
-Here we use the basic method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model.
-We model marginals distributions as coming from the same probability family (NB) yet are hetereogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.
+<!-- 
+TYPESTE MoM estimators here.
+your comment -->
 
-TYPESTE MoM estimators here.
-
-```{r ch050-nbHelpers}
+```{r ch050-nbHelpers, echo=TRUE}
 make_nbinom_alist <- function(sizes, probs) {
@@
 
-Now specify the desired multivariate negative binomial distribution.
+Once the functions are defined to complete marginal estimation, we specify the desired multivariate negative binomial distribution and generate the desired random vectors:
 
-```{r ch050-runBRCA, echo=TRUE, results=TRUE, cache = TRUE}
+```{r ch050-estMargins, echo = TRUE, eval = TRUE}
+sizes <- apply( unname(as.matrix(brca)), 2, nbinom_mom )[1, ]
+probs <- apply( unname(as.matrix(brca)), 2, nbinom_mom )[2, ]
+```
+
+Once the functions are defined/executed to complete marginal estimation, we specify targets and generate the desired random vectors using `rvec`:
+
+```{r ch050-runBRCA-echo, echo=TRUE, eval=FALSE, cache = FALSE}
+## Set the number of random vectors
+n <- 10000
+## construct margins
+nb_margins <- make_nbinom_alist(sizes, probs)
+## run sims
+sim_nbinom <- rvec(n, nb_Rho, nb_margins, type = corType,
+                   ensure_PSD = TRUE, cores = cores) 
+colnames(sim_nbinom) <- names(brca)
+```
+
+```{r ch050-runBRCA, echo=FALSE, results=FALSE, cache = FALSE}
+brcaSimRDS <- paste0( "results/brca", round(myProb*100), "sim.rds" )
+if( !file.exists(brcaSimRDS) ) {
     ## Set the number of random vectors
@@
     colnames(sim_nbinom) <- names(brca)
+    saveRDS( sim_nbinom, brcaSimRDS )
+}
+sim_nbinom <- readRDS( brcaSimRDS )
 ```
 
-Figure X.
+Figure \@ref(fig:ch050-simDataFig) displays the simulated counts and pairwise relationships for our example genes in \@ref(fig:ch010-realDataFig).
+Simulated counts roughly mimick the observed data but with a smoothier appearance due to the assumed parameter form.
 
-```{r ch050-simDataFig, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Simulated data for 3 randomly selected high-expressing genes, replicating the real data structure describing in the Introduction.'}
+```{r ch050-simDataFig, echo = FALSE, eval=TRUE, cache=T, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Simulated data for 3 selected high-expressing genes, replicating the observed data structure.'}
 ## retrieve the examples from an earlier chapter
@@
 
-Figure X shows the results of our simulation by comparing the specified target parameter (horizontal axes) with the corresponding quantities estimated from the simulated data (vertical axes).
+Figure \@ref(fig:ch050-figBRCA) displays the aggregated results of our simulation by comparing the specified target parameter (horizontal axes) with the corresponding quantities estimated from the simulated data (vertical axes).
 The evaluation shows that the simulated counts approximately match the target parameters and exhibit the full range of estimated correlation from the data.
-Utilizing 15 CPU threads in a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, the simulation completed in less than 30 seconds.
+Utilizing 15 CPU threads in a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, the simulation completed just shy of 7.5 minutes.
 
-```{r ch050-figBRCA, echo=F, eval=T, fig.cap="bigsimr produces simulated random vectors from a multivariate negative binomial that replicate the estimated structure from an RNA-seq data set.
-The dashed red lines indicated equality between estimated parameters (vertical axes; derived from the simulated data) and the specified target parameters (horizontal axes)."}
+```{r ch050-figBRCA, echo=F, eval=T, cache=T, fig.cap="`bigsimr` produces simulated random vectors from a multivariate negative binomial that replicate the estimated structure from an RNA-seq data set.
+The dashed red lines indicated equality between estimated parameters from simulated data(vertical axes) and the specified target parameters (horizontal axes)."}
 par(mfrow=c(1,3))
@@
 
-To conduct statistical inference, via maximum likelihood for example, a critical task is to evaluate the joint probability mass (or density) function:
+To conduct statistical inference a critical task is to evaluate the joint probability mass (or density) function:
 
 $$
-P( \bf{Y} = \bf{y} ), y_i \in \chi_i.
+P( {\bf Y} = {\bf y} ), y_i \in \chi_i.
 $$
@@
 Given a large number simulated vectors as produced above, estimated probabilities are readily given by counting the proportion of simulated vectors meeting the desired condition.
-In our motivating application, one may ask what is the probability that all genes expressed greater than a certain threshold value $\bf{y_0}$.
+In our motivating application, one may ask what is the probability that all genes expressed greater than a certain threshold value ${ \bf y}_0$.
 
@@
 $$
-\hat{P}( \bf{Y} >= \bf{ y_0 } ) = \sum_{b=1}^B I( \bf{Y^{(b)}} > \bf{ y_0 } ) / B
+\hat{P}( {\bf Y} >= {\bf y_0 } ) = \sum_{b=1}^B I( {\bf Y^{(b) }} > {\bf y_0 } ) / B
 $$
 
-where $\bf{Y^{(b)}}$ is the $b^{th}$ simulated vector in a total of $B$ simulation replicates and $I(\dot)$ is the indicator function.
-For example, we can estimate from our $B=10,000$ simulated vectors the probability that all genes express more than $\bf{y_0} = 1000$.
+where ${\bf Y^{(b)} }$ is the $b^{th}$ simulated vector in a total of $B$ simulation replicates and $I ( )$ is the indicator function.
+For example, we can estimate from our $B=10,000$ simulated vectors the probability that all genes are expressed (i.e., ${\bf y}_i \geq 1, \forall \; i$).
 
-```{r ch050-densityEvaluation, echo=TRUE, eval=TRUE}
+```{r ch050-densityEvaluation, echo=TRUE, eval=TRUE, cache=F}
 d <- ncol(sim_nbinom)
 B <- nrow(sim_nbinom)
-threshold <- rep( 1000, d)
+threshold <- rep( 1, d)
 mean(apply( sim_nbinom,  1,
@@
 
+<!-- 
 ## Monte Carlo estimation of MSE
@@
 
-```{r ch050-mom-mse, echo=TRUE, eval=TRUE, cache=FALSE}
+```{r ch050-mom-mse, echo=TRUE, eval=FALSE, cache=FALSE}
 ## system( "rm results/mseSizes.rds")
-if ( !file.exists( 'results/mseSizes.rds') )  {
+brcaSimMSEsizesRDS <- paste0( "results/brca", round(myProb*100), "simMSEsizes.rds" )
+if ( !file.exists( brcaSimMSEsizesRDS) )  {
     n <- nrow(brca)
@@
     momSizes <- replicate(n = m, expr =
-                                     { tmpSim <- rvec(n, nb_Rho, nb_margins, type = corType, ensure_PSD = TRUE, cores = cores)
+                                     { tmpSim <- rvec(n, nb_Rho, brcaMargins, type = corType, ensure_PSD = TRUE, cores = cores)
                                          tmpMom <- apply( unname(as.matrix(tmpSim)), 2, estimateNegBinMoM )
@@
     mseSizes <- rowMeans( sqDiffSizes )
-    saveRDS(mseSizes, 'results/mseSizes.rds')
+    saveRDS(mseSizes, brcaSimMSEsizesRDS)
 }
+mseSizes <- readRDS( brcaSimMSEsizesRDS )
 ```
@@
 
-```{r ch050-fig-mseSizes, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Real data for the 3 selected high-expressing genes'}
-mseSizes <- readRDS(file = 'results/mseSizes.rds')
+```{r ch050-fig-mseSizes, echo = FALSE, eval=FALSE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Real data for the 3 selected high-expressing genes'}
+mseSizes <- readRDS(file = brcaSimMSEsizesRDS)
 qplot(mseSizes, bins = 20) + theme_bw()
@@
 One could then compare the MSE of the baisc MoM estimation scheme to other strategies ( such as MLE, trimmed approaches, etc.) and decide whether a more computational costly approach is warrented or if the MoM will suffice.
+
+your comment -->

### 2020-11-04 index.Rmd
edit overall. Cut RNA-seq MSE sizes. Revise RNA-seq example heavily.

@@
 subtitle: "using the bigsimr R Package"
-author:
-  - name: Alexander D. Knudson, Tomasz J. Kozubowski, Anna K. Panorska, Juli Petereit, and Alfred G. Schissler
-bibliography:
-  - bigsimr.bib
-  - packages.bib
-site: bookdown::bookdown_site
-#output: distill::distill_article
-abstract: |
-  It is critical to realistically simulate data when conducting Monte Carlo studies and methods.
+author: "Alexander D. Knudson, Tomasz J. Kozubowski, Anna K. Panorska, Juli Petereit, and Alfred G. Schissler"
+bibliography: ["bigsimr.bib", "packages.bib"]
+abstract:  It is critical to realistically simulate data when conducting Monte Carlo studies and methods.
 But measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments.
@@
 library(knitr)
+knitr::opts_chunk$set(eval = TRUE, echo = FALSE, cache = TRUE)
 library(rmarkdown)

### 2020-11-04 050-examples.Rmd
revise order of code chunks/text so have dynamic inline R
draft quad loss example

@@
 See @Efron2012 for a detailed discussion with related methods and see @Wu2012b, @Schissler2018, @Schissler2019 for applied examples.
-The following subsections apply `bigsimr`'s methods to real RNA-seq data, including replicated an estimated parameteric structure, MC probability estimation, and MC correlation estimation efficiency assessment.
+The following subsections apply `bigsimr`'s methods to real RNA-seq data, including replicated an estimated parameteric structure, MC probability estimation, and MC evaluation of correlation estimation efficiency.
 
@@
 ## full workflow simulating RNA-seq data
+## brcaRDS <- 'data/brca99.rds'
 brca <- readRDS(file = brcaRDS)
@@
 This procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions. 
-Notably, the marginal NB probabilities $\hat{p}_i's$ are small --- ranging in $[`r min(probs)` , `r max(probs)`]$.
-This gives rise to highly variable counts and, typically, less restriction on potential pairwise correlation pairs.
-To complete the simulation algorithm inputs, we estimate the Spearman correlation matrix $\bf{R^{Spearman}_Y}$ and set that as the target correlation.
+To specify the simulation algorithm inputs, we estimate the Spearman correlation matrix $\bf{R^{Spearman}_Y}$ and marginal NB parameters.
 
@@
 ```{r ch050-estRhoBRCA, echo=TRUE, eval=TRUE}
-## 1. Estimate Spearman's correlation on the count data
-## corType <- 'pearson'
+## Estimate Spearman's correlation on the count data
 corType <- 'spearman'
@@
 
+Notably, the marginal NB probabilities $\hat{p}_i's$ are small --- ranging in $[`r min(probs)` , `r max(probs)`]$.
+This gives rise to highly variable counts and, typically, less restriction on potential pairwise correlation pairs.
 Once the functions are defined/executed to complete marginal estimation, we specify targets and generate the desired random vectors using `rvec`:
@@
 ```{r ch050-runBRCA, echo=FALSE, results=FALSE, cache = FALSE}
+nb_margins <- make_nbinom_alist(sizes, probs)
 brcaSimRDS <- paste0( "results/brca", round(myProb*100), "sim.rds" )
@@
     ## construct margins
-    nb_margins <- make_nbinom_alist(sizes, probs)
     ## run sims
@@
 
+## MC evaluation of correlation estimation efficiency
+
+```{r ch010-myM, include=FALSE, cache = TRUE}
+m <- 100
+```
+
+MC methods are routinely used in many statistical inferential tasks including estimation, hypothesis testing, error rates, and empirical interval coverage rates.
+For an concise introduction to these methods, see, for example @Rizzo2007, Ch. 6.
+Now, we demonstrate how `bigsimr` can be used evaluate estimation efficiency.
+In particular, we'd like to assess the error our correlation estimation above.
+We used a conventional method, based on classical statistical theory. 
+Yet this method was not designed for high-dimensional data.
+Indeed, high-dimensional covariance estimation (and precision matrices) is an active area of statistical science (see, for example, @Won2013g and @VanWieringen2016).
+In this example, we simulate $m=`r m`$ data sets with the number of simulated vectors matching the number of patients in the BRCA data set, $N=`r nrow(brca)`$.
+At each iteration we estimate the quadratic loss from the specified $\bf{R^{Spearman}_Y}$ and average over the $m$ iteration.
+
+```{r ch050-quadloss, echo=TRUE, eval=TRUE, cache=FALSE}
+quadLossRDS <- paste0( "results/brca", round(myProb*100), "quadLoss.rds" )
+if ( !file.exists( quadLossRDS) )  {
+    n <- nrow(brca)
+    simRho <- replicate(n = m, expr = { tmpSim <- rvec(n = nrow(brca) , nb_Rho, nb_margins,
+                                                       type = corType, ensure_PSD = TRUE, cores = cores)
+                                                       tmpRho <- bigsimr::cor_fast( x = tmpSim )} ,
+                        simplify = FALSE)
+
+    quadLoss <- unlist( lapply( simRho, rags2ridges::loss, T = nb_Rho, type = "quadratic"  ) )
+    saveRDS(quadLoss, quadLossRDS)
+}
+quadLoss <- readRDS( quadLossRDS )
+```
+
+The standard `R` summary function supplies the mean-augmented five-number summary of the quadratic loss distribution computed above.
+
+```{r ch050-quadloss-summary, echo=TRUE, eval=TRUE, cache=FALSE}
+summary(quadLoss)
+```
+
+This distribution could be compared to more elaborate, high-dimensional estimators to guide analysts in deciding whether the additional complexity and computation time are warranted for their application.
+
 <!-- 

### 2020-11-05 index.Rmd
2nd draft: revise abstract and introduction

@@
 This article introduces the `bigsimr` R package that provides a flexible, scaleable procedure to simulate high-dimensional random vectors with given marginal characteristics and dependency measures.
-We'll describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation matrices, and find close positive semi-definite matrices.
+We'll describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and find close positive semi-definite matrices.
 Finally, we demonstrate the power of `bigsimr` by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d>1000$.
@@
 Massive high-dimensional data sets are now commonplace in many areas of scientific inquiry.
-This is particularly true when simulating massive *multivariate*, *non-normal* distributions, arising naturally in many fields of study in this era of big data.
 As new methods are developed for these data, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness --- often in comparison to existing methods.
 Further, efficient simulation empowers statistical computing strategies, such as the parameteric bootstrap [@Rizzo2007] to simulate from a hypothesized null model, providing inference in analytically challenging settings.
-Such Monte Carlo techniques become difficult for high-dimensional data with the current existing algorithms and tools.
+Such Monte Carlo (MC) techniques become difficult for high-dimensional data with the current existing algorithms and tools.
+This is particularly true when simulating massive *multivariate*, *non-normal* distributions, arising naturally in many fields of study.
 
 As others have noted, it can be vexing to simulate dependent, non-normal/discrete data ---  even for low dimensional settings [@MB13; @XZ19].
-For continous non--normal data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula [@Nelsen2007] approaches are well-studied with flexible, robust software available [@Yan2007; @Chen2001].
-Yet these approaches do not scale in a timely fashion to high-dimensional problems (see @Li2019gpu). 
-For discrete data, multiple techniques have been proposed.
-But some schemes have major flaws such as failing to obtain the full range of possible dependencies (for example, feasible for only positive correlations @Park1996). 
-While more recent approaches [@MB13; @Xia17; @BF17] have largely rememdied this issue for low dimesional problems, the existing tools are not designed to scale to high dimensions.
+For continous non--normal multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula [@Nelsen2007] approaches are well-studied with flexible, robust software available [@Yan2007; @Chen2001].
+Yet these approaches do not scale in a timely fashion to high-dimensional problems [@Li2019gpu]. 
+For discrete data, early simulation strategies had major flaws --- such as failing to obtain the full range of possible dependencies (e.g., admitting only positive correlations @Park1996). 
+While more recent approaches [@MB13; @Xia17; @BF17] have largely rememdied this issue for low-dimesional problems, the existing tools are not designed to scale to high dimensions.
 
@@
 <!-- Here we propose a scheme that ignores the analytic goal --- and, instead, aims to match the empirically-derived estimated correlation data and marginal characteristics.  -->
-For normal data, the Pearson product-moment correlation describes the dependency completely.
+For normal data, the Pearson product-moment correlation describes the dependency perfectly.
 As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix exactly is computationally intense [@Chen2001; @Xia17]. 
-On the other hand, an analyst may consider the use of nonparametric correlation measures to better characterize monotone dependency, such as Spearman's $\rho$ and Kendall's $\tau$.
-Throughout, we'll emphasize matching these nonparametric dependency measures as our aim lies in modeling non-normal data.
+On the other hand, an analyst may consider the use of nonparametric correlation measures to better characterize monotone, non-linear dependency, such as Spearman's $\rho$ and Kendall's $\tau$.
+Throughout, we'll emphasize matching these nonparametric dependency measures as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology.
 
-In the study, we present scalable, flexible multivariate simulation algorithms. 
+With all this mind, we present a scalable, flexible multivariate simulation algorithm. 
 The crux of the method lies in the construction of a Gaussian copula, in the spirit of the NORTA procedure.
-Further, we introduce the `bigsimr` R package that provides parallelized, high-performance software.
-The algorithm design relies on useful properties of nonparameteric correlation measures, namely invariance under monotone transformation and well-known closed form relationships between dependency measures for the multivariate normal (MVN) distribution.
+Further, we introduce the `bigsimr` R package that provides parallelized, high-performance software implemented our NORTA-inspirsed algorithm.
+The algorithm design leverages useful properties of nonparameteric correlation measures, namely invariance under monotone transformation and well-known closed form relationships between dependency measures for the multivariate normal (MVN) distribution.
 
-The study proceeds with background and notation, including a description of our motivating example in RNA-sequencing (RNA-seq) breast cancer data.
+The study proceeds by providing background information, including a description of our motivating example application --- RNA-sequencing (RNA-seq) breast cancer data.
 Then we describe and justify our simulation methodology and related algorithms.
-Next, a detailed illustrative low-dimensional example of basic and advanced use of the `bigsimr` R package is provided.
-Then we proceed with extensive Monte Carlo studies under various distributional assumptions.
-After evaluating *in silico*, we revisit our high-dimensional motivating RNA-seq example and employ our methods to commonplace statistical computing tasks.
+Next, we detail an illustrative low-dimensional example of basic and advanced use of the `bigsimr` R package.
+Then we proceed with Monte Carlo studies under various distributional assumptions to assess accuracy and scaleable.
+After the MC evaluations, we revisit our high-dimensional motivating example and employ our methods to commonplace statistical computing tasks.
 Finally, we'll discuss the method's utility, limitations, and future directions.

### 2020-11-05 040-monte-carlo-evaluation.Rmd
2nd draft. Combine bivariate plots and create unifying figure

@@
 We vary $\rho$ across the entire possible range of correlations for each correlation type.
+<!-- 
 Figure \@ref(fig:ch040-biNormPlot) `bigsimr` recovers the Pearson specified correlations for MVN.
+ -->
 
-```{r ch040-biNormPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.height = 5, fig.width= 8, fig.align='center', fig.cap = "`bigsimr` recovers the Pearson specified correlations for MVN."}
+```{r ch040-biNormPlot, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, fig.height = 5, fig.width= 8, fig.align='center', fig.cap = "`bigsimr` recovers the Pearson specified correlations for MVN."}
 dat <- readRDS("./results/biNorm_sims.rds")
@@
 We vary $\rho$ across the entire possible range of correlations for each correlation type.
-Figure \@ref(fig:ch040-biGammaPlot) `bigsimr` recovers the Pearson specified correlations for bivariate gamma.
+<!--  Figure \@ref(fig:ch040-biGammaPlot) `bigsimr` recovers the Pearson specified correlations for bivariate gamma. -->
 
-```{r ch040-biGammaPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.height = 5, fig.width= 8, fig.align='center', fig.cap = "`bigsimr` recovers the Pearson specified correlations for Bivariate Gamma."}
+```{r ch040-biGammaPlot, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, fig.height = 5, fig.width= 8, fig.align='center', fig.cap = "`bigsimr` recovers the Pearson specified correlations for Bivariate Gamma."}
 dat <- readRDS("./results/biGamma_sims.rds")
@@
 We vary $\rho$ across the entire possible range of correlations for each correlation type.
-Figure \@ref(fig:ch040-biNegBinPlot) `bigsimr` recovers the Pearson specified correlations for bivariate negative binomial.
+<!-- Figure \@ref(fig:ch040-biNegBinPlot) `bigsimr` recovers the Pearson specified correlations for bivariate negative binomial. -->
 
-```{r ch040-biNegBinPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.align='center', fig.width = 8, fig.height=5,fig.cap = "`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
+```{r ch040-biNegBinPlot, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.align='center', fig.width = 8, fig.height=5,fig.cap = "`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
 dat <- readRDS("./results/biNB_sims.rds")
@@
 
-```{r ch040-biMSEtable}
+```{r ch040-combineBiSims}
 ## compare relative differences in MSE for each dependency, compared to Pearson
@@
 allDat$margins <- factor(allDat$margins, levels = c( 'norm', 'gamma', 'nbinom') )
-allDat$type <- factor(allDat$type, levels = c( 'kendall', 'spearman', 'pearson' ) )
+allDat$type <- factor(allDat$type, levels = c( 'pearson', 'kendall', 'spearman' ) )
 allDat$N <- factor(allDat$N, levels = c( 1000, 10000, 100000) )
+```
+
+```{r ch040-bPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.asp = 1.30, fig.weight = 4, fig.align='center', fig.cap = "Bivariate simulations match specified correlations.
+The horizontal axis plots the specified target correlations across the entire range of possible correlations for each bivariate margin.
+Normal margins are plotted in dark grey, gamma in medium grey, and negative binomial in light grey.
+As the number of simulated vectors $B$ increases from left to right, the variation in estimated correlations (vertical axis) decreases.
+The dashed line indicates equality between the specified and estimated correlations.
+Only the normal margins match the Pearson correlations (top row) exactly across the entire range of correlations.
+The two non-normal bivariate random vectors experience attentuation (downward bias) for extreme Pearson correlations (due to limitations in our algorithm) and more restriction (due to the Frechet bounds).
+The rank-based correlations (bottom two rows) are matched exactly for all margins and obtain the full range of possible dependencies."}
+# https://www.datanovia.com/en/blog/how-to-change-ggplot-facet-labels/
+# New facet label names
+repsLabs <- paste0("B=", c("1,000", "10,000", "100,000") )
+names(repsLabs) <- c(1000, 10000, 100000)
+typeLabs <- c( 'Pearson', 'Kendall', 'Spearman' ) 
+names(typeLabs) <- c( 'pearson', 'kendall', 'spearman' )
+
+# Set colors
+## RColorBrewer::display.brewer.all()
+numColors <- 4
+numGroups <- length(levels(allDat$margins))
+myColors <- rev( RColorBrewer::brewer.pal(n = numColors, name = 'Greys')[ ((numColors - numGroups) + 1): numColors  ] )
+
+allDat %>%
+    ## ggplot(aes(x = rho, y = rho_hat, color = margins, shape = margins)) +
+    ggplot(aes(x = rho, y = rho_hat, color = margins)) +
+    ## ggplot(aes(rho, rho_hat)) +
+    ## geom_point(alpha = 1, size = 2) +
+    geom_point(size = 2) +
+    ## scale_shape_manual(values=c(1, 2, 3))+
+    ## scale_color_manual(values=c('#999999','#E69F00', '#56B4E9')) +
+    scale_color_manual(values = myColors ) +
+    geom_abline(slope = 1, linetype = 'dashed') +
+    labs(x = "Specified Correlation", y = "Estimated Correlation") +
+    ## facet_wrap(~ + N) + theme_bw()
+    facet_wrap(~ type + N, labeller = labeller(N = repsLabs, type = typeLabs)) +
+    theme_bw() +
+    theme(legend.position = "bottom", legend.direction = "horizontal")
+
+## allDat %>%
+##     ggplot(aes(x = rho, y = rho_hat, color = type, shape = type)) +
+##     ## ggplot(aes(rho, rho_hat)) +
+##     geom_point(alpha = 0.75) +
+##     geom_abline(slope = 1) +
+##     labs(x = "Specified Correlation", y = "Estimated Correlation") + 
+##     ## facet_wrap(~ + N) + theme_bw()
+##     facet_wrap(~ margins + N) + theme_bw()
+
+## ggsave('fig/plot-biNormPlot.pdf')
+```
+
+```{r ch040-biMSEtable}
 ## tabMSE  <- allDat %>%

### 2020-11-05 040-monte-carlo-evaluation.Rmd
2nd draft. polish bivariate plot

@@
         ## I'll create a minimal working example and open an issue
-        ## tmp_bounds <- cor_bounds(margins, type = type)
-        ## cor_lo <- tmp_bounds$lower[1,2] + eps
-        ## cor_hi <- tmp_bounds$upper[1,2] - eps
-        cor_lo <- -0.75
-        cor_hi <- 0.95
+        ## work around for now
+        if ( type == 'pearson' ) {
+            tmp_bounds <- cor_bounds(margins, type = type)
+            cor_lo <- tmp_bounds$lower[1,2] + eps
+            cor_hi <- tmp_bounds$upper[1,2] - eps
+        } else {
+            cor_lo <- -1 + eps
+            cor_hi <- 1 - eps
+        }
         cor_seq <- seq(cor_lo, cor_hi, length.out = grid_steps)
@@
 
-```{r ch040-combineBiSims}
+```{r ch040-combineBiSims, cache=F}
 ## compare relative differences in MSE for each dependency, compared to Pearson
@@
 
-```{r ch040-bPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.asp = 1.30, fig.weight = 4, fig.align='center', fig.cap = "Bivariate simulations match specified correlations.
+```{r ch040-bPlot, cache=F, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.asp = 1.30, fig.weight = 4, fig.align='center', fig.cap = "Bivariate simulations match specified correlations.
 The horizontal axis plots the specified target correlations across the entire range of possible correlations for each bivariate margin.
@@
 numGroups <- length(levels(allDat$margins))
-myColors <- rev( RColorBrewer::brewer.pal(n = numColors, name = 'Greys')[ ((numColors - numGroups) + 1): numColors  ] )
+## myColors <- rev( RColorBrewer::brewer.pal(n = numColors, name = 'Greys')[ ((numColors - numGroups) + 1): numColors  ] )
+myColors <- rev( RColorBrewer::brewer.pal(n = numColors, name = 'Blues')[ ((numColors - numGroups) + 1): numColors  ] )
 

### 2020-11-06 010-background.Rmd
2nd. Draft. Revise background

@@
 
-The article presents several algorithms to work with high-dimesional multivariate data, but all `bigsimr` algorithms were originally designed to support a single task:
-to generate random vectors drawn from multivariate probability distribution with given marginal distributions and component-wise dependency metrics.
-Specifically, our goal is to efficiently simulate a large number, $B$, of random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with **correlated** components and hetereogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i$, and $d$ can be very large.
+The `bigsimr` `R` package provides multiple algorithms to work with high-dimesional multivariate data, but all these algorithms were originally designed to support a single task:
+to generate random vectors drawn from multivariate probability distribution with given marginal distributions and dependency metrics.
+Specifically, our goal is to efficiently simulate a large number, $B$, of random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with **correlated** components and hetereogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i$, where $d$ can be very large and still be computed in practically useful times.
 
 When designing this methodology, we developed the following properties to guide our effort.
-We divide the properties into two categories: (1) basic properties (**BP**) and "scaleability" properties (**SP**). 
+We divide the properties into two categories: (1) basic properties (BP) and "scaleability" properties (SP). 
 The BPs are adapted from an existing criteria due to @Nik13a.
@@
 
-To fix ideas and provide example applications enabled via `bigsimr`, the next section describes a motivating data set that originally inspired the authors` interest in developing this methodology.
+To fix ideas and provide examples applications enabled via `bigsimr`, the next section describes a motivating data set that originally inspired the authors' interest in developing this methodology.
 
@@
 RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data.
-This results in count data corresponding to over 20,000 genes (coding genomic regions) or even over 77,000 isoforms when alternating spliced mRNA are counted (for human derived samples).
-Yet due to inherent biological processes, gene expression data exhibits correlation (co-expression) across genes [@BE07; @Schissler2018].
+For human derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted.
+Importantly, due to inherent biological processes, gene expression data exhibits correlation --- co-expression --- across genes [@BE07; @Schissler2018].
 
 ```{r ch010-myProb, include=FALSE, cache = TRUE}
-myProb <- 0.99
+myProb <- 0.95
 ```
@@
 
-Specifically, we'll illustrate our methodology using the Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgements).
-For simplicity, we begin by filtering to retain the top `r (1-myProb)*100`% highest expressing (in terms of median expression) genes of the 20,501 gene measurements from $N=`r nrow(brca)`$ patients' tumor samples, resulting in $d=`r d`$ genes.
-This results in $`r choose(d,2)`$ pairwise dependencies among the bivariate marginals.
-To illustrate our methodology's flexible/robust simulation scheme and better align with the actual data-generating process, we remove the statistical adjustment for read alignment ambiguity (RNA-seq by Expectation Maximization; RSEM; @Li2011c) by simple rounding to integer values.
-Table \@ref(tab:ch010-realDataTab) displays counts for three selected genes for the first five patients' breast tumor samples.
+We'll illustrate our methodology using the well-studied Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgements).
+For simplicity, we only consider high expressing genes.
+In turn, we begin by filtering to retain the top `r (1-myProb)*100`% highest- expressing genes (in terms of median expression) of the 20,501 gene measurements from $N=`r nrow(brca)`$ patients' tumor samples, resulting in $d=`r d`$ genes.
+This gives a massive number of pairwise dependencies among the marginals (specifically, $`r choose(d,2)`$ correlation parameters).
+We further process the data to illustrate our methodology's flexible and robust simulation scheme, while better aligning with the actual data-generating process, by rounding @Li2011c's RNA-seq by Expectation Maximization (RSEM) values to counts.
+Table \@ref(tab:ch010-realDataTab) displays counts for three selected high-expressing genes for the first five patients' breast tumor samples.
 
@@
 
-Figure \@ref(fig:ch010-realDataFig) plots the marginal distributions and estimated Spearman's correlations (see Equation \@ref(eq:spearman) below).
+To help visualize the bivariate relationships for these three selected genes across all patients, Figure \@ref(fig:ch010-realDataFig) plots the marginal distributions and estimated Spearman's correlations (see Equation \@ref(eq:spearman) below).
 
-```{r ch010-realDataFig, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's $rho$ for three example genes.
+```{r ch010-realDataFig, cache=F, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's $\rho_s$ for three example genes.
 The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations.
@@
 For non-normal marginals with monotone correlation patterns, $\rho_P$ suffers some drawbacks and may mislead or fail to capture important relationships (@MK01).
-Alternatively, in these settings, analysts often prefer rank-based correlation measures to describe the degree of monotonic association.
+Alternatively in these settings, analysts often prefer rank-based correlation measures to describe the degree of monotonic association.
 
 Two nonparameteric, rank-based measures common in practice are Spearman's correlation (denoted $\rho_S$) and Kendall's $\tau$. 
-Formally, define
+Define
 
@@
 (\#eq:spearman)
-\rho_S(X,Y) = 3 \left[ P\left[ (X_1 - X_2)(Y_1-Y_3) > 0 \right] - P\left[ (X_1 - X_2)(Y_1-Y_3) < 0 \right] \right]
+\rho_S(X,Y) = 3 \left[ P\left[ (X_1 - X_2)(Y_1-Y_3) > 0 \right] - P\left[ (X_1 - X_2)(Y_1-Y_3) < 0 \right] \right],
 \end{equation}
@@
 \noindent where $(X_1, Y_1) \overset{d}{=} (X,Y), X_2 \overset{d}{=} X, Y_3 \overset{d}{=} Y$ with $X_2$ and $Y_3$ are independent of one other and of $(X_1, Y_1)$. 
-Spearman's $\rho_S$ has an appealing correspondence as the Pearson's correlaiton coefficient on *ranks* of the values, thereby captures nonlinear yet monotone relationships.
+Spearman's $\rho_S$ has an appealing correspondence as the Pearson's correlation coefficient on *ranks* of the values, thereby captures nonlinear yet monotone relationships.
 
@@
 Both $\tau$ and $\rho_S$ are **invariant to under monotone transformations** of the underlying random variates.
-As we will describe more fully in the [Algorithms](#algorithms) section, this property is preferred in our scheme to satistify properties SP1 and SP2, above.
+As we will describe more fully in the [Algorithms](#algorithms) section, this property is essential to scaleable match rank-based correlations with speed (SP1) and accuracy (SP2).
 
@@
 
-Also we'll note that for a discrete random variable $Y_i$, some care must be taken to define the quantile function $F_{i}^{-1}$.
+Also, for a discrete random variable $Y_i$, some care must be taken to define the quantile function $F_{i}^{-1}$.
 Let
@@
 The so-called *Frechet-Hoeffding bounds* are well-studied (for example, see @Nelsen2007 and @BF17).
-This situation gives strict restraints on the possible bounds and cannot be overcome through algorithm design.
-In general the bounds are given by
+This situation gives strict restraints on the possible correlations and cannot be overcome through algorithm design.
+In general, the bounds are given by
 
@@
 
-There is a strong connection of our simulation strategy to Gaussian **copulas** (for a technical introduction see @Nelsen2007).
+There is a strong connection of our simulation strategy to Gaussian **copulas** (see @Nelsen2007 for a technical introduction).
 A copula is a distribution function on $[0,1]^d$ that describes a multivariate probability distribution with standard uniform marginals.

### 2020-11-06 index.Rmd
2nd. Draft. Revise background

@@
 We'll describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and find close positive semi-definite matrices.
-Finally, we demonstrate the power of `bigsimr` by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d>1000$.
+Finally, we demonstrate the power of `bigsimr` by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d =1026$.
 ---
@@
 
-```{r indexSetup, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
+```{r indexSetup, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide", cache=F}
 library(tidyverse)

### 2020-11-06 020-simulation-algorithm.Rmd
2nd. Draft. Revise Algorithms

--- 2020-10-27 020-simulation-algorithm.Rmd
@@
 Each $Y_i$ has a specified marginal CDF $F_i$ and its inverse $F^{-1}_i$.
-To characterize dependency, every pair $(Y_i, Y_j)$ has a given Pearson correlation $\rho_P$, (rescaled) Spearman correlation $\rho_S$, and/or Kendall's $\tau$.
-The method is best understand as a **high-performance Gaussian copula** (Equation \@ref(eq:gauss)) providing a high-dimensianl NORTA-inspired algorithm.
+To characterize dependency, every pair $(Y_i, Y_j)$ has a given Pearson correlation $\rho_P$, Spearman correlation $\rho_S$, and/or Kendall's $\tau$.
+The method is best understand as a **high-performance Gaussian copula** (Equation \@ref(eq:gauss)) providing a high-dimensional NORTA-inspired algorithm.
 
@@
 2. Determine the input matrix $\Sigma_{\bf Z}$ that corresponds with the specified output $\Sigma_{\bf Y}$.
-3. Produce the Cholesky factor $M$ of $\Sigma_{\bf Z}$ so that $M M^{\prime}=\Sigma_{\bf Z}$.
+3. Produce a Cholesky factor $M$ of $\Sigma_{\bf Z}$ so that $M M^{\prime}=\Sigma_{\bf Z}$.
 4. Set $X$ by $X \gets MZ$.
@@
 
-With modern parallelized computing, steps 1, 3, 4, 5 are readily implemented as high-performance (multicore and/or graphical-processing-unit (GPU) acceleration) algorithms, providing the greatest scaleability in a practical manner.
+With modern parallelized computing, steps 1, 3, 4, 5 are readily implemented as high-performance, multicore and/or graphical-processing-unit (GPU) accelerated, algorithms --- providing the fast scaleability using readily-available hardware.
 
@@
 In general, there is no closed form correspondence between the components of the input $\Sigma_{\bf Z}$ and target $\Sigma_{\bf Y}$.
-Matching the correlations involves evaluating or approximating $\binom{d}{2}$ integrals of the form $EY_iY_j = \int \int y_i y_j f_{X|r}(F_i^{-1}(\Phi(z_i)), F_j^{-1}(\Phi(z_j))dy_idy_j$, for $i,j=1,2,\ldots,d$.
-For high-dimensional data, these evaluations are often too costly to enable feasible comprehensive MC studies.
-For low-dimensional problems, methods and tools exist to match exactly (see @Chen2001; @Xia17; @MB13), including the publicly available `nortaRA` R package.
+Matching the correlations involves evaluating or approximating $\binom{d}{2}$ integrals of the form $EY_iY_j = \int \int y_i y_j f_{X}(F_i^{-1}(\Phi(z_i)), F_j^{-1}(\Phi(z_j))dy_idy_j$, for $i,j=1,2,\ldots,d, \; i \neq j$.
+For high-dimensional data, these evaluations are often too costly to enable feasible simulation studies.
+For low-dimensional problems, methods and tools exist to match Pearson correlations precisely (see @Chen2001; @Xia17; @MB13), including the publicly available `nortaRA` R package.
 
-Our solution is to essentially avoid this complication in Pearson matching.
+To maintain scaleability (SP1), our solution is to essentially avoid this complication in Pearson matching.
 Since our goal is to simulate non-normal marginals, we greatly prefer the use of rank-based measures $\rho_S$ and $\tau$ from a modeling standpoint.
-Further, $\rho_S$ and $\tau$ invariance under monotone transformation (see [Background](#background)), perserves the correlation coefficients through steps 3, 4, and 5 in the NORTA algorithm above.
+Further, $\rho_S$ and $\tau$'s invariance under monotone transformation (see [Background](#background)), perserves the correlation coefficients through steps 3, 4, and 5 in the NORTA algorithm above.
 This eliminates the need for computing the $\binom{d}{2}$ integrals to match exactly (nothing is for free, however, as discussed in below in Section \@ref(rand-vec-gen)).
 
-Despite all this, if one does desire to characterize dependency using Pearson correlations, simply using the target Pearson correlation matrix as the initial conditions to our proposed algorithm will lead to approximate matching in the resultant distribution (@Song00).
-The quality of this approximation depends on the modeling goal, but in practice, for high-dimensional count data we find the accuracy to be adequate.
+Despite all this, if one does desire to characterize dependency using Pearson correlations, simply using the target Pearson correlation matrix as the initial conditions to our proposed algorithm will lead to approximate matching in the resultant distribution (@Song00) in many practical applications.
+The quality of this approximation depends on the setting, but in practice, for high-dimensional count data we find the accuracy to be adequate.
 Later, we'll study the robustness of our method to this limitation in selected [Monte Carlo evalutations]({#simulations}).
 
-## Random vector generation via `bigsimr::rvec()` {#rand-vec-gen}
+## Random vector generation via `bigsimr::rvec` {#rand-vec-gen}
 
-Now we describe `bigsimr::rvec()`, our algorithm to generate random vectors.
+Now we describe `bigsimr::rvec`, our algorithm to generate random vectors.
 It mirrors the classical NORTA algorithm above with some modifications for rank-based dependency matching:
@@
 	(ii) Check that ${\bf R_{Pearson}}$ is semi-positive definite.  
-	(iii) If not compute a close semi-positive ${\bf \widetilde{R}_{Pearson}}$.  
+	(iii) If not compute a close semi-positive definite correlation matrix ${\bf \widetilde{R}_{Pearson}}$.  
 2. Gaussian copula construction.  
-	(i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_{Pearson}})$ ;  
-	(ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$;  
+	(i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_{Pearson}})$.  
+	(ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
 3. Quantile evaluations.  
-	Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;
+	(i) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;
 
-The pre-processing (Step1) takes advantage of the closed form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertSpearman) or \@ref(eq:convertKendall), respectively (implemented as `bigsimr::cor_covert()`).
+The pre-processing (Step1) takes advantage of the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertSpearman) or \@ref(eq:convertKendall), respectively (implemented as `bigsimr::cor_covert`).
 
-A complication often arises at this stage: the parallelized pairwise conversions may become non-positive definite, especially in high dimensions.
-Researchers working in multivariate computation frequently encounter such difficulties and need to find a closest positive definiteness matrix.
-The most widely used routine for this task in R is called `matrix::nearPD()` and is not suitable for high dimensions.
-To overcome this issue, we've implemented `bigsimr::nearPSD()`, a quadractically convergent Netwon method for finding the nearest correlation matrix introduced by @QS2006.
+A complication often arises at this stage: the parallelized, pairwise conversions may create a (spuriously) non-positive definite correlation matrix, especially in high dimensions.
+Researchers working in multivariate computation frequently encounter such difficulties and need to find a close positive (semi-)definiteness matrix.
+The most widely-available routine for this task in `R` is called `matrix::nearPD` and is not suitable for high dimensions.
+To overcome this issue, we've developed `bigsimr::nearPSD`, a quadractically-convergent Netwon method for finding the nearest correlation matrix, developed by @QS2006.
 We hope that this routine could be useful in many applications aside from our primary goal of random vector generation.
@@
 Once the target margins and algorithm inputs are determined, steps 2 and 3 are essentially a NORTA algorithm with modern high-performance computing implementations.
-Specifically, step 2i uses either an efficient multicore multivariate normal simulator `mvnfast`[@Fasiolo2016] or a using Google's JAX python library for graphical-processing-unit (GPU) acceleration of the Cholesky factorization and matrix multiplication (steps 3, 4 in the NORTA algorithm in the preceeding section).
+Specifically, step 2i uses either an efficient multicore multivariate normal simulator (the `R` package `mvnfast`[@Fasiolo2016]) or a using Google's JAX python library NumPy for graphical-processing-unit (GPU) acceleration of the Cholesky factorization and matrix multiplication (steps 3, 4 in the NORTA algorithm in the preceeding section).
+(note that JAX is a not acronym and can be thought of a high-performance NumPy).
 

### 2020-11-06 030-bigsimr-package.Rmd
revise MC studies rev3; example rev1
2nd draft. Revise bigsimr package chapter

@@
 The `bigsimr` package is a high-performance implementation of the proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms) for details).
-When designing `bigsimr`, we aimed to conveniently provide parallelized computation, through multi-core and GPU acceleration.
+When designing `bigsimr`, we aimed to conveniently provide parallelized computation, through multi-core and GPU acceleration, while allowing advanced users to customize and automate worflows.
 
-This section describes basic use of `bigsimr`, by stepping through a low-dimensional (2D) simulation workflow for the often-used example data set `airquality`.
+This subsections below describe the basic use of `bigsimr`, by stepping through a low-dimensional (2D) simulation workflow for the often-used example data set `airquality`.
 This workflow proceeds from data, to estimation, simulation configuration, random vector generation, and result vizualization.
@@
 We'll demonstrate the basic use and syntax of `bigsimr` through an example workflow applied to the New York air quality data set (`airquality`) included in the R `datasets` package.
-First, we load the `bigsimr` library and a few other convenient data science packages, including the syntacically-pleasing `tidyverse` suite of `R` packages.
+First, we load the `bigsimr` library and a few other convenient data science packages, including the syntacically-elegant `tidyverse` suite of `R` packages.
 
@@
 We aim to simulate random two-component vectors mimicking this structure. 
-The margins are not normally distributed, particulary the ozone level exhibits a strong left skew.
+The margins are not normally distributed, particulary the ozone level exhibits a strong positive skew.
 
@@
 
-Next we must specify the marginal distributions and correlation coefficient (type and magnitude).
+Next, we specify the marginal distributions and correlation coefficient (both its type and magnitude).
 Here the analyst is free to be creative.
-For this example, we will not take up goodness-of-fit considerations.
-But it seems sensible without domain knowledge to estimate these quantities from the data and we have a suite of functions designed for this task.
+For this example, we will not take up goodness-of-fit considerations to determine the marginal distributions.
+But it seems sensible without domain knowledge to estimate these quantities from the data and `bigsimr` contains fast functions designed for this task.
 
 *Specifying marginal distributions*.
-Of course, we don't know the true distribution of ozone levels or air temperatures.
 Based on the estimated densities in Figure \@ref(fig:ch030-aq-joint-dist), we'll assume `Temp` is normally ditsributed and `Ozone` is log-normally distributed since its values are positive and skewed.
-We'll use the well-known unbiased estimators for normal distribution and maximum likelihood estimators for the lognormal parameters:
+We'll use the well-known, unbiased estimators for the normal distribution's parameters and maximum likelihood estimators for the lognormal parameters:
 
@@
 
-Next, we'll configure the input marginals later input into `bigsimr::rvec`.
+Next, we'll configure the input marginals for later input into `bigsimr::rvec`.
 The marginal distributions are specifying using `R`'s special `alist` function.
@@
 This implementation strategy supports all `R` `base` probability distributions.
-And allows flexible extensions using other `R` packages that adhere to conventions, such as `extraDistr`
+And allows flexible extensions using other `R` packages that adhere to conventions, such as `extraDistr`.
 Further, by using `alist`, users can specify their own custom distributions (see below in [creating custom margins](#custom-margins)).
 
-It is a bit inconvenient to have to fill in the parameter values manually each time, so we provide a convenience function called `mlist` which behaves similarly to `alist`, except that it will evaluate the right hand side of argument values within the list.
-This is intended to help when scaling up your code to high dimensions.
+It is a bit inconvenient to have to fill in the parameter values manually each time, so we provide a convenience function `mlist` which behaves similarly to `alist`, except that it will evaluate the right hand side of argument values within the list.
+This is intended to help when scaling up your code to high dimensions when many marginals to specify.
 
@@
 *Specfying correlation*.
-Again the user must decide how to describe correlation, based on the particulars of the problem.
+As mentioned, the user must decide how to describe correlation, based on the particulars of the problem.
 For non-normal data and for improved simulation accuracy in our scheme, we advocate the use of rank-based correlations Spearman's $\rho_S$ and Kendall's $\tau$.
-But we also support approximate Pearson correlation coefficient matching, while cautioning the user to check the performance for their parameteric multivariate model (see [Monte Carlo evaluations](#simulations) for evaluation strategies).
-To aid in correlation specification, and estimation in general, we provide a high-performance function `bigsimr::cor_fast` which can estimate Pearson, Spearman, or Kendall correlation using the fastest methods available.
+But we also support approximate Pearson correlation coefficient matching, while cautioning the user to check the performance for their parameteric multivariate model (see [Monte Carlo evaluations](#simulations) for evaluation strategies and guidance).
+To aid in correlation specification, and estimation in general, we provide a high-performance function `bigsimr::cor_fast` which estimates Pearson, Spearman, or Kendall correlation using the fastest methods available.
 (Anyone who has tried estimating Kendall's $\tau$ using `stats::cor` can attest that the routine does not scale to even moderate dimensions).
@@
 As discussed in Section \@ref(background), given a pair of marignal distributions the possible correlations are not free to vary between $[-1, 1]$. 
-To ensure that the simulation is not configured to impossible settings, we provide the `bigsimr::cor_bounds` fuction provides Monte Carlo estimated theoretical lower and upper bounds (using the Generate, Sort, and Correlate algorithm of @DH2011).
+To ensure that the simulation is not configured to impossible settings, we provide the `bigsimr::cor_bounds` fuction provides MC estimated theoretical lower and upper bounds (using the Generate, Sort, and Correlate algorithm of @DH2011).
 
@@
 
-Since our estimated Spearm correlation $\hat{ \rho}_S$ is within the theoretical bounds, the correlation is valid as input to `bigsimr:rvec`.
+Since our estimated Spearman correlation $\hat{ \rho}_S$ is within the theoretical bounds, the correlation is valid as input to `bigsimr:rvec`.
 By comparison, the Pearson correlation $\rho_P$ is restricted to [-0.8648, 0.8727] for these margins (see `bigsimr::cor_bounds`output below).
@@
 Finally, we arrive at the main function of `bigsimr`, `rvec`.
-Let's now simulate $B=10,000$ observations from the assumed joint distribution of Ozone levels and Temp.
+Let's now simulate $B=10,000$ random vectors from the assumed joint distribution of Ozone levels and Temp.
 
@@
 
-## Advanced use
+## Advanced use {#advanced-use}
 
 *Creating a custom marginal distribution*.
-Because `bigsimr` uses an `alist` to store the margins, any kind of probability distribution can be used including custom marginal distributions not provided in base `R`. 
-This accommodation is possible because the main function `rvec` uses the inverse CDF of a margin to transform from correlated uniform margins to the target distribution.
-Therefore, to specify a marginal distribution absent from an existing packge, the user simply needs to provide a closed form expression from the corresponding quantile function.
+Because `bigsimr` uses an `alist` to store the margins, any probability distribution with a well-defined inverse CDF can be used including custom marginal distributions not provided in base `R`. 
+Therefore, to specify a marginal distribution absent from an existing packge, the user needs to provide a closed-form expression to form the corresponding quantile function.
 
@@
 
-Next we define the Pareto quantile function in R.
+Next we define the Pareto quantile function in R:
 
@@
 
-For example, writing random number generating function for our target marginal can be accomplished by calling the quantile function on a uniformly distributed random variable (the inverse transform method @Rizzo2007).
+Writing random number generating function for our target marginal can be accomplished by calling the quantile function on a uniformly distributed random variable (the inverse transform method @Rizzo2007).
 
@@
 
-Now with the quantile and RNG Pareto functions, we can use the distribution in `bigsimr` just like the other built-in distributions.
+Now with the `qpareto` and `rpareto` functions, we can use the distribution in `bigsimr` just like the other built-in distributions.
 
@@
 This example assumes that `bigsimr` is installed on a system with a slurm scheduler installed.
+A single run of `bigsimr:rvec` using `rslurm` can be code as:
 
@@
 Now, let's show off the real power of combining `bigsimr` and `rslurm` by simulating many correlation structures.
-The `rslurm::slumr_map` syntax mirrors the familiar `base::lapply` and `purrr::map` functions.
+The `rslurm::slumr_map` syntax mirrors the `base::lapply` and `purrr::map` functions.
 
@@
 simReps <- 100
-rhoList <- replicate( n = simReps, bigsimr::cor_randPSD(d = length(margins)), simplify=FALSE )
+rhoList <- replicate( n = simReps, bigsimr::cor_randPSD(d = length(margins)),
+                     simplify=FALSE )
 sjob <- slurm_map(x = rhoList,

### 2020-11-06 050-examples.Rmd
2nd draft. Revise bigsimr package chapter

@@
 ## full workflow simulating RNA-seq data
-## brcaRDS <- 'data/brca99.rds'
+## brcaRDS <- 'data/brca95.rds'
+## myProb = 0.95
 brca <- readRDS(file = brcaRDS)
@@
 
-```{r ch050-quadloss, echo=TRUE, eval=TRUE, cache=FALSE}
+```{r ch050-quadloss, echo=TRUE, eval=FALSE, cache=FALSE}
 quadLossRDS <- paste0( "results/brca", round(myProb*100), "quadLoss.rds" )
@@
 
-```{r ch050-quadloss-summary, echo=TRUE, eval=TRUE, cache=FALSE}
+```{r ch050-quadloss-summary, echo=TRUE, eval=FALSE, cache=FALSE}
 summary(quadLoss)

### 2020-11-06 050-examples.Rmd
2nd draft. scale up d in quad loss, drop m to 10

@@
 
-## Simulation-based UHD joint probability calculations
+## Simulation-based joint probability calculations
 
@@
 ```{r ch010-myM, include=FALSE, cache = TRUE}
-m <- 100
+m <- 10
 ```
@@
 In this example, we simulate $m=`r m`$ data sets with the number of simulated vectors matching the number of patients in the BRCA data set, $N=`r nrow(brca)`$.
-At each iteration we estimate the quadratic loss from the specified $\bf{R^{Spearman}_Y}$ and average over the $m$ iteration.
+Using the fact that our simulation is much faster for the Pearson correlation type (see Figure \@ref(fig:ch040-gpuVScpuFig), we convert the Spearman correlation matrix and ensure PSD outside the `rvec` call.
+At each iteration, we estimate the quadratic loss from the specified ${\bf R}^{Spearman}_{ \bf Y}$, producing a distribution of loss values.
 
-```{r ch050-quadloss, echo=TRUE, eval=FALSE, cache=FALSE}
+```{r ch050-quadloss, echo=TRUE, eval=T, cache=FALSE}
 quadLossRDS <- paste0( "results/brca", round(myProb*100), "quadLoss.rds" )
@@
     n <- nrow(brca)
-    simRho <- replicate(n = m, expr = { tmpSim <- rvec(n = nrow(brca) , nb_Rho, nb_margins,
-                                                       type = corType, ensure_PSD = TRUE, cores = cores)
-                                                       tmpRho <- bigsimr::cor_fast( x = tmpSim )} ,
+    ## convert outside for faster simulation
+    nb_Rho_p <- bigsimr::cor_convert( rho = nb_Rho, from = corType, to = "pearson" )
+    ## ensure PSD
+    nb_Rho_p <- bigsimr::cor_nearPSD( G = nb_Rho_p )
+    ## create m random vectors and estimate correlation
+    simRho <- replicate(n = m,
+                        expr = { tmpSim <- rvec(n = n , nb_Rho_p, nb_margins,
+                                                type = 'pearson', ensure_PSD = FALSE, cores = cores);
+                                                bigsimr::cor_fast( x = tmpSim, method = corType )} ,
                         simplify = FALSE)
-
+    ## find quadratic loss at each rep
     quadLoss <- unlist( lapply( simRho, rags2ridges::loss, T = nb_Rho, type = "quadratic"  ) )
@@
 
-```{r ch050-quadloss-summary, echo=TRUE, eval=FALSE, cache=FALSE}
+```{r ch050-quadloss-summary, echo=TRUE, eval=T, cache=FALSE}
 summary(quadLoss)

### 2020-11-09 040-monte-carlo-evaluation.Rmd
2nd draft. scale up d in quad loss, drop m to 10
2nd draft. Revise MC chapter

@@
 Before applying our methodology to real data simulation, we conduct several Monte Carlo studies to investigate method performance.
-Since marginal parameter matching in our scheme is essentially a sequence of univariate inverse probability transforms, the challenging aspects are the accuracy of dependency matching and computational efficiency.
-And, so, we'll focus the numerical experiments on assessing how well the procedure scales to high dimension with respect to reasonable computation times (property S1 above) and accurately m
-atching marginal and dependency parameters.
-
-Further, we select simulaiton configurations to ultimately simulate discrete-valued RNA-seq data and the simulations will proceed in increasing complexity, leading to our motivating application in Section \@ref(examples).
-We begin wiith empirically evaluating the dependency matching across all three supported correlations --- Pearson's, Spearman's, and Kendall's ---- in identical, bivariate marginal configurations.
-The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (mimicking RNA-seq counts).
+Since marginal parameter matching in our scheme is essentially a sequence of univariate inverse probability transforms, the challenging aspects are the accuracy of dependency matching and computational efficiency at high dimensions.
+To evaluate our methods in those respects, we design the following numerical experiments to first assess accuracy in match dependency parameters in bivariate simulations and then time the procedure in increasingly large dimension $d$.
 
 ## Bivariate experiments
+
+We select bivariate simulation configurations to ultimately simulate discrete-valued RNA-seq data and, so, proceed in increasing complexity, leading to the model in our motivating application in Section \@ref(examples).
+We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson's, Spearman's, and Kendall's ---- in identical, bivariate marginal configurations.
+For each pair of identical margins, we vary the correlations across the entire possible range of values to evaluate the simulation's ability to obtain the theoretic bounds.
+The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (mimicking RNA-seq counts).
 
@@
 
+Table XX lists our identical-marginal, bivariate simulation configurations.
+We increase the simulate replicates $B$ to check that our results converge to the target correlations and gauge statistical efficiency.
+We select distributions beginning with a standard multivariate normal (MVN) as we expect the performance to be exact (up to MC error) for all correlation types.
+Then, we select a non-symmetric continuous distribution: a standard (rate =1) two-component multivariate gamma (MVG).
+Finally, we select distributions and marginal parameter values that are motivated by our RNA-seq data, namely values proximal to probabilities and sizes estimated from the data (see [Example applications for our motivating data](examples) for estimation details).
+Thus we arrive at a multivariate negative binomial (MVNB) $p_1 = p_2 = `r prob`, r_1 = r_2 = 4,\rho$ ).
+
+| Simulation Reps ($B$) | Correlation Types | Distribution |
+|---------------------|:------------------:|--------------:|
+|$\{ 1000, 10000, 100000 \}$ | $\{ \rho_P, \rho_S, \tau \}$ | $\{ { \bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho ), \\ { \bf Y} \sim MVG( shape = 10, rate = 1, \rho ), \\ { \bf Y} \sim MVNB(p = `r prob`, r = 4,\rho) \}$.
+|
+
+For each of the unique 9 simulation configurations above, we estimate the correlation bounds and vary $\rho$ along a sequence of 100 points evenly placed within the bounds (minus an adjustment factor $\epsilon=0.01$ to handle numeric issues when the bound is specified exactly).
+
+<!-- 
 *Bivariate Normal*.
@@
 We vary $\rho$ across the entire possible range of correlations for each correlation type.
-<!-- 
 Figure \@ref(fig:ch040-biNormPlot) `bigsimr` recovers the Pearson specified correlations for MVN.
+
+*Bivariate Gamma*. 
+Similarly, let's check the performance for a non-symmetric continuous distribution: a standard (rate =1) bivariate gamma.
+Here we have a Bivariate Gamma with $shape_1 = shape_2 = 10, \rho_{type}$.
+We vary $\rho$ across the entire possible range of correlations for each correlation type.
+Figure \@ref(fig:ch040-biGammaPlot) `bigsimr` recovers the Pearson specified correlations for bivariate gamma.
+*Bivariate Negative Binomial*. 
+Figure \@ref(fig:ch040-biNegBinPlot) `bigsimr` recovers the Pearson specified correlations for bivariate negative binomial.
+
  -->
@@
 ```
-
-*Bivariate Gamma*. 
-Similarly, let's check the performance for a non-symmetric continuous distribution: a standard (rate =1) bivariate gamma.
-Here we have a Bivariate Gamma with $shape_1 = shape_2 = 10, \rho_{type}$.
-We vary $\rho$ across the entire possible range of correlations for each correlation type.
-<!--  Figure \@ref(fig:ch040-biGammaPlot) `bigsimr` recovers the Pearson specified correlations for bivariate gamma. -->
 
@@
 
-*Bivariate Negative Binomial*. 
-Let's check the performance for a discrete distribution: a bivariate negative binomial.
-We use values that are motivated by our RNA-seq data, namely the mean probability and sizes estimated from the data ( see [Example applications for our motivating data](examples) for estimation details): Bivariate Negative Binomial ( $prob_1 = prob_2 = `r prob`, size_1 = size_2 = 4,\rho_{type}$ ).
-We vary $\rho$ across the entire possible range of correlations for each correlation type.
-<!-- Figure \@ref(fig:ch040-biNegBinPlot) `bigsimr` recovers the Pearson specified correlations for bivariate negative binomial. -->
-
 ```{r ch040-biNegBinPlot, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.align='center', fig.width = 8, fig.height=5,fig.cap = "`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
@@
 
-Taken together, the studies show our methology are accuracy across the entire range of possible correlation values for the rank-based dependency measures in continuous marginals.
-For the two non-normal bivariate marginals, the Pearson correlation matching is approximate.
-Table XX contains the average absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.
-For discrete margins, matching the dependency measures was somewhat less accurate, even for the rank-based metrics, especially near the correlation bounds.
-Since the bivariate normal exactly matches all dependency measures in our scheme, we also report relative MSE of the non-normal cases to the normal simulations.
-
 ```{r ch040-combineBiSims, cache=F}
@@
 allDat <- rbind( allDat, dat )
-allDat$margins <- factor(allDat$margins, levels = c( 'norm', 'gamma', 'nbinom') )
-allDat$type <- factor(allDat$type, levels = c( 'pearson', 'kendall', 'spearman' ) )
+allDat$margins <- factor(allDat$margins, levels = c( 'norm', 'gamma', 'nbinom'),
+                         labels = c( 'MVN', 'MVG', 'MVNB' ) )
+allDat$type <- factor(allDat$type, levels = c( 'pearson', 'spearman', 'kendall' ),
+                      labels = c( 'Pearson', 'Spearman', 'Kendall' ) )
 allDat$N <- factor(allDat$N, levels = c( 1000, 10000, 100000) )
@@
 The horizontal axis plots the specified target correlations across the entire range of possible correlations for each bivariate margin.
-Normal margins are plotted in dark grey, gamma in medium grey, and negative binomial in light grey.
+Normal margins are plotted in dark blue, gamma in medium blue, and negative binomial in light blue.
 As the number of simulated vectors $B$ increases from left to right, the variation in estimated correlations (vertical axis) decreases.
@@
 names(repsLabs) <- c(1000, 10000, 100000)
-typeLabs <- c( 'Pearson', 'Kendall', 'Spearman' ) 
-names(typeLabs) <- c( 'pearson', 'kendall', 'spearman' )
+typeLabs <- c( 'Pearson', 'Spearman', 'Kendall' )
+names(typeLabs) <- c( 'Pearson', 'Spearman', 'Kendall' )
 
@@
 
-```{r ch040-biMSEtable}
+```{r ch040-biMSEtable, results='asis'}
 ## tabMSE  <- allDat %>%
@@
 ##     summarize( mse = mean( ( rho - rho_hat )^2 ) ) %>%
-    summarize( mae = mean( abs( rho - rho_hat )  ) )
+    summarize( MAE = mean( abs( rho - rho_hat )  ) )
 ## normMSE <- as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ] ) 
@@
 ## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) ) / tabMSE$mse
-knitr::kable(
-  tabMSE, booktabs = TRUE,
-  caption = 'Average abolute error in matching the target dependency and  '
+knitr::kable(tabMSE,
+             col.names = c("No. of random vectors",
+                           "Correlation type",
+                           "Distribution",
+                           "Mean abs. error"),
+             booktabs = TRUE,
+             caption = 'Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.'
 )
@@
 
-Generally, the accuracy seems adequate for most applications, at least in these limited simulation settings.
-Of course, one could choose marginals that lead to poor matching by construction.
-In partice, we recommend users to always evaluate the accuracy for their application, using methods similar to those presented above.
-See [Discusion](discussion) for future directions for fast Pearson matching and discrete-specific modifications.
+Taken together, the studies show our methodology is generally accuracy across the entire range of possible correlation values for the rank-based dependency measures, at least in these limited simulation settings for the rank-based correlations.
+For the two non-normal bivariate marginals, the Pearson correlation matching is approximate.
+Table \@ref(tab:ch040-biMSEtable) contains the average absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.
+For discrete margins, matching the dependency measures was somewhat less accurate, even for the rank-based metrics, and particularly inaccurate near the lower bound of Pearson correlations.
 
-## Scale up to Ultra-High Dimensions
+The accuracy appears adequate for many applications.
+In practice, we recommend users to always evaluate the accuracy for their application, using methods similar to those presented above.
+See [Discussion](discussion) for future directions for fast Pearson matching and discrete-specific modifications.
+
+## Scale up to High Dimensions
 

### 2020-11-09 040-monte-carlo-evaluation.Rmd
2nd draft. Revise MC chapter v2

@@
 
-Table XX lists our identical-marginal, bivariate simulation configurations.
+Table \@ref(tab:sims) lists our identical-marginal, bivariate simulation configurations.
 We increase the simulate replicates $B$ to check that our results converge to the target correlations and gauge statistical efficiency.
@@
 Thus we arrive at a multivariate negative binomial (MVNB) $p_1 = p_2 = `r prob`, r_1 = r_2 = 4,\rho$ ).
+
+Table: (\#tab:sims) Identical margin, bivariate simulation configurations to evaluate correlation matching accuracy and efficiency.
 
@@
 
+Figure \@ref(fig:ch040-bPlot) displays the aggregated bivariate simulation results with a detailed summarization.
+Table \@ref(tab:ch040-biMAEtable) contains the mean absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.
+Taken together, the studies show our methodology is generally accuracy across the entire range of possible correlation values for the rank-based dependency measures, at least in these limited simulation settings for the rank-based correlations.
+For the two non-normal bivariate marginals, the Pearson correlation matching is approximate.
+For discrete margins, matching the dependency measures was somewhat less accurate, even for the rank-based metrics, and particularly inaccurate near the lower bound of Pearson correlations.
+
 ```{r ch040-bPlot, cache=F, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.asp = 1.30, fig.weight = 4, fig.align='center', fig.cap = "Bivariate simulations match specified correlations.
@@
 
-```{r ch040-biMSEtable, results='asis'}
+```{r ch040-biMAEtable, results='asis'}
 ## tabMSE  <- allDat %>%
@@
 
-Taken together, the studies show our methodology is generally accuracy across the entire range of possible correlation values for the rank-based dependency measures, at least in these limited simulation settings for the rank-based correlations.
-For the two non-normal bivariate marginals, the Pearson correlation matching is approximate.
-Table \@ref(tab:ch040-biMSEtable) contains the average absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.
-For discrete margins, matching the dependency measures was somewhat less accurate, even for the rank-based metrics, and particularly inaccurate near the lower bound of Pearson correlations.
-
 The accuracy appears adequate for many applications.
@@
 
-With some assurance of the method's accuracy from the low-dimensional perspective in the bivariate setting, we now turn to assessing whether the `bigsimr` can scale to large dimesional problems. 
-Now we seek evidence to determine whether the method can scale to higher dimensions in a practical time.
-Using our motivating RNA-seq data, described in [Background](background), we filtered the orignal 20,501 genes to the high-expressing genes at increasing percentiles, $1, 5, 10, 15, 20, 25\%$, to obtain $d=\{206, 1026, 2051, 3076, 4101, 5127\}$ marginals and $\binom{d}{2}$ pairwise correlations at each setting.
+With information of our method's accuracy from a low-dimensional perspective, we now turn to assessing whether the `bigsimr` can scale to larger dimensional problems. 
+Specifically, we seek evidence to determine whether the method can scale to higher dimensions in a practical time.
+Using our motivating RNA-seq data, described in [Background](background), we filtered the original 20,501 genes to the high-expressing genes at increasing percentiles, $1, 5, 10, 15, 20, 25\%$, to obtain $d=\{206, 1026, 2051, 3076, 4101, 5127\}$ marginals and $\binom{d}{2}$ pairwise correlations at each setting.
 For example, for $d=5127$ there are $13,140,501$ correlation coefficients.
 We estimated the marginal negative binomial parameters and the correlation coefficients from the RNA-seq data to seed our simulations.
-( See [Example applications for our motivating data](examples) for a detailed descrition of estimation).
-Figure \@ref(fig:ch040-gpuVScpuFig) displays computation times using various high-performance settings to produce $B=10,000$ random vectors.
-The Pearson simulations are much faster since the correlation conversion steps are avoided (pre-processing step; see [Algorithms](algorithms)), but as we known from above the accuracy will suffer slightly, especially near the negative boundary of the possible correlations.
-Matching Spearman's correlation at larger $d$ gets costly if one wants to produce $B=10,000$ random vectors at many different simulation settings.
-We recommend computing this type of simulations on a computing cluster (possibly using `rslurm`; see the [Advance Use](advanced-use)).
+(See [Example applications for our motivating data](examples) for a detailed description of estimation).
+
+Figure \@ref(fig:ch040-gpuVScpuFig) displays computation times using various high-performance settings (1 central processing unit, CPU-1, versus twenty CPUs, CPU-20; with and without GPU acceleration, GPU-1; GPU-20) to produce $B=10,000$ random vectors.
+The Pearson simulations are much faster since the correlation conversion steps are avoided (pre-processing step; see [Algorithms](algorithms)), but as we know from above the accuracy will suffer slightly, especially near the negative boundary of the possible correlations.
+Matching Spearman's correlation at larger $d$ gets costly if one wants to produce $B=10,000$ random vectors at many different simulation settings, but the conversion steps need only be computed once (see [MC evaluation of correlation estimation efficiency](MC-evaluation-of-correlation-estimation-efficiency) for an example of this strategy).
 

### 2020-11-11 010-background.Rmd
2nd draft. Revise example 3D example results

@@
 # Background {#background}
+
+```{r ch010-background, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
+library(tidyverse)
+library(GGally)
+knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
+```
 
@@
 exampleGenes <- readRDS('./data/exampleGenes.rds')
-knitr::kable(round(head( brca[, exampleGenes], n = 5)),
+small <-  brca %>% select( exampleGenes )
+knitr::kable(round(head( small, n = 5)),
       booktabs = TRUE,
@@
 Modeling these data motivate our simulation methodology."}
-## Describe and plot real data
-## NEED TO CHANGE TO SPEARMANS
-GGally::ggpairs(data = as.data.frame(brca[ ,exampleGenes] ))
+GGally::ggpairs(
+  data = small, 
+  upper = list(
+    continuous = wrap('cor', method = "spearman")
+  ), 
+) + theme_bw()
 ```

### 2020-11-11 050-examples.Rmd
2nd draft. Revise example 3D example results

@@
 ```{r ch050-simDataFig, echo = FALSE, eval=TRUE, cache=T, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Simulated data for 3 selected high-expressing genes, replicating the observed data structure.'}
-## retrieve the examples from an earlier chapter
-## there is probably a better way to do this
 exampleGenes <- readRDS(file = './data/exampleGenes.rds')
-GGally::ggpairs(data = as.data.frame(sim_nbinom[ ,exampleGenes] ))
+GGally::ggpairs(
+  data = as.data.frame( sim_nbinom[ ,exampleGenes] ), 
+  upper = list(
+    continuous = wrap('cor', method = "spearman")
+  ), 
+) + theme_bw()
 ```

### 2020-11-11 010-background.Rmd
2nd draft. Begin to fix scales, but maybe not needed.

--- 2020-11-06 010-background.Rmd
@@
 # Background {#background}
+
+```{r ch010-background, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide", cache=F}
+library(tidyverse)
+library(GGally)
+knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
+```
 
@@
 )
-## GGally::ggpairs(data = as.data.frame(brca[ ,exampleGenes] ))
 ```
@@
 Modeling these data motivate our simulation methodology."}
-## Describe and plot real data
-## NEED TO CHANGE TO SPEARMANS
-GGally::ggpairs(data = as.data.frame(brca[ ,exampleGenes] ))
+# lowerfun <- function(data,mapping){
+#   ggplot(data = data, mapping = mapping)+
+#     geom_point()+
+#     scale_x_continuous(limits = c(-1.5,1.5))+
+#     scale_y_continuous(limits = c(-1.5,1.5))
+# }
+
+GGally::ggpairs( data = small,
+                # lower = list( continuous = wrap(lowerfun)),
+                upper = list(
+                    continuous = wrap('cor', method = "spearman")
+                ), 
+                ) + theme_bw()
 ```

### 2020-11-11 030-bigsimr-package.Rmd
2nd draft. Polish ch040 figures and tables

@@
 
-```{r ch030-basic-example-setup, echo=TRUE, message=FALSE}
+```{r ch030-basic-example-setup, echo=TRUE, message=FALSE, cache=F}
 library(bigsimr)
@@
 Temperature with estimated marginal densities.
-The data are left skewed tails and appear to be correlation."}
+The data are left skewed tails and appear to be correlation.", cache=F}
 

### 2020-11-11 040-monte-carlo-evaluation.Rmd
2nd draft. Revise MC chapter v3
2nd draft. Polish ch040 figures and tables

@@
 
-Table \@ref(tab:sims) lists our identical-marginal, bivariate simulation configurations.
-We increase the simulate replicates $B$ to check that our results converge to the target correlations and gauge statistical efficiency.
-We select distributions beginning with a standard multivariate normal (MVN) as we expect the performance to be exact (up to MC error) for all correlation types.
-Then, we select a non-symmetric continuous distribution: a standard (rate =1) two-component multivariate gamma (MVG).
-Finally, we select distributions and marginal parameter values that are motivated by our RNA-seq data, namely values proximal to probabilities and sizes estimated from the data (see [Example applications for our motivating data](examples) for estimation details).
-Thus we arrive at a multivariate negative binomial (MVNB) $p_1 = p_2 = `r prob`, r_1 = r_2 = 4,\rho$ ).
-
-Table: (\#tab:sims) Identical margin, bivariate simulation configurations to evaluate correlation matching accuracy and efficiency.
-
-| Simulation Reps ($B$) | Correlation Types | Distribution |
-|---------------------|:------------------:|--------------:|
-|$\{ 1000, 10000, 100000 \}$ | $\{ \rho_P, \rho_S, \tau \}$ | $\{ { \bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho ), \\ { \bf Y} \sim MVG( shape = 10, rate = 1, \rho ), \\ { \bf Y} \sim MVNB(p = `r prob`, r = 4,\rho) \}$.
-|
-
-For each of the unique 9 simulation configurations above, we estimate the correlation bounds and vary $\rho$ along a sequence of 100 points evenly placed within the bounds (minus an adjustment factor $\epsilon=0.01$ to handle numeric issues when the bound is specified exactly).
-
-<!-- 
-*Bivariate Normal*.
-Let's simulate a bivariate normal and check our correlation matching performance as N increases.
-Standard bivariate normal 
-Here we have BVN( $\mu_1 = \mu_2 = 0, \rho_{type}$ ).
-We vary $\rho$ across the entire possible range of correlations for each correlation type.
-Figure \@ref(fig:ch040-biNormPlot) `bigsimr` recovers the Pearson specified correlations for MVN.
-
-*Bivariate Gamma*. 
-Similarly, let's check the performance for a non-symmetric continuous distribution: a standard (rate =1) bivariate gamma.
-Here we have a Bivariate Gamma with $shape_1 = shape_2 = 10, \rho_{type}$.
-We vary $\rho$ across the entire possible range of correlations for each correlation type.
-Figure \@ref(fig:ch040-biGammaPlot) `bigsimr` recovers the Pearson specified correlations for bivariate gamma.
-*Bivariate Negative Binomial*. 
-Figure \@ref(fig:ch040-biNegBinPlot) `bigsimr` recovers the Pearson specified correlations for bivariate negative binomial.
-
- -->
-
-```{r ch040-biNormPlot, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, fig.height = 5, fig.width= 8, fig.align='center', fig.cap = "`bigsimr` recovers the Pearson specified correlations for MVN."}
-dat <- readRDS("./results/biNorm_sims.rds")
-dat %>%
-    filter(cores == cores) %>%
-    ggplot(aes(rho, rho_hat, color = type)) +
-    ## ggplot(aes(rho, rho_hat)) +
-    geom_point() +
-    geom_abline(slope = 1) +
-    ## facet_wrap(~ + N) + theme_bw()
-    facet_wrap(~ type + N) + theme_bw()
-## ggsave('fig/plot-biNormPlot.pdf')
-```
-
 ```{r ch040-biGamma, echo = FALSE, eval = TRUE}
@@
 }
-```
-
-```{r ch040-biGammaPlot, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, fig.height = 5, fig.width= 8, fig.align='center', fig.cap = "`bigsimr` recovers the Pearson specified correlations for Bivariate Gamma."}
-dat <- readRDS("./results/biGamma_sims.rds")
-dat %>%
-    filter(cores == cores) %>%
-    ggplot(aes(rho, rho_hat, color = type)) +
-    ## ggplot(aes(rho, rho_hat)) +
-    geom_point() +
-    geom_abline(slope = 1) +
-    ## facet_wrap(~ + N) + theme_bw()
-    facet_wrap(~ type + N) + theme_bw()
-## ggsave('fig/plot-biGammaPlot.pdf')
 ```
@@
 
+Table \@ref(tab:sims) lists our identical-marginal, bivariate simulation configurations.
+We increase the simulate replicates $B$ to check that our results converge to the target correlations and gauge statistical efficiency.
+We select distributions beginning with a standard multivariate normal (MVN) as we expect the performance to be exact (up to MC error) for all correlation types.
+Then, we select a non-symmetric continuous distribution: a standard (rate =1) two-component multivariate gamma (MVG).
+Finally, we select distributions and marginal parameter values that are motivated by our RNA-seq data, namely values proximal to probabilities and sizes estimated from the data (see [Example applications for our motivating data](examples) for estimation details).
+Thus we arrive at a multivariate negative binomial (MVNB) $p_1 = p_2 = `r prob`, r_1 = r_2 = 4,\rho$ ).
+
+Table: (\#tab:sims) Identical margin, bivariate simulation configurations to evaluate correlation matching accuracy and efficiency.
+
+| Simulation Reps ($B$) | Correlation Types | Identical-margin 2D distribution |
+|-------------|:--------------:|----------------------:|
+|$1000$ | Pearson ($\rho_P$) | ${ \bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho )$ |
+|$10,000$ | Spearman ($\rho_S$) | ${ \bf Y} \sim MVG( shape = 10, rate = 1, \rho )$ |
+| $100,000$| Kendall ($\tau$)| ${ \bf Y} \sim MVNB(p = `r prob`, r = 4,\rho)$ |
+
+For each of the unique 9 simulation configurations above, we estimate the correlation bounds and vary $\rho$ along a sequence of 100 points evenly placed within the bounds (minus an adjustment factor of $\epsilon=0.01$ to handle numeric issues when the bound is specified exactly).
+
+<!-- 
+*Bivariate Normal*.
+Let's simulate a bivariate normal and check our correlation matching performance as N increases.
+Standard bivariate normal 
+Here we have BVN( $\mu_1 = \mu_2 = 0, \rho_{type}$ ).
+We vary $\rho$ across the entire possible range of correlations for each correlation type.
+Figure \@ref(fig:ch040-biNormPlot) `bigsimr` recovers the Pearson specified correlations for MVN.
+
+*Bivariate Gamma*. 
+Similarly, let's check the performance for a non-symmetric continuous distribution: a standard (rate =1) bivariate gamma.
+Here we have a Bivariate Gamma with $shape_1 = shape_2 = 10, \rho_{type}$.
+We vary $\rho$ across the entire possible range of correlations for each correlation type.
+Figure \@ref(fig:ch040-biGammaPlot) `bigsimr` recovers the Pearson specified correlations for bivariate gamma.
+*Bivariate Negative Binomial*. 
+Figure \@ref(fig:ch040-biNegBinPlot) `bigsimr` recovers the Pearson specified correlations for bivariate negative binomial.
+
+ -->
+
+```{r ch040-biNormPlot, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, fig.height = 5, fig.width= 8, fig.align='center', fig.cap = "`bigsimr` recovers the Pearson specified correlations for MVN."}
+dat <- readRDS("./results/biNorm_sims.rds")
+dat %>%
+    filter(cores == cores) %>%
+    ggplot(aes(rho, rho_hat, color = type)) +
+    ## ggplot(aes(rho, rho_hat)) +
+    geom_point() +
+    geom_abline(slope = 1) +
+    ## facet_wrap(~ + N) + theme_bw()
+    facet_wrap(~ type + N) + theme_bw()
+## ggsave('fig/plot-biNormPlot.pdf')
+```
+
+```{r ch040-biGammaPlot, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, fig.height = 5, fig.width= 8, fig.align='center', fig.cap = "`bigsimr` recovers the Pearson specified correlations for Bivariate Gamma."}
+dat <- readRDS("./results/biGamma_sims.rds")
+dat %>%
+    filter(cores == cores) %>%
+    ggplot(aes(rho, rho_hat, color = type)) +
+    ## ggplot(aes(rho, rho_hat)) +
+    geom_point() +
+    geom_abline(slope = 1) +
+    ## facet_wrap(~ + N) + theme_bw()
+    facet_wrap(~ type + N) + theme_bw()
+## ggsave('fig/plot-biGammaPlot.pdf')
+```
+
 ```{r ch040-biNegBinPlot, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.align='center', fig.width = 8, fig.height=5,fig.cap = "`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
@@
 
-```{r ch040-biMAEtable, results='asis'}
+```{r ch040-biMAEtable, results='asis', cache=F}
 ## tabMSE  <- allDat %>%
@@
 ## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) ) / tabMSE$mse
+
 knitr::kable(tabMSE,
+             format = 'latex',
              col.names = c("No. of random vectors",
@@
              booktabs = TRUE,
-             caption = 'Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.'
+             caption = 'Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.',
+             linesep = c('', '',  '\\addlinespace')
 )

### 2020-11-11 050-examples.Rmd
2nd draft. Polish ch040 figures and tables

@@
 
-```{r ch050-readBRCA, echo=FALSE, eval=TRUE}
+```{r ch050-readBRCA, echo=FALSE, eval=TRUE, cache=F}
 ## full workflow simulating RNA-seq data

### 2020-11-11 050-examples.Rmd
2nd draft. Revise ch050-examples

--- 2020-11-11 050-examples.Rmd
@@
-# Example applications for RNA-seq data {#examples}
+# RNA-seq data example applications  {#examples}
 
 ```{r ch050-LoadLib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide", cache=F}
-knitr::opts_chunk$set(echo = FALSE, cache = FALSE)
+## knitr::opts_chunk$set(echo = FALSE, cache = FALSE)
 # reticulate::use_condaenv("bigsimr")
@@
 This section demonstrates how to simulate multivariate data using `bigsimr`, aiming to replicate the structure of high-dimensional dependent count data.
-Simulating RNA-sequencing (RNA-seq) data is a primary motivating application of the proposed methodology, seeking scaleable Monte Carlo (MC) methods for realistic multivariate simulation for these data.
-Intergene correlation is an inherent part of biological processes.
-Yet many models do not account for this, leading to major distruptions to the operating characteristics of statistical estimation, testing, and prediction.
+In an illustration of our proposed methodology applied to real data, we seek to simulate RNA-sequencing data by producing simulated random vectors with assumed marginal distributions with estimated parameters that mimick the observed data and its generating process.
+We seek scaleable realistic multivariate simulation to enable Monte Carlo (MC) methods for thes data.
+Modeling RNA-seq using multivariate probabilities distributions is natural as inter-gene correlation is an inherent part of biological processes.
+Yet many models do not account for this, leading to major disruptions to the operating characteristics of statistical estimation, testing, and prediction.
 See @Efron2012 for a detailed discussion with related methods and see @Wu2012b, @Schissler2018, @Schissler2019 for applied examples.
-The following subsections apply `bigsimr`'s methods to real RNA-seq data, including replicated an estimated parameteric structure, MC probability estimation, and MC evaluation of correlation estimation efficiency.
+The following subsections apply `bigsimr`'s methods to real RNA-seq data, including replicated an estimated parametric structure, MC probability estimation, and MC evaluation of correlation estimation efficiency.
 
@@
 
-```{r ch050-readBRCA, echo=FALSE, eval=TRUE}
+```{r ch050-readBRCA, echo=FALSE, eval=TRUE, cache=F}
 ## full workflow simulating RNA-seq data
-## brcaRDS <- 'data/brca95.rds'
-## myProb = 0.95
+brcaRDS <- 'data/brca95.rds'
+myProb = 0.95
 brca <- readRDS(file = brcaRDS)
+d <- ncol(brca)
 ## smaller set for prototyping
@@
 
-In an illustration of our proposed methodology applied to real data, we seek to simulate RNA-sequencing data by producing simulated random vectors with assumed marginal distributions with estimated parameters.
-Our goal is to replicate the structure of a breast cancer data set (BRCA data set from The Cancer Genome Atlas).
-Specifically, we will simulate $B=10,000$ random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$
-All these genes exhibit over-dispersion and, so, we proceed to estimate the NB parameters $(r_i, p_i), i=1,\ldots,d$ to determine the target marginal PMFs $g_i(y_i)$ (via method of moments).
+Our first goal is to replicate the structure of the TCGA BRCA RNA-seq data set (see [Background](background)).
+Ultimately, we will simulate $B=10,000$ random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with $d=`r d`$.
 Often researchers posit a negative binomial (NB) model as RNA-seq counts are often over-dispersed that a Poisson model would suggest.
+All $d$ selected genes exhibit over-dispersion (data not shown) and, so, we proceed to estimate the NB parameters $(r_i, p_i), i=1,\ldots,d$ to determine the target marginal PMFs $f_i$ (via method of moments).
+To complete specification of the simulation algorithm inputs, we'll estimate the Spearman correlation matrix ${ \bf R}_{Spearman}$ to characterize dependency.
+
+<!--
 This suggests simulating high-dimensional multivariate NB (MVB) with heterogeneous marginals would be useful tool in the development and evaluation of RNA-seq analytics.
 This procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions. 
-To specify the simulation algorithm inputs, we estimate the Spearman correlation matrix $\bf{R^{Spearman}_Y}$ and marginal NB parameters.
-
-<!--
 All these genes exhibit over-dispersion and, so, we proceed to estimate the NB parameters $(r_i, p_i), i=1,\ldots,d$ to determine the target marginal PMFs $g_i(y_i)$ (via method of moments).
@@
 
-```{r ch050-estRhoBRCA, echo=TRUE, eval=TRUE}
+```{r ch050-estRhoBRCA, echo=TRUE, eval=TRUE, cache=F}
 ## Estimate Spearman's correlation on the count data
@@
 
-Next we estimate the marginal parameters. 
-Here we use the basic method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model.
-We model the marginals distributions as coming from the same probability family (NB) yet are hetereogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.
-The code below features some advanced `R` utilities for concise, rapid, and generalizable target marginal specification:
+Next, we estimate the marginal parameters. 
+We use method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model.
+The marginal distributions are from the same probability family (NB) yet are hetereogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.
+The functions below help achieve this estimation and specifying the inputs for use in `bigsimr::rvec`.
 
@@
 
-Once the functions are defined to complete marginal estimation, we specify the desired multivariate negative binomial distribution and generate the desired random vectors:
+Now we estimate the marginal parameters using the built-in `apply`function:
 
-```{r ch050-estMargins, echo = TRUE, eval = TRUE}
+```{r ch050-estMargins, echo = TRUE, eval = TRUE, cache=F}
 sizes <- apply( unname(as.matrix(brca)), 2, nbinom_mom )[1, ]
@@
     ## run sims
+    set.seed(44)
     system.time( sim_nbinom <- rvec(n, nb_Rho, nb_margins, type = corType,
@@
 
-Figure \@ref(fig:ch050-simDataFig) displays the simulated counts and pairwise relationships for our example genes in \@ref(fig:ch010-realDataFig).
-Simulated counts roughly mimick the observed data but with a smoothier appearance due to the assumed parameter form.
+Figure \@ref(fig:ch050-simDataFig) displays the simulated counts and pairwise relationships for our example genes in Table \@ref(tab:ch010-realDataFig).
+Simulated counts roughly mimick the observed data but with a smoothier appearance due to the assumed parameter form and with less extreme points then the observed data (c.f.
+Figure \@ref(fig:ch010-realDataFig).)
 
@@
   upper = list(
-    continuous = wrap('cor', method = "spearman")
+    continuous = GGally::wrap('cor', method = "spearman")
   ), 
@@
 The evaluation shows that the simulated counts approximately match the target parameters and exhibit the full range of estimated correlation from the data.
-Utilizing 15 CPU threads in a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, the simulation completed just shy of 7.5 minutes.
+Utilizing 15 CPU threads in a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, the simulation completed just over of 2 minutes.
 
 ```{r ch050-figBRCA, echo=F, eval=T, cache=T, fig.cap="`bigsimr` produces simulated random vectors from a multivariate negative binomial that replicate the estimated structure from an RNA-seq data set.
-The dashed red lines indicated equality between estimated parameters from simulated data(vertical axes) and the specified target parameters (horizontal axes)."}
-par(mfrow=c(1,3))
+The dashed red lines indicated equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes)."}
 ## estimate rho from sims and plot
 nb_Rho_hat <- cor_fast(sim_nbinom, method = corType)
+## triplot
+par(mfrow=c(1,3))
 plot(nb_Rho[lower.tri(nb_Rho)], nb_Rho_hat[lower.tri(nb_Rho_hat)],
      xlim = c(-1, 1), ylim = c(-1, 1),
-     pch = 4, cex = 1, col = rgb(0, 0, 1, 0.05),
+     pch = 1, cex = 1, col = rgb(0, 0, 0, 0.05),
      xlab = "Target Correlation", ylab = "Estimated Correlation",
-     main = "Correlation Matching")
+     main = "Spearman Cor. Matching")
 abline(a = 0, b = 1, col="red", lty="dashed")
@@
 plot(sizes, nbinom_size_prob_hat[1,],
-     pch = 4, cex = 1, col = rgb(0, 0, 1, 1),
+     pch = 1, cex = 1, col = rgb(0, 0, 0, 1),
      xlab = "Target Size", ylab = "Estimated Size",
@@
 plot(log(probs), log(nbinom_size_prob_hat[2,]),
-     pch = 4, cex = 1, col = rgb(0, 0, 1, 1),
+     pch = 1, cex = 1, col = rgb(0, 0, 0, 1),
      xlab = "Target Log Probability", ylab = "Estimated Log Probability",
-     main = "Marginal NB Probability Matching")
+     main = "Marginal NB Prob Matching")
 abline(a=0, b=1, col="red", lty="dashed")
@@
 
-## MC evaluation of correlation estimation efficiency
+## Evaluation of correlation estimation efficiency
 
@@
 MC methods are routinely used in many statistical inferential tasks including estimation, hypothesis testing, error rates, and empirical interval coverage rates.
-For an concise introduction to these methods, see, for example @Rizzo2007, Ch. 6.
-Now, we demonstrate how `bigsimr` can be used evaluate estimation efficiency.
-In particular, we'd like to assess the error our correlation estimation above.
+For an concise introduction to these methods, see @Rizzo2007, Ch. 6.
+To conclude the example applications, we demonstrate how `bigsimr` can be used evaluate estimation efficiency.
+In particular, we'd like to assess the error in our correlation estimation above.
 We used a conventional method, based on classical statistical theory. 
 Yet this method was not designed for high-dimensional data.
-Indeed, high-dimensional covariance estimation (and precision matrices) is an active area of statistical science (see, for example, @Won2013g and @VanWieringen2016).
-In this example, we simulate $m=`r m`$ data sets with the number of simulated vectors matching the number of patients in the BRCA data set, $N=`r nrow(brca)`$.
-Using the fact that our simulation is much faster for the Pearson correlation type (see Figure \@ref(fig:ch040-gpuVScpuFig), we convert the Spearman correlation matrix and ensure PSD outside the `rvec` call.
-At each iteration, we estimate the quadratic loss from the specified ${\bf R}^{Spearman}_{ \bf Y}$, producing a distribution of loss values.
+Indeed, high-dimensional covariance estimation (and precision matrices) is an active area of statistical science (see, for example, [@Won2013g; @VanWieringen2016].
 
-```{r ch050-quadloss, echo=TRUE, eval=T, cache=FALSE}
-quadLossRDS <- paste0( "results/brca", round(myProb*100), "quadLoss.rds" )
-if ( !file.exists( quadLossRDS) )  {
+In this small example, we simulate $m=`r m`$ data sets with the number of simulated vectors matching the number of patients in the BRCA data set, $N=`r nrow(brca)`$.
+Since our simulation is much faster for the Pearson correlation type (see Figure \@ref(fig:ch040-gpuVScpuFig)), we only convert the Spearman correlation matrix once (and ensure its PSD).
+At each iteration, we estimate the quadratic loss from the specified ${\bf R}_{Spearman}$, producing a distribution of loss values.
+
+```{r ch050-quadlossECHO, echo=T, eval=F, cache=FALSE}
+## Simulate random vectors equal to the sample size
 n <- nrow(brca)
 ## convert outside for faster simulation
-    nb_Rho_p <- bigsimr::cor_convert( rho = nb_Rho, from = corType, to = "pearson" )
+nb_Rho_p <- bigsimr::cor_convert( rho = nb_Rho,
+                                 from = corType, to = "pearson" )
 ## ensure PSD
@@
 ## find quadratic loss at each rep
-    quadLoss <- unlist( lapply( simRho, rags2ridges::loss, T = nb_Rho, type = "quadratic"  ) )
+quadLoss <- unlist( lapply( simRho, rags2ridges::loss,
+                           T = nb_Rho, type = "quadratic"))
+
+```
+
+```{r ch050-quadloss, echo=F, eval=T, cache=FALSE}
+quadLossRDS <- paste0( "results/brca", round(myProb*100), "quadLoss.rds" )
+if ( !file.exists( quadLossRDS) )  {
+    n <- nrow(brca)
+    ## convert outside for faster simulation
+    nb_Rho_p <- bigsimr::cor_convert( rho = nb_Rho,
+                                     from = corType, to = "pearson" )
+    ## ensure PSD
+    nb_Rho_p <- bigsimr::cor_nearPSD( G = nb_Rho_p )
+    ## create m random vectors and estimate correlation
+    simRho <- replicate(n = m,
+        expr = { tmpSim <- rvec(n = n , nb_Rho_p, nb_margins,
+          type = 'pearson', ensure_PSD = FALSE, cores = cores);
+          bigsimr::cor_fast( x = tmpSim, method = corType )} ,
+     simplify = FALSE)
+    ## find quadratic loss at each rep
+    quadLoss <- unlist( lapply( simRho, rags2ridges::loss,
+                               T = nb_Rho, type = "quadratic"))
     saveRDS(quadLoss, quadLossRDS)
@@
 
-This distribution could be compared to more elaborate, high-dimensional estimators to guide analysts in deciding whether the additional complexity and computation time are warranted for their application.
+This distribution could be compared to high-dimensional designed covariance estimators to guide to help decide whether the additional complexity and computation time are warranted.
 

### 2020-11-11 060-conclusion-discussion.Rmd
2nd draft. Revise ch060 conclusions

@@
 
-We've introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a high-performance implementation called [`bigsimr`](https://schisslergroup.github.io/bigsimr/). 
-The high-performance --- efficient, multi-core and GPU enabled --- algorithms and software provide 
+We've introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a high-performance `R` package called [`bigsimr`](https://schisslergroup.github.io/bigsimr/).
 The random vector generation method is largely inspired by NORTA (@Cario1997) and Gaussian copula-based approaches (@MB13, @BF17, @Xia17).
-The major contributions of our work presented here involve high-dimensional scalability, model flexibility, and high-performance implementation with broad potential data analytic applications for modern big data challenges.
+The major contributions of this work are high-dimensional scalability, flexible modeling of dependency, and an high-performance implementation with broad potential data analytic applications for modern, big-data statistical computing.
 
-It is usually customary to compare new tools and algorthms directly to existing competing methods.
-In our MC studies, however, we only employ our proposed methodology, since our previous work has show that existing tools are simply not designed or feasible to meet our HD goal (see @Li2019gpu for evalutions of the `R` `copula`package and others).
+It is customary to compare new tools and algorthms directly to existing competing methods and software.
+In this study, however, we only employ our proposed methodology, since our previous work has show that existing tools are simply not designed or feasible to meet our high-dimensional goal (see @Li2019gpu for evalutions of the `R` `copula`package and others).
 For the bivariate simulations, existing packages such as [`nortaRA`](https://github.com/cran/NORTARA/blob/master/inst/doc/NORTARA.R) work well to match Pearson correlations exactly.
@@
 There are limitations to the methodology and implementation.
-The most obvious missing feature the proposed methodology is the inability to match a Pearson correlation matrix exactly.
-As discussed in [Algorithms](algorithms) and extensively by @XZ19, this is a computational intense procedure and not a natural choice for characterizing dependency for non-normal marginals, especially in the high-dimensional setting.
-While we do not provide an implementation directly supporting Pearson matching, user can always supply their own input Pearson after using a supplementary matching scheme [@Cario1997; @XZ19].
+The most obvious missing feature of the proposed methodology is the inability to match a Pearson correlation matrix exactly.
+As discussed in [Algorithms](algorithms) and extensively by @XZ19, this is a computationally intense procedure and Pearson's correlation is not a natural choice to describe dependency for non-normal marginals.
+While we do not provide an implementation directly supporting Pearson matching, users may supply their own input Pearson correlation after using a supplementary matching scheme [@Cario1997; @XZ19].
 
@@
 
-Future work includes developing scalable algorithms to match the Pearson correlation matrix more precisely, discrete-margin specific modifications including fast Spearman's correlation rescaling (see Equation \@ref(eq:spearmanRescaled)), and *bona fide* high-dimensional correlation estimation (for example, see @Won2013g).
-From an implementation standpoint, `bigsimr` only supports Nvidia GPUs and refactoring the code using OpenCL would broaden the user base.
-As problems grow even larger, multi-GPU support is a logical progression in this highly parallelization algorthmic scheme and warrant further development.
+Future work includes developing scaleable algorithms to match the Pearson correlation matrix more precisely, discrete-margin specific modifications including fast Spearman's correlation rescaling (see Equation \@ref(eq:spearmanRescaled)), and high-dimensional covariance estimation.
+From an implementation standpoint, `bigsimr` only supports Nvidia GPUs and redesigning the code using OpenCL would broaden the users who would benefit.
+As data-analytic problems grow to even larger dimension, multi-GPU support is a promising hardware-based future direction.

### 2020-11-12 010-background.Rmd
3rd draft. Finalizing pdf output and small issues

@@
 
-```{r ch010-realDataFig, cache=F, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's $\rho_s$ for three example genes.
+```{r ch010-realDataFig, cache=F, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes.
 The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations.

### 2020-11-12 020-simulation-algorithm.Rmd
3rd draft. Finalizing pdf output and small issues

@@
 For high-dimensional data, these evaluations are often too costly to enable feasible simulation studies.
-For low-dimensional problems, methods and tools exist to match Pearson correlations precisely (see @Chen2001; @Xia17; @MB13), including the publicly available `nortaRA` R package.
+For low-dimensional problems, methods and tools exist to match Pearson correlations precisely, see @Chen2001; @Xia17; @MB13 and the publicly available `nortaRA` R package.
 
@@
 
-Despite all this, if one does desire to characterize dependency using Pearson correlations, simply using the target Pearson correlation matrix as the initial conditions to our proposed algorithm will lead to approximate matching in the resultant distribution (@Song00) in many practical applications.
+Despite all this, if one does desire to characterize dependency using Pearson correlations, simply using the target Pearson correlation matrix as the initial conditions to our proposed algorithm will lead to approximate matching in the resultant distribution [@Song00] in many practical applications.
 The quality of this approximation depends on the setting, but in practice, for high-dimensional count data we find the accuracy to be adequate.
@@
 
-The pre-processing (Step1) takes advantage of the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertSpearman) or \@ref(eq:convertKendall), respectively (implemented as `bigsimr::cor_covert`).
+The pre-processing (Step 1) takes advantage of the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertSpearman) or \@ref(eq:convertKendall), respectively (implemented as `bigsimr::cor_covert`).
 

### 2020-11-12 050-examples.Rmd
3rd draft. Finalizing pdf output and small issues

@@
 ## knitr::opts_chunk$set(echo = FALSE, cache = FALSE)
+knitr::opts_chunk$set(dev = "png", dev.args = list(type = "cairo-png"))
 # reticulate::use_condaenv("bigsimr")
@@
 
-Figure \@ref(fig:ch050-simDataFig) displays the simulated counts and pairwise relationships for our example genes in Table \@ref(tab:ch010-realDataFig).
+Figure \@ref(fig:ch050-simDataFig) displays the simulated counts and pairwise relationships for our example genes in Table \@ref(tab:ch010-realDataTab).
 Simulated counts roughly mimick the observed data but with a smoothier appearance due to the assumed parameter form and with less extreme points then the observed data (c.f.
@@
 
-```{r ch050-simDataFig, echo = FALSE, eval=TRUE, cache=T, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Simulated data for 3 selected high-expressing genes, replicating the observed data structure.'}
+```{r ch050-simDataFig, echo = FALSE, eval=TRUE, cache=T, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Simulated data for 3 selected high-expressing genes, replicating the observed data structure.', dpi=36}
 exampleGenes <- readRDS(file = './data/exampleGenes.rds')
@@
 
-```{r ch050-figBRCA, echo=F, eval=T, cache=T, fig.cap="`bigsimr` produces simulated random vectors from a multivariate negative binomial that replicate the estimated structure from an RNA-seq data set.
-The dashed red lines indicated equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes)."}
+```{r ch050-figBRCA, echo=F, eval=T, cache=T,fig.cap="`bigsimr` produces simulated random vectors from a multivariate negative binomial that replicate the estimated structure from an RNA-seq data set.
+The dashed red lines indicated equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes).", dpi=36}
 ## estimate rho from sims and plot
@@
 where ${\bf Y^{(b)} }$ is the $b^{th}$ simulated vector in a total of $B$ simulation replicates and $I ( )$ is the indicator function.
-For example, we can estimate from our $B=10,000$ simulated vectors the probability that all genes are expressed (i.e., ${\bf y}_i \geq 1, \forall \; i$).
+For example, we can estimate from our $B=10,000$ simulated vectors the probability that all genes are expressed (i.e., ${\bf y}_i \geq 1, \forall \; i$) is $0.1708$.
 

### 2020-11-12 index.Rmd
3rd draft. Finalizing pdf output and small issues

@@
 subtitle: "using the bigsimr R Package"
-author: "Alexander D. Knudson, Tomasz J. Kozubowski, Anna K. Panorska, Juli Petereit, and Alfred G. Schissler"
+author:
+  - Alfred G. Schissler
+  - Alexander D. Knudson
+  - Tomasz J. Kozubowski
+  - Anna K. Panorska
+  - Juli Petereit
 bibliography: ["bigsimr.bib", "packages.bib"]

### 2020-12-01 020-simulation-algorithm.Rmd
Draft, revise NORTA in higher d, cite GH02

@@
 
-A complication often arises at this stage: the parallelized, pairwise conversions may create a (spuriously) non-positive definite correlation matrix, especially in high dimensions.
-Researchers working in multivariate computation frequently encounter such difficulties and need to find a close positive (semi-)definiteness matrix.
-The most widely-available routine for this task in `R` is called `matrix::nearPD` and is not suitable for high dimensions.
+A complication often arises at this stage: the resultant matrix may become indefinite, either through numerical error or naturally occurring in the correlation conversion.
+Researchers working in multivariate computation frequently encounter such difficulties and need to find a close positive (semi-)definite matrix.
+A widely-available routine for this task in `R` is called `matrix::nearPD`, though it is not suitable for high dimensions.
 To overcome this issue, we've developed `bigsimr::nearPSD`, a quadractically-convergent Netwon method for finding the nearest correlation matrix, developed by @QS2006.
@@
 Specifically, step 2i uses either an efficient multicore multivariate normal simulator (the `R` package `mvnfast`[@Fasiolo2016]) or a using Google's JAX python library NumPy for graphical-processing-unit (GPU) acceleration of the Cholesky factorization and matrix multiplication (steps 3, 4 in the NORTA algorithm in the preceeding section).
-(note that JAX is a not acronym and can be thought of a high-performance NumPy).
+
+## NORTA in higher dimensions
+
+Sklar's theorem provides a useful characterization of multivariate distributions through copulas.
+Yet the choice of copula-based simulation algorithm affects which joint distributions may be simulated.
+Even in low-dimensional spaces (e.g., $d=3$), there exists valid multivariate distributions with *feasible* Pearson correlation matrices that NORTA cannot match exactly [@LH75].
+This occurs when the bivariate transformations are applied to find the input correlation matrix yet when combined the resultant matrix is indefinite.
+These situations do occur, even using exact analytic calculations. Such problematic target correlation matrices are termed *NORTA defective*.
+
+@GH02 conducted a Monte Carlo study to estimate the probability of NORTA defective matrices while increasing the dimension $d$. 
+They found at what is now considered low-to-moderate dimensions ($d \approx 20$) that almost *all* feasible matrices are NORTA defective.
+This stems from the concentration of measure near the boundary of the space of all possible correlation matrices as dimension increases.
+Unfortunately, it is precisely near this boundary that NORTA defective matrices reside.
+
+There is hope, however, as @GH02 also show that by augmenting the NORTA procedure by replacing the indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$.
+This provides evidence that our nearest positive semidefinite (PSD) augmented approach will maintain reasonable accuracy if our input matching scheme returns an indefinite matrix, at least for the rank-based matching scheme described above.
 

### 2020-12-01 050-examples.Rmd
Draft, revise NORTA in higher d, cite GH02

--- 2020-11-11 050-examples.Rmd
@@
 
-```{r ch050-simDataFig, echo = FALSE, eval=TRUE, cache=T, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Simulated data for 3 selected high-expressing genes, replicating the observed data structure.'}
+```{r ch050-simDataFig, echo = FALSE, eval=T, cache=F, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Simulated data for 3 selected high-expressing genes, replicating the observed data structure.'}
 exampleGenes <- readRDS(file = './data/exampleGenes.rds')
@@
 
-```{r ch050-figBRCA, echo=F, eval=T, cache=T, fig.cap="`bigsimr` produces simulated random vectors from a multivariate negative binomial that replicate the estimated structure from an RNA-seq data set.
+```{r ch050-figBRCA, echo=F, eval=T, cache=F, fig.cap="Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set.
 The dashed red lines indicated equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes)."}

### 2020-12-01 050-examples.Rmd
reduce size of large figures

@@
 
-```{r ch050-simDataFig, echo = FALSE, eval=T, cache=F, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Simulated data for 3 selected high-expressing genes, replicating the observed data structure.'}
+```{r ch050-makeSimDataFig, echo = FALSE, eval=T, cache=F, message=FALSE, warning=FALSE}
+if( !file.exists("fig/ch050-simDataFig.png") ) {
     exampleGenes <- readRDS(file = './data/exampleGenes.rds')
-GGally::ggpairs(
+    p <- GGally::ggpairs(
                      data = as.data.frame( sim_nbinom[ ,exampleGenes] ), 
@@
                          continuous = GGally::wrap('cor', method = "spearman")
-  ), 
-) + theme_bw()
+                     )) + theme_bw()
+    ## ggsave("fig/ch050-simDataFig.png", plot = p, width = 5, height = 4, dpi = 300)
+    ggsave("fig/ch050-simDataFig.png", plot = p, dpi = 300) 
+}
+```
+
+```{r ch050-simDataFig, echo=F, fig.align='center', fig.cap='Simulated data for three selected high-expressing genes replicates the estimated data structure.'}
+knitr::include_graphics('fig/ch050-simDataFig.png')
 ```
@@
 
-```{r ch050-figBRCA, echo=F, eval=T, cache=F, fig.cap="Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set.
-The dashed red lines indicated equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes)."}
+```{r ch050-makeFigBRCA, echo=F, eval=T, cache=F}
+if( !file.exists("fig/ch050-figBRCA.png") ) {
     ## estimate rho from sims and plot
@@
     ## triplot
+    ## png("fig/ch050-figBRCA.png", width = 4, res = 100, units = 'in')
+    png("fig/ch050-figBRCA.png", res = 300, width = 5, height = 4, units = 'in')
     par(mfrow=c(1,3))
@@
          pch = 1, cex = 1, col = rgb(0, 0, 0, 0.05),
-     xlab = "Target Correlation", ylab = "Estimated Correlation",
-     main = "Spearman Cor. Matching")
+         xlab = "Target Correlation", ylab = "Estimated Correlation")
+         ## main = "Spearman Cor. Matching")
     abline(a = 0, b = 1, col="red", lty="dashed")
@@
          pch = 1, cex = 1, col = rgb(0, 0, 0, 1),
-     xlab = "Target Size", ylab = "Estimated Size",
-     main = "Marginal NB Size Matching")
+         xlab = "Target Size", ylab = "Estimated Size")
+         ## main = "Marginal NB Size Matching")
     abline(a=0, b=1, col="red", lty="dashed")
@@
          pch = 1, cex = 1, col = rgb(0, 0, 0, 1),
-     xlab = "Target Log Probability", ylab = "Estimated Log Probability",
-     main = "Marginal NB Prob Matching")
+         xlab = "Target Log Probability", ylab = "Estimated Log Probability")
+         ## main = "Marginal NB Prob Matching")
     abline(a=0, b=1, col="red", lty="dashed")
+    dev.off()
     par(mfrow=c(1,1))
+}
+```
+
+```{r ch050-figBRCA, echo=F, out.width = '80%', fig.align='center', fig.cap="Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set.
+The dashed red lines indicated equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes)."}
+knitr::include_graphics('fig/ch050-figBRCA.png')
 ```
@@
 
-The standard `R` summary function supplies the mean-augmented five-number summary of the quadratic loss distribution computed above.
+The `R` summary function supplies the mean-augmented five-number summary of the quadratic loss distribution computed above.
 

### 2020-12-01 010-background.Rmd
3rd draft. run spell checking and edit

@@
 
-The `bigsimr` `R` package provides multiple algorithms to work with high-dimesional multivariate data, but all these algorithms were originally designed to support a single task:
+The `bigsimr` `R` package provides multiple algorithms to work with high-dimensional multivariate data, but all these algorithms were originally designed to support a single task:
 to generate random vectors drawn from multivariate probability distribution with given marginal distributions and dependency metrics.
-Specifically, our goal is to efficiently simulate a large number, $B$, of random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with **correlated** components and hetereogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i$, where $d$ can be very large and still be computed in practically useful times.
+Specifically, our goal is to efficiently simulate a large number, $B$, of random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with **correlated** components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i$, where $d$ can be very large and still be computed in practically useful times.
 
@@
 
-Simulating high-dimensional, non-normal, correleated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
+Simulating high-dimensional, non-normal, correlated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
 The RNA-seq data-generating process involves counting how often a particular messenger RNA (mRNA) is expressed in a biological sample.
@@
 
-We'll illustrate our methodology using the well-studied Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgements).
+We'll illustrate our methodology using the well-studied Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments).
 For simplicity, we only consider high expressing genes.
@@
 This gives a massive number of pairwise dependencies among the marginals (specifically, $`r choose(d,2)`$ correlation parameters).
-We further process the data to illustrate our methodology's flexible and robust simulation scheme, while better aligning with the actual data-generating process, by rounding @Li2011c's RNA-seq by Expectation Maximization (RSEM) values to counts.
+We further process the data to illustrate our methodology's flexible and robust simulation scheme, while better aligning with the actual data-generating process, by rounding @Li2011 RNA-seq by Expectation Maximization (RSEM) values to counts.
 Table \@ref(tab:ch010-realDataTab) displays counts for three selected high-expressing genes for the first five patients' breast tumor samples.
@@
 
-Two nonparameteric, rank-based measures common in practice are Spearman's correlation (denoted $\rho_S$) and Kendall's $\tau$. 
+Two nonparametric, rank-based measures common in practice are Spearman's correlation (denoted $\rho_S$) and Kendall's $\tau$. 
 Define
@@
 Kendall's $\tau$, on the other hand, is the difference in probabilities of concordant and discordant pairs of observations $(X_i, Y_i)$ and $(X_j, Y_j)$.
-By concordance we mean that orderings have the same direction (e.g., if $X_i < X_j$, then $Y_i < Y_j$) and is determinded by the ranks of the values, not the values themselves.
+By concordance we mean that orderings have the same direction (e.g., if $X_i < X_j$, then $Y_i < Y_j$) and is determined by the ranks of the values, not the values themselves.
 
@@
 *Correspondence among Pearson, Spearman, $\tau$ correlations*.
-This is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficent, as the marginal distributions $F_i$ are intrinstic in their calculation.
+This is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficient, as the marginal distributions $F_i$ are intrinsic in their calculation.
 But for **bivariate normal vectors**, however, the correspondence is well-known:
@@
 
-\noindent where $U$ is a uniform random variable in $(0,1)$, and $F^{-1}_1, F^{-1}_2$ are the inverse cdf of random variables $X_1$ and $X_2$, respectively. 
+\noindent where $U$ is a uniform random variable in $(0,1)$, and $F^{-1}_1, F^{-1}_2$ are the inverse CDF of random variables $X_1$ and $X_2$, respectively. 
 For discrete random variables, define $F^{-1}$ as in Equation \@ref(eq:inverseCDF).
@@
 
-A Gaussian copula is the case where all marginal CDFs $F_i$ are the standard normal cdf, $\Phi$. 
+A Gaussian copula is the case where all marginal CDFs $F_i$ are the standard normal CDF, $\Phi$. 
 This representation corresponds to a multivariate normal distribution with standard normal marginal distributions and covariance matrix ${\bf R_P}$.
@@
 
-Sklar's Theorem [@Sklar1959; @Ubeda-Flores2017] guarantees that given inverse CDFs $F_i^{-1}$s and a valid correlation matrix (within the Frechet bounds) *any* random vector can be obtained via transformations involving copula functions.
+Sklar's Theorem [@Sklar1959; @Ubeda-Flores2017] guarantees that given inverse CDFs $F_i^{-1}$s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions.
 For example, using Gaussian copulas, we can construct a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)$ with $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$, via ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$ povides $Y_i \sim F_i, \, \forall \, i$ .

### 2020-12-01 020-simulation-algorithm.Rmd
3rd draft. run spell checking and edit

@@
 
-With modern parallelized computing, steps 1, 3, 4, 5 are readily implemented as high-performance, multicore and/or graphical-processing-unit (GPU) accelerated, algorithms --- providing the fast scaleability using readily-available hardware.
+With modern parallelized computing, steps 1, 3, 4, 5 are readily implemented as high-performance, multi-core and/or graphical-processing-unit (GPU) accelerated, algorithms --- providing the fast scaleability using readily-available hardware.
 
@@
 Since our goal is to simulate non-normal marginals, we greatly prefer the use of rank-based measures $\rho_S$ and $\tau$ from a modeling standpoint.
-Further, $\rho_S$ and $\tau$'s invariance under monotone transformation (see [Background](#background)), perserves the correlation coefficients through steps 3, 4, and 5 in the NORTA algorithm above.
+Further, $\rho_S$ and $\tau$'s invariance under monotone transformation (see [Background](#background)), preserves the correlation coefficients through steps 3, 4, and 5 in the NORTA algorithm above.
 This eliminates the need for computing the $\binom{d}{2}$ integrals to match exactly (nothing is for free, however, as discussed in below in Section \@ref(rand-vec-gen)).
@@
 The quality of this approximation depends on the setting, but in practice, for high-dimensional count data we find the accuracy to be adequate.
-Later, we'll study the robustness of our method to this limitation in selected [Monte Carlo evalutations]({#simulations}).
+Later, we'll study the robustness of our method to this limitation in selected [Monte Carlo evaluations]({#simulations}).
 
@@
 
-1. Pre-processing for nonparameteric dependency matching.  
+1. Pre-processing for nonparametric dependency matching.  
 	(i) Convert from either ${\bf R_{Spearman}}$ or ${\bf R_{Kendall}}$ into the corresponding MVN input correlation ${\bf R_{Pearson}}$.
@@
 A widely-available routine for this task in `R` is called `matrix::nearPD`, though it is not suitable for high dimensions.
-To overcome this issue, we've developed `bigsimr::nearPSD`, a quadractically-convergent Netwon method for finding the nearest correlation matrix, developed by @QS2006.
+To overcome this issue, we've developed `bigsimr::nearPSD`, a quadractically-convergent Newton method for finding the nearest correlation matrix, developed by @QS2006.
 We hope that this routine could be useful in many applications aside from our primary goal of random vector generation.
@@
 Once the target margins and algorithm inputs are determined, steps 2 and 3 are essentially a NORTA algorithm with modern high-performance computing implementations.
-Specifically, step 2i uses either an efficient multicore multivariate normal simulator (the `R` package `mvnfast`[@Fasiolo2016]) or a using Google's JAX python library NumPy for graphical-processing-unit (GPU) acceleration of the Cholesky factorization and matrix multiplication (steps 3, 4 in the NORTA algorithm in the preceeding section).
+Specifically, step 2i uses either an efficient multi-core multivariate normal simulator (the `R` package `mvnfast`[@Fasiolo2016]) or a using Google's JAX python library NumPy for graphical-processing-unit (GPU) acceleration of the Cholesky factorization and matrix multiplication (steps 3, 4 in the NORTA algorithm in the preceding section).
 
-## NORTA in higher dimensions
+## A note on NORTA in higher dimensions
 
@@
 There is hope, however, as @GH02 also show that by augmenting the NORTA procedure by replacing the indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$.
-This provides evidence that our nearest positive semidefinite (PSD) augmented approach will maintain reasonable accuracy if our input matching scheme returns an indefinite matrix, at least for the rank-based matching scheme described above.
+This provides evidence that our nearest positive semi-definite (PSD) augmented approach will maintain reasonable accuracy if our input matching scheme returns an indefinite matrix, at least for the rank-based matching scheme described above.
 

### 2020-12-01 030-bigsimr-package.Rmd
3rd draft. run spell checking and edit

@@
 The `bigsimr` package is a high-performance implementation of the proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms) for details).
-When designing `bigsimr`, we aimed to conveniently provide parallelized computation, through multi-core and GPU acceleration, while allowing advanced users to customize and automate worflows.
+When designing `bigsimr`, we aimed to conveniently provide parallelized computation, through multi-core and GPU acceleration, while allowing advanced users to customize and automate workflows.
 
 This subsections below describe the basic use of `bigsimr`, by stepping through a low-dimensional (2D) simulation workflow for the often-used example data set `airquality`.
-This workflow proceeds from data, to estimation, simulation configuration, random vector generation, and result vizualization.
-For this low-dimensional setting, we compute using a single central processing unit (CPU) as the overhead in forking the tasks to multiple cores outways the computational gains. 
+This workflow proceeds from data, to estimation, simulation configuration, random vector generation, and result visualization.
+For this low-dimensional setting, we compute using a single central processing unit (CPU) as the overhead in forking the tasks to multiple cores outweighs the computational gains. 
 Then we transition to advanced use where we briefly describe some of the high-performance features and syntax.
@@
 We'll demonstrate the basic use and syntax of `bigsimr` through an example workflow applied to the New York air quality data set (`airquality`) included in the R `datasets` package.
-First, we load the `bigsimr` library and a few other convenient data science packages, including the syntacically-elegant `tidyverse` suite of `R` packages.
+First, we load the `bigsimr` library and a few other convenient data science packages, including the syntactically-elegant `tidyverse` suite of `R` packages.
 
@@
 We aim to simulate random two-component vectors mimicking this structure. 
-The margins are not normally distributed, particulary the ozone level exhibits a strong positive skew.
+The margins are not normally distributed, particularly the ozone level exhibits a strong positive skew.
 
@@
 *Specifying marginal distributions*.
-Based on the estimated densities in Figure \@ref(fig:ch030-aq-joint-dist), we'll assume `Temp` is normally ditsributed and `Ozone` is log-normally distributed since its values are positive and skewed.
-We'll use the well-known, unbiased estimators for the normal distribution's parameters and maximum likelihood estimators for the lognormal parameters:
+Based on the estimated densities in Figure \@ref(fig:ch030-aq-joint-dist), we'll assume `Temp` is normally distributed and `Ozone` is log-normally distributed since its values are positive and skewed.
+We'll use the well-known, unbiased estimators for the normal distribution's parameters and maximum likelihood estimators for the log-normal parameters:
 
@@
 
-Notice that we use the *quantile* function for the marginals, as that is how the marginal distributions $F_i$ enter into the `bigsimr::rvec` algorthm.
+Notice that we use the *quantile* function for the marginals, as that is how the marginal distributions $F_i$ enter into the `bigsimr::rvec` algorithm.
 This implementation strategy supports all `R` `base` probability distributions.
@@
 
-*Specfying correlation*.
+*Specifying correlation*.
 As mentioned, the user must decide how to describe correlation, based on the particulars of the problem.
 For non-normal data and for improved simulation accuracy in our scheme, we advocate the use of rank-based correlations Spearman's $\rho_S$ and Kendall's $\tau$.
-But we also support approximate Pearson correlation coefficient matching, while cautioning the user to check the performance for their parameteric multivariate model (see [Monte Carlo evaluations](#simulations) for evaluation strategies and guidance).
+But we also support approximate Pearson correlation coefficient matching, while cautioning the user to check the performance for their parametric multivariate model (see [Monte Carlo evaluations](#simulations) for evaluation strategies and guidance).
 To aid in correlation specification, and estimation in general, we provide a high-performance function `bigsimr::cor_fast` which estimates Pearson, Spearman, or Kendall correlation using the fastest methods available.
@@
 *Checking the theoretical correlation bounds*
-As discussed in Section \@ref(background), given a pair of marignal distributions the possible correlations are not free to vary between $[-1, 1]$. 
-To ensure that the simulation is not configured to impossible settings, we provide the `bigsimr::cor_bounds` fuction provides MC estimated theoretical lower and upper bounds (using the Generate, Sort, and Correlate algorithm of @DH2011).
+As discussed in Section \@ref(background), given a pair of marginal distributions the possible correlations are not free to vary between $[-1, 1]$. 
+To ensure that the simulation is not configured to impossible settings, we provide the `bigsimr::cor_bounds` function provides MC estimated theoretical lower and upper bounds (using the Generate, Sort, and Correlate algorithm of @DH2011).
 
@@
 ```{r ch030-plot-sim, fig.cap="Contour plot and marginal densities for the simulated bivariate distribution of Air Quality Temperatures and Ozone levels.
-The simulated points mimick the observed data with respect to both the marginal characteristics and bivariate association."}
+The simulated points mimic the observed data with respect to both the marginal characteristics and bivariate association."}
 p1 <- df_sim %>%
@@
 Because `bigsimr` uses an `alist` to store the margins, any probability distribution with a well-defined inverse CDF can be used including custom marginal distributions not provided in base `R`. 
-Therefore, to specify a marginal distribution absent from an existing packge, the user needs to provide a closed-form expression to form the corresponding quantile function.
+Therefore, to specify a marginal distribution absent from an existing package, the user needs to provide a closed-form expression to form the corresponding quantile function.
 

### 2020-12-01 040-monte-carlo-evaluation.Rmd
3rd draft. run spell checking and edit

@@
 
-Figure \@ref(fig:ch040-bPlot) displays the aggregated bivariate simulation results with a detailed summarization.
+```{r ch040-biMAEtable, results='asis', cache=F}
+## tabMSE  <- allDat %>%
+##    group_by( N, margins ) %>%
+##    summarize( mse = mean( ( rho - rho_hat )^2 ) )
+tabMSE  <- allDat %>%
+    ## filter( N == 100000) %>%
+    group_by( N, type, margins ) %>%
+##     summarize( mse = mean( ( rho - rho_hat )^2 ) ) %>%
+    summarize( MAE = mean( abs( rho - rho_hat )  ) )
+## normMSE <- as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ] ) 
+## ( tabMSE$mse - normMSE ) / normMSE
+## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) )
+## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) ) / tabMSE$mse
+
+knitr::kable(tabMSE,
+             format = 'latex',
+             col.names = c("No. of random vectors",
+                           "Correlation type",
+                           "Distribution",
+                           "Mean abs. error"),
+             booktabs = TRUE,
+             caption = 'Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.',
+             linesep = c('', '',  '\\addlinespace')
+)
+```
+
+Figure \@ref(fig:ch040-bPlot) displays the aggregated bivariate simulation results.
 Table \@ref(tab:ch040-biMAEtable) contains the mean absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.
@@
 Only the normal margins match the Pearson correlations (top row) exactly across the entire range of correlations.
-The two non-normal bivariate random vectors experience attentuation (downward bias) for extreme Pearson correlations (due to limitations in our algorithm) and more restriction (due to the Frechet bounds).
+The two non-normal bivariate random vectors experience attenuation (downward bias) for extreme Pearson correlations (due to limitations in our algorithm) and more restriction (due to the Frechet bounds).
 The rank-based correlations (bottom two rows) are matched exactly for all margins and obtain the full range of possible dependencies."}
@@
 
-```{r ch040-biMAEtable, results='asis', cache=F}
-## tabMSE  <- allDat %>%
-##    group_by( N, margins ) %>%
-##    summarize( mse = mean( ( rho - rho_hat )^2 ) )
-tabMSE  <- allDat %>%
-    ## filter( N == 100000) %>%
-    group_by( N, type, margins ) %>%
-##     summarize( mse = mean( ( rho - rho_hat )^2 ) ) %>%
-    summarize( MAE = mean( abs( rho - rho_hat )  ) )
-## normMSE <- as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ] ) 
-## ( tabMSE$mse - normMSE ) / normMSE
-## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) )
-## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) ) / tabMSE$mse
-
-knitr::kable(tabMSE,
-             format = 'latex',
-             col.names = c("No. of random vectors",
-                           "Correlation type",
-                           "Distribution",
-                           "Mean abs. error"),
-             booktabs = TRUE,
-             caption = 'Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.',
-             linesep = c('', '',  '\\addlinespace')
-)
-```
-
 The accuracy appears adequate for many applications.

### 2020-12-01 050-examples.Rmd
3rd draft. run spell checking and edit

@@
 This section demonstrates how to simulate multivariate data using `bigsimr`, aiming to replicate the structure of high-dimensional dependent count data.
-In an illustration of our proposed methodology applied to real data, we seek to simulate RNA-sequencing data by producing simulated random vectors with assumed marginal distributions with estimated parameters that mimick the observed data and its generating process.
-We seek scaleable realistic multivariate simulation to enable Monte Carlo (MC) methods for thes data.
+In an illustration of our proposed methodology applied to real data, we seek to simulate RNA-sequencing data by producing simulated random vectors with assumed marginal distributions with estimated parameters that mimic the observed data and its generating process.
+We seek scaleable realistic multivariate simulation to enable Monte Carlo (MC) methods for these data.
 Modeling RNA-seq using multivariate probabilities distributions is natural as inter-gene correlation is an inherent part of biological processes.
@@
 We use method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model.
-The marginal distributions are from the same probability family (NB) yet are hetereogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.
+The marginal distributions are from the same probability family (NB) yet are heterogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.
 The functions below help achieve this estimation and specifying the inputs for use in `bigsimr::rvec`.
@@
 Figure \@ref(fig:ch050-simDataFig) displays the simulated counts and pairwise relationships for our example genes in Table \@ref(tab:ch010-realDataTab).
-Simulated counts roughly mimick the observed data but with a smoothier appearance due to the assumed parameter form and with less extreme points then the observed data (c.f.
+Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parameter form and with less extreme points then the observed data (c.f.
 Figure \@ref(fig:ch010-realDataFig).)
@@
 where $\chi_i$ is the sample space for the $i^{th}$ component of the random vector $\bf{Y}$.
-Compact representations with convenient computational forms are rare for high-dimensional constructions, especially with hetereogeneous, correlated marginal distributions (or margins of mixed data types).
+Compact representations with convenient computational forms are rare for high-dimensional constructions, especially with heterogeneous, correlated marginal distributions (or margins of mixed data types).
 Given a large number simulated vectors as produced above, estimated probabilities are readily given by counting the proportion of simulated vectors meeting the desired condition.

### 2020-12-01 060-conclusion-discussion.Rmd
3rd draft. run spell checking and edit

@@
 The random vector generation method is largely inspired by NORTA (@Cario1997) and Gaussian copula-based approaches (@MB13, @BF17, @Xia17).
-The major contributions of this work are high-dimensional scalability, flexible modeling of dependency, and an high-performance implementation with broad potential data analytic applications for modern, big-data statistical computing.
+The major contributions of this work are high-dimensional scaleability, flexible modeling of dependency, and an high-performance implementation with broad potential data analytic applications for modern, big-data statistical computing.
 
-It is customary to compare new tools and algorthms directly to existing competing methods and software.
-In this study, however, we only employ our proposed methodology, since our previous work has show that existing tools are simply not designed or feasible to meet our high-dimensional goal (see @Li2019gpu for evalutions of the `R` `copula`package and others).
+It is customary to compare new tools and algorithms directly to existing competing methods and software.
+In this study, however, we only employ our proposed methodology, since our previous work has show that existing tools are simply not designed or feasible to meet our high-dimensional goal (see @Li2019gpu for evaluations of the `R` `copula`package and others).
 For the bivariate simulations, existing packages such as [`nortaRA`](https://github.com/cran/NORTARA/blob/master/inst/doc/NORTARA.R) work well to match Pearson correlations exactly.

### 2020-12-01 070-supplementary-materials.Rmd
edit overall. Cut RNA-seq MSE sizes. Revise RNA-seq example heavily.
3rd draft. Finalizing pdf output and small issues
Draft, revise NORTA in higher d, cite GH02
3rd draft. run spell checking and edit

@@
-# Supplementary Materials {#misc}
+# Supplementary Materials {.unnumbered #misc}
 
-We provide an open-source implementation of our methology as the `bigsimr` R package, hosted on github, https://schisslergroup.github.io/bigsimr/.
+We provide an open-source implementation of our methodology as the `bigsimr` R package, hosted on github, https://schisslergroup.github.io/bigsimr/.
 
-# Acknowledgement(s)
+```{r, out.width='5%', fig.align='center'}
+knitr::include_graphics('images/hex-bigsimr.png')
+```
+
+# Acknowledgment(s) {.unnumbered #acknowledgments}
 
@@
 
-# Disclosure statement
+# Disclosure statement {.unnumbered #coi}
 
@@
 
-# Funding
+# Funding {.unnumbered #funding}
 

### 2020-12-01 index.Rmd
3rd draft. run spell checking and edit

@@
 As new methods are developed for these data, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness --- often in comparison to existing methods.
-Further, efficient simulation empowers statistical computing strategies, such as the parameteric bootstrap [@Rizzo2007] to simulate from a hypothesized null model, providing inference in analytically challenging settings.
+Further, efficient simulation empowers statistical computing strategies, such as the parametric bootstrap [@Rizzo2007] to simulate from a hypothesized null model, providing inference in analytically challenging settings.
 Such Monte Carlo (MC) techniques become difficult for high-dimensional data with the current existing algorithms and tools.
@@
 As others have noted, it can be vexing to simulate dependent, non-normal/discrete data ---  even for low dimensional settings [@MB13; @XZ19].
-For continous non--normal multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula [@Nelsen2007] approaches are well-studied with flexible, robust software available [@Yan2007; @Chen2001].
+For continuous non--normal multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula [@Nelsen2007] approaches are well-studied with flexible, robust software available [@Yan2007; @Chen2001].
 Yet these approaches do not scale in a timely fashion to high-dimensional problems [@Li2019gpu]. 
 For discrete data, early simulation strategies had major flaws --- such as failing to obtain the full range of possible dependencies (e.g., admitting only positive correlations @Park1996). 
-While more recent approaches [@MB13; @Xia17; @BF17] have largely rememdied this issue for low-dimesional problems, the existing tools are not designed to scale to high dimensions.
+While more recent approaches [@MB13; @Xia17; @BF17] have largely remedied this issue for low-dimensional problems, the existing tools are not designed to scale to high dimensions.
 
@@
 
-With all this mind, we present a scalable, flexible multivariate simulation algorithm. 
+With all this mind, we present a scaleable, flexible multivariate simulation algorithm. 
 The crux of the method lies in the construction of a Gaussian copula, in the spirit of the NORTA procedure.
-Further, we introduce the `bigsimr` R package that provides parallelized, high-performance software implemented our NORTA-inspirsed algorithm.
-The algorithm design leverages useful properties of nonparameteric correlation measures, namely invariance under monotone transformation and well-known closed form relationships between dependency measures for the multivariate normal (MVN) distribution.
+Further, we introduce the `bigsimr` R package that provides parallelized, high-performance software implemented our NORTA-inspired algorithm.
+The algorithm design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed form relationships between dependency measures for the multivariate normal (MVN) distribution.
 

### 2020-12-02 010-background.Rmd
pdf render before sharing with co-authors

@@
 This gives a massive number of pairwise dependencies among the marginals (specifically, $`r choose(d,2)`$ correlation parameters).
-We further process the data to illustrate our methodology's flexible and robust simulation scheme, while better aligning with the actual data-generating process, by rounding @Li2011 RNA-seq by Expectation Maximization (RSEM) values to counts.
+We further process the data to illustrate our methodology's flexible and robust simulation scheme, while better aligning with the actual data-generating process, by rounding @Li2011c RNA-seq by Expectation Maximization (RSEM) values to counts.
 Table \@ref(tab:ch010-realDataTab) displays counts for three selected high-expressing genes for the first five patients' breast tumor samples.

### 2020-12-02 notebooks/cmstats2020_talk.rmd
1st draft CMStats 2020 talk

---
title: "Simulating Ultra High-Dimensional Multivariate Data"
subtitle: "Using the `bigsimr` R package"
author:
  - Alfred G. Schissler
  - Alexander D. Knudson
  - Tomasz J. Kozubowski
  - Anna K. Panorska
  - Juli Petereit
institute: |
    "Department of Mathematics & Statistics" |
    "University of Nevada, Reno"
date: "21 Dec 2020 (updated: `r Sys.Date()`)"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
    logo: ../images/hex-bigsimr-60x60.png
bibliography: ../bigsimr.bib
abstract: "It is critical to realistically simulate data when conducting Monte Carlo studies and methods. But measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers may resort to simulation designs that posit independence or perform arbitrary data transformations. This greatly diminishes insights into the empirical operating characteristics of any proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness. This article introduces the `bigsimr` R package that provides a flexible, scaleable procedure to simulate high-dimensional random vectors with given marginal characteristics and dependency measures. We'll describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and find close positive semi-definite matrices. Finally, we demonstrate the power of `bigsimr` by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d =1026$"
---

```{r LoadLib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
library(tidyverse)
set.seed(12212020)
## devtools::install_github("ggobi/GGally")
## for ggpairs
library(GGally)
```

<!-- 20-25 minutes -->

# 1 – Introduction

## Problem we're addressing

- In this era of Big Data, it is critical to realistically simulate data.
- This is computationally challenging when data are correlated and (ultra-) high dimensional.
- Many domains are producing these type of data.
- A prototypical example is RNA-sequencing data derived from high-throughput biomedical experiments.

## Our goal

- Our goal is to simulate $N$ random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with **correlated** components.
- The simulation replicates $N$ can be large (~100,000).
- And the dimension $d$ can be very large (~20,000).
- We propose a scaleable NORmal To Anything approach to this task .
- We provide a high-performance and GPU-accelerated implementation in the `R` package `bigsimr`.

# 2 - Background

## Motivating example: RNA-seq data

Simulating high-dimensional, non-normal, correlated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
The RNA-seq data-generating process involves counting how often a particular messenger RNA (mRNA) is expressed in a biological sample.
RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data.
For human derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted.
Importantly, due to inherent biological processes, gene expression data exhibits correlation --- co-expression --- across genes [@BE07; @Schissler2018].

```{r ch010-myProb, include=FALSE, cache = TRUE}
myProb <- 0.95
```

```{r ch010-processBRCA, echo=FALSE, eval=TRUE, results='none', cache=F}
brcaRDS <- paste0( "../data/brca", round(myProb*100), ".rds" )
if ( !file.exists( brcaRDS ) )  {
    ## full workflow simulating RNA-seq data
    allDat <- readRDS( file = "~/Downloads/complete_processed_tcga2stat_RNASeq2_with_clinical.rds" )
    lastClinical <- which( names(allDat) == 'tumorsize' )
    brca <- allDat[ allDat$disease == "BRCA", (lastClinical+1):ncol(allDat) ]
    dim(brca) ## num of genes 20501
    ## compute the median expression! avoid the 0s
    brcaMedian <- apply(brca, 2, median)
    ## retain top (1-myProb)*100% highest expressing genes for illustration
    cutPoint <- quantile( x = brcaMedian, probs = myProb )
    ## genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
    genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
    brca <- brca[ , genesToKeep ]
    ## ncol(brca) / 20501
    ## print(d <- length(genesToKeep))
    ## save data as rds
    saveRDS(brca, brcaRDS)
}
brca <- readRDS(file = brcaRDS)
d <- ncol(brca)
```

---

```{r ch010-realDataTab, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
## Describe and plot real data
if ( !file.exists( '../data/exampleGenes.rds' ) )  {
    set.seed(10192020)
    exampleGenes <- colnames(brca)[sort(sample(ncol(brca), 3))]
    saveRDS(exampleGenes, file = '../data/exampleGenes.rds')
}
exampleGenes <- readRDS('../data/exampleGenes.rds')
small <-  brca %>% select( exampleGenes )
knitr::kable(round(head( small, n = 5)),
      booktabs = TRUE,
      caption = 'mRNA counts for three selected high-expressing genes from the first five observations of the BRCA data set.'
)
```

---

```{r ch010-realDataFig, cache=F, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes. The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations. Modeling these data motivate our simulation methodology."}
# lowerfun <- function(data,mapping){
#   ggplot(data = data, mapping = mapping)+
#     geom_point()+
#     scale_x_continuous(limits = c(-1.5,1.5))+
#     scale_y_continuous(limits = c(-1.5,1.5))
# }

GGally::ggpairs( data = small,
                # lower = list( continuous = wrap(lowerfun)),
                upper = list(
                    continuous = wrap('cor', method = "spearman")
                ), 
                ) + theme_bw()
```

## Desired high-dimensional multivariate simulation properties

Our purposed methodology should possess the following properties  
(adapted from [@Nik13a]):

* BP1: A wide range of dependences, allowing both positive and negative values, and, ideally, admitting the full range of possible values.
* BP2: Flexible dependence, meaning that the number of bivariate marginals can be equal to the number of dependence parameters.
* BP3: Flexible marginal modeling, generating heterogeneous data --- possibly from differing probability families.

Moreover, the simulation method must **scale** to high dimensions:

* SP1: Procedure must scale to high dimensions, computable in a reasonable amount time.
* SP2: Procedure must scale to high dimensions while maintaining accuracy.

## Measures of dependency

In multivariate analysis, an analyst must select a metric to quantify dependency.
The most widely-known is the Pearson (product-moment) correlation coefficient that describes the linear association between two random variables $X$ and $Y$, and, it is given by

\begin{equation}
(\#eq:pearson)
\rho_P(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ var(X)var(Y)\right]^{1/2}}.
\end{equation}

<!-- 
As @MB13 and @MK01 discuss, for a bivariate normal $(X,Y)$ random vector, the Pearson correlation completely describes the dependency between the components. 
For non-normal marginals with monotone correlation patterns, $\rho_P$ suffers some drawbacks and may mislead or fail to capture important relationships (@MK01).
Alternatively in these settings, analysts often prefer rank-based correlation measures to describe the degree of monotonic association.
-->

Two nonparametric, rank-based measures common in practice are Spearman's correlation (denoted $\rho_S$) and Kendall's $\tau$. 
Define

\begin{equation}
(\#eq:spearman)
\rho_S(X,Y) = 3 \left[ P\left[ (X_1 - X_2)(Y_1-Y_3) > 0 \right] - P\left[ (X_1 - X_2)(Y_1-Y_3) < 0 \right] \right],
\end{equation}

\noindent where $(X_1, Y_1) \overset{d}{=} (X,Y), X_2 \overset{d}{=} X, Y_3 \overset{d}{=} Y$ with $X_2$ and $Y_3$ are independent of one other and of $(X_1, Y_1)$. 
Spearman's $\rho_S$ has an appealing correspondence as the Pearson's correlation coefficient on *ranks* of the values, thereby captures nonlinear yet monotone relationships.

Kendall's $\tau$, on the other hand, is the difference in probabilities of concordant and discordant pairs of observations $(X_i, Y_i)$ and $(X_j, Y_j)$.
By concordance we mean that orderings have the same direction (e.g., if $X_i < X_j$, then $Y_i < Y_j$) and is determined by the ranks of the values, not the values themselves.

Both $\tau$ and $\rho_S$ are **invariant to under monotone transformations** of the underlying random variates.
As we will describe more fully in the [Algorithms](#algorithms) section, this property is essential to scaleable match rank-based correlations with speed (SP1) and accuracy (SP2).

## Correspondence among Pearson, Spearman, $\tau$ correlations

This is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficient, as the marginal distributions $F_i$ are intrinsic in their calculation.
But for **bivariate normal vectors**, however, the correspondence is well-known:

\begin{equation}
(\#eq:convertKendall)
\rho_{P} = sin \left( \tau \times \frac{\pi}{2} \right), 
\end{equation}

\noindent and similarly for Spearman's $\rho$ [@K58],

\begin{equation}
(\#eq:convertSpearman)
\rho_P = 2 \times sin \left( \rho_S \times \frac{\pi}{6} \right).
\end{equation}

## Gaussian copulas

There is a strong connection of our simulation strategy to Gaussian **copulas** (see @Nelsen2007 for a technical introduction).
A copula is a distribution function on $[0,1]^d$ that describes a multivariate probability distribution with standard uniform marginals.
This definition provides a powerful, natural way characterize joint probability structure.
Consequently, the study of copulas is an important and active area of statistical theory and practice.

For any random vector ${\bf X}=(X_1, \ldots, X_d)$ with CDF $F$ and marginal CDFs $F_i$ there is a copula function 
$C(u_1, \ldots, u_d)$ so that

\[
F(x_1, \ldots,x_d) = \mathbb P(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), \,\,\, x_i\in \mathbb R, i=1,\ldots,d. 
\]

A Gaussian copula is the case where all marginal CDFs $F_i$ are the standard normal CDF, $\Phi$. 
This representation corresponds to a multivariate normal distribution with standard normal marginal distributions and covariance matrix ${\bf R_P}$.
But since the marginals are standardized to have unit variance, this ${\bf R_P}$ is a Pearson correlation matrix. 
If $F_{{\bf R}}$ is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula $C_{{\bf R}}$ is defined through

\begin{equation}
(\#eq:gauss)
F_{{\bf R}}(x_1, \ldots, x_d) = C_{{\bf R}}(\Phi(x_1), \ldots, \Phi(x_d)),
\end{equation}

where $\Phi(\cdot)$ is the standard normal CDF.
Note that the copula $C_{{\bf R}}$ is the familar multivariate normal CDF of the random vector $(\Phi(X_1), \ldots, \Phi(X_d))$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})$.

Sklar's Theorem [@Sklar1959; @Ubeda-Flores2017] guarantees that given inverse CDFs $F_i^{-1}$s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions.
For example, using Gaussian copulas, we can construct a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)$ with $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$, via ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$ povides $Y_i \sim F_i, \, \forall \, i$ .

## NORmal To Anything algorithm

The well-known NORTA algorithm [@Cario1997] can be used simulate a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$.
Specifically, the NORTA algorithm follows like this:

1. Simulate a random vector $\bf Z$ with $d$ **independent** and **identical** standard normal components.
2. Determine the input matrix $\Sigma_{\bf Z}$ that corresponds with the specified output $\Sigma_{\bf Y}$.
3. Produce a Cholesky factor $M$ of $\Sigma_{\bf Z}$ so that $M M^{\prime}=\Sigma_{\bf Z}$.
4. Set $X$ by $X \gets MZ$.
5. $\text{Return} \; Y \; \text{where} \; Y_i \gets F_{Y_i}^{-1}[\Phi(X_i)], \; i=1,2,...,d$.

## Other `R` packages for simulating random vectors

There are a few R packages for multivariate simulation.

* `copula` Highly flexible copula specification [@Yan2007]
* `nortaRA` Implements exact Pearson matching (step 2 in NORTA) [@Chen2001] 
* `Genord` Simulated correlated discrete variables [@BF17]
* `mvnfast` High-performance multivariate normal simulator [@Fasiolo2016]

# 3 - Methodology/Algorithm

## Random vector generation via `bigsimr::rvec` {#rand-vec-gen}

Now we describe `bigsimr::rvec`, our algorithm to generate random vectors.
It mirrors the classical NORTA algorithm above with some modifications for rank-based dependency matching:

1. Pre-processing for nonparametric dependency matching.  
	(i) Convert from either ${\bf R_{Spearman}}$ or ${\bf R_{Kendall}}$ into the corresponding MVN input correlation ${\bf R_{Pearson}}$.
	(ii) Check that ${\bf R_{Pearson}}$ is semi-positive definite.  
	(iii) If not compute a close semi-positive definite correlation matrix ${\bf \widetilde{R}_{Pearson}}$.  
2. Gaussian copula construction.  
	(i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_{Pearson}})$.  
	(ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
3. Quantile evaluations.  
	(i) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;

The pre-processing (Step 1) takes advantage of the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertSpearman) or \@ref(eq:convertKendall), respectively (implemented as `bigsimr::cor_covert`).

## Algorithm notes

- Discrete margins pose some difficulties and matching *unadjusted* correlation
  measures is not exact (empirically approximate).
- Exact Pearson solutions are possible but require computing $d \choose 2$
   double integrals $EY_iY_j = \int \int y_i y_j f_{X|r}(F_i^{-1}(\Phi(z_i)),
   F_j^{-1}(\Phi(z_j))dy_idy_j$. We have yet to accelerate this process.
- We GPU accelerate whenever possible during our proposed High-Performance NORTA
  algorithm.
- One thorny problem when dealing with high-dimensional correlation matrices is
they can become non-positive definite through either during estimation or
bivariate transformations.
- When converting nonparametric correlations to Pearson (first step above) the
  resultant correlation may not be positive definite (PD).
- In that case, we replace ${\bf R}_{Pearson}$ with a "close" PD matrix ${\bf \tilde{R}}_{Pearson}$.
- In practice, this loss in accuracy typically has little impact on performance,
  but the algorithm needs acceleration (`Matrix::nearPD()` violates property S1).

# 3 The `bigsimr` R package

## Installation and setup

- `bigsimr` is on `Github`: [https://github.com/SchisslerGroup/bigsimr](https://github.com/SchisslerGroup/bigsimr).
- Use `devtools` to install.
- `bigsimr` uses `JAX` python libraries for GPU acceleration and other high
  performance algorithms.

## Basic use illustrated through a minimal example

```{r ch030-basic-example-setup, echo=TRUE, message=FALSE, cache=F}
library(bigsimr)
library(tidyverse)
library(patchwork)
```

For simplicity and to provide a minimal working example, we'll consider bivariate simulation of temperature, in degrees Fahrenheit, and ozone level, in parts per billion.

```{r ch030-air-quality-data, echo=TRUE}
df <- airquality %>%
  select(Temp, Ozone) %>%
  drop_na()
```

```{r ch030-aq-glimpse}
glimpse(df)
```

Figure \@ref(fig:ch030-aq-joint-dist) visualizes the bivariate relationship between Ozone and Temperature.
We aim to simulate random two-component vectors mimicking this structure. 
The margins are not normally distributed, particularly the ozone level exhibits a strong positive skew.

```{r ch030-aq-joint-dist, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Bivariate scatterplot of Ozone vs. Temperature with estimated marginal densities. The data are left skewed tails and appear to be correlation.", cache=F}

p0 <- ggplot(df, aes(Temp, Ozone)) +
  geom_point(size = 1) +
  theme(legend.position = "none") +
  labs(x = "Temperature")

pTemp <- ggplot(df, aes(Temp)) + 
  geom_density() +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())

pOzone <- ggplot(df, aes(Ozone)) + 
    geom_density() +
  theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank()) +
  coord_flip()

pTemp + plot_spacer() + p0 + pOzone + 
  plot_layout(widths = c(3,1), heights = c(1, 3)) 
```

Next, we specify the marginal distributions and correlation coefficient (both its type and magnitude).
Here the analyst is free to be creative.
For this example, we will not take up goodness-of-fit considerations to determine the marginal distributions.
But it seems sensible without domain knowledge to estimate these quantities from the data and `bigsimr` contains fast functions designed for this task.

## Specifying marginal distributions

```{r ch030-aq-temp-pars, echo=TRUE, eval=TRUE}
df %>% 
  select(Temp) %>% 
  summarise_all(.funs = c(mean = mean, sd = sd))
```

```{r ch030-aq-ozone-pars, echo=TRUE}
mle_mean <- function(x) mean(log(x))
mle_sd <- function(x) mean( (log(x) - mean(log(x)))^2 )

df %>% 
  select(Ozone) %>% 
  summarise_all(.funs = c(meanlog = mle_mean, sdlog = mle_sd))
```

```{r ch030-margins-alist, echo=TRUE}
margins <- alist(
    qnorm(mean = 77.871, sd = 9.4855),
    qlnorm(meanlog = 3.419, sdlog = 0.7426),
)
```

```{r ch030-margins-mlist, echo=TRUE}
margins <- mlist(
  qnorm(mean = mean(df$Temp), sd = sd(df$Temp)),
  qlnorm(meanlog = mle_mean(df$Ozone), sdlog = mle_sd(df$Ozone))
)
margins
```

## Specifying correlation

```{r ch030-aq-cor, echo=TRUE}
type <- 'spearman'
(rho <- cor_fast(df, method = "spearman"))
```

## Checking the theoretical correlation bounds

```{r ch030-cor-bounds, echo=TRUE}
cor_bounds(margins = margins, type = type)
```

Since our estimated Spearman correlation $\hat{ \rho}_S$ is within the theoretical bounds, the correlation is valid as input to `bigsimr:rvec`.
By comparison, the Pearson correlation $\rho_P$ is restricted to [-0.8648, 0.8727] for these margins (see `bigsimr::cor_bounds`output below).

```{r ch030-cor-bounds-pearson, echo=TRUE}
cor_bounds(margins = margins, type = 'pearson')
```

## Simulating random vectors

Finally, we arrive at the main function of `bigsimr`, `rvec`.
Let's now simulate $B=10,000$ random vectors from the assumed joint distribution of Ozone levels and Temp.

```{r ch030-sim-margins, echo=TRUE}
x <- rvec(10000, rho, margins, type)
df_sim <- as.data.frame(x)
colnames(df_sim) <- colnames(df)
```

```{r ch030-plot-sim, fig.cap="Contour plot and marginal densities for the simulated bivariate distribution of Air Quality Temperatures and Ozone levels. The simulated points mimic the observed data with respect to both the marginal characteristics and bivariate association."}
p1 <- df_sim %>%
  ggplot(aes(Temp, Ozone)) +
  geom_density_2d_filled() +
  theme(legend.position = "none") +
  labs(x = "Simulated Temperature", y = "Simulated Ozone") +
  scale_y_continuous(limits = c(0, max(df$Ozone))) +
  scale_x_continuous(limits = range(df$Temp))

p1Temp <- ggplot(df_sim, aes(Temp)) + 
  geom_density(alpha = 0.5, fill = "lightseagreen") +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())

p1Ozone <- ggplot(df_sim, aes(Ozone)) + 
  geom_density(alpha = 0.5, fill = "lightseagreen") + 
  theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank()) +
  coord_flip()

p1Temp + plot_spacer() + p1 + p1Ozone + 
  plot_layout(widths = c(3,1), heights = c(1, 3))
```
---

# Monte carlo experiments

## Bivariate experiments

```{r ch040-biNegBin, echo = FALSE, eval = TRUE}
size <- 4
prob <- 3e-04
```

Table: Identical margin, bivariate simulation configurations to evaluate correlation matching accuracy and efficiency.

| Simulation Reps ($B$) | Correlation Types | Identical-margin 2D distribution |
|-------------|:--------------:|----------------------:|
|$1000$ | Pearson ($\rho_P$) | ${ \bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho )$ |
|$10,000$ | Spearman ($\rho_S$) | ${ \bf Y} \sim MVG( shape = 10, rate = 1, \rho )$ |
| $100,000$| Kendall ($\tau$)| ${ \bf Y} \sim MVNB(p = `r prob`, r = 4,\rho)$ |

For each of the unique 9 simulation configurations above, we estimate the correlation bounds and vary $\rho$ along a sequence of 100 points evenly placed within the bounds (minus an adjustment factor of $\epsilon=0.01$ to handle numeric issues when the bound is specified exactly).

## Bivariate result table

```{r ch040-combineBiSims, cache=F}
## compare relative differences in MSE for each dependency, compared to Pearson
allDat <- NULL
dat <- readRDS("../results/biNorm_sims.rds") %>% select( margins, type, N, rho, rho_hat  ) 
allDat <- rbind( allDat, dat )
dat <- readRDS("../results/biGamma_sims.rds") %>% select( margins, type, N, rho, rho_hat  )
allDat <- rbind( allDat, dat )
dat <- readRDS("../results/biNB_sims.rds") %>% select( margins, type, N, rho, rho_hat  )
allDat <- rbind( allDat, dat )
allDat$margins <- factor(allDat$margins, levels = c( 'norm', 'gamma', 'nbinom'),
                         labels = c( 'MVN', 'MVG', 'MVNB' ) )
allDat$type <- factor(allDat$type, levels = c( 'pearson', 'spearman', 'kendall' ),
                      labels = c( 'Pearson', 'Spearman', 'Kendall' ) )
allDat$N <- factor(allDat$N, levels = c( 1000, 10000, 100000) )
```

```{r ch040-biMAEtable, results='asis', cache=F}
## tabMSE  <- allDat %>%
##    group_by( N, margins ) %>%
##    summarize( mse = mean( ( rho - rho_hat )^2 ) )
tabMSE  <- allDat %>%
    ## filter( N == 100000) %>%
    group_by( N, type, margins ) %>%
##     summarize( mse = mean( ( rho - rho_hat )^2 ) ) %>%
    summarize( MAE = mean( abs( rho - rho_hat )  ) )
## normMSE <- as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ] ) 
## ( tabMSE$mse - normMSE ) / normMSE
## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) )
## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) ) / tabMSE$mse

knitr::kable(tabMSE,
             format = 'latex',
             col.names = c("No. of random vectors",
                           "Correlation type",
                           "Distribution",
                           "Mean abs. error"),
             booktabs = TRUE,
             caption = 'Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.',
             linesep = c('', '',  '\\addlinespace')
)
```

## Bivariate results figure

Displayed are the aggregated bivariate simulation results.
```{r ch040-bPlot, cache=F, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.asp = 1.30, fig.weight = 4, fig.align='center', fig.cap = "Bivariate simulations match specified correlations. The horizontal axis plots the specified target correlations across the entire range of possible correlations for each bivariate margin. Normal margins are plotted in dark blue, gamma in medium blue, and negative binomial in light blue. As the number of simulated vectors $B$ increases from left to right, the variation in estimated correlations (vertical axis) decreases. The dashed line indicates equality between the specified and estimated correlations. Only the normal margins match the Pearson correlations (top row) exactly across the entire range of correlations. The two non-normal bivariate random vectors experience attenuation (downward bias) for extreme Pearson correlations (due to limitations in our algorithm) and more restriction (due to the Frechet bounds). The rank-based correlations (bottom two rows) are matched exactly for all margins and obtain the full range of possible dependencies."}
# https://www.datanovia.com/en/blog/how-to-change-ggplot-facet-labels/
# New facet label names
repsLabs <- paste0("B=", c("1,000", "10,000", "100,000") )
names(repsLabs) <- c(1000, 10000, 100000)
typeLabs <- c( 'Pearson', 'Spearman', 'Kendall' )
names(typeLabs) <- c( 'Pearson', 'Spearman', 'Kendall' )

# Set colors
## RColorBrewer::display.brewer.all()
numColors <- 4
numGroups <- length(levels(allDat$margins))
## myColors <- rev( RColorBrewer::brewer.pal(n = numColors, name = 'Greys')[ ((numColors - numGroups) + 1): numColors  ] )
myColors <- rev( RColorBrewer::brewer.pal(n = numColors, name = 'Blues')[ ((numColors - numGroups) + 1): numColors  ] )

allDat %>%
    ## ggplot(aes(x = rho, y = rho_hat, color = margins, shape = margins)) +
    ggplot(aes(x = rho, y = rho_hat, color = margins)) +
    ## ggplot(aes(rho, rho_hat)) +
    ## geom_point(alpha = 1, size = 2) +
    geom_point(size = 2) +
    ## scale_shape_manual(values=c(1, 2, 3))+
    ## scale_color_manual(values=c('#999999','#E69F00', '#56B4E9')) +
    scale_color_manual(values = myColors ) +
    geom_abline(slope = 1, linetype = 'dashed') +
    labs(x = "Specified Correlation", y = "Estimated Correlation") +
    ## facet_wrap(~ + N) + theme_bw()
    facet_wrap(~ type + N, labeller = labeller(N = repsLabs, type = typeLabs)) +
    theme_bw() +
    theme(legend.position = "bottom", legend.direction = "horizontal")

## allDat %>%
##     ggplot(aes(x = rho, y = rho_hat, color = type, shape = type)) +
##     ## ggplot(aes(rho, rho_hat)) +
##     geom_point(alpha = 0.75) +
##     geom_abline(slope = 1) +
##     labs(x = "Specified Correlation", y = "Estimated Correlation") + 
##     ## facet_wrap(~ + N) + theme_bw()
##     facet_wrap(~ margins + N) + theme_bw()

## ggsave('fig/plot-biNormPlot.pdf')
```

## Scale up to High dimensions using RNA-seq data

- We have RNA-sequencing data for N=1212 breast cancer patients (TCGA BRCA).  
- There are 20,501 genes that have count data quantifying the expression of the
  each gene's mRNA.  
- We simulate N random vectors to simulate a *fake dataset* with heterogeneous
  negative binomial margins and correlations estimated from real data.

```{r, echo=F, out.width='80%', fig.align='center', fig.cap='Computation times as d increases. We filter to the top 1, 5, 10, 15, 20, 25% expressing genes (in terms of median expression'}
knitr::include_graphics("../fig/cpu-gpu-times.png")
```

# 4 - Examples

## Simulating High-Dimensional RNA-seq data

```{r ch050-readBRCA, echo=FALSE, eval=TRUE, cache=F}
## full workflow simulating RNA-seq data
brcaRDS <- '../data/brca95.rds'
myProb = 0.95
brca <- readRDS(file = brcaRDS)
d <- ncol(brca)
## smaller set for prototyping
## d=5
## brca <- brca[ , 1:d]
```

First estimate the desired correlation matrix using the fast implementation provided by `bigsimr`:

```{r ch050-estRhoBRCA, echo=TRUE, eval=TRUE, cache=F}
## Estimate Spearman's correlation on the count data
corType <- 'spearman'
system.time( nb_Rho <- bigsimr::cor_fast( brca, method = corType ) )
```
## Prepare to estimate the marginal parameters.

We use method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model.
The marginal distributions are from the same probability family (NB) yet are heterogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.
The functions below help achieve this estimation and specifying the inputs for use in `bigsimr::rvec`.

```{r ch050-nbHelpers, echo=TRUE}
make_nbinom_alist <- function(sizes, probs) {
  lapply(1:length(sizes), function(i) {
    substitute(qnbinom(size = s, prob = p), 
               list(s = sizes[i], p = probs[i]))
  })
}
## make_nbinom_alist(c(20, 21, 22), c(0.3, 0.4, 0.5))
nbinom_mom <- function(x) {
  m <- mean(x)
  s <- sd(x)
  s2 <- s^2
  p <- m/s2
  r <- m^2 / (s2 - m)
  c(r, p)
}
```

## Estimate the marginal parameters

```{r ch050-estMargins, echo = TRUE, eval = TRUE, cache=F}
sizes <- apply( unname(as.matrix(brca)), 2, nbinom_mom )[1, ]
probs <- apply( unname(as.matrix(brca)), 2, nbinom_mom )[2, ]
```

Notably, the marginal NB probabilities $\hat{p}_i's$ are small --- ranging in $[`r min(probs)` , `r max(probs)`]$.
This gives rise to highly variable counts and, typically, less restriction on potential pairwise correlation pairs.
Once the functions are defined/executed to complete marginal estimation, we specify targets and generate the desired random vectors using `rvec`:

## Run `rvec`

```{r ch050-runBRCA-echo, echo=TRUE, eval=FALSE, cache = FALSE}
## Set the number of random vectors
n <- 10000
## construct margins
nb_margins <- make_nbinom_alist(sizes, probs)
## run sims
sim_nbinom <- rvec(n, nb_Rho, nb_margins, type = corType,
                   ensure_PSD = TRUE, cores = cores) 
colnames(sim_nbinom) <- names(brca)
```

```{r ch050-runBRCA, echo=FALSE, results=FALSE, cache = FALSE}
nb_margins <- make_nbinom_alist(sizes, probs)
brcaSimRDS <- paste0( "../results/brca", round(myProb*100), "sim.rds" )
sim_nbinom <- readRDS( brcaSimRDS )
```

## Results for three example genes

Figure \@ref(fig:ch050-simDataFig) displays the simulated counts and pairwise relationships for our example genes in Table \@ref(tab:ch010-realDataTab).
Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parameter form and with less extreme points then the observed data (c.f. Figure \@ref(fig:ch010-realDataFig).)

```{r ch050-simDataFig, echo=F, fig.align='center', fig.cap='Simulated data for three selected high-expressing genes replicates the estimated data structure.'}
knitr::include_graphics('../fig/ch050-simDataFig.png')
```

## Results for all genes

Displayed are the aggregated results of our simulation by comparing the specified target parameter (horizontal axes) with the corresponding quantities estimated from the simulated data (vertical axes).
The evaluation shows that the simulated counts approximately match the target parameters and exhibit the full range of estimated correlation from the data.
Utilizing 15 CPU threads in a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, the simulation completed just over of 2 minutes.

```{r ch050-figBRCA, echo=F, out.width = '80%', fig.align='center', fig.cap="Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set. The dashed red lines indicated equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes)."}
knitr::include_graphics('../fig/ch050-figBRCA.png')
```

## Simulation-based joint probability calculations

To conduct statistical inference a critical task is to evaluate the joint probability mass (or density) function:

$$
P( {\bf Y} = {\bf y} ), y_i \in \chi_i.
$$

So we estimate

$$
\hat{P}( {\bf Y} >= {\bf y_0 } ) = \sum_{b=1}^B I( {\bf Y^{(b) }} > {\bf y_0 } ) / B
$$

where ${\bf Y^{(b)} }$ is the $b^{th}$ simulated vector in a total of $B$ simulation replicates and $I ( )$ is the indicator function.
For example, we can estimate from our $B=10,000$ simulated vectors the probability that all genes are expressed (i.e., ${\bf y}_i \geq 1, \forall \; i$) is $0.1708$.

```{r ch050-densityEvaluation, echo=TRUE, eval=TRUE, cache=F}
d <- ncol(sim_nbinom)
B <- nrow(sim_nbinom)
threshold <- rep( 1, d)
mean(apply( sim_nbinom,  1,
           function(X, y0=threshold) {
               all( X > y0) }
           ))
```

## Evaluation of correlation estimation efficiency

```{r ch010-myM, include=FALSE, cache = TRUE}
m <- 10
```

MC methods are routinely used in many statistical inferential tasks including estimation, hypothesis testing, error rates, and empirical interval coverage rates.
For an concise introduction to these methods, see @Rizzo2007, Ch. 6.
To conclude the example applications, we demonstrate how `bigsimr` can be used evaluate estimation efficiency.
In particular, we'd like to assess the error in our correlation estimation above.
We used a conventional method, based on classical statistical theory. 
Yet this method was not designed for high-dimensional data.
Indeed, high-dimensional covariance estimation (and precision matrices) is an active area of statistical science (see, for example, [@Won2013g; @VanWieringen2016].

In this small example, we simulate $m=`r m`$ data sets with the number of simulated vectors matching the number of patients in the BRCA data set, $N=`r nrow(brca)`$.
Since our simulation is much faster for the Pearson correlation type (see Figure \@ref(fig:ch040-gpuVScpuFig)), we only convert the Spearman correlation matrix once (and ensure its PSD).
At each iteration, we estimate the quadratic loss from the specified ${\bf R}_{Spearman}$, producing a distribution of loss values.

```{r ch050-quadlossECHO, echo=T, eval=F, cache=FALSE}
## Simulate random vectors equal to the sample size
n <- nrow(brca)
## convert outside for faster simulation
nb_Rho_p <- bigsimr::cor_convert( rho = nb_Rho,
                                 from = corType, to = "pearson" )
## ensure PSD
nb_Rho_p <- bigsimr::cor_nearPSD( G = nb_Rho_p )
## create m random vectors and estimate correlation
simRho <- replicate(n = m,
  expr = { tmpSim <- rvec(n = n , nb_Rho_p, nb_margins,
    type = 'pearson', ensure_PSD = FALSE, cores = cores);
    bigsimr::cor_fast( x = tmpSim, method = corType )} ,
simplify = FALSE)
## find quadratic loss at each rep
quadLoss <- unlist( lapply( simRho, rags2ridges::loss,
                           T = nb_Rho, type = "quadratic"))

```

```{r ch050-quadloss, echo=F, eval=T, cache=FALSE}
quadLossRDS <- paste0( "../results/brca", round(myProb*100), "quadLoss.rds" )
quadLoss <- readRDS( quadLossRDS )
```

The `R` summary function supplies the mean-augmented five-number summary of the quadratic loss distribution computed above.

```{r ch050-quadloss-summary, echo=TRUE, eval=T, cache=FALSE}
summary(quadLoss)
```

This distribution could be compared to high-dimensional designed covariance estimators to guide to help decide whether the additional complexity and computation time are warranted.

# 5 - Conclusions

## Concluding remarks

* Our scaleable NORmal To Anything provides a flexible way to simulate
  ultra high-dimensional *fake datasets*.

### Pros
* An easy-to-use tool to produce multivariate data of dimension larger than any
  other existing tool.
* Flexible modeling providing highly hetereogenous specification of joint probability models.
* The R `bigsimr`package is a high-performance and GPU-accelerated implementation
  of algorithm and other fast correlation utilities.

### Cons
* Our method requires a well-defined inverse CDF.
* Discrete margins can slightly disrupt the matching algorithm (due the probability of ties). 
* Not all algorithms are currently optimized.

## Future Work

* Investigate *rescaling* the correlations to adjust for discrete margins.
* GPU accelerate the `normal2marginal()`.
* Better handling of non-positive definiteness (fast checks for PD and faster
  implementation/algorithms of nearest PD).
* High-performance implementations for exact Pearson matching.
* Match *partial correlations*.
* Implement user-defined inverse CDFs.
* Implement high-dimensional covariance estimation algorithms.
* Take advantage of sparsity.
* Explore different simulation algorithms.

<style>
slides > slide { overflow: scroll; }
slides > slide:not(.nobackground):before {
  background: none;
}
</style>

## References

### 2020-12-03 notebooks/cmstats2020_talk.rmd
2nd draft CMStats 2020 talk

@@
 title: "Simulating Ultra High-Dimensional Multivariate Data"
-subtitle: "Using the `bigsimr` R package"
 author:
@@
     smaller: true
-    logo: ../images/hex-bigsimr-60x60.png
+    logo: ../images/hex-bigsimr.png
 bibliography: ../bigsimr.bib
@@
 library(GGally)
+## xaringan::inf_mr('cmstats2020_talk.rmd')
 ```
@@
 
-## Problem we're addressing
+## Big Multivariate Data
 
-- In this era of Big Data, it is critical to realistically simulate data.
-- This is computationally challenging when data are correlated and (ultra-) high dimensional.
-- Many domains are producing these type of data.
-- A prototypical example is RNA-sequencing data derived from high-throughput biomedical experiments.
+- It is critical to realistically simulate data when conducting Monte Carlo studies and methods.
+- But measurements are often **correlated**, **non-normal**, and **high dimensional** in this era of big data.
+- One example are *RNA-seq data* obtained through high-throughput biomedical experiments.
+- 3 out of 20,501 mRNA measurements from 1212 breast cancer patients (TCGA BRCA):
 
-## Our goal
+```{r, echo=F, out.width='60%', fig.align='center'}
+knitr::include_graphics("../fig/realDataFig.png")
+```
+
+## Desired high-dimensional multivariate simulation properties
 
 - Our goal is to simulate $N$ random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with **correlated** components.
-- The simulation replicates $N$ can be large (~100,000).
-- And the dimension $d$ can be very large (~20,000).
-- We propose a scaleable NORmal To Anything approach to this task .
-- We provide a high-performance and GPU-accelerated implementation in the `R` package `bigsimr`.
+
+Our purposed methodology should possess the following properties  
+(adapted from @Nik13a):
+
+* BP1: A wide range of dependences, allowing both positive and negative values, and, ideally, admitting the full range of possible values.
+* BP2: Flexible dependence, meaning that the number of bivariate marginals can be equal to the number of dependence parameters.
+* BP3: Flexible marginal modeling, generating heterogeneous data --- possibly from differing probability families.
+
+Moreover, the simulation method must **scale** to high dimensions:
+
+* SP1: Procedure must scale to high dimensions, computable in a reasonable amount time.
+* SP2: Procedure must scale to high dimensions while maintaining accuracy.
+
+## Our contribution
+
+- We propose a Gaussian-copula random vector algorithm; an augmented NORmal To Anything (NORTA) @Cario1997 method. 
+- We provide a high-performance and GPU-accelerated implementation in the `R` package `bigsimr`, along with related tools.
 
@@
 
-Simulating high-dimensional, non-normal, correlated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
-The RNA-seq data-generating process involves counting how often a particular messenger RNA (mRNA) is expressed in a biological sample.
-RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data.
-For human derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted.
-Importantly, due to inherent biological processes, gene expression data exhibits correlation --- co-expression --- across genes [@BE07; @Schissler2018].
+- Simulating high-dimensional, non-normal, correlated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
+- The RNA-seq data-generating process involves counting how often a particular messenger RNA (mRNA) is expressed in a biological sample.
+- RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data.
+- For human derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted.
+- Importantly, due to inherent biological processes, gene expression data exhibits correlation --- co-expression --- across genes [@BE07; @Schissler2018].
 
@@
 
-GGally::ggpairs( data = small,
+if ( !file.exists( '../fig/realDataFig.png' ) ) {
+    p <- GGally::ggpairs( data = small,
                                         # lower = list( continuous = wrap(lowerfun)),
@@
                              continuous = wrap('cor', method = "spearman")
-                ), 
-                ) + theme_bw()
+                         )) + theme_bw()
+    ggsave('../fig/realDataFig.png', plot = p, dpi = 300)
+}
+knitr::include_graphics("../fig/realDataFig.png")
 ```
-
-## Desired high-dimensional multivariate simulation properties
-
-Our purposed methodology should possess the following properties  
-(adapted from [@Nik13a]):
-
-* BP1: A wide range of dependences, allowing both positive and negative values, and, ideally, admitting the full range of possible values.
-* BP2: Flexible dependence, meaning that the number of bivariate marginals can be equal to the number of dependence parameters.
-* BP3: Flexible marginal modeling, generating heterogeneous data --- possibly from differing probability families.
-
-Moreover, the simulation method must **scale** to high dimensions:
-
-* SP1: Procedure must scale to high dimensions, computable in a reasonable amount time.
-* SP2: Procedure must scale to high dimensions while maintaining accuracy.
 
@@
 
-In multivariate analysis, an analyst must select a metric to quantify dependency.
-The most widely-known is the Pearson (product-moment) correlation coefficient that describes the linear association between two random variables $X$ and $Y$, and, it is given by
+- In multivariate analysis, an analyst must select a metric to quantify dependency.
+- The most widely-known is the Pearson (product-moment) correlation coefficient that describes the linear association between two random variables $X$ and $Y$, and, it is given by
 
-\begin{equation}
-(\#eq:pearson)
+$$
 \rho_P(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ var(X)var(Y)\right]^{1/2}}.
-\end{equation}
+$$
 
@@
 
-Two nonparametric, rank-based measures common in practice are Spearman's correlation (denoted $\rho_S$) and Kendall's $\tau$. 
-Define
+## Rank-based measures of dependency
 
-\begin{equation}
-(\#eq:spearman)
+Two nonparametric, rank-based measures common in practice are **Spearman's correlation** (denoted $\rho_S$) and **Kendall's $\tau$**.
+
+- Spearman's:
+
+$$
 \rho_S(X,Y) = 3 \left[ P\left[ (X_1 - X_2)(Y_1-Y_3) > 0 \right] - P\left[ (X_1 - X_2)(Y_1-Y_3) < 0 \right] \right],
-\end{equation}
+$$
 
 \noindent where $(X_1, Y_1) \overset{d}{=} (X,Y), X_2 \overset{d}{=} X, Y_3 \overset{d}{=} Y$ with $X_2$ and $Y_3$ are independent of one other and of $(X_1, Y_1)$.
-Spearman's $\rho_S$ has an appealing correspondence as the Pearson's correlation coefficient on *ranks* of the values, thereby captures nonlinear yet monotone relationships.
 
-Kendall's $\tau$, on the other hand, is the difference in probabilities of concordant and discordant pairs of observations $(X_i, Y_i)$ and $(X_j, Y_j)$.
-By concordance we mean that orderings have the same direction (e.g., if $X_i < X_j$, then $Y_i < Y_j$) and is determined by the ranks of the values, not the values themselves.
+-  Kendall $\tau$, on the other hand, is the difference in probabilities of concordant and discordant pairs of observations $(X_i, Y_i)$ and $(X_j, Y_j)$.
 
 Both $\tau$ and $\rho_S$ are **invariant to under monotone transformations** of the underlying random variates.
-As we will describe more fully in the [Algorithms](#algorithms) section, this property is essential to scaleable match rank-based correlations with speed (SP1) and accuracy (SP2).
 
-## Correspondence among Pearson, Spearman, $\tau$ correlations
+## Correspondence among Pearson, Spearman, $\tau$
 
-This is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficient, as the marginal distributions $F_i$ are intrinsic in their calculation.
-But for **bivariate normal vectors**, however, the correspondence is well-known:
+- There is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficient, as the marginal distributions $F_i$ are intrinsic in their calculation.
+- But for **bivariate normal vectors**, however, the correspondence is well-known for Kendall's $\tau$:
 
-\begin{equation}
-(\#eq:convertKendall)
+$$
 \rho_{P} = sin \left( \tau \times \frac{\pi}{2} \right), 
-\end{equation}
+$$
 
-\noindent and similarly for Spearman's $\rho$ [@K58],
+- and similarly for Spearman's $\rho$ [@K58],
 
-\begin{equation}
-(\#eq:convertSpearman)
+$$
 \rho_P = 2 \times sin \left( \rho_S \times \frac{\pi}{6} \right).
-\end{equation}
-
-## Gaussian copulas
-
-There is a strong connection of our simulation strategy to Gaussian **copulas** (see @Nelsen2007 for a technical introduction).
-A copula is a distribution function on $[0,1]^d$ that describes a multivariate probability distribution with standard uniform marginals.
-This definition provides a powerful, natural way characterize joint probability structure.
-Consequently, the study of copulas is an important and active area of statistical theory and practice.
-
-For any random vector ${\bf X}=(X_1, \ldots, X_d)$ with CDF $F$ and marginal CDFs $F_i$ there is a copula function 
-$C(u_1, \ldots, u_d)$ so that
-
-\[
-F(x_1, \ldots,x_d) = \mathbb P(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), \,\,\, x_i\in \mathbb R, i=1,\ldots,d. 
-\]
-
-A Gaussian copula is the case where all marginal CDFs $F_i$ are the standard normal CDF, $\Phi$. 
-This representation corresponds to a multivariate normal distribution with standard normal marginal distributions and covariance matrix ${\bf R_P}$.
-But since the marginals are standardized to have unit variance, this ${\bf R_P}$ is a Pearson correlation matrix. 
-If $F_{{\bf R}}$ is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula $C_{{\bf R}}$ is defined through
-
-\begin{equation}
-(\#eq:gauss)
-F_{{\bf R}}(x_1, \ldots, x_d) = C_{{\bf R}}(\Phi(x_1), \ldots, \Phi(x_d)),
-\end{equation}
-
-where $\Phi(\cdot)$ is the standard normal CDF.
-Note that the copula $C_{{\bf R}}$ is the familar multivariate normal CDF of the random vector $(\Phi(X_1), \ldots, \Phi(X_d))$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})$.
-
-Sklar's Theorem [@Sklar1959; @Ubeda-Flores2017] guarantees that given inverse CDFs $F_i^{-1}$s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions.
-For example, using Gaussian copulas, we can construct a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)$ with $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$, via ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$ povides $Y_i \sim F_i, \, \forall \, i$ .
+$$
 
@@
 
-The well-known NORTA algorithm [@Cario1997] can be used simulate a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$.
-Specifically, the NORTA algorithm follows like this:
+- The well-known NORTA algorithm [@Cario1997] can be used simulate a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$.
+- Specifically, the NORTA algorithm follows like this:
 
@@
 
+- Exact Pearson solutions are possible but require computing $d \choose 2$ double integrals $EY_iY_j = \int \int y_i y_j f_{X|r}(F_i^{-1}(\Phi(z_i)), F_j^{-1}(\Phi(z_j))dy_idy_j$.
+- But even when closed form solutions matching exactly can be impossible (*NORTA-defective* correlation matrices) and the probability of this increases with $d$ @GH02.
+
 ## Other `R` packages for simulating random vectors
 
-There are a few R packages for multivariate simulation.
+Here are some other  R packages for multivariate simulation.
 
@@
 ## Random vector generation via `bigsimr::rvec` {#rand-vec-gen}
-
-Now we describe `bigsimr::rvec`, our algorithm to generate random vectors.
-It mirrors the classical NORTA algorithm above with some modifications for rank-based dependency matching:
 
@@
 
-The pre-processing (Step 1) takes advantage of the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertSpearman) or \@ref(eq:convertKendall), respectively (implemented as `bigsimr::cor_covert`).
+# 4 - Monte carlo experiments and examples
 
-## Algorithm notes
-
-- Discrete margins pose some difficulties and matching *unadjusted* correlation
-  measures is not exact (empirically approximate).
-- Exact Pearson solutions are possible but require computing $d \choose 2$
-   double integrals $EY_iY_j = \int \int y_i y_j f_{X|r}(F_i^{-1}(\Phi(z_i)),
-   F_j^{-1}(\Phi(z_j))dy_idy_j$. We have yet to accelerate this process.
-- We GPU accelerate whenever possible during our proposed High-Performance NORTA
-  algorithm.
-- One thorny problem when dealing with high-dimensional correlation matrices is
-they can become non-positive definite through either during estimation or
-bivariate transformations.
-- When converting nonparametric correlations to Pearson (first step above) the
-  resultant correlation may not be positive definite (PD).
-- In that case, we replace ${\bf R}_{Pearson}$ with a "close" PD matrix ${\bf \tilde{R}}_{Pearson}$.
-- In practice, this loss in accuracy typically has little impact on performance,
-  but the algorithm needs acceleration (`Matrix::nearPD()` violates property S1).
-
-# 3 The `bigsimr` R package
-
-## Installation and setup
-
-- `bigsimr` is on `Github`: [https://github.com/SchisslerGroup/bigsimr](https://github.com/SchisslerGroup/bigsimr).
-- Use `devtools` to install.
-- `bigsimr` uses `JAX` python libraries for GPU acceleration and other high
-  performance algorithms.
-
-## Basic use illustrated through a minimal example
-
-```{r ch030-basic-example-setup, echo=TRUE, message=FALSE, cache=F}
-library(bigsimr)
-library(tidyverse)
-library(patchwork)
-```
-
-For simplicity and to provide a minimal working example, we'll consider bivariate simulation of temperature, in degrees Fahrenheit, and ozone level, in parts per billion.
-
-```{r ch030-air-quality-data, echo=TRUE}
-df <- airquality %>%
-  select(Temp, Ozone) %>%
-  drop_na()
-```
-
-```{r ch030-aq-glimpse}
-glimpse(df)
-```
-
-Figure \@ref(fig:ch030-aq-joint-dist) visualizes the bivariate relationship between Ozone and Temperature.
-We aim to simulate random two-component vectors mimicking this structure. 
-The margins are not normally distributed, particularly the ozone level exhibits a strong positive skew.
-
-```{r ch030-aq-joint-dist, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Bivariate scatterplot of Ozone vs.
-Temperature with estimated marginal densities.
-The data are left skewed tails and appear to be correlation.", cache=F}
-
-p0 <- ggplot(df, aes(Temp, Ozone)) +
-  geom_point(size = 1) +
-  theme(legend.position = "none") +
-  labs(x = "Temperature")
-
-pTemp <- ggplot(df, aes(Temp)) + 
-  geom_density() +
-  theme(axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        axis.text = element_blank(),
-        axis.ticks = element_blank())
-
-pOzone <- ggplot(df, aes(Ozone)) + 
-    geom_density() +
-  theme(axis.title.y = element_blank(),
-        axis.title.x = element_blank(),
-        axis.ticks = element_blank(),
-        axis.text = element_blank()) +
-  coord_flip()
-
-pTemp + plot_spacer() + p0 + pOzone + 
-  plot_layout(widths = c(3,1), heights = c(1, 3)) 
-```
-
-Next, we specify the marginal distributions and correlation coefficient (both its type and magnitude).
-Here the analyst is free to be creative.
-For this example, we will not take up goodness-of-fit considerations to determine the marginal distributions.
-But it seems sensible without domain knowledge to estimate these quantities from the data and `bigsimr` contains fast functions designed for this task.
-
-## Specifying marginal distributions
-
-```{r ch030-aq-temp-pars, echo=TRUE, eval=TRUE}
-df %>% 
-  select(Temp) %>% 
-  summarise_all(.funs = c(mean = mean, sd = sd))
-```
-
-```{r ch030-aq-ozone-pars, echo=TRUE}
-mle_mean <- function(x) mean(log(x))
-mle_sd <- function(x) mean( (log(x) - mean(log(x)))^2 )
-
-df %>% 
-  select(Ozone) %>% 
-  summarise_all(.funs = c(meanlog = mle_mean, sdlog = mle_sd))
-```
-
-```{r ch030-margins-alist, echo=TRUE}
-margins <- alist(
-    qnorm(mean = 77.871, sd = 9.4855),
-    qlnorm(meanlog = 3.419, sdlog = 0.7426),
-)
-```
-
-```{r ch030-margins-mlist, echo=TRUE}
-margins <- mlist(
-  qnorm(mean = mean(df$Temp), sd = sd(df$Temp)),
-  qlnorm(meanlog = mle_mean(df$Ozone), sdlog = mle_sd(df$Ozone))
-)
-margins
-```
-
-## Specifying correlation
-
-```{r ch030-aq-cor, echo=TRUE}
-type <- 'spearman'
-(rho <- cor_fast(df, method = "spearman"))
-```
-
-## Checking the theoretical correlation bounds
-
-```{r ch030-cor-bounds, echo=TRUE}
-cor_bounds(margins = margins, type = type)
-```
-
-Since our estimated Spearman correlation $\hat{ \rho}_S$ is within the theoretical bounds, the correlation is valid as input to `bigsimr:rvec`.
-By comparison, the Pearson correlation $\rho_P$ is restricted to [-0.8648, 0.8727] for these margins (see `bigsimr::cor_bounds`output below).
-
-```{r ch030-cor-bounds-pearson, echo=TRUE}
-cor_bounds(margins = margins, type = 'pearson')
-```
-
-## Simulating random vectors
-
-Finally, we arrive at the main function of `bigsimr`, `rvec`.
-Let's now simulate $B=10,000$ random vectors from the assumed joint distribution of Ozone levels and Temp.
-
-```{r ch030-sim-margins, echo=TRUE}
-x <- rvec(10000, rho, margins, type)
-df_sim <- as.data.frame(x)
-colnames(df_sim) <- colnames(df)
-```
-
-```{r ch030-plot-sim, fig.cap="Contour plot and marginal densities for the simulated bivariate distribution of Air Quality Temperatures and Ozone levels.
-The simulated points mimic the observed data with respect to both the marginal characteristics and bivariate association."}
-p1 <- df_sim %>%
-  ggplot(aes(Temp, Ozone)) +
-  geom_density_2d_filled() +
-  theme(legend.position = "none") +
-  labs(x = "Simulated Temperature", y = "Simulated Ozone") +
-  scale_y_continuous(limits = c(0, max(df$Ozone))) +
-  scale_x_continuous(limits = range(df$Temp))
-
-p1Temp <- ggplot(df_sim, aes(Temp)) + 
-  geom_density(alpha = 0.5, fill = "lightseagreen") +
-  theme(axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        axis.text = element_blank(),
-        axis.ticks = element_blank())
-
-p1Ozone <- ggplot(df_sim, aes(Ozone)) + 
-  geom_density(alpha = 0.5, fill = "lightseagreen") + 
-  theme(axis.title.y = element_blank(),
-        axis.title.x = element_blank(),
-        axis.ticks = element_blank(),
-        axis.text = element_blank()) +
-  coord_flip()
-
-p1Temp + plot_spacer() + p1 + p1Ozone + 
-  plot_layout(widths = c(3,1), heights = c(1, 3))
-```
----
-
-# Monte carlo experiments
-
-## Bivariate experiments
+## 4A - Bivariate studies
 
@@
 
-Table: Identical margin, bivariate simulation configurations to evaluate correlation matching accuracy and efficiency.
+- Identical margin, bivariate simulation configurations to provide a low-dimensional evaluation:
 
@@
 
-For each of the unique 9 simulation configurations above, we estimate the correlation bounds and vary $\rho$ along a sequence of 100 points evenly placed within the bounds (minus an adjustment factor of $\epsilon=0.01$ to handle numeric issues when the bound is specified exactly).
+- For each of the unique 9 simulation configurations above, we estimate the correlation bounds and vary $\rho$ along a sequence of 100 points evenly placed within the bounds.
 
-## Bivariate result table
+---
 
-```{r ch040-combineBiSims, cache=F}
-## compare relative differences in MSE for each dependency, compared to Pearson
-allDat <- NULL
-dat <- readRDS("../results/biNorm_sims.rds") %>% select( margins, type, N, rho, rho_hat  ) 
-allDat <- rbind( allDat, dat )
-dat <- readRDS("../results/biGamma_sims.rds") %>% select( margins, type, N, rho, rho_hat  )
-allDat <- rbind( allDat, dat )
-dat <- readRDS("../results/biNB_sims.rds") %>% select( margins, type, N, rho, rho_hat  )
-allDat <- rbind( allDat, dat )
-allDat$margins <- factor(allDat$margins, levels = c( 'norm', 'gamma', 'nbinom'),
-                         labels = c( 'MVN', 'MVG', 'MVNB' ) )
-allDat$type <- factor(allDat$type, levels = c( 'pearson', 'spearman', 'kendall' ),
-                      labels = c( 'Pearson', 'Spearman', 'Kendall' ) )
-allDat$N <- factor(allDat$N, levels = c( 1000, 10000, 100000) )
-```
-
-```{r ch040-biMAEtable, results='asis', cache=F}
-## tabMSE  <- allDat %>%
-##    group_by( N, margins ) %>%
-##    summarize( mse = mean( ( rho - rho_hat )^2 ) )
-tabMSE  <- allDat %>%
-    ## filter( N == 100000) %>%
-    group_by( N, type, margins ) %>%
-##     summarize( mse = mean( ( rho - rho_hat )^2 ) ) %>%
-    summarize( MAE = mean( abs( rho - rho_hat )  ) )
-## normMSE <- as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ] ) 
-## ( tabMSE$mse - normMSE ) / normMSE
-## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) )
-## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) ) / tabMSE$mse
-
-knitr::kable(tabMSE,
-             format = 'latex',
-             col.names = c("No. of random vectors",
-                           "Correlation type",
-                           "Distribution",
-                           "Mean abs. error"),
-             booktabs = TRUE,
-             caption = 'Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.',
-             linesep = c('', '',  '\\addlinespace')
-)
-```
-
-## Bivariate results figure
-
-Displayed are the aggregated bivariate simulation results.
-```{r ch040-bPlot, cache=F, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.asp = 1.30, fig.weight = 4, fig.align='center', fig.cap = "Bivariate simulations match specified correlations.
-The horizontal axis plots the specified target correlations across the entire range of possible correlations for each bivariate margin.
-Normal margins are plotted in dark blue, gamma in medium blue, and negative binomial in light blue.
-As the number of simulated vectors $B$ increases from left to right, the variation in estimated correlations (vertical axis) decreases.
-The dashed line indicates equality between the specified and estimated correlations.
-Only the normal margins match the Pearson correlations (top row) exactly across the entire range of correlations.
-The two non-normal bivariate random vectors experience attenuation (downward bias) for extreme Pearson correlations (due to limitations in our algorithm) and more restriction (due to the Frechet bounds).
-The rank-based correlations (bottom two rows) are matched exactly for all margins and obtain the full range of possible dependencies."}
+```{r ch040-bPlot, cache=F, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.width = 7, fig.height = 6, fig.align='center'}
 # https://www.datanovia.com/en/blog/how-to-change-ggplot-facet-labels/
@@
 
-## Scale up to High dimensions using RNA-seq data
+## 4A - Average absolute error in bivariate studies
 
-- We have RNA-sequencing data for N=1212 breast cancer patients (TCGA BRCA).  
-- There are 20,501 genes that have count data quantifying the expression of the
-  each gene's mRNA.  
-- We simulate N random vectors to simulate a *fake dataset* with heterogeneous
-  negative binomial margins and correlations estimated from real data.
-
-```{r, echo=F, out.width='80%', fig.align='center', fig.cap='Computation times as d increases. We filter to the top 1, 5, 10, 15, 20, 25% expressing genes (in terms of median expression'}
-knitr::include_graphics("../fig/cpu-gpu-times.png")
+```{r ch040-combineBiSims, cache=F, echo=F}
+## compare relative differences in MSE for each dependency, compared to Pearson
+allDat <- NULL
+dat <- readRDS("../results/biNorm_sims.rds") %>% select( margins, type, N, rho, rho_hat  ) 
+allDat <- rbind( allDat, dat )
+dat <- readRDS("../results/biGamma_sims.rds") %>% select( margins, type, N, rho, rho_hat  )
+allDat <- rbind( allDat, dat )
+dat <- readRDS("../results/biNB_sims.rds") %>% select( margins, type, N, rho, rho_hat  )
+allDat <- rbind( allDat, dat )
+allDat$margins <- factor(allDat$margins, levels = c( 'norm', 'gamma', 'nbinom'),
+                         labels = c( 'MVN', 'MVG', 'MVNB' ) )
+allDat$type <- factor(allDat$type, levels = c( 'pearson', 'spearman', 'kendall' ),
+                      labels = c( 'Pearson', 'Spearman', 'Kendall' ) )
+allDat$N <- factor(allDat$N, levels = c( 1000, 10000, 100000) )
 ```
 
-# 4 - Examples
+```{r ch040-biMAEtable, cache=F, echo=F, warning=F, message=F}
+## tabMSE  <- allDat %>%
+##    group_by( N, margins ) %>%
+##    summarize( mse = mean( ( rho - rho_hat )^2 ) )
+tabMSE  <- allDat %>%
+    filter( N == 100000) %>%
+    group_by( N, type, margins ) %>%
+##     summarize( mse = mean( ( rho - rho_hat )^2 ) ) %>%
+    summarize( MAE = mean( abs( rho - rho_hat )  ) )
+## normMSE <- as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ] ) 
+## ( tabMSE$mse - normMSE ) / normMSE
+## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) )
+## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) ) / tabMSE$mse
 
-## Simulating High-Dimensional RNA-seq data
+knitr::kable(tabMSE,
+             col.names = c("No. of random vectors",
+                           "Correlation type",
+                           "Distribution",
+                           "Mean abs. error"),
+             caption = '',
+)
+```
+
+## 4B - Simulating High-Dimensional RNA-seq data
 
@@
 ```
-## Prepare to estimate the marginal parameters.
+## 4B - Estimate the marginal parameters
 
-We use method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model.
-The marginal distributions are from the same probability family (NB) yet are heterogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.
-The functions below help achieve this estimation and specifying the inputs for use in `bigsimr::rvec`.
+- We use method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model.
+- The marginal distributions are from the same probability family (NB) yet are heterogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.
 
@@
 
-## Estimate the marginal parameters
+## 4B - Estimate the marginal parameters
 
@@
 
-Notably, the marginal NB probabilities $\hat{p}_i's$ are small --- ranging in $[`r min(probs)` , `r max(probs)`]$.
-This gives rise to highly variable counts and, typically, less restriction on potential pairwise correlation pairs.
-Once the functions are defined/executed to complete marginal estimation, we specify targets and generate the desired random vectors using `rvec`:
+- Notably, the marginal NB probabilities $\hat{p}_i's$ are small --- ranging in $[`r min(probs)` , `r max(probs)`]$.
+- This gives rise to highly variable counts and, typically, less restriction on potential pairwise correlation pairs.
+- Once the functions are defined/executed to complete marginal estimation, we specify targets and generate the desired random vectors using `rvec`.
 
-## Run `rvec`
+## 4B - Simulate RNA-seq data via `bigsimr::rvec`
 
@@
                    ensure_PSD = TRUE, cores = cores) 
-colnames(sim_nbinom) <- names(brca)
 ```
@@
 
-## Results for three example genes
+## 4B -  Results for the three example genes
 
-Figure \@ref(fig:ch050-simDataFig) displays the simulated counts and pairwise relationships for our example genes in Table \@ref(tab:ch010-realDataTab).
-Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parameter form and with less extreme points then the observed data (c.f.
-Figure \@ref(fig:ch010-realDataFig).)
-
-```{r ch050-simDataFig, echo=F, fig.align='center', fig.cap='Simulated data for three selected high-expressing genes replicates the estimated data structure.'}
+```{r ch050-simDataFig, echo=F, fig.align='center', fig.cap='Simulated data for three selected high-expressing genes replicates the estimated data structure.', fig.show = 'hold', out.width='50%'}
+knitr::include_graphics("../fig/realDataFig.png")
 knitr::include_graphics('../fig/ch050-simDataFig.png')
@@
 
-## Results for all genes
+## 4B - Results for all genes
 
+<!-- 
 Displayed are the aggregated results of our simulation by comparing the specified target parameter (horizontal axes) with the corresponding quantities estimated from the simulated data (vertical axes).
@@
 Utilizing 15 CPU threads in a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, the simulation completed just over of 2 minutes.
+-->
 
-```{r ch050-figBRCA, echo=F, out.width = '80%', fig.align='center', fig.cap="Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set.
-The dashed red lines indicated equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes)."}
+```{r ch050-figBRCA, echo=F, fig.align='center', out.width='60%'}
 knitr::include_graphics('../fig/ch050-figBRCA.png')
@@
 
-## Simulation-based joint probability calculations
+## 4C - Simulation-based joint probability calculations
 
-To conduct statistical inference a critical task is to evaluate the joint probability mass (or density) function:
+- To conduct statistical inference a critical task is to evaluate the joint probability mass (or density) function:
 
@@
 
-So we estimate
+- So we may estimate via:
 
@@
 
-where ${\bf Y^{(b)} }$ is the $b^{th}$ simulated vector in a total of $B$ simulation replicates and $I ( )$ is the indicator function.
-For example, we can estimate from our $B=10,000$ simulated vectors the probability that all genes are expressed (i.e., ${\bf y}_i \geq 1, \forall \; i$) is $0.1708$.
+- where ${\bf Y^{(b)} }$ is the $b^{th}$ simulated vector in a total of $B$ simulation replicates and $I ( )$ is the indicator function.
+
+## 4C - Simulation-based joint probability calculations
+
+- For example, we can estimate from our  simulated vectors the probability that all $`r d`$ genes are expressed (i.e., ${\bf y}_i \geq 1, \forall \; i$) :
 
@@
 
-## Evaluation of correlation estimation efficiency
+## 4D - Assessing estimation efficiency
 
-```{r ch010-myM, include=FALSE, cache = TRUE}
-m <- 10
-```
-
-MC methods are routinely used in many statistical inferential tasks including estimation, hypothesis testing, error rates, and empirical interval coverage rates.
-For an concise introduction to these methods, see @Rizzo2007, Ch. 6.
-To conclude the example applications, we demonstrate how `bigsimr` can be used evaluate estimation efficiency.
-In particular, we'd like to assess the error in our correlation estimation above.
-We used a conventional method, based on classical statistical theory. 
-Yet this method was not designed for high-dimensional data.
-Indeed, high-dimensional covariance estimation (and precision matrices) is an active area of statistical science (see, for example, [@Won2013g; @VanWieringen2016].
-
-In this small example, we simulate $m=`r m`$ data sets with the number of simulated vectors matching the number of patients in the BRCA data set, $N=`r nrow(brca)`$.
-Since our simulation is much faster for the Pearson correlation type (see Figure \@ref(fig:ch040-gpuVScpuFig)), we only convert the Spearman correlation matrix once (and ensure its PSD).
-At each iteration, we estimate the quadratic loss from the specified ${\bf R}_{Spearman}$, producing a distribution of loss values.
+-  To assess the error in our correlation estimation above:
 
 ```{r ch050-quadlossECHO, echo=T, eval=F, cache=FALSE}
+## Number of experiments
+m <- 10
 ## Simulate random vectors equal to the sample size
@@
 
+## 4D - Assessing estimation efficiency
+
 The `R` summary function supplies the mean-augmented five-number summary of the quadratic loss distribution computed above.
@@
 
-This distribution could be compared to high-dimensional designed covariance estimators to guide to help decide whether the additional complexity and computation time are warranted.
+- This distribution could be compared to high-dimensional designed covariance estimators to guide to help decide whether the additional complexity and computation time are warranted.
+
+## 4E - Scale up to high dimensional RNA-seq data
+
+- Lastly, we provide computation times at increasing $d$ to assess scaling feasibility:
+
+```{r, echo=F, out.width='80%', fig.align='center'}
+knitr::include_graphics("../fig/cpu-gpu-times.png")
+```
 
@@
 
-* Our scaleable NORmal To Anything provides a flexible way to simulate
-  ultra high-dimensional *fake datasets*.
+- Our scaleable NORmal To Anything provides a flexible way to simulate high-dimensional *fake datasets*.
+- `bigsimr` is on `Github`: [https://github.com/SchisslerGroup/bigsimr](https://github.com/SchisslerGroup/bigsimr).
+- Use `devtools` to install.
+- `bigsimr` uses `JAX` python libraries for GPU acceleration and other high
+  performance algorithms.
 
 ### Pros
+
 * An easy-to-use tool to produce multivariate data of dimension larger than any
@@
 
+## Acknowledgements
+
+- Ed
+- Walt
+- Heather
+- Eric Li
+- Fred Harris
+- CP3 grant
+- Co-aothers
+
 <style>

### 2020-12-03 notebooks/cmstats2020_talk.rmd
3rd draft CMStats 2020 talk

@@
 ---
-title: "Simulating Ultra High-Dimensional Multivariate Data"
+subtitle: "Simulating Ultra High-Dimensional Multivariate Data"
 author:
-  - Alfred G. Schissler
-  - Alexander D. Knudson
-  - Tomasz J. Kozubowski
-  - Anna K. Panorska
-  - Juli Petereit
+  - "Alfred G. **Schissler**, PhD"
+  - "Alexander D. **Knudson**, MS"
+  - "Tomasz J. **Kozubowski**, PhD"
+  - "Anna K. **Panorska**, PhD"
+  - "Juli **Petereit**, PhD"
+  - "*University of Nevada, Reno (UNR)*"
 institute: |
@@
     "University of Nevada, Reno"
-date: "21 Dec 2020 (updated: `r Sys.Date()`)"
+date: "21 Dec 2020 (*updated: `r Sys.Date()`)*"
 output:
@@
 bibliography: ../bigsimr.bib
-abstract: "It is critical to realistically simulate data when conducting Monte Carlo studies and methods.
-But measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments.
-Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers may resort to simulation designs that posit independence or perform arbitrary data transformations.
-This greatly diminishes insights into the empirical operating characteristics of any proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness.
-This article introduces the `bigsimr` R package that provides a flexible, scaleable procedure to simulate high-dimensional random vectors with given marginal characteristics and dependency measures.
-We'll describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and find close positive semi-definite matrices.
-Finally, we demonstrate the power of `bigsimr` by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d =1026$"
 ---
@@
 
-Our purposed methodology should possess the following properties  
+Our proposed methodology should possess the following properties  
 (adapted from @Nik13a):
@@
 - We provide a high-performance and GPU-accelerated implementation in the `R` package `bigsimr`, along with related tools.
+- An article motivating, describing, and applying the methods; archived [here](???).
 
@@
 
-```{r ch010-realDataFig, cache=F, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes.
+```{r ch010-realDataFig, cache=F, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, out.width='60%', fig.align='center', fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes.
 The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations.
@@
 
-```{r ch050-simDataFig, echo=F, fig.align='center', fig.cap='Simulated data for three selected high-expressing genes replicates the estimated data structure.', fig.show = 'hold', out.width='50%'}
-knitr::include_graphics("../fig/realDataFig.png")
+```{r ch050-simDataFig, echo=F, fig.align='center', fig.cap='Simulated data for three selected high-expressing genes replicates the estimated data structure.', fig.show = 'hold', out.width='40%', fig.asp=1}
+## knitr::include_graphics("../fig/realDataFig.png")
 knitr::include_graphics('../fig/ch050-simDataFig.png')
@@
 
-## Concluding remarks
+## 5 - Concluding remarks
 
-- Our scaleable NORmal To Anything provides a flexible way to simulate high-dimensional *fake datasets*.
+- Our scaleable simulation algorithm provides a flexible way to simulate high-dimensional *fake datasets*.
 - `bigsimr` is on `Github`: [https://github.com/SchisslerGroup/bigsimr](https://github.com/SchisslerGroup/bigsimr).
@@
 ### Cons
-* Our method requires a well-defined inverse CDF.
 * Discrete margins can slightly disrupt the matching algorithm (due the probability of ties). 
-* Not all algorithms are currently optimized.
+* (Nearly) exact Pearson matching is not supported.
 
@@
 * Investigate *rescaling* the correlations to adjust for discrete margins.
-* GPU accelerate the `normal2marginal()`.
-* Better handling of non-positive definiteness (fast checks for PD and faster
-  implementation/algorithms of nearest PD).
-* High-performance implementations for exact Pearson matching.
-* Match *partial correlations*.
-* Implement user-defined inverse CDFs.
+* High-performance implementations, algorithms for Pearson matching.
 * Implement high-dimensional covariance estimation algorithms.
-* Take advantage of sparsity.
-* Explore different simulation algorithms.
 
-## Acknowledgements
+## Gratefully acknowledge
 
-- Ed
-- Walt
-- Heather
-- Eric Li
-- Fred Harris
-- CP3 grant
-- Co-aothers
+- Walter W. Piegorsch, PhD
+- Edward J. Bedrick, PhD
+- Yves A. Lussier, MD
+- Heather Knudson
+- Eric Li, MS
+- Fred Harris, PhD
+
+### Funding and grants
+
+- Research reported in this publication was supported by MW-CTR-IN of the National Institutes of Health under award number NIH 1U54GM104944.
+- Research & Innovation, UNR
+
+### CFE-CMSTATS 2020 organizers
+
+ - Thanks to everyone who makes this event possible!
+ - Professor Anna Panorska for the invitation to speak today.
+
+</div>
 

### 2020-12-11 030-bigsimr-package.Rmd
4th draft. fix mle of log normal var

@@
 mle_mean <- function(x) mean(log(x))
-mle_sd <- function(x) mean( (log(x) - mean(log(x)))^2 )
+mle_sd <- function(x) mean( sqrt( (log(x) - mean(log(x)))^2 ) )
 
@@
     qnorm(mean = 77.871, sd = 9.4855),
-    qlnorm(meanlog = 3.419, sdlog = 0.7426),
+    qlnorm(meanlog = 3.419, sdlog = 0.6967),
 )
@@
 ```{r ch030-cor-bounds-pearson, echo=TRUE}
-cor_bounds(margins = margins, type = 'pearson')
+cor_bounds(margins = margins, type = 'pearson', reps=1e6)
 ```

### 2020-12-11 index.Rmd
Add new co-authors

@@
   - Alfred G. Schissler
+  - Edward J. Bedrick
   - Alexander D. Knudson
   - Tomasz J. Kozubowski
+  - Tin Nguyen
   - Anna K. Panorska
   - Juli Petereit
+  - Walter W. Piegorsch
+  - Duc Tran
 bibliography: ["bigsimr.bib", "packages.bib"]

### 2020-12-11 notebooks/cor_boundsBug.R
Add cor_boundsBug notebook

@@ -1,0 +1,12 @@
+library(bigsimr)
+size <- 4
+prob <- 3e-04
+type <- "spearman"
+n <- 1e3
+
+margins <- alist(
+        qnbinom(size = size, prob = prob),
+        qnbinom(size = size, prob = prob)
+)
+
+tmp_bounds <- cor_bounds(margins, type = type)

### 2020-12-15 pdf/article_bigsimr.tex
4th draft. Check length at double spacing

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\usepackage{hyperref}
\hypersetup{
            pdftitle={Simulating High-Dimensional Multivariate Data},
            pdfauthor={Alfred G. Schissler; Edward J. Bedrick; Alexander D. Knudson; Tomasz J. Kozubowski; Tin Nguyen; Anna K. Panorska; Juli Petereit; Walter W. Piegorsch; Duc Tran},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{longtable,booktabs}
% Fix footnotes in tables (requires footnote package)
\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{longtable}}{}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

% Place anything extra you'd like in the preamble
\usepackage{setspace}
\doublespacing
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\usepackage[]{natbib}
\bibliographystyle{plainnat}

\title{Simulating High-Dimensional Multivariate Data}
\providecommand{\subtitle}[1]{}
\subtitle{using the bigsimr R Package}
\author{Alfred G. Schissler \and Edward J. Bedrick \and Alexander D. Knudson \and Tomasz J. Kozubowski \and Tin Nguyen \and Anna K. Panorska \and Juli Petereit \and Walter W. Piegorsch \and Duc Tran}
\date{}

\begin{document}
\maketitle
\begin{abstract}
It is critical to realistically simulate data when conducting Monte Carlo studies and methods. But measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers may resort to simulation designs that posit independence or perform arbitrary data transformations. This greatly diminishes insights into the empirical operating characteristics of any proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness. This article introduces the \texttt{bigsimr} R package that provides a flexible, scaleable procedure to simulate high-dimensional random vectors with given marginal characteristics and dependency measures. We'll describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and find close positive semi-definite matrices. Finally, we demonstrate the power of \texttt{bigsimr} by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size \(n=1212\) patients and dimension \(d =1026\).
\end{abstract}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Massive high-dimensional data sets are now commonplace in many areas of scientific inquiry.
As new methods are developed for these data, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness --- often in comparison to existing methods.
Further, efficient simulation empowers statistical computing strategies, such as the parametric bootstrap \citep{Rizzo2007} to simulate from a hypothesized null model, providing inference in analytically challenging settings.
Such Monte Carlo (MC) techniques become difficult for high-dimensional data with the current existing algorithms and tools.
This is particularly true when simulating massive \emph{multivariate}, \emph{non-normal} distributions, arising naturally in many fields of study.

As others have noted, it can be vexing to simulate dependent, non-normal/discrete data --- even for low dimensional settings \citep{MB13, XZ19}.
For continuous non--normal multivariate data, the well-known NORmal To Anything (NORTA) algorithm \citep{Cario1997} and other copula \citep{Nelsen2007} approaches are well-studied with flexible, robust software available \citep{Yan2007, Chen2001}.
Yet these approaches do not scale in a timely fashion to high-dimensional problems \citep{Li2019gpu}.
For discrete data, early simulation strategies had major flaws --- such as failing to obtain the full range of possible dependencies (e.g., admitting only positive correlations \citet{Park1996}).
While more recent approaches \citep{MB13, Xia17, BF17} have largely remedied this issue for low-dimensional problems, the existing tools are not designed to scale to high dimensions.

Another central issue lies characterizing dependency between components in the high-dimensional random vector.
The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g.~non-normal, discrete, infinite support, etc).
For normal data, the Pearson product-moment correlation describes the dependency perfectly.
As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix exactly is computationally intense \citep{Chen2001, Xia17}.
On the other hand, an analyst may consider the use of nonparametric correlation measures to better characterize monotone, non-linear dependency, such as Spearman's \(\rho\) and Kendall's \(\tau\).
Throughout, we'll emphasize matching these nonparametric dependency measures as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology.

With all this mind, we present a scaleable, flexible multivariate simulation algorithm.
The crux of the method lies in the construction of a Gaussian copula, in the spirit of the NORTA procedure.
Further, we introduce the \texttt{bigsimr} R package that provides parallelized, high-performance software implemented our NORTA-inspired algorithm.
The algorithm design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed form relationships between dependency measures for the multivariate normal (MVN) distribution.

The study proceeds by providing background information, including a description of our motivating example application --- RNA-sequencing (RNA-seq) breast cancer data.
Then we describe and justify our simulation methodology and related algorithms.
Next, we detail an illustrative low-dimensional example of basic and advanced use of the \texttt{bigsimr} R package.
Then we proceed with Monte Carlo studies under various distributional assumptions to assess accuracy and scaleable.
After the MC evaluations, we revisit our high-dimensional motivating example and employ our methods to commonplace statistical computing tasks.
Finally, we'll discuss the method's utility, limitations, and future directions.

\hypertarget{background}{%
\section{Background}\label{background}}

The \texttt{bigsimr} \texttt{R} package provides multiple algorithms to work with high-dimensional multivariate data, but all these algorithms were originally designed to support a single task:
to generate random vectors drawn from multivariate probability distribution with given marginal distributions and dependency metrics.
Specifically, our goal is to efficiently simulate a large number, \(B\), of random vectors \({\bf Y}=(Y_1, \ldots, Y_d)^\top\) with \textbf{correlated} components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) \(F_i\), where \(d\) can be very large and still be computed in practically useful times.

When designing this methodology, we developed the following properties to guide our effort.
We divide the properties into two categories: (1) basic properties (BP) and ``scaleability'' properties (SP).
The BPs are adapted from an existing criteria due to \citet{Nik13a}.
Our simulation strategy should allow:

\begin{itemize}
\tightlist
\item
  BP1: A wide range of dependences, allowing both positive and negative values, and, ideally, admitting the full range of possible values.
\item
  BP2: Flexible dependence, meaning that the number of bivariate marginals can be equal to the number of dependence parameters.
\item
  BP3: Flexible marginal modeling, generating heterogeneous data --- possibly from differing probability families.
\end{itemize}

Moreover, the simulation method must \textbf{scale} to high dimensions:

\begin{itemize}
\tightlist
\item
  SP1: Procedure must scale to high dimensions, computable in a reasonable amount time.
\item
  SP2: Procedure must scale to high dimensions while maintaining accuracy.
\end{itemize}

To fix ideas and provide examples applications enabled via \texttt{bigsimr}, the next section describes a motivating data set that originally inspired the authors' interest in developing this methodology.

\hypertarget{motivating-example-rna-seq-data}{%
\subsection{Motivating example: RNA-seq data}\label{motivating-example-rna-seq-data}}

Simulating high-dimensional, non-normal, correlated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data \citep{Wang2009b, Conesa2016b} derived from breast cancer patients.
The RNA-seq data-generating process involves counting how often a particular messenger RNA (mRNA) is expressed in a biological sample.
RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data.
For human derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted.
Importantly, due to inherent biological processes, gene expression data exhibits correlation --- co-expression --- across genes \citep{BE07, Schissler2018}.

We'll illustrate our methodology using the well-studied Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments).
For simplicity, we only consider high expressing genes.
In turn, we begin by filtering to retain the top 5\% highest- expressing genes (in terms of median expression) of the 20,501 gene measurements from \(N=1212\) patients' tumor samples, resulting in \(d=1026\) genes.
This gives a massive number of pairwise dependencies among the marginals (specifically, \(\ensuremath{5.2582\times 10^{5}}\) correlation parameters).
We further process the data to illustrate our methodology's flexible and robust simulation scheme, while better aligning with the actual data-generating process, by rounding \citet{Li2011c} RNA-seq by Expectation Maximization (RSEM) values to counts.
Table \ref{tab:ch010-realDataTab} displays counts for three selected high-expressing genes for the first five patients' breast tumor samples.

\begin{table}

\caption{\label{tab:ch010-realDataTab}mRNA counts for three selected high-expressing genes from the first five observations of the BRCA data set.}
\centering
\begin{tabular}[t]{rrr}
\toprule
RPL5 & TXNIP & VIM\\
\midrule
20283 & 13401 & 26883\\
18614 & 11365 & 28806\\
31378 & 5365 & 22221\\
37861 & 5873 & 26871\\
17902 & 12564 & 15985\\
\bottomrule
\end{tabular}
\end{table}

To help visualize the bivariate relationships for these three selected genes across all patients, Figure \ref{fig:ch010-realDataFig} plots the marginal distributions and estimated Spearman's correlations (see Equation \eqref{eq:spearman} below).

\begin{figure}

{\centering \includegraphics[width=0.85\linewidth]{article_bigsimr_files/figure-latex/ch010-realDataFig-1}

}

\caption{Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes. The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations. Modeling these data motivate our simulation methodology.}\label{fig:ch010-realDataFig}
\end{figure}

\hypertarget{measures-of-dependency}{%
\subsection{Measures of dependency}\label{measures-of-dependency}}

In multivariate analysis, an analyst must select a metric to quantify dependency.
The most widely-known is the Pearson (product-moment) correlation coefficient that describes the linear association between two random variables \(X\) and \(Y\), and, it is given by

\begin{equation}
\label{eq:pearson}
\rho_P(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ var(X)var(Y)\right]^{1/2}}.
\end{equation}

As \citet{MB13} and \citet{MK01} discuss, for a bivariate normal \((X,Y)\) random vector, the Pearson correlation completely describes the dependency between the components.
For non-normal marginals with monotone correlation patterns, \(\rho_P\) suffers some drawbacks and may mislead or fail to capture important relationships (\citet{MK01}).
Alternatively in these settings, analysts often prefer rank-based correlation measures to describe the degree of monotonic association.

Two nonparametric, rank-based measures common in practice are Spearman's correlation (denoted \(\rho_S\)) and Kendall's \(\tau\).
Define

\begin{equation}
\label{eq:spearman}
\rho_S(X,Y) = 3 \left[ P\left[ (X_1 - X_2)(Y_1-Y_3) > 0 \right] - P\left[ (X_1 - X_2)(Y_1-Y_3) < 0 \right] \right],
\end{equation}

\noindent where \((X_1, Y_1) \overset{d}{=} (X,Y), X_2 \overset{d}{=} X, Y_3 \overset{d}{=} Y\) with \(X_2\) and \(Y_3\) are independent of one other and of \((X_1, Y_1)\).
Spearman's \(\rho_S\) has an appealing correspondence as the Pearson's correlation coefficient on \emph{ranks} of the values, thereby captures nonlinear yet monotone relationships.

Kendall's \(\tau\), on the other hand, is the difference in probabilities of concordant and discordant pairs of observations \((X_i, Y_i)\) and \((X_j, Y_j)\).
By concordance we mean that orderings have the same direction (e.g., if \(X_i < X_j\), then \(Y_i < Y_j\)) and is determined by the ranks of the values, not the values themselves.

Both \(\tau\) and \(\rho_S\) are \textbf{invariant to under monotone transformations} of the underlying random variates.
As we will describe more fully in the \protect\hyperlink{algorithms}{Algorithms} section, this property is essential to scaleable match rank-based correlations with speed (SP1) and accuracy (SP2).

\emph{Correspondence among Pearson, Spearman, \(\tau\) correlations}.
This is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficient, as the marginal distributions \(F_i\) are intrinsic in their calculation.
But for \textbf{bivariate normal vectors}, however, the correspondence is well-known:

\begin{equation}
\label{eq:convertKendall}
\rho_{P} = sin \left( \tau \times \frac{\pi}{2} \right), 
\end{equation}

\noindent and similarly for Spearman's \(\rho\) \citep{K58},

\begin{equation}
\label{eq:convertSpearman}
\rho_P = 2 \times sin \left( \rho_S \times \frac{\pi}{6} \right).
\end{equation}

These facts are also critical in our simulation algorithm to broaden the dependency measures supported by \texttt{bigsimr}, in a computationally effective manner.

\emph{Discrete marginal considerations}.
Spearman's correlation for discrete marginal suffers some issues due to the nonzero probability of ties (for example, the Spearman's correlation of a discrete-valued random variable \(X\) with itself could be less than 1; \citet{MB13}).
One remedy for this issue is to rescale Equation \eqref{eq:spearman}.
For two random variables \(X,Y\) with probability mass functions (PMFs) or probability densities functions (PDFs) \(p(x)\) and \(q(y)\), respectively, define the rescaled Spearman's correlation as

\begin{equation}
\label{eq:spearmanRescaled}
\rho_{RS}(X,Y) = \frac{\rho_s(X,Y)}{ \left[ \left[ 1 - \sum_x p(x)^3 \right] \left[ 1 - \sum_y q(y)^3 \right] \right]^{1/2}}.
\end{equation}

Note that with continuous marginals the rescaling returns \(\rho_S\).
For discrete marginals with large or infinite support, computing the adjustment factors \(\sum_x p(x)^3\), \(\sum_y p(y)^3\) over all large number of pairs becomes expensive (often violating desired property SP1).
And further the infinite sums must be approximately for count-valued data and potentially violating desired property SP2.

Also, for a discrete random variable \(Y_i\), some care must be taken to define the quantile function \(F_{i}^{-1}\).
Let

\begin{equation}
F_{i}^{-1} = inf\{y:F_{i}(y) \geq u \}.
\label{eq:inverseCDF}
\end{equation}

\emph{Marginal-dependent bivariate correlation bounds}.
Given two marginal distributions, \(\rho_P\) is not free to vary the entire range of possible correlations \([-1,1]\).
The so-called \emph{Frechet-Hoeffding bounds} are well-studied (for example, see \citet{Nelsen2007} and \citet{BF17}).
This situation gives strict restraints on the possible correlations and cannot be overcome through algorithm design.
In general, the bounds are given by

\begin{equation}
\label{eq:frechet}
\rho_P^{max} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (U) \right), \quad \rho_P^{min} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (1 - U) \right)
\end{equation}

\noindent where \(U\) is a uniform random variable in \((0,1)\), and \(F^{-1}_1, F^{-1}_2\) are the inverse CDF of random variables \(X_1\) and \(X_2\), respectively.
For discrete random variables, define \(F^{-1}\) as in Equation \eqref{eq:inverseCDF}.

\hypertarget{gaussian-copulas}{%
\subsection{Gaussian copulas}\label{gaussian-copulas}}

There is a strong connection of our simulation strategy to Gaussian \textbf{copulas} (see \citet{Nelsen2007} for a technical introduction).
A copula is a distribution function on \([0,1]^d\) that describes a multivariate probability distribution with standard uniform marginals.
This definition provides a powerful, natural way characterize joint probability structure.
Consequently, the study of copulas is an important and active area of statistical theory and practice.

For any random vector \({\bf X}=(X_1, \ldots, X_d)\) with CDF \(F\) and marginal CDFs \(F_i\) there is a copula function
\(C(u_1, \ldots, u_d)\) so that

\[
F(x_1, \ldots,x_d) = \mathbb P(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), \,\,\, x_i\in \mathbb R, i=1,\ldots,d. 
\]

A Gaussian copula is the case where all marginal CDFs \(F_i\) are the standard normal CDF, \(\Phi\).
This representation corresponds to a multivariate normal distribution with standard normal marginal distributions and covariance matrix \({\bf R_P}\).
But since the marginals are standardized to have unit variance, this \({\bf R_P}\) is a Pearson correlation matrix.
If \(F_{{\bf R}}\) is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula \(C_{{\bf R}}\) is defined through

\begin{equation}
\label{eq:gauss}
F_{{\bf R}}(x_1, \ldots, x_d) = C_{{\bf R}}(\Phi(x_1), \ldots, \Phi(x_d)),
\end{equation}

where \(\Phi(\cdot)\) is the standard normal CDF.
Note that the copula \(C_{{\bf R}}\) is the familar multivariate normal CDF of the random vector \((\Phi(X_1), \ldots, \Phi(X_d))\), where \((X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})\).

Sklar's Theorem \citep{Sklar1959, Ubeda-Flores2017} guarantees that given inverse CDFs \(F_i^{-1}\)s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions.
For example, using Gaussian copulas, we can construct a random vector \({\bf Y} = (Y_1, \ldots, Y_d)\) with \(Y_i=F_i^{-1}(U_i)\), \(i=1, \ldots, d\), via \({\bf U} = (U_1, \ldots, U_d)\) viz \(U_i=\Phi(X_i)\), \(i=1, \ldots, d\) povides \(Y_i \sim F_i, \, \forall \, i\) .

\hypertarget{algorithms}{%
\section{Algorithms}\label{algorithms}}

This section describes our methods involved in simulating a random vector \(\bf Y\) with \(Y_i\) components for \(i=1,2,\ldots,d\).
Each \(Y_i\) has a specified marginal CDF \(F_i\) and its inverse \(F^{-1}_i\).
To characterize dependency, every pair \((Y_i, Y_j)\) has a given Pearson correlation \(\rho_P\), Spearman correlation \(\rho_S\), and/or Kendall's \(\tau\).
The method is best understand as a \textbf{high-performance Gaussian copula} (Equation \eqref{eq:gauss}) providing a high-dimensional NORTA-inspired algorithm.

\hypertarget{normal-to-anything-norta}{%
\subsection{NORmal To Anything (NORTA)}\label{normal-to-anything-norta}}

The well-known NORTA algorithm \citep{Cario1997} can be used simulate a random vector \(\bf Y\) with variance-covariance matrix \(\Sigma_{\bf Y}\).
Specifically, the NORTA algorithm follows like this:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Simulate a random vector \(\bf Z\) with \(d\) \textbf{independent} and \textbf{identical} standard normal components.
\item
  Determine the input matrix \(\Sigma_{\bf Z}\) that corresponds with the specified output \(\Sigma_{\bf Y}\).
\item
  Produce a Cholesky factor \(M\) of \(\Sigma_{\bf Z}\) so that \(M M^{\prime}=\Sigma_{\bf Z}\).
\item
  Set \(X\) by \(X \gets MZ\).
\item
  \(\text{Return} \; Y \; \text{where} \; Y_i \gets F_{Y_i}^{-1}[\Phi(X_i)], \; i=1,2,...,d\).
\end{enumerate}

With modern parallelized computing, steps 1, 3, 4, 5 are readily implemented as high-performance, multi-core and/or graphical-processing-unit (GPU) accelerated, algorithms --- providing the fast scaleability using readily-available hardware.

Matching specified Pearson correlation coefficients exactly (step 2 above), however, is problematic.
In general, there is no closed form correspondence between the components of the input \(\Sigma_{\bf Z}\) and target \(\Sigma_{\bf Y}\).
Matching the correlations involves evaluating or approximating \(\binom{d}{2}\) integrals of the form \(EY_iY_j = \int \int y_i y_j f_{X}(F_i^{-1}(\Phi(z_i)), F_j^{-1}(\Phi(z_j))dy_idy_j\), for \(i,j=1,2,\ldots,d, \; i \neq j\).
For high-dimensional data, these evaluations are often too costly to enable feasible simulation studies.
For low-dimensional problems, methods and tools exist to match Pearson correlations precisely, see \citet{Chen2001}; \citet{Xia17}; \citet{MB13} and the publicly available \texttt{nortaRA} R package.

To maintain scaleability (SP1), our solution is to essentially avoid this complication in Pearson matching.
Since our goal is to simulate non-normal marginals, we greatly prefer the use of rank-based measures \(\rho_S\) and \(\tau\) from a modeling standpoint.
Further, \(\rho_S\) and \(\tau\)'s invariance under monotone transformation (see \protect\hyperlink{background}{Background}), preserves the correlation coefficients through steps 3, 4, and 5 in the NORTA algorithm above.
This eliminates the need for computing the \(\binom{d}{2}\) integrals to match exactly (nothing is for free, however, as discussed in below in Section \ref{rand-vec-gen}).

Despite all this, if one does desire to characterize dependency using Pearson correlations, simply using the target Pearson correlation matrix as the initial conditions to our proposed algorithm will lead to approximate matching in the resultant distribution \citep{Song00} in many practical applications.
The quality of this approximation depends on the setting, but in practice, for high-dimensional count data we find the accuracy to be adequate.
Later, we'll study the robustness of our method to this limitation in selected \href{\%7B\#simulations\%7D}{Monte Carlo evaluations}.

\hypertarget{rand-vec-gen}{%
\subsection{\texorpdfstring{Random vector generation via \texttt{bigsimr::rvec}}{Random vector generation via bigsimr::rvec}}\label{rand-vec-gen}}

Now we describe \texttt{bigsimr::rvec}, our algorithm to generate random vectors.
It mirrors the classical NORTA algorithm above with some modifications for rank-based dependency matching:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Pre-processing for nonparametric dependency matching.

  \begin{enumerate}
  \def\labelenumii{(\roman{enumii})}
  \tightlist
  \item
    Convert from either \({\bf R_{Spearman}}\) or \({\bf R_{Kendall}}\) into the corresponding MVN input correlation \({\bf R_{Pearson}}\).
  \item
    Check that \({\bf R_{Pearson}}\) is semi-positive definite.\\
  \item
    If not compute a close semi-positive definite correlation matrix \({\bf \widetilde{R}_{Pearson}}\).\\
  \end{enumerate}
\item
  Gaussian copula construction.

  \begin{enumerate}
  \def\labelenumii{(\roman{enumii})}
  \tightlist
  \item
    Generate \({\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_{Pearson}})\).\\
  \item
    Transform \({\bf X}\) to \({\bf U} = (U_1, \ldots, U_d)\) viz \(U_i=\Phi(X_i)\), \(i=1, \ldots, d\).\\
  \end{enumerate}
\item
  Quantile evaluations.

  \begin{enumerate}
  \def\labelenumii{(\roman{enumii})}
  \tightlist
  \item
    Return \({\bf Y} = (Y_1, \ldots, Y_d)\), where \(Y_i=F_i^{-1}(U_i)\), \(i=1, \ldots, d\);
  \end{enumerate}
\end{enumerate}

The pre-processing (Step 1) takes advantage of the closed-form relationships between \(\rho_S\) and \(\tau\) with \(\rho_P\) for bivariate normal random variables via Equations \eqref{eq:convertSpearman} or \eqref{eq:convertKendall}, respectively (implemented as \texttt{bigsimr::cor\_covert}).

A complication often arises at this stage: the resultant matrix may become indefinite, either through numerical error or naturally occurring in the correlation conversion.
Researchers working in multivariate computation frequently encounter such difficulties and need to find a close positive (semi-)definite matrix.
A widely-available routine for this task in \texttt{R} is called \texttt{matrix::nearPD}, though it is not suitable for high dimensions.
To overcome this issue, we've developed \texttt{bigsimr::nearPSD}, a quadractically-convergent Newton method for finding the nearest correlation matrix, developed by \citet{QS2006}.
We hope that this routine could be useful in many applications aside from our primary goal of random vector generation.

Once the target margins and algorithm inputs are determined, steps 2 and 3 are essentially a NORTA algorithm with modern high-performance computing implementations.
Specifically, step 2i uses either an efficient multi-core multivariate normal simulator (the \texttt{R} package \texttt{mvnfast}\citep{Fasiolo2016}) or a using Google's JAX python library NumPy for graphical-processing-unit (GPU) acceleration of the Cholesky factorization and matrix multiplication (steps 3, 4 in the NORTA algorithm in the preceding section).

\hypertarget{a-note-on-norta-in-higher-dimensions}{%
\subsection{A note on NORTA in higher dimensions}\label{a-note-on-norta-in-higher-dimensions}}

Sklar's theorem provides a useful characterization of multivariate distributions through copulas.
Yet the choice of copula-based simulation algorithm affects which joint distributions may be simulated.
Even in low-dimensional spaces (e.g., \(d=3\)), there exists valid multivariate distributions with \emph{feasible} Pearson correlation matrices that NORTA cannot match exactly \citep{LH75}.
This occurs when the bivariate transformations are applied to find the input correlation matrix yet when combined the resultant matrix is indefinite.
These situations do occur, even using exact analytic calculations. Such problematic target correlation matrices are termed \emph{NORTA defective}.

\citet{GH02} conducted a Monte Carlo study to estimate the probability of NORTA defective matrices while increasing the dimension \(d\).
They found at what is now considered low-to-moderate dimensions (\(d \approx 20\)) that almost \emph{all} feasible matrices are NORTA defective.
This stems from the concentration of measure near the boundary of the space of all possible correlation matrices as dimension increases. Unfortunately, it is precisely near this boundary that NORTA defective matrices reside.

There is hope, however, as \citet{GH02} also show that by augmenting the NORTA procedure by replacing the indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate \(d\).
This provides evidence that our nearest positive semi-definite (PSD) augmented approach will maintain reasonable accuracy if our input matching scheme returns an indefinite matrix, at least for the rank-based matching scheme described above.

\hypertarget{the-bigsimr-r-package}{%
\section{\texorpdfstring{The \texttt{bigsimr} R package}{The bigsimr R package}}\label{the-bigsimr-r-package}}

The \texttt{bigsimr} package is a high-performance implementation of the proposed random vector generation algorithm and associated functions (see Section \ref{algorithms} for details).
When designing \texttt{bigsimr}, we aimed to conveniently provide parallelized computation, through multi-core and GPU acceleration, while allowing advanced users to customize and automate workflows.

This subsections below describe the basic use of \texttt{bigsimr}, by stepping through a low-dimensional (2D) simulation workflow for the often-used example data set \texttt{airquality}.
This workflow proceeds from data, to estimation, simulation configuration, random vector generation, and result visualization.
For this low-dimensional setting, we compute using a single central processing unit (CPU) as the overhead in forking the tasks to multiple cores outweighs the computational gains.
Then we transition to advanced use where we briefly describe some of the high-performance features and syntax.
The section concludes with a short description of how to use \texttt{bigsimr} on a computer clusters through slurm scheduling via the \texttt{rslurm} package.

\hypertarget{basic-use-illustrated-through-a-minimal-example}{%
\subsection{Basic use illustrated through a minimal example}\label{basic-use-illustrated-through-a-minimal-example}}

We'll demonstrate the basic use and syntax of \texttt{bigsimr} through an example workflow applied to the New York air quality data set (\texttt{airquality}) included in the R \texttt{datasets} package.
First, we load the \texttt{bigsimr} library and a few other convenient data science packages, including the syntactically-elegant \texttt{tidyverse} suite of \texttt{R} packages.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(bigsimr)}
\KeywordTok{library}\NormalTok{(tidyverse)}
\KeywordTok{library}\NormalTok{(patchwork)}
\end{Highlighting}
\end{Shaded}

For simplicity and to provide a minimal working example, we'll consider bivariate simulation of temperature, in degrees Fahrenheit, and ozone level, in parts per billion.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{df <-}\StringTok{ }\NormalTok{airquality }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{select}\NormalTok{(Temp, Ozone) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{drop_na}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
Rows: 116
Columns: 2
$ Temp  <int> 67, 72, 74, 62, 66, 65, 59, 61, 74, 69, 66, 68, 58, 64, 66, 5...
$ Ozone <int> 41, 36, 12, 18, 28, 23, 19, 8, 7, 16, 11, 14, 18, 14, 34, 6, ...
\end{verbatim}

Figure \ref{fig:ch030-aq-joint-dist} visualizes the bivariate relationship between Ozone and Temperature.
We aim to simulate random two-component vectors mimicking this structure.
The margins are not normally distributed, particularly the ozone level exhibits a strong positive skew.

\begin{figure}

{\centering \includegraphics[width=0.85\linewidth]{article_bigsimr_files/figure-latex/ch030-aq-joint-dist-1}

}

\caption{Bivariate scatterplot of Ozone vs. Temperature with estimated marginal densities. The data are left skewed tails and appear to be correlation.}\label{fig:ch030-aq-joint-dist}
\end{figure}

Next, we specify the marginal distributions and correlation coefficient (both its type and magnitude).
Here the analyst is free to be creative.
For this example, we will not take up goodness-of-fit considerations to determine the marginal distributions.
But it seems sensible without domain knowledge to estimate these quantities from the data and \texttt{bigsimr} contains fast functions designed for this task.

\emph{Specifying marginal distributions}.
Based on the estimated densities in Figure \ref{fig:ch030-aq-joint-dist}, we'll assume \texttt{Temp} is normally distributed and \texttt{Ozone} is log-normally distributed since its values are positive and skewed.
We'll use the well-known, unbiased estimators for the normal distribution's parameters and maximum likelihood estimators for the log-normal parameters:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{df }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{select}\NormalTok{(Temp) }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{summarise_all}\NormalTok{(}\DataTypeTok{.funs =} \KeywordTok{c}\NormalTok{(}\DataTypeTok{mean =}\NormalTok{ mean, }\DataTypeTok{sd =}\NormalTok{ sd))}
\NormalTok{   mean    sd}
\DecValTok{1} \FloatTok{77.87} \FloatTok{9.485}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{mle_mean <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x) }\KeywordTok{mean}\NormalTok{(}\KeywordTok{log}\NormalTok{(x))}
\NormalTok{mle_sd <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x) }\KeywordTok{mean}\NormalTok{( }\KeywordTok{sqrt}\NormalTok{( (}\KeywordTok{log}\NormalTok{(x) }\OperatorTok{-}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{log}\NormalTok{(x)))}\OperatorTok{^}\DecValTok{2}\NormalTok{ ) )}

\NormalTok{df }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{select}\NormalTok{(Ozone) }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{summarise_all}\NormalTok{(}\DataTypeTok{.funs =} \KeywordTok{c}\NormalTok{(}\DataTypeTok{meanlog =}\NormalTok{ mle_mean, }\DataTypeTok{sdlog =}\NormalTok{ mle_sd))}
\NormalTok{  meanlog  sdlog}
\DecValTok{1}   \FloatTok{3.419} \FloatTok{0.6967}
\end{Highlighting}
\end{Shaded}

Next, we'll configure the input marginals for later input into \texttt{bigsimr::rvec}.
The marginal distributions are specifying using \texttt{R}'s special \texttt{alist} function.
This allows one to enter the distributions without evaluating anything (yet).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{margins <-}\StringTok{ }\KeywordTok{alist}\NormalTok{(}
    \KeywordTok{qnorm}\NormalTok{(}\DataTypeTok{mean =} \FloatTok{77.871}\NormalTok{, }\DataTypeTok{sd =} \FloatTok{9.4855}\NormalTok{),}
    \KeywordTok{qlnorm}\NormalTok{(}\DataTypeTok{meanlog =} \FloatTok{3.419}\NormalTok{, }\DataTypeTok{sdlog =} \FloatTok{0.6967}\NormalTok{),}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Notice that we use the \emph{quantile} function for the marginals, as that is how the marginal distributions \(F_i\) enter into the \texttt{bigsimr::rvec} algorithm.
This implementation strategy supports all \texttt{R} \texttt{base} probability distributions.
And allows flexible extensions using other \texttt{R} packages that adhere to conventions, such as \texttt{extraDistr}.
Further, by using \texttt{alist}, users can specify their own custom distributions (see below in \protect\hyperlink{custom-margins}{creating custom margins}).

It is a bit inconvenient to have to fill in the parameter values manually each time, so we provide a convenience function \texttt{mlist} which behaves similarly to \texttt{alist}, except that it will evaluate the right hand side of argument values within the list.
This is intended to help when scaling up your code to high dimensions when many marginals to specify.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{margins <-}\StringTok{ }\KeywordTok{mlist}\NormalTok{(}
  \KeywordTok{qnorm}\NormalTok{(}\DataTypeTok{mean =} \KeywordTok{mean}\NormalTok{(df}\OperatorTok{$}\NormalTok{Temp), }\DataTypeTok{sd =} \KeywordTok{sd}\NormalTok{(df}\OperatorTok{$}\NormalTok{Temp)),}
  \KeywordTok{qlnorm}\NormalTok{(}\DataTypeTok{meanlog =} \KeywordTok{mle_mean}\NormalTok{(df}\OperatorTok{$}\NormalTok{Ozone), }\DataTypeTok{sdlog =} \KeywordTok{mle_sd}\NormalTok{(df}\OperatorTok{$}\NormalTok{Ozone))}
\NormalTok{)}
\NormalTok{margins}
\NormalTok{[[}\DecValTok{1}\NormalTok{]]}
\KeywordTok{qnorm}\NormalTok{(}\DataTypeTok{mean =} \FloatTok{77.8706896551724}\NormalTok{, }\DataTypeTok{sd =} \FloatTok{9.48548563759966}\NormalTok{)}

\NormalTok{[[}\DecValTok{2}\NormalTok{]]}
\KeywordTok{qlnorm}\NormalTok{(}\DataTypeTok{meanlog =} \FloatTok{3.41851510081201}\NormalTok{, }\DataTypeTok{sdlog =} \FloatTok{0.696668863646896}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\emph{Specifying correlation}.
As mentioned, the user must decide how to describe correlation, based on the particulars of the problem.
For non-normal data and for improved simulation accuracy in our scheme, we advocate the use of rank-based correlations Spearman's \(\rho_S\) and Kendall's \(\tau\).
But we also support approximate Pearson correlation coefficient matching, while cautioning the user to check the performance for their parametric multivariate model (see \protect\hyperlink{simulations}{Monte Carlo evaluations} for evaluation strategies and guidance).
To aid in correlation specification, and estimation in general, we provide a high-performance function \texttt{bigsimr::cor\_fast} which estimates Pearson, Spearman, or Kendall correlation using the fastest methods available.
(Anyone who has tried estimating Kendall's \(\tau\) using \texttt{stats::cor} can attest that the routine does not scale to even moderate dimensions).
Notably, these estimation methods are the standard approaches, not designed specifically designed for high-dimensional correlation estimation (see \href{\%7B\#discussion}{Conclusion and Discussion} for more on this).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{type <-}\StringTok{ 'spearman'}
\NormalTok{(rho <-}\StringTok{ }\KeywordTok{cor_fast}\NormalTok{(df, }\DataTypeTok{method =} \StringTok{"spearman"}\NormalTok{))}
\NormalTok{       Temp Ozone}
\NormalTok{Temp  }\FloatTok{1.000} \FloatTok{0.774}
\NormalTok{Ozone }\FloatTok{0.774} \FloatTok{1.000}
\end{Highlighting}
\end{Shaded}

\emph{Checking the theoretical correlation bounds}
As discussed in Section \ref{background}, given a pair of marginal distributions the possible correlations are not free to vary between \([-1, 1]\).
To ensure that the simulation is not configured to impossible settings, we provide the \texttt{bigsimr::cor\_bounds} function provides MC estimated theoretical lower and upper bounds (using the Generate, Sort, and Correlate algorithm of \citet{DH2011}).

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{cor_bounds}\NormalTok{(}\DataTypeTok{margins =}\NormalTok{ margins, }\DataTypeTok{type =}\NormalTok{ type)}
\OperatorTok{$}\NormalTok{lower}
\NormalTok{     [,}\DecValTok{1}\NormalTok{] [,}\DecValTok{2}\NormalTok{]}
\NormalTok{[}\DecValTok{1}\NormalTok{,]    }\DecValTok{1}   \DecValTok{-1}
\NormalTok{[}\DecValTok{2}\NormalTok{,]   }\DecValTok{-1}    \DecValTok{1}

\OperatorTok{$}\NormalTok{upper}
\NormalTok{     [,}\DecValTok{1}\NormalTok{] [,}\DecValTok{2}\NormalTok{]}
\NormalTok{[}\DecValTok{1}\NormalTok{,]    }\DecValTok{1}    \DecValTok{1}
\NormalTok{[}\DecValTok{2}\NormalTok{,]    }\DecValTok{1}    \DecValTok{1}
\end{Highlighting}
\end{Shaded}

Since our estimated Spearman correlation \(\hat{ \rho}_S\) is within the theoretical bounds, the correlation is valid as input to \texttt{bigsimr:rvec}.
On the other hand, the bounds on the Pearson correlation coefficient \(\rho_P\) between these margins is

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{cor_bounds}\NormalTok{(}\DataTypeTok{margins =}\NormalTok{ margins, }\DataTypeTok{type =} \StringTok{'pearson'}\NormalTok{, }\DataTypeTok{reps=}\FloatTok{1e6}\NormalTok{)}
\OperatorTok{$}\NormalTok{lower}
\NormalTok{       [,}\DecValTok{1}\NormalTok{]   [,}\DecValTok{2}\NormalTok{]}
\NormalTok{[}\DecValTok{1}\NormalTok{,]  }\FloatTok{1.000} \FloatTok{-0.881}
\NormalTok{[}\DecValTok{2}\NormalTok{,] }\FloatTok{-0.881}  \FloatTok{1.000}

\OperatorTok{$}\NormalTok{upper}
\NormalTok{       [,}\DecValTok{1}\NormalTok{]   [,}\DecValTok{2}\NormalTok{]}
\NormalTok{[}\DecValTok{1}\NormalTok{,] }\FloatTok{1.0000} \FloatTok{0.8806}
\NormalTok{[}\DecValTok{2}\NormalTok{,] }\FloatTok{0.8806} \FloatTok{1.0000}
\end{Highlighting}
\end{Shaded}

\emph{Simulating random vectors}.
Finally, we arrive at the main function of \texttt{bigsimr}, \texttt{rvec}.
Let's now simulate \(B=10,000\) random vectors from the assumed joint distribution of Ozone levels and Temp.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{x <-}\StringTok{ }\KeywordTok{rvec}\NormalTok{(}\DecValTok{10000}\NormalTok{, rho, margins, type)}
\NormalTok{df_sim <-}\StringTok{ }\KeywordTok{as.data.frame}\NormalTok{(x)}
\KeywordTok{colnames}\NormalTok{(df_sim) <-}\StringTok{ }\KeywordTok{colnames}\NormalTok{(df)}
\end{Highlighting}
\end{Shaded}

Figure \ref{fig:ch030-plot-sim} plots the 10,000 simulated points.

\begin{figure}

{\centering \includegraphics[width=0.85\linewidth]{article_bigsimr_files/figure-latex/ch030-plot-sim-1}

}

\caption{Contour plot and marginal densities for the simulated bivariate distribution of Air Quality Temperatures and Ozone levels. The simulated points mimic the observed data with respect to both the marginal characteristics and bivariate association.}\label{fig:ch030-plot-sim}
\end{figure}

\hypertarget{advanced-use}{%
\subsection{Advanced use}\label{advanced-use}}

\emph{Creating a custom marginal distribution}.
Because \texttt{bigsimr} uses an \texttt{alist} to store the margins, any probability distribution with a well-defined inverse CDF can be used including custom marginal distributions not provided in base \texttt{R}.
Therefore, to specify a marginal distribution absent from an existing package, the user needs to provide a closed-form expression to form the corresponding quantile function.

It is important to follow \texttt{R}'s naming convention for probability distributions.
Users should prefix distributions with \texttt{r} and \texttt{q} for \emph{random} and \emph{quantile} respectively.
Only the quantile function is necessary to simulate random vectors, but to compute the theoretical correlation bounds, it is required to supply univariate random number generator as well.

For an example, let's provide a custom \emph{Pareto} distribution for use with \texttt{bigsimr}.
The Pareto CDF is

\[
F(x) = 1 - \left(\frac{x_m}{x}\right)^\alpha
\]

for scale \(x_m > 0\) and shape \(\alpha > 0\), with support \(x \in \left[x_m, \infty\right)\).
From the CDF, we compute the inverse CDF

\[
F^{-1}(p) = \frac{x_m}{\left(1 - p\right)^{1/\alpha}}
\]

Next we define the Pareto quantile function in R:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{qpareto <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(p, scale, shape) \{}
\NormalTok{  scale }\OperatorTok{/}\StringTok{ }\NormalTok{(}\DecValTok{1} \OperatorTok{-}\StringTok{ }\NormalTok{p)}\OperatorTok{^}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\NormalTok{shape)}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

Writing random number generating function for our target marginal can be accomplished by calling the quantile function on a uniformly distributed random variable (the inverse transform method \citet{Rizzo2007}).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{rpareto <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(n, scale, shape) \{}
  \KeywordTok{qpareto}\NormalTok{(}\KeywordTok{runif}\NormalTok{(n), scale, shape)}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

Now with the \texttt{qpareto} and \texttt{rpareto} functions, we can use the distribution in \texttt{bigsimr} just like the other built-in distributions.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{margins <-}\StringTok{ }\KeywordTok{alist}\NormalTok{(}
  \KeywordTok{qnorm}\NormalTok{(}\DataTypeTok{mean =} \FloatTok{3.14}\NormalTok{, }\DataTypeTok{sd =} \FloatTok{0.1}\NormalTok{),}
  \KeywordTok{qbeta}\NormalTok{(}\DataTypeTok{shape1 =} \DecValTok{1}\NormalTok{, }\DataTypeTok{shape2 =} \DecValTok{4}\NormalTok{),}
  \KeywordTok{qnbinom}\NormalTok{(}\DataTypeTok{size =} \DecValTok{10}\NormalTok{, }\DataTypeTok{prob =} \FloatTok{0.75}\NormalTok{),}
  \KeywordTok{qpareto}\NormalTok{(}\DataTypeTok{scale =} \FloatTok{1.11}\NormalTok{, }\DataTypeTok{shape =} \FloatTok{5.55}\NormalTok{)}
\NormalTok{)}
\KeywordTok{cor_bounds}\NormalTok{(margins, }\StringTok{"pearson"}\NormalTok{)}
\NormalTok{rho <-}\StringTok{ }\KeywordTok{cor_randPD}\NormalTok{(}\DecValTok{4}\NormalTok{)}
\NormalTok{x <-}\StringTok{ }\KeywordTok{rvec}\NormalTok{(}\DecValTok{10}\NormalTok{, rho, margins)}
\end{Highlighting}
\end{Shaded}

\emph{Using \texttt{bigsimr} on a computing cluster via \texttt{rslurm}}.
Though \texttt{bigsimr} runs quickly, at large \(d\) users may want to run jobs on a shared computing server.
The R package \texttt{rslurm} makes it easy to run embarrassingly large parallel \texttt{rvec} calls.
This example assumes that \texttt{bigsimr} is installed on a system with a slurm scheduler installed.
A single run of \texttt{bigsimr:rvec} using \texttt{rslurm} can be code as:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(rslurm)}
\NormalTok{sjob <-}\StringTok{ }\KeywordTok{slurm_call}\NormalTok{(rvec, }\DataTypeTok{jobname =} \StringTok{'rvec'}\NormalTok{,}
                   \KeywordTok{alist}\NormalTok{(}\DataTypeTok{n=}\FloatTok{1e6}\NormalTok{,}
                         \DataTypeTok{rho =}\NormalTok{ rho,}
                        \DataTypeTok{margins =}\NormalTok{ margins,}
                        \DataTypeTok{type =} \StringTok{"spearman"}\NormalTok{),}
                   \DataTypeTok{submit =} \OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Now, let's show off the real power of combining \texttt{bigsimr} and \texttt{rslurm} by simulating many correlation structures.
The \texttt{rslurm::slumr\_map} syntax mirrors the \texttt{base::lapply} and \texttt{purrr::map} functions.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(rslurm)}
\NormalTok{simReps <-}\StringTok{ }\DecValTok{100}
\NormalTok{rhoList <-}\StringTok{ }\KeywordTok{replicate}\NormalTok{( }\DataTypeTok{n =}\NormalTok{ simReps, bigsimr}\OperatorTok{::}\KeywordTok{cor_randPSD}\NormalTok{(}\DataTypeTok{d =} \KeywordTok{length}\NormalTok{(margins)),}
                     \DataTypeTok{simplify=}\OtherTok{FALSE}\NormalTok{ )}
\NormalTok{sjob <-}\StringTok{ }\KeywordTok{slurm_map}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ rhoList,}
                  \DataTypeTok{f =}\NormalTok{ rvec,}
                  \DataTypeTok{jobname =} \StringTok{'rvecMap'}\NormalTok{,}
                  \DataTypeTok{n=}\FloatTok{1e6}\NormalTok{,}
                  \DataTypeTok{margins =}\NormalTok{ margins,}
                  \DataTypeTok{type =} \StringTok{"spearman"}\NormalTok{,}
                  \DataTypeTok{nodes =} \DecValTok{4}\NormalTok{,}
                  \DataTypeTok{cpus_per_node =} \DecValTok{4}\NormalTok{,}
                  \DataTypeTok{cores =} \DecValTok{4}\NormalTok{,}
                  \DataTypeTok{submit =} \OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

On a cluster carrying 24 nodes with 48 threads, these 100 jobs completed in about a minute.

\hypertarget{simulations}{%
\section{Monte Carlo evaluations}\label{simulations}}

Before applying our methodology to real data simulation, we conduct several Monte Carlo studies to investigate method performance.
Since marginal parameter matching in our scheme is essentially a sequence of univariate inverse probability transforms, the challenging aspects are the accuracy of dependency matching and computational efficiency at high dimensions.
To evaluate our methods in those respects, we design the following numerical experiments to first assess accuracy in match dependency parameters in bivariate simulations and then time the procedure in increasingly large dimension \(d\).

\hypertarget{bivariate-experiments}{%
\subsection{Bivariate experiments}\label{bivariate-experiments}}

We select bivariate simulation configurations to ultimately simulate discrete-valued RNA-seq data and, so, proceed in increasing complexity, leading to the model in our motivating application in Section \ref{examples}.
We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson's, Spearman's, and Kendall's ---- in identical, bivariate marginal configurations.
For each pair of identical margins, we vary the correlations across the entire possible range of values to evaluate the simulation's ability to obtain the theoretic bounds.
The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (mimicking RNA-seq counts).

Table \ref{tab:sims} lists our identical-marginal, bivariate simulation configurations.
We increase the simulate replicates \(B\) to check that our results converge to the target correlations and gauge statistical efficiency.
We select distributions beginning with a standard multivariate normal (MVN) as we expect the performance to be exact (up to MC error) for all correlation types.
Then, we select a non-symmetric continuous distribution: a standard (rate =1) two-component multivariate gamma (MVG).
Finally, we select distributions and marginal parameter values that are motivated by our RNA-seq data, namely values proximal to probabilities and sizes estimated from the data (see \href{examples}{Example applications for our motivating data} for estimation details).
Thus we arrive at a multivariate negative binomial (MVNB) \(p_1 = p_2 = \ensuremath{3\times 10^{-4}}, r_1 = r_2 = 4,\rho\) ).

\begin{longtable}[]{@{}lcr@{}}
\caption{\label{tab:sims} Identical margin, bivariate simulation configurations to evaluate correlation matching accuracy and efficiency.}\tabularnewline
\toprule
\begin{minipage}[b]{0.23\columnwidth}\raggedright
Simulation Reps (\(B\))\strut
\end{minipage} & \begin{minipage}[b]{0.28\columnwidth}\centering
Correlation Types\strut
\end{minipage} & \begin{minipage}[b]{0.40\columnwidth}\raggedleft
Identical-margin 2D distribution\strut
\end{minipage}\tabularnewline
\midrule
\endfirsthead
\toprule
\begin{minipage}[b]{0.23\columnwidth}\raggedright
Simulation Reps (\(B\))\strut
\end{minipage} & \begin{minipage}[b]{0.28\columnwidth}\centering
Correlation Types\strut
\end{minipage} & \begin{minipage}[b]{0.40\columnwidth}\raggedleft
Identical-margin 2D distribution\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.23\columnwidth}\raggedright
\(1000\)\strut
\end{minipage} & \begin{minipage}[t]{0.28\columnwidth}\centering
Pearson (\(\rho_P\))\strut
\end{minipage} & \begin{minipage}[t]{0.40\columnwidth}\raggedleft
\({ \bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho )\)\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.23\columnwidth}\raggedright
\(10,000\)\strut
\end{minipage} & \begin{minipage}[t]{0.28\columnwidth}\centering
Spearman (\(\rho_S\))\strut
\end{minipage} & \begin{minipage}[t]{0.40\columnwidth}\raggedleft
\({ \bf Y} \sim MVG( shape = 10, rate = 1, \rho )\)\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.23\columnwidth}\raggedright
\(100,000\)\strut
\end{minipage} & \begin{minipage}[t]{0.28\columnwidth}\centering
Kendall (\(\tau\))\strut
\end{minipage} & \begin{minipage}[t]{0.40\columnwidth}\raggedleft
\({ \bf Y} \sim MVNB(p = \ensuremath{3\times 10^{-4}}, r = 4,\rho)\)\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

For each of the unique 9 simulation configurations above, we estimate the correlation bounds and vary \(\rho\) along a sequence of 100 points evenly placed within the bounds (minus an adjustment factor of \(\epsilon=0.01\) to handle numeric issues when the bound is specified exactly).

\begin{table}

\caption{\label{tab:ch040-biMAEtable}Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.}
\centering
\begin{tabular}[t]{lllr}
\toprule
No. of random vectors & Correlation type & Distribution & Mean abs. error\\
\midrule
1000 & Pearson & MVN & 0.0151\\
1000 & Pearson & MVG & 0.0217\\
1000 & Pearson & MVNB & 0.0326\\
\addlinespace
1000 & Spearman & MVN & 0.0182\\
1000 & Spearman & MVG & 0.0169\\
1000 & Spearman & MVNB & 0.0159\\
\addlinespace
1000 & Kendall & MVN & 0.0111\\
1000 & Kendall & MVG & 0.0122\\
1000 & Kendall & MVNB & 0.0114\\
\addlinespace
10000 & Pearson & MVN & 0.0055\\
10000 & Pearson & MVG & 0.0120\\
10000 & Pearson & MVNB & 0.0246\\
\addlinespace
10000 & Spearman & MVN & 0.0056\\
10000 & Spearman & MVG & 0.0056\\
10000 & Spearman & MVNB & 0.0050\\
\addlinespace
10000 & Kendall & MVN & 0.0040\\
10000 & Kendall & MVG & 0.0033\\
10000 & Kendall & MVNB & 0.0031\\
\addlinespace
1e+05 & Pearson & MVN & 0.0017\\
1e+05 & Pearson & MVG & 0.0102\\
1e+05 & Pearson & MVNB & 0.0227\\
\addlinespace
1e+05 & Spearman & MVN & 0.0017\\
1e+05 & Spearman & MVG & 0.0018\\
1e+05 & Spearman & MVNB & 0.0018\\
\addlinespace
1e+05 & Kendall & MVN & 0.0012\\
1e+05 & Kendall & MVG & 0.0011\\
1e+05 & Kendall & MVNB & 0.0010\\
\bottomrule
\end{tabular}
\end{table}

Figure \ref{fig:ch040-bPlot} displays the aggregated bivariate simulation results.
Table \ref{tab:ch040-biMAEtable} contains the mean absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.
Taken together, the studies show our methodology is generally accuracy across the entire range of possible correlation values for the rank-based dependency measures, at least in these limited simulation settings for the rank-based correlations.
For the two non-normal bivariate marginals, the Pearson correlation matching is approximate.
For discrete margins, matching the dependency measures was somewhat less accurate, even for the rank-based metrics, and particularly inaccurate near the lower bound of Pearson correlations.

\begin{figure}

{\centering \includegraphics[width=0.85\linewidth]{article_bigsimr_files/figure-latex/ch040-bPlot-1}

}

\caption{Bivariate simulations match specified correlations. The horizontal axis plots the specified target correlations across the entire range of possible correlations for each bivariate margin. Normal margins are plotted in dark blue, gamma in medium blue, and negative binomial in light blue. As the number of simulated vectors $B$ increases from left to right, the variation in estimated correlations (vertical axis) decreases. The dashed line indicates equality between the specified and estimated correlations. Only the normal margins match the Pearson correlations (top row) exactly across the entire range of correlations. The two non-normal bivariate random vectors experience attenuation (downward bias) for extreme Pearson correlations (due to limitations in our algorithm) and more restriction (due to the Frechet bounds). The rank-based correlations (bottom two rows) are matched exactly for all margins and obtain the full range of possible dependencies.}\label{fig:ch040-bPlot}
\end{figure}

The accuracy appears adequate for many applications.
In practice, we recommend users to always evaluate the accuracy for their application, using methods similar to those presented above.
See \href{discussion}{Discussion} for future directions for fast Pearson matching and discrete-specific modifications.

\hypertarget{scale-up-to-high-dimensions}{%
\subsection{Scale up to High Dimensions}\label{scale-up-to-high-dimensions}}

With information of our method's accuracy from a low-dimensional perspective, we now turn to assessing whether the \texttt{bigsimr} can scale to larger dimensional problems.
Specifically, we seek evidence to determine whether the method can scale to higher dimensions in a practical time.
Using our motivating RNA-seq data, described in \href{background}{Background}, we filtered the original 20,501 genes to the high-expressing genes at increasing percentiles, \(1, 5, 10, 15, 20, 25\%\), to obtain \(d=\{206, 1026, 2051, 3076, 4101, 5127\}\) marginals and \(\binom{d}{2}\) pairwise correlations at each setting.
For example, for \(d=5127\) there are \(13,140,501\) correlation coefficients.
We estimated the marginal negative binomial parameters and the correlation coefficients from the RNA-seq data to seed our simulations.
(See \href{examples}{Example applications for our motivating data} for a detailed description of estimation).

Figure \ref{fig:ch040-gpuVScpuFig} displays computation times using various high-performance settings (1 central processing unit, CPU-1, versus twenty CPUs, CPU-20; with and without GPU acceleration, GPU-1; GPU-20) to produce \(B=10,000\) random vectors.
The Pearson simulations are much faster since the correlation conversion steps are avoided (pre-processing step; see \href{algorithms}{Algorithms}), but as we know from above the accuracy will suffer slightly, especially near the negative boundary of the possible correlations.
Matching Spearman's correlation at larger \(d\) gets costly if one wants to produce \(B=10,000\) random vectors at many different simulation settings, but the conversion steps need only be computed once (see \href{MC-evaluation-of-correlation-estimation-efficiency}{MC evaluation of correlation estimation efficiency} for an example of this strategy).

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{fig/cpu-gpu-times}

}

\caption{Computation times as d increases. We filter to the top 1, 5, 10, 15, 20, 25\% expressing genes (in terms of median expression.)}\label{fig:ch040-gpuVScpuFig}
\end{figure}

\hypertarget{examples}{%
\section{RNA-seq data example applications}\label{examples}}

This section demonstrates how to simulate multivariate data using \texttt{bigsimr}, aiming to replicate the structure of high-dimensional dependent count data.
In an illustration of our proposed methodology applied to real data, we seek to simulate RNA-sequencing data by producing simulated random vectors with assumed marginal distributions with estimated parameters that mimic the observed data and its generating process.
We seek scaleable realistic multivariate simulation to enable Monte Carlo (MC) methods for these data.
Modeling RNA-seq using multivariate probabilities distributions is natural as inter-gene correlation is an inherent part of biological processes.
Yet many models do not account for this, leading to major disruptions to the operating characteristics of statistical estimation, testing, and prediction.
See \citet{Efron2012} for a detailed discussion with related methods and see \citet{Wu2012b}, \citet{Schissler2018}, \citet{Schissler2019} for applied examples.
The following subsections apply \texttt{bigsimr}'s methods to real RNA-seq data, including replicated an estimated parametric structure, MC probability estimation, and MC evaluation of correlation estimation efficiency.

\hypertarget{simulating-high-dimensional-rna-seq-data}{%
\subsection{Simulating High-Dimensional RNA-seq data}\label{simulating-high-dimensional-rna-seq-data}}

Our first goal is to replicate the structure of the TCGA BRCA RNA-seq data set (see \href{background}{Background}).
Ultimately, we will simulate \(B=10,000\) random vectors \({\bf Y}=(Y_1, \ldots, Y_d)^\top\) with \(d=1026\).
Often researchers posit a negative binomial (NB) model as RNA-seq counts are often over-dispersed that a Poisson model would suggest.
All \(d\) selected genes exhibit over-dispersion (data not shown) and, so, we proceed to estimate the NB parameters \((r_i, p_i), i=1,\ldots,d\) to determine the target marginal PMFs \(f_i\) (via method of moments).
To complete specification of the simulation algorithm inputs, we'll estimate the Spearman correlation matrix \({ \bf R}_{Spearman}\) to characterize dependency.

With our goal in mind, we first estimate the desired correlation matrix using the fast implementation provided by \texttt{bigsimr}:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Estimate Spearman's correlation on the count data}
\NormalTok{corType <-}\StringTok{ 'spearman'}
\KeywordTok{system.time}\NormalTok{( nb_Rho <-}\StringTok{ }\NormalTok{bigsimr}\OperatorTok{::}\KeywordTok{cor_fast}\NormalTok{( brca, }\DataTypeTok{method =}\NormalTok{ corType ) )}
\NormalTok{   user  system elapsed }
  \FloatTok{0.621}   \FloatTok{0.000}   \FloatTok{0.622} 
\end{Highlighting}
\end{Shaded}

Next, we estimate the marginal parameters.
We use method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model.
The marginal distributions are from the same probability family (NB) yet are heterogeneous in terms of the parameters probability and size \((p_i, n_i)\) for \(i,\ldots,d\).
The functions below help achieve this estimation and specifying the inputs for use in \texttt{bigsimr::rvec}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{make_nbinom_alist <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(sizes, probs) \{}
  \KeywordTok{lapply}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\KeywordTok{length}\NormalTok{(sizes), }\ControlFlowTok{function}\NormalTok{(i) \{}
    \KeywordTok{substitute}\NormalTok{(}\KeywordTok{qnbinom}\NormalTok{(}\DataTypeTok{size =}\NormalTok{ s, }\DataTypeTok{prob =}\NormalTok{ p), }
               \KeywordTok{list}\NormalTok{(}\DataTypeTok{s =}\NormalTok{ sizes[i], }\DataTypeTok{p =}\NormalTok{ probs[i]))}
\NormalTok{  \})}
\NormalTok{\}}
\CommentTok{## make_nbinom_alist(c(20, 21, 22), c(0.3, 0.4, 0.5))}
\NormalTok{nbinom_mom <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x) \{}
\NormalTok{  m <-}\StringTok{ }\KeywordTok{mean}\NormalTok{(x)}
\NormalTok{  s <-}\StringTok{ }\KeywordTok{sd}\NormalTok{(x)}
\NormalTok{  s2 <-}\StringTok{ }\NormalTok{s}\OperatorTok{^}\DecValTok{2}
\NormalTok{  p <-}\StringTok{ }\NormalTok{m}\OperatorTok{/}\NormalTok{s2}
\NormalTok{  r <-}\StringTok{ }\NormalTok{m}\OperatorTok{^}\DecValTok{2} \OperatorTok{/}\StringTok{ }\NormalTok{(s2 }\OperatorTok{-}\StringTok{ }\NormalTok{m)}
  \KeywordTok{c}\NormalTok{(r, p)}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

Now we estimate the marginal parameters using the built-in \texttt{apply}function:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{sizes <-}\StringTok{ }\KeywordTok{apply}\NormalTok{( }\KeywordTok{unname}\NormalTok{(}\KeywordTok{as.matrix}\NormalTok{(brca)), }\DecValTok{2}\NormalTok{, nbinom_mom )[}\DecValTok{1}\NormalTok{, ]}
\NormalTok{probs <-}\StringTok{ }\KeywordTok{apply}\NormalTok{( }\KeywordTok{unname}\NormalTok{(}\KeywordTok{as.matrix}\NormalTok{(brca)), }\DecValTok{2}\NormalTok{, nbinom_mom )[}\DecValTok{2}\NormalTok{, ]}
\end{Highlighting}
\end{Shaded}

Notably, the marginal NB probabilities \(\hat{p}_i's\) are small --- ranging in \([\ensuremath{3.9342\times 10^{-6}} , 0.0122]\).
This gives rise to highly variable counts and, typically, less restriction on potential pairwise correlation pairs.
Once the functions are defined/executed to complete marginal estimation, we specify targets and generate the desired random vectors using \texttt{rvec}:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Set the number of random vectors}
\NormalTok{n <-}\StringTok{ }\DecValTok{10000}
\CommentTok{## construct margins}
\NormalTok{nb_margins <-}\StringTok{ }\KeywordTok{make_nbinom_alist}\NormalTok{(sizes, probs)}
\CommentTok{## run sims}
\NormalTok{sim_nbinom <-}\StringTok{ }\KeywordTok{rvec}\NormalTok{(n, nb_Rho, nb_margins, }\DataTypeTok{type =}\NormalTok{ corType,}
                   \DataTypeTok{ensure_PSD =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{cores =}\NormalTok{ cores) }
\KeywordTok{colnames}\NormalTok{(sim_nbinom) <-}\StringTok{ }\KeywordTok{names}\NormalTok{(brca)}
\end{Highlighting}
\end{Shaded}

Figure \ref{fig:ch050-simDataFig} displays the simulated counts and pairwise relationships for our example genes in Table \ref{tab:ch010-realDataTab}.
Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parameter form and with less extreme points then the observed data (c.f. Figure \ref{fig:ch010-realDataFig}.)

\begin{figure}

{\centering \includegraphics[width=0.85\linewidth]{fig/ch050-simDataFig}

}

\caption{Simulated data for three selected high-expressing genes replicates the estimated data structure.}\label{fig:ch050-simDataFig}
\end{figure}

Figure \ref{fig:ch050-figBRCA} displays the aggregated results of our simulation by comparing the specified target parameter (horizontal axes) with the corresponding quantities estimated from the simulated data (vertical axes).
The evaluation shows that the simulated counts approximately match the target parameters and exhibit the full range of estimated correlation from the data.
Utilizing 15 CPU threads in a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, the simulation completed just over of 2 minutes.

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{fig/ch050-figBRCA}

}

\caption{Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set. The dashed red lines indicated equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes).}\label{fig:ch050-figBRCA}
\end{figure}

\hypertarget{simulation-based-joint-probability-calculations}{%
\subsection{Simulation-based joint probability calculations}\label{simulation-based-joint-probability-calculations}}

To conduct statistical inference a critical task is to evaluate the joint probability mass (or density) function:

\[
P( {\bf Y} = {\bf y} ), y_i \in \chi_i.
\]

where \(\chi_i\) is the sample space for the \(i^{th}\) component of the random vector \(\bf{Y}\).
Compact representations with convenient computational forms are rare for high-dimensional constructions, especially with heterogeneous, correlated marginal distributions (or margins of mixed data types).
Given a large number simulated vectors as produced above, estimated probabilities are readily given by counting the proportion of simulated vectors meeting the desired condition.
In our motivating application, one may ask what is the probability that all genes expressed greater than a certain threshold value \({ \bf y}_0\).

Then we estimate

\[
\hat{P}( {\bf Y} >= {\bf y_0 } ) = \sum_{b=1}^B I( {\bf Y^{(b) }} > {\bf y_0 } ) / B
\]

where \({\bf Y^{(b)} }\) is the \(b^{th}\) simulated vector in a total of \(B\) simulation replicates and \(I ( )\) is the indicator function.
For example, we can estimate from our \(B=10,000\) simulated vectors the probability that all genes are expressed (i.e., \({\bf y}_i \geq 1, \forall \; i\)) is \(0.1708\).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{d <-}\StringTok{ }\KeywordTok{ncol}\NormalTok{(sim_nbinom)}
\NormalTok{B <-}\StringTok{ }\KeywordTok{nrow}\NormalTok{(sim_nbinom)}
\NormalTok{threshold <-}\StringTok{ }\KeywordTok{rep}\NormalTok{( }\DecValTok{1}\NormalTok{, d)}
\KeywordTok{mean}\NormalTok{(}\KeywordTok{apply}\NormalTok{( sim_nbinom,  }\DecValTok{1}\NormalTok{,}
           \ControlFlowTok{function}\NormalTok{(X, }\DataTypeTok{y0=}\NormalTok{threshold) \{}
               \KeywordTok{all}\NormalTok{( X }\OperatorTok{>}\StringTok{ }\NormalTok{y0) \}}
\NormalTok{           ))}
\NormalTok{[}\DecValTok{1}\NormalTok{] }\FloatTok{0.1708}
\end{Highlighting}
\end{Shaded}

\hypertarget{evaluation-of-correlation-estimation-efficiency}{%
\subsection{Evaluation of correlation estimation efficiency}\label{evaluation-of-correlation-estimation-efficiency}}

MC methods are routinely used in many statistical inferential tasks including estimation, hypothesis testing, error rates, and empirical interval coverage rates.
For an concise introduction to these methods, see \citet{Rizzo2007}, Ch. 6.
To conclude the example applications, we demonstrate how \texttt{bigsimr} can be used evaluate estimation efficiency.
In particular, we'd like to assess the error in our correlation estimation above.
We used a conventional method, based on classical statistical theory.
Yet this method was not designed for high-dimensional data.
Indeed, high-dimensional covariance estimation (and precision matrices) is an active area of statistical science (see, for example, \citep{Won2013g, VanWieringen2016}.

In this small example, we simulate \(m=10\) data sets with the number of simulated vectors matching the number of patients in the BRCA data set, \(N=1212\).
Since our simulation is much faster for the Pearson correlation type (see Figure \ref{fig:ch040-gpuVScpuFig}), we only convert the Spearman correlation matrix once (and ensure its PSD).
At each iteration, we estimate the quadratic loss from the specified \({\bf R}_{Spearman}\), producing a distribution of loss values.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Simulate random vectors equal to the sample size}
\NormalTok{n <-}\StringTok{ }\KeywordTok{nrow}\NormalTok{(brca)}
\CommentTok{## convert outside for faster simulation}
\NormalTok{nb_Rho_p <-}\StringTok{ }\NormalTok{bigsimr}\OperatorTok{::}\KeywordTok{cor_convert}\NormalTok{( }\DataTypeTok{rho =}\NormalTok{ nb_Rho,}
                                 \DataTypeTok{from =}\NormalTok{ corType, }\DataTypeTok{to =} \StringTok{"pearson"}\NormalTok{ )}
\CommentTok{## ensure PSD}
\NormalTok{nb_Rho_p <-}\StringTok{ }\NormalTok{bigsimr}\OperatorTok{::}\KeywordTok{cor_nearPSD}\NormalTok{( }\DataTypeTok{G =}\NormalTok{ nb_Rho_p )}
\CommentTok{## create m random vectors and estimate correlation}
\NormalTok{simRho <-}\StringTok{ }\KeywordTok{replicate}\NormalTok{(}\DataTypeTok{n =}\NormalTok{ m,}
  \DataTypeTok{expr =}\NormalTok{ \{ tmpSim <-}\StringTok{ }\KeywordTok{rvec}\NormalTok{(}\DataTypeTok{n =}\NormalTok{ n , nb_Rho_p, nb_margins,}
    \DataTypeTok{type =} \StringTok{'pearson'}\NormalTok{, }\DataTypeTok{ensure_PSD =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{cores =}\NormalTok{ cores);}
\NormalTok{    bigsimr}\OperatorTok{::}\KeywordTok{cor_fast}\NormalTok{( }\DataTypeTok{x =}\NormalTok{ tmpSim, }\DataTypeTok{method =}\NormalTok{ corType )\} ,}
\DataTypeTok{simplify =} \OtherTok{FALSE}\NormalTok{)}
\CommentTok{## find quadratic loss at each rep}
\NormalTok{quadLoss <-}\StringTok{ }\KeywordTok{unlist}\NormalTok{( }\KeywordTok{lapply}\NormalTok{( simRho, rags2ridges}\OperatorTok{::}\NormalTok{loss,}
                           \DataTypeTok{T =}\NormalTok{ nb_Rho, }\DataTypeTok{type =} \StringTok{"quadratic"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

The \texttt{R} summary function supplies the mean-augmented five-number summary of the quadratic loss distribution computed above.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{summary}\NormalTok{(quadLoss)}
\NormalTok{   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. }
\DecValTok{1225076} \DecValTok{1251629} \DecValTok{1277417} \DecValTok{1321153} \DecValTok{1390574} \DecValTok{1461040} 
\end{Highlighting}
\end{Shaded}

This distribution could be compared to high-dimensional designed covariance estimators to guide to help decide whether the additional complexity and computation time are warranted.

\hypertarget{discussion}{%
\section{Conclusion and discussion}\label{discussion}}

We've introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a high-performance \texttt{R} package called \href{https://schisslergroup.github.io/bigsimr/}{\texttt{bigsimr}}.
The random vector generation method is largely inspired by NORTA (\citet{Cario1997}) and Gaussian copula-based approaches (\citet{MB13}, \citet{BF17}, \citet{Xia17}).
The major contributions of this work are high-dimensional scaleability, flexible modeling of dependency, and an high-performance implementation with broad potential data analytic applications for modern, big-data statistical computing.

It is customary to compare new tools and algorithms directly to existing competing methods and software.
In this study, however, we only employ our proposed methodology, since our previous work has show that existing tools are simply not designed or feasible to meet our high-dimensional goal (see \citet{Li2019gpu} for evaluations of the \texttt{R} \texttt{copula}package and others).
For the bivariate simulations, existing packages such as \href{https://github.com/cran/NORTARA/blob/master/inst/doc/NORTARA.R}{\texttt{nortaRA}} work well to match Pearson correlations exactly.

There are limitations to the methodology and implementation.
The most obvious missing feature of the proposed methodology is the inability to match a Pearson correlation matrix exactly.
As discussed in \href{algorithms}{Algorithms} and extensively by \citet{XZ19}, this is a computationally intense procedure and Pearson's correlation is not a natural choice to describe dependency for non-normal marginals.
While we do not provide an implementation directly supporting Pearson matching, users may supply their own input Pearson correlation after using a supplementary matching scheme \citep{Cario1997, XZ19}.

Future work includes developing scaleable algorithms to match the Pearson correlation matrix more precisely, discrete-margin specific modifications including fast Spearman's correlation rescaling (see Equation \eqref{eq:spearmanRescaled}), and high-dimensional covariance estimation.
From an implementation standpoint, \texttt{bigsimr} only supports Nvidia GPUs and redesigning the code using OpenCL would broaden the users who would benefit.
As data-analytic problems grow to even larger dimension, multi-GPU support is a promising hardware-based future direction.

\hypertarget{misc}{%
\section*{Supplementary Materials}\label{misc}}
\addcontentsline{toc}{section}{Supplementary Materials}

We provide an open-source implementation of our methodology as the \texttt{bigsimr} R package, hosted on github, \url{https://schisslergroup.github.io/bigsimr/}.

\begin{center}\includegraphics[width=0.05\linewidth]{images/hex-bigsimr} \end{center}

\hypertarget{acknowledgments}{%
\section*{Acknowledgment(s)}\label{acknowledgments}}
\addcontentsline{toc}{section}{Acknowledgment(s)}

The authors gratefully acknowledge the helpful discussions with University of Arizona's Professor Walter W. Piegorsch and Professor Edward J. Bedrick during this project's conception.
We also gratefully acknowledge Heather Knudson's graphic design for the \texttt{bigsimr} R Package.
The results published here are in whole or part based upon data generated by the TCGA Research Network: \url{https://www.cancer.gov/tcga}.

\hypertarget{coi}{%
\section*{Disclosure statement}\label{coi}}
\addcontentsline{toc}{section}{Disclosure statement}

The authors report no conflict of interest.

\hypertarget{funding}{%
\section*{Funding}\label{funding}}
\addcontentsline{toc}{section}{Funding}

Research reported in this publication was supported by MW-CTR-IN of the National Institutes of Health under award number NIH 1U54GM104944.

\bibliography{bigsimr.bib,packages.bib}

\end{document}

### 2020-12-15 index.Rmd
4th draft. Address WWP abstract/introduction

@@
 bibliography: ["bigsimr.bib", "packages.bib"]
-abstract:  It is critical to realistically simulate data when conducting Monte Carlo studies and methods.
-But measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments.
+abstract:  It is critical to realistically simulate data when conducting Monte Carlo studies and evaluating quantitative methods.
+But measurements are often correlated and high dimensional in this era of big data, however, such as data obtained through high-throughput biomedical experiments.
 Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers may resort to simulation designs that posit independence or perform arbitrary data transformations.
@@
 This article introduces the `bigsimr` R package that provides a flexible, scaleable procedure to simulate high-dimensional random vectors with given marginal characteristics and dependency measures.
-We'll describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and find close positive semi-definite matrices.
+We describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and find close positive semi-definite matrices.
 Finally, we demonstrate the power of `bigsimr` by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d =1026$.
@@
 Massive high-dimensional data sets are now commonplace in many areas of scientific inquiry.
-As new methods are developed for these data, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness --- often in comparison to existing methods.
-Further, efficient simulation empowers statistical computing strategies, such as the parametric bootstrap [@Rizzo2007] to simulate from a hypothesized null model, providing inference in analytically challenging settings.
-Such Monte Carlo (MC) techniques become difficult for high-dimensional data with the current existing algorithms and tools.
+As new methods are developed for data, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of analyzing such proposed methodology --- such as false positive rates, statistical power, interval coverage, and robustness --- often with comparison to existing methods.
+Further, efficient simulation empowers statistical computing strategies, such as the parametric bootstrap [@Chernick2008] to simulate from a hypothesized null model, providing inference in analytically challenging settings.
+Such Monte Carlo (MC) techniques become difficult for high-dimensional data using existing algorithms and tools.
 This is particularly true when simulating massive *multivariate*, *non-normal* distributions, arising naturally in many fields of study.
 
-As others have noted, it can be vexing to simulate dependent, non-normal/discrete data ---  even for low dimensional settings [@MB13; @XZ19].
-For continuous non--normal multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula [@Nelsen2007] approaches are well-studied with flexible, robust software available [@Yan2007; @Chen2001].
+As many have noted, it can be vexing to simulate dependent, non-normal/discrete data ---  even for low dimensional settings [@MB13; @XZ19].
+For continuous non--normal multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula approaches [@Nelsen2007] are well-studied, with flexible, robust software available [@Yan2007; @Chen2001].
 Yet these approaches do not scale in a timely fashion to high-dimensional problems [@Li2019gpu]. 
-For discrete data, early simulation strategies had major flaws --- such as failing to obtain the full range of possible dependencies (e.g., admitting only positive correlations @Park1996). 
+For discrete data, early simulation strategies had major flaws --- such as failing to obtain the full range of possible dependencies (e.g., admitting only positive correlations: see @Park1996). 
 While more recent approaches [@MB13; @Xia17; @BF17] have largely remedied this issue for low-dimensional problems, the existing tools are not designed to scale to high dimensions.
 
-Another central issue lies characterizing dependency between components in the high-dimensional random vector.
-The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g. non-normal, discrete, infinite support, etc). 
+Another central issue lies in characterizing dependency between components in the high-dimensional random vector.
+The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc). 
 <!-- Here we propose a scheme that ignores the analytic goal --- and, instead, aims to match the empirically-derived estimated correlation data and marginal characteristics.  -->
@@
 As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix exactly is computationally intense [@Chen2001; @Xia17]. 
-On the other hand, an analyst may consider the use of nonparametric correlation measures to better characterize monotone, non-linear dependency, such as Spearman's $\rho$ and Kendall's $\tau$.
-Throughout, we'll emphasize matching these nonparametric dependency measures as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology.
+On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependency, such as Spearman's $\rho$ and Kendall's $\tau$.
+Throughout, we focus on matching these nonparametric dependency measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology.
 
@@
 The crux of the method lies in the construction of a Gaussian copula, in the spirit of the NORTA procedure.
-Further, we introduce the `bigsimr` R package that provides parallelized, high-performance software implemented our NORTA-inspired algorithm.
-The algorithm design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed form relationships between dependency measures for the multivariate normal (MVN) distribution.
+Further, we introduce the `bigsimr` R package that provides parallelized, high-performance software implemented in our NORTA-inspired algorithm.
+The algorithm design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed-form relationships between dependency measures for the multivariate normal (MVN) distribution.
 
-The study proceeds by providing background information, including a description of our motivating example application --- RNA-sequencing (RNA-seq) breast cancer data.
+Our study proceeds by providing background information, including a description of our motivating example application --- RNA-sequencing (RNA-seq) breast cancer data.
 Then we describe and justify our simulation methodology and related algorithms.
@@
 Then we proceed with Monte Carlo studies under various distributional assumptions to assess accuracy and scaleable.
-After the MC evaluations, we revisit our high-dimensional motivating example and employ our methods to commonplace statistical computing tasks.
-Finally, we'll discuss the method's utility, limitations, and future directions.
+After the MC evaluations, we revisit our high-dimensional motivating example and employ our methods in commonplace statistical computing tasks.
+Finally, we discuss the method's utility, limitations, and future directions.

### 2020-12-15 010-background.Rmd
4th draft. Address WWP background/ch010-realDataTab

@@
 
-The `bigsimr` `R` package provides multiple algorithms to work with high-dimensional multivariate data, but all these algorithms were originally designed to support a single task:
-to generate random vectors drawn from multivariate probability distribution with given marginal distributions and dependency metrics.
-Specifically, our goal is to efficiently simulate a large number, $B$, of random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with **correlated** components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i$, where $d$ can be very large and still be computed in practically useful times.
+The `bigsimr` `R` package presented here provides multiple algorithms that operate with high-dimensional multivariate data; however, all these algorithms were originally designed to support a single task:
+to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics.
+Specifically, our goal is to efficiently simulate a large number, $B$, of random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with *correlated* components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i$, where $d$ can be very large and still be computed in practically useful timescale.
 
@@
 
-To fix ideas and provide examples applications enabled via `bigsimr`, the next section describes a motivating data set that originally inspired the authors' interest in developing this methodology.
+To fix ideas and provide example applications enabled via `bigsimr`, the next section describes a motivating data set.
 
@@
 Simulating high-dimensional, non-normal, correlated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
-The RNA-seq data-generating process involves counting how often a particular messenger RNA (mRNA) is expressed in a biological sample.
+The RNA-seq data-generating process involves counting how often a particular form of human messenger RNA (mRNA) is expressed in a biological sample.
 RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data.
-For human derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted.
+For human-derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted [@Schissler2019].
 Importantly, due to inherent biological processes, gene expression data exhibits correlation --- co-expression --- across genes [@BE07; @Schissler2018].
@@
 brcaRDS <- paste0( "data/brca", round(myProb*100), ".rds" )
-if ( !file.exists( brcaRDS ) )  {
+if ( !file.exists( brcaRDS ) | !file.exists( './data/examplePatients.txt' ))  {
     ## full workflow simulating RNA-seq data
@@
     lastClinical <- which( names(allDat) == 'tumorsize' )
-    brca <- allDat[ allDat$disease == "BRCA", (lastClinical+1):ncol(allDat) ]
+    brca <- allDat[ allDat$disease == "BRCA", ]
+    ## Patient ID
+    writeLines( as.character( brca$patient ), './data/examplePatients.txt' ) 
+    brca <- brca[ , (lastClinical+1):ncol(allDat) ]
     dim(brca) ## num of genes 20501
@@
 We further process the data to illustrate our methodology's flexible and robust simulation scheme, while better aligning with the actual data-generating process, by rounding @Li2011c RNA-seq by Expectation Maximization (RSEM) values to counts.
-Table \@ref(tab:ch010-realDataTab) displays counts for three selected high-expressing genes for the first five patients' breast tumor samples.
+Table \@ref(tab:ch010-realDataTab) displays RNA-seq counts for three selected high-expressing genes for the first five patients' breast tumor samples.
 
@@
 exampleGenes <- readRDS('./data/exampleGenes.rds')
-small <-  brca %>% select( exampleGenes )
-knitr::kable(round(head( small, n = 5)),
+## Add row names per WWP comment
+small <-  brca %>% select( exampleGenes ) %>% as.data.frame
+numPatients <- 5
+smallForTable <- head( small, n = numPatients )
+smallForTable$PatientID <- head( readLines('./data/examplePatients.txt'), n = numPatients )
+smallForTable <- smallForTable[ , c(4, 1:3 ) ]
+smallForTable[ , 2:4 ] <- round( smallForTable[ , 2:4 ] )
+knitr::kable(smallForTable ,
       booktabs = TRUE,
-      caption = 'mRNA counts for three selected high-expressing genes from the first five observations of the BRCA data set.'
+      caption = "mRNA express for three selected high-expressing genes (*RPL5*, *TKNIP*, and *VIM*) for the first five patients in the TCGA BRCA data set."
 )

### 2020-12-16 010-background.Rmd
4th draft. Address WWP background part 2

@@
 
-We'll illustrate our methodology using the well-studied Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments).
+We illustrate our methodology using a well-studied Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments).
 For simplicity, we only consider high expressing genes.
-In turn, we begin by filtering to retain the top `r (1-myProb)*100`% highest- expressing genes (in terms of median expression) of the 20,501 gene measurements from $N=`r nrow(brca)`$ patients' tumor samples, resulting in $d=`r d`$ genes.
+In turn, we begin by filtering to retain the top `r (1-myProb)*100`% of highest- expressing genes (in terms of median expression) of the 20,501 gene measurements from $N=`r nrow(brca)`$ patients' tumor samples, resulting in $d=`r d`$ genes.
 This gives a massive number of pairwise dependencies among the marginals (specifically, $`r choose(d,2)`$ correlation parameters).
@@
       booktabs = TRUE,
-      caption = "mRNA express for three selected high-expressing genes (*RPL5*, *TKNIP*, and *VIM*) for the first five patients in the TCGA BRCA data set."
+      caption = "mRNA expression for three selected high-expressing genes (*RPL5*, *TKNIP*, and *VIM*) for the first five patients in the TCGA BRCA data set."
 )
@@
 
-```{r ch010-realDataFig, cache=F, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes.
+```{r ch010-realDataFig, cache=F, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes from Table 1.
 The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations.
@@
 As @MB13 and @MK01 discuss, for a bivariate normal $(X,Y)$ random vector, the Pearson correlation completely describes the dependency between the components. 
-For non-normal marginals with monotone correlation patterns, $\rho_P$ suffers some drawbacks and may mislead or fail to capture important relationships (@MK01).
+For non-normal marginals with monotone correlation patterns, however, $\rho_P$ suffers some drawbacks and may mislead or fail to capture important relationships [@MK01].
 Alternatively in these settings, analysts often prefer rank-based correlation measures to describe the degree of monotonic association.
@@
 *Correspondence among Pearson, Spearman, $\tau$ correlations*.
-This is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficient, as the marginal distributions $F_i$ are intrinsic in their calculation.
-But for **bivariate normal vectors**, however, the correspondence is well-known:
+There is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficient, as the marginal distributions $F_i$ are intrinsic in their calculation.
+For *bivariate normal vectors*, however, the correspondence is well-known:
 
@@
 
-These facts are also critical in our simulation algorithm to broaden the dependency measures supported by `bigsimr`, in a computationally effective manner.
+These features are critical in our simulation algorithm to broaden the dependency measures supported by `bigsimr`, in a computationally effective manner.
 
 *Discrete marginal considerations*.
-Spearman's correlation for discrete marginal suffers some issues due to the nonzero probability of ties (for example, the Spearman's correlation of a discrete-valued random variable $X$ with itself could be less than 1; @MB13).
+Spearman's correlation for discrete marginal suffers issues whenever there is a nonzero probability of ties (for example, the Spearman's correlation of a discrete-valued random variable $X$ with itself could be less than 1; @MB13).
 One remedy for this issue is to rescale Equation \@ref(eq:spearman). 
@@
 For discrete marginals with large or infinite support, computing the adjustment factors $\sum_x p(x)^3$, $\sum_y p(y)^3$ over all large number of pairs becomes expensive (often violating desired property SP1).
-And further the infinite sums must be approximately for count-valued data and potentially violating desired property SP2.
+Further the infinite sums must be approximated for count-valued data.
 
@@
 *Marginal-dependent bivariate correlation bounds*. 
-Given two marginal distributions, $\rho_P$ is not free to vary the entire range of possible correlations $[-1,1]$.
-The so-called *Frechet-Hoeffding bounds* are well-studied (for example, see @Nelsen2007 and @BF17).
-This situation gives strict restraints on the possible correlations and cannot be overcome through algorithm design.
+Given two marginal distributions, $\rho_P$ is not free to vary over the entire range of possible correlations $[-1,1]$.
+The so-called *Frechet-Hoeffding bounds* are well-studied (see @Nelsen2007 and @BF17).
+This gives strict restraints on the possible correlations and cannot be overcome through algorithm design.
 In general, the bounds are given by
@@
 
-\noindent where $U$ is a uniform random variable in $(0,1)$, and $F^{-1}_1, F^{-1}_2$ are the inverse CDF of random variables $X_1$ and $X_2$, respectively. 
+\noindent where $U$ is a uniform random variable in $(0,1)$, and $F^{-1}_1, F^{-1}_2$ are the inverse CDFs of random variables $X_1$ and $X_2$, respectively. 
 For discrete random variables, define $F^{-1}$ as in Equation \@ref(eq:inverseCDF).
@@
 A copula is a distribution function on $[0,1]^d$ that describes a multivariate probability distribution with standard uniform marginals.
-This definition provides a powerful, natural way characterize joint probability structure.
+This provides a powerful, natural way to characterize joint probability structures.
 Consequently, the study of copulas is an important and active area of statistical theory and practice.
@@
 For any random vector ${\bf X}=(X_1, \ldots, X_d)$ with CDF $F$ and marginal CDFs $F_i$ there is a copula function 
-$C(u_1, \ldots, u_d)$ so that
+$C(u_1, \ldots, u_d)$ satisfying
 
@@
 
-A Gaussian copula is the case where all marginal CDFs $F_i$ are the standard normal CDF, $\Phi$. 
+A Gaussian copula has marginal CDFs as the normal CDF, $\Phi$. 
 This representation corresponds to a multivariate normal distribution with standard normal marginal distributions and covariance matrix ${\bf R_P}$.
-But since the marginals are standardized to have unit variance, this ${\bf R_P}$ is a Pearson correlation matrix. 
+As the marginals are standardized to have unit variance, however, ${\bf R_P}$ is a Pearson correlation matrix. 
 If $F_{{\bf R}}$ is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula $C_{{\bf R}}$ is defined through

### 2020-12-16 020-simulation-algorithm.Rmd
4th draft. Address WWP ch020

@@
 The well-known NORTA algorithm [@Cario1997] can be used simulate a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$.
-Specifically, the NORTA algorithm follows like this:
+Specifically, NORTA algorithm proceedings as follows:
 
@@
 2. Determine the input matrix $\Sigma_{\bf Z}$ that corresponds with the specified output $\Sigma_{\bf Y}$.
-3. Produce a Cholesky factor $M$ of $\Sigma_{\bf Z}$ so that $M M^{\prime}=\Sigma_{\bf Z}$.
+3. Produce a Cholesky factor $M$ of $\Sigma_{\bf Z}$ such that $M M^{\prime}=\Sigma_{\bf Z}$.
 4. Set $X$ by $X \gets MZ$.
@@
 
-With modern parallelized computing, steps 1, 3, 4, 5 are readily implemented as high-performance, multi-core and/or graphical-processing-unit (GPU) accelerated, algorithms --- providing the fast scaleability using readily-available hardware.
+With modern parallelized computing, steps 1, 3, 4, 5 are readily implemented as high-performance, multi-core and/or graphical-processing-unit (GPU) accelerated, algorithms --- providing fast scaleability using readily-available hardware.
 
 Matching specified Pearson correlation coefficients exactly (step 2 above), however, is problematic.
-In general, there is no closed form correspondence between the components of the input $\Sigma_{\bf Z}$ and target $\Sigma_{\bf Y}$.
+In general, there is no closed-form correspondence between the components of the input $\Sigma_{\bf Z}$ and target $\Sigma_{\bf Y}$.
 Matching the correlations involves evaluating or approximating $\binom{d}{2}$ integrals of the form $EY_iY_j = \int \int y_i y_j f_{X}(F_i^{-1}(\Phi(z_i)), F_j^{-1}(\Phi(z_j))dy_idy_j$, for $i,j=1,2,\ldots,d, \; i \neq j$.
 For high-dimensional data, these evaluations are often too costly to enable feasible simulation studies.
-For low-dimensional problems, methods and tools exist to match Pearson correlations precisely, see @Chen2001; @Xia17; @MB13 and the publicly available `nortaRA` R package.
+For low-dimensional problems, however, methods and tools exist to match Pearson correlations precisely, see @Chen2001; @Xia17; @MB13 and the publicly available `nortaRA` R package.
 
 To maintain scaleability (SP1), our solution is to essentially avoid this complication in Pearson matching.
-Since our goal is to simulate non-normal marginals, we greatly prefer the use of rank-based measures $\rho_S$ and $\tau$ from a modeling standpoint.
+Since our goal is to simulate non-normal marginals, we greatly prefer the use of rank-based measures, such as $\rho_S$ and $\tau$, from a modeling standpoint.
 Further, $\rho_S$ and $\tau$'s invariance under monotone transformation (see [Background](#background)), preserves the correlation coefficients through steps 3, 4, and 5 in the NORTA algorithm above.
-This eliminates the need for computing the $\binom{d}{2}$ integrals to match exactly (nothing is for free, however, as discussed in below in Section \@ref(rand-vec-gen)).
+This eliminates the need for computing the $\binom{d}{2}$ integrals to match exactly (nothing is for free, however, as discussed below in Section \@ref(rand-vec-gen)).
 
-Despite all this, if one does desire to characterize dependency using Pearson correlations, simply using the target Pearson correlation matrix as the initial conditions to our proposed algorithm will lead to approximate matching in the resultant distribution [@Song00] in many practical applications.
+Now, if one does desire to characterize dependency using Pearson correlations, simply using the target Pearson correlation matrix as the initial conditions to our proposed algorithm will lead to approximate matching in the resultant distribution [@Song00] in many practical applications.
 The quality of this approximation depends on the setting, but in practice, for high-dimensional count data we find the accuracy to be adequate.
-Later, we'll study the robustness of our method to this limitation in selected [Monte Carlo evaluations]({#simulations}).
+Later, we study robustness of our method to this limitation in selected [Monte Carlo evaluations]({#simulations}).
 
@@
 
-Now we describe `bigsimr::rvec`, our algorithm to generate random vectors.
+We now describe `bigsimr::rvec`, our algorithm to generate random vectors.
 It mirrors the classical NORTA algorithm above with some modifications for rank-based dependency matching:
@@
 	(ii) Check that ${\bf R_{Pearson}}$ is semi-positive definite.  
-	(iii) If not compute a close semi-positive definite correlation matrix ${\bf \widetilde{R}_{Pearson}}$.  
+	(iii) If not, compute the nearest correlation matrix ${\bf \widetilde{R}_{Pearson}}$ to ${\bf R_{Pearson}}$ in the Frobenius norm.  
 2. Gaussian copula construction.  
@@
 Researchers working in multivariate computation frequently encounter such difficulties and need to find a close positive (semi-)definite matrix.
-A widely-available routine for this task in `R` is called `matrix::nearPD`, though it is not suitable for high dimensions.
-To overcome this issue, we've developed `bigsimr::nearPSD`, a quadractically-convergent Newton method for finding the nearest correlation matrix, developed by @QS2006.
-We hope that this routine could be useful in many applications aside from our primary goal of random vector generation.
+A widely-available routine for this task in `R` is `matrix::nearPD`, though it is not suitable for high dimensions.
+To overcome this issue, we introduce `bigsimr::nearPSD` --- an implementation of @QS2006's method that computes the nearest correlation matrix to a given symmetric matrix in the Frobenius norm.
+We hope that this routine will be useful in many applications aside from our primary goal of random vector generation.
 
 Once the target margins and algorithm inputs are determined, steps 2 and 3 are essentially a NORTA algorithm with modern high-performance computing implementations.
-Specifically, step 2i uses either an efficient multi-core multivariate normal simulator (the `R` package `mvnfast`[@Fasiolo2016]) or a using Google's JAX python library NumPy for graphical-processing-unit (GPU) acceleration of the Cholesky factorization and matrix multiplication (steps 3, 4 in the NORTA algorithm in the preceding section).
+Specifically, step 2i uses either an efficient multi-core multivariate normal simulator (the `R` package `mvnfast`[@Fasiolo2016]) or Google's JAX python library NumPy for graphical-processing-unit (GPU) acceleration of the Cholesky factorization and matrix multiplication (steps 3, 4 in the NORTA algorithm from the preceding section).
 
@@
 Yet the choice of copula-based simulation algorithm affects which joint distributions may be simulated.
-Even in low-dimensional spaces (e.g., $d=3$), there exists valid multivariate distributions with *feasible* Pearson correlation matrices that NORTA cannot match exactly [@LH75].
-This occurs when the bivariate transformations are applied to find the input correlation matrix yet when combined the resultant matrix is indefinite.
+Even in low-dimensional spaces (e.g., $d=3$), there exist valid multivariate distributions with *feasible* Pearson correlation matrices that NORTA cannot match exactly [@LH75].
+This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined the resultant matrix is indefinite.
 These situations do occur, even using exact analytic calculations. Such problematic target correlation matrices are termed *NORTA defective*.
 
-@GH02 conducted a Monte Carlo study to estimate the probability of NORTA defective matrices while increasing the dimension $d$. 
+@GH02 conducted a Monte Carlo study to estimate the probability of encountering NORTA defective matrices while increasing the dimension $d$. 
 They found at what is now considered low-to-moderate dimensions ($d \approx 20$) that almost *all* feasible matrices are NORTA defective.
@@
 
-There is hope, however, as @GH02 also show that by augmenting the NORTA procedure by replacing the indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$.
+There is hope, however, as @GH02 also showed that replicating an indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$.
 This provides evidence that our nearest positive semi-definite (PSD) augmented approach will maintain reasonable accuracy if our input matching scheme returns an indefinite matrix, at least for the rank-based matching scheme described above.

### 2020-12-16 010-background.Rmd
4th draft. Add packages.bib to address WWP ch030 part 1

@@
 
-Moreover, the simulation method must **scale** to high dimensions:
+Moreover, the simulation method must *scale* to high dimensions:
 
@@
 
-Both $\tau$ and $\rho_S$ are **invariant to under monotone transformations** of the underlying random variates.
+Both $\tau$ and $\rho_S$ are *invariant to under monotone transformations* of the underlying random variates.
 As we will describe more fully in the [Algorithms](#algorithms) section, this property is essential to scaleable match rank-based correlations with speed (SP1) and accuracy (SP2).
@@
 
-There is a strong connection of our simulation strategy to Gaussian **copulas** (see @Nelsen2007 for a technical introduction).
+There is a strong connection of our simulation strategy to Gaussian *copulas* (see @Nelsen2007 for a technical introduction).
 A copula is a distribution function on $[0,1]^d$ that describes a multivariate probability distribution with standard uniform marginals.

### 2020-12-16 020-simulation-algorithm.Rmd
4th draft. Add packages.bib to address WWP ch030 part 1

@@
 To characterize dependency, every pair $(Y_i, Y_j)$ has a given Pearson correlation $\rho_P$, Spearman correlation $\rho_S$, and/or Kendall's $\tau$.
-The method is best understand as a **high-performance Gaussian copula** (Equation \@ref(eq:gauss)) providing a high-dimensional NORTA-inspired algorithm.
+The method is best understand as a *high-performance Gaussian copula* (Equation \@ref(eq:gauss)) providing a high-dimensional NORTA-inspired algorithm.
 
@@
 
-1. Simulate a random vector $\bf Z$ with $d$ **independent** and **identical** standard normal components.
+1. Simulate a random vector $\bf Z$ with $d$ *independent* and *identical* standard normal components.
 2. Determine the input matrix $\Sigma_{\bf Z}$ that corresponds with the specified output $\Sigma_{\bf Y}$.

### 2020-12-16 030-bigsimr-package.Rmd
4th draft. Fix the (X, e^X) bound calculation
4th draft. Add packages.bib to address WWP ch030 part 1

@@
 
-The `bigsimr` package is a high-performance implementation of the proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms) for details).
+The `bigsimr` package is a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms)).
 When designing `bigsimr`, we aimed to conveniently provide parallelized computation, through multi-core and GPU acceleration, while allowing advanced users to customize and automate workflows.
 
-This subsections below describe the basic use of `bigsimr`, by stepping through a low-dimensional (2D) simulation workflow for the often-used example data set `airquality`.
+This subsections below describe the basic use of `bigsimr`, by stepping through a low-dimensional (2D) simulation workflow for the often-used example data set `airquality` that contains daily air quality measurements in New York, May to September 1973 [@Chambers1983].
 This workflow proceeds from data, to estimation, simulation configuration, random vector generation, and result visualization.
-For this low-dimensional setting, we compute using a single central processing unit (CPU) as the overhead in forking the tasks to multiple cores outweighs the computational gains. 
+For this low-dimensional setting, we operate using a single central processing unit (CPU), as the overhead in forking the tasks to multiple cores outweighs the computational gains. 
 Then we transition to advanced use where we briefly describe some of the high-performance features and syntax.
-The section concludes with a short description of how to use `bigsimr` on a computer clusters through slurm scheduling via the `rslurm` package.
+The section concludes with a short description of how to use `bigsimr` on a computer clusters through slurm scheduling via the `rslurm` package [@R-rslurm].
 
@@
 Since our estimated Spearman correlation $\hat{ \rho}_S$ is within the theoretical bounds, the correlation is valid as input to `bigsimr:rvec`.
-By comparison, the Pearson correlation $\rho_P$ is restricted to [-0.8648, 0.8727] for these margins (see `bigsimr::cor_bounds`output below).
+On the other hand, the bounds on the Pearson correlation coefficient $\rho_P$ between these margins is
 

### 2020-12-16 index.Rmd
4th draft. Add packages.bib to address WWP ch030 part 1

@@
 knitr::write_bib(c(
-  .packages(), 'bookdown', 'knitr', 'rmarkdown'
+  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'rslurm'
 ), 'packages.bib')

### 2020-12-17 030-bigsimr-package.Rmd
4th draft. Address WWP ch030 part 2

@@
 
-We'll demonstrate the basic use and syntax of `bigsimr` through an example workflow applied to the New York air quality data set (`airquality`) included in the R `datasets` package.
+We demonstrate the basic use and syntax of `bigsimr` through an example workflow applied to the New York air quality data set (`airquality`) included in the R `datasets` package.
 First, we load the `bigsimr` library and a few other convenient data science packages, including the syntactically-elegant `tidyverse` suite of `R` packages.
@@
 
-For simplicity and to provide a minimal working example, we'll consider bivariate simulation of temperature, in degrees Fahrenheit, and ozone level, in parts per billion.
+For simplicity and to provide a minimal working example, we consider bivariate simulation of the two `airquality` variables `Temperature`, in degrees Fahrenheit, and `Ozone` level, in parts per billion.
 
@@
 We aim to simulate random two-component vectors mimicking this structure. 
-The margins are not normally distributed, particularly the ozone level exhibits a strong positive skew.
+The margins are not normally distributed; particularly the Ozone level exhibits a strong positive skew.
 
 ```{r ch030-aq-joint-dist, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Bivariate scatterplot of Ozone vs.
-Temperature with estimated marginal densities.
-The data are left skewed tails and appear to be correlation.", cache=F}
+Temp with estimated marginal densities.
+We model the Ozone data as marginally log-normal and the Temperature data as normal.", cache=F}
 
@@
 
-Next, we specify the marginal distributions and correlation coefficient (both its type and magnitude).
+Next, we specify the marginal distributions and correlation coefficient (both type and magnitude).
 Here the analyst is free to be creative.
-For this example, we will not take up goodness-of-fit considerations to determine the marginal distributions.
+For this example, we do not take up goodness-of-fit considerations to determine the marginal distributions.
 But it seems sensible without domain knowledge to estimate these quantities from the data and `bigsimr` contains fast functions designed for this task.
@@
 *Specifying marginal distributions*.
-Based on the estimated densities in Figure \@ref(fig:ch030-aq-joint-dist), we'll assume `Temp` is normally distributed and `Ozone` is log-normally distributed since its values are positive and skewed.
-We'll use the well-known, unbiased estimators for the normal distribution's parameters and maximum likelihood estimators for the log-normal parameters:
+Based on the estimated densities in Figure \@ref(fig:ch030-aq-joint-dist), we assume `Temp` is normally distributed and `Ozone` is log-normally distributed, as the latter values are positive and skewed.
+We use the well-known, unbiased estimators for the normal distribution's parameters and maximum likelihood estimators for the log-normal parameters:
 
@@
 
-Next, we'll configure the input marginals for later input into `bigsimr::rvec`.
-The marginal distributions are specifying using `R`'s special `alist` function.
+Next, we configure the input marginals for later input into `bigsimr::rvec`.
+The marginal distributions are specified using `R`'s special `alist` function.
 This allows one to enter the distributions without evaluating anything (yet).
@@
 This implementation strategy supports all `R` `base` probability distributions.
-And allows flexible extensions using other `R` packages that adhere to conventions, such as `extraDistr`.
+It allows flexible extensions using other `R` packages that adhere to conventions, such as `extraDistr`.
 Further, by using `alist`, users can specify their own custom distributions (see below in [creating custom margins](#custom-margins)).
 
-It is a bit inconvenient to have to fill in the parameter values manually each time, so we provide a convenience function `mlist` which behaves similarly to `alist`, except that it will evaluate the right hand side of argument values within the list.
-This is intended to help when scaling up your code to high dimensions when many marginals to specify.
+It is a bit inconvenient to repeatedly fill in the parameter values manually, so we provide a convenience function `mlist` which behaves similarly to `alist`, except that it evaluates the right side of argument values within the list.
+This is intended to help with scaling up your code to high dimensions when many marginals to specify.
 
@@
 For non-normal data and for improved simulation accuracy in our scheme, we advocate the use of rank-based correlations Spearman's $\rho_S$ and Kendall's $\tau$.
-But we also support approximate Pearson correlation coefficient matching, while cautioning the user to check the performance for their parametric multivariate model (see [Monte Carlo evaluations](#simulations) for evaluation strategies and guidance).
+But we also support approximate Pearson correlation coefficient matching, while cautioning the user to check the performance for their parametric multivariate model (see [Monte Carlo evaluations](#simulations) below for evaluation strategies and guidance).
 To aid in correlation specification, and estimation in general, we provide a high-performance function `bigsimr::cor_fast` which estimates Pearson, Spearman, or Kendall correlation using the fastest methods available.
 (Anyone who has tried estimating Kendall's $\tau$ using `stats::cor` can attest that the routine does not scale to even moderate dimensions).
-Notably, these estimation methods are the standard approaches, not designed specifically designed for high-dimensional correlation estimation (see [Conclusion and Discussion]({#discussion) for more on this).
+These estimation methods are the standard approaches, not designed specifically designed for high-dimensional correlation estimation (see the [Conclusion and Discussion]({#discussion) sections for more on this).
 
@@
 
-*Checking the theoretical correlation bounds*
+*Checking the theoretical correlation bounds*.
 As discussed in Section \@ref(background), given a pair of marginal distributions the possible correlations are not free to vary between $[-1, 1]$. 
-To ensure that the simulation is not configured to impossible settings, we provide the `bigsimr::cor_bounds` function provides MC estimated theoretical lower and upper bounds (using the Generate, Sort, and Correlate algorithm of @DH2011).
+To ensure that the simulation is not configured to impossible settings, the `bigsimr::cor_bounds` function provides MC estimated theoretical lower and upper bounds (using the Generate, Sort, and Correlate algorithm of @DH2011).
 
@@
 Finally, we arrive at the main function of `bigsimr`, `rvec`.
-Let's now simulate $B=10,000$ random vectors from the assumed joint distribution of Ozone levels and Temp.
+We now simulate $B=10,000$ random vectors from the assumed joint distribution of Ozone levels and Temp.
 
@@
 *Creating a custom marginal distribution*.
-Because `bigsimr` uses an `alist` to store the margins, any probability distribution with a well-defined inverse CDF can be used including custom marginal distributions not provided in base `R`. 
-Therefore, to specify a marginal distribution absent from an existing package, the user needs to provide a closed-form expression to form the corresponding quantile function.
+Because `bigsimr` uses an `alist` to store the margins, any probability distribution with a well-defined inverse CDF can be employed, including custom marginal distributions not provided in base `R`. 
+Therefore, to specify a marginal distribution absent from an existing package, the user must provide a closed-form expression to form the corresponding quantile function.
 
@@
 Users should prefix distributions with `r` and `q` for *random* and *quantile* respectively.
-Only the quantile function is necessary to simulate random vectors, but to compute the theoretical correlation bounds, it is required to supply univariate random number generator as well.
+Only the quantile function is necessary to simulate random vectors, but to compute the theoretical correlation bounds, one must to supply a univariate random number generator as well.
 
-For an example, let's provide a custom *Pareto* distribution for use with `bigsimr`.
+For an example, consider a custom *Pareto* distribution for use with `bigsimr`.
 The Pareto CDF is
@@
 
-Now with the `qpareto` and `rpareto` functions, we can use the distribution in `bigsimr` just like the other built-in distributions.
+Now with the `qpareto` and `rpareto` functions, we can use the distribution in `bigsimr` similar to the other built-in distributions.
 
@@
 *Using `bigsimr` on a computing cluster via `rslurm`*.
-Though `bigsimr` runs quickly, at large $d$ users may want to run jobs on a shared computing server.
+Though `bigsimr` runs quickly, for large $d$ users may wish to run jobs on a shared computing server.
 The R package `rslurm` makes it easy to run embarrassingly large parallel `rvec` calls.
-This example assumes that `bigsimr` is installed on a system with a slurm scheduler installed.
-A single run of `bigsimr:rvec` using `rslurm` can be code as:
+This example assumes that `bigsimr` is installed on a system with a slurm scheduler.
+A single run of `bigsimr:rvec` using `rslurm` can be coded as:
 
@@
 
-Now, let's show off the real power of combining `bigsimr` and `rslurm` by simulating many correlation structures.
+The power of combining `bigsimr` and `rslurm` can be illustrated by simulating many correlation structures.
 The `rslurm::slumr_map` syntax mirrors the `base::lapply` and `purrr::map` functions.
@@
 
-On a cluster carrying 24 nodes with 48 threads, these 100 jobs completed in about a minute.
+On a cluster carrying 24 nodes with 48 threads, these 100 jobs complete in about a minute.
 

### 2020-12-17 040-monte-carlo-evaluation.Rmd
4th draft. Address WWP ch040

@@
 Since marginal parameter matching in our scheme is essentially a sequence of univariate inverse probability transforms, the challenging aspects are the accuracy of dependency matching and computational efficiency at high dimensions.
-To evaluate our methods in those respects, we design the following numerical experiments to first assess accuracy in match dependency parameters in bivariate simulations and then time the procedure in increasingly large dimension $d$.
+To evaluate our methods in those respects, we design the following numerical experiments to first assess accuracy of matching dependency parameters in bivariate simulations and then time the procedure in increasingly large dimension $d$.
 
@@
 
-We select bivariate simulation configurations to ultimately simulate discrete-valued RNA-seq data and, so, proceed in increasing complexity, leading to the model in our motivating application in Section \@ref(examples).
+We select bivariate simulation configurations to ultimately simulate our motivating discrete-valued RNA-seq example and, so, proceed in increasing complexity, leading to the model in our motivating application in Section \@ref(examples).
 We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson's, Spearman's, and Kendall's ---- in identical, bivariate marginal configurations.
-For each pair of identical margins, we vary the correlations across the entire possible range of values to evaluate the simulation's ability to obtain the theoretic bounds.
+For each pair of identical margins, we vary the target correlation across $\Omega$, the set of possible admissible values for correlation type, to evaluate the simulation's ability to obtain the theoretic bounds.
 The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (mimicking RNA-seq counts).
@@
 Then, we select a non-symmetric continuous distribution: a standard (rate =1) two-component multivariate gamma (MVG).
-Finally, we select distributions and marginal parameter values that are motivated by our RNA-seq data, namely values proximal to probabilities and sizes estimated from the data (see [Example applications for our motivating data](examples) for estimation details).
-Thus we arrive at a multivariate negative binomial (MVNB) $p_1 = p_2 = `r prob`, r_1 = r_2 = 4,\rho$ ).
+Finally, we select distributions and marginal parameter values that are motivated by our RNA-seq data, namely values proximal to probabilities and sizes estimated from the data (see [Example applications](examples) for estimation details).
+Thus we arrive at a multivariate negative binomial (MVNB) $p_1 = p_2 = `r prob`, r_1 = r_2 = 4,\rho \in \Omega$.
 
@@
 |-------------|:--------------:|----------------------:|
-|$1000$ | Pearson ($\rho_P$) | ${ \bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho )$ |
-|$10,000$ | Spearman ($\rho_S$) | ${ \bf Y} \sim MVG( shape = 10, rate = 1, \rho )$ |
-| $100,000$| Kendall ($\tau$)| ${ \bf Y} \sim MVNB(p = `r prob`, r = 4,\rho)$ |
+|$1000$ | Pearson ($\rho_P$) | ${ \bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho_i ), i=1,\ldots,100$ |
+|$10,000$ | Spearman ($\rho_S$) | ${ \bf Y} \sim MVG( shape = 10, rate = 1, \rho_i ), i=1,\ldots,100$ |
+| $100,000$| Kendall ($\tau$)| ${ \bf Y} \sim MVNB(p = `r prob`, r = 4,\rho_i), i=1,\ldots,100$ |
 
-For each of the unique 9 simulation configurations above, we estimate the correlation bounds and vary $\rho$ along a sequence of 100 points evenly placed within the bounds (minus an adjustment factor of $\epsilon=0.01$ to handle numeric issues when the bound is specified exactly).
+For each of the unique 9 simulation configurations described above, we estimate the correlation bounds and vary the correlations along a sequence of 100 points evenly placed within the bounds, aiming to explore $\Omega$.
+Specifically, we set correlations $\{ \rho_1 = ( \hat{l} + \epsilon), \rho_2 = (\hat{l} + \epsilon) + \delta, \ldots, \rho_{100} = (\hat{u} - \epsilon) \}$, with $\hat{l}$ and $\hat{u}$ being the estimated lower and upper bounds, respectively, and increment value $\delta$.
+The adjustment factor, $\epsilon=0.01$, is introduced to handle numeric issues when the bound is specified exactly.
 
@@
 Table \@ref(tab:ch040-biMAEtable) contains the mean absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.
-Taken together, the studies show our methodology is generally accuracy across the entire range of possible correlation values for the rank-based dependency measures, at least in these limited simulation settings for the rank-based correlations.
+Taken together, the studies show that our methodology is generally accurate across the entire range of possible correlation values for the rank-based dependency measures, at least in these limited simulation settings for the rank-based correlations.
 For the two non-normal bivariate marginals, the Pearson correlation matching is approximate.
@@
 The accuracy appears adequate for many applications.
-In practice, we recommend users to always evaluate the accuracy for their application, using methods similar to those presented above.
+In practice, we recommend evaluating the accuracy in each application setting, using methods similar to those presented above.
+
+<!-- 
 See [Discussion](discussion) for future directions for fast Pearson matching and discrete-specific modifications.
+your comment -->
 
@@
 Specifically, we seek evidence to determine whether the method can scale to higher dimensions in a practical time.
-Using our motivating RNA-seq data, described in [Background](background), we filtered the original 20,501 genes to the high-expressing genes at increasing percentiles, $1, 5, 10, 15, 20, 25\%$, to obtain $d=\{206, 1026, 2051, 3076, 4101, 5127\}$ marginals and $\binom{d}{2}$ pairwise correlations at each setting.
+Using our motivating RNA-seq data, we filtered the original 20,501 genes to the high-expressing genes at increasing percentiles, $1, 5, 10, 15, 20, 25\%$, to obtain $d=\{206, 1026, 2051, 3076, 4101, 5127\}$ marginals and $\binom{d}{2}$ pairwise correlations at each setting.
 For example, for $d=5127$ there are $13,140,501$ correlation coefficients.
 We estimated the marginal negative binomial parameters and the correlation coefficients from the RNA-seq data to seed our simulations.
-(See [Example applications for our motivating data](examples) for a detailed description of estimation).
+(See [Example applications](examples) for a detailed description of the estimation).
 
@@
 The Pearson simulations are much faster since the correlation conversion steps are avoided (pre-processing step; see [Algorithms](algorithms)), but as we know from above the accuracy will suffer slightly, especially near the negative boundary of the possible correlations.
-Matching Spearman's correlation at larger $d$ gets costly if one wants to produce $B=10,000$ random vectors at many different simulation settings, but the conversion steps need only be computed once (see [MC evaluation of correlation estimation efficiency](MC-evaluation-of-correlation-estimation-efficiency) for an example of this strategy).
+Matching Spearman's correlation at larger $d$ becomes costly if one desires to produce $B=10,000$ random vectors at many different simulation settings, but the conversion steps need only be computed once (see [MC evaluation of correlation estimation efficiency](MC-evaluation-of-correlation-estimation-efficiency) for an example of this strategy).
 

### 2020-12-17 050-examples.Rmd
4th draft. Address WWP ch050

@@
-# RNA-seq data example applications  {#examples}
+# RNA-seq data applications  {#examples}
 
@@
 This section demonstrates how to simulate multivariate data using `bigsimr`, aiming to replicate the structure of high-dimensional dependent count data.
-In an illustration of our proposed methodology applied to real data, we seek to simulate RNA-sequencing data by producing simulated random vectors with assumed marginal distributions with estimated parameters that mimic the observed data and its generating process.
+In an illustration of our proposed methodology, we seek to simulate RNA-sequencing data by producing simulated random vectors with assumed marginal distributions and with estimated parameters that mimic the observed data and its generating process.
 We seek scaleable realistic multivariate simulation to enable Monte Carlo (MC) methods for these data.
-Modeling RNA-seq using multivariate probabilities distributions is natural as inter-gene correlation is an inherent part of biological processes.
+Modeling RNA-seq using multivariate probabilities distributions is natural as inter-gene correlation is an inherent part of biological processes [@Wang2009b].
 Yet many models do not account for this, leading to major disruptions to the operating characteristics of statistical estimation, testing, and prediction.
-See @Efron2012 for a detailed discussion with related methods and see @Wu2012b, @Schissler2018, @Schissler2019 for applied examples.
-The following subsections apply `bigsimr`'s methods to real RNA-seq data, including replicated an estimated parametric structure, MC probability estimation, and MC evaluation of correlation estimation efficiency.
+See @Efron2012 for a detailed discussion with related methods and @Wu2012b, @Schissler2018, @Schissler2019 for applied examples.
+The following subsections apply `bigsimr`'s methods to real RNA-seq data, including replicating an estimated parametric structure, MC probability estimation, and MC evaluation of correlation estimation efficiency.
 
@@
 
-Our first goal is to replicate the structure of the TCGA BRCA RNA-seq data set (see [Background](background)).
+Our initial goal is to replicate the structure of the TCGA BRCA RNA-seq data set (see [Background](background)).
 Ultimately, we will simulate $B=10,000$ random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with $d=`r d`$.
-Often researchers posit a negative binomial (NB) model as RNA-seq counts are often over-dispersed that a Poisson model would suggest.
-All $d$ selected genes exhibit over-dispersion (data not shown) and, so, we proceed to estimate the NB parameters $(r_i, p_i), i=1,\ldots,d$ to determine the target marginal PMFs $f_i$ (via method of moments).
-To complete specification of the simulation algorithm inputs, we'll estimate the Spearman correlation matrix ${ \bf R}_{Spearman}$ to characterize dependency.
+Often researchers posit a negative binomial (NB) model as RNA-seq counts are often over-dispersed.
+All $d$ selected genes exhibit over-dispersion (data not shown) and, so, we proceed to estimate the NB parameters $(r_i, p_i), i=1,\ldots,d$ to determine the target marginal PMFs $f_i$ (via the method of moments).
+To complete specification of the simulation algorithm inputs, we estimate the Spearman correlation matrix ${ \bf R}_{Spearman}$ to characterize dependency.
 
@@
 
-With our goal in mind, we first estimate the desired correlation matrix using the fast implementation provided by `bigsimr`:
+With this goal in mind, we first estimate the desired correlation matrix using the fast implementation provided by `bigsimr`:
 
@@
 Next, we estimate the marginal parameters. 
-We use method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model.
+We use the method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model.
 The marginal distributions are from the same probability family (NB) yet are heterogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.
@@
 
-Now we estimate the marginal parameters using the built-in `apply`function:
+Now we estimate the marginal parameters using the built-in `apply` function:
 
@@
 
-Notably, the marginal NB probabilities $\hat{p}_i's$ are small --- ranging in $[`r min(probs)` , `r max(probs)`]$.
+Notably, the marginal NB probabilities $\hat{p}_i's$ are small --- ranging in the interval $[`r min(probs)` , `r max(probs)`]$.
 This gives rise to highly variable counts and, typically, less restriction on potential pairwise correlation pairs.
@@
 
-Figure \@ref(fig:ch050-simDataFig) displays the simulated counts and pairwise relationships for our example genes in Table \@ref(tab:ch010-realDataTab).
-Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parameter form and with less extreme points then the observed data (c.f.
-Figure \@ref(fig:ch010-realDataFig).)
+Figure \@ref(fig:ch050-simDataFig) displays the simulated counts and pairwise relationships for our example genes from Table \@ref(tab:ch010-realDataTab).
+Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parameter form and with less extreme points then the observed data in Figure \@ref(fig:ch010-realDataFig).
 
@@
 The evaluation shows that the simulated counts approximately match the target parameters and exhibit the full range of estimated correlation from the data.
-Utilizing 15 CPU threads in a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, the simulation completed just over of 2 minutes.
+Utilizing 15 CPU threads in a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, the simulation completed in just over 2 minutes.
 
@@
 
-where ${\bf Y^{(b)} }$ is the $b^{th}$ simulated vector in a total of $B$ simulation replicates and $I ( )$ is the indicator function.
-For example, we can estimate from our $B=10,000$ simulated vectors the probability that all genes are expressed (i.e., ${\bf y}_i \geq 1, \forall \; i$) is $0.1708$.
+```{r ch050-densityEvaluation, include=FALSE, eval=TRUE, cache=F}
+d <- ncol(sim_nbinom)
+B <- nrow(sim_nbinom)
+threshold <- rep( 1, d)
+pHat <- mean(apply( sim_nbinom,  1,
+           function(X, y0=threshold) {
+               all( X >= y0) }
+           ))
+```
 
-```{r ch050-densityEvaluation, echo=TRUE, eval=TRUE, cache=F}
+where ${\bf Y^{(b)} }$ is the $b^{th}$ simulated vector from a total of $B$ simulation replicates and $I ( )$ is the indicator function.
+For example, we can estimate from our $B=10,000$ simulated vectors that the probability of all genes expressing (i.e., ${\bf y}_i \geq 1, \forall \; i$) is $`r pHat`$.
+
+```{r ch050-densityEvaluationECHO, echo=TRUE, eval=TRUE, cache=F}
 d <- ncol(sim_nbinom)
@@
            function(X, y0=threshold) {
-               all( X > y0) }
+               all( X >= y0) }
            ))
@@
 MC methods are routinely used in many statistical inferential tasks including estimation, hypothesis testing, error rates, and empirical interval coverage rates.
-For an concise introduction to these methods, see @Rizzo2007, Ch. 6.
 To conclude the example applications, we demonstrate how `bigsimr` can be used evaluate estimation efficiency.
-In particular, we'd like to assess the error in our correlation estimation above.
-We used a conventional method, based on classical statistical theory. 
-Yet this method was not designed for high-dimensional data.
-Indeed, high-dimensional covariance estimation (and precision matrices) is an active area of statistical science (see, for example, [@Won2013g; @VanWieringen2016].
+In particular, we wish to assess the error in our correlation estimation above.
+We used a conventional method, based on classical statistical theory, which unfortunately, was not designed for high-dimensional data.
+Indeed, high-dimensional covariance estimation (and precision matrices) is an active area of statistical science (.e.g., @Won2013g and @VanWieringen2016).
 
@@
 
-This distribution could be compared to high-dimensional designed covariance estimators to guide to help decide whether the additional complexity and computation time are warranted.
+This distribution could be compared to high-dimensional designed covariance estimators for guidance on whether the additional complexity and computation time are warranted.
 

### 2020-12-17 060-conclusion-discussion.Rmd
4th draft. Address WWP ch060

@@
 
-We've introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a high-performance `R` package called [`bigsimr`](https://schisslergroup.github.io/bigsimr/).
+We have introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a high-performance `R` package named [`bigsimr`](https://schisslergroup.github.io/bigsimr/).
 The random vector generation method is largely inspired by NORTA (@Cario1997) and Gaussian copula-based approaches (@MB13, @BF17, @Xia17).
@@
 It is customary to compare new tools and algorithms directly to existing competing methods and software.
-In this study, however, we only employ our proposed methodology, since our previous work has show that existing tools are simply not designed or feasible to meet our high-dimensional goal (see @Li2019gpu for evaluations of the `R` `copula`package and others).
+In this study, however, we only employ our proposed methodology, as our previous work has show that existing tools are simply not designed nor feasible to meet our high-dimensional goal (see @Li2019gpu for evaluations of the `R` `copula`package and others).
 For the bivariate simulations, existing packages such as [`nortaRA`](https://github.com/cran/NORTARA/blob/master/inst/doc/NORTARA.R) work well to match Pearson correlations exactly.
@@
 The most obvious missing feature of the proposed methodology is the inability to match a Pearson correlation matrix exactly.
-As discussed in [Algorithms](algorithms) and extensively by @XZ19, this is a computationally intense procedure and Pearson's correlation is not a natural choice to describe dependency for non-normal marginals.
-While we do not provide an implementation directly supporting Pearson matching, users may supply their own input Pearson correlation after using a supplementary matching scheme [@Cario1997; @XZ19].
+As discussed in the [Algorithms](algorithms) section and extensively by @XZ19, this is a computationally intense procedure and Pearson's correlation is not a natural choice to describe dependency for non-normal marginals.
+While we do not provide an implementation directly supporting Pearson matching, users may supply their own input Pearson correlation after employing a supplementary matching scheme [@Cario1997; @XZ19].
 
@@
 
+<!-- 
 Future work includes developing scaleable algorithms to match the Pearson correlation matrix more precisely, discrete-margin specific modifications including fast Spearman's correlation rescaling (see Equation \@ref(eq:spearmanRescaled)), and high-dimensional covariance estimation.
@@
 As data-analytic problems grow to even larger dimension, multi-GPU support is a promising hardware-based future direction.
+your comment -->

### 2020-12-17 010-background.Rmd
4th draft. Address ADK feedback

@@
 
-* SP1: Procedure must scale to high dimensions, computable in a reasonable amount time.
+* SP1: Procedure must scale to high dimensions with practical computable times.
 * SP2: Procedure must scale to high dimensions while maintaining accuracy.
@@
 *Discrete marginal considerations*.
-Spearman's correlation for discrete marginal suffers issues whenever there is a nonzero probability of ties (for example, the Spearman's correlation of a discrete-valued random variable $X$ with itself could be less than 1; @MB13).
+Spearman's correlation for discrete marginal suffers issues whenever there is a nonzero probability of ties.
+For example, the Spearman's correlation of a discrete-valued random variable $X$ with itself could be less than 1 [@MB13].
 One remedy for this issue is to rescale Equation \@ref(eq:spearman). 
@@
 Given two marginal distributions, $\rho_P$ is not free to vary over the entire range of possible correlations $[-1,1]$.
-The so-called *Frechet-Hoeffding bounds* are well-studied (see @Nelsen2007 and @BF17).
-This gives strict restraints on the possible correlations and cannot be overcome through algorithm design.
+The so-called *Frechet-Hoeffding bounds* are well-studied ([@Nelsen2007; @BF17]).
+These constraints cannot be overcome through algorithm design.
 In general, the bounds are given by

### 2020-12-17 020-simulation-algorithm.Rmd
4th draft. Address ADK feedback

@@
 To characterize dependency, every pair $(Y_i, Y_j)$ has a given Pearson correlation $\rho_P$, Spearman correlation $\rho_S$, and/or Kendall's $\tau$.
-The method is best understand as a *high-performance Gaussian copula* (Equation \@ref(eq:gauss)) providing a high-dimensional NORTA-inspired algorithm.
+The method can be described as a *high-performance Gaussian copula* (Equation \@ref(eq:gauss)) providing a high-dimensional NORTA-inspired algorithm.
 
@@
 Once the target margins and algorithm inputs are determined, steps 2 and 3 are essentially a NORTA algorithm with modern high-performance computing implementations.
-Specifically, step 2i uses either an efficient multi-core multivariate normal simulator (the `R` package `mvnfast`[@Fasiolo2016]) or Google's JAX python library NumPy for graphical-processing-unit (GPU) acceleration of the Cholesky factorization and matrix multiplication (steps 3, 4 in the NORTA algorithm from the preceding section).
+Specifically, step 2i uses either the efficient multi-core multivariate normal simulator `mvnfast`[@Fasiolo2016] or Google's JAX python library for GPU acceleration of the Cholesky factorization and matrix multiplication (steps 3, 4 in the NORTA algorithm from the preceding section).
 

### 2020-12-17 030-bigsimr-package.Rmd
4th draft. Address ADK feedback

@@
 
-This subsections below describe the basic use of `bigsimr`, by stepping through a low-dimensional (2D) simulation workflow for the often-used example data set `airquality` that contains daily air quality measurements in New York, May to September 1973 [@Chambers1983].
+The subsections below describe the basic use of `bigsimr`, by stepping through a low-dimensional (2D) simulation workflow for the readily-available data set `airquality` that contains daily air quality measurements in New York, May to September 1973 [@Chambers1983].
 This workflow proceeds from data, to estimation, simulation configuration, random vector generation, and result visualization.
@@
 It is a bit inconvenient to repeatedly fill in the parameter values manually, so we provide a convenience function `mlist` which behaves similarly to `alist`, except that it evaluates the right side of argument values within the list.
-This is intended to help with scaling up your code to high dimensions when many marginals to specify.
 
@@
 (Anyone who has tried estimating Kendall's $\tau$ using `stats::cor` can attest that the routine does not scale to even moderate dimensions).
-These estimation methods are the standard approaches, not designed specifically designed for high-dimensional correlation estimation (see the [Conclusion and Discussion]({#discussion) sections for more on this).
+These estimation methods are classical approaches, not designed for high-dimensional correlation estimation (see the [Conclusion and Discussion]({#discussion) sections for more on this).
 
@@
 sjob <- slurm_call(rvec, jobname = 'rvec',
-                   alist(n=1e6,
-                         rho = rho,
-                        margins = margins,
-                        type = "spearman"),
+                   alist(n=1e6, rho = rho, margins = margins, type = "spearman"),
                    submit = TRUE)
@@
 simReps <- 100
-rhoList <- replicate( n = simReps, bigsimr::cor_randPSD(d = length(margins)),
-                     simplify=FALSE )
+rhoList <- replicate(n = simReps, bigsimr::cor_randPSD(d = length(margins)),
+                     simplify=FALSE)
 sjob <- slurm_map(x = rhoList,

### 2020-12-17 040-monte-carlo-evaluation.Rmd
4th draft. Address ADK feedback

@@
 We select bivariate simulation configurations to ultimately simulate our motivating discrete-valued RNA-seq example and, so, proceed in increasing complexity, leading to the model in our motivating application in Section \@ref(examples).
-We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson's, Spearman's, and Kendall's ---- in identical, bivariate marginal configurations.
+We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson's, Spearman's, and Kendall's --- in identical, bivariate marginal configurations.
 For each pair of identical margins, we vary the target correlation across $\Omega$, the set of possible admissible values for correlation type, to evaluate the simulation's ability to obtain the theoretic bounds.

### 2020-12-17 050-examples.Rmd
4th draft. Address ADK feedback

@@
 In an illustration of our proposed methodology, we seek to simulate RNA-sequencing data by producing simulated random vectors with assumed marginal distributions and with estimated parameters that mimic the observed data and its generating process.
-We seek scaleable realistic multivariate simulation to enable Monte Carlo (MC) methods for these data.
-Modeling RNA-seq using multivariate probabilities distributions is natural as inter-gene correlation is an inherent part of biological processes [@Wang2009b].
+We seek realistic multivariate simulation to enable Monte Carlo (MC) methods for these data.
+Modeling RNA-seq using multivariate probability distributions is natural as inter-gene correlation is an inherent part of biological processes [@Wang2009b].
 Yet many models do not account for this, leading to major disruptions to the operating characteristics of statistical estimation, testing, and prediction.
@@
 
-With this goal in mind, we first estimate the desired correlation matrix using the fast implementation provided by `bigsimr`:
+With this goal in mind, we first estimate the desired correlation matrix using the fast implementation provided by `bigsimr`.:
 
@@
 corType <- 'spearman'
-system.time( nb_Rho <- bigsimr::cor_fast( brca, method = corType ) )
+nb_Rho <- bigsimr::cor_fast(brca, method = corType)
 ```
@@
 ```{r ch050-runBRCA-echo, echo=TRUE, eval=FALSE, cache = FALSE}
-## Set the number of random vectors
+## Set the number of random vectors (B in text)
 n <- 10000
@@
 
-To conduct statistical inference a critical task is to evaluate the joint probability mass (or density) function:
+Many statistical tasks require evaluation of a joint probability mass (or density) function:
 
@@
 $$
-\hat{P}( {\bf Y} >= {\bf y_0 } ) = \sum_{b=1}^B I( {\bf Y^{(b) }} > {\bf y_0 } ) / B
+\hat{P}( {\bf Y} >= {\bf y_0 } ) = \frac{1}{B} \sum_{b=1}^B I( {\bf Y^{(b) }} >= {\bf y_0 } )
 $$
@@
 threshold <- rep( 1, d)
-pHat <- mean(apply( sim_nbinom,  1,
+pHat <- mean(apply(sim_nbinom,  1,
            function(X, y0=threshold) {
-               all( X >= y0) }
+               all(X >= y0)}
            ))

### 2020-12-17 060-conclusion-discussion.Rmd
4th draft. Address ADK feedback

@@
 It is customary to compare new tools and algorithms directly to existing competing methods and software.
-In this study, however, we only employ our proposed methodology, as our previous work has show that existing tools are simply not designed nor feasible to meet our high-dimensional goal (see @Li2019gpu for evaluations of the `R` `copula`package and others).
+In this study, however, we only employ our proposed methodology, as our previous work has shown that existing tools are simply not designed nor feasible to meet our high-dimensional goal (see @Li2019gpu for evaluations of the `R` `copula` package and others).
 For the bivariate simulations, existing packages such as [`nortaRA`](https://github.com/cran/NORTARA/blob/master/inst/doc/NORTARA.R) work well to match Pearson correlations exactly.

### 2020-12-17 index.Rmd
4th draft. Address ADK feedback

@@
 This article introduces the `bigsimr` R package that provides a flexible, scaleable procedure to simulate high-dimensional random vectors with given marginal characteristics and dependency measures.
-We describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and find close positive semi-definite matrices.
+We describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix.
 Finally, we demonstrate the power of `bigsimr` by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d =1026$.
@@
 The crux of the method lies in the construction of a Gaussian copula, in the spirit of the NORTA procedure.
-Further, we introduce the `bigsimr` R package that provides parallelized, high-performance software implemented in our NORTA-inspired algorithm.
+Further, we introduce the `bigsimr` R package that provides high-performance software implementing our algorithm.
 The algorithm design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed-form relationships between dependency measures for the multivariate normal (MVN) distribution.
@@
 Next, we detail an illustrative low-dimensional example of basic and advanced use of the `bigsimr` R package.
-Then we proceed with Monte Carlo studies under various distributional assumptions to assess accuracy and scaleable.
+Then we proceed with Monte Carlo studies under various distributional assumptions to assess accuracy and computation times for increasingly large $d$.
 After the MC evaluations, we revisit our high-dimensional motivating example and employ our methods in commonplace statistical computing tasks.

### 2020-12-19 010-background.Rmd
4th draft. Address JP feedback

@@
 When designing this methodology, we developed the following properties to guide our effort.
-We divide the properties into two categories: (1) basic properties (BP) and "scaleability" properties (SP). 
+We divide the properties into two categories: (1) basic properties (BP) and "scalability" properties (SP). 
 The BPs are adapted from an existing criteria due to @Nik13a.
-Our simulation strategy should allow:
+A suitable simulation strategy should possess the following properties:
 
@@
 
-To fix ideas and provide example applications enabled via `bigsimr`, the next section describes a motivating data set.
+<!-- 
+The next section describes the motivating data set we use to provide example applications.
+your comment -->
 
@@
 For human-derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted [@Schissler2019].
-Importantly, due to inherent biological processes, gene expression data exhibits correlation --- co-expression --- across genes [@BE07; @Schissler2018].
+Importantly, due to inherent biological processes, gene expression data exhibit correlation --- co-expression --- across genes [@BE07; @Schissler2018].
 
@@
 Both $\tau$ and $\rho_S$ are *invariant to under monotone transformations* of the underlying random variates.
-As we will describe more fully in the [Algorithms](#algorithms) section, this property is essential to scaleable match rank-based correlations with speed (SP1) and accuracy (SP2).
+As we will describe more fully in the [Algorithms](#algorithms) section, this property is essential to scalable match rank-based correlations with speed (SP1) and accuracy (SP2).
 
@@
 
-\[
-F(x_1, \ldots,x_d) = \mathbb P(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), \,\,\, x_i\in \mathbb R, i=1,\ldots,d. 
-\]
+\begin{equation}
+(\#eq:copula)
+F(x_1, \ldots,x_d) = \mathbb P(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), 
+\end{equation}
 
-A Gaussian copula has marginal CDFs as the normal CDF, $\Phi$. 
-This representation corresponds to a multivariate normal distribution with standard normal marginal distributions and covariance matrix ${\bf R_P}$.
+$x_i \in \mathbb R, i=1,\ldots,d.$
+
+A Gaussian copula has marginal CDFs that are all normal, $F_i = \Phi, \forall \, i$. 
+This representation corresponds to a multivariate normal (MVN) distribution with standard normal marginal distributions and covariance matrix ${\bf R_P}$.
 As the marginals are standardized to have unit variance, however, ${\bf R_P}$ is a Pearson correlation matrix. 

### 2020-12-19 020-simulation-algorithm.Rmd
4th draft. Address JP feedback

@@
 
-The well-known NORTA algorithm [@Cario1997] can be used simulate a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$.
+The well-known NORTA algorithm [@Cario1997] simulates a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$.
 Specifically, NORTA algorithm proceedings as follows:
@@
 
-With modern parallelized computing, steps 1, 3, 4, 5 are readily implemented as high-performance, multi-core and/or graphical-processing-unit (GPU) accelerated, algorithms --- providing fast scaleability using readily-available hardware.
+With modern parallelized computing, steps 1, 3, 4, 5 are readily implemented as high-performance, multi-core and/or graphical-processing-unit (GPU) accelerated algorithms --- providing fast scalability using readily-available hardware.
 
@@
 For high-dimensional data, these evaluations are often too costly to enable feasible simulation studies.
-For low-dimensional problems, however, methods and tools exist to match Pearson correlations precisely, see @Chen2001; @Xia17; @MB13 and the publicly available `nortaRA` R package.
+For low-dimensional problems, however, methods and tools exist to match Pearson correlations precisely [@Xia17] and the publicly available `nortaRA` R package [@Chen2001].
 
-To maintain scaleability (SP1), our solution is to essentially avoid this complication in Pearson matching.
+To maintain scalability (SP1), our solution is to essentially avoid this complication in Pearson matching.
 Since our goal is to simulate non-normal marginals, we greatly prefer the use of rank-based measures, such as $\rho_S$ and $\tau$, from a modeling standpoint.
@@
 
-## Random vector generation via `bigsimr::rvec` {#rand-vec-gen}
+## `bigsimr::rvec` for random vector generation  {#rand-vec-gen}
 
-We now describe `bigsimr::rvec`, our algorithm to generate random vectors.
-It mirrors the classical NORTA algorithm above with some modifications for rank-based dependency matching:
+We now describe `bigsimr::rvec`, our algorithm to generate random vectors, which mirrors the classical NORTA algorithm above with some modifications for rank-based dependency matching:
 
 1. Pre-processing for nonparametric dependency matching.  
-	(i) Convert from either ${\bf R_{Spearman}}$ or ${\bf R_{Kendall}}$ into the corresponding MVN input correlation ${\bf R_{Pearson}}$.
+	(i) Convert ${\bf R_{Kendall}}$ or ${\bf R_{Spearman}}$ to the corresponding MVN correlation ${\bf R_{Pearson}}$.
 	(ii) Check that ${\bf R_{Pearson}}$ is semi-positive definite.  
@@
 
-The pre-processing (Step 1) takes advantage of the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertSpearman) or \@ref(eq:convertKendall), respectively (implemented as `bigsimr::cor_covert`).
+The pre-processing (Step 1) takes advantage of the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertKendall) or \@ref(eq:convertSpearman), respectively (implemented as `bigsimr::cor_covert`).
 
@@
 To overcome this issue, we introduce `bigsimr::nearPSD` --- an implementation of @QS2006's method that computes the nearest correlation matrix to a given symmetric matrix in the Frobenius norm.
-We hope that this routine will be useful in many applications aside from our primary goal of random vector generation.
+This routine as many applications beyond our primary goal of random vector generation.
 
-Once the target margins and algorithm inputs are determined, steps 2 and 3 are essentially a NORTA algorithm with modern high-performance computing implementations.
-Specifically, step 2i uses either the efficient multi-core multivariate normal simulator `mvnfast`[@Fasiolo2016] or Google's JAX python library for GPU acceleration of the Cholesky factorization and matrix multiplication (steps 3, 4 in the NORTA algorithm from the preceding section).
+Once the target margins and algorithm inputs are determined, Steps 2 and 3 are essentially a NORTA algorithm with modern high-performance computing implementations.
+Specifically, step 2i uses either the efficient multi-core multivariate normal simulator `mvnfast`[@Fasiolo2016] or Google's JAX python library for GPU acceleration of the Cholesky factorization and matrix multiplication.
 
@@
 @GH02 conducted a Monte Carlo study to estimate the probability of encountering NORTA defective matrices while increasing the dimension $d$. 
-They found at what is now considered low-to-moderate dimensions ($d \approx 20$) that almost *all* feasible matrices are NORTA defective.
+They found for what is now considered low-to-moderate dimensions ($d \approx 20$) that almost *all* feasible matrices are NORTA defective.
 This stems from the concentration of measure near the boundary of the space of all possible correlation matrices as dimension increases.

### 2020-12-19 030-bigsimr-package.Rmd
4th draft. Address JP feedback

@@
 
-Notice that we use the *quantile* function for the marginals, as that is how the marginal distributions $F_i$ enter into the `bigsimr::rvec` algorithm.
+Notice that we use the *quantile* function for the marginals, as that is how the marginal distributions $F_i$ are specified the `bigsimr::rvec` algorithm.
 This implementation strategy supports all `R` `base` probability distributions.
@@
 
-It is a bit inconvenient to repeatedly fill in the parameter values manually, so we provide a convenience function `mlist` which behaves similarly to `alist`, except that it evaluates the right side of argument values within the list.
+To avoid manual entry, `mlist` presents an alternative to `alist` with on-the-fly evaluation.
 
@@
 As discussed in Section \@ref(background), given a pair of marginal distributions the possible correlations are not free to vary between $[-1, 1]$. 
-To ensure that the simulation is not configured to impossible settings, the `bigsimr::cor_bounds` function provides MC estimated theoretical lower and upper bounds (using the Generate, Sort, and Correlate algorithm of @DH2011).
+`bigsimr::cor_bounds` estimates the theoretic lower and upper correlation bounds, using the Generate, Sort, and Correlate algorithm of @DH2011.
 
@@
 
-It is important to follow `R`'s naming convention for probability distributions.
 Users should prefix distributions with `r` and `q` for *random* and *quantile* respectively.
@@
 *Using `bigsimr` on a computing cluster via `rslurm`*.
-Though `bigsimr` runs quickly, for large $d$ users may wish to run jobs on a shared computing server.
-The R package `rslurm` makes it easy to run embarrassingly large parallel `rvec` calls.
+Although `bigsimr` runs quickly, for large $d$ users may wish to run jobs on a shared computing server.
+The R package `rslurm` enables parallel `rvec` calls.
 This example assumes that `bigsimr` is installed on a system with a slurm scheduler.

### 2020-12-19 040-monte-carlo-evaluation.Rmd
4th draft. Address JP feedback

@@
 
-We select bivariate simulation configurations to ultimately simulate our motivating discrete-valued RNA-seq example and, so, proceed in increasing complexity, leading to the model in our motivating application in Section \@ref(examples).
+We select bivariate simulation configurations to ultimately simulate our motivating discrete-valued RNA-seq example, and, so we proceed in increasing complexity, leading to the model in our motivating application in Section \@ref(examples).
 We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson's, Spearman's, and Kendall's --- in identical, bivariate marginal configurations.
@@
 Table \@ref(tab:ch040-biMAEtable) contains the mean absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.
-Taken together, the studies show that our methodology is generally accurate across the entire range of possible correlation values for the rank-based dependency measures, at least in these limited simulation settings for the rank-based correlations.
+Overall, the studies show that our methodology is generally accurate across the entire range of possible correlation values for the rank-based dependency measures, at least in these limited simulation settings for the rank-based correlations.
 For the two non-normal bivariate marginals, the Pearson correlation matching is approximate.

### 2020-12-19 050-examples.Rmd
4th draft. Address JP feedback

@@
 The marginal distributions are from the same probability family (NB) yet are heterogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.
-The functions below help achieve this estimation and specifying the inputs for use in `bigsimr::rvec`.
+The functions below support this estimation for later use in `bigsimr::rvec`.
 
@@
 MC methods are routinely used in many statistical inferential tasks including estimation, hypothesis testing, error rates, and empirical interval coverage rates.
-To conclude the example applications, we demonstrate how `bigsimr` can be used evaluate estimation efficiency.
+To conclude the example applications, we demonstrate how `bigsimr` can be used to evaluate estimation efficiency.
 In particular, we wish to assess the error in our correlation estimation above.

### 2020-12-19 060-conclusion-discussion.Rmd
4th draft. Address JP feedback

@@
 
-We have introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a high-performance `R` package named [`bigsimr`](https://schisslergroup.github.io/bigsimr/).
+We have introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a user-friendly, high-performance `R` package named [`bigsimr`](https://schisslergroup.github.io/bigsimr/).
 The random vector generation method is largely inspired by NORTA (@Cario1997) and Gaussian copula-based approaches (@MB13, @BF17, @Xia17).
-The major contributions of this work are high-dimensional scaleability, flexible modeling of dependency, and an high-performance implementation with broad potential data analytic applications for modern, big-data statistical computing.
+The major contributions of this work are high-dimensional scalability, flexible modeling of dependency, and an high-performance implementation with broad potential data analytic applications for modern, big-data statistical computing.
 

### 2020-12-19 070-supplementary-materials.Rmd
4th draft. Add packages.bib to address WWP ch030 part 1
4th draft. Remove naive References heading
4th draft. Address JP feedback

@@
 
-The authors gratefully acknowledge the helpful discussions with University of Arizona's Professor Walter W. Piegorsch and Professor Edward J. Bedrick during this project's conception.
-We also gratefully acknowledge Heather Knudson's graphic design for the `bigsimr` R Package.
+The authors gratefully acknowledge Heather Knudson's graphic design for the `bigsimr` R Package.
 The results published here are in whole or part based upon data generated by the TCGA Research Network: https://www.cancer.gov/tcga.
@@
 
-Research reported in this publication was supported by MW-CTR-IN of the National Institutes of Health under award number NIH 1U54GM104944.
+Research reported in this publication was supported by grants from NV INBRE and the National Institutes of General Medical Sciences (NIH 5U54GM104944, GM103440).
 

### 2020-12-22 030-bigsimr-package.Rmd

@@
 ```{r ch030-cor-bounds-pearson, echo=TRUE}
-cor_bounds(margins = margins, type = 'pearson', reps=1e6)
+cor_bounds(margins = margins, type = 'pearson', reps = 1e6)
 ```
@@
 sjob <- slurm_call(rvec, jobname = 'rvec',
-                   list(n=1e6,
+                   list(n = 1e6,
                         rho = rho,
@@
 sjob <- slurm_call(rvec, jobname = 'rvec',
-                   alist(n=1e6, rho = rho, margins = margins, type = "spearman"),
+                   alist(n = 1e6, 
+                         rho = rho, 
+                         margins = margins, 
+                         type = "spearman"),
                    submit = TRUE)
@@
                   jobname = 'rvecMap',
-                  n=1e6,
+                  n = 1e6,
                   margins = margins,
@@
 res <- rslurm::get_slurm_out(sjob, outtype = 'raw')
-rhoHats <- lapply( res, bigsimr::fastCor, method = 'spearman' )
+rhoHats <- lapply(res, bigsimr::fastCor, method = 'spearman')
 resData <- NULL
@@
     avgRhoHat <- tmpRhoHat[lower.tri(tmpRhoHat)]
-    tmpPair <- paste0( "pair", apply( combn(x = length(margins), 2), 2, paste, collapse = "_" ) )
-    tmpResData <- data.frame( simNum=i, rho=rhoList[[i]][1,2], rhoHat=tmpRhoHat[lower.tri(tmpRhoHat)], pair=tmpPair )
+    tmpPair <- paste0(
+      "pair", apply(combn(x = length(margins), 2), 2, paste, collapse = "_")
+    )
+    tmpResData <- data.frame(
+      simNum = i,
+      rho    = rhoList[[i]][1,2], 
+      rhoHat = tmpRhoHat[lower.tri(tmpRhoHat)], 
+      pair   = tmpPair
+    )
     resData <- rbind(resData, tmpResData)
@@
 ## save for viz
-saveRDS( res, file = 'data/rvecRslurmMapRes.rds' )
+saveRDS(res, file = 'data/rvecRslurmMapRes.rds')
 ## cleanup careful this deletes output also

### 2020-12-22 050-examples.Rmd
4th draft. Finalize CMStats2020 talk and fix >= typesetting

@@
 $$
-\hat{P}( {\bf Y} >= {\bf y_0 } ) = \frac{1}{B} \sum_{b=1}^B I( {\bf Y^{(b) }} >= {\bf y_0 } )
+\hat{P}( {\bf Y} \ge {\bf y_0 } ) = \frac{1}{B} \sum_{b=1}^B I( {\bf Y^{(b) }} \ge {\bf y_0 } )
 $$
@@
 B <- nrow(sim_nbinom)
-threshold <- rep( 1, d)
-pHat <- mean(apply(sim_nbinom,  1,
-           function(X, y0=threshold) {
-               all(X >= y0)}
-           ))
+threshold <- rep(1, d)
+pHat <- mean(apply(sim_nbinom, 1, function(X, y0 = threshold) {
+  all(X >= y0)
+}))
 ```
@@
 B <- nrow(sim_nbinom)
-threshold <- rep( 1, d)
-mean(apply( sim_nbinom,  1,
-           function(X, y0=threshold) {
-               all( X >= y0) }
-           ))
+threshold <- rep(1, d)
+mean(apply(sim_nbinom, 1, function(X, y0 = threshold) {
+  all( X >= y0)
+}))
 ```
@@
 ## create m random vectors and estimate correlation
-simRho <- replicate(n = m,
-  expr = { tmpSim <- rvec(n = n , nb_Rho_p, nb_margins,
-    type = 'pearson', ensure_PSD = FALSE, cores = cores);
-    bigsimr::cor_fast( x = tmpSim, method = corType )} ,
-simplify = FALSE)
+simRho <- replicate(n = m, expr = {
+  tmpSim <- rvec(n = n , nb_Rho_p, nb_margins,
+                 type = 'pearson', ensure_PSD = FALSE, cores = cores)
+  bigsimr::cor_fast(x = tmpSim, method = corType)
+}, simplify = FALSE)
 ## find quadratic loss at each rep
-quadLoss <- unlist( lapply( simRho, rags2ridges::loss,
-                           T = nb_Rho, type = "quadratic"))
-
+quadLoss <- sapply(simRho, rags2ridges::loss, T = nb_Rho, type = "quadratic")
 ```

### 2020-12-22 070-supplementary-materials.Rmd
4th draft. Fix funding information.

@@
 
-Research reported in this publication was supported by grants from NV INBRE and the National Institutes of General Medical Sciences (NIH 5U54GM104944, GM103440).
+Research reported in this publication was supported by grants from the National Institutes of General Medical Sciences (5 U54 GM104944, GM103440) from the National Institutes of Health.
 

### 2020-12-22 040-monte-carlo-evaluation.Rmd
4th draft. Remove latex styling

@@
 ## devtools::install_github("SchisslerGroup/bigsimr", ref = 'develop')
-knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
+## knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
 # reticulate::use_condaenv("bigsimr")
@@
 
-```{r ch040-biMAEtable, results='asis', cache=F}
+Figure \@ref(fig:ch040-bPlot) displays the aggregated bivariate simulation results.
+Table \@ref(tab:ch040-BiError) contains the mean absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.
+
+```{r ch040-BiError, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
 ## tabMSE  <- allDat %>%
@@
 ##    summarize( mse = mean( ( rho - rho_hat )^2 ) )
-tabMSE  <- allDat %>%
+tabMAE  <- allDat %>%
     ## filter( N == 100000) %>%
@@
 
-knitr::kable(tabMSE,
-             format = 'latex',
+## AGS 12/22/20: Removing the latex formating fixes the html render.
+## But how to improve latex table formating? Better to use the template
+knitr::kable(tabMAE,
+             ## format = 'latex',
              col.names = c("No. of random vectors",
@@
                            "Mean abs. error"),
-             booktabs = TRUE,
+             ## booktabs = TRUE,
              caption = 'Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.',
-             linesep = c('', '',  '\\addlinespace')
+             ## linesep = c('', '',  '\\addlinespace')
 )
@@
 
-Figure \@ref(fig:ch040-bPlot) displays the aggregated bivariate simulation results.
-Table \@ref(tab:ch040-biMAEtable) contains the mean absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.
 Overall, the studies show that our methodology is generally accurate across the entire range of possible correlation values for the rank-based dependency measures, at least in these limited simulation settings for the rank-based correlations.

### 2020-12-23 020-simulation-algorithm.Rmd
4th draft. Address Ed comments Part I.

@@
 
-## `bigsimr::rvec` for random vector generation  {#rand-vec-gen}
-
-We now describe `bigsimr::rvec`, our algorithm to generate random vectors, which mirrors the classical NORTA algorithm above with some modifications for rank-based dependency matching:
-
-1. Pre-processing for nonparametric dependency matching.  
-	(i) Convert ${\bf R_{Kendall}}$ or ${\bf R_{Spearman}}$ to the corresponding MVN correlation ${\bf R_{Pearson}}$.
-	(ii) Check that ${\bf R_{Pearson}}$ is semi-positive definite.  
-	(iii) If not, compute the nearest correlation matrix ${\bf \widetilde{R}_{Pearson}}$ to ${\bf R_{Pearson}}$ in the Frobenius norm.  
-2. Gaussian copula construction.  
-	(i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_{Pearson}})$.  
-	(ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
-3. Quantile evaluations.  
-	(i) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;
-
-The pre-processing (Step 1) takes advantage of the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertKendall) or \@ref(eq:convertSpearman), respectively (implemented as `bigsimr::cor_covert`).
-
-A complication often arises at this stage: the resultant matrix may become indefinite, either through numerical error or naturally occurring in the correlation conversion.
-Researchers working in multivariate computation frequently encounter such difficulties and need to find a close positive (semi-)definite matrix.
-A widely-available routine for this task in `R` is `matrix::nearPD`, though it is not suitable for high dimensions.
-To overcome this issue, we introduce `bigsimr::nearPSD` --- an implementation of @QS2006's method that computes the nearest correlation matrix to a given symmetric matrix in the Frobenius norm.
-This routine as many applications beyond our primary goal of random vector generation.
-
-Once the target margins and algorithm inputs are determined, Steps 2 and 3 are essentially a NORTA algorithm with modern high-performance computing implementations.
-Specifically, step 2i uses either the efficient multi-core multivariate normal simulator `mvnfast`[@Fasiolo2016] or Google's JAX python library for GPU acceleration of the Cholesky factorization and matrix multiplication.
-
-## A note on NORTA in higher dimensions
+## NORTA in higher dimensions
 
@@
 This provides evidence that our nearest positive semi-definite (PSD) augmented approach will maintain reasonable accuracy if our input matching scheme returns an indefinite matrix, at least for the rank-based matching scheme described above.
+
+## `bigsimr::rvec` for random vector generation  {#rand-vec-gen}
+
+- Discuss how the HD NORTA defective problem and the scalability issues lead to our algorithm
+We now describe `bigsimr::rvec`, our algorithm to generate random vectors, which mirrors the classical NORTA algorithm above with some modifications for rank-based dependency matching:
+
+1. Mapping step  
+   i. Convert the target Spearman correlation matrix ${\bf R_{Spearman}}$ to the corresponding MVN Pearson correlation ${\bf R_X}$.   
+   or  
+i$^\prime$. Convert the target Kendall $\tau$ matrix ${\bf R_{Kendall}}$ to the corresponding MVN Pearson correlation ${\bf R_X}$.
+2. Checking admissibility and nearest correlation computation
+    (i) Check that ${\bf R_{X}}$ is a correlation matrix, a PSD matrix with 1's along the diagonal.   
+	(ii) If ${\bf R_{X}}$ is a correlation matrix, the input matrix is *admissible* in this scheme.  
+	Otherwise,  
+(ii$^\prime$) Replace ${\bf R_{X}}$ with the nearest correlation matrix ${\bf \widetilde{R}_{X}}$, in the Frobenius norm.
+3. NORTA step
+	(i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_{X}})$.  
+	(ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
+	(iii) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;
+
+The *Mapping step* takes advantage of the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertKendall) or \@ref(eq:convertSpearman), respectively (implemented as `bigsimr::cor_covert`).
+
+Our algorithm allows the generation high-dimensional data with arbitrary marginal distributions with a broad class of admissible Spearman correlation matrices and Kendall tau matrices.
+The admissible classes consist of the matrices that map to a Pearson correlation matrix.
+In particular, if we let X be MVN with N components and set
+
+$S_P = \{ R_p : R_p is a correlation matrix for X \}$
+$S_K = \{R_K : R_K is a Kendall tau matrix for a MVN \}$
+$S_S = \{R_s : R_s is a Spearman correlation matrix for a MVN\}$
+
+then there are 1-1 mappings between these sets.
+
+So if a user of your package says "I want a 100 dimensional RV with NB marginals where R_s is the target Spearman correlation matrix", you can easily check whether R_S is an element of the "admissible set" S_S.
+The same if they say they want the same marginals but a Kendall matrix R_K.
+As is, you check (at least based on paper) whether the Spearman satisfies 2-D bounds which is ok for 2-D problems.
+So you can provide higher dimensional information about whether something can be done using your algorithm.
+Furthermore, I would guess that these sets of admissible S_K and S_S are not that restrictive - In particular R_p approx R_s for a MVN so that would suggest the admissible set S_S is not very restrictive because R_p can be any psd matrix with 1's on diagonal.
+
+The final step implements a high-performance NORTA algorithm.
+Specifically, step 3i uses either the efficient multi-core multivariate normal simulator `mvnfast`[@Fasiolo2016] or Google's JAX python library for GPU acceleration of the Cholesky factorization and matrix multiplication and steps 3ii, 3iii are parallelized over CPU cores.
 

### 2020-12-23 060-conclusion-discussion.Rmd
4th draft. Address Ed comments Part I.

@@
 The major contributions of this work are high-dimensional scalability, flexible modeling of dependency, and an high-performance implementation with broad potential data analytic applications for modern, big-data statistical computing.
+
+Researchers working in multivariate computation frequently encounter such difficulties and need to find a close positive (semi-)definite matrix.
+A widely-available routine for this task in `R` is `matrix::nearPD`, though it is not suitable for high dimensions.
+To overcome this issue, we introduce `bigsimr::nearPSD` --- an implementation of @QS2006's method that computes the nearest correlation matrix to a given symmetric matrix in the Frobenius norm.
+This routine as many applications beyond our primary goal of random vector generation.
 

### 2020-12-23 index.Rmd
4th draft. Address JP feedback
4th draft. Address Ed comments Part I.

@@
 Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers may resort to simulation designs that posit independence or perform arbitrary data transformations.
-This greatly diminishes insights into the empirical operating characteristics of any proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness.
-This article introduces the `bigsimr` R package that provides a flexible, scaleable procedure to simulate high-dimensional random vectors with given marginal characteristics and dependency measures.
+This article introduces the `bigsimr` R package that provides a flexible, scaleable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions and a broad class of admissible Spearman correlation and Kendall $\tau$ matrices.
 We describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix.
@@
 
-With all this mind, we present a scaleable, flexible multivariate simulation algorithm. 
+With all this in mind, we present a scaleable, flexible multivariate simulation algorithm. 
 The crux of the method lies in the construction of a Gaussian copula, in the spirit of the NORTA procedure.

### 2020-12-24 020-simulation-algorithm.Rmd
4th draft. Address Ed comments Part II

@@
 
-## NORTA in higher dimensions
-
-Sklar's theorem provides a useful characterization of multivariate distributions through copulas.
+*NORTA in higher dimensions*. Sklar's theorem provides a useful characterization of multivariate distributions through copulas.
 Yet the choice of copula-based simulation algorithm affects which joint distributions may be simulated.
@@
 
-The *Mapping step* takes advantage of the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertKendall) or \@ref(eq:convertSpearman), respectively (implemented as `bigsimr::cor_covert`).
+The "mapping step" takes advantage of the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertKendall) or \@ref(eq:convertSpearman), respectively (implemented as `bigsimr::cor_covert`).
 
-Our algorithm allows the generation high-dimensional data with arbitrary marginal distributions with a broad class of admissible Spearman correlation matrices and Kendall tau matrices.
-The admissible classes consist of the matrices that map to a Pearson correlation matrix.
-In particular, if we let X be MVN with N components and set
+Our algorithm allows the generation of high-dimensional multivariate data with arbitrary marginal distributions with a broad class of admissible Spearman correlation matrices and Kendall $\tau$ matrices.
+The admissible classes consist of the matrices that map to a Pearson correlation matrix for a MVN.
 
-$S_P = \{ R_p : R_p is a correlation matrix for X \}$
-$S_K = \{R_K : R_K is a Kendall tau matrix for a MVN \}$
-$S_S = \{R_s : R_s is a Spearman correlation matrix for a MVN\}$
+In particular, if we let $X$ be MVN with $d$ components and set
+
+$$
+\begin{aligned}
+\Omega_P &= \{ R_P : R_P \,  is \, a \,  correlation \,  matrix \,  for \,  X \} \\
+\Omega_K &= \{ R_K : R_K \,  is \,  a \,  Kendall \,  \tau  \, matrix \,  for \,  X \} \\
+\Omega_S &= \{ R_S : R_S \,  is \,  a \,  Spearman \,  correlation \,  matrix \,  for \,  X\} \\
+\end{aligned}
+$$
 
 then there are 1-1 mappings between these sets.
+We conjecture informally that the sets of admissible $R_S$ and $R_K$ are not highly restrictive.
+In particular, $R_P$ is approximately $R_S$ for a MVN, suggesting that the admissible set $\Omega_S$ should be flexible as $R_P$ can be any PSD matrix with 1's along the diagonal. 
+We provide methods to check whether a target $R_S$ is an element of the *admissible set* $\Omega_S$ (and similarly for $R_K$) and also for checking that each pairwise correlation value is admissible (via the MC estimation of the 2-dimensional Frechet bounds).
 
-So if a user of your package says "I want a 100 dimensional RV with NB marginals where R_s is the target Spearman correlation matrix", you can easily check whether R_S is an element of the "admissible set" S_S.
-The same if they say they want the same marginals but a Kendall matrix R_K.
-As is, you check (at least based on paper) whether the Spearman satisfies 2-D bounds which is ok for 2-D problems.
-So you can provide higher dimensional information about whether something can be done using your algorithm.
-Furthermore, I would guess that these sets of admissible S_K and S_S are not that restrictive - In particular R_p approx R_s for a MVN so that would suggest the admissible set S_S is not very restrictive because R_p can be any psd matrix with 1's on diagonal.
+Yet this scheme still suffers similar issues as the original Pearson-matching NORTA algorithm regarding the increasing probability of encountering a non-admissible correlation matrix as dimension increases.
+In our experience, the *mapping step* for large $d$ almost always produces a $R_X$ that is not in $\Omega_P$ (or even a correlation matrix).
+The second step handles this issue by computing the nearest correlation matrix to $R_X$.
+In the RNA-seq example (Section \@ref(examples)), we demonstrate this and show how to quantify and control the approximation error.
 

### 2020-12-25 030-bigsimr-package.Rmd
4th draft. Address Ed comments Part I.
double spacing and updated bigsimr code

@@
 As discussed in Section \@ref(background), given a pair of marginal distributions the possible correlations are not free to vary between $[-1, 1]$. 
-`bigsimr::cor_bounds` estimates the theoretic lower and upper correlation bounds, using the Generate, Sort, and Correlate algorithm of @DH2011.
+`bigsimr::cor_bounds` estimates the pairwise lower and upper correlation bounds, using the Generate, Sort, and Correlate algorithm of @DH2011.
 
 ```{r ch030-cor-bounds, echo=TRUE}
-cor_bounds(margins = margins, type = type)
+cor_bounds(margins = margins, method = type)
 ```
@@
 ```{r ch030-cor-bounds-pearson, echo=TRUE}
-cor_bounds(margins = margins, type = 'pearson', reps = 1e6)
+cor_bounds(margins = margins, method = 'pearson', reps = 1e6)
 ```
@@
 ```{r ch030-sim-margins, echo=TRUE}
-x <- rvec(10000, rho, margins, type)
+x <- rvec(10000, rho, margins)
 df_sim <- as.data.frame(x)

### 2020-12-25 050-examples.Rmd
double spacing and updated bigsimr code

@@
         expr = { tmpSim <- rvec(n = n , nb_Rho_p, nb_margins,
-          type = 'pearson', ensure_PSD = FALSE, cores = cores);
+          ensure_PSD = FALSE, cores = cores);
           bigsimr::cor_fast( x = tmpSim, method = corType )} ,

### 2020-12-25 index.Rmd
double spacing and updated bigsimr code

@@
   - Duc Tran
+site: bookdown::bookdown_site
 bibliography: ["bigsimr.bib", "packages.bib"]
@@
 We describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix.
-Finally, we demonstrate the power of `bigsimr` by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d =1026$.
+Finally, we demonstrate the power of `bigsimr` by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d=1026$.
 ---

### 2020-12-25 notebooks/bigsimr.Rmd
double spacing and updated bigsimr code

---
title: "Get Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{tidyverse, patchwork}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## Installation

Please see the [Installation and Setup](https://schisslergroup.github.io/bigsimr/articles/install-bigsimr.html) page for getting `bigsimr` up and running. `bigsimr` requires a working version of Python that can be found in the path variable. It is also required that the version of python has `numpy` and `jax` installed.

## Usage

```{r setup}
library(bigsimr)
library(tidyverse)
library(patchwork)
```

We're going to show the basic use and syntax of `bigsimr` by using the New York air quality data set (`airquality`) included in the R `datasets` package. We will focus specifically on the temperature (degrees Fahrenheit) and ozone level (parts per billion).

```{r}
df <- airquality %>%
  select(Ozone, Temp) %>%
  drop_na()

glimpse(df)
```

Let's look at the joint distribution of the Ozone and Temperature

```{r}
p0 <- ggplot(df, aes(Temp, Ozone)) +
  geom_point(size = 1) +
  theme(legend.position = "none") +
  labs(x = "Temperature")

pTemp <- ggplot(df, aes(Temp)) + 
  geom_density() +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())

pOzone <- ggplot(df, aes(Ozone)) + 
  geom_density() + 
  theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank()) +
  coord_flip()

pTemp + plot_spacer() + p0 + pOzone + 
  plot_layout(widths = c(3,1), heights = c(1, 3))
```

We can see that not all margins are normally distributed; the ozone level is highly skewed. Though we don't know the true distribution of ozone levels, we can go forward assuming that it is log-normally distributed.

To simulate observations from this joint distribution, we need to estimate the correlation and the marginal parameters.

### Estimating Correlation

To estimate the correlation, we can use `R`'s function `cor` or we can use the convenience function `bigsimr::cor_fast` which can estimate Pearson, Spearman, or Kendall correlation using the fastest methods available from other packages.

```{r}
(rho <- cor_fast(df, method = "pearson"))
```

### Defining Marginal Distributions

Next we can estimate the marginal parameters. Assuming that `Temperature` is normally distributed, it has parameters

```{r}
df %>% select(Temp) %>% summarise_all(.funs = c(mean = mean, sd = sd))
```

and assuming that `Ozone` is log-normally distributed, it has parameters

```{r}
mle_mean <- function(x) mean(log(x))
mle_sd <- function(x) mean( (log(x) - mean(log(x)))^2 )

df %>% select(Ozone) %>% 
  summarise_all(.funs = c(meanlog = mle_mean, sdlog = mle_sd))
```

Finally we take the parameters and put them into an `alist` of margins.

```{r}
margins <- alist(
  qlnorm(meanlog = 3.42, sdlog = 0.743),
  qnorm(mean = 77.9, sd = 9.49)
)
```

Right now the marginal parameters must be filled in manually with the quantile function of the distribution. If the margins are homogeneous, then some automation can be implemented.

### Correlation Bounds

Given a list of margins, the theoretical lower and upper correlation coefficients can be estimated.

```{r}
cor_bounds(margins = margins, method = "pearson")
```

Since our estimated correlation is within the theoretical bounds (assuming the distributions are correct), we should be able to achieve a similar correlation in the simulated data.

### Simulating Multivariate Distributions

Let's now simulate 10,000 observations from the joint distribution using `bigsimr::rvec`

```{r}
x <- rvec(10000, rho, margins, "pearson")

df_sim <- as.data.frame(x)
colnames(df_sim) <- colnames(df)
```

and visualize the joint distribution

```{r}
p1 <- df_sim %>%
  ggplot(aes(Temp, Ozone)) +
  geom_density_2d_filled() +
  theme(legend.position = "none") +
  labs(x = "Temperature", y = "Ozone") +
  scale_y_continuous(limits = c(0, max(df$Ozone))) +
  scale_x_continuous(limits = range(df$Temp))

p1Temp <- ggplot(df_sim, aes(Temp)) + 
  geom_density(alpha = 0.5, fill = "lightseagreen") +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())

p1Ozone <- ggplot(df_sim, aes(Ozone)) + 
  geom_density(alpha = 0.5, fill = "lightseagreen") + 
  theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank()) +
  coord_flip()

p1Temp + plot_spacer() + p1 + p1Ozone + 
  plot_layout(widths = c(3,1), heights = c(1, 3))
```

## Comparison to Uncorrelated Samples

We can compare the bivariate distribution above to one where no correlation is taken into account.

```{r}
df_sim2 <- data.frame(
  Ozone = rlnorm(10000, mle_mean(df$Ozone), mle_sd(df$Ozone)),
  Temp = rnorm(10000, mean(df$Temp), sd(df$Temp))
)
```

```{r, echo=FALSE, warning=FALSE}
p2 <- df_sim2 %>%
  ggplot(aes(Temp, Ozone)) +
  geom_density_2d_filled() +
  theme(legend.position = "none") +
  labs(x = "Temperature", y = "Ozone") +
  scale_y_continuous(limits = c(0, max(df$Ozone))) +
  scale_x_continuous(limits = range(df$Temp))

p2Temp <- ggplot(df_sim2, aes(Temp)) + 
  geom_density(alpha = 0.5, fill = "lightseagreen") +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())

p2Ozone <- ggplot(df_sim2, aes(Ozone)) + 
  geom_density(alpha = 0.5, fill = "lightseagreen") + 
  theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank()) +
  coord_flip()

p2Temp + plot_spacer() + p2 + p2Ozone + 
  plot_layout(widths = c(3,1), heights = c(1, 3))
```

Notice that this uncorrelated bivariate distribution has the same marginal distributions as the correlated example above.

### 2020-12-28 030-bigsimr-package.Rmd
4th draft. working on Ed comments

--- 2020-12-22 030-bigsimr-package.Rmd
@@
-# The `bigsimr` R package
+# The `bigsimr` R package {#package}
 
@@
 Here the analyst is free to be creative.
-For this example, we do not take up goodness-of-fit considerations to determine the marginal distributions.
+For this example, we avoid goodness-of-fit considerations to determine the marginal distributions.
 But it seems sensible without domain knowledge to estimate these quantities from the data and `bigsimr` contains fast functions designed for this task.
@@
 
-To avoid manual entry, `mlist` presents an alternative to `alist` with on-the-fly evaluation.
+To circumvent manual entry, `mlist` presents an alternative to `alist` with on-the-fly evaluation.
 
@@
 As mentioned, the user must decide how to describe correlation, based on the particulars of the problem.
-For non-normal data and for improved simulation accuracy in our scheme, we advocate the use of rank-based correlations Spearman's $\rho_S$ and Kendall's $\tau$.
-But we also support approximate Pearson correlation coefficient matching, while cautioning the user to check the performance for their parametric multivariate model (see [Monte Carlo evaluations](#simulations) below for evaluation strategies and guidance).
-To aid in correlation specification, and estimation in general, we provide a high-performance function `bigsimr::cor_fast` which estimates Pearson, Spearman, or Kendall correlation using the fastest methods available.
+For non-normal data and for improved simulation accuracy/scalability in our scheme, we advocate the use of Spearman correlation matrix $R_S$ and Kendall $\tau$ matrix $R_K$.
+We also support approximate Pearson correlation coefficient matching, while cautioning the user to check the performance for their parametric multivariate model (see [Monte Carlo evaluations](#simulations) below for evaluation strategies and guidance).
+To aid in correlation specification, and estimation in general, we provide a high-performance function `bigsimr::cor_fast` which estimates Spearman, Kendall, or Pearson correlation using the fastest methods available.
 (Anyone who has tried estimating Kendall's $\tau$ using `stats::cor` can attest that the routine does not scale to even moderate dimensions).
@@
 type <- 'spearman'
-(rho <- cor_fast(df, method = "spearman"))
+(R_S <- cor_fast(df, method = "spearman"))
 ```
 
-*Checking the theoretical correlation bounds*.
-As discussed in Section \@ref(background), given a pair of marginal distributions the possible correlations are not free to vary between $[-1, 1]$. 
-`bigsimr::cor_bounds` estimates the theoretic lower and upper correlation bounds, using the Generate, Sort, and Correlate algorithm of @DH2011.
+*Checking target correlation matrix admissibility*.
+First we use `bigsimr::cor_bounds` to ensure that the pairwise target correlation values are valid prior to mapping step which constructs $R_X$ for the MVN input into NORTA (see Section \@ref(algorithms)).
+`cor_bounds` estimates the pairwise lower and upper correlation bounds using the Generate, Sort, and Correlate algorithm of @DH2011.
 
@@
 
-Since our estimated Spearman correlation $\hat{ \rho}_S$ is within the theoretical bounds, the correlation is valid as input to `bigsimr:rvec`.
-On the other hand, the bounds on the Pearson correlation coefficient $\rho_P$ between these margins is
+Since our our single estimated Spearman correlation is within the theoretical bounds, the correlation is valid as input to `bigsimr:rvec`.
+But even if the 2-dimensional bounds are satified for each pair of margins, this does not guarantee the feasibility of a $d-$variate distribution [@BF17].
+
+To provide higher dimensional feasibility/admissibility checking, we begin by mapping using `bigsimr::cor_convert` and check admissibility.
+
+```{r ch030-cor-convert, echo=TRUE}
+## Step 1. Mapping
+(R_X <- cor_convert(rho = R_S, from = type, to = 'pearson'))
+## Step 2. Check admissibily
+is_valid_correlation <- function(rho) {
+  all(diag(rho) == 1) && matrixcalc::is.positive.semi.definite(rho)
+}
+is_valid_correlation( R_X )
+matrixcalc::is.symmetric.matrix( R_X )
+all(diag(R_X) == 1)
+cor_nearPSD( G = R_X )
+```
+
+We note that the bounds on the Pearson correlation coefficient $R_P$ between these margins is restricted as seen below:
 
 ```{r ch030-cor-bounds-pearson, echo=TRUE}
-cor_bounds(margins = margins, type = 'pearson', reps = 1e6)
+cor_bounds(margins = margins, type = 'pearson', reps = 1e5)
 ```
@@
 ```{r ch030-sim-margins, echo=TRUE}
-x <- rvec(10000, rho, margins, type)
+x <- rvec(10000, R_S, margins, type)
 df_sim <- as.data.frame(x)

### 2020-12-28 050-examples.Rmd
4th draft. working on Ed comments

--- 2020-12-22 050-examples.Rmd
@@
 ## Estimate Spearman's correlation on the count data
-corType <- 'spearman'
-nb_Rho <- bigsimr::cor_fast(brca, method = corType)
+type <- 'spearman'
+R_S <- bigsimr::cor_fast(brca, method = type)
 ```
@@
 Once the functions are defined/executed to complete marginal estimation, we specify targets and generate the desired random vectors using `rvec`:
+
+```{r ch050-cor-check, echo=TRUE, eval=FALSE, cache = FALSE}
+## 1. Check 2D bounds
+## OUR CODE HERE
+
+## 2. Check admissibility
+R_X <- cor_convert(rho = R_S, from = type, to = 'pearson')
+## diag(R_X) <- 1
+is_valid_correlation <- function(rho) {
+  all(diag(rho) == 1) && matrixcalc::is.positive.semi.definite(rho)
+}
+is_valid_correlation( R_X )
+
+system.time( R_Xtilde <- cor_nearPSD( G = R_X ) )
+## diag(R_X) <- 1
+is_valid_correlation( R_X )
+
+norm( x = R_X, type = "F" )
+norm( x = R_Xtilde, type = "F" )
+norm( x = R_X - R_Xtilde, type = "F" )
+
+```
 

### 2020-12-29 020-simulation-algorithm.Rmd

@@
 1. Mapping step  
-   i. Convert the target Spearman correlation matrix ${\bf R_{Spearman}}$ to the corresponding MVN Pearson correlation ${\bf R_X}$.   
+   i. Convert the target Spearman correlation matrix $R_S$ to the corresponding MVN Pearson correlation $R_X$.   
    or  
-i$^\prime$. Convert the target Kendall $\tau$ matrix ${\bf R_{Kendall}}$ to the corresponding MVN Pearson correlation ${\bf R_X}$.
+i$^\prime$. Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$.
 2. Checking admissibility and nearest correlation computation
-    (i) Check that ${\bf R_{X}}$ is a correlation matrix, a PSD matrix with 1's along the diagonal.   
-	(ii) If ${\bf R_{X}}$ is a correlation matrix, the input matrix is *admissible* in this scheme.  
+    (i) Check that $R_X$ is a correlation matrix, a PSD matrix with 1's along the diagonal.   
+	(ii) If $R_X$ is a correlation matrix, the input matrix is *admissible* in this scheme.  
 	Otherwise,  
-(ii$^\prime$) Replace ${\bf R_{X}}$ with the nearest correlation matrix ${\bf \widetilde{R}_{X}}$, in the Frobenius norm.
+(ii$^\prime$) Replace $R_X$ with the nearest correlation matrix $R_X$, in the Frobenius norm.
 3. NORTA step
-	(i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_{X}})$.  
+	(i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)$.  
 	(ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
@@
 The second step handles this issue by computing the nearest correlation matrix to $R_X$.
-In the RNA-seq example (Section \@ref(examples)), we demonstrate this and show how to quantify and control the approximation error.
+In Section \@ref(package) we provide a basic description of how to quantify and potentially control the approximation error.
+Further, the RNA-seq example in Section \@ref(examples) provides an illustration of this in practice.
 

### 2020-12-29 030-bigsimr-package.Rmd

@@
 ```{r ch030-cor-bounds, echo=TRUE}
-cor_bounds(margins = margins, type = type)
+cor_bounds(margins = margins, method = type)
 ```
@@
 ```{r ch030-cor-bounds-pearson, echo=TRUE}
-cor_bounds(margins = margins, type = 'pearson', reps = 1e5)
+cor_bounds(margins = margins, method = 'pearson', reps = 1e6)
 ```
@@
 ```{r ch030-sim-margins, echo=TRUE}
-x <- rvec(10000, R_S, margins, type)
+x <- rvec(10000, rho, margins)
 df_sim <- as.data.frame(x)

### 2020-12-29 040-monte-carlo-evaluation.Rmd
4th draft. Address Ed comments Part I.

@@
         adjustForDiscrete <- sim_pars$adjustForDiscrete[i]
-        tmp_bounds <- cor_bounds(margins, type = type)
+        tmp_bounds <- cor_bounds(margins, method = type)
         cor_lo <- tmp_bounds$lower[1,2] + eps
@@
 
-        tmp_bounds <- cor_bounds(margins, type = type)
+        tmp_bounds <- cor_bounds(margins, method = type)
         cor_lo <- tmp_bounds$lower[1,2] + eps
@@
         if ( type == 'pearson' ) {
-            tmp_bounds <- cor_bounds(margins, type = type)
+            tmp_bounds <- cor_bounds(margins, method = type)
             cor_lo <- tmp_bounds$lower[1,2] + eps
@@
 
-## AGS 12/22/20: Removing the latex formating fixes the html render.
-## But how to improve latex table formating? Better to use the template
+## AGS 12/22/20: Removing the latex formatting fixes the html render.
+## But how to improve latex table formatting? Better to use the template
 knitr::kable(tabMAE,

### 2020-12-29 030-bigsimr-package.Rmd
Begin rewriting after moving to bigsimr develop branch

--- 2020-12-28 030-bigsimr-package.Rmd
@@
 is_valid_correlation( R_X )
-matrixcalc::is.symmetric.matrix( R_X )
-all(diag(R_X) == 1)
-cor_nearPSD( G = R_X )
 ```
@@
 ```{r ch030-cor-bounds-pearson, echo=TRUE}
-cor_bounds(margins = margins, type = 'pearson', reps = 1e5)
+cor_bounds(margins = margins, method = 'pearson', reps = 1e5)
 ```
@@
 ```{r ch030-sim-margins, echo=TRUE}
-x <- rvec(10000, R_S, margins, type)
+x <- rvec(n = 10000, rho = R_X, margins)
 df_sim <- as.data.frame(x)

### 2020-12-29 index.Rmd
Begin rewriting after moving to bigsimr develop branch

--- 2020-12-23 index.Rmd
@@
 ```{r indexSetup, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide", cache=F}
+library(bigsimr)
 library(tidyverse)

### 2020-12-29 030-bigsimr-package.Rmd

--- 2020-12-29 030-bigsimr-package.Rmd
@@
 cor_bounds(margins = margins, method = 'pearson', reps = 1e6)
+
 ```

### 2020-12-29 index.Rmd
Update code to bigsimr_0.9.3

@@
 ## bookdown::preview_chapter("index.Rmd")
+rmarkdown::render_site(encoding = 'UTF-8')
 bookdown::serve_book()

### 2021-01-11 010-background.Rmd
2nd draft. Begin integrating new RNA-seq data

@@
 
-```{r ch010-myProb, include=FALSE, cache = TRUE}
-myProb <- 0.95
+```{r ch010-myD, include=FALSE, cache = TRUE, eval=T}
+d <- 1000
 ```
@@
 ```{r ch010-processBRCA, echo=FALSE, eval=TRUE, results='none', cache=F}
-brcaRDS <- paste0( "data/brca", round(myProb*100), ".rds" )
-if ( !file.exists( brcaRDS ) | !file.exists( './data/examplePatients.txt' ))  {
-    ## full workflow simulating RNA-seq data
-    allDat <- readRDS( file = "~/Downloads/complete_processed_tcga2stat_RNASeq2_with_clinical.rds" )
-    lastClinical <- which( names(allDat) == 'tumorsize' )
-    brca <- allDat[ allDat$disease == "BRCA", ]
-    ## Patient ID
-    writeLines( as.character( brca$patient ), './data/examplePatients.txt' ) 
-    brca <- brca[ , (lastClinical+1):ncol(allDat) ]
-    dim(brca) ## num of genes 20501
-    ## compute the median expression! avoid the 0s
-    brcaMedian <- apply(brca, 2, median)
-    ## retain top (1-myProb)*100% highest expressing genes for illustration
-    cutPoint <- quantile( x = brcaMedian, probs = myProb )
-    ## genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
-    genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
-    brca <- brca[ , genesToKeep ]
-    ## ncol(brca) / 20501
-    ## print(d <- length(genesToKeep))
-    ## save data as rds
+brcaRDS <- paste0( "data/brca", d, ".rds" )
+if ( !file.exists( brcaRDS ) | !file.exists( './data/examplePatients.txt' ) )  {
+    brca0 <- TCGA2STAT::getTCGA(disease = "BRCA",
+                            data.type = "RNASeq",
+                            p = 6L)
+    brca1 <- as.data.frame(t(brca0$dat))
+    gene_median <- apply(brca1, 2, median)
+    gene_median_order <- order(gene_median, decreasing = TRUE)
+    brca  <- brca1[,head(gene_median_order, d)]
+    writeLines( substring( row.names( brca ), first=1, last= 12 ), './data/examplePatients.txt' ) 
     saveRDS(brca, brcaRDS)
@@
 brca <- readRDS(file = brcaRDS)
-d <- ncol(brca)
 ```
@@
 For simplicity, we only consider high expressing genes.
-In turn, we begin by filtering to retain the top `r (1-myProb)*100`% of highest- expressing genes (in terms of median expression) of the 20,501 gene measurements from $N=`r nrow(brca)`$ patients' tumor samples, resulting in $d=`r d`$ genes.
-This gives a massive number of pairwise dependencies among the marginals (specifically, $`r choose(d,2)`$ correlation parameters).
-We further process the data to illustrate our methodology's flexible and robust simulation scheme, while better aligning with the actual data-generating process, by rounding @Li2011c RNA-seq by Expectation Maximization (RSEM) values to counts.
+In turn, we begin by filtering to retain the top `r d`% of highest-expressing genes (in terms of median expression) of the 20,502 gene measurements from $N=`r nrow(brca)`$ patients' tumor samples.
+This gives a great number of pairwise dependencies among the marginals (specifically, $`r choose(d,2)`$ correlation parameters).
 Table \@ref(tab:ch010-realDataTab) displays RNA-seq counts for three selected high-expressing genes for the first five patients' breast tumor samples.
@@
 if ( !file.exists( './data/exampleGenes.rds' ) )  {
-    set.seed(10192020)
+    set.seed(01112021)
     exampleGenes <- colnames(brca)[sort(sample(ncol(brca), 3))]

### 2021-01-11 index.Rmd
2nd draft. Begin integrating new RNA-seq data

@@
 abstract:  It is critical to realistically simulate data when conducting Monte Carlo studies and evaluating quantitative methods.
-But measurements are often correlated and high dimensional in this era of big data, however, such as data obtained through high-throughput biomedical experiments.
-Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers may resort to simulation designs that posit independence or perform arbitrary data transformations.
+Measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments.
+Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations.
 This article introduces the `bigsimr` R package that provides a flexible, scaleable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions and a broad class of admissible Spearman correlation and Kendall $\tau$ matrices.
 We describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix.
-Finally, we demonstrate the power of `bigsimr` by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d=1026$.
+Finally, we demonstrate example applications enabled via `bigsimr` by applying these functions to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=878$ patients and dimension $d=1000$.
 ---

### 2021-01-13 010-background.Rmd
2nd draft. Polishing ch050 after new RNA-seq

@@
 if ( !file.exists( './data/exampleGenes.rds' ) )  {
-    set.seed(01112021)
+    set.seed(01122021)
     exampleGenes <- colnames(brca)[sort(sample(ncol(brca), 3))]

### 2021-01-13 020-simulation-algorithm.Rmd
2nd draft. Polishing ch050 after new RNA-seq

@@
 i$^\prime$. Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$.
-2. Checking admissibility and nearest correlation computation
+2. Checking admissibility and, if needed, nearest correlation computation
     (i) Check that $R_X$ is a correlation matrix, a PSD matrix with 1's along the diagonal.   

### 2021-01-13 050-examples.Rmd
2nd draft. Polishing ch050 after new RNA-seq

--- 2020-12-25 050-examples.Rmd
@@
 ```{r ch050-readBRCA, echo=FALSE, eval=TRUE, cache=F}
-## full workflow simulating RNA-seq data
-brcaRDS <- 'data/brca95.rds'
-myProb = 0.95
+d <- 1000
+brcaRDS <- paste0( "data/brca", d, ".rds" )
 brca <- readRDS(file = brcaRDS)
-d <- ncol(brca)
 ## smaller set for prototyping
@@
 probs <- apply( unname(as.matrix(brca)), 2, nbinom_mom )[2, ]
+## construct margins
+nb_margins <- make_nbinom_alist(sizes, probs)
 ```
@@
 
+Now we check admissibility of specified correlation matrix.
+We begin with the pairwise, 2D MC estimated bounds.
+And then follow the `r d`D bounds.
+
+```{r ch050-cor-check, echo=TRUE, eval=TRUE, cache = FALSE}
+## 1. Mapping step first
+R_X <- cor_convert(rho = R_S, from = type, to = 'pearson')
+## 2a. Check admissibility
+is_valid_correlation <- function(rho) {
+  all(diag(rho) == 1) && matrixcalc::is.positive.semi.definite(rho)
+}
+is_valid_correlation( R_X )
+
+## 2b. compute nearest correlation
+if ( ! is_valid_correlation( R_X ) ) {
+    ## system.time( R_Xtilde <- cor_nearPSD( G = R_X )  )
+    R_Xtilde <- cor_nearPSD( G = R_X )
+    ## Quantify the error
+    targets <- R_X[ lower.tri( R_X ) ]
+    approximates <- R_Xtilde[ lower.tri( R_Xtilde ) ]
+    ## R_Xtilde[1:5, 1:5]
+    ## norm( x = R_X - R_Xtilde, type = "F" )
+    R_X  <- R_Xtilde
+}
+## qplot( x = targets, y = approximates ) + geom_abline(slope=1,intercept = 0)
+## qplot( x = targets - approximates ) + geom_abline(slope=1,intercept = 0)
+summary(targets - approximates)
+```
+
+While the exact dxd Spearman correlation matrix is not strictly admissible in our scheme (seen by the non-PSD result above), the approximation is of sufficient quality for most applications with a maximum pairwise of `r max( targets - approximations)` and average error of `r mean( targets - approximations)`.
+
 ```{r ch050-runBRCA-echo, echo=TRUE, eval=FALSE, cache = FALSE}
-## Set the number of random vectors (B in text)
-n <- 10000
-## construct margins
-nb_margins <- make_nbinom_alist(sizes, probs)
-## run sims
-sim_nbinom <- rvec(n, nb_Rho, nb_margins, type = corType,
-                   ensure_PSD = TRUE, cores = cores) 
+sim_nbinom <- rvec(10000, R_X, nb_margins, ensure_PSD = FALSE, cores = cores) 
 colnames(sim_nbinom) <- names(brca)
@@
 ```{r ch050-runBRCA, echo=FALSE, results=FALSE, cache = FALSE}
-nb_margins <- make_nbinom_alist(sizes, probs)
-brcaSimRDS <- paste0( "results/brca", round(myProb*100), "sim.rds" )
+brcaSimRDS <- paste0( "results/brca", d, "sim.rds" )
 if( !file.exists(brcaSimRDS) ) {
@@
     set.seed(44)
-    system.time( sim_nbinom <- rvec(n, nb_Rho, nb_margins, type = corType,
-                                    ensure_PSD = TRUE, cores = cores) )
-    colnames(sim_nbinom) <- names(brca)
+    sim_nbinom <- rvec(10000, R_X, nb_margins, ensure_PSD = FALSE, cores = cores)
     saveRDS( sim_nbinom, brcaSimRDS )
@@
 
-```{r ch050-simDataFig, echo=F, fig.align='center', fig.cap='Simulated data for three selected high-expressing genes replicates the estimated data structure.'}
+```{r ch050-simDataFig, echo=F, fig.align='center', fig.cap='Simulated data for three selected high-expressing genes generally replicates the estimated data structure.
+The data do not exhibit outlying points, but do possess the desired Spearman correlations, central tendencies, and discrete values.'}
 knitr::include_graphics('fig/ch050-simDataFig.png')
@@
     ## estimate rho from sims and plot
-    nb_Rho_hat <- cor_fast(sim_nbinom, method = corType)
+    R_S_hat <- cor_fast(sim_nbinom, method = type)
     ## triplot
@@
     par(mfrow=c(1,3))
-    plot(nb_Rho[lower.tri(nb_Rho)], nb_Rho_hat[lower.tri(nb_Rho_hat)],
+    plot(R_S[lower.tri(R_S)], R_S_hat[lower.tri(R_S_hat)],
          xlim = c(-1, 1), ylim = c(-1, 1),

### 2021-01-15 050-examples.Rmd
2nd draft. revise ch050 with new brca data.

@@
 
-While the exact dxd Spearman correlation matrix is not strictly admissible in our scheme (seen by the non-PSD result above), the approximation is of sufficient quality for most applications with a maximum pairwise of `r max( targets - approximations)` and average error of `r mean( targets - approximations)`.
+While the exact dxd Spearman correlation matrix is not strictly admissible in our scheme (seen by the non-PSD result above), the approximation is of sufficient quality for most applications with a maximum pairwise of `r max( targets - approximates)` and average error of `r mean( targets - approximates)`.
 
@@
 ## Simulate random vectors equal to the sample size
-n <- nrow(brca)
-## convert outside for faster simulation
-nb_Rho_p <- bigsimr::cor_convert( rho = nb_Rho,
-                                 from = corType, to = "pearson" )
-## ensure PSD
-nb_Rho_p <- bigsimr::cor_nearPSD( G = nb_Rho_p )
+N <- nrow(brca)
 ## create m random vectors and estimate correlation
 simRho <- replicate(n = m, expr = {
-  tmpSim <- rvec(n = n , nb_Rho_p, nb_margins,
-                 type = 'pearson', ensure_PSD = FALSE, cores = cores)
-  bigsimr::cor_fast(x = tmpSim, method = corType)
+  tmpSim <- rvec(n = N , R_X, nb_margins, cores = cores)
+  bigsimr::cor_fast(x = tmpSim, method = type)
 }, simplify = FALSE)
 ## find quadratic loss at each rep
-quadLoss <- sapply(simRho, rags2ridges::loss, T = nb_Rho, type = "quadratic")
+quadLoss <- sapply(simRho, rags2ridges::loss, T = R_S, type = "quadratic")
 ```
 
-```{r ch050-quadloss, echo=F, eval=T, cache=FALSE}
-quadLossRDS <- paste0( "results/brca", round(myProb*100), "quadLoss.rds" )
-if ( !file.exists( quadLossRDS) )  {
-    n <- nrow(brca)
-    ## convert outside for faster simulation
-    nb_Rho_p <- bigsimr::cor_convert( rho = nb_Rho,
-                                     from = corType, to = "pearson" )
-    ## ensure PSD
-    nb_Rho_p <- bigsimr::cor_nearPSD( G = nb_Rho_p )
+```{r ch050-lossResults, echo=F, eval=T, cache=FALSE}
+## quadLossRDS <- paste0( "results/brca", d, "_quadLoss.rds" )
+frobeniusRDS <- paste0( "results/brca", d, "_frobenius.rds" )
+if ( !file.exists( frobeniusRDS) )  {
+    N <- nrow(brca)
     ## create m random vectors and estimate correlation
-    simRho <- replicate(n = m,
-        expr = { tmpSim <- rvec(n = n , nb_Rho_p, nb_margins,
-          ensure_PSD = FALSE, cores = cores);
-          bigsimr::cor_fast( x = tmpSim, method = corType )} ,
-     simplify = FALSE)
+    simRho <- replicate(n = m, expr = {
+        tmpSim <- rvec(n = N , R_X, nb_margins, cores = cores)
+        bigsimr::cor_fast(x = tmpSim, method = type)
+    }, simplify = FALSE)
     ## find quadratic loss at each rep
-    quadLoss <- unlist( lapply( simRho, rags2ridges::loss,
-                               T = nb_Rho, type = "quadratic"))
-    saveRDS(quadLoss, quadLossRDS)
+    ## simRho[[1]][1:5,1:5]
+    ## R_S[1:5,1:5]
+    ## rags2ridges::loss(simRho[[1]], R_S, type = 'frobenius')
+    frobenius <- sapply(simRho, rags2ridges::loss, T = R_S, type = "frobenius")
+    saveRDS(frobenius, frobeniusRDS)
 }
-quadLoss <- readRDS( quadLossRDS )
+frobenius <- readRDS( frobeniusRDS )
 ```
@@
 
-```{r ch050-quadloss-summary, echo=TRUE, eval=T, cache=FALSE}
-summary(quadLoss)
+```{r ch050-loss-summary, echo=TRUE, eval=T, cache=FALSE}
+summary(frobenius)
 ```
@@
 This distribution could be compared to high-dimensional designed covariance estimators for guidance on whether the additional complexity and computation time are warranted.
-
-<!-- 
-## Monte Carlo estimation of MSE
-
-Monte Carlo (MC) methods offer flexible and straightfoward statistical inferenec in this typically difficult analytic setting.
-MC methods are routinely used many statistical inferential tasks including estimation, hypothesis testing, error rates, and empirical interval coverage rates.
-For an concise introduction to these Methos, see, for example @Rizzo2007, Ch. 6.
-As a simple example, let's now employ `bigsimr` to estimate the mean squared error (MSE) of the method of moments estimators we used in section X.
-
-Specifically, 
-TYPESET THE MATH HERE. REF to MoM above.
-
-```{r ch050-mom-mse, echo=TRUE, eval=FALSE, cache=FALSE}
-## system( "rm results/mseSizes.rds")
-brcaSimMSEsizesRDS <- paste0( "results/brca", round(myProb*100), "simMSEsizes.rds" )
-if ( !file.exists( brcaSimMSEsizesRDS) )  {
-    n <- nrow(brca)
-    m <- 100
-    momSizes <- replicate(n = m, expr =
-                                     { tmpSim <- rvec(n, nb_Rho, brcaMargins, type = corType, ensure_PSD = TRUE, cores = cores)
-                                         tmpMom <- apply( unname(as.matrix(tmpSim)), 2, estimateNegBinMoM )
-                                         tmpSizes <- unlist( lapply( X = tmpMom, function(x) x$size ) )
-                                     }
-                          )
-    sqDiffSizes <- apply( momSizes, 2, function(x) { (x  - sizes)^2 } )
-    mseSizes <- rowMeans( sqDiffSizes )
-    saveRDS(mseSizes, brcaSimMSEsizesRDS)
-}
-mseSizes <- readRDS( brcaSimMSEsizesRDS )
-```
-
-Figure X.
-
-```{r ch050-fig-mseSizes, echo = FALSE, eval=FALSE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Real data for the 3 selected high-expressing genes'}
-mseSizes <- readRDS(file = brcaSimMSEsizesRDS)
-qplot(mseSizes, bins = 20) + theme_bw()
-```
-
-One could then compare the MSE of the baisc MoM estimation scheme to other strategies ( such as MLE, trimmed approaches, etc.) and decide whether a more computational costly approach is warrented or if the MoM will suffice.
-
-your comment -->

### 2021-01-21 030-bigsimr-package.Rmd
Update code to bigsimr_0.9.3
2nd draft. Insert Tom / Ania bivariate example

@@
 
-## Basic use illustrated through a minimal example
+## Bivariate example of basic use
 
@@
 
-```{r ch030-margins-alist, echo=TRUE}
+```{r ch030-margins-alist, echo=TRUE, cache=T}
 margins <- alist(
     qnorm(mean = 77.871, sd = 9.4855),
-    qlnorm(meanlog = 3.419, sdlog = 0.6967),
+    qlnorm(meanlog = 3.419, sdlog = 0.6967)
 )
@@
 ```{r ch030-margins-mlist, echo=TRUE}
-margins <- mlist(
+mlist(
   qnorm(mean = mean(df$Temp), sd = sd(df$Temp)),
@@
 )
-margins
 ```
@@
 
-We note that the bounds on the Pearson correlation coefficient $R_P$ between these margins is restricted as seen below:
+The bounds on the Pearson correlation coefficient $R_P$ between these margins is restricted as seen below in our MC estimated correlation bounds:
 
@@
 cor_bounds(margins = margins, method = 'pearson', reps = 1e6)
+```
 
-```
+Our MC estimate of the bounds slightly underestimates the theoretic bounds of $(-0.881, 0.881)$. An analytic derivation is presented as an Appendix.
 
@@
 ```{r ch030-sim-margins, echo=TRUE}
-x <- rvec(10000, rho, margins)
+x <- rvec(10000, R_X, margins)
 df_sim <- as.data.frame(x)

### 2021-01-21 080-appendix.Rmd
2nd draft. Insert Tom / Ania bivariate example

# Appendix {.unnumbered #appendix}

\noindent Consider a bivariate example where we have a correlated $(Y_1, Y_2)^\top$ with $Y_1\sim N(\mu_1, \sigma_1^2)$ (normal) and $Y_2\sim LN(\mu_2, \sigma_2^2)$ (lognormal).

\vspace{0.1in}

\noindent First, let us note that when we get such a vector from BigSimR package then in fact it can be represented as

\begin{equation}
\label{kram1}
(Y_1, Y_2)^\top \stackrel{d}{=} \left(X_1, e^{X_2}\right)^\top
\end{equation}

where $(X_1, X_2)^\top \sim N_2(\boldsymbol \mu, \boldsymbol \Sigma)$, which is bivariate normal with mean vector $\boldsymbol \mu = (\mu_1, \mu_2)^\top$ and variance-covariance matrix

\begin{equation}
\label{kram2}
\boldsymbol \Sigma = 
\left[
\begin{array}{cc}
\sigma_1^2 & \rho \sigma_1\sigma_2\\
\rho \sigma_1\sigma_2 & \sigma_2^2
\end{array}
\right].
\end{equation}

To see this, consider the three steps of the NORTA construction:

\begin{enumerate}

\item Generate $(Z_1, Z_2)^\top \sim N_2(\boldsymbol 0, \boldsymbol R)$, where

\begin{equation}
\label{kram3}
\boldsymbol R = 
\left[
\begin{array}{cc}
1 & \rho \\
\rho & 1
\end{array}
\right].
\end{equation}

\item Transform $(Z_1, Z_2)^\top$ to $(U_1, U_2)^\top$ viz. $U_i =\Phi(Z_i)$,  $i=1,2$, where $\Phi(\cdot)$ is the standard normal CDF.

\item Return $(Y_1, Y_2)^\top$, where $Y_i=F_i^{-1}(U_i)$, $i=1,2$, where $F_i(\cdot)$ is the CDF of $Y_i$ and $F_i^{-1}(\cdot)$ is its inverse (the quantile function).

\end{enumerate}

In this example, $F_1$ is the CDF of $N(\mu_1, \sigma_1^2)$ while $F_2$ is the CDF of $LN(\mu_2, \sigma_2^2)$, so that the two required quantile functions are

\begin{equation}
\label{kram4}
F_1^{-1}(u) = \mu_1+\sigma_1 \Phi^{-1}(u)\,\,\, \mbox{and} \,\,\, F_2^{-1}(u) = e^{\mu_2+\sigma_2 \Phi^{-1}(u)}, 
\end{equation}

as can be seen by standard calculation. Thus, when we apply Step 3 of the NORTA algorithm, we get

\begin{equation}
\label{kram5}
F_1^{-1}(U_1) = \mu_1+\sigma_1 \Phi^{-1}(\Phi(Z_1)) =  \mu_1+\sigma_1 Z_1 \,\,\, \mbox{and} \,\,\, F_2^{-1}(U_2) = e^{\mu_2+\sigma_2 \Phi^{-1}(\Phi(Z_2))} = e^{\mu_2+\sigma_2 Z_2}, 
\end{equation}

where $(X_1, X_2)^\top =  (\mu_1+\sigma_1 Z_1, \mu_2+\sigma_2 Z_2)^\top \sim N_2(\boldsymbol \mu, \boldsymbol \Sigma)$. Consequently, the vector $(Y_1, Y_2)^\top$ obtained in Step 3 of the algorithm has the structure provided by (\ref{kram1}).

\vspace{0.1in}

\noindent Next, we provide the exact covariance structure of the random vector $(Y_1, Y_2)^\top$ given by (\ref{kram1}), and relate the correlation of $Y_1$ and $Y_2$ to $\rho$, which is the correlation of the normal variables $Z_1$, $Z_2$ (and also $X_1$ and $X_2$). A straightforward albeit somewhat tedious algebra produces the following result.

\begin{lemma}
Let ${\bf Y} = (Y_1, Y_2)^\top$ admit the stochastic representation (\ref{kram1}), where ${\bf X} = (X_1, X_2)^\top \sim N_2(\boldsymbol \mu, \boldsymbol \Sigma)$ with $\boldsymbol \mu$ and $\boldsymbol \Sigma$ as above. Then the mean vector and the variance-covariance matrix of ${\bf Y}$ are given by

\begin{equation}
\label{kram6}
\boldsymbol \mu_{{\bf Y}} = \left(\mu_1, e^{\mu_2+\sigma_2^2/2}\right)^\top
\end{equation}

and

\begin{equation}
\label{kram7}
\boldsymbol \Sigma = 
\left[
\begin{array}{cc}
\sigma_1^2 & \rho \sigma_1\sigma_2  e^{\mu_2+\sigma_2^2/2}  \\
\rho \sigma_1\sigma_2 e^{\mu_2+\sigma_2^2/2} & \left[ e^{\mu_2+\sigma_2^2/2}\right]^2 \left[e^{\sigma_2^2} -1 \right],
\end{array}
\right].
\end{equation}

respectively. 
\end{lemma}

\begin{proof}
The values of the means and the variances are obtained immediately from normal marginal distribution of $Y_1$ and lognormal marginal distribution of $Y_2$. It remains to establish the covariance of $Y_1$ and $Y_2$,

\begin{equation}
\label{covy1y2}
\mbox{Cov}(Y_1, Y_2) = \mathbb E(Y_1 Y_2) - \mathbb E(Y_1)\mathbb E(Y_2), 
\end{equation}

and in particular the first expectation on the right-hand-side above. By using the tower property of expectations, the latter expectation can be expressed as

\begin{equation}
\label{kre1}
\mathbb E(Y_1 Y_2)  =  \mathbb E\left(X_1 e^{X_2}\right)  =  \mathbb E \left\{ \mathbb E \left(X_1 e^{X_2} | X_1 \right) \right\} = \mathbb E \left\{ X_1 \mathbb E \left(e^{X_2} | X_1 \right) \right\}.
\end{equation}

Further, by using the fact that the conditional distribution of $X_2$ given $X_1=x_1$ is normal with mean $\mathbb E(X_2|X_1=x_1) = \mu_2+\rho\sigma_2(x_1-\mu_1)/\sigma_1$ and variance $\sigma_2^2(1-\rho^2)$, one can relate the inner expectation on the far right in (\ref{kre1}) to the moment generating function $M(t)$ of this conditional distribution evaluated at $t=1$, leading to

\begin{equation}
\label{kre2}
E \left(e^{X_2} | X_1 \right) = e^{ \mu_2+\rho\sigma_2(x_1-\mu_1)/\sigma_1 + \sigma_2^2(1-\rho^2)/2},
\end{equation}

so that

\begin{equation}
\label{kre3}
\mathbb E(Y_1 Y_2)  =  e^{ \mu_2+ \frac{1}{2} \sigma_2^2(1-\rho^2)} \mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\}.
\end{equation}

Since $X_1$ is normal with mean $\mu_1$ and variance $\sigma_1^2$, the expectation in (\ref{kre3}) becomes

\begin{equation}
\label{kre4}
\mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = \int_{-\infty}^\infty x_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (x_1-\mu_1)} \frac{1}{\sqrt{2\pi}\sigma_1} e^{-\frac{1}{2\sigma_1^2}(x_1-\mu_1)^2}.
\end{equation}

Upon substituting $u=x_1-\mu_1$, followed by some algebra, we arrive at

\begin{equation}
\label{kre5}
\mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = e^{\frac{1}{2}\sigma_2^2\rho^2} \int_{-\infty}^\infty (u+\mu_1)  \frac{1}{\sqrt{2\pi}\sigma_1} e^{-\frac{1}{2\sigma_1^2}(u-\rho\sigma_1\sigma_2)^2}.
\end{equation}

Since the integral in (\ref{kre5}) is the expectation $\mathbb E(X+\mu_1)$, where $X$ is normal with mean $\rho\sigma_1\sigma_2$ and variance $\sigma_1^2$, we conclude that

\begin{equation}
\label{kre6}
\mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = e^{\frac{1}{2}\sigma_2^2\rho^2} (\rho\sigma_1\sigma_2 +\mu_1),
\end{equation}

which, in view of (\ref{kre3}), leads to

\begin{equation}
\label{kre7}
\mathbb E (Y_1Y_2) = e^{\mu_2 + \frac{1}{2}\sigma_2^2} (\rho\sigma_1\sigma_2 +\mu_1).
\end{equation}

Finally, (\ref{covy1y2}), along with the expressions for the means of $Y_1$ and $Y_2$, produce the covariance of $Y_1$ and $Y_2$:

\begin{equation}
\label{kre8}
\mbox{Cov}(Y_1, Y_2) = e^{\mu_2 + \frac{1}{2}\sigma_2^2} (\rho\sigma_1\sigma_2 +\mu_1) - \mu_1 e^{\mu_2 + \frac{1}{2}\sigma_2^2} = \rho\sigma_1\sigma_2 e^{\mu_2 + \frac{1}{2}\sigma_2^2}.
\end{equation}

This concludes the proof
\end{proof}

Using the above result, we can directly relate the correlation coefficient of $Y_1$ and $Y_2$ with that of $X_1$ and $X_2$,

\begin{equation}
\label{kram8}
\rho_{Y_1, Y_2} = \rho \frac{\sigma_2}{\sqrt{e^{\sigma_2^2} -1}},
\end{equation}

where $\rho$ is the correlation of $X_1$ and $X_2$. The above result is useful when studying the possible range of correlation of $Y_1$ and $Y_2$ in the above example. It can be shown that the factor on the far right in (\ref{kram8}) is a monotonically decreasing function of $\sigma_2$ on $(0,\infty)$, with the limits of 1 and 0 at zero and infinity, respectively. Thus, in principle, the range of correlation of $Y_1$ and $Y_2$ is the same as that of $X_1$ and $X_2$, as the factor on the far right in (\ref{kram8}) can be made arbitrarily close to 1. However, by changing this factor we may affect the marginal distributions of $Y_1$ and $Y_2$. It can be shown that if the marginal distributions of $Y_1$ and $Y_2$ are fixed, then the relation (\ref{kram8}) becomes

\begin{equation}
\label{kram9}
\rho_{Y_1, Y_2} = \rho \sqrt{\frac{\log(1+c_2^2)}{c_2^2}},
\end{equation}

where $c_2=\sigma_{Y_2}/\mu_{Y_2}$ is the {\it coefficient of variation} (CV) of the variable $Y_2$. Thus, the range of possible correlations in this model is not affected by the distribution of $Y_1$, and is determined by the CV of $Y_2$ as follows:

\begin{equation}
\label{kram10}
- \sqrt{\frac{\log(1+c_2^2)}{c_2^2}} \leq \rho_{Y_1, Y_2} \leq \sqrt{\frac{\log(1+c_2^2)}{c_2^2}}. 
\end{equation}

 \vspace{0.1in}

 \noindent {\bf Further notes on the basic example in Section 4.1}. Based on the above analysis, one may propose a model of the form (\ref{kram1}) for these data, where $Y_1$ is the temperature (normal) and $Y_2$ is the Ozone variable (lognormal). Then, $X_1=Y_1$ and $X_2=\log Y_2$ will be bivariate normal, and the parameters can be estimated  based on the bivariate normal model. One can again use the marginal estimates as discussed in the paper, although one could actually use the exact same method for both variables (based on normality, after converting the Ozone variable to normal through the logarithmic transformation). According to the analysis presented in the paper, the values of 3.419 and 0.6967 are the estimates (the MLEs) of the mean and the variance of the underlying normal random variable $X_2$, that is $\hat{\mu}_2 = 0.3.419$ and $\hat{\sigma}_2=0.6967$. When this estimate of $\sigma_2$ is used in (\ref{kram8}), the factor on the far right becomes

\begin{equation}
\label{kram11}
\frac{\hat{\sigma}_2}{\sqrt{e^{\hat{\sigma}_2^2} -1}} = 0.881.
\end{equation}

Thus, the possible range of correlation becomes $(-0.881, 0.881)$. This approximately agrees with the MC results in Section \@ref(package).

### 2021-02-12 notebooks/bigsimr-julia-getting-started.Rmd
test julia version

--- 2020-12-25 notebooks/bigsimr.Rmd
@@
 ---
-title: "Get Started"
+title: "Get Started with `bigsimr` Julia Version"
 output: rmarkdown::html_vignette
@@
 ```{r setup}
+## devtools::install_github("SchisslerGroup/bigsimr", ref="julia")
 library(bigsimr)
+Sys.setenv(JULIA_NUM_THREADS=parallel::detectCores())
+## Either let JuliaCall find your installation of Julia or set manually
+bs <- bigsimr_setup(JULIA_HOME = "/usr/local/bin/")
+dist <- distributions_setup() # Julia distributions
 library(tidyverse)
@@
 ```{r}
-(rho <- cor_fast(df, method = "pearson"))
+p <- bs$cor(as.matrix(df), bs$Pearson)
+p
 ```
@@
 ```{r}
-df %>% select(Temp) %>% summarise_all(.funs = c(mean = mean, sd = sd))
+mu_temp <- mean(df$Temp)
+sd_temp <- sd(df$Temp)
 ```
@@
 ```{r}
-mle_mean <- function(x) mean(log(x))
-mle_sd <- function(x) mean( (log(x) - mean(log(x)))^2 )
-
-df %>% select(Ozone) %>% 
-  summarise_all(.funs = c(meanlog = mle_mean, sdlog = mle_sd))
+mu_ozone <- mean(log(df$Ozone))
+sd_ozone <- sqrt(mean((log(df$Ozone) - mean(log(df$Ozone)))^2))
 ```
@@
 ```{r}
-margins <- alist(
-  qlnorm(meanlog = 3.42, sdlog = 0.743),
-  qnorm(mean = 77.9, sd = 9.49)
+margins <- c(
+  dist$Normal(mu_temp, sd_temp),
+  dist$LogNormal(mu_ozone, sd_ozone)
 )
+margins
+```
+
+```{r}
+D <- bs$MvDistribution(p, margins, bs$Pearson)
+D
 ```
@@
 ```{r}
-cor_bounds(margins = margins, method = "pearson")
+bounds <- bs$cor_bounds(D)
+bounds$lower
+bounds$upper
+```
+
+```{r}
+bounds <- bs$pearson_bounds(D)
+bounds$lower
+bounds$upper
 ```
@@
 ```{r}
-x <- rvec(10000, rho, margins, "pearson")
-
-df_sim <- as.data.frame(x)
-colnames(df_sim) <- colnames(df)
+x <- bs$rvec(10000L, p, margins)
+head(x)
 ```
@@
 Notice that this uncorrelated bivariate distribution has the same marginal distributions as the correlated example above.
+
+# Pearson matching
+
+## Correlation conversion
+
+```{r}
+ps <- bs$cor(as.matrix(df), bs$Spearman)
+ps
+```
+
+```{r}
+bs$cor_bounds(margins[1], margins[2], bs$Spearman)
+```
+
+```{r}
+D2 <- bs$MvDistribution(ps, margins, bs$Spearman)
+x2 <- bs$rand(D2, 1000000L)
+bs$cor(x2, bs$Spearman)
+```
+
+```{r}
+pp <- bs$cor_convert(ps, bs$Spearman, bs$Pearson)
+D3 <- bs$MvDistribution(pp, margins, bs$Pearson)
+x3 <- bs$rand(D3, 1000000L)
+bs$cor(x3, bs$Pearson)
+bs$cor(x3, bs$Spearman)
+```
+
+```{r}
+D4 <- bs$pearson_match(D2);
+bs$cor(D4)
+```
+
+```{r}
+x4 <- bs$rand(D4, 1000000L)
+bs$cor(x4, bs$Pearson)
+```
+
+```{r}
+bs$cor(bs$rand(D, 1000000L))
+p
+```
+
+```{r}
+D5 <- bs$pearson_match(D)
+x5 <- bs$rand(D5, 1000000L)
+bs$cor(x5)
+```
+
+```{r}
+m <- as.matrix(brca)
+str(m)
+```
+
+```{r}
+fit_mom <- function(x) {
+  m <- mean(x)
+  s <- sd(x)
+  r <- m^2 / (s^2 - m)
+  p <- m / s^2
+  dist$NegativeBinomial(r, p)
+}
+margins <- apply(m, 2, fit_mom)
+margins <- do.call(c, unname(margins))
+```
+
+```{r}
+tau <- bs$cor(m, bs$Spearman)
+rho <- bs$cor_convert(tau, bs$Spearman, bs$Pearson)
+matrixcalc::is.positive.definite(rho)
+```
+
+```{r}
+bs$rvec(10L, rho, margins)
+```
+
+```{r}
+rho_tilde <- bs$cor_nearPD(rho)
+matrixcalc::is.positive.definite(rho_tilde)
+```
+
+```{r}
+x <- bs$rvec(10L, rho_tilde, margins)
+str(x)
+```

### 2021-02-23 010-background.Rmd
Polishing for CSDA submission, addressing review criteria

@@
              booktabs = TRUE,
-      caption = "mRNA expression for three selected high-expressing genes (*RPL5*, *TKNIP*, and *VIM*) for the first five patients in the TCGA BRCA data set."
+             row.names = F,
+      caption = paste0( "mRNA expression for three selected high-expressing genes, ", paste( names(smallForTable)[-1], collapse = ", " ), ", for the first five patients in the TCGA BRCA data set.")
 )

### 2021-02-23 030-bigsimr-package.Rmd
Polishing for CSDA submission, addressing review criteria

@@
 
-The `bigsimr` package is a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms)).
+The following descrption is directed towards towards `R` users via `bigsimr` R package.
+This package provides an interface to the native code written in Julia, registered as the `Bigsimr`Julia package PROVIDE LINK.
+In addition to the native Julia `Bigsimr` package and `R` interface `bigsimr`, we also provide a python interface `PyBigsimr` at URL that interfaces with the Julia `Bigsimr` package.
+The Julia implementation provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms)).
 When designing `bigsimr`, we aimed to conveniently provide parallelized computation, through multi-core and GPU acceleration, while allowing advanced users to customize and automate workflows.
@@
 ```{r ch030-basic-example-setup, echo=TRUE, message=FALSE, cache=F}
+## devtools::install_github("SchisslerGroup/bigsimr", ref="dev-julia")
 library(bigsimr)
+Sys.setenv(JULIA_NUM_THREADS=parallel::detectCores())
+## Either let JuliaCall find your installation of Julia or set manually
+bs <- bigsimr_setup(JULIA_HOME = "/usr/local/bin/")
+dist <- distributions_setup() # Julia distributions
 library(tidyverse)
@@
 type <- 'spearman'
-(R_S <- cor_fast(df, method = "spearman"))
+(R_S <- bs$cor(as.matrix(df), bs$Spearman))
 ```
@@
 ```{r ch030-cor-bounds, echo=TRUE}
-cor_bounds(margins = margins, method = type)
+bounds <- bs$cor_bounds(D)
+bounds$lower
 ```

### 2021-02-23 040-monte-carlo-evaluation.Rmd
Polishing for CSDA submission, addressing review criteria

@@
 
-```{r ch040-LoadLib040, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
-## devtools::install_github("SchisslerGroup/bigsimr", ref = 'develop')
-## knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
-# reticulate::use_condaenv("bigsimr")
+```{r ch040-LoadLib040, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, results = "hide"}
+## devtools::install_github("SchisslerGroup/bigsimr", ref="dev-julia")
 library(bigsimr)
+Sys.setenv(JULIA_NUM_THREADS=parallel::detectCores())
+## Either let JuliaCall find your installation of Julia or set manually
+bs <- bigsimr_setup(JULIA_HOME = "/usr/local/bin/")
+dist <- distributions_setup() # Julia distributions
 library(tidyverse)
@@
 ```
+
+*Limitations, conclusions, and recommendations.*
+In the bivariate studies, we chose arbitrary simulation parameters for three distributions, moving from the Gaussian to the discrete and non-normal, multivariate negative binomial.
+Under these conditions, the simulated random vectors sample the desired bivariate distribution across the entire range of pairwise correlations for the three dependency measures.
+The simulation results could differ for other choices of simulation settings.
+Specifying extreme correlations near the boundary or Frechet bounds could result in poor simulation performance.
+Fortunately, it is straightforward to evaluate simulation performance by using strategies similar to those completed above.
+We expect our random vector generation to perform well for the vast majority of NORTA-feasible correlation matrices, but advise to check the performance before making inferences/further analyses.

### 2021-02-23 050-examples.Rmd
2nd draft. Insert Tom / Ania bivariate example
Polishing for CSDA submission, addressing review criteria

@@
 
-```{r ch050-cor-check, echo=TRUE, eval=TRUE, cache = FALSE}
+```{r ch050-cor-check, echo=TRUE, eval=TRUE, cache = TRUE}
 ## 1. Mapping step first
@@
 
+*Limitations, conclusions, and recommendations.*
+The results show good simulation performance for our choice of parameters settings.
+Our settings were motivated by modeling high-expressing genes from TCGA BRCA data set.
+In general, the ability to match marginal and dependence parameters depends on the particular joint probability model.
+We recommend to evaluate and tune your simulation until you can be assured of the accuracy.
+
 ## Simulation-based joint probability calculations

### 2021-02-23 index.Rmd
Polishing for CSDA submission, addressing review criteria

@@
 ```{r indexSetup, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide", cache=F}
-library(bigsimr)
+## devtools::install_github("SchisslerGroup/bigsimr", ref="dev-julia")
+## library(bigsimr)
+## Sys.setenv(JULIA_NUM_THREADS=parallel::detectCores())
+## Either let JuliaCall find your installation of Julia or set manually
+## bs <- bigsimr_setup(JULIA_HOME = "/usr/local/bin/")
+## dist <- distributions_setup() # Julia distributions
 library(tidyverse)
 library(knitr)
+library(printr)
 knitr::opts_chunk$set(eval = TRUE, echo = FALSE, cache = TRUE)

### 2021-02-24 010-background.Rmd
begin work converting to julia

@@
     brca0 <- TCGA2STAT::getTCGA(disease = "BRCA",
-                            data.type = "RNASeq",
-                            p = 6L)
+                                data.type = "RNASeq")
     brca1 <- as.data.frame(t(brca0$dat))
@@
     gene_median_order <- order(gene_median, decreasing = TRUE)
-    brca  <- brca1[,head(gene_median_order, d)]
-    writeLines( substring( row.names( brca ), first=1, last= 12 ), './data/examplePatients.txt' ) 
+    brca  <- brca1[, head(gene_median_order, d)]
+    writeLines(substring(row.names(brca), first = 1, last = 12), './data/examplePatients.txt') 
     saveRDS(brca, brcaRDS)

### 2021-02-24 030-bigsimr-package.Rmd
begin work converting to julia

@@
+```{r ch030-basic-example-setup, echo=TRUE, message=FALSE, cache=F}
+## devtools::install_github("SchisslerGroup/bigsimr", ref="dev-julia")
+library(bigsimr)
+Sys.setenv(JULIA_NUM_THREADS=parallel::detectCores())
+## Either let JuliaCall find your installation of Julia or set manually
+# bs <- bigsimr_setup(JULIA_HOME = "/usr/local/bin/")
+bs <- bigsimr_setup(JULIA_HOME = "/home/alex/julia-1.5.3/bin/", pkg_check = FALSE)
+dist <- distributions_setup() # Julia distributions
+library(tidyverse)
+library(patchwork)
+```
+
 # The `bigsimr` R package {#package}
@@
 First, we load the `bigsimr` library and a few other convenient data science packages, including the syntactically-elegant `tidyverse` suite of `R` packages.
-
-```{r ch030-basic-example-setup, echo=TRUE, message=FALSE, cache=F}
-## devtools::install_github("SchisslerGroup/bigsimr", ref="dev-julia")
-library(bigsimr)
-Sys.setenv(JULIA_NUM_THREADS=parallel::detectCores())
-## Either let JuliaCall find your installation of Julia or set manually
-bs <- bigsimr_setup(JULIA_HOME = "/usr/local/bin/")
-dist <- distributions_setup() # Julia distributions
-library(tidyverse)
-library(patchwork)
-```
 
@@
 
-Next, we configure the input marginals for later input into `bigsimr::rvec`.
-The marginal distributions are specified using `R`'s special `alist` function.
-This allows one to enter the distributions without evaluating anything (yet).
+Next, we configure the input marginals for later input into `rvec`.
+The marginal distributions are specified using `Julias`'s `Distributions.jl` package and stored in an `R` vector.
 
 ```{r ch030-margins-alist, echo=TRUE, cache=T}
-margins <- alist(
-    qnorm(mean = 77.871, sd = 9.4855),
-    qlnorm(meanlog = 3.419, sdlog = 0.6967)
-)
-```
-
-Notice that we use the *quantile* function for the marginals, as that is how the marginal distributions $F_i$ are specified the `bigsimr::rvec` algorithm.
-This implementation strategy supports all `R` `base` probability distributions.
-It allows flexible extensions using other `R` packages that adhere to conventions, such as `extraDistr`.
-Further, by using `alist`, users can specify their own custom distributions (see below in [creating custom margins](#custom-margins)).
-
-To circumvent manual entry, `mlist` presents an alternative to `alist` with on-the-fly evaluation.
-
-```{r ch030-margins-mlist, echo=TRUE}
-mlist(
-  qnorm(mean = mean(df$Temp), sd = sd(df$Temp)),
-  qlnorm(meanlog = mle_mean(df$Ozone), sdlog = mle_sd(df$Ozone))
+margins <- c(
+    dist$Normal(mean(df$Temp), sd(df$Temp)),
+    dist$LogNormal(mle_mean(df$Ozone), mle_sd(df$Ozone))
 )
@@
 As mentioned, the user must decide how to describe correlation, based on the particulars of the problem.
-For non-normal data and for improved simulation accuracy/scalability in our scheme, we advocate the use of Spearman correlation matrix $R_S$ and Kendall $\tau$ matrix $R_K$.
+For non-normal data and for improved simulation accuracy/scalability in our scheme, we advocate the use of Spearman's $\rho$ correlation matrix $R_S$ and Kendall's $\tau$ correlation matrix $R_K$.
 We also support approximate Pearson correlation coefficient matching, while cautioning the user to check the performance for their parametric multivariate model (see [Monte Carlo evaluations](#simulations) below for evaluation strategies and guidance).
-To aid in correlation specification, and estimation in general, we provide a high-performance function `bigsimr::cor_fast` which estimates Spearman, Kendall, or Pearson correlation using the fastest methods available.
+To aid in correlation specification, and estimation in general, we provide a high-performance function `cor_fast` which estimates Spearman, Kendall, or Pearson correlation of large data sets by computing pairwise correlations in parallel.
 (Anyone who has tried estimating Kendall's $\tau$ using `stats::cor` can attest that the routine does not scale to even moderate dimensions).
@@
 ```{r ch030-aq-cor, echo=TRUE}
-type <- 'spearman'
 (R_S <- bs$cor(as.matrix(df), bs$Spearman))
@@
 *Checking target correlation matrix admissibility*.
-First we use `bigsimr::cor_bounds` to ensure that the pairwise target correlation values are valid prior to mapping step which constructs $R_X$ for the MVN input into NORTA (see Section \@ref(algorithms)).
+First we use `cor_bounds` to ensure that the pairwise target correlation values are valid prior to mapping step which constructs $R_X$ for the MVN input into NORTA (see Section \@ref(algorithms)).
 `cor_bounds` estimates the pairwise lower and upper correlation bounds using the Generate, Sort, and Correlate algorithm of @DH2011.
@@
 ```{r ch030-cor-bounds, echo=TRUE}
-bounds <- bs$cor_bounds(D)
+bounds <- bs$cor_bounds(margins)
 bounds$lower
@@
 
-Since our our single estimated Spearman correlation is within the theoretical bounds, the correlation is valid as input to `bigsimr:rvec`.
-But even if the 2-dimensional bounds are satified for each pair of margins, this does not guarantee the feasibility of a $d-$variate distribution [@BF17].
+Since our single estimated Spearman correlation is within the theoretical bounds, the correlation is valid as input to `rvec`.
+But even if the 2-dimensional bounds are satisfied for each pair of margins, this does not guarantee the feasibility of a $d-$variate distribution [@BF17].
 
-To provide higher dimensional feasibility/admissibility checking, we begin by mapping using `bigsimr::cor_convert` and check admissibility.
+To provide higher dimensional feasibility/admissibility checking, we begin by mapping using `cor_convert` and check admissibility.
 
@@
 ## Step 1. Mapping
-(R_X <- cor_convert(rho = R_S, from = type, to = 'pearson'))
+(R_X <- bs$cor_convert(R_S, bs$Spearman, bs$Pearson))
 ## Step 2. Check admissibily
@@
 }
-is_valid_correlation( R_X )
+is_valid_correlation(R_X)
 ```
@@
 ```{r ch030-cor-bounds-pearson, echo=TRUE}
-cor_bounds(margins = margins, method = 'pearson', reps = 1e6)
+bs$cor_bounds(margins[1], margins[2], bs$Pearson, n_samples = 1e6)
 ```
@@
 ```{r ch030-sim-margins, echo=TRUE}
-x <- rvec(10000, R_X, margins)
+x <- bs$rvec(10000, R_X, margins)
 df_sim <- as.data.frame(x)
@@
 
-*Creating a custom marginal distribution*.
-Because `bigsimr` uses an `alist` to store the margins, any probability distribution with a well-defined inverse CDF can be employed, including custom marginal distributions not provided in base `R`. 
-Therefore, to specify a marginal distribution absent from an existing package, the user must provide a closed-form expression to form the corresponding quantile function.
-
-Users should prefix distributions with `r` and `q` for *random* and *quantile* respectively.
-Only the quantile function is necessary to simulate random vectors, but to compute the theoretical correlation bounds, one must to supply a univariate random number generator as well.
-
-For an example, consider a custom *Pareto* distribution for use with `bigsimr`.
-The Pareto CDF is
-
-$$
-F(x) = 1 - \left(\frac{x_m}{x}\right)^\alpha
-$$
-
-for scale $x_m > 0$ and shape $\alpha > 0$, with support $x \in \left[x_m, \infty\right)$.
-From the CDF, we compute the inverse CDF
-
-$$
-F^{-1}(p) = \frac{x_m}{\left(1 - p\right)^{1/\alpha}}
-$$
-
-Next we define the Pareto quantile function in R:
-
-```{r ch030-define-qpareto, echo=TRUE}
-qpareto <- function(p, scale, shape) {
-  scale / (1 - p)^(1/shape)
-}
-```
-
-Writing random number generating function for our target marginal can be accomplished by calling the quantile function on a uniformly distributed random variable (the inverse transform method @Rizzo2007).
-
-```{r ch030-define-rpareto, echo=TRUE}
-rpareto <- function(n, scale, shape) {
-  qpareto(runif(n), scale, shape)
-}
-```
-
-Now with the `qpareto` and `rpareto` functions, we can use the distribution in `bigsimr` similar to the other built-in distributions.
-
-```{r ch030-cust-dist-example, echo=TRUE, eval=FALSE}
-margins <- alist(
-  qnorm(mean = 3.14, sd = 0.1),
-  qbeta(shape1 = 1, shape2 = 4),
-  qnbinom(size = 10, prob = 0.75),
-  qpareto(scale = 1.11, shape = 5.55)
-)
-cor_bounds(margins, "pearson")
-rho <- cor_randPD(4)
-x <- rvec(10, rho, margins)
-```
-
 *Using `bigsimr` on a computing cluster via `rslurm`*.
@@
 ## conda env config vars list
-sjob <- slurm_call(rvec, jobname = 'rvec',
-                   list(n = 1e6,
-                        rho = rho,
-                        margins = margins,
-                        type = "spearman"),
+sjob <- slurm_call(bs$rvec, jobname = 'rvec',
+                   list(1e6,
+                        rho,
+                        margins),
                    submit = TRUE)
@@
 library(rslurm)
-sjob <- slurm_call(rvec, jobname = 'rvec',
-                   alist(n = 1e6, 
-                         rho = rho, 
-                         margins = margins, 
-                         type = "spearman"),
+sjob <- slurm_call(bs$rvec, jobname = 'rvec',
+                   list(1e6, 
+                        rho, 
+                        margins),
                    submit = TRUE)
@@
 simReps <- 100
-rhoList <- replicate(n = simReps, bigsimr::cor_randPSD(d = length(margins)),
+rhoList <- replicate(n = simReps, bs::cor_randPSD(length(margins)),
                      simplify=FALSE)
 sjob <- slurm_map(x = rhoList,
-                  f = rvec,
+                  f = bs$rvec,
                   jobname = 'rvecMap',
-                  n = 1e6,
-                  margins = margins,
-                  type = "spearman",
+                  1e6,
+                  margins,
                   nodes = 4,
@@
                   submit = TRUE)
-
 ```

### 2021-02-24 040-monte-carlo-evaluation.Rmd
begin work converting to julia

@@
-# Monte Carlo evaluations {#simulations}
-
 ```{r ch040-LoadLib040, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, results = "hide"}
@@
 ## Either let JuliaCall find your installation of Julia or set manually
-bs <- bigsimr_setup(JULIA_HOME = "/usr/local/bin/")
+# bs <- bigsimr_setup(JULIA_HOME = "/usr/local/bin/")
+bs <- bigsimr_setup(JULIA_HOME = "/home/alex/julia-1.5.3/bin/", pkg_check = FALSE)
 dist <- distributions_setup() # Julia distributions
@@
 ```
+
+# Monte Carlo evaluations {#simulations}
 
@@
 sigma <- 1
-if ( !file.exists( 'results/biNorm_sims.rds' ) )  {
+if (!file.exists( 'results/biNorm_sims.rds'))  {
     mom_norm <- function(x) {
@@
     sigma <- 1
-    margins <- alist(
-        qnorm(mean = mu, sd = sigma),
-        qnorm(mean = mu, sd = sigma)
+    margins <- c(
+        dist$Normal(mu, sigma),
+        dist$Normal(mu, sigma)
     )
-    type <- c("pearson", "spearman", "kendall")
+    types <- c("Pearson", "Spearman", "Kendall")
     cores <- as.integer( parallel::detectCores() - 1 )
@@
     ## n <- c(1e1, 1e2, 1e3)
-    adjustForDiscrete <- c(FALSE)
     eps <- 1e-2
@@
     ## grid_steps <- 10
-    sim_pars <- expand.grid(type = type, cores = cores, n = n,
-                            stringsAsFactors = FALSE,
-                            adjustForDiscrete = adjustForDiscrete)
+    sim_pars <- expand.grid(type = types, cores = cores, n = n, stringsAsFactors = FALSE)
     res <- data.frame()
@@
         n <- sim_pars$n[i]
-        adjustForDiscrete <- sim_pars$adjustForDiscrete[i]
-        tmp_bounds <- cor_bounds(margins, method = type)
+        tmp_bounds <- bs$cor_bounds(margins, bs[[type]])
         cor_lo <- tmp_bounds$lower[1,2] + eps
@@
             diag(Rho) <- 1.0
-            ## Rho <- convertCor( rho = Rho, from = type, to = 'pearson' )
+            Rho <- bs$cor_convert(Rho, bs[[type]], bs$Pearson)
             time_data <- system.time({
@@
                 ## use rvec to test our software
-                x <- rvec(n = n,
-                          rho = Rho,
-                          margins = margins,
-                          type = type,
-                          cores = cores)
+                x <- bs$rvec(n, Rho, margins)
             })
@@
                 "-Cor", type,
-                "-adj", as.character(adjustForDiscrete),
                 "-dev", "1CORE",
@@
             ## Estimate statistics
-            Rho_hat <- cor_fast(x, method = type)
+            Rho_hat <- bs$cor_fast(x, bs[[type]])
             rho_hat <- Rho_hat[1, 2]
@@
                                   margins = "norm",
-                                  adjustForDiscrete = adjustForDiscrete,
                                   d = 2,
@@
 rate <- 1
-if ( !file.exists( 'results/biGamma_sims.rds' ) )  {
+if (!file.exists( 'results/biGamma_sims.rds'))  {
     mom_gamma <- function(x) {
@@
     }
-    margins <- alist(
-        qgamma(shape = shape, rate = rate),
-        qgamma(shape = shape, rate = rate)
+    margins <- c(
+        dist$Gamma(shape, rate),
+        dist$Gamma(shape, rate)
     )
 
-    type <- c("pearson", "spearman", "kendall")
+    types <- c("Pearson", "Spearman", "Kendall")
     ## cores <- c(1L)
@@
     ## n <- c(1e1, 1e2, 1e3)
-    adjustForDiscrete <- c(FALSE)
 
@@
     ## grid_steps <- 10
-    sim_pars <- expand.grid(type = type, cores = cores, n = n,
-                            stringsAsFactors = FALSE,
-                            adjustForDiscrete = adjustForDiscrete)
+    sim_pars <- expand.grid(type = types, cores = cores, n = n,
+                            stringsAsFactors = FALSE)
 
@@
         n <- sim_pars$n[i]
-        adjustForDiscrete <- sim_pars$adjustForDiscrete[i]
 
-        tmp_bounds <- cor_bounds(margins, method = type)
+        tmp_bounds <- bs$cor_bounds(margins, bs[[type]])
         cor_lo <- tmp_bounds$lower[1,2] + eps
@@
             diag(Rho) <- 1.0
+            Rho <- bs$cor_convert(Rho, bs[[type]], bs$Pearson)
 
             time_data <- system.time({
-                x <- rvec(n = n,
-                          rho = Rho,
-                          margins = margins,
-                          cores = cores,
-                          type = type,
-                          ensure_PSD = TRUE)
+                x <- bs$rvec(n, Rho, margins)
             })
@@
                 "-Cor", type,
-                "-adj", as.character(adjustForDiscrete),
                 "-dev", "CPU",
@@
                                   margins = "gamma",
-                                  adjustForDiscrete = adjustForDiscrete,
                                   d = 2,
@@
 prob <- 3e-04
-if ( !file.exists( 'results/biNB_sims.rds' ) )  {
+if (!file.exists( 'results/biNB_sims.rds'))  {
     mom_nbinom <- function(x) {
@@
 
-    margins <- alist(
-        qnbinom(size = size, prob = prob),
-        qnbinom(size = size, prob = prob)
+    margins <- c(
+        dist$NegativeBinomial(size, prob),
+        dist$NegativeBinomial(size, prob)
     )
 
-    type <- c("pearson", "spearman", "kendall")
+    types <- c("Pearson", "Spearman", "Kendall")
     cores <- as.integer( parallel::detectCores() - 1 )
@@
         n <- sim_pars$n[i]
-        adjustForDiscrete <- sim_pars$adjustForDiscrete[i]
-        ## AGS. 17 Oct 2020. cor_bounds failed.
-        ## I'll create a minimal working example and open an issue
-        ## work around for now
-        if ( type == 'pearson' ) {
-            tmp_bounds <- cor_bounds(margins, method = type)
+
+        tmp_bounds <- bs$cor_bounds(margins, bs[[type]])
         cor_lo <- tmp_bounds$lower[1,2] + eps
         cor_hi <- tmp_bounds$upper[1,2] - eps
-        } else {
-            cor_lo <- -1 + eps
-            cor_hi <- 1 - eps
-        }
         cor_seq <- seq(cor_lo, cor_hi, length.out = grid_steps)
@@
                                   margins = "nbinom",
-                                  adjustForDiscrete = adjustForDiscrete,
                                   d = 2,

### 2021-02-24 050-examples.Rmd
begin work converting to julia

@@
 library(bigsimr)
+# bs <- bigsimr_setup(JULIA_HOME = "/usr/local/bin/")
+bs <- bigsimr_setup()
+dist <- distributions_setup() # Julia distributions
 library(tidyverse)

### 2021-02-24 index.Rmd
added DESCRIPTION file to manage dependencies
see the README for instructions on installing deps

--- 2021-01-11 index.Rmd
@@
 
-```{r indexSetup, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide", cache=F}
-library(bigsimr)
+```{r indexSetup, include=FALSE, cache=FALSE}
+knitr::opts_chunk$set(eval = TRUE, 
+                      echo = FALSE,
+                      message = FALSE,
+                      warning = FALSE,
+                      cache = TRUE)
+
 library(tidyverse)
 library(knitr)
-knitr::opts_chunk$set(eval = TRUE, echo = FALSE, cache = TRUE)
+library(printr)
 library(rmarkdown)
 library(bookdown)
-## setwd('~/article_bigsimr')
+
 set.seed(09212020)
@@
 
-```{r copyBib, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, results = "hide"}
-system( 'cp ~/bib/bigsimr.bib ./.' )
+```{r copyBib, include=FALSE, eval=FALSE}
+system('cp ~/bib/bigsimr.bib ./.')
 ```
@@
 
-```{r countWords, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, results = "hide"}
+```{r countWords, include=FALSE, eval=FALSE}
 rmdFiles <- dir(pattern = 'Rmd$', ignore.case= T)
@@
 The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc).
+
 <!-- Here we propose a scheme that ignores the analytic goal --- and, instead, aims to match the empirically-derived estimated correlation data and marginal characteristics.  -->
+
 For normal data, the Pearson product-moment correlation describes the dependency perfectly.

### 2021-02-24 010-background.Rmd
updated all bigsimr code
moved generated data into separate scripts

--- 2021-01-13 010-background.Rmd
@@
 
-```{r ch010-background, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide", cache=F}
+```{r ch010-background, include=FALSE, cache=FALSE}
+devtools::load_all()
 library(tidyverse)
-library(GGally)
-knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
 ```
 
-The `bigsimr` `R` package presented here provides multiple algorithms that operate with high-dimensional multivariate data; however, all these algorithms were originally designed to support a single task:
-to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics.
+The `bigsimr` `R` package presented here provides multiple algorithms that operate with high-dimensional multivariate data; however, all these algorithms were originally designed to support a single task: to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics.
 Specifically, our goal is to efficiently simulate a large number, $B$, of random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with *correlated* components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i$, where $d$ can be very large and still be computed in practically useful timescale.
@@
 
-* SP1: Procedure must scale to high dimensions with practical computable times.
+* SP1: Procedure must scale to high dimensions with practical compute times.
 * SP2: Procedure must scale to high dimensions while maintaining accuracy.
@@
 The next section describes the motivating data set we use to provide example applications.
-your comment -->
+-->
 
@@
 
-```{r ch010-myD, include=FALSE, cache = TRUE, eval=T}
-d <- 1000
-```
-
-```{r ch010-processBRCA, echo=FALSE, eval=TRUE, results='none', cache=F}
-brcaRDS <- paste0( "data/brca", d, ".rds" )
-if ( !file.exists( brcaRDS ) | !file.exists( './data/examplePatients.txt' ) )  {
-    brca0 <- TCGA2STAT::getTCGA(disease = "BRCA",
-                            data.type = "RNASeq",
-                            p = 6L)
-    brca1 <- as.data.frame(t(brca0$dat))
-    gene_median <- apply(brca1, 2, median)
-    gene_median_order <- order(gene_median, decreasing = TRUE)
-    brca  <- brca1[,head(gene_median_order, d)]
-    writeLines( substring( row.names( brca ), first=1, last= 12 ), './data/examplePatients.txt' ) 
-    saveRDS(brca, brcaRDS)
-}
-brca <- readRDS(file = brcaRDS)
-```
-
 We illustrate our methodology using a well-studied Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments).
@@
 
-```{r ch010-realDataTab, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
-## Describe and plot real data
-if ( !file.exists( './data/exampleGenes.rds' ) )  {
+```{r ch010-realDataTab}
 set.seed(01122021)
-    exampleGenes <- colnames(brca)[sort(sample(ncol(brca), 3))]
-    saveRDS(exampleGenes, file = './data/exampleGenes.rds')
-}
-exampleGenes <- readRDS('./data/exampleGenes.rds')
-## Add row names per WWP comment
-small <-  brca %>% select( exampleGenes ) %>% as.data.frame
-numPatients <- 5
-smallForTable <- head( small, n = numPatients )
-smallForTable$PatientID <- head( readLines('./data/examplePatients.txt'), n = numPatients )
-smallForTable <- smallForTable[ , c(4, 1:3 ) ]
-smallForTable[ , 2:4 ] <- round( smallForTable[ , 2:4 ] )
-knitr::kable(smallForTable ,
-      booktabs = TRUE,
-      caption = "mRNA expression for three selected high-expressing genes (*RPL5*, *TKNIP*, and *VIM*) for the first five patients in the TCGA BRCA data set."
-)
+num_genes    <- 3
+num_patients <- 5
+gene_sample  <- sample(example_genes, num_genes)
+
+cap <- paste0("mRNA expression for three selected high-expressing genes, ",
+              paste(gene_sample, collapse = ", "), 
+              ", for the first five patients in the TCGA BRCA data set.")
+
+small_brca <- brca %>%
+  select(all_of(gene_sample)) %>%
+  head(n = num_patients) %>%
+  as_tibble(rownames = "Patient ID") %>%
+  mutate(`Patient ID` = str_sub(`Patient ID`, end = 12))
+
+small_brca %>%
+  knitr::kable(booktabs = TRUE, caption = cap)
 ```
@@
 
-```{r ch010-realDataFig, cache=F, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes from Table 1.
+```{r ch010-realDataFig, cache=FALSE, fig.width=8, fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes from Table 1.
 The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations.
 Modeling these data motivate our simulation methodology."}
-# lowerfun <- function(data,mapping){
-#   ggplot(data = data, mapping = mapping)+
-#     geom_point()+
-#     scale_x_continuous(limits = c(-1.5,1.5))+
-#     scale_y_continuous(limits = c(-1.5,1.5))
-# }
-
-GGally::ggpairs( data = small,
-                # lower = list( continuous = wrap(lowerfun)),
-                upper = list(
-                    continuous = wrap('cor', method = "spearman")
-                ), 
+GGally::ggpairs(
+  data = small_brca[,-1],
+  upper = list(continuous = GGally::wrap('cor', method = "spearman"))
 ) + theme_bw()
@@
 
-Two nonparametric, rank-based measures common in practice are Spearman's correlation (denoted $\rho_S$) and Kendall's $\tau$. 
-Define
+Two nonparametric, rank-based measures common in practice are Spearman's correlation (denoted $\rho_S$) and Kendall's $\tau$. Define
 
@@
 
-Both $\tau$ and $\rho_S$ are *invariant to under monotone transformations* of the underlying random variates.
+Both $\tau$ and $\rho_S$ are *invariant under monotone transformations* of the underlying random variates.
 As we will describe more fully in the [Algorithms](#algorithms) section, this property is essential to scalable match rank-based correlations with speed (SP1) and accuracy (SP2).
@@
 
-Also, for a discrete random variable $Y_i$, some care must be taken to define the quantile function $F_{i}^{-1}$.
-Let
+Also, for a discrete random variable $Y_i$, some care must be taken to define the quantile function $F_{i}^{-1}$. Let
 
@@
 \noindent where $U$ is a uniform random variable in $(0,1)$, and $F^{-1}_1, F^{-1}_2$ are the inverse CDFs of random variables $X_1$ and $X_2$, respectively.
+
 For discrete random variables, define $F^{-1}$ as in Equation \@ref(eq:inverseCDF).
@@
 
-For any random vector ${\bf X}=(X_1, \ldots, X_d)$ with CDF $F$ and marginal CDFs $F_i$ there is a copula function 
-$C(u_1, \ldots, u_d)$ satisfying
+For any random vector ${\bf X}=(X_1, \ldots, X_d)$ with CDF $F$ and marginal CDFs $F_i$ there is a copula function $C(u_1, \ldots, u_d)$ satisfying
 
@@
 where $\Phi(\cdot)$ is the standard normal CDF.
-Note that the copula $C_{{\bf R}}$ is the familar multivariate normal CDF of the random vector $(\Phi(X_1), \ldots, \Phi(X_d))$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})$.
+Note that the copula $C_{{\bf R}}$ is the familiar multivariate normal CDF of the random vector $(\Phi(X_1), \ldots, \Phi(X_d))$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})$.
 
 Sklar's Theorem [@Sklar1959; @Ubeda-Flores2017] guarantees that given inverse CDFs $F_i^{-1}$s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions.
-For example, using Gaussian copulas, we can construct a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)$ with $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$, via ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$ povides $Y_i \sim F_i, \, \forall \, i$ .
+For example, using Gaussian copulas, we can construct a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)$ with $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$, via ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$ provides $Y_i \sim F_i, \, \forall \, i$.

### 2021-02-24 030-bigsimr-package.Rmd
updated all bigsimr code
moved generated data into separate scripts

@@
-```{r ch030-basic-example-setup, echo=TRUE, message=FALSE, cache=F}
-## devtools::install_github("SchisslerGroup/bigsimr", ref="dev-julia")
-library(bigsimr)
-Sys.setenv(JULIA_NUM_THREADS=parallel::detectCores())
-## Either let JuliaCall find your installation of Julia or set manually
-# bs <- bigsimr_setup(JULIA_HOME = "/usr/local/bin/")
-bs <- bigsimr_setup(JULIA_HOME = "/home/alex/julia-1.5.3/bin/", pkg_check = FALSE)
-dist <- distributions_setup() # Julia distributions
+# The `bigsimr` R package {#package}
+
+```{r ch030-basic-example-setup, echo=TRUE, cache=FALSE}
 library(tidyverse)
 library(patchwork)
+library(bigsimr)
+
+Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
+bs <- bigsimr_setup()
+dist <- distributions_setup()
 ```
 
-# The `bigsimr` R package {#package}
-
-The following descrption is directed towards towards `R` users via `bigsimr` R package.
-This package provides an interface to the native code written in Julia, registered as the `Bigsimr`Julia package PROVIDE LINK.
-In addition to the native Julia `Bigsimr` package and `R` interface `bigsimr`, we also provide a python interface `PyBigsimr` at URL that interfaces with the Julia `Bigsimr` package.
+The following description is directed towards towards `R` users via the `bigsimr` `R` package.
+This package provides an interface to the native code written in Julia, registered as the `Bigsimr` Julia package PROVIDE URL In addition to the native Julia `Bigsimr` package and `R` interface `bigsimr`, we also provide a python interface `bigsimr` at PROVIDE URL that interfaces with the Julia `Bigsimr` package.
 The Julia implementation provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms)).
@@
 
-```{r ch030-aq-joint-dist, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Bivariate scatterplot of Ozone vs.
+```{r ch030-aq-joint-dist, cache=FALSE, fig.width= 8, fig.cap="Bivariate scatterplot of Ozone vs.
 Temp with estimated marginal densities.
-We model the Ozone data as marginally log-normal and the Temperature data as normal.", cache=F}
-
+We model the Ozone data as marginally log-normal and the Temperature data as normal."}
 p0 <- ggplot(df, aes(Temp, Ozone)) +
@@
 
-```{r ch030-aq-temp-pars, echo=TRUE, eval=TRUE}
+```{r ch030-aq-temp-pars, echo=TRUE}
 df %>% 
@@
 
-Next, we configure the input marginals for later input into `rvec`.
-The marginal distributions are specified using `Julias`'s `Distributions.jl` package and stored in an `R` vector.
+Next, we configure the input marginals for later input into `rvec`. The marginal distributions are specified using `Julia`'s `Distributions` package and stored in a vector.
 
-```{r ch030-margins-alist, echo=TRUE, cache=T}
+```{r ch030-margins-alist, echo=TRUE}
 margins <- c(
@@
 ```{r ch030-cor-convert, echo=TRUE}
-## Step 1. Mapping
+# Step 1. Mapping
 (R_X <- bs$cor_convert(R_S, bs$Spearman, bs$Pearson))
-## Step 2. Check admissibily
+
+# Step 2. Check admissibility
 is_valid_correlation <- function(rho) {
@@
 *Simulating random vectors*.
-Finally, we arrive at the main function of `bigsimr`, `rvec`.
-We now simulate $B=10,000$ random vectors from the assumed joint distribution of Ozone levels and Temp.
+Finally, we arrive at the main function of `bigsimr`, `rvec`. We now simulate $B=10,000$ random vectors from the assumed joint distribution of Ozone levels and Temp.
 
@@
 Notice that this uncorrelated bivariate distribution has the same marginal distributions as the correlated example above.
-
-your comment -->
+-->
 
@@
 This example assumes that `bigsimr` is installed on a system with a slurm scheduler.
-A single run of `bigsimr:rvec` using `rslurm` can be coded as:
+A single run of `rvec` using `rslurm` can be coded as:
 
@@
 <!-- 
-Once your jobs complete, `rslurm` conveniently stores simulation results in a standardized format.
-Retriving the data is accomplished by
+Once your jobs complete, `rslurm` conveniently stores simulation results in a standardized format. Retriving the data is accomplished by
 -->
@@
 
-The power of combining `bigsimr` and `rslurm` can be illustrated by simulating many correlation structures.
-The `rslurm::slumr_map` syntax mirrors the `base::lapply` and `purrr::map` functions.
+The power of combining `bigsimr` and `rslurm` can be illustrated by simulating many correlation structures. The `rslurm::slumr_map` syntax mirrors the `base::lapply` and `purrr::map` functions.
 
@@
 Let's evaluate the quality of simulation performance.
-
-your comment -->
+-->
 
@@
 
-```{r ch030-rvecRslurmMapFig, echo = FALSE, eval=FALSE, message=FALSE, warning=FALSE, out.width= '75%', fig.align='center', fig.cap='The continous marginal pairs exact reproduce the specified correlation whereas the discrete pairs (green nad blue) show a downward bias.
+```{r ch030-rvecRslurmMapFig, eval=FALSE, out.width= '75%', fig.cap='The continous marginal pairs exact reproduce the specified correlation whereas the discrete pairs (green nad blue) show a downward bias.
 Future work with employ a discrete adjusted procedures using *rescaling* on the traditional correlation coefficient.'}

### 2021-02-24 040-monte-carlo-evaluation.Rmd
updated all bigsimr code
moved generated data into separate scripts

@@
-```{r ch040-LoadLib040, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, results = "hide"}
-## devtools::install_github("SchisslerGroup/bigsimr", ref="dev-julia")
-library(bigsimr)
-Sys.setenv(JULIA_NUM_THREADS=parallel::detectCores())
-## Either let JuliaCall find your installation of Julia or set manually
-# bs <- bigsimr_setup(JULIA_HOME = "/usr/local/bin/")
-bs <- bigsimr_setup(JULIA_HOME = "/home/alex/julia-1.5.3/bin/", pkg_check = FALSE)
-dist <- distributions_setup() # Julia distributions
-library(tidyverse)
-cores <- as.integer( parallel::detectCores() - 1 )
-## cores <- as.integer( 1 )
-set.seed(10162020)
-```
-
 # Monte Carlo evaluations {#simulations}
 
-```{r ch040-preview, echo = FALSE, eval = FALSE}
-bookdown::preview_chapter('040-monte-carlo-evaluation.Rmd')
+```{r ch040-LoadLib040, include=FALSE}
+library(tidyverse)
+library(bigsimr)
+
+Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
+bs <- bigsimr_setup()
+dist <- distributions_setup()
+
+cores <- as.integer(parallel::detectCores() - 1)
+set.seed(10162020)
 ```
@@
 
-```{r ch040-biNormal, echo = FALSE, eval = TRUE}
-mu <- 0
-sigma <- 1
-if (!file.exists( 'results/biNorm_sims.rds'))  {
-    mom_norm <- function(x) {
-        m <- mean(x)
-        s <- sd(x)
-        list(mean = m, sd = s)
-    }
-    mu <- 0
-    sigma <- 1
-    margins <- c(
-        dist$Normal(mu, sigma),
-        dist$Normal(mu, sigma)
-    )
-    types <- c("Pearson", "Spearman", "Kendall")
-    cores <- as.integer( parallel::detectCores() - 1 )
-    n <- c(1e3, 1e4, 1e5)
-    ## n <- c(1e1, 1e2, 1e3)
-    eps <- 1e-2
-    grid_steps <- 100
-    ## grid_steps <- 10
-    sim_pars <- expand.grid(type = types, cores = cores, n = n, stringsAsFactors = FALSE)
-    res <- data.frame()
-    dir.create("./results/biNorm_sims", showWarnings = TRUE, recursive = TRUE)
-    for (i in 1:nrow(sim_pars)) {
-        type <- sim_pars$type[i]
-        cores <- sim_pars$cores[i]
-        n <- sim_pars$n[i]
-        tmp_bounds <- bs$cor_bounds(margins, bs[[type]])
-        cor_lo <- tmp_bounds$lower[1,2] + eps
-        cor_hi <- tmp_bounds$upper[1,2] - eps
-        cor_seq <- seq(cor_lo, cor_hi, length.out = grid_steps)
-        ## rho = cor_seq[1]
-        for (rho in cor_seq) {
-            Rho <- matrix(rho, 2, 2)
-            diag(Rho) <- 1.0
-            Rho <- bs$cor_convert(Rho, bs[[type]], bs$Pearson)
-            time_data <- system.time({
-                ## x <- rmvn(n = n, mu = rep( mu, 2 ), sigma = Rho)
-                ## use rvec to test our software
-                x <- bs$rvec(n, Rho, margins)
-            })
-            ## Save the sims in case
-            id <- paste0(
-                "d", 2,
-                "-N", n,
-                "-c", cores,
-                "-r", rho,
-                "-Cor", type,
-                "-dev", "1CORE",
-                "-lib", "bigsimr"
-            )
-            ## save sims in their own subdirectory
-            saveRDS(x, file = paste0("./results/biNorm_sims/", id, ".rds"))
-            ## Estimate statistics
-            Rho_hat <- bs$cor_fast(x, bs[[type]])
-            rho_hat <- Rho_hat[1, 2]
-            norm_args_hat <- mom_norm(x[,1])
-            mu_hat <- norm_args_hat$mean
-            sigma_hat <- norm_args_hat$sd
-            ## Save the results
-            res <- rbind(res, data.frame(
-                                  method = "bigsimr",
-                                  device = "CPU",
-                                  type = type,
-                                  cores = cores,
-                                  margins = "norm",
-                                  d = 2,
-                                  N = n,
-                                  rho = rho,
-                                  rho_hat = rho_hat,
-                                  mean = mu,
-                                  sd = sigma,
-                                  mean_hat = mu_hat,
-                                  sd_hat = sigma_hat,
-                                  sim_time = unname(time_data["elapsed"])
-                              ))
-        }
-    }
-    res$type <- factor(res$type, levels = c("pearson",  "spearman", "kendall" ) )
-    saveRDS(object = res, "./results/biNorm_sims.rds")
-}
-```
-
-```{r ch040-biGamma, echo = FALSE, eval = TRUE}
-shape <- 10
-rate <- 1
-if (!file.exists( 'results/biGamma_sims.rds'))  {
-    mom_gamma <- function(x) {
-        m <- mean(x)
-        s <- sd(x)
-        list(shape = m^2 / s^2, rate = m / s^2)
-    }
-    margins <- c(
-        dist$Gamma(shape, rate),
-        dist$Gamma(shape, rate)
-    )
-
-    types <- c("Pearson", "Spearman", "Kendall")
-    ## cores <- c(1L)
-    cores <- as.integer( parallel::detectCores() - 1 )
-    n <- c(1e3, 1e4, 1e5)
-    ## n <- c(1e1, 1e2, 1e3)
-
-    eps <- 1e-2
-    grid_steps <- 100
-    ## grid_steps <- 10
-    sim_pars <- expand.grid(type = types, cores = cores, n = n,
-                            stringsAsFactors = FALSE)
-
-    dir.create("./results/biGamma_sims", showWarnings = TRUE, recursive = TRUE)
-    res <- data.frame()
-    for (i in 1:nrow(sim_pars)) {
-
-        type <- sim_pars$type[i]
-        cores <- sim_pars$cores[i]
-        n <- sim_pars$n[i]
-
-        tmp_bounds <- bs$cor_bounds(margins, bs[[type]])
-        cor_lo <- tmp_bounds$lower[1,2] + eps
-        cor_hi <- tmp_bounds$upper[1,2] - eps
-        cor_seq <- seq(cor_lo, cor_hi, length.out = grid_steps)
-
-        ## rho = cor_seq[1]
-        for (rho in cor_seq) {
-            Rho <- matrix(rho, 2, 2)
-            diag(Rho) <- 1.0
-            Rho <- bs$cor_convert(Rho, bs[[type]], bs$Pearson)
-
-            time_data <- system.time({
-                x <- bs$rvec(n, Rho, margins)
-            })
-
-            ## Save the sims in case
-            id <- paste0(
-                "d", 2,
-                "-N", n,
-                "-c", cores,
-                "-r", rho,
-                "-Cor", type,
-                "-dev", "CPU",
-                "-lib", "bigsimr"
-            )
-            saveRDS(x, file = paste0("./results/biGamma_sims/", id, ".rds"))
-
-            ## Estimate statistics
-            Rho_hat <- bs$cor_fast(x, bs[[type]])
-            rho_hat <- Rho_hat[1, 2]
-            gamma_args_hat <- mom_gamma(x[,1])
-            shape_hat <- gamma_args_hat$shape
-            rate_hat <- gamma_args_hat$rate
-
-            ## Save the results
-            res <- rbind(res, data.frame(
-                                  method = "bigsimr",
-                                  device = "CPU",
-                                  type = type,
-                                  cores = cores,
-                                  margins = "gamma",
-                                  d = 2,
-                                  N = n,
-                                  rho = rho,
-                                  rho_hat = rho_hat,
-                                  shape = shape,
-                                  rate = rate,
-                                  shape_hat = shape_hat,
-                                  rate_hat = rate_hat,
-                                  sim_time = unname(time_data["elapsed"])
-                              ))
-        }
-    }
-    res$type <- factor(res$type, levels = c("pearson",  "spearman", "kendall" ) )
-    saveRDS(object = res, "./results/biGamma_sims.rds")
-}
-```
-
-```{r ch040-biNegBin, echo = FALSE, eval = TRUE}
-size <- 4
-prob <- 3e-04
-if (!file.exists( 'results/biNB_sims.rds'))  {
-    mom_nbinom <- function(x) {
-        m <- mean(x)
-        s <- sd(x)
-        list(size = m^2 / (s^2 - m), prob = m / s^2)
-    }
-
-    margins <- c(
-        dist$NegativeBinomial(size, prob),
-        dist$NegativeBinomial(size, prob)
-    )
-
-    types <- c("Pearson", "Spearman", "Kendall")
-    cores <- as.integer( parallel::detectCores() - 1 )
-    ## cores <- cores
-    n <- c(1e3, 1e4, 1e5)
-    ## n <- c(1e1, 1e2, 1e3)
-
-    eps <- 1e-2
-    grid_steps <- 100
-    ## grid_steps <- 10
-    sim_pars <- expand.grid(type = types, cores = cores, n = n,
-                            stringsAsFactors = FALSE)
-
-    dir.create("./results/biNB_sims", showWarnings = TRUE, recursive = TRUE)
-    res <- data.frame()
-    for (i in 1:nrow(sim_pars)) {
-        type <- sim_pars$type[i]
-        cores <- sim_pars$cores[i]
-        n <- sim_pars$n[i]
-
-        tmp_bounds <- bs$cor_bounds(margins, bs[[type]])
-        cor_lo <- tmp_bounds$lower[1,2] + eps
-        cor_hi <- tmp_bounds$upper[1,2] - eps
-        cor_seq <- seq(cor_lo, cor_hi, length.out = grid_steps)
-
-        for (rho in cor_seq) {
-            Rho <- matrix(rho, 2, 2)
-            diag(Rho) <- 1.0
-            Rho <- bs$cor_convert(Rho, bs[[type]], bs$Pearson)
-
-            time_data <- system.time({
-                x <- bs$rvec(n, Rho, margins)
-            })
-
-            ## Save the sims in case
-            id <- paste0(
-                "d", 2,
-                "-N", n,
-                "-c", cores,
-                "-r", rho,
-                "-Cor", type,
-                "-dev", "CPU",
-                "-lib", "bigsimr"
-            )
-            saveRDS(x, file = paste0("./results/biNB_sims/", id, ".rds"))
-
-            ## Estimate statistics
-            Rho_hat <- bs$cor_fast(x, bs[[type]])
-            rho_hat <- Rho_hat[1, 2]
-            nbinom_args_hat <- mom_nbinom(x[,1])
-            size_hat <- nbinom_args_hat$size
-            prob_hat <- nbinom_args_hat$prob
-
-            ## Save the results
-            res <- rbind(res, data.frame(
-                                  method = "bigsimr",
-                                  device = "CPU",
-                                  type = type,
-                                  cores = cores,
-                                  margins = "nbinom",
-                                  d = 2,
-                                  N = n,
-                                  rho = rho,
-                                  rho_hat = rho_hat,
-                                  size = size,
-                                  prob = prob,
-                                  size_hat = size_hat,
-                                  prob_hat = prob_hat,
-                                  sim_time = unname(time_data["elapsed"])
-                              ))
-        }
-    }
-    res$type <- factor(res$type, levels = c("pearson",  "spearman", "kendall" ) )
-    saveRDS(object = res, "./results/biNB_sims.rds")
-}
-```
-
 Table \@ref(tab:sims) lists our identical-marginal, bivariate simulation configurations.
@@
 Finally, we select distributions and marginal parameter values that are motivated by our RNA-seq data, namely values proximal to probabilities and sizes estimated from the data (see [Example applications](examples) for estimation details).
-Thus we arrive at a multivariate negative binomial (MVNB) $p_1 = p_2 = `r prob`, r_1 = r_2 = 4,\rho \in \Omega$.
+Thus we arrive at a multivariate negative binomial (MVNB) $p_1 = p_2 = 3e-04, r_1 = r_2 = 4, \rho \in \Omega$.
 
@@
 Figure \@ref(fig:ch040-biNegBinPlot) `bigsimr` recovers the Pearson specified correlations for bivariate negative binomial.
-
  -->
 
-```{r ch040-biNormPlot, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, fig.height = 5, fig.width= 8, fig.align='center', fig.cap = "`bigsimr` recovers the Pearson specified correlations for MVN."}
+```{r ch040-biNormPlot, eval=FALSE, fig.height=5, fig.width=8, fig.cap="`bigsimr` recovers the Pearson specified correlations for MVN."}
 dat <- readRDS("./results/biNorm_sims.rds")
@@
 
-```{r ch040-biGammaPlot, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, fig.height = 5, fig.width= 8, fig.align='center', fig.cap = "`bigsimr` recovers the Pearson specified correlations for Bivariate Gamma."}
+```{r ch040-biGammaPlot, eval=FALSE, fig.height=5, fig.width=8, fig.cap="`bigsimr` recovers the Pearson specified correlations for Bivariate Gamma."}
 dat <- readRDS("./results/biGamma_sims.rds")
@@
 
-```{r ch040-biNegBinPlot, echo = FALSE, eval = FALSE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.align='center', fig.width = 8, fig.height=5,fig.cap = "`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
+```{r ch040-biNegBinPlot, eval=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.width=8, fig.height=5, fig.cap="`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
 dat <- readRDS("./results/biNB_sims.rds")
@@
 
-```{r ch040-combineBiSims, cache=F}
+```{r ch040-combineBiSims, cache=FALSE}
 ## compare relative differences in MSE for each dependency, compared to Pearson
@@
 
-```{r ch040-BiError, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
+```{r ch040-BiError}
 ## tabMSE  <- allDat %>%
@@
 
-```{r ch040-bPlot, cache=F, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.asp = 1.30, fig.weight = 4, fig.align='center', fig.cap = "Bivariate simulations match specified correlations.
+```{r ch040-bPlot, cache=FALSE, fig.asp=1.30, fig.weight=4, fig.cap = "Bivariate simulations match specified correlations.
 The horizontal axis plots the specified target correlations across the entire range of possible correlations for each bivariate margin.
@@
 
-The accuracy appears adequate for many applications.
-In practice, we recommend evaluating the accuracy in each application setting, using methods similar to those presented above.
+The accuracy appears adequate for many applications. In practice, we recommend evaluating the accuracy in each application setting, using methods similar to those presented above.
 
@@
 See [Discussion](discussion) for future directions for fast Pearson matching and discrete-specific modifications.
-your comment -->
+-->
 

### 2021-02-24 060-conclusion-discussion.Rmd
updated all bigsimr code
moved generated data into separate scripts

@@
 A widely-available routine for this task in `R` is `matrix::nearPD`, though it is not suitable for high dimensions.
-To overcome this issue, we introduce `bigsimr::nearPSD` --- an implementation of @QS2006's method that computes the nearest correlation matrix to a given symmetric matrix in the Frobenius norm.
+To overcome this issue, we introduce `cor_nearPD` --- an implementation of @QS2006's method that computes the nearest correlation matrix to a given symmetric matrix in the Frobenius norm.
 This routine as many applications beyond our primary goal of random vector generation.
@@
 As data-analytic problems grow to even larger dimension, multi-GPU support is a promising hardware-based future direction.
-your comment -->
+-->

### 2021-02-24 070-supplementary-materials.Rmd
updated all bigsimr code
moved generated data into separate scripts

@@
 
-```{r, out.width='5%', fig.align='center'}
+```{r, out.width='5%'}
 knitr::include_graphics('images/hex-bigsimr.png')
@@
 <!-- 
-
 Nomenclature/Notation
-
-```{r ch070-all-dists, eval=FALSE}
-all_dists <- list(
-  list(dist = "beta", shape1, shape2),
-  list(dist = "binom", size, prob),
-  list(dist = "cauchy", location, scale),
-  list(dist = "chisq", df),
-  list(dist = "exp", rate),
-  list(dist = "f", df1, df2),
-  list(dist = "gamma", shape, rate),
-  list(dist = "geom", prob),
-  list(dist = "hyper", m, n, k),
-  list(dist = "logis", location, scale),
-  list(dist = "lnorm", meanlog, sdlog),
-  list(dist = "nbinom", size, prob),
-  list(dist = "norm", mean, sd),
-  list(dist = "pois", lambda),
-  list(dist = "t", df),
-  list(dist = "unif", min, max),
-  list(dist = "weibull", shape, scale),
-  list(dist = "wilcox", m, n),
-  list(dist = "signrank", n)
-)
-```
-
-your comment -->
+-->

### 2021-02-24 data-raw/bivariate_nbinom_sims.R

@@ -1,0 +1,91 @@
+## code to prepare `bivariate_nbinom_sims` dataset goes here
+
+library(bigsimr)
+Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
+bs <- bigsimr_setup()
+dist <- distributions_setup()
+
+mom_nbinom <- function(x) {
+  m <- mean(x)
+  s <- sd(x)
+  list(size = m^2 / (s^2 - m), prob = m / s^2)
+}
+
+size <- 4
+prob <- 3e-04
+
+margins <- c(dist$NegativeBinomial(size, prob),
+             dist$NegativeBinomial(size, prob))
+
+types <- c("Pearson", "Spearman", "Kendall")
+n <- c(1e3, 1e4, 1e5)
+eps <- 1e-2
+grid_steps <- 100
+
+sim_pars <- expand.grid(
+  type = types,
+  n = n,
+  stringsAsFactors = FALSE
+)
+
+res <- data.frame()
+
+for (i in 1:nrow(sim_pars)) {
+  type <- sim_pars$type[i]
+  n <- sim_pars$n[i]
+
+  tmp_bounds <- bs$cor_bounds(margins, bs[[type]])
+
+  cor_lo <- tmp_bounds$lower[1, 2] + eps
+  cor_hi <- tmp_bounds$upper[1, 2] - eps
+  cor_seq <- seq(cor_lo, cor_hi, length.out = grid_steps)
+
+  for (rho in cor_seq) {
+    Rho <- matrix(rho, 2, 2)
+    diag(Rho) <- 1.0
+    Rho <- bs$cor_convert(Rho, bs[[type]], bs$Pearson)
+
+    time_data <- system.time(x <- bs$rvec(n, Rho, margins))
+
+    id <- paste0(
+      "d", 2,
+      "-N", n,
+      "-c", Sys.getenv("JULIA_NUM_THREADS"),
+      "-r", rho,
+      "-Cor", type,
+      "-dev", "cpu",
+      "-lib", "bigsimr"
+    )
+
+    ## Estimate statistics
+    Rho_hat <- bs$cor_fast(x, bs[[type]])
+    rho_hat <- Rho_hat[1, 2]
+
+    nbinom_args_hat <- mom_nbinom(x[,1])
+    size_hat <- nbinom_args_hat$size
+    prob_hat <- nbinom_args_hat$prob
+
+    ## Save the results
+    res <- rbind(res, data.frame(
+      method = "bigsimr",
+      device = "CPU",
[…]

### 2021-02-24 data-raw/brca.R

@@ -1,0 +1,16 @@
+## code to prepare `brca` dataset goes here
+
+options(timeout = 240) # my internet is sloooowww apparently
+
+brca0 <- TCGA2STAT::getTCGA(disease = "BRCA", data.type = "RNASeq")
+brca1 <- as.data.frame(t(brca0$dat))
+gene_median <- apply(brca1, 2, median)
+gene_median_order <- order(gene_median, decreasing = TRUE)
+
+example_brca     <- brca1[, head(gene_median_order, 10000)]
+example_patients <- substring(row.names(example_brca), first = 1, last = 12)
+example_genes    <- colnames(example_brca)
+
+usethis::use_data(example_brca, overwrite = TRUE)
+usethis::use_data(example_patients, overwrite = TRUE)
+usethis::use_data(example_genes, overwrite = TRUE)

### 2021-02-24 040-monte-carlo-evaluation.Rmd
document renders

@@
 ```{r ch040-LoadLib040, include=FALSE}
+devtools::load_all()
 library(tidyverse)
@@
 set.seed(10162020)
+prob <- 3e-04
 ```
@@
 Finally, we select distributions and marginal parameter values that are motivated by our RNA-seq data, namely values proximal to probabilities and sizes estimated from the data (see [Example applications](examples) for estimation details).
-Thus we arrive at a multivariate negative binomial (MVNB) $p_1 = p_2 = 3e-04, r_1 = r_2 = 4, \rho \in \Omega$.
+Thus we arrive at a multivariate negative binomial (MVNB) $p_1 = p_2 = 3\times10^{-4}, r_1 = r_2 = 4, \rho \in \Omega$.
 
@@
 |-------------|:--------------:|----------------------:|
-|$1000$ | Pearson ($\rho_P$) | ${ \bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho_i ), i=1,\ldots,100$ |
-|$10,000$ | Spearman ($\rho_S$) | ${ \bf Y} \sim MVG( shape = 10, rate = 1, \rho_i ), i=1,\ldots,100$ |
-| $100,000$| Kendall ($\tau$)| ${ \bf Y} \sim MVNB(p = `r prob`, r = 4,\rho_i), i=1,\ldots,100$ |
+| $1000$    | Pearson ($\rho_P$) | ${ \bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho_i ), i=1,\ldots,100$ |
+| $10,000$  | Spearman ($\rho_S$) | ${ \bf Y} \sim MVG( shape = 10, rate = 1, \rho_i ), i=1,\ldots,100$ |
+| $100,000$ | Kendall ($\tau$)| ${ \bf Y} \sim MVNB(p = 3\times10^{-4}, r = 4,\rho_i), i=1,\ldots,100$ |
 
@@
 ```{r ch040-biNormPlot, eval=FALSE, fig.height=5, fig.width=8, fig.cap="`bigsimr` recovers the Pearson specified correlations for MVN."}
-dat <- readRDS("./results/biNorm_sims.rds")
-dat %>%
-    filter(cores == cores) %>%
+bivariate_normal_sims %>%
     ggplot(aes(rho, rho_hat, color = type)) +
-    ## ggplot(aes(rho, rho_hat)) +
     geom_point() +
     geom_abline(slope = 1) +
-    ## facet_wrap(~ + N) + theme_bw()
-    facet_wrap(~ type + N) + theme_bw()
-## ggsave('fig/plot-biNormPlot.pdf')
+    facet_wrap(~ type + N) + 
+    theme_bw()
 ```
@@
 ```{r ch040-biGammaPlot, eval=FALSE, fig.height=5, fig.width=8, fig.cap="`bigsimr` recovers the Pearson specified correlations for Bivariate Gamma."}
-dat <- readRDS("./results/biGamma_sims.rds")
-dat %>%
-    filter(cores == cores) %>%
+bivariate_gamma_sims %>%
     ggplot(aes(rho, rho_hat, color = type)) +
-    ## ggplot(aes(rho, rho_hat)) +
     geom_point() +
     geom_abline(slope = 1) +
-    ## facet_wrap(~ + N) + theme_bw()
-    facet_wrap(~ type + N) + theme_bw()
+    facet_wrap(~ type + N) + 
+    theme_bw()
 ## ggsave('fig/plot-biGammaPlot.pdf')
@@
 
-```{r ch040-biNegBinPlot, eval=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.width=8, fig.height=5, fig.cap="`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
-dat <- readRDS("./results/biNB_sims.rds")
-dat %>%
-    filter(cores == cores) %>%
+```{r ch040-biNegBinPlot, eval=FALSE, fig.width=8, fig.height=5, fig.cap="`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
+bivariate_nbinom_sims %>%
     ggplot(aes(rho, rho_hat, color = type)) +
-    ## ggplot(aes(rho, rho_hat)) +
     geom_point() +
     geom_abline(slope = 1) +
-    ## facet_wrap(~ + N) + theme_bw()
-    facet_wrap(~ type + N) + theme_bw()
-## ggsave('fig/plot-biNegBinPlot.pdf')
+    facet_wrap(~ type + N) + 
+    theme_bw()
 ```
@@
 ```{r ch040-combineBiSims, cache=FALSE}
-## compare relative differences in MSE for each dependency, compared to Pearson
-allDat <- NULL
-dat <- readRDS("./results/biNorm_sims.rds") %>% select( margins, type, N, rho, rho_hat  ) 
-allDat <- rbind( allDat, dat )
-dat <- readRDS("./results/biGamma_sims.rds") %>% select( margins, type, N, rho, rho_hat  )
-allDat <- rbind( allDat, dat )
-dat <- readRDS("./results/biNB_sims.rds") %>% select( margins, type, N, rho, rho_hat  )
-allDat <- rbind( allDat, dat )
-allDat$margins <- factor(allDat$margins, levels = c( 'norm', 'gamma', 'nbinom'),
-                         labels = c( 'MVN', 'MVG', 'MVNB' ) )
-allDat$type <- factor(allDat$type, levels = c( 'pearson', 'spearman', 'kendall' ),
-                      labels = c( 'Pearson', 'Spearman', 'Kendall' ) )
-allDat$N <- factor(allDat$N, levels = c( 1000, 10000, 100000) )
+allDat <- bind_rows(
+    select(bivariate_normal_sims, margins, type, N, rho, rho_hat),
+    select(bivariate_gamma_sims, margins, type, N, rho, rho_hat),
+    select(bivariate_nbinom_sims, margins, type, N, rho, rho_hat)
+) %>%
+    mutate(margins = factor(margins, levels=c("norm", "gamma", "nbinom")),
+           type = factor(type, levels=c("Pearson", "Spearman", "Kendall")),
+           N = factor(N))
 ```
@@
 ```{r ch040-BiError}
-## tabMSE  <- allDat %>%
-##    group_by( N, margins ) %>%
-##    summarize( mse = mean( ( rho - rho_hat )^2 ) )
 tabMAE  <- allDat %>%
-    ## filter( N == 100000) %>%
-    group_by( N, type, margins ) %>%
-##     summarize( mse = mean( ( rho - rho_hat )^2 ) ) %>%
-    summarize( MAE = mean( abs( rho - rho_hat )  ) )
-## normMSE <- as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ] ) 
-## ( tabMSE$mse - normMSE ) / normMSE
-## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) )
-## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) ) / tabMSE$mse
+    group_by(N, type, margins) %>%
+    summarize(MAE = mean(abs(rho - rho_hat))) %>%
+    ungroup()
 
@@
 ## But how to improve latex table formatting? Better to use the template
-knitr::kable(tabMAE,
-             ## format = 'latex',
+tabMAE %>%
+    knitr::kable(booktabs = TRUE,
                  col.names = c("No. of random vectors",
@@
                                "Mean abs. error"),
-             ## booktabs = TRUE,
-             caption = 'Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.',
-             ## linesep = c('', '',  '\\addlinespace')
-)
+                 caption = 'Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.')
 ```
@@
 
-```{r ch040-bPlot, cache=FALSE, fig.asp=1.30, fig.weight=4, fig.cap = "Bivariate simulations match specified correlations.
+```{r ch040-bPlot, cache=FALSE, fig.asp=1.30, fig.weight=4, fig.cap="Bivariate simulations match specified correlations.
 The horizontal axis plots the specified target correlations across the entire range of possible correlations for each bivariate margin.
@@
 names(repsLabs) <- c(1000, 10000, 100000)
-typeLabs <- c( 'Pearson', 'Spearman', 'Kendall' )
-names(typeLabs) <- c( 'Pearson', 'Spearman', 'Kendall' )
+typeLabs <- c("Pearson", "Spearman", "Kendall")
+names(typeLabs) <- c("Pearson", "Spearman", "Kendall")
 
@@
 numGroups <- length(levels(allDat$margins))
-## myColors <- rev( RColorBrewer::brewer.pal(n = numColors, name = 'Greys')[ ((numColors - numGroups) + 1): numColors  ] )
-myColors <- rev( RColorBrewer::brewer.pal(n = numColors, name = 'Blues')[ ((numColors - numGroups) + 1): numColors  ] )
+myColors <- rev(RColorBrewer::brewer.pal(n = numColors, name = "Blues")[ ((numColors - numGroups) + 1):numColors])
 
 allDat %>%
-    ## ggplot(aes(x = rho, y = rho_hat, color = margins, shape = margins)) +
     ggplot(aes(x = rho, y = rho_hat, color = margins)) +
-    ## ggplot(aes(rho, rho_hat)) +
-    ## geom_point(alpha = 1, size = 2) +
     geom_point(size = 2) +
-    ## scale_shape_manual(values=c(1, 2, 3))+
-    ## scale_color_manual(values=c('#999999','#E69F00', '#56B4E9')) +
     scale_color_manual(values = myColors ) +
-    geom_abline(slope = 1, linetype = 'dashed') +
+    geom_abline(slope = 1, linetype = "dashed") +
     labs(x = "Specified Correlation", y = "Estimated Correlation") +
-    ## facet_wrap(~ + N) + theme_bw()
     facet_wrap(~ type + N, labeller = labeller(N = repsLabs, type = typeLabs)) +
@@
     theme(legend.position = "bottom", legend.direction = "horizontal")
-
-## allDat %>%
-##     ggplot(aes(x = rho, y = rho_hat, color = type, shape = type)) +
-##     ## ggplot(aes(rho, rho_hat)) +
-##     geom_point(alpha = 0.75) +
-##     geom_abline(slope = 1) +
-##     labs(x = "Specified Correlation", y = "Estimated Correlation") + 
-##     ## facet_wrap(~ + N) + theme_bw()
-##     facet_wrap(~ margins + N) + theme_bw()
-
-## ggsave('fig/plot-biNormPlot.pdf')
 ```

### 2021-02-24 010-background.Rmd

@@
 library(tidyverse)
+d <- 1000
 ```
@@
 For simplicity, we only consider high expressing genes.
-In turn, we begin by filtering to retain the top `r d`% of highest-expressing genes (in terms of median expression) of the 20,502 gene measurements from $N=`r nrow(brca)`$ patients' tumor samples.
+In turn, we begin by filtering to retain the top `r d`% of highest-expressing genes (in terms of median expression) of the 20,502 gene measurements from $N=`r nrow(example_brca)`$ patients' tumor samples.
 This gives a great number of pairwise dependencies among the marginals (specifically, $`r choose(d,2)`$ correlation parameters).
@@
 
-small_brca <- brca %>%
+small_brca <- example_brca %>%
   select(all_of(gene_sample)) %>%

### 2021-02-24 020-simulation-algorithm.Rmd

@@
 
-```{r ch020-preview, echo = FALSE, eval = FALSE}
+```{r ch020-preview, eval=FALSE}
 bookdown::preview_chapter('020-simulation-algorithm.Rmd')
@@
 
-The well-known NORTA algorithm [@Cario1997] simulates a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$.
-Specifically, NORTA algorithm proceedings as follows:
+The well-known NORTA algorithm [@Cario1997] simulates a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$. Specifically, NORTA algorithm proceedings as follows:
 
@@
 
-*NORTA in higher dimensions*. Sklar's theorem provides a useful characterization of multivariate distributions through copulas.
+*NORTA in higher dimensions*. 
+Sklar's theorem provides a useful characterization of multivariate distributions through copulas.
 Yet the choice of copula-based simulation algorithm affects which joint distributions may be simulated.
@@
 This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined the resultant matrix is indefinite.
-These situations do occur, even using exact analytic calculations. Such problematic target correlation matrices are termed *NORTA defective*.
+These situations do occur, even using exact analytic calculations.
+Such problematic target correlation matrices are termed *NORTA defective*.
 
@@
 
-## `bigsimr::rvec` for random vector generation  {#rand-vec-gen}
+## `rvec` for random vector generation {#rand-vec-gen}
 
 - Discuss how the HD NORTA defective problem and the scalability issues lead to our algorithm
-We now describe `bigsimr::rvec`, our algorithm to generate random vectors, which mirrors the classical NORTA algorithm above with some modifications for rank-based dependency matching:
+We now describe `rvec`, our algorithm to generate random vectors, which mirrors the classical NORTA algorithm above with some modifications for rank-based dependency matching:
 
@@
 
-The "mapping step" takes advantage of the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertKendall) or \@ref(eq:convertSpearman), respectively (implemented as `bigsimr::cor_covert`).
+The "mapping step" takes advantage of the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertKendall) or \@ref(eq:convertSpearman), respectively (implemented as `cor_covert`).
 
@@
 \begin{aligned}
-\Omega_P &= \{ R_P : R_P \,  is \, a \,  correlation \,  matrix \,  for \,  X \} \\
-\Omega_K &= \{ R_K : R_K \,  is \,  a \,  Kendall \,  \tau  \, matrix \,  for \,  X \} \\
-\Omega_S &= \{ R_S : R_S \,  is \,  a \,  Spearman \,  correlation \,  matrix \,  for \,  X\} \\
+\Omega_P &= \{ R_P : R_P \textrm{ is a correlation matrix for } X \} \\
+\Omega_K &= \{ R_K : R_K \textrm{ is a correlation matrix for } X \} \\
+\Omega_S &= \{ R_S : R_S \textrm{ is a correlation matrix for } X \} \\
 \end{aligned}
@@
 
-The final step implements a high-performance NORTA algorithm.
-Specifically, step 3i uses either the efficient multi-core multivariate normal simulator `mvnfast`[@Fasiolo2016] or Google's JAX python library for GPU acceleration of the Cholesky factorization and matrix multiplication and steps 3ii, 3iii are parallelized over CPU cores.
+The final step implements a high-performance NORTA algorithm. Specifically, step 3i uses a fast multi-core multivariate normal simulator, and steps 3ii, 3iii are parallelized over CPU cores.
 

### 2021-02-24 050-examples.Rmd

--- 2021-02-23 050-examples.Rmd
@@
 
-```{r ch050-LoadLib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide", cache=F}
-## knitr::opts_chunk$set(echo = FALSE, cache = FALSE)
-# reticulate::use_condaenv("bigsimr")
+```{r ch050-LoadLib, include=FALSE, cache=FALSE}
+devtools::load_all()
+library(tidyverse)
 library(bigsimr)
-library(tidyverse)
+
+Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
+bs <- bigsimr_setup()
+dist <- distributions_setup()
+
 cores <- as.integer( parallel::detectCores() - 1 )
 set.seed(10142020)
-```
-
-```{r ch050-preview, echo = FALSE, eval = FALSE}
-bookdown::preview_chapter('050-examples.Rmd')
 ```
@@
 
-```{r ch050-readBRCA, echo=FALSE, eval=TRUE, cache=F}
+```{r ch050-readBRCA, cache=FALSE}
 d <- 1000
-brcaRDS <- paste0( "data/brca", d, ".rds" )
-brca <- readRDS(file = brcaRDS)
-## smaller set for prototyping
-## d=5
-## brca <- brca[ , 1:d]
+brca1000 <- example_brca %>%
+  select(all_of(1:d)) %>%
+  mutate(across(everything(), as.double)) # cor_fast currently doesn't work with integers
 ```
@@
 
-With this goal in mind, we first estimate the desired correlation matrix using the fast implementation provided by `bigsimr`.:
+With this goal in mind, we first estimate the desired correlation matrix using the fast implementation provided by `bigsimr`:
 
-```{r ch050-estRhoBRCA, echo=TRUE, eval=TRUE, cache=F}
-## Estimate Spearman's correlation on the count data
-type <- 'spearman'
-R_S <- bigsimr::cor_fast(brca, method = type)
+```{r ch050-estRhoBRCA, echo=TRUE, cache=FALSE}
+# Estimate Spearman's correlation on the count data
+type <- 'Spearman'
+R_S <- bs$cor_fast(as.matrix(brca1000), bs[[type]])
 ```
@@
 The marginal distributions are from the same probability family (NB) yet are heterogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.
-The functions below support this estimation for later use in `bigsimr::rvec`.
+The functions below support this estimation for later use in `rvec`.
 
@@
 TYPESTE MoM estimators here.
-your comment -->
+-->
 
 ```{r ch050-nbHelpers, echo=TRUE}
-make_nbinom_alist <- function(sizes, probs) {
-  lapply(1:length(sizes), function(i) {
-    substitute(qnbinom(size = s, prob = p), 
-               list(s = sizes[i], p = probs[i]))
+make_nbinom_margins <- function(sizes, probs) {
+  margins <- lapply(1:length(sizes), function(i) {
+    dist$NegativeBinomial(sizes[i], probs[i])
   })
+  do.call(c, margins)
 }
-## make_nbinom_alist(c(20, 21, 22), c(0.3, 0.4, 0.5))
+
 nbinom_mom <- function(x) {
@@
   r <- m^2 / (s2 - m)
-  c(r, p)
+  c(size = r, prob = p)
 }
@@
 ```{r ch050-estMargins, echo = TRUE, eval = TRUE, cache=F}
-sizes <- apply( unname(as.matrix(brca)), 2, nbinom_mom )[1, ]
-probs <- apply( unname(as.matrix(brca)), 2, nbinom_mom )[2, ]
-## construct margins
-nb_margins <- make_nbinom_alist(sizes, probs)
+nbinom_fit <- apply(brca1000, 2, nbinom_mom)
+sizes <- nbinom_fit["size",]
+probs <- nbinom_fit["prob",]
+
+nb_margins <- make_nbinom_margins(sizes, probs)
 ```
@@
 
-Now we check admissibility of specified correlation matrix.
-We begin with the pairwise, 2D MC estimated bounds.
-And then follow the `r d`D bounds.
+Now we check admissibility of specified correlation matrix. We begin with the pairwise, 2D MC estimated bounds. And then follow the `r d`D bounds.
 
-```{r ch050-cor-check, echo=TRUE, eval=TRUE, cache = TRUE}
-## 1. Mapping step first
-R_X <- cor_convert(rho = R_S, from = type, to = 'pearson')
-## 2a. Check admissibility
+```{r ch050-cor-check, echo=TRUE}
+# 1. Mapping step first
+R_X <- bs$cor_convert(R_S, bs[[type]], bs$Pearson)
+
+# 2a. Check admissibility
 is_valid_correlation <- function(rho) {
@@
 }
-is_valid_correlation( R_X )
+is_valid_correlation(R_X)
 
-## 2b. compute nearest correlation
-if ( ! is_valid_correlation( R_X ) ) {
-    ## system.time( R_Xtilde <- cor_nearPSD( G = R_X )  )
-    R_Xtilde <- cor_nearPSD( G = R_X )
+# 2b. compute nearest correlation
+if (!is_valid_correlation(R_X)) {
+  R_Xtilde <- bs$cor_nearPD(R_X)
   ## Quantify the error
-    targets <- R_X[ lower.tri( R_X ) ]
-    approximates <- R_Xtilde[ lower.tri( R_Xtilde ) ]
+  targets <- R_X[lower.tri(R_X)]
+  approximates <- R_Xtilde[lower.tri(R_Xtilde)]
   ## R_Xtilde[1:5, 1:5]
@@
 }
-## qplot( x = targets, y = approximates ) + geom_abline(slope=1,intercept = 0)
-## qplot( x = targets - approximates ) + geom_abline(slope=1,intercept = 0)
 summary(targets - approximates)
@@
 
-While the exact dxd Spearman correlation matrix is not strictly admissible in our scheme (seen by the non-PSD result above), the approximation is of sufficient quality for most applications with a maximum pairwise of `r max( targets - approximates)` and average error of `r mean( targets - approximates)`.
+While the exact $d \times d$ Spearman correlation matrix is not strictly admissible in our scheme (seen by the non-PSD result above), the approximation is of sufficient quality for most applications with a maximum pairwise of `r max( targets - approximates)` and average error of `r mean( targets - approximates)`.
 
-```{r ch050-runBRCA-echo, echo=TRUE, eval=FALSE, cache = FALSE}
-sim_nbinom <- rvec(10000, R_X, nb_margins, ensure_PSD = FALSE, cores = cores) 
-colnames(sim_nbinom) <- names(brca)
+```{r ch050-runBRCA-echo, echo=TRUE, eval=FALSE, cache=FALSE}
+sim_nbinom <- bs$rvec(10000, R_X, nb_margins) 
+colnames(sim_nbinom) <- colnames(brca1000)
 ```
 
-```{r ch050-runBRCA, echo=FALSE, results=FALSE, cache = FALSE}
-brcaSimRDS <- paste0( "results/brca", d, "sim.rds" )
-if( !file.exists(brcaSimRDS) ) {
-    ## Set the number of random vectors
+```{r ch050-runBRCA, echo=FALSE, results=FALSE, cache=FALSE}
+dir.create("results", showWarnings = FALSE, recursive = TRUE)
+brcaSimRDS <- paste0("results/brca", d, "sim.rds")
+if(!file.exists(brcaSimRDS)) {
+  set.seed(44)
   n <- 10000
-    ## construct margins
-    ## run sims
-    set.seed(44)
-    sim_nbinom <- rvec(10000, R_X, nb_margins, ensure_PSD = FALSE, cores = cores)
-    saveRDS( sim_nbinom, brcaSimRDS )
+  sim_nbinom <- bs$rvec(n, R_X, nb_margins) 
+  colnames(sim_nbinom) <- colnames(brca1000)
+  saveRDS(sim_nbinom, brcaSimRDS)
 }
-sim_nbinom <- readRDS( brcaSimRDS )
+sim_nbinom <- readRDS(brcaSimRDS)
 ```
@@
 
-```{r ch050-makeSimDataFig, echo = FALSE, eval=T, cache=F, message=FALSE, warning=FALSE}
-if( !file.exists("fig/ch050-simDataFig.png") ) {
-    exampleGenes <- readRDS(file = './data/exampleGenes.rds')
+```{r ch050-makeSimDataFig, cache=FALSE}
+if(!file.exists("fig/ch050-simDataFig.png")) {
   p <- GGally::ggpairs(
-                     data = as.data.frame( sim_nbinom[ ,exampleGenes] ), 
-                     upper = list(
-                         continuous = GGally::wrap('cor', method = "spearman")
-                     )) + theme_bw()
-    ## ggsave("fig/ch050-simDataFig.png", plot = p, width = 5, height = 4, dpi = 300)
+    data = as.data.frame(sim_nbinom[,example_genes[1:3]]),
+    upper = list(continuous = GGally::wrap('cor', method = "spearman"))
+  ) + theme_bw()
   ggsave("fig/ch050-simDataFig.png", plot = p, dpi = 300) 
@@
 
-```{r ch050-simDataFig, echo=F, fig.align='center', fig.cap='Simulated data for three selected high-expressing genes generally replicates the estimated data structure.
+```{r ch050-simDataFig, fig.cap='Simulated data for three selected high-expressing genes generally replicates the estimated data structure.
 The data do not exhibit outlying points, but do possess the desired Spearman correlations, central tendencies, and discrete values.'}
@@
 
-```{r ch050-makeFigBRCA, echo=F, eval=T, cache=F}
-if( !file.exists("fig/ch050-figBRCA.png") ) {
+```{r ch050-makeFigBRCA, cache=FALSE}
+if(!file.exists("fig/ch050-figBRCA.png")) {
     ## estimate rho from sims and plot
@@
 
-```{r ch050-figBRCA, echo=F, out.width = '80%', fig.align='center', fig.cap="Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set.
+```{r ch050-figBRCA, out.width='80%', fig.cap="Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set.
 The dashed red lines indicated equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes)."}
@@
 
-```{r ch050-densityEvaluation, include=FALSE, eval=TRUE, cache=F}
+```{r ch050-densityEvaluation, include=FALSE, cache=FALSE}
 d <- ncol(sim_nbinom)
@@
 
-where ${\bf Y^{(b)} }$ is the $b^{th}$ simulated vector from a total of $B$ simulation replicates and $I ( )$ is the indicator function.
+where ${\bf Y^{(b)} }$ is the $b^{th}$ simulated vector from a total of $B$ simulation replicates and $I(\cdot)$ is the indicator function.
 For example, we can estimate from our $B=10,000$ simulated vectors that the probability of all genes expressing (i.e., ${\bf y}_i \geq 1, \forall \; i$) is $`r pHat`$.
 
-```{r ch050-densityEvaluationECHO, echo=TRUE, eval=TRUE, cache=F}
+```{r ch050-densityEvaluationECHO, echo=TRUE, cache=FALSE}
 d <- ncol(sim_nbinom)
@@
 
-```{r ch010-myM, include=FALSE, cache = TRUE}
+```{r ch010-myM, include=FALSE}
 m <- 10
@@
 
-```{r ch050-quadlossECHO, echo=T, eval=F, cache=FALSE}
+```{r ch050-quadlossECHO, echo=TRUE, eval=FALSE, cache=FALSE}
 ## Simulate random vectors equal to the sample size
@@
 simRho <- replicate(n = m, expr = {
-  tmpSim <- rvec(n = N , R_X, nb_margins, cores = cores)
-  bigsimr::cor_fast(x = tmpSim, method = type)
+  tmpSim <- bs$rvec(N , R_X, nb_margins)
+  bs::cor_fast(tmpSim, bs[[type]])
 }, simplify = FALSE)
@@
 
-```{r ch050-lossResults, echo=F, eval=T, cache=FALSE}
-## quadLossRDS <- paste0( "results/brca", d, "_quadLoss.rds" )
-frobeniusRDS <- paste0( "results/brca", d, "_frobenius.rds" )
-if ( !file.exists( frobeniusRDS) )  {
-    N <- nrow(brca)
+```{r ch050-lossResults, cache=FALSE}
+frobeniusRDS <- paste0("results/brca", d, "_frobenius.rds")
+if (!file.exists(frobeniusRDS)) {
+    N <- nrow(brca1000)
     ## create m random vectors and estimate correlation
     simRho <- replicate(n = m, expr = {
-        tmpSim <- rvec(n = N , R_X, nb_margins, cores = cores)
-        bigsimr::cor_fast(x = tmpSim, method = type)
+        tmpSim <- bs$rvec(N , R_X, nb_margins)
+        bs$cor_fast(tmpSim, bs[[type]])
     }, simplify = FALSE)
-    ## find quadratic loss at each rep
-    ## simRho[[1]][1:5,1:5]
-    ## R_S[1:5,1:5]
-    ## rags2ridges::loss(simRho[[1]], R_S, type = 'frobenius')
     frobenius <- sapply(simRho, rags2ridges::loss, T = R_S, type = "frobenius")
@@
 }
-frobenius <- readRDS( frobeniusRDS )
+frobenius <- readRDS(frobeniusRDS)
 ```
@@
 
-```{r ch050-loss-summary, echo=TRUE, eval=T, cache=FALSE}
+```{r ch050-loss-summary, echo=TRUE, eval=TRUE, cache=FALSE}
 summary(frobenius)

### 2021-02-24 index.Rmd

@@
 ```{r indexSetup, include=FALSE, cache=FALSE}
-knitr::opts_chunk$set(eval = TRUE, 
-                      echo = FALSE,
-                      message = FALSE,
-                      warning = FALSE,
-                      cache = TRUE)
-
 library(tidyverse)
@@
 ```{r serveHelper, include=FALSE, eval=FALSE}
-## rmote::start_rmote()
-## library(ggplot2); qplot(mpg, wt, data=mtcars, colour=cyl)
-## bookdown::render_book("index.Rmd", "bookdown::pdf_book", output_dir = "pdf")
-## bookdown::preview_chapter("index.Rmd")
+# rmote::start_rmote()
+# library(ggplot2); qplot(mpg, wt, data=mtcars, colour=cyl)
+# bookdown::render_book("index.Rmd", "bookdown::pdf_book", output_dir = "pdf")
+# bookdown::preview_chapter("index.Rmd")
 rmarkdown::render_site(encoding = 'UTF-8')
@@
 rmdFiles <- dir(pattern = 'Rmd$', ignore.case= T)
-## rmdFiles <- rmdFiles[-grep('^_', rmdFiles)]
+# rmdFiles <- rmdFiles[-grep('^_', rmdFiles)]
 rmdFiles <- rmdFiles[-grep('references', rmdFiles)]
-## rmdFiles <- rmdFiles[-grep('README', rmdFiles)]
-## rmdFiles <- rmdFiles[-grep('article', rmdFiles)] ## after generating a single document
-## wordcountaddin::word_count( rmdFiles[11] )
+# rmdFiles <- rmdFiles[-grep('README', rmdFiles)]
+# rmdFiles <- rmdFiles[-grep('article', rmdFiles)] ## after generating a single document
+# wordcountaddin::word_count( rmdFiles[11] )
 (counts <- sapply( rmdFiles, wordcountaddin::word_count, simplify = T ))
@@
 
-<!-- Here we propose a scheme that ignores the analytic goal --- and, instead, aims to match the empirically-derived estimated correlation data and marginal characteristics.  -->
+<!-- 
+Here we propose a scheme that ignores the analytic goal --- and, instead, aims to match the empirically-derived estimated correlation data and marginal characteristics.
+-->
 

### 2021-02-24 index.Rmd
Revise index

@@
 title: "Simulating High-Dimensional Multivariate Data"
-subtitle: "using the bigsimr R Package"
+subtitle: "using the `Bigsimr` Package"
 author:
@@
 bibliography: ["bigsimr.bib", "packages.bib"]
-abstract:  It is critical to realistically simulate data when conducting Monte Carlo studies and evaluating quantitative methods.
+abstract:  It is critical to realistically simulate data when employing Monte Carlo techinques and evaluating statistical methodology.
 Measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments.
 Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations.
-This article introduces the `bigsimr` R package that provides a flexible, scaleable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions and a broad class of admissible Spearman correlation and Kendall $\tau$ matrices.
-We describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix.
+This article introduces the `bigsimr` R interface to the Julia `Bigsimr` package.
+This high-package provides a flexible, scaleable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions with specified Pearson, Kendall, or Spearman correlation matrices.
+It also contains additional high-performance features, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix.
 Finally, we demonstrate example applications enabled via `bigsimr` by applying these functions to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=878$ patients and dimension $d=1000$.
@@
 
-Massive high-dimensional data sets are now commonplace in many areas of scientific inquiry.
+Massive high-dimensional (HD) data sets are now commonplace in many areas of scientific inquiry.
 As new methods are developed for data, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of analyzing such proposed methodology --- such as false positive rates, statistical power, interval coverage, and robustness --- often with comparison to existing methods.
@@
 
-Another central issue lies in characterizing dependency between components in the high-dimensional random vector.
-The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc).
-
 <!-- 
@@
 -->
-
+Another central issue lies in characterizing dependency between components in the high-dimensional random vector.
+The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc).
 For normal data, the Pearson product-moment correlation describes the dependency perfectly.
@@
 Throughout, we focus on matching these nonparametric dependency measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology.
+We do, however, provide Pearson matching with some caveats.
 
@@
 Then we describe and justify our simulation methodology and related algorithms.
-Next, we detail an illustrative low-dimensional example of basic and advanced use of the `bigsimr` R package.
-Then we proceed with Monte Carlo studies under various distributional assumptions to assess accuracy and computation times for increasingly large $d$.
-After the MC evaluations, we revisit our high-dimensional motivating example and employ our methods in commonplace statistical computing tasks.
+Next, we detail an illustrative low-dimensional example of basic use of the `bigsimr` R package.
+Then we proceed with Monte Carlo studies under various bivariate distributional assumptions to assess accuracy.
+We conclude the Monte Carlo evaluations by summarizing the computation time for increasingly higher dimensional vectors.
+After the MC evaluations, we simulate random vectors motivated by our RNA-seq example, evaluate the accuracy, and provide example statistical computing tasks, namely a MC joint probability calculation and HD-correlation estimation efficiency.
 Finally, we discuss the method's utility, limitations, and future directions.

### 2021-02-24 010-background.Rmd
Revise 3 example genes code and scatterplot mat

@@
 ```{r ch010-realDataTab}
-set.seed(01122021)
+set.seed(02242021)
 num_genes    <- 3
@@
 GGally::ggpairs(
-  data = small_brca[,-1],
+  data = example_brca[, gene_sample],
   upper = list(continuous = GGally::wrap('cor', method = "spearman"))

### 2021-02-24 010-background.Rmd
Revise 010 prior to CSDA

@@
 For simplicity, we only consider high expressing genes.
-In turn, we begin by filtering to retain the top `r d`% of highest-expressing genes (in terms of median expression) of the 20,502 gene measurements from $N=`r nrow(example_brca)`$ patients' tumor samples.
+In turn, we begin by filtering to retain the top `r d` of the highest-expressing genes (in terms of median expression) of the oover 20,000 gene measurements from $N=`r nrow(example_brca)`$ patients' tumor samples.
 This gives a great number of pairwise dependencies among the marginals (specifically, $`r choose(d,2)`$ correlation parameters).
@@
 Both $\tau$ and $\rho_S$ are *invariant under monotone transformations* of the underlying random variates.
-As we will describe more fully in the [Algorithms](#algorithms) section, this property is essential to scalable match rank-based correlations with speed (SP1) and accuracy (SP2).
+As we will describe more fully in the [Algorithms](#algorithms) section, this property enables matching rank-based correlations with speed (SP1) and accuracy (SP2).
 
@@
 \end{equation}
-
-These features are critical in our simulation algorithm to broaden the dependency measures supported by `bigsimr`, in a computationally effective manner.
 

### 2021-02-24 020-simulation-algorithm.Rmd
Add new section 3.3

@@
 
-## `rvec` for random vector generation {#rand-vec-gen}
+## Random vector generator {#rand-vec-gen}
 
-- Discuss how the HD NORTA defective problem and the scalability issues lead to our algorithm
-We now describe `rvec`, our algorithm to generate random vectors, which mirrors the classical NORTA algorithm above with some modifications for rank-based dependency matching:
+We now describe our algorithm to generate random vectors, which mirrors the classical NORTA algorithm above with some modifications for rank-based dependency matching:
 
@@
 
+## Algorithm details
+
+*Correlation Conversion*
+
+Given:
+
+* A target Spearman's $\rho$ or Kendall's $\tau$ correlation
+
+Do:
+
+* Determine the Pearson correlation for use in a Gaussian copula such that the resulting random samples using the NORTA algorithm have an estimated Spearman or Kendall correlation that matches the target correlation
+
+From Proposition 22 (Lebrun and Dutfow, 2009), we have the result for Gaussian copula that the copula correlation $r$ is related to Spearman's $\rho$ and Kendall's $\tau$ by the following relations:
+
+$r = 2 \sin \left(\frac{\pi}{6} \rho\right) \\
+r = \sin \left(\frac{\pi}{2} \tau\right)$
+
+* Lebrun, R., & Dutfoy, A. (2009). An innovating analysis of the Nataf transformation from the copula viewpoint. Probabilistic Engineering Mechanics, 24(3), 312-320.
+
+*Pearson Matching*
+
+Given:
+
+* A target Pearson correlation matrix, $R$
+* A list of marginal distributions, $F$
+
+Do:
+
+* Determine the input correlation, $\tilde{P}$, such that the resulting random samples using the NORTA algorithm have an estimated Pearson correlation that matches $P$
+
+This method relies on Equations 19, 39, and 49 (Xiao and Zhou, 2019).
+
+* Xiao, Q., & Zhou, S. (2019). Matching a correlation coefficient by a Gaussian copula. Communications in Statistics-Theory and Methods, 48(7), 1728-1747.
+
+*Nearest Correlation Matrix*
+
+This algorithm is trying to find a correlation matrix $X$ close to $G$ in the convex optimization problem:
+
+$\begin{aligned}
+    \mathrm{min}\quad & \frac{1}{2} \Vert G - X \Vert^2 \\
+    \mathrm{s.t.}\quad & X_{ii} = 1, \quad i = 1, \ldots , n, \\
+    & X \in S_{+}^{n}
+\end{aligned}$
+
+* Qi, H., & Sun, D. (2006). A quadratically convergent Newton method for computing the nearest correlation matrix. SIAM journal on matrix analysis and applications, 28(2), 360-385.
+
 <!-- 

### 2021-02-24 data-raw/benchmark_dependences.R
benchmark correlation types workflow
Three correlation types tested/benched. Times four steps: correlation estimation, mapping step, admissibility step, simulation step.

@@ -1,0 +1,129 @@
+## code to prepare `benchmark_dependences` dataset goes here
+
+library(tidyverse)
+library(bigsimr)
+Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
+bs <- bigsimr_setup(pkg_check = FALSE)
+dist <- distributions_setup()
+
+# "preheat" the bigsimr functions
+bs$cor(rnorm(10), rnorm(10), bs$Pearson)
+bs$cor(rnorm(10), rnorm(10), bs$Spearman)
+bs$cor(rnorm(10), rnorm(10), bs$Kendall)
+
+bs$cor_fast(matrix(rnorm(100), 10, 10), bs$Pearson)
+bs$cor_fast(matrix(rnorm(100), 10, 10), bs$Spearman)
+bs$cor_fast(matrix(rnorm(100), 10, 10), bs$Kendall)
+
+bs$cor_convert(matrix(rnorm(100), 10, 10), bs$Spearman, bs$Pearson)
+bs$cor_convert(matrix(rnorm(100), 10, 10), bs$Kendall, bs$Pearson)
+
+d <- dist$NegativeBinomial(20, 0.2)
+bs$pearson_match(0.5, d, d)
+
+d <- dist$NegativeBinomial(20, 0.002)
+bs$pearson_bounds(d, d)
+bs$pearson_match(0.5, d, d)
+
+d <- dist$Gamma(2, 0.002)
+bs$pearson_bounds(d, d)
+bs$pearson_match(0.5, d, d)
+
+s <- bs$cor_randPSD(100)
+p <- bs$cor_convert(s, bs$Spearman, bs$Pearson)
+r <- bs$cor_nearPD(p)
+JuliaCall::julia_call("Bigsimr.iscorrelation", r)
+
+
+# Prepare the data
+load("data/example_brca.rda")
+brca_mat <- do.call(cbind, lapply(example_brca, as.double))
+
+mom_gamma <- function(x) {
+  m <- mean(x)
+  s <- sd(x)
+  c(shape = m^2 / s^2, rate = m / s^2)
+}
+
+make_margins <- function(par1, par2) {
+  margins <- lapply(1:length(par1), function(i) {
+    dist$Gamma(par1[i], par2[i])
+  })
+  do.call(c, margins)
+}
+
+params <- apply(brca_mat, 2, mom_gamma)
+margins <- make_margins(params[1, 1:3], params[2, 1:3])
[…]

### 2021-02-25 030-bigsimr-package.Rmd
Revise 030, cut rslurm.

@@
 # The `bigsimr` R package {#package}
+
+This section describes a low-dimensional (2D) random vector simulation workflow via the `bigsimr` `R` package.
+`bigsimr` provides an interface to the native code written in Julia, registered as the `Bigsimr` Julia package PROVIDE URL.
+In addition to the native Julia `Bigsimr` package and `R` interface `bigsimr`, we also provide a python interface `bigsimr` (PROVIDE URL) that interfaces with the Julia `Bigsimr` package. 
+The Julia implementation provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms)).
+When designing `bigsimr`, we aimed to conveniently provide parallelized computation, through multi-core and GPU acceleration, while allowing advanced users to customize.
+
+The subsections below describe the basic use of `bigsimr`, by stepping through an example workflow using the data set `airquality` that contains daily air quality measurements in New York, May to September 1973 [@Chambers1983].
+This workflow proceeds from computing environment setup, data wrangling, estimation, simulation configuration, random vector generation, and, finally, result visualization.
+The code chunk below prepares the computing environment:
+
+<!-- 
+`bigsimr` is an R package for simulating high-dimensional multivariate data with arbitrary marginal distributions.
+The efficiency behind generating multivariate samples comes from the ability to utilize a GPU during the generation of multivariate normal samples and the subsequent transformation to uniform samples (see section \@ref(rand-vec-gen)).
+A GPU is not necessary in order to use `bigsimr`, and the speed benefits generally only come when simulating very high dimensional data ($d > 10000$).
+
+The next computational step after obtaining the correlated uniform marginals is the inverse transform into the target marginal distributions.
+Since this consists of $d$ independent transformations, we utilize parallelization to achieve higher throughput.
+As with many parallel algorithms, there is overhead associated with forking the task to utilize multiple cores, however cost is negligible compared to the total run-time.
+-->
 
@@
 
-The following description is directed towards towards `R` users via the `bigsimr` `R` package.
-This package provides an interface to the native code written in Julia, registered as the `Bigsimr` Julia package PROVIDE URL In addition to the native Julia `Bigsimr` package and `R` interface `bigsimr`, we also provide a python interface `bigsimr` at PROVIDE URL that interfaces with the Julia `Bigsimr` package.
-The Julia implementation provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms)).
-When designing `bigsimr`, we aimed to conveniently provide parallelized computation, through multi-core and GPU acceleration, while allowing advanced users to customize and automate workflows.
+## Bivariate example description
 
-The subsections below describe the basic use of `bigsimr`, by stepping through a low-dimensional (2D) simulation workflow for the readily-available data set `airquality` that contains daily air quality measurements in New York, May to September 1973 [@Chambers1983].
-This workflow proceeds from data, to estimation, simulation configuration, random vector generation, and result visualization.
-For this low-dimensional setting, we operate using a single central processing unit (CPU), as the overhead in forking the tasks to multiple cores outweighs the computational gains.
-Then we transition to advanced use where we briefly describe some of the high-performance features and syntax.
-The section concludes with a short description of how to use `bigsimr` on a computer clusters through slurm scheduling via the `rslurm` package [@R-rslurm].
-
-<!-- 
-`bigsimr` is an R package for simulating high-dimensional multivariate data with arbitrary marginal distributions.
-The efficiency behind generating multivariate samples comes from the ability to utilize a GPU during the generation of multivariate normal samples and the subsequent transformation to uniform samples (see section \@ref(rand-vec-gen)).
-A GPU is not necessary in order to use `bigsimr`, and the speed benefits generally only come when simulating very high dimensional data ($d > 10000$).
-
-The next computational step after obtaining the correlated uniform marginals is the inverse transform into the target marginal distributions.
-Since this consists of $d$ independent transformations, we utilize parallelization to achieve higher throughput.
-As with many parallel algorithms, there is overhead associated with forking the task to utilize multiple cores, however cost is negligible compared to the total run-time.
--->
-
-## Bivariate example of basic use
-
-We demonstrate the basic use and syntax of `bigsimr` through an example workflow applied to the New York air quality data set (`airquality`) included in the R `datasets` package.
+We illustrate the use of `bigsimr` motivated by the New York air quality data set (`airquality`) included in the R `datasets` package.
 First, we load the `bigsimr` library and a few other convenient data science packages, including the syntactically-elegant `tidyverse` suite of `R` packages.
@@
 
-*Specifying marginal distributions*.
+## Specifying marginal distributions
+
 Based on the estimated densities in Figure \@ref(fig:ch030-aq-joint-dist), we assume `Temp` is normally distributed and `Ozone` is log-normally distributed, as the latter values are positive and skewed.
@@
 
-*Specifying correlation*.
+## Specifying correlation
+
 As mentioned, the user must decide how to describe correlation, based on the particulars of the problem.
 For non-normal data and for improved simulation accuracy/scalability in our scheme, we advocate the use of Spearman's $\rho$ correlation matrix $R_S$ and Kendall's $\tau$ correlation matrix $R_K$.
-We also support approximate Pearson correlation coefficient matching, while cautioning the user to check the performance for their parametric multivariate model (see [Monte Carlo evaluations](#simulations) below for evaluation strategies and guidance).
+We also support Pearson correlation coefficient matching, while cautioning the user to check the performance for their parametric multivariate model (see [Monte Carlo evaluations](#simulations) below for evaluation strategies and guidance).
 To aid in correlation specification, and estimation in general, we provide a high-performance function `cor_fast` which estimates Spearman, Kendall, or Pearson correlation of large data sets by computing pairwise correlations in parallel.
@@
 
-*Checking target correlation matrix admissibility*.
+## Checking target correlation matrix admissibility
+
 First we use `cor_bounds` to ensure that the pairwise target correlation values are valid prior to mapping step which constructs $R_X$ for the MVN input into NORTA (see Section \@ref(algorithms)).
@@
 
-*Simulating random vectors*.
+## Simulating random vectors
 Finally, we arrive at the main function of `bigsimr`, `rvec`. We now simulate $B=10,000$ random vectors from the assumed joint distribution of Ozone levels and Temp.
@@
 -->
-
-## Advanced use {#advanced-use}
-
-*Using `bigsimr` on a computing cluster via `rslurm`*.
-Although `bigsimr` runs quickly, for large $d$ users may wish to run jobs on a shared computing server.
-The R package `rslurm` enables parallel `rvec` calls.
-This example assumes that `bigsimr` is installed on a system with a slurm scheduler.
-A single run of `rvec` using `rslurm` can be coded as:
-
-```{r ch030-rvecRslurm-with-comments, echo=FALSE, eval=FALSE}
-library(rslurm)
-## a single call of rvec
-## had to edit rslurm/templates/submit_single_sh.txt
-## source ~/.bashrc
-## conda activate bigsimr
-## conda env config vars list
-sjob <- slurm_call(bs$rvec, jobname = 'rvec',
-                   list(1e6,
-                        rho,
-                        margins),
-                   submit = TRUE)
-```
-
-```{r ch030-rvecRslurm, echo=TRUE, eval=FALSE}
-library(rslurm)
-sjob <- slurm_call(bs$rvec, jobname = 'rvec',
-                   list(1e6, 
-                        rho, 
-                        margins),
-                   submit = TRUE)
-```
-
-<!-- 
-Once your jobs complete, `rslurm` conveniently stores simulation results in a standardized format. Retriving the data is accomplished by
--->
-
-```{r ch030-rvecRslurmRes, echo=FALSE, eval=FALSE}
-res <- readRDS(file.path('_rslurm_rvec', list.files('_rslurm_rvec', 'results')))
-head(res, 3)
-```
-
-The power of combining `bigsimr` and `rslurm` can be illustrated by simulating many correlation structures. The `rslurm::slumr_map` syntax mirrors the `base::lapply` and `purrr::map` functions.
-
-```{r ch030-rvecRslurmMap-comments, echo=FALSE, eval=FALSE}
-library(rslurm)
-## a single call of rvec
-## had to edit rslurm/templates/submit_sh.txt
-## source ~/.bashrc
-## conda activate bigsimr
-## conda env config vars list
-set.seed(06202020)
-simReps <- 100
-rhoList <- replicate(n = simReps, bs::cor_randPSD(length(margins)),
-                     simplify=FALSE)
-sjob <- slurm_map(x = rhoList,
-                  f = bs$rvec,
-                  jobname = 'rvecMap',
-                  1e6,
-                  margins,
-                  nodes = 4,
-                  cpus_per_node = 4,
-                  submit = TRUE)
-```
-
-```{r ch030-rvecRslurmMap, echo=TRUE, eval=FALSE}
-library(rslurm)
-simReps <- 100
-rhoList <- replicate(n = simReps, bs::cor_randPSD(length(margins)),
-                     simplify=FALSE)
-sjob <- slurm_map(x = rhoList,
-                  f = bs$rvec,
-                  jobname = 'rvecMap',
-                  1e6,
-                  margins,
-                  nodes = 4,
-                  cpus_per_node = 4,
-                  cores = 4,
-                  submit = TRUE)
-```
-
-On a cluster carrying 24 nodes with 48 threads, these 100 jobs complete in about a minute.
-
-<!-- 
-Let's evaluate the quality of simulation performance.
--->
-
-```{r ch030-rvecRslurmMapRes, echo=FALSE, eval=FALSE}
-list.files('_rslurm_rvecMap', 'results')
-res <- rslurm::get_slurm_out(sjob, outtype = 'raw')
-rhoHats <- lapply(res, bigsimr::fastCor, method = 'spearman')
-resData <- NULL
-simreps <- length(rhoHats)
-for (i in 1:simreps){
-    tmpRhoHat <- rhoHats[[i]]
-    ## average together
-    margins
-    avgRhoHat <- tmpRhoHat[lower.tri(tmpRhoHat)]
-    tmpPair <- paste0(
-      "pair", apply(combn(x = length(margins), 2), 2, paste, collapse = "_")
-    )
-    tmpResData <- data.frame(
-      simNum = i,
-      rho    = rhoList[[i]][1,2], 
-      rhoHat = tmpRhoHat[lower.tri(tmpRhoHat)], 
-      pair   = tmpPair
-    )
-    resData <- rbind(resData, tmpResData)
-}
-## "wrangle" into a tibble
-library(tidyverse)
-res <- as_tibble(resData)
-## save for viz
-saveRDS(res, file = 'data/rvecRslurmMapRes.rds')
-## cleanup careful this deletes output also
-## cleanup_files(sjob) 
-```
-
-```{r ch030-rvecRslurmMapFig, eval=FALSE, out.width= '75%', fig.cap='The continous marginal pairs exact reproduce the specified correlation whereas the discrete pairs (green nad blue) show a downward bias.
-Future work with employ a discrete adjusted procedures using *rescaling* on the traditional correlation coefficient.'}
-res <- readRDS( 'data/rvecRslurmMapRes.rds' )
-res
-res %>%
-    ggplot( aes( x = rho, y = rhoHat, color = pair ) )  +
-    geom_point() +
-    geom_abline( slope = 1, intercept = 0 )
-```

### 2021-02-25 bench_figure.R
updated bivariate

@@ -1,0 +1,19 @@
+devtools::load_all()
+library(tidyverse)
+
+dat_long <- benchmark_dependences %>%
+  select(-total_time, -n_sim, -needed_near_pd) %>%
+  pivot_longer(cols = c(corr_time:sim_time), names_to = "Step", values_to = "Time") %>%
+  mutate(dim = factor(dim),
+         Step = factor(Step,
+                       levels = c("corr_time", "adj_time", "admiss_time", "sim_time"),
+                       labels = c("Compute Correlation",
+                                  "Adjust Correlation",
+                                  "Check Admissibility",
+                                  "Simulate Data"))) %>%
+  rename(Correlation = corr_type, Dimensions = dim)
+
+ggplot(dat_long, aes(Dimensions, Time, fill=Step)) +
+  geom_bar(position = "stack", stat = "identity") +
+  labs(y = "Time (Seconds)") +
+  facet_grid(Correlation ~ ., scales = "free")

### 2021-02-25 data-raw/benchmark_dependences.R

@@ -34,6 +34,9 @@ p <- bs$cor_convert(s, bs$Spearman, bs$Pearson)
 r <- bs$cor_nearPD(p)
 JuliaCall::julia_call("Bigsimr.iscorrelation", r)
 
+d <- dist$Gamma(3, 0.01)
+bs$rvec(10, bs$cor_randPD(2), c(d, d))
+
 
 # Prepare the data
 load("data/example_brca.rda")
@@ -53,8 +56,10 @@ make_margins <- function(par1, par2) {
 }
 
 params <- apply(brca_mat, 2, mom_gamma)
-margins <- make_margins(params[1, 1:3], params[2, 1:3])
-bs$pearson_bounds(margins)
+tmp_cor <- bs$cor(brca_mat[,1:100])
+tmp_margins <- make_margins(params[1, 1:100], params[2, 1:100])
+
+bs$pearson_match(tmp_cor, tmp_margins)
 
 # steps
 # 1. estimate pearson correlation: cor(data, cor_type)
@@ -108,17 +113,24 @@ bench <- function(type, d, n=1000) {
   )
 }
 
+
 dims <- c(100, 250, 500, 1000)
 pearson_bench <- lapply(dims, function(d) bench("Pearson", d)) %>%
   do.call(what = bind_rows)
+usethis::use_data(pearson_bench, overwrite = TRUE)
+
 
 dims <- c(100, 250, 500, 1000, 2500, 5000, 10000)
 spearman_bench <- lapply(dims, function(d) bench("Spearman", d)) %>%
   do.call(what = bind_rows)
+usethis::use_data(spearman_bench, overwrite = TRUE)
+
 
 dims <- c(100, 250, 500, 1000, 2500, 5000, 10000)
 kendall_bench <- lapply(dims, function(d) bench("Kendall", d)) %>%
   do.call(what = bind_rows)
+usethis::use_data(kendall_bench, overwrite = TRUE)
+
 
 benchmark_dependences <- bind_rows(pearson_bench, spearman_bench, kendall_bench) %>%
   mutate(corr_type = factor(corr_type, levels = c("Pearson", "Spearman", "Kendall")),

### 2021-02-25 030-bigsimr-package.Rmd
removed ref to cor_fast

@@
 We also support Pearson correlation coefficient matching, while cautioning the user to check the performance for their parametric multivariate model (see [Monte Carlo evaluations](#simulations) below for evaluation strategies and guidance).
-To aid in correlation specification, and estimation in general, we provide a high-performance function `cor_fast` which estimates Spearman, Kendall, or Pearson correlation of large data sets by computing pairwise correlations in parallel.
-(Anyone who has tried estimating Kendall's $\tau$ using `stats::cor` can attest that the routine does not scale to even moderate dimensions).
 These estimation methods are classical approaches, not designed for high-dimensional correlation estimation (see the [Conclusion and Discussion]({#discussion) sections for more on this).

### 2021-02-25 050-examples.Rmd
removed ref to cor_fast

@@
   select(all_of(1:d)) %>%
-  mutate(across(everything(), as.double)) # cor_fast currently doesn't work with integers
+  mutate(across(everything(), as.double))
 ```
@@
 type <- 'Spearman'
-R_S <- bs$cor_fast(as.matrix(brca1000), bs[[type]])
+R_S <- bs$cor(as.matrix(brca1000), bs[[type]])
 ```
@@
     ## estimate rho from sims and plot
-    R_S_hat <- cor_fast(sim_nbinom, method = type)
+    R_S_hat <- bs$cor(as.matrix(sim_nbinom), bs[[type]])
     ## triplot
@@
   tmpSim <- bs$rvec(N , R_X, nb_margins)
-  bs::cor_fast(tmpSim, bs[[type]])
+  bs$cor(tmpSim, bs[[type]])
 }, simplify = FALSE)
@@
         tmpSim <- bs$rvec(N , R_X, nb_margins)
-        bs$cor_fast(tmpSim, bs[[type]])
+        bs$cor(tmpSim, bs[[type]])
     }, simplify = FALSE)

### 2021-02-25 bench_figure.R
benchmark data using synthetic data

@@ -13,7 +13,22 @@ dat_long <- benchmark_dependences %>%
                                   "Simulate Data"))) %>%
   rename(Correlation = corr_type, Dimensions = dim)
 
-ggplot(dat_long, aes(Dimensions, Time, fill=Step)) +
+dat_long %>%
+  filter(Dimensions %in% c(100, 250, 500, 1000)) %>%
+  ggplot(aes(Correlation, Time, fill=Step)) +
   geom_bar(position = "stack", stat = "identity") +
   labs(y = "Time (Seconds)") +
-  facet_grid(Correlation ~ ., scales = "free")
+  facet_grid(. ~ Dimensions, scales = "free") +
+  theme(axis.text.x = element_text(angle = -90))
+
+
+dat_long %>%
+  filter(Dimensions %in% c(2500, 5000, 10000, 20000)) %>%
+  mutate(`Time (minutes)` = Time / 60) %>%
+  ggplot(aes(Correlation, `Time (minutes)`, fill=Step)) +
+  geom_bar(position = "stack", stat = "identity") +
+  scale_y_continuous(breaks = c(0, 5, 10, 15, 20, 30, 45),
+                     minor_breaks = NULL) +
+  facet_grid(. ~ Dimensions) +
+  theme(axis.text.x = element_text(angle = -90))
+

### 2021-02-25 data-raw/benchmark_dependences.R

@@ -6,10 +6,25 @@ Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
 bs <- bigsimr_setup(pkg_check = FALSE)
 dist <- distributions_setup()
 
+
+make_margins <- function(par1, par2) {
+  margins <- lapply(1:length(par1), function(i) {
+    dist$Gamma(par1[i], par2[i])
+  })
+  do.call(c, margins)
+}
+shapes <- runif(20000, 1, 10)
+rates  <- rexp(20000, 1/5)
+
+
 # "preheat" the bigsimr functions
 bs$cor(rnorm(10), rnorm(10), bs$Pearson)
 bs$cor(rnorm(10), rnorm(10), bs$Spearman)
 bs$cor(rnorm(10), rnorm(10), bs$Kendall)
+
+bs$cor(matrix(rnorm(100), 10, 10), bs$Pearson)
+bs$cor(matrix(rnorm(100), 10, 10), bs$Spearman)
+bs$cor(matrix(rnorm(100), 10, 10), bs$Kendall)
 
 bs$cor_fast(matrix(rnorm(100), 10, 10), bs$Pearson)
 bs$cor_fast(matrix(rnorm(100), 10, 10), bs$Spearman)
@@ -37,29 +52,28 @@ JuliaCall::julia_call("Bigsimr.iscorrelation", r)
 d <- dist$Gamma(3, 0.01)
 bs$rvec(10, bs$cor_randPD(2), c(d, d))
 
+tmp_cor <- bs$cor_randPD(100)
+tmp_margins <- make_margins(shapes[1:100], rates[1:100])
+bs$pearson_match(tmp_cor, tmp_margins)
+
+rm(p, r, s, d, tmp_cor, tmp_margins, shapes, rates)
+gc()
+
 
 # Prepare the data
-load("data/example_brca.rda")
-brca_mat <- do.call(cbind, lapply(example_brca, as.double))
+# margins <- make_margins(shapes, rates)
+# system.time(corr <- bs$cor_randPD(20000))
+# data <- bs$rvec(1000, corr, margins)
+#
+# saveRDS(data,   "data/synthetic_data.rds")
+# saveRDS(shapes, "data/synthetic_shapes.rds")
+# saveRDS(rates,  "data/synthetic_rates.rds")
+# rm(corr)
+# gc()
 
-mom_gamma <- function(x) {
-  m <- mean(x)
-  s <- sd(x)
-  c(shape = m^2 / s^2, rate = m / s^2)
-}
-
[…]

### 2021-02-25 010-background.Rmd
small changes up to 020

@@
 The `bigsimr` `R` package presented here provides multiple algorithms that operate with high-dimensional multivariate data; however, all these algorithms were originally designed to support a single task: to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics.
-Specifically, our goal is to efficiently simulate a large number, $B$, of random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with *correlated* components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i$, where $d$ can be very large and still be computed in practically useful timescale.
+Specifically, our goal is to efficiently simulate a large number, $B$, of random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with *correlated* components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i$, where $d$ can be very large and still be computed in a practical timescale.
 
@@
 (\#eq:pearson)
-\rho_P(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ var(X)var(Y)\right]^{1/2}}.
+\rho_P(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ \mathrm{Var}(X)\mathrm{Var}(Y)\right]^{1/2}}.
 \end{equation}
@@
 
-*Correspondence among Pearson, Spearman, $\tau$ correlations*.
+*Correspondence among Pearson, Spearman, Kendall correlations*.
 There is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficient, as the marginal distributions $F_i$ are intrinsic in their calculation.
@@
 (\#eq:convertKendall)
-\rho_{P} = sin \left( \tau \times \frac{\pi}{2} \right), 
+\rho_{P} = \sin \left( \tau \times \frac{\pi}{2} \right), 
 \end{equation}
@@
 (\#eq:convertSpearman)
-\rho_P = 2 \times sin \left( \rho_S \times \frac{\pi}{6} \right).
+\rho_P = 2 \times \sin \left( \rho_S \times \frac{\pi}{6} \right).
 \end{equation}
@@
 \begin{equation}
-F_{i}^{-1} = inf\{y:F_{i}(y) \geq u \}.
+F_{i}^{-1} = \inf\{y:F_{i}(y) \geq u \}.
 (\#eq:inverseCDF)
@@
 
-For any random vector ${\bf X}=(X_1, \ldots, X_d)$ with CDF $F$ and marginal CDFs $F_i$ there is a copula function $C(u_1, \ldots, u_d)$ satisfying
+For any random vector ${\bf X}=(X_1, \ldots, X_d)^\top$ with CDF $F$ and marginal CDFs $F_i$ there is a copula function $C(u_1, \ldots, u_d)$ satisfying
 
@@
 Sklar's Theorem [@Sklar1959; @Ubeda-Flores2017] guarantees that given inverse CDFs $F_i^{-1}$s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions.
-For example, using Gaussian copulas, we can construct a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)$ with $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$, via ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$ provides $Y_i \sim F_i, \, \forall \, i$.
+For example, using Gaussian copulas, we can construct a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)^\top$ with $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$, via ${\bf U} = (U_1, \ldots,  U_d)^\top$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$ provides $Y_i \sim F_i, \, \forall \, i$.

### 2021-02-25 020-simulation-algorithm.Rmd
small changes up to 020

@@
 
-1. Simulate a random vector $\bf Z$ with $d$ *independent* and *identical* standard normal components.
+1. Simulate a random vector $\bf Z$ with $d$ *independent and identically distributed* (iid) standard normal components.
 2. Determine the input matrix $\Sigma_{\bf Z}$ that corresponds with the specified output $\Sigma_{\bf Y}$.
@@
 4. Set $X$ by $X \gets MZ$.
-5. $\text{Return} \; Y \; \text{where} \; Y_i \gets F_{Y_i}^{-1}[\Phi(X_i)], \; i=1,2,...,d$.
+5. $\text{Return} \; Y \; \text{where} \; Y_i \gets F_{Y_i}^{-1}[\Phi(X_i)], \; i=1,...,d$.
 
@@
 In general, there is no closed-form correspondence between the components of the input $\Sigma_{\bf Z}$ and target $\Sigma_{\bf Y}$.
-Matching the correlations involves evaluating or approximating $\binom{d}{2}$ integrals of the form $EY_iY_j = \int \int y_i y_j f_{X}(F_i^{-1}(\Phi(z_i)), F_j^{-1}(\Phi(z_j))dy_idy_j$, for $i,j=1,2,\ldots,d, \; i \neq j$.
+Matching the correlations involves evaluating or approximating $\binom{d}{2}$ integrals of the form $EY_iY_j = \int \int y_i y_j f_{X}(F_i^{-1}(\Phi(z_i)), F_j^{-1}(\Phi(z_j))dy_idy_j$, for $i,j=1,\ldots, d, \; i \neq j$.
 For high-dimensional data, these evaluations are often too costly to enable feasible simulation studies.
@@
 There is hope, however, as @GH02 also showed that replicating an indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$.
-This provides evidence that our nearest positive semi-definite (PSD) augmented approach will maintain reasonable accuracy if our input matching scheme returns an indefinite matrix, at least for the rank-based matching scheme described above.
+This provides evidence that our nearest positive definite (PD) augmented approach will maintain reasonable accuracy if our input matching scheme returns an indefinite matrix, at least for the rank-based matching scheme described above.
 
@@
 1. Mapping step
-   i. Convert the target Spearman correlation matrix $R_S$ to the corresponding MVN Pearson correlation $R_X$.   
+  (i) Convert the target Spearman correlation matrix $R_S$ to the corresponding MVN Pearson correlation $R_X$.
   or
-i$^\prime$. Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$.
-2. Checking admissibility and, if needed, nearest correlation computation
-    (i) Check that $R_X$ is a correlation matrix, a PSD matrix with 1's along the diagonal.   
+  (i$^\prime$) Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$.
+2. Check admissibility and, if needed, compute the nearest correlation matrix.
+  (i) Check that $R_X$ is a correlation matrix, a positive [semi]definite matrix with 1's along the diagonal.   
 	(ii) If $R_X$ is a correlation matrix, the input matrix is *admissible* in this scheme.
@@
   (ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
-	(iii) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;
+  (iii) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$.
 
@@
 Yet this scheme still suffers similar issues as the original Pearson-matching NORTA algorithm regarding the increasing probability of encountering a non-admissible correlation matrix as dimension increases.
-In our experience, the *mapping step* for large $d$ almost always produces a $R_X$ that is not in $\Omega_P$ (or even a correlation matrix).
+In our experience, the mapping step for large $d$ almost always produces a $R_X$ that is not in $\Omega_P$ (or even a correlation matrix).
 The second step handles this issue by computing the nearest correlation matrix to $R_X$.
@@
 
-$r = 2 \sin \left(\frac{\pi}{6} \rho\right) \\
-r = \sin \left(\frac{\pi}{2} \tau\right)$
+$$
+r = 2 \sin \left(\frac{\pi}{6} \rho\right) \\
+r = \sin \left(\frac{\pi}{2} \tau\right)
+$$
 
@@
 
-* Determine the input correlation, $\tilde{P}$, such that the resulting random samples using the NORTA algorithm have an estimated Pearson correlation that matches $P$
+* Determine the input correlation, $\tilde{R}$, such that the resulting random samples using the NORTA algorithm have an estimated Pearson correlation that matches $R$
 

### 2021-02-25 index.Rmd
small changes up to 020

@@
 This article introduces the `bigsimr` R interface to the Julia `Bigsimr` package.
-This high-package provides a flexible, scaleable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions with specified Pearson, Kendall, or Spearman correlation matrices.
-It also contains additional high-performance features, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix.
+This high-level package provides a flexible and scaleable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions with specified Pearson, Spearman, or Kendall correlation matrices.
+It also contains additional high-performance features, including multi-core algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix.
 Finally, we demonstrate example applications enabled via `bigsimr` by applying these functions to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=878$ patients and dimension $d=1000$.
@@
 library(tidyverse)
-library(knitr)
-library(printr)
-library(rmarkdown)
-library(bookdown)
 
@@
 ```{r countWords, include=FALSE, eval=FALSE}
-rmdFiles <- dir(pattern = 'Rmd$', ignore.case= T)
+rmdFiles <- dir(pattern = 'Rmd$', ignore.case = TRUE)
 # rmdFiles <- rmdFiles[-grep('^_', rmdFiles)]

### 2021-02-25 data-raw/benchmark_dependences.R
final round of updates to benchmark data

@@ -47,7 +47,18 @@ bs$pearson_match(0.5, d, d)
 s <- bs$cor_randPSD(100)
 p <- bs$cor_convert(s, bs$Spearman, bs$Pearson)
 r <- bs$cor_nearPD(p)
-JuliaCall::julia_call("Bigsimr.iscorrelation", r)
+bs$cor_nearPD(p, 1e-6)
+bs$cor_nearPD(p, 1e-6, 1e-6)
+
+k <- bs$cor_randPSD(10000)
+p <- bs$cor_convert(s, bs$Kendall, bs$Pearson)
+JuliaCall::julia_call("Bigsimr.iscorrelation", p)
+system.time(r1 <- bs$cor_nearPD(p, 1e-6, 1e-2))
+system.time(r2 <- bs$cor_nearPD(p, 1e-6, 1e-6))
+system.time(r3 <- bs$cor_fastPD(p))
+JuliaCall::julia_call("Bigsimr.iscorrelation", r1)
+JuliaCall::julia_call("Bigsimr.iscorrelation", r2)
+JuliaCall::julia_call("Bigsimr.iscorrelation", r3)
 
 d <- dist$Gamma(3, 0.01)
 bs$rvec(10, bs$cor_randPD(2), c(d, d))
@@ -56,7 +67,7 @@ tmp_cor <- bs$cor_randPD(100)
 tmp_margins <- make_margins(shapes[1:100], rates[1:100])
 bs$pearson_match(tmp_cor, tmp_margins)
 
-rm(p, r, s, d, tmp_cor, tmp_margins, shapes, rates)
+rm(p, r, r1, r2, r3, s, d, tmp_cor, tmp_margins, shapes, rates)
 gc()
 
 
@@ -108,7 +119,7 @@ bench <- function(type, d, n=1000) {
     needed_near_pd <- FALSE
     if (!JuliaCall::julia_call("Bigsimr.iscorrelation", adj_corr)) {
       needed_near_pd <- TRUE
-      adj_corr_pd <- bs$cor_nearPD(adj_corr)
+      adj_corr_pd <- bs$cor_nearPD(adj_corr, 1e-6, 0.005)
     }
   })
 
@@ -129,20 +140,32 @@ bench <- function(type, d, n=1000) {
 
 
 dims <- c(100, 250, 500, 1000, 2500, 5000, 10000)
-pearson_bench <- lapply(dims, function(d) bench("Pearson", d)) %>%
-  do.call(what = bind_rows)
+pearson_bench <- local({
[…]

### 2021-02-25 010-background.Rmd
using renv and box for package management

@@
 ```{r ch010-background, include=FALSE, cache=FALSE}
-devtools::load_all()
-library(tidyverse)
+box::use(
+  dplyr[...],
+  ggplot2[...],
+  stringr[str_sub],
+  GGally[ggpairs, wrap],
+  knitr[kable]
+)
+load("data/example_genes.rda")
+load("data/example_brca.rda")
 d <- 1000
@@
 num_patients <- 5
-gene_sample  <- sample(example_genes, num_genes)
+gene_sample  <- sample(example_genes[1:1000], num_genes)
 
@@
 small_brca %>%
-  knitr::kable(booktabs = TRUE, caption = cap)
+  kable(booktabs = TRUE, caption = cap)
 ```
@@
 Modeling these data motivate our simulation methodology."}
-GGally::ggpairs(
+ggpairs(
   data = example_brca[, gene_sample],
-  upper = list(continuous = GGally::wrap('cor', method = "spearman"))
+  upper = list(continuous = wrap('cor', method = "spearman"))
 ) + theme_bw()

### 2021-02-25 030-bigsimr-package.Rmd
using renv and box for package management

@@
 # The `bigsimr` R package {#package}
+
+```{r ch030-basic-example-setup, echo=TRUE, cache=FALSE}
+box::use(
+  patchwork[...],
+  dplyr[...],
+  ggplot2[...],
+  tidyr[drop_na],
+  bigsimr[bigsimr_setup, distributions_setup],
+  JuliaCall[julia_call]
+)
+
+Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
+bs <- bigsimr_setup(pkg_check = FALSE)
+dist <- distributions_setup()
+```
 
@@
 -->
-
-```{r ch030-basic-example-setup, echo=TRUE, cache=FALSE}
-library(tidyverse)
-library(patchwork)
-library(bigsimr)
-
-Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
-bs <- bigsimr_setup()
-dist <- distributions_setup()
-```
 
@@
 # Step 2. Check admissibility
-is_valid_correlation <- function(rho) {
-  all(diag(rho) == 1) && matrixcalc::is.positive.semi.definite(rho)
-}
-is_valid_correlation(R_X)
+julia_call("Bigsimr.iscorrelation", R_X)
 ```

### 2021-02-25 040-monte-carlo-evaluation.Rmd
using renv and box for package management

@@
 ```{r ch040-LoadLib040, include=FALSE}
-devtools::load_all()
-library(tidyverse)
-library(bigsimr)
+box::use(
+    dplyr[...],
+    ggplot2[...],
+    patchwork[...],
+    tidyr[pivot_longer],
+    bigsimr[bigsimr_setup, distributions_setup],
+    knitr[include_graphics]
+)
+load("data/bivariate_normal_sims.rda")
+load("data/bivariate_gamma_sims.rda")
+load("data/bivariate_nbinom_sims.rda")
+
+load("data/benchmark_dependences.rda")
 
 Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
-bs <- bigsimr_setup()
+bs <- bigsimr_setup(pkg_check = FALSE)
 dist <- distributions_setup()
-
-cores <- as.integer(parallel::detectCores() - 1)
 set.seed(10162020)
@@
 
-```{r ch040-gpuVScpuFig, echo=F, out.width='80%', fig.align='center', fig.cap="Computation times as d increases.
-We filter to the top 1, 5, 10, 15, 20, 25\\% expressing genes (in terms of median expression.)"}
-knitr::include_graphics("fig/cpu-gpu-times.png")
+```{r ch040-gpuVScpuFig, echo=FALSE, out.width='80%', fig.asp=1, fig.cap="Computation times as d increases."}
+dat_long <- benchmark_dependences %>%
+  select(-total_time, -n_sim, -needed_near_pd) %>%
+  pivot_longer(cols = c(corr_time:sim_time), names_to = "Step", values_to = "Time") %>%
+  mutate(dim_size = if_else(dim <= 1000, "Low Dimensional", "High Dimensional"),
+         dim_size = factor(dim_size, levels = c("Low Dimensional", "High Dimensional")),
+         dim = factor(dim),
+         Step = factor(Step,
+                       levels = c("corr_time",
+                                  "adj_time",
+                                  "admiss_time",
+                                  "sim_time"),
+                       labels = c("Compute Correlation",
+                                  "Adjust Correlation",
+                                  "Check Admissibility",
+                                  "Simulate Data"))) %>%
+  rename(Correlation = corr_type, Dimensions = dim)
+
+ggplot(dat_long, aes(Correlation, Time, fill=Step)) +
+  geom_bar(position = "stack", stat = "identity") +
+  labs(y = "Time (Seconds)") +
+  facet_grid(dim_size ~ Dimensions, scales = "free") +
+  theme(axis.text.x = element_text(angle = -90))
 ```

### 2021-02-25 050-examples.Rmd
using renv and box for package management

@@
 ```{r ch050-LoadLib, include=FALSE, cache=FALSE}
-devtools::load_all()
-library(tidyverse)
-library(bigsimr)
+box::use(
+  dplyr[...],
+  ggplot2[...],
+  patchwork[...],
+  bigsimr[...],
+  knitr[include_graphics]
+)
+load("data/example_brca.rda")
+load("data/example_genes.rda")
 
 Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
-bs <- bigsimr_setup()
+bs <- bigsimr_setup(pkg_check = FALSE)
 dist <- distributions_setup()
 
-cores <- as.integer( parallel::detectCores() - 1 )
 set.seed(10142020)
@@
 # Estimate Spearman's correlation on the count data
-type <- 'Spearman'
-R_S <- bs$cor(as.matrix(brca1000), bs[[type]])
+R_S <- bs$cor(as.matrix(brca1000), bs$Spearman)
 ```
@@
   s2 <- s^2
-  p <- m/s2
-  r <- m^2 / (s2 - m)
-  c(size = r, prob = p)
+  c(size = m^2 / (s2 - m), prob = m/s2)
 }
@@
 probs <- nbinom_fit["prob",]
-
 nb_margins <- make_nbinom_margins(sizes, probs)
@@
 This gives rise to highly variable counts and, typically, less restriction on potential pairwise correlation pairs.
-Once the functions are defined/executed to complete marginal estimation, we specify targets and generate the desired random vectors using `rvec`:
-
-Now we check admissibility of specified correlation matrix. We begin with the pairwise, 2D MC estimated bounds. And then follow the `r d`D bounds.
+Once the functions are defined/executed to complete marginal estimation, we specify targets and generate the desired random vectors using `rvec`.
+Now we check admissibility of specified correlation matrix.
+We begin with the pairwise, 2D MC estimated bounds.
+And then follow the `r d`-D bounds.
 
@@
 # 1. Mapping step first
-R_X <- bs$cor_convert(R_S, bs[[type]], bs$Pearson)
+R_X <- bs$cor_convert(R_S, bs$Spearman, bs$Pearson)
 
 # 2a. Check admissibility
-is_valid_correlation <- function(rho) {
-  all(diag(rho) == 1) && matrixcalc::is.positive.semi.definite(rho)
-}
-is_valid_correlation(R_X)
+(is_valid_corr <- JuliaCall::julia_call("Bigsimr.iscorrelation", R_X))
 
 # 2b. compute nearest correlation
-if (!is_valid_correlation(R_X)) {
-  R_Xtilde <- bs$cor_nearPD(R_X)
+if (!is_valid_corr) {
+  R_X_pd <- bs$cor_nearPD(R_X)
   ## Quantify the error
-  targets <- R_X[lower.tri(R_X)]
-  approximates <- R_Xtilde[lower.tri(R_Xtilde)]
-  ## R_Xtilde[1:5, 1:5]
-  ## norm( x = R_X - R_Xtilde, type = "F" )
-  R_X  <- R_Xtilde
+  targets      <- R_X[lower.tri(R_X, diag = FALSE)]
+  approximates <- R_X_pd[lower.tri(R_X_pd, diag = FALSE)]
+  R_X  <- R_X_pd
 }
-summary(targets - approximates)
+summary(abs(targets - approximates))
 ```
 
-While the exact $d \times d$ Spearman correlation matrix is not strictly admissible in our scheme (seen by the non-PSD result above), the approximation is of sufficient quality for most applications with a maximum pairwise of `r max( targets - approximates)` and average error of `r mean( targets - approximates)`.
+While the exact $d \times d$ Spearman correlation matrix is not strictly admissible in our scheme (seen by the non-positive definite result above), the approximation is of sufficient quality for most applications with a maximum pairwise of `r max( abs(targets - approximates))` and average absolute error of `r mean(abs(targets - approximates))`.
 
@@
 dir.create("results", showWarnings = FALSE, recursive = TRUE)
-brcaSimRDS <- paste0("results/brca", d, "sim.rds")
-if(!file.exists(brcaSimRDS)) {
+brca_sim_path <- paste0("results/brca_", d, "_sim.rds")
+if(!file.exists(brca_sim_path)) {
   set.seed(44)
-  n <- 10000
-  sim_nbinom <- bs$rvec(n, R_X, nb_margins) 
+  sim_nbinom <- bs$rvec(10000, R_X, nb_margins) 
   colnames(sim_nbinom) <- colnames(brca1000)
-  saveRDS(sim_nbinom, brcaSimRDS)
+  saveRDS(sim_nbinom, brca_sim_path)
 }
-sim_nbinom <- readRDS(brcaSimRDS)
+sim_nbinom <- readRDS(brca_sim_path)
 ```
@@
 
-```{r ch050-makeSimDataFig, cache=FALSE}
-if(!file.exists("fig/ch050-simDataFig.png")) {
-  p <- GGally::ggpairs(
-    data = as.data.frame(sim_nbinom[,example_genes[1:3]]),
+```{r ch050-simDataFig, cache=FALSE, fig.cap="Simulated data for three selected high-expressing genes generally replicates the estimated data structure.
+The data do not exhibit outlying points, but do possess the desired Spearman correlations, central tendencies, and discrete values."}
+set.seed(02242021)
+num_genes    <- 3
+gene_sample  <- sample(example_genes[1:1000], num_genes)
+
+GGally::ggpairs(
+  data = as.data.frame(sim_nbinom[, gene_sample]),
   upper = list(continuous = GGally::wrap('cor', method = "spearman"))
 ) + theme_bw()
-  ggsave("fig/ch050-simDataFig.png", plot = p, dpi = 300) 
-}
-```
-
-```{r ch050-simDataFig, fig.cap='Simulated data for three selected high-expressing genes generally replicates the estimated data structure.
-The data do not exhibit outlying points, but do possess the desired Spearman correlations, central tendencies, and discrete values.'}
-knitr::include_graphics('fig/ch050-simDataFig.png')
 ```
@@
 ```{r ch050-makeFigBRCA, cache=FALSE}
+R_S_hat <- bs$cor(as.matrix(sim_nbinom), bs$Spearman)
+sim_nbinom_fit <- apply(sim_nbinom, 2, nbinom_mom)
+corr_compare <- tibble(
+  `Target Correlation` = R_S[lower.tri(R_S, diag = FALSE)],
+  `Estimated Correlation` = R_S_hat[lower.tri(R_S_hat, diag = FALSE)]
+)
+target_compare <- tibble(
+  `Target Size` = sizes,
+  `Estimated Size` = sim_nbinom_fit[1,]
+)
+logprob_compare <- tibble(
+  `Target Log Probability` = log(probs),
+  `Estimated Log Probability` = log(sim_nbinom_fit[2,])
+)
+
 if(!file.exists("fig/ch050-figBRCA.png")) {
-    ## estimate rho from sims and plot
-    R_S_hat <- bs$cor(as.matrix(sim_nbinom), bs[[type]])
-    ## triplot
-    ## png("fig/ch050-figBRCA.png", width = 4, res = 100, units = 'in')
-    png("fig/ch050-figBRCA.png", res = 300, width = 5, height = 4, units = 'in')
-    par(mfrow=c(1,3))
-    plot(R_S[lower.tri(R_S)], R_S_hat[lower.tri(R_S_hat)],
-         xlim = c(-1, 1), ylim = c(-1, 1),
-         pch = 1, cex = 1, col = rgb(0, 0, 0, 0.05),
-         xlab = "Target Correlation", ylab = "Estimated Correlation")
-         ## main = "Spearman Cor. Matching")
-    abline(a = 0, b = 1, col="red", lty="dashed")
-    ## estimate prob,size from sims and plot
-    nbinom_size_prob_hat <- apply(sim_nbinom, 2, nbinom_mom)
-    plot(sizes, nbinom_size_prob_hat[1,],
-         pch = 1, cex = 1, col = rgb(0, 0, 0, 1),
-         xlab = "Target Size", ylab = "Estimated Size")
-         ## main = "Marginal NB Size Matching")
-    abline(a=0, b=1, col="red", lty="dashed")
-    plot(log(probs), log(nbinom_size_prob_hat[2,]),
-         pch = 1, cex = 1, col = rgb(0, 0, 0, 1),
-         xlab = "Target Log Probability", ylab = "Estimated Log Probability")
-         ## main = "Marginal NB Prob Matching")
-    abline(a=0, b=1, col="red", lty="dashed")
-    dev.off()
-    par(mfrow=c(1,1))
+  p1 <- ggplot(corr_compare, aes(`Target Correlation`, `Estimated Correlation`)) +
+    geom_point(alpha = 0.05) + 
+    geom_abline(slope = 1, col = "red", lty = "dashed") +
+    theme_bw() + coord_equal()
+
+  p2 <- ggplot(target_compare, aes(`Target Size`, `Estimated Size`)) +
+    geom_point(alpha = 0.85) + 
+    geom_abline(slope = 1, col = "red", lty = "dashed") +
+    theme_bw() + coord_equal()
+
+  p3 <- ggplot(logprob_compare, 
+               aes(`Target Log Probability`, `Estimated Log Probability`)) +
+    geom_point(alpha = 0.85) + 
+    geom_abline(slope = 1, col = "red", lty = "dashed") +
+    theme_bw() + coord_equal()
+
+  p <- p1 + p2 + p3
+
+  ggsave(filename = "fig/ch050-figBRCA.png",
+         plot = p,
+         device = "png",
+         width = 8,
+         height = 4,
+         units = "in")
 }
@@
 The dashed red lines indicated equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes)."}
-knitr::include_graphics('fig/ch050-figBRCA.png')
+include_graphics('fig/ch050-figBRCA.png')
 ```
@@
 B <- nrow(sim_nbinom)
-threshold <- rep(1, d)
-pHat <- mean(apply(sim_nbinom, 1, function(X, y0 = threshold) {
-  all(X >= y0)
+y0 <- 1
+pHat <- mean(apply(sim_nbinom, 1, function(Y) {
+  all(Y >= y0)
 }))
@@
 B <- nrow(sim_nbinom)
-threshold <- rep(1, d)
-mean(apply(sim_nbinom, 1, function(X, y0 = threshold) {
-  all( X >= y0)
+y0 <- 1
+mean(apply(sim_nbinom, 1, function(Y) {
+  all(Y >= y0)
 }))
@@
 
-In this small example, we simulate $m=`r m`$ data sets with the number of simulated vectors matching the number of patients in the BRCA data set, $N=`r nrow(brca)`$.
-Since our simulation is much faster for the Pearson correlation type (see Figure \@ref(fig:ch040-gpuVScpuFig)), we only convert the Spearman correlation matrix once (and ensure its PSD).
-At each iteration, we estimate the quadratic loss from the specified ${\bf R}_{Spearman}$, producing a distribution of loss values.
+In this small example, we simulate $m=`r m`$ data sets with the number of simulated vectors matching the number of patients in the BRCA data set, $N=`r nrow(example_brca)`$.
+Since our simulation is much faster for the Pearson correlation type (see Figure \@ref(fig:ch040-gpuVScpuFig)), we only convert the Spearman correlation matrix once (and ensure it is positive definite).
+At each iteration, we estimate the quadratic loss (residual sum of squared errors) from the specified ${\bf R}_{Spearman}$, producing a distribution of loss values.
 
 ```{r ch050-quadlossECHO, echo=TRUE, eval=FALSE, cache=FALSE}
-## Simulate random vectors equal to the sample size
-N <- nrow(brca)
-## create m random vectors and estimate correlation
+# Simulate random vectors equal to the sample size
+N <- nrow(example_brca)
+# create m random vectors and estimate correlation
 simRho <- replicate(n = m, expr = {
   tmpSim <- bs$rvec(N , R_X, nb_margins)
-  bs$cor(tmpSim, bs[[type]])
+  bs$cor(tmpSim, bs$Spearman)
 }, simplify = FALSE)
-## find quadratic loss at each rep
-quadLoss <- sapply(simRho, rags2ridges::loss, T = R_S, type = "quadratic")
+# Evaluate the residual sum of squared error
+sapply(simRho, function(R) sum((R - R_S)^2))
+```
+
+```{r loss_examples, include=FALSE, eval=FALSE}
+sapply(simRho, function(R) norm(R - R_S))           # norm of diff
+sapply(simRho, function(R) norm(R) - norm(R_S))     # diff of norms
+sapply(simRho, function(R) sum((R - R_S)^2))        # RSS
+sapply(simRho, function(R) mean((R - R_S)^2))       # MSE
+sapply(simRho, function(R) sqrt(mean((R - R_S)^2))) # RMSE
+sapply(simRho, function(R) mean(abs(R - R_S)))      # MAE
 ```
@@
 ```{r ch050-lossResults, cache=FALSE}
-frobeniusRDS <- paste0("results/brca", d, "_frobenius.rds")
-if (!file.exists(frobeniusRDS)) {
-    N <- nrow(brca1000)
+frobenius_path <- paste0("results/brca_", d, "_frobenius.rds")
+if (!file.exists(frobenius_path)) {
+    N <- nrow(example_brca)
     ## create m random vectors and estimate correlation
     simRho <- replicate(n = m, expr = {
-        tmpSim <- bs$rvec(N , R_X, nb_margins)
-        bs$cor(tmpSim, bs[[type]])
+        tmpSim <- bs$rvec(N, R_X, nb_margins)
+        bs$cor(tmpSim, bs$Spearman)
     }, simplify = FALSE)
-    frobenius <- sapply(simRho, rags2ridges::loss, T = R_S, type = "frobenius")
-    saveRDS(frobenius, frobeniusRDS)
+    frobenius <- sapply(simRho, function(R) sum((R - R_S)^2))
+    saveRDS(frobenius, frobenius_path)
 }
-frobenius <- readRDS(frobeniusRDS)
+frobenius_loss <- readRDS(frobenius_path)
 ```
@@
 ```{r ch050-loss-summary, echo=TRUE, eval=TRUE, cache=FALSE}
-summary(frobenius)
+summary(frobenius_loss)
 ```

### 2021-02-25 index.Rmd
using renv and box for package management

@@
 ```{r indexSetup, include=FALSE, cache=FALSE}
-library(tidyverse)
-
 set.seed(09212020)
@@
 knitr::write_bib(c(
-  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'rslurm'
+  .packages(), 'bookdown', 'knitr', 'rmarkdown'
 ), 'packages.bib')
@@
 -->
+
 Another central issue lies in characterizing dependency between components in the high-dimensional random vector.

### 2021-02-25 _outgoing.R

@@ -1,5 +1,4 @@
 # Remove all objects
-# rm(list = ls(all.names = TRUE))
+rm(list = ls(all.names = TRUE))
 
 # Unload all packages
-# pacman::p_unload(pacman::p_loaded(), character.only = TRUE)

### 2021-02-25 050-examples.Rmd

@@
 ```{r ch050-LoadLib, include=FALSE, cache=FALSE}
-box::use(
-  dplyr[...],
-  ggplot2[...],
-  patchwork[...],
-  bigsimr[...],
-  knitr[include_graphics]
-)
-load("data/example_brca.rda")
-load("data/example_genes.rda")
+devtools::load_all()
+library(tidyverse)
+library(patchwork)
+library(bigsimr)
 
@@
 
+cores <- as.integer(parallel::detectCores() - 1)
 set.seed(10142020)
@@
 The dashed red lines indicated equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes)."}
-include_graphics('fig/ch050-figBRCA.png')
+knitr::include_graphics('fig/ch050-figBRCA.png')
 ```

### 2021-02-25 R/method_of_moments.R
remove data that is auto-generated

@@ -1,0 +1,18 @@
+#' @export
+mom_norm <- function(x) {
+  c(mean = mean(x), sd = sd(x))
+}
+
+#' @export
+mom_nbinom <- function(x) {
+  m <- mean(x)
+  s <- sd(x)
+  c(size = m^2 / (s^2 - m), prob = m / s^2)
+}
+
+#' @export
+mom_gamma <- function(x) {
+  m <- mean(x)
+  s <- sd(x)
+  c(shape = m^2 / s^2, rate = m / s^2)
+}

### 2021-02-25 data-raw/bivariate_normal_sims.R

@@ -1,6 +1,10 @@
 ## code to prepare `bivariate_normal_sims` dataset goes here
+box::use(
+  bigsimr[...],
+  usethis[use_data],
+  ./R/method_of_moments[mom_norm]
+)
 
-library(bigsimr)
 Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
 bs <- bigsimr_setup(pkg_check = FALSE)
 dist <- distributions_setup()

### 2021-02-25 data-raw/brca.R

@@ -7,10 +7,10 @@ brca1 <- as.data.frame(t(brca0$dat))
 gene_median <- apply(brca1, 2, median)
 gene_median_order <- order(gene_median, decreasing = TRUE)
 
-example_brca     <- brca1[, head(gene_median_order, 10000)]
+example_brca     <- brca1[, gene_median_order]
 example_patients <- substring(row.names(example_brca), first = 1, last = 12)
 example_genes    <- colnames(example_brca)
 
-usethis::use_data(example_brca, overwrite = TRUE)
-usethis::use_data(example_patients, overwrite = TRUE)
-usethis::use_data(example_genes, overwrite = TRUE)
+save(example_brca, file = "data/example_brca.rda")
+save(example_patients, file = "data/example_patients.rda")
+save(example_genes, file = "data/example_genes.rda")

### 2021-02-25 040-monte-carlo-evaluation.Rmd
revise 040

--- 2021-02-24 040-monte-carlo-evaluation.Rmd
@@
 library(tidyverse)
-library(bigsimr)
-
-Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
-bs <- bigsimr_setup()
-dist <- distributions_setup()
-
-cores <- as.integer(parallel::detectCores() - 1)
-set.seed(10162020)
-prob <- 3e-04
+## library(bigsimr)
+## Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
+## bs <- bigsimr_setup()
+## dist <- distributions_setup()
 ```
@@
 
-Overall, the studies show that our methodology is generally accurate across the entire range of possible correlation values for the rank-based dependency measures, at least in these limited simulation settings for the rank-based correlations.
-For the two non-normal bivariate marginals, the Pearson correlation matching is approximate.
-For discrete margins, matching the dependency measures was somewhat less accurate, even for the rank-based metrics, and particularly inaccurate near the lower bound of Pearson correlations.
+Overall, the studies show that our methodology is generally accurate across the entire range of possible correlation values all three dependency measures, at least in these limited simulation settings for the rank-based correlations.
+Our Pearson matching performs nearly as well as Spearman or Kendall, except a slight increase in error for negative binomial case.
+This is due the particularly large counts generated with our choice of parameters for $p$ and $r$.
 
-```{r ch040-bPlot, cache=FALSE, fig.asp=1.30, fig.weight=4, fig.cap="Bivariate simulations match specified correlations.
-The horizontal axis plots the specified target correlations across the entire range of possible correlations for each bivariate margin.
-Normal margins are plotted in dark blue, gamma in medium blue, and negative binomial in light blue.
+```{r ch040-bPlot, cache=FALSE, fig.asp=1.30, fig.weight=4, fig.cap="Bivariate simulations match target correlations across the entire range of feasible correlations.
+The horizontal axis plots the specified target correlations for each bivariate margin.
+Normal margins are plotted in dark dark grey, gamma in medium grey, and negative binomial in light grey.
 As the number of simulated vectors $B$ increases from left to right, the variation in estimated correlations (vertical axis) decreases.
-The dashed line indicates equality between the specified and estimated correlations.
-Only the normal margins match the Pearson correlations (top row) exactly across the entire range of correlations.
-The two non-normal bivariate random vectors experience attenuation (downward bias) for extreme Pearson correlations (due to limitations in our algorithm) and more restriction (due to the Frechet bounds).
-The rank-based correlations (bottom two rows) are matched exactly for all margins and obtain the full range of possible dependencies."}
+The dashed line indicates equality between the specified and estimated correlations."}
 # https://www.datanovia.com/en/blog/how-to-change-ggplot-facet-labels/
@@
 numGroups <- length(levels(allDat$margins))
-myColors <- rev(RColorBrewer::brewer.pal(n = numColors, name = "Blues")[ ((numColors - numGroups) + 1):numColors])
+myColors <- rev(RColorBrewer::brewer.pal(n = numColors, name = "Greys")[ ((numColors - numGroups) + 1):numColors])
 
@@
 
-The accuracy appears adequate for many applications. In practice, we recommend evaluating the accuracy in each application setting, using methods similar to those presented above.
-
 <!-- 
@@
 
-With information of our method's accuracy from a low-dimensional perspective, we now turn to assessing whether the `bigsimr` can scale to larger dimensional problems.
-Specifically, we seek evidence to determine whether the method can scale to higher dimensions in a practical time.
-Using our motivating RNA-seq data, we filtered the original 20,501 genes to the high-expressing genes at increasing percentiles, $1, 5, 10, 15, 20, 25\%$, to obtain $d=\{206, 1026, 2051, 3076, 4101, 5127\}$ marginals and $\binom{d}{2}$ pairwise correlations at each setting.
-For example, for $d=5127$ there are $13,140,501$ correlation coefficients.
-We estimated the marginal negative binomial parameters and the correlation coefficients from the RNA-seq data to seed our simulations.
-(See [Example applications](examples) for a detailed description of the estimation).
+With information of our method's accuracy from a low-dimensional perspective, we now turn to assessing whether the `bigsimr` can scale to larger dimensional problems with practical computation times.
+Specifically, we ultimately generate $B=1,000$ random vectors for $d=\{100, 250, 500, 1000, 2500, 5000, 10000\}$ for each correlation type, $\{Pearson, Spearman, Kendall\}$ while timing the algorithm's major steps.
 
-Figure \@ref(fig:ch040-gpuVScpuFig) displays computation times using various high-performance settings (1 central processing unit, CPU-1, versus twenty CPUs, CPU-20; with and without GPU acceleration, GPU-1; GPU-20) to produce $B=10,000$ random vectors.
-The Pearson simulations are much faster since the correlation conversion steps are avoided (pre-processing step; see [Algorithms](algorithms)), but as we know from above the accuracy will suffer slightly, especially near the negative boundary of the possible correlations.
-Matching Spearman's correlation at larger $d$ becomes costly if one desires to produce $B=10,000$ random vectors at many different simulation settings, but the conversion steps need only be computed once (see [MC evaluation of correlation estimation efficiency](MC-evaluation-of-correlation-estimation-efficiency) for an example of this strategy).
+To mimick the workflow, we first produce a synthetic "data set" by completing the following steps:  
+1.
+Produce heterogeneous gamma marginals by randomly selecting the $j^{th}$ gamma shape parameter from $U_j \sim uniform(1,10), j=1,\ldots,d$ and the $j^{th}$ rate parameter from $V_j \sim exp(1/5), j=1,\ldots,d$, with the constant parameters determined arbitrarily.
+2. Produce a random full-rank Pearson correlation matrix via `cor_randPD` of size $d \times d$.  
+3. Simulate a "data set" of $1,000 \times d$ random vectors via `rvec` (without matching the Pearson exactly).
 
-```{r ch040-gpuVScpuFig, echo=F, out.width='80%', fig.align='center', fig.cap="Computation times as d increases.
-We filter to the top 1, 5, 10, 15, 20, 25\\% expressing genes (in terms of median expression.)"}
-knitr::include_graphics("fig/cpu-gpu-times.png")
+With the synthetic data set in hand, we complete and time the following 4 steps:  
+i. Estimate the correlation matrix from the "data" in the *Compute Correlation* step.  
+1. Map the correlations to initialize the algorithm (Pearson to Pearson, Spearman to Pearson, or Kendall to Pearson) in the *Adjust Correlation* step.  
+2. Check whether the mapping produces a valid correlation matrix and, if not, find the nearest PD correlation matrix in the *Check Admissibility* step.  
+3. Simulate $1,000$ vectors via the Inverse Transform method in the *Simulate Data* step.
+
+The results for $d \leq 500$ are fast with all times under 3 seconds.
+Table \@ref(tab:ch040-moderateDtab) displays the total computation time of the algorithms steps 1through 3, including estimation step i.
+
+```{r ch040-moderateDtab}
+dat_long <- benchmark_dependences %>%
+  select(-total_time, -n_sim, -needed_near_pd) %>%
+  pivot_longer(cols = c(corr_time:sim_time), names_to = "Step", values_to = "Time") %>%
+  mutate(dim = factor(dim),
+         Step = factor(Step,
+                       levels = c("corr_time", "adj_time", "admiss_time", "sim_time"),
+                       labels = c("Compute Correlation",
+                                  "Adjust Correlation",
+                                  "Check Admissibility",
+                                  "Simulate Data"))) %>%
+    rename(Correlation = corr_type, Dimensions = dim)
+
+tabTimes  <- dat_long %>%
+    filter(Dimensions %in% c(100, 250, 500)) %>%
+    group_by(Dimensions, Correlation) %>%
+    summarize('Total Time(Seconds)' = sum(Time)) %>%
+    ungroup()
+
+tabTimes %>%
+    knitr::kable(booktabs = TRUE,
+                 col.names = c("Dimension",
+                               "Correlation type",
+                               "Total Time (Seconds)" ),
+                 caption = 'Total time to produce 10,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.')
+```
+
+The results larger dimensional vectors show scalebility to ultra-high dimensions for all three correlation types, although the total times do become much larger. 
+Figure \@ref(ch040-largeDfig) displays computation times for $d=\{1000, 2500, 5000, 10000\}$.
+For $d$ of 1000 and 2500, the total time is under a couple of minutes.
+At $d$ of 5000 and 10,000, Pearson correlation matching in the *Adjust Correlation* step becomes costly.
+Interestingly, Pearson is actually faster than Kendall for $d=10,000$ due to bottlenecks in *Compute Correlation* and *Check Admissibility*.
+Uniformly, matching Spearman correlations is faster with total times under 5 minutes for $d=10,000$, making Spearman correlations the most computationally-friendly dependency type.
+With this in mind, we scaled the simulation to $d=20,000$ for the Spearman type and obtaned the 1,000 vectors in under an hour (data not shown).
+In principle, this would enable the simulation of an entire human-derived RNA-seq data set.
+We note that for a given target correlation matrix and margins, steps i, 1, and 2 only need to be computed once and the third step, *Simulate Data*, is fast under all schemes for all $d$ under consideration.
+
+```{r ch040-largeDfig, echo=F, fig.cap="Computation times as d increases."}
+# Set colors
+## RColorBrewer::display.brewer.all()
+numColors <- 5
+numGroups <- 4
+myColors <- rev(RColorBrewer::brewer.pal(n = numColors, name = "Greys")[ ((numColors - numGroups) + 1):numColors])
+
+dat_long %>%
+    filter(Dimensions %in% c(1000, 2500, 5000, 10000, 20000)) %>%
+    mutate(`Time (minutes)` = Time / 60) %>%
+    ggplot(aes(Correlation, `Time (minutes)`, fill=Step)) +
+    geom_bar(position = "stack", stat = "identity") +
+    scale_fill_manual(values = myColors ) +
+    scale_y_continuous(breaks = c(0, 5, 10, 15, 20, 25, 30),
+                     minor_breaks = NULL) +
+    facet_grid(. ~ Dimensions) +
+    theme(axis.text.x = element_text(angle = -90))
 ```
@@
 We expect our random vector generation to perform well for the vast majority of NORTA-feasible correlation matrices, but advise to check the performance before making inferences/further analyses.
+Finally, we note that one could use our single-pass algorithm `cor_fastPD` to produce a "close" (not nearest PD) to scale to even higher dimensions with greater loss of accuracy.

### 2021-02-25 050-examples.Rmd
changed object name

@@
     }, simplify = FALSE)
-    frobenius <- sapply(simRho, function(R) sum((R - R_S)^2))
-    saveRDS(frobenius, frobenius_path)
+    frobenius_loss <- sapply(simRho, function(R) sum((R - R_S)^2))
+    saveRDS(frobenius_loss, frobenius_path)
 }

### 2021-02-25 050-examples.Rmd
revise 050

--- 2021-02-25 050-examples.Rmd
@@
 This section demonstrates how to simulate multivariate data using `bigsimr`, aiming to replicate the structure of high-dimensional dependent count data.
-In an illustration of our proposed methodology, we seek to simulate RNA-sequencing data by producing simulated random vectors with assumed marginal distributions and with estimated parameters that mimic the observed data and its generating process.
-We seek realistic multivariate simulation to enable Monte Carlo (MC) methods for these data.
+In an illustration of our proposed methodology, we seek to simulate RNA-sequencing data by producing simulated random vectors mimicking the observed data and its generating process.
 Modeling RNA-seq using multivariate probability distributions is natural as inter-gene correlation is an inherent part of biological processes [@Wang2009b].
@@
 
-Our initial goal is to replicate the structure of the TCGA BRCA RNA-seq data set (see [Background](background)).
+We begin by estimating the structure of the TCGA BRCA RNA-seq data set (see [Background](background)).
 Ultimately, we will simulate $B=10,000$ random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with $d=`r d`$.
-Often researchers posit a negative binomial (NB) model as RNA-seq counts are often over-dispersed.
-All $d$ selected genes exhibit over-dispersion (data not shown) and, so, we proceed to estimate the NB parameters $(r_i, p_i), i=1,\ldots,d$ to determine the target marginal PMFs $f_i$ (via the method of moments).
+We assume a multivariate negative binomial (MVNB) model as RNA-seq counts are often over-dispersed and correlated.
+All $d$ selected genes exhibit over-dispersion (data not shown) and, so, we proceed to estimate the NB parameters $(r_i, p_i), i=1,\ldots,d$ to determine the target marginal PMFs $f_i$.
 To complete specification of the simulation algorithm inputs, we estimate the Spearman correlation matrix ${ \bf R}_{Spearman}$ to characterize dependency.
@@
 
-Now we estimate the marginal parameters using the built-in `apply` function:
+We apply these estimators to all `r d` genes across the `r nrow(brca1000)` patients:
 
@@
 Now we check admissibility of specified correlation matrix.
-We begin with the pairwise, 2D MC estimated bounds.
-And then follow the `r d`-D bounds.
 
@@
 
-While the exact $d \times d$ Spearman correlation matrix is not strictly admissible in our scheme (seen by the non-positive definite result above), the approximation is of sufficient quality for most applications with a maximum pairwise of `r max( abs(targets - approximates))` and average absolute error of `r mean(abs(targets - approximates))`.
+While the exact $d \times d$ Spearman correlation matrix is not strictly admissible in our scheme (seen by the non-positive definite result above), the approximation is close with a maximum absolute error of `r max( abs(targets - approximates))` and average absolute error of `r mean(abs(targets - approximates))` across the `r choose(d,2)` correlations.
 
@@
 The evaluation shows that the simulated counts approximately match the target parameters and exhibit the full range of estimated correlation from the data.
+
+<!-- 
 Utilizing 15 CPU threads in a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, the simulation completed in just over 2 minutes.
+your comment -->
 

### 2021-02-26 040-monte-carlo-evaluation.Rmd

@@
 ```{r ch040-LoadLib040, include=FALSE}
-devtools::load_all()
-library(tidyverse)
-## library(bigsimr)
-## Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
-## bs <- bigsimr_setup()
-## dist <- distributions_setup()
+box::use(
+    dplyr[...],
+    ggplot2[...],
+    patchwork[...],
+    tidyr[pivot_longer],
+    bigsimr[bigsimr_setup, distributions_setup],
+    knitr[include_graphics]
+)
+load("data/bivariate_normal_sims.rda")
+load("data/bivariate_gamma_sims.rda")
+load("data/bivariate_nbinom_sims.rda")
+
+load("data/benchmark_dependences.rda")
+
+Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
+bs <- bigsimr_setup(pkg_check = FALSE)
+dist <- distributions_setup()
+set.seed(10162020)
+prob <- 3e-04
 ```

### 2021-02-26 050-examples.Rmd

@@
 ```{r ch050-LoadLib, include=FALSE, cache=FALSE}
-devtools::load_all()
-library(tidyverse)
-library(patchwork)
-library(bigsimr)
+box::use(
+  dplyr[...],
+  ggplot2[...],
+  patchwork[...],
+  bigsimr[...],
+  knitr[include_graphics]
+)
+load("data/example_brca.rda")
+load("data/example_genes.rda")
 
@@
 
-cores <- as.integer(parallel::detectCores() - 1)
 set.seed(10142020)

### 2021-02-26 040-monte-carlo-evaluation.Rmd
renv+box fully implemented

--- 2021-02-25 040-monte-carlo-evaluation.Rmd
@@
 ```{r ch040-LoadLib040, include=FALSE}
-devtools::load_all()
-library(tidyverse)
-## library(bigsimr)
-## Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
-## bs <- bigsimr_setup()
-## dist <- distributions_setup()
+box::use(
+    dplyr[...],
+    ggplot2[...],
+    patchwork[...],
+    tidyr[pivot_longer],
+    bigsimr[bigsimr_setup, distributions_setup],
+    knitr[kable],
+    RColorBrewer[brewer.pal]
+)
+load("data/bivariate_normal_sims.rda")
+load("data/bivariate_gamma_sims.rda")
+load("data/bivariate_nbinom_sims.rda")
+load("data/benchmark_dependences.rda")
+
+Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
+bs <- bigsimr_setup(pkg_check = FALSE)
+dist <- distributions_setup()
+prob <- 3e-04
 ```
@@
     theme_bw()
-## ggsave('fig/plot-biGammaPlot.pdf')
 ```
@@
 
-## AGS 12/22/20: Removing the latex formatting fixes the html render.
-## But how to improve latex table formatting? Better to use the template
-tabMAE %>%
-    knitr::kable(booktabs = TRUE,
+kable(tabMAE, booktabs = TRUE,
       col.names = c("No. of random vectors",
@@
                     "Mean abs. error"),
-                 caption = 'Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.')
+      caption = "Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.")
 ```
@@
 To mimick the workflow, we first produce a synthetic "data set" by completing the following steps:
+
 1.
@@
 With the synthetic data set in hand, we complete and time the following 4 steps:
+
 i. Estimate the correlation matrix from the "data" in the *Compute Correlation* step.  
@@
 tabTimes %>%
-    knitr::kable(booktabs = TRUE,
+    kable(booktabs = TRUE,
           col.names = c("Dimension",
                         "Correlation type",
-                               "Total Time (Seconds)" ),
+                        "Total Time (Seconds)"),
           caption = 'Total time to produce 10,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.')
@@
 The results larger dimensional vectors show scalebility to ultra-high dimensions for all three correlation types, although the total times do become much larger. 
-Figure \@ref(ch040-largeDfig) displays computation times for $d=\{1000, 2500, 5000, 10000\}$.
+Figure \@ref(fig:ch040-largeDfig) displays computation times for $d=\{1000, 2500, 5000, 10000\}$.
 For $d$ of 1000 and 2500, the total time is under a couple of minutes.
@@
 
-```{r ch040-largeDfig, echo=F, fig.cap="Computation times as d increases."}
+```{r ch040-largeDfig, fig.cap="Computation times as d increases."}
 # Set colors
-## RColorBrewer::display.brewer.all()
 numColors <- 5
 numGroups <- 4
-myColors <- rev(RColorBrewer::brewer.pal(n = numColors, name = "Greys")[ ((numColors - numGroups) + 1):numColors])
+myColors <- rev(brewer.pal(n=numColors, name="Greys")[((numColors-numGroups)+1):numColors])
 

### 2021-02-26 050-examples.Rmd
renv+box fully implemented

@@
   ggplot2[...],
+  GGally[ggpairs, wrap],
   patchwork[...],
   bigsimr[...],
-  knitr[include_graphics]
+  knitr[include_graphics],
+  JuliaCall[julia_call],
+  ./R/utils[mom_nbinom]
 )
@@
 load("data/example_genes.rda")
+sim_nbinom <- readRDS("results/brca_1000_sim.rds")
+simRho <- readRDS("results/brca_1000_frobenius.rds")
 
@@
 dist <- distributions_setup()
-
-set.seed(10142020)
 ```
@@
 }
-
-nbinom_mom <- function(x) {
-  m <- mean(x)
-  s <- sd(x)
-  s2 <- s^2
-  c(size = m^2 / (s2 - m), prob = m/s2)
-}
 ```
@@
 
-```{r ch050-estMargins, echo = TRUE, eval = TRUE, cache=F}
-nbinom_fit <- apply(brca1000, 2, nbinom_mom)
+```{r ch050-estMargins, echo=TRUE}
+nbinom_fit <- apply(brca1000, 2, mom_nbinom)
 sizes <- nbinom_fit["size",]
@@
 # 2a. Check admissibility
-(is_valid_corr <- JuliaCall::julia_call("Bigsimr.iscorrelation", R_X))
+(is_valid_corr <- julia_call("Bigsimr.iscorrelation", R_X))
 
@@
 
-```{r ch050-runBRCA-echo, echo=TRUE, eval=FALSE, cache=FALSE}
+```{r ch050-runBRCA-echo, echo=TRUE, eval=FALSE}
 sim_nbinom <- bs$rvec(10000, R_X, nb_margins) 
 colnames(sim_nbinom) <- colnames(brca1000)
-```
-
-```{r ch050-runBRCA, echo=FALSE, results=FALSE, cache=FALSE}
-dir.create("results", showWarnings = FALSE, recursive = TRUE)
-brca_sim_path <- paste0("results/brca_", d, "_sim.rds")
-if(!file.exists(brca_sim_path)) {
-  set.seed(44)
-  sim_nbinom <- bs$rvec(10000, R_X, nb_margins) 
-  colnames(sim_nbinom) <- colnames(brca1000)
-  saveRDS(sim_nbinom, brca_sim_path)
-}
-sim_nbinom <- readRDS(brca_sim_path)
 ```
@@
 The data do not exhibit outlying points, but do possess the desired Spearman correlations, central tendencies, and discrete values."}
-set.seed(02242021)
+set.seed(2020-02-25)
 num_genes <- 3
@@
 
-GGally::ggpairs(
-  data = as.data.frame(sim_nbinom[, gene_sample]),
-  upper = list(continuous = GGally::wrap('cor', method = "spearman"))
-) + theme_bw()
+ggpairs(data = as.data.frame(sim_nbinom[, gene_sample]),
+        upper = list(continuous = wrap('cor', method = "spearman"))) + 
+  theme_bw()
 ```
@@
 Utilizing 15 CPU threads in a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, the simulation completed in just over 2 minutes.
-your comment -->
-
-```{r ch050-makeFigBRCA, cache=FALSE}
-R_S_hat <- bs$cor(as.matrix(sim_nbinom), bs$Spearman)
-sim_nbinom_fit <- apply(sim_nbinom, 2, nbinom_mom)
-corr_compare <- tibble(
-  `Target Correlation` = R_S[lower.tri(R_S, diag = FALSE)],
-  `Estimated Correlation` = R_S_hat[lower.tri(R_S_hat, diag = FALSE)]
-)
-target_compare <- tibble(
-  `Target Size` = sizes,
-  `Estimated Size` = sim_nbinom_fit[1,]
-)
-logprob_compare <- tibble(
-  `Target Log Probability` = log(probs),
-  `Estimated Log Probability` = log(sim_nbinom_fit[2,])
-)
-
-if(!file.exists("fig/ch050-figBRCA.png")) {
-  p1 <- ggplot(corr_compare, aes(`Target Correlation`, `Estimated Correlation`)) +
-    geom_point(alpha = 0.05) + 
-    geom_abline(slope = 1, col = "red", lty = "dashed") +
-    theme_bw() + coord_equal()
-
-  p2 <- ggplot(target_compare, aes(`Target Size`, `Estimated Size`)) +
-    geom_point(alpha = 0.85) + 
-    geom_abline(slope = 1, col = "red", lty = "dashed") +
-    theme_bw() + coord_equal()
-
-  p3 <- ggplot(logprob_compare, 
-               aes(`Target Log Probability`, `Estimated Log Probability`)) +
-    geom_point(alpha = 0.85) + 
-    geom_abline(slope = 1, col = "red", lty = "dashed") +
-    theme_bw() + coord_equal()
-
-  p <- p1 + p2 + p3
-
-  ggsave(filename = "fig/ch050-figBRCA.png",
-         plot = p,
-         device = "png",
-         width = 8,
-         height = 4,
-         units = "in")
-}
-```
+-->
 
@@
 
-```{r ch050-densityEvaluation, include=FALSE, cache=FALSE}
+```{r ch050-densityEvaluation, include=FALSE}
 d <- ncol(sim_nbinom)
@@
 
-```{r ch050-densityEvaluationECHO, echo=TRUE, cache=FALSE}
-d <- ncol(sim_nbinom)
-B <- nrow(sim_nbinom)
-y0 <- 1
-mean(apply(sim_nbinom, 1, function(Y) {
-  all(Y >= y0)
-}))
+```{r ch050-densityEvaluationECHO, echo=TRUE}
+<<ch050-densityEvaluation>>
+pHat
 ```
@@
 ## Evaluation of correlation estimation efficiency
-
-```{r ch010-myM, include=FALSE}
-m <- 10
-```
 
@@
 
-In this small example, we simulate $m=`r m`$ data sets with the number of simulated vectors matching the number of patients in the BRCA data set, $N=`r nrow(example_brca)`$.
-Since our simulation is much faster for the Pearson correlation type (see Figure \@ref(fig:ch040-gpuVScpuFig)), we only convert the Spearman correlation matrix once (and ensure it is positive definite).
+In this small example, we simulate $m=30$ data sets with the number of simulated vectors matching the number of patients in the BRCA data set, $N=`r nrow(example_brca)`$.
+Since our simulation is much faster for the Pearson correlation type (see Figure \@ref(fig:ch040-largeDfig)), we only convert the Spearman correlation matrix once (and ensure it is positive definite).
 At each iteration, we estimate the quadratic loss (residual sum of squared errors) from the specified ${\bf R}_{Spearman}$, producing a distribution of loss values.
 
-```{r ch050-quadlossECHO, echo=TRUE, eval=FALSE, cache=FALSE}
+```{r ch050-quadlossECHO, echo=TRUE, eval=FALSE}
 # Simulate random vectors equal to the sample size
@@
 
-```{r loss_examples, include=FALSE, eval=FALSE}
-sapply(simRho, function(R) norm(R - R_S))           # norm of diff
-sapply(simRho, function(R) norm(R) - norm(R_S))     # diff of norms
-sapply(simRho, function(R) sum((R - R_S)^2))        # RSS
-sapply(simRho, function(R) mean((R - R_S)^2))       # MSE
-sapply(simRho, function(R) sqrt(mean((R - R_S)^2))) # RMSE
-sapply(simRho, function(R) mean(abs(R - R_S)))      # MAE
-```
-
-```{r ch050-lossResults, cache=FALSE}
-frobenius_path <- paste0("results/brca_", d, "_frobenius.rds")
-if (!file.exists(frobenius_path)) {
-    N <- nrow(example_brca)
-    ## create m random vectors and estimate correlation
-    simRho <- replicate(n = m, expr = {
-        tmpSim <- bs$rvec(N, R_X, nb_margins)
-        bs$cor(tmpSim, bs$Spearman)
-    }, simplify = FALSE)
+```{r ch050-lossResults}
 frobenius_loss <- sapply(simRho, function(R) sum((R - R_S)^2))
-    saveRDS(frobenius_loss, frobenius_path)
-}
-frobenius_loss <- readRDS(frobenius_path)
 ```
@@
 
-```{r ch050-loss-summary, echo=TRUE, eval=TRUE, cache=FALSE}
+```{r ch050-loss-summary, echo=TRUE}
 summary(frobenius_loss)

### 2021-02-26 index.Rmd
Revise 040
renv+box fully implemented

@@
 
-```{r indexSetup, include=FALSE, cache=FALSE}
-set.seed(09212020)
-```
-
-```{r copyBib, include=FALSE, eval=FALSE}
-system('cp ~/bib/bigsimr.bib ./.')
-```
-
 ```{r include=FALSE}
@@
 ), 'packages.bib')
-```
-
-```{r serveHelper, include=FALSE, eval=FALSE}
-# rmote::start_rmote()
-# library(ggplot2); qplot(mpg, wt, data=mtcars, colour=cyl)
-# bookdown::render_book("index.Rmd", "bookdown::pdf_book", output_dir = "pdf")
-# bookdown::preview_chapter("index.Rmd")
-rmarkdown::render_site(encoding = 'UTF-8')
-bookdown::serve_book()
-```
-
-```{r countWords, include=FALSE, eval=FALSE}
-rmdFiles <- dir(pattern = 'Rmd$', ignore.case = TRUE)
-# rmdFiles <- rmdFiles[-grep('^_', rmdFiles)]
-rmdFiles <- rmdFiles[-grep('references', rmdFiles)]
-# rmdFiles <- rmdFiles[-grep('README', rmdFiles)]
-# rmdFiles <- rmdFiles[-grep('article', rmdFiles)] ## after generating a single document
-# wordcountaddin::word_count( rmdFiles[11] )
-(counts <- sapply( rmdFiles, wordcountaddin::word_count, simplify = T ))
-sum(counts)
 ```

### 2021-02-26 R/utils.R

@@ -1,10 +1,12 @@
 #' @export
 mom_norm <- function(x) {
+  box::use(stats[sd])
   c(mean = mean(x), sd = sd(x))
 }
 
 #' @export
 mom_nbinom <- function(x) {
+  box::use(stats[sd])
   m <- mean(x)
   s <- sd(x)
   c(size = m^2 / (s^2 - m), prob = m / s^2)
@@ -12,6 +14,7 @@ mom_nbinom <- function(x) {
 
 #' @export
 mom_gamma <- function(x) {
+  box::use(stats[sd])
   m <- mean(x)
   s <- sd(x)
   c(shape = m^2 / s^2, rate = m / s^2)

### 2021-02-26 _common.R

@@ -1,4 +1,3 @@
-set.seed(12)
 options(digits = 4)
 
 knitr::opts_chunk$set(
@@ -24,3 +23,6 @@ knitr::opts_chunk$set(
 )
 
 # options(dplyr.print_min = 6, dplyr.print_max = 6)
+
+seed_number <- 2020-02-25
+set.seed(seed = seed_number)

### 2021-02-26 data-raw/bivariate_gamma_sims.R

@@ -1,15 +1,13 @@
 ## code to prepare `bivariate_gamma_sims` dataset goes here
 
-library(bigsimr)
+box::use(
+  bigsimr[...],
+  ./R/utils[mom_gamma]
+)
+
 Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
 bs <- bigsimr_setup(pkg_check = FALSE)
 dist <- distributions_setup()
-
-mom_gamma <- function(x) {
-  m <- mean(x)
-  s <- sd(x)
-  list(shape = m^2 / s^2, rate = m / s^2)
-}
 
 shape <- 10
 rate <- 1
@@ -65,8 +63,8 @@ for (i in 1:nrow(sim_pars)) {
     rho_hat <- Rho_hat[1, 2]
 
     gamma_args_hat <- mom_gamma(x[,1])
-    shape_hat <- gamma_args_hat$shape
-    rate_hat <- gamma_args_hat$rate
+    shape_hat <- gamma_args_hat["shape"]
+    rate_hat  <- gamma_args_hat["rate"]
 
     ## Save the results
     res <- rbind(res, data.frame(
@@ -87,8 +85,7 @@ for (i in 1:nrow(sim_pars)) {
     ))
   }
 }
+
 res$type <- factor(res$type, levels = c("Pearson", "Spearman", "Kendall"))
-
 bivariate_gamma_sims <- res
-
-usethis::use_data(bivariate_gamma_sims, overwrite = TRUE)
+save(bivariate_gamma_sims, file = "data/bivariate_gamma_sims.rda")

### 2021-02-26 data-raw/bivariate_nbinom_sims.R

@@ -1,19 +1,16 @@
 ## code to prepare `bivariate_nbinom_sims` dataset goes here
 
-library(bigsimr)
+box::use(
+  bigsimr[...],
+  ./R/utils[mom_nbinom]
+)
+
 Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
 bs <- bigsimr_setup(pkg_check = FALSE)
 dist <- distributions_setup()
 
-mom_nbinom <- function(x) {
-  m <- mean(x)
-  s <- sd(x)
-  list(size = m^2 / (s^2 - m), prob = m / s^2)
-}
-
 size <- 4
 prob <- 3e-04
-
 margins <- c(dist$NegativeBinomial(size, prob),
              dist$NegativeBinomial(size, prob))
 
@@ -66,8 +63,8 @@ for (i in 1:nrow(sim_pars)) {
     rho_hat <- Rho_hat[1, 2]
 
     nbinom_args_hat <- mom_nbinom(x[,1])
-    size_hat <- nbinom_args_hat$size
-    prob_hat <- nbinom_args_hat$prob
+    size_hat <- nbinom_args_hat["size"]
+    prob_hat <- nbinom_args_hat["prob"]
 
     ## Save the results
     res <- rbind(res, data.frame(
@@ -88,8 +85,7 @@ for (i in 1:nrow(sim_pars)) {
     ))
   }
 }
+
 res$type <- factor(res$type, levels = c("Pearson", "Spearman", "Kendall"))
-
 bivariate_nbinom_sims <- res
-
-usethis::use_data(bivariate_nbinom_sims, overwrite = TRUE)
+save(bivariate_nbinom_sims, file = "data/bivariate_nbinom_sims.rda")

### 2021-02-26 data-raw/bivariate_normal_sims.R

@@ -1,24 +1,19 @@
 ## code to prepare `bivariate_normal_sims` dataset goes here
+
 box::use(
   bigsimr[...],
-  usethis[use_data],
-  ./R/method_of_moments[mom_norm]
+  ./R/utils[mom_norm]
 )
 
 Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
 bs <- bigsimr_setup(pkg_check = FALSE)
 dist <- distributions_setup()
 
-mom_norm <- function(x) {
-  m <- mean(x)
-  s <- sd(x)
-  list(mean = m, sd = s)
-}
-
 mu <- 0
 sigma <- 1
 margins <- c(dist$Normal(mu, sigma),
              dist$Normal(mu, sigma))
+
 types <- c("Pearson", "Spearman", "Kendall")
 n <- c(1e3, 1e4, 1e5)
 eps <- 1e-2
@@ -68,8 +63,8 @@ for (i in 1:nrow(sim_pars)) {
     rho_hat <- Rho_hat[1, 2]
 
     norm_args_hat <- mom_norm(x[, 1])
-    mu_hat <- norm_args_hat$mean
-    sigma_hat <- norm_args_hat$sd
+    mu_hat <- norm_args_hat["mean"]
+    sd_hat <- norm_args_hat["sd"]
 
     ## Save the results
     res <- rbind(res, data.frame(
@@ -85,13 +80,12 @@ for (i in 1:nrow(sim_pars)) {
       mean = mu,
       sd = sigma,
       mean_hat = mu_hat,
-      sd_hat = sigma_hat,
+      sd_hat = sd_hat,
       sim_time = unname(time_data["elapsed"])
     ))
   }
 }
+
 res$type <- factor(res$type, levels = c("Pearson", "Spearman", "Kendall"))
-
 bivariate_normal_sims <- res
-
-usethis::use_data(bivariate_normal_sims, overwrite = TRUE)
+save(bivariate_normal_sims, file = "data/bivariate_normal_sims.rda")

### 2021-02-26 data-raw/make_fig_brca.R

@@ -1,0 +1,62 @@
+box::use(
+  dplyr[...],
+  ggplot2[...],
+  patchwork[...],
+  bigsimr[...],
+  ./R/utils[mom_nbinom]
+)
+
+Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
+bs <- bigsimr_setup(pkg_check = FALSE)
+dist <- distributions_setup()
+
+sim_nbinom <- readRDS("results/brca_1000_sim.rds")
+load("data/example_brca.rda")
+
+brca1000 <- example_brca %>%
+  select(all_of(1:1000)) %>%
+  mutate(across(everything(), as.double))
+R_S <- bs$cor(as.matrix(brca1000), bs$Spearman)
+nbinom_fit <- apply(brca1000, 2, mom_nbinom)
+sizes <- nbinom_fit["size",]
+probs <- nbinom_fit["prob",]
+
+R_S_hat <- bs$cor(as.matrix(sim_nbinom), bs$Spearman)
+sim_nbinom_fit <- apply(sim_nbinom, 2, mom_nbinom)
+corr_compare <- tibble(
+  `Target Correlation` = R_S[lower.tri(R_S, diag = FALSE)],
+  `Estimated Correlation` = R_S_hat[lower.tri(R_S_hat, diag = FALSE)]
+)
+target_compare <- tibble(
+  `Target Size` = sizes,
+  `Estimated Size` = sim_nbinom_fit[1,]
+)
+logprob_compare <- tibble(
+  `Target Log Probability` = log(probs),
+  `Estimated Log Probability` = log(sim_nbinom_fit[2,])
+)
+
+p1 <- ggplot(corr_compare, aes(`Target Correlation`, `Estimated Correlation`)) +
+  geom_point(alpha = 0.05) +
+  geom_abline(slope = 1, col = "red", lty = "dashed") +
+  theme_bw() + coord_equal()
+
+p2 <- ggplot(target_compare, aes(`Target Size`, `Estimated Size`)) +
+  geom_point(alpha = 0.85) +
+  geom_abline(slope = 1, col = "red", lty = "dashed") +
+  theme_bw() + coord_equal()
+
+p3 <- ggplot(logprob_compare,
+             aes(`Target Log Probability`, `Estimated Log Probability`)) +
+  geom_point(alpha = 0.85) +
+  geom_abline(slope = 1, col = "red", lty = "dashed") +
+  theme_bw() + coord_equal()
+
[…]

### 2021-02-26 data-raw/sim_nbinom.R

@@ -1,0 +1,37 @@
+set.seed(2020-02-25)
+
+box::use(
+  dplyr[...],
+  bigsimr[...],
+  ./R/utils[mom_nbinom]
+)
+
+Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
+bs <- bigsimr_setup(pkg_check = FALSE)
+dist <- distributions_setup()
+
+make_nbinom_margins <- function(sizes, probs) {
+  margins <- lapply(1:length(sizes), function(i) {
+    dist$NegativeBinomial(sizes[i], probs[i])
+  })
+  do.call(c, margins)
+}
+
+load("data/example_brca.rda")
+
+brca1000 <- example_brca %>%
+  select(all_of(1:1000)) %>%
+  mutate(across(everything(), as.double))
+
+R_S <- bs$cor(as.matrix(brca1000), bs$Spearman)
+R_X <- bs$cor_convert(R_S, bs$Spearman, bs$Pearson)
+R_X <- bs$cor_nearPD(R_X)
+
+nbinom_fit <- apply(brca1000, 2, mom_nbinom)
+sizes <- nbinom_fit["size",]
+probs <- nbinom_fit["prob",]
+nb_margins <- make_nbinom_margins(sizes, probs)
+sim_nbinom <- bs$rvec(10000, R_X, nb_margins)
+colnames(sim_nbinom) <- colnames(brca1000)
+
+saveRDS(sim_nbinom, file = paste0("results/brca_", d, "_sim.rds"))

### 2021-02-26 data-raw/sim_nbinom_corr.R

@@ -1,0 +1,44 @@
+set.seed(2020-02-25)
+
+box::use(
+  dplyr[...],
+  bigsimr[...],
+  ./R/utils[mom_nbinom]
+)
+
+Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
+bs <- bigsimr_setup(pkg_check = FALSE)
+dist <- distributions_setup()
+
+make_nbinom_margins <- function(sizes, probs) {
+  margins <- lapply(1:length(sizes), function(i) {
+    dist$NegativeBinomial(sizes[i], probs[i])
+  })
+  do.call(c, margins)
+}
+
+load("data/example_brca.rda")
+
+brca1000 <- example_brca %>%
+  select(all_of(1:1000)) %>%
+  mutate(across(everything(), as.double))
+
+R_S <- bs$cor(as.matrix(brca1000), bs$Spearman)
+R_X <- bs$cor_convert(R_S, bs$Spearman, bs$Pearson)
+R_X <- bs$cor_nearPD(R_X)
+
+nbinom_fit <- apply(brca1000, 2, mom_nbinom)
+sizes <- nbinom_fit["size",]
+probs <- nbinom_fit["prob",]
+nb_margins <- make_nbinom_margins(sizes, probs)
+
+
+N <- nrow(example_brca)
+m <- 30
+## create m random vectors and estimate correlation
+simRho <- replicate(n = m, expr = {
+  tmpSim <- bs$rvec(N, R_X, nb_margins)
+  bs$cor(tmpSim, bs$Spearman)
+}, simplify = FALSE)
+
+saveRDS(simRho, file = "results/brca_1000_frobenius.rds")

### 2021-02-26 setup.R

@@ -1,0 +1,40 @@
+if (!file.exists("data/example_brca.rda")) {
+  e <- new.env()
+  source("data-raw/brca.R", local = e)
+  rm(e)
+}
+if (!file.exists("data/bivariate_normal_sims.rda")) {
+  e <- new.env()
+  source("data-raw/bivariate_normal_sims.R", local = e)
+  rm(e)
+}
+if (!file.exists("data/bivariate_gamma_sims.rda")) {
+  e <- new.env()
+  source("data-raw/bivariate_gamma_sims.R", local = e)
+  rm(e)
+}
+if (!file.exists("data/bivariate_nbinom_sims.rda")) {
+  e <- new.env()
+  source("data-raw/bivariate_nbinom_sims.R", local = e)
+  rm(e)
+}
+if (!file.exists("data/benchmark_dependences.rda")) {
+  e <- new.env()
+  source("data-raw/benchmark_dependences.R", local = e)
+  rm(e)
+}
+if (!file.exists("results/brca_1000_sim.rds")) {
+  e <- new.env()
+  source("data-raw/sim_nbinom.R", local = e)
+  rm(e)
+}
+if (!file.exists("fig/ch050-figBRCA.png")) {
+  e <- new.env()
+  source("data-raw/make_fig_brca.R", local = e)
+  rm(e)
+}
+if (!file.exists("results/brca_1000_frobenius.rds")) {
+  e <- new.env()
+  source("data-raw/sim_nbinom_corr.R", local = e)
+  rm(e)
+}

### 2021-02-27 020-simulation-algorithm.Rmd
renv+box fully implemented
removed notebooks

@@
 
-```{r ch020-preview, eval=FALSE}
-bookdown::preview_chapter('020-simulation-algorithm.Rmd')
-```
-
-This section describes our methods involved in simulating a random vector $\bf Y$ with $Y_i$ components for $i=1,2,\ldots,d$.
+This section describes our methods involved in simulating a random vector $\bf Y$ with $Y_i$ components for $i=1,\ldots,d$.
 Each $Y_i$ has a specified marginal CDF $F_i$ and its inverse $F^{-1}_i$.
@@
 4. Set $X$ by $X \gets MZ$.
-5. $\text{Return} \; Y \; \text{where} \; Y_i \gets F_{Y_i}^{-1}[\Phi(X_i)], \; i=1,...,d$.
+5. $\text{Return} \; Y \; \text{where} \; Y_i \gets F_i^{-1}[\Phi(X_i)], \; i=1,...,d$.
 
-With modern parallelized computing, steps 1, 3, 4, 5 are readily implemented as high-performance, multi-core and/or graphical-processing-unit (GPU) accelerated algorithms --- providing fast scalability using readily-available hardware.
+With modern parallel computing, steps 1, 3, 4, 5 are readily implemented as high-performance, multi-core and/or graphical-processing-unit (GPU) accelerated algorithms --- providing fast scalability using readily-available hardware.
 
@@
 
-There is hope, however, as @GH02 also showed that replicating an indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$.
+There is hope, however, as @GH02 also showed that replacing an indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$.
 This provides evidence that our nearest positive definite (PD) augmented approach will maintain reasonable accuracy if our input matching scheme returns an indefinite matrix, at least for the rank-based matching scheme described above.
@@
     (i) Convert the target Spearman correlation matrix $R_S$ to the corresponding MVN Pearson correlation $R_X$.
+
     or
+
     (i$^\prime$) Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$.
 2. Check admissibility and, if needed, compute the nearest correlation matrix.
-  (i) Check that $R_X$ is a correlation matrix, a positive [semi]definite matrix with 1's along the diagonal.   
+    (i) Check that $R_X$ is a correlation matrix, a positive definite matrix with 1's along the diagonal.   
 	(ii) If $R_X$ is a correlation matrix, the input matrix is *admissible* in this scheme.
 	Otherwise,
-  (ii$^\prime$) Replace $R_X$ with the nearest correlation matrix $R_X$, in the Frobenius norm.
+    (ii$^\prime$) Replace $R_X$ with the nearest correlation matrix $\tilde{R}_X$, in the Frobenius norm.
 3. NORTA step
@@
 
-$$
-\begin{aligned}
+\begin{align*}
 \Omega_P &= \{ R_P : R_P \textrm{ is a correlation matrix for } X \} \\
@@
 \Omega_S &= \{ R_S : R_S \textrm{ is a correlation matrix for } X \} \\
-\end{aligned}
-$$
+\end{align*}
 
@@
 We conjecture informally that the sets of admissible $R_S$ and $R_K$ are not highly restrictive.
-In particular, $R_P$ is approximately $R_S$ for a MVN, suggesting that the admissible set $\Omega_S$ should be flexible as $R_P$ can be any PSD matrix with 1's along the diagonal.
+In particular, $R_P$ is approximately $R_S$ for a MVN, suggesting that the admissible set $\Omega_S$ should be flexible as $R_P$ can be any PD matrix with 1's along the diagonal.
 We provide methods to check whether a target $R_S$ is an element of the *admissible set* $\Omega_S$ (and similarly for $R_K$) and also for checking that each pairwise correlation value is admissible (via the MC estimation of the 2-dimensional Frechet bounds).
@@
 
-The final step implements a high-performance NORTA algorithm. Specifically, step 3i uses a fast multi-core multivariate normal simulator, and steps 3ii, 3iii are parallelized over CPU cores.
+The final step implements a high-performance NORTA algorithm. Specifically, step 3i uses a fast multi-core multivariate normal simulator, and steps 3ii, 3iii are distributed over CPU cores.
 
@@
 
-*Correlation Conversion*
+*Pearson Matching*
+Given a desired target correlation $\rho_Y$ between a pair of marginal distributions, the correlation coefficient $\rho_Z$ used in the Gaussian copula must be specified so as to satisfy equation \@ref(eq:pearsonIntegralRelation).
 
-Given:
+\begin{equation}
+    \mathrm{E}\left[Y_i Y_j\right] = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} F_i^{-1}\left[\Phi(z_i)\right] F_j^{-1}\left[\Phi(z_j)\right] \phi(z_i, z_j, \rho_z) dz_i dz_j
+    (\#eq:pearsonIntegralRelation)
+\end{equation}
 
-* A target Spearman's $\rho$ or Kendall's $\tau$ correlation
-
-Do:
-
-* Determine the Pearson correlation for use in a Gaussian copula such that the resulting random samples using the NORTA algorithm have an estimated Spearman or Kendall correlation that matches the target correlation
-
-From Proposition 22 (Lebrun and Dutfow, 2009), we have the result for Gaussian copula that the copula correlation $r$ is related to Spearman's $\rho$ and Kendall's $\tau$ by the following relations:
-
-$$
-r = 2 \sin \left(\frac{\pi}{6} \rho\right) \\
-r = \sin \left(\frac{\pi}{2} \tau\right)
-$$
-
-* Lebrun, R., & Dutfoy, A. (2009). An innovating analysis of the Nataf transformation from the copula viewpoint. Probabilistic Engineering Mechanics, 24(3), 312-320.
-
-*Pearson Matching*
-
-Given:
-
-* A target Pearson correlation matrix, $R$
-* A list of marginal distributions, $F$
-
-Do:
-
-* Determine the input correlation, $\tilde{R}$, such that the resulting random samples using the NORTA algorithm have an estimated Pearson correlation that matches $R$
-
-This method relies on Equations 19, 39, and 49 (Xiao and Zhou, 2019).
-
-* Xiao, Q., & Zhou, S. (2019). Matching a correlation coefficient by a Gaussian copula. Communications in Statistics-Theory and Methods, 48(7), 1728-1747.
+where $\phi(\cdot)$ is the joint probability distribution function of two correlated standard normal variables.
+In general, the double integral in \@ref(eq:pearsonIntegralRelation) is difficult or impossible to solve analytically, so numerical methods are required.
+Our implementation is based off the work of @xiao2019matching where the marginal transformation is approximated by a weighted sum of Hermite polynomials.
 
@@
 
-$\begin{aligned}
+\begin{align*}
     \mathrm{min}\quad & \frac{1}{2} \Vert G - X \Vert^2 \\
@@
     & X \in S_{+}^{n}
-\end{aligned}$
+\end{align*}
 
 * Qi, H., & Sun, D. (2006). A quadratically convergent Newton method for computing the nearest correlation matrix. SIAM journal on matrix analysis and applications, 28(2), 360-385.
-
-<!-- 
-## Other high-performance `bigsimr` algorithms
-
-<mark> ALEX, briefly discuss other algorithms here </mark>
-- `cor_bounds()`  
-- `cor_fast()`  
-- `cor_nearPSD()`
-
-For discrete marginals, achieving a target Spearman correlation under this scheme is possible by using components from Equation \@ref(eq:spearmanRescaled) to further adjust the input correlation matrix.
-Let the unscaled Spearman correlation coefficients be $\rho_{s} \left(Y_{i}, Y_{i^\prime}\right)$ for two marginal distributions and divide the target correlation by the product in the denominator of Equation \@ref(eq:spearmanRescaled).
-Let these adjustment factors be denoted as $a_i = \left[ 1 - \sum_y p_i(y)^3 \right]^{1/2}$ and specifically rescale the target Spearman correlation matrix by
-
-\begin{equation}
-(\#eq:convertSpearmanDiscrete)
-\rho_{rs} \left(Y_{i}, Y_{i^\prime}\right) = \frac{\rho_{s} \left(Y_{i}, Y_{i^\prime}\right)}{a_i \times a_{i^\prime}}.
-\end{equation}
-
-In a similar fashion, we rescale Kendall's $\tau$ to adjust the input correlation matrix. The conversion formula is given by
-
-\begin{equation}
-(\#eq:convertKendallDiscrete)
-\rho_{rs} \left(Y_{i}, Y_{i^\prime}\right) = \frac{\rho_{s} \left(Y_{i}, Y_{i^\prime}\right)}{a_i \times a_{i^\prime}}.
-\end{equation}
-
-your comment -->

### 2021-02-27 index.Rmd
removed notebooks

@@
 site: bookdown::bookdown_site
-bibliography: ["bigsimr.bib", "packages.bib"]
+bibliography: ["bigsimr.bib", "packages.bib", "alex.bib"]
 abstract:  It is critical to realistically simulate data when employing Monte Carlo techinques and evaluating statistical methodology.

### 2021-02-27 data-raw/sim_nbinom.R

@@ -34,4 +34,4 @@ nb_margins <- make_nbinom_margins(sizes, probs)
 sim_nbinom <- bs$rvec(10000, R_X, nb_margins)
 colnames(sim_nbinom) <- colnames(brca1000)
 
-saveRDS(sim_nbinom, file = paste0("results/brca_", d, "_sim.rds"))
+saveRDS(sim_nbinom, file = "results/brca_1000_sim.rds")

### 2021-02-27 020-simulation-algorithm.Rmd
fixed italic bug

@@
 
-    (i$^\prime$) Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$.
+    (i') Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$.
 2. Check admissibility and, if needed, compute the nearest correlation matrix.
@@
 *Pearson Matching*
+
 Given a desired target correlation $\rho_Y$ between a pair of marginal distributions, the correlation coefficient $\rho_Z$ used in the Gaussian copula must be specified so as to satisfy equation \@ref(eq:pearsonIntegralRelation).
@@
 
-This algorithm is trying to find a correlation matrix $X$ close to $G$ in the convex optimization problem:
+Finding the nearest correlation matrix to an *almost* correlation matrix is a problem that has been researched since the 1990's, and improved upon in the 2000's by Nick Higham who presented an alternating projection algorithm (@higham2002computing) which is used by R (`Matrix::nearPD`), and then by Qi and Sun who developed a quadratically convergent algorithm based on the theory of strongly semismooth matrix functions (@qi2006quadratically).
+Our algorithm is transcription of the `R` code shared by Sun to the Julia language.
+The nearest correlation matrix problem can be written down as the following convex optimization problem:
 
@@
 \end{align*}
-
-* Qi, H., & Sun, D. (2006). A quadratically convergent Newton method for computing the nearest correlation matrix. SIAM journal on matrix analysis and applications, 28(2), 360-385.

### 2021-02-28 notebooks/bigsimr-julia-getting-started.Rmd
re-add notebooks so Grant can check out new branch without losing files

@@
 ## Either let JuliaCall find your installation of Julia or set manually
-bs <- bigsimr_setup(JULIA_HOME = "/usr/local/bin/")
+## bs <- bigsimr_setup(JULIA_HOME = "/usr/local/bin/")
+bs <- bigsimr_setup()
 dist <- distributions_setup() # Julia distributions

### 2021-02-28 notebooks/bivariateGamma.rmd
re-add notebooks so Grant can check out new branch without losing files

@@
 set.seed(06032020)
+## devtools::install_github( "adknudson/bigsimr", ref = 'develop')
 library(bigsimr)
-CORES <- parallel::detectCores() - 1
-## CORES <- 1
 ## devtools::install_github( "superdesolator/NORTARA" )
 library(nortaRA)
+## to parallize
+library(parallel)
+CORES <- parallel::detectCores() - 1
 ```
@@
 
-```{r, globalParams}
+```{r, globalParams, results = "hide"}
 ## bivariate gamma
@@
 simRes <- NULL
+
+## parallelize outside of bigsimr
+cl <- makeCluster(CORES)
+clusterEvalQ(cl, library("bigsimr"))
+```
+
+```{r runSim}
+runSim <- function( r, N, margins, cores, simMethod, corType, invcdfnames = NULL) {
+    d <- length(margins)
+    print( paste0( 'Working on true rho of ', r ) )
+    rho <- matrix(r, nrow = d, ncol = d)
+    diag(rho) <- 1.0
+    if (simMethod == 'bigsimr') {
+        simData  <- as.data.frame( rvec(n = N, rho = rho, params = margins, type = corType, cores = cores) )
+    }
+    if (simMethod == 'nortaRA') {
+        simData <- nortaRA::gennortaRA(n = N, cor_matrix = rho, invcdfnames, margins)
+    } 
+    tmpSimData <- data.frame( shape1 = margins[[1]]$shape, shape2 = margins[[2]]$shape, trueRho = r, estRho = fastCor(simData, method = type)[ 1, 2], method = simMethod , type = corType)
+    return(tmpSimData)
+}
+## runSim( r = 0.5, N = 1e2, margins = margins, cores = 1, simMethod = 'bigsimr', corType = 'pearson')
 ```
@@
 
-```{r, boundsKendall, eval=FALSE}
+```{r, boundsKendall, eval=TRUE}
 type <- 'kendall'
@@
 ```{r bigsimrPearson, eval = TRUE, results = 'hide'}
+simMethod <- 'bigsimr'
 type <- 'pearson'
+
 ## now show that our simulation methods works
@@
 tmpUpper <- boundsDat$upper[ boundsDat$type == type ]
-rVec <- seq( tmpLower, tmpUpper, length.out = gridSteps)  ## loose approximation for now
+## rVec <- c(-0.4, 0.1, 0.5)
+## rVec <- seq( tmpLower, tmpUpper, length.out = gridSteps)
+## loose on the boundary for now
+epsilon <- 1e-2
+rVec <- seq( tmpLower+epsilon, tmpUpper-epsilon, length.out = gridSteps )
 
-## loop for simplicity see if fast enough
-for (r in rVec) {
-    print( paste0( 'Working on true rho of ', r ) )
-    rho <- matrix(r, nrow = d, ncol = d)
-    diag(rho) <- 1.0
-    simData  <- as.data.frame( rvec(n = N, rho = rho, params = margins, type = type, cores = CORES, adjustForDiscrete = FALSE) )
-    tmpSimData <- data.frame( shape1 = shape1, shape2 = shape2, trueRho = r, estRho = cor(simData, method = type)[ 1, 2], method = 'bigsimr' , type = type)
-    simRes <- rbind( simRes, tmpSimData )
-}
+clusterExport(cl, varlist = c( 'simMethod', 'type') )
+tmpSimRes <- as.data.frame( do.call( "rbind", parSapply( cl = cl, rVec, FUN = runSim, N = N, margins = margins, cores = 1, simMethod = simMethod, corType = type, simplify = FALSE) ) )
+simRes <- rbind( simRes, tmpSimRes )
+
 ```
@@
 type <- 'spearman'
+simMethod <- 'bigsimr'
+
 ## now show that our simulation methods works
@@
 
-```{r bigsimrKendall, eval = FALSE}
-type <- 'kendall'
+```{r bigsimrKendall, eval = TRUE}
+type <-'kendall'
+simMethod <- 'bigsimr'
+
 ## now show that our simulation methods works
@@
 tmpUpper <- boundsDat$upper[ boundsDat$type == type ]
-rVec <- seq( tmpLower, tmpUpper, length.out = gridSteps )
+## rVec <- c(-0.4, 0.1, 0.5)
+## rVec <- seq( tmpLower, tmpUpper, length.out = gridSteps)
+## loose on the boundary for now
+epsilon <- 1e-2
+rVec <- seq( tmpLower+epsilon, tmpUpper-epsilon, length.out = gridSteps )
 
-## loop for simplicity see if fast enough
-for (r in rVec) {
-    print( paste0( 'Working on true rho of ', r ) )
-    rho <- matrix(r, nrow = d, ncol = d)
-    diag(rho) <- 1.0
-    simData  <- as.data.frame( rvec(n = N, rho = rho, params = margins, type = type, cores = CORES, adjustForDiscrete = FALSE) )
-    tmpSimData <- data.frame( shape1 = shape1, shape2 = shape2, trueRho = r, estRho = cor(simData, method = type)[ 1, 2], method = 'bigsimr' , type = type)
-    simRes <- rbind( simRes, tmpSimData )
-}
+clusterExport(cl, varlist = c( 'simMethod', 'type') )
+tmpSimRes <- as.data.frame( do.call( "rbind", parSapply( cl = cl, rVec, FUN = runSim, N = N, margins = margins, cores = 1, simMethod = simMethod, corType = type, simplify = FALSE) ) )
+simRes <- rbind( simRes, tmpSimRes )
 ```
 
-## `NORTARA` simulations
+## `nortaRA` simulations
 
-```{r nortaraSetup}
+```{r nortaRAsetup, eval = TRUE}
 ## slightly different input format than bigsimr
@@
 
-```{r nortaraPearson, eval = TRUE, warning=FALSE, results='hide'}
+```{r nortaRApearson, eval = TRUE, warning=FALSE, results='hide'}
 type <- 'pearson'
+simMethod <- 'nortaRA'
+
 ## now show that our simulation methods works
@@
 tmpUpper <- boundsDat$upper[ boundsDat$type == type ]
-esp <- 1e-3
-rVec <- seq( tmpLower+esp, tmpUpper-esp, length.out = gridSteps ) 
+## rVec <- c(-0.4, 0.1, 0.5)
+## rVec <- seq( tmpLower, tmpUpper, length.out = gridSteps)
+## loose on the boundary for now
+epsilon <- 1e-2
+rVec <- seq( tmpLower+epsilon, tmpUpper-epsilon, length.out = gridSteps )
+
 ## drop N for this one
 tmpN <- 1e4
-## loop for simplicity see if fast enough
-for (r in rVec) {
-    print( paste0( 'Working on true rho of ', r ) )
-    rho <- matrix(r, nrow = d, ncol = d)
-    diag(rho) <- 1.0
-    simData  <- nortaRA::gennortaRA(n = tmpN, cor_matrix = rho, invcdfnames, paramslists)
-    tmpSimData <- data.frame( shape1 = shape1, shape2 = shape2, trueRho = r, estRho = cor(simData, method = type)[ 1, 2], method = 'nortara' , type = type)
-    simRes <- rbind( simRes, tmpSimData )
-}
+
+## runSim( r = 0.4, N = tmpN, margins = paramslists, cores = 1, simMethod = simMethod, corType = type, invcdfnames = invcdfnames)
+clusterExport(cl, varlist = c( 'simMethod', 'type') )
+tmpSimRes <- as.data.frame( do.call( "rbind", parSapply( cl = cl, rVec, FUN = runSim, N = N, margins = paramslists, cores = 1, simMethod = simMethod, corType = type, invcdfnames = invcdfnames, simplify = FALSE) ) )
+simRes <- rbind( simRes, tmpSimRes )
 ```
@@
 
-```{r empVspecPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.path='fig/plot-', out.width= '75%', fig.align='center', fig.ext='pdf', fig.cap = 'Plot of the specified (true) versus empirical (computed) correlations for bivariate lognormal data with $p_1 = p_2 = 0.2$ generated by the standard approach, bigsimr, and T-Poisson.
+```{r empVspecPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.path='fig/plot-', out.width= '75%', fig.align='center', fig.ext='pdf', fig.cap = 'Plot of the specified versus empirical correlations for standard bivariate gamma data with shape1 = shape2 = 5 generated by various approaches.
 Empirical correlations are the estimated correlations from the simulated samples.'}
@@
 ```
+
+```{r cleanUP, echo = FALSE}
+## When you're done, clean up the cluster
+## stopCluster(cl)
+```

### 2021-02-28 notebooks/cmstats2019-Schissler.Rmd
re-add notebooks so Grant can check out new branch without losing files

---
title: "Simulating High-Dimensional Multivariate Data"
author: "A. Grant Schissler and Alex Knudson"
institute: "Department of Mathematics and Statistics University of Nevada, Reno"
bibliography: /Users/alfred/Dropbox/bib/MulNB.bib
theme: shadow
output: beamer_presentation
aspectratio: 43
---

```{r setup, include=FALSE}
library(ggplot2)
knitr::opts_chunk$set(cache=TRUE, autodep=TRUE, cache.comments=TRUE)
set.seed(44)
```

# 1.1 – Overview

- In this era of Big Data, it is critical to realistically simulate data to conduct informative Monte Carlo studies. 
- This is often problematic when data are inherently multivariate while at the same time are (ultra-) high dimensional.
- This situation appears frequently in observational data found on online and in high-throughput biomedical experiments (e.g., RNA-sequencing).
- Help researchers think/simulate their data *generatively* from simple joint probability models rather than assuming their statistical model, etc.

# 1.2 - Desired Properties

We seek a general-purpose algorithm that possesses the following properities [adapted from @Nik13a]:

* P1: Wide range of dependence, allowing both positive and negative dependence
* P2: Flexible dependence, meaning that the number of bivariate marginals is (approximately) equal to the number of dependence parameters.
* P3: Flexible marginal modeling, generating hetereogenous data --- possibly from differing (discrete) probability families.

Moreover, in simulation method must scale to high dimensions:

* S1: Procedure must scale to high dimensions, computable in a reasonable amount time.
* S2: Procedure must scale to high dimensions while maintaining accuracy.

# 2 - Motivating example: large-scale hypothesis testing (Efron 2007) of RNA-seq data

\begin{columns}[T]
  \begin{column}{0.5\columnwidth}
	\begin{itemize}
\item RNA-sequencing data derived from breast cancer patients (n=1093), with 20501 genes measured (from TCGA).
\item We filter to genes with average expression greater than 10000 counts (d = 4777) across both groups.
	\item Consider basic differentially expressed genes (DEG) analysis between surviving (n0 = 941) and deceased (n1 = 152) patients.
\end{itemize}

  \end{column}
    \begin{column}{0.5\columnwidth}
	\begin{itemize}
\item This setting can be described as large-scale simultaneous hypothesis testing, under correlation.
\item We should evaluate existing and new methodology using simulations that reflect dependency among variables.
\item But multivariate simulation tools often scale poorly to high dimension or do not model the full range of dependency.

\end{itemize}
	\end{column}
	\end{columns}

```{r load, echo=FALSE, eval=TRUE}
load("~/OneDrive - University of Nevada, Reno/Research/Mul_NB/brca/brca_rnaseq.RData")
## str(brca_rnaseq)
## params <- readRDS("~/OneDrive - University of Nevada, Reno/Research/Mul_NB/brca/brca_nb_params_all.RDS")
## round to remove RSEM adjustment
brca_rnaseq <- round(brca_rnaseq)
## brca_rnaseq[1:5, 1:5]
## get genes from simulated data
sim_kendall_data0 <- readRDS("~/OneDrive - University of Nevada, Reno/Research/Mul_NB/brca/results/sim_data_brca_kendall_MulNB_threshold=1000_d=4777_N=941_vital=0.rds")
sim_kendall_data1 <- readRDS("~/OneDrive - University of Nevada, Reno/Research/Mul_NB/brca/results/sim_data_brca_kendall_MulNB_threshold=1000_d=4777_N=152_vital=1.rds")
## str(sim_data)
d <- ncol(sim_kendall_data1)
## Use sim data for 
tmp_genes <- colnames(sim_kendall_data1)
## head(tmp_genes)
## j = 4777
## tmp_genes[j]
## summary(brca_rnaseq[,tmp_genes[j]])

#### Get cinical data
## brca_all <- TCGA2STAT::getTCGA(disease = "BRCA", clinical = TRUE)
## str(brca_all)
## brca_clinical <- brca_all$clinical[row.names(brca_rnaseq),]
## saveRDS(brca_clinical, "~/OneDrive - University of Nevada, Reno/Research/Mul_NB/brca/brca_clinical.rds")
brca_clinical <- readRDS("~/OneDrive - University of Nevada, Reno/Research/Mul_NB/brca/brca_clinical.rds")
## colnames(brca_clinical)

## Create two groups via a clinical variable

## try vital status
group <- as.numeric(brca_clinical[ , "vitalstatus"])
n1 <- sum(group) ## 152 deaths
n0 <- sum(1 - group) ## 941 survived

## susbest to the genes of interest
tmp_rnaseq <- brca_rnaseq[ ,colnames(brca_rnaseq) %in% tmp_genes]

## str(tmp_rnaseq)

## estimate the correlation
## reps <- 1e3
## pairs_mat <- combn(x = d, m = 2)
## index <- sample(ncol(pairs_mat), size = reps)
## tmp_index <- pairs_mat[,j]
## 
## real_cor_values0 <- apply(pairs_mat[ ,index], 2, function(tmp_index){
##     cor(tmp_rnaseq[ group == 0, tmp_index[1]], tmp_rnaseq[group == 0 ,tmp_index[2]])
## })
## summary(real_cor_values0)
## 
## real_cor_values1 <- apply(pairs_mat[ ,index], 2, function(tmp_index){
##     cor(tmp_rnaseq[ group == 1, tmp_index[1]], tmp_rnaseq[group == 1 ,tmp_index[2]])
## })
## summary(real_cor_values1)
## 
## on average the correlation is weak --- may need a more striking example data set

```

```{r empirical, echo = FALSE, eval=TRUE}
## 1. Empirical case
## j = 1
empirical_z_values <- sapply(1:d, function(j) {
    tmp_t <- t.test(alternative = "two.sided", x = tmp_rnaseq[group == 0, j], y = tmp_rnaseq[group == 1, j])
    ## close to z already for large samples
    tmp_z <- qnorm( p = pt(q = tmp_t$statistic, df = tmp_t$parameter) )
    return(unname(tmp_z))
})
## summary(empirical_z_values)
## sd(empirical_z_values)
## qqnorm(empirical_z_values); abline(0,1)

## compute p-values
empirical_p_values <- sapply(1:d, function(j) {
    tmp_t <- t.test(alternative = "two.sided", x = tmp_rnaseq[group == 0, j], y = tmp_rnaseq[group == 1, j])
    ## close to z already for large samples
    tmp_p <- tmp_t$p.value
    return(unname(tmp_p))
})
lvl <- 0.05
## summary(empirical_p_values)
## sum(empirical_p_values <= lvl)
## sum(empirical_p_values <= lvl)/ d
## Holm's adjustment
fdr_lvl <- holm_lvl <- 0.1
empirical_holm_values <- p.adjust(p = empirical_p_values, method = "holm")
## summary(empirical_holm_values)
## sum(empirical_holm_values <= fdr_lvl)
## sum(empirical_holm_values <= fdr_lvl)/ d
## fdr
empirical_fdr_values <- p.adjust(p = empirical_p_values, method = "fdr")
## fdr_lvl = 0.1
## summary(empirical_fdr_values)
## sum(empirical_fdr_values <= fdr_lvl)
## sum(empirical_fdr_values <= fdr_lvl)/ d

## ALL genes were null then this should be N(0,1) (the theoretical null)

```{r independent, echo = FALSE, eval=TRUE}

#### Simulations

## 2. Independence case

## this is quick
params0 <- readRDS("~/OneDrive - University of Nevada, Reno/Research/Mul_NB/brca/brca_nb_params_vital=0.RDS")
params0 <- params0[ ,colnames(params0) %in% tmp_genes]
params1 <- readRDS("~/OneDrive - University of Nevada, Reno/Research/Mul_NB/brca/brca_nb_params_vital=1.RDS")
params1 <- params1[ ,colnames(params1) %in% tmp_genes]

sim_ind0 <- sapply(1:d, function(j) {
    res <- rnbinom(n = n0, size = params0["alpha",j], prob = (1 + params0["lambda",j])^(-1))
    if (any(is.nan(res))) print(j)
    return(res)
})
## any(is.na(sim_ind0))

## j = 4777
sim_ind1 <- sapply(1:d, function(j) {
    res <- rnbinom(n = n1, size = params1["alpha",j], prob = (1 + params1["lambda",j])^(-1))
    if (any(is.nan(res))) print(j)
    return(res)
})

sim_ind_z_values <- sapply(1:d, function(j) {
    tmp_t <- t.test(alternative = "two.sided", x = sim_ind0[, j], y = sim_ind1[, j])
    ## close to z already for large samples
    tmp_z <- qnorm( p = pt(q = tmp_t$statistic, df = tmp_t$parameter) )
    return(unname(tmp_z))
})
## summary(sim_ind_z_values)
## sd(sim_ind_z_values)
## qqnorm(sim_ind_z_values); abline(0,1)

## compute p-values
sim_ind_p_values <- sapply(1:d, function(j) {
    tmp_t <- t.test(alternative = "two.sided", x = sim_ind0[, j], y = sim_ind1[, j])
    ## close to z already for large samples
    tmp_p <- tmp_t$p.value
    return(unname(tmp_p))
})
lvl <- 0.05
## summary(sim_ind_p_values)
## sum(sim_ind_p_values <= lvl)
## sum(sim_ind_p_values <= lvl)/ d
## Holm's adjustment
holm_lvl <- 0.1
sim_ind_holm_values <- p.adjust(p = sim_ind_p_values, method = "holm")
## summary(sim_ind_holm_values)
## sum(sim_ind_holm_values <= fdr_lvl)
## sum(sim_ind_holm_values <= fdr_lvl)/ d
## fdr
sim_ind_fdr_values <- p.adjust(p = sim_ind_p_values, method = "fdr")
## fdr_lvl = 0.1
## summary(sim_ind_fdr_values)
## sum(sim_ind_fdr_values <= fdr_lvl)
## sum(sim_ind_fdr_values <= fdr_lvl)/ d

```

```{r kendall, echo = FALSE, eval=TRUE}
## 3. Simulated multivariate (show it better mimics the empirical data)

## j = 1
sim_kendall_z_values <- sapply(1:d, function(j) {
    tmp_t <- t.test(alternative = "two.sided", x = sim_kendall_data0[, j], y = sim_kendall_data1[, j])
    ## close to z already for large samples
    tmp_z <- qnorm( p = pt(q = tmp_t$statistic, df = tmp_t$parameter) )
    return(unname(tmp_z))
})
## summary(sim_kendall_z_values)
## sd(sim_kendall_z_values)
## qqnorm(sim_kendall_z_values); abline(0,1)

## compute p-values
sim_kendall_p_values <- sapply(1:d, function(j) {
    tmp_t <- t.test(alternative = "two.sided", x = sim_kendall_data0[, j], y = sim_kendall_data1[, j])
    ## close to z already for large samples
    tmp_p <- tmp_t$p.value
    return(unname(tmp_p))
})

lvl <- 0.05
## summary(sim_kendall_p_values)
## sum(sim_kendall_p_values <= lvl)
## sum(sim_kendall_p_values <= lvl)/ d
## ## Holm's adjustment
holm_lvl <- 0.1
sim_kendall_holm_values <- p.adjust(p = sim_kendall_p_values, method = "holm")
## summary(sim_kendall_holm_values)
## sum(sim_kendall_holm_values <= fdr_lvl)
## sum(sim_kendall_holm_values <= fdr_lvl)/ d
## fdr
sim_kendall_fdr_values <- p.adjust(p = sim_kendall_p_values, method = "fdr")
## fdr_lvl = 0.1
## summary(sim_kendall_fdr_values)
## sum(sim_kendall_fdr_values <= fdr_lvl)
## sum(sim_kendall_fdr_values <= fdr_lvl)/ d

```

```{r exactR, echo = FALSE, eval=TRUE}
## 4. Exact Pearson matching via NORTARA.

```

```{r condreg, echo = FALSE, eval=TRUE, include=FALSE}
## Using unadjusted input
sim_condreg_data0 <- readRDS("~/OneDrive - University of Nevada, Reno/Research/Mul_NB/brca/results/sim_data_brca_condreg_direct_threshold=1000_d=4777_N=941_vital=0.rds")
sim_condreg_data1 <- readRDS("~/OneDrive - University of Nevada, Reno/Research/Mul_NB/brca/results/sim_data_brca_condreg_direct_threshold=1000_d=4777_N=152_vital=1.rds")

## j = 1
sim_condreg_z_values <- sapply(1:d, function(j) {
    tmp_t <- t.test(alternative = "two.sided", x = sim_condreg_data0[, j], y = sim_condreg_data1[, j])
    ## close to z already for large samples
    tmp_z <- qnorm( p = pt(q = tmp_t$statistic, df = tmp_t$parameter) )
    return(unname(tmp_z))
})
## summary(sim_condreg_z_values)
## sd(sim_condreg_z_values)
## qqnorm(sim_condreg_z_values); abline(0,1)

## compute p-values
sim_condreg_p_values <- sapply(1:d, function(j) {
    tmp_t <- t.test(alternative = "two.sided", x = sim_condreg_data0[, j], y = sim_condreg_data1[, j])
    ## close to z already for large samples
    tmp_p <- tmp_t$p.value
    return(unname(tmp_p))
})

lvl <- 0.05
## summary(sim_condreg_p_values)
## sum(sim_condreg_p_values <= lvl)
## sum(sim_condreg_p_values <= lvl)/ d
## ## Holm's adjustment
holm_lvl <- 0.1
sim_condreg_holm_values <- p.adjust(p = sim_condreg_p_values, method = "holm")
## summary(sim_condreg_holm_values)
## sum(sim_condreg_holm_values <= fdr_lvl)
## sum(sim_condreg_holm_values <= fdr_lvl)/ d
## fdr
sim_condreg_fdr_values <- p.adjust(p = sim_condreg_p_values, method = "fdr")
## fdr_lvl = 0.1
## summary(sim_condreg_fdr_values)
## sum(sim_condreg_fdr_values <= fdr_lvl)
## sum(sim_condreg_fdr_values <= fdr_lvl)/ d

```

```{r approxR, echo = FALSE, eval=TRUE}
## Using unadjusted input
sim_approxR_data0 <- readRDS("~/OneDrive - University of Nevada, Reno/Research/Mul_NB/brca/results/sim_data_brca_pearson_threshold=1000_d=4777_N=941_vital=0.rds")
sim_approxR_data1 <- readRDS("~/OneDrive - University of Nevada, Reno/Research/Mul_NB/brca/results/sim_data_brca_pearson_threshold=1000_d=4777_N=152_vital=1.rds")

## j = 1
sim_approxR_z_values <- sapply(1:d, function(j) {
    tmp_t <- t.test(alternative = "two.sided", x = sim_approxR_data0[, j], y = sim_approxR_data1[, j])
    ## close to z already for large samples
    tmp_z <- qnorm( p = pt(q = tmp_t$statistic, df = tmp_t$parameter) )
    return(unname(tmp_z))
})
## summary(sim_approxR_z_values)
## sd(sim_approxR_z_values)
## qqnorm(sim_approxR_z_values); abline(0,1)

## compute p-values
sim_approxR_p_values <- sapply(1:d, function(j) {
    tmp_t <- t.test(alternative = "two.sided", x = sim_approxR_data0[, j], y = sim_approxR_data1[, j])
    ## close to z already for large samples
    tmp_p <- tmp_t$p.value
    return(unname(tmp_p))
})

lvl <- 0.05
## summary(sim_approxR_p_values)
## sum(sim_approxR_p_values <= lvl)
## sum(sim_approxR_p_values <= lvl)/ d
## ## Holm's adjustment
holm_lvl <- 0.1
sim_approxR_holm_values <- p.adjust(p = sim_approxR_p_values, method = "holm")
## summary(sim_approxR_holm_values)
## sum(sim_approxR_holm_values <= fdr_lvl)
## sum(sim_approxR_holm_values <= fdr_lvl)/ d
## fdr
sim_approxR_fdr_values <- p.adjust(p = sim_approxR_p_values, method = "fdr")
## fdr_lvl = 0.1
## summary(sim_approxR_fdr_values)
## sum(sim_approxR_fdr_values <= fdr_lvl)
## sum(sim_approxR_fdr_values <= fdr_lvl)/ d

```

```{r spearman, echo = FALSE, eval=TRUE}
sim_spearman_data0 <- readRDS("~/OneDrive - University of Nevada, Reno/Research/Mul_NB/brca/results/sim_data_brca_spearman_direct_threshold=1000_d=4777_N=941_vital=0.rds")
sim_spearman_data1 <- readRDS("~/OneDrive - University of Nevada, Reno/Research/Mul_NB/brca/results/sim_data_brca_spearman_direct_threshold=1000_d=4777_N=152_vital=1.rds")

## j = 1
sim_spearman_z_values <- sapply(1:d, function(j) {
    tmp_t <- t.test(alternative = "two.sided", x = sim_spearman_data0[, j], y = sim_spearman_data1[, j])
    ## close to z already for large samples
    tmp_z <- qnorm( p = pt(q = tmp_t$statistic, df = tmp_t$parameter) )
    return(unname(tmp_z))
})
## summary(sim_spearman_z_values)
## sd(sim_spearman_z_values)
## qqnorm(sim_spearman_z_values); abline(0,1)

## compute p-values
sim_spearman_p_values <- sapply(1:d, function(j) {
    tmp_t <- t.test(alternative = "two.sided", x = sim_spearman_data0[, j], y = sim_spearman_data1[, j])
    ## close to z already for large samples
    tmp_p <- tmp_t$p.value
    return(unname(tmp_p))
})

lvl <- 0.05
## summary(sim_spearman_p_values)
## sum(sim_spearman_p_values <= lvl)
## sum(sim_spearman_p_values <= lvl)/ d
## ## Holm's adjustment
holm_lvl <- 0.1
sim_spearman_holm_values <- p.adjust(p = sim_spearman_p_values, method = "holm")
## summary(sim_spearman_holm_values)
## sum(sim_spearman_holm_values <= fdr_lvl)
## sum(sim_spearman_holm_values <= fdr_lvl)/ d
## fdr
sim_spearman_fdr_values <- p.adjust(p = sim_spearman_p_values, method = "fdr")
## fdr_lvl = 0.1
## summary(sim_spearman_fdr_values)
## sum(sim_spearman_fdr_values <= fdr_lvl)
## sum(sim_spearman_fdr_values <= fdr_lvl)/ d

```

```{r combine, echo = FALSE, eval=TRUE}
## empirical is efancy for empirical data
## method_names <- c("empirical", "independent", "approxR", "condreg", "Spearman", "Kendall")
method_names <- c("Empirical", "Independent", "ApproxR", "Spearman", "Kendall")
num_methods <- length(method_names)
all_dat <- data.frame(dat_source = rep( method_names, each = d),
                      index = rep(1:d, times = num_methods))

## aggregate z scores (order matters!)
all_dat$z_score <- c( sort(empirical_z_values, decreasing = T),
                     sort(sim_ind_z_values, decreasing = T),
                     sort(sim_approxR_z_values, decreasing = T),
                     ## sort(sim_condreg_z_values, decreasing = T),
                     sort(sim_spearman_z_values, decreasing = T),
                     sort(sim_kendall_z_values, decreasing = T) )

## aggregate p values (order matters
all_dat$p_value <- c( sort(empirical_p_values),
                     sort(sim_ind_p_values),
                     sort(sim_approxR_p_values),
                     ## sort(sim_condreg_p_values),
                     sort(sim_spearman_p_values),
                     sort(sim_kendall_p_values) )
## all_dat$holm_value <- c(empirical_holm_values, sim_ind_holm_values, sim_mul_holm_values)
## all_dat$fdr_value <- c(empirical_fdr_values, sim_ind_fdr_values, sim_mul_fdr_values)
## str(all_dat)

## add labels
## method_names <- c("empirical", "independent", "approxR", "condreg", "Spearman", "Kendall")
## method_labs <- c("Kendall Copula", "Empirical", "Independent", "Null ref: N(0,1) ")
## factor(as.character(all_dat$dat_source), labels = )

```

# Correspondence between Pearson and rank correlation

For bivariate Normal r.v's there is a closed form relationship between Pearson's correlation and Spearman's / Kendall's correlation:

\begin{equation}
\label{convertSpearman}
\rho_{Pearson} = 2 \times sin \left( \rho_{Spearman} \times \frac{\pi}{6} \right).
\end{equation}

\begin{equation}
\label{convertKendall}
r_{Pearson} = sin \left( \tau_{Kendall} \times \frac{\pi}{2} \right), 
\end{equation}

# 3.1 - Methods: Scalable copula-based methods for simulating RNA-seq data as multivariate negative binomial

\begin{columns}[T]
  \begin{column}{0.5\columnwidth}
 \includegraphics[width=0.5\paperwidth]{../figures/title_fig.pdf} 
  \end{column}
    \begin{column}{0.5\columnwidth}
	\begin{itemize}
	\item Model genes as marginally negative binomial with heterogenous genewise parameters
\item Compute sample correlations (using desired dependency measure) for the 11,407,476 genes pairs, for each group.
\item We implemented a {\bf parallelized Gaussian copula}, with that produces the correct margins while matching the target correlations.
\end{itemize}
	\end{column}
	\end{columns}

# A remark about discrete marginals

It's known [@MB13] that when the marginals are discrete it constraints the possible correlations for Spearman's correlation. One can rescale to allow the full range of dependency. Call these adjustment factors $a_i = \left[ 1 - \sum_y p_i(y)^3 \right]^{1/2}$ for marginal $F_i$.

\begin{equation}
\label{convertSpearmanDiscrete}
\rho_{rs} \left(Y_{i}, Y_{i^\prime}\right) = \frac{\rho_{s} \left(Y_{i}, Y_{i^\prime}\right)}{a_i \times a_{i^\prime}}.
\end{equation}

This is difficult to compute for discrete marginals with infinite support (such as negative binomial in our example). So one must approximate with a finite sum. The error is known and controlled if done carefully. More details are in the paper.

# Algorithm input
(1) Marginal characteristics including the same of distributional family and parameter values, denotes as $F_i$ for marginal component random variable $Y_i$ for $i=1,\ldots,d$.
(2) A target (specified) Spearman's correlation matrix ${\bf R_{Spearman}}$ with each $\rho_{Spearman}$ within the Frechet limits. Alternatively, the rescaled Spearman's correlation can be provided.
(3) An error tolerance $\epsilon$ for finding the nearest positive definite matrix a transformed correlation matrix (see step iii below). Alternatively, a maximum number of iterations can be supplied.

# Algorithm steps and return value
(i) Compute Spearman's $\rho_{rs}$ from the specified $\rho_{spearman}$ for each pair of marginal random variables $(Y_i,Y_{i^\prime})$ with $i \neq i^\prime$ when either marginal is discrete. 
(ii) Convert ${\bf R_{Spearman}}$ into ${\bf R_{Pearson}}$ via $\rho_{Pearson} = 2 \times sin \left( \rho_{rs} \times \frac{\pi}{6} \right)$.   
(iii) Finally, ensure that ${\bf R_{Pearson}}$ is positive definite, by finding the nearest positive definite correlation matrix ${\bf R}$ --- using the `R` routine `nearPD` in the `Matrix` package --- within an error tolerance of $\epsilon$.
(iv) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R})$;  
(v) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$;  
(vi) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;

# Key proposition

The simulation method produces random vectors $(Y_1, \ldots, Y_d)$ with marginal distribution functions $F_1,\ldots,F_d$ matching a given Spearman's/Kendall's correlation matrix for random variables.

- Appeal to Sklar's Theorem (1959) that any joint distribution has a copula representaton and that conversion steps are continuous functions.
- Approximation may be required when using the `Matrix::nearPD` routine (or simply replaced negative eigenvalues with small positive valuse). 
- The montone transformations in steps (v) and (vi) guarantee that the rank-based correlations are preserved for continuous marginals. 
- For discrete marginals with infinite support, the rescaling of Spearman's $\rho$ introduces approxixmation error.

# Implemtation `bigsimr`

We'll release the `bigsimr` `R` package to simulate random vectors. A few notes:  
- The most computationally extensive step lies in the second step of our algorithm.  
- We employ a optimized/parallelized multivariate normal simulator within the `R` package `mvnfast`.  
- The rest of the transformation are also parallelized, as the calculations can be done independently and results aggregated.

```{r repData, echo = FALSE, eval=TRUE}
sim_ttests <- readRDS("~/OneDrive - University of Nevada, Reno/Research/Mul_NB/results/sim_ttests_brca_spearman_rescaled_rvec_vital_d4777_rep1000_14nov2019.rds")
## sim_ttests <- readRDS("~/OneDrive - University of Nevada, Reno/Research/Mul_NB/results/sim_ttests_brca_spearman_rescaled_rvec_vital_d4777_rep100_14nov2019.rds")
## sim_ttests <- readRDS("~/OneDrive - University of Nevada, Reno/Research/Mul_NB/results/sim_ttests_brca_spearman_rescaled_rvec_vital_d4777_rep100_14nov2019.rds")
## add empirical data
## str(sim_ttests)
## which(!(1:100 %in% unique(sim_ttests$rep)))
sim_ttests$type = "sim"
emp_data <- data.frame(Gene = unique(sim_ttests$Gene), z = empirical_z_values, pvalue = empirical_p_values, rep = NA, type = "empirical", stringsAsFactors = FALSE)
sim_ttests <- rbind(sim_ttests, emp_data)
```

```{r repInd, echo=FALSE, eval=TRUE}
## Replicate with independence assumption
## set up parallel
## library(parallel)
## cl <- makeCluster(detectCores()-1)
## ## ## simulation replicates
## simreps <- 100
## parReplicate <- function(cl, n, expr, simplify=TRUE, USE.NAMES=TRUE)
##     parSapply(cl, integer(n), function(i, ex) eval(ex, envir=.GlobalEnv),
##               substitute(expr), simplify=simplify, USE.NAMES=USE.NAMES)
## # export functions/objects to the cores
## clusterExport(cl,c("sim_independent", "simreps", "n0", "n1", "params0", "params1", "d", "tmp_rnaseq"))
##  
## ## replicate data sets
## sim_ind_list <- parReplicate(cl, simreps, sim_independent(n0 = n0, n1 = n1, params0 = params0, params1 = params1, d = d), simplify = FALSE)
## str(sim_ind_list)
## ## stop the cluster
## stopCluster(cl)
## 
## ## put into long format for analysis
## rep_vec <- rep(1:simreps, each = d)
## sim_ind_ttests <- do.call("rbind", sim_ind_list)
## sim_ind_ttests$rep <- rep_vec
## sim_ind_ttests$type <- "ind"
## str(sim_ind_ttests)
## saveRDS(sim_ind_ttests, file = "sim_ind_ttests_simrep=100_twoSided.rds")
sim_ind_ttests <- readRDS("~/OneDrive - University of Nevada, Reno/Research/Mul_NB/JcompNgraphStats_submission/simDepHDdata_tf_rticle/sim_ind_ttests_simrep=100_twoSided.rds")
## combine
all_ttests <- rbind(sim_ttests, sim_ind_ttests)
all_ttests$type <- factor(all_ttests$type)
## table(all_ttests$type)
## table(all_ttests$rep)
```

# 4.2 Results

```{r repZplot, echo=FALSE, eval=TRUE, warning=FALSE, out.width='80%', fig.cap="Curves illustrating z-values from 4777 t-tests under the observed statistics ('Empirical'; blue) and simulations."}
## str(sim_ttests)
## plot
## p0 <- ggplot(all_ttests, aes(x = z, group=rep, color = type))
p0 <- ggplot(all_ttests[all_ttests$type != "ind", ], aes(x = z, group=rep, color = type))
p1 <- p0 + geom_density() + theme_bw()
p1 + geom_density(data=all_ttests[all_ttests$type != "sim", ], aes(x = z, group=rep, color = type))
## ggplot(aes = aes( x = empirical_z_values)) + geom_density()
```

# 5. Conclusions / Future Work

- Correlations matter during testing involving high dimensional data [see @Efron2007a].
- High performance algorithms enable high-dimensional copula-based [e.g., @Li2019gpu] multivariate simulations.
- Release `bigsimr` R package implementing this method.
- Explore the role of both row and column correlation (instead of just columnwise dependence).
- Explore the scalability of exact Pearson-matching methods (for example, see @Chen2001 and @Xia17).
- Explore the role of covariance estimation (sparsity, low-rank, etc) in the simulation process.

# 6.1 - Acknowledgements

\begin{columns}[T]
  \begin{column}{0.55\columnwidth}
    \begin{block}{University of Nevada, Reno}
      Alex Knudson, MS Stat student\\
      Anna Panorska, PhD\\
      Tomasz Kozubowski, PhD
    \end{block}
    \begin{block}{University of Arizona}
      Walter W.~Piegorsch, PhD\\
      Edward J.~Bedrick, PhD
      %%D.~Dean Billheimer, PhD\\
      %%Hao Helen Zhang, PhD\\
      %%Qike Li, MS
    \end{block}

    %%\begin{block}{Biologists}
    %%  Ikbel Achour, PhD\\
    %%  Joanne Berghout, PhD
    %%\end{block}
  \end{column}
  \begin{column}{0.45\columnwidth}
%%    \begin{block}{Medical doctors}
%%      Yves A.~Lussier, MD
%%    \end{block}
    \begin{block}{Grants/travel awards}
      Research \& Innovation, UNR
    \end{block}
    \begin{block}{CFE-CMSTATS 2019 organizers}
		Mihye Ahn, PhD\\
		Chae Ryon Kang, PhD\\
		All others involved
    \end{block}
  \end{column}
\end{columns}

# 6.2 References

### 2021-02-28 notebooks/cmstats2020_talk.rmd
re-add notebooks so Grant can check out new branch without losing files

@@
 - We provide a high-performance and GPU-accelerated implementation in the `R` package `bigsimr`, along with related tools.
-- An article motivating, describing, and applying the methods; archived [here](???).
+- An article motivating, describing, and applying the methods; in development.
 
@@
 |-------------|:--------------:|----------------------:|
-|$1000$ | Pearson ($\rho_P$) | ${ \bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho )$ |
-|$10,000$ | Spearman ($\rho_S$) | ${ \bf Y} \sim MVG( shape = 10, rate = 1, \rho )$ |
-| $100,000$| Kendall ($\tau$)| ${ \bf Y} \sim MVNB(p = `r prob`, r = 4,\rho)$ |
+|$1000$ | Pearson ($\rho_P$) | ${ \bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho_i ), i=1,\ldots,100$ |
+|$10,000$ | Spearman ($\rho_S$) | ${ \bf Y} \sim MVG( shape = 10, rate = 1, \rho_i ), i=1,\ldots,100$ |
+| $100,000$| Kendall ($\tau$)| ${ \bf Y} \sim MVNB(p = `r prob`, r = 4,\rho_i), i=1,\ldots,100$ |
 
-- For each of the unique 9 simulation configurations above, we estimate the correlation bounds and vary $\rho$ along a sequence of 100 points evenly placed within the bounds.
+- For each of the unique 9 simulation configurations described above, we estimate the correlation bounds and vary the correlations along a sequence of 100 points evenly placed within the bounds, aiming to explore $\Omega$.
+- Specifically, we set correlations $\{ \rho_1 = ( \hat{l} + \epsilon), \rho_2 = (\hat{l} + \epsilon) + \delta, \ldots, \rho_{100} = (\hat{u} - \epsilon) \}$, with $\hat{l}$ and $\hat{u}$ being the estimated lower and upper bounds, respectively, and increment value $\delta$.
+- The adjustment factor, $\epsilon=0.01$, is introduced to handle numeric issues when the bound is specified exactly.
+
+```{r ch040-combineBiSims, cache=F, echo=F}
+## compare relative differences in MSE for each dependency, compared to Pearson
+allDat <- NULL
+dat <- readRDS("../results/biNorm_sims.rds") %>% select( margins, type, N, rho, rho_hat  ) 
+allDat <- rbind( allDat, dat )
+dat <- readRDS("../results/biGamma_sims.rds") %>% select( margins, type, N, rho, rho_hat  )
+allDat <- rbind( allDat, dat )
+dat <- readRDS("../results/biNB_sims.rds") %>% select( margins, type, N, rho, rho_hat  )
+allDat <- rbind( allDat, dat )
+allDat$margins <- factor(allDat$margins, levels = c( 'norm', 'gamma', 'nbinom'),
+                         labels = c( 'MVN', 'MVG', 'MVNB' ) )
+allDat$type <- factor(allDat$type, levels = c( 'pearson', 'spearman', 'kendall' ),
+                      labels = c( 'Pearson', 'Spearman', 'Kendall' ) )
+allDat$N <- factor(allDat$N, levels = c( 1000, 10000, 100000) )
+```
 
@@
 
-```{r ch040-combineBiSims, cache=F, echo=F}
-## compare relative differences in MSE for each dependency, compared to Pearson
-allDat <- NULL
-dat <- readRDS("../results/biNorm_sims.rds") %>% select( margins, type, N, rho, rho_hat  ) 
-allDat <- rbind( allDat, dat )
-dat <- readRDS("../results/biGamma_sims.rds") %>% select( margins, type, N, rho, rho_hat  )
-allDat <- rbind( allDat, dat )
-dat <- readRDS("../results/biNB_sims.rds") %>% select( margins, type, N, rho, rho_hat  )
-allDat <- rbind( allDat, dat )
-allDat$margins <- factor(allDat$margins, levels = c( 'norm', 'gamma', 'nbinom'),
-                         labels = c( 'MVN', 'MVG', 'MVNB' ) )
-allDat$type <- factor(allDat$type, levels = c( 'pearson', 'spearman', 'kendall' ),
-                      labels = c( 'Pearson', 'Spearman', 'Kendall' ) )
-allDat$N <- factor(allDat$N, levels = c( 1000, 10000, 100000) )
-```
-
 ```{r ch040-biMAEtable, cache=F, echo=F, warning=F, message=F}
@@
 ```
+
 ## 4B - Estimate the marginal parameters
@@
 $$
-\hat{P}( {\bf Y} >= {\bf y_0 } ) = \sum_{b=1}^B I( {\bf Y^{(b) }} > {\bf y_0 } ) / B
+\hat{P}( {\bf Y} \geq {\bf y_0 } ) = \frac{1}{B} \sum_{b=1}^B I( {\bf Y^{(b) }} \geq {\bf y_0 } )
 $$

### 2021-02-28 notebooks/esco2020_talk.rmd
re-add notebooks so Grant can check out new branch without losing files

@@
     smaller: true
-    incremental: true
     logo: ../man/figures/logo.png
@@
 - Use `devtools` to install.
-- `bigsimr` uses `JAX` python libraries for GPU acceleration.
+- `bigsimr` uses `JAX` python libraries for GPU acceleration and other high
+  performance algorithms.
 

### 2021-02-28 notebooks/gradseminar10feb2021_talk.rmd
re-add notebooks so Grant can check out new branch without losing files

--- 2021-02-28 notebooks/cmstats2020_talk.rmd
@@
 ---
-subtitle: "Simulating Ultra High-Dimensional Multivariate Data"
-author:
-  - "Alfred G. **Schissler**, PhD"
-  - "Alexander D. **Knudson**, MS"
-  - "Tomasz J. **Kozubowski**, PhD"
-  - "Anna K. **Panorska**, PhD"
-  - "Juli **Petereit**, PhD"
-  - "*University of Nevada, Reno (UNR)*"
+title: "Simulating High-Dimensional Multivariate Data"
+subtitle: "using the bigsimr R Package"
+author: "Alfred G. Schissler, Alexander D. Knudson, Edward J. Bedrick, , Tomasz J. Kozubowski, Tin Nguyen, Anna K. Panorska, Juli Petereit, Walter W. Piegorsch, Duc Tran"
 institute: |
@@
     "University of Nevada, Reno"
-date: "21 Dec 2020 (*updated: `r Sys.Date()`)*"
+date: "10 Feb 2021 (*updated: `r Sys.Date()`)*"
 output:
@@
 library(tidyverse)
-set.seed(12212020)
+library(rstanarm)
+set.seed( 02102021 )
 ## devtools::install_github("ggobi/GGally")
@@
 
+```{r ch010-myD, include=FALSE, cache = TRUE, eval=T}
+d <- 1000
+```
+
+```{r ch010-processBRCA, echo=FALSE, eval=TRUE, results='none', cache=F}
+brcaRDS <- paste0( "../data/brca", d, ".rds" )
+if ( !file.exists( brcaRDS ) | !file.exists( '../data/examplePatients.txt' ) )  {
+    brca0 <- TCGA2STAT::getTCGA(disease = "BRCA",
+                            data.type = "RNASeq",
+                            p = 6L)
+    brca1 <- as.data.frame(t(brca0$dat))
+    gene_median <- apply(brca1, 2, median)
+    gene_median_order <- order(gene_median, decreasing = TRUE)
+    brca  <- brca1[,head(gene_median_order, d)]
+    writeLines( substring( row.names( brca ), first=1, last= 12 ), '../data/examplePatients.txt' ) 
+    saveRDS(brca, brcaRDS)
+}
+brca <- readRDS(file = brcaRDS)
+```
+
 ## Big Multivariate Data
@@
 - But measurements are often **correlated**, **non-normal**, and **high dimensional** in this era of big data.
-- One example are *RNA-seq data* obtained through high-throughput biomedical experiments.
-- 3 out of 20,501 mRNA measurements from 1212 breast cancer patients (TCGA BRCA):
+- One example are *RNA-seq data* obtained through high-throughput biomedical experiments with Over  20,000 mRNA measurements from `r dim(brca)[1]` breast cancer patients (TCGA BRCA).
 
-```{r, echo=F, out.width='60%', fig.align='center'}
-knitr::include_graphics("../fig/realDataFig.png")
+## Motivating example: RNA-seq data
+
+- Simulating high-dimensional, non-normal, correlated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
+- The RNA-seq data-generating process involves counting how often a particular messenger RNA (mRNA) is expressed in a biological sample.
+- RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data.
+- For human derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted.
+- Importantly, due to inherent biological processes, gene expression data exhibits correlation --- co-expression --- across genes [@BE07; @Schissler2018].
+
+## Intergeneic correlation
+
+```{r ch010-realDataFig, cache=F, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes from Table 1.
+The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations.
+Modeling these data motivate our simulation methodology."}
+# lowerfun <- function(data,mapping){
+#   ggplot(data = data, mapping = mapping)+
+#     geom_point()+
+#     scale_x_continuous(limits = c(-1.5,1.5))+
+#     scale_y_continuous(limits = c(-1.5,1.5))
+# }  
+GGally::ggpairs( data = small,
+                # lower = list( continuous = wrap(lowerfun)),
+                upper = list(
+                    continuous = wrap('cor', method = "spearman")
+                ), 
+                ) + theme_bw()
 ```
@@
 - We propose a Gaussian-copula random vector algorithm; an augmented NORmal To Anything (NORTA) @Cario1997 method. 
-- We provide a high-performance and GPU-accelerated implementation in the `R` package `bigsimr`, along with related tools.
+- We provide a high-performance and GPU-accelerated implementation in the `R` package [`bigsimr`](https://github.com/SchisslerGroup/bigsimr), along with related tools.
 - An article motivating, describing, and applying the methods; in development.
@@
 # 2 - Background
-
-## Motivating example: RNA-seq data
-
-- Simulating high-dimensional, non-normal, correlated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
-- The RNA-seq data-generating process involves counting how often a particular messenger RNA (mRNA) is expressed in a biological sample.
-- RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data.
-- For human derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted.
-- Importantly, due to inherent biological processes, gene expression data exhibits correlation --- co-expression --- across genes [@BE07; @Schissler2018].
-
-```{r ch010-myProb, include=FALSE, cache = TRUE}
-myProb <- 0.95
-```
-
-```{r ch010-processBRCA, echo=FALSE, eval=TRUE, results='none', cache=F}
-brcaRDS <- paste0( "../data/brca", round(myProb*100), ".rds" )
-if ( !file.exists( brcaRDS ) )  {
-    ## full workflow simulating RNA-seq data
-    allDat <- readRDS( file = "~/Downloads/complete_processed_tcga2stat_RNASeq2_with_clinical.rds" )
-    lastClinical <- which( names(allDat) == 'tumorsize' )
-    brca <- allDat[ allDat$disease == "BRCA", (lastClinical+1):ncol(allDat) ]
-    dim(brca) ## num of genes 20501
-    ## compute the median expression! avoid the 0s
-    brcaMedian <- apply(brca, 2, median)
-    ## retain top (1-myProb)*100% highest expressing genes for illustration
-    cutPoint <- quantile( x = brcaMedian, probs = myProb )
-    ## genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
-    genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
-    brca <- brca[ , genesToKeep ]
-    ## ncol(brca) / 20501
-    ## print(d <- length(genesToKeep))
-    ## save data as rds
-    saveRDS(brca, brcaRDS)
-}
-brca <- readRDS(file = brcaRDS)
-d <- ncol(brca)
-```
-
----
-
-```{r ch010-realDataTab, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
-## Describe and plot real data
-if ( !file.exists( '../data/exampleGenes.rds' ) )  {
-    set.seed(10192020)
-    exampleGenes <- colnames(brca)[sort(sample(ncol(brca), 3))]
-    saveRDS(exampleGenes, file = '../data/exampleGenes.rds')
-}
-exampleGenes <- readRDS('../data/exampleGenes.rds')
-small <-  brca %>% select( exampleGenes )
-knitr::kable(round(head( small, n = 5)),
-      booktabs = TRUE,
-      caption = 'mRNA counts for three selected high-expressing genes from the first five observations of the BRCA data set.'
-)
-```
-
----
-
-```{r ch010-realDataFig, cache=F, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, out.width='60%', fig.align='center', fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes.
-The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations.
-Modeling these data motivate our simulation methodology."}
-# lowerfun <- function(data,mapping){
-#   ggplot(data = data, mapping = mapping)+
-#     geom_point()+
-#     scale_x_continuous(limits = c(-1.5,1.5))+
-#     scale_y_continuous(limits = c(-1.5,1.5))
-# }
-
-if ( !file.exists( '../fig/realDataFig.png' ) ) {
-    p <- GGally::ggpairs( data = small,
-                                        # lower = list( continuous = wrap(lowerfun)),
-                         upper = list(
-                             continuous = wrap('cor', method = "spearman")
-                         )) + theme_bw()
-    ggsave('../fig/realDataFig.png', plot = p, dpi = 300)
-}
-knitr::include_graphics("../fig/realDataFig.png")
-```
 
@@
 
-1. Pre-processing for nonparametric dependency matching.  
-	(i) Convert from either ${\bf R_{Spearman}}$ or ${\bf R_{Kendall}}$ into the corresponding MVN input correlation ${\bf R_{Pearson}}$.
-	(ii) Check that ${\bf R_{Pearson}}$ is semi-positive definite.  
-	(iii) If not compute a close semi-positive definite correlation matrix ${\bf \widetilde{R}_{Pearson}}$.  
-2. Gaussian copula construction.  
-	(i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_{Pearson}})$.  
+1. Mapping step  
+   i. Convert the target Spearman correlation matrix $R_S$ to the corresponding MVN Pearson correlation $R_X$.   
+   or  
+i$^\prime$. Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$.
+2. Checking admissibility and, if needed, nearest correlation computation
+    (i) Check that $R_X$ is a correlation matrix, a PSD matrix with 1's along the diagonal.   
+	(ii) If $R_X$ is a correlation matrix, the input matrix is *admissible* in this scheme.  
+	Otherwise,  
+(ii$^\prime$) Replace $R_X$ with the nearest correlation matrix $R_X$, in the Frobenius norm.
+3. NORTA step
+	(i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)$.  
 	(ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
-3. Quantile evaluations.  
-	(i) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;
+	(iii) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;
 
@@
 
-```{r ch040-biMAEtable, cache=F, echo=F, warning=F, message=F}
+```{r ch040-BiError, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
 ## tabMSE  <- allDat %>%
@@
 ##    summarize( mse = mean( ( rho - rho_hat )^2 ) )
-tabMSE  <- allDat %>%
-    filter( N == 100000) %>%
+tabMAE  <- allDat %>%
+    ## filter( N == 100000) %>%
     group_by( N, type, margins ) %>%
@@
 
-knitr::kable(tabMSE,
+## AGS 12/22/20: Removing the latex formatting fixes the html render.
+## But how to improve latex table formatting? Better to use the template
+knitr::kable(tabMAE,
+             ## format = 'latex',
              col.names = c("No. of random vectors",
@@
                            "Mean abs. error"),
-             caption = '',
+             ## booktabs = TRUE,
+             caption = 'Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.',
+             ## linesep = c('', '',  '\\addlinespace')
 )
@@
 
-```{r ch050-readBRCA, echo=FALSE, eval=TRUE, cache=F}
-## full workflow simulating RNA-seq data
-brcaRDS <- '../data/brca95.rds'
-myProb = 0.95
-brca <- readRDS(file = brcaRDS)
-d <- ncol(brca)
-## smaller set for prototyping
-## d=5
-## brca <- brca[ , 1:d]
-```
-
 First estimate the desired correlation matrix using the fast implementation provided by `bigsimr`:
@@
 ## Estimate Spearman's correlation on the count data
-corType <- 'spearman'
-system.time( nb_Rho <- bigsimr::cor_fast( brca, method = corType ) )
+type <- 'spearman'
+R_S <- bigsimr::cor_fast(brca, method = type)
 ```
@@
 ```{r ch050-runBRCA-echo, echo=TRUE, eval=FALSE, cache = FALSE}
-## Set the number of random vectors
-n <- 10000
-## construct margins
-nb_margins <- make_nbinom_alist(sizes, probs)
-## run sims
-sim_nbinom <- rvec(n, nb_Rho, nb_margins, type = corType,
-                   ensure_PSD = TRUE, cores = cores) 
+sim_nbinom <- rvec(10000, R_X, nb_margins, ensure_PSD = FALSE, cores = cores) 
+colnames(sim_nbinom) <- names(brca)
 ```
@@
 ```{r ch050-runBRCA, echo=FALSE, results=FALSE, cache = FALSE}
-nb_margins <- make_nbinom_alist(sizes, probs)
-brcaSimRDS <- paste0( "../results/brca", round(myProb*100), "sim.rds" )
+brcaSimRDS <- paste0( "../results/brca", d, "sim.rds" )
+if( !file.exists(brcaSimRDS) ) {
+    set.seed(44)
+    sim_nbinom <- rvec(10000, R_X, nb_margins, ensure_PSD = FALSE, cores = cores)
+    saveRDS( sim_nbinom, brcaSimRDS )
+}
 sim_nbinom <- readRDS( brcaSimRDS )
@@
 
-```{r ch050-simDataFig, echo=F, fig.align='center', fig.cap='Simulated data for three selected high-expressing genes replicates the estimated data structure.', fig.show = 'hold', out.width='40%', fig.asp=1}
+```{r ch050-simDataFig, echo=F, fig.align='center', fig.cap='Simulated data for three selected high-expressing genes replicates the estimated data structure.', out.width='50%'}
 ## knitr::include_graphics("../fig/realDataFig.png")
@@
 
-## 4D - Assessing estimation efficiency
-
--  To assess the error in our correlation estimation above:
-
-```{r ch050-quadlossECHO, echo=T, eval=F, cache=FALSE}
-## Number of experiments
-m <- 10
-## Simulate random vectors equal to the sample size
-n <- nrow(brca)
-## convert outside for faster simulation
-nb_Rho_p <- bigsimr::cor_convert( rho = nb_Rho,
-                                 from = corType, to = "pearson" )
-## ensure PSD
-nb_Rho_p <- bigsimr::cor_nearPSD( G = nb_Rho_p )
-## create m random vectors and estimate correlation
-simRho <- replicate(n = m,
-  expr = { tmpSim <- rvec(n = n , nb_Rho_p, nb_margins,
-    type = 'pearson', ensure_PSD = FALSE, cores = cores);
-    bigsimr::cor_fast( x = tmpSim, method = corType )} ,
-simplify = FALSE)
-## find quadratic loss at each rep
-quadLoss <- unlist( lapply( simRho, rags2ridges::loss,
-                           T = nb_Rho, type = "quadratic"))
-
-```
-
-```{r ch050-quadloss, echo=F, eval=T, cache=FALSE}
-quadLossRDS <- paste0( "../results/brca", round(myProb*100), "quadLoss.rds" )
-quadLoss <- readRDS( quadLossRDS )
-```
-
-## 4D - Assessing estimation efficiency
-
-The `R` summary function supplies the mean-augmented five-number summary of the quadratic loss distribution computed above.
-
-```{r ch050-quadloss-summary, echo=TRUE, eval=T, cache=FALSE}
-summary(quadLoss)
-```
-
-- This distribution could be compared to high-dimensional designed covariance estimators to guide to help decide whether the additional complexity and computation time are warranted.
-
-## 4E - Scale up to high dimensional RNA-seq data
+## 4D - Scale up to high dimensional RNA-seq data
 
@@
 * Discrete margins can slightly disrupt the matching algorithm (due the probability of ties). 
-* (Nearly) exact Pearson matching is not supported.
+* ~~exact Pearson matching is not supported.~~ EDIT: Pearson matching added via Hermite polynomial in `Julia`
 
@@
 
-- Walter W. Piegorsch, PhD
-- Edward J. Bedrick, PhD
-- Yves A. Lussier, MD
-- Heather Knudson
+- Heather Knudson (graphic design)
 - Eric Li, MS
@@
 
-### CFE-CMSTATS 2020 organizers
+## Ongoing project HD multilevel survival model
 
- - Thanks to everyone who makes this event possible!
- - Professor Anna Panorska for the invitation to speak today.
+* precision medicine of cancer 
+* finding biomarkers from underrepresented patients
+
+```{r readFits, echo=F}
+fit <- fit0group <- readRDS( '~/article_survHD/results/fitOLDDATA_group_0_hs_ms_4_4_02162021_2000.rds')
+## summary(fit)
+## print(fit, digits=2)
+## ps_check(fit)
+```
+
+```{r psCheckFinalFit0group, echo=F, cache=T}
+raceLevels <- c('white', gsub( 'race', '', names(coef(fit0group))[2:3]) )
+nd <- data.frame( race= raceLevels )
+ps <- posterior_survfit(fit0group,
+                        newdata = nd,
+                        times = 0,
+                        extrapolate = TRUE, control = list(edist = 8000))
+panel_labels  <- c( '1' = "white", '2' = "black or african american", '3'= "asian")
+pps <- plot(ps) + facet_wrap( ~ id, labeller = labeller(id=panel_labels))
+pps
+```
+
+## Ongoing projects 2
+
+- Other simulation and PMF calculations (*T-Poisson* article under minor revisions)
+- Bayesian modeling of high-resolution NFL data (Erick Luerken, PhD student)
+- Bayesian multilevel modeling of the pyschometric function 
+- Bayesian multilevel modeling of US Electoral College (Andrey Sarantsev, Richard Foote).
+- Bayesian imputation of missing single-cell RNA-seq data.
 

### 2021-02-28 notebooks/cor_boundsBug.R

@@ -1,0 +1,12 @@
+library(bigsimr)
+size <- 4
+prob <- 3e-04
+type <- "spearman"
+n <- 1e3
+
+margins <- alist(
+        qnbinom(size = size, prob = prob),
+        qnbinom(size = size, prob = prob)
+)
+
+tmp_bounds <- cor_bounds(margins, method = type)

### 2021-02-28 010-background.Rmd
renv+box fully implemented
Addressing TJK edits

--- 2021-02-25 010-background.Rmd
@@
 
-To help visualize the bivariate relationships for these three selected genes across all patients, Figure \@ref(fig:ch010-realDataFig) plots the marginal distributions and estimated Spearman's correlations (see Equation \@ref(eq:spearman) below).
+To help visualize the bivariate relationships for these three selected genes across all patients, Figure \@ref(fig:ch010-realDataFig) displays the marginal distributions and estimated Spearman's correlations (see Equation \@ref(eq:spearman) below).
 

### 2021-02-28 030-bigsimr-package.Rmd
renv+box fully implemented
Addressing TJK edits

--- 2021-02-25 030-bigsimr-package.Rmd
@@
 
-This section describes a low-dimensional (2D) random vector simulation workflow via the `bigsimr` `R` package.
-`bigsimr` provides an interface to the native code written in Julia, registered as the `Bigsimr` Julia package PROVIDE URL.
-In addition to the native Julia `Bigsimr` package and `R` interface `bigsimr`, we also provide a python interface `bigsimr` (PROVIDE URL) that interfaces with the Julia `Bigsimr` package. 
+This section describes a low-dimensional (2D) random vector simulation workflow via the `bigsimr` `R` package (https://github.com/SchisslerGroup/r-bigsimr).
+`bigsimr` provides an interface to the native code written in Julia, registered as the `Bigsimr` Julia package (https://github.com/adknudson/Bigsimr.jl).
+In addition to the native Julia `Bigsimr` package and `R` interface `bigsimr`, we also provide a python interface `bigsimr` (https://github.com/SchisslerGroup/python-bigsimr/) that interfaces with the Julia `Bigsimr` package.
 The Julia implementation provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms)).

### 2021-02-28 index.Rmd
Addressing TJK edits

--- 2021-02-25 index.Rmd
@@
   - Walter W. Piegorsch
-  - Duc Tran
+  - Duc Tranrmarkdown::render_site(encoding = 'UTF-8')
 site: bookdown::bookdown_site
@@
 It also contains additional high-performance features, including multi-core algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix.
-Finally, we demonstrate example applications enabled via `bigsimr` by applying these functions to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=878$ patients and dimension $d=1000$.
+Monte carlo studies quantify the accuracy and scalability of our approach, up to $d=20,000$.
+Finally, we demonstrate example applications enabled via `bigsimr` by applying these methods to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=878$ patients and dimension $d=1000$.
 ---
 
-<!-- keywords simulation multivariate models high-dimensional data R packages -->
+*Keywords:* High-dimensional data, Big-data computing, Simulation, Nonparametric correlation, Gaussian copula, RNA-sequencing data
 
 ```{r indexSetup, include=FALSE, cache=FALSE}
+library(tidyverse)
 set.seed(09212020)
@@
 -->
-
 Another central issue lies in characterizing dependency between components in the high-dimensional random vector.

### 2021-02-28 010-background.Rmd

--- 2021-02-25 010-background.Rmd
@@
 ```{r ch010-realDataTab}
-set.seed(02242021)
+set.seed(2020-02-25)
 num_genes    <- 3

### 2021-02-28 index.Rmd

--- 2021-02-26 index.Rmd
@@
   - Walter W. Piegorsch
-  - Duc Tran
+  - Duc Tranrmarkdown::render_site(encoding = 'UTF-8')
 site: bookdown::bookdown_site
-bibliography: ["bigsimr.bib", "packages.bib"]
+bibliography: ["bigsimr.bib", "packages.bib", "alex.bib"]
 abstract:  It is critical to realistically simulate data when employing Monte Carlo techinques and evaluating statistical methodology.
@@
 It also contains additional high-performance features, including multi-core algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix.
-Finally, we demonstrate example applications enabled via `bigsimr` by applying these functions to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=878$ patients and dimension $d=1000$.
+Monte carlo studies quantify the accuracy and scalability of our approach, up to $d=20,000$.
+Finally, we demonstrate example applications enabled via `bigsimr` by applying these methods to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=878$ patients and dimension $d=1000$.
 ---
 
-<!-- keywords simulation multivariate models high-dimensional data R packages -->
+*Keywords:* High-dimensional data, Big-data computing, Simulation, Nonparametric correlation, Gaussian copula, RNA-sequencing data
 

### 2021-02-28 index.Rmd
begin corresponding author information

--- 2021-02-28 index.Rmd
@@
 author:
-  - Alfred G. Schissler
+  - Alfred G. Schissler*
   - Edward J. Bedrick
@@
   - Walter W. Piegorsch
-  - Duc Tranrmarkdown::render_site(encoding = 'UTF-8')
+  - Duc Tran
 site: bookdown::bookdown_site
@@
 *Keywords:* High-dimensional data, Big-data computing, Simulation, Nonparametric correlation, Gaussian copula, RNA-sequencing data
+
+*=Corresponding author. aschissler@unr.edu. 1664 N Virginia St, Reno, NV 89557
 

### 2021-03-01 index.Rmd
Revise index prior to CSDA submission

@@
 Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations.
-This article introduces the `bigsimr` R interface to the Julia `Bigsimr` package.
-This high-level package provides a flexible and scaleable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions with specified Pearson, Spearman, or Kendall correlation matrices.
-It also contains additional high-performance features, including multi-core algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix.
+This article introduces the `bigsimr` R package.
+This high-level package provides a flexible and scaleable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions a known Pearson, Spearman, or Kendall correlation matrix.
+`bigsimr` contains additional high-performance features, including multi-core algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix.
 Monte carlo studies quantify the accuracy and scalability of our approach, up to $d=20,000$.
-Finally, we demonstrate example applications enabled via `bigsimr` by applying these methods to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=878$ patients and dimension $d=1000$.
+Finally, we demonstrate example applications enabled via `bigsimr` by applying these methods to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples.
 ---
 
-*Keywords:* High-dimensional data, Big-data computing, Simulation, Nonparametric correlation, Gaussian copula, RNA-sequencing data
+*Keywords:*  Multivariate simulation, High-dimensional data, Nonparametric correlation, Gaussian copula, RNA-sequencing data, breast cancer
 
@@
 Such Monte Carlo (MC) techniques become difficult for high-dimensional data using existing algorithms and tools.
-This is particularly true when simulating massive *multivariate*, *non-normal* distributions, arising naturally in many fields of study.
+This is particularly true when simulating massive multivariate, non-normal distributions, arising naturally in many fields of study.
 
@@
 For normal data, the Pearson product-moment correlation describes the dependency perfectly.
-As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix exactly is computationally intense [@Chen2001; @Xia17].
+As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix is computationally intense [@Chen2001; @Xia17].
 On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependency, such as Spearman's $\rho$ and Kendall's $\tau$.
@@
 
-With all this in mind, we present a scaleable, flexible multivariate simulation algorithm.
+With all this in mind, we present a scalable, flexible multivariate simulation algorithm.
 The crux of the method lies in the construction of a Gaussian copula, in the spirit of the NORTA procedure.
@@
 
-Our study proceeds by providing background information, including a description of our motivating example application --- RNA-sequencing (RNA-seq) breast cancer data.
+Our study proceeds by providing background information, including a description of a motivating example application --- RNA-sequencing (RNA-seq) breast cancer data.
 Then we describe and justify our simulation methodology and related algorithms.
@@
 We conclude the Monte Carlo evaluations by summarizing the computation time for increasingly higher dimensional vectors.
-After the MC evaluations, we simulate random vectors motivated by our RNA-seq example, evaluate the accuracy, and provide example statistical computing tasks, namely a MC joint probability calculation and HD-correlation estimation efficiency.
+After the MC evaluations, we simulate random vectors motivated by our RNA-seq example, evaluate the accuracy, and provide example statistical computing tasks, namely MC estimation of joint probabilities and evaluateing HD correlation estimation efficiency.
 Finally, we discuss the method's utility, limitations, and future directions.

### 2021-03-01 index.Rmd
Revise index prior to CSDA submission

@@
 We conclude the Monte Carlo evaluations by summarizing the computation time for increasingly higher dimensional vectors.
-After the MC evaluations, we simulate random vectors motivated by our RNA-seq example, evaluate the accuracy, and provide example statistical computing tasks, namely MC estimation of joint probabilities and evaluateing HD correlation estimation efficiency.
+After the MC evaluations, we simulate random vectors motivated by our RNA-seq example, evaluate the accuracy, and provide example statistical computing tasks, namely MC estimation of joint probabilities and evaluating HD correlation estimation efficiency.
 Finally, we discuss the method's utility, limitations, and future directions.

### 2021-03-01 010-background.Rmd
Revise 010 prior to CSDA submission

@@
 
-Two nonparametric, rank-based measures common in practice are Spearman's correlation (denoted $\rho_S$) and Kendall's $\tau$. Define
-
-\begin{equation}
-(\#eq:spearman)
-\rho_S(X,Y) = 3 \left[ P\left[ (X_1 - X_2)(Y_1-Y_3) > 0 \right] - P\left[ (X_1 - X_2)(Y_1-Y_3) < 0 \right] \right],
-\end{equation}
-
-\noindent where $(X_1, Y_1) \overset{d}{=} (X,Y), X_2 \overset{d}{=} X, Y_3 \overset{d}{=} Y$ with $X_2$ and $Y_3$ are independent of one other and of $(X_1, Y_1)$.
-Spearman's $\rho_S$ has an appealing correspondence as the Pearson's correlation coefficient on *ranks* of the values, thereby captures nonlinear yet monotone relationships.
-
-Kendall's $\tau$, on the other hand, is the difference in probabilities of concordant and discordant pairs of observations $(X_i, Y_i)$ and $(X_j, Y_j)$.
-By concordance we mean that orderings have the same direction (e.g., if $X_i < X_j$, then $Y_i < Y_j$) and is determined by the ranks of the values, not the values themselves.
+Two nonparametric, rank-based measures common in practice are Spearman's correlation (denoted $\rho_S$) and Kendall's $\tau$.
+Spearman's $\rho_S$ has an appealing correspondence as the Pearson correlation coefficient on *ranks* of the values, thereby captures nonlinear yet monotone relationships.
+Kendall's $\tau$, on the other hand, is the difference in probabilities of concordant and discordant pairs of observations, $(X_i, Y_i)$ and $(X_j, Y_j)$, with concordance meaning that orderings have the same direction (e.g., if $X_i < X_j$, then $Y_i < Y_j$).
+Note that concordance is determined by the ranks of the values, not the values themselves.
 
@@
 
-*Correspondence among Pearson, Spearman, Kendall correlations*.
+*Correspondence among Pearson, Spearman, and Kendall correlations*.
 There is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficient, as the marginal distributions $F_i$ are intrinsic in their calculation.
@@
 For discrete marginals with large or infinite support, computing the adjustment factors $\sum_x p(x)^3$, $\sum_y p(y)^3$ over all large number of pairs becomes expensive (often violating desired property SP1).
-Further the infinite sums must be approximated for count-valued data.
+Further, the infinite sums must be approximated for count-valued data.
 
-Also, for a discrete random variable $Y_i$, some care must be taken to define the quantile function $F_{i}^{-1}$. Let
+Also, for a discrete random variable $Y_i$, some care must be taken to define the quantile function $F_{i}^{-1}$. A standard practice is to set
 
@@
 Given two marginal distributions, $\rho_P$ is not free to vary over the entire range of possible correlations $[-1,1]$.
-The so-called *Frechet-Hoeffding bounds* are well-studied ([@Nelsen2007; @BF17]).
+The so-called *Frechet-Hoeffding bounds* are well-studied [@Nelsen2007; @BF17].
 These constraints cannot be overcome through algorithm design.
@@
 
-\noindent where $U$ is a uniform random variable in $(0,1)$, and $F^{-1}_1, F^{-1}_2$ are the inverse CDFs of random variables $X_1$ and $X_2$, respectively.
-
-For discrete random variables, define $F^{-1}$ as in Equation \@ref(eq:inverseCDF).
+\noindent where $U$ is a uniform random variable on $(0,1)$ and $F^{-1}_1, F^{-1}_2$ are the inverse CDFs of $X_1$ and $X_2$, respectively, definitely by \@ref(eq:inverseCDF) when the variables are discrete.
 
@@
 
-A Gaussian copula has marginal CDFs that are all normal, $F_i = \Phi, \forall \, i$.
+A Gaussian copula has marginal CDFs that are all standard normal, $F_i = \Phi, \forall \, i$.
 This representation corresponds to a multivariate normal (MVN) distribution with standard normal marginal distributions and covariance matrix ${\bf R_P}$.
@@
 Sklar's Theorem [@Sklar1959; @Ubeda-Flores2017] guarantees that given inverse CDFs $F_i^{-1}$s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions.
-For example, using Gaussian copulas, we can construct a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)^\top$ with $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$, via ${\bf U} = (U_1, \ldots,  U_d)^\top$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$ provides $Y_i \sim F_i, \, \forall \, i$.
+For example, using Gaussian copulas, we can construct a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)^\top$ with $Y_i \sim F_i$, viz.
+$Y_i = F_i^{-1}(\Phi(X_i)), i=1, \ldots, d$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})$.

### 2021-03-01 020-simulation-algorithm.Rmd
Revise 020 prior to CSDA submission

--- 2021-02-27 020-simulation-algorithm.Rmd
@@
 
-The well-known NORTA algorithm [@Cario1997] simulates a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$. Specifically, NORTA algorithm proceedings as follows:
+The well-known NORTA algorithm [@Cario1997] simulates a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$. Specifically, NORTA algorithm proceeds as follows:
 
@@
 
-Matching specified Pearson correlation coefficients exactly (step 2 above), however, is problematic.
+Matching specified Pearson correlation coefficients exactly (step 2 above), however, is computationally costly.
 In general, there is no closed-form correspondence between the components of the input $\Sigma_{\bf Z}$ and target $\Sigma_{\bf Y}$.
-Matching the correlations involves evaluating or approximating $\binom{d}{2}$ integrals of the form $EY_iY_j = \int \int y_i y_j f_{X}(F_i^{-1}(\Phi(z_i)), F_j^{-1}(\Phi(z_j))dy_idy_j$, for $i,j=1,\ldots, d, \; i \neq j$.
-For high-dimensional data, these evaluations are often too costly to enable feasible simulation studies.
+Matching the correlations involves evaluating or approximating $\binom{d}{2}$ integrals of the form
+
+\begin{equation}
+    \mathrm{E}\left[Y_i Y_j\right] = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} F_i^{-1}\left[\Phi(z_i)\right] F_j^{-1}\left[\Phi(z_j)\right] \phi(z_i, z_j, \rho_z) dz_i dz_j,
+    (\#eq:pearsonIntegralRelation)
+\end{equation}
+
+where $\phi(\cdot)$ is the joint probability distribution function of two correlated standard normal variables.
+For HD data, (nearly) exact evaluation may become too costly to enable practical simulation studies.
 For low-dimensional problems, however, methods and tools exist to match Pearson correlations precisely [@Xia17] and the publicly available `nortaRA` R package [@Chen2001].
-
-To maintain scalability (SP1), our solution is to essentially avoid this complication in Pearson matching.
-Since our goal is to simulate non-normal marginals, we greatly prefer the use of rank-based measures, such as $\rho_S$ and $\tau$, from a modeling standpoint.
-Further, $\rho_S$ and $\tau$'s invariance under monotone transformation (see [Background](#background)), preserves the correlation coefficients through steps 3, 4, and 5 in the NORTA algorithm above.
-This eliminates the need for computing the $\binom{d}{2}$ integrals to match exactly (nothing is for free, however, as discussed below in Section \@ref(rand-vec-gen)).
-
-Now, if one does desire to characterize dependency using Pearson correlations, simply using the target Pearson correlation matrix as the initial conditions to our proposed algorithm will lead to approximate matching in the resultant distribution [@Song00] in many practical applications.
-The quality of this approximation depends on the setting, but in practice, for high-dimensional count data we find the accuracy to be adequate.
-Later, we study robustness of our method to this limitation in selected [Monte Carlo evaluations]({#simulations}).
+As described later, to enable HD Pearson matching we approximate these integrals.
 
@@
 @GH02 conducted a Monte Carlo study to estimate the probability of encountering NORTA defective matrices while increasing the dimension $d$.
-They found for what is now considered low-to-moderate dimensions ($d \approx 20$) that almost *all* feasible matrices are NORTA defective.
+They found that for what is now considered low-to-moderate dimensions ($d \approx 20$), almost *all* feasible matrices are NORTA defective.
 This stems from the concentration of measure near the boundary of the space of all possible correlation matrices as dimension increases.
@@
 1. Mapping step
-    (i) Convert the target Spearman correlation matrix $R_S$ to the corresponding MVN Pearson correlation $R_X$.
-
-    or
-
-    (i$^\prime$) Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$.
+    (i) Convert the target Spearman correlation matrix $R_S$ to the corresponding MVN Pearson correlation $R_X$. Alternatively,  
+    (i') Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$. Alternatively,  
+    (i'') Convert the target Pearson correlation matrix to $R_P$ to the corresponding, approximate MVN Pearson correlation $R_X$.  
 2. Check admissibility and, if needed, compute the nearest correlation matrix.
@@
     (ii$^\prime$) Replace $R_X$ with the nearest correlation matrix $\tilde{R}_X$, in the Frobenius norm.  
-3. NORTA step
+3. Gaussian copula
     (i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)$.  
-    (ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
+    (ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz. $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
     (iii) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$.
 
-The "mapping step" takes advantage of the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertKendall) or \@ref(eq:convertSpearman), respectively (implemented as `cor_covert`).
+*Step 1*.
+The first two descriptions of the *Mapping step* employ the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertKendall) and \@ref(eq:convertSpearman), respectively (implemented as `cor_covert`).
+Initializing our algorthim to match the nonparametric correlations by computing these equations for all pairs is computationally trivial.
+For the computationally expensive process to match Pearson correlations, we approximate Equation \@ref(eq:pearsonIntegralRelation) for all pairs of margins.
+To this end, we implement the approximation scheme introduced by @xiao2019matching.
+Briefly, the many double integrals of the form in \@ref(eq:pearsonIntegralRelation) are approimated by weighted sums of Hermite polynomials.
 
-Our algorithm allows the generation of high-dimensional multivariate data with arbitrary marginal distributions with a broad class of admissible Spearman correlation matrices and Kendall $\tau$ matrices.
+*Step 2*.
+Once $R_X$ has been determined, we check admissibility of the target correlation via the steps described above.
+If $R_X$ is not a valid correlation matrix, then we compute the nearest correlation matrix.
+Finding the nearest correlation matrix is a common statistical computing problem.
+The defacto default function in `R` is `Matrix::nearPD`, an alternating projection algorithm due to @higham2002computing.
+As implemented the function fails to scale to HD.
+Instead, we provide the quadratically convergent algorithm based on the theory of strongly semi-smooth matrix functions (@qi2006quadratically).
+The nearest correlation matrix problem can be written down as the following convex optimization problem:
+
+\begin{align*}
+    \mathrm{min}\quad & \frac{1}{2} \Vert G - X \Vert^2 \\
+    \mathrm{s.t.}\quad & X_{ii} = 1, \quad i = 1, \ldots , n, \\
+    & X \in S_{+}^{n}
+\end{align*}
+
+For nonparametric correlation measures, our algorithm allows the generation of high-dimensional multivariate data with arbitrary marginal distributions with a broad class of admissible Spearman correlation matrices and Kendall $\tau$ matrices.
 The admissible classes consist of the matrices that map to a Pearson correlation matrix for a MVN.
-
 In particular, if we let $X$ be MVN with $d$ components and set
@@
 In particular, $R_P$ is approximately $R_S$ for a MVN, suggesting that the admissible set $\Omega_S$ should be flexible as $R_P$ can be any PD matrix with 1's along the diagonal.
-We provide methods to check whether a target $R_S$ is an element of the *admissible set* $\Omega_S$ (and similarly for $R_K$) and also for checking that each pairwise correlation value is admissible (via the MC estimation of the 2-dimensional Frechet bounds).
+We provide methods to check whether a target $R_S$ is an element of the *admissible set* $\Omega_S$ (and similarly for $R_K$).
 
-Yet this scheme still suffers similar issues as the original Pearson-matching NORTA algorithm regarding the increasing probability of encountering a non-admissible correlation matrix as dimension increases.
-In our experience, the mapping step for large $d$ almost always produces a $R_X$ that is not in $\Omega_P$ (or even a correlation matrix).
-The second step handles this issue by computing the nearest correlation matrix to $R_X$.
-In Section \@ref(package) we provide a basic description of how to quantify and potentially control the approximation error.
+There is an increasing probability of encountering a non-admissible correlation matrix as dimension increases.
+In our experience, the mapping step for large $d$ almost always produces a $R_X$ that is not a correlation matrix.
+In Section \@ref(package) we provide a basic description of how to quantify and control the approximation error.
 Further, the RNA-seq example in Section \@ref(examples) provides an illustration of this in practice.
 
-The final step implements a high-performance NORTA algorithm. Specifically, step 3i uses a fast multi-core multivariate normal simulator, and steps 3ii, 3iii are distributed over CPU cores.
-
-## Algorithm details
-
-*Pearson Matching*
-Given a desired target correlation $\rho_Y$ between a pair of marginal distributions, the correlation coefficient $\rho_Z$ used in the Gaussian copula must be specified so as to satisfy equation \@ref(eq:pearsonIntegralRelation).
-
-\begin{equation}
-    \mathrm{E}\left[Y_i Y_j\right] = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} F_i^{-1}\left[\Phi(z_i)\right] F_j^{-1}\left[\Phi(z_j)\right] \phi(z_i, z_j, \rho_z) dz_i dz_j
-    (\#eq:pearsonIntegralRelation)
-\end{equation}
-
-where $\phi(\cdot)$ is the joint probability distribution function of two correlated standard normal variables.
-In general, the double integral in \@ref(eq:pearsonIntegralRelation) is difficult or impossible to solve analytically, so numerical methods are required.
-Our implementation is based off the work of @xiao2019matching where the marginal transformation is approximated by a weighted sum of Hermite polynomials.
-
-*Nearest Correlation Matrix*
-
-This algorithm is trying to find a correlation matrix $X$ close to $G$ in the convex optimization problem:
-
-\begin{align*}
-    \mathrm{min}\quad & \frac{1}{2} \Vert G - X \Vert^2 \\
-    \mathrm{s.t.}\quad & X_{ii} = 1, \quad i = 1, \ldots , n, \\
-    & X \in S_{+}^{n}
-\end{align*}
-
-* Qi, H., & Sun, D. (2006). A quadratically convergent Newton method for computing the nearest correlation matrix. SIAM journal on matrix analysis and applications, 28(2), 360-385.
+*Step 3* The final step implements a NORTA-inspired, Gaussian copula approach to produce the desired margins.
+Steps 1 and 2 determine the MVN Pearson correlation values that will eventually match the target correlation.
+Then all that is required is a fast MVN simulator, a standard normal CDF, and well-defined quantile functions for marginals.
+The MVN is transformed to a copula (distribution with standard uniform margins) by applying the normal CDF $\phi(\cdot)$.
+The quantile functions $F^{-1}$ are applied across the margins to return the desired random vector ${\bf Y}$.

### 2021-03-01 030-bigsimr-package.Rmd
Revise 030 prior to CSDA submission

@@
 This section describes a low-dimensional (2D) random vector simulation workflow via the `bigsimr` `R` package (https://github.com/SchisslerGroup/r-bigsimr).
-`bigsimr` provides an interface to the native code written in Julia, registered as the `Bigsimr` Julia package (https://github.com/adknudson/Bigsimr.jl).
+The package `bigsimr` provides an interface to the native code written in Julia, registered as the `Bigsimr` Julia package (https://github.com/adknudson/Bigsimr.jl).
 In addition to the native Julia `Bigsimr` package and `R` interface `bigsimr`, we also provide a python interface `bigsimr` (https://github.com/SchisslerGroup/python-bigsimr/) that interfaces with the Julia `Bigsimr` package.
@@
 
-The bounds on the Pearson correlation coefficient $R_P$ between these margins is restricted as seen below in our MC estimated correlation bounds:
+The bounds on the Pearson correlation coefficient $R_P$ between these margins are restricted as seen below in our MC estimated correlation bounds:
 

### 2021-03-01 040-monte-carlo-evaluation.Rmd
Revise 040 prior to CSDA submission

@@
 The results for $d \leq 500$ are fast with all times under 3 seconds.
-Table \@ref(tab:ch040-moderateDtab) displays the total computation time of the algorithms steps 1through 3, including estimation step i.
+Table \@ref(tab:ch040-moderateDtab) displays the total computation time of the algorithms steps 1 through 3, including estimation step i.
 
@@
 Figure \@ref(fig:ch040-largeDfig) displays computation times for $d=\{1000, 2500, 5000, 10000\}$.
-For $d$ of 1000 and 2500, the total time is under a couple of minutes.
+For $d$ equal to 1000 and 2500, the total time is under a couple of minutes.
 At $d$ of 5000 and 10,000, Pearson correlation matching in the *Adjust Correlation* step becomes costly.
 Interestingly, Pearson is actually faster than Kendall for $d=10,000$ due to bottlenecks in *Compute Correlation* and *Check Admissibility*.
-Uniformly, matching Spearman correlations is faster with total times under 5 minutes for $d=10,000$, making Spearman correlations the most computationally-friendly dependency type.
+Uniformly, matching Spearman correlations is faster, with total times under 5 minutes for $d=10,000$, making Spearman correlations the most computationally-friendly dependency type.
 With this in mind, we scaled the simulation to $d=20,000$ for the Spearman type and obtaned the 1,000 vectors in under an hour (data not shown).

### 2021-03-01 050-examples.Rmd
Revise 050 prior to CSDA submission

@@
 Modeling RNA-seq using multivariate probability distributions is natural as inter-gene correlation is an inherent part of biological processes [@Wang2009b].
-Yet many models do not account for this, leading to major disruptions to the operating characteristics of statistical estimation, testing, and prediction.
-See @Efron2012 for a detailed discussion with related methods and @Wu2012b, @Schissler2018, @Schissler2019 for applied examples.
+And yet, many models do not account for this, leading to major disruptions to the operating characteristics of statistical estimation, testing, and prediction.
+See @Efron2012 for a detailed discussion with related methods and @Wu2012b, @Schissler2018; @Schissler2019 for applied examples.
 The following subsections apply `bigsimr`'s methods to real RNA-seq data, including replicating an estimated parametric structure, MC probability estimation, and MC evaluation of correlation estimation efficiency.
@@
 We assume a multivariate negative binomial (MVNB) model as RNA-seq counts are often over-dispersed and correlated.
-All $d$ selected genes exhibit over-dispersion (data not shown) and, so, we proceed to estimate the NB parameters $(r_i, p_i), i=1,\ldots,d$ to determine the target marginal PMFs $f_i$.
-To complete specification of the simulation algorithm inputs, we estimate the Spearman correlation matrix ${ \bf R}_{Spearman}$ to characterize dependency.
+Since all $d$ selected genes exhibit over-dispersion (data not shown), we proceed to estimate the NB parameters $(r_i, p_i), i=1,\ldots,d$, to determine the target marginal PMFs $f_i$.
+To complete specification of the simulation algorithm inputs, we estimate the Spearman correlation matrix ${ \bf R}_{S}$ to characterize dependency.
 
@@
 We use the method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model.
-The marginal distributions are from the same probability family (NB) yet are heterogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.
+While, marginal distributions are from the same probability family (NB), they are heterogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.
 The functions below support this estimation for later use in `rvec`.
@@
 
-Notably, the marginal NB probabilities $\hat{p}_i's$ are small --- ranging in the interval $[`r min(probs)` , `r max(probs)`]$.
+Notably, the estimated marginal NB probabilities $\{ \hat{p}_i \}$ are small --- ranging in the interval $[`r min(probs)` , `r max(probs)`]$.
 This gives rise to highly variable counts and, typically, less restriction on potential pairwise correlation pairs.
@@
 
-While the exact $d \times d$ Spearman correlation matrix is not strictly admissible in our scheme (seen by the non-positive definite result above), the approximation is close with a maximum absolute error of `r max( abs(targets - approximates))` and average absolute error of `r mean(abs(targets - approximates))` across the `r choose(d,2)` correlations.
+While the exact $d \times d$ Spearman correlation matrix is not strictly admissible in our scheme (as seen by the non-positive definite result above), the approximation is close with a maximum absolute error of `r max( abs(targets - approximates))` and average absolute error of `r mean(abs(targets - approximates))` across the `r choose(d,2)` correlations.
 
@@
 Figure \@ref(fig:ch050-simDataFig) displays the simulated counts and pairwise relationships for our example genes from Table \@ref(tab:ch010-realDataTab).
-Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parameter form and with less extreme points then the observed data in Figure \@ref(fig:ch010-realDataFig).
+Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parametric form and with less extreme points then the observed data in Figure \@ref(fig:ch010-realDataFig).
 
@@
 ```{r ch050-figBRCA, out.width='80%', fig.cap="Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set.
-The dashed red lines indicated equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes)."}
+The dashed red lines indicate equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes)."}
 include_graphics('fig/ch050-figBRCA.png')
@@
 *Limitations, conclusions, and recommendations.*
-The results show good simulation performance for our choice of parameters settings.
+The results show overall good simulation performance for our choice of parameters settings.
 Our settings were motivated by modeling high-expressing genes from TCGA BRCA data set.
@@
 $$
-P( {\bf Y} = {\bf y} ), y_i \in \chi_i.
+P( {\bf Y} = {\bf y} ), \: y_i \in \chi_i.
 $$
@@
 Compact representations with convenient computational forms are rare for high-dimensional constructions, especially with heterogeneous, correlated marginal distributions (or margins of mixed data types).
-Given a large number simulated vectors as produced above, estimated probabilities are readily given by counting the proportion of simulated vectors meeting the desired condition.
+Given a large number of simulated vectors as produced above, estimated probabilities are readily given by counting the proportion of simulated vectors meeting the desired condition.
 In our motivating application, one may ask what is the probability that all genes expressed greater than a certain threshold value ${ \bf y}_0$.
-
-Then we estimate
+This can be estimated as
 
 $$
-\hat{P}( {\bf Y} \ge {\bf y_0 } ) = \frac{1}{B} \sum_{b=1}^B I( {\bf Y^{(b) }} \ge {\bf y_0 } )
+\hat{P}( {\bf Y} \ge {\bf y_0 } ) = \frac{1}{B} \sum_{b=1}^B I( {\bf Y}^{ (b) } \ge {\bf y_0 } ),
 $$
@@
 In particular, we wish to assess the error in our correlation estimation above.
-We used a conventional method, based on classical statistical theory, which unfortunately, was not designed for high-dimensional data.
-Indeed, high-dimensional covariance estimation (and precision matrices) is an active area of statistical science (.e.g., @Won2013g and @VanWieringen2016).
+We used a conventional method, based on classical statistical theory which was not designed for high-dimensional data.
+Indeed, high-dimensional covariance estimation (and precision matrices) is an active area of statistical science, .e.g., [@Won2013g; @VanWieringen2016].
 
@@
 Since our simulation is much faster for the Pearson correlation type (see Figure \@ref(fig:ch040-largeDfig)), we only convert the Spearman correlation matrix once (and ensure it is positive definite).
-At each iteration, we estimate the quadratic loss (residual sum of squared errors) from the specified ${\bf R}_{Spearman}$, producing a distribution of loss values.
+At each iteration, we estimate the quadratic loss (residual sum of squared errors) from the specified ${\bf R}_{S}$, producing a distribution of loss values.
 

### 2021-03-01 080-appendix.Rmd
updated all bigsimr code
moved generated data into separate scripts
Addressing TJK edits
Revise 080 prior to CSDA submission

@@
 \noindent Consider a bivariate example where we have a correlated $(Y_1, Y_2)^\top$ with $Y_1\sim N(\mu_1, \sigma_1^2)$ (normal) and $Y_2\sim LN(\mu_2, \sigma_2^2)$ (lognormal).
-
-\vspace{0.1in}
-
-\noindent First, let us note that when we get such a vector from BigSimR package then in fact it can be represented as
+First, let us note that when we get such a vector from BigSimR package then in fact it can be represented as
 
@@
 \label{kram1}
-(Y_1, Y_2)^\top \stackrel{d}{=} \left(X_1, e^{X_2}\right)^\top
+(Y_1, Y_2)^\top \stackrel{d}{=} \left(X_1, e^{X_2}\right)^\top,
 \end{equation}
@@
 
-\item Return $(Y_1, Y_2)^\top$, where $Y_i=F_i^{-1}(U_i)$, $i=1,2$, where $F_i(\cdot)$ is the CDF of $Y_i$ and $F_i^{-1}(\cdot)$ is its inverse (the quantile function).
+\item Return $(Y_1, Y_2)^\top$, where $Y_i=F_i^{-1}(U_i)$, $i=1,2$, and $F_i(\cdot)$ is the CDF of $Y_i$ and $F_i^{-1}(\cdot)$ is its inverse (the quantile function).
 
@@
 
-where $c_2=\sigma_{Y_2}/\mu_{Y_2}$ is the {\it coefficient of variation} (CV) of the variable $Y_2$.
+where $c_2=\sigma_{Y_2}/\mu_{Y_2}$ is the *coefficient of variation* (CV) of the variable $Y_2$.
 Thus, the range of possible correlations in this model is not affected by the distribution of $Y_1$, and is determined by the CV of $Y_2$ as follows:
@@
 
- \vspace{0.1in}
-
- \noindent {\bf Further notes on the basic example in Section 4.1}.
-Based on the above analysis, one may propose a model of the form (\ref{kram1}) for these data, where $Y_1$ is the temperature (normal) and $Y_2$ is the Ozone variable (lognormal).
-Then, $X_1=Y_1$ and $X_2=\log Y_2$ will be bivariate normal, and the parameters can be estimated  based on the bivariate normal model.
-One can again use the marginal estimates as discussed in the paper, although one could actually use the exact same method for both variables (based on normality, after converting the Ozone variable to normal through the logarithmic transformation).
-According to the analysis presented in the paper, the values of 3.419 and 0.6967 are the estimates (the MLEs) of the mean and the variance of the underlying normal random variable $X_2$, that is $\hat{\mu}_2 = 0.3.419$ and $\hat{\sigma}_2=0.6967$.
-When this estimate of $\sigma_2$ is used in (\ref{kram8}), the factor on the far right becomes
+Plugging in the estimands of these quantities from Section \@ref(package), we see that
 
@@
 
-Thus, the possible range of correlation becomes $(-0.881, 0.881)$. This approximately agrees with the MC results in Section \@ref(package).
+Thus, the possible range of correlation becomes $(-0.881, 0.881)$. This approximately agrees with the MC results provided.

### 2021-03-01 060-conclusion-discussion.Rmd
Revise 060 prior to CSDA submission

@@
 
-We have introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a user-friendly, high-performance `R` package named [`bigsimr`](https://schisslergroup.github.io/bigsimr/).
-The random vector generation method is largely inspired by NORTA (@Cario1997) and Gaussian copula-based approaches (@MB13, @BF17, @Xia17).
-The major contributions of this work are high-dimensional scalability, flexible modeling of dependency, and an high-performance implementation with broad potential data analytic applications for modern, big-data statistical computing.
+We have introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a user-friendly, high-performance `R` package [`bigsimr`].
+The random vector generation method is inspired by NORTA [@Cario1997] and Gaussian copula-based approaches [@MB13, @BF17, @Xia17].
+The major contributions of this work are methods and software for flexible, scalable simulation of HD multivariate probability distributions with broad potential data analytic applications for modern, big-data statistical computing.
+For example, one could simulate high-resolution time series data, such as those consistent with an auto-regressive moving average model exhibiting a specified Spearman structure.
+Or our methods could be used to simulate sparsely correlated data, as many HD methods assume, via specifying a *spiked correlation matrix*.
 
-Researchers working in multivariate computation frequently encounter such difficulties and need to find a close positive (semi-)definite matrix.
+Researchers working in multivariate computation frequently encounter such difficulties and need to find a close or nearest correlation matrix.
 A widely-available routine for this task in `R` is `matrix::nearPD`, though it is not suitable for high dimensions.
@@
 It is customary to compare new tools and algorithms directly to existing competing methods and software.
-In this study, however, we only employ our proposed methodology, as our previous work has shown that existing tools are simply not designed nor feasible to meet our high-dimensional goal (see @Li2019gpu for evaluations of the `R` `copula` package and others).
+In this study, however, we only employ our proposed methodology, as our previous work has shown that existing `R` tools are simply not designed to meet our high-dimensional goal (see @Li2019gpu for evaluations of the `R` `copula` package and others).
 For the bivariate simulations, existing packages such as [`nortaRA`](https://github.com/cran/NORTARA/blob/master/inst/doc/NORTARA.R) work well to match Pearson correlations exactly.
@@
 There are limitations to the methodology and implementation.
+We could only investigate selected multivariate distributions in our Monte Carlo studies.
+There may be instances that the methods do not perform well.
+Along those lines, we expect that correlation values close to the boundary of the feasible region could result in algorithm failure.
+Another issue is that for discrete distributions, we use continuous approximations when the support set is large.
+This could limit the scalability/accuracy for particular joint distributions.
+
+<!-- 
 The most obvious missing feature of the proposed methodology is the inability to match a Pearson correlation matrix exactly.
@@
 While we do not provide an implementation directly supporting Pearson matching, users may supply their own input Pearson correlation after employing a supplementary matching scheme [@Cario1997; @XZ19].
+-->
 

### 2021-03-01 010-background.Rmd
Finish revising prior to passing to Alex

@@
 The `bigsimr` `R` package presented here provides multiple algorithms that operate with high-dimensional multivariate data; however, all these algorithms were originally designed to support a single task: to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics.
-Specifically, our goal is to efficiently simulate a large number, $B$, of random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with *correlated* components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i$, where $d$ can be very large and still be computed in a practical timescale.
+Specifically, our goal is to efficiently simulate a large number, $B$, of HD random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with *correlated* components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i$.
 

### 2021-03-01 index.Rmd
Finish revising prior to passing to Alex

@@
 Yet these approaches do not scale in a timely fashion to high-dimensional problems [@Li2019gpu].
-For discrete data, early simulation strategies had major flaws --- such as failing to obtain the full range of possible dependencies (e.g., admitting only positive correlations: see @Park1996).
+For discrete data, early simulation strategies had major flaws --- such as failing to obtain the full range of possible correlations (e.g., admitting only positive correlations: see @Park1996).
 While more recent approaches [@MB13; @Xia17; @BF17] have largely remedied this issue for low-dimensional problems, the existing tools are not designed to scale to high dimensions.
@@
 
-Another central issue lies in characterizing dependency between components in the high-dimensional random vector.
+Another central issue lies in characterizing dependence between components in the high-dimensional random vector.
 The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc).
@@
 As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix is computationally intense [@Chen2001; @Xia17].
-On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependency, such as Spearman's $\rho$ and Kendall's $\tau$.
-Throughout, we focus on matching these nonparametric dependency measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology.
+On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependence, such as Spearman's $\rho$ and Kendall's $\tau$.
+Throughout, we focus on matching these nonparametric dependence measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology.
 We do, however, provide Pearson matching with some caveats.
@@
 Further, we introduce the `bigsimr` R package that provides high-performance software implementing our algorithm.
-The algorithm design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed-form relationships between dependency measures for the multivariate normal (MVN) distribution.
+The algorithm design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed-form relationships between dependence measures for the multivariate normal (MVN) distribution.
 

### 2021-03-01 010-background.Rmd
added GSDist ref. Did built-in spell check

@@
 For simplicity, we only consider high expressing genes.
-In turn, we begin by filtering to retain the top `r d` of the highest-expressing genes (in terms of median expression) of the oover 20,000 gene measurements from $N=`r nrow(example_brca)`$ patients' tumor samples.
+In turn, we begin by filtering to retain the top `r d` of the highest-expressing genes (in terms of median expression) of the over 20,000 gene measurements from $N=`r nrow(example_brca)`$ patients' tumor samples.
 This gives a great number of pairwise dependencies among the marginals (specifically, $`r choose(d,2)`$ correlation parameters).

### 2021-03-01 020-simulation-algorithm.Rmd
added GSDist ref. Did built-in spell check

@@
 The first two descriptions of the *Mapping step* employ the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertKendall) and \@ref(eq:convertSpearman), respectively (implemented as `cor_covert`).
-Initializing our algorthim to match the nonparametric correlations by computing these equations for all pairs is computationally trivial.
+Initializing our algorithm to match the nonparametric correlations by computing these equations for all pairs is computationally trivial.
+
 For the computationally expensive process to match Pearson correlations, we approximate Equation \@ref(eq:pearsonIntegralRelation) for all pairs of margins.
 To this end, we implement the approximation scheme introduced by @xiao2019matching.
-Briefly, the many double integrals of the form in \@ref(eq:pearsonIntegralRelation) are approimated by weighted sums of Hermite polynomials.
+Briefly, the many double integrals of the form in \@ref(eq:pearsonIntegralRelation) are approximated by weighted sums of Hermite polynomials.
+Matching coefficients for pairs of continuous distributions is made tractable by this method, but for discrete distributions (especially discrete distributions with large support sets or infinite support), the problem is not so simple or efficient.
+
+Our solution is to approximate discrete distributions by a continuous distribution.
+However the question is then what distribution to use.
+We found that Generalized S-Distributions [@muino2006gs] solve this problem by approximating a wide range of unimodal distributions, both continuous and discrete.
+Using the GS-Distribution approximation of discrete distributions in Pearson matching scheme above yields favorable results.
 

### 2021-03-01 080-appendix.Rmd
added GSDist ref. Did built-in spell check

@@
 \noindent Consider a bivariate example where we have a correlated $(Y_1, Y_2)^\top$ with $Y_1\sim N(\mu_1, \sigma_1^2)$ (normal) and $Y_2\sim LN(\mu_2, \sigma_2^2)$ (lognormal).
-First, let us note that when we get such a vector from BigSimR package then in fact it can be represented as
+First, let us note that when we get such a vector from bigsimr package then in fact it can be represented as
 
@@
 
-Plugging in the estimands of these quantities from Section \@ref(package), we see that
+Plugging in the estimates of these quantities from Section \@ref(package), we see that
 

### 2021-03-01 010-background.Rmd
revised through 020

@@
 
-<!-- 
-The next section describes the motivating data set we use to provide example applications.
--->
-
 ## Motivating example: RNA-seq data
@@
 For human-derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted [@Schissler2019].
-Importantly, due to inherent biological processes, gene expression data exhibit correlation --- co-expression --- across genes [@BE07; @Schissler2018].
+Importantly, due to inherent biological processes, gene expression data exhibit correlation (co-expression) across genes [@BE07; @Schissler2018].
 
@@
 
-*Correspondence among Pearson, Spearman, and Kendall correlations*.
+*Correspondence among Pearson, Spearman, and Kendall correlations*
+
 There is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficient, as the marginal distributions $F_i$ are intrinsic in their calculation.
@@
 
-*Discrete marginal considerations*.
+*Discrete marginal considerations*
+
 Spearman's correlation for discrete marginal suffers issues whenever there is a nonzero probability of ties.
@@
 
-*Marginal-dependent bivariate correlation bounds*. 
+*Marginal-dependent bivariate correlation bounds*
+
 Given two marginal distributions, $\rho_P$ is not free to vary over the entire range of possible correlations $[-1,1]$.

### 2021-03-01 letter_bigsimr_CSDA.Rmd
CSDA submission stuff

---
institution:
  name: "University of Nevada, Reno"
  department: "Mathematics and Statistics"
  address:
    line1: "1664 N. Virginia Street / 0084"
    line2: "Reno, Nevada 89557"
  phone: "(775) 784-6773"
  fax: "(775) 784-6378"
  url: https://www.unr.edu/math/
  logo: "logos/blue.png"
author:
  first: "Alfred"
  last: "Schissler"
  short: "Grant"
  title: "Ph.D."
  position: "Assistant Professor, Statistics"
  signature: "signature.jpg"
to:
  first: "Alexander"
  last: "Knudson"
  short: "Alex"
date: "`r format(Sys.Date(), format='%d %B %Y')`"
opening: "Dear CSDA Annals of Statistical Data Science Editor(s):"
closing: "Sincerely,"
output:
  pdf_document:
    template: templateSingleSignature.tex
---

As an invited speaker at CMStatistics 2020, I was delighted to learn of the opportunity to have our work included in the first issue of CSDA Annals of Statistical Data Science.

On behalf of my co-authors, I'm pleased to submit our manuscript entitled *Simulating High-Dimensional Multivariate Data using the `Bigsimr` Package* for your consideration.
We feel that the work is well suited to the indicated themes and hope that it will be considered a valuable contribution to this prestigious journal.

I hope that everything is in order. If not, or if you have any additional needs, please do not hesitate to contact me. Thank you and all those involved in this process.

Best regards,

### 2021-03-01 020-simulation-algorithm.Rmd

@@
 For HD data, (nearly) exact evaluation may become too costly to enable practical simulation studies.
-For low-dimensional problems, however, methods and tools exist to match Pearson correlations precisely [@Xia17] and the publicly available `nortaRA` R package [@Chen2001].
+For low-dimensional problems, however, methods and tools exist to match Pearson correlations precisely; see [@Xia17] and the publicly available `nortaRA` R package [@Chen2001].
 As described later, to enable HD Pearson matching we approximate these integrals.
 
-*NORTA in higher dimensions*. 
+*NORTA in higher dimensions*
+
 Sklar's theorem provides a useful characterization of multivariate distributions through copulas.
@@
     (i) Check that $R_X$ is a correlation matrix, a positive definite matrix with 1's along the diagonal.   
-	(ii) If $R_X$ is a correlation matrix, the input matrix is *admissible* in this scheme.  
-	Otherwise,
+	(ii) If $R_X$ is a correlation matrix, the input matrix is *admissible* in this scheme. Otherwise
     (ii$^\prime$) Replace $R_X$ with the nearest correlation matrix $\tilde{R}_X$, in the Frobenius norm.  
@@
 
-*Step 1*.
+*Step 1*
+
 The first two descriptions of the *Mapping step* employ the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertKendall) and \@ref(eq:convertSpearman), respectively (implemented as `cor_covert`).
@@
 Our solution is to approximate discrete distributions by a continuous distribution.
-However the question is then what distribution to use.
+The question then becomes that of which distribution to use.
 We found that Generalized S-Distributions [@muino2006gs] solve this problem by approximating a wide range of unimodal distributions, both continuous and discrete.
@@
 
-*Step 2*.
-Once $R_X$ has been determined, we check admissibility of the target correlation via the steps described above.
+*Step 2*
+
+Once $R_X$ has been determined, we check admissibility of the adjusted correlation via the steps described above.
 If $R_X$ is not a valid correlation matrix, then we compute the nearest correlation matrix.
@@
 
-*Step 3* The final step implements a NORTA-inspired, Gaussian copula approach to produce the desired margins.
+*Step 3*
+
+The final step implements a NORTA-inspired, Gaussian copula approach to produce the desired margins.
 Steps 1 and 2 determine the MVN Pearson correlation values that will eventually match the target correlation.

### 2021-03-01 030-bigsimr-package.Rmd

@@
 ```
-
-<!-- 
-*Comparison to Uncorrelated Samples*.
-We can compare the bivariate distribution above to one where no correlation is taken into account.
-
-```{r ch030-compare-uncorrelated, eval=FALSE}
-df_sim2 <- data.frame(
-  Temp = rnorm(10000, mean(df$Temp), sd(df$Temp)),
-  Ozone = rlnorm(10000, mle_mean(df$Ozone), mle_sd(df$Ozone))
-)
-
-p2 <- df_sim2 %>%
-  ggplot(aes(Temp, Ozone)) +
-  geom_density_2d_filled() +
-  theme(legend.position = "none") +
-  labs(x = "Temperature", y = "Ozone") +
-  scale_y_continuous(limits = c(0, max(df$Ozone))) +
-  scale_x_continuous(limits = range(df$Temp))
-
-p2Temp <- ggplot(df_sim2, aes(Temp)) + 
-  geom_density(alpha = 0.5, fill = "lightseagreen") +
-  theme(axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        axis.text = element_blank(),
-        axis.ticks = element_blank())
-
-p2Ozone <- ggplot(df_sim2, aes(Ozone)) + 
-  geom_density(alpha = 0.5, fill = "lightseagreen") + 
-  theme(axis.title.y = element_blank(),
-        axis.title.x = element_blank(),
-        axis.ticks = element_blank(),
-        axis.text = element_blank()) +
-  coord_flip()
-
-p2Temp + plot_spacer() + p2 + p2Ozone + 
-  plot_layout(widths = c(3,1), heights = c(1, 3))
-```
-
-Notice that this uncorrelated bivariate distribution has the same marginal distributions as the correlated example above.
--->

### 2021-03-01 040-monte-carlo-evaluation.Rmd
Revise 040 to include AGS specs

@@
 The adjustment factor, $\epsilon=0.01$, is introduced to handle numeric issues when the bound is specified exactly.
-
-<!-- 
-*Bivariate Normal*.
-Let's simulate a bivariate normal and check our correlation matching performance as N increases.
-Standard bivariate normal 
-Here we have BVN( $\mu_1 = \mu_2 = 0, \rho_{type}$ ).
-We vary $\rho$ across the entire possible range of correlations for each correlation type.
-Figure \@ref(fig:ch040-biNormPlot) `bigsimr` recovers the Pearson specified correlations for MVN.
-
-*Bivariate Gamma*. 
-Similarly, let's check the performance for a non-symmetric continuous distribution: a standard (rate =1) bivariate gamma.
-Here we have a Bivariate Gamma with $shape_1 = shape_2 = 10, \rho_{type}$.
-We vary $\rho$ across the entire possible range of correlations for each correlation type.
-Figure \@ref(fig:ch040-biGammaPlot) `bigsimr` recovers the Pearson specified correlations for bivariate gamma.
-*Bivariate Negative Binomial*. 
-Figure \@ref(fig:ch040-biNegBinPlot) `bigsimr` recovers the Pearson specified correlations for bivariate negative binomial.
- -->
 
@@
 
-To mimick the workflow, we first produce a synthetic "data set" by completing the following steps:
+To mimic the workflow, we first produce a synthetic "data set" by completing the following steps:
 
@@
 
+The experiments are conducted on a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, with all 16 threads employed during computation.
 The results for $d \leq 500$ are fast with all times under 3 seconds.
@@
 
-The results larger dimensional vectors show scalebility to ultra-high dimensions for all three correlation types, although the total times do become much larger. 
+The results larger dimensional vectors show scalability to ultra-high dimensions for all three correlation types, although the total times do become much larger. 
 Figure \@ref(fig:ch040-largeDfig) displays computation times for $d=\{1000, 2500, 5000, 10000\}$.
@@
 Uniformly, matching Spearman correlations is faster, with total times under 5 minutes for $d=10,000$, making Spearman correlations the most computationally-friendly dependency type.
-With this in mind, we scaled the simulation to $d=20,000$ for the Spearman type and obtaned the 1,000 vectors in under an hour (data not shown).
+With this in mind, we scaled the simulation to $d=20,000$ for the Spearman type and obtained the 1,000 vectors in under an hour (data not shown).
 In principle, this would enable the simulation of an entire human-derived RNA-seq data set.

### 2021-03-01 index.Rmd

@@
 This article introduces the `bigsimr` R package.
-This high-level package provides a flexible and scaleable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions a known Pearson, Spearman, or Kendall correlation matrix.
+This high-level package provides a flexible and scalable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions and known Pearson, Spearman, or Kendall correlation matrix.
 `bigsimr` contains additional high-performance features, including multi-core algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix.
-Monte carlo studies quantify the accuracy and scalability of our approach, up to $d=20,000$.
+Monte carlo studies quantify the accuracy and scalability of our approach, up to $d=10,000$.
 Finally, we demonstrate example applications enabled via `bigsimr` by applying these methods to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples.
@@
 
-As many have noted, it can be vexing to simulate dependent, non-normal/discrete data ---  even for low dimensional settings [@MB13; @XZ19].
-For continuous non--normal multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula approaches [@Nelsen2007] are well-studied, with flexible, robust software available [@Yan2007; @Chen2001].
+As many have noted, it can be vexing to simulate dependent, non-normal/discrete data, even for low dimensional settings [@MB13; @XZ19].
+For continuous non-normal multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula approaches [@Nelsen2007] are well-studied, with flexible, robust software available [@Yan2007; @Chen2001].
 Yet these approaches do not scale in a timely fashion to high-dimensional problems [@Li2019gpu].
-For discrete data, early simulation strategies had major flaws --- such as failing to obtain the full range of possible correlations (e.g., admitting only positive correlations: see @Park1996).
+For discrete data, early simulation strategies had major flaws, such as failing to obtain the full range of possible correlations (e.g., admitting only positive correlations: see @Park1996).
 While more recent approaches [@MB13; @Xia17; @BF17] have largely remedied this issue for low-dimensional problems, the existing tools are not designed to scale to high dimensions.
-
-<!-- 
-Here we propose a scheme that ignores the analytic goal --- and, instead, aims to match the empirically-derived estimated correlation data and marginal characteristics.
--->
 
@@
 With all this in mind, we present a scalable, flexible multivariate simulation algorithm.
-The crux of the method lies in the construction of a Gaussian copula, in the spirit of the NORTA procedure.
+The crux of the method lies in the construction of a Gaussian copula in the spirit of the NORTA procedure.
 Further, we introduce the `bigsimr` R package that provides high-performance software implementing our algorithm.
@@
 
-Our study proceeds by providing background information, including a description of a motivating example application --- RNA-sequencing (RNA-seq) breast cancer data.
+Our study proceeds by providing background information, including a description of a motivating example application: RNA-sequencing (RNA-seq) breast cancer data.
 Then we describe and justify our simulation methodology and related algorithms.

### 2021-03-01 index.Rmd
Polish abstract

@@
 bibliography: ["bigsimr.bib", "packages.bib", "alex.bib"]
-abstract:  It is critical to realistically simulate data when employing Monte Carlo techinques and evaluating statistical methodology.
+abstract:  It is critical to realistically simulate data when employing Monte Carlo techniques and evaluating statistical methodology.
 Measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments.
@@
 `bigsimr` contains additional high-performance features, including multi-core algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix.
-Monte carlo studies quantify the accuracy and scalability of our approach, up to $d=10,000$.
+Monte Carlo studies quantify the accuracy and scalability of our approach, up to $d=10,000$.
 Finally, we demonstrate example applications enabled via `bigsimr` by applying these methods to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples.

### 2021-03-01 030-bigsimr-package.Rmd

@@
 In addition to the native Julia `Bigsimr` package and `R` interface `bigsimr`, we also provide a python interface `bigsimr` (https://github.com/SchisslerGroup/python-bigsimr/) that interfaces with the Julia `Bigsimr` package.
-The Julia implementation provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms)).
-When designing `bigsimr`, we aimed to conveniently provide parallelized computation, through multi-core and GPU acceleration, while allowing advanced users to customize.
+The Julia package provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms)).
 
-The subsections below describe the basic use of `bigsimr`, by stepping through an example workflow using the data set `airquality` that contains daily air quality measurements in New York, May to September 1973 [@Chambers1983].
-This workflow proceeds from computing environment setup, data wrangling, estimation, simulation configuration, random vector generation, and, finally, result visualization.
-The code chunk below prepares the computing environment:
-
-<!-- 
-`bigsimr` is an R package for simulating high-dimensional multivariate data with arbitrary marginal distributions.
-The efficiency behind generating multivariate samples comes from the ability to utilize a GPU during the generation of multivariate normal samples and the subsequent transformation to uniform samples (see section \@ref(rand-vec-gen)).
-A GPU is not necessary in order to use `bigsimr`, and the speed benefits generally only come when simulating very high dimensional data ($d > 10000$).
-
-The next computational step after obtaining the correlated uniform marginals is the inverse transform into the target marginal distributions.
-Since this consists of $d$ independent transformations, we utilize parallelization to achieve higher throughput.
-As with many parallel algorithms, there is overhead associated with forking the task to utilize multiple cores, however cost is negligible compared to the total run-time.
--->
+The subsections below describe the basic use of `bigsimr` by stepping through an example workflow using the data set `airquality` that contains daily air quality measurements in New York, May to September 1973 [@Chambers1983].
+This workflow proceeds from setting up the computing environment, to data wrangling, estimation, simulation configuration, random vector generation, and, finally, result visualization.
 
@@
 First, we load the `bigsimr` library and a few other convenient data science packages, including the syntactically-elegant `tidyverse` suite of `R` packages.
+The code chunk below prepares the computing environment:
+
+```{r ch030-computeEnvironmentSetup, eval=FALSE, echo=TRUE}
+library("tidyverse")
+library("bigsimr")
+# Activate multithreading in Julia
+Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
+# Load the Bigsimr and Distributions Julia packages
+bs <- bigsimr_setup()
+dist <- distributions_setup()
+```
 
@@
 For this example, we avoid goodness-of-fit considerations to determine the marginal distributions.
-But it seems sensible without domain knowledge to estimate these quantities from the data and `bigsimr` contains fast functions designed for this task.
+But it seems sensible without domain knowledge to estimate these quantities from the data, and `bigsimr` contains fast functions designed for this task.
 
@@
 
-As mentioned, the user must decide how to describe correlation, based on the particulars of the problem.
+As mentioned, the user must decide how to describe correlation based on the particulars of the problem.
 For non-normal data and for improved simulation accuracy/scalability in our scheme, we advocate the use of Spearman's $\rho$ correlation matrix $R_S$ and Kendall's $\tau$ correlation matrix $R_K$.
@@
 bounds$lower
+bounds$upper
 ```
@@
 ## Simulating random vectors
-Finally, we arrive at the main function of `bigsimr`, `rvec`. We now simulate $B=10,000$ random vectors from the assumed joint distribution of Ozone levels and Temp.
+
+Finally, we arrive at the main function of `bigsimr`: `rvec`. We now simulate $B=10,000$ random vectors from the assumed joint distribution of Ozone levels and Temp.
 

### 2021-03-01 010-background.Rmd
the document renders on my PC

@@
 
-To help visualize the bivariate relationships for these three selected genes across all patients, Figure \@ref(fig:ch010-realDataFig) displays the marginal distributions and estimated Spearman's correlations (see Equation \@ref(eq:spearman) below).
+To help visualize the bivariate relationships for these three selected genes across all patients, Figure \@ref(fig:ch010-realDataFig) displays the marginal distributions and estimated Spearman's correlations.
 
@@
 \begin{equation}
+\rho_P(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ \mathrm{Var}(X)\mathrm{Var}(Y)\right]^{1/2}}.
 (\#eq:pearson)
-\rho_P(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ \mathrm{Var}(X)\mathrm{Var}(Y)\right]^{1/2}}.
 \end{equation}
@@
 For example, the Spearman's correlation of a discrete-valued random variable $X$ with itself could be less than 1 [@MB13].
-One remedy for this issue is to rescale Equation \@ref(eq:spearman).
+One remedy for this issue is to rescale the Spearman correlation.
 For two random variables $X,Y$ with probability mass functions (PMFs) or probability densities functions (PDFs) $p(x)$ and $q(y)$, respectively, define the rescaled Spearman's correlation as

### 2021-03-01 030-bigsimr-package.Rmd
the document renders on my PC

@@
 
-```{r ch030-basic-example-setup, echo=TRUE, cache=FALSE}
+```{r ch030-basic-example-setup, cache=FALSE}
 box::use(
@@
 ```{r ch030-air-quality-data, echo=TRUE}
-df <- airquality %>%
-  select(Temp, Ozone) %>%
-  drop_na()
+df <- airquality %>% select(Temp, Ozone) %>% drop_na()
 ```
@@
 ```{r ch030-aq-temp-pars, echo=TRUE}
-df %>% 
-  select(Temp) %>% 
+df %>% select(Temp) %>% 
   summarise_all(.funs = c(mean = mean, sd = sd))
@@
 mle_sd <- function(x) mean( sqrt( (log(x) - mean(log(x)))^2 ) )
-
 df %>% 
@@
 ```{r ch030-margins-alist, echo=TRUE}
-margins <- c(
-  dist$Normal(mean(df$Temp), sd(df$Temp)),
-  dist$LogNormal(mle_mean(df$Ozone), mle_sd(df$Ozone))
-)
+margins <- c(dist$Normal(mean(df$Temp), sd(df$Temp)),
+             dist$LogNormal(mle_mean(df$Ozone), mle_sd(df$Ozone)))
 ```
@@
 (R_X <- bs$cor_convert(R_S, bs$Spearman, bs$Pearson))
-
 # Step 2. Check admissibility
-julia_call("Bigsimr.iscorrelation", R_X)
+bs$iscorrelation(R_X)
 ```

### 2021-03-01 040-monte-carlo-evaluation.Rmd
the document renders on my PC

@@
 The experiments are conducted on a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, with all 16 threads employed during computation.
-The results for $d \leq 500$ are fast with all times under 3 seconds.
+The results for $d \leq 500$ are fast with all times (except Kendall) under 3 seconds.
 Table \@ref(tab:ch040-moderateDtab) displays the total computation time of the algorithms steps 1 through 3, including estimation step i.
@@
 
-*Limitations, conclusions, and recommendations.*
+*Limitations, conclusions, and recommendations*
+
 In the bivariate studies, we chose arbitrary simulation parameters for three distributions, moving from the Gaussian to the discrete and non-normal, multivariate negative binomial.

### 2021-03-01 050-examples.Rmd
the document renders on my PC

@@
 
-<!--
-This suggests simulating high-dimensional multivariate NB (MVB) with heterogeneous marginals would be useful tool in the development and evaluation of RNA-seq analytics.
-This procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions.
-All these genes exhibit over-dispersion and, so, we proceed to estimate the NB parameters $(r_i, p_i), i=1,\ldots,d$ to determine the target marginal PMFs $g_i(y_i)$ (via method of moments).
-Notably, the $\hat{p}_i's$ are small --- ranging in $[3.934 \times 10^{-6} , 1.217 \times 10^{-2}]$.
-To complete the simulation algorithm inputs, we estimate the Pearson correlation matrix $\bf{R_Y}$ and set that as the target correlation.
--->
-
-<!--
-With the simulation targets specified, we proceed to simulate $N=10,000$ random vectors $\bf{Y}$ $=( Y_1,\ldots,Y_d)$ with target Pearson correlation $\bf{R_Y}$ and marginal PMFs $g_i(y_i)$ using a $\bf{T}$-Poisson hierarchy of Kind II.
-Specifically, we first employ the \emph{direct Gaussian copula} approach to generate $N$ random vectors following a standard multivariate Gamma distribution $\bf{T}$ with shape parameters $r_i$ equal to the target $n_i$ and Pearson correlation matrix $\bf{R_T}$.
-Care must be taken when setting the specifying $\bf{R}$ (refer to Equation \ref{gay.cop.pdf}) --- we employ Equation \ref{mix.poi.corr} to compute the scaling factors $c_{i,j}$ and adjust the underlying correlations to ultimately match the target $\bf{R_Y}$.
-Notably, of the $525,825$ pairwise correlations from the $1026$ genes, no scale factor was less than $0.9907$, indicating the model can produce essentially the entire range of possible correlations.
-Here we are satisfied with approximate matching of the specified Gamma correlation and set $\bf{R}$ = $\bf{R_T}$ in our Gaussian copula scheme ($\bf{R}$ indicating the specified multivariate Gaussian correlation matrix).
-Finally, we generate the desired random vector $Y_i=N_i(t_i)$ by simulating Poisson counts with expected value $\mu_i=\lambda_i \times T_i$, for $i=1,\ldots,d$ (with $\lambda_i=\frac{(1-p_i)}{p_i}$) and repeat $N=10,000$ times.
--->
-
 With this goal in mind, we first estimate the desired correlation matrix using the fast implementation provided by `bigsimr`:
@@
 The functions below support this estimation for later use in `rvec`.
-
-<!-- 
-TYPESTE MoM estimators here.
--->
 
@@
 R_X <- bs$cor_convert(R_S, bs$Spearman, bs$Pearson)
-
 # 2a. Check admissibility
-(is_valid_corr <- julia_call("Bigsimr.iscorrelation", R_X))
-
+(is_valid_corr <- bs$iscorrelation(R_X))
 # 2b. compute nearest correlation
@@
 
-<!-- 
-Utilizing 15 CPU threads in a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, the simulation completed in just over 2 minutes.
--->
-
 ```{r ch050-figBRCA, out.width='80%', fig.cap="Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set.
@@
 
-*Limitations, conclusions, and recommendations.*
+*Limitations, conclusions, and recommendations*
+
 The results show overall good simulation performance for our choice of parameters settings.

### 2021-03-01 060-conclusion-discussion.Rmd
the document renders on my PC

@@
 In this study, however, we only employ our proposed methodology, as our previous work has shown that existing `R` tools are simply not designed to meet our high-dimensional goal (see @Li2019gpu for evaluations of the `R` `copula` package and others).
-For the bivariate simulations, existing packages such as [`nortaRA`](https://github.com/cran/NORTARA/blob/master/inst/doc/NORTARA.R) work well to match Pearson correlations exactly.
-
-<!-- 
-We advocate the use of Kendall's $\tau$ as it better captures correlation among components of non-normal, as well as non-linear patterns of association.
-Moreover, the matching Kendall's $\tau$ is nearly trivial due to the invariant of monotone transformations, after an adjustment to the input correlation matrix.
--->
-
-<!-- 
-We also show utility in our methodology through an application to differential gene expression analysis from RNA-sequencing data.
-The application results show that correlations indeed matter in the large-scale hypothesis testing, as many others have noted (for example, see @BE07, @Wu2012b).
-We also hope that the application provides an example workflow --- and other strategy to use in simulation design.
-Even the best-performing simulations we provide show a gap from the empirical distribution.
-To keep the demonstration straightforward, we did not attempt to match the empirical distribution of test statistics as precisely as possible.
-Yet our methodology could be more creatively applied to meet that goal.
-One could consider marginal distributions from different families, such as Poisson for a subset of genes.
-One could imagine finding a best fitting probability distribution among a class of distributions for each gene and this could perhaps a better fit.
-One could imagine additional structures/features in the data that an analyst could model to improve the correspondence with the empirical values, for example row correlations and high-dimensional covariance estimators (see @Won2013g).
--->
+For the bivariate simulations, existing packages such as [`nortaRA`](https://github.com/superdesolator/NORTARA) work well to match Pearson correlations exactly.
 
@@
 This could limit the scalability/accuracy for particular joint distributions.
-
-<!-- 
-The most obvious missing feature of the proposed methodology is the inability to match a Pearson correlation matrix exactly.
-As discussed in the [Algorithms](algorithms) section and extensively by @XZ19, this is a computationally intense procedure and Pearson's correlation is not a natural choice to describe dependency for non-normal marginals.
-While we do not provide an implementation directly supporting Pearson matching, users may supply their own input Pearson correlation after employing a supplementary matching scheme [@Cario1997; @XZ19].
--->
-
-<!-- 
-Further, while we provide the ability of the user to specify discrete marginals, exact matching a desired dependency measure is not yet obtained.
-One potential solution for this is *rescale* to account for ties [ref].
-Finally we note, that the algorithm requires invertible marginal cdfs.
-This gives some restriction to the available marginal probability distributions (DOES IT?
-FOR EXAMPLE?).
-From a practical computing standpoint, the user's computing resource provides the ultimate limit on how large a dimension can be simulated using the `bigsimr` R package.
-A user must consider carefully the available memory and cores when conducting a Monte Carlo experiment.
--->
-
-<!-- 
-Future work includes developing scaleable algorithms to match the Pearson correlation matrix more precisely, discrete-margin specific modifications including fast Spearman's correlation rescaling (see Equation \@ref(eq:spearmanRescaled)), and high-dimensional covariance estimation.
-From an implementation standpoint, `bigsimr` only supports Nvidia GPUs and redesigning the code using OpenCL would broaden the users who would benefit.
-As data-analytic problems grow to even larger dimension, multi-GPU support is a promising hardware-based future direction.
--->

### 2021-03-01 index.Rmd
the document renders on my PC

@@
 
-```{r include=FALSE}
+```{r 000-setup, include=FALSE}
 # automatically create a bib database for R packages

### 2021-03-01 article_bigsimr.Rmd
revise 050

--- 
title: "Simulating High-Dimensional Multivariate Data"
subtitle: "using the `Bigsimr` Package"
author:
  - Alfred G. Schissler*
  - Edward J. Bedrick
  - Alexander D. Knudson
  - Tomasz J. Kozubowski
  - Tin Nguyen
  - Anna K. Panorska
  - Juli Petereit
  - Walter W. Piegorsch
  - Duc Tran
site: bookdown::bookdown_site
bibliography: ["bigsimr.bib", "packages.bib", "alex.bib"]
abstract:  It is critical to realistically simulate data when employing Monte Carlo techinques and evaluating statistical methodology. Measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations. This article introduces the `bigsimr` R package. This high-level package provides a flexible and scalable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions and known Pearson, Spearman, or Kendall correlation matrix. `bigsimr` contains additional high-performance features, including multi-core algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix. Monte carlo studies quantify the accuracy and scalability of our approach, up to $d=10,000$. Finally, we demonstrate example applications enabled via `bigsimr` by applying these methods to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples.
---
```{r include=FALSE, cache=FALSE}
options(digits = 4)

knitr::opts_chunk$set(
  # Text Results
  eval = TRUE,
  echo = FALSE,
  results = 'markup',
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  # Code Decoration
  comment = NA,
  background = '#F7F7F7',
  # Cache
  cache = TRUE,
  # Plots
  fig.show = "hold",
  fig.width = 6.5,
  fig.asp = 0.618,  # 1 / phi
  out.width = "85%",
  fig.align = 'center'
)

# options(dplyr.print_min = 6, dplyr.print_max = 6)

seed_number <- 2020-02-25
set.seed(seed = seed_number)
```

*Keywords:*  Multivariate simulation, High-dimensional data, Nonparametric correlation, Gaussian copula, RNA-sequencing data, breast cancer

*=Corresponding author. aschissler@unr.edu. 1664 N Virginia St, Reno, NV 89557

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

# Introduction

Massive high-dimensional (HD) data sets are now commonplace in many areas of scientific inquiry. As new methods are developed for data, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of analyzing such proposed methodology --- such as false positive rates, statistical power, interval coverage, and robustness --- often with comparison to existing methods. Further, efficient simulation empowers statistical computing strategies, such as the parametric bootstrap [@Chernick2008] to simulate from a hypothesized null model, providing inference in analytically challenging settings. Such Monte Carlo (MC) techniques become difficult for high-dimensional data using existing algorithms and tools. This is particularly true when simulating massive multivariate, non-normal distributions, arising naturally in many fields of study.

As many have noted, it can be vexing to simulate dependent, non-normal/discrete data, even for low dimensional settings [@MB13; @XZ19]. For continuous non-normal multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula approaches [@Nelsen2007] are well-studied, with flexible, robust software available [@Yan2007; @Chen2001]. Yet these approaches do not scale in a timely fashion to high-dimensional problems [@Li2019gpu]. For discrete data, early simulation strategies had major flaws, such as failing to obtain the full range of possible correlations (e.g., admitting only positive correlations: see @Park1996). While more recent approaches [@MB13; @Xia17; @BF17] have largely remedied this issue for low-dimensional problems, the existing tools are not designed to scale to high dimensions.

Another central issue lies in characterizing dependence between components in the high-dimensional random vector. The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc). For normal data, the Pearson product-moment correlation describes the dependency perfectly. As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix is computationally intense [@Chen2001; @Xia17]. On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependence, such as Spearman's $\rho$ and Kendall's $\tau$. Throughout, we focus on matching these nonparametric dependence measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology. We do, however, provide Pearson matching with some caveats.

With all this in mind, we present a scalable, flexible multivariate simulation algorithm. The crux of the method lies in the construction of a Gaussian copula in the spirit of the NORTA procedure. Further, we introduce the `bigsimr` R package that provides high-performance software implementing our algorithm. The algorithm design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed-form relationships between dependence measures for the multivariate normal (MVN) distribution.

Our study proceeds by providing background information, including a description of a motivating example application: RNA-sequencing (RNA-seq) breast cancer data. Then we describe and justify our simulation methodology and related algorithms. Next, we detail an illustrative low-dimensional example of basic use of the `bigsimr` R package. Then we proceed with Monte Carlo studies under various bivariate distributional assumptions to assess accuracy. We conclude the Monte Carlo evaluations by summarizing the computation time for increasingly higher dimensional vectors. After the MC evaluations, we simulate random vectors motivated by our RNA-seq example, evaluate the accuracy, and provide example statistical computing tasks, namely MC estimation of joint probabilities and evaluating HD correlation estimation efficiency. Finally, we discuss the method's utility, limitations, and future directions.
```{r include=FALSE, cache=FALSE}
# Remove all objects
rm(list = ls(all.names = TRUE))

# Unload all packages
```

<!--chapter:end:index.Rmd-->

```{r include=FALSE, cache=FALSE}
options(digits = 4)

knitr::opts_chunk$set(
  # Text Results
  eval = TRUE,
  echo = FALSE,
  results = 'markup',
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  # Code Decoration
  comment = NA,
  background = '#F7F7F7',
  # Cache
  cache = TRUE,
  # Plots
  fig.show = "hold",
  fig.width = 6.5,
  fig.asp = 0.618,  # 1 / phi
  out.width = "85%",
  fig.align = 'center'
)

# options(dplyr.print_min = 6, dplyr.print_max = 6)

seed_number <- 2020-02-25
set.seed(seed = seed_number)
```
# Background {#background}

```{r ch010-background, include=FALSE, cache=FALSE}
box::use(
  dplyr[...],
  ggplot2[...],
  stringr[str_sub],
  GGally[ggpairs, wrap],
  knitr[kable]
)
load("data/example_genes.rda")
load("data/example_brca.rda")
d <- 1000
```

The `bigsimr` `R` package presented here provides multiple algorithms that operate with high-dimensional multivariate data; however, all these algorithms were originally designed to support a single task: to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics. Specifically, our goal is to efficiently simulate a large number, $B$, of HD random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with *correlated* components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i$.

When designing this methodology, we developed the following properties to guide our effort. We divide the properties into two categories: (1) basic properties (BP) and "scalability" properties (SP). The BPs are adapted from an existing criteria due to @Nik13a. A suitable simulation strategy should possess the following properties:

* BP1: A wide range of dependences, allowing both positive and negative values, and, ideally, admitting the full range of possible values.
* BP2: Flexible dependence, meaning that the number of bivariate marginals can be equal to the number of dependence parameters.
* BP3: Flexible marginal modeling, generating heterogeneous data --- possibly from differing probability families.

Moreover, the simulation method must *scale* to high dimensions:

* SP1: Procedure must scale to high dimensions with practical compute times.
* SP2: Procedure must scale to high dimensions while maintaining accuracy.

## Motivating example: RNA-seq data

Simulating high-dimensional, non-normal, correlated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. The RNA-seq data-generating process involves counting how often a particular form of human messenger RNA (mRNA) is expressed in a biological sample. RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data. For human-derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted [@Schissler2019]. Importantly, due to inherent biological processes, gene expression data exhibit correlation (co-expression) across genes [@BE07; @Schissler2018].

We illustrate our methodology using a well-studied Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments).
For simplicity, we only consider high expressing genes. In turn, we begin by filtering to retain the top `r d` of the highest-expressing genes (in terms of median expression) of the over 20,000 gene measurements from $N=`r nrow(example_brca)`$ patients' tumor samples. This gives a great number of pairwise dependencies among the marginals (specifically, $`r choose(d,2)`$ correlation parameters). Table \@ref(tab:ch010-realDataTab) displays RNA-seq counts for three selected high-expressing genes for the first five patients' breast tumor samples.

```{r ch010-realDataTab}
set.seed(2020-02-25)
num_genes    <- 3
num_patients <- 5
gene_sample  <- sample(example_genes[1:1000], num_genes)

cap <- paste0("mRNA expression for three selected high-expressing genes, ",
              paste(gene_sample, collapse = ", "), 
              ", for the first five patients in the TCGA BRCA data set.")

small_brca <- example_brca %>%
  select(all_of(gene_sample)) %>%
  head(n = num_patients) %>%
  as_tibble(rownames = "Patient ID") %>%
  mutate(`Patient ID` = str_sub(`Patient ID`, end = 12))

small_brca %>%
  kable(booktabs = TRUE, caption = cap)
```

To help visualize the bivariate relationships for these three selected genes across all patients, Figure \@ref(fig:ch010-realDataFig) displays the marginal distributions and estimated Spearman's correlations (see Equation \@ref(eq:spearman) below).

```{r ch010-realDataFig, cache=FALSE, fig.width=8, fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes from Table 1. The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations. Modeling these data motivate our simulation methodology."}
ggpairs(
  data = example_brca[, gene_sample],
  upper = list(continuous = wrap('cor', method = "spearman"))
) + theme_bw()
```

## Measures of dependency

In multivariate analysis, an analyst must select a metric to quantify dependency.
The most widely-known is the Pearson (product-moment) correlation coefficient that describes the linear association between two random variables $X$ and $Y$, and, it is given by

\begin{equation}
(\#eq:pearson)
\rho_P(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ \mathrm{Var}(X)\mathrm{Var}(Y)\right]^{1/2}}.
\end{equation}

As @MB13 and @MK01 discuss, for a bivariate normal $(X,Y)$ random vector, the Pearson correlation completely describes the dependency between the components. For non-normal marginals with monotone correlation patterns, however, $\rho_P$ suffers some drawbacks and may mislead or fail to capture important relationships [@MK01]. Alternatively in these settings, analysts often prefer rank-based correlation measures to describe the degree of monotonic association.

Two nonparametric, rank-based measures common in practice are Spearman's correlation (denoted $\rho_S$) and Kendall's $\tau$. Spearman's $\rho_S$ has an appealing correspondence as the Pearson correlation coefficient on *ranks* of the values, thereby captures nonlinear yet monotone relationships. Kendall's $\tau$, on the other hand, is the difference in probabilities of concordant and discordant pairs of observations, $(X_i, Y_i)$ and $(X_j, Y_j)$, with concordance meaning that orderings have the same direction (e.g., if $X_i < X_j$, then $Y_i < Y_j$). Note that concordance is determined by the ranks of the values, not the values themselves.

Both $\tau$ and $\rho_S$ are *invariant under monotone transformations* of the underlying random variates. As we will describe more fully in the [Algorithms](#algorithms) section, this property enables matching rank-based correlations with speed (SP1) and accuracy (SP2).

*Correspondence among Pearson, Spearman, and Kendall correlations*

There is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficient, as the marginal distributions $F_i$ are intrinsic in their calculation. For *bivariate normal vectors*, however, the correspondence is well-known:

\begin{equation}
(\#eq:convertKendall)
\rho_{P} = \sin \left( \tau \times \frac{\pi}{2} \right), 
\end{equation}

\noindent and similarly for Spearman's $\rho$ [@K58],

\begin{equation}
(\#eq:convertSpearman)
\rho_P = 2 \times \sin \left( \rho_S \times \frac{\pi}{6} \right).
\end{equation}

*Discrete marginal considerations*

Spearman's correlation for discrete marginal suffers issues whenever there is a nonzero probability of ties. For example, the Spearman's correlation of a discrete-valued random variable $X$ with itself could be less than 1 [@MB13]. One remedy for this issue is to rescale Equation \@ref(eq:spearman). For two random variables $X,Y$ with probability mass functions (PMFs) or probability densities functions (PDFs) $p(x)$ and $q(y)$, respectively, define the rescaled Spearman's correlation as

\begin{equation}
(\#eq:spearmanRescaled)
\rho_{RS}(X,Y) = \frac{\rho_s(X,Y)}{ \left[ \left[ 1 - \sum_x p(x)^3 \right] \left[ 1 - \sum_y q(y)^3 \right] \right]^{1/2}}.
\end{equation}

Note that with continuous marginals the rescaling returns $\rho_S$. For discrete marginals with large or infinite support, computing the adjustment factors $\sum_x p(x)^3$, $\sum_y p(y)^3$ over all large number of pairs becomes expensive (often violating desired property SP1). Further, the infinite sums must be approximated for count-valued data.

Also, for a discrete random variable $Y_i$, some care must be taken to define the quantile function $F_{i}^{-1}$. A standard practice is to set

\begin{equation}
F_{i}^{-1} = \inf\{y:F_{i}(y) \geq u \}.
(\#eq:inverseCDF)
\end{equation}

*Marginal-dependent bivariate correlation bounds*

Given two marginal distributions, $\rho_P$ is not free to vary over the entire range of possible correlations $[-1,1]$. The so-called *Frechet-Hoeffding bounds* are well-studied [@Nelsen2007; @BF17]. These constraints cannot be overcome through algorithm design. In general, the bounds are given by

\begin{equation}
(\#eq:frechet)
\rho_P^{max} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (U) \right), \quad \rho_P^{min} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (1 - U) \right)
\end{equation}

\noindent where $U$ is a uniform random variable on $(0,1)$ and $F^{-1}_1, F^{-1}_2$ are the inverse CDFs of $X_1$ and $X_2$, respectively, definitely by \@ref(eq:inverseCDF) when the variables are discrete.

## Gaussian copulas

There is a strong connection of our simulation strategy to Gaussian *copulas* (see @Nelsen2007 for a technical introduction). A copula is a distribution function on $[0,1]^d$ that describes a multivariate probability distribution with standard uniform marginals. This provides a powerful, natural way to characterize joint probability structures. Consequently, the study of copulas is an important and active area of statistical theory and practice.

For any random vector ${\bf X}=(X_1, \ldots, X_d)^\top$ with CDF $F$ and marginal CDFs $F_i$ there is a copula function $C(u_1, \ldots, u_d)$ satisfying

\begin{equation}
(\#eq:copula)
F(x_1, \ldots,x_d) = \mathbb P(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), 
\end{equation}

$x_i \in \mathbb R, i=1,\ldots,d.$

A Gaussian copula has marginal CDFs that are all standard normal, $F_i = \Phi, \forall \, i$. This representation corresponds to a multivariate normal (MVN) distribution with standard normal marginal distributions and covariance matrix ${\bf R_P}$. As the marginals are standardized to have unit variance, however, ${\bf R_P}$ is a Pearson correlation matrix. If $F_{{\bf R}}$ is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula $C_{{\bf R}}$ is defined through

\begin{equation}
(\#eq:gauss)
F_{{\bf R}}(x_1, \ldots, x_d) = C_{{\bf R}}(\Phi(x_1), \ldots, \Phi(x_d)),
\end{equation}

where $\Phi(\cdot)$ is the standard normal CDF. Note that the copula $C_{{\bf R}}$ is the familiar multivariate normal CDF of the random vector $(\Phi(X_1), \ldots, \Phi(X_d))$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})$.

Sklar's Theorem [@Sklar1959; @Ubeda-Flores2017] guarantees that given inverse CDFs $F_i^{-1}$s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions. For example, using Gaussian copulas, we can construct a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)^\top$ with $Y_i \sim F_i$, viz. $Y_i = F_i^{-1}(\Phi(X_i)), i=1, \ldots, d$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})$.
```{r include=FALSE, cache=FALSE}
# Remove all objects
rm(list = ls(all.names = TRUE))

# Unload all packages
```

<!--chapter:end:010-background.Rmd-->

```{r include=FALSE, cache=FALSE}
options(digits = 4)

knitr::opts_chunk$set(
  # Text Results
  eval = TRUE,
  echo = FALSE,
  results = 'markup',
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  # Code Decoration
  comment = NA,
  background = '#F7F7F7',
  # Cache
  cache = TRUE,
  # Plots
  fig.show = "hold",
  fig.width = 6.5,
  fig.asp = 0.618,  # 1 / phi
  out.width = "85%",
  fig.align = 'center'
)

# options(dplyr.print_min = 6, dplyr.print_max = 6)

seed_number <- 2020-02-25
set.seed(seed = seed_number)
```
# Algorithms {#algorithms}

This section describes our methods involved in simulating a random vector $\bf Y$ with $Y_i$ components for $i=1,\ldots,d$. Each $Y_i$ has a specified marginal CDF $F_i$ and its inverse $F^{-1}_i$. To characterize dependency, every pair $(Y_i, Y_j)$ has a given Pearson correlation $\rho_P$, Spearman correlation $\rho_S$, and/or Kendall's $\tau$. The method can be described as a *high-performance Gaussian copula* (Equation \@ref(eq:gauss)) providing a high-dimensional NORTA-inspired algorithm.

## NORmal To Anything (NORTA)

The well-known NORTA algorithm [@Cario1997] simulates a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$. Specifically, NORTA algorithm proceeds as follows:

1. Simulate a random vector $\bf Z$ with $d$ *independent and identically distributed* (iid) standard normal components.
2. Determine the input matrix $\Sigma_{\bf Z}$ that corresponds with the specified output $\Sigma_{\bf Y}$.
3. Produce a Cholesky factor $M$ of $\Sigma_{\bf Z}$ such that $M M^{\prime}=\Sigma_{\bf Z}$.
4. Set $X$ by $X \gets MZ$.
5. $\text{Return} \; Y \; \text{where} \; Y_i \gets F_i^{-1}[\Phi(X_i)], \; i=1,...,d$.

With modern parallel computing, steps 1, 3, 4, 5 are readily implemented as high-performance, multi-core and/or graphical-processing-unit (GPU) accelerated algorithms --- providing fast scalability using readily-available hardware.

Matching specified Pearson correlation coefficients exactly (step 2 above), however, is computationally costly. In general, there is no closed-form correspondence between the components of the input $\Sigma_{\bf Z}$ and target $\Sigma_{\bf Y}$. Matching the correlations involves evaluating or approximating $\binom{d}{2}$ integrals of the form

\begin{equation}
    \mathrm{E}\left[Y_i Y_j\right] = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} F_i^{-1}\left[\Phi(z_i)\right] F_j^{-1}\left[\Phi(z_j)\right] \phi(z_i, z_j, \rho_z) dz_i dz_j,
    (\#eq:pearsonIntegralRelation)
\end{equation}

where $\phi(\cdot)$ is the joint probability distribution function of two correlated standard normal variables. For HD data, (nearly) exact evaluation may become too costly to enable practical simulation studies. For low-dimensional problems, however, methods and tools exist to match Pearson correlations precisely; see [@Xia17] and the publicly available `nortaRA` R package [@Chen2001]. As described later, to enable HD Pearson matching we approximate these integrals.

*NORTA in higher dimensions*

Sklar's theorem provides a useful characterization of multivariate distributions through copulas. Yet the choice of copula-based simulation algorithm affects which joint distributions may be simulated. Even in low-dimensional spaces (e.g., $d=3$), there exist valid multivariate distributions with *feasible* Pearson correlation matrices that NORTA cannot match exactly [@LH75]. This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined the resultant matrix is indefinite. These situations do occur, even using exact analytic calculations. Such problematic target correlation matrices are termed *NORTA defective*.

@GH02 conducted a Monte Carlo study to estimate the probability of encountering NORTA defective matrices while increasing the dimension $d$. They found that for what is now considered low-to-moderate dimensions ($d \approx 20$), almost *all* feasible matrices are NORTA defective. This stems from the concentration of measure near the boundary of the space of all possible correlation matrices as dimension increases. Unfortunately, it is precisely near this boundary that NORTA defective matrices reside.

There is hope, however, as @GH02 also showed that replacing an indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$. This provides evidence that our nearest positive definite (PD) augmented approach will maintain reasonable accuracy if our input matching scheme returns an indefinite matrix, at least for the rank-based matching scheme described above.

## Random vector generator {#rand-vec-gen}

We now describe our algorithm to generate random vectors, which mirrors the classical NORTA algorithm above with some modifications for rank-based dependency matching:

1. Mapping step
    (i) Convert the target Spearman correlation matrix $R_S$ to the corresponding MVN Pearson correlation $R_X$. Alternatively,  
    (i') Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$. Alternatively,  
    (i'') Convert the target Pearson correlation matrix to $R_P$ to the corresponding, approximate MVN Pearson correlation $R_X$.  
2. Check admissibility and, if needed, compute the nearest correlation matrix.
    (i) Check that $R_X$ is a correlation matrix, a positive definite matrix with 1's along the diagonal.   
	(ii) If $R_X$ is a correlation matrix, the input matrix is *admissible* in this scheme. Otherwise
    (ii$^\prime$) Replace $R_X$ with the nearest correlation matrix $\tilde{R}_X$, in the Frobenius norm.  
3. Gaussian copula
    (i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)$.  
    (ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz. $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
    (iii) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$.

*Step 1*

The first two descriptions of the *Mapping step* employ the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertKendall) and \@ref(eq:convertSpearman), respectively (implemented as `cor_covert`). Initializing our algorithm to match the nonparametric correlations by computing these equations for all pairs is computationally trivial.

For the computationally expensive process to match Pearson correlations, we approximate Equation \@ref(eq:pearsonIntegralRelation) for all pairs of margins. To this end, we implement the approximation scheme introduced by @xiao2019matching. Briefly, the many double integrals of the form in \@ref(eq:pearsonIntegralRelation) are approximated by weighted sums of Hermite polynomials. Matching coefficients for pairs of continuous distributions is made tractable by this method, but for discrete distributions (especially discrete distributions with large support sets or infinite support), the problem is not so simple or efficient.

Our solution is to approximate discrete distributions by a continuous distribution. The question then becomes that of which distribution to use. We found that Generalized S-Distributions [@muino2006gs] solve this problem by approximating a wide range of unimodal distributions, both continuous and discrete. Using the GS-Distribution approximation of discrete distributions in Pearson matching scheme above yields favorable results.

*Step 2*

Once $R_X$ has been determined, we check admissibility of the adjusted correlation via the steps described above. If $R_X$ is not a valid correlation matrix, then we compute the nearest correlation matrix. Finding the nearest correlation matrix is a common statistical computing problem. The defacto default function in `R` is `Matrix::nearPD`, an alternating projection algorithm due to @higham2002computing. As implemented the function fails to scale to HD. Instead, we provide the quadratically convergent algorithm based on the theory of strongly semi-smooth matrix functions (@qi2006quadratically). The nearest correlation matrix problem can be written down as the following convex optimization problem:

\begin{align*}
    \mathrm{min}\quad & \frac{1}{2} \Vert G - X \Vert^2 \\
    \mathrm{s.t.}\quad & X_{ii} = 1, \quad i = 1, \ldots , n, \\
    & X \in S_{+}^{n}
\end{align*}

For nonparametric correlation measures, our algorithm allows the generation of high-dimensional multivariate data with arbitrary marginal distributions with a broad class of admissible Spearman correlation matrices and Kendall $\tau$ matrices. The admissible classes consist of the matrices that map to a Pearson correlation matrix for a MVN. In particular, if we let $X$ be MVN with $d$ components and set

\begin{align*}
\Omega_P &= \{ R_P : R_P \textrm{ is a correlation matrix for } X \} \\
\Omega_K &= \{ R_K : R_K \textrm{ is a correlation matrix for } X \} \\
\Omega_S &= \{ R_S : R_S \textrm{ is a correlation matrix for } X \} \\
\end{align*}

then there are 1-1 mappings between these sets. We conjecture informally that the sets of admissible $R_S$ and $R_K$ are not highly restrictive. In particular, $R_P$ is approximately $R_S$ for a MVN, suggesting that the admissible set $\Omega_S$ should be flexible as $R_P$ can be any PD matrix with 1's along the diagonal. We provide methods to check whether a target $R_S$ is an element of the *admissible set* $\Omega_S$ (and similarly for $R_K$).

There is an increasing probability of encountering a non-admissible correlation matrix as dimension increases. In our experience, the mapping step for large $d$ almost always produces a $R_X$ that is not a correlation matrix. In Section \@ref(package) we provide a basic description of how to quantify and control the approximation error. Further, the RNA-seq example in Section \@ref(examples) provides an illustration of this in practice.

*Step 3*

The final step implements a NORTA-inspired, Gaussian copula approach to produce the desired margins. Steps 1 and 2 determine the MVN Pearson correlation values that will eventually match the target correlation. Then all that is required is a fast MVN simulator, a standard normal CDF, and well-defined quantile functions for marginals. The MVN is transformed to a copula (distribution with standard uniform margins) by applying the normal CDF $\phi(\cdot)$. The quantile functions $F^{-1}$ are applied across the margins to return the desired random vector ${\bf Y}$.

```{r include=FALSE, cache=FALSE}
# Remove all objects
rm(list = ls(all.names = TRUE))

# Unload all packages
```

<!--chapter:end:020-simulation-algorithm.Rmd-->

```{r include=FALSE, cache=FALSE}
options(digits = 4)

knitr::opts_chunk$set(
  # Text Results
  eval = TRUE,
  echo = FALSE,
  results = 'markup',
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  # Code Decoration
  comment = NA,
  background = '#F7F7F7',
  # Cache
  cache = TRUE,
  # Plots
  fig.show = "hold",
  fig.width = 6.5,
  fig.asp = 0.618,  # 1 / phi
  out.width = "85%",
  fig.align = 'center'
)

# options(dplyr.print_min = 6, dplyr.print_max = 6)

seed_number <- 2020-02-25
set.seed(seed = seed_number)
```
# The `bigsimr` R package {#package}

```{r ch030-basic-example-setup, echo=TRUE, cache=FALSE}
box::use(
  patchwork[...],
  dplyr[...],
  ggplot2[...],
  tidyr[drop_na],
  bigsimr[bigsimr_setup, distributions_setup],
  JuliaCall[julia_call]
)

Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
bs <- bigsimr_setup(pkg_check = FALSE)
dist <- distributions_setup()
```

This section describes a low-dimensional (2D) random vector simulation workflow via the `bigsimr` `R` package (https://github.com/SchisslerGroup/r-bigsimr).
The package `bigsimr` provides an interface to the native code written in Julia, registered as the `Bigsimr` Julia package (https://github.com/adknudson/Bigsimr.jl).
In addition to the native Julia `Bigsimr` package and `R` interface `bigsimr`, we also provide a python interface `bigsimr` (https://github.com/SchisslerGroup/python-bigsimr/) that interfaces with the Julia `Bigsimr` package. 
The Julia implementation provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms)).
When designing `bigsimr`, we aimed to conveniently provide parallelized computation, through multi-core and GPU acceleration, while allowing advanced users to customize.

The subsections below describe the basic use of `bigsimr`, by stepping through an example workflow using the data set `airquality` that contains daily air quality measurements in New York, May to September 1973 [@Chambers1983].
This workflow proceeds from computing environment setup, data wrangling, estimation, simulation configuration, random vector generation, and, finally, result visualization.
The code chunk below prepares the computing environment:

<!-- 
`bigsimr` is an R package for simulating high-dimensional multivariate data with arbitrary marginal distributions. The efficiency behind generating multivariate samples comes from the ability to utilize a GPU during the generation of multivariate normal samples and the subsequent transformation to uniform samples (see section \@ref(rand-vec-gen)). A GPU is not necessary in order to use `bigsimr`, and the speed benefits generally only come when simulating very high dimensional data ($d > 10000$).

The next computational step after obtaining the correlated uniform marginals is the inverse transform into the target marginal distributions. Since this consists of $d$ independent transformations, we utilize parallelization to achieve higher throughput. As with many parallel algorithms, there is overhead associated with forking the task to utilize multiple cores, however cost is negligible compared to the total run-time.
-->

## Bivariate example description

We illustrate the use of `bigsimr` motivated by the New York air quality data set (`airquality`) included in the R `datasets` package.
First, we load the `bigsimr` library and a few other convenient data science packages, including the syntactically-elegant `tidyverse` suite of `R` packages.

For simplicity and to provide a minimal working example, we consider bivariate simulation of the two `airquality` variables `Temperature`, in degrees Fahrenheit, and `Ozone` level, in parts per billion.

```{r ch030-air-quality-data, echo=TRUE}
df <- airquality %>%
  select(Temp, Ozone) %>%
  drop_na()
```

```{r ch030-aq-glimpse}
glimpse(df)
```

Figure \@ref(fig:ch030-aq-joint-dist) visualizes the bivariate relationship between Ozone and Temperature. We aim to simulate random two-component vectors mimicking this structure. The margins are not normally distributed; particularly the Ozone level exhibits a strong positive skew.

```{r ch030-aq-joint-dist, cache=FALSE, fig.width= 8, fig.cap="Bivariate scatterplot of Ozone vs. Temp with estimated marginal densities. We model the Ozone data as marginally log-normal and the Temperature data as normal."}
p0 <- ggplot(df, aes(Temp, Ozone)) +
  geom_point(size = 1) +
  theme(legend.position = "none") + 
  labs(x = "Temperature")

pTemp <- ggplot(df, aes(Temp)) + 
  geom_density() +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())

pOzone <- ggplot(df, aes(Ozone)) + 
    geom_density() +
  theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank()) +
  coord_flip()

pTemp + plot_spacer() + p0 + pOzone + 
  plot_layout(widths = c(3,1), heights = c(1, 3)) 
```

Next, we specify the marginal distributions and correlation coefficient (both type and magnitude). Here the analyst is free to be creative. For this example, we avoid goodness-of-fit considerations to determine the marginal distributions. But it seems sensible without domain knowledge to estimate these quantities from the data and `bigsimr` contains fast functions designed for this task.

## Specifying marginal distributions

Based on the estimated densities in Figure \@ref(fig:ch030-aq-joint-dist), we assume `Temp` is normally distributed and `Ozone` is log-normally distributed, as the latter values are positive and skewed. We use the well-known, unbiased estimators for the normal distribution's parameters and maximum likelihood estimators for the log-normal parameters:

```{r ch030-aq-temp-pars, echo=TRUE}
df %>% 
  select(Temp) %>% 
  summarise_all(.funs = c(mean = mean, sd = sd))
```

```{r ch030-aq-ozone-pars, echo=TRUE}
mle_mean <- function(x) mean(log(x))
mle_sd <- function(x) mean( sqrt( (log(x) - mean(log(x)))^2 ) )

df %>% 
  select(Ozone) %>% 
  summarise_all(.funs = c(meanlog = mle_mean, sdlog = mle_sd))
```

Next, we configure the input marginals for later input into `rvec`. The marginal distributions are specified using `Julia`'s `Distributions` package and stored in a vector.

```{r ch030-margins-alist, echo=TRUE}
margins <- c(
  dist$Normal(mean(df$Temp), sd(df$Temp)),
  dist$LogNormal(mle_mean(df$Ozone), mle_sd(df$Ozone))
)
```

## Specifying correlation

As mentioned, the user must decide how to describe correlation, based on the particulars of the problem. For non-normal data and for improved simulation accuracy/scalability in our scheme, we advocate the use of Spearman's $\rho$ correlation matrix $R_S$ and Kendall's $\tau$ correlation matrix $R_K$. We also support Pearson correlation coefficient matching, while cautioning the user to check the performance for their parametric multivariate model (see [Monte Carlo evaluations](#simulations) below for evaluation strategies and guidance). These estimation methods are classical approaches, not designed for high-dimensional correlation estimation (see the [Conclusion and Discussion]({#discussion) sections for more on this).

```{r ch030-aq-cor, echo=TRUE}
(R_S <- bs$cor(as.matrix(df), bs$Spearman))
```

## Checking target correlation matrix admissibility

First we use `cor_bounds` to ensure that the pairwise target correlation values are valid prior to mapping step which constructs $R_X$ for the MVN input into NORTA (see Section \@ref(algorithms)). `cor_bounds` estimates the pairwise lower and upper correlation bounds using the Generate, Sort, and Correlate algorithm of @DH2011.

```{r ch030-cor-bounds, echo=TRUE}
bounds <- bs$cor_bounds(margins)
bounds$lower
```

Since our single estimated Spearman correlation is within the theoretical bounds, the correlation is valid as input to `rvec`. But even if the 2-dimensional bounds are satisfied for each pair of margins, this does not guarantee the feasibility of a $d-$variate distribution [@BF17].

To provide higher dimensional feasibility/admissibility checking, we begin by mapping using `cor_convert` and check admissibility.

```{r ch030-cor-convert, echo=TRUE}
# Step 1. Mapping
(R_X <- bs$cor_convert(R_S, bs$Spearman, bs$Pearson))

# Step 2. Check admissibility
julia_call("Bigsimr.iscorrelation", R_X)
```

The bounds on the Pearson correlation coefficient $R_P$ between these margins are restricted as seen below in our MC estimated correlation bounds:

```{r ch030-cor-bounds-pearson, echo=TRUE}
bs$cor_bounds(margins[1], margins[2], bs$Pearson, n_samples = 1e6)
```

Our MC estimate of the bounds slightly underestimates the theoretic bounds of $(-0.881, 0.881)$. An analytic derivation is presented as an Appendix.

## Simulating random vectors
Finally, we arrive at the main function of `bigsimr`, `rvec`. We now simulate $B=10,000$ random vectors from the assumed joint distribution of Ozone levels and Temp.

```{r ch030-sim-margins, echo=TRUE}
x <- bs$rvec(10000, R_X, margins)
df_sim <- as.data.frame(x)
colnames(df_sim) <- colnames(df)
```

Figure \@ref(fig:ch030-plot-sim) plots the 10,000 simulated points.

```{r ch030-plot-sim, fig.cap="Contour plot and marginal densities for the simulated bivariate distribution of Air Quality Temperatures and Ozone levels. The simulated points mimic the observed data with respect to both the marginal characteristics and bivariate association."}
p1 <- df_sim %>%
  ggplot(aes(Temp, Ozone)) +
  geom_density_2d_filled() +
  theme(legend.position = "none") +
  labs(x = "Simulated Temperature", y = "Simulated Ozone") +
  scale_y_continuous(limits = c(0, max(df$Ozone))) +
  scale_x_continuous(limits = range(df$Temp))

p1Temp <- ggplot(df_sim, aes(Temp)) + 
  geom_density(alpha = 0.5, fill = "lightseagreen") +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())

p1Ozone <- ggplot(df_sim, aes(Ozone)) + 
  geom_density(alpha = 0.5, fill = "lightseagreen") + 
  theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank()) +
  coord_flip()

p1Temp + plot_spacer() + p1 + p1Ozone + 
  plot_layout(widths = c(3,1), heights = c(1, 3))
```
```{r include=FALSE, cache=FALSE}
# Remove all objects
rm(list = ls(all.names = TRUE))

# Unload all packages
```

<!--chapter:end:030-bigsimr-package.Rmd-->

```{r include=FALSE, cache=FALSE}
options(digits = 4)

knitr::opts_chunk$set(
  # Text Results
  eval = TRUE,
  echo = FALSE,
  results = 'markup',
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  # Code Decoration
  comment = NA,
  background = '#F7F7F7',
  # Cache
  cache = TRUE,
  # Plots
  fig.show = "hold",
  fig.width = 6.5,
  fig.asp = 0.618,  # 1 / phi
  out.width = "85%",
  fig.align = 'center'
)

# options(dplyr.print_min = 6, dplyr.print_max = 6)

seed_number <- 2020-02-25
set.seed(seed = seed_number)
```
# Monte Carlo evaluations {#simulations}

```{r ch040-LoadLib040, include=FALSE}
box::use(
    dplyr[...],
    ggplot2[...],
    patchwork[...],
    tidyr[pivot_longer],
    bigsimr[bigsimr_setup, distributions_setup],
    knitr[kable],
    RColorBrewer[brewer.pal]
)
load("data/bivariate_normal_sims.rda")
load("data/bivariate_gamma_sims.rda")
load("data/bivariate_nbinom_sims.rda")
load("data/benchmark_dependences.rda")

Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
bs <- bigsimr_setup(pkg_check = FALSE)
dist <- distributions_setup()
prob <- 3e-04
```

Before applying our methodology to real data simulation, we conduct several Monte Carlo studies to investigate method performance. Since marginal parameter matching in our scheme is essentially a sequence of univariate inverse probability transforms, the challenging aspects are the accuracy of dependency matching and computational efficiency at high dimensions. To evaluate our methods in those respects, we design the following numerical experiments to first assess accuracy of matching dependency parameters in bivariate simulations and then time the procedure in increasingly large dimension $d$.

## Bivariate experiments

We select bivariate simulation configurations to ultimately simulate our motivating discrete-valued RNA-seq example, and, so we proceed in increasing complexity, leading to the model in our motivating application in Section \@ref(examples). We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson's, Spearman's, and Kendall's --- in identical, bivariate marginal configurations. For each pair of identical margins, we vary the target correlation across $\Omega$, the set of possible admissible values for correlation type, to evaluate the simulation's ability to obtain the theoretic bounds. The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (mimicking RNA-seq counts).

Table \@ref(tab:sims) lists our identical-marginal, bivariate simulation configurations. We increase the simulate replicates $B$ to check that our results converge to the target correlations and gauge statistical efficiency. We select distributions beginning with a standard multivariate normal (MVN) as we expect the performance to be exact (up to MC error) for all correlation types. Then, we select a non-symmetric continuous distribution: a standard (rate =1) two-component multivariate gamma (MVG). Finally, we select distributions and marginal parameter values that are motivated by our RNA-seq data, namely values proximal to probabilities and sizes estimated from the data (see [Example applications](examples) for estimation details). Thus we arrive at a multivariate negative binomial (MVNB) $p_1 = p_2 = 3\times10^{-4}, r_1 = r_2 = 4, \rho \in \Omega$.

Table: (\#tab:sims) Identical margin, bivariate simulation configurations to evaluate correlation matching accuracy and efficiency.

| Simulation Reps ($B$) | Correlation Types | Identical-margin 2D distribution |
|-------------|:--------------:|----------------------:|
| $1000$    | Pearson ($\rho_P$) | ${ \bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho_i ), i=1,\ldots,100$ |
| $10,000$  | Spearman ($\rho_S$) | ${ \bf Y} \sim MVG( shape = 10, rate = 1, \rho_i ), i=1,\ldots,100$ |
| $100,000$ | Kendall ($\tau$)| ${ \bf Y} \sim MVNB(p = 3\times10^{-4}, r = 4,\rho_i), i=1,\ldots,100$ |

For each of the unique 9 simulation configurations described above, we estimate the correlation bounds and vary the correlations along a sequence of 100 points evenly placed within the bounds, aiming to explore $\Omega$. Specifically, we set correlations $\{ \rho_1 = ( \hat{l} + \epsilon), \rho_2 = (\hat{l} + \epsilon) + \delta, \ldots, \rho_{100} = (\hat{u} - \epsilon) \}$, with $\hat{l}$ and $\hat{u}$ being the estimated lower and upper bounds, respectively, and increment value $\delta$. The adjustment factor, $\epsilon=0.01$, is introduced to handle numeric issues when the bound is specified exactly.

```{r ch040-biNormPlot, eval=FALSE, fig.height=5, fig.width=8, fig.cap="`bigsimr` recovers the Pearson specified correlations for MVN."}
bivariate_normal_sims %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    geom_point() +
    geom_abline(slope = 1) +
    facet_wrap(~ type + N) + 
    theme_bw()
```

```{r ch040-biGammaPlot, eval=FALSE, fig.height=5, fig.width=8, fig.cap="`bigsimr` recovers the Pearson specified correlations for Bivariate Gamma."}
bivariate_gamma_sims %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    geom_point() +
    geom_abline(slope = 1) +
    facet_wrap(~ type + N) + 
    theme_bw()
```

```{r ch040-biNegBinPlot, eval=FALSE, fig.width=8, fig.height=5, fig.cap="`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
bivariate_nbinom_sims %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    geom_point() +
    geom_abline(slope = 1) +
    facet_wrap(~ type + N) + 
    theme_bw()
```

```{r ch040-combineBiSims, cache=FALSE}
allDat <- bind_rows(
    select(bivariate_normal_sims, margins, type, N, rho, rho_hat),
    select(bivariate_gamma_sims, margins, type, N, rho, rho_hat),
    select(bivariate_nbinom_sims, margins, type, N, rho, rho_hat)
) %>%
    mutate(margins = factor(margins, levels=c("norm", "gamma", "nbinom")),
           type = factor(type, levels=c("Pearson", "Spearman", "Kendall")),
           N = factor(N))
```

Figure \@ref(fig:ch040-bPlot) displays the aggregated bivariate simulation results. Table \@ref(tab:ch040-BiError) contains the mean absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.

```{r ch040-BiError}
tabMAE <- allDat %>%
    group_by(N, type, margins) %>%
    summarize(MAE = mean(abs(rho - rho_hat))) %>%
    ungroup()

kable(tabMAE, booktabs = TRUE,
      col.names = c("No. of random vectors",
                    "Correlation type",
                    "Distribution",
                    "Mean abs. error"),
      caption = "Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.")
```

Overall, the studies show that our methodology is generally accurate across the entire range of possible correlation values all three dependency measures, at least in these limited simulation settings for the rank-based correlations. Our Pearson matching performs nearly as well as Spearman or Kendall, except a slight increase in error for negative binomial case. This is due the particularly large counts generated with our choice of parameters for $p$ and $r$.

```{r ch040-bPlot, cache=FALSE, fig.asp=1.30, fig.weight=4, fig.cap="Bivariate simulations match target correlations across the entire range of feasible correlations. The horizontal axis plots the specified target correlations for each bivariate margin. Normal margins are plotted in dark dark grey, gamma in medium grey, and negative binomial in light grey. As the number of simulated vectors $B$ increases from left to right, the variation in estimated correlations (vertical axis) decreases. The dashed line indicates equality between the specified and estimated correlations."}
# https://www.datanovia.com/en/blog/how-to-change-ggplot-facet-labels/
# New facet label names
repsLabs <- paste0("B=", c("1,000", "10,000", "100,000") )
names(repsLabs) <- c(1000, 10000, 100000)
typeLabs <- c("Pearson", "Spearman", "Kendall")
names(typeLabs) <- c("Pearson", "Spearman", "Kendall")

# Set colors
## RColorBrewer::display.brewer.all()
numColors <- 4
numGroups <- length(levels(allDat$margins))
myColors <- rev(RColorBrewer::brewer.pal(n = numColors, name = "Greys")[ ((numColors - numGroups) + 1):numColors])

allDat %>%
    ggplot(aes(x = rho, y = rho_hat, color = margins)) +
    geom_point(size = 2) +
    scale_color_manual(values = myColors ) +
    geom_abline(slope = 1, linetype = "dashed") +
    labs(x = "Specified Correlation", y = "Estimated Correlation") +
    facet_wrap(~ type + N, labeller = labeller(N = repsLabs, type = typeLabs)) +
    theme_bw() +
    theme(legend.position = "bottom", legend.direction = "horizontal")
```

<!-- 
See [Discussion](discussion) for future directions for fast Pearson matching and discrete-specific modifications.
-->

## Scale up to High Dimensions

With information of our method's accuracy from a low-dimensional perspective, we now turn to assessing whether the `bigsimr` can scale to larger dimensional problems with practical computation times. Specifically, we ultimately generate $B=1,000$ random vectors for $d=\{100, 250, 500, 1000, 2500, 5000, 10000\}$ for each correlation type, $\{Pearson, Spearman, Kendall\}$ while timing the algorithm's major steps.

To mimic the workflow, we first produce a synthetic "data set" by completing the following steps:

1. Produce heterogeneous gamma marginals by randomly selecting the $j^{th}$ gamma shape parameter from $U_j \sim uniform(1,10), j=1,\ldots,d$ and the $j^{th}$ rate parameter from $V_j \sim exp(1/5), j=1,\ldots,d$, with the constant parameters determined arbitrarily.  
2. Produce a random full-rank Pearson correlation matrix via `cor_randPD` of size $d \times d$.  
3. Simulate a "data set" of $1,000 \times d$ random vectors via `rvec` (without matching the Pearson exactly).

With the synthetic data set in hand, we complete and time the following 4 steps:

i. Estimate the correlation matrix from the "data" in the *Compute Correlation* step.  
1. Map the correlations to initialize the algorithm (Pearson to Pearson, Spearman to Pearson, or Kendall to Pearson) in the *Adjust Correlation* step.  
2. Check whether the mapping produces a valid correlation matrix and, if not, find the nearest PD correlation matrix in the *Check Admissibility* step.  
3. Simulate $1,000$ vectors via the Inverse Transform method in the *Simulate Data* step.

The experiments are conducted on a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, with all 16 threads employed during computation. The results for $d \leq 500$ are fast with all times under 3 seconds. Table \@ref(tab:ch040-moderateDtab) displays the total computation time of the algorithms steps 1 through 3, including estimation step i.

```{r ch040-moderateDtab}
dat_long <- benchmark_dependences %>%
  select(-total_time, -n_sim, -needed_near_pd) %>%
  pivot_longer(cols = c(corr_time:sim_time), names_to = "Step", values_to = "Time") %>%
  mutate(dim = factor(dim),
         Step = factor(Step,
                       levels = c("corr_time", "adj_time", "admiss_time", "sim_time"),
                       labels = c("Compute Correlation",
                                  "Adjust Correlation",
                                  "Check Admissibility",
                                  "Simulate Data"))) %>%
    rename(Correlation = corr_type, Dimensions = dim)

tabTimes  <- dat_long %>%
    filter(Dimensions %in% c(100, 250, 500)) %>%
    group_by(Dimensions, Correlation) %>%
    summarize('Total Time(Seconds)' = sum(Time)) %>%
    ungroup()

tabTimes %>%
    kable(booktabs = TRUE,
          col.names = c("Dimension",
                        "Correlation type",
                        "Total Time (Seconds)"),
          caption = 'Total time to produce 10,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.')
```

The results larger dimensional vectors show scalability to ultra-high dimensions for all three correlation types, although the total times do become much larger. 
Figure \@ref(fig:ch040-largeDfig) displays computation times for $d=\{1000, 2500, 5000, 10000\}$. For $d$ equal to 1000 and 2500, the total time is under a couple of minutes. At $d$ of 5000 and 10,000, Pearson correlation matching in the *Adjust Correlation* step becomes costly. Interestingly, Pearson is actually faster than Kendall for $d=10,000$ due to bottlenecks in *Compute Correlation* and *Check Admissibility*. Uniformly, matching Spearman correlations is faster, with total times under 5 minutes for $d=10,000$, making Spearman correlations the most computationally-friendly dependency type. With this in mind, we scaled the simulation to $d=20,000$ for the Spearman type and obtained the 1,000 vectors in under an hour (data not shown). In principle, this would enable the simulation of an entire human-derived RNA-seq data set. We note that for a given target correlation matrix and margins, steps i, 1, and 2 only need to be computed once and the third step, *Simulate Data*, is fast under all schemes for all $d$ under consideration.

```{r ch040-largeDfig, fig.cap="Computation times as d increases."}
# Set colors
numColors <- 5
numGroups <- 4
myColors <- rev(brewer.pal(n=numColors, name="Greys")[((numColors-numGroups)+1):numColors])

dat_long %>%
    filter(Dimensions %in% c(1000, 2500, 5000, 10000, 20000)) %>%
    mutate(`Time (minutes)` = Time / 60) %>%
    ggplot(aes(Correlation, `Time (minutes)`, fill=Step)) +
    geom_bar(position = "stack", stat = "identity") +
    scale_fill_manual(values = myColors ) +
    scale_y_continuous(breaks = c(0, 5, 10, 15, 20, 25, 30),
                     minor_breaks = NULL) +
    facet_grid(. ~ Dimensions) +
    theme(axis.text.x = element_text(angle = -90))
```

*Limitations, conclusions, and recommendations.*
In the bivariate studies, we chose arbitrary simulation parameters for three distributions, moving from the Gaussian to the discrete and non-normal, multivariate negative binomial. Under these conditions, the simulated random vectors sample the desired bivariate distribution across the entire range of pairwise correlations for the three dependency measures. The simulation results could differ for other choices of simulation settings. Specifying extreme correlations near the boundary or Frechet bounds could result in poor simulation performance. Fortunately, it is straightforward to evaluate simulation performance by using strategies similar to those completed above. We expect our random vector generation to perform well for the vast majority of NORTA-feasible correlation matrices, but advise to check the performance before making inferences/further analyses.
Finally, we note that one could use our single-pass algorithm `cor_fastPD` to produce a "close" (not nearest PD) to scale to even higher dimensions with greater loss of accuracy.
```{r include=FALSE, cache=FALSE}
# Remove all objects
rm(list = ls(all.names = TRUE))

# Unload all packages
```

<!--chapter:end:040-monte-carlo-evaluation.Rmd-->

```{r include=FALSE, cache=FALSE}
options(digits = 4)

knitr::opts_chunk$set(
  # Text Results
  eval = TRUE,
  echo = FALSE,
  results = 'markup',
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  # Code Decoration
  comment = NA,
  background = '#F7F7F7',
  # Cache
  cache = TRUE,
  # Plots
  fig.show = "hold",
  fig.width = 6.5,
  fig.asp = 0.618,  # 1 / phi
  out.width = "85%",
  fig.align = 'center'
)

# options(dplyr.print_min = 6, dplyr.print_max = 6)

seed_number <- 2020-02-25
set.seed(seed = seed_number)
```
# RNA-seq data applications  {#examples}

```{r ch050-LoadLib, include=FALSE, cache=FALSE}
box::use(
  dplyr[...],
  ggplot2[...],
  GGally[ggpairs, wrap],
  patchwork[...],
  bigsimr[...],
  knitr[include_graphics],
  JuliaCall[julia_call],
  ./R/utils[mom_nbinom]
)
load("data/example_brca.rda")
load("data/example_genes.rda")
sim_nbinom <- readRDS("results/brca_1000_sim.rds")
simRho <- readRDS("results/brca_1000_frobenius.rds")

Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
bs <- bigsimr_setup(pkg_check = FALSE)
dist <- distributions_setup()
```

This section demonstrates how to simulate multivariate data using `bigsimr`, aiming to replicate the structure of high-dimensional dependent count data. In an illustration of our proposed methodology, we seek to simulate RNA-sequencing data by producing simulated random vectors mimicking the observed data and its generating process. Modeling RNA-seq using multivariate probability distributions is natural as inter-gene correlation is an inherent part of biological processes [@Wang2009b]. And yet, many models do not account for this, leading to major disruptions to the operating characteristics of statistical estimation, testing, and prediction. See @Efron2012 for a detailed discussion with related methods and @Wu2012b, @Schissler2018; @Schissler2019 for applied examples. The following subsections apply `bigsimr`'s methods to real RNA-seq data, including replicating an estimated parametric structure, MC probability estimation, and MC evaluation of correlation estimation efficiency.

## Simulating High-Dimensional RNA-seq data

```{r ch050-readBRCA, cache=FALSE}
d <- 1000
brca1000 <- example_brca %>%
  select(all_of(1:d)) %>%
  mutate(across(everything(), as.double))
```

We begin by estimating the structure of the TCGA BRCA RNA-seq data set (see [Background](background)). Ultimately, we will simulate $B=10,000$ random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with $d=`r d`$. We assume a multivariate negative binomial (MVNB) model as RNA-seq counts are often over-dispersed and correlated. Since all $d$ selected genes exhibit over-dispersion (data not shown), we proceed to estimate the NB parameters $(r_i, p_i), i=1,\ldots,d$, to determine the target marginal PMFs $f_i$. To complete specification of the simulation algorithm inputs, we estimate the Spearman correlation matrix ${ \bf R}_{S}$ to characterize dependency.

<!--
This suggests simulating high-dimensional multivariate NB (MVB) with heterogeneous marginals would be useful tool in the development and evaluation of RNA-seq analytics. This procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions. All these genes exhibit over-dispersion and, so, we proceed to estimate the NB parameters $(r_i, p_i), i=1,\ldots,d$ to determine the target marginal PMFs $g_i(y_i)$ (via method of moments). Notably, the $\hat{p}_i's$ are small --- ranging in $[3.934 \times 10^{-6} , 1.217 \times 10^{-2}]$. To complete the simulation algorithm inputs, we estimate the Pearson correlation matrix $\bf{R_Y}$ and set that as the target correlation.
-->

<!--
With the simulation targets specified, we proceed to simulate $N=10,000$ random vectors $\bf{Y}$ $=( Y_1,\ldots,Y_d)$ with target Pearson correlation $\bf{R_Y}$ and marginal PMFs $g_i(y_i)$ using a $\bf{T}$-Poisson hierarchy of Kind II. Specifically, we first employ the \emph{direct Gaussian copula} approach to generate $N$ random vectors following a standard multivariate Gamma distribution $\bf{T}$ with shape parameters $r_i$ equal to the target $n_i$ and Pearson correlation matrix $\bf{R_T}$. Care must be taken when setting the specifying $\bf{R}$ (refer to Equation \ref{gay.cop.pdf}) --- we employ Equation \ref{mix.poi.corr} to compute the scaling factors $c_{i,j}$ and adjust the underlying correlations to ultimately match the target $\bf{R_Y}$. Notably, of the $525,825$ pairwise correlations from the $1026$ genes, no scale factor was less than $0.9907$, indicating the model can produce essentially the entire range of possible correlations. Here we are satisfied with approximate matching of the specified Gamma correlation and set $\bf{R}$ = $\bf{R_T}$ in our Gaussian copula scheme ($\bf{R}$ indicating the specified multivariate Gaussian correlation matrix). Finally, we generate the desired random vector $Y_i=N_i(t_i)$ by simulating Poisson counts with expected value $\mu_i=\lambda_i \times T_i$, for $i=1,\ldots,d$ (with $\lambda_i=\frac{(1-p_i)}{p_i}$) and repeat $N=10,000$ times.
-->

With this goal in mind, we first estimate the desired correlation matrix using the fast implementation provided by `bigsimr`:

```{r ch050-estRhoBRCA, echo=TRUE, cache=FALSE}
# Estimate Spearman's correlation on the count data
R_S <- bs$cor(as.matrix(brca1000), bs$Spearman)
```

Next, we estimate the marginal parameters. We use the method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model. While, marginal distributions are from the same probability family (NB), they are heterogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$. The functions below support this estimation for later use in `rvec`.

<!-- 
TYPESTE MoM estimators here.
-->

```{r ch050-nbHelpers, echo=TRUE}
make_nbinom_margins <- function(sizes, probs) {
  margins <- lapply(1:length(sizes), function(i) {
    dist$NegativeBinomial(sizes[i], probs[i])
  })
  do.call(c, margins)
}
```

We apply these estimators to all `r d` genes across the `r nrow(brca1000)` patients:

```{r ch050-estMargins, echo=TRUE}
nbinom_fit <- apply(brca1000, 2, mom_nbinom)
sizes <- nbinom_fit["size",]
probs <- nbinom_fit["prob",]
nb_margins <- make_nbinom_margins(sizes, probs)
```

Notably, the estimated marginal NB probabilities $\{ \hat{p}_i \}$ are small --- ranging in the interval $[`r min(probs)` , `r max(probs)`]$. This gives rise to highly variable counts and, typically, less restriction on potential pairwise correlation pairs. Once the functions are defined/executed to complete marginal estimation, we specify targets and generate the desired random vectors using `rvec`. Now we check admissibility of specified correlation matrix.

```{r ch050-cor-check, echo=TRUE}
# 1. Mapping step first
R_X <- bs$cor_convert(R_S, bs$Spearman, bs$Pearson)

# 2a. Check admissibility
(is_valid_corr <- julia_call("Bigsimr.iscorrelation", R_X))

# 2b. compute nearest correlation
if (!is_valid_corr) {
  R_X_pd <- bs$cor_nearPD(R_X)
  ## Quantify the error
  targets      <- R_X[lower.tri(R_X, diag = FALSE)]
  approximates <- R_X_pd[lower.tri(R_X_pd, diag = FALSE)]
  R_X          <- R_X_pd
}
summary(abs(targets - approximates))
```

While the exact $d \times d$ Spearman correlation matrix is not strictly admissible in our scheme (as seen by the non-positive definite result above), the approximation is close with a maximum absolute error of `r max( abs(targets - approximates))` and average absolute error of `r mean(abs(targets - approximates))` across the `r choose(d,2)` correlations.

```{r ch050-runBRCA-echo, echo=TRUE, eval=FALSE}
sim_nbinom <- bs$rvec(10000, R_X, nb_margins) 
colnames(sim_nbinom) <- colnames(brca1000)
```

Figure \@ref(fig:ch050-simDataFig) displays the simulated counts and pairwise relationships for our example genes from Table \@ref(tab:ch010-realDataTab).
Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parametric form and with less extreme points then the observed data in Figure \@ref(fig:ch010-realDataFig).

```{r ch050-simDataFig, cache=FALSE, fig.cap="Simulated data for three selected high-expressing genes generally replicates the estimated data structure. The data do not exhibit outlying points, but do possess the desired Spearman correlations, central tendencies, and discrete values."}
set.seed(2020-02-25)
num_genes <- 3
gene_sample <- sample(example_genes[1:1000], num_genes)

ggpairs(data = as.data.frame(sim_nbinom[, gene_sample]),
        upper = list(continuous = wrap('cor', method = "spearman"))) + 
  theme_bw()
```

Figure \@ref(fig:ch050-figBRCA) displays the aggregated results of our simulation by comparing the specified target parameter (horizontal axes) with the corresponding quantities estimated from the simulated data (vertical axes). The evaluation shows that the simulated counts approximately match the target parameters and exhibit the full range of estimated correlation from the data.

<!-- 
Utilizing 15 CPU threads in a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, the simulation completed in just over 2 minutes.
-->

```{r ch050-figBRCA, out.width='80%', fig.cap="Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set. The dashed red lines indicate equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes)."}
include_graphics('fig/ch050-figBRCA.png')
```

*Limitations, conclusions, and recommendations.*
The results show overall good simulation performance for our choice of parameters settings. Our settings were motivated by modeling high-expressing genes from TCGA BRCA data set. In general, the ability to match marginal and dependence parameters depends on the particular joint probability model. We recommend to evaluate and tune your simulation until you can be assured of the accuracy.

## Simulation-based joint probability calculations

Many statistical tasks require evaluation of a joint probability mass (or density) function:

$$
P( {\bf Y} = {\bf y} ), \: y_i \in \chi_i.
$$

where $\chi_i$ is the sample space for the $i^{th}$ component of the random vector $\bf{Y}$. Compact representations with convenient computational forms are rare for high-dimensional constructions, especially with heterogeneous, correlated marginal distributions (or margins of mixed data types). Given a large number of simulated vectors as produced above, estimated probabilities are readily given by counting the proportion of simulated vectors meeting the desired condition. In our motivating application, one may ask what is the probability that all genes expressed greater than a certain threshold value ${ \bf y}_0$. This can be estimated as

$$
\hat{P}( {\bf Y} \ge {\bf y_0 } ) = \frac{1}{B} \sum_{b=1}^B I( {\bf Y}^{ (b) } \ge {\bf y_0 } ),
$$

```{r ch050-densityEvaluation, include=FALSE}
d <- ncol(sim_nbinom)
B <- nrow(sim_nbinom)
y0 <- 1
pHat <- mean(apply(sim_nbinom, 1, function(Y) {
  all(Y >= y0)
}))
```

where ${\bf Y^{(b)} }$ is the $b^{th}$ simulated vector from a total of $B$ simulation replicates and $I(\cdot)$ is the indicator function. For example, we can estimate from our $B=10,000$ simulated vectors that the probability of all genes expressing (i.e., ${\bf y}_i \geq 1, \forall \; i$) is $`r pHat`$.

```{r ch050-densityEvaluationECHO, echo=TRUE}
<<ch050-densityEvaluation>>
pHat
```

## Evaluation of correlation estimation efficiency

MC methods are routinely used in many statistical inferential tasks including estimation, hypothesis testing, error rates, and empirical interval coverage rates. To conclude the example applications, we demonstrate how `bigsimr` can be used to evaluate estimation efficiency. In particular, we wish to assess the error in our correlation estimation above. We used a conventional method, based on classical statistical theory which was not designed for high-dimensional data. Indeed, high-dimensional covariance estimation (and precision matrices) is an active area of statistical science, .e.g., [@Won2013g; @VanWieringen2016].

In this small example, we simulate $m=30$ data sets with the number of simulated vectors matching the number of patients in the BRCA data set, $N=`r nrow(example_brca)`$. Since our simulation is much faster for the Pearson correlation type (see Figure \@ref(fig:ch040-largeDfig)), we only convert the Spearman correlation matrix once (and ensure it is positive definite). At each iteration, we estimate the quadratic loss (residual sum of squared errors) from the specified ${\bf R}_{S}$, producing a distribution of loss values.

```{r ch050-quadlossECHO, echo=TRUE, eval=FALSE}
# Simulate random vectors equal to the sample size
N <- nrow(example_brca)
# create m random vectors and estimate correlation
simRho <- replicate(n = m, expr = {
  tmpSim <- bs$rvec(N , R_X, nb_margins)
  bs$cor(tmpSim, bs$Spearman)
}, simplify = FALSE)
# Evaluate the residual sum of squared error
sapply(simRho, function(R) sum((R - R_S)^2))
```

```{r ch050-lossResults}
frobenius_loss <- sapply(simRho, function(R) sum((R - R_S)^2))
```

The `R` summary function supplies the mean-augmented five-number summary of the quadratic loss distribution computed above.

```{r ch050-loss-summary, echo=TRUE}
summary(frobenius_loss)
```

This distribution could be compared to high-dimensional designed covariance estimators for guidance on whether the additional complexity and computation time are warranted.
```{r include=FALSE, cache=FALSE}
# Remove all objects
rm(list = ls(all.names = TRUE))

# Unload all packages
```

<!--chapter:end:050-examples.Rmd-->

```{r include=FALSE, cache=FALSE}
options(digits = 4)

knitr::opts_chunk$set(
  # Text Results
  eval = TRUE,
  echo = FALSE,
  results = 'markup',
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  # Code Decoration
  comment = NA,
  background = '#F7F7F7',
  # Cache
  cache = TRUE,
  # Plots
  fig.show = "hold",
  fig.width = 6.5,
  fig.asp = 0.618,  # 1 / phi
  out.width = "85%",
  fig.align = 'center'
)

# options(dplyr.print_min = 6, dplyr.print_max = 6)

seed_number <- 2020-02-25
set.seed(seed = seed_number)
```
# Conclusion and discussion {#discussion}

We have introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a user-friendly, high-performance `R` package [`bigsimr`]. The random vector generation method is inspired by NORTA [@Cario1997] and Gaussian copula-based approaches [@MB13, @BF17, @Xia17]. The major contributions of this work are methods and software for flexible, scalable simulation of HD multivariate probability distributions with broad potential data analytic applications for modern, big-data statistical computing. For example, one could simulate high-resolution time series data, such as those consistent with an auto-regressive moving average model exhibiting a specified Spearman structure. Or our methods could be used to simulate sparsely correlated data, as many HD methods assume, via specifying a *spiked correlation matrix*.

Researchers working in multivariate computation frequently encounter such difficulties and need to find a close or nearest correlation matrix. A widely-available routine for this task in `R` is `matrix::nearPD`, though it is not suitable for high dimensions. To overcome this issue, we introduce `cor_nearPD` --- an implementation of @QS2006's method that computes the nearest correlation matrix to a given symmetric matrix in the Frobenius norm. This routine as many applications beyond our primary goal of random vector generation.

It is customary to compare new tools and algorithms directly to existing competing methods and software. In this study, however, we only employ our proposed methodology, as our previous work has shown that existing `R` tools are simply not designed to meet our high-dimensional goal (see @Li2019gpu for evaluations of the `R` `copula` package and others). For the bivariate simulations, existing packages such as [`nortaRA`](https://github.com/cran/NORTARA/blob/master/inst/doc/NORTARA.R) work well to match Pearson correlations exactly.

<!-- 
We advocate the use of Kendall's $\tau$ as it better captures correlation among components of non-normal, as well as non-linear patterns of association. Moreover, the matching Kendall's $\tau$ is nearly trivial due to the invariant of monotone transformations, after an adjustment to the input correlation matrix.
-->

<!-- 
We also show utility in our methodology through an application to differential gene expression analysis from RNA-sequencing data. The application results show that correlations indeed matter in the large-scale hypothesis testing, as many others have noted (for example, see @BE07, @Wu2012b). We also hope that the application provides an example workflow --- and other strategy to use in simulation design. Even the best-performing simulations we provide show a gap from the empirical distribution. To keep the demonstration straightforward, we did not attempt to match the empirical distribution of test statistics as precisely as possible. Yet our methodology could be more creatively applied to meet that goal. One could consider marginal distributions from different families, such as Poisson for a subset of genes. One could imagine finding a best fitting probability distribution among a class of distributions for each gene and this could perhaps a better fit. One could imagine additional structures/features in the data that an analyst could model to improve the correspondence with the empirical values, for example row correlations and high-dimensional covariance estimators (see @Won2013g).
-->

There are limitations to the methodology and implementation. We could only investigate selected multivariate distributions in our Monte Carlo studies. There may be instances that the methods do not perform well. Along those lines, we expect that correlation values close to the boundary of the feasible region could result in algorithm failure. Another issue is that for discrete distributions, we use continuous approximations when the support set is large. This could limit the scalability/accuracy for particular joint distributions.

<!-- 
The most obvious missing feature of the proposed methodology is the inability to match a Pearson correlation matrix exactly. As discussed in the [Algorithms](algorithms) section and extensively by @XZ19, this is a computationally intense procedure and Pearson's correlation is not a natural choice to describe dependency for non-normal marginals. While we do not provide an implementation directly supporting Pearson matching, users may supply their own input Pearson correlation after employing a supplementary matching scheme [@Cario1997; @XZ19].
-->

<!-- 
Further, while we provide the ability of the user to specify discrete marginals, exact matching a desired dependency measure is not yet obtained. One potential solution for this is *rescale* to account for ties [ref]. Finally we note, that the algorithm requires invertible marginal cdfs. This gives some restriction to the available marginal probability distributions (DOES IT? FOR EXAMPLE?). From a practical computing standpoint, the user's computing resource provides the ultimate limit on how large a dimension can be simulated using the `bigsimr` R package. A user must consider carefully the available memory and cores when conducting a Monte Carlo experiment. 
-->

<!-- 
Future work includes developing scaleable algorithms to match the Pearson correlation matrix more precisely, discrete-margin specific modifications including fast Spearman's correlation rescaling (see Equation \@ref(eq:spearmanRescaled)), and high-dimensional covariance estimation. From an implementation standpoint, `bigsimr` only supports Nvidia GPUs and redesigning the code using OpenCL would broaden the users who would benefit. As data-analytic problems grow to even larger dimension, multi-GPU support is a promising hardware-based future direction.
-->
```{r include=FALSE, cache=FALSE}
# Remove all objects
rm(list = ls(all.names = TRUE))

# Unload all packages
```

<!--chapter:end:060-conclusion-discussion.Rmd-->

```{r include=FALSE, cache=FALSE}
options(digits = 4)

knitr::opts_chunk$set(
  # Text Results
  eval = TRUE,
  echo = FALSE,
  results = 'markup',
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  # Code Decoration
  comment = NA,
  background = '#F7F7F7',
  # Cache
  cache = TRUE,
  # Plots
  fig.show = "hold",
  fig.width = 6.5,
  fig.asp = 0.618,  # 1 / phi
  out.width = "85%",
  fig.align = 'center'
)

# options(dplyr.print_min = 6, dplyr.print_max = 6)

seed_number <- 2020-02-25
set.seed(seed = seed_number)
```
# Supplementary Materials {.unnumbered #misc}

We provide an open-source implementation of our methodology as the `bigsimr` R package, hosted on github, https://github.com/SchisslerGroup/r-bigsimr.

```{r, out.width='5%'}
knitr::include_graphics('images/hex-bigsimr.png')
```

# Acknowledgment(s) {.unnumbered #acknowledgments}

The authors gratefully acknowledge Heather Knudson's graphic design for the `bigsimr` R Package. The results published here are in whole or part based upon data generated by the TCGA Research Network: https://www.cancer.gov/tcga.

# Disclosure statement {.unnumbered #coi}

The authors report no conflict of interest.

# Funding {.unnumbered #funding}

Research reported in this publication was supported by grants from the National Institutes of General Medical Sciences (5 U54 GM104944, GM103440) from the National Institutes of Health.

<!-- 
Nomenclature/Notation
-->
```{r include=FALSE, cache=FALSE}
# Remove all objects
rm(list = ls(all.names = TRUE))

# Unload all packages
```

<!--chapter:end:070-supplementary-materials.Rmd-->

```{r include=FALSE, cache=FALSE}
options(digits = 4)

knitr::opts_chunk$set(
  # Text Results
  eval = TRUE,
  echo = FALSE,
  results = 'markup',
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  # Code Decoration
  comment = NA,
  background = '#F7F7F7',
  # Cache
  cache = TRUE,
  # Plots
  fig.show = "hold",
  fig.width = 6.5,
  fig.asp = 0.618,  # 1 / phi
  out.width = "85%",
  fig.align = 'center'
)

# options(dplyr.print_min = 6, dplyr.print_max = 6)

seed_number <- 2020-02-25
set.seed(seed = seed_number)
```
# Appendix {.unnumbered #appendix}

\noindent Consider a bivariate example where we have a correlated $(Y_1, Y_2)^\top$ with $Y_1\sim N(\mu_1, \sigma_1^2)$ (normal) and $Y_2\sim LN(\mu_2, \sigma_2^2)$ (lognormal). First, let us note that when we get such a vector from bigsimr package then in fact it can be represented as

\begin{equation}
\label{kram1}
(Y_1, Y_2)^\top \stackrel{d}{=} \left(X_1, e^{X_2}\right)^\top,
\end{equation}

where $(X_1, X_2)^\top \sim N_2(\boldsymbol \mu, \boldsymbol \Sigma)$, which is bivariate normal with mean vector $\boldsymbol \mu = (\mu_1, \mu_2)^\top$ and variance-covariance matrix

\begin{equation}
\label{kram2}
\boldsymbol \Sigma = 
\left[
\begin{array}{cc}
\sigma_1^2 & \rho \sigma_1\sigma_2\\
\rho \sigma_1\sigma_2 & \sigma_2^2
\end{array}
\right].
\end{equation}

To see this, consider the three steps of the NORTA construction:

\begin{enumerate}

\item Generate $(Z_1, Z_2)^\top \sim N_2(\boldsymbol 0, \boldsymbol R)$, where

\begin{equation}
\label{kram3}
\boldsymbol R = 
\left[
\begin{array}{cc}
1 & \rho \\
\rho & 1
\end{array}
\right].
\end{equation}

\item Transform $(Z_1, Z_2)^\top$ to $(U_1, U_2)^\top$ viz. $U_i =\Phi(Z_i)$,  $i=1,2$, where $\Phi(\cdot)$ is the standard normal CDF.

\item Return $(Y_1, Y_2)^\top$, where $Y_i=F_i^{-1}(U_i)$, $i=1,2$, and $F_i(\cdot)$ is the CDF of $Y_i$ and $F_i^{-1}(\cdot)$ is its inverse (the quantile function).

\end{enumerate}

In this example, $F_1$ is the CDF of $N(\mu_1, \sigma_1^2)$ while $F_2$ is the CDF of $LN(\mu_2, \sigma_2^2)$, so that the two required quantile functions are

\begin{equation}
\label{kram4}
F_1^{-1}(u) = \mu_1+\sigma_1 \Phi^{-1}(u)\,\,\, \mbox{and} \,\,\, F_2^{-1}(u) = e^{\mu_2+\sigma_2 \Phi^{-1}(u)}, 
\end{equation}

as can be seen by standard calculation. Thus, when we apply Step 3 of the NORTA algorithm, we get

\begin{equation}
\label{kram5}
F_1^{-1}(U_1) = \mu_1+\sigma_1 \Phi^{-1}(\Phi(Z_1)) =  \mu_1+\sigma_1 Z_1 \,\,\, \mbox{and} \,\,\, F_2^{-1}(U_2) = e^{\mu_2+\sigma_2 \Phi^{-1}(\Phi(Z_2))} = e^{\mu_2+\sigma_2 Z_2}, 
\end{equation}

where $(X_1, X_2)^\top =  (\mu_1+\sigma_1 Z_1, \mu_2+\sigma_2 Z_2)^\top \sim N_2(\boldsymbol \mu, \boldsymbol \Sigma)$. Consequently, the vector $(Y_1, Y_2)^\top$ obtained in Step 3 of the algorithm has the structure provided by (\ref{kram1}).

\vspace{0.1in}

\noindent Next, we provide the exact covariance structure of the random vector $(Y_1, Y_2)^\top$ given by (\ref{kram1}), and relate the correlation of $Y_1$ and $Y_2$ to $\rho$, which is the correlation of the normal variables $Z_1$, $Z_2$ (and also $X_1$ and $X_2$). A straightforward albeit somewhat tedious algebra produces the following result.

\begin{lemma}
Let ${\bf Y} = (Y_1, Y_2)^\top$ admit the stochastic representation (\ref{kram1}), where ${\bf X} = (X_1, X_2)^\top \sim N_2(\boldsymbol \mu, \boldsymbol \Sigma)$ with $\boldsymbol \mu$ and $\boldsymbol \Sigma$ as above. Then the mean vector and the variance-covariance matrix of ${\bf Y}$ are given by

\begin{equation}
\label{kram6}
\boldsymbol \mu_{{\bf Y}} = \left(\mu_1, e^{\mu_2+\sigma_2^2/2}\right)^\top
\end{equation}

and

\begin{equation}
\label{kram7}
\boldsymbol \Sigma = 
\left[
\begin{array}{cc}
\sigma_1^2 & \rho \sigma_1\sigma_2  e^{\mu_2+\sigma_2^2/2}  \\
\rho \sigma_1\sigma_2 e^{\mu_2+\sigma_2^2/2} & \left[ e^{\mu_2+\sigma_2^2/2}\right]^2 \left[e^{\sigma_2^2} -1 \right],
\end{array}
\right].
\end{equation}

respectively. 
\end{lemma}

\begin{proof}
The values of the means and the variances are obtained immediately from normal marginal distribution of $Y_1$ and lognormal marginal distribution of $Y_2$. It remains to establish the covariance of $Y_1$ and $Y_2$,

\begin{equation}
\label{covy1y2}
\mbox{Cov}(Y_1, Y_2) = \mathbb E(Y_1 Y_2) - \mathbb E(Y_1)\mathbb E(Y_2), 
\end{equation}

and in particular the first expectation on the right-hand-side above. By using the tower property of expectations, the latter expectation can be expressed as

\begin{equation}
\label{kre1}
\mathbb E(Y_1 Y_2)  =  \mathbb E\left(X_1 e^{X_2}\right)  =  \mathbb E \left\{ \mathbb E \left(X_1 e^{X_2} | X_1 \right) \right\} = \mathbb E \left\{ X_1 \mathbb E \left(e^{X_2} | X_1 \right) \right\}.
\end{equation}

Further, by using the fact that the conditional distribution of $X_2$ given $X_1=x_1$ is normal with mean $\mathbb E(X_2|X_1=x_1) = \mu_2+\rho\sigma_2(x_1-\mu_1)/\sigma_1$ and variance $\sigma_2^2(1-\rho^2)$, one can relate the inner expectation on the far right in (\ref{kre1}) to the moment generating function $M(t)$ of this conditional distribution evaluated at $t=1$, leading to

\begin{equation}
\label{kre2}
E \left(e^{X_2} | X_1 \right) = e^{ \mu_2+\rho\sigma_2(x_1-\mu_1)/\sigma_1 + \sigma_2^2(1-\rho^2)/2},
\end{equation}

so that

\begin{equation}
\label{kre3}
\mathbb E(Y_1 Y_2)  =  e^{ \mu_2+ \frac{1}{2} \sigma_2^2(1-\rho^2)} \mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\}.
\end{equation}

Since $X_1$ is normal with mean $\mu_1$ and variance $\sigma_1^2$, the expectation in (\ref{kre3}) becomes

\begin{equation}
\label{kre4}
\mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = \int_{-\infty}^\infty x_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (x_1-\mu_1)} \frac{1}{\sqrt{2\pi}\sigma_1} e^{-\frac{1}{2\sigma_1^2}(x_1-\mu_1)^2}.
\end{equation}

Upon substituting $u=x_1-\mu_1$, followed by some algebra, we arrive at

\begin{equation}
\label{kre5}
\mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = e^{\frac{1}{2}\sigma_2^2\rho^2} \int_{-\infty}^\infty (u+\mu_1)  \frac{1}{\sqrt{2\pi}\sigma_1} e^{-\frac{1}{2\sigma_1^2}(u-\rho\sigma_1\sigma_2)^2}.
\end{equation}

Since the integral in (\ref{kre5}) is the expectation $\mathbb E(X+\mu_1)$, where $X$ is normal with mean $\rho\sigma_1\sigma_2$ and variance $\sigma_1^2$, we conclude that

\begin{equation}
\label{kre6}
\mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = e^{\frac{1}{2}\sigma_2^2\rho^2} (\rho\sigma_1\sigma_2 +\mu_1),
\end{equation}

which, in view of (\ref{kre3}), leads to

\begin{equation}
\label{kre7}
\mathbb E (Y_1Y_2) = e^{\mu_2 + \frac{1}{2}\sigma_2^2} (\rho\sigma_1\sigma_2 +\mu_1).
\end{equation}

Finally, (\ref{covy1y2}), along with the expressions for the means of $Y_1$ and $Y_2$, produce the covariance of $Y_1$ and $Y_2$:

\begin{equation}
\label{kre8}
\mbox{Cov}(Y_1, Y_2) = e^{\mu_2 + \frac{1}{2}\sigma_2^2} (\rho\sigma_1\sigma_2 +\mu_1) - \mu_1 e^{\mu_2 + \frac{1}{2}\sigma_2^2} = \rho\sigma_1\sigma_2 e^{\mu_2 + \frac{1}{2}\sigma_2^2}.
\end{equation}

This concludes the proof
\end{proof}

Using the above result, we can directly relate the correlation coefficient of $Y_1$ and $Y_2$ with that of $X_1$ and $X_2$,

\begin{equation}
\label{kram8}
\rho_{Y_1, Y_2} = \rho \frac{\sigma_2}{\sqrt{e^{\sigma_2^2} -1}},
\end{equation}

where $\rho$ is the correlation of $X_1$ and $X_2$. The above result is useful when studying the possible range of correlation of $Y_1$ and $Y_2$ in the above example. It can be shown that the factor on the far right in (\ref{kram8}) is a monotonically decreasing function of $\sigma_2$ on $(0,\infty)$, with the limits of 1 and 0 at zero and infinity, respectively. Thus, in principle, the range of correlation of $Y_1$ and $Y_2$ is the same as that of $X_1$ and $X_2$, as the factor on the far right in (\ref{kram8}) can be made arbitrarily close to 1. However, by changing this factor we may affect the marginal distributions of $Y_1$ and $Y_2$. It can be shown that if the marginal distributions of $Y_1$ and $Y_2$ are fixed, then the relation (\ref{kram8}) becomes

\begin{equation}
\label{kram9}
\rho_{Y_1, Y_2} = \rho \sqrt{\frac{\log(1+c_2^2)}{c_2^2}},
\end{equation}

where $c_2=\sigma_{Y_2}/\mu_{Y_2}$ is the *coefficient of variation* (CV) of the variable $Y_2$. Thus, the range of possible correlations in this model is not affected by the distribution of $Y_1$, and is determined by the CV of $Y_2$ as follows:

\begin{equation}
\label{kram10}
- \sqrt{\frac{\log(1+c_2^2)}{c_2^2}} \leq \rho_{Y_1, Y_2} \leq \sqrt{\frac{\log(1+c_2^2)}{c_2^2}}. 
\end{equation}

Plugging in the estimates of these quantities from Section \@ref(package), we see that

\begin{equation}
\label{kram11}
\frac{\hat{\sigma}_2}{\sqrt{e^{\hat{\sigma}_2^2} -1}} = 0.881.
\end{equation}

Thus, the possible range of correlation becomes $(-0.881, 0.881)$. This approximately agrees with the MC results provided.
```{r include=FALSE, cache=FALSE}
# Remove all objects
rm(list = ls(all.names = TRUE))

# Unload all packages
```

<!--chapter:end:080-appendix.Rmd-->

```{r include=FALSE, cache=FALSE}
options(digits = 4)

knitr::opts_chunk$set(
  # Text Results
  eval = TRUE,
  echo = FALSE,
  results = 'markup',
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  # Code Decoration
  comment = NA,
  background = '#F7F7F7',
  # Cache
  cache = TRUE,
  # Plots
  fig.show = "hold",
  fig.width = 6.5,
  fig.asp = 0.618,  # 1 / phi
  out.width = "85%",
  fig.align = 'center'
)

# options(dplyr.print_min = 6, dplyr.print_max = 6)

seed_number <- 2020-02-25
set.seed(seed = seed_number)
```
`r if (knitr:::is_html_output()) '# References {-}'`
```{r include=FALSE, cache=FALSE}
# Remove all objects
rm(list = ls(all.names = TRUE))

# Unload all packages
```

<!--chapter:end:300-references.Rmd-->

---
institution:
  name: "University of Nevada, Reno"
  department: "Mathematics and Statistics"
  address:
    line1: "1664 N. Virginia Street / 0084"
    line2: "Reno, Nevada 89557"
  phone: "(775) 784-6773"
  fax: "(775) 784-6378"
  url: https://www.unr.edu/math/
  logo: "logos/blue.png"
author:
  first: "Alfred"
  last: "Schissler"
  short: "Grant"
  title: "Ph.D."
  position: "Assistant Professor, Statistics"
  signature: "signature.jpg"
to:
  first: "Alexander"
  last: "Knudson"
  short: "Alex"
date: "`r format(Sys.Date(), format='%d %B %Y')`"
opening: "Dear CSDA Annals of Statistical Data Science Editor(s):"
closing: "Sincerely,"
output:
  pdf_document:
    template: templateSingleSignature.tex
---
```{r include=FALSE, cache=FALSE}
options(digits = 4)

knitr::opts_chunk$set(
  # Text Results
  eval = TRUE,
  echo = FALSE,
  results = 'markup',
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  # Code Decoration
  comment = NA,
  background = '#F7F7F7',
  # Cache
  cache = TRUE,
  # Plots
  fig.show = "hold",
  fig.width = 6.5,
  fig.asp = 0.618,  # 1 / phi
  out.width = "85%",
  fig.align = 'center'
)

# options(dplyr.print_min = 6, dplyr.print_max = 6)

seed_number <- 2020-02-25
set.seed(seed = seed_number)
```

As an invited speaker at CMStatistics 2020, I was delighted to learn of the opportunity to have our work included in the first issue of CSDA Annals of Statistical Data Science.

On behalf of my co-authors, I'm pleased to submit our manuscript entitled *Simulating High-Dimensional Multivariate Data using the `Bigsimr` Package* for your consideration.
We feel that the work is well suited to the indicated themes and hope that it will be considered a valuable contribution to this prestigious journal.

I hope that everything is in order. If not, or if you have any additional needs, please do not hesitate to contact me. Thank you and all those involved in this process.

Best regards,
```{r include=FALSE, cache=FALSE}
# Remove all objects
rm(list = ls(all.names = TRUE))

# Unload all packages
```

<!--chapter:end:letter_bigsimr_CSDA.Rmd-->

### 2021-03-01 letter_bigsimr_CSDA.Rmd
fix double ending to letter

@@
 opening: "Dear CSDA Annals of Statistical Data Science Editor(s):"
-closing: "Sincerely,"
+closing: "Best regards,"
 output:
@@
 I hope that everything is in order. If not, or if you have any additional needs, please do not hesitate to contact me. Thank you and all those involved in this process.
-
-Best regards,

### 2021-03-01 010-background.Rmd
fixed formatting

@@
   upper = list(continuous = wrap('cor', method = "spearman"))
-) + theme_bw()
+) + 
+  theme_bw() +
+  theme(axis.text.x = element_text(angle = -90, vjust = 0.5))
 ```
@@
 
-Spearman's correlation for discrete marginal suffers issues whenever there is a nonzero probability of ties.
+Spearman's correlation for discrete marginal distributions suffers issues whenever there is a nonzero probability of ties.
 For example, the Spearman's correlation of a discrete-valued random variable $X$ with itself could be less than 1 [@MB13].
@@
 Note that with continuous marginals the rescaling returns $\rho_S$.
-For discrete marginals with large or infinite support, computing the adjustment factors $\sum_x p(x)^3$, $\sum_y p(y)^3$ over all large number of pairs becomes expensive (often violating desired property SP1).
+For discrete marginals with large or infinite support, computing the adjustment factors $\sum_x p(x)^3$, $\sum_y p(y)^3$ over a large number of pairs becomes expensive (often violating desired property SP1).
 Further, the infinite sums must be approximated for count-valued data.

### 2021-03-01 030-bigsimr-package.Rmd
fix comparison with theory in 030
fixed formatting

@@
 
-First we use `cor_bounds` to ensure that the pairwise target correlation values are valid prior to mapping step which constructs $R_X$ for the MVN input into NORTA (see Section \@ref(algorithms)).
+First we use `cor_bounds` to ensure that the pairwise target correlation values are valid prior to the mapping step which constructs $R_X$ for the MVN input into NORTA (see Section \@ref(algorithms)).
 `cor_bounds` estimates the pairwise lower and upper correlation bounds using the Generate, Sort, and Correlate algorithm of @DH2011.

### 2021-03-01 040-monte-carlo-evaluation.Rmd
fixed formatting

@@
 
-```{r ch040-LoadLib040, include=FALSE}
+```{r ch040-LoadLib040, include=FALSE, cache=FALSE}
 box::use(
@@
 kable(tabMAE, booktabs = TRUE,
+      linesep = c("", "", "\\addlinespace"),
       col.names = c("No. of random vectors",
@@
 Overall, the studies show that our methodology is generally accurate across the entire range of possible correlation values all three dependency measures, at least in these limited simulation settings for the rank-based correlations.
-Our Pearson matching performs nearly as well as Spearman or Kendall, except a slight increase in error for negative binomial case.
+Our Pearson matching performs nearly as well as Spearman or Kendall, except for a slight increase in error for negative binomial case.
 This is due the particularly large counts generated with our choice of parameters for $p$ and $r$.
@@
 
-<!-- 
-See [Discussion](discussion) for future directions for fast Pearson matching and discrete-specific modifications.
--->
-
 ## Scale up to High Dimensions
 
-With information of our method's accuracy from a low-dimensional perspective, we now turn to assessing whether the `bigsimr` can scale to larger dimensional problems with practical computation times.
+With information of our method's accuracy from a low-dimensional perspective, we now turn to assessing whether `bigsimr` can scale to larger dimensional problems with practical computation times.
 Specifically, we ultimately generate $B=1,000$ random vectors for $d=\{100, 250, 500, 1000, 2500, 5000, 10000\}$ for each correlation type, $\{Pearson, Spearman, Kendall\}$ while timing the algorithm's major steps.
@@
     kable(booktabs = TRUE,
+          linesep = c("", "", "\\addlinespace"),
           col.names = c("Dimension",
@@
 
-The results larger dimensional vectors show scalability to ultra-high dimensions for all three correlation types, although the total times do become much larger. 
+The results with larger dimensional vectors show scalability to ultra-high dimensions for all three correlation types, although the total times do become much larger. 
 Figure \@ref(fig:ch040-largeDfig) displays computation times for $d=\{1000, 2500, 5000, 10000\}$.

### 2021-03-01 050-examples.Rmd
fix 050 typo
revise 050
fixed formatting

@@
 Once the functions are defined/executed to complete marginal estimation, we specify targets and generate the desired random vectors using `rvec`.
-Now we check admissibility of specified correlation matrix.
+Now we check admissibility of the specified correlation matrix.
 
@@
 Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parametric form and with less extreme points then the observed data in Figure \@ref(fig:ch010-realDataFig).
+Figure \@ref(fig:ch050-figBRCA) displays the aggregated results of our simulation by comparing the specified target parameter (horizontal axes) with the corresponding quantities estimated from the simulated data (vertical axes).
+The evaluation shows that the simulated counts approximately match the target parameters and exhibit the full range of estimated correlation from the data.
 
@@
 
-Figure \@ref(fig:ch050-figBRCA) displays the aggregated results of our simulation by comparing the specified target parameter (horizontal axes) with the corresponding quantities estimated from the simulated data (vertical axes).
-The evaluation shows that the simulated counts approximately match the target parameters and exhibit the full range of estimated correlation from the data.
-
 ```{r ch050-figBRCA, out.width='80%', fig.cap="Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set.
@@
 The results show overall good simulation performance for our choice of parameters settings.
-Our settings were motivated by modeling high-expressing genes from TCGA BRCA data set.
+Our settings were motivated by modeling high-expressing genes from the TCGA BRCA data set.
 In general, the ability to match marginal and dependence parameters depends on the particular joint probability model.
@@
 
-where ${\bf Y^{(b)} }$ is the $b^{th}$ simulated vector from a total of $B$ simulation replicates and $I(\cdot)$ is the indicator function.
+where ${\bf Y}^{(b)} $ is the $b^{th}$ simulated vector from a total of $B$ simulation replicates and $I(\cdot)$ is the indicator function.
 For example, we can estimate from our $B=10,000$ simulated vectors that the probability of all genes expressing (i.e., ${\bf y}_i \geq 1, \forall \; i$) is $`r pHat`$.

### 2021-03-01 040-monte-carlo-evaluation.Rmd
fixed breaks on fig 5

@@
     scale_fill_manual(values = myColors ) +
-    scale_y_continuous(breaks = c(0, 5, 10, 15, 20, 25, 30),
+    scale_y_continuous(breaks = seq(0, 80, 10),
                      minor_breaks = NULL) +

### 2021-03-01 040-monte-carlo-evaluation.Rmd
"fixed" list spacing
fixed lists by removing two spaces at end of lnes
say more about Kendall matching

--- 2021-03-01 040-monte-carlo-evaluation.Rmd
@@
 We expect our random vector generation to perform well for the vast majority of NORTA-feasible correlation matrices, but advise to check the performance before making inferences/further analyses.
+
+Somewhat surprising, Kendall estimation and nearest PD computation scale poorly compared to Spearman and, even, approximate Pearson matching.
+In our experience, Kendall computation times are sensitive to the number of cores, benefiting from multi-core parallelization.
+This could mitigate some of the current algorithmic/implementation shortcomings.
+Despite this, Kendall matching is still feasible for most high-dimension data sets.
 Finally, we note that one could use our single-pass algorithm `cor_fastPD` to produce a "close" (not nearest PD) to scale to even higher dimensions with greater loss of accuracy.

### 2021-03-01 020-simulation-algorithm.Rmd
fixed formatting
fixed breaks on fig 5

@@
 
-The well-known NORTA algorithm [@Cario1997] simulates a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$. Specifically, NORTA algorithm proceeds as follows:
+The well-known NORTA algorithm [@Cario1997] simulates a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$. Specifically, NORTA algorithm proceeds as:
+
+\setstretch{1.5}
 
@@
 5. $\text{Return} \; Y \; \text{where} \; Y_i \gets F_i^{-1}[\Phi(X_i)], \; i=1,...,d$.
+
+\setstretch{2.0}
 
@@
 
+\setstretch{1.5}
+
 1. Mapping step
-    (i) Convert the target Spearman correlation matrix $R_S$ to the corresponding MVN Pearson correlation $R_X$. Alternatively,  
-    (i') Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$. Alternatively,  
-    (i'') Convert the target Pearson correlation matrix to $R_P$ to the corresponding, approximate MVN Pearson correlation $R_X$.  
+    * (i) Convert the target Spearman correlation matrix $R_S$ to the corresponding MVN Pearson correlation $R_X$. Alternatively,
+    * (i') Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$. Alternatively,
+    * (i'') Convert the target Pearson correlation matrix to $R_P$ to the corresponding, approximate MVN Pearson correlation $R_X$.
 2. Check admissibility and, if needed, compute the nearest correlation matrix.
-    (i) Check that $R_X$ is a correlation matrix, a positive definite matrix with 1's along the diagonal.   
-	(ii) If $R_X$ is a correlation matrix, the input matrix is *admissible* in this scheme. Otherwise
-    (ii$^\prime$) Replace $R_X$ with the nearest correlation matrix $\tilde{R}_X$, in the Frobenius norm.  
+    * (i) Check that $R_X$ is a correlation matrix, a positive definite matrix with 1's along the diagonal.
+    * (ii) If $R_X$ is a correlation matrix, the input matrix is *admissible* in this scheme. Otherwise
+    * (ii') Replace $R_X$ with the nearest correlation matrix $\tilde{R}_X$, in the Frobenius norm.
 3. Gaussian copula
-    (i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)$.  
-    (ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz. $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
-    (iii) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$.
+    * (i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)$.
+    * (ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz. $U_i=\Phi(X_i)$, $i=1, \ldots, d$.
+    * (iii) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$.
+
+\setstretch{2.0}
 

### 2021-03-01 040-monte-carlo-evaluation.Rmd

--- 2021-03-01 040-monte-carlo-evaluation.Rmd
@@
 
+\setstretch{1.5}
 1.
@@
 3. Simulate a "data set" of $1,000 \times d$ random vectors via `rvec` (without matching the Pearson exactly).
+\setstretch{2.0}
 
@@
 
+\setstretch{1.5}
 i. Estimate the correlation matrix from the "data" in the *Compute Correlation* step.
@@
 3. Simulate $1,000$ vectors via the Inverse Transform method in the *Simulate Data* step.
+\setstretch{2.0}
 

### 2021-03-01 050-examples.Rmd
fixed issue with math environment

@@
 Given a large number of simulated vectors as produced above, estimated probabilities are readily given by counting the proportion of simulated vectors meeting the desired condition.
-In our motivating application, one may ask what is the probability that all genes expressed greater than a certain threshold value ${ \bf y}_0$.
+In our motivating application, one may ask what is the probability that all genes expressed greater than a certain threshold value ${\bf y}_0$.
 This can be estimated as
@@
 $$
-\hat{P}( {\bf Y} \ge {\bf y_0 } ) = \frac{1}{B} \sum_{b=1}^B I( {\bf Y}^{ (b) } \ge {\bf y_0 } ),
+\hat{P}( {\bf Y} \ge {\bf y_0 } ) = \frac{1}{B} \sum_{b=1}^B I( {\bf Y}^{ (b) } \ge {\bf y_0} ),
 $$
@@
 
-where ${\bf Y}^{(b)} $ is the $b^{th}$ simulated vector from a total of $B$ simulation replicates and $I(\cdot)$ is the indicator function.
+where ${\bf Y}^{(b)}$ is the $b^{th}$ simulated vector from a total of $B$ simulation replicates and $I(\cdot)$ is the indicator function.
 For example, we can estimate from our $B=10,000$ simulated vectors that the probability of all genes expressing (i.e., ${\bf y}_i \geq 1, \forall \; i$) is $`r pHat`$.

### 2021-03-01 010-background.Rmd
"fixed" list spacing
fixed lists by removing two spaces at end of lnes

@@
 
+\setstretch{1.5}
+
 * BP1: A wide range of dependences, allowing both positive and negative values, and, ideally, admitting the full range of possible values.
@@
 
+\setstretch{2.0}
+
 Moreover, the simulation method must *scale* to high dimensions:
+
+\setstretch{1.5}
 
@@
 * SP2: Procedure must scale to high dimensions while maintaining accuracy.
+
+\setstretch{2.0}
 
@@
 We illustrate our methodology using a well-studied Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments).
-For simplicity, we only consider high expressing genes.
+For ease of modeling and simplicity of exposition, we only consider high expressing genes.
 In turn, we begin by filtering to retain the top `r d` of the highest-expressing genes (in terms of median expression) of the over 20,000 gene measurements from $N=`r nrow(example_brca)`$ patients' tumor samples.
@@
 Table \@ref(tab:ch010-realDataTab) displays RNA-seq counts for three selected high-expressing genes for the first five patients' breast tumor samples.
+To help visualize the bivariate relationships for these three selected genes across all patients, Figure \@ref(fig:ch010-realDataFig) displays the marginal distributions and estimated Spearman's correlations.
 
@@
 
-To help visualize the bivariate relationships for these three selected genes across all patients, Figure \@ref(fig:ch010-realDataFig) displays the marginal distributions and estimated Spearman's correlations.
-
 ```{r ch010-realDataFig, cache=FALSE, fig.width=8, fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes from Table 1.
-The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations.
+The data possess outliers, heavy-right tails, are discrete, and have non-trivial intergene correlations.
 Modeling these data motivate our simulation methodology."}
@@
 
-*Discrete marginal considerations*
-
-Spearman's correlation for discrete marginal distributions suffers issues whenever there is a nonzero probability of ties.
-For example, the Spearman's correlation of a discrete-valued random variable $X$ with itself could be less than 1 [@MB13].
-One remedy for this issue is to rescale the Spearman correlation.
-For two random variables $X,Y$ with probability mass functions (PMFs) or probability densities functions (PDFs) $p(x)$ and $q(y)$, respectively, define the rescaled Spearman's correlation as
-
-\begin{equation}
-(\#eq:spearmanRescaled)
-\rho_{RS}(X,Y) = \frac{\rho_s(X,Y)}{ \left[ \left[ 1 - \sum_x p(x)^3 \right] \left[ 1 - \sum_y q(y)^3 \right] \right]^{1/2}}.
-\end{equation}
-
-Note that with continuous marginals the rescaling returns $\rho_S$.
-For discrete marginals with large or infinite support, computing the adjustment factors $\sum_x p(x)^3$, $\sum_y p(y)^3$ over a large number of pairs becomes expensive (often violating desired property SP1).
-Further, the infinite sums must be approximated for count-valued data.
-
-Also, for a discrete random variable $Y_i$, some care must be taken to define the quantile function $F_{i}^{-1}$. A standard practice is to set
-
-\begin{equation}
-F_{i}^{-1} = \inf\{y:F_{i}(y) \geq u \}.
-(\#eq:inverseCDF)
-\end{equation}
-
 *Marginal-dependent bivariate correlation bounds*
@@
 \noindent where $U$ is a uniform random variable on $(0,1)$ and $F^{-1}_1, F^{-1}_2$ are the inverse CDFs of $X_1$ and $X_2$, respectively, definitely by \@ref(eq:inverseCDF) when the variables are discrete.
+
+\begin{equation}
+F_{i}^{-1} = \inf\{y:F_{i}(y) \geq u \}.
+(\#eq:inverseCDF)
+\end{equation}
 

### 2021-03-01 040-monte-carlo-evaluation.Rmd

@@
 This could mitigate some of the current algorithmic/implementation shortcomings.
-Despite this, Kendall matching is still feasible for most high-dimension data sets.
-Finally, we note that one could use our single-pass algorithm `cor_fastPD` to produce a "close" (not nearest PD) to scale to even higher dimensions with greater loss of accuracy.
+Despite this, Kendall matching is still feasible for most high-dimensional data sets.
+Finally, we note that one could use our single-pass algorithm `cor_fastPD` to produce a 'close' (not nearest PD) to scale to even higher dimensions with greater loss of accuracy.

### 2021-03-01 060-conclusion-discussion.Rmd
fixed formatting

@@
 Or our methods could be used to simulate sparsely correlated data, as many HD methods assume, via specifying a *spiked correlation matrix*.
-
-Researchers working in multivariate computation frequently encounter such difficulties and need to find a close or nearest correlation matrix.
-A widely-available routine for this task in `R` is `matrix::nearPD`, though it is not suitable for high dimensions.
-To overcome this issue, we introduce `cor_nearPD` --- an implementation of @QS2006's method that computes the nearest correlation matrix to a given symmetric matrix in the Frobenius norm.
-This routine as many applications beyond our primary goal of random vector generation.
 
@@
 Another issue is that for discrete distributions, we use continuous approximations when the support set is large.
-This could limit the scalability/accuracy for particular joint distributions.
+This could limit the scalability/accuracy for particular discrete/mixed multivariate distributions.
+Our method would also benefit computationally from faster Kendall estimation and more finely tuned nearest PD calculation for non-admissible Kendall matrices.

### 2021-03-01 070-supplementary-materials.Rmd

@@
 
-We provide an open-source implementation of our methodology as the `bigsimr` R package, hosted on github, https://schisslergroup.github.io/bigsimr/.
+We provide an open-source implementation of our methodology as the `Bigsimr.jl` Julia package, hosted on github at https://github.com/adknudson/Bigsimr.jl.
+Additionally we provide R and Python interfaces to `Bigsimr`, respectively at https://github.com/SchisslerGroup/r-bigsimr and https://github.com/SchisslerGroup/python-bigsimr.
+`Bigsimr.jl` is an ongoing project, and feature requests or issues can be submitted to https://github.com/adknudson/Bigsimr.jl/issues.
 

### 2021-03-27 000-introduction.Rmd
idk

--- 2021-03-01 article_bigsimr.Rmd
@@
---- 
-title: "Simulating High-Dimensional Multivariate Data"
-subtitle: "using the `Bigsimr` Package"
-author:
-  - Alfred G. Schissler*
-  - Edward J. Bedrick
-  - Alexander D. Knudson
-  - Tomasz J. Kozubowski
-  - Tin Nguyen
-  - Anna K. Panorska
-  - Juli Petereit
-  - Walter W. Piegorsch
-  - Duc Tran
-site: bookdown::bookdown_site
-bibliography: ["bigsimr.bib", "packages.bib", "alex.bib"]
-abstract:  It is critical to realistically simulate data when employing Monte Carlo techinques and evaluating statistical methodology.
-Measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments.
-Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations.
-This article introduces the `bigsimr` R package.
-This high-level package provides a flexible and scalable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions and known Pearson, Spearman, or Kendall correlation matrix.
-`bigsimr` contains additional high-performance features, including multi-core algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix.
-Monte carlo studies quantify the accuracy and scalability of our approach, up to $d=10,000$.
-Finally, we demonstrate example applications enabled via `bigsimr` by applying these methods to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples.
----
-```{r include=FALSE, cache=FALSE}
-options(digits = 4)
-
-knitr::opts_chunk$set(
-  # Text Results
-  eval = TRUE,
-  echo = FALSE,
-  results = 'markup',
-  collapse = TRUE,
-  message = FALSE,
-  warning = FALSE,
-  error = FALSE,
-  # Code Decoration
-  comment = NA,
-  background = '#F7F7F7',
-  # Cache
-  cache = TRUE,
-  # Plots
-  fig.show = "hold",
-  fig.width = 6.5,
-  fig.asp = 0.618,  # 1 / phi
-  out.width = "85%",
-  fig.align = 'center'
-)
-
-# options(dplyr.print_min = 6, dplyr.print_max = 6)
-
-seed_number <- 2020-02-25
-set.seed(seed = seed_number)
-```
-
-*Keywords:*  Multivariate simulation, High-dimensional data, Nonparametric correlation, Gaussian copula, RNA-sequencing data, breast cancer
-
-*=Corresponding author. aschissler@unr.edu. 1664 N Virginia St, Reno, NV 89557
-
-```{r include=FALSE}
-# automatically create a bib database for R packages
-knitr::write_bib(c(
-  .packages(), 'bookdown', 'knitr', 'rmarkdown'
-), 'packages.bib')
-```
-
 # Introduction
@@
 Finally, we discuss the method's utility, limitations, and future directions.
-```{r include=FALSE, cache=FALSE}
-# Remove all objects
-rm(list = ls(all.names = TRUE))
-
-# Unload all packages
-```
-
-<!--chapter:end:index.Rmd-->
-
-```{r include=FALSE, cache=FALSE}
-options(digits = 4)
-
-knitr::opts_chunk$set(
-  # Text Results
-  eval = TRUE,
-  echo = FALSE,
-  results = 'markup',
-  collapse = TRUE,
-  message = FALSE,
-  warning = FALSE,
-  error = FALSE,
-  # Code Decoration
-  comment = NA,
-  background = '#F7F7F7',
-  # Cache
-  cache = TRUE,
-  # Plots
-  fig.show = "hold",
-  fig.width = 6.5,
-  fig.asp = 0.618,  # 1 / phi
-  out.width = "85%",
-  fig.align = 'center'
-)
-
-# options(dplyr.print_min = 6, dplyr.print_max = 6)
-
-seed_number <- 2020-02-25
-set.seed(seed = seed_number)
-```
-# Background {#background}
-
-```{r ch010-background, include=FALSE, cache=FALSE}
-box::use(
-  dplyr[...],
-  ggplot2[...],
-  stringr[str_sub],
-  GGally[ggpairs, wrap],
-  knitr[kable]
-)
-load("data/example_genes.rda")
-load("data/example_brca.rda")
-d <- 1000
-```
-
-The `bigsimr` `R` package presented here provides multiple algorithms that operate with high-dimensional multivariate data; however, all these algorithms were originally designed to support a single task: to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics.
-Specifically, our goal is to efficiently simulate a large number, $B$, of HD random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with *correlated* components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i$.
-
-When designing this methodology, we developed the following properties to guide our effort.
-We divide the properties into two categories: (1) basic properties (BP) and "scalability" properties (SP).
-The BPs are adapted from an existing criteria due to @Nik13a.
-A suitable simulation strategy should possess the following properties:
-
-* BP1: A wide range of dependences, allowing both positive and negative values, and, ideally, admitting the full range of possible values.
-* BP2: Flexible dependence, meaning that the number of bivariate marginals can be equal to the number of dependence parameters.
-* BP3: Flexible marginal modeling, generating heterogeneous data --- possibly from differing probability families.
-
-Moreover, the simulation method must *scale* to high dimensions:
-
-* SP1: Procedure must scale to high dimensions with practical compute times.
-* SP2: Procedure must scale to high dimensions while maintaining accuracy.
-
-## Motivating example: RNA-seq data
-
-Simulating high-dimensional, non-normal, correlated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients.
-The RNA-seq data-generating process involves counting how often a particular form of human messenger RNA (mRNA) is expressed in a biological sample.
-RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data.
-For human-derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted [@Schissler2019].
-Importantly, due to inherent biological processes, gene expression data exhibit correlation (co-expression) across genes [@BE07; @Schissler2018].
-
-We illustrate our methodology using a well-studied Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments).
-For simplicity, we only consider high expressing genes.
-In turn, we begin by filtering to retain the top `r d` of the highest-expressing genes (in terms of median expression) of the over 20,000 gene measurements from $N=`r nrow(example_brca)`$ patients' tumor samples.
-This gives a great number of pairwise dependencies among the marginals (specifically, $`r choose(d,2)`$ correlation parameters).
-Table \@ref(tab:ch010-realDataTab) displays RNA-seq counts for three selected high-expressing genes for the first five patients' breast tumor samples.
-
-```{r ch010-realDataTab}
-set.seed(2020-02-25)
-num_genes    <- 3
-num_patients <- 5
-gene_sample  <- sample(example_genes[1:1000], num_genes)
-
-cap <- paste0("mRNA expression for three selected high-expressing genes, ",
-              paste(gene_sample, collapse = ", "), 
-              ", for the first five patients in the TCGA BRCA data set.")
-
-small_brca <- example_brca %>%
-  select(all_of(gene_sample)) %>%
-  head(n = num_patients) %>%
-  as_tibble(rownames = "Patient ID") %>%
-  mutate(`Patient ID` = str_sub(`Patient ID`, end = 12))
-
-small_brca %>%
-  kable(booktabs = TRUE, caption = cap)
-```
-
-To help visualize the bivariate relationships for these three selected genes across all patients, Figure \@ref(fig:ch010-realDataFig) displays the marginal distributions and estimated Spearman's correlations (see Equation \@ref(eq:spearman) below).
-
-```{r ch010-realDataFig, cache=FALSE, fig.width=8, fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes from Table 1.
-The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations.
-Modeling these data motivate our simulation methodology."}
-ggpairs(
-  data = example_brca[, gene_sample],
-  upper = list(continuous = wrap('cor', method = "spearman"))
-) + theme_bw()
-```
-
-## Measures of dependency
-
-In multivariate analysis, an analyst must select a metric to quantify dependency.
-The most widely-known is the Pearson (product-moment) correlation coefficient that describes the linear association between two random variables $X$ and $Y$, and, it is given by
-
-\begin{equation}
-(\#eq:pearson)
-\rho_P(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ \mathrm{Var}(X)\mathrm{Var}(Y)\right]^{1/2}}.
-\end{equation}
-
-As @MB13 and @MK01 discuss, for a bivariate normal $(X,Y)$ random vector, the Pearson correlation completely describes the dependency between the components.
-For non-normal marginals with monotone correlation patterns, however, $\rho_P$ suffers some drawbacks and may mislead or fail to capture important relationships [@MK01].
-Alternatively in these settings, analysts often prefer rank-based correlation measures to describe the degree of monotonic association.
-
-Two nonparametric, rank-based measures common in practice are Spearman's correlation (denoted $\rho_S$) and Kendall's $\tau$.
-Spearman's $\rho_S$ has an appealing correspondence as the Pearson correlation coefficient on *ranks* of the values, thereby captures nonlinear yet monotone relationships.
-Kendall's $\tau$, on the other hand, is the difference in probabilities of concordant and discordant pairs of observations, $(X_i, Y_i)$ and $(X_j, Y_j)$, with concordance meaning that orderings have the same direction (e.g., if $X_i < X_j$, then $Y_i < Y_j$).
-Note that concordance is determined by the ranks of the values, not the values themselves.
-
-Both $\tau$ and $\rho_S$ are *invariant under monotone transformations* of the underlying random variates.
-As we will describe more fully in the [Algorithms](#algorithms) section, this property enables matching rank-based correlations with speed (SP1) and accuracy (SP2).
-
-*Correspondence among Pearson, Spearman, and Kendall correlations*
-
-There is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficient, as the marginal distributions $F_i$ are intrinsic in their calculation.
-For *bivariate normal vectors*, however, the correspondence is well-known:
-
-\begin{equation}
-(\#eq:convertKendall)
-\rho_{P} = \sin \left( \tau \times \frac{\pi}{2} \right), 
-\end{equation}
-
-\noindent and similarly for Spearman's $\rho$ [@K58],
-
-\begin{equation}
-(\#eq:convertSpearman)
-\rho_P = 2 \times \sin \left( \rho_S \times \frac{\pi}{6} \right).
-\end{equation}
-
-*Discrete marginal considerations*
-
-Spearman's correlation for discrete marginal suffers issues whenever there is a nonzero probability of ties.
-For example, the Spearman's correlation of a discrete-valued random variable $X$ with itself could be less than 1 [@MB13].
-One remedy for this issue is to rescale Equation \@ref(eq:spearman).
-For two random variables $X,Y$ with probability mass functions (PMFs) or probability densities functions (PDFs) $p(x)$ and $q(y)$, respectively, define the rescaled Spearman's correlation as
-
-\begin{equation}
-(\#eq:spearmanRescaled)
-\rho_{RS}(X,Y) = \frac{\rho_s(X,Y)}{ \left[ \left[ 1 - \sum_x p(x)^3 \right] \left[ 1 - \sum_y q(y)^3 \right] \right]^{1/2}}.
-\end{equation}
-
-Note that with continuous marginals the rescaling returns $\rho_S$.
-For discrete marginals with large or infinite support, computing the adjustment factors $\sum_x p(x)^3$, $\sum_y p(y)^3$ over all large number of pairs becomes expensive (often violating desired property SP1).
-Further, the infinite sums must be approximated for count-valued data.
-
-Also, for a discrete random variable $Y_i$, some care must be taken to define the quantile function $F_{i}^{-1}$. A standard practice is to set
-
-\begin{equation}
-F_{i}^{-1} = \inf\{y:F_{i}(y) \geq u \}.
-(\#eq:inverseCDF)
-\end{equation}
-
-*Marginal-dependent bivariate correlation bounds*
-
-Given two marginal distributions, $\rho_P$ is not free to vary over the entire range of possible correlations $[-1,1]$.
-The so-called *Frechet-Hoeffding bounds* are well-studied [@Nelsen2007; @BF17].
-These constraints cannot be overcome through algorithm design.
-In general, the bounds are given by
-
-\begin{equation}
-(\#eq:frechet)
-\rho_P^{max} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (U) \right), \quad \rho_P^{min} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (1 - U) \right)
-\end{equation}
-
-\noindent where $U$ is a uniform random variable on $(0,1)$ and $F^{-1}_1, F^{-1}_2$ are the inverse CDFs of $X_1$ and $X_2$, respectively, definitely by \@ref(eq:inverseCDF) when the variables are discrete.
-
-## Gaussian copulas
-
-There is a strong connection of our simulation strategy to Gaussian *copulas* (see @Nelsen2007 for a technical introduction).
-A copula is a distribution function on $[0,1]^d$ that describes a multivariate probability distribution with standard uniform marginals.
-This provides a powerful, natural way to characterize joint probability structures.
-Consequently, the study of copulas is an important and active area of statistical theory and practice.
-
-For any random vector ${\bf X}=(X_1, \ldots, X_d)^\top$ with CDF $F$ and marginal CDFs $F_i$ there is a copula function $C(u_1, \ldots, u_d)$ satisfying
-
-\begin{equation}
-(\#eq:copula)
-F(x_1, \ldots,x_d) = \mathbb P(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), 
-\end{equation}
-
-$x_i \in \mathbb R, i=1,\ldots,d.$
-
-A Gaussian copula has marginal CDFs that are all standard normal, $F_i = \Phi, \forall \, i$.
-This representation corresponds to a multivariate normal (MVN) distribution with standard normal marginal distributions and covariance matrix ${\bf R_P}$.
-As the marginals are standardized to have unit variance, however, ${\bf R_P}$ is a Pearson correlation matrix.
-If $F_{{\bf R}}$ is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula $C_{{\bf R}}$ is defined through
-
-\begin{equation}
-(\#eq:gauss)
-F_{{\bf R}}(x_1, \ldots, x_d) = C_{{\bf R}}(\Phi(x_1), \ldots, \Phi(x_d)),
-\end{equation}
-
-where $\Phi(\cdot)$ is the standard normal CDF.
-Note that the copula $C_{{\bf R}}$ is the familiar multivariate normal CDF of the random vector $(\Phi(X_1), \ldots, \Phi(X_d))$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})$.
-
-Sklar's Theorem [@Sklar1959; @Ubeda-Flores2017] guarantees that given inverse CDFs $F_i^{-1}$s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions.
-For example, using Gaussian copulas, we can construct a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)^\top$ with $Y_i \sim F_i$, viz.
-$Y_i = F_i^{-1}(\Phi(X_i)), i=1, \ldots, d$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})$.
-```{r include=FALSE, cache=FALSE}
-# Remove all objects
-rm(list = ls(all.names = TRUE))
-
-# Unload all packages
-```
-
-<!--chapter:end:010-background.Rmd-->
-
-```{r include=FALSE, cache=FALSE}
-options(digits = 4)
-
-knitr::opts_chunk$set(
-  # Text Results
-  eval = TRUE,
-  echo = FALSE,
-  results = 'markup',
-  collapse = TRUE,
-  message = FALSE,
-  warning = FALSE,
-  error = FALSE,
-  # Code Decoration
-  comment = NA,
-  background = '#F7F7F7',
-  # Cache
-  cache = TRUE,
-  # Plots
-  fig.show = "hold",
-  fig.width = 6.5,
-  fig.asp = 0.618,  # 1 / phi
-  out.width = "85%",
-  fig.align = 'center'
-)
-
-# options(dplyr.print_min = 6, dplyr.print_max = 6)
-
-seed_number <- 2020-02-25
-set.seed(seed = seed_number)
-```
-# Algorithms {#algorithms}
-
-This section describes our methods involved in simulating a random vector $\bf Y$ with $Y_i$ components for $i=1,\ldots,d$.
-Each $Y_i$ has a specified marginal CDF $F_i$ and its inverse $F^{-1}_i$.
-To characterize dependency, every pair $(Y_i, Y_j)$ has a given Pearson correlation $\rho_P$, Spearman correlation $\rho_S$, and/or Kendall's $\tau$.
-The method can be described as a *high-performance Gaussian copula* (Equation \@ref(eq:gauss)) providing a high-dimensional NORTA-inspired algorithm.
-
-## NORmal To Anything (NORTA)
-
-The well-known NORTA algorithm [@Cario1997] simulates a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$. Specifically, NORTA algorithm proceeds as follows:
-
-1. Simulate a random vector $\bf Z$ with $d$ *independent and identically distributed* (iid) standard normal components.
-2. Determine the input matrix $\Sigma_{\bf Z}$ that corresponds with the specified output $\Sigma_{\bf Y}$.
-3. Produce a Cholesky factor $M$ of $\Sigma_{\bf Z}$ such that $M M^{\prime}=\Sigma_{\bf Z}$.
-4. Set $X$ by $X \gets MZ$.
-5. $\text{Return} \; Y \; \text{where} \; Y_i \gets F_i^{-1}[\Phi(X_i)], \; i=1,...,d$.
-
-With modern parallel computing, steps 1, 3, 4, 5 are readily implemented as high-performance, multi-core and/or graphical-processing-unit (GPU) accelerated algorithms --- providing fast scalability using readily-available hardware.
-
-Matching specified Pearson correlation coefficients exactly (step 2 above), however, is computationally costly.
-In general, there is no closed-form correspondence between the components of the input $\Sigma_{\bf Z}$ and target $\Sigma_{\bf Y}$.
-Matching the correlations involves evaluating or approximating $\binom{d}{2}$ integrals of the form
-
-\begin{equation}
-    \mathrm{E}\left[Y_i Y_j\right] = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} F_i^{-1}\left[\Phi(z_i)\right] F_j^{-1}\left[\Phi(z_j)\right] \phi(z_i, z_j, \rho_z) dz_i dz_j,
-    (\#eq:pearsonIntegralRelation)
-\end{equation}
-
-where $\phi(\cdot)$ is the joint probability distribution function of two correlated standard normal variables.
-For HD data, (nearly) exact evaluation may become too costly to enable practical simulation studies.
-For low-dimensional problems, however, methods and tools exist to match Pearson correlations precisely; see [@Xia17] and the publicly available `nortaRA` R package [@Chen2001].
-As described later, to enable HD Pearson matching we approximate these integrals.
-
-*NORTA in higher dimensions*
-
-Sklar's theorem provides a useful characterization of multivariate distributions through copulas.
-Yet the choice of copula-based simulation algorithm affects which joint distributions may be simulated.
-Even in low-dimensional spaces (e.g., $d=3$), there exist valid multivariate distributions with *feasible* Pearson correlation matrices that NORTA cannot match exactly [@LH75].
-This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined the resultant matrix is indefinite.
-These situations do occur, even using exact analytic calculations.
-Such problematic target correlation matrices are termed *NORTA defective*.
-
-@GH02 conducted a Monte Carlo study to estimate the probability of encountering NORTA defective matrices while increasing the dimension $d$.
-They found that for what is now considered low-to-moderate dimensions ($d \approx 20$), almost *all* feasible matrices are NORTA defective.
-This stems from the concentration of measure near the boundary of the space of all possible correlation matrices as dimension increases.
-Unfortunately, it is precisely near this boundary that NORTA defective matrices reside.
-
-There is hope, however, as @GH02 also showed that replacing an indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$.
-This provides evidence that our nearest positive definite (PD) augmented approach will maintain reasonable accuracy if our input matching scheme returns an indefinite matrix, at least for the rank-based matching scheme described above.
-
-## Random vector generator {#rand-vec-gen}
-
-We now describe our algorithm to generate random vectors, which mirrors the classical NORTA algorithm above with some modifications for rank-based dependency matching:
-
-1. Mapping step
-    (i) Convert the target Spearman correlation matrix $R_S$ to the corresponding MVN Pearson correlation $R_X$. Alternatively,  
-    (i') Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$. Alternatively,  
-    (i'') Convert the target Pearson correlation matrix to $R_P$ to the corresponding, approximate MVN Pearson correlation $R_X$.  
-2. Check admissibility and, if needed, compute the nearest correlation matrix.
-    (i) Check that $R_X$ is a correlation matrix, a positive definite matrix with 1's along the diagonal.   
-	(ii) If $R_X$ is a correlation matrix, the input matrix is *admissible* in this scheme. Otherwise
-    (ii$^\prime$) Replace $R_X$ with the nearest correlation matrix $\tilde{R}_X$, in the Frobenius norm.  
-3. Gaussian copula
-    (i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)$.  
-    (ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz. $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
-    (iii) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$.
-
-*Step 1*
-
-The first two descriptions of the *Mapping step* employ the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertKendall) and \@ref(eq:convertSpearman), respectively (implemented as `cor_covert`).
-Initializing our algorithm to match the nonparametric correlations by computing these equations for all pairs is computationally trivial.
-
-For the computationally expensive process to match Pearson correlations, we approximate Equation \@ref(eq:pearsonIntegralRelation) for all pairs of margins.
-To this end, we implement the approximation scheme introduced by @xiao2019matching.
-Briefly, the many double integrals of the form in \@ref(eq:pearsonIntegralRelation) are approximated by weighted sums of Hermite polynomials.
-Matching coefficients for pairs of continuous distributions is made tractable by this method, but for discrete distributions (especially discrete distributions with large support sets or infinite support), the problem is not so simple or efficient.
-
-Our solution is to approximate discrete distributions by a continuous distribution.
-The question then becomes that of which distribution to use.
-We found that Generalized S-Distributions [@muino2006gs] solve this problem by approximating a wide range of unimodal distributions, both continuous and discrete.
-Using the GS-Distribution approximation of discrete distributions in Pearson matching scheme above yields favorable results.
-
-*Step 2*
-
-Once $R_X$ has been determined, we check admissibility of the adjusted correlation via the steps described above.
-If $R_X$ is not a valid correlation matrix, then we compute the nearest correlation matrix.
-Finding the nearest correlation matrix is a common statistical computing problem.
-The defacto default function in `R` is `Matrix::nearPD`, an alternating projection algorithm due to @higham2002computing.
-As implemented the function fails to scale to HD.
-Instead, we provide the quadratically convergent algorithm based on the theory of strongly semi-smooth matrix functions (@qi2006quadratically).
-The nearest correlation matrix problem can be written down as the following convex optimization problem:
-
-\begin{align*}
-    \mathrm{min}\quad & \frac{1}{2} \Vert G - X \Vert^2 \\
-    \mathrm{s.t.}\quad & X_{ii} = 1, \quad i = 1, \ldots , n, \\
-    & X \in S_{+}^{n}
-\end{align*}
-
-For nonparametric correlation measures, our algorithm allows the generation of high-dimensional multivariate data with arbitrary marginal distributions with a broad class of admissible Spearman correlation matrices and Kendall $\tau$ matrices.
-The admissible classes consist of the matrices that map to a Pearson correlation matrix for a MVN.
-In particular, if we let $X$ be MVN with $d$ components and set
-
-\begin{align*}
-\Omega_P &= \{ R_P : R_P \textrm{ is a correlation matrix for } X \} \\
-\Omega_K &= \{ R_K : R_K \textrm{ is a correlation matrix for } X \} \\
-\Omega_S &= \{ R_S : R_S \textrm{ is a correlation matrix for } X \} \\
-\end{align*}
-
-then there are 1-1 mappings between these sets.
-We conjecture informally that the sets of admissible $R_S$ and $R_K$ are not highly restrictive.
-In particular, $R_P$ is approximately $R_S$ for a MVN, suggesting that the admissible set $\Omega_S$ should be flexible as $R_P$ can be any PD matrix with 1's along the diagonal.
-We provide methods to check whether a target $R_S$ is an element of the *admissible set* $\Omega_S$ (and similarly for $R_K$).
-
-There is an increasing probability of encountering a non-admissible correlation matrix as dimension increases.
-In our experience, the mapping step for large $d$ almost always produces a $R_X$ that is not a correlation matrix.
-In Section \@ref(package) we provide a basic description of how to quantify and control the approximation error.
-Further, the RNA-seq example in Section \@ref(examples) provides an illustration of this in practice.
-
-*Step 3*
-
-The final step implements a NORTA-inspired, Gaussian copula approach to produce the desired margins.
-Steps 1 and 2 determine the MVN Pearson correlation values that will eventually match the target correlation.
-Then all that is required is a fast MVN simulator, a standard normal CDF, and well-defined quantile functions for marginals.
-The MVN is transformed to a copula (distribution with standard uniform margins) by applying the normal CDF $\phi(\cdot)$.
-The quantile functions $F^{-1}$ are applied across the margins to return the desired random vector ${\bf Y}$.
-
-```{r include=FALSE, cache=FALSE}
-# Remove all objects
-rm(list = ls(all.names = TRUE))
-
-# Unload all packages
-```
-
-<!--chapter:end:020-simulation-algorithm.Rmd-->
-
-```{r include=FALSE, cache=FALSE}
-options(digits = 4)
-
-knitr::opts_chunk$set(
-  # Text Results
-  eval = TRUE,
-  echo = FALSE,
-  results = 'markup',
-  collapse = TRUE,
-  message = FALSE,
-  warning = FALSE,
-  error = FALSE,
-  # Code Decoration
-  comment = NA,
-  background = '#F7F7F7',
-  # Cache
-  cache = TRUE,
-  # Plots
-  fig.show = "hold",
-  fig.width = 6.5,
-  fig.asp = 0.618,  # 1 / phi
-  out.width = "85%",
-  fig.align = 'center'
-)
-
-# options(dplyr.print_min = 6, dplyr.print_max = 6)
-
-seed_number <- 2020-02-25
-set.seed(seed = seed_number)
-```
-# The `bigsimr` R package {#package}
-
-```{r ch030-basic-example-setup, echo=TRUE, cache=FALSE}
-box::use(
-  patchwork[...],
-  dplyr[...],
-  ggplot2[...],
-  tidyr[drop_na],
-  bigsimr[bigsimr_setup, distributions_setup],
-  JuliaCall[julia_call]
-)
-
-Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
-bs <- bigsimr_setup(pkg_check = FALSE)
-dist <- distributions_setup()
-```
-
-This section describes a low-dimensional (2D) random vector simulation workflow via the `bigsimr` `R` package (https://github.com/SchisslerGroup/r-bigsimr).
-The package `bigsimr` provides an interface to the native code written in Julia, registered as the `Bigsimr` Julia package (https://github.com/adknudson/Bigsimr.jl).
-In addition to the native Julia `Bigsimr` package and `R` interface `bigsimr`, we also provide a python interface `bigsimr` (https://github.com/SchisslerGroup/python-bigsimr/) that interfaces with the Julia `Bigsimr` package.
-The Julia implementation provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms)).
-When designing `bigsimr`, we aimed to conveniently provide parallelized computation, through multi-core and GPU acceleration, while allowing advanced users to customize.
-
-The subsections below describe the basic use of `bigsimr`, by stepping through an example workflow using the data set `airquality` that contains daily air quality measurements in New York, May to September 1973 [@Chambers1983].
-This workflow proceeds from computing environment setup, data wrangling, estimation, simulation configuration, random vector generation, and, finally, result visualization.
-The code chunk below prepares the computing environment:
-
-<!-- 
-`bigsimr` is an R package for simulating high-dimensional multivariate data with arbitrary marginal distributions.
-The efficiency behind generating multivariate samples comes from the ability to utilize a GPU during the generation of multivariate normal samples and the subsequent transformation to uniform samples (see section \@ref(rand-vec-gen)).
-A GPU is not necessary in order to use `bigsimr`, and the speed benefits generally only come when simulating very high dimensional data ($d > 10000$).
-
-The next computational step after obtaining the correlated uniform marginals is the inverse transform into the target marginal distributions.
-Since this consists of $d$ independent transformations, we utilize parallelization to achieve higher throughput.
-As with many parallel algorithms, there is overhead associated with forking the task to utilize multiple cores, however cost is negligible compared to the total run-time.
--->
-
-## Bivariate example description
-
-We illustrate the use of `bigsimr` motivated by the New York air quality data set (`airquality`) included in the R `datasets` package.
-First, we load the `bigsimr` library and a few other convenient data science packages, including the syntactically-elegant `tidyverse` suite of `R` packages.
-
-For simplicity and to provide a minimal working example, we consider bivariate simulation of the two `airquality` variables `Temperature`, in degrees Fahrenheit, and `Ozone` level, in parts per billion.
-
-```{r ch030-air-quality-data, echo=TRUE}
-df <- airquality %>%
-  select(Temp, Ozone) %>%
-  drop_na()
-```
-
-```{r ch030-aq-glimpse}
-glimpse(df)
-```
-
-Figure \@ref(fig:ch030-aq-joint-dist) visualizes the bivariate relationship between Ozone and Temperature.
-We aim to simulate random two-component vectors mimicking this structure.
-The margins are not normally distributed; particularly the Ozone level exhibits a strong positive skew.
-
-```{r ch030-aq-joint-dist, cache=FALSE, fig.width= 8, fig.cap="Bivariate scatterplot of Ozone vs.
-Temp with estimated marginal densities.
-We model the Ozone data as marginally log-normal and the Temperature data as normal."}
-p0 <- ggplot(df, aes(Temp, Ozone)) +
-  geom_point(size = 1) +
-  theme(legend.position = "none") + 
-  labs(x = "Temperature")
-
-pTemp <- ggplot(df, aes(Temp)) + 
-  geom_density() +
-  theme(axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        axis.text = element_blank(),
-        axis.ticks = element_blank())
-
-pOzone <- ggplot(df, aes(Ozone)) + 
-    geom_density() +
-  theme(axis.title.y = element_blank(),
-        axis.title.x = element_blank(),
-        axis.ticks = element_blank(),
-        axis.text = element_blank()) +
-  coord_flip()
-
-pTemp + plot_spacer() + p0 + pOzone + 
-  plot_layout(widths = c(3,1), heights = c(1, 3)) 
-```
-
-Next, we specify the marginal distributions and correlation coefficient (both type and magnitude).
-Here the analyst is free to be creative.
-For this example, we avoid goodness-of-fit considerations to determine the marginal distributions.
-But it seems sensible without domain knowledge to estimate these quantities from the data and `bigsimr` contains fast functions designed for this task.
-
-## Specifying marginal distributions
-
-Based on the estimated densities in Figure \@ref(fig:ch030-aq-joint-dist), we assume `Temp` is normally distributed and `Ozone` is log-normally distributed, as the latter values are positive and skewed.
-We use the well-known, unbiased estimators for the normal distribution's parameters and maximum likelihood estimators for the log-normal parameters:
-
-```{r ch030-aq-temp-pars, echo=TRUE}
-df %>% 
-  select(Temp) %>% 
-  summarise_all(.funs = c(mean = mean, sd = sd))
-```
-
-```{r ch030-aq-ozone-pars, echo=TRUE}
-mle_mean <- function(x) mean(log(x))
-mle_sd <- function(x) mean( sqrt( (log(x) - mean(log(x)))^2 ) )
-
-df %>% 
-  select(Ozone) %>% 
-  summarise_all(.funs = c(meanlog = mle_mean, sdlog = mle_sd))
-```
-
-Next, we configure the input marginals for later input into `rvec`. The marginal distributions are specified using `Julia`'s `Distributions` package and stored in a vector.
-
-```{r ch030-margins-alist, echo=TRUE}
-margins <- c(
-  dist$Normal(mean(df$Temp), sd(df$Temp)),
-  dist$LogNormal(mle_mean(df$Ozone), mle_sd(df$Ozone))
-)
-```
-
-## Specifying correlation
-
-As mentioned, the user must decide how to describe correlation, based on the particulars of the problem.
-For non-normal data and for improved simulation accuracy/scalability in our scheme, we advocate the use of Spearman's $\rho$ correlation matrix $R_S$ and Kendall's $\tau$ correlation matrix $R_K$.
-We also support Pearson correlation coefficient matching, while cautioning the user to check the performance for their parametric multivariate model (see [Monte Carlo evaluations](#simulations) below for evaluation strategies and guidance).
-These estimation methods are classical approaches, not designed for high-dimensional correlation estimation (see the [Conclusion and Discussion]({#discussion) sections for more on this).
-
-```{r ch030-aq-cor, echo=TRUE}
-(R_S <- bs$cor(as.matrix(df), bs$Spearman))
-```
-
-## Checking target correlation matrix admissibility
-
-First we use `cor_bounds` to ensure that the pairwise target correlation values are valid prior to mapping step which constructs $R_X$ for the MVN input into NORTA (see Section \@ref(algorithms)).
-`cor_bounds` estimates the pairwise lower and upper correlation bounds using the Generate, Sort, and Correlate algorithm of @DH2011.
-
-```{r ch030-cor-bounds, echo=TRUE}
-bounds <- bs$cor_bounds(margins)
-bounds$lower
-```
-
-Since our single estimated Spearman correlation is within the theoretical bounds, the correlation is valid as input to `rvec`.
-But even if the 2-dimensional bounds are satisfied for each pair of margins, this does not guarantee the feasibility of a $d-$variate distribution [@BF17].
-
-To provide higher dimensional feasibility/admissibility checking, we begin by mapping using `cor_convert` and check admissibility.
-
-```{r ch030-cor-convert, echo=TRUE}
-# Step 1. Mapping
-(R_X <- bs$cor_convert(R_S, bs$Spearman, bs$Pearson))
-
-# Step 2. Check admissibility
-julia_call("Bigsimr.iscorrelation", R_X)
-```
-
-The bounds on the Pearson correlation coefficient $R_P$ between these margins are restricted as seen below in our MC estimated correlation bounds:
-
-```{r ch030-cor-bounds-pearson, echo=TRUE}
-bs$cor_bounds(margins[1], margins[2], bs$Pearson, n_samples = 1e6)
-```
-
-Our MC estimate of the bounds slightly underestimates the theoretic bounds of $(-0.881, 0.881)$. An analytic derivation is presented as an Appendix.
-
-## Simulating random vectors
-Finally, we arrive at the main function of `bigsimr`, `rvec`. We now simulate $B=10,000$ random vectors from the assumed joint distribution of Ozone levels and Temp.
-
-```{r ch030-sim-margins, echo=TRUE}
-x <- bs$rvec(10000, R_X, margins)
-df_sim <- as.data.frame(x)
-colnames(df_sim) <- colnames(df)
-```
-
-Figure \@ref(fig:ch030-plot-sim) plots the 10,000 simulated points.
-
-```{r ch030-plot-sim, fig.cap="Contour plot and marginal densities for the simulated bivariate distribution of Air Quality Temperatures and Ozone levels.
-The simulated points mimic the observed data with respect to both the marginal characteristics and bivariate association."}
-p1 <- df_sim %>%
-  ggplot(aes(Temp, Ozone)) +
-  geom_density_2d_filled() +
-  theme(legend.position = "none") +
-  labs(x = "Simulated Temperature", y = "Simulated Ozone") +
-  scale_y_continuous(limits = c(0, max(df$Ozone))) +
-  scale_x_continuous(limits = range(df$Temp))
-
-p1Temp <- ggplot(df_sim, aes(Temp)) + 
-  geom_density(alpha = 0.5, fill = "lightseagreen") +
-  theme(axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        axis.text = element_blank(),
-        axis.ticks = element_blank())
-
-p1Ozone <- ggplot(df_sim, aes(Ozone)) + 
-  geom_density(alpha = 0.5, fill = "lightseagreen") + 
-  theme(axis.title.y = element_blank(),
-        axis.title.x = element_blank(),
-        axis.ticks = element_blank(),
-        axis.text = element_blank()) +
-  coord_flip()
-
-p1Temp + plot_spacer() + p1 + p1Ozone + 
-  plot_layout(widths = c(3,1), heights = c(1, 3))
-```
-```{r include=FALSE, cache=FALSE}
-# Remove all objects
-rm(list = ls(all.names = TRUE))
-
-# Unload all packages
-```
-
-<!--chapter:end:030-bigsimr-package.Rmd-->
-
-```{r include=FALSE, cache=FALSE}
-options(digits = 4)
-
-knitr::opts_chunk$set(
-  # Text Results
-  eval = TRUE,
-  echo = FALSE,
-  results = 'markup',
-  collapse = TRUE,
-  message = FALSE,
-  warning = FALSE,
-  error = FALSE,
-  # Code Decoration
-  comment = NA,
-  background = '#F7F7F7',
-  # Cache
-  cache = TRUE,
-  # Plots
-  fig.show = "hold",
-  fig.width = 6.5,
-  fig.asp = 0.618,  # 1 / phi
-  out.width = "85%",
-  fig.align = 'center'
-)
-
-# options(dplyr.print_min = 6, dplyr.print_max = 6)
-
-seed_number <- 2020-02-25
-set.seed(seed = seed_number)
-```
-# Monte Carlo evaluations {#simulations}
-
-```{r ch040-LoadLib040, include=FALSE}
-box::use(
-    dplyr[...],
-    ggplot2[...],
-    patchwork[...],
-    tidyr[pivot_longer],
-    bigsimr[bigsimr_setup, distributions_setup],
-    knitr[kable],
-    RColorBrewer[brewer.pal]
-)
-load("data/bivariate_normal_sims.rda")
-load("data/bivariate_gamma_sims.rda")
-load("data/bivariate_nbinom_sims.rda")
-load("data/benchmark_dependences.rda")
-
-Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
-bs <- bigsimr_setup(pkg_check = FALSE)
-dist <- distributions_setup()
-prob <- 3e-04
-```
-
-Before applying our methodology to real data simulation, we conduct several Monte Carlo studies to investigate method performance.
-Since marginal parameter matching in our scheme is essentially a sequence of univariate inverse probability transforms, the challenging aspects are the accuracy of dependency matching and computational efficiency at high dimensions.
-To evaluate our methods in those respects, we design the following numerical experiments to first assess accuracy of matching dependency parameters in bivariate simulations and then time the procedure in increasingly large dimension $d$.
-
-## Bivariate experiments
-
-We select bivariate simulation configurations to ultimately simulate our motivating discrete-valued RNA-seq example, and, so we proceed in increasing complexity, leading to the model in our motivating application in Section \@ref(examples).
-We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson's, Spearman's, and Kendall's --- in identical, bivariate marginal configurations.
-For each pair of identical margins, we vary the target correlation across $\Omega$, the set of possible admissible values for correlation type, to evaluate the simulation's ability to obtain the theoretic bounds.
-The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (mimicking RNA-seq counts).
-
-Table \@ref(tab:sims) lists our identical-marginal, bivariate simulation configurations.
-We increase the simulate replicates $B$ to check that our results converge to the target correlations and gauge statistical efficiency.
-We select distributions beginning with a standard multivariate normal (MVN) as we expect the performance to be exact (up to MC error) for all correlation types.
-Then, we select a non-symmetric continuous distribution: a standard (rate =1) two-component multivariate gamma (MVG).
-Finally, we select distributions and marginal parameter values that are motivated by our RNA-seq data, namely values proximal to probabilities and sizes estimated from the data (see [Example applications](examples) for estimation details).
-Thus we arrive at a multivariate negative binomial (MVNB) $p_1 = p_2 = 3\times10^{-4}, r_1 = r_2 = 4, \rho \in \Omega$.
-
-Table: (\#tab:sims) Identical margin, bivariate simulation configurations to evaluate correlation matching accuracy and efficiency.
-
-| Simulation Reps ($B$) | Correlation Types | Identical-margin 2D distribution |
-|-------------|:--------------:|----------------------:|
-| $1000$    | Pearson ($\rho_P$) | ${ \bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho_i ), i=1,\ldots,100$ |
-| $10,000$  | Spearman ($\rho_S$) | ${ \bf Y} \sim MVG( shape = 10, rate = 1, \rho_i ), i=1,\ldots,100$ |
-| $100,000$ | Kendall ($\tau$)| ${ \bf Y} \sim MVNB(p = 3\times10^{-4}, r = 4,\rho_i), i=1,\ldots,100$ |
-
-For each of the unique 9 simulation configurations described above, we estimate the correlation bounds and vary the correlations along a sequence of 100 points evenly placed within the bounds, aiming to explore $\Omega$.
-Specifically, we set correlations $\{ \rho_1 = ( \hat{l} + \epsilon), \rho_2 = (\hat{l} + \epsilon) + \delta, \ldots, \rho_{100} = (\hat{u} - \epsilon) \}$, with $\hat{l}$ and $\hat{u}$ being the estimated lower and upper bounds, respectively, and increment value $\delta$.
-The adjustment factor, $\epsilon=0.01$, is introduced to handle numeric issues when the bound is specified exactly.
-
-```{r ch040-biNormPlot, eval=FALSE, fig.height=5, fig.width=8, fig.cap="`bigsimr` recovers the Pearson specified correlations for MVN."}
-bivariate_normal_sims %>%
-    ggplot(aes(rho, rho_hat, color = type)) +
-    geom_point() +
-    geom_abline(slope = 1) +
-    facet_wrap(~ type + N) + 
-    theme_bw()
-```
-
-```{r ch040-biGammaPlot, eval=FALSE, fig.height=5, fig.width=8, fig.cap="`bigsimr` recovers the Pearson specified correlations for Bivariate Gamma."}
-bivariate_gamma_sims %>%
-    ggplot(aes(rho, rho_hat, color = type)) +
-    geom_point() +
-    geom_abline(slope = 1) +
-    facet_wrap(~ type + N) + 
-    theme_bw()
-```
-
-```{r ch040-biNegBinPlot, eval=FALSE, fig.width=8, fig.height=5, fig.cap="`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
-bivariate_nbinom_sims %>%
-    ggplot(aes(rho, rho_hat, color = type)) +
-    geom_point() +
-    geom_abline(slope = 1) +
-    facet_wrap(~ type + N) + 
-    theme_bw()
-```
-
-```{r ch040-combineBiSims, cache=FALSE}
-allDat <- bind_rows(
-    select(bivariate_normal_sims, margins, type, N, rho, rho_hat),
-    select(bivariate_gamma_sims, margins, type, N, rho, rho_hat),
-    select(bivariate_nbinom_sims, margins, type, N, rho, rho_hat)
-) %>%
-    mutate(margins = factor(margins, levels=c("norm", "gamma", "nbinom")),
-           type = factor(type, levels=c("Pearson", "Spearman", "Kendall")),
-           N = factor(N))
-```
-
-Figure \@ref(fig:ch040-bPlot) displays the aggregated bivariate simulation results.
-Table \@ref(tab:ch040-BiError) contains the mean absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.
-
-```{r ch040-BiError}
-tabMAE <- allDat %>%
-    group_by(N, type, margins) %>%
-    summarize(MAE = mean(abs(rho - rho_hat))) %>%
-    ungroup()
-
-kable(tabMAE, booktabs = TRUE,
-      col.names = c("No. of random vectors",
-                    "Correlation type",
-                    "Distribution",
-                    "Mean abs. error"),
-      caption = "Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.")
-```
-
-Overall, the studies show that our methodology is generally accurate across the entire range of possible correlation values all three dependency measures, at least in these limited simulation settings for the rank-based correlations.
-Our Pearson matching performs nearly as well as Spearman or Kendall, except a slight increase in error for negative binomial case.
-This is due the particularly large counts generated with our choice of parameters for $p$ and $r$.
-
-```{r ch040-bPlot, cache=FALSE, fig.asp=1.30, fig.weight=4, fig.cap="Bivariate simulations match target correlations across the entire range of feasible correlations.
-The horizontal axis plots the specified target correlations for each bivariate margin.
-Normal margins are plotted in dark dark grey, gamma in medium grey, and negative binomial in light grey.
-As the number of simulated vectors $B$ increases from left to right, the variation in estimated correlations (vertical axis) decreases.
-The dashed line indicates equality between the specified and estimated correlations."}
-# https://www.datanovia.com/en/blog/how-to-change-ggplot-facet-labels/
-# New facet label names
-repsLabs <- paste0("B=", c("1,000", "10,000", "100,000") )
-names(repsLabs) <- c(1000, 10000, 100000)
-typeLabs <- c("Pearson", "Spearman", "Kendall")
-names(typeLabs) <- c("Pearson", "Spearman", "Kendall")
-
-# Set colors
-## RColorBrewer::display.brewer.all()
-numColors <- 4
-numGroups <- length(levels(allDat$margins))
-myColors <- rev(RColorBrewer::brewer.pal(n = numColors, name = "Greys")[ ((numColors - numGroups) + 1):numColors])
-
-allDat %>%
-    ggplot(aes(x = rho, y = rho_hat, color = margins)) +
-    geom_point(size = 2) +
-    scale_color_manual(values = myColors ) +
-    geom_abline(slope = 1, linetype = "dashed") +
-    labs(x = "Specified Correlation", y = "Estimated Correlation") +
-    facet_wrap(~ type + N, labeller = labeller(N = repsLabs, type = typeLabs)) +
-    theme_bw() +
-    theme(legend.position = "bottom", legend.direction = "horizontal")
-```
-
-<!-- 
-See [Discussion](discussion) for future directions for fast Pearson matching and discrete-specific modifications.
--->
-
-## Scale up to High Dimensions
-
-With information of our method's accuracy from a low-dimensional perspective, we now turn to assessing whether the `bigsimr` can scale to larger dimensional problems with practical computation times.
-Specifically, we ultimately generate $B=1,000$ random vectors for $d=\{100, 250, 500, 1000, 2500, 5000, 10000\}$ for each correlation type, $\{Pearson, Spearman, Kendall\}$ while timing the algorithm's major steps.
-
-To mimic the workflow, we first produce a synthetic "data set" by completing the following steps:
-
-1.
-Produce heterogeneous gamma marginals by randomly selecting the $j^{th}$ gamma shape parameter from $U_j \sim uniform(1,10), j=1,\ldots,d$ and the $j^{th}$ rate parameter from $V_j \sim exp(1/5), j=1,\ldots,d$, with the constant parameters determined arbitrarily.
-2. Produce a random full-rank Pearson correlation matrix via `cor_randPD` of size $d \times d$.  
-3. Simulate a "data set" of $1,000 \times d$ random vectors via `rvec` (without matching the Pearson exactly).
-
-With the synthetic data set in hand, we complete and time the following 4 steps:
-
-i. Estimate the correlation matrix from the "data" in the *Compute Correlation* step.  
-1. Map the correlations to initialize the algorithm (Pearson to Pearson, Spearman to Pearson, or Kendall to Pearson) in the *Adjust Correlation* step.  
-2. Check whether the mapping produces a valid correlation matrix and, if not, find the nearest PD correlation matrix in the *Check Admissibility* step.  
-3. Simulate $1,000$ vectors via the Inverse Transform method in the *Simulate Data* step.
-
-The experiments are conducted on a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, with all 16 threads employed during computation.
-The results for $d \leq 500$ are fast with all times under 3 seconds.
-Table \@ref(tab:ch040-moderateDtab) displays the total computation time of the algorithms steps 1 through 3, including estimation step i.
-
-```{r ch040-moderateDtab}
-dat_long <- benchmark_dependences %>%
-  select(-total_time, -n_sim, -needed_near_pd) %>%
-  pivot_longer(cols = c(corr_time:sim_time), names_to = "Step", values_to = "Time") %>%
-  mutate(dim = factor(dim),
-         Step = factor(Step,
-                       levels = c("corr_time", "adj_time", "admiss_time", "sim_time"),
-                       labels = c("Compute Correlation",
-                                  "Adjust Correlation",
-                                  "Check Admissibility",
-                                  "Simulate Data"))) %>%
-    rename(Correlation = corr_type, Dimensions = dim)
-
-tabTimes  <- dat_long %>%
-    filter(Dimensions %in% c(100, 250, 500)) %>%
-    group_by(Dimensions, Correlation) %>%
-    summarize('Total Time(Seconds)' = sum(Time)) %>%
-    ungroup()
-
-tabTimes %>%
-    kable(booktabs = TRUE,
-          col.names = c("Dimension",
-                        "Correlation type",
-                        "Total Time (Seconds)"),
-          caption = 'Total time to produce 10,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.')
-```
-
-The results larger dimensional vectors show scalability to ultra-high dimensions for all three correlation types, although the total times do become much larger. 
-Figure \@ref(fig:ch040-largeDfig) displays computation times for $d=\{1000, 2500, 5000, 10000\}$.
-For $d$ equal to 1000 and 2500, the total time is under a couple of minutes.
-At $d$ of 5000 and 10,000, Pearson correlation matching in the *Adjust Correlation* step becomes costly.
-Interestingly, Pearson is actually faster than Kendall for $d=10,000$ due to bottlenecks in *Compute Correlation* and *Check Admissibility*.
-Uniformly, matching Spearman correlations is faster, with total times under 5 minutes for $d=10,000$, making Spearman correlations the most computationally-friendly dependency type.
-With this in mind, we scaled the simulation to $d=20,000$ for the Spearman type and obtained the 1,000 vectors in under an hour (data not shown).
-In principle, this would enable the simulation of an entire human-derived RNA-seq data set.
-We note that for a given target correlation matrix and margins, steps i, 1, and 2 only need to be computed once and the third step, *Simulate Data*, is fast under all schemes for all $d$ under consideration.
-
-```{r ch040-largeDfig, fig.cap="Computation times as d increases."}
-# Set colors
-numColors <- 5
-numGroups <- 4
-myColors <- rev(brewer.pal(n=numColors, name="Greys")[((numColors-numGroups)+1):numColors])
-
-dat_long %>%
-    filter(Dimensions %in% c(1000, 2500, 5000, 10000, 20000)) %>%
-    mutate(`Time (minutes)` = Time / 60) %>%
-    ggplot(aes(Correlation, `Time (minutes)`, fill=Step)) +
-    geom_bar(position = "stack", stat = "identity") +
-    scale_fill_manual(values = myColors ) +
-    scale_y_continuous(breaks = c(0, 5, 10, 15, 20, 25, 30),
-                     minor_breaks = NULL) +
-    facet_grid(. ~ Dimensions) +
-    theme(axis.text.x = element_text(angle = -90))
-```
-
-*Limitations, conclusions, and recommendations.*
-In the bivariate studies, we chose arbitrary simulation parameters for three distributions, moving from the Gaussian to the discrete and non-normal, multivariate negative binomial.
-Under these conditions, the simulated random vectors sample the desired bivariate distribution across the entire range of pairwise correlations for the three dependency measures.
-The simulation results could differ for other choices of simulation settings.
-Specifying extreme correlations near the boundary or Frechet bounds could result in poor simulation performance.
-Fortunately, it is straightforward to evaluate simulation performance by using strategies similar to those completed above.
-We expect our random vector generation to perform well for the vast majority of NORTA-feasible correlation matrices, but advise to check the performance before making inferences/further analyses.
-Finally, we note that one could use our single-pass algorithm `cor_fastPD` to produce a "close" (not nearest PD) to scale to even higher dimensions with greater loss of accuracy.
-```{r include=FALSE, cache=FALSE}
-# Remove all objects
-rm(list = ls(all.names = TRUE))
-
-# Unload all packages
-```
-
-<!--chapter:end:040-monte-carlo-evaluation.Rmd-->
-
-```{r include=FALSE, cache=FALSE}
-options(digits = 4)
-
-knitr::opts_chunk$set(
-  # Text Results
-  eval = TRUE,
-  echo = FALSE,
-  results = 'markup',
-  collapse = TRUE,
-  message = FALSE,
-  warning = FALSE,
-  error = FALSE,
-  # Code Decoration
-  comment = NA,
-  background = '#F7F7F7',
-  # Cache
-  cache = TRUE,
-  # Plots
-  fig.show = "hold",
-  fig.width = 6.5,
-  fig.asp = 0.618,  # 1 / phi
-  out.width = "85%",
-  fig.align = 'center'
-)
-
-# options(dplyr.print_min = 6, dplyr.print_max = 6)
-
-seed_number <- 2020-02-25
-set.seed(seed = seed_number)
-```
-# RNA-seq data applications  {#examples}
-
-```{r ch050-LoadLib, include=FALSE, cache=FALSE}
-box::use(
-  dplyr[...],
-  ggplot2[...],
-  GGally[ggpairs, wrap],
-  patchwork[...],
-  bigsimr[...],
-  knitr[include_graphics],
-  JuliaCall[julia_call],
-  ./R/utils[mom_nbinom]
-)
-load("data/example_brca.rda")
-load("data/example_genes.rda")
-sim_nbinom <- readRDS("results/brca_1000_sim.rds")
-simRho <- readRDS("results/brca_1000_frobenius.rds")
-
-Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
-bs <- bigsimr_setup(pkg_check = FALSE)
-dist <- distributions_setup()
-```
-
-This section demonstrates how to simulate multivariate data using `bigsimr`, aiming to replicate the structure of high-dimensional dependent count data.
-In an illustration of our proposed methodology, we seek to simulate RNA-sequencing data by producing simulated random vectors mimicking the observed data and its generating process.
-Modeling RNA-seq using multivariate probability distributions is natural as inter-gene correlation is an inherent part of biological processes [@Wang2009b].
-And yet, many models do not account for this, leading to major disruptions to the operating characteristics of statistical estimation, testing, and prediction.
-See @Efron2012 for a detailed discussion with related methods and @Wu2012b, @Schissler2018; @Schissler2019 for applied examples.
-The following subsections apply `bigsimr`'s methods to real RNA-seq data, including replicating an estimated parametric structure, MC probability estimation, and MC evaluation of correlation estimation efficiency.
-
-## Simulating High-Dimensional RNA-seq data
-
-```{r ch050-readBRCA, cache=FALSE}
-d <- 1000
-brca1000 <- example_brca %>%
-  select(all_of(1:d)) %>%
-  mutate(across(everything(), as.double))
-```
-
-We begin by estimating the structure of the TCGA BRCA RNA-seq data set (see [Background](background)).
-Ultimately, we will simulate $B=10,000$ random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with $d=`r d`$.
-We assume a multivariate negative binomial (MVNB) model as RNA-seq counts are often over-dispersed and correlated.
-Since all $d$ selected genes exhibit over-dispersion (data not shown), we proceed to estimate the NB parameters $(r_i, p_i), i=1,\ldots,d$, to determine the target marginal PMFs $f_i$.
-To complete specification of the simulation algorithm inputs, we estimate the Spearman correlation matrix ${ \bf R}_{S}$ to characterize dependency.
-
-<!--
-This suggests simulating high-dimensional multivariate NB (MVB) with heterogeneous marginals would be useful tool in the development and evaluation of RNA-seq analytics.
-This procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions.
-All these genes exhibit over-dispersion and, so, we proceed to estimate the NB parameters $(r_i, p_i), i=1,\ldots,d$ to determine the target marginal PMFs $g_i(y_i)$ (via method of moments).
-Notably, the $\hat{p}_i's$ are small --- ranging in $[3.934 \times 10^{-6} , 1.217 \times 10^{-2}]$.
-To complete the simulation algorithm inputs, we estimate the Pearson correlation matrix $\bf{R_Y}$ and set that as the target correlation.
--->
-
-<!--
-With the simulation targets specified, we proceed to simulate $N=10,000$ random vectors $\bf{Y}$ $=( Y_1,\ldots,Y_d)$ with target Pearson correlation $\bf{R_Y}$ and marginal PMFs $g_i(y_i)$ using a $\bf{T}$-Poisson hierarchy of Kind II.
-Specifically, we first employ the \emph{direct Gaussian copula} approach to generate $N$ random vectors following a standard multivariate Gamma distribution $\bf{T}$ with shape parameters $r_i$ equal to the target $n_i$ and Pearson correlation matrix $\bf{R_T}$.
-Care must be taken when setting the specifying $\bf{R}$ (refer to Equation \ref{gay.cop.pdf}) --- we employ Equation \ref{mix.poi.corr} to compute the scaling factors $c_{i,j}$ and adjust the underlying correlations to ultimately match the target $\bf{R_Y}$.
-Notably, of the $525,825$ pairwise correlations from the $1026$ genes, no scale factor was less than $0.9907$, indicating the model can produce essentially the entire range of possible correlations.
-Here we are satisfied with approximate matching of the specified Gamma correlation and set $\bf{R}$ = $\bf{R_T}$ in our Gaussian copula scheme ($\bf{R}$ indicating the specified multivariate Gaussian correlation matrix).
-Finally, we generate the desired random vector $Y_i=N_i(t_i)$ by simulating Poisson counts with expected value $\mu_i=\lambda_i \times T_i$, for $i=1,\ldots,d$ (with $\lambda_i=\frac{(1-p_i)}{p_i}$) and repeat $N=10,000$ times.
--->
-
-With this goal in mind, we first estimate the desired correlation matrix using the fast implementation provided by `bigsimr`:
-
-```{r ch050-estRhoBRCA, echo=TRUE, cache=FALSE}
-# Estimate Spearman's correlation on the count data
-R_S <- bs$cor(as.matrix(brca1000), bs$Spearman)
-```
-
-Next, we estimate the marginal parameters.
-We use the method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model.
-While, marginal distributions are from the same probability family (NB), they are heterogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.
-The functions below support this estimation for later use in `rvec`.
-
-<!-- 
-TYPESTE MoM estimators here.
--->
-
-```{r ch050-nbHelpers, echo=TRUE}
-make_nbinom_margins <- function(sizes, probs) {
-  margins <- lapply(1:length(sizes), function(i) {
-    dist$NegativeBinomial(sizes[i], probs[i])
-  })
-  do.call(c, margins)
-}
-```
-
-We apply these estimators to all `r d` genes across the `r nrow(brca1000)` patients:
-
-```{r ch050-estMargins, echo=TRUE}
-nbinom_fit <- apply(brca1000, 2, mom_nbinom)
-sizes <- nbinom_fit["size",]
-probs <- nbinom_fit["prob",]
-nb_margins <- make_nbinom_margins(sizes, probs)
-```
-
-Notably, the estimated marginal NB probabilities $\{ \hat{p}_i \}$ are small --- ranging in the interval $[`r min(probs)` , `r max(probs)`]$.
-This gives rise to highly variable counts and, typically, less restriction on potential pairwise correlation pairs.
-Once the functions are defined/executed to complete marginal estimation, we specify targets and generate the desired random vectors using `rvec`.
-Now we check admissibility of specified correlation matrix.
-
-```{r ch050-cor-check, echo=TRUE}
-# 1. Mapping step first
-R_X <- bs$cor_convert(R_S, bs$Spearman, bs$Pearson)
-
-# 2a. Check admissibility
-(is_valid_corr <- julia_call("Bigsimr.iscorrelation", R_X))
-
-# 2b. compute nearest correlation
-if (!is_valid_corr) {
-  R_X_pd <- bs$cor_nearPD(R_X)
-  ## Quantify the error
-  targets      <- R_X[lower.tri(R_X, diag = FALSE)]
-  approximates <- R_X_pd[lower.tri(R_X_pd, diag = FALSE)]
-  R_X          <- R_X_pd
-}
-summary(abs(targets - approximates))
-```
-
-While the exact $d \times d$ Spearman correlation matrix is not strictly admissible in our scheme (as seen by the non-positive definite result above), the approximation is close with a maximum absolute error of `r max( abs(targets - approximates))` and average absolute error of `r mean(abs(targets - approximates))` across the `r choose(d,2)` correlations.
-
-```{r ch050-runBRCA-echo, echo=TRUE, eval=FALSE}
-sim_nbinom <- bs$rvec(10000, R_X, nb_margins) 
-colnames(sim_nbinom) <- colnames(brca1000)
-```
-
-Figure \@ref(fig:ch050-simDataFig) displays the simulated counts and pairwise relationships for our example genes from Table \@ref(tab:ch010-realDataTab).
-Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parametric form and with less extreme points then the observed data in Figure \@ref(fig:ch010-realDataFig).
-
-```{r ch050-simDataFig, cache=FALSE, fig.cap="Simulated data for three selected high-expressing genes generally replicates the estimated data structure.
-The data do not exhibit outlying points, but do possess the desired Spearman correlations, central tendencies, and discrete values."}
-set.seed(2020-02-25)
-num_genes <- 3
-gene_sample <- sample(example_genes[1:1000], num_genes)
-
-ggpairs(data = as.data.frame(sim_nbinom[, gene_sample]),
-        upper = list(continuous = wrap('cor', method = "spearman"))) + 
-  theme_bw()
-```
-
-Figure \@ref(fig:ch050-figBRCA) displays the aggregated results of our simulation by comparing the specified target parameter (horizontal axes) with the corresponding quantities estimated from the simulated data (vertical axes).
-The evaluation shows that the simulated counts approximately match the target parameters and exhibit the full range of estimated correlation from the data.
-
-<!-- 
-Utilizing 15 CPU threads in a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, the simulation completed in just over 2 minutes.
--->
-
-```{r ch050-figBRCA, out.width='80%', fig.cap="Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set.
-The dashed red lines indicate equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes)."}
-include_graphics('fig/ch050-figBRCA.png')
-```
-
-*Limitations, conclusions, and recommendations.*
-The results show overall good simulation performance for our choice of parameters settings.
-Our settings were motivated by modeling high-expressing genes from TCGA BRCA data set.
-In general, the ability to match marginal and dependence parameters depends on the particular joint probability model.
-We recommend to evaluate and tune your simulation until you can be assured of the accuracy.
-
-## Simulation-based joint probability calculations
-
-Many statistical tasks require evaluation of a joint probability mass (or density) function:
-
-$$
-P( {\bf Y} = {\bf y} ), \: y_i \in \chi_i.
-$$
-
-where $\chi_i$ is the sample space for the $i^{th}$ component of the random vector $\bf{Y}$.
-Compact representations with convenient computational forms are rare for high-dimensional constructions, especially with heterogeneous, correlated marginal distributions (or margins of mixed data types).
-Given a large number of simulated vectors as produced above, estimated probabilities are readily given by counting the proportion of simulated vectors meeting the desired condition.
-In our motivating application, one may ask what is the probability that all genes expressed greater than a certain threshold value ${ \bf y}_0$.
-This can be estimated as
-
-$$
-\hat{P}( {\bf Y} \ge {\bf y_0 } ) = \frac{1}{B} \sum_{b=1}^B I( {\bf Y}^{ (b) } \ge {\bf y_0 } ),
-$$
-
-```{r ch050-densityEvaluation, include=FALSE}
-d <- ncol(sim_nbinom)
-B <- nrow(sim_nbinom)
-y0 <- 1
-pHat <- mean(apply(sim_nbinom, 1, function(Y) {
-  all(Y >= y0)
-}))
-```
-
-where ${\bf Y^{(b)} }$ is the $b^{th}$ simulated vector from a total of $B$ simulation replicates and $I(\cdot)$ is the indicator function.
-For example, we can estimate from our $B=10,000$ simulated vectors that the probability of all genes expressing (i.e., ${\bf y}_i \geq 1, \forall \; i$) is $`r pHat`$.
-
-```{r ch050-densityEvaluationECHO, echo=TRUE}
-<<ch050-densityEvaluation>>
-pHat
-```
-
-## Evaluation of correlation estimation efficiency
-
-MC methods are routinely used in many statistical inferential tasks including estimation, hypothesis testing, error rates, and empirical interval coverage rates.
-To conclude the example applications, we demonstrate how `bigsimr` can be used to evaluate estimation efficiency.
-In particular, we wish to assess the error in our correlation estimation above.
-We used a conventional method, based on classical statistical theory which was not designed for high-dimensional data.
-Indeed, high-dimensional covariance estimation (and precision matrices) is an active area of statistical science, .e.g., [@Won2013g; @VanWieringen2016].
-
-In this small example, we simulate $m=30$ data sets with the number of simulated vectors matching the number of patients in the BRCA data set, $N=`r nrow(example_brca)`$.
-Since our simulation is much faster for the Pearson correlation type (see Figure \@ref(fig:ch040-largeDfig)), we only convert the Spearman correlation matrix once (and ensure it is positive definite).
-At each iteration, we estimate the quadratic loss (residual sum of squared errors) from the specified ${\bf R}_{S}$, producing a distribution of loss values.
-
-```{r ch050-quadlossECHO, echo=TRUE, eval=FALSE}
-# Simulate random vectors equal to the sample size
-N <- nrow(example_brca)
-# create m random vectors and estimate correlation
-simRho <- replicate(n = m, expr = {
-  tmpSim <- bs$rvec(N , R_X, nb_margins)
-  bs$cor(tmpSim, bs$Spearman)
-}, simplify = FALSE)
-# Evaluate the residual sum of squared error
-sapply(simRho, function(R) sum((R - R_S)^2))
-```
-
-```{r ch050-lossResults}
-frobenius_loss <- sapply(simRho, function(R) sum((R - R_S)^2))
-```
-
-The `R` summary function supplies the mean-augmented five-number summary of the quadratic loss distribution computed above.
-
-```{r ch050-loss-summary, echo=TRUE}
-summary(frobenius_loss)
-```
-
-This distribution could be compared to high-dimensional designed covariance estimators for guidance on whether the additional complexity and computation time are warranted.
-```{r include=FALSE, cache=FALSE}
-# Remove all objects
-rm(list = ls(all.names = TRUE))
-
-# Unload all packages
-```
-
-<!--chapter:end:050-examples.Rmd-->
-
-```{r include=FALSE, cache=FALSE}
-options(digits = 4)
-
-knitr::opts_chunk$set(
-  # Text Results
-  eval = TRUE,
-  echo = FALSE,
-  results = 'markup',
-  collapse = TRUE,
-  message = FALSE,
-  warning = FALSE,
-  error = FALSE,
-  # Code Decoration
-  comment = NA,
-  background = '#F7F7F7',
-  # Cache
-  cache = TRUE,
-  # Plots
-  fig.show = "hold",
-  fig.width = 6.5,
-  fig.asp = 0.618,  # 1 / phi
-  out.width = "85%",
-  fig.align = 'center'
-)
-
-# options(dplyr.print_min = 6, dplyr.print_max = 6)
-
-seed_number <- 2020-02-25
-set.seed(seed = seed_number)
-```
-# Conclusion and discussion {#discussion}
-
-We have introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a user-friendly, high-performance `R` package [`bigsimr`].
-The random vector generation method is inspired by NORTA [@Cario1997] and Gaussian copula-based approaches [@MB13, @BF17, @Xia17].
-The major contributions of this work are methods and software for flexible, scalable simulation of HD multivariate probability distributions with broad potential data analytic applications for modern, big-data statistical computing.
-For example, one could simulate high-resolution time series data, such as those consistent with an auto-regressive moving average model exhibiting a specified Spearman structure.
-Or our methods could be used to simulate sparsely correlated data, as many HD methods assume, via specifying a *spiked correlation matrix*.
-
-Researchers working in multivariate computation frequently encounter such difficulties and need to find a close or nearest correlation matrix.
-A widely-available routine for this task in `R` is `matrix::nearPD`, though it is not suitable for high dimensions.
-To overcome this issue, we introduce `cor_nearPD` --- an implementation of @QS2006's method that computes the nearest correlation matrix to a given symmetric matrix in the Frobenius norm.
-This routine as many applications beyond our primary goal of random vector generation.
-
-It is customary to compare new tools and algorithms directly to existing competing methods and software.
-In this study, however, we only employ our proposed methodology, as our previous work has shown that existing `R` tools are simply not designed to meet our high-dimensional goal (see @Li2019gpu for evaluations of the `R` `copula` package and others).
-For the bivariate simulations, existing packages such as [`nortaRA`](https://github.com/cran/NORTARA/blob/master/inst/doc/NORTARA.R) work well to match Pearson correlations exactly.
-
-<!-- 
-We advocate the use of Kendall's $\tau$ as it better captures correlation among components of non-normal, as well as non-linear patterns of association.
-Moreover, the matching Kendall's $\tau$ is nearly trivial due to the invariant of monotone transformations, after an adjustment to the input correlation matrix.
--->
-
-<!-- 
-We also show utility in our methodology through an application to differential gene expression analysis from RNA-sequencing data.
-The application results show that correlations indeed matter in the large-scale hypothesis testing, as many others have noted (for example, see @BE07, @Wu2012b).
-We also hope that the application provides an example workflow --- and other strategy to use in simulation design.
-Even the best-performing simulations we provide show a gap from the empirical distribution.
-To keep the demonstration straightforward, we did not attempt to match the empirical distribution of test statistics as precisely as possible.
-Yet our methodology could be more creatively applied to meet that goal.
-One could consider marginal distributions from different families, such as Poisson for a subset of genes.
-One could imagine finding a best fitting probability distribution among a class of distributions for each gene and this could perhaps a better fit.
-One could imagine additional structures/features in the data that an analyst could model to improve the correspondence with the empirical values, for example row correlations and high-dimensional covariance estimators (see @Won2013g).
--->
-
-There are limitations to the methodology and implementation.
-We could only investigate selected multivariate distributions in our Monte Carlo studies.
-There may be instances that the methods do not perform well.
-Along those lines, we expect that correlation values close to the boundary of the feasible region could result in algorithm failure.
-Another issue is that for discrete distributions, we use continuous approximations when the support set is large.
-This could limit the scalability/accuracy for particular joint distributions.
-
-<!-- 
-The most obvious missing feature of the proposed methodology is the inability to match a Pearson correlation matrix exactly.
-As discussed in the [Algorithms](algorithms) section and extensively by @XZ19, this is a computationally intense procedure and Pearson's correlation is not a natural choice to describe dependency for non-normal marginals.
-While we do not provide an implementation directly supporting Pearson matching, users may supply their own input Pearson correlation after employing a supplementary matching scheme [@Cario1997; @XZ19].
--->
-
-<!-- 
-Further, while we provide the ability of the user to specify discrete marginals, exact matching a desired dependency measure is not yet obtained.
-One potential solution for this is *rescale* to account for ties [ref].
-Finally we note, that the algorithm requires invertible marginal cdfs.
-This gives some restriction to the available marginal probability distributions (DOES IT?
-FOR EXAMPLE?).
-From a practical computing standpoint, the user's computing resource provides the ultimate limit on how large a dimension can be simulated using the `bigsimr` R package.
-A user must consider carefully the available memory and cores when conducting a Monte Carlo experiment.
--->
-
-<!-- 
-Future work includes developing scaleable algorithms to match the Pearson correlation matrix more precisely, discrete-margin specific modifications including fast Spearman's correlation rescaling (see Equation \@ref(eq:spearmanRescaled)), and high-dimensional covariance estimation.
-From an implementation standpoint, `bigsimr` only supports Nvidia GPUs and redesigning the code using OpenCL would broaden the users who would benefit.
-As data-analytic problems grow to even larger dimension, multi-GPU support is a promising hardware-based future direction.
--->
-```{r include=FALSE, cache=FALSE}
-# Remove all objects
-rm(list = ls(all.names = TRUE))
-
-# Unload all packages
-```
-
-<!--chapter:end:060-conclusion-discussion.Rmd-->
-
-```{r include=FALSE, cache=FALSE}
-options(digits = 4)
-
-knitr::opts_chunk$set(
-  # Text Results
-  eval = TRUE,
-  echo = FALSE,
-  results = 'markup',
-  collapse = TRUE,
-  message = FALSE,
-  warning = FALSE,
-  error = FALSE,
-  # Code Decoration
-  comment = NA,
-  background = '#F7F7F7',
-  # Cache
-  cache = TRUE,
-  # Plots
-  fig.show = "hold",
-  fig.width = 6.5,
-  fig.asp = 0.618,  # 1 / phi
-  out.width = "85%",
-  fig.align = 'center'
-)
-
-# options(dplyr.print_min = 6, dplyr.print_max = 6)
-
-seed_number <- 2020-02-25
-set.seed(seed = seed_number)
-```
-# Supplementary Materials {.unnumbered #misc}
-
-We provide an open-source implementation of our methodology as the `bigsimr` R package, hosted on github, https://github.com/SchisslerGroup/r-bigsimr.
-
-```{r, out.width='5%'}
-knitr::include_graphics('images/hex-bigsimr.png')
-```
-
-# Acknowledgment(s) {.unnumbered #acknowledgments}
-
-The authors gratefully acknowledge Heather Knudson's graphic design for the `bigsimr` R Package.
-The results published here are in whole or part based upon data generated by the TCGA Research Network: https://www.cancer.gov/tcga.
-
-# Disclosure statement {.unnumbered #coi}
-
-The authors report no conflict of interest.
-
-# Funding {.unnumbered #funding}
-
-Research reported in this publication was supported by grants from the National Institutes of General Medical Sciences (5 U54 GM104944, GM103440) from the National Institutes of Health.
-
-<!-- 
-Nomenclature/Notation
--->
-```{r include=FALSE, cache=FALSE}
-# Remove all objects
-rm(list = ls(all.names = TRUE))
-
-# Unload all packages
-```
-
-<!--chapter:end:070-supplementary-materials.Rmd-->
-
-```{r include=FALSE, cache=FALSE}
-options(digits = 4)
-
-knitr::opts_chunk$set(
-  # Text Results
-  eval = TRUE,
-  echo = FALSE,
-  results = 'markup',
-  collapse = TRUE,
-  message = FALSE,
-  warning = FALSE,
-  error = FALSE,
-  # Code Decoration
-  comment = NA,
-  background = '#F7F7F7',
-  # Cache
-  cache = TRUE,
-  # Plots
-  fig.show = "hold",
-  fig.width = 6.5,
-  fig.asp = 0.618,  # 1 / phi
-  out.width = "85%",
-  fig.align = 'center'
-)
-
-# options(dplyr.print_min = 6, dplyr.print_max = 6)
-
-seed_number <- 2020-02-25
-set.seed(seed = seed_number)
-```
-# Appendix {.unnumbered #appendix}
-
-\noindent Consider a bivariate example where we have a correlated $(Y_1, Y_2)^\top$ with $Y_1\sim N(\mu_1, \sigma_1^2)$ (normal) and $Y_2\sim LN(\mu_2, \sigma_2^2)$ (lognormal).
-First, let us note that when we get such a vector from bigsimr package then in fact it can be represented as
-
-\begin{equation}
-\label{kram1}
-(Y_1, Y_2)^\top \stackrel{d}{=} \left(X_1, e^{X_2}\right)^\top,
-\end{equation}
-
-where $(X_1, X_2)^\top \sim N_2(\boldsymbol \mu, \boldsymbol \Sigma)$, which is bivariate normal with mean vector $\boldsymbol \mu = (\mu_1, \mu_2)^\top$ and variance-covariance matrix
-
-\begin{equation}
-\label{kram2}
-\boldsymbol \Sigma = 
-\left[
-\begin{array}{cc}
-\sigma_1^2 & \rho \sigma_1\sigma_2\\
-\rho \sigma_1\sigma_2 & \sigma_2^2
-\end{array}
-\right].
-\end{equation}
-
-To see this, consider the three steps of the NORTA construction:
-
-\begin{enumerate}
-
-\item Generate $(Z_1, Z_2)^\top \sim N_2(\boldsymbol 0, \boldsymbol R)$, where
-
-\begin{equation}
-\label{kram3}
-\boldsymbol R = 
-\left[
-\begin{array}{cc}
-1 & \rho \\
-\rho & 1
-\end{array}
-\right].
-\end{equation}
-
-\item Transform $(Z_1, Z_2)^\top$ to $(U_1, U_2)^\top$ viz. $U_i =\Phi(Z_i)$,  $i=1,2$, where $\Phi(\cdot)$ is the standard normal CDF.
-
-\item Return $(Y_1, Y_2)^\top$, where $Y_i=F_i^{-1}(U_i)$, $i=1,2$, and $F_i(\cdot)$ is the CDF of $Y_i$ and $F_i^{-1}(\cdot)$ is its inverse (the quantile function).
-
-\end{enumerate}
-
-In this example, $F_1$ is the CDF of $N(\mu_1, \sigma_1^2)$ while $F_2$ is the CDF of $LN(\mu_2, \sigma_2^2)$, so that the two required quantile functions are
-
-\begin{equation}
-\label{kram4}
-F_1^{-1}(u) = \mu_1+\sigma_1 \Phi^{-1}(u)\,\,\, \mbox{and} \,\,\, F_2^{-1}(u) = e^{\mu_2+\sigma_2 \Phi^{-1}(u)}, 
-\end{equation}
-
-as can be seen by standard calculation. Thus, when we apply Step 3 of the NORTA algorithm, we get
-
-\begin{equation}
-\label{kram5}
-F_1^{-1}(U_1) = \mu_1+\sigma_1 \Phi^{-1}(\Phi(Z_1)) =  \mu_1+\sigma_1 Z_1 \,\,\, \mbox{and} \,\,\, F_2^{-1}(U_2) = e^{\mu_2+\sigma_2 \Phi^{-1}(\Phi(Z_2))} = e^{\mu_2+\sigma_2 Z_2}, 
-\end{equation}
-
-where $(X_1, X_2)^\top =  (\mu_1+\sigma_1 Z_1, \mu_2+\sigma_2 Z_2)^\top \sim N_2(\boldsymbol \mu, \boldsymbol \Sigma)$.
-Consequently, the vector $(Y_1, Y_2)^\top$ obtained in Step 3 of the algorithm has the structure provided by (\ref{kram1}).
-
-\vspace{0.1in}
-
-\noindent Next, we provide the exact covariance structure of the random vector $(Y_1, Y_2)^\top$ given by (\ref{kram1}), and relate the correlation of $Y_1$ and $Y_2$ to $\rho$, which is the correlation of the normal variables $Z_1$, $Z_2$ (and also $X_1$ and $X_2$).
-A straightforward albeit somewhat tedious algebra produces the following result.
-
-\begin{lemma}
-Let ${\bf Y} = (Y_1, Y_2)^\top$ admit the stochastic representation (\ref{kram1}), where ${\bf X} = (X_1, X_2)^\top \sim N_2(\boldsymbol \mu, \boldsymbol \Sigma)$ with $\boldsymbol \mu$ and $\boldsymbol \Sigma$ as above.
-Then the mean vector and the variance-covariance matrix of ${\bf Y}$ are given by
-
-\begin{equation}
-\label{kram6}
-\boldsymbol \mu_{{\bf Y}} = \left(\mu_1, e^{\mu_2+\sigma_2^2/2}\right)^\top
-\end{equation}
-
-and
-
-\begin{equation}
-\label{kram7}
-\boldsymbol \Sigma = 
-\left[
-\begin{array}{cc}
-\sigma_1^2 & \rho \sigma_1\sigma_2  e^{\mu_2+\sigma_2^2/2}  \\
-\rho \sigma_1\sigma_2 e^{\mu_2+\sigma_2^2/2} & \left[ e^{\mu_2+\sigma_2^2/2}\right]^2 \left[e^{\sigma_2^2} -1 \right],
-\end{array}
-\right].
-\end{equation}
-
-respectively. 
-\end{lemma}
-
-\begin{proof}
-The values of the means and the variances are obtained immediately from normal marginal distribution of $Y_1$ and lognormal marginal distribution of $Y_2$.
-It remains to establish the covariance of $Y_1$ and $Y_2$,
-
-\begin{equation}
-\label{covy1y2}
-\mbox{Cov}(Y_1, Y_2) = \mathbb E(Y_1 Y_2) - \mathbb E(Y_1)\mathbb E(Y_2), 
-\end{equation}
-
-and in particular the first expectation on the right-hand-side above. By using the tower property of expectations, the latter expectation can be expressed as
-
-\begin{equation}
-\label{kre1}
-\mathbb E(Y_1 Y_2)  =  \mathbb E\left(X_1 e^{X_2}\right)  =  \mathbb E \left\{ \mathbb E \left(X_1 e^{X_2} | X_1 \right) \right\} = \mathbb E \left\{ X_1 \mathbb E \left(e^{X_2} | X_1 \right) \right\}.
-\end{equation}
-
-Further, by using the fact that the conditional distribution of $X_2$ given $X_1=x_1$ is normal with mean $\mathbb E(X_2|X_1=x_1) = \mu_2+\rho\sigma_2(x_1-\mu_1)/\sigma_1$ and variance $\sigma_2^2(1-\rho^2)$, one can relate the inner expectation on the far right in (\ref{kre1}) to the moment generating function $M(t)$ of this conditional distribution evaluated at $t=1$, leading to
-
-\begin{equation}
-\label{kre2}
-E \left(e^{X_2} | X_1 \right) = e^{ \mu_2+\rho\sigma_2(x_1-\mu_1)/\sigma_1 + \sigma_2^2(1-\rho^2)/2},
-\end{equation}
-
-so that
-
-\begin{equation}
-\label{kre3}
-\mathbb E(Y_1 Y_2)  =  e^{ \mu_2+ \frac{1}{2} \sigma_2^2(1-\rho^2)} \mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\}.
-\end{equation}
-
-Since $X_1$ is normal with mean $\mu_1$ and variance $\sigma_1^2$, the expectation in (\ref{kre3}) becomes
-
-\begin{equation}
-\label{kre4}
-\mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = \int_{-\infty}^\infty x_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (x_1-\mu_1)} \frac{1}{\sqrt{2\pi}\sigma_1} e^{-\frac{1}{2\sigma_1^2}(x_1-\mu_1)^2}.
-\end{equation}
-
-Upon substituting $u=x_1-\mu_1$, followed by some algebra, we arrive at
-
-\begin{equation}
-\label{kre5}
-\mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = e^{\frac{1}{2}\sigma_2^2\rho^2} \int_{-\infty}^\infty (u+\mu_1)  \frac{1}{\sqrt{2\pi}\sigma_1} e^{-\frac{1}{2\sigma_1^2}(u-\rho\sigma_1\sigma_2)^2}.
-\end{equation}
-
-Since the integral in (\ref{kre5}) is the expectation $\mathbb E(X+\mu_1)$, where $X$ is normal with mean $\rho\sigma_1\sigma_2$ and variance $\sigma_1^2$, we conclude that
-
-\begin{equation}
-\label{kre6}
-\mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = e^{\frac{1}{2}\sigma_2^2\rho^2} (\rho\sigma_1\sigma_2 +\mu_1),
-\end{equation}
-
-which, in view of (\ref{kre3}), leads to
-
-\begin{equation}
-\label{kre7}
-\mathbb E (Y_1Y_2) = e^{\mu_2 + \frac{1}{2}\sigma_2^2} (\rho\sigma_1\sigma_2 +\mu_1).
-\end{equation}
-
-Finally, (\ref{covy1y2}), along with the expressions for the means of $Y_1$ and $Y_2$, produce the covariance of $Y_1$ and $Y_2$:
-
-\begin{equation}
-\label{kre8}
-\mbox{Cov}(Y_1, Y_2) = e^{\mu_2 + \frac{1}{2}\sigma_2^2} (\rho\sigma_1\sigma_2 +\mu_1) - \mu_1 e^{\mu_2 + \frac{1}{2}\sigma_2^2} = \rho\sigma_1\sigma_2 e^{\mu_2 + \frac{1}{2}\sigma_2^2}.
-\end{equation}
-
-This concludes the proof
-\end{proof}
-
-Using the above result, we can directly relate the correlation coefficient of $Y_1$ and $Y_2$ with that of $X_1$ and $X_2$,
-
-\begin{equation}
-\label{kram8}
-\rho_{Y_1, Y_2} = \rho \frac{\sigma_2}{\sqrt{e^{\sigma_2^2} -1}},
-\end{equation}
-
-where $\rho$ is the correlation of $X_1$ and $X_2$.
-The above result is useful when studying the possible range of correlation of $Y_1$ and $Y_2$ in the above example.
-It can be shown that the factor on the far right in (\ref{kram8}) is a monotonically decreasing function of $\sigma_2$ on $(0,\infty)$, with the limits of 1 and 0 at zero and infinity, respectively.
-Thus, in principle, the range of correlation of $Y_1$ and $Y_2$ is the same as that of $X_1$ and $X_2$, as the factor on the far right in (\ref{kram8}) can be made arbitrarily close to 1.
-However, by changing this factor we may affect the marginal distributions of $Y_1$ and $Y_2$.
-It can be shown that if the marginal distributions of $Y_1$ and $Y_2$ are fixed, then the relation (\ref{kram8}) becomes
-
-\begin{equation}
-\label{kram9}
-\rho_{Y_1, Y_2} = \rho \sqrt{\frac{\log(1+c_2^2)}{c_2^2}},
-\end{equation}
-
-where $c_2=\sigma_{Y_2}/\mu_{Y_2}$ is the *coefficient of variation* (CV) of the variable $Y_2$.
-Thus, the range of possible correlations in this model is not affected by the distribution of $Y_1$, and is determined by the CV of $Y_2$ as follows:
-
-\begin{equation}
-\label{kram10}
-- \sqrt{\frac{\log(1+c_2^2)}{c_2^2}} \leq \rho_{Y_1, Y_2} \leq \sqrt{\frac{\log(1+c_2^2)}{c_2^2}}. 
-\end{equation}
-
-Plugging in the estimates of these quantities from Section \@ref(package), we see that
-
-\begin{equation}
-\label{kram11}
-\frac{\hat{\sigma}_2}{\sqrt{e^{\hat{\sigma}_2^2} -1}} = 0.881.
-\end{equation}
-
-Thus, the possible range of correlation becomes $(-0.881, 0.881)$. This approximately agrees with the MC results provided.
-```{r include=FALSE, cache=FALSE}
-# Remove all objects
-rm(list = ls(all.names = TRUE))
-
-# Unload all packages
-```
-
-<!--chapter:end:080-appendix.Rmd-->
-
-```{r include=FALSE, cache=FALSE}
-options(digits = 4)
-
-knitr::opts_chunk$set(
-  # Text Results
-  eval = TRUE,
-  echo = FALSE,
-  results = 'markup',
-  collapse = TRUE,
-  message = FALSE,
-  warning = FALSE,
-  error = FALSE,
-  # Code Decoration
-  comment = NA,
-  background = '#F7F7F7',
-  # Cache
-  cache = TRUE,
-  # Plots
-  fig.show = "hold",
-  fig.width = 6.5,
-  fig.asp = 0.618,  # 1 / phi
-  out.width = "85%",
-  fig.align = 'center'
-)
-
-# options(dplyr.print_min = 6, dplyr.print_max = 6)
-
-seed_number <- 2020-02-25
-set.seed(seed = seed_number)
-```
-`r if (knitr:::is_html_output()) '# References {-}'`
-```{r include=FALSE, cache=FALSE}
-# Remove all objects
-rm(list = ls(all.names = TRUE))
-
-# Unload all packages
-```
-
-<!--chapter:end:300-references.Rmd-->
-
----
-institution:
-  name: "University of Nevada, Reno"
-  department: "Mathematics and Statistics"
-  address:
-    line1: "1664 N. Virginia Street / 0084"
-    line2: "Reno, Nevada 89557"
-  phone: "(775) 784-6773"
-  fax: "(775) 784-6378"
-  url: https://www.unr.edu/math/
-  logo: "logos/blue.png"
-author:
-  first: "Alfred"
-  last: "Schissler"
-  short: "Grant"
-  title: "Ph.D."
-  position: "Assistant Professor, Statistics"
-  signature: "signature.jpg"
-to:
-  first: "Alexander"
-  last: "Knudson"
-  short: "Alex"
-date: "`r format(Sys.Date(), format='%d %B %Y')`"
-opening: "Dear CSDA Annals of Statistical Data Science Editor(s):"
-closing: "Sincerely,"
-output:
-  pdf_document:
-    template: templateSingleSignature.tex
----
-```{r include=FALSE, cache=FALSE}
-options(digits = 4)
-
-knitr::opts_chunk$set(
-  # Text Results
-  eval = TRUE,
-  echo = FALSE,
-  results = 'markup',
-  collapse = TRUE,
-  message = FALSE,
-  warning = FALSE,
-  error = FALSE,
-  # Code Decoration
-  comment = NA,
-  background = '#F7F7F7',
-  # Cache
-  cache = TRUE,
-  # Plots
-  fig.show = "hold",
-  fig.width = 6.5,
-  fig.asp = 0.618,  # 1 / phi
-  out.width = "85%",
-  fig.align = 'center'
-)
-
-# options(dplyr.print_min = 6, dplyr.print_max = 6)
-
-seed_number <- 2020-02-25
-set.seed(seed = seed_number)
-```
-
-As an invited speaker at CMStatistics 2020, I was delighted to learn of the opportunity to have our work included in the first issue of CSDA Annals of Statistical Data Science.
-
-On behalf of my co-authors, I'm pleased to submit our manuscript entitled *Simulating High-Dimensional Multivariate Data using the `Bigsimr` Package* for your consideration.
-We feel that the work is well suited to the indicated themes and hope that it will be considered a valuable contribution to this prestigious journal.
-
-I hope that everything is in order. If not, or if you have any additional needs, please do not hesitate to contact me. Thank you and all those involved in this process.
-
-Best regards,
-```{r include=FALSE, cache=FALSE}
-# Remove all objects
-rm(list = ls(all.names = TRUE))
-
-# Unload all packages
-```
-
-<!--chapter:end:letter_bigsimr_CSDA.Rmd-->

### 2021-03-27 010-background.Rmd
idk

@@
 )
+
 load("data/example_genes.rda")
@@
 small_brca %>%
-  kable(booktabs = TRUE, caption = cap)
+  kable(format = "latex", booktabs = TRUE, caption = cap)
 ```
@@
 \begin{equation}
+F(x_1, \ldots, x_d) = {\mathbb P}(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), 
 (\#eq:copula)
-F(x_1, \ldots,x_d) = \mathbb P(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), 
 \end{equation}
 
-$x_i \in \mathbb R, i=1,\ldots,d.$
+$x_i \in {\mathbb R}, i=1,\ldots,d.$
 

### 2021-03-27 040-monte-carlo-evaluation.Rmd
idk

@@
 )
+
 load("data/bivariate_normal_sims.rda")
@@
 
-| Simulation Reps ($B$) | Correlation Types | Identical-margin 2D distribution |
-|-------------|:--------------:|----------------------:|
-| $1000$    | Pearson ($\rho_P$) | ${ \bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho_i ), i=1,\ldots,100$ |
-| $10,000$  | Spearman ($\rho_S$) | ${ \bf Y} \sim MVG( shape = 10, rate = 1, \rho_i ), i=1,\ldots,100$ |
-| $100,000$ | Kendall ($\tau$)| ${ \bf Y} \sim MVNB(p = 3\times10^{-4}, r = 4,\rho_i), i=1,\ldots,100$ |
+\begin{table}[]
+\begin{tabular}{@{}lcr@{}}
+\toprule
+Simulation Reps $B$ & Correlation Types & Identical-margin 2D distribution \\ \midrule
+$1000$ & Pearson ($\rho_P$) & ${\bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho_i ), i=1,\ldots,100$ \\
+$10,000$ & Spearman ($\rho_S$) & ${\bf Y} \sim MVG( shape = 10, rate = 1, \rho_i ), i=1,\ldots,100$ \\
+$100,000$ & Kendall ($\tau$) & ${\bf Y} \sim MVNB(p = 3\times10^{-4}, r = 4,\rho_i), i=1,\ldots,100$ \\ \bottomrule
+\end{tabular}
+\end{table}
 
@@
 
-kable(tabMAE, booktabs = TRUE,
+kable(tabMAE, booktabs = TRUE, format = "latex",
       linesep = c("", "", "\\addlinespace"),

### 2021-03-27 _article/article_bigsimr.tex
idk

\documentclass[
]{jss}

\usepackage[utf8]{inputenc}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\author{
\\
}
\title{Simulating High-Dimensional Multivariate Data using the \pkg{Bigsimr} Package}

\Plainauthor{}
\Plaintitle{Simulating High-Dimensional Multivariate Data using the Bigsimr Package}
\Shorttitle{\pkg{Bigsimr}: Simulate High-Dimensional Data}

\Abstract{
It is critical to realistically simulate data when employing Monte Carlo techniques and evaluating statistical methodology. Measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations. This article introduces the \textbackslash pkg\{Bigsimr\} R package. This high-level package provides a flexible and scalable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions and known Pearson, Spearman, or Kendall correlation matrix. \textbackslash pkg\{Bigsimr\} contains additional high-performance features, including multi-core algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix. Monte Carlo studies quantify the accuracy and scalability of our approach, up to \(d=10,000\). Finally, we demonstrate example applications enabled via \textbackslash pkg\{Bigsimr\} by applying these methods to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples.
}

\Keywords{multivariate simulation, high-dimensional data, nonparametric correlation, Gaussian copula, RNA-sequencing data, breast cancer}
\Plainkeywords{multivariate simulation, high-dimensional data, nonparametric correlation, Gaussian copula, RNA-sequencing data, breast cancer}

%% publication information
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{}
%% \Acceptdate{2012-06-04}

\Address{
  }

% Pandoc citation processing

% Pandoc header
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amsthm}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{ragged2e}

\usepackage{setspace}
\setstretch{2.0}

\begin{document}

\newpage

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Massive high-dimensional (HD) data sets are now commonplace in many areas of scientific inquiry. As new methods are developed for data, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of analyzing such proposed methodology --- such as false positive rates, statistical power, interval coverage, and robustness --- often with comparison to existing methods. Further, efficient simulation empowers statistical computing strategies, such as the parametric bootstrap \citep{Chernick2008} to simulate from a hypothesized null model, providing inference in analytically challenging settings. Such Monte Carlo (MC) techniques become difficult for high-dimensional data using existing algorithms and tools. This is particularly true when simulating massive multivariate, non-normal distributions, arising naturally in many fields of study.

As many have noted, it can be vexing to simulate dependent, non-normal/discrete data, even for low dimensional settings \citep{MB13, XZ19}. For continuous non-normal multivariate data, the well-known NORmal To Anything (NORTA) algorithm \citep{Cario1997} and other copula approaches \citep{Nelsen2007} are well-studied, with flexible, robust software available \citep{Yan2007, Chen2001}. Yet these approaches do not scale in a timely fashion to high-dimensional problems \citep{Li2019gpu}. For discrete data, early simulation strategies had major flaws, such as failing to obtain the full range of possible correlations (e.g., admitting only positive correlations: see \citet{Park1996}). While more recent approaches \citep{MB13, Xia17, BF17} have largely remedied this issue for low-dimensional problems, the existing tools are not designed to scale to high dimensions.

Another central issue lies in characterizing dependence between components in the high-dimensional random vector. The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc). For normal data, the Pearson product-moment correlation describes the dependency perfectly. As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix is computationally intense \citep{Chen2001, Xia17}. On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependence, such as Spearman's \(\rho\) and Kendall's \(\tau\). Throughout, we focus on matching these nonparametric dependence measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology. We do, however, provide Pearson matching with some caveats.

With all this in mind, we present a scalable, flexible multivariate simulation algorithm. The crux of the method lies in the construction of a Gaussian copula in the spirit of the NORTA procedure. Further, we introduce the \texttt{bigsimr} R package that provides high-performance software implementing our algorithm. The algorithm design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed-form relationships between dependence measures for the multivariate normal (MVN) distribution.

Our study proceeds by providing background information, including a description of a motivating example application: RNA-sequencing (RNA-seq) breast cancer data. Then we describe and justify our simulation methodology and related algorithms. Next, we detail an illustrative low-dimensional example of basic use of the \texttt{bigsimr} R package. Then we proceed with Monte Carlo studies under various bivariate distributional assumptions to assess accuracy. We conclude the Monte Carlo evaluations by summarizing the computation time for increasingly higher dimensional vectors. After the MC evaluations, we simulate random vectors motivated by our RNA-seq example, evaluate the accuracy, and provide example statistical computing tasks, namely MC estimation of joint probabilities and evaluating HD correlation estimation efficiency. Finally, we discuss the method's utility, limitations, and future directions.

\hypertarget{background}{%
\section{Background}\label{background}}

The \texttt{bigsimr} \texttt{R} package presented here provides multiple algorithms that operate with high-dimensional multivariate data; however, all these algorithms were originally designed to support a single task: to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics. Specifically, our goal is to efficiently simulate a large number, \(B\), of HD random vectors \({\bf Y}=(Y_1, \ldots, Y_d)^\top\) with \emph{correlated} components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) \(F_i\).

When designing this methodology, we developed the following properties to guide our effort. We divide the properties into two categories: (1) basic properties (BP) and ``scalability'' properties (SP). The BPs are adapted from an existing criteria due to \citet{Nik13a}. A suitable simulation strategy should possess the following properties:

\setstretch{1.5}

\begin{itemize}
\tightlist
\item
  BP1: A wide range of dependences, allowing both positive and negative values, and, ideally, admitting the full range of possible values.
\item
  BP2: Flexible dependence, meaning that the number of bivariate marginals can be equal to the number of dependence parameters.
\item
  BP3: Flexible marginal modeling, generating heterogeneous data --- possibly from differing probability families.
\end{itemize}

\setstretch{2.0}

Moreover, the simulation method must \emph{scale} to high dimensions:

\setstretch{1.5}

\begin{itemize}
\tightlist
\item
  SP1: Procedure must scale to high dimensions with practical compute times.
\item
  SP2: Procedure must scale to high dimensions while maintaining accuracy.
\end{itemize}

\setstretch{2.0}

\hypertarget{motivating-example-rna-seq-data}{%
\subsection{Motivating example: RNA-seq data}\label{motivating-example-rna-seq-data}}

Simulating high-dimensional, non-normal, correlated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data \citep{Wang2009b, Conesa2016b} derived from breast cancer patients. The RNA-seq data-generating process involves counting how often a particular form of human messenger RNA (mRNA) is expressed in a biological sample. RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data. For human-derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted \citep{Schissler2019}. Importantly, due to inherent biological processes, gene expression data exhibit correlation (co-expression) across genes \citep{BE07, Schissler2018}.

We illustrate our methodology using a well-studied Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments).
For ease of modeling and simplicity of exposition, we only consider high expressing genes. In turn, we begin by filtering to retain the top 1000 of the highest-expressing genes (in terms of median expression) of the over 20,000 gene measurements from \(N=878\) patients' tumor samples. This gives a great number of pairwise dependencies among the marginals (specifically, \(\ensuremath{4.995\times 10^{5}}\) correlation parameters). Table \ref{tab:ch010-realDataTab} displays RNA-seq counts for three selected high-expressing genes for the first five patients' breast tumor samples. To help visualize the bivariate relationships for these three selected genes across all patients, Figure \ref{fig:ch010-realDataFig} displays the marginal distributions and estimated Spearman's correlations.

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(2020-02-25)
R> num_genes    <- 3
R> num_patients <- 5
R> gene_sample  <- sample(example_genes[1:1000], num_genes)
R> 
R> cap <- paste0("mRNA expression for three selected high-expressing genes, ",
+               paste(gene_sample, collapse = ", "), 
+               ", for the first five patients in the TCGA BRCA data set.")
R> 
R> small_brca <- example_brca %>%
+   select(all_of(gene_sample)) %>%
+   head(n = num_patients) %>%
+   as_tibble(rownames = "Patient ID") %>%
+   mutate(`Patient ID` = str_sub(`Patient ID`, end = 12))
R> 
R> small_brca %>%
+   kable(format = "latex", booktabs = TRUE, caption = cap)
\end{CodeInput}
\begin{table}

\caption{\label{tab:ch010-realDataTab}mRNA expression for three selected high-expressing genes, STAU1, FKBP1A, NME2, for the first five patients in the TCGA BRCA data set.}
\centering
\begin{tabular}[t]{lrrr}
\toprule
Patient ID & STAU1 & FKBP1A & NME2\\
\midrule
TCGA-A1-A0SB & 10440 & 11354 & 17655\\
TCGA-A1-A0SD & 21523 & 20221 & 14653\\
TCGA-A1-A0SE & 21733 & 22937 & 35251\\
TCGA-A1-A0SF & 11866 & 19650 & 16551\\
TCGA-A1-A0SG & 12486 & 12089 & 10434\\
\bottomrule
\end{tabular}
\end{table}

\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> ggpairs(
+   data = example_brca[, gene_sample],
+   upper = list(continuous = wrap('cor', method = "spearman"))
+ ) + 
+   theme_bw() +
+   theme(axis.text.x = element_text(angle = -90, vjust = 0.5))
\end{CodeInput}
\begin{CodeOutput}
Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
exact p-value with ties
\end{CodeOutput}
\begin{CodeOutput}
Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
exact p-value with ties
\end{CodeOutput}
\begin{CodeOutput}
Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
exact p-value with ties
\end{CodeOutput}
\begin{figure}

{\centering \includegraphics{article_bigsimr_files/figure-latex/ch010-realDataFig-1}

}

\caption[Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes from Table 1]{Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes from Table 1. The data possess outliers, heavy-right tails, are discrete, and have non-trivial intergene correlations. Modeling these data motivate our simulation methodology.}\label{fig:ch010-realDataFig}
\end{figure}
\end{CodeChunk}

\hypertarget{measures-of-dependency}{%
\subsection{Measures of dependency}\label{measures-of-dependency}}

In multivariate analysis, an analyst must select a metric to quantify dependency.
The most widely-known is the Pearson (product-moment) correlation coefficient that describes the linear association between two random variables \(X\) and \(Y\), and, it is given by

\begin{equation}
\rho_P(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ \mathrm{Var}(X)\mathrm{Var}(Y)\right]^{1/2}}.
\label{eq:pearson}
\end{equation}

As \citet{MB13} and \citet{MK01} discuss, for a bivariate normal \((X,Y)\) random vector, the Pearson correlation completely describes the dependency between the components. For non-normal marginals with monotone correlation patterns, however, \(\rho_P\) suffers some drawbacks and may mislead or fail to capture important relationships \citep{MK01}. Alternatively in these settings, analysts often prefer rank-based correlation measures to describe the degree of monotonic association.

Two nonparametric, rank-based measures common in practice are Spearman's correlation (denoted \(\rho_S\)) and Kendall's \(\tau\). Spearman's \(\rho_S\) has an appealing correspondence as the Pearson correlation coefficient on \emph{ranks} of the values, thereby captures nonlinear yet monotone relationships. Kendall's \(\tau\), on the other hand, is the difference in probabilities of concordant and discordant pairs of observations, \((X_i, Y_i)\) and \((X_j, Y_j)\), with concordance meaning that orderings have the same direction (e.g., if \(X_i < X_j\), then \(Y_i < Y_j\)). Note that concordance is determined by the ranks of the values, not the values themselves.

Both \(\tau\) and \(\rho_S\) are \emph{invariant under monotone transformations} of the underlying random variates. As we will describe more fully in the \protect\hyperlink{algorithms}{Algorithms} section, this property enables matching rank-based correlations with speed (SP1) and accuracy (SP2).

\emph{Correspondence among Pearson, Spearman, and Kendall correlations}

There is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficient, as the marginal distributions \(F_i\) are intrinsic in their calculation. For \emph{bivariate normal vectors}, however, the correspondence is well-known:

\begin{equation}
\label{eq:convertKendall}
\rho_{P} = \sin \left( \tau \times \frac{\pi}{2} \right), 
\end{equation}

\noindent and similarly for Spearman's \(\rho\) \citep{K58},

\begin{equation}
\label{eq:convertSpearman}
\rho_P = 2 \times \sin \left( \rho_S \times \frac{\pi}{6} \right).
\end{equation}

\emph{Marginal-dependent bivariate correlation bounds}

Given two marginal distributions, \(\rho_P\) is not free to vary over the entire range of possible correlations \([-1,1]\). The so-called \emph{Frechet-Hoeffding bounds} are well-studied \citep{Nelsen2007, BF17}. These constraints cannot be overcome through algorithm design. In general, the bounds are given by

\begin{equation}
\label{eq:frechet}
\rho_P^{max} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (U) \right), \quad \rho_P^{min} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (1 - U) \right)
\end{equation}

\noindent where \(U\) is a uniform random variable on \((0,1)\) and \(F^{-1}_1, F^{-1}_2\) are the inverse CDFs of \(X_1\) and \(X_2\), respectively, definitely by \eqref{eq:inverseCDF} when the variables are discrete.

\begin{equation}
F_{i}^{-1} = \inf\{y:F_{i}(y) \geq u \}.
\label{eq:inverseCDF}
\end{equation}

\hypertarget{gaussian-copulas}{%
\subsection{Gaussian copulas}\label{gaussian-copulas}}

There is a strong connection of our simulation strategy to Gaussian \emph{copulas} (see \citet{Nelsen2007} for a technical introduction). A copula is a distribution function on \([0,1]^d\) that describes a multivariate probability distribution with standard uniform marginals. This provides a powerful, natural way to characterize joint probability structures. Consequently, the study of copulas is an important and active area of statistical theory and practice.

For any random vector \({\bf X}=(X_1, \ldots, X_d)^\top\) with CDF \(F\) and marginal CDFs \(F_i\) there is a copula function \(C(u_1, \ldots, u_d)\) satisfying

\begin{equation}
F(x_1, \ldots, x_d) = {\mathbb P}(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), 
\label{eq:copula}
\end{equation}

\(x_i \in {\mathbb R}, i=1,\ldots,d.\)

A Gaussian copula has marginal CDFs that are all standard normal, \(F_i = \Phi, \forall \, i\). This representation corresponds to a multivariate normal (MVN) distribution with standard normal marginal distributions and covariance matrix \({\bf R_P}\). As the marginals are standardized to have unit variance, however, \({\bf R_P}\) is a Pearson correlation matrix. If \(F_{{\bf R}}\) is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula \(C_{{\bf R}}\) is defined through

\begin{equation}
\label{eq:gauss}
F_{{\bf R}}(x_1, \ldots, x_d) = C_{{\bf R}}(\Phi(x_1), \ldots, \Phi(x_d)),
\end{equation}

where \(\Phi(\cdot)\) is the standard normal CDF. Note that the copula \(C_{{\bf R}}\) is the familiar multivariate normal CDF of the random vector \((\Phi(X_1), \ldots, \Phi(X_d))\), where \((X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})\).

Sklar's Theorem \citep{Sklar1959, Ubeda-Flores2017} guarantees that given inverse CDFs \(F_i^{-1}\)s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions. For example, using Gaussian copulas, we can construct a random vector \({\bf Y} = (Y_1, \ldots, Y_d)^\top\) with \(Y_i \sim F_i\), viz.~\(Y_i = F_i^{-1}(\Phi(X_i)), i=1, \ldots, d\), where \((X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})\).

\hypertarget{algorithms}{%
\section{Algorithms}\label{algorithms}}

This section describes our methods involved in simulating a random vector \(\bf Y\) with \(Y_i\) components for \(i=1,\ldots,d\). Each \(Y_i\) has a specified marginal CDF \(F_i\) and its inverse \(F^{-1}_i\). To characterize dependency, every pair \((Y_i, Y_j)\) has a given Pearson correlation \(\rho_P\), Spearman correlation \(\rho_S\), and/or Kendall's \(\tau\). The method can be described as a \emph{high-performance Gaussian copula} (Equation \eqref{eq:gauss}) providing a high-dimensional NORTA-inspired algorithm.

\hypertarget{normal-to-anything-norta}{%
\subsection{NORmal To Anything (NORTA)}\label{normal-to-anything-norta}}

The well-known NORTA algorithm \citep{Cario1997} simulates a random vector \(\bf Y\) with variance-covariance matrix \(\Sigma_{\bf Y}\). Specifically, NORTA algorithm proceeds as:

\setstretch{1.5}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Simulate a random vector \(\bf Z\) with \(d\) \emph{independent and identically distributed} (iid) standard normal components.
\item
  Determine the input matrix \(\Sigma_{\bf Z}\) that corresponds with the specified output \(\Sigma_{\bf Y}\).
\item
  Produce a Cholesky factor \(M\) of \(\Sigma_{\bf Z}\) such that \(M M^{\prime}=\Sigma_{\bf Z}\).
\item
  Set \(X\) by \(X \gets MZ\).
\item
  \(\text{Return} \; Y \; \text{where} \; Y_i \gets F_i^{-1}[\Phi(X_i)], \; i=1,...,d\).
\end{enumerate}

\setstretch{2.0}

With modern parallel computing, steps 1, 3, 4, 5 are readily implemented as high-performance, multi-core and/or graphical-processing-unit (GPU) accelerated algorithms --- providing fast scalability using readily-available hardware.

Matching specified Pearson correlation coefficients exactly (step 2 above), however, is computationally costly. In general, there is no closed-form correspondence between the components of the input \(\Sigma_{\bf Z}\) and target \(\Sigma_{\bf Y}\). Matching the correlations involves evaluating or approximating \(\binom{d}{2}\) integrals of the form

\begin{equation}
    \mathrm{E}\left[Y_i Y_j\right] = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} F_i^{-1}\left[\Phi(z_i)\right] F_j^{-1}\left[\Phi(z_j)\right] \phi(z_i, z_j, \rho_z) dz_i dz_j,
    \label{eq:pearsonIntegralRelation}
\end{equation}

where \(\phi(\cdot)\) is the joint probability distribution function of two correlated standard normal variables. For HD data, (nearly) exact evaluation may become too costly to enable practical simulation studies. For low-dimensional problems, however, methods and tools exist to match Pearson correlations precisely; see \citep{Xia17} and the publicly available \texttt{nortaRA} R package \citep{Chen2001}. As described later, to enable HD Pearson matching we approximate these integrals.

\emph{NORTA in higher dimensions}

Sklar's theorem provides a useful characterization of multivariate distributions through copulas. Yet the choice of copula-based simulation algorithm affects which joint distributions may be simulated. Even in low-dimensional spaces (e.g., \(d=3\)), there exist valid multivariate distributions with \emph{feasible} Pearson correlation matrices that NORTA cannot match exactly \citep{LH75}. This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined the resultant matrix is indefinite. These situations do occur, even using exact analytic calculations. Such problematic target correlation matrices are termed \emph{NORTA defective}.

\citet{GH02} conducted a Monte Carlo study to estimate the probability of encountering NORTA defective matrices while increasing the dimension \(d\). They found that for what is now considered low-to-moderate dimensions (\(d \approx 20\)), almost \emph{all} feasible matrices are NORTA defective. This stems from the concentration of measure near the boundary of the space of all possible correlation matrices as dimension increases. Unfortunately, it is precisely near this boundary that NORTA defective matrices reside.

There is hope, however, as \citet{GH02} also showed that replacing an indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate \(d\). This provides evidence that our nearest positive definite (PD) augmented approach will maintain reasonable accuracy if our input matching scheme returns an indefinite matrix, at least for the rank-based matching scheme described above.

\hypertarget{rand-vec-gen}{%
\subsection{Random vector generator}\label{rand-vec-gen}}

We now describe our algorithm to generate random vectors, which mirrors the classical NORTA algorithm above with some modifications for rank-based dependency matching:

\setstretch{1.5}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Mapping step

  \begin{itemize}
  \item
    \begin{enumerate}
    \def\labelenumii{(\roman{enumii})}
    \tightlist
    \item
      Convert the target Spearman correlation matrix \(R_S\) to the corresponding MVN Pearson correlation \(R_X\). Alternatively,
    \end{enumerate}
  \item
    (i') Convert the target Kendall \(\tau\) matrix \(R_K\) to the corresponding MVN Pearson correlation \(R_X\). Alternatively,
  \item
    (i'\,') Convert the target Pearson correlation matrix to \(R_P\) to the corresponding, approximate MVN Pearson correlation \(R_X\).
  \end{itemize}
\item
  Check admissibility and, if needed, compute the nearest correlation matrix.

  \begin{itemize}
  \item
    \begin{enumerate}
    \def\labelenumii{(\roman{enumii})}
    \tightlist
    \item
      Check that \(R_X\) is a correlation matrix, a positive definite matrix with 1's along the diagonal.
    \end{enumerate}
  \item
    \begin{enumerate}
    \def\labelenumii{(\roman{enumii})}
    \setcounter{enumii}{1}
    \tightlist
    \item
      If \(R_X\) is a correlation matrix, the input matrix is \emph{admissible} in this scheme. Otherwise
    \end{enumerate}
  \item
    (ii') Replace \(R_X\) with the nearest correlation matrix \(\tilde{R}_X\), in the Frobenius norm.
  \end{itemize}
\item
  Gaussian copula

  \begin{itemize}
  \item
    \begin{enumerate}
    \def\labelenumii{(\roman{enumii})}
    \tightlist
    \item
      Generate \({\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)\).
    \end{enumerate}
  \item
    \begin{enumerate}
    \def\labelenumii{(\roman{enumii})}
    \setcounter{enumii}{1}
    \tightlist
    \item
      Transform \({\bf X}\) to \({\bf U} = (U_1, \ldots, U_d)\) viz.~\(U_i=\Phi(X_i)\), \(i=1, \ldots, d\).
    \end{enumerate}
  \item
    \begin{enumerate}
    \def\labelenumii{(\roman{enumii})}
    \setcounter{enumii}{2}
    \tightlist
    \item
      Return \({\bf Y} = (Y_1, \ldots, Y_d)\), where \(Y_i=F_i^{-1}(U_i)\), \(i=1, \ldots, d\).
    \end{enumerate}
  \end{itemize}
\end{enumerate}

\setstretch{2.0}

\emph{Step 1}

The first two descriptions of the \emph{Mapping step} employ the closed-form relationships between \(\rho_S\) and \(\tau\) with \(\rho_P\) for bivariate normal random variables via Equations \eqref{eq:convertKendall} and \eqref{eq:convertSpearman}, respectively (implemented as \texttt{cor\_covert}). Initializing our algorithm to match the nonparametric correlations by computing these equations for all pairs is computationally trivial.

For the computationally expensive process to match Pearson correlations, we approximate Equation \eqref{eq:pearsonIntegralRelation} for all pairs of margins. To this end, we implement the approximation scheme introduced by \citet{xiao2019matching}. Briefly, the many double integrals of the form in \eqref{eq:pearsonIntegralRelation} are approximated by weighted sums of Hermite polynomials. Matching coefficients for pairs of continuous distributions is made tractable by this method, but for discrete distributions (especially discrete distributions with large support sets or infinite support), the problem is not so simple or efficient.

Our solution is to approximate discrete distributions by a continuous distribution. The question then becomes that of which distribution to use. We found that Generalized S-Distributions \citep{muino2006gs} solve this problem by approximating a wide range of unimodal distributions, both continuous and discrete. Using the GS-Distribution approximation of discrete distributions in Pearson matching scheme above yields favorable results.

\emph{Step 2}

Once \(R_X\) has been determined, we check admissibility of the adjusted correlation via the steps described above. If \(R_X\) is not a valid correlation matrix, then we compute the nearest correlation matrix. Finding the nearest correlation matrix is a common statistical computing problem. The defacto default function in \texttt{R} is \texttt{Matrix::nearPD}, an alternating projection algorithm due to \citet{higham2002computing}. As implemented the function fails to scale to HD. Instead, we provide the quadratically convergent algorithm based on the theory of strongly semi-smooth matrix functions (\citet{qi2006quadratically}). The nearest correlation matrix problem can be written down as the following convex optimization problem:

\begin{align*}
    \mathrm{min}\quad & \frac{1}{2} \Vert G - X \Vert^2 \\
    \mathrm{s.t.}\quad & X_{ii} = 1, \quad i = 1, \ldots , n, \\
    & X \in S_{+}^{n}
\end{align*}

For nonparametric correlation measures, our algorithm allows the generation of high-dimensional multivariate data with arbitrary marginal distributions with a broad class of admissible Spearman correlation matrices and Kendall \(\tau\) matrices. The admissible classes consist of the matrices that map to a Pearson correlation matrix for a MVN. In particular, if we let \(X\) be MVN with \(d\) components and set

\begin{align*}
\Omega_P &= \{ R_P : R_P \textrm{ is a correlation matrix for } X \} \\
\Omega_K &= \{ R_K : R_K \textrm{ is a correlation matrix for } X \} \\
\Omega_S &= \{ R_S : R_S \textrm{ is a correlation matrix for } X \} \\
\end{align*}

then there are 1-1 mappings between these sets. We conjecture informally that the sets of admissible \(R_S\) and \(R_K\) are not highly restrictive. In particular, \(R_P\) is approximately \(R_S\) for a MVN, suggesting that the admissible set \(\Omega_S\) should be flexible as \(R_P\) can be any PD matrix with 1's along the diagonal. We provide methods to check whether a target \(R_S\) is an element of the \emph{admissible set} \(\Omega_S\) (and similarly for \(R_K\)).

There is an increasing probability of encountering a non-admissible correlation matrix as dimension increases. In our experience, the mapping step for large \(d\) almost always produces a \(R_X\) that is not a correlation matrix. In Section \ref{package} we provide a basic description of how to quantify and control the approximation error. Further, the RNA-seq example in Section \ref{examples} provides an illustration of this in practice.

\emph{Step 3}

The final step implements a NORTA-inspired, Gaussian copula approach to produce the desired margins. Steps 1 and 2 determine the MVN Pearson correlation values that will eventually match the target correlation. Then all that is required is a fast MVN simulator, a standard normal CDF, and well-defined quantile functions for marginals. The MVN is transformed to a copula (distribution with standard uniform margins) by applying the normal CDF \(\phi(\cdot)\). The quantile functions \(F^{-1}\) are applied across the margins to return the desired random vector \({\bf Y}\).

\section[bigsimr-pkg]{The \pkg{bigsimr} R package}\label{package}

\begin{CodeChunk}
\begin{CodeInput}
R> box::use(
+   patchwork[...],
+   dplyr[...],
+   ggplot2[...],
+   tidyr[drop_na],
+   bigsimr[bigsimr_setup, distributions_setup],
+   JuliaCall[julia_call]
+ )
R> 
R> Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
R> bs <- bigsimr_setup(pkg_check = FALSE)
\end{CodeInput}
\begin{CodeOutput}
Julia version 1.5.4 at location /home/alex/julia-1.5.4/bin will be used.
\end{CodeOutput}
\begin{CodeOutput}
Loading setup script for JuliaCall...
\end{CodeOutput}
\begin{CodeOutput}
Finish loading setup script for JuliaCall.
\end{CodeOutput}
\begin{CodeInput}
R> dist <- distributions_setup()
\end{CodeInput}
\end{CodeChunk}

This section describes a low-dimensional (2D) random vector simulation workflow via the \texttt{bigsimr} \texttt{R} package (\url{https://github.com/SchisslerGroup/r-bigsimr}). The package \texttt{bigsimr} provides an interface to the native code written in Julia, registered as the \texttt{Bigsimr} Julia package (\url{https://github.com/adknudson/Bigsimr.jl}). In addition to the native Julia \texttt{Bigsimr} package and \texttt{R} interface \texttt{bigsimr}, we also provide a python interface \texttt{bigsimr} (\url{https://github.com/SchisslerGroup/python-bigsimr/}) that interfaces with the Julia \texttt{Bigsimr} package. The Julia package provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \ref{algorithms}).

The subsections below describe the basic use of \texttt{bigsimr} by stepping through an example workflow using the data set \texttt{airquality} that contains daily air quality measurements in New York, May to September 1973 \citep{Chambers1983}. This workflow proceeds from setting up the computing environment, to data wrangling, estimation, simulation configuration, random vector generation, and, finally, result visualization.

\hypertarget{bivariate-example-description}{%
\subsection{Bivariate example description}\label{bivariate-example-description}}

We illustrate the use of \texttt{bigsimr} motivated by the New York air quality data set (\texttt{airquality}) included in the R \texttt{datasets} package. First, we load the \texttt{bigsimr} library and a few other convenient data science packages, including the syntactically-elegant \texttt{tidyverse} suite of \texttt{R} packages. The code chunk below prepares the computing environment:

\begin{CodeChunk}
\begin{CodeInput}
R> library("tidyverse")
R> library("bigsimr")
R> # Activate multithreading in Julia
R> Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
R> # Load the Bigsimr and Distributions Julia packages
R> bs <- bigsimr_setup()
R> dist <- distributions_setup()
\end{CodeInput}
\end{CodeChunk}

For simplicity and to provide a minimal working example, we consider bivariate simulation of the two \texttt{airquality} variables \texttt{Temperature}, in degrees Fahrenheit, and \texttt{Ozone} level, in parts per billion.

\begin{CodeChunk}
\begin{CodeInput}
R> df <- airquality %>% select(Temp, Ozone) %>% drop_na()
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> glimpse(df)
\end{CodeInput}
\begin{CodeOutput}
Rows: 116
Columns: 2
$ Temp  <int> 67, 72, 74, 62, 66, 65, 59, 61, 74, 69, 66, 68, 58, 64, 66, 57, ~
$ Ozone <int> 41, 36, 12, 18, 28, 23, 19, 8, 7, 16, 11, 14, 18, 14, 34, 6, 30,~
\end{CodeOutput}
\end{CodeChunk}

Figure \ref{fig:ch030-aq-joint-dist} visualizes the bivariate relationship between Ozone and Temperature. We aim to simulate random two-component vectors mimicking this structure. The margins are not normally distributed; particularly the Ozone level exhibits a strong positive skew.

\begin{CodeChunk}
\begin{CodeInput}
R> p0 <- ggplot(df, aes(Temp, Ozone)) +
+   geom_point(size = 1) +
+   theme(legend.position = "none") + 
+   labs(x = "Temperature")
R> 
R> pTemp <- ggplot(df, aes(Temp)) + 
+   geom_density() +
+   theme(axis.title.x = element_blank(),
+         axis.title.y = element_blank(),
+         axis.text = element_blank(),
+         axis.ticks = element_blank())
R> 
R> pOzone <- ggplot(df, aes(Ozone)) + 
+     geom_density() +
+   theme(axis.title.y = element_blank(),
+         axis.title.x = element_blank(),
+         axis.ticks = element_blank(),
+         axis.text = element_blank()) +
+   coord_flip()
R> 
R> pTemp + plot_spacer() + p0 + pOzone + 
+   plot_layout(widths = c(3,1), heights = c(1, 3)) 
\end{CodeInput}
\begin{figure}

{\centering \includegraphics{article_bigsimr_files/figure-latex/ch030-aq-joint-dist-1}

}

\caption[Bivariate scatterplot of Ozone vs]{Bivariate scatterplot of Ozone vs. Temp with estimated marginal densities. We model the Ozone data as marginally log-normal and the Temperature data as normal.}\label{fig:ch030-aq-joint-dist}
\end{figure}
\end{CodeChunk}

Next, we specify the marginal distributions and correlation coefficient (both type and magnitude). Here the analyst is free to be creative. For this example, we avoid goodness-of-fit considerations to determine the marginal distributions. But it seems sensible without domain knowledge to estimate these quantities from the data, and \texttt{bigsimr} contains fast functions designed for this task.

\hypertarget{specifying-marginal-distributions}{%
\subsection{Specifying marginal distributions}\label{specifying-marginal-distributions}}

Based on the estimated densities in Figure \ref{fig:ch030-aq-joint-dist}, we assume \texttt{Temp} is normally distributed and \texttt{Ozone} is log-normally distributed, as the latter values are positive and skewed. We use the well-known, unbiased estimators for the normal distribution's parameters and maximum likelihood estimators for the log-normal parameters:

\begin{CodeChunk}
\begin{CodeInput}
R> df %>% select(Temp) %>% 
+   summarise_all(.funs = c(mean = mean, sd = sd))
\end{CodeInput}
\begin{CodeOutput}
   mean    sd
1 77.87 9.485
\end{CodeOutput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> mle_mean <- function(x) mean(log(x))
R> mle_sd <- function(x) mean( sqrt( (log(x) - mean(log(x)))^2 ) )
R> df %>% 
+   select(Ozone) %>% 
+   summarise_all(.funs = c(meanlog = mle_mean, sdlog = mle_sd))
\end{CodeInput}
\begin{CodeOutput}
  meanlog  sdlog
1   3.419 0.6967
\end{CodeOutput}
\end{CodeChunk}

Next, we configure the input marginals for later input into \texttt{rvec}. The marginal distributions are specified using \texttt{Julia}'s \texttt{Distributions} package and stored in a vector.

\begin{CodeChunk}
\begin{CodeInput}
R> margins <- c(dist$Normal(mean(df$Temp), sd(df$Temp)),
+              dist$LogNormal(mle_mean(df$Ozone), mle_sd(df$Ozone)))
\end{CodeInput}
\end{CodeChunk}

\hypertarget{specifying-correlation}{%
\subsection{Specifying correlation}\label{specifying-correlation}}

As mentioned, the user must decide how to describe correlation based on the particulars of the problem. For non-normal data and for improved simulation accuracy/scalability in our scheme, we advocate the use of Spearman's \(\rho\) correlation matrix \(R_S\) and Kendall's \(\tau\) correlation matrix \(R_K\). We also support Pearson correlation coefficient matching, while cautioning the user to check the performance for their parametric multivariate model (see \protect\hyperlink{simulations}{Monte Carlo evaluations} below for evaluation strategies and guidance). These estimation methods are classical approaches, not designed for high-dimensional correlation estimation (see the \href{\%7B\#discussion}{Conclusion and Discussion} sections for more on this).

\begin{CodeChunk}
\begin{CodeInput}
R> (R_S <- bs$cor(as.matrix(df), bs$Spearman))
\end{CodeInput}
\begin{CodeOutput}
      [,1]  [,2]
[1,] 1.000 0.774
[2,] 0.774 1.000
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{checking-target-correlation-matrix-admissibility}{%
\subsection{Checking target correlation matrix admissibility}\label{checking-target-correlation-matrix-admissibility}}

First we use \texttt{cor\_bounds} to ensure that the pairwise target correlation values are valid prior to the mapping step which constructs \(R_X\) for the MVN input into NORTA (see Section \ref{algorithms}). \texttt{cor\_bounds} estimates the pairwise lower and upper correlation bounds using the Generate, Sort, and Correlate algorithm of \citet{DH2011}.

\begin{CodeChunk}
\begin{CodeInput}
R> bounds <- bs$cor_bounds(margins)
R> bounds$lower
\end{CodeInput}
\begin{CodeOutput}
        [,1]    [,2]
[1,]  1.0000 -0.8902
[2,] -0.8902  1.0000
\end{CodeOutput}
\begin{CodeInput}
R> bounds$upper
\end{CodeInput}
\begin{CodeOutput}
       [,1]   [,2]
[1,] 1.0000 0.8876
[2,] 0.8876 1.0000
\end{CodeOutput}
\end{CodeChunk}

Since our single estimated Spearman correlation is within the theoretical bounds, the correlation is valid as input to \texttt{rvec}. But even if the 2-dimensional bounds are satisfied for each pair of margins, this does not guarantee the feasibility of a \(d-\)variate distribution \citep{BF17}.

To provide higher dimensional feasibility/admissibility checking, we begin by mapping using \texttt{cor\_convert} and check admissibility.

\begin{CodeChunk}
\begin{CodeInput}
R> # Step 1. Mapping
R> (R_X <- bs$cor_convert(R_S, bs$Spearman, bs$Pearson))
\end{CodeInput}
\begin{CodeOutput}
       [,1]   [,2]
[1,] 1.0000 0.7886
[2,] 0.7886 1.0000
\end{CodeOutput}
\begin{CodeInput}
R> # Step 2. Check admissibility
R> bs$iscorrelation(R_X)
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\end{CodeChunk}

The bounds on the Pearson correlation coefficient \(R_P\) between these margins are restricted as seen below in our MC estimated correlation bounds:

\begin{CodeChunk}
\begin{CodeInput}
R> bs$cor_bounds(margins[1], margins[2], bs$Pearson, n_samples = 1e6)
\end{CodeInput}
\begin{CodeOutput}
Julia Object of type NamedTuple{(:lower, :upper),Tuple{Float64,Float64}}.
(lower = -0.8822730883942045, upper = 0.8833146839651883)
\end{CodeOutput}
\end{CodeChunk}

Our MC estimate of the bounds slightly overestimates the theoretic bounds of \((-0.881, 0.881)\). An analytic derivation is presented as an Appendix.

\hypertarget{simulating-random-vectors}{%
\subsection{Simulating random vectors}\label{simulating-random-vectors}}

Finally, we arrive at the main function of \texttt{bigsimr}: \texttt{rvec}. We now simulate \(B=10,000\) random vectors from the assumed joint distribution of Ozone levels and Temp.

\begin{CodeChunk}
\begin{CodeInput}
R> x <- bs$rvec(10000, R_X, margins)
R> df_sim <- as.data.frame(x)
R> colnames(df_sim) <- colnames(df)
\end{CodeInput}
\end{CodeChunk}

Figure \ref{fig:ch030-plot-sim} plots the 10,000 simulated points.

\begin{CodeChunk}
\begin{CodeInput}
R> p1 <- df_sim %>%
+   ggplot(aes(Temp, Ozone)) +
+   geom_density_2d_filled() +
+   theme(legend.position = "none") +
+   labs(x = "Simulated Temperature", y = "Simulated Ozone") +
+   scale_y_continuous(limits = c(0, max(df$Ozone))) +
+   scale_x_continuous(limits = range(df$Temp))
R> 
R> p1Temp <- ggplot(df_sim, aes(Temp)) + 
+   geom_density(alpha = 0.5, fill = "lightseagreen") +
+   theme(axis.title.x = element_blank(),
+         axis.title.y = element_blank(),
+         axis.text = element_blank(),
+         axis.ticks = element_blank())
R> 
R> p1Ozone <- ggplot(df_sim, aes(Ozone)) + 
+   geom_density(alpha = 0.5, fill = "lightseagreen") + 
+   theme(axis.title.y = element_blank(),
+         axis.title.x = element_blank(),
+         axis.ticks = element_blank(),
+         axis.text = element_blank()) +
+   coord_flip() 
R> 
R> p1Temp + plot_spacer() + p1 + p1Ozone + 
+   plot_layout(widths = c(3,1), heights = c(1, 3))
\end{CodeInput}
\begin{CodeOutput}
Warning: Removed 342 rows containing non-finite values (stat_density2d_filled).
\end{CodeOutput}
\begin{figure}

{\centering \includegraphics{article_bigsimr_files/figure-latex/ch030-plot-sim-1}

}

\caption[Contour plot and marginal densities for the simulated bivariate distribution of Air Quality Temperatures and Ozone levels]{Contour plot and marginal densities for the simulated bivariate distribution of Air Quality Temperatures and Ozone levels. The simulated points mimic the observed data with respect to both the marginal characteristics and bivariate association.}\label{fig:ch030-plot-sim}
\end{figure}
\end{CodeChunk}

\hypertarget{simulations}{%
\section{Monte Carlo evaluations}\label{simulations}}

Before applying our methodology to real data simulation, we conduct several Monte Carlo studies to investigate method performance. Since marginal parameter matching in our scheme is essentially a sequence of univariate inverse probability transforms, the challenging aspects are the accuracy of dependency matching and computational efficiency at high dimensions. To evaluate our methods in those respects, we design the following numerical experiments to first assess accuracy of matching dependency parameters in bivariate simulations and then time the procedure in increasingly large dimension \(d\).

\hypertarget{bivariate-experiments}{%
\subsection{Bivariate experiments}\label{bivariate-experiments}}

We select bivariate simulation configurations to ultimately simulate our motivating discrete-valued RNA-seq example, and, so we proceed in increasing complexity, leading to the model in our motivating application in Section \ref{examples}. We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson's, Spearman's, and Kendall's --- in identical, bivariate marginal configurations. For each pair of identical margins, we vary the target correlation across \(\Omega\), the set of possible admissible values for correlation type, to evaluate the simulation's ability to obtain the theoretic bounds. The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (mimicking RNA-seq counts).

Table \ref{tab:sims} lists our identical-marginal, bivariate simulation configurations. We increase the simulate replicates \(B\) to check that our results converge to the target correlations and gauge statistical efficiency. We select distributions beginning with a standard multivariate normal (MVN) as we expect the performance to be exact (up to MC error) for all correlation types. Then, we select a non-symmetric continuous distribution: a standard (rate =1) two-component multivariate gamma (MVG). Finally, we select distributions and marginal parameter values that are motivated by our RNA-seq data, namely values proximal to probabilities and sizes estimated from the data (see \href{examples}{Example applications} for estimation details). Thus we arrive at a multivariate negative binomial (MVNB) \(p_1 = p_2 = 3\times10^{-4}, r_1 = r_2 = 4, \rho \in \Omega\).

Table: \label{tab:sims} Identical margin, bivariate simulation configurations to evaluate correlation matching accuracy and efficiency.

\begin{table}[]
\begin{tabular}{@{}lcr@{}}
\toprule
Simulation Reps $B$ & Correlation Types & Identical-margin 2D distribution \\ \midrule
$1000$ & Pearson ($\rho_P$) & ${\bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho_i ), i=1,\ldots,100$ \\
$10,000$ & Spearman ($\rho_S$) & ${\bf Y} \sim MVG( shape = 10, rate = 1, \rho_i ), i=1,\ldots,100$ \\
$100,000$ & Kendall ($\tau$) & ${\bf Y} \sim MVNB(p = 3\times10^{-4}, r = 4,\rho_i), i=1,\ldots,100$ \\ \bottomrule
\end{tabular}
\end{table}

For each of the unique 9 simulation configurations described above, we estimate the correlation bounds and vary the correlations along a sequence of 100 points evenly placed within the bounds, aiming to explore \(\Omega\). Specifically, we set correlations \(\{ \rho_1 = ( \hat{l} + \epsilon), \rho_2 = (\hat{l} + \epsilon) + \delta, \ldots, \rho_{100} = (\hat{u} - \epsilon) \}\), with \(\hat{l}\) and \(\hat{u}\) being the estimated lower and upper bounds, respectively, and increment value \(\delta\). The adjustment factor, \(\epsilon=0.01\), is introduced to handle numeric issues when the bound is specified exactly.

\begin{CodeChunk}
\begin{CodeInput}
R> bivariate_normal_sims %>%
+     ggplot(aes(rho, rho_hat, color = type)) +
+     geom_point() +
+     geom_abline(slope = 1) +
+     facet_wrap(~ type + N) + 
+     theme_bw()
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> bivariate_gamma_sims %>%
+     ggplot(aes(rho, rho_hat, color = type)) +
+     geom_point() +
+     geom_abline(slope = 1) +
+     facet_wrap(~ type + N) + 
+     theme_bw()
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> bivariate_nbinom_sims %>%
+     ggplot(aes(rho, rho_hat, color = type)) +
+     geom_point() +
+     geom_abline(slope = 1) +
+     facet_wrap(~ type + N) + 
+     theme_bw()
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> allDat <- bind_rows(
+     select(bivariate_normal_sims, margins, type, N, rho, rho_hat),
+     select(bivariate_gamma_sims, margins, type, N, rho, rho_hat),
+     select(bivariate_nbinom_sims, margins, type, N, rho, rho_hat)
+ ) %>%
+     mutate(margins = factor(margins, levels=c("norm", "gamma", "nbinom")),
+            type = factor(type, levels=c("Pearson", "Spearman", "Kendall")),
+            N = factor(N))
\end{CodeInput}
\end{CodeChunk}

Figure \ref{fig:ch040-bPlot} displays the aggregated bivariate simulation results. Table \ref{tab:ch040-BiError} contains the mean absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.

\begin{CodeChunk}
\begin{CodeInput}
R> tabMAE <- allDat %>%
+     group_by(N, type, margins) %>%
+     summarize(MAE = mean(abs(rho - rho_hat))) %>%
+     ungroup()
\end{CodeInput}
\begin{CodeOutput}
`summarise()` has grouped output by 'N', 'type'. You can override using the `.groups` argument.
\end{CodeOutput}
\begin{CodeInput}
R> kable(tabMAE, booktabs = TRUE, format = "latex",
+       linesep = c("", "", "\\addlinespace"),
+       col.names = c("No. of random vectors",
+                     "Correlation type",
+                     "Distribution",
+                     "Mean abs. error"),
+       caption = "Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.")
\end{CodeInput}
\begin{table}

\caption{\label{tab:ch040-BiError}Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.}
\centering
\begin{tabular}[t]{lllr}
\toprule
No. of random vectors & Correlation type & Distribution & Mean abs. error\\
\midrule
1000 & Pearson & norm & 0.0168\\
1000 & Pearson & gamma & 0.0159\\
1000 & Pearson & nbinom & 0.0188\\
\addlinespace
1000 & Spearman & norm & 0.0203\\
1000 & Spearman & gamma & 0.0180\\
1000 & Spearman & nbinom & 0.0165\\
\addlinespace
1000 & Kendall & norm & 0.0118\\
1000 & Kendall & gamma & 0.0099\\
1000 & Kendall & nbinom & 0.0089\\
\addlinespace
10000 & Pearson & norm & 0.0048\\
10000 & Pearson & gamma & 0.0055\\
10000 & Pearson & nbinom & 0.0066\\
\addlinespace
10000 & Spearman & norm & 0.0058\\
10000 & Spearman & gamma & 0.0054\\
10000 & Spearman & nbinom & 0.0060\\
\addlinespace
10000 & Kendall & norm & 0.0034\\
10000 & Kendall & gamma & 0.0031\\
10000 & Kendall & nbinom & 0.0037\\
\addlinespace
1e+05 & Pearson & norm & 0.0016\\
1e+05 & Pearson & gamma & 0.0016\\
1e+05 & Pearson & nbinom & 0.0032\\
\addlinespace
1e+05 & Spearman & norm & 0.0018\\
1e+05 & Spearman & gamma & 0.0019\\
1e+05 & Spearman & nbinom & 0.0013\\
\addlinespace
1e+05 & Kendall & norm & 0.0011\\
1e+05 & Kendall & gamma & 0.0010\\
1e+05 & Kendall & nbinom & 0.0010\\
\bottomrule
\end{tabular}
\end{table}

\end{CodeChunk}

Overall, the studies show that our methodology is generally accurate across the entire range of possible correlation values all three dependency measures, at least in these limited simulation settings for the rank-based correlations. Our Pearson matching performs nearly as well as Spearman or Kendall, except for a slight increase in error for negative binomial case. This is due the particularly large counts generated with our choice of parameters for \(p\) and \(r\).

\begin{CodeChunk}
\begin{CodeInput}
R> # https://www.datanovia.com/en/blog/how-to-change-ggplot-facet-labels/
R> # New facet label names
R> repsLabs <- paste0("B=", c("1,000", "10,000", "100,000") )
R> names(repsLabs) <- c(1000, 10000, 100000)
R> typeLabs <- c("Pearson", "Spearman", "Kendall")
R> names(typeLabs) <- c("Pearson", "Spearman", "Kendall")
R> 
R> # Set colors
R> ## RColorBrewer::display.brewer.all()
R> numColors <- 4
R> numGroups <- length(levels(allDat$margins))
R> myColors <- rev(RColorBrewer::brewer.pal(n = numColors, name = "Greys")[ ((numColors - numGroups) + 1):numColors])
R> 
R> allDat %>%
+     ggplot(aes(x = rho, y = rho_hat, color = margins)) +
+     geom_point(size = 2) +
+     scale_color_manual(values = myColors ) +
+     geom_abline(slope = 1, linetype = "dashed") +
+     labs(x = "Specified Correlation", y = "Estimated Correlation") +
+     facet_wrap(~ type + N, labeller = labeller(N = repsLabs, type = typeLabs)) +
+     theme_bw() +
+     theme(legend.position = "bottom", legend.direction = "horizontal")
\end{CodeInput}
\begin{figure}

{\centering \includegraphics{article_bigsimr_files/figure-latex/ch040-bPlot-1}

}

\caption[Bivariate simulations match target correlations across the entire range of feasible correlations]{Bivariate simulations match target correlations across the entire range of feasible correlations. The horizontal axis plots the specified target correlations for each bivariate margin. Normal margins are plotted in dark dark grey, gamma in medium grey, and negative binomial in light grey. As the number of simulated vectors $B$ increases from left to right, the variation in estimated correlations (vertical axis) decreases. The dashed line indicates equality between the specified and estimated correlations.}\label{fig:ch040-bPlot}
\end{figure}
\end{CodeChunk}

\hypertarget{scale-up-to-high-dimensions}{%
\subsection{Scale up to High Dimensions}\label{scale-up-to-high-dimensions}}

With information of our method's accuracy from a low-dimensional perspective, we now turn to assessing whether \texttt{bigsimr} can scale to larger dimensional problems with practical computation times. Specifically, we ultimately generate \(B=1,000\) random vectors for \(d=\{100, 250, 500, 1000, 2500, 5000, 10000\}\) for each correlation type, \(\{Pearson, Spearman, Kendall\}\) while timing the algorithm's major steps.

To mimic the workflow, we first produce a synthetic ``data set'' by completing the following steps:

\setstretch{1.5}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Produce heterogeneous gamma marginals by randomly selecting the \(j^{th}\) gamma shape parameter from \(U_j \sim uniform(1,10), j=1,\ldots,d\) and the \(j^{th}\) rate parameter from \(V_j \sim exp(1/5), j=1,\ldots,d\), with the constant parameters determined arbitrarily.
\item
  Produce a random full-rank Pearson correlation matrix via \texttt{cor\_randPD} of size \(d \times d\).
\item
  Simulate a ``data set'' of \(1,000 \times d\) random vectors via \texttt{rvec} (without matching the Pearson exactly).
  \setstretch{2.0}
\end{enumerate}

With the synthetic data set in hand, we complete and time the following 4 steps:

\setstretch{1.5}

\begin{enumerate}
\def\labelenumi{\roman{enumi}.}
\tightlist
\item
  Estimate the correlation matrix from the ``data'' in the \emph{Compute Correlation} step.
\end{enumerate}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Map the correlations to initialize the algorithm (Pearson to Pearson, Spearman to Pearson, or Kendall to Pearson) in the \emph{Adjust Correlation} step.
\item
  Check whether the mapping produces a valid correlation matrix and, if not, find the nearest PD correlation matrix in the \emph{Check Admissibility} step.
\item
  Simulate \(1,000\) vectors via the Inverse Transform method in the \emph{Simulate Data} step.
  \setstretch{2.0}
\end{enumerate}

The experiments are conducted on a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, with all 16 threads employed during computation. The results for \(d \leq 500\) are fast with all times (except Kendall) under 3 seconds. Table \ref{tab:ch040-moderateDtab} displays the total computation time of the algorithms steps 1 through 3, including estimation step i.

\begin{CodeChunk}
\begin{CodeInput}
R> dat_long <- benchmark_dependences %>%
+   select(-total_time, -n_sim, -needed_near_pd) %>%
+   pivot_longer(cols = c(corr_time:sim_time), names_to = "Step", values_to = "Time") %>%
+   mutate(dim = factor(dim),
+          Step = factor(Step,
+                        levels = c("corr_time", "adj_time", "admiss_time", "sim_time"),
+                        labels = c("Compute Correlation",
+                                   "Adjust Correlation",
+                                   "Check Admissibility",
+                                   "Simulate Data"))) %>%
+     rename(Correlation = corr_type, Dimensions = dim)
R> 
R> tabTimes  <- dat_long %>%
+     filter(Dimensions %in% c(100, 250, 500)) %>%
+     group_by(Dimensions, Correlation) %>%
+     summarize('Total Time(Seconds)' = sum(Time)) %>%
+     ungroup()
\end{CodeInput}
\begin{CodeOutput}
`summarise()` has grouped output by 'Dimensions'. You can override using the `.groups` argument.
\end{CodeOutput}
\begin{CodeInput}
R> tabTimes %>%
+     kable(booktabs = TRUE,
+           linesep = c("", "", "\\addlinespace"),
+           col.names = c("Dimension",
+                         "Correlation type",
+                         "Total Time (Seconds)"),
+           caption = 'Total time to produce 10,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.')
\end{CodeInput}
\begin{table}

\caption{\label{tab:ch040-moderateDtab}Total time to produce 10,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.}
\centering
\begin{tabular}[t]{llr}
\toprule
Dimension & Correlation type & Total Time (Seconds)\\
\midrule
100 & Pearson & 0.080\\
100 & Spearman & 0.036\\
100 & Kendall & 0.116\\
\addlinespace
250 & Pearson & 0.286\\
250 & Spearman & 0.371\\
250 & Kendall & 0.366\\
\addlinespace
500 & Pearson & 2.932\\
500 & Spearman & 0.176\\
500 & Kendall & 2.183\\
\bottomrule
\end{tabular}
\end{table}

\end{CodeChunk}

The results with larger dimensional vectors show scalability to ultra-high dimensions for all three correlation types, although the total times do become much larger.
Figure \ref{fig:ch040-largeDfig} displays computation times for \(d=\{1000, 2500, 5000, 10000\}\). For \(d\) equal to 1000 and 2500, the total time is under a couple of minutes. At \(d\) of 5000 and 10,000, Pearson correlation matching in the \emph{Adjust Correlation} step becomes costly. Interestingly, Pearson is actually faster than Kendall for \(d=10,000\) due to bottlenecks in \emph{Compute Correlation} and \emph{Check Admissibility}. Uniformly, matching Spearman correlations is faster, with total times under 5 minutes for \(d=10,000\), making Spearman correlations the most computationally-friendly dependency type. With this in mind, we scaled the simulation to \(d=20,000\) for the Spearman type and obtained the 1,000 vectors in under an hour (data not shown). In principle, this would enable the simulation of an entire human-derived RNA-seq data set. We note that for a given target correlation matrix and margins, steps i, 1, and 2 only need to be computed once and the third step, \emph{Simulate Data}, is fast under all schemes for all \(d\) under consideration.

\begin{CodeChunk}
\begin{CodeInput}
R> # Set colors
R> numColors <- 5
R> numGroups <- 4
R> myColors <- rev(brewer.pal(n=numColors, name="Greys")[((numColors-numGroups)+1):numColors])
R> 
R> dat_long %>%
+     filter(Dimensions %in% c(1000, 2500, 5000, 10000, 20000)) %>%
+     mutate(`Time (minutes)` = Time / 60) %>%
+     ggplot(aes(Correlation, `Time (minutes)`, fill=Step)) +
+     geom_bar(position = "stack", stat = "identity") +
+     scale_fill_manual(values = myColors ) +
+     scale_y_continuous(breaks = seq(0, 80, 10),
+                      minor_breaks = NULL) +
+     facet_grid(. ~ Dimensions) +
+     theme(axis.text.x = element_text(angle = -90))
\end{CodeInput}
\begin{figure}

{\centering \includegraphics{article_bigsimr_files/figure-latex/ch040-largeDfig-1}

}

\caption[Computation times as d increases]{Computation times as d increases.}\label{fig:ch040-largeDfig}
\end{figure}
\end{CodeChunk}

\emph{Limitations, conclusions, and recommendations}

In the bivariate studies, we chose arbitrary simulation parameters for three distributions, moving from the Gaussian to the discrete and non-normal, multivariate negative binomial. Under these conditions, the simulated random vectors sample the desired bivariate distribution across the entire range of pairwise correlations for the three dependency measures. The simulation results could differ for other choices of simulation settings. Specifying extreme correlations near the boundary or Frechet bounds could result in poor simulation performance. Fortunately, it is straightforward to evaluate simulation performance by using strategies similar to those completed above. We expect our random vector generation to perform well for the vast majority of NORTA-feasible correlation matrices, but advise to check the performance before making inferences/further analyses.

Somewhat surprising, Kendall estimation and nearest PD computation scale poorly compared to Spearman and, even, approximate Pearson matching. In our experience, Kendall computation times are sensitive to the number of cores, benefiting from multi-core parallelization. This could mitigate some of the current algorithmic/implementation shortcomings. Despite this, Kendall matching is still feasible for most high-dimensional data sets. Finally, we note that one could use our single-pass algorithm \texttt{cor\_fastPD} to produce a `close' (not nearest PD) to scale to even higher dimensions with greater loss of accuracy.

\hypertarget{examples}{%
\section{RNA-seq data applications}\label{examples}}

This section demonstrates how to simulate multivariate data using \texttt{bigsimr}, aiming to replicate the structure of high-dimensional dependent count data. In an illustration of our proposed methodology, we seek to simulate RNA-sequencing data by producing simulated random vectors mimicking the observed data and its generating process. Modeling RNA-seq using multivariate probability distributions is natural as inter-gene correlation is an inherent part of biological processes \citep{Wang2009b}. And yet, many models do not account for this, leading to major disruptions to the operating characteristics of statistical estimation, testing, and prediction. See \citet{Efron2012} for a detailed discussion with related methods and \citet{Wu2012b}, \citet{Schissler2018}; \citet{Schissler2019} for applied examples. The following subsections apply \texttt{bigsimr}'s methods to real RNA-seq data, including replicating an estimated parametric structure, MC probability estimation, and MC evaluation of correlation estimation efficiency.

\hypertarget{simulating-high-dimensional-rna-seq-data}{%
\subsection{Simulating High-Dimensional RNA-seq data}\label{simulating-high-dimensional-rna-seq-data}}

\begin{CodeChunk}
\begin{CodeInput}
R> d <- 1000
R> brca1000 <- example_brca %>%
+   select(all_of(1:d)) %>%
+   mutate(across(everything(), as.double))
\end{CodeInput}
\end{CodeChunk}

We begin by estimating the structure of the TCGA BRCA RNA-seq data set (see \href{background}{Background}). Ultimately, we will simulate \(B=10,000\) random vectors \({\bf Y}=(Y_1, \ldots, Y_d)^\top\) with \(d=1000\). We assume a multivariate negative binomial (MVNB) model as RNA-seq counts are often over-dispersed and correlated. Since all \(d\) selected genes exhibit over-dispersion (data not shown), we proceed to estimate the NB parameters \((r_i, p_i), i=1,\ldots,d\), to determine the target marginal PMFs \(f_i\). To complete specification of the simulation algorithm inputs, we estimate the Spearman correlation matrix \({ \bf R}_{S}\) to characterize dependency.

With this goal in mind, we first estimate the desired correlation matrix using the fast implementation provided by \texttt{bigsimr}:

\begin{CodeChunk}
\begin{CodeInput}
R> # Estimate Spearman's correlation on the count data
R> R_S <- bs$cor(as.matrix(brca1000), bs$Spearman)
\end{CodeInput}
\end{CodeChunk}

Next, we estimate the marginal parameters. We use the method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model. While, marginal distributions are from the same probability family (NB), they are heterogeneous in terms of the parameters probability and size \((p_i, n_i)\) for \(i,\ldots,d\). The functions below support this estimation for later use in \texttt{rvec}.

\begin{CodeChunk}
\begin{CodeInput}
R> make_nbinom_margins <- function(sizes, probs) {
+   margins <- lapply(1:length(sizes), function(i) {
+     dist$NegativeBinomial(sizes[i], probs[i])
+   })
+   do.call(c, margins)
+ }
\end{CodeInput}
\end{CodeChunk}

We apply these estimators to all 1000 genes across the 878 patients:

\begin{CodeChunk}
\begin{CodeInput}
R> nbinom_fit <- apply(brca1000, 2, mom_nbinom)
R> sizes <- nbinom_fit["size",]
R> probs <- nbinom_fit["prob",]
R> nb_margins <- make_nbinom_margins(sizes, probs)
\end{CodeInput}
\end{CodeChunk}

Notably, the estimated marginal NB probabilities \(\{ \hat{p}_i \}\) are small --- ranging in the interval \([\ensuremath{7.5305\times 10^{-7}} , \ensuremath{5.6592\times 10^{-4}}]\). This gives rise to highly variable counts and, typically, less restriction on potential pairwise correlation pairs. Once the functions are defined/executed to complete marginal estimation, we specify targets and generate the desired random vectors using \texttt{rvec}. Now we check admissibility of the specified correlation matrix.

\begin{CodeChunk}
\begin{CodeInput}
R> # 1. Mapping step first
R> R_X <- bs$cor_convert(R_S, bs$Spearman, bs$Pearson)
R> # 2a. Check admissibility
R> (is_valid_corr <- bs$iscorrelation(R_X))
\end{CodeInput}
\begin{CodeOutput}
[1] FALSE
\end{CodeOutput}
\begin{CodeInput}
R> # 2b. compute nearest correlation
R> if (!is_valid_corr) {
+   R_X_pd <- bs$cor_nearPD(R_X)
+   ## Quantify the error
+   targets      <- R_X[lower.tri(R_X, diag = FALSE)]
+   approximates <- R_X_pd[lower.tri(R_X_pd, diag = FALSE)]
+   R_X          <- R_X_pd
+ }
R> summary(abs(targets - approximates))
\end{CodeInput}
\begin{CodeOutput}
    Min.  1st Qu.   Median     Mean  3rd Qu.     Max. 
0.000000 0.000201 0.000432 0.000534 0.000753 0.006931 
\end{CodeOutput}
\end{CodeChunk}

While the exact \(d \times d\) Spearman correlation matrix is not strictly admissible in our scheme (as seen by the non-positive definite result above), the approximation is close with a maximum absolute error of 0.0069 and average absolute error of \ensuremath{5.3366\times 10^{-4}} across the \ensuremath{4.995\times 10^{5}} correlations.

\begin{CodeChunk}
\begin{CodeInput}
R> sim_nbinom <- bs$rvec(10000, R_X, nb_margins) 
R> colnames(sim_nbinom) <- colnames(brca1000)
\end{CodeInput}
\end{CodeChunk}

Figure \ref{fig:ch050-simDataFig} displays the simulated counts and pairwise relationships for our example genes from Table \ref{tab:ch010-realDataTab}.
Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parametric form and with less extreme points then the observed data in Figure \ref{fig:ch010-realDataFig}. Figure \ref{fig:ch050-figBRCA} displays the aggregated results of our simulation by comparing the specified target parameter (horizontal axes) with the corresponding quantities estimated from the simulated data (vertical axes). The evaluation shows that the simulated counts approximately match the target parameters and exhibit the full range of estimated correlation from the data.

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(2020-02-25)
R> num_genes <- 3
R> gene_sample <- sample(example_genes[1:1000], num_genes)
R> 
R> ggpairs(data = as.data.frame(sim_nbinom[, gene_sample]),
+         upper = list(continuous = wrap('cor', method = "spearman"))) + 
+   theme_bw()
\end{CodeInput}
\begin{CodeOutput}
Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
exact p-value with ties
\end{CodeOutput}
\begin{CodeOutput}
Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
exact p-value with ties
\end{CodeOutput}
\begin{CodeOutput}
Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
exact p-value with ties
\end{CodeOutput}
\begin{figure}

{\centering \includegraphics{article_bigsimr_files/figure-latex/ch050-simDataFig-1}

}

\caption[Simulated data for three selected high-expressing genes generally replicates the estimated data structure]{Simulated data for three selected high-expressing genes generally replicates the estimated data structure. The data do not exhibit outlying points, but do possess the desired Spearman correlations, central tendencies, and discrete values.}\label{fig:ch050-simDataFig}
\end{figure}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> include_graphics('fig/ch050-figBRCA.png')
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{fig/ch050-figBRCA}

}

\caption[Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set]{Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set. The dashed red lines indicate equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes).}\label{fig:ch050-figBRCA}
\end{figure}
\end{CodeChunk}

\emph{Limitations, conclusions, and recommendations}

The results show overall good simulation performance for our choice of parameters settings. Our settings were motivated by modeling high-expressing genes from the TCGA BRCA data set. In general, the ability to match marginal and dependence parameters depends on the particular joint probability model. We recommend to evaluate and tune your simulation until you can be assured of the accuracy.

\hypertarget{simulation-based-joint-probability-calculations}{%
\subsection{Simulation-based joint probability calculations}\label{simulation-based-joint-probability-calculations}}

Many statistical tasks require evaluation of a joint probability mass (or density) function:

\[
P( {\bf Y} = {\bf y} ), \: y_i \in \chi_i.
\]

where \(\chi_i\) is the sample space for the \(i^{th}\) component of the random vector \(\bf{Y}\). Compact representations with convenient computational forms are rare for high-dimensional constructions, especially with heterogeneous, correlated marginal distributions (or margins of mixed data types). Given a large number of simulated vectors as produced above, estimated probabilities are readily given by counting the proportion of simulated vectors meeting the desired condition. In our motivating application, one may ask what is the probability that all genes expressed greater than a certain threshold value \({\bf y}_0\). This can be estimated as

\[
\hat{P}( {\bf Y} \ge {\bf y_0 } ) = \frac{1}{B} \sum_{b=1}^B I( {\bf Y}^{ (b) } \ge {\bf y_0} ),
\]

where \({\bf Y}^{(b)}\) is the \(b^{th}\) simulated vector from a total of \(B\) simulation replicates and \(I(\cdot)\) is the indicator function. For example, we can estimate from our \(B=10,000\) simulated vectors that the probability of all genes expressing (i.e., \({\bf y}_i \geq 1, \forall \; i\)) is \(0.5979\).

\begin{CodeChunk}
\begin{CodeInput}
R> d <- ncol(sim_nbinom)
R> B <- nrow(sim_nbinom)
R> y0 <- 1
R> pHat <- mean(apply(sim_nbinom, 1, function(Y) {
+   all(Y >= y0)
+ }))
R> pHat
\end{CodeInput}
\begin{CodeOutput}
[1] 0.5979
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{evaluation-of-correlation-estimation-efficiency}{%
\subsection{Evaluation of correlation estimation efficiency}\label{evaluation-of-correlation-estimation-efficiency}}

MC methods are routinely used in many statistical inferential tasks including estimation, hypothesis testing, error rates, and empirical interval coverage rates. To conclude the example applications, we demonstrate how \texttt{bigsimr} can be used to evaluate estimation efficiency. In particular, we wish to assess the error in our correlation estimation above. We used a conventional method, based on classical statistical theory which was not designed for high-dimensional data. Indeed, high-dimensional covariance estimation (and precision matrices) is an active area of statistical science, .e.g., \citep{Won2013g, VanWieringen2016}.

In this small example, we simulate \(m=30\) data sets with the number of simulated vectors matching the number of patients in the BRCA data set, \(N=878\). Since our simulation is much faster for the Pearson correlation type (see Figure \ref{fig:ch040-largeDfig}), we only convert the Spearman correlation matrix once (and ensure it is positive definite). At each iteration, we estimate the quadratic loss (residual sum of squared errors) from the specified \({\bf R}_{S}\), producing a distribution of loss values.

\begin{CodeChunk}
\begin{CodeInput}
R> # Simulate random vectors equal to the sample size
R> N <- nrow(example_brca)
R> # create m random vectors and estimate correlation
R> simRho <- replicate(n = m, expr = {
+   tmpSim <- bs$rvec(N , R_X, nb_margins)
+   bs$cor(tmpSim, bs$Spearman)
+ }, simplify = FALSE)
R> # Evaluate the residual sum of squared error
R> sapply(simRho, function(R) sum((R - R_S)^2))
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> frobenius_loss <- sapply(simRho, function(R) sum((R - R_S)^2))
\end{CodeInput}
\end{CodeChunk}

The \texttt{R} summary function supplies the mean-augmented five-number summary of the quadratic loss distribution computed above.

\begin{CodeChunk}
\begin{CodeInput}
R> summary(frobenius_loss)
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
    691     829     890     925     986    1224 
\end{CodeOutput}
\end{CodeChunk}

This distribution could be compared to high-dimensional designed covariance estimators for guidance on whether the additional complexity and computation time are warranted.

\hypertarget{discussion}{%
\section{Conclusion and discussion}\label{discussion}}

We have introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a user-friendly, high-performance \texttt{R} package {[}\texttt{bigsimr}{]}. The random vector generation method is inspired by NORTA \citep{Cario1997} and Gaussian copula-based approaches \citep[\citet{BF17}, \citet{Xia17}]{MB13}. The major contributions of this work are methods and software for flexible, scalable simulation of HD multivariate probability distributions with broad potential data analytic applications for modern, big-data statistical computing. For example, one could simulate high-resolution time series data, such as those consistent with an auto-regressive moving average model exhibiting a specified Spearman structure. Or our methods could be used to simulate sparsely correlated data, as many HD methods assume, via specifying a \emph{spiked correlation matrix}.

It is customary to compare new tools and algorithms directly to existing competing methods and software. In this study, however, we only employ our proposed methodology, as our previous work has shown that existing \texttt{R} tools are simply not designed to meet our high-dimensional goal (see \citet{Li2019gpu} for evaluations of the \texttt{R} \texttt{copula} package and others). For the bivariate simulations, existing packages such as \href{https://github.com/superdesolator/NORTARA}{\texttt{nortaRA}} work well to match Pearson correlations exactly.

There are limitations to the methodology and implementation. We could only investigate selected multivariate distributions in our Monte Carlo studies. There may be instances that the methods do not perform well. Along those lines, we expect that correlation values close to the boundary of the feasible region could result in algorithm failure. Another issue is that for discrete distributions, we use continuous approximations when the support set is large. This could limit the scalability/accuracy for particular discrete/mixed multivariate distributions. Our method would also benefit computationally from faster Kendall estimation and more finely tuned nearest PD calculation for non-admissible Kendall matrices.

\hypertarget{misc}{%
\section*{Supplementary Materials}\label{misc}}
\addcontentsline{toc}{section}{Supplementary Materials}

We provide an open-source implementation of our methodology as the \texttt{Bigsimr.jl} Julia package, hosted on github at \url{https://github.com/adknudson/Bigsimr.jl}. Additionally we provide R and Python interfaces to \texttt{Bigsimr}, respectively at \url{https://github.com/SchisslerGroup/r-bigsimr} and \url{https://github.com/SchisslerGroup/python-bigsimr}. \texttt{Bigsimr.jl} is an ongoing project, and feature requests or issues can be submitted to \url{https://github.com/adknudson/Bigsimr.jl/issues}.

\begin{CodeChunk}
\begin{CodeInput}
R> knitr::include_graphics('images/hex-bigsimr.png')
\end{CodeInput}

\begin{center}\includegraphics[width=0.05\linewidth]{images/hex-bigsimr} \end{center}

\end{CodeChunk}

\hypertarget{acknowledgments}{%
\section*{Acknowledgment(s)}\label{acknowledgments}}
\addcontentsline{toc}{section}{Acknowledgment(s)}

The authors gratefully acknowledge Heather Knudson's graphic design for the \texttt{bigsimr} R Package. The results published here are in whole or part based upon data generated by the TCGA Research Network: \url{https://www.cancer.gov/tcga}.

\hypertarget{coi}{%
\section*{Disclosure statement}\label{coi}}
\addcontentsline{toc}{section}{Disclosure statement}

The authors report no conflict of interest.

\hypertarget{funding}{%
\section*{Funding}\label{funding}}
\addcontentsline{toc}{section}{Funding}

Research reported in this publication was supported by grants from the National Institutes of General Medical Sciences (5 U54 GM104944, GM103440) from the National Institutes of Health.

\hypertarget{appendix}{%
\section*{Appendix}\label{appendix}}
\addcontentsline{toc}{section}{Appendix}

\noindent Consider a bivariate example where we have a correlated \((Y_1, Y_2)^\top\) with \(Y_1\sim N(\mu_1, \sigma_1^2)\) (normal) and \(Y_2\sim LN(\mu_2, \sigma_2^2)\) (lognormal). First, let us note that when we get such a vector from the \texttt{bigsimr} package then in fact it can be represented as

\begin{equation}
(Y_1, Y_2)^\top \stackrel{d}{=} \left(X_1, e^{X_2}\right)^\top,
\label{eq:kram1}
\end{equation}

where \((X_1, X_2)^\top \sim N_2(\boldsymbol \mu, \boldsymbol \Sigma)\), which is bivariate normal with mean vector \(\boldsymbol \mu = (\mu_1, \mu_2)^\top\) and variance-covariance matrix

\begin{equation}
\boldsymbol \Sigma = 
\begin{bmatrix}
\sigma_1^2 & \rho \sigma_1\sigma_2\\
\rho \sigma_1\sigma_2 & \sigma_2^2
\end{bmatrix}
\label{eq:kram2}
\end{equation}

To see this, consider the three steps of the NORTA construction:

\begin{enumerate}

\item Generate $(Z_1, Z_2)^\top \sim N_2(\boldsymbol 0, \boldsymbol R)$, where

\begin{equation}
\boldsymbol R = 
\left[
\begin{array}{cc}
1 & \rho \\
\rho & 1
\end{array}
\right].
\label{eq:kram3}
\end{equation}

\item Transform $(Z_1, Z_2)^\top$ to $(U_1, U_2)^\top$ viz. $U_i =\Phi(Z_i)$,  $i=1,2$, where $\Phi(\cdot)$ is the standard normal CDF.

\item Return $(Y_1, Y_2)^\top$, where $Y_i=F_i^{-1}(U_i)$, $i=1,2$, and $F_i(\cdot)$ is the CDF of $Y_i$ and $F_i^{-1}(\cdot)$ is its inverse (the quantile function).

\end{enumerate}

In this example, \(F_1\) is the CDF of \(N(\mu_1, \sigma_1^2)\) while \(F_2\) is the CDF of \(LN(\mu_2, \sigma_2^2)\), so that the two required quantile functions are

\begin{equation}
F_1^{-1}(u) = \mu_1+\sigma_1 \Phi^{-1}(u)\,\,\, \mbox{and} \,\,\, F_2^{-1}(u) = e^{\mu_2+\sigma_2 \Phi^{-1}(u)}, 
\label{eq:kram4}
\end{equation}

as can be seen by standard calculation. Thus, when we apply Step 3 of the NORTA algorithm, we get

\begin{equation}
F_1^{-1}(U_1) = \mu_1+\sigma_1 \Phi^{-1}(\Phi(Z_1)) =  \mu_1+\sigma_1 Z_1 \,\,\, \mbox{and} \,\,\, F_2^{-1}(U_2) = e^{\mu_2+\sigma_2 \Phi^{-1}(\Phi(Z_2))} = e^{\mu_2+\sigma_2 Z_2}, 
\label{eq:kram5}
\end{equation}

where \((X_1, X_2)^\top = (\mu_1+\sigma_1 Z_1, \mu_2+\sigma_2 Z_2)^\top \sim N_2(\boldsymbol \mu, \boldsymbol \Sigma)\). Consequently, the vector \((Y_1, Y_2)^\top\) obtained in Step 3 of the algorithm has the structure provided by (\ref{kram1}).

\vspace{0.1in}

\noindent Next, we provide the exact covariance structure of the random vector \((Y_1, Y_2)^\top\) given by (\ref{kram1}), and relate the correlation of \(Y_1\) and \(Y_2\) to \(\rho\), which is the correlation of the normal variables \(Z_1\), \(Z_2\) (and also \(X_1\) and \(X_2\)). A straightforward albeit somewhat tedious algebra produces the following result.

\begin{lemma}
Let ${\bf Y} = (Y_1, Y_2)^\top$ admit the stochastic representation (\ref{kram1}), where ${\bf X} = (X_1, X_2)^\top \sim N_2(\boldsymbol \mu, \boldsymbol \Sigma)$ with $\boldsymbol \mu$ and $\boldsymbol \Sigma$ as above. Then the mean vector and the variance-covariance matrix of ${\bf Y}$ are given by

\begin{equation}
\boldsymbol \mu_{{\bf Y}} = \left(\mu_1, e^{\mu_2+\sigma_2^2/2}\right)^\top
\label{eq:kram6}
\end{equation}

and

\begin{equation}
\boldsymbol \Sigma = 
\begin{bmatrix}
\sigma_1^2 & \rho \sigma_1\sigma_2  e^{\mu_2+\sigma_2^2/2}  \\
\rho \sigma_1\sigma_2 e^{\mu_2+\sigma_2^2/2} & \left[ e^{\mu_2+\sigma_2^2/2}\right]^2 \left[e^{\sigma_2^2} -1 \right],
\end{bmatrix}
\label{eq:kram7}
\end{equation}

respectively. 
\end{lemma}

\begin{proof}
The values of the means and the variances are obtained immediately from normal marginal distribution of $Y_1$ and lognormal marginal distribution of $Y_2$. It remains to establish the covariance of $Y_1$ and $Y_2$,

\begin{equation}
\mbox{Cov}(Y_1, Y_2) = \mathbb E(Y_1 Y_2) - \mathbb E(Y_1)\mathbb E(Y_2), 
\label{eq:covy1y2}
\end{equation}

and in particular the first expectation on the right-hand-side above. By using the tower property of expectations, the latter expectation can be expressed as

\begin{equation}
\mathbb E(Y_1 Y_2)  =  \mathbb E\left(X_1 e^{X_2}\right)  =  \mathbb E \left\{ \mathbb E \left(X_1 e^{X_2} | X_1 \right) \right\} = \mathbb E \left\{ X_1 \mathbb E \left(e^{X_2} | X_1 \right) \right\}.
\label{eq:kre1}
\end{equation}

Further, by using the fact that the conditional distribution of $X_2$ given $X_1=x_1$ is normal with mean $\mathbb E(X_2|X_1=x_1) = \mu_2+\rho\sigma_2(x_1-\mu_1)/\sigma_1$ and variance $\sigma_2^2(1-\rho^2)$, one can relate the inner expectation on the far right in (\ref{kre1}) to the moment generating function $M(t)$ of this conditional distribution evaluated at $t=1$, leading to

\begin{equation}
E \left(e^{X_2} | X_1 \right) = e^{ \mu_2+\rho\sigma_2(x_1-\mu_1)/\sigma_1 + \sigma_2^2(1-\rho^2)/2},
\label{eq:kre2}
\end{equation}

so that

\begin{equation}
\mathbb E(Y_1 Y_2)  =  e^{ \mu_2+ \frac{1}{2} \sigma_2^2(1-\rho^2)} \mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\}.
\label{eq:kre3}
\end{equation}

Since $X_1$ is normal with mean $\mu_1$ and variance $\sigma_1^2$, the expectation in (\ref{kre3}) becomes

\begin{equation}
\mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = \int_{-\infty}^\infty x_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (x_1-\mu_1)} \frac{1}{\sqrt{2\pi}\sigma_1} e^{-\frac{1}{2\sigma_1^2}(x_1-\mu_1)^2}.
\label{eq:kre4}
\end{equation}

Upon substituting $u=x_1-\mu_1$, followed by some algebra, we arrive at

\begin{equation}
\mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = e^{\frac{1}{2}\sigma_2^2\rho^2} \int_{-\infty}^\infty (u+\mu_1)  \frac{1}{\sqrt{2\pi}\sigma_1} e^{-\frac{1}{2\sigma_1^2}(u-\rho\sigma_1\sigma_2)^2}.
\label{eq:kre5}
\end{equation}

Since the integral in (\ref{kre5}) is the expectation $\mathbb E(X+\mu_1)$, where $X$ is normal with mean $\rho\sigma_1\sigma_2$ and variance $\sigma_1^2$, we conclude that

\begin{equation}
\mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = e^{\frac{1}{2}\sigma_2^2\rho^2} (\rho\sigma_1\sigma_2 +\mu_1),
\label{eq:kre6}
\end{equation}

which, in view of (\ref{kre3}), leads to

\begin{equation}
\mathbb E (Y_1Y_2) = e^{\mu_2 + \frac{1}{2}\sigma_2^2} (\rho\sigma_1\sigma_2 +\mu_1).
\label{eq:kre7}
\end{equation}

Finally, (\ref{covy1y2}), along with the expressions for the means of $Y_1$ and $Y_2$, produce the covariance of $Y_1$ and $Y_2$:

\begin{equation}
\mbox{Cov}(Y_1, Y_2) = e^{\mu_2 + \frac{1}{2}\sigma_2^2} (\rho\sigma_1\sigma_2 +\mu_1) - \mu_1 e^{\mu_2 + \frac{1}{2}\sigma_2^2} = \rho\sigma_1\sigma_2 e^{\mu_2 + \frac{1}{2}\sigma_2^2}.
\label{eq:kre8}
\end{equation}

This concludes the proof
\end{proof}

Using the above result, we can directly relate the correlation coefficient of \(Y_1\) and \(Y_2\) with that of \(X_1\) and \(X_2\),

\begin{equation}
\rho_{Y_1, Y_2} = \rho \frac{\sigma_2}{\sqrt{e^{\sigma_2^2} -1}},
\label{eq:kram8}
\end{equation}

where \(\rho\) is the correlation of \(X_1\) and \(X_2\). The above result is useful when studying the possible range of correlation of \(Y_1\) and \(Y_2\) in the above example. It can be shown that the factor on the far right in (\ref{kram8}) is a monotonically decreasing function of \(\sigma_2\) on \((0,\infty)\), with the limits of 1 and 0 at zero and infinity, respectively. Thus, in principle, the range of correlation of \(Y_1\) and \(Y_2\) is the same as that of \(X_1\) and \(X_2\), as the factor on the far right in (\ref{kram8}) can be made arbitrarily close to 1. However, by changing this factor we may affect the marginal distributions of \(Y_1\) and \(Y_2\). It can be shown that if the marginal distributions of \(Y_1\) and \(Y_2\) are fixed, then the relation (\ref{kram8}) becomes

\begin{equation}
\rho_{Y_1, Y_2} = \rho \sqrt{\frac{\log(1+c_2^2)}{c_2^2}},
\label{eq:kram9}
\end{equation}

where \(c_2=\sigma_{Y_2}/\mu_{Y_2}\) is the \emph{coefficient of variation} (CV) of the variable \(Y_2\). Thus, the range of possible correlations in this model is not affected by the distribution of \(Y_1\), and is determined by the CV of \(Y_2\) as follows:

\begin{equation}
- \sqrt{\frac{\log(1+c_2^2)}{c_2^2}} \leq \rho_{Y_1, Y_2} \leq \sqrt{\frac{\log(1+c_2^2)}{c_2^2}}. 
\label{eq:kram10}
\end{equation}

Plugging in the estimates of these quantities from Section \ref{package}, we see that

\begin{equation}
\frac{\hat{\sigma}_2}{\sqrt{e^{\hat{\sigma}_2^2} -1}} = 0.881.
\label{eq:kram11}
\end{equation}

Thus, the possible range of correlation becomes \((-0.881, 0.881)\). This approximately agrees with the MC results provided.

As an invited speaker at CMStatistics 2020, I was delighted to learn of the opportunity to have our work included in the first issue of CSDA Annals of Statistical Data Science.

On behalf of my co-authors, I'm pleased to submit our manuscript entitled \emph{Simulating High-Dimensional Multivariate Data using the \texttt{Bigsimr} Package} for your consideration.
We feel that the work is well suited to the indicated themes and hope that it will be considered a valuable contribution to this prestigious journal.

The entire article can be reproduced from code and is housed at \url{https://github.com/SchisslerGroup/article_bigsimr}, which we are glad to make available upon request to the reviewers and, after acceptance, the general public.

I hope that everything is in order. If not, or if you have any additional needs, please do not hesitate to contact me. Thank you and all those involved in this process.

\bibliography{bigsimr.bib,packages.bib,alex.bib}

\end{document}

### 2021-03-27 _book/_main.tex
idk

--- 2021-03-27 _article/article_bigsimr.tex
@@
 \begin{document}
-
-\newpage
 
@@
 
-{\centering \includegraphics{article_bigsimr_files/figure-latex/ch010-realDataFig-1}
+{\centering \includegraphics{_main_files/figure-latex/ch010-realDataFig-1}
 
@@
 
-{\centering \includegraphics{article_bigsimr_files/figure-latex/ch030-aq-joint-dist-1}
+{\centering \includegraphics{_main_files/figure-latex/ch030-aq-joint-dist-1}
 
@@
         [,1]    [,2]
-[1,]  1.0000 -0.8902
-[2,] -0.8902  1.0000
+[1,]  1.0000 -0.8821
+[2,] -0.8821  1.0000
 \end{CodeOutput}
@@
        [,1]   [,2]
-[1,] 1.0000 0.8876
-[2,] 0.8876 1.0000
+[1,] 1.0000 0.8812
+[2,] 0.8812 1.0000
 \end{CodeOutput}
@@
 Julia Object of type NamedTuple{(:lower, :upper),Tuple{Float64,Float64}}.
-(lower = -0.8822730883942045, upper = 0.8833146839651883)
+(lower = -0.8814915390483397, upper = 0.8817369743444273)
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-Warning: Removed 342 rows containing non-finite values (stat_density2d_filled).
+Warning: Removed 392 rows containing non-finite values (stat_density2d_filled).
 \end{CodeOutput}
@@
 
-{\centering \includegraphics{article_bigsimr_files/figure-latex/ch030-plot-sim-1}
+{\centering \includegraphics{_main_files/figure-latex/ch030-plot-sim-1}
 
@@
 
-{\centering \includegraphics{article_bigsimr_files/figure-latex/ch040-bPlot-1}
+{\centering \includegraphics{_main_files/figure-latex/ch040-bPlot-1}
 
@@
 
-{\centering \includegraphics{article_bigsimr_files/figure-latex/ch040-largeDfig-1}
+{\centering \includegraphics{_main_files/figure-latex/ch040-largeDfig-1}
 
@@
 
-{\centering \includegraphics{article_bigsimr_files/figure-latex/ch050-simDataFig-1}
+{\centering \includegraphics{_main_files/figure-latex/ch050-simDataFig-1}
 

### 2021-03-27 index.Rmd
idk

--- 2021-03-01 index.Rmd
@@
 ---
-title: "Simulating High-Dimensional Multivariate Data"
-subtitle: "using the `Bigsimr` Package"
+documentclass: jss
+site: bookdown::bookdown_site
+output:
+  bookdown::pdf_book:
+    base_format: rticles::jss_article
+    keep_tex: true
+    includes:
+      in_header: "preamble.tex"
 author:
-  - Alfred G. Schissler*
-  - Edward J. Bedrick
-  - Alexander D. Knudson
-  - Tomasz J. Kozubowski
-  - Tin Nguyen
-  - Anna K. Panorska
-  - Juli Petereit
-  - Walter W. Piegorsch
-  - Duc Tran
-site: bookdown::bookdown_site
-bibliography: ["bigsimr.bib", "packages.bib", "alex.bib"]
-abstract:  It is critical to realistically simulate data when employing Monte Carlo techniques and evaluating statistical methodology.
+  - name: Alfred G. Schissler
+    affiliation: University of Nevada
+    address: |
+      | 1664 N Virginia St.
+      | Reno, NV 89557
+    email: \email{aschissler@unr.edu}
+  - name: Edward J. Bedrick
+    affiliation: University of Nevada
+  - name: Alexander D. Knudson
+    affiliation: University of Nevada \AND
+  - name: Tomasz J. Kozubowski
+    affiliation: University of Nevada
+  - name: Tin Nguyen
+    affiliation: University of Nevada
+  - name: Anna K. Panorska
+    affiliation: University of Nevada \AND
+  - name: Juli Petereit
+    affiliation: University of Nevada
+  - name: Walter W. Piegorsch
+    affiliation: University of Nevada
+  - name: Duc Tran
+    affiliation: University of Nevada
+title:
+  formatted: "Simulating High-Dimensional Multivariate Data using the \\pkg{Bigsimr} Package"
+  plain:     "Simulating High-Dimensional Multivariate Data using the Bigsimr Package"
+  short:     "\\pkg{Bigsimr}: Simulate High-Dimensional Data"
+abstract: >
+  It is critical to realistically simulate data when employing Monte Carlo techniques and evaluating statistical methodology.
 Measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments.
 Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations.
-This article introduces the `bigsimr` R package.
+This article introduces the \\pkg{Bigsimr} R package.
 This high-level package provides a flexible and scalable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions and known Pearson, Spearman, or Kendall correlation matrix.
-`bigsimr` contains additional high-performance features, including multi-core algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix.
+\\pkg{Bigsimr} contains additional high-performance features, including multi-core algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix.
 Monte Carlo studies quantify the accuracy and scalability of our approach, up to $d=10,000$.
-Finally, we demonstrate example applications enabled via `bigsimr` by applying these methods to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples.
+Finally, we demonstrate example applications enabled via \\pkg{Bigsimr} by applying these methods to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples.
+keywords:
+  formatted: [multivariate simulation, high-dimensional data, nonparametric correlation, Gaussian copula, RNA-sequencing data, breast cancer]
+  plain:     [multivariate simulation, high-dimensional data, nonparametric correlation, Gaussian copula, RNA-sequencing data, breast cancer]
+bibliography: [bigsimr.bib, packages.bib, alex.bib]
+biblio-style: jss
+link-citations: yes
 ---
-
-*Keywords:*  Multivariate simulation, High-dimensional data, Nonparametric correlation, Gaussian copula, RNA-sequencing data, breast cancer
-
-*=Corresponding author. aschissler@unr.edu. 1664 N Virginia St, Reno, NV 89557
 
@@
 ), 'packages.bib')
+
+options(prompt = 'R> ', 
+        continue = '+ ',
+        digits = 4)
 ```
 
-# Introduction
-
-Massive high-dimensional (HD) data sets are now commonplace in many areas of scientific inquiry.
-As new methods are developed for data, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of analyzing such proposed methodology --- such as false positive rates, statistical power, interval coverage, and robustness --- often with comparison to existing methods.
-Further, efficient simulation empowers statistical computing strategies, such as the parametric bootstrap [@Chernick2008] to simulate from a hypothesized null model, providing inference in analytically challenging settings.
-Such Monte Carlo (MC) techniques become difficult for high-dimensional data using existing algorithms and tools.
-This is particularly true when simulating massive multivariate, non-normal distributions, arising naturally in many fields of study.
-
-As many have noted, it can be vexing to simulate dependent, non-normal/discrete data, even for low dimensional settings [@MB13; @XZ19].
-For continuous non-normal multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula approaches [@Nelsen2007] are well-studied, with flexible, robust software available [@Yan2007; @Chen2001].
-Yet these approaches do not scale in a timely fashion to high-dimensional problems [@Li2019gpu].
-For discrete data, early simulation strategies had major flaws, such as failing to obtain the full range of possible correlations (e.g., admitting only positive correlations: see @Park1996).
-While more recent approaches [@MB13; @Xia17; @BF17] have largely remedied this issue for low-dimensional problems, the existing tools are not designed to scale to high dimensions.
-
-Another central issue lies in characterizing dependence between components in the high-dimensional random vector.
-The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc).
-For normal data, the Pearson product-moment correlation describes the dependency perfectly.
-As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix is computationally intense [@Chen2001; @Xia17].
-On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependence, such as Spearman's $\rho$ and Kendall's $\tau$.
-Throughout, we focus on matching these nonparametric dependence measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology.
-We do, however, provide Pearson matching with some caveats.
-
-With all this in mind, we present a scalable, flexible multivariate simulation algorithm.
-The crux of the method lies in the construction of a Gaussian copula in the spirit of the NORTA procedure.
-Further, we introduce the `bigsimr` R package that provides high-performance software implementing our algorithm.
-The algorithm design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed-form relationships between dependence measures for the multivariate normal (MVN) distribution.
-
-Our study proceeds by providing background information, including a description of a motivating example application: RNA-sequencing (RNA-seq) breast cancer data.
-Then we describe and justify our simulation methodology and related algorithms.
-Next, we detail an illustrative low-dimensional example of basic use of the `bigsimr` R package.
-Then we proceed with Monte Carlo studies under various bivariate distributional assumptions to assess accuracy.
-We conclude the Monte Carlo evaluations by summarizing the computation time for increasingly higher dimensional vectors.
-After the MC evaluations, we simulate random vectors motivated by our RNA-seq example, evaluate the accuracy, and provide example statistical computing tasks, namely MC estimation of joint probabilities and evaluating HD correlation estimation efficiency.
-Finally, we discuss the method's utility, limitations, and future directions.
+\newpage

### 2021-03-27 _common.R

@@ -24,5 +24,5 @@ knitr::opts_chunk$set(
 
 # options(dplyr.print_min = 6, dplyr.print_max = 6)
 
-seed_number <- 2020-02-25
+seed_number <- 2021-03-23
 set.seed(seed = seed_number)

### 2021-03-27 data-raw/benchmark_dependences.R

@@ -52,14 +52,14 @@ s <- bs$cor_randPSD(100)
 p <- bs$cor_convert(s, bs$Spearman, bs$Pearson)
 r <- bs$cor_nearPD(p)
 bs$cor_nearPD(p, 1e-6)
-bs$cor_nearPD(p, 1e-6, 1e-6)
+bs$cor_nearPD(p, 1e-6, tol=1e-6)
 bs$cor_fastPD(p)
 
 k <- bs$cor_randPSD(10000)
 p <- bs$cor_convert(s, bs$Kendall, bs$Pearson)
 julia_call("Bigsimr.iscorrelation", p)
-system.time(r1 <- bs$cor_nearPD(p, 1e-6, 1e-2))
-system.time(r2 <- bs$cor_nearPD(p, 1e-6, 1e-6))
+system.time(r1 <- bs$cor_nearPD(p, 1e-6, tol=1e-2))
+system.time(r2 <- bs$cor_nearPD(p, 1e-6, tol=1e-6))
 system.time(r3 <- bs$cor_fastPD(p))
 julia_call("Bigsimr.iscorrelation", r1)
 julia_call("Bigsimr.iscorrelation", r2)
@@ -143,7 +143,7 @@ bench <- function(type, d, n=1000) {
     needed_near_pd <- FALSE
     if (!julia_call("Bigsimr.iscorrelation", adj_corr)) {
       needed_near_pd <- TRUE
-      adj_corr_pd <- bs$cor_nearPD(adj_corr, 1e-6, 0.005)
+      adj_corr_pd <- bs$cor_nearPD(adj_corr, 1e-6, tol=0.005)
     }
   })
 

### 2021-03-27 notebooks/cor_boundsBug.R

@@ -1,12 +1,0 @@
-library(bigsimr)
-size <- 4
-prob <- 3e-04
-type <- "spearman"
-n <- 1e3
-
-margins <- alist(
-        qnbinom(size = size, prob = prob),
-        qnbinom(size = size, prob = prob)
-)
-
-tmp_bounds <- cor_bounds(margins, method = type)

### 2021-03-28 article_bigsimr_new.Rmd
template using rticles

--- 2021-03-27 index.Rmd
@@
 documentclass: jss
-site: bookdown::bookdown_site
-output:
-  bookdown::pdf_book:
-    base_format: rticles::jss_article
-    keep_tex: true
-    includes:
-      in_header: "preamble.tex"
 author:
@@
 biblio-style: jss
-link-citations: yes
+preamble: >
+  \usepackage{amsmath}
+output: rticles::jss_article
 ---
 
-```{r include=FALSE}
-# automatically create a bib database for R packages
+```{r, setup, include=FALSE}
 knitr::write_bib(c(
@@
         digits = 4)
+
+get_rmd <- function(str_no = "") list.files(path = ".", pattern = str_no)
+
+knitr::opts_chunk$set(
+  # Text Results
+  eval = TRUE,
+  echo = FALSE,
+  results = 'markup',
+  collapse = TRUE,
+  message = FALSE,
+  warning = FALSE,
+  error = FALSE,
+  # Code Decoration
+  comment = NA,
+  background = '#F7F7F7',
+  # Cache
+  cache = TRUE,
+  # Plots
+  fig.show = "hold",
+  fig.width = 6.5,
+  fig.asp = 0.618,  # 1 / phi
+  out.width = "85%",
+  fig.align = 'center'
+)
+
+# options(dplyr.print_min = 6, dplyr.print_max = 6)
+
+seed_number <- 2021-03-23
+set.seed(seed = seed_number)
 ```
@@
 \newpage
+
+```{r, child=get_rmd("000")}
+```

### 2021-03-28 _common.R

@@ -1,28 +1,0 @@
-options(digits = 4)
-
-knitr::opts_chunk$set(
-  # Text Results
-  eval = TRUE,
-  echo = FALSE,
-  results = 'markup',
-  collapse = TRUE,
-  message = FALSE,
-  warning = FALSE,
-  error = FALSE,
-  # Code Decoration
-  comment = NA,
-  background = '#F7F7F7',
-  # Cache
-  cache = TRUE,
-  # Plots
-  fig.show = "hold",
-  fig.width = 6.5,
-  fig.asp = 0.618,  # 1 / phi
-  out.width = "85%",
-  fig.align = 'center'
-)
-
-# options(dplyr.print_min = 6, dplyr.print_max = 6)
-
-seed_number <- 2021-03-23
-set.seed(seed = seed_number)

### 2021-03-28 _outgoing.R

@@ -1,4 +1,0 @@
-# Remove all objects
-rm(list = ls(all.names = TRUE))
-
-# Unload all packages

### 2021-04-01 080-appendix.Rmd
final spell check
idk

@@
 \noindent Consider a bivariate example where we have a correlated $(Y_1, Y_2)^\top$ with $Y_1\sim N(\mu_1, \sigma_1^2)$ (normal) and $Y_2\sim LN(\mu_2, \sigma_2^2)$ (lognormal).
-First, let us note that when we get such a vector from bigsimr package then in fact it can be represented as
+First, let us note that when we get such a vector from the `bigsimr` package then in fact it can be represented as
 
 \begin{equation}
-\label{kram1}
+\label{eq:kram1}
 (Y_1, Y_2)^\top \stackrel{d}{=} \left(X_1, e^{X_2}\right)^\top,
@@
 \begin{equation}
-\label{kram2}
+\label{eq:kram2}
 \boldsymbol \Sigma = 
-\left[
-\begin{array}{cc}
+\begin{bmatrix}
 \sigma_1^2 & \rho \sigma_1\sigma_2\\
 \rho \sigma_1\sigma_2 & \sigma_2^2
-\end{array}
-\right].
+\end{bmatrix}
 \end{equation}
@@
 \begin{equation}
-\label{kram3}
+\label{eq:kram3}
 \boldsymbol R = 
@@
 \begin{equation}
-\label{kram4}
+\label{eq:kram4}
 F_1^{-1}(u) = \mu_1+\sigma_1 \Phi^{-1}(u)\,\,\, \mbox{and} \,\,\, F_2^{-1}(u) = e^{\mu_2+\sigma_2 \Phi^{-1}(u)}, 
@@
 \begin{equation}
-\label{kram5}
+\label{eq:kram5}
 F_1^{-1}(U_1) = \mu_1+\sigma_1 \Phi^{-1}(\Phi(Z_1)) =  \mu_1+\sigma_1 Z_1 \,\,\, \mbox{and} \,\,\, F_2^{-1}(U_2) = e^{\mu_2+\sigma_2 \Phi^{-1}(\Phi(Z_2))} = e^{\mu_2+\sigma_2 Z_2}, 
@@
 where $(X_1, X_2)^\top = (\mu_1+\sigma_1 Z_1, \mu_2+\sigma_2 Z_2)^\top \sim N_2(\boldsymbol \mu, \boldsymbol \Sigma)$.
-Consequently, the vector $(Y_1, Y_2)^\top$ obtained in Step 3 of the algorithm has the structure provided by (\ref{kram1}).
+Consequently, the vector $(Y_1, Y_2)^\top$ obtained in Step 3 of the algorithm has the structure provided by (\ref{eq:kram1}).
 
@@
 
-\noindent Next, we provide the exact covariance structure of the random vector $(Y_1, Y_2)^\top$ given by (\ref{kram1}), and relate the correlation of $Y_1$ and $Y_2$ to $\rho$, which is the correlation of the normal variables $Z_1$, $Z_2$ (and also $X_1$ and $X_2$).
+\noindent Next, we provide the exact covariance structure of the random vector $(Y_1, Y_2)^\top$ given by (\ref{eq:kram1}), and relate the correlation of $Y_1$ and $Y_2$ to $\rho$, which is the correlation of the normal variables $Z_1$, $Z_2$ (and also $X_1$ and $X_2$).
 A straightforward albeit somewhat tedious algebra produces the following result.
@@
 \begin{lemma}
-Let ${\bf Y} = (Y_1, Y_2)^\top$ admit the stochastic representation (\ref{kram1}), where ${\bf X} = (X_1, X_2)^\top \sim N_2(\boldsymbol \mu, \boldsymbol \Sigma)$ with $\boldsymbol \mu$ and $\boldsymbol \Sigma$ as above.
+Let ${\bf Y} = (Y_1, Y_2)^\top$ admit the stochastic representation (\ref{eq:kram1}), where ${\bf X} = (X_1, X_2)^\top \sim N_2(\boldsymbol \mu, \boldsymbol \Sigma)$ with $\boldsymbol \mu$ and $\boldsymbol \Sigma$ as above.
 Then the mean vector and the variance-covariance matrix of ${\bf Y}$ are given by
@@
 \begin{equation}
-\label{kram6}
+\label{eq:kram6}
 \boldsymbol \mu_{{\bf Y}} = \left(\mu_1, e^{\mu_2+\sigma_2^2/2}\right)^\top
@@
 \begin{equation}
-\label{kram7}
+\label{eq:kram7}
 \boldsymbol \Sigma = 
-\left[
-\begin{array}{cc}
+\begin{bmatrix}
 \sigma_1^2 & \rho \sigma_1\sigma_2  e^{\mu_2+\sigma_2^2/2}  \\
 \rho \sigma_1\sigma_2 e^{\mu_2+\sigma_2^2/2} & \left[ e^{\mu_2+\sigma_2^2/2}\right]^2 \left[e^{\sigma_2^2} -1 \right],
-\end{array}
-\right].
+\end{bmatrix}
 \end{equation}
@@
 \begin{equation}
-\label{covy1y2}
+\label{eq:covy1y2}
 \mbox{Cov}(Y_1, Y_2) = \mathbb E(Y_1 Y_2) - \mathbb E(Y_1)\mathbb E(Y_2), 
@@
 \begin{equation}
-\label{kre1}
+\label{eq:kre1}
 \mathbb E(Y_1 Y_2)  =  \mathbb E\left(X_1 e^{X_2}\right)  =  \mathbb E \left\{ \mathbb E \left(X_1 e^{X_2} | X_1 \right) \right\} = \mathbb E \left\{ X_1 \mathbb E \left(e^{X_2} | X_1 \right) \right\}.
@@
 
-Further, by using the fact that the conditional distribution of $X_2$ given $X_1=x_1$ is normal with mean $\mathbb E(X_2|X_1=x_1) = \mu_2+\rho\sigma_2(x_1-\mu_1)/\sigma_1$ and variance $\sigma_2^2(1-\rho^2)$, one can relate the inner expectation on the far right in (\ref{kre1}) to the moment generating function $M(t)$ of this conditional distribution evaluated at $t=1$, leading to
+Further, by using the fact that the conditional distribution of $X_2$ given $X_1=x_1$ is normal with mean $\mathbb E(X_2|X_1=x_1) = \mu_2+\rho\sigma_2(x_1-\mu_1)/\sigma_1$ and variance $\sigma_2^2(1-\rho^2)$, one can relate the inner expectation on the far right in (\ref{eq:kre1}) to the moment generating function $M(t)$ of this conditional distribution evaluated at $t=1$, leading to
 
 \begin{equation}
-\label{kre2}
+\label{eq:kre2}
 E \left(e^{X_2} | X_1 \right) = e^{ \mu_2+\rho\sigma_2(x_1-\mu_1)/\sigma_1 + \sigma_2^2(1-\rho^2)/2},
@@
 \begin{equation}
-\label{kre3}
+\label{eq:kre3}
 \mathbb E(Y_1 Y_2)  =  e^{ \mu_2+ \frac{1}{2} \sigma_2^2(1-\rho^2)} \mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\}.
@@
 
-Since $X_1$ is normal with mean $\mu_1$ and variance $\sigma_1^2$, the expectation in (\ref{kre3}) becomes
+Since $X_1$ is normal with mean $\mu_1$ and variance $\sigma_1^2$, the expectation in (\ref{eq:kre3}) becomes
 
 \begin{equation}
-\label{kre4}
+\label{eq:kre4}
 \mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = \int_{-\infty}^\infty x_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (x_1-\mu_1)} \frac{1}{\sqrt{2\pi}\sigma_1} e^{-\frac{1}{2\sigma_1^2}(x_1-\mu_1)^2}.
@@
 \begin{equation}
-\label{kre5}
+\label{eq:kre5}
 \mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = e^{\frac{1}{2}\sigma_2^2\rho^2} \int_{-\infty}^\infty (u+\mu_1)  \frac{1}{\sqrt{2\pi}\sigma_1} e^{-\frac{1}{2\sigma_1^2}(u-\rho\sigma_1\sigma_2)^2}.
@@
 
-Since the integral in (\ref{kre5}) is the expectation $\mathbb E(X+\mu_1)$, where $X$ is normal with mean $\rho\sigma_1\sigma_2$ and variance $\sigma_1^2$, we conclude that
+Since the integral in (\ref{eq:kre5}) is the expectation $\mathbb E(X+\mu_1)$, where $X$ is normal with mean $\rho\sigma_1\sigma_2$ and variance $\sigma_1^2$, we conclude that
 
 \begin{equation}
-\label{kre6}
+\label{eq:kre6}
 \mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = e^{\frac{1}{2}\sigma_2^2\rho^2} (\rho\sigma_1\sigma_2 +\mu_1),
@@
 
-which, in view of (\ref{kre3}), leads to
+which, in view of (\ref{eq:kre3}), leads to
 
 \begin{equation}
-\label{kre7}
+\label{eq:kre7}
 \mathbb E (Y_1Y_2) = e^{\mu_2 + \frac{1}{2}\sigma_2^2} (\rho\sigma_1\sigma_2 +\mu_1).
@@
 
-Finally, (\ref{covy1y2}), along with the expressions for the means of $Y_1$ and $Y_2$, produce the covariance of $Y_1$ and $Y_2$:
+Finally, (\ref{eq:covy1y2}), along with the expressions for the means of $Y_1$ and $Y_2$, produce the covariance of $Y_1$ and $Y_2$:
 
 \begin{equation}
-\label{kre8}
+\label{eq:kre8}
 \mbox{Cov}(Y_1, Y_2) = e^{\mu_2 + \frac{1}{2}\sigma_2^2} (\rho\sigma_1\sigma_2 +\mu_1) - \mu_1 e^{\mu_2 + \frac{1}{2}\sigma_2^2} = \rho\sigma_1\sigma_2 e^{\mu_2 + \frac{1}{2}\sigma_2^2}.
@@
 \begin{equation}
-\label{kram8}
+\label{eq:kram8}
 \rho_{Y_1, Y_2} = \rho \frac{\sigma_2}{\sqrt{e^{\sigma_2^2} -1}},
@@
 The above result is useful when studying the possible range of correlation of $Y_1$ and $Y_2$ in the above example.
-It can be shown that the factor on the far right in (\ref{kram8}) is a monotonically decreasing function of $\sigma_2$ on $(0,\infty)$, with the limits of 1 and 0 at zero and infinity, respectively.
-Thus, in principle, the range of correlation of $Y_1$ and $Y_2$ is the same as that of $X_1$ and $X_2$, as the factor on the far right in (\ref{kram8}) can be made arbitrarily close to 1.
+It can be shown that the factor on the far right in (\ref{eq:kram8}) is a monotonically decreasing function of $\sigma_2$ on $(0,\infty)$, with the limits of 1 and 0 at zero and infinity, respectively.
+Thus, in principle, the range of correlation of $Y_1$ and $Y_2$ is the same as that of $X_1$ and $X_2$, as the factor on the far right in (\ref{eq:kram8}) can be made arbitrarily close to 1.
 However, by changing this factor we may affect the marginal distributions of $Y_1$ and $Y_2$.
-It can be shown that if the marginal distributions of $Y_1$ and $Y_2$ are fixed, then the relation (\ref{kram8}) becomes
+It can be shown that if the marginal distributions of $Y_1$ and $Y_2$ are fixed, then the relation (\ref{eq:kram8}) becomes
 
 \begin{equation}
-\label{kram9}
+\label{eq:kram9}
 \rho_{Y_1, Y_2} = \rho \sqrt{\frac{\log(1+c_2^2)}{c_2^2}},
@@
 \begin{equation}
-\label{kram10}
+\label{eq:kram10}
 - \sqrt{\frac{\log(1+c_2^2)}{c_2^2}} \leq \rho_{Y_1, Y_2} \leq \sqrt{\frac{\log(1+c_2^2)}{c_2^2}}. 
@@
 \begin{equation}
-\label{kram11}
+\label{eq:kram11}
 \frac{\hat{\sigma}_2}{\sqrt{e^{\hat{\sigma}_2^2} -1}} = 0.881.

### 2021-04-01 article_bigsimr.Rmd

@@
 ---
-title: "Simulating High-Dimensional Multivariate Data"
-subtitle: "using the `Bigsimr` Package"
+documentclass: jss
 author:
-  - Alfred G. Schissler*
-  - Edward J. Bedrick
-  - Alexander D. Knudson
-  - Tomasz J. Kozubowski
-  - Tin Nguyen
-  - Anna K. Panorska
-  - Juli Petereit
-  - Walter W. Piegorsch
-  - Duc Tran
-site: bookdown::bookdown_site
-bibliography: ["bigsimr.bib", "packages.bib", "alex.bib"]
-abstract:  It is critical to realistically simulate data when employing Monte Carlo techinques and evaluating statistical methodology.
+  - name: Alfred G. Schissler
+    affiliation: University of Nevada
+    address: |
+      | 1664 N Virginia St.
+      | Reno, NV 89557
+    email: \email{aschissler@unr.edu}
+  - name: Edward J. Bedrick
+    affiliation: University of Nevada
+  - name: Alexander D. Knudson
+    affiliation: University of Nevada \AND
+  - name: Tomasz J. Kozubowski
+    affiliation: University of Nevada
+  - name: Tin Nguyen
+    affiliation: University of Nevada
+  - name: Anna K. Panorska
+    affiliation: University of Nevada \AND
+  - name: Juli Petereit
+    affiliation: University of Nevada
+  - name: Walter W. Piegorsch
+    affiliation: University of Nevada
+  - name: Duc Tran
+    affiliation: University of Nevada
+title:
+  formatted: "Simulating High-Dimensional Multivariate Data using the \\pkg{Bigsimr} Package"
+  plain:     "Simulating High-Dimensional Multivariate Data using the Bigsimr Package"
+  short:     "\\pkg{Bigsimr}: Simulate High-Dimensional Data"
+abstract: >
+  It is critical to realistically simulate data when employing Monte Carlo techniques and evaluating statistical methodology.
 Measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments.
 Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations.
-This article introduces the `bigsimr` R package.
+This article introduces the \\pkg{Bigsimr} R package.
 This high-level package provides a flexible and scalable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions and known Pearson, Spearman, or Kendall correlation matrix.
-`bigsimr` contains additional high-performance features, including multi-core algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix.
-Monte carlo studies quantify the accuracy and scalability of our approach, up to $d=10,000$.
-Finally, we demonstrate example applications enabled via `bigsimr` by applying these methods to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples.
+\\pkg{Bigsimr} contains additional high-performance features, including multi-core algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix.
+Monte Carlo studies quantify the accuracy and scalability of our approach, up to $d=10,000$.
+Finally, we demonstrate example applications enabled via \\pkg{Bigsimr} by applying these methods to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples.
+keywords:
+  formatted: [multivariate simulation, high-dimensional data, nonparametric correlation, Gaussian copula, RNA-sequencing data, breast cancer]
+  plain:     [multivariate simulation, high-dimensional data, nonparametric correlation, Gaussian copula, RNA-sequencing data, breast cancer]
+bibliography: [bigsimr.bib, packages.bib, alex.bib]
+biblio-style: jss
+preamble: >
+  \usepackage{amsmath}
+  \usepackage{amssymb}
+  \usepackage{amsfonts}
+  \usepackage{amsthm}
+  \newtheorem{theorem}{Theorem}
+  \newtheorem{lemma}[theorem]{Lemma}
+  \usepackage{booktabs}
+  \usepackage{longtable}
+  \usepackage{array}
+  \usepackage{ragged2e}
+  \usepackage{setspace}
+output: rticles::jss_article
 ---
-```{r include=FALSE, cache=FALSE}
-options(digits = 4)
+
+```{r, setup, include=FALSE}
+knitr::write_bib(c(
+  .packages(), 'bookdown', 'knitr', 'rmarkdown'
+), 'packages.bib')
+
+options(prompt = 'R> ', 
+        continue = '+ ',
+        digits = 4)
+
+get_rmd <- function(str_no = "") list.files(path = ".", pattern = str_no)
 
@@
 
-seed_number <- 2020-02-25
+seed_number <- 2021-03-23
 set.seed(seed = seed_number)
@@
 
-*Keywords:*  Multivariate simulation, High-dimensional data, Nonparametric correlation, Gaussian copula, RNA-sequencing data, breast cancer
+<!-- New page after title and abstract -->
+\newpage
 
-*=Corresponding author. aschissler@unr.edu. 1664 N Virginia St, Reno, NV 89557
+<!-- Set double spaces -->
+\setstretch{2.0}
 
-```{r include=FALSE}
-# automatically create a bib database for R packages
-knitr::write_bib(c(
-  .packages(), 'bookdown', 'knitr', 'rmarkdown'
-), 'packages.bib')
+<!-- Main content -->
+
+```{r, child=get_rmd("000")}
 ```
 
-# Introduction
-
-Massive high-dimensional (HD) data sets are now commonplace in many areas of scientific inquiry.
-As new methods are developed for data, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of analyzing such proposed methodology --- such as false positive rates, statistical power, interval coverage, and robustness --- often with comparison to existing methods.
-Further, efficient simulation empowers statistical computing strategies, such as the parametric bootstrap [@Chernick2008] to simulate from a hypothesized null model, providing inference in analytically challenging settings.
-Such Monte Carlo (MC) techniques become difficult for high-dimensional data using existing algorithms and tools.
-This is particularly true when simulating massive multivariate, non-normal distributions, arising naturally in many fields of study.
-
-As many have noted, it can be vexing to simulate dependent, non-normal/discrete data, even for low dimensional settings [@MB13; @XZ19].
-For continuous non-normal multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula approaches [@Nelsen2007] are well-studied, with flexible, robust software available [@Yan2007; @Chen2001].
-Yet these approaches do not scale in a timely fashion to high-dimensional problems [@Li2019gpu].
-For discrete data, early simulation strategies had major flaws, such as failing to obtain the full range of possible correlations (e.g., admitting only positive correlations: see @Park1996).
-While more recent approaches [@MB13; @Xia17; @BF17] have largely remedied this issue for low-dimensional problems, the existing tools are not designed to scale to high dimensions.
-
-Another central issue lies in characterizing dependence between components in the high-dimensional random vector.
-The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc).
-For normal data, the Pearson product-moment correlation describes the dependency perfectly.
-As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix is computationally intense [@Chen2001; @Xia17].
-On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependence, such as Spearman's $\rho$ and Kendall's $\tau$.
-Throughout, we focus on matching these nonparametric dependence measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology.
-We do, however, provide Pearson matching with some caveats.
-
-With all this in mind, we present a scalable, flexible multivariate simulation algorithm.
-The crux of the method lies in the construction of a Gaussian copula in the spirit of the NORTA procedure.
-Further, we introduce the `bigsimr` R package that provides high-performance software implementing our algorithm.
-The algorithm design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed-form relationships between dependence measures for the multivariate normal (MVN) distribution.
-
-Our study proceeds by providing background information, including a description of a motivating example application: RNA-sequencing (RNA-seq) breast cancer data.
-Then we describe and justify our simulation methodology and related algorithms.
-Next, we detail an illustrative low-dimensional example of basic use of the `bigsimr` R package.
-Then we proceed with Monte Carlo studies under various bivariate distributional assumptions to assess accuracy.
-We conclude the Monte Carlo evaluations by summarizing the computation time for increasingly higher dimensional vectors.
-After the MC evaluations, we simulate random vectors motivated by our RNA-seq example, evaluate the accuracy, and provide example statistical computing tasks, namely MC estimation of joint probabilities and evaluating HD correlation estimation efficiency.
-Finally, we discuss the method's utility, limitations, and future directions.
-```{r include=FALSE, cache=FALSE}
-# Remove all objects
-rm(list = ls(all.names = TRUE))
-
-# Unload all packages
+```{r, child=get_rmd("010")}
 ```
 
-<!--chapter:end:index.Rmd-->
-
-```{r include=FALSE, cache=FALSE}
-options(digits = 4)
-
-knitr::opts_chunk$set(
-  # Text Results
-  eval = TRUE,
-  echo = FALSE,
-  results = 'markup',
-  collapse = TRUE,
-  message = FALSE,
-  warning = FALSE,
-  error = FALSE,
-  # Code Decoration
-  comment = NA,
-  background = '#F7F7F7',
-  # Cache
-  cache = TRUE,
-  # Plots
-  fig.show = "hold",
-  fig.width = 6.5,
-  fig.asp = 0.618,  # 1 / phi
-  out.width = "85%",
-  fig.align = 'center'
-)
-
-# options(dplyr.print_min = 6, dplyr.print_max = 6)
-
-seed_number <- 2020-02-25
-set.seed(seed = seed_number)
-```
-# Background {#background}
-
-```{r ch010-background, include=FALSE, cache=FALSE}
-box::use(
-  dplyr[...],
-  ggplot2[...],
-  stringr[str_sub],
-  GGally[ggpairs, wrap],
-  knitr[kable]
-)
-load("data/example_genes.rda")
-load("data/example_brca.rda")
-d <- 1000
+```{r, child=get_rmd("020")}
 ```
 
-The `bigsimr` `R` package presented here provides multiple algorithms that operate with high-dimensional multivariate data; however, all these algorithms were originally designed to support a single task: to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics.
-Specifically, our goal is to efficiently simulate a large number, $B$, of HD random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with *correlated* components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i$.
-
-When designing this methodology, we developed the following properties to guide our effort.
-We divide the properties into two categories: (1) basic properties (BP) and "scalability" properties (SP).
-The BPs are adapted from an existing criteria due to @Nik13a.
-A suitable simulation strategy should possess the following properties:
-
-* BP1: A wide range of dependences, allowing both positive and negative values, and, ideally, admitting the full range of possible values.
-* BP2: Flexible dependence, meaning that the number of bivariate marginals can be equal to the number of dependence parameters.
-* BP3: Flexible marginal modeling, generating heterogeneous data --- possibly from differing probability families.
-
-Moreover, the simulation method must *scale* to high dimensions:
-
-* SP1: Procedure must scale to high dimensions with practical compute times.
-* SP2: Procedure must scale to high dimensions while maintaining accuracy.
-
-## Motivating example: RNA-seq data
-
-Simulating high-dimensional, non-normal, correlated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients.
-The RNA-seq data-generating process involves counting how often a particular form of human messenger RNA (mRNA) is expressed in a biological sample.
-RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data.
-For human-derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted [@Schissler2019].
-Importantly, due to inherent biological processes, gene expression data exhibit correlation (co-expression) across genes [@BE07; @Schissler2018].
-
-We illustrate our methodology using a well-studied Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments).
-For simplicity, we only consider high expressing genes.
-In turn, we begin by filtering to retain the top `r d` of the highest-expressing genes (in terms of median expression) of the over 20,000 gene measurements from $N=`r nrow(example_brca)`$ patients' tumor samples.
-This gives a great number of pairwise dependencies among the marginals (specifically, $`r choose(d,2)`$ correlation parameters).
-Table \@ref(tab:ch010-realDataTab) displays RNA-seq counts for three selected high-expressing genes for the first five patients' breast tumor samples.
-
-```{r ch010-realDataTab}
-set.seed(2020-02-25)
-num_genes    <- 3
-num_patients <- 5
-gene_sample  <- sample(example_genes[1:1000], num_genes)
-
-cap <- paste0("mRNA expression for three selected high-expressing genes, ",
-              paste(gene_sample, collapse = ", "), 
-              ", for the first five patients in the TCGA BRCA data set.")
-
-small_brca <- example_brca %>%
-  select(all_of(gene_sample)) %>%
-  head(n = num_patients) %>%
-  as_tibble(rownames = "Patient ID") %>%
-  mutate(`Patient ID` = str_sub(`Patient ID`, end = 12))
-
-small_brca %>%
-  kable(booktabs = TRUE, caption = cap)
+```{r, child=get_rmd("030")}
 ```
 
-To help visualize the bivariate relationships for these three selected genes across all patients, Figure \@ref(fig:ch010-realDataFig) displays the marginal distributions and estimated Spearman's correlations (see Equation \@ref(eq:spearman) below).
-
-```{r ch010-realDataFig, cache=FALSE, fig.width=8, fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes from Table 1.
-The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations.
-Modeling these data motivate our simulation methodology."}
-ggpairs(
-  data = example_brca[, gene_sample],
-  upper = list(continuous = wrap('cor', method = "spearman"))
-) + theme_bw()
+```{r, child=get_rmd("040")}
 ```
 
-## Measures of dependency
-
-In multivariate analysis, an analyst must select a metric to quantify dependency.
-The most widely-known is the Pearson (product-moment) correlation coefficient that describes the linear association between two random variables $X$ and $Y$, and, it is given by
-
-\begin{equation}
-(\#eq:pearson)
-\rho_P(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ \mathrm{Var}(X)\mathrm{Var}(Y)\right]^{1/2}}.
-\end{equation}
-
-As @MB13 and @MK01 discuss, for a bivariate normal $(X,Y)$ random vector, the Pearson correlation completely describes the dependency between the components.
-For non-normal marginals with monotone correlation patterns, however, $\rho_P$ suffers some drawbacks and may mislead or fail to capture important relationships [@MK01].
-Alternatively in these settings, analysts often prefer rank-based correlation measures to describe the degree of monotonic association.
-
-Two nonparametric, rank-based measures common in practice are Spearman's correlation (denoted $\rho_S$) and Kendall's $\tau$.
-Spearman's $\rho_S$ has an appealing correspondence as the Pearson correlation coefficient on *ranks* of the values, thereby captures nonlinear yet monotone relationships.
-Kendall's $\tau$, on the other hand, is the difference in probabilities of concordant and discordant pairs of observations, $(X_i, Y_i)$ and $(X_j, Y_j)$, with concordance meaning that orderings have the same direction (e.g., if $X_i < X_j$, then $Y_i < Y_j$).
-Note that concordance is determined by the ranks of the values, not the values themselves.
-
-Both $\tau$ and $\rho_S$ are *invariant under monotone transformations* of the underlying random variates.
-As we will describe more fully in the [Algorithms](#algorithms) section, this property enables matching rank-based correlations with speed (SP1) and accuracy (SP2).
-
-*Correspondence among Pearson, Spearman, and Kendall correlations*
-
-There is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficient, as the marginal distributions $F_i$ are intrinsic in their calculation.
-For *bivariate normal vectors*, however, the correspondence is well-known:
-
-\begin{equation}
-(\#eq:convertKendall)
-\rho_{P} = \sin \left( \tau \times \frac{\pi}{2} \right), 
-\end{equation}
-
-\noindent and similarly for Spearman's $\rho$ [@K58],
-
-\begin{equation}
-(\#eq:convertSpearman)
-\rho_P = 2 \times \sin \left( \rho_S \times \frac{\pi}{6} \right).
-\end{equation}
-
-*Discrete marginal considerations*
-
-Spearman's correlation for discrete marginal suffers issues whenever there is a nonzero probability of ties.
-For example, the Spearman's correlation of a discrete-valued random variable $X$ with itself could be less than 1 [@MB13].
-One remedy for this issue is to rescale Equation \@ref(eq:spearman).
-For two random variables $X,Y$ with probability mass functions (PMFs) or probability densities functions (PDFs) $p(x)$ and $q(y)$, respectively, define the rescaled Spearman's correlation as
-
-\begin{equation}
-(\#eq:spearmanRescaled)
-\rho_{RS}(X,Y) = \frac{\rho_s(X,Y)}{ \left[ \left[ 1 - \sum_x p(x)^3 \right] \left[ 1 - \sum_y q(y)^3 \right] \right]^{1/2}}.
-\end{equation}
-
-Note that with continuous marginals the rescaling returns $\rho_S$.
-For discrete marginals with large or infinite support, computing the adjustment factors $\sum_x p(x)^3$, $\sum_y p(y)^3$ over all large number of pairs becomes expensive (often violating desired property SP1).
-Further, the infinite sums must be approximated for count-valued data.
-
-Also, for a discrete random variable $Y_i$, some care must be taken to define the quantile function $F_{i}^{-1}$. A standard practice is to set
-
-\begin{equation}
-F_{i}^{-1} = \inf\{y:F_{i}(y) \geq u \}.
-(\#eq:inverseCDF)
-\end{equation}
-
-*Marginal-dependent bivariate correlation bounds*
-
-Given two marginal distributions, $\rho_P$ is not free to vary over the entire range of possible correlations $[-1,1]$.
-The so-called *Frechet-Hoeffding bounds* are well-studied [@Nelsen2007; @BF17].
-These constraints cannot be overcome through algorithm design.
-In general, the bounds are given by
-
-\begin{equation}
-(\#eq:frechet)
-\rho_P^{max} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (U) \right), \quad \rho_P^{min} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (1 - U) \right)
-\end{equation}
-
-\noindent where $U$ is a uniform random variable on $(0,1)$ and $F^{-1}_1, F^{-1}_2$ are the inverse CDFs of $X_1$ and $X_2$, respectively, definitely by \@ref(eq:inverseCDF) when the variables are discrete.
-
-## Gaussian copulas
-
-There is a strong connection of our simulation strategy to Gaussian *copulas* (see @Nelsen2007 for a technical introduction).
-A copula is a distribution function on $[0,1]^d$ that describes a multivariate probability distribution with standard uniform marginals.
-This provides a powerful, natural way to characterize joint probability structures.
-Consequently, the study of copulas is an important and active area of statistical theory and practice.
-
-For any random vector ${\bf X}=(X_1, \ldots, X_d)^\top$ with CDF $F$ and marginal CDFs $F_i$ there is a copula function $C(u_1, \ldots, u_d)$ satisfying
-
-\begin{equation}
-(\#eq:copula)
-F(x_1, \ldots,x_d) = \mathbb P(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), 
-\end{equation}
-
-$x_i \in \mathbb R, i=1,\ldots,d.$
-
-A Gaussian copula has marginal CDFs that are all standard normal, $F_i = \Phi, \forall \, i$.
-This representation corresponds to a multivariate normal (MVN) distribution with standard normal marginal distributions and covariance matrix ${\bf R_P}$.
-As the marginals are standardized to have unit variance, however, ${\bf R_P}$ is a Pearson correlation matrix.
-If $F_{{\bf R}}$ is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula $C_{{\bf R}}$ is defined through
-
-\begin{equation}
-(\#eq:gauss)
-F_{{\bf R}}(x_1, \ldots, x_d) = C_{{\bf R}}(\Phi(x_1), \ldots, \Phi(x_d)),
-\end{equation}
-
-where $\Phi(\cdot)$ is the standard normal CDF.
-Note that the copula $C_{{\bf R}}$ is the familiar multivariate normal CDF of the random vector $(\Phi(X_1), \ldots, \Phi(X_d))$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})$.
-
-Sklar's Theorem [@Sklar1959; @Ubeda-Flores2017] guarantees that given inverse CDFs $F_i^{-1}$s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions.
-For example, using Gaussian copulas, we can construct a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)^\top$ with $Y_i \sim F_i$, viz.
-$Y_i = F_i^{-1}(\Phi(X_i)), i=1, \ldots, d$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})$.
-```{r include=FALSE, cache=FALSE}
-# Remove all objects
-rm(list = ls(all.names = TRUE))
-
-# Unload all packages
+```{r, child=get_rmd("050")}
 ```
 
-<!--chapter:end:010-background.Rmd-->
-
-```{r include=FALSE, cache=FALSE}
-options(digits = 4)
-
-knitr::opts_chunk$set(
-  # Text Results
-  eval = TRUE,
-  echo = FALSE,
-  results = 'markup',
-  collapse = TRUE,
-  message = FALSE,
-  warning = FALSE,
-  error = FALSE,
-  # Code Decoration
-  comment = NA,
-  background = '#F7F7F7',
-  # Cache
-  cache = TRUE,
-  # Plots
-  fig.show = "hold",
-  fig.width = 6.5,
-  fig.asp = 0.618,  # 1 / phi
-  out.width = "85%",
-  fig.align = 'center'
-)
-
-# options(dplyr.print_min = 6, dplyr.print_max = 6)
-
-seed_number <- 2020-02-25
-set.seed(seed = seed_number)
-```
-# Algorithms {#algorithms}
-
-This section describes our methods involved in simulating a random vector $\bf Y$ with $Y_i$ components for $i=1,\ldots,d$.
-Each $Y_i$ has a specified marginal CDF $F_i$ and its inverse $F^{-1}_i$.
-To characterize dependency, every pair $(Y_i, Y_j)$ has a given Pearson correlation $\rho_P$, Spearman correlation $\rho_S$, and/or Kendall's $\tau$.
-The method can be described as a *high-performance Gaussian copula* (Equation \@ref(eq:gauss)) providing a high-dimensional NORTA-inspired algorithm.
-
-## NORmal To Anything (NORTA)
-
-The well-known NORTA algorithm [@Cario1997] simulates a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$. Specifically, NORTA algorithm proceeds as follows:
-
-1. Simulate a random vector $\bf Z$ with $d$ *independent and identically distributed* (iid) standard normal components.
-2. Determine the input matrix $\Sigma_{\bf Z}$ that corresponds with the specified output $\Sigma_{\bf Y}$.
-3. Produce a Cholesky factor $M$ of $\Sigma_{\bf Z}$ such that $M M^{\prime}=\Sigma_{\bf Z}$.
-4. Set $X$ by $X \gets MZ$.
-5. $\text{Return} \; Y \; \text{where} \; Y_i \gets F_i^{-1}[\Phi(X_i)], \; i=1,...,d$.
-
-With modern parallel computing, steps 1, 3, 4, 5 are readily implemented as high-performance, multi-core and/or graphical-processing-unit (GPU) accelerated algorithms --- providing fast scalability using readily-available hardware.
-
-Matching specified Pearson correlation coefficients exactly (step 2 above), however, is computationally costly.
-In general, there is no closed-form correspondence between the components of the input $\Sigma_{\bf Z}$ and target $\Sigma_{\bf Y}$.
-Matching the correlations involves evaluating or approximating $\binom{d}{2}$ integrals of the form
-
-\begin{equation}
-    \mathrm{E}\left[Y_i Y_j\right] = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} F_i^{-1}\left[\Phi(z_i)\right] F_j^{-1}\left[\Phi(z_j)\right] \phi(z_i, z_j, \rho_z) dz_i dz_j,
-    (\#eq:pearsonIntegralRelation)
-\end{equation}
-
-where $\phi(\cdot)$ is the joint probability distribution function of two correlated standard normal variables.
-For HD data, (nearly) exact evaluation may become too costly to enable practical simulation studies.
-For low-dimensional problems, however, methods and tools exist to match Pearson correlations precisely; see [@Xia17] and the publicly available `nortaRA` R package [@Chen2001].
-As described later, to enable HD Pearson matching we approximate these integrals.
-
-*NORTA in higher dimensions*
-
-Sklar's theorem provides a useful characterization of multivariate distributions through copulas.
-Yet the choice of copula-based simulation algorithm affects which joint distributions may be simulated.
-Even in low-dimensional spaces (e.g., $d=3$), there exist valid multivariate distributions with *feasible* Pearson correlation matrices that NORTA cannot match exactly [@LH75].
-This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined the resultant matrix is indefinite.
-These situations do occur, even using exact analytic calculations.
-Such problematic target correlation matrices are termed *NORTA defective*.
-
-@GH02 conducted a Monte Carlo study to estimate the probability of encountering NORTA defective matrices while increasing the dimension $d$.
-They found that for what is now considered low-to-moderate dimensions ($d \approx 20$), almost *all* feasible matrices are NORTA defective.
-This stems from the concentration of measure near the boundary of the space of all possible correlation matrices as dimension increases.
-Unfortunately, it is precisely near this boundary that NORTA defective matrices reside.
-
-There is hope, however, as @GH02 also showed that replacing an indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$.
-This provides evidence that our nearest positive definite (PD) augmented approach will maintain reasonable accuracy if our input matching scheme returns an indefinite matrix, at least for the rank-based matching scheme described above.
-
-## Random vector generator {#rand-vec-gen}
-
-We now describe our algorithm to generate random vectors, which mirrors the classical NORTA algorithm above with some modifications for rank-based dependency matching:
-
-1. Mapping step
-    (i) Convert the target Spearman correlation matrix $R_S$ to the corresponding MVN Pearson correlation $R_X$. Alternatively,  
-    (i') Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$. Alternatively,  
-    (i'') Convert the target Pearson correlation matrix to $R_P$ to the corresponding, approximate MVN Pearson correlation $R_X$.  
-2. Check admissibility and, if needed, compute the nearest correlation matrix.
-    (i) Check that $R_X$ is a correlation matrix, a positive definite matrix with 1's along the diagonal.   
-	(ii) If $R_X$ is a correlation matrix, the input matrix is *admissible* in this scheme. Otherwise
-    (ii$^\prime$) Replace $R_X$ with the nearest correlation matrix $\tilde{R}_X$, in the Frobenius norm.  
-3. Gaussian copula
-    (i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)$.  
-    (ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz. $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
-    (iii) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$.
-
-*Step 1*
-
-The first two descriptions of the *Mapping step* employ the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertKendall) and \@ref(eq:convertSpearman), respectively (implemented as `cor_covert`).
-Initializing our algorithm to match the nonparametric correlations by computing these equations for all pairs is computationally trivial.
-
-For the computationally expensive process to match Pearson correlations, we approximate Equation \@ref(eq:pearsonIntegralRelation) for all pairs of margins.
-To this end, we implement the approximation scheme introduced by @xiao2019matching.
-Briefly, the many double integrals of the form in \@ref(eq:pearsonIntegralRelation) are approximated by weighted sums of Hermite polynomials.
-Matching coefficients for pairs of continuous distributions is made tractable by this method, but for discrete distributions (especially discrete distributions with large support sets or infinite support), the problem is not so simple or efficient.
-
-Our solution is to approximate discrete distributions by a continuous distribution.
-The question then becomes that of which distribution to use.
-We found that Generalized S-Distributions [@muino2006gs] solve this problem by approximating a wide range of unimodal distributions, both continuous and discrete.
-Using the GS-Distribution approximation of discrete distributions in Pearson matching scheme above yields favorable results.
-
-*Step 2*
-
-Once $R_X$ has been determined, we check admissibility of the adjusted correlation via the steps described above.
-If $R_X$ is not a valid correlation matrix, then we compute the nearest correlation matrix.
-Finding the nearest correlation matrix is a common statistical computing problem.
-The defacto default function in `R` is `Matrix::nearPD`, an alternating projection algorithm due to @higham2002computing.
-As implemented the function fails to scale to HD.
-Instead, we provide the quadratically convergent algorithm based on the theory of strongly semi-smooth matrix functions (@qi2006quadratically).
-The nearest correlation matrix problem can be written down as the following convex optimization problem:
-
-\begin{align*}
-    \mathrm{min}\quad & \frac{1}{2} \Vert G - X \Vert^2 \\
-    \mathrm{s.t.}\quad & X_{ii} = 1, \quad i = 1, \ldots , n, \\
-    & X \in S_{+}^{n}
-\end{align*}
-
-For nonparametric correlation measures, our algorithm allows the generation of high-dimensional multivariate data with arbitrary marginal distributions with a broad class of admissible Spearman correlation matrices and Kendall $\tau$ matrices.
-The admissible classes consist of the matrices that map to a Pearson correlation matrix for a MVN.
-In particular, if we let $X$ be MVN with $d$ components and set
-
-\begin{align*}
-\Omega_P &= \{ R_P : R_P \textrm{ is a correlation matrix for } X \} \\
-\Omega_K &= \{ R_K : R_K \textrm{ is a correlation matrix for } X \} \\
-\Omega_S &= \{ R_S : R_S \textrm{ is a correlation matrix for } X \} \\
-\end{align*}
-
-then there are 1-1 mappings between these sets.
-We conjecture informally that the sets of admissible $R_S$ and $R_K$ are not highly restrictive.
-In particular, $R_P$ is approximately $R_S$ for a MVN, suggesting that the admissible set $\Omega_S$ should be flexible as $R_P$ can be any PD matrix with 1's along the diagonal.
-We provide methods to check whether a target $R_S$ is an element of the *admissible set* $\Omega_S$ (and similarly for $R_K$).
-
-There is an increasing probability of encountering a non-admissible correlation matrix as dimension increases.
-In our experience, the mapping step for large $d$ almost always produces a $R_X$ that is not a correlation matrix.
-In Section \@ref(package) we provide a basic description of how to quantify and control the approximation error.
-Further, the RNA-seq example in Section \@ref(examples) provides an illustration of this in practice.
-
-*Step 3*
-
-The final step implements a NORTA-inspired, Gaussian copula approach to produce the desired margins.
-Steps 1 and 2 determine the MVN Pearson correlation values that will eventually match the target correlation.
-Then all that is required is a fast MVN simulator, a standard normal CDF, and well-defined quantile functions for marginals.
-The MVN is transformed to a copula (distribution with standard uniform margins) by applying the normal CDF $\phi(\cdot)$.
-The quantile functions $F^{-1}$ are applied across the margins to return the desired random vector ${\bf Y}$.
-
-```{r include=FALSE, cache=FALSE}
-# Remove all objects
-rm(list = ls(all.names = TRUE))
-
-# Unload all packages
+```{r, child=get_rmd("060")}
 ```
 
-<!--chapter:end:020-simulation-algorithm.Rmd-->
+<!-- Supplementary materials -->
+\newpage
 
-```{r include=FALSE, cache=FALSE}
-options(digits = 4)
-
-knitr::opts_chunk$set(
-  # Text Results
-  eval = TRUE,
-  echo = FALSE,
-  results = 'markup',
-  collapse = TRUE,
-  message = FALSE,
-  warning = FALSE,
-  error = FALSE,
-  # Code Decoration
-  comment = NA,
-  background = '#F7F7F7',
-  # Cache
-  cache = TRUE,
-  # Plots
-  fig.show = "hold",
-  fig.width = 6.5,
-  fig.asp = 0.618,  # 1 / phi
-  out.width = "85%",
-  fig.align = 'center'
-)
-
-# options(dplyr.print_min = 6, dplyr.print_max = 6)
-
-seed_number <- 2020-02-25
-set.seed(seed = seed_number)
-```
-# The `bigsimr` R package {#package}
-
-```{r ch030-basic-example-setup, echo=TRUE, cache=FALSE}
-box::use(
-  patchwork[...],
-  dplyr[...],
-  ggplot2[...],
-  tidyr[drop_na],
-  bigsimr[bigsimr_setup, distributions_setup],
-  JuliaCall[julia_call]
-)
-
-Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
-bs <- bigsimr_setup(pkg_check = FALSE)
-dist <- distributions_setup()
+```{r, child=get_rmd("070")}
 ```
 
-This section describes a low-dimensional (2D) random vector simulation workflow via the `bigsimr` `R` package (https://github.com/SchisslerGroup/r-bigsimr).
-The package `bigsimr` provides an interface to the native code written in Julia, registered as the `Bigsimr` Julia package (https://github.com/adknudson/Bigsimr.jl).
-In addition to the native Julia `Bigsimr` package and `R` interface `bigsimr`, we also provide a python interface `bigsimr` (https://github.com/SchisslerGroup/python-bigsimr/) that interfaces with the Julia `Bigsimr` package.
-The Julia implementation provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms)).
-When designing `bigsimr`, we aimed to conveniently provide parallelized computation, through multi-core and GPU acceleration, while allowing advanced users to customize.
+<!-- Appendix -->
+\newpage
 
-The subsections below describe the basic use of `bigsimr`, by stepping through an example workflow using the data set `airquality` that contains daily air quality measurements in New York, May to September 1973 [@Chambers1983].
-This workflow proceeds from computing environment setup, data wrangling, estimation, simulation configuration, random vector generation, and, finally, result visualization.
-The code chunk below prepares the computing environment:
-
-<!-- 
-`bigsimr` is an R package for simulating high-dimensional multivariate data with arbitrary marginal distributions.
-The efficiency behind generating multivariate samples comes from the ability to utilize a GPU during the generation of multivariate normal samples and the subsequent transformation to uniform samples (see section \@ref(rand-vec-gen)).
-A GPU is not necessary in order to use `bigsimr`, and the speed benefits generally only come when simulating very high dimensional data ($d > 10000$).
-
-The next computational step after obtaining the correlated uniform marginals is the inverse transform into the target marginal distributions.
-Since this consists of $d$ independent transformations, we utilize parallelization to achieve higher throughput.
-As with many parallel algorithms, there is overhead associated with forking the task to utilize multiple cores, however cost is negligible compared to the total run-time.
--->
-
-## Bivariate example description
-
-We illustrate the use of `bigsimr` motivated by the New York air quality data set (`airquality`) included in the R `datasets` package.
-First, we load the `bigsimr` library and a few other convenient data science packages, including the syntactically-elegant `tidyverse` suite of `R` packages.
-
-For simplicity and to provide a minimal working example, we consider bivariate simulation of the two `airquality` variables `Temperature`, in degrees Fahrenheit, and `Ozone` level, in parts per billion.
-
-```{r ch030-air-quality-data, echo=TRUE}
-df <- airquality %>%
-  select(Temp, Ozone) %>%
-  drop_na()
+```{r, child=get_rmd("080")}
 ```
 
-```{r ch030-aq-glimpse}
-glimpse(df)
-```
-
-Figure \@ref(fig:ch030-aq-joint-dist) visualizes the bivariate relationship between Ozone and Temperature.
-We aim to simulate random two-component vectors mimicking this structure.
-The margins are not normally distributed; particularly the Ozone level exhibits a strong positive skew.
-
-```{r ch030-aq-joint-dist, cache=FALSE, fig.width= 8, fig.cap="Bivariate scatterplot of Ozone vs.
-Temp with estimated marginal densities.
-We model the Ozone data as marginally log-normal and the Temperature data as normal."}
-p0 <- ggplot(df, aes(Temp, Ozone)) +
-  geom_point(size = 1) +
-  theme(legend.position = "none") + 
-  labs(x = "Temperature")
-
-pTemp <- ggplot(df, aes(Temp)) + 
-  geom_density() +
-  theme(axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        axis.text = element_blank(),
-        axis.ticks = element_blank())
-
-pOzone <- ggplot(df, aes(Ozone)) + 
-    geom_density() +
-  theme(axis.title.y = element_blank(),
-        axis.title.x = element_blank(),
-        axis.ticks = element_blank(),
-        axis.text = element_blank()) +
-  coord_flip()
-
-pTemp + plot_spacer() + p0 + pOzone + 
-  plot_layout(widths = c(3,1), heights = c(1, 3)) 
-```
-
-Next, we specify the marginal distributions and correlation coefficient (both type and magnitude).
-Here the analyst is free to be creative.
-For this example, we avoid goodness-of-fit considerations to determine the marginal distributions.
-But it seems sensible without domain knowledge to estimate these quantities from the data and `bigsimr` contains fast functions designed for this task.
-
-## Specifying marginal distributions
-
-Based on the estimated densities in Figure \@ref(fig:ch030-aq-joint-dist), we assume `Temp` is normally distributed and `Ozone` is log-normally distributed, as the latter values are positive and skewed.
-We use the well-known, unbiased estimators for the normal distribution's parameters and maximum likelihood estimators for the log-normal parameters:
-
-```{r ch030-aq-temp-pars, echo=TRUE}
-df %>% 
-  select(Temp) %>% 
-  summarise_all(.funs = c(mean = mean, sd = sd))
-```
-
-```{r ch030-aq-ozone-pars, echo=TRUE}
-mle_mean <- function(x) mean(log(x))
-mle_sd <- function(x) mean( sqrt( (log(x) - mean(log(x)))^2 ) )
-
-df %>% 
-  select(Ozone) %>% 
-  summarise_all(.funs = c(meanlog = mle_mean, sdlog = mle_sd))
-```
-
-Next, we configure the input marginals for later input into `rvec`. The marginal distributions are specified using `Julia`'s `Distributions` package and stored in a vector.
-
-```{r ch030-margins-alist, echo=TRUE}
-margins <- c(
-  dist$Normal(mean(df$Temp), sd(df$Temp)),
-  dist$LogNormal(mle_mean(df$Ozone), mle_sd(df$Ozone))
-)
-```
-
-## Specifying correlation
-
-As mentioned, the user must decide how to describe correlation, based on the particulars of the problem.
-For non-normal data and for improved simulation accuracy/scalability in our scheme, we advocate the use of Spearman's $\rho$ correlation matrix $R_S$ and Kendall's $\tau$ correlation matrix $R_K$.
-We also support Pearson correlation coefficient matching, while cautioning the user to check the performance for their parametric multivariate model (see [Monte Carlo evaluations](#simulations) below for evaluation strategies and guidance).
-These estimation methods are classical approaches, not designed for high-dimensional correlation estimation (see the [Conclusion and Discussion]({#discussion) sections for more on this).
-
-```{r ch030-aq-cor, echo=TRUE}
-(R_S <- bs$cor(as.matrix(df), bs$Spearman))
-```
-
-## Checking target correlation matrix admissibility
-
-First we use `cor_bounds` to ensure that the pairwise target correlation values are valid prior to mapping step which constructs $R_X$ for the MVN input into NORTA (see Section \@ref(algorithms)).
-`cor_bounds` estimates the pairwise lower and upper correlation bounds using the Generate, Sort, and Correlate algorithm of @DH2011.
-
-```{r ch030-cor-bounds, echo=TRUE}
-bounds <- bs$cor_bounds(margins)
-bounds$lower
-```
-
-Since our single estimated Spearman correlation is within the theoretical bounds, the correlation is valid as input to `rvec`.
-But even if the 2-dimensional bounds are satisfied for each pair of margins, this does not guarantee the feasibility of a $d-$variate distribution [@BF17].
-
-To provide higher dimensional feasibility/admissibility checking, we begin by mapping using `cor_convert` and check admissibility.
-
-```{r ch030-cor-convert, echo=TRUE}
-# Step 1. Mapping
-(R_X <- bs$cor_convert(R_S, bs$Spearman, bs$Pearson))
-
-# Step 2. Check admissibility
-julia_call("Bigsimr.iscorrelation", R_X)
-```
-
-The bounds on the Pearson correlation coefficient $R_P$ between these margins are restricted as seen below in our MC estimated correlation bounds:
-
-```{r ch030-cor-bounds-pearson, echo=TRUE}
-bs$cor_bounds(margins[1], margins[2], bs$Pearson, n_samples = 1e6)
-```
-
-Our MC estimate of the bounds slightly underestimates the theoretic bounds of $(-0.881, 0.881)$. An analytic derivation is presented as an Appendix.
-
-## Simulating random vectors
-Finally, we arrive at the main function of `bigsimr`, `rvec`. We now simulate $B=10,000$ random vectors from the assumed joint distribution of Ozone levels and Temp.
-
-```{r ch030-sim-margins, echo=TRUE}
-x <- bs$rvec(10000, R_X, margins)
-df_sim <- as.data.frame(x)
-colnames(df_sim) <- colnames(df)
-```
-
-Figure \@ref(fig:ch030-plot-sim) plots the 10,000 simulated points.
-
-```{r ch030-plot-sim, fig.cap="Contour plot and marginal densities for the simulated bivariate distribution of Air Quality Temperatures and Ozone levels.
-The simulated points mimic the observed data with respect to both the marginal characteristics and bivariate association."}
-p1 <- df_sim %>%
-  ggplot(aes(Temp, Ozone)) +
-  geom_density_2d_filled() +
-  theme(legend.position = "none") +
-  labs(x = "Simulated Temperature", y = "Simulated Ozone") +
-  scale_y_continuous(limits = c(0, max(df$Ozone))) +
-  scale_x_continuous(limits = range(df$Temp))
-
-p1Temp <- ggplot(df_sim, aes(Temp)) + 
-  geom_density(alpha = 0.5, fill = "lightseagreen") +
-  theme(axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        axis.text = element_blank(),
-        axis.ticks = element_blank())
-
-p1Ozone <- ggplot(df_sim, aes(Ozone)) + 
-  geom_density(alpha = 0.5, fill = "lightseagreen") + 
-  theme(axis.title.y = element_blank(),
-        axis.title.x = element_blank(),
-        axis.ticks = element_blank(),
-        axis.text = element_blank()) +
-  coord_flip()
-
-p1Temp + plot_spacer() + p1 + p1Ozone + 
-  plot_layout(widths = c(3,1), heights = c(1, 3))
-```
-```{r include=FALSE, cache=FALSE}
-# Remove all objects
-rm(list = ls(all.names = TRUE))
-
-# Unload all packages
-```
-
-<!--chapter:end:030-bigsimr-package.Rmd-->
-
-```{r include=FALSE, cache=FALSE}
-options(digits = 4)
-
-knitr::opts_chunk$set(
-  # Text Results
-  eval = TRUE,
-  echo = FALSE,
-  results = 'markup',
-  collapse = TRUE,
-  message = FALSE,
-  warning = FALSE,
-  error = FALSE,
-  # Code Decoration
-  comment = NA,
-  background = '#F7F7F7',
-  # Cache
-  cache = TRUE,
-  # Plots
-  fig.show = "hold",
-  fig.width = 6.5,
-  fig.asp = 0.618,  # 1 / phi
-  out.width = "85%",
-  fig.align = 'center'
-)
-
-# options(dplyr.print_min = 6, dplyr.print_max = 6)
-
-seed_number <- 2020-02-25
-set.seed(seed = seed_number)
-```
-# Monte Carlo evaluations {#simulations}
-
-```{r ch040-LoadLib040, include=FALSE}
-box::use(
-    dplyr[...],
-    ggplot2[...],
-    patchwork[...],
-    tidyr[pivot_longer],
-    bigsimr[bigsimr_setup, distributions_setup],
-    knitr[kable],
-    RColorBrewer[brewer.pal]
-)
-load("data/bivariate_normal_sims.rda")
-load("data/bivariate_gamma_sims.rda")
-load("data/bivariate_nbinom_sims.rda")
-load("data/benchmark_dependences.rda")
-
-Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
-bs <- bigsimr_setup(pkg_check = FALSE)
-dist <- distributions_setup()
-prob <- 3e-04
-```
-
-Before applying our methodology to real data simulation, we conduct several Monte Carlo studies to investigate method performance.
-Since marginal parameter matching in our scheme is essentially a sequence of univariate inverse probability transforms, the challenging aspects are the accuracy of dependency matching and computational efficiency at high dimensions.
-To evaluate our methods in those respects, we design the following numerical experiments to first assess accuracy of matching dependency parameters in bivariate simulations and then time the procedure in increasingly large dimension $d$.
-
-## Bivariate experiments
-
-We select bivariate simulation configurations to ultimately simulate our motivating discrete-valued RNA-seq example, and, so we proceed in increasing complexity, leading to the model in our motivating application in Section \@ref(examples).
-We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson's, Spearman's, and Kendall's --- in identical, bivariate marginal configurations.
-For each pair of identical margins, we vary the target correlation across $\Omega$, the set of possible admissible values for correlation type, to evaluate the simulation's ability to obtain the theoretic bounds.
-The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (mimicking RNA-seq counts).
-
-Table \@ref(tab:sims) lists our identical-marginal, bivariate simulation configurations.
-We increase the simulate replicates $B$ to check that our results converge to the target correlations and gauge statistical efficiency.
-We select distributions beginning with a standard multivariate normal (MVN) as we expect the performance to be exact (up to MC error) for all correlation types.
-Then, we select a non-symmetric continuous distribution: a standard (rate =1) two-component multivariate gamma (MVG).
-Finally, we select distributions and marginal parameter values that are motivated by our RNA-seq data, namely values proximal to probabilities and sizes estimated from the data (see [Example applications](examples) for estimation details).
-Thus we arrive at a multivariate negative binomial (MVNB) $p_1 = p_2 = 3\times10^{-4}, r_1 = r_2 = 4, \rho \in \Omega$.
-
-Table: (\#tab:sims) Identical margin, bivariate simulation configurations to evaluate correlation matching accuracy and efficiency.
-
-| Simulation Reps ($B$) | Correlation Types | Identical-margin 2D distribution |
-|-------------|:--------------:|----------------------:|
-| $1000$    | Pearson ($\rho_P$) | ${ \bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho_i ), i=1,\ldots,100$ |
-| $10,000$  | Spearman ($\rho_S$) | ${ \bf Y} \sim MVG( shape = 10, rate = 1, \rho_i ), i=1,\ldots,100$ |
-| $100,000$ | Kendall ($\tau$)| ${ \bf Y} \sim MVNB(p = 3\times10^{-4}, r = 4,\rho_i), i=1,\ldots,100$ |
-
-For each of the unique 9 simulation configurations described above, we estimate the correlation bounds and vary the correlations along a sequence of 100 points evenly placed within the bounds, aiming to explore $\Omega$.
-Specifically, we set correlations $\{ \rho_1 = ( \hat{l} + \epsilon), \rho_2 = (\hat{l} + \epsilon) + \delta, \ldots, \rho_{100} = (\hat{u} - \epsilon) \}$, with $\hat{l}$ and $\hat{u}$ being the estimated lower and upper bounds, respectively, and increment value $\delta$.
-The adjustment factor, $\epsilon=0.01$, is introduced to handle numeric issues when the bound is specified exactly.
-
-```{r ch040-biNormPlot, eval=FALSE, fig.height=5, fig.width=8, fig.cap="`bigsimr` recovers the Pearson specified correlations for MVN."}
-bivariate_normal_sims %>%
-    ggplot(aes(rho, rho_hat, color = type)) +
-    geom_point() +
-    geom_abline(slope = 1) +
-    facet_wrap(~ type + N) + 
-    theme_bw()
-```
-
-```{r ch040-biGammaPlot, eval=FALSE, fig.height=5, fig.width=8, fig.cap="`bigsimr` recovers the Pearson specified correlations for Bivariate Gamma."}
-bivariate_gamma_sims %>%
-    ggplot(aes(rho, rho_hat, color = type)) +
-    geom_point() +
-    geom_abline(slope = 1) +
-    facet_wrap(~ type + N) + 
-    theme_bw()
-```
-
-```{r ch040-biNegBinPlot, eval=FALSE, fig.width=8, fig.height=5, fig.cap="`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
-bivariate_nbinom_sims %>%
-    ggplot(aes(rho, rho_hat, color = type)) +
-    geom_point() +
-    geom_abline(slope = 1) +
-    facet_wrap(~ type + N) + 
-    theme_bw()
-```
-
-```{r ch040-combineBiSims, cache=FALSE}
-allDat <- bind_rows(
-    select(bivariate_normal_sims, margins, type, N, rho, rho_hat),
-    select(bivariate_gamma_sims, margins, type, N, rho, rho_hat),
-    select(bivariate_nbinom_sims, margins, type, N, rho, rho_hat)
-) %>%
-    mutate(margins = factor(margins, levels=c("norm", "gamma", "nbinom")),
-           type = factor(type, levels=c("Pearson", "Spearman", "Kendall")),
-           N = factor(N))
-```
-
-Figure \@ref(fig:ch040-bPlot) displays the aggregated bivariate simulation results.
-Table \@ref(tab:ch040-BiError) contains the mean absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.
-
-```{r ch040-BiError}
-tabMAE <- allDat %>%
-    group_by(N, type, margins) %>%
-    summarize(MAE = mean(abs(rho - rho_hat))) %>%
-    ungroup()
-
-kable(tabMAE, booktabs = TRUE,
-      col.names = c("No. of random vectors",
-                    "Correlation type",
-                    "Distribution",
-                    "Mean abs. error"),
-      caption = "Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.")
-```
-
-Overall, the studies show that our methodology is generally accurate across the entire range of possible correlation values all three dependency measures, at least in these limited simulation settings for the rank-based correlations.
-Our Pearson matching performs nearly as well as Spearman or Kendall, except a slight increase in error for negative binomial case.
-This is due the particularly large counts generated with our choice of parameters for $p$ and $r$.
-
-```{r ch040-bPlot, cache=FALSE, fig.asp=1.30, fig.weight=4, fig.cap="Bivariate simulations match target correlations across the entire range of feasible correlations.
-The horizontal axis plots the specified target correlations for each bivariate margin.
-Normal margins are plotted in dark dark grey, gamma in medium grey, and negative binomial in light grey.
-As the number of simulated vectors $B$ increases from left to right, the variation in estimated correlations (vertical axis) decreases.
-The dashed line indicates equality between the specified and estimated correlations."}
-# https://www.datanovia.com/en/blog/how-to-change-ggplot-facet-labels/
-# New facet label names
-repsLabs <- paste0("B=", c("1,000", "10,000", "100,000") )
-names(repsLabs) <- c(1000, 10000, 100000)
-typeLabs <- c("Pearson", "Spearman", "Kendall")
-names(typeLabs) <- c("Pearson", "Spearman", "Kendall")
-
-# Set colors
-## RColorBrewer::display.brewer.all()
-numColors <- 4
-numGroups <- length(levels(allDat$margins))
-myColors <- rev(RColorBrewer::brewer.pal(n = numColors, name = "Greys")[ ((numColors - numGroups) + 1):numColors])
-
-allDat %>%
-    ggplot(aes(x = rho, y = rho_hat, color = margins)) +
-    geom_point(size = 2) +
-    scale_color_manual(values = myColors ) +
-    geom_abline(slope = 1, linetype = "dashed") +
-    labs(x = "Specified Correlation", y = "Estimated Correlation") +
-    facet_wrap(~ type + N, labeller = labeller(N = repsLabs, type = typeLabs)) +
-    theme_bw() +
-    theme(legend.position = "bottom", legend.direction = "horizontal")
-```
-
-<!-- 
-See [Discussion](discussion) for future directions for fast Pearson matching and discrete-specific modifications.
--->
-
-## Scale up to High Dimensions
-
-With information of our method's accuracy from a low-dimensional perspective, we now turn to assessing whether the `bigsimr` can scale to larger dimensional problems with practical computation times.
-Specifically, we ultimately generate $B=1,000$ random vectors for $d=\{100, 250, 500, 1000, 2500, 5000, 10000\}$ for each correlation type, $\{Pearson, Spearman, Kendall\}$ while timing the algorithm's major steps.
-
-To mimic the workflow, we first produce a synthetic "data set" by completing the following steps:
-
-1.
-Produce heterogeneous gamma marginals by randomly selecting the $j^{th}$ gamma shape parameter from $U_j \sim uniform(1,10), j=1,\ldots,d$ and the $j^{th}$ rate parameter from $V_j \sim exp(1/5), j=1,\ldots,d$, with the constant parameters determined arbitrarily.
-2. Produce a random full-rank Pearson correlation matrix via `cor_randPD` of size $d \times d$.  
-3. Simulate a "data set" of $1,000 \times d$ random vectors via `rvec` (without matching the Pearson exactly).
-
-With the synthetic data set in hand, we complete and time the following 4 steps:
-
-i. Estimate the correlation matrix from the "data" in the *Compute Correlation* step.  
-1. Map the correlations to initialize the algorithm (Pearson to Pearson, Spearman to Pearson, or Kendall to Pearson) in the *Adjust Correlation* step.  
-2. Check whether the mapping produces a valid correlation matrix and, if not, find the nearest PD correlation matrix in the *Check Admissibility* step.  
-3. Simulate $1,000$ vectors via the Inverse Transform method in the *Simulate Data* step.
-
-The experiments are conducted on a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, with all 16 threads employed during computation.
-The results for $d \leq 500$ are fast with all times under 3 seconds.
-Table \@ref(tab:ch040-moderateDtab) displays the total computation time of the algorithms steps 1 through 3, including estimation step i.
-
-```{r ch040-moderateDtab}
-dat_long <- benchmark_dependences %>%
-  select(-total_time, -n_sim, -needed_near_pd) %>%
-  pivot_longer(cols = c(corr_time:sim_time), names_to = "Step", values_to = "Time") %>%
-  mutate(dim = factor(dim),
-         Step = factor(Step,
-                       levels = c("corr_time", "adj_time", "admiss_time", "sim_time"),
-                       labels = c("Compute Correlation",
-                                  "Adjust Correlation",
-                                  "Check Admissibility",
-                                  "Simulate Data"))) %>%
-    rename(Correlation = corr_type, Dimensions = dim)
-
-tabTimes  <- dat_long %>%
-    filter(Dimensions %in% c(100, 250, 500)) %>%
-    group_by(Dimensions, Correlation) %>%
-    summarize('Total Time(Seconds)' = sum(Time)) %>%
-    ungroup()
-
-tabTimes %>%
-    kable(booktabs = TRUE,
-          col.names = c("Dimension",
-                        "Correlation type",
-                        "Total Time (Seconds)"),
-          caption = 'Total time to produce 10,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.')
-```
-
-The results larger dimensional vectors show scalability to ultra-high dimensions for all three correlation types, although the total times do become much larger. 
-Figure \@ref(fig:ch040-largeDfig) displays computation times for $d=\{1000, 2500, 5000, 10000\}$.
-For $d$ equal to 1000 and 2500, the total time is under a couple of minutes.
-At $d$ of 5000 and 10,000, Pearson correlation matching in the *Adjust Correlation* step becomes costly.
-Interestingly, Pearson is actually faster than Kendall for $d=10,000$ due to bottlenecks in *Compute Correlation* and *Check Admissibility*.
-Uniformly, matching Spearman correlations is faster, with total times under 5 minutes for $d=10,000$, making Spearman correlations the most computationally-friendly dependency type.
-With this in mind, we scaled the simulation to $d=20,000$ for the Spearman type and obtained the 1,000 vectors in under an hour (data not shown).
-In principle, this would enable the simulation of an entire human-derived RNA-seq data set.
-We note that for a given target correlation matrix and margins, steps i, 1, and 2 only need to be computed once and the third step, *Simulate Data*, is fast under all schemes for all $d$ under consideration.
-
-```{r ch040-largeDfig, fig.cap="Computation times as d increases."}
-# Set colors
-numColors <- 5
-numGroups <- 4
-myColors <- rev(brewer.pal(n=numColors, name="Greys")[((numColors-numGroups)+1):numColors])
-
-dat_long %>%
-    filter(Dimensions %in% c(1000, 2500, 5000, 10000, 20000)) %>%
-    mutate(`Time (minutes)` = Time / 60) %>%
-    ggplot(aes(Correlation, `Time (minutes)`, fill=Step)) +
-    geom_bar(position = "stack", stat = "identity") +
-    scale_fill_manual(values = myColors ) +
-    scale_y_continuous(breaks = c(0, 5, 10, 15, 20, 25, 30),
-                     minor_breaks = NULL) +
-    facet_grid(. ~ Dimensions) +
-    theme(axis.text.x = element_text(angle = -90))
-```
-
-*Limitations, conclusions, and recommendations.*
-In the bivariate studies, we chose arbitrary simulation parameters for three distributions, moving from the Gaussian to the discrete and non-normal, multivariate negative binomial.
-Under these conditions, the simulated random vectors sample the desired bivariate distribution across the entire range of pairwise correlations for the three dependency measures.
-The simulation results could differ for other choices of simulation settings.
-Specifying extreme correlations near the boundary or Frechet bounds could result in poor simulation performance.
-Fortunately, it is straightforward to evaluate simulation performance by using strategies similar to those completed above.
-We expect our random vector generation to perform well for the vast majority of NORTA-feasible correlation matrices, but advise to check the performance before making inferences/further analyses.
-Finally, we note that one could use our single-pass algorithm `cor_fastPD` to produce a "close" (not nearest PD) to scale to even higher dimensions with greater loss of accuracy.
-```{r include=FALSE, cache=FALSE}
-# Remove all objects
-rm(list = ls(all.names = TRUE))
-
-# Unload all packages
-```
-
-<!--chapter:end:040-monte-carlo-evaluation.Rmd-->
-
-```{r include=FALSE, cache=FALSE}
-options(digits = 4)
-
-knitr::opts_chunk$set(
-  # Text Results
-  eval = TRUE,
-  echo = FALSE,
-  results = 'markup',
-  collapse = TRUE,
-  message = FALSE,
-  warning = FALSE,
-  error = FALSE,
-  # Code Decoration
-  comment = NA,
-  background = '#F7F7F7',
-  # Cache
-  cache = TRUE,
-  # Plots
-  fig.show = "hold",
-  fig.width = 6.5,
-  fig.asp = 0.618,  # 1 / phi
-  out.width = "85%",
-  fig.align = 'center'
-)
-
-# options(dplyr.print_min = 6, dplyr.print_max = 6)
-
-seed_number <- 2020-02-25
-set.seed(seed = seed_number)
-```
-# RNA-seq data applications  {#examples}
-
-```{r ch050-LoadLib, include=FALSE, cache=FALSE}
-box::use(
-  dplyr[...],
-  ggplot2[...],
-  GGally[ggpairs, wrap],
-  patchwork[...],
-  bigsimr[...],
-  knitr[include_graphics],
-  JuliaCall[julia_call],
-  ./R/utils[mom_nbinom]
-)
-load("data/example_brca.rda")
-load("data/example_genes.rda")
-sim_nbinom <- readRDS("results/brca_1000_sim.rds")
-simRho <- readRDS("results/brca_1000_frobenius.rds")
-
-Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
-bs <- bigsimr_setup(pkg_check = FALSE)
-dist <- distributions_setup()
-```
-
-This section demonstrates how to simulate multivariate data using `bigsimr`, aiming to replicate the structure of high-dimensional dependent count data.
-In an illustration of our proposed methodology, we seek to simulate RNA-sequencing data by producing simulated random vectors mimicking the observed data and its generating process.
-Modeling RNA-seq using multivariate probability distributions is natural as inter-gene correlation is an inherent part of biological processes [@Wang2009b].
-And yet, many models do not account for this, leading to major disruptions to the operating characteristics of statistical estimation, testing, and prediction.
-See @Efron2012 for a detailed discussion with related methods and @Wu2012b, @Schissler2018; @Schissler2019 for applied examples.
-The following subsections apply `bigsimr`'s methods to real RNA-seq data, including replicating an estimated parametric structure, MC probability estimation, and MC evaluation of correlation estimation efficiency.
-
-## Simulating High-Dimensional RNA-seq data
-
-```{r ch050-readBRCA, cache=FALSE}
-d <- 1000
-brca1000 <- example_brca %>%
-  select(all_of(1:d)) %>%
-  mutate(across(everything(), as.double))
-```
-
-We begin by estimating the structure of the TCGA BRCA RNA-seq data set (see [Background](background)).
-Ultimately, we will simulate $B=10,000$ random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with $d=`r d`$.
-We assume a multivariate negative binomial (MVNB) model as RNA-seq counts are often over-dispersed and correlated.
-Since all $d$ selected genes exhibit over-dispersion (data not shown), we proceed to estimate the NB parameters $(r_i, p_i), i=1,\ldots,d$, to determine the target marginal PMFs $f_i$.
-To complete specification of the simulation algorithm inputs, we estimate the Spearman correlation matrix ${ \bf R}_{S}$ to characterize dependency.
-
-<!--
-This suggests simulating high-dimensional multivariate NB (MVB) with heterogeneous marginals would be useful tool in the development and evaluation of RNA-seq analytics.
-This procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions.
-All these genes exhibit over-dispersion and, so, we proceed to estimate the NB parameters $(r_i, p_i), i=1,\ldots,d$ to determine the target marginal PMFs $g_i(y_i)$ (via method of moments).
-Notably, the $\hat{p}_i's$ are small --- ranging in $[3.934 \times 10^{-6} , 1.217 \times 10^{-2}]$.
-To complete the simulation algorithm inputs, we estimate the Pearson correlation matrix $\bf{R_Y}$ and set that as the target correlation.
--->
-
-<!--
-With the simulation targets specified, we proceed to simulate $N=10,000$ random vectors $\bf{Y}$ $=( Y_1,\ldots,Y_d)$ with target Pearson correlation $\bf{R_Y}$ and marginal PMFs $g_i(y_i)$ using a $\bf{T}$-Poisson hierarchy of Kind II.
-Specifically, we first employ the \emph{direct Gaussian copula} approach to generate $N$ random vectors following a standard multivariate Gamma distribution $\bf{T}$ with shape parameters $r_i$ equal to the target $n_i$ and Pearson correlation matrix $\bf{R_T}$.
-Care must be taken when setting the specifying $\bf{R}$ (refer to Equation \ref{gay.cop.pdf}) --- we employ Equation \ref{mix.poi.corr} to compute the scaling factors $c_{i,j}$ and adjust the underlying correlations to ultimately match the target $\bf{R_Y}$.
-Notably, of the $525,825$ pairwise correlations from the $1026$ genes, no scale factor was less than $0.9907$, indicating the model can produce essentially the entire range of possible correlations.
-Here we are satisfied with approximate matching of the specified Gamma correlation and set $\bf{R}$ = $\bf{R_T}$ in our Gaussian copula scheme ($\bf{R}$ indicating the specified multivariate Gaussian correlation matrix).
-Finally, we generate the desired random vector $Y_i=N_i(t_i)$ by simulating Poisson counts with expected value $\mu_i=\lambda_i \times T_i$, for $i=1,\ldots,d$ (with $\lambda_i=\frac{(1-p_i)}{p_i}$) and repeat $N=10,000$ times.
--->
-
-With this goal in mind, we first estimate the desired correlation matrix using the fast implementation provided by `bigsimr`:
-
-```{r ch050-estRhoBRCA, echo=TRUE, cache=FALSE}
-# Estimate Spearman's correlation on the count data
-R_S <- bs$cor(as.matrix(brca1000), bs$Spearman)
-```
-
-Next, we estimate the marginal parameters.
-We use the method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model.
-While, marginal distributions are from the same probability family (NB), they are heterogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.
-The functions below support this estimation for later use in `rvec`.
-
-<!-- 
-TYPESTE MoM estimators here.
--->
-
-```{r ch050-nbHelpers, echo=TRUE}
-make_nbinom_margins <- function(sizes, probs) {
-  margins <- lapply(1:length(sizes), function(i) {
-    dist$NegativeBinomial(sizes[i], probs[i])
-  })
-  do.call(c, margins)
-}
-```
-
-We apply these estimators to all `r d` genes across the `r nrow(brca1000)` patients:
-
-```{r ch050-estMargins, echo=TRUE}
-nbinom_fit <- apply(brca1000, 2, mom_nbinom)
-sizes <- nbinom_fit["size",]
-probs <- nbinom_fit["prob",]
-nb_margins <- make_nbinom_margins(sizes, probs)
-```
-
-Notably, the estimated marginal NB probabilities $\{ \hat{p}_i \}$ are small --- ranging in the interval $[`r min(probs)` , `r max(probs)`]$.
-This gives rise to highly variable counts and, typically, less restriction on potential pairwise correlation pairs.
-Once the functions are defined/executed to complete marginal estimation, we specify targets and generate the desired random vectors using `rvec`.
-Now we check admissibility of specified correlation matrix.
-
-```{r ch050-cor-check, echo=TRUE}
-# 1. Mapping step first
-R_X <- bs$cor_convert(R_S, bs$Spearman, bs$Pearson)
-
-# 2a. Check admissibility
-(is_valid_corr <- julia_call("Bigsimr.iscorrelation", R_X))
-
-# 2b. compute nearest correlation
-if (!is_valid_corr) {
-  R_X_pd <- bs$cor_nearPD(R_X)
-  ## Quantify the error
-  targets      <- R_X[lower.tri(R_X, diag = FALSE)]
-  approximates <- R_X_pd[lower.tri(R_X_pd, diag = FALSE)]
-  R_X          <- R_X_pd
-}
-summary(abs(targets - approximates))
-```
-
-While the exact $d \times d$ Spearman correlation matrix is not strictly admissible in our scheme (as seen by the non-positive definite result above), the approximation is close with a maximum absolute error of `r max( abs(targets - approximates))` and average absolute error of `r mean(abs(targets - approximates))` across the `r choose(d,2)` correlations.
-
-```{r ch050-runBRCA-echo, echo=TRUE, eval=FALSE}
-sim_nbinom <- bs$rvec(10000, R_X, nb_margins) 
-colnames(sim_nbinom) <- colnames(brca1000)
-```
-
-Figure \@ref(fig:ch050-simDataFig) displays the simulated counts and pairwise relationships for our example genes from Table \@ref(tab:ch010-realDataTab).
-Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parametric form and with less extreme points then the observed data in Figure \@ref(fig:ch010-realDataFig).
-
-```{r ch050-simDataFig, cache=FALSE, fig.cap="Simulated data for three selected high-expressing genes generally replicates the estimated data structure.
-The data do not exhibit outlying points, but do possess the desired Spearman correlations, central tendencies, and discrete values."}
-set.seed(2020-02-25)
-num_genes <- 3
-gene_sample <- sample(example_genes[1:1000], num_genes)
-
-ggpairs(data = as.data.frame(sim_nbinom[, gene_sample]),
-        upper = list(continuous = wrap('cor', method = "spearman"))) + 
-  theme_bw()
-```
-
-Figure \@ref(fig:ch050-figBRCA) displays the aggregated results of our simulation by comparing the specified target parameter (horizontal axes) with the corresponding quantities estimated from the simulated data (vertical axes).
-The evaluation shows that the simulated counts approximately match the target parameters and exhibit the full range of estimated correlation from the data.
-
-<!-- 
-Utilizing 15 CPU threads in a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, the simulation completed in just over 2 minutes.
--->
-
-```{r ch050-figBRCA, out.width='80%', fig.cap="Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set.
-The dashed red lines indicate equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes)."}
-include_graphics('fig/ch050-figBRCA.png')
-```
-
-*Limitations, conclusions, and recommendations.*
-The results show overall good simulation performance for our choice of parameters settings.
-Our settings were motivated by modeling high-expressing genes from TCGA BRCA data set.
-In general, the ability to match marginal and dependence parameters depends on the particular joint probability model.
-We recommend to evaluate and tune your simulation until you can be assured of the accuracy.
-
-## Simulation-based joint probability calculations
-
-Many statistical tasks require evaluation of a joint probability mass (or density) function:
-
-$$
-P( {\bf Y} = {\bf y} ), \: y_i \in \chi_i.
-$$
-
-where $\chi_i$ is the sample space for the $i^{th}$ component of the random vector $\bf{Y}$.
-Compact representations with convenient computational forms are rare for high-dimensional constructions, especially with heterogeneous, correlated marginal distributions (or margins of mixed data types).
-Given a large number of simulated vectors as produced above, estimated probabilities are readily given by counting the proportion of simulated vectors meeting the desired condition.
-In our motivating application, one may ask what is the probability that all genes expressed greater than a certain threshold value ${ \bf y}_0$.
-This can be estimated as
-
-$$
-\hat{P}( {\bf Y} \ge {\bf y_0 } ) = \frac{1}{B} \sum_{b=1}^B I( {\bf Y}^{ (b) } \ge {\bf y_0 } ),
-$$
-
-```{r ch050-densityEvaluation, include=FALSE}
-d <- ncol(sim_nbinom)
-B <- nrow(sim_nbinom)
-y0 <- 1
-pHat <- mean(apply(sim_nbinom, 1, function(Y) {
-  all(Y >= y0)
-}))
-```
-
-where ${\bf Y^{(b)} }$ is the $b^{th}$ simulated vector from a total of $B$ simulation replicates and $I(\cdot)$ is the indicator function.
-For example, we can estimate from our $B=10,000$ simulated vectors that the probability of all genes expressing (i.e., ${\bf y}_i \geq 1, \forall \; i$) is $`r pHat`$.
-
-```{r ch050-densityEvaluationECHO, echo=TRUE}
-<<ch050-densityEvaluation>>
-pHat
-```
-
-## Evaluation of correlation estimation efficiency
-
-MC methods are routinely used in many statistical inferential tasks including estimation, hypothesis testing, error rates, and empirical interval coverage rates.
-To conclude the example applications, we demonstrate how `bigsimr` can be used to evaluate estimation efficiency.
-In particular, we wish to assess the error in our correlation estimation above.
-We used a conventional method, based on classical statistical theory which was not designed for high-dimensional data.
-Indeed, high-dimensional covariance estimation (and precision matrices) is an active area of statistical science, .e.g., [@Won2013g; @VanWieringen2016].
-
-In this small example, we simulate $m=30$ data sets with the number of simulated vectors matching the number of patients in the BRCA data set, $N=`r nrow(example_brca)`$.
-Since our simulation is much faster for the Pearson correlation type (see Figure \@ref(fig:ch040-largeDfig)), we only convert the Spearman correlation matrix once (and ensure it is positive definite).
-At each iteration, we estimate the quadratic loss (residual sum of squared errors) from the specified ${\bf R}_{S}$, producing a distribution of loss values.
-
-```{r ch050-quadlossECHO, echo=TRUE, eval=FALSE}
-# Simulate random vectors equal to the sample size
-N <- nrow(example_brca)
-# create m random vectors and estimate correlation
-simRho <- replicate(n = m, expr = {
-  tmpSim <- bs$rvec(N , R_X, nb_margins)
-  bs$cor(tmpSim, bs$Spearman)
-}, simplify = FALSE)
-# Evaluate the residual sum of squared error
-sapply(simRho, function(R) sum((R - R_S)^2))
-```
-
-```{r ch050-lossResults}
-frobenius_loss <- sapply(simRho, function(R) sum((R - R_S)^2))
-```
-
-The `R` summary function supplies the mean-augmented five-number summary of the quadratic loss distribution computed above.
-
-```{r ch050-loss-summary, echo=TRUE}
-summary(frobenius_loss)
-```
-
-This distribution could be compared to high-dimensional designed covariance estimators for guidance on whether the additional complexity and computation time are warranted.
-```{r include=FALSE, cache=FALSE}
-# Remove all objects
-rm(list = ls(all.names = TRUE))
-
-# Unload all packages
-```
-
-<!--chapter:end:050-examples.Rmd-->
-
-```{r include=FALSE, cache=FALSE}
-options(digits = 4)
-
-knitr::opts_chunk$set(
-  # Text Results
-  eval = TRUE,
-  echo = FALSE,
-  results = 'markup',
-  collapse = TRUE,
-  message = FALSE,
-  warning = FALSE,
-  error = FALSE,
-  # Code Decoration
-  comment = NA,
-  background = '#F7F7F7',
-  # Cache
-  cache = TRUE,
-  # Plots
-  fig.show = "hold",
-  fig.width = 6.5,
-  fig.asp = 0.618,  # 1 / phi
-  out.width = "85%",
-  fig.align = 'center'
-)
-
-# options(dplyr.print_min = 6, dplyr.print_max = 6)
-
-seed_number <- 2020-02-25
-set.seed(seed = seed_number)
-```
-# Conclusion and discussion {#discussion}
-
-We have introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a user-friendly, high-performance `R` package [`bigsimr`].
-The random vector generation method is inspired by NORTA [@Cario1997] and Gaussian copula-based approaches [@MB13, @BF17, @Xia17].
-The major contributions of this work are methods and software for flexible, scalable simulation of HD multivariate probability distributions with broad potential data analytic applications for modern, big-data statistical computing.
-For example, one could simulate high-resolution time series data, such as those consistent with an auto-regressive moving average model exhibiting a specified Spearman structure.
-Or our methods could be used to simulate sparsely correlated data, as many HD methods assume, via specifying a *spiked correlation matrix*.
-
-Researchers working in multivariate computation frequently encounter such difficulties and need to find a close or nearest correlation matrix.
-A widely-available routine for this task in `R` is `matrix::nearPD`, though it is not suitable for high dimensions.
-To overcome this issue, we introduce `cor_nearPD` --- an implementation of @QS2006's method that computes the nearest correlation matrix to a given symmetric matrix in the Frobenius norm.
-This routine as many applications beyond our primary goal of random vector generation.
-
-It is customary to compare new tools and algorithms directly to existing competing methods and software.
-In this study, however, we only employ our proposed methodology, as our previous work has shown that existing `R` tools are simply not designed to meet our high-dimensional goal (see @Li2019gpu for evaluations of the `R` `copula` package and others).
-For the bivariate simulations, existing packages such as [`nortaRA`](https://github.com/cran/NORTARA/blob/master/inst/doc/NORTARA.R) work well to match Pearson correlations exactly.
-
-<!-- 
-We advocate the use of Kendall's $\tau$ as it better captures correlation among components of non-normal, as well as non-linear patterns of association.
-Moreover, the matching Kendall's $\tau$ is nearly trivial due to the invariant of monotone transformations, after an adjustment to the input correlation matrix.
--->
-
-<!-- 
-We also show utility in our methodology through an application to differential gene expression analysis from RNA-sequencing data.
-The application results show that correlations indeed matter in the large-scale hypothesis testing, as many others have noted (for example, see @BE07, @Wu2012b).
-We also hope that the application provides an example workflow --- and other strategy to use in simulation design.
-Even the best-performing simulations we provide show a gap from the empirical distribution.
-To keep the demonstration straightforward, we did not attempt to match the empirical distribution of test statistics as precisely as possible.
-Yet our methodology could be more creatively applied to meet that goal.
-One could consider marginal distributions from different families, such as Poisson for a subset of genes.
-One could imagine finding a best fitting probability distribution among a class of distributions for each gene and this could perhaps a better fit.
-One could imagine additional structures/features in the data that an analyst could model to improve the correspondence with the empirical values, for example row correlations and high-dimensional covariance estimators (see @Won2013g).
--->
-
-There are limitations to the methodology and implementation.
-We could only investigate selected multivariate distributions in our Monte Carlo studies.
-There may be instances that the methods do not perform well.
-Along those lines, we expect that correlation values close to the boundary of the feasible region could result in algorithm failure.
-Another issue is that for discrete distributions, we use continuous approximations when the support set is large.
-This could limit the scalability/accuracy for particular joint distributions.
-
-<!-- 
-The most obvious missing feature of the proposed methodology is the inability to match a Pearson correlation matrix exactly.
-As discussed in the [Algorithms](algorithms) section and extensively by @XZ19, this is a computationally intense procedure and Pearson's correlation is not a natural choice to describe dependency for non-normal marginals.
-While we do not provide an implementation directly supporting Pearson matching, users may supply their own input Pearson correlation after employing a supplementary matching scheme [@Cario1997; @XZ19].
--->
-
-<!-- 
-Further, while we provide the ability of the user to specify discrete marginals, exact matching a desired dependency measure is not yet obtained.
-One potential solution for this is *rescale* to account for ties [ref].
-Finally we note, that the algorithm requires invertible marginal cdfs.
-This gives some restriction to the available marginal probability distributions (DOES IT?
-FOR EXAMPLE?).
-From a practical computing standpoint, the user's computing resource provides the ultimate limit on how large a dimension can be simulated using the `bigsimr` R package.
-A user must consider carefully the available memory and cores when conducting a Monte Carlo experiment.
--->
-
-<!-- 
-Future work includes developing scaleable algorithms to match the Pearson correlation matrix more precisely, discrete-margin specific modifications including fast Spearman's correlation rescaling (see Equation \@ref(eq:spearmanRescaled)), and high-dimensional covariance estimation.
-From an implementation standpoint, `bigsimr` only supports Nvidia GPUs and redesigning the code using OpenCL would broaden the users who would benefit.
-As data-analytic problems grow to even larger dimension, multi-GPU support is a promising hardware-based future direction.
--->
-```{r include=FALSE, cache=FALSE}
-# Remove all objects
-rm(list = ls(all.names = TRUE))
-
-# Unload all packages
-```
-
-<!--chapter:end:060-conclusion-discussion.Rmd-->
-
-```{r include=FALSE, cache=FALSE}
-options(digits = 4)
-
-knitr::opts_chunk$set(
-  # Text Results
-  eval = TRUE,
-  echo = FALSE,
-  results = 'markup',
-  collapse = TRUE,
-  message = FALSE,
-  warning = FALSE,
-  error = FALSE,
-  # Code Decoration
-  comment = NA,
-  background = '#F7F7F7',
-  # Cache
-  cache = TRUE,
-  # Plots
-  fig.show = "hold",
-  fig.width = 6.5,
-  fig.asp = 0.618,  # 1 / phi
-  out.width = "85%",
-  fig.align = 'center'
-)
-
-# options(dplyr.print_min = 6, dplyr.print_max = 6)
-
-seed_number <- 2020-02-25
-set.seed(seed = seed_number)
-```
-# Supplementary Materials {.unnumbered #misc}
-
-We provide an open-source implementation of our methodology as the `bigsimr` R package, hosted on github, https://github.com/SchisslerGroup/r-bigsimr.
-
-```{r, out.width='5%'}
-knitr::include_graphics('images/hex-bigsimr.png')
-```
-
-# Acknowledgment(s) {.unnumbered #acknowledgments}
-
-The authors gratefully acknowledge Heather Knudson's graphic design for the `bigsimr` R Package.
-The results published here are in whole or part based upon data generated by the TCGA Research Network: https://www.cancer.gov/tcga.
-
-# Disclosure statement {.unnumbered #coi}
-
-The authors report no conflict of interest.
-
-# Funding {.unnumbered #funding}
-
-Research reported in this publication was supported by grants from the National Institutes of General Medical Sciences (5 U54 GM104944, GM103440) from the National Institutes of Health.
-
-<!-- 
-Nomenclature/Notation
--->
-```{r include=FALSE, cache=FALSE}
-# Remove all objects
-rm(list = ls(all.names = TRUE))
-
-# Unload all packages
-```
-
-<!--chapter:end:070-supplementary-materials.Rmd-->
-
-```{r include=FALSE, cache=FALSE}
-options(digits = 4)
-
-knitr::opts_chunk$set(
-  # Text Results
-  eval = TRUE,
-  echo = FALSE,
-  results = 'markup',
-  collapse = TRUE,
-  message = FALSE,
-  warning = FALSE,
-  error = FALSE,
-  # Code Decoration
-  comment = NA,
-  background = '#F7F7F7',
-  # Cache
-  cache = TRUE,
-  # Plots
-  fig.show = "hold",
-  fig.width = 6.5,
-  fig.asp = 0.618,  # 1 / phi
-  out.width = "85%",
-  fig.align = 'center'
-)
-
-# options(dplyr.print_min = 6, dplyr.print_max = 6)
-
-seed_number <- 2020-02-25
-set.seed(seed = seed_number)
-```
-# Appendix {.unnumbered #appendix}
-
-\noindent Consider a bivariate example where we have a correlated $(Y_1, Y_2)^\top$ with $Y_1\sim N(\mu_1, \sigma_1^2)$ (normal) and $Y_2\sim LN(\mu_2, \sigma_2^2)$ (lognormal).
-First, let us note that when we get such a vector from bigsimr package then in fact it can be represented as
-
-\begin{equation}
-\label{kram1}
-(Y_1, Y_2)^\top \stackrel{d}{=} \left(X_1, e^{X_2}\right)^\top,
-\end{equation}
-
-where $(X_1, X_2)^\top \sim N_2(\boldsymbol \mu, \boldsymbol \Sigma)$, which is bivariate normal with mean vector $\boldsymbol \mu = (\mu_1, \mu_2)^\top$ and variance-covariance matrix
-
-\begin{equation}
-\label{kram2}
-\boldsymbol \Sigma = 
-\left[
-\begin{array}{cc}
-\sigma_1^2 & \rho \sigma_1\sigma_2\\
-\rho \sigma_1\sigma_2 & \sigma_2^2
-\end{array}
-\right].
-\end{equation}
-
-To see this, consider the three steps of the NORTA construction:
-
-\begin{enumerate}
-
-\item Generate $(Z_1, Z_2)^\top \sim N_2(\boldsymbol 0, \boldsymbol R)$, where
-
-\begin{equation}
-\label{kram3}
-\boldsymbol R = 
-\left[
-\begin{array}{cc}
-1 & \rho \\
-\rho & 1
-\end{array}
-\right].
-\end{equation}
-
-\item Transform $(Z_1, Z_2)^\top$ to $(U_1, U_2)^\top$ viz. $U_i =\Phi(Z_i)$,  $i=1,2$, where $\Phi(\cdot)$ is the standard normal CDF.
-
-\item Return $(Y_1, Y_2)^\top$, where $Y_i=F_i^{-1}(U_i)$, $i=1,2$, and $F_i(\cdot)$ is the CDF of $Y_i$ and $F_i^{-1}(\cdot)$ is its inverse (the quantile function).
-
-\end{enumerate}
-
-In this example, $F_1$ is the CDF of $N(\mu_1, \sigma_1^2)$ while $F_2$ is the CDF of $LN(\mu_2, \sigma_2^2)$, so that the two required quantile functions are
-
-\begin{equation}
-\label{kram4}
-F_1^{-1}(u) = \mu_1+\sigma_1 \Phi^{-1}(u)\,\,\, \mbox{and} \,\,\, F_2^{-1}(u) = e^{\mu_2+\sigma_2 \Phi^{-1}(u)}, 
-\end{equation}
-
-as can be seen by standard calculation. Thus, when we apply Step 3 of the NORTA algorithm, we get
-
-\begin{equation}
-\label{kram5}
-F_1^{-1}(U_1) = \mu_1+\sigma_1 \Phi^{-1}(\Phi(Z_1)) =  \mu_1+\sigma_1 Z_1 \,\,\, \mbox{and} \,\,\, F_2^{-1}(U_2) = e^{\mu_2+\sigma_2 \Phi^{-1}(\Phi(Z_2))} = e^{\mu_2+\sigma_2 Z_2}, 
-\end{equation}
-
-where $(X_1, X_2)^\top =  (\mu_1+\sigma_1 Z_1, \mu_2+\sigma_2 Z_2)^\top \sim N_2(\boldsymbol \mu, \boldsymbol \Sigma)$.
-Consequently, the vector $(Y_1, Y_2)^\top$ obtained in Step 3 of the algorithm has the structure provided by (\ref{kram1}).
-
-\vspace{0.1in}
-
-\noindent Next, we provide the exact covariance structure of the random vector $(Y_1, Y_2)^\top$ given by (\ref{kram1}), and relate the correlation of $Y_1$ and $Y_2$ to $\rho$, which is the correlation of the normal variables $Z_1$, $Z_2$ (and also $X_1$ and $X_2$).
-A straightforward albeit somewhat tedious algebra produces the following result.
-
-\begin{lemma}
-Let ${\bf Y} = (Y_1, Y_2)^\top$ admit the stochastic representation (\ref{kram1}), where ${\bf X} = (X_1, X_2)^\top \sim N_2(\boldsymbol \mu, \boldsymbol \Sigma)$ with $\boldsymbol \mu$ and $\boldsymbol \Sigma$ as above.
-Then the mean vector and the variance-covariance matrix of ${\bf Y}$ are given by
-
-\begin{equation}
-\label{kram6}
-\boldsymbol \mu_{{\bf Y}} = \left(\mu_1, e^{\mu_2+\sigma_2^2/2}\right)^\top
-\end{equation}
-
-and
-
-\begin{equation}
-\label{kram7}
-\boldsymbol \Sigma = 
-\left[
-\begin{array}{cc}
-\sigma_1^2 & \rho \sigma_1\sigma_2  e^{\mu_2+\sigma_2^2/2}  \\
-\rho \sigma_1\sigma_2 e^{\mu_2+\sigma_2^2/2} & \left[ e^{\mu_2+\sigma_2^2/2}\right]^2 \left[e^{\sigma_2^2} -1 \right],
-\end{array}
-\right].
-\end{equation}
-
-respectively. 
-\end{lemma}
-
-\begin{proof}
-The values of the means and the variances are obtained immediately from normal marginal distribution of $Y_1$ and lognormal marginal distribution of $Y_2$.
-It remains to establish the covariance of $Y_1$ and $Y_2$,
-
-\begin{equation}
-\label{covy1y2}
-\mbox{Cov}(Y_1, Y_2) = \mathbb E(Y_1 Y_2) - \mathbb E(Y_1)\mathbb E(Y_2), 
-\end{equation}
-
-and in particular the first expectation on the right-hand-side above. By using the tower property of expectations, the latter expectation can be expressed as
-
-\begin{equation}
-\label{kre1}
-\mathbb E(Y_1 Y_2)  =  \mathbb E\left(X_1 e^{X_2}\right)  =  \mathbb E \left\{ \mathbb E \left(X_1 e^{X_2} | X_1 \right) \right\} = \mathbb E \left\{ X_1 \mathbb E \left(e^{X_2} | X_1 \right) \right\}.
-\end{equation}
-
-Further, by using the fact that the conditional distribution of $X_2$ given $X_1=x_1$ is normal with mean $\mathbb E(X_2|X_1=x_1) = \mu_2+\rho\sigma_2(x_1-\mu_1)/\sigma_1$ and variance $\sigma_2^2(1-\rho^2)$, one can relate the inner expectation on the far right in (\ref{kre1}) to the moment generating function $M(t)$ of this conditional distribution evaluated at $t=1$, leading to
-
-\begin{equation}
-\label{kre2}
-E \left(e^{X_2} | X_1 \right) = e^{ \mu_2+\rho\sigma_2(x_1-\mu_1)/\sigma_1 + \sigma_2^2(1-\rho^2)/2},
-\end{equation}
-
-so that
-
-\begin{equation}
-\label{kre3}
-\mathbb E(Y_1 Y_2)  =  e^{ \mu_2+ \frac{1}{2} \sigma_2^2(1-\rho^2)} \mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\}.
-\end{equation}
-
-Since $X_1$ is normal with mean $\mu_1$ and variance $\sigma_1^2$, the expectation in (\ref{kre3}) becomes
-
-\begin{equation}
-\label{kre4}
-\mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = \int_{-\infty}^\infty x_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (x_1-\mu_1)} \frac{1}{\sqrt{2\pi}\sigma_1} e^{-\frac{1}{2\sigma_1^2}(x_1-\mu_1)^2}.
-\end{equation}
-
-Upon substituting $u=x_1-\mu_1$, followed by some algebra, we arrive at
-
-\begin{equation}
-\label{kre5}
-\mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = e^{\frac{1}{2}\sigma_2^2\rho^2} \int_{-\infty}^\infty (u+\mu_1)  \frac{1}{\sqrt{2\pi}\sigma_1} e^{-\frac{1}{2\sigma_1^2}(u-\rho\sigma_1\sigma_2)^2}.
-\end{equation}
-
-Since the integral in (\ref{kre5}) is the expectation $\mathbb E(X+\mu_1)$, where $X$ is normal with mean $\rho\sigma_1\sigma_2$ and variance $\sigma_1^2$, we conclude that
-
-\begin{equation}
-\label{kre6}
-\mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = e^{\frac{1}{2}\sigma_2^2\rho^2} (\rho\sigma_1\sigma_2 +\mu_1),
-\end{equation}
-
-which, in view of (\ref{kre3}), leads to
-
-\begin{equation}
-\label{kre7}
-\mathbb E (Y_1Y_2) = e^{\mu_2 + \frac{1}{2}\sigma_2^2} (\rho\sigma_1\sigma_2 +\mu_1).
-\end{equation}
-
-Finally, (\ref{covy1y2}), along with the expressions for the means of $Y_1$ and $Y_2$, produce the covariance of $Y_1$ and $Y_2$:
-
-\begin{equation}
-\label{kre8}
-\mbox{Cov}(Y_1, Y_2) = e^{\mu_2 + \frac{1}{2}\sigma_2^2} (\rho\sigma_1\sigma_2 +\mu_1) - \mu_1 e^{\mu_2 + \frac{1}{2}\sigma_2^2} = \rho\sigma_1\sigma_2 e^{\mu_2 + \frac{1}{2}\sigma_2^2}.
-\end{equation}
-
-This concludes the proof
-\end{proof}
-
-Using the above result, we can directly relate the correlation coefficient of $Y_1$ and $Y_2$ with that of $X_1$ and $X_2$,
-
-\begin{equation}
-\label{kram8}
-\rho_{Y_1, Y_2} = \rho \frac{\sigma_2}{\sqrt{e^{\sigma_2^2} -1}},
-\end{equation}
-
-where $\rho$ is the correlation of $X_1$ and $X_2$.
-The above result is useful when studying the possible range of correlation of $Y_1$ and $Y_2$ in the above example.
-It can be shown that the factor on the far right in (\ref{kram8}) is a monotonically decreasing function of $\sigma_2$ on $(0,\infty)$, with the limits of 1 and 0 at zero and infinity, respectively.
-Thus, in principle, the range of correlation of $Y_1$ and $Y_2$ is the same as that of $X_1$ and $X_2$, as the factor on the far right in (\ref{kram8}) can be made arbitrarily close to 1.
-However, by changing this factor we may affect the marginal distributions of $Y_1$ and $Y_2$.
-It can be shown that if the marginal distributions of $Y_1$ and $Y_2$ are fixed, then the relation (\ref{kram8}) becomes
-
-\begin{equation}
-\label{kram9}
-\rho_{Y_1, Y_2} = \rho \sqrt{\frac{\log(1+c_2^2)}{c_2^2}},
-\end{equation}
-
-where $c_2=\sigma_{Y_2}/\mu_{Y_2}$ is the *coefficient of variation* (CV) of the variable $Y_2$.
-Thus, the range of possible correlations in this model is not affected by the distribution of $Y_1$, and is determined by the CV of $Y_2$ as follows:
-
-\begin{equation}
-\label{kram10}
-- \sqrt{\frac{\log(1+c_2^2)}{c_2^2}} \leq \rho_{Y_1, Y_2} \leq \sqrt{\frac{\log(1+c_2^2)}{c_2^2}}. 
-\end{equation}
-
-Plugging in the estimates of these quantities from Section \@ref(package), we see that
-
-\begin{equation}
-\label{kram11}
-\frac{\hat{\sigma}_2}{\sqrt{e^{\hat{\sigma}_2^2} -1}} = 0.881.
-\end{equation}
-
-Thus, the possible range of correlation becomes $(-0.881, 0.881)$. This approximately agrees with the MC results provided.
-```{r include=FALSE, cache=FALSE}
-# Remove all objects
-rm(list = ls(all.names = TRUE))
-
-# Unload all packages
-```
-
-<!--chapter:end:080-appendix.Rmd-->
-
-```{r include=FALSE, cache=FALSE}
-options(digits = 4)
-
-knitr::opts_chunk$set(
-  # Text Results
-  eval = TRUE,
-  echo = FALSE,
-  results = 'markup',
-  collapse = TRUE,
-  message = FALSE,
-  warning = FALSE,
-  error = FALSE,
-  # Code Decoration
-  comment = NA,
-  background = '#F7F7F7',
-  # Cache
-  cache = TRUE,
-  # Plots
-  fig.show = "hold",
-  fig.width = 6.5,
-  fig.asp = 0.618,  # 1 / phi
-  out.width = "85%",
-  fig.align = 'center'
-)
-
-# options(dplyr.print_min = 6, dplyr.print_max = 6)
-
-seed_number <- 2020-02-25
-set.seed(seed = seed_number)
-```
-`r if (knitr:::is_html_output()) '# References {-}'`
-```{r include=FALSE, cache=FALSE}
-# Remove all objects
-rm(list = ls(all.names = TRUE))
-
-# Unload all packages
-```
-
-<!--chapter:end:300-references.Rmd-->
-
----
-institution:
-  name: "University of Nevada, Reno"
-  department: "Mathematics and Statistics"
-  address:
-    line1: "1664 N. Virginia Street / 0084"
-    line2: "Reno, Nevada 89557"
-  phone: "(775) 784-6773"
-  fax: "(775) 784-6378"
-  url: https://www.unr.edu/math/
-  logo: "logos/blue.png"
-author:
-  first: "Alfred"
-  last: "Schissler"
-  short: "Grant"
-  title: "Ph.D."
-  position: "Assistant Professor, Statistics"
-  signature: "signature.jpg"
-to:
-  first: "Alexander"
-  last: "Knudson"
-  short: "Alex"
-date: "`r format(Sys.Date(), format='%d %B %Y')`"
-opening: "Dear CSDA Annals of Statistical Data Science Editor(s):"
-closing: "Sincerely,"
-output:
-  pdf_document:
-    template: templateSingleSignature.tex
----
-```{r include=FALSE, cache=FALSE}
-options(digits = 4)
-
-knitr::opts_chunk$set(
-  # Text Results
-  eval = TRUE,
-  echo = FALSE,
-  results = 'markup',
-  collapse = TRUE,
-  message = FALSE,
-  warning = FALSE,
-  error = FALSE,
-  # Code Decoration
-  comment = NA,
-  background = '#F7F7F7',
-  # Cache
-  cache = TRUE,
-  # Plots
-  fig.show = "hold",
-  fig.width = 6.5,
-  fig.asp = 0.618,  # 1 / phi
-  out.width = "85%",
-  fig.align = 'center'
-)
-
-# options(dplyr.print_min = 6, dplyr.print_max = 6)
-
-seed_number <- 2020-02-25
-set.seed(seed = seed_number)
-```
-
-As an invited speaker at CMStatistics 2020, I was delighted to learn of the opportunity to have our work included in the first issue of CSDA Annals of Statistical Data Science.
-
-On behalf of my co-authors, I'm pleased to submit our manuscript entitled *Simulating High-Dimensional Multivariate Data using the `Bigsimr` Package* for your consideration.
-We feel that the work is well suited to the indicated themes and hope that it will be considered a valuable contribution to this prestigious journal.
-
-I hope that everything is in order. If not, or if you have any additional needs, please do not hesitate to contact me. Thank you and all those involved in this process.
-
-Best regards,
-```{r include=FALSE, cache=FALSE}
-# Remove all objects
-rm(list = ls(all.names = TRUE))
-
-# Unload all packages
-```
-
-<!--chapter:end:letter_bigsimr_CSDA.Rmd-->
+<!-- Turn off double spaces -->
+\setstretch{1.0}

### 2021-04-01 article_bigsimr.tex

--- 2021-03-27 _article/article_bigsimr.tex
@@
 \author{
-\\
+Alfred G. Schissler\\University of Nevada \And Edward J.
+Bedrick\\University of Nevada \And Alexander D. Knudson\\University of
+Nevada \AND Tomasz J. Kozubowski\\University of Nevada \And Tin
+Nguyen\\University of Nevada \And Anna K. Panorska\\University of Nevada
+\AND Juli Petereit\\University of Nevada \And Walter W.
+Piegorsch\\University of Nevada \And Duc Tran\\University of Nevada
 }
-\title{Simulating High-Dimensional Multivariate Data using the \pkg{Bigsimr} Package}
-
-\Plainauthor{}
-\Plaintitle{Simulating High-Dimensional Multivariate Data using the Bigsimr Package}
+\title{Simulating High-Dimensional Multivariate Data using the
+\pkg{Bigsimr} Package}
+
+\Plainauthor{Alfred G. Schissler, Edward J. Bedrick, Alexander D.
+Knudson, Tomasz J. Kozubowski, Tin Nguyen, Anna K. Panorska, Juli
+Petereit, Walter W. Piegorsch, Duc Tran}
+\Plaintitle{Simulating High-Dimensional Multivariate Data using the
+Bigsimr Package}
 \Shorttitle{\pkg{Bigsimr}: Simulate High-Dimensional Data}
@@
 \Abstract{
-It is critical to realistically simulate data when employing Monte Carlo techniques and evaluating statistical methodology.
-Measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments.
-Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations.
-This article introduces the \textbackslash pkg\{Bigsimr\} R package.
-This high-level package provides a flexible and scalable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions and known Pearson, Spearman, or Kendall correlation matrix.
-\textbackslash pkg\{Bigsimr\} contains additional high-performance features, including multi-core algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix.
-Monte Carlo studies quantify the accuracy and scalability of our approach, up to \(d=10,000\).
-Finally, we demonstrate example applications enabled via \textbackslash pkg\{Bigsimr\} by applying these methods to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples.
+It is critical to realistically simulate data when employing Monte Carlo
+techniques and evaluating statistical methodology. Measurements are
+often correlated and high dimensional in this era of big data, such as
+data obtained through high-throughput biomedical experiments. Due to
+computational complexity and a lack of user-friendly software available
+to simulate these massive multivariate constructions, researchers resort
+to simulation designs that posit independence or perform arbitrary data
+transformations. This article introduces the
+\textbackslash pkg\{Bigsimr\} R package. This high-level package
+provides a flexible and scalable procedure to simulate high-dimensional
+random vectors with arbitrary marginal distributions and known Pearson,
+Spearman, or Kendall correlation matrix. \textbackslash pkg\{Bigsimr\}
+contains additional high-performance features, including multi-core
+algorithms to simulate random vectors, estimate correlation, and compute
+the nearest correlation matrix. Monte Carlo studies quantify the
+accuracy and scalability of our approach, up to \(d=10,000\). Finally,
+we demonstrate example applications enabled via
+\textbackslash pkg\{Bigsimr\} by applying these methods to a motivating
+dataset --- RNA-sequencing data obtained from breast cancer tumor
+samples.
 }
 
-\Keywords{multivariate simulation, high-dimensional data, nonparametric correlation, Gaussian copula, RNA-sequencing data, breast cancer}
-\Plainkeywords{multivariate simulation, high-dimensional data, nonparametric correlation, Gaussian copula, RNA-sequencing data, breast cancer}
+\Keywords{multivariate simulation, high-dimensional data, nonparametric
+correlation, Gaussian copula, RNA-sequencing data, breast cancer}
+\Plainkeywords{multivariate simulation, high-dimensional
+data, nonparametric correlation, Gaussian copula, RNA-sequencing
+data, breast cancer}
 
@@
 \Address{
+    Alfred G. Schissler\\
+    University of Nevada\\
+    1664 N Virginia St.\\
+Reno, NV 89557\\
+  E-mail: \email{aschissler@unr.edu}\\
+
                   }
@@
 % Pandoc header
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{amsfonts}
-\usepackage{amsthm}
-
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}[theorem]{Lemma}
-
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{ragged2e}
-
-\usepackage{setspace}
+
+\usepackage{amsmath} \usepackage{amssymb} \usepackage{amsfonts} \usepackage{amsthm} \newtheorem{theorem}{Theorem} \newtheorem{lemma}[theorem]{Lemma} \usepackage{booktabs} \usepackage{longtable} \usepackage{array} \usepackage{ragged2e} \usepackage{setspace}
+
+\begin{document}
+
+\newpage
+
 \setstretch{2.0}
-
-\begin{document}
-
-\newpage
 
@@
 
-Massive high-dimensional (HD) data sets are now commonplace in many areas of scientific inquiry.
-As new methods are developed for data, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of analyzing such proposed methodology --- such as false positive rates, statistical power, interval coverage, and robustness --- often with comparison to existing methods.
-Further, efficient simulation empowers statistical computing strategies, such as the parametric bootstrap \citep{Chernick2008} to simulate from a hypothesized null model, providing inference in analytically challenging settings.
-Such Monte Carlo (MC) techniques become difficult for high-dimensional data using existing algorithms and tools.
-This is particularly true when simulating massive multivariate, non-normal distributions, arising naturally in many fields of study.
-
-As many have noted, it can be vexing to simulate dependent, non-normal/discrete data, even for low dimensional settings \citep{MB13, XZ19}.
-For continuous non-normal multivariate data, the well-known NORmal To Anything (NORTA) algorithm \citep{Cario1997} and other copula approaches \citep{Nelsen2007} are well-studied, with flexible, robust software available \citep{Yan2007, Chen2001}.
-Yet these approaches do not scale in a timely fashion to high-dimensional problems \citep{Li2019gpu}.
-For discrete data, early simulation strategies had major flaws, such as failing to obtain the full range of possible correlations (e.g., admitting only positive correlations: see \citet{Park1996}).
-While more recent approaches \citep{MB13, Xia17, BF17} have largely remedied this issue for low-dimensional problems, the existing tools are not designed to scale to high dimensions.
-
-Another central issue lies in characterizing dependence between components in the high-dimensional random vector.
-The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc).
-For normal data, the Pearson product-moment correlation describes the dependency perfectly.
-As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix is computationally intense \citep{Chen2001, Xia17}.
-On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependence, such as Spearman's \(\rho\) and Kendall's \(\tau\).
-Throughout, we focus on matching these nonparametric dependence measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology.
-We do, however, provide Pearson matching with some caveats.
-
-With all this in mind, we present a scalable, flexible multivariate simulation algorithm.
-The crux of the method lies in the construction of a Gaussian copula in the spirit of the NORTA procedure.
-Further, we introduce the \texttt{bigsimr} R package that provides high-performance software implementing our algorithm.
-The algorithm design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed-form relationships between dependence measures for the multivariate normal (MVN) distribution.
-
-Our study proceeds by providing background information, including a description of a motivating example application: RNA-sequencing (RNA-seq) breast cancer data.
-Then we describe and justify our simulation methodology and related algorithms.
-Next, we detail an illustrative low-dimensional example of basic use of the \texttt{bigsimr} R package.
-Then we proceed with Monte Carlo studies under various bivariate distributional assumptions to assess accuracy.
-We conclude the Monte Carlo evaluations by summarizing the computation time for increasingly higher dimensional vectors.
-After the MC evaluations, we simulate random vectors motivated by our RNA-seq example, evaluate the accuracy, and provide example statistical computing tasks, namely MC estimation of joint probabilities and evaluating HD correlation estimation efficiency.
-Finally, we discuss the method's utility, limitations, and future directions.
+Massive high-dimensional (HD) data sets are now commonplace in many
+areas of scientific inquiry. As new methods are developed for data, a
+fundamental challenge lies in designing and conducting simulation
+studies to assess the operating characteristics of analyzing such
+proposed methodology --- such as false positive rates, statistical
+power, interval coverage, and robustness --- often with comparison to
+existing methods. Further, efficient simulation empowers statistical
+computing strategies, such as the parametric bootstrap
+\citep{Chernick2008} to simulate from a hypothesized null model,
+providing inference in analytically challenging settings. Such Monte
+Carlo (MC) techniques become difficult for high-dimensional data using
+existing algorithms and tools. This is particularly true when simulating
+massive multivariate, non-normal distributions, arising naturally in
+many fields of study.
+
+As many have noted, it can be vexing to simulate dependent,
+non-normal/discrete data, even for low dimensional settings
+\citep{MB13, XZ19}. For continuous non-normal multivariate data, the
+well-known NORmal To Anything (NORTA) algorithm \citep{Cario1997} and
+other copula approaches \citep{Nelsen2007} are well-studied, with
+flexible, robust software available \citep{Yan2007, Chen2001}. Yet these
+approaches do not scale in a timely fashion to high-dimensional problems
+\citep{Li2019gpu}. For discrete data, early simulation strategies had
+major flaws, such as failing to obtain the full range of possible
+correlations (e.g., admitting only positive correlations: see
+\citet{Park1996}). While more recent approaches
+\citep{MB13, Xia17, BF17} have largely remedied this issue for
+low-dimensional problems, the existing tools are not designed to scale
+to high dimensions.
+
+Another central issue lies in characterizing dependence between
+components in the high-dimensional random vector. The choice of
+correlation in practice usually relates to the eventual analytic goal
+and distributional assumptions of the data (e.g., non-normal, discrete,
+infinite support, etc). For normal data, the Pearson product-moment
+correlation describes the dependency perfectly. As we will see, however,
+simulating arbitrary random vectors that match a target Pearson
+correlation matrix is computationally intense \citep{Chen2001, Xia17}.
+On the other hand, an analyst might consider use of nonparametric
+correlation measures to better characterize monotone, non-linear
+dependence, such as Spearman's \(\rho\) and Kendall's \(\tau\).
+Throughout, we focus on matching these nonparametric dependence
+measures, as our aim lies in modeling non-normal data and these
+rank-based measures possess invariance properties favorable in our
+proposed methodology. We do, however, provide Pearson matching with some
+caveats.
+
+With all this in mind, we present a scalable, flexible multivariate
+simulation algorithm. The crux of the method lies in the construction of
+a Gaussian copula in the spirit of the NORTA procedure. Further, we
+introduce the \texttt{bigsimr} R package that provides high-performance
+software implementing our algorithm. The algorithm design leverages
+useful properties of nonparametric correlation measures, namely
+invariance under monotone transformation and well-known closed-form
+relationships between dependence measures for the multivariate normal
+(MVN) distribution.
+
+Our study proceeds by providing background information, including a
+description of a motivating example application: RNA-sequencing
+(RNA-seq) breast cancer data. Then we describe and justify our
+simulation methodology and related algorithms. Next, we detail an
+illustrative low-dimensional example of basic use of the
+\texttt{bigsimr} R package. Then we proceed with Monte Carlo studies
+under various bivariate distributional assumptions to assess accuracy.
+We conclude the Monte Carlo evaluations by summarizing the computation
+time for increasingly higher dimensional vectors. After the MC
+evaluations, we simulate random vectors motivated by our RNA-seq
+example, evaluate the accuracy, and provide example statistical
+computing tasks, namely MC estimation of joint probabilities and
+evaluating HD correlation estimation efficiency. Finally, we discuss the
+method's utility, limitations, and future directions.
 
@@
 
-The \texttt{bigsimr} \texttt{R} package presented here provides multiple algorithms that operate with high-dimensional multivariate data; however, all these algorithms were originally designed to support a single task: to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics.
-Specifically, our goal is to efficiently simulate a large number, \(B\), of HD random vectors \({\bf Y}=(Y_1, \ldots, Y_d)^\top\) with \emph{correlated} components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) \(F_i\).
-
-When designing this methodology, we developed the following properties to guide our effort.
-We divide the properties into two categories: (1) basic properties (BP) and ``scalability'' properties (SP).
-The BPs are adapted from an existing criteria due to \citet{Nik13a}.
-A suitable simulation strategy should possess the following properties:
+The \texttt{bigsimr} \texttt{R} package presented here provides multiple
+algorithms that operate with high-dimensional multivariate data;
+however, all these algorithms were originally designed to support a
+single task: to generate random vectors drawn from multivariate
+probability distributions with given marginal distributions and
+dependency metrics. Specifically, our goal is to efficiently simulate a
+large number, \(B\), of HD random vectors
+\({\bf Y}=(Y_1, \ldots, Y_d)^\top\) with \emph{correlated} components
+and heterogeneous marginal distributions, described via cumulative
+distribution functions (CDFs) \(F_i\).
+
+When designing this methodology, we developed the following properties
+to guide our effort. We divide the properties into two categories: (1)
+basic properties (BP) and ``scalability'' properties (SP). The BPs are
+adapted from an existing criteria due to \citet{Nik13a}. A suitable
+simulation strategy should possess the following properties:
 
@@
 \item
-  BP1: A wide range of dependences, allowing both positive and negative values, and, ideally, admitting the full range of possible values.
-\item
-  BP2: Flexible dependence, meaning that the number of bivariate marginals can be equal to the number of dependence parameters.
-\item
-  BP3: Flexible marginal modeling, generating heterogeneous data --- possibly from differing probability families.
+  BP1: A wide range of dependences, allowing both positive and negative
+  values, and, ideally, admitting the full range of possible values.
+\item
+  BP2: Flexible dependence, meaning that the number of bivariate
+  marginals can be equal to the number of dependence parameters.
+\item
+  BP3: Flexible marginal modeling, generating heterogeneous data ---
+  possibly from differing probability families.
 \end{itemize}
@@
 \item
-  SP1: Procedure must scale to high dimensions with practical compute times.
-\item
-  SP2: Procedure must scale to high dimensions while maintaining accuracy.
+  SP1: Procedure must scale to high dimensions with practical compute
+  times.
+\item
+  SP2: Procedure must scale to high dimensions while maintaining
+  accuracy.
 \end{itemize}
@@
 \hypertarget{motivating-example-rna-seq-data}{%
-\subsection{Motivating example: RNA-seq data}\label{motivating-example-rna-seq-data}}
-
-Simulating high-dimensional, non-normal, correlated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data \citep{Wang2009b, Conesa2016b} derived from breast cancer patients.
-The RNA-seq data-generating process involves counting how often a particular form of human messenger RNA (mRNA) is expressed in a biological sample.
-RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data.
-For human-derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted \citep{Schissler2019}.
-Importantly, due to inherent biological processes, gene expression data exhibit correlation (co-expression) across genes \citep{BE07, Schissler2018}.
-
-We illustrate our methodology using a well-studied Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments).
-For ease of modeling and simplicity of exposition, we only consider high expressing genes.
-In turn, we begin by filtering to retain the top 1000 of the highest-expressing genes (in terms of median expression) of the over 20,000 gene measurements from \(N=878\) patients' tumor samples.
-This gives a great number of pairwise dependencies among the marginals (specifically, \(\ensuremath{4.995\times 10^{5}}\) correlation parameters).
-Table \ref{tab:ch010-realDataTab} displays RNA-seq counts for three selected high-expressing genes for the first five patients' breast tumor samples.
-To help visualize the bivariate relationships for these three selected genes across all patients, Figure \ref{fig:ch010-realDataFig} displays the marginal distributions and estimated Spearman's correlations.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> set.seed(2020-02-25)
-R> num_genes    <- 3
-R> num_patients <- 5
-R> gene_sample  <- sample(example_genes[1:1000], num_genes)
-R> 
-R> cap <- paste0("mRNA expression for three selected high-expressing genes, ",
-+               paste(gene_sample, collapse = ", "), 
-+               ", for the first five patients in the TCGA BRCA data set.")
-R> 
-R> small_brca <- example_brca %>%
-+   select(all_of(gene_sample)) %>%
-+   head(n = num_patients) %>%
-+   as_tibble(rownames = "Patient ID") %>%
-+   mutate(`Patient ID` = str_sub(`Patient ID`, end = 12))
-R> 
-R> small_brca %>%
-+   kable(format = "latex", booktabs = TRUE, caption = cap)
-\end{CodeInput}
+\subsection{Motivating example: RNA-seq
+data}\label{motivating-example-rna-seq-data}}
+
+Simulating high-dimensional, non-normal, correlated data motivates this
+work --- in pursuit of modeling RNA-sequencing (RNA-seq) data
+\citep{Wang2009b, Conesa2016b} derived from breast cancer patients. The
+RNA-seq data-generating process involves counting how often a particular
+form of human messenger RNA (mRNA) is expressed in a biological sample.
+RNA-seq platforms typically quantify the entire transcriptome in one
+experimental run, resulting in high-dimensional data. For human-derived
+samples, this results in count data corresponding to over 20,000 genes
+(protein-coding genomic regions) or even over 77,000 isoforms when
+alternatively-spliced mRNA are counted \citep{Schissler2019}.
+Importantly, due to inherent biological processes, gene expression data
+exhibit correlation (co-expression) across genes
+\citep{BE07, Schissler2018}.
+
+We illustrate our methodology using a well-studied Breast Invasive
+Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see
+Acknowledgments). For ease of modeling and simplicity of exposition, we
+only consider high expressing genes. In turn, we begin by filtering to
+retain the top 1000 of the highest-expressing genes (in terms of median
+expression) of the over 20,000 gene measurements from \(N=878\)
+patients' tumor samples. This gives a great number of pairwise
+dependencies among the marginals (specifically,
+\(\ensuremath{4.995\times 10^{5}}\) correlation parameters). Table
+@ref(tab:ch010-realDataTab) displays RNA-seq counts for three selected
+high-expressing genes for the first five patients' breast tumor samples.
+To help visualize the bivariate relationships for these three selected
+genes across all patients, Figure @ref(fig:ch010-realDataFig) displays
+the marginal distributions and estimated Spearman's correlations.
+
+\begin{CodeChunk}
 \begin{table}
@@
 \begin{CodeChunk}
-\begin{CodeInput}
-R> ggpairs(
-+   data = example_brca[, gene_sample],
-+   upper = list(continuous = wrap('cor', method = "spearman"))
-+ ) + 
-+   theme_bw() +
-+   theme(axis.text.x = element_text(angle = -90, vjust = 0.5))
-\end{CodeInput}
-\begin{CodeOutput}
-Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
-exact p-value with ties
-\end{CodeOutput}
-\begin{CodeOutput}
-Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
-exact p-value with ties
-\end{CodeOutput}
-\begin{CodeOutput}
-Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
-exact p-value with ties
-\end{CodeOutput}
 \begin{figure}
 
-{\centering \includegraphics{article_bigsimr_files/figure-latex/ch010-realDataFig-1}
+{\centering \includegraphics[width=0.85\linewidth]{article_bigsimr_files/figure-latex/ch010-realDataFig-1}
 
@@
 
-In multivariate analysis, an analyst must select a metric to quantify dependency.
-The most widely-known is the Pearson (product-moment) correlation coefficient that describes the linear association between two random variables \(X\) and \(Y\), and, it is given by
+In multivariate analysis, an analyst must select a metric to quantify
+dependency. The most widely-known is the Pearson (product-moment)
+correlation coefficient that describes the linear association between
+two random variables \(X\) and \(Y\), and, it is given by
 
@@
 \rho_P(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ \mathrm{Var}(X)\mathrm{Var}(Y)\right]^{1/2}}.
-\label{eq:pearson}
-\end{equation}
-
-As \citet{MB13} and \citet{MK01} discuss, for a bivariate normal \((X,Y)\) random vector, the Pearson correlation completely describes the dependency between the components.
-For non-normal marginals with monotone correlation patterns, however, \(\rho_P\) suffers some drawbacks and may mislead or fail to capture important relationships \citep{MK01}.
-Alternatively in these settings, analysts often prefer rank-based correlation measures to describe the degree of monotonic association.
-
-Two nonparametric, rank-based measures common in practice are Spearman's correlation (denoted \(\rho_S\)) and Kendall's \(\tau\).
-Spearman's \(\rho_S\) has an appealing correspondence as the Pearson correlation coefficient on \emph{ranks} of the values, thereby captures nonlinear yet monotone relationships.
-Kendall's \(\tau\), on the other hand, is the difference in probabilities of concordant and discordant pairs of observations, \((X_i, Y_i)\) and \((X_j, Y_j)\), with concordance meaning that orderings have the same direction (e.g., if \(X_i < X_j\), then \(Y_i < Y_j\)).
-Note that concordance is determined by the ranks of the values, not the values themselves.
-
-Both \(\tau\) and \(\rho_S\) are \emph{invariant under monotone transformations} of the underlying random variates.
-As we will describe more fully in the \protect\hyperlink{algorithms}{Algorithms} section, this property enables matching rank-based correlations with speed (SP1) and accuracy (SP2).
+(\#eq:pearson)
+\end{equation}
+
+As \citet{MB13} and \citet{MK01} discuss, for a bivariate normal
+\((X,Y)\) random vector, the Pearson correlation completely describes
+the dependency between the components. For non-normal marginals with
+monotone correlation patterns, however, \(\rho_P\) suffers some
+drawbacks and may mislead or fail to capture important relationships
+\citep{MK01}. Alternatively in these settings, analysts often prefer
+rank-based correlation measures to describe the degree of monotonic
+association.
+
+Two nonparametric, rank-based measures common in practice are Spearman's
+correlation (denoted \(\rho_S\)) and Kendall's \(\tau\). Spearman's
+\(\rho_S\) has an appealing correspondence as the Pearson correlation
+coefficient on \emph{ranks} of the values, thereby captures nonlinear
+yet monotone relationships. Kendall's \(\tau\), on the other hand, is
+the difference in probabilities of concordant and discordant pairs of
+observations, \((X_i, Y_i)\) and \((X_j, Y_j)\), with concordance
+meaning that orderings have the same direction (e.g., if \(X_i < X_j\),
+then \(Y_i < Y_j\)). Note that concordance is determined by the ranks of
+the values, not the values themselves.
+
+Both \(\tau\) and \(\rho_S\) are \emph{invariant under monotone
+transformations} of the underlying random variates. As we will describe
+more fully in the \protect\hyperlink{algorithms}{Algorithms} section,
+this property enables matching rank-based correlations with speed (SP1)
+and accuracy (SP2).
 
@@
 
-There is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficient, as the marginal distributions \(F_i\) are intrinsic in their calculation.
-For \emph{bivariate normal vectors}, however, the correspondence is well-known:
-
-\begin{equation}
-\label{eq:convertKendall}
+There is no closed form, general correspondence among the rank-based
+measures and the Pearson correlation coefficient, as the marginal
+distributions \(F_i\) are intrinsic in their calculation. For
+\emph{bivariate normal vectors}, however, the correspondence is
+well-known:
+
+\begin{equation}
+(\#eq:convertKendall)
 \rho_{P} = \sin \left( \tau \times \frac{\pi}{2} \right), 
@@
 \begin{equation}
-\label{eq:convertSpearman}
+(\#eq:convertSpearman)
 \rho_P = 2 \times \sin \left( \rho_S \times \frac{\pi}{6} \right).
@@
 
-Given two marginal distributions, \(\rho_P\) is not free to vary over the entire range of possible correlations \([-1,1]\).
-The so-called \emph{Frechet-Hoeffding bounds} are well-studied \citep{Nelsen2007, BF17}.
-These constraints cannot be overcome through algorithm design.
-In general, the bounds are given by
-
-\begin{equation}
-\label{eq:frechet}
+Given two marginal distributions, \(\rho_P\) is not free to vary over
+the entire range of possible correlations \([-1,1]\). The so-called
+\emph{Frechet-Hoeffding bounds} are well-studied
+\citep{Nelsen2007, BF17}. These constraints cannot be overcome through
+algorithm design. In general, the bounds are given by
+
+\begin{equation}
+(\#eq:frechet)
 \rho_P^{max} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (U) \right), \quad \rho_P^{min} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (1 - U) \right)
@@
 
-\noindent where \(U\) is a uniform random variable on \((0,1)\) and \(F^{-1}_1, F^{-1}_2\) are the inverse CDFs of \(X_1\) and \(X_2\), respectively, definitely by \eqref{eq:inverseCDF} when the variables are discrete.
+\noindent where \(U\) is a uniform random variable on \((0,1)\) and
+\(F^{-1}_1, F^{-1}_2\) are the inverse CDFs of \(X_1\) and \(X_2\),
+respectively, definitely by @ref(eq:inverseCDF) when the variables are
+discrete.
 
@@
 F_{i}^{-1} = \inf\{y:F_{i}(y) \geq u \}.
-\label{eq:inverseCDF}
+(\#eq:inverseCDF)
 \end{equation}
@@
 
-There is a strong connection of our simulation strategy to Gaussian \emph{copulas} (see \citet{Nelsen2007} for a technical introduction).
-A copula is a distribution function on \([0,1]^d\) that describes a multivariate probability distribution with standard uniform marginals.
-This provides a powerful, natural way to characterize joint probability structures.
-Consequently, the study of copulas is an important and active area of statistical theory and practice.
-
-For any random vector \({\bf X}=(X_1, \ldots, X_d)^\top\) with CDF \(F\) and marginal CDFs \(F_i\) there is a copula function \(C(u_1, \ldots, u_d)\) satisfying
+There is a strong connection of our simulation strategy to Gaussian
+\emph{copulas} (see \citet{Nelsen2007} for a technical introduction). A
+copula is a distribution function on \([0,1]^d\) that describes a
+multivariate probability distribution with standard uniform marginals.
+This provides a powerful, natural way to characterize joint probability
+structures. Consequently, the study of copulas is an important and
+active area of statistical theory and practice.
+
+For any random vector \({\bf X}=(X_1, \ldots, X_d)^\top\) with CDF \(F\)
+and marginal CDFs \(F_i\) there is a copula function
+\(C(u_1, \ldots, u_d)\) satisfying
 
@@
 F(x_1, \ldots, x_d) = {\mathbb P}(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), 
-\label{eq:copula}
+(\#eq:copula)
 \end{equation}
@@
 
-A Gaussian copula has marginal CDFs that are all standard normal, \(F_i = \Phi, \forall \, i\).
-This representation corresponds to a multivariate normal (MVN) distribution with standard normal marginal distributions and covariance matrix \({\bf R_P}\).
-As the marginals are standardized to have unit variance, however, \({\bf R_P}\) is a Pearson correlation matrix.
-If \(F_{{\bf R}}\) is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula \(C_{{\bf R}}\) is defined through
-
-\begin{equation}
-\label{eq:gauss}
+A Gaussian copula has marginal CDFs that are all standard normal,
+\(F_i = \Phi, \forall \, i\). This representation corresponds to a
+multivariate normal (MVN) distribution with standard normal marginal
+distributions and covariance matrix \({\bf R_P}\). As the marginals are
+standardized to have unit variance, however, \({\bf R_P}\) is a Pearson
+correlation matrix. If \(F_{{\bf R}}\) is the CDF of such a multivariate
+normal distribution, then the corresponding Gaussian copula
+\(C_{{\bf R}}\) is defined through
+
+\begin{equation}
+(\#eq:gauss)
 F_{{\bf R}}(x_1, \ldots, x_d) = C_{{\bf R}}(\Phi(x_1), \ldots, \Phi(x_d)),
@@
 
-where \(\Phi(\cdot)\) is the standard normal CDF.
-Note that the copula \(C_{{\bf R}}\) is the familiar multivariate normal CDF of the random vector \((\Phi(X_1), \ldots, \Phi(X_d))\), where \((X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})\).
-
-Sklar's Theorem \citep{Sklar1959, Ubeda-Flores2017} guarantees that given inverse CDFs \(F_i^{-1}\)s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions.
-For example, using Gaussian copulas, we can construct a random vector \({\bf Y} = (Y_1, \ldots, Y_d)^\top\) with \(Y_i \sim F_i\), viz.~\(Y_i = F_i^{-1}(\Phi(X_i)), i=1, \ldots, d\), where \((X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})\).
+where \(\Phi(\cdot)\) is the standard normal CDF. Note that the copula
+\(C_{{\bf R}}\) is the familiar multivariate normal CDF of the random
+vector \((\Phi(X_1), \ldots, \Phi(X_d))\), where
+\((X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})\).
+
+Sklar's Theorem \citep{Sklar1959, Ubeda-Flores2017} guarantees that
+given inverse CDFs \(F_i^{-1}\)s and a valid correlation matrix (within
+the Frechet bounds) a random vector can be obtained via transformations
+involving copula functions. For example, using Gaussian copulas, we can
+construct a random vector \({\bf Y} = (Y_1, \ldots, Y_d)^\top\) with
+\(Y_i \sim F_i\), viz.~\(Y_i = F_i^{-1}(\Phi(X_i)), i=1, \ldots, d\),
+where \((X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})\).
 
@@
 
-This section describes our methods involved in simulating a random vector \(\bf Y\) with \(Y_i\) components for \(i=1,\ldots,d\).
-Each \(Y_i\) has a specified marginal CDF \(F_i\) and its inverse \(F^{-1}_i\).
-To characterize dependency, every pair \((Y_i, Y_j)\) has a given Pearson correlation \(\rho_P\), Spearman correlation \(\rho_S\), and/or Kendall's \(\tau\).
-The method can be described as a \emph{high-performance Gaussian copula} (Equation \eqref{eq:gauss}) providing a high-dimensional NORTA-inspired algorithm.
+This section describes our methods involved in simulating a random
+vector \(\bf Y\) with \(Y_i\) components for \(i=1,\ldots,d\). Each
+\(Y_i\) has a specified marginal CDF \(F_i\) and its inverse
+\(F^{-1}_i\). To characterize dependency, every pair \((Y_i, Y_j)\) has
+a given Pearson correlation \(\rho_P\), Spearman correlation \(\rho_S\),
+and/or Kendall's \(\tau\). The method can be described as a
+\emph{high-performance Gaussian copula} (Equation @ref(eq:gauss))
+providing a high-dimensional NORTA-inspired algorithm.
 
@@
 
-The well-known NORTA algorithm \citep{Cario1997} simulates a random vector \(\bf Y\) with variance-covariance matrix \(\Sigma_{\bf Y}\). Specifically, NORTA algorithm proceeds as:
+The well-known NORTA algorithm \citep{Cario1997} simulates a random
+vector \(\bf Y\) with variance-covariance matrix \(\Sigma_{\bf Y}\).
+Specifically, NORTA algorithm proceeds as:
 
@@
 \item
-  Simulate a random vector \(\bf Z\) with \(d\) \emph{independent and identically distributed} (iid) standard normal components.
-\item
-  Determine the input matrix \(\Sigma_{\bf Z}\) that corresponds with the specified output \(\Sigma_{\bf Y}\).
-\item
-  Produce a Cholesky factor \(M\) of \(\Sigma_{\bf Z}\) such that \(M M^{\prime}=\Sigma_{\bf Z}\).
+  Simulate a random vector \(\bf Z\) with \(d\) \emph{independent and
+  identically distributed} (iid) standard normal components.
+\item
+  Determine the input matrix \(\Sigma_{\bf Z}\) that corresponds with
+  the specified output \(\Sigma_{\bf Y}\).
+\item
+  Produce a Cholesky factor \(M\) of \(\Sigma_{\bf Z}\) such that
+  \(M M^{\prime}=\Sigma_{\bf Z}\).
 \item
@@
 
-With modern parallel computing, steps 1, 3, 4, 5 are readily implemented as high-performance, multi-core and/or graphical-processing-unit (GPU) accelerated algorithms --- providing fast scalability using readily-available hardware.
-
-Matching specified Pearson correlation coefficients exactly (step 2 above), however, is computationally costly.
-In general, there is no closed-form correspondence between the components of the input \(\Sigma_{\bf Z}\) and target \(\Sigma_{\bf Y}\).
-Matching the correlations involves evaluating or approximating \(\binom{d}{2}\) integrals of the form
+With modern parallel computing, steps 1, 3, 4, 5 are readily implemented
+as high-performance, multi-core and/or graphical-processing-unit (GPU)
+accelerated algorithms --- providing fast scalability using
+readily-available hardware.
+
+Matching specified Pearson correlation coefficients exactly (step 2
+above), however, is computationally costly. In general, there is no
+closed-form correspondence between the components of the input
+\(\Sigma_{\bf Z}\) and target \(\Sigma_{\bf Y}\). Matching the
+correlations involves evaluating or approximating \(\binom{d}{2}\)
+integrals of the form
 
@@
     \mathrm{E}\left[Y_i Y_j\right] = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} F_i^{-1}\left[\Phi(z_i)\right] F_j^{-1}\left[\Phi(z_j)\right] \phi(z_i, z_j, \rho_z) dz_i dz_j,
-    \label{eq:pearsonIntegralRelation}
-\end{equation}
-
-where \(\phi(\cdot)\) is the joint probability distribution function of two correlated standard normal variables.
-For HD data, (nearly) exact evaluation may become too costly to enable practical simulation studies.
-For low-dimensional problems, however, methods and tools exist to match Pearson correlations precisely; see \citep{Xia17} and the publicly available \texttt{nortaRA} R package \citep{Chen2001}.
-As described later, to enable HD Pearson matching we approximate these integrals.
+    (\#eq:pearsonIntegralRelation)
+\end{equation}
+
+where \(\phi(\cdot)\) is the joint probability distribution function of
+two correlated standard normal variables. For HD data, (nearly) exact
+evaluation may become too costly to enable practical simulation studies.
+For low-dimensional problems, however, methods and tools exist to match
+Pearson correlations precisely; see \citep{Xia17} and the publicly
+available \texttt{nortaRA} R package \citep{Chen2001}. As described
+later, to enable HD Pearson matching we approximate these integrals.
 
@@
 
-Sklar's theorem provides a useful characterization of multivariate distributions through copulas.
-Yet the choice of copula-based simulation algorithm affects which joint distributions may be simulated.
-Even in low-dimensional spaces (e.g., \(d=3\)), there exist valid multivariate distributions with \emph{feasible} Pearson correlation matrices that NORTA cannot match exactly \citep{LH75}.
-This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined the resultant matrix is indefinite.
-These situations do occur, even using exact analytic calculations.
-Such problematic target correlation matrices are termed \emph{NORTA defective}.
-
-\citet{GH02} conducted a Monte Carlo study to estimate the probability of encountering NORTA defective matrices while increasing the dimension \(d\).
-They found that for what is now considered low-to-moderate dimensions (\(d \approx 20\)), almost \emph{all} feasible matrices are NORTA defective.
-This stems from the concentration of measure near the boundary of the space of all possible correlation matrices as dimension increases.
-Unfortunately, it is precisely near this boundary that NORTA defective matrices reside.
-
-There is hope, however, as \citet{GH02} also showed that replacing an indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate \(d\).
-This provides evidence that our nearest positive definite (PD) augmented approach will maintain reasonable accuracy if our input matching scheme returns an indefinite matrix, at least for the rank-based matching scheme described above.
+Sklar's theorem provides a useful characterization of multivariate
+distributions through copulas. Yet the choice of copula-based simulation
+algorithm affects which joint distributions may be simulated. Even in
+low-dimensional spaces (e.g., \(d=3\)), there exist valid multivariate
+distributions with \emph{feasible} Pearson correlation matrices that
+NORTA cannot match exactly \citep{LH75}. This occurs when the bivariate
+transformations are applied to find the input correlation matrix, yet
+when combined the resultant matrix is indefinite. These situations do
+occur, even using exact analytic calculations. Such problematic target
+correlation matrices are termed \emph{NORTA defective}.
+
+\citet{GH02} conducted a Monte Carlo study to estimate the probability
+of encountering NORTA defective matrices while increasing the dimension
+\(d\). They found that for what is now considered low-to-moderate
+dimensions (\(d \approx 20\)), almost \emph{all} feasible matrices are
+NORTA defective. This stems from the concentration of measure near the
+boundary of the space of all possible correlation matrices as dimension
+increases. Unfortunately, it is precisely near this boundary that NORTA
+defective matrices reside.
+
+There is hope, however, as \citet{GH02} also showed that replacing an
+indefinite input correlation matrix with a close proxy will give
+approximate matching to the target --- with adequate performance for
+moderate \(d\). This provides evidence that our nearest positive
+definite (PD) augmented approach will maintain reasonable accuracy if
+our input matching scheme returns an indefinite matrix, at least for the
+rank-based matching scheme described above.
 
@@
 
-We now describe our algorithm to generate random vectors, which mirrors the classical NORTA algorithm above with some modifications for rank-based dependency matching:
+We now describe our algorithm to generate random vectors, which mirrors
+the classical NORTA algorithm above with some modifications for
+rank-based dependency matching:
 
@@
     \item
-      Convert the target Spearman correlation matrix \(R_S\) to the corresponding MVN Pearson correlation \(R_X\). Alternatively,
+      Convert the target Spearman correlation matrix \(R_S\) to the
+      corresponding MVN Pearson correlation \(R_X\). Alternatively,
     \end{enumerate}
   \item
-    (i') Convert the target Kendall \(\tau\) matrix \(R_K\) to the corresponding MVN Pearson correlation \(R_X\). Alternatively,
-  \item
-    (i'\,') Convert the target Pearson correlation matrix to \(R_P\) to the corresponding, approximate MVN Pearson correlation \(R_X\).
+    (i') Convert the target Kendall \(\tau\) matrix \(R_K\) to the
+    corresponding MVN Pearson correlation \(R_X\). Alternatively,
+  \item
+    (i'\,') Convert the target Pearson correlation matrix to \(R_P\) to
+    the corresponding, approximate MVN Pearson correlation \(R_X\).
   \end{itemize}
 \item
-  Check admissibility and, if needed, compute the nearest correlation matrix.
+  Check admissibility and, if needed, compute the nearest correlation
+  matrix.
 
@@
     \item
-      Check that \(R_X\) is a correlation matrix, a positive definite matrix with 1's along the diagonal.
+      Check that \(R_X\) is a correlation matrix, a positive definite
+      matrix with 1's along the diagonal.
     \end{enumerate}
@@
     \item
-      If \(R_X\) is a correlation matrix, the input matrix is \emph{admissible} in this scheme. Otherwise
+      If \(R_X\) is a correlation matrix, the input matrix is
+      \emph{admissible} in this scheme. Otherwise
     \end{enumerate}
   \item
-    (ii') Replace \(R_X\) with the nearest correlation matrix \(\tilde{R}_X\), in the Frobenius norm.
+    (ii') Replace \(R_X\) with the nearest correlation matrix
+    \(\tilde{R}_X\), in the Frobenius norm.
   \end{itemize}
@@
     \item
-      Transform \({\bf X}\) to \({\bf U} = (U_1, \ldots, U_d)\) viz.~\(U_i=\Phi(X_i)\), \(i=1, \ldots, d\).
+      Transform \({\bf X}\) to \({\bf U} = (U_1, \ldots, U_d)\)
+      viz.~\(U_i=\Phi(X_i)\), \(i=1, \ldots, d\).
     \end{enumerate}
@@
     \item
-      Return \({\bf Y} = (Y_1, \ldots, Y_d)\), where \(Y_i=F_i^{-1}(U_i)\), \(i=1, \ldots, d\).
+      Return \({\bf Y} = (Y_1, \ldots, Y_d)\), where
+      \(Y_i=F_i^{-1}(U_i)\), \(i=1, \ldots, d\).
     \end{enumerate}
@@
 
-The first two descriptions of the \emph{Mapping step} employ the closed-form relationships between \(\rho_S\) and \(\tau\) with \(\rho_P\) for bivariate normal random variables via Equations \eqref{eq:convertKendall} and \eqref{eq:convertSpearman}, respectively (implemented as \texttt{cor\_covert}).
-Initializing our algorithm to match the nonparametric correlations by computing these equations for all pairs is computationally trivial.
-
-For the computationally expensive process to match Pearson correlations, we approximate Equation \eqref{eq:pearsonIntegralRelation} for all pairs of margins.
-To this end, we implement the approximation scheme introduced by \citet{xiao2019matching}.
-Briefly, the many double integrals of the form in \eqref{eq:pearsonIntegralRelation} are approximated by weighted sums of Hermite polynomials.
-Matching coefficients for pairs of continuous distributions is made tractable by this method, but for discrete distributions (especially discrete distributions with large support sets or infinite support), the problem is not so simple or efficient.
-
-Our solution is to approximate discrete distributions by a continuous distribution.
-The question then becomes that of which distribution to use.
-We found that Generalized S-Distributions \citep{muino2006gs} solve this problem by approximating a wide range of unimodal distributions, both continuous and discrete.
-Using the GS-Distribution approximation of discrete distributions in Pearson matching scheme above yields favorable results.
+The first two descriptions of the \emph{Mapping step} employ the
+closed-form relationships between \(\rho_S\) and \(\tau\) with
+\(\rho_P\) for bivariate normal random variables via Equations
+@ref(eq:convertKendall) and @ref(eq:convertSpearman), respectively
+(implemented as \texttt{cor\_covert}). Initializing our algorithm to
+match the nonparametric correlations by computing these equations for
+all pairs is computationally trivial.
+
+For the computationally expensive process to match Pearson correlations,
+we approximate Equation @ref(eq:pearsonIntegralRelation) for all pairs
+of margins. To this end, we implement the approximation scheme
+introduced by \citet{xiao2019matching}. Briefly, the many double
+integrals of the form in @ref(eq:pearsonIntegralRelation) are
+approximated by weighted sums of Hermite polynomials. Matching
+coefficients for pairs of continuous distributions is made tractable by
+this method, but for discrete distributions (especially discrete
+distributions with large support sets or infinite support), the problem
+is not so simple or efficient.
+
+Our solution is to approximate discrete distributions by a continuous
+distribution. The question then becomes that of which distribution to
+use. We found that Generalized S-Distributions \citep{muino2006gs} solve
+this problem by approximating a wide range of unimodal distributions,
+both continuous and discrete. Using the GS-Distribution approximation of
+discrete distributions in Pearson matching scheme above yields favorable
+results.
 
@@
 
-Once \(R_X\) has been determined, we check admissibility of the adjusted correlation via the steps described above.
-If \(R_X\) is not a valid correlation matrix, then we compute the nearest correlation matrix.
-Finding the nearest correlation matrix is a common statistical computing problem.
-The defacto default function in \texttt{R} is \texttt{Matrix::nearPD}, an alternating projection algorithm due to \citet{higham2002computing}.
-As implemented the function fails to scale to HD.
-Instead, we provide the quadratically convergent algorithm based on the theory of strongly semi-smooth matrix functions (\citet{qi2006quadratically}).
-The nearest correlation matrix problem can be written down as the following convex optimization problem:
+Once \(R_X\) has been determined, we check admissibility of the adjusted
+correlation via the steps described above. If \(R_X\) is not a valid
+correlation matrix, then we compute the nearest correlation matrix.
+Finding the nearest correlation matrix is a common statistical computing
+problem. The defacto default function in \texttt{R} is
+\texttt{Matrix::nearPD}, an alternating projection algorithm due to
+\citet{higham2002computing}. As implemented the function fails to scale
+to HD. Instead, we provide the quadratically convergent algorithm based
+on the theory of strongly semi-smooth matrix functions
+(\citet{qi2006quadratically}). The nearest correlation matrix problem
+can be written down as the following convex optimization problem:
 
@@
 
-For nonparametric correlation measures, our algorithm allows the generation of high-dimensional multivariate data with arbitrary marginal distributions with a broad class of admissible Spearman correlation matrices and Kendall \(\tau\) matrices.
-The admissible classes consist of the matrices that map to a Pearson correlation matrix for a MVN.
-In particular, if we let \(X\) be MVN with \(d\) components and set
+For nonparametric correlation measures, our algorithm allows the
+generation of high-dimensional multivariate data with arbitrary marginal
+distributions with a broad class of admissible Spearman correlation
+matrices and Kendall \(\tau\) matrices. The admissible classes consist
+of the matrices that map to a Pearson correlation matrix for a MVN. In
+particular, if we let \(X\) be MVN with \(d\) components and set
 
@@
 
-then there are 1-1 mappings between these sets.
-We conjecture informally that the sets of admissible \(R_S\) and \(R_K\) are not highly restrictive.
-In particular, \(R_P\) is approximately \(R_S\) for a MVN, suggesting that the admissible set \(\Omega_S\) should be flexible as \(R_P\) can be any PD matrix with 1's along the diagonal.
-We provide methods to check whether a target \(R_S\) is an element of the \emph{admissible set} \(\Omega_S\) (and similarly for \(R_K\)).
-
-There is an increasing probability of encountering a non-admissible correlation matrix as dimension increases.
-In our experience, the mapping step for large \(d\) almost always produces a \(R_X\) that is not a correlation matrix.
-In Section \ref{package} we provide a basic description of how to quantify and control the approximation error.
-Further, the RNA-seq example in Section \ref{examples} provides an illustration of this in practice.
+then there are 1-1 mappings between these sets. We conjecture informally
+that the sets of admissible \(R_S\) and \(R_K\) are not highly
+restrictive. In particular, \(R_P\) is approximately \(R_S\) for a MVN,
+suggesting that the admissible set \(\Omega_S\) should be flexible as
+\(R_P\) can be any PD matrix with 1's along the diagonal. We provide
+methods to check whether a target \(R_S\) is an element of the
+\emph{admissible set} \(\Omega_S\) (and similarly for \(R_K\)).
+
+There is an increasing probability of encountering a non-admissible
+correlation matrix as dimension increases. In our experience, the
+mapping step for large \(d\) almost always produces a \(R_X\) that is
+not a correlation matrix. In Section @ref(package) we provide a basic
+description of how to quantify and control the approximation error.
+Further, the RNA-seq example in Section @ref(examples) provides an
+illustration of this in practice.
 
@@
 
-The final step implements a NORTA-inspired, Gaussian copula approach to produce the desired margins.
-Steps 1 and 2 determine the MVN Pearson correlation values that will eventually match the target correlation.
-Then all that is required is a fast MVN simulator, a standard normal CDF, and well-defined quantile functions for marginals.
-The MVN is transformed to a copula (distribution with standard uniform margins) by applying the normal CDF \(\phi(\cdot)\).
-The quantile functions \(F^{-1}\) are applied across the margins to return the desired random vector \({\bf Y}\).
+The final step implements a NORTA-inspired, Gaussian copula approach to
+produce the desired margins. Steps 1 and 2 determine the MVN Pearson
+correlation values that will eventually match the target correlation.
+Then all that is required is a fast MVN simulator, a standard normal
+CDF, and well-defined quantile functions for marginals. The MVN is
+transformed to a copula (distribution with standard uniform margins) by
+applying the normal CDF \(\phi(\cdot)\). The quantile functions
+\(F^{-1}\) are applied across the margins to return the desired random
+vector \({\bf Y}\).
 
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> box::use(
-+   patchwork[...],
-+   dplyr[...],
-+   ggplot2[...],
-+   tidyr[drop_na],
-+   bigsimr[bigsimr_setup, distributions_setup],
-+   JuliaCall[julia_call]
-+ )
-R> 
-R> Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
-R> bs <- bigsimr_setup(pkg_check = FALSE)
-\end{CodeInput}
-\begin{CodeOutput}
-Julia version 1.5.4 at location /home/alex/julia-1.5.4/bin will be used.
-\end{CodeOutput}
-\begin{CodeOutput}
-Loading setup script for JuliaCall...
-\end{CodeOutput}
-\begin{CodeOutput}
-Finish loading setup script for JuliaCall.
-\end{CodeOutput}
-\begin{CodeInput}
-R> dist <- distributions_setup()
-\end{CodeInput}
-\end{CodeChunk}
-
-This section describes a low-dimensional (2D) random vector simulation workflow via the \texttt{bigsimr} \texttt{R} package (\url{https://github.com/SchisslerGroup/r-bigsimr}).
-The package \texttt{bigsimr} provides an interface to the native code written in Julia, registered as the \texttt{Bigsimr} Julia package (\url{https://github.com/adknudson/Bigsimr.jl}).
-In addition to the native Julia \texttt{Bigsimr} package and \texttt{R} interface \texttt{bigsimr}, we also provide a python interface \texttt{bigsimr} (\url{https://github.com/SchisslerGroup/python-bigsimr/}) that interfaces with the Julia \texttt{Bigsimr} package.
-The Julia package provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \ref{algorithms}).
-
-The subsections below describe the basic use of \texttt{bigsimr} by stepping through an example workflow using the data set \texttt{airquality} that contains daily air quality measurements in New York, May to September 1973 \citep{Chambers1983}.
-This workflow proceeds from setting up the computing environment, to data wrangling, estimation, simulation configuration, random vector generation, and, finally, result visualization.
+This section describes a low-dimensional (2D) random vector simulation
+workflow via the \texttt{bigsimr} \texttt{R} package
+(\url{https://github.com/SchisslerGroup/r-bigsimr}). The package
+\texttt{bigsimr} provides an interface to the native code written in
+Julia, registered as the \texttt{Bigsimr} Julia package
+(\url{https://github.com/adknudson/Bigsimr.jl}). In addition to the
+native Julia \texttt{Bigsimr} package and \texttt{R} interface
+\texttt{bigsimr}, we also provide a python interface \texttt{bigsimr}
+(\url{https://github.com/SchisslerGroup/python-bigsimr/}) that
+interfaces with the Julia \texttt{Bigsimr} package. The Julia package
+provides a high-performance implementation of our proposed random vector
+generation algorithm and associated functions (see Section
+@ref(algorithms)).
+
+The subsections below describe the basic use of \texttt{bigsimr} by
+stepping through an example workflow using the data set
+\texttt{airquality} that contains daily air quality measurements in New
+York, May to September 1973 \citep{Chambers1983}. This workflow proceeds
+from setting up the computing environment, to data wrangling,
+estimation, simulation configuration, random vector generation, and,
+finally, result visualization.
 
 \hypertarget{bivariate-example-description}{%
-\subsection{Bivariate example description}\label{bivariate-example-description}}
-
-We illustrate the use of \texttt{bigsimr} motivated by the New York air quality data set (\texttt{airquality}) included in the R \texttt{datasets} package.
-First, we load the \texttt{bigsimr} library and a few other convenient data science packages, including the syntactically-elegant \texttt{tidyverse} suite of \texttt{R} packages.
+\subsection{Bivariate example
+description}\label{bivariate-example-description}}
+
+We illustrate the use of \texttt{bigsimr} motivated by the New York air
+quality data set (\texttt{airquality}) included in the R
+\texttt{datasets} package. First, we load the \texttt{bigsimr} library
+and a few other convenient data science packages, including the
+syntactically-elegant \texttt{tidyverse} suite of \texttt{R} packages.
 The code chunk below prepares the computing environment:
@@
 
-For simplicity and to provide a minimal working example, we consider bivariate simulation of the two \texttt{airquality} variables \texttt{Temperature}, in degrees Fahrenheit, and \texttt{Ozone} level, in parts per billion.
+For simplicity and to provide a minimal working example, we consider
+bivariate simulation of the two \texttt{airquality} variables
+\texttt{Temperature}, in degrees Fahrenheit, and \texttt{Ozone} level,
+in parts per billion.
 
@@
 \begin{CodeChunk}
-\begin{CodeInput}
-R> glimpse(df)
-\end{CodeInput}
 \begin{CodeOutput}
@@
 
-Figure \ref{fig:ch030-aq-joint-dist} visualizes the bivariate relationship between Ozone and Temperature.
-We aim to simulate random two-component vectors mimicking this structure.
-The margins are not normally distributed; particularly the Ozone level exhibits a strong positive skew.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> p0 <- ggplot(df, aes(Temp, Ozone)) +
-+   geom_point(size = 1) +
-+   theme(legend.position = "none") + 
-+   labs(x = "Temperature")
-R> 
-R> pTemp <- ggplot(df, aes(Temp)) + 
-+   geom_density() +
-+   theme(axis.title.x = element_blank(),
-+         axis.title.y = element_blank(),
-+         axis.text = element_blank(),
-+         axis.ticks = element_blank())
-R> 
-R> pOzone <- ggplot(df, aes(Ozone)) + 
-+     geom_density() +
-+   theme(axis.title.y = element_blank(),
-+         axis.title.x = element_blank(),
-+         axis.ticks = element_blank(),
-+         axis.text = element_blank()) +
-+   coord_flip()
-R> 
-R> pTemp + plot_spacer() + p0 + pOzone + 
-+   plot_layout(widths = c(3,1), heights = c(1, 3)) 
-\end{CodeInput}
+Figure @ref(fig:ch030-aq-joint-dist) visualizes the bivariate
+relationship between Ozone and Temperature. We aim to simulate random
+two-component vectors mimicking this structure. The margins are not
+normally distributed; particularly the Ozone level exhibits a strong
+positive skew.
+
+\begin{CodeChunk}
 \begin{figure}
 
-{\centering \includegraphics{article_bigsimr_files/figure-latex/ch030-aq-joint-dist-1}
+{\centering \includegraphics[width=0.85\linewidth]{article_bigsimr_files/figure-latex/ch030-aq-joint-dist-1}
 
@@
 
-Next, we specify the marginal distributions and correlation coefficient (both type and magnitude).
-Here the analyst is free to be creative.
-For this example, we avoid goodness-of-fit considerations to determine the marginal distributions.
-But it seems sensible without domain knowledge to estimate these quantities from the data, and \texttt{bigsimr} contains fast functions designed for this task.
+Next, we specify the marginal distributions and correlation coefficient
+(both type and magnitude). Here the analyst is free to be creative. For
+this example, we avoid goodness-of-fit considerations to determine the
+marginal distributions. But it seems sensible without domain knowledge
+to estimate these quantities from the data, and \texttt{bigsimr}
+contains fast functions designed for this task.
 
 \hypertarget{specifying-marginal-distributions}{%
-\subsection{Specifying marginal distributions}\label{specifying-marginal-distributions}}
-
-Based on the estimated densities in Figure \ref{fig:ch030-aq-joint-dist}, we assume \texttt{Temp} is normally distributed and \texttt{Ozone} is log-normally distributed, as the latter values are positive and skewed.
-We use the well-known, unbiased estimators for the normal distribution's parameters and maximum likelihood estimators for the log-normal parameters:
+\subsection{Specifying marginal
+distributions}\label{specifying-marginal-distributions}}
+
+Based on the estimated densities in Figure
+@ref(fig:ch030-aq-joint-dist), we assume \texttt{Temp} is normally
+distributed and \texttt{Ozone} is log-normally distributed, as the
+latter values are positive and skewed. We use the well-known, unbiased
+estimators for the normal distribution's parameters and maximum
+likelihood estimators for the log-normal parameters:
 
@@
 
-Next, we configure the input marginals for later input into \texttt{rvec}. The marginal distributions are specified using \texttt{Julia}'s \texttt{Distributions} package and stored in a vector.
+Next, we configure the input marginals for later input into
+\texttt{rvec}. The marginal distributions are specified using
+\texttt{Julia}'s \texttt{Distributions} package and stored in a vector.
 
@@
 
-As mentioned, the user must decide how to describe correlation based on the particulars of the problem.
-For non-normal data and for improved simulation accuracy/scalability in our scheme, we advocate the use of Spearman's \(\rho\) correlation matrix \(R_S\) and Kendall's \(\tau\) correlation matrix \(R_K\).
-We also support Pearson correlation coefficient matching, while cautioning the user to check the performance for their parametric multivariate model (see \protect\hyperlink{simulations}{Monte Carlo evaluations} below for evaluation strategies and guidance).
-These estimation methods are classical approaches, not designed for high-dimensional correlation estimation (see the \href{\%7B\#discussion}{Conclusion and Discussion} sections for more on this).
+As mentioned, the user must decide how to describe correlation based on
+the particulars of the problem. For non-normal data and for improved
+simulation accuracy/scalability in our scheme, we advocate the use of
+Spearman's \(\rho\) correlation matrix \(R_S\) and Kendall's \(\tau\)
+correlation matrix \(R_K\). We also support Pearson correlation
+coefficient matching, while cautioning the user to check the performance
+for their parametric multivariate model (see
+\protect\hyperlink{simulations}{Monte Carlo evaluations} below for
+evaluation strategies and guidance). These estimation methods are
+classical approaches, not designed for high-dimensional correlation
+estimation (see the \href{\%7B\#discussion}{Conclusion and Discussion}
+sections for more on this).
 
@@
 \hypertarget{checking-target-correlation-matrix-admissibility}{%
-\subsection{Checking target correlation matrix admissibility}\label{checking-target-correlation-matrix-admissibility}}
-
-First we use \texttt{cor\_bounds} to ensure that the pairwise target correlation values are valid prior to the mapping step which constructs \(R_X\) for the MVN input into NORTA (see Section \ref{algorithms}).
-\texttt{cor\_bounds} estimates the pairwise lower and upper correlation bounds using the Generate, Sort, and Correlate algorithm of \citet{DH2011}.
+\subsection{Checking target correlation matrix
+admissibility}\label{checking-target-correlation-matrix-admissibility}}
+
+First we use \texttt{cor\_bounds} to ensure that the pairwise target
+correlation values are valid prior to the mapping step which constructs
+\(R_X\) for the MVN input into NORTA (see Section @ref(algorithms)).
+\texttt{cor\_bounds} estimates the pairwise lower and upper correlation
+bounds using the Generate, Sort, and Correlate algorithm of
+\citet{DH2011}.
 
@@
         [,1]    [,2]
-[1,]  1.0000 -0.8902
-[2,] -0.8902  1.0000
+[1,]  1.0000 -0.8745
+[2,] -0.8745  1.0000
 \end{CodeOutput}
@@
        [,1]   [,2]
-[1,] 1.0000 0.8876
-[2,] 0.8876 1.0000
+[1,] 1.0000 0.8748
+[2,] 0.8748 1.0000
 \end{CodeOutput}
@@
 
-Since our single estimated Spearman correlation is within the theoretical bounds, the correlation is valid as input to \texttt{rvec}.
-But even if the 2-dimensional bounds are satisfied for each pair of margins, this does not guarantee the feasibility of a \(d-\)variate distribution \citep{BF17}.
-
-To provide higher dimensional feasibility/admissibility checking, we begin by mapping using \texttt{cor\_convert} and check admissibility.
+Since our single estimated Spearman correlation is within the
+theoretical bounds, the correlation is valid as input to \texttt{rvec}.
+But even if the 2-dimensional bounds are satisfied for each pair of
+margins, this does not guarantee the feasibility of a \(d-\)variate
+distribution \citep{BF17}.
+
+To provide higher dimensional feasibility/admissibility checking, we
+begin by mapping using \texttt{cor\_convert} and check admissibility.
 
@@
 
-The bounds on the Pearson correlation coefficient \(R_P\) between these margins are restricted as seen below in our MC estimated correlation bounds:
+The bounds on the Pearson correlation coefficient \(R_P\) between these
+margins are restricted as seen below in our MC estimated correlation
+bounds:
 
@@
 Julia Object of type NamedTuple{(:lower, :upper),Tuple{Float64,Float64}}.
-(lower = -0.8822730883942045, upper = 0.8833146839651883)
+(lower = -0.8799819883897544, upper = 0.8798667127065275)
 \end{CodeOutput}
@@
 
-Our MC estimate of the bounds slightly overestimates the theoretic bounds of \((-0.881, 0.881)\). An analytic derivation is presented as an Appendix.
+Our MC estimate of the bounds slightly overestimates the theoretic
+bounds of \((-0.881, 0.881)\). An analytic derivation is presented as an
+Appendix.
 
@@
 
-Finally, we arrive at the main function of \texttt{bigsimr}: \texttt{rvec}. We now simulate \(B=10,000\) random vectors from the assumed joint distribution of Ozone levels and Temp.
+Finally, we arrive at the main function of \texttt{bigsimr}:
+\texttt{rvec}. We now simulate \(B=10,000\) random vectors from the
+assumed joint distribution of Ozone levels and Temp.
 
@@
 
-Figure \ref{fig:ch030-plot-sim} plots the 10,000 simulated points.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> p1 <- df_sim %>%
-+   ggplot(aes(Temp, Ozone)) +
-+   geom_density_2d_filled() +
-+   theme(legend.position = "none") +
-+   labs(x = "Simulated Temperature", y = "Simulated Ozone") +
-+   scale_y_continuous(limits = c(0, max(df$Ozone))) +
-+   scale_x_continuous(limits = range(df$Temp))
-R> 
-R> p1Temp <- ggplot(df_sim, aes(Temp)) + 
-+   geom_density(alpha = 0.5, fill = "lightseagreen") +
-+   theme(axis.title.x = element_blank(),
-+         axis.title.y = element_blank(),
-+         axis.text = element_blank(),
-+         axis.ticks = element_blank())
-R> 
-R> p1Ozone <- ggplot(df_sim, aes(Ozone)) + 
-+   geom_density(alpha = 0.5, fill = "lightseagreen") + 
-+   theme(axis.title.y = element_blank(),
-+         axis.title.x = element_blank(),
-+         axis.ticks = element_blank(),
-+         axis.text = element_blank()) +
-+   coord_flip() 
-R> 
-R> p1Temp + plot_spacer() + p1 + p1Ozone + 
-+   plot_layout(widths = c(3,1), heights = c(1, 3))
-\end{CodeInput}
-\begin{CodeOutput}
-Warning: Removed 342 rows containing non-finite values (stat_density2d_filled).
-\end{CodeOutput}
+Figure @ref(fig:ch030-plot-sim) plots the 10,000 simulated points.
+
+\begin{CodeChunk}
 \begin{figure}
 
-{\centering \includegraphics{article_bigsimr_files/figure-latex/ch030-plot-sim-1}
+{\centering \includegraphics[width=0.85\linewidth]{article_bigsimr_files/figure-latex/ch030-plot-sim-1}
 
@@
 
-Before applying our methodology to real data simulation, we conduct several Monte Carlo studies to investigate method performance.
-Since marginal parameter matching in our scheme is essentially a sequence of univariate inverse probability transforms, the challenging aspects are the accuracy of dependency matching and computational efficiency at high dimensions.
-To evaluate our methods in those respects, we design the following numerical experiments to first assess accuracy of matching dependency parameters in bivariate simulations and then time the procedure in increasingly large dimension \(d\).
+Before applying our methodology to real data simulation, we conduct
+several Monte Carlo studies to investigate method performance. Since
+marginal parameter matching in our scheme is essentially a sequence of
+univariate inverse probability transforms, the challenging aspects are
+the accuracy of dependency matching and computational efficiency at high
+dimensions. To evaluate our methods in those respects, we design the
+following numerical experiments to first assess accuracy of matching
+dependency parameters in bivariate simulations and then time the
+procedure in increasingly large dimension \(d\).
 
@@
 
-We select bivariate simulation configurations to ultimately simulate our motivating discrete-valued RNA-seq example, and, so we proceed in increasing complexity, leading to the model in our motivating application in Section \ref{examples}.
-We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson's, Spearman's, and Kendall's --- in identical, bivariate marginal configurations.
-For each pair of identical margins, we vary the target correlation across \(\Omega\), the set of possible admissible values for correlation type, to evaluate the simulation's ability to obtain the theoretic bounds.
-The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (mimicking RNA-seq counts).
-
-Table \ref{tab:sims} lists our identical-marginal, bivariate simulation configurations.
-We increase the simulate replicates \(B\) to check that our results converge to the target correlations and gauge statistical efficiency.
-We select distributions beginning with a standard multivariate normal (MVN) as we expect the performance to be exact (up to MC error) for all correlation types.
-Then, we select a non-symmetric continuous distribution: a standard (rate =1) two-component multivariate gamma (MVG).
-Finally, we select distributions and marginal parameter values that are motivated by our RNA-seq data, namely values proximal to probabilities and sizes estimated from the data (see \href{examples}{Example applications} for estimation details).
-Thus we arrive at a multivariate negative binomial (MVNB) \(p_1 = p_2 = 3\times10^{-4}, r_1 = r_2 = 4, \rho \in \Omega\).
-
-Table: \label{tab:sims} Identical margin, bivariate simulation configurations to evaluate correlation matching accuracy and efficiency.
+We select bivariate simulation configurations to ultimately simulate our
+motivating discrete-valued RNA-seq example, and, so we proceed in
+increasing complexity, leading to the model in our motivating
+application in Section @ref(examples). We begin with empirically
+evaluating the dependency matching across all three supported
+correlations --- Pearson's, Spearman's, and Kendall's --- in identical,
+bivariate marginal configurations. For each pair of identical margins,
+we vary the target correlation across \(\Omega\), the set of possible
+admissible values for correlation type, to evaluate the simulation's
+ability to obtain the theoretic bounds. The simulations progress from
+bivariate normal, to bivariate gamma (non-normal yet continuous), and
+bivariate negative binomial (mimicking RNA-seq counts).
+
+Table @ref(tab:sims) lists our identical-marginal, bivariate simulation
+configurations. We increase the simulate replicates \(B\) to check that
+our results converge to the target correlations and gauge statistical
+efficiency. We select distributions beginning with a standard
+multivariate normal (MVN) as we expect the performance to be exact (up
+to MC error) for all correlation types. Then, we select a non-symmetric
+continuous distribution: a standard (rate =1) two-component multivariate
+gamma (MVG). Finally, we select distributions and marginal parameter
+values that are motivated by our RNA-seq data, namely values proximal to
+probabilities and sizes estimated from the data (see
+\href{examples}{Example applications} for estimation details). Thus we
+arrive at a multivariate negative binomial (MVNB)
+\(p_1 = p_2 = 3\times10^{-4}, r_1 = r_2 = 4, \rho \in \Omega\).
+
+Table: (\#tab:sims) Identical margin, bivariate simulation
+configurations to evaluate correlation matching accuracy and efficiency.
 
@@
 
-For each of the unique 9 simulation configurations described above, we estimate the correlation bounds and vary the correlations along a sequence of 100 points evenly placed within the bounds, aiming to explore \(\Omega\).
-Specifically, we set correlations \(\{ \rho_1 = ( \hat{l} + \epsilon), \rho_2 = (\hat{l} + \epsilon) + \delta, \ldots, \rho_{100} = (\hat{u} - \epsilon) \}\), with \(\hat{l}\) and \(\hat{u}\) being the estimated lower and upper bounds, respectively, and increment value \(\delta\).
-The adjustment factor, \(\epsilon=0.01\), is introduced to handle numeric issues when the bound is specified exactly.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> bivariate_normal_sims %>%
-+     ggplot(aes(rho, rho_hat, color = type)) +
-+     geom_point() +
-+     geom_abline(slope = 1) +
-+     facet_wrap(~ type + N) + 
-+     theme_bw()
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> bivariate_gamma_sims %>%
-+     ggplot(aes(rho, rho_hat, color = type)) +
-+     geom_point() +
-+     geom_abline(slope = 1) +
-+     facet_wrap(~ type + N) + 
-+     theme_bw()
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> bivariate_nbinom_sims %>%
-+     ggplot(aes(rho, rho_hat, color = type)) +
-+     geom_point() +
-+     geom_abline(slope = 1) +
-+     facet_wrap(~ type + N) + 
-+     theme_bw()
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> allDat <- bind_rows(
-+     select(bivariate_normal_sims, margins, type, N, rho, rho_hat),
-+     select(bivariate_gamma_sims, margins, type, N, rho, rho_hat),
-+     select(bivariate_nbinom_sims, margins, type, N, rho, rho_hat)
-+ ) %>%
-+     mutate(margins = factor(margins, levels=c("norm", "gamma", "nbinom")),
-+            type = factor(type, levels=c("Pearson", "Spearman", "Kendall")),
-+            N = factor(N))
-\end{CodeInput}
-\end{CodeChunk}
-
-Figure \ref{fig:ch040-bPlot} displays the aggregated bivariate simulation results.
-Table \ref{tab:ch040-BiError} contains the mean absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> tabMAE <- allDat %>%
-+     group_by(N, type, margins) %>%
-+     summarize(MAE = mean(abs(rho - rho_hat))) %>%
-+     ungroup()
-\end{CodeInput}
-\begin{CodeOutput}
-`summarise()` has grouped output by 'N', 'type'. You can override using the `.groups` argument.
-\end{CodeOutput}
-\begin{CodeInput}
-R> kable(tabMAE, booktabs = TRUE, format = "latex",
-+       linesep = c("", "", "\\addlinespace"),
-+       col.names = c("No. of random vectors",
-+                     "Correlation type",
-+                     "Distribution",
-+                     "Mean abs. error"),
-+       caption = "Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.")
-\end{CodeInput}
+For each of the unique 9 simulation configurations described above, we
+estimate the correlation bounds and vary the correlations along a
+sequence of 100 points evenly placed within the bounds, aiming to
+explore \(\Omega\). Specifically, we set correlations
+\(\{ \rho_1 = ( \hat{l} + \epsilon), \rho_2 = (\hat{l} + \epsilon) + \delta, \ldots, \rho_{100} = (\hat{u} - \epsilon) \}\),
+with \(\hat{l}\) and \(\hat{u}\) being the estimated lower and upper
+bounds, respectively, and increment value \(\delta\). The adjustment
+factor, \(\epsilon=0.01\), is introduced to handle numeric issues when
+the bound is specified exactly.
+
+Figure @ref(fig:ch040-bPlot) displays the aggregated bivariate
+simulation results. Table @ref(tab:ch040-BiError) contains the mean
+absolute error (MAE) in reproducing the desired dependency measures for
+the three bivariate scenarios.
+
+\begin{CodeChunk}
 \begin{table}
@@
 
-Overall, the studies show that our methodology is generally accurate across the entire range of possible correlation values all three dependency measures, at least in these limited simulation settings for the rank-based correlations.
-Our Pearson matching performs nearly as well as Spearman or Kendall, except for a slight increase in error for negative binomial case.
-This is due the particularly large counts generated with our choice of parameters for \(p\) and \(r\).
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> # https://www.datanovia.com/en/blog/how-to-change-ggplot-facet-labels/
-R> # New facet label names
-R> repsLabs <- paste0("B=", c("1,000", "10,000", "100,000") )
-R> names(repsLabs) <- c(1000, 10000, 100000)
-R> typeLabs <- c("Pearson", "Spearman", "Kendall")
-R> names(typeLabs) <- c("Pearson", "Spearman", "Kendall")
-R> 
-R> # Set colors
-R> ## RColorBrewer::display.brewer.all()
-R> numColors <- 4
-R> numGroups <- length(levels(allDat$margins))
-R> myColors <- rev(RColorBrewer::brewer.pal(n = numColors, name = "Greys")[ ((numColors - numGroups) + 1):numColors])
-R> 
-R> allDat %>%
-+     ggplot(aes(x = rho, y = rho_hat, color = margins)) +
-+     geom_point(size = 2) +
-+     scale_color_manual(values = myColors ) +
-+     geom_abline(slope = 1, linetype = "dashed") +
-+     labs(x = "Specified Correlation", y = "Estimated Correlation") +
-+     facet_wrap(~ type + N, labeller = labeller(N = repsLabs, type = typeLabs)) +
-+     theme_bw() +
-+     theme(legend.position = "bottom", legend.direction = "horizontal")
-\end{CodeInput}
+Overall, the studies show that our methodology is generally accurate
+across the entire range of possible correlation values all three
+dependency measures, at least in these limited simulation settings for
+the rank-based correlations. Our Pearson matching performs nearly as
+well as Spearman or Kendall, except for a slight increase in error for
+negative binomial case. This is due the particularly large counts
+generated with our choice of parameters for \(p\) and \(r\).
+
+\begin{CodeChunk}
 \begin{figure}
 
-{\centering \includegraphics{article_bigsimr_files/figure-latex/ch040-bPlot-1}
+{\centering \includegraphics[width=0.85\linewidth]{article_bigsimr_files/figure-latex/ch040-bPlot-1}
 
@@
 \hypertarget{scale-up-to-high-dimensions}{%
-\subsection{Scale up to High Dimensions}\label{scale-up-to-high-dimensions}}
-
-With information of our method's accuracy from a low-dimensional perspective, we now turn to assessing whether \texttt{bigsimr} can scale to larger dimensional problems with practical computation times.
-Specifically, we ultimately generate \(B=1,000\) random vectors for \(d=\{100, 250, 500, 1000, 2500, 5000, 10000\}\) for each correlation type, \(\{Pearson, Spearman, Kendall\}\) while timing the algorithm's major steps.
-
-To mimic the workflow, we first produce a synthetic ``data set'' by completing the following steps:
+\subsection{Scale up to High
+Dimensions}\label{scale-up-to-high-dimensions}}
+
+With information of our method's accuracy from a low-dimensional
+perspective, we now turn to assessing whether \texttt{bigsimr} can scale
+to larger dimensional problems with practical computation times.
+Specifically, we ultimately generate \(B=1,000\) random vectors for
+\(d=\{100, 250, 500, 1000, 2500, 5000, 10000\}\) for each correlation
+type, \(\{Pearson, Spearman, Kendall\}\) while timing the algorithm's
+major steps.
+
+To mimic the workflow, we first produce a synthetic ``data set'' by
+completing the following steps:
 
@@
 \item
-  Produce heterogeneous gamma marginals by randomly selecting the \(j^{th}\) gamma shape parameter from \(U_j \sim uniform(1,10), j=1,\ldots,d\) and the \(j^{th}\) rate parameter from \(V_j \sim exp(1/5), j=1,\ldots,d\), with the constant parameters determined arbitrarily.
-\item
-  Produce a random full-rank Pearson correlation matrix via \texttt{cor\_randPD} of size \(d \times d\).
-\item
-  Simulate a ``data set'' of \(1,000 \times d\) random vectors via \texttt{rvec} (without matching the Pearson exactly).
-  \setstretch{2.0}
+  Produce heterogeneous gamma marginals by randomly selecting the
+  \(j^{th}\) gamma shape parameter from
+  \(U_j \sim uniform(1,10), j=1,\ldots,d\) and the \(j^{th}\) rate
+  parameter from \(V_j \sim exp(1/5), j=1,\ldots,d\), with the constant
+  parameters determined arbitrarily.
+\item
+  Produce a random full-rank Pearson correlation matrix via
+  \texttt{cor\_randPD} of size \(d \times d\).
+\item
+  Simulate a ``data set'' of \(1,000 \times d\) random vectors via
+  \texttt{rvec} (without matching the Pearson exactly). \setstretch{2.0}
 \end{enumerate}
 
-With the synthetic data set in hand, we complete and time the following 4 steps:
+With the synthetic data set in hand, we complete and time the following
+4 steps:
 
@@
 \item
-  Estimate the correlation matrix from the ``data'' in the \emph{Compute Correlation} step.
+  Estimate the correlation matrix from the ``data'' in the \emph{Compute
+  Correlation} step.
 \end{enumerate}
@@
 \item
-  Map the correlations to initialize the algorithm (Pearson to Pearson, Spearman to Pearson, or Kendall to Pearson) in the \emph{Adjust Correlation} step.
-\item
-  Check whether the mapping produces a valid correlation matrix and, if not, find the nearest PD correlation matrix in the \emph{Check Admissibility} step.
-\item
-  Simulate \(1,000\) vectors via the Inverse Transform method in the \emph{Simulate Data} step.
-  \setstretch{2.0}
+  Map the correlations to initialize the algorithm (Pearson to Pearson,
+  Spearman to Pearson, or Kendall to Pearson) in the \emph{Adjust
+  Correlation} step.
+\item
+  Check whether the mapping produces a valid correlation matrix and, if
+  not, find the nearest PD correlation matrix in the \emph{Check
+  Admissibility} step.
+\item
+  Simulate \(1,000\) vectors via the Inverse Transform method in the
+  \emph{Simulate Data} step. \setstretch{2.0}
 \end{enumerate}
 
-The experiments are conducted on a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, with all 16 threads employed during computation.
-The results for \(d \leq 500\) are fast with all times (except Kendall) under 3 seconds.
-Table \ref{tab:ch040-moderateDtab} displays the total computation time of the algorithms steps 1 through 3, including estimation step i.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> dat_long <- benchmark_dependences %>%
-+   select(-total_time, -n_sim, -needed_near_pd) %>%
-+   pivot_longer(cols = c(corr_time:sim_time), names_to = "Step", values_to = "Time") %>%
-+   mutate(dim = factor(dim),
-+          Step = factor(Step,
-+                        levels = c("corr_time", "adj_time", "admiss_time", "sim_time"),
-+                        labels = c("Compute Correlation",
-+                                   "Adjust Correlation",
-+                                   "Check Admissibility",
-+                                   "Simulate Data"))) %>%
-+     rename(Correlation = corr_type, Dimensions = dim)
-R> 
-R> tabTimes  <- dat_long %>%
-+     filter(Dimensions %in% c(100, 250, 500)) %>%
-+     group_by(Dimensions, Correlation) %>%
-+     summarize('Total Time(Seconds)' = sum(Time)) %>%
-+     ungroup()
-\end{CodeInput}
-\begin{CodeOutput}
-`summarise()` has grouped output by 'Dimensions'. You can override using the `.groups` argument.
-\end{CodeOutput}
-\begin{CodeInput}
-R> tabTimes %>%
-+     kable(booktabs = TRUE,
-+           linesep = c("", "", "\\addlinespace"),
-+           col.names = c("Dimension",
-+                         "Correlation type",
-+                         "Total Time (Seconds)"),
-+           caption = 'Total time to produce 10,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.')
-\end{CodeInput}
-\begin{table}
-
-\caption{\label{tab:ch040-moderateDtab}Total time to produce 10,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.}
-\centering
-\begin{tabular}[t]{llr}
-\toprule
-Dimension & Correlation type & Total Time (Seconds)\\
-\midrule
-100 & Pearson & 0.080\\
-100 & Spearman & 0.036\\
-100 & Kendall & 0.116\\
-\addlinespace
-250 & Pearson & 0.286\\
-250 & Spearman & 0.371\\
-250 & Kendall & 0.366\\
-\addlinespace
-500 & Pearson & 2.932\\
-500 & Spearman & 0.176\\
-500 & Kendall & 2.183\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\end{CodeChunk}
-
-The results with larger dimensional vectors show scalability to ultra-high dimensions for all three correlation types, although the total times do become much larger.
-Figure \ref{fig:ch040-largeDfig} displays computation times for \(d=\{1000, 2500, 5000, 10000\}\).
-For \(d\) equal to 1000 and 2500, the total time is under a couple of minutes.
-At \(d\) of 5000 and 10,000, Pearson correlation matching in the \emph{Adjust Correlation} step becomes costly.
-Interestingly, Pearson is actually faster than Kendall for \(d=10,000\) due to bottlenecks in \emph{Compute Correlation} and \emph{Check Admissibility}.
-Uniformly, matching Spearman correlations is faster, with total times under 5 minutes for \(d=10,000\), making Spearman correlations the most computationally-friendly dependency type.
-With this in mind, we scaled the simulation to \(d=20,000\) for the Spearman type and obtained the 1,000 vectors in under an hour (data not shown).
-In principle, this would enable the simulation of an entire human-derived RNA-seq data set.
-We note that for a given target correlation matrix and margins, steps i, 1, and 2 only need to be computed once and the third step, \emph{Simulate Data}, is fast under all schemes for all \(d\) under consideration.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> # Set colors
-R> numColors <- 5
-R> numGroups <- 4
-R> myColors <- rev(brewer.pal(n=numColors, name="Greys")[((numColors-numGroups)+1):numColors])
-R> 
-R> dat_long %>%
-+     filter(Dimensions %in% c(1000, 2500, 5000, 10000, 20000)) %>%
-+     mutate(`Time (minutes)` = Time / 60) %>%
-+     ggplot(aes(Correlation, `Time (minutes)`, fill=Step)) +
-+     geom_bar(position = "stack", stat = "identity") +
-+     scale_fill_manual(values = myColors ) +
-+     scale_y_continuous(breaks = seq(0, 80, 10),
-+                      minor_breaks = NULL) +
-+     facet_grid(. ~ Dimensions) +
-+     theme(axis.text.x = element_text(angle = -90))
-\end{CodeInput}
+The experiments are conducted on a MacBook Pro carrying a 2.4 GHz 8-Core
+Intel Core i9 processor, with all 16 threads employed during
+computation. The results for \(d \leq 500\) are fast with all times
+(except Kendall) under 3 seconds. Table @ref(tab:ch040-moderateDtab)
+displays the total computation time of the algorithms steps 1 through 3,
+including estimation step i.
+
+\begin{CodeChunk}
+
+Table: Total time to produce 10,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.
+
+|Dimension |Correlation type | Total Time (Seconds)|
+|:---------|:----------------|--------------------:|
+|100       |Pearson          |                0.080|
+|100       |Spearman         |                0.036|
+|100       |Kendall          |                0.116|
+|250       |Pearson          |                0.286|
+|250       |Spearman         |                0.371|
+|250       |Kendall          |                0.366|
+|500       |Pearson          |                2.932|
+|500       |Spearman         |                0.176|
+|500       |Kendall          |                2.183|
+
+\end{CodeChunk}
+
+The results with larger dimensional vectors show scalability to
+ultra-high dimensions for all three correlation types, although the
+total times do become much larger. Figure @ref(fig:ch040-largeDfig)
+displays computation times for \(d=\{1000, 2500, 5000, 10000\}\). For
+\(d\) equal to 1000 and 2500, the total time is under a couple of
+minutes. At \(d\) of 5000 and 10,000, Pearson correlation matching in
+the \emph{Adjust Correlation} step becomes costly. Interestingly,
+Pearson is actually faster than Kendall for \(d=10,000\) due to
+bottlenecks in \emph{Compute Correlation} and \emph{Check
+Admissibility}. Uniformly, matching Spearman correlations is faster,
+with total times under 5 minutes for \(d=10,000\), making Spearman
+correlations the most computationally-friendly dependency type. With
+this in mind, we scaled the simulation to \(d=20,000\) for the Spearman
+type and obtained the 1,000 vectors in under an hour (data not shown).
+In principle, this would enable the simulation of an entire
+human-derived RNA-seq data set. We note that for a given target
+correlation matrix and margins, steps i, 1, and 2 only need to be
+computed once and the third step, \emph{Simulate Data}, is fast under
+all schemes for all \(d\) under consideration.
+
+\begin{CodeChunk}
 \begin{figure}
 
-{\centering \includegraphics{article_bigsimr_files/figure-latex/ch040-largeDfig-1}
+{\centering \includegraphics[width=0.85\linewidth]{article_bigsimr_files/figure-latex/ch040-largeDfig-1}
 
@@
 
-In the bivariate studies, we chose arbitrary simulation parameters for three distributions, moving from the Gaussian to the discrete and non-normal, multivariate negative binomial.
-Under these conditions, the simulated random vectors sample the desired bivariate distribution across the entire range of pairwise correlations for the three dependency measures.
-The simulation results could differ for other choices of simulation settings.
-Specifying extreme correlations near the boundary or Frechet bounds could result in poor simulation performance.
-Fortunately, it is straightforward to evaluate simulation performance by using strategies similar to those completed above.
-We expect our random vector generation to perform well for the vast majority of NORTA-feasible correlation matrices, but advise to check the performance before making inferences/further analyses.
-
-Somewhat surprising, Kendall estimation and nearest PD computation scale poorly compared to Spearman and, even, approximate Pearson matching.
-In our experience, Kendall computation times are sensitive to the number of cores, benefiting from multi-core parallelization.
-This could mitigate some of the current algorithmic/implementation shortcomings.
-Despite this, Kendall matching is still feasible for most high-dimensional data sets.
-Finally, we note that one could use our single-pass algorithm \texttt{cor\_fastPD} to produce a `close' (not nearest PD) to scale to even higher dimensions with greater loss of accuracy.
+In the bivariate studies, we chose arbitrary simulation parameters for
+three distributions, moving from the Gaussian to the discrete and
+non-normal, multivariate negative binomial. Under these conditions, the
+simulated random vectors sample the desired bivariate distribution
+across the entire range of pairwise correlations for the three
+dependency measures. The simulation results could differ for other
+choices of simulation settings. Specifying extreme correlations near the
+boundary or Frechet bounds could result in poor simulation performance.
+Fortunately, it is straightforward to evaluate simulation performance by
+using strategies similar to those completed above. We expect our random
+vector generation to perform well for the vast majority of
+NORTA-feasible correlation matrices, but advise to check the performance
+before making inferences/further analyses.
+
+Somewhat surprising, Kendall estimation and nearest PD computation scale
+poorly compared to Spearman and, even, approximate Pearson matching. In
+our experience, Kendall computation times are sensitive to the number of
+cores, benefiting from multi-core parallelization. This could mitigate
+some of the current algorithmic/implementation shortcomings. Despite
+this, Kendall matching is still feasible for most high-dimensional data
+sets. Finally, we note that one could use our single-pass algorithm
+\texttt{cor\_fastPD} to produce a `close' (not nearest PD) to scale to
+even higher dimensions with greater loss of accuracy.
 
@@
 
-This section demonstrates how to simulate multivariate data using \texttt{bigsimr}, aiming to replicate the structure of high-dimensional dependent count data.
-In an illustration of our proposed methodology, we seek to simulate RNA-sequencing data by producing simulated random vectors mimicking the observed data and its generating process.
-Modeling RNA-seq using multivariate probability distributions is natural as inter-gene correlation is an inherent part of biological processes \citep{Wang2009b}.
-And yet, many models do not account for this, leading to major disruptions to the operating characteristics of statistical estimation, testing, and prediction.
-See \citet{Efron2012} for a detailed discussion with related methods and \citet{Wu2012b}, \citet{Schissler2018}; \citet{Schissler2019} for applied examples.
-The following subsections apply \texttt{bigsimr}'s methods to real RNA-seq data, including replicating an estimated parametric structure, MC probability estimation, and MC evaluation of correlation estimation efficiency.
+This section demonstrates how to simulate multivariate data using
+\texttt{bigsimr}, aiming to replicate the structure of high-dimensional
+dependent count data. In an illustration of our proposed methodology, we
+seek to simulate RNA-sequencing data by producing simulated random
+vectors mimicking the observed data and its generating process. Modeling
+RNA-seq using multivariate probability distributions is natural as
+inter-gene correlation is an inherent part of biological processes
+\citep{Wang2009b}. And yet, many models do not account for this, leading
+to major disruptions to the operating characteristics of statistical
+estimation, testing, and prediction. See \citet{Efron2012} for a
+detailed discussion with related methods and \citet{Wu2012b},
+\citet{Schissler2018}; \citet{Schissler2019} for applied examples. The
+following subsections apply \texttt{bigsimr}'s methods to real RNA-seq
+data, including replicating an estimated parametric structure, MC
+probability estimation, and MC evaluation of correlation estimation
+efficiency.
 
 \hypertarget{simulating-high-dimensional-rna-seq-data}{%
-\subsection{Simulating High-Dimensional RNA-seq data}\label{simulating-high-dimensional-rna-seq-data}}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> d <- 1000
-R> brca1000 <- example_brca %>%
-+   select(all_of(1:d)) %>%
-+   mutate(across(everything(), as.double))
-\end{CodeInput}
-\end{CodeChunk}
-
-We begin by estimating the structure of the TCGA BRCA RNA-seq data set (see \href{background}{Background}).
-Ultimately, we will simulate \(B=10,000\) random vectors \({\bf Y}=(Y_1, \ldots, Y_d)^\top\) with \(d=1000\).
-We assume a multivariate negative binomial (MVNB) model as RNA-seq counts are often over-dispersed and correlated.
-Since all \(d\) selected genes exhibit over-dispersion (data not shown), we proceed to estimate the NB parameters \((r_i, p_i), i=1,\ldots,d\), to determine the target marginal PMFs \(f_i\).
-To complete specification of the simulation algorithm inputs, we estimate the Spearman correlation matrix \({ \bf R}_{S}\) to characterize dependency.
-
-With this goal in mind, we first estimate the desired correlation matrix using the fast implementation provided by \texttt{bigsimr}:
+\subsection{Simulating High-Dimensional RNA-seq
+data}\label{simulating-high-dimensional-rna-seq-data}}
+
+We begin by estimating the structure of the TCGA BRCA RNA-seq data set
+(see \href{background}{Background}). Ultimately, we will simulate
+\(B=10,000\) random vectors \({\bf Y}=(Y_1, \ldots, Y_d)^\top\) with
+\(d=1000\). We assume a multivariate negative binomial (MVNB) model as
+RNA-seq counts are often over-dispersed and correlated. Since all \(d\)
+selected genes exhibit over-dispersion (data not shown), we proceed to
+estimate the NB parameters \((r_i, p_i), i=1,\ldots,d\), to determine
+the target marginal PMFs \(f_i\). To complete specification of the
+simulation algorithm inputs, we estimate the Spearman correlation matrix
+\({ \bf R}_{S}\) to characterize dependency.
+
+With this goal in mind, we first estimate the desired correlation matrix
+using the fast implementation provided by \texttt{bigsimr}:
 
@@
 
-Next, we estimate the marginal parameters.
-We use the method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model.
-While, marginal distributions are from the same probability family (NB), they are heterogeneous in terms of the parameters probability and size \((p_i, n_i)\) for \(i,\ldots,d\).
-The functions below support this estimation for later use in \texttt{rvec}.
+Next, we estimate the marginal parameters. We use the method of moments
+(MoM) to estimate the marginal parameters for the multivariate negative
+binomial model. While, marginal distributions are from the same
+probability family (NB), they are heterogeneous in terms of the
+parameters probability and size \((p_i, n_i)\) for \(i,\ldots,d\). The
+functions below support this estimation for later use in \texttt{rvec}.
 
@@
 
-Notably, the estimated marginal NB probabilities \(\{ \hat{p}_i \}\) are small --- ranging in the interval \([\ensuremath{7.5305\times 10^{-7}} , \ensuremath{5.6592\times 10^{-4}}]\).
-This gives rise to highly variable counts and, typically, less restriction on potential pairwise correlation pairs.
-Once the functions are defined/executed to complete marginal estimation, we specify targets and generate the desired random vectors using \texttt{rvec}.
-Now we check admissibility of the specified correlation matrix.
+Notably, the estimated marginal NB probabilities \(\{ \hat{p}_i \}\) are
+small --- ranging in the interval
+\([\ensuremath{7.5305\times 10^{-7}} , \ensuremath{5.6592\times 10^{-4}}]\).
+This gives rise to highly variable counts and, typically, less
+restriction on potential pairwise correlation pairs. Once the functions
+are defined/executed to complete marginal estimation, we specify targets
+and generate the desired random vectors using \texttt{rvec}. Now we
+check admissibility of the specified correlation matrix.
 
@@
 
-While the exact \(d \times d\) Spearman correlation matrix is not strictly admissible in our scheme (as seen by the non-positive definite result above), the approximation is close with a maximum absolute error of 0.0069 and average absolute error of \ensuremath{5.3366\times 10^{-4}} across the \ensuremath{4.995\times 10^{5}} correlations.
+While the exact \(d \times d\) Spearman correlation matrix is not
+strictly admissible in our scheme (as seen by the non-positive definite
+result above), the approximation is close with a maximum absolute error
+of 0.0069 and average absolute error of
+\ensuremath{5.3366\times 10^{-4}} across the
+\ensuremath{4.995\times 10^{5}} correlations.
 
@@
 
-Figure \ref{fig:ch050-simDataFig} displays the simulated counts and pairwise relationships for our example genes from Table \ref{tab:ch010-realDataTab}.
-Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parametric form and with less extreme points then the observed data in Figure \ref{fig:ch010-realDataFig}.
-Figure \ref{fig:ch050-figBRCA} displays the aggregated results of our simulation by comparing the specified target parameter (horizontal axes) with the corresponding quantities estimated from the simulated data (vertical axes).
-The evaluation shows that the simulated counts approximately match the target parameters and exhibit the full range of estimated correlation from the data.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> set.seed(2020-02-25)
-R> num_genes <- 3
-R> gene_sample <- sample(example_genes[1:1000], num_genes)
-R> 
-R> ggpairs(data = as.data.frame(sim_nbinom[, gene_sample]),
-+         upper = list(continuous = wrap('cor', method = "spearman"))) + 
-+   theme_bw()
-\end{CodeInput}
-\begin{CodeOutput}
-Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
-exact p-value with ties
-\end{CodeOutput}
-\begin{CodeOutput}
-Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
-exact p-value with ties
-\end{CodeOutput}
-\begin{CodeOutput}
-Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
-exact p-value with ties
-\end{CodeOutput}
+Figure @ref(fig:ch050-simDataFig) displays the simulated counts and
+pairwise relationships for our example genes from Table
+@ref(tab:ch010-realDataTab). Simulated counts roughly mimic the observed
+data but with a smoother appearance due to the assumed parametric form
+and with less extreme points then the observed data in Figure
+@ref(fig:ch010-realDataFig). Figure @ref(fig:ch050-figBRCA) displays the
+aggregated results of our simulation by comparing the specified target
+parameter (horizontal axes) with the corresponding quantities estimated
+from the simulated data (vertical axes). The evaluation shows that the
+simulated counts approximately match the target parameters and exhibit
+the full range of estimated correlation from the data.
+
+\begin{CodeChunk}
 \begin{figure}
 
-{\centering \includegraphics{article_bigsimr_files/figure-latex/ch050-simDataFig-1}
+{\centering \includegraphics[width=0.85\linewidth]{article_bigsimr_files/figure-latex/ch050-simDataFig-1}
 
@@
 \begin{CodeChunk}
-\begin{CodeInput}
-R> include_graphics('fig/ch050-figBRCA.png')
-\end{CodeInput}
 \begin{figure}
@@
 
-The results show overall good simulation performance for our choice of parameters settings.
-Our settings were motivated by modeling high-expressing genes from the TCGA BRCA data set.
-In general, the ability to match marginal and dependence parameters depends on the particular joint probability model.
-We recommend to evaluate and tune your simulation until you can be assured of the accuracy.
+The results show overall good simulation performance for our choice of
+parameters settings. Our settings were motivated by modeling
+high-expressing genes from the TCGA BRCA data set. In general, the
+ability to match marginal and dependence parameters depends on the
+particular joint probability model. We recommend to evaluate and tune
+your simulation until you can be assured of the accuracy.
 
 \hypertarget{simulation-based-joint-probability-calculations}{%
-\subsection{Simulation-based joint probability calculations}\label{simulation-based-joint-probability-calculations}}
-
-Many statistical tasks require evaluation of a joint probability mass (or density) function:
+\subsection{Simulation-based joint probability
+calculations}\label{simulation-based-joint-probability-calculations}}
+
+Many statistical tasks require evaluation of a joint probability mass
+(or density) function:
 
@@
 
-where \(\chi_i\) is the sample space for the \(i^{th}\) component of the random vector \(\bf{Y}\).
-Compact representations with convenient computational forms are rare for high-dimensional constructions, especially with heterogeneous, correlated marginal distributions (or margins of mixed data types).
-Given a large number of simulated vectors as produced above, estimated probabilities are readily given by counting the proportion of simulated vectors meeting the desired condition.
-In our motivating application, one may ask what is the probability that all genes expressed greater than a certain threshold value \({\bf y}_0\).
+where \(\chi_i\) is the sample space for the \(i^{th}\) component of the
+random vector \(\bf{Y}\). Compact representations with convenient
+computational forms are rare for high-dimensional constructions,
+especially with heterogeneous, correlated marginal distributions (or
+margins of mixed data types). Given a large number of simulated vectors
+as produced above, estimated probabilities are readily given by counting
+the proportion of simulated vectors meeting the desired condition. In
+our motivating application, one may ask what is the probability that all
+genes expressed greater than a certain threshold value \({\bf y}_0\).
 This can be estimated as
@@
 
-where \({\bf Y}^{(b)}\) is the \(b^{th}\) simulated vector from a total of \(B\) simulation replicates and \(I(\cdot)\) is the indicator function.
-For example, we can estimate from our \(B=10,000\) simulated vectors that the probability of all genes expressing (i.e., \({\bf y}_i \geq 1, \forall \; i\)) is \(0.5979\).
+where \({\bf Y}^{(b)}\) is the \(b^{th}\) simulated vector from a total
+of \(B\) simulation replicates and \(I(\cdot)\) is the indicator
+function. For example, we can estimate from our \(B=10,000\) simulated
+vectors that the probability of all genes expressing (i.e.,
+\({\bf y}_i \geq 1, \forall \; i\)) is \(0.5979\).
 
@@
 \hypertarget{evaluation-of-correlation-estimation-efficiency}{%
-\subsection{Evaluation of correlation estimation efficiency}\label{evaluation-of-correlation-estimation-efficiency}}
-
-MC methods are routinely used in many statistical inferential tasks including estimation, hypothesis testing, error rates, and empirical interval coverage rates.
-To conclude the example applications, we demonstrate how \texttt{bigsimr} can be used to evaluate estimation efficiency.
-In particular, we wish to assess the error in our correlation estimation above.
-We used a conventional method, based on classical statistical theory which was not designed for high-dimensional data.
-Indeed, high-dimensional covariance estimation (and precision matrices) is an active area of statistical science, .e.g., \citep{Won2013g, VanWieringen2016}.
-
-In this small example, we simulate \(m=30\) data sets with the number of simulated vectors matching the number of patients in the BRCA data set, \(N=878\).
-Since our simulation is much faster for the Pearson correlation type (see Figure \ref{fig:ch040-largeDfig}), we only convert the Spearman correlation matrix once (and ensure it is positive definite).
-At each iteration, we estimate the quadratic loss (residual sum of squared errors) from the specified \({\bf R}_{S}\), producing a distribution of loss values.
+\subsection{Evaluation of correlation estimation
+efficiency}\label{evaluation-of-correlation-estimation-efficiency}}
+
+MC methods are routinely used in many statistical inferential tasks
+including estimation, hypothesis testing, error rates, and empirical
+interval coverage rates. To conclude the example applications, we
+demonstrate how \texttt{bigsimr} can be used to evaluate estimation
+efficiency. In particular, we wish to assess the error in our
+correlation estimation above. We used a conventional method, based on
+classical statistical theory which was not designed for high-dimensional
+data. Indeed, high-dimensional covariance estimation (and precision
+matrices) is an active area of statistical science, .e.g.,
+\citep{Won2013g, VanWieringen2016}.
+
+In this small example, we simulate \(m=30\) data sets with the number of
+simulated vectors matching the number of patients in the BRCA data set,
+\(N=878\). Since our simulation is much faster for the Pearson
+correlation type (see Figure @ref(fig:ch040-largeDfig)), we only convert
+the Spearman correlation matrix once (and ensure it is positive
+definite). At each iteration, we estimate the quadratic loss (residual
+sum of squared errors) from the specified \({\bf R}_{S}\), producing a
+distribution of loss values.
 
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> frobenius_loss <- sapply(simRho, function(R) sum((R - R_S)^2))
-\end{CodeInput}
-\end{CodeChunk}
-
-The \texttt{R} summary function supplies the mean-augmented five-number summary of the quadratic loss distribution computed above.
+The \texttt{R} summary function supplies the mean-augmented five-number
+summary of the quadratic loss distribution computed above.
 
@@
 
-This distribution could be compared to high-dimensional designed covariance estimators for guidance on whether the additional complexity and computation time are warranted.
+This distribution could be compared to high-dimensional designed
+covariance estimators for guidance on whether the additional complexity
+and computation time are warranted.
 
@@
 
-We have introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a user-friendly, high-performance \texttt{R} package {[}\texttt{bigsimr}{]}.
-The random vector generation method is inspired by NORTA \citep{Cario1997} and Gaussian copula-based approaches \citep[\citet{BF17}, \citet{Xia17}]{MB13}.
-The major contributions of this work are methods and software for flexible, scalable simulation of HD multivariate probability distributions with broad potential data analytic applications for modern, big-data statistical computing.
-For example, one could simulate high-resolution time series data, such as those consistent with an auto-regressive moving average model exhibiting a specified Spearman structure.
-Or our methods could be used to simulate sparsely correlated data, as many HD methods assume, via specifying a \emph{spiked correlation matrix}.
-
-It is customary to compare new tools and algorithms directly to existing competing methods and software.
-In this study, however, we only employ our proposed methodology, as our previous work has shown that existing \texttt{R} tools are simply not designed to meet our high-dimensional goal (see \citet{Li2019gpu} for evaluations of the \texttt{R} \texttt{copula} package and others).
-For the bivariate simulations, existing packages such as \href{https://github.com/superdesolator/NORTARA}{\texttt{nortaRA}} work well to match Pearson correlations exactly.
-
-There are limitations to the methodology and implementation.
-We could only investigate selected multivariate distributions in our Monte Carlo studies.
-There may be instances that the methods do not perform well.
-Along those lines, we expect that correlation values close to the boundary of the feasible region could result in algorithm failure.
-Another issue is that for discrete distributions, we use continuous approximations when the support set is large.
-This could limit the scalability/accuracy for particular discrete/mixed multivariate distributions.
-Our method would also benefit computationally from faster Kendall estimation and more finely tuned nearest PD calculation for non-admissible Kendall matrices.
+We have introduced a general-purpose high-dimensional multivariate
+simulation algorithm and provide a user-friendly, high-performance
+\texttt{R} package {[}\texttt{bigsimr}{]}. The random vector generation
+method is inspired by NORTA \citep{Cario1997} and Gaussian copula-based
+approaches \citep[\citet{BF17}, \citet{Xia17}]{MB13}. The major
+contributions of this work are methods and software for flexible,
+scalable simulation of HD multivariate probability distributions with
+broad potential data analytic applications for modern, big-data
+statistical computing. For example, one could simulate high-resolution
+time series data, such as those consistent with an auto-regressive
+moving average model exhibiting a specified Spearman structure. Or our
+methods could be used to simulate sparsely correlated data, as many HD
+methods assume, via specifying a \emph{spiked correlation matrix}.
+
+It is customary to compare new tools and algorithms directly to existing
+competing methods and software. In this study, however, we only employ
+our proposed methodology, as our previous work has shown that existing
+\texttt{R} tools are simply not designed to meet our high-dimensional
+goal (see \citet{Li2019gpu} for evaluations of the \texttt{R}
+\texttt{copula} package and others). For the bivariate simulations,
+existing packages such as
+\href{https://github.com/superdesolator/NORTARA}{\texttt{nortaRA}} work
+well to match Pearson correlations exactly.
+
+There are limitations to the methodology and implementation. We could
+only investigate selected multivariate distributions in our Monte Carlo
+studies. There may be instances that the methods do not perform well.
+Along those lines, we expect that correlation values close to the
+boundary of the feasible region could result in algorithm failure.
+Another issue is that for discrete distributions, we use continuous
+approximations when the support set is large. This could limit the
+scalability/accuracy for particular discrete/mixed multivariate
+distributions. Our method would also benefit computationally from faster
+Kendall estimation and more finely tuned nearest PD calculation for
+non-admissible Kendall matrices.
+
+\newpage
 
@@
 
-We provide an open-source implementation of our methodology as the \texttt{Bigsimr.jl} Julia package, hosted on github at \url{https://github.com/adknudson/Bigsimr.jl}.
-Additionally we provide R and Python interfaces to \texttt{Bigsimr}, respectively at \url{https://github.com/SchisslerGroup/r-bigsimr} and \url{https://github.com/SchisslerGroup/python-bigsimr}.
-\texttt{Bigsimr.jl} is an ongoing project, and feature requests or issues can be submitted to \url{https://github.com/adknudson/Bigsimr.jl/issues}.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> knitr::include_graphics('images/hex-bigsimr.png')
-\end{CodeInput}
+We provide an open-source implementation of our methodology as the
+\texttt{Bigsimr.jl} Julia package, hosted on github at
+\url{https://github.com/adknudson/Bigsimr.jl}. Additionally we provide R
+and Python interfaces to \texttt{Bigsimr}, respectively at
+\url{https://github.com/SchisslerGroup/r-bigsimr} and
+\url{https://github.com/SchisslerGroup/python-bigsimr}.
+\texttt{Bigsimr.jl} is an ongoing project, and feature requests or
+issues can be submitted to
+\url{https://github.com/adknudson/Bigsimr.jl/issues}.
+
+\begin{CodeChunk}
 
@@
 
-The authors gratefully acknowledge Heather Knudson's graphic design for the \texttt{bigsimr} R Package.
-The results published here are in whole or part based upon data generated by the TCGA Research Network: \url{https://www.cancer.gov/tcga}.
+The authors gratefully acknowledge Heather Knudson's graphic design for
+the \texttt{bigsimr} R Package. The results published here are in whole
+or part based upon data generated by the TCGA Research Network:
+\url{https://www.cancer.gov/tcga}.
 
@@
 
-Research reported in this publication was supported by grants from the National Institutes of General Medical Sciences (5 U54 GM104944, GM103440) from the National Institutes of Health.
+Research reported in this publication was supported by grants from the
+National Institutes of General Medical Sciences (5 U54 GM104944,
+GM103440) from the National Institutes of Health.
+
+\newpage
 
@@
 
-\noindent Consider a bivariate example where we have a correlated \((Y_1, Y_2)^\top\) with \(Y_1\sim N(\mu_1, \sigma_1^2)\) (normal) and \(Y_2\sim LN(\mu_2, \sigma_2^2)\) (lognormal).
-First, let us note that when we get such a vector from the \texttt{bigsimr} package then in fact it can be represented as
+\noindent Consider a bivariate example where we have a correlated
+\((Y_1, Y_2)^\top\) with \(Y_1\sim N(\mu_1, \sigma_1^2)\) (normal) and
+\(Y_2\sim LN(\mu_2, \sigma_2^2)\) (lognormal). First, let us note that
+when we get such a vector from the \texttt{bigsimr} package then in fact
+it can be represented as
 
@@
 (Y_1, Y_2)^\top \stackrel{d}{=} \left(X_1, e^{X_2}\right)^\top,
-\label{eq:kram1}
-\end{equation}
-
-where \((X_1, X_2)^\top \sim N_2(\boldsymbol \mu, \boldsymbol \Sigma)\), which is bivariate normal with mean vector \(\boldsymbol \mu = (\mu_1, \mu_2)^\top\) and variance-covariance matrix
+(\#eq:kram1)
+\end{equation}
+
+where \((X_1, X_2)^\top \sim N_2(\boldsymbol \mu, \boldsymbol \Sigma)\),
+which is bivariate normal with mean vector
+\(\boldsymbol \mu = (\mu_1, \mu_2)^\top\) and variance-covariance matrix
 
@@
 \end{bmatrix}
-\label{eq:kram2}
+(\#eq:kram2)
 \end{equation}
@@
 \right].
-\label{eq:kram3}
+(\#eq:kram3)
 \end{equation}
@@
 
-In this example, \(F_1\) is the CDF of \(N(\mu_1, \sigma_1^2)\) while \(F_2\) is the CDF of \(LN(\mu_2, \sigma_2^2)\), so that the two required quantile functions are
+In this example, \(F_1\) is the CDF of \(N(\mu_1, \sigma_1^2)\) while
+\(F_2\) is the CDF of \(LN(\mu_2, \sigma_2^2)\), so that the two
+required quantile functions are
 
@@
 F_1^{-1}(u) = \mu_1+\sigma_1 \Phi^{-1}(u)\,\,\, \mbox{and} \,\,\, F_2^{-1}(u) = e^{\mu_2+\sigma_2 \Phi^{-1}(u)}, 
-\label{eq:kram4}
-\end{equation}
-
-as can be seen by standard calculation. Thus, when we apply Step 3 of the NORTA algorithm, we get
+(\#eq:kram4)
+\end{equation}
+
+as can be seen by standard calculation. Thus, when we apply Step 3 of
+the NORTA algorithm, we get
 
@@
 F_1^{-1}(U_1) = \mu_1+\sigma_1 \Phi^{-1}(\Phi(Z_1)) =  \mu_1+\sigma_1 Z_1 \,\,\, \mbox{and} \,\,\, F_2^{-1}(U_2) = e^{\mu_2+\sigma_2 \Phi^{-1}(\Phi(Z_2))} = e^{\mu_2+\sigma_2 Z_2}, 
-\label{eq:kram5}
-\end{equation}
-
-where \((X_1, X_2)^\top = (\mu_1+\sigma_1 Z_1, \mu_2+\sigma_2 Z_2)^\top \sim N_2(\boldsymbol \mu, \boldsymbol \Sigma)\).
-Consequently, the vector \((Y_1, Y_2)^\top\) obtained in Step 3 of the algorithm has the structure provided by (\ref{kram1}).
+(\#eq:kram5)
+\end{equation}
+
+where
+\((X_1, X_2)^\top = (\mu_1+\sigma_1 Z_1, \mu_2+\sigma_2 Z_2)^\top \sim N_2(\boldsymbol \mu, \boldsymbol \Sigma)\).
+Consequently, the vector \((Y_1, Y_2)^\top\) obtained in Step 3 of the
+algorithm has the structure provided by (\ref{kram1}).
 
@@
 
-\noindent Next, we provide the exact covariance structure of the random vector \((Y_1, Y_2)^\top\) given by (\ref{kram1}), and relate the correlation of \(Y_1\) and \(Y_2\) to \(\rho\), which is the correlation of the normal variables \(Z_1\), \(Z_2\) (and also \(X_1\) and \(X_2\)).
-A straightforward albeit somewhat tedious algebra produces the following result.
+\noindent Next, we provide the exact covariance structure of the random
+vector \((Y_1, Y_2)^\top\) given by (\ref{kram1}), and relate the
+correlation of \(Y_1\) and \(Y_2\) to \(\rho\), which is the correlation
+of the normal variables \(Z_1\), \(Z_2\) (and also \(X_1\) and \(X_2\)).
+A straightforward albeit somewhat tedious algebra produces the following
+result.
 
@@
 \boldsymbol \mu_{{\bf Y}} = \left(\mu_1, e^{\mu_2+\sigma_2^2/2}\right)^\top
-\label{eq:kram6}
+(\#eq:kram6)
 \end{equation}
@@
 \end{bmatrix}
-\label{eq:kram7}
+(\#eq:kram7)
 \end{equation}
@@
 \mbox{Cov}(Y_1, Y_2) = \mathbb E(Y_1 Y_2) - \mathbb E(Y_1)\mathbb E(Y_2), 
-\label{eq:covy1y2}
+(\#eq:covy1y2)
 \end{equation}
@@
 \mathbb E(Y_1 Y_2)  =  \mathbb E\left(X_1 e^{X_2}\right)  =  \mathbb E \left\{ \mathbb E \left(X_1 e^{X_2} | X_1 \right) \right\} = \mathbb E \left\{ X_1 \mathbb E \left(e^{X_2} | X_1 \right) \right\}.
-\label{eq:kre1}
+(\#eq:kre1)
 \end{equation}
@@
 E \left(e^{X_2} | X_1 \right) = e^{ \mu_2+\rho\sigma_2(x_1-\mu_1)/\sigma_1 + \sigma_2^2(1-\rho^2)/2},
-\label{eq:kre2}
+(\#eq:kre2)
 \end{equation}
@@
 \mathbb E(Y_1 Y_2)  =  e^{ \mu_2+ \frac{1}{2} \sigma_2^2(1-\rho^2)} \mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\}.
-\label{eq:kre3}
+(\#eq:kre3)
 \end{equation}
@@
 \mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = \int_{-\infty}^\infty x_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (x_1-\mu_1)} \frac{1}{\sqrt{2\pi}\sigma_1} e^{-\frac{1}{2\sigma_1^2}(x_1-\mu_1)^2}.
-\label{eq:kre4}
+(\#eq:kre4)
 \end{equation}
@@
 \mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = e^{\frac{1}{2}\sigma_2^2\rho^2} \int_{-\infty}^\infty (u+\mu_1)  \frac{1}{\sqrt{2\pi}\sigma_1} e^{-\frac{1}{2\sigma_1^2}(u-\rho\sigma_1\sigma_2)^2}.
-\label{eq:kre5}
+(\#eq:kre5)
 \end{equation}
@@
 \mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = e^{\frac{1}{2}\sigma_2^2\rho^2} (\rho\sigma_1\sigma_2 +\mu_1),
-\label{eq:kre6}
+(\#eq:kre6)
 \end{equation}
@@
 \mathbb E (Y_1Y_2) = e^{\mu_2 + \frac{1}{2}\sigma_2^2} (\rho\sigma_1\sigma_2 +\mu_1).
-\label{eq:kre7}
+(\#eq:kre7)
 \end{equation}
@@
 \mbox{Cov}(Y_1, Y_2) = e^{\mu_2 + \frac{1}{2}\sigma_2^2} (\rho\sigma_1\sigma_2 +\mu_1) - \mu_1 e^{\mu_2 + \frac{1}{2}\sigma_2^2} = \rho\sigma_1\sigma_2 e^{\mu_2 + \frac{1}{2}\sigma_2^2}.
-\label{eq:kre8}
+(\#eq:kre8)
 \end{equation}
@@
 
-Using the above result, we can directly relate the correlation coefficient of \(Y_1\) and \(Y_2\) with that of \(X_1\) and \(X_2\),
+Using the above result, we can directly relate the correlation
+coefficient of \(Y_1\) and \(Y_2\) with that of \(X_1\) and \(X_2\),
 
@@
 \rho_{Y_1, Y_2} = \rho \frac{\sigma_2}{\sqrt{e^{\sigma_2^2} -1}},
-\label{eq:kram8}
-\end{equation}
-
-where \(\rho\) is the correlation of \(X_1\) and \(X_2\).
-The above result is useful when studying the possible range of correlation of \(Y_1\) and \(Y_2\) in the above example.
-It can be shown that the factor on the far right in (\ref{kram8}) is a monotonically decreasing function of \(\sigma_2\) on \((0,\infty)\), with the limits of 1 and 0 at zero and infinity, respectively.
-Thus, in principle, the range of correlation of \(Y_1\) and \(Y_2\) is the same as that of \(X_1\) and \(X_2\), as the factor on the far right in (\ref{kram8}) can be made arbitrarily close to 1.
-However, by changing this factor we may affect the marginal distributions of \(Y_1\) and \(Y_2\).
-It can be shown that if the marginal distributions of \(Y_1\) and \(Y_2\) are fixed, then the relation (\ref{kram8}) becomes
+(\#eq:kram8)
+\end{equation}
+
+where \(\rho\) is the correlation of \(X_1\) and \(X_2\). The above
+result is useful when studying the possible range of correlation of
+\(Y_1\) and \(Y_2\) in the above example. It can be shown that the
+factor on the far right in (\ref{kram8}) is a monotonically decreasing
+function of \(\sigma_2\) on \((0,\infty)\), with the limits of 1 and 0
+at zero and infinity, respectively. Thus, in principle, the range of
+correlation of \(Y_1\) and \(Y_2\) is the same as that of \(X_1\) and
+\(X_2\), as the factor on the far right in (\ref{kram8}) can be made
+arbitrarily close to 1. However, by changing this factor we may affect
+the marginal distributions of \(Y_1\) and \(Y_2\). It can be shown that
+if the marginal distributions of \(Y_1\) and \(Y_2\) are fixed, then the
+relation (\ref{kram8}) becomes
 
@@
 \rho_{Y_1, Y_2} = \rho \sqrt{\frac{\log(1+c_2^2)}{c_2^2}},
-\label{eq:kram9}
-\end{equation}
-
-where \(c_2=\sigma_{Y_2}/\mu_{Y_2}\) is the \emph{coefficient of variation} (CV) of the variable \(Y_2\).
-Thus, the range of possible correlations in this model is not affected by the distribution of \(Y_1\), and is determined by the CV of \(Y_2\) as follows:
+(\#eq:kram9)
+\end{equation}
+
+where \(c_2=\sigma_{Y_2}/\mu_{Y_2}\) is the \emph{coefficient of
+variation} (CV) of the variable \(Y_2\). Thus, the range of possible
+correlations in this model is not affected by the distribution of
+\(Y_1\), and is determined by the CV of \(Y_2\) as follows:
 
@@
 - \sqrt{\frac{\log(1+c_2^2)}{c_2^2}} \leq \rho_{Y_1, Y_2} \leq \sqrt{\frac{\log(1+c_2^2)}{c_2^2}}. 
-\label{eq:kram10}
-\end{equation}
-
-Plugging in the estimates of these quantities from Section \ref{package}, we see that
+(\#eq:kram10)
+\end{equation}
+
+Plugging in the estimates of these quantities from Section
+@ref(package), we see that
 
@@
 \frac{\hat{\sigma}_2}{\sqrt{e^{\hat{\sigma}_2^2} -1}} = 0.881.
-\label{eq:kram11}
-\end{equation}
-
-Thus, the possible range of correlation becomes \((-0.881, 0.881)\). This approximately agrees with the MC results provided.
-
-As an invited speaker at CMStatistics 2020, I was delighted to learn of the opportunity to have our work included in the first issue of CSDA Annals of Statistical Data Science.
-
-On behalf of my co-authors, I'm pleased to submit our manuscript entitled \emph{Simulating High-Dimensional Multivariate Data using the \texttt{Bigsimr} Package} for your consideration.
-We feel that the work is well suited to the indicated themes and hope that it will be considered a valuable contribution to this prestigious journal.
-
-The entire article can be reproduced from code and is housed at \url{https://github.com/SchisslerGroup/article_bigsimr}, which we are glad to make available upon request to the reviewers and, after acceptance, the general public.
-
-I hope that everything is in order. If not, or if you have any additional needs, please do not hesitate to contact me. Thank you and all those involved in this process.
+(\#eq:kram11)
+\end{equation}
+
+Thus, the possible range of correlation becomes \((-0.881, 0.881)\).
+This approximately agrees with the MC results provided.
+
+\setstretch{1.0}
 

### 2021-05-06 010-background.Rmd
working template
Drafting comparison table in background

@@
 \rho_P(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ \mathrm{Var}(X)\mathrm{Var}(Y)\right]^{1/2}}.
-(\#eq:pearson)
+\label{eq:pearson}
 \end{equation}
@@
 \begin{equation}
-(\#eq:convertKendall)
+\label{eq:convertKendall}
 \rho_{P} = \sin \left( \tau \times \frac{\pi}{2} \right), 
@@
 \begin{equation}
-(\#eq:convertSpearman)
+\label{eq:convertSpearman}
 \rho_P = 2 \times \sin \left( \rho_S \times \frac{\pi}{6} \right).
@@
 \begin{equation}
-(\#eq:frechet)
+\label{eq:frechet}
 \rho_P^{max} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (U) \right), \quad \rho_P^{min} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (1 - U) \right)
@@
 F_{i}^{-1} = \inf\{y:F_{i}(y) \geq u \}.
-(\#eq:inverseCDF)
+\label{eq:inverseCDF}
 \end{equation}
@@
 F(x_1, \ldots, x_d) = {\mathbb P}(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), 
-(\#eq:copula)
+\label{eq:copula}
 \end{equation}
@@
 \begin{equation}
-(\#eq:gauss)
+\label{eq:gauss}
 F_{{\bf R}}(x_1, \ldots, x_d) = C_{{\bf R}}(\Phi(x_1), \ldots, \Phi(x_d)),
@@
 $Y_i = F_i^{-1}(\Phi(X_i)), i=1, \ldots, d$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})$.
+
+## Relevant Multivariate Simulation software Comparison
+
+```{r tab:compare-table}
+knitr::kable(
+           head(mtcars[, 1:8], 10),  booktabs = TRUE, format = "latex",
+                 linesep = c("", "", "\\addlinespace"),
+           caption = 'A table of the first 10 rows of the mtcars data.'
+  )
+```
+
+<!-- 
+| Software | Designed for | Scales to Large P | Hetereogenous Margins | Flexible Dependence |
+|----------|--------------|-------------------|-----------------------|---------------------|
+|`copula`|              |                   |                       |                     |
+|`genord`|              |                   |                       |                     |
+|`mvnfast`|              |                   |                       |                     |
+
+your comment -->

### 2021-05-06 article_bigsimr.tex
Drafting comparison table in background

@@
 where \((X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})\).
+
+\hypertarget{relevant-multivariate-simulation-software-comparison}{%
+\subsection{Relevant Multivariate Simulation software
+Comparison}\label{relevant-multivariate-simulation-software-comparison}}
+
+\begin{CodeChunk}
+\begin{table}
+
+\caption{\label{tab:tab:compare-table}A table of the first 10 rows of the mtcars data.}
+\centering
+\begin{tabular}[t]{lrrrrrrrr}
+\toprule
+  & mpg & cyl & disp & hp & drat & wt & qsec & vs\\
+\midrule
+Mazda RX4 & 21.0 & 6 & 160.0 & 110 & 3.90 & 2.620 & 16.46 & 0\\
+Mazda RX4 Wag & 21.0 & 6 & 160.0 & 110 & 3.90 & 2.875 & 17.02 & 0\\
+Datsun 710 & 22.8 & 4 & 108.0 & 93 & 3.85 & 2.320 & 18.61 & 1\\
+\addlinespace
+Hornet 4 Drive & 21.4 & 6 & 258.0 & 110 & 3.08 & 3.215 & 19.44 & 1\\
+Hornet Sportabout & 18.7 & 8 & 360.0 & 175 & 3.15 & 3.440 & 17.02 & 0\\
+Valiant & 18.1 & 6 & 225.0 & 105 & 2.76 & 3.460 & 20.22 & 1\\
+\addlinespace
+Duster 360 & 14.3 & 8 & 360.0 & 245 & 3.21 & 3.570 & 15.84 & 0\\
+Merc 240D & 24.4 & 4 & 146.7 & 62 & 3.69 & 3.190 & 20.00 & 1\\
+Merc 230 & 22.8 & 4 & 140.8 & 95 & 3.92 & 3.150 & 22.90 & 1\\
+\addlinespace
+Merc 280 & 19.2 & 6 & 167.6 & 123 & 3.92 & 3.440 & 18.30 & 1\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\end{CodeChunk}
 
@@
     \mathrm{E}\left[Y_i Y_j\right] = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} F_i^{-1}\left[\Phi(z_i)\right] F_j^{-1}\left[\Phi(z_j)\right] \phi(z_i, z_j, \rho_z) dz_i dz_j,
-    (\#eq:pearsonIntegralRelation)
+    \label{eq:pearsonIntegralRelation}
 \end{equation}
@@
        [,1]   [,2]
-[1,]  1.0000 -0.8745
-[2,] -0.8745  1.0000
+[1,]  1.000 -0.878
+[2,] -0.878  1.000
 \end{CodeOutput}
@@
       [,1]  [,2]
-[1,] 1.0000 0.8748
-[2,] 0.8748 1.0000
+[1,] 1.000 0.879
+[2,] 0.879 1.000
 \end{CodeOutput}
@@
 Julia Object of type NamedTuple{(:lower, :upper),Tuple{Float64,Float64}}.
-(lower = -0.8799819883897544, upper = 0.8798667127065275)
+(lower = -0.8801373451421908, upper = 0.8801357032571077)
 \end{CodeOutput}
@@
 \midrule
-1000 & Pearson & norm & 0.0168\\
+1000 & Pearson & norm & 0.0156\\
 1000 & Pearson & gamma & 0.0159\\
-1000 & Pearson & nbinom & 0.0188\\
+1000 & Pearson & nbinom & 0.0162\\
 \addlinespace
-1000 & Spearman & norm & 0.0203\\
-1000 & Spearman & gamma & 0.0180\\
-1000 & Spearman & nbinom & 0.0165\\
+1000 & Spearman & norm & 0.0175\\
+1000 & Spearman & gamma & 0.0182\\
+1000 & Spearman & nbinom & 0.0151\\
 \addlinespace
-1000 & Kendall & norm & 0.0118\\
-1000 & Kendall & gamma & 0.0099\\
-1000 & Kendall & nbinom & 0.0089\\
+1000 & Kendall & norm & 0.0130\\
+1000 & Kendall & gamma & 0.0115\\
+1000 & Kendall & nbinom & 0.0126\\
 \addlinespace
-10000 & Pearson & norm & 0.0048\\
-10000 & Pearson & gamma & 0.0055\\
-10000 & Pearson & nbinom & 0.0066\\
+10000 & Pearson & norm & 0.0060\\
+10000 & Pearson & gamma & 0.0058\\
+10000 & Pearson & nbinom & 0.0058\\
 \addlinespace
-10000 & Spearman & norm & 0.0058\\
-10000 & Spearman & gamma & 0.0054\\
-10000 & Spearman & nbinom & 0.0060\\
+10000 & Spearman & norm & 0.0062\\
+10000 & Spearman & gamma & 0.0056\\
+10000 & Spearman & nbinom & 0.0049\\
 \addlinespace
-10000 & Kendall & norm & 0.0034\\
-10000 & Kendall & gamma & 0.0031\\
-10000 & Kendall & nbinom & 0.0037\\
+10000 & Kendall & norm & 0.0033\\
+10000 & Kendall & gamma & 0.0038\\
+10000 & Kendall & nbinom & 0.0033\\
 \addlinespace
-1e+05 & Pearson & norm & 0.0016\\
-1e+05 & Pearson & gamma & 0.0016\\
-1e+05 & Pearson & nbinom & 0.0032\\
+1e+05 & Pearson & norm & 0.0018\\
+1e+05 & Pearson & gamma & 0.0017\\
+1e+05 & Pearson & nbinom & 0.0030\\
 \addlinespace
-1e+05 & Spearman & norm & 0.0018\\
-1e+05 & Spearman & gamma & 0.0019\\
-1e+05 & Spearman & nbinom & 0.0013\\
+1e+05 & Spearman & norm & 0.0017\\
+1e+05 & Spearman & gamma & 0.0016\\
+1e+05 & Spearman & nbinom & 0.0015\\
 \addlinespace
-1e+05 & Kendall & norm & 0.0011\\
-1e+05 & Kendall & gamma & 0.0010\\
-1e+05 & Kendall & nbinom & 0.0010\\
+1e+05 & Kendall & norm & 0.0010\\
+1e+05 & Kendall & gamma & 0.0011\\
+1e+05 & Kendall & nbinom & 0.0012\\
 \bottomrule
@@
 |:---------|:----------------|--------------------:|
-|100       |Pearson          |                0.080|
-|100       |Spearman         |                0.036|
-|100       |Kendall          |                0.116|
-|250       |Pearson          |                0.286|
-|250       |Spearman         |                0.371|
-|250       |Kendall          |                0.366|
-|500       |Pearson          |                2.932|
-|500       |Spearman         |                0.176|
-|500       |Kendall          |                2.183|
+|100       |Pearson          |                0.081|
+|100       |Spearman         |                0.027|
+|100       |Kendall          |                0.075|
+|250       |Pearson          |                0.378|
+|250       |Spearman         |                0.062|
+|250       |Kendall          |                0.375|
+|500       |Pearson          |                1.425|
+|500       |Spearman         |                0.135|
+|500       |Kendall          |                1.495|
 
@@
 vectors that the probability of all genes expressing (i.e.,
-\({\bf y}_i \geq 1, \forall \; i\)) is \(0.5979\).
+\({\bf y}_i \geq 1, \forall \; i\)) is \(0.6015\).
 
@@
 \begin{CodeOutput}
-[1] 0.5979
+[1] 0.6015
 \end{CodeOutput}
@@
    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-    691     829     890     925     986    1224 
+    675     813     974     998    1080    1851 
 \end{CodeOutput}
@@
 \begin{equation}
+\label{eq:kram1}
 (Y_1, Y_2)^\top \stackrel{d}{=} \left(X_1, e^{X_2}\right)^\top,
-(\#eq:kram1)
 \end{equation}
@@
 \begin{equation}
+\label{eq:kram2}
 \boldsymbol \Sigma = 
@@
 \end{bmatrix}
-(\#eq:kram2)
 \end{equation}
@@
 \begin{equation}
+\label{eq:kram3}
 \boldsymbol R = 
@@
 \right].
-(\#eq:kram3)
 \end{equation}
@@
 \begin{equation}
+\label{eq:kram4}
 F_1^{-1}(u) = \mu_1+\sigma_1 \Phi^{-1}(u)\,\,\, \mbox{and} \,\,\, F_2^{-1}(u) = e^{\mu_2+\sigma_2 \Phi^{-1}(u)}, 
-(\#eq:kram4)
 \end{equation}
@@
 \begin{equation}
+\label{eq:kram5}
 F_1^{-1}(U_1) = \mu_1+\sigma_1 \Phi^{-1}(\Phi(Z_1)) =  \mu_1+\sigma_1 Z_1 \,\,\, \mbox{and} \,\,\, F_2^{-1}(U_2) = e^{\mu_2+\sigma_2 \Phi^{-1}(\Phi(Z_2))} = e^{\mu_2+\sigma_2 Z_2}, 
-(\#eq:kram5)
 \end{equation}
@@
 Consequently, the vector \((Y_1, Y_2)^\top\) obtained in Step 3 of the
-algorithm has the structure provided by (\ref{kram1}).
+algorithm has the structure provided by (\ref{eq:kram1}).
 
@@
 \noindent Next, we provide the exact covariance structure of the random
-vector \((Y_1, Y_2)^\top\) given by (\ref{kram1}), and relate the
+vector \((Y_1, Y_2)^\top\) given by (\ref{eq:kram1}), and relate the
 correlation of \(Y_1\) and \(Y_2\) to \(\rho\), which is the correlation
@@
 \begin{equation}
+\label{eq:kram6}
 \boldsymbol \mu_{{\bf Y}} = \left(\mu_1, e^{\mu_2+\sigma_2^2/2}\right)^\top
-(\#eq:kram6)
 \end{equation}
@@
 \begin{equation}
+\label{eq:kram7}
 \boldsymbol \Sigma = 
@@
 \end{bmatrix}
-(\#eq:kram7)
 \end{equation}
@@
 \begin{equation}
+\label{eq:covy1y2}
 \mbox{Cov}(Y_1, Y_2) = \mathbb E(Y_1 Y_2) - \mathbb E(Y_1)\mathbb E(Y_2), 
-(\#eq:covy1y2)
 \end{equation}
@@
 \begin{equation}
+\label{eq:kre1}
 \mathbb E(Y_1 Y_2)  =  \mathbb E\left(X_1 e^{X_2}\right)  =  \mathbb E \left\{ \mathbb E \left(X_1 e^{X_2} | X_1 \right) \right\} = \mathbb E \left\{ X_1 \mathbb E \left(e^{X_2} | X_1 \right) \right\}.
-(\#eq:kre1)
-\end{equation}
-
-Further, by using the fact that the conditional distribution of $X_2$ given $X_1=x_1$ is normal with mean $\mathbb E(X_2|X_1=x_1) = \mu_2+\rho\sigma_2(x_1-\mu_1)/\sigma_1$ and variance $\sigma_2^2(1-\rho^2)$, one can relate the inner expectation on the far right in (\ref{kre1}) to the moment generating function $M(t)$ of this conditional distribution evaluated at $t=1$, leading to
-
-\begin{equation}
+\end{equation}
+
+Further, by using the fact that the conditional distribution of $X_2$ given $X_1=x_1$ is normal with mean $\mathbb E(X_2|X_1=x_1) = \mu_2+\rho\sigma_2(x_1-\mu_1)/\sigma_1$ and variance $\sigma_2^2(1-\rho^2)$, one can relate the inner expectation on the far right in (\ref{eq:kre1}) to the moment generating function $M(t)$ of this conditional distribution evaluated at $t=1$, leading to
+
+\begin{equation}
+\label{eq:kre2}
 E \left(e^{X_2} | X_1 \right) = e^{ \mu_2+\rho\sigma_2(x_1-\mu_1)/\sigma_1 + \sigma_2^2(1-\rho^2)/2},
-(\#eq:kre2)
 \end{equation}
@@
 \begin{equation}
+\label{eq:kre3}
 \mathbb E(Y_1 Y_2)  =  e^{ \mu_2+ \frac{1}{2} \sigma_2^2(1-\rho^2)} \mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\}.
-(\#eq:kre3)
-\end{equation}
-
-Since $X_1$ is normal with mean $\mu_1$ and variance $\sigma_1^2$, the expectation in (\ref{kre3}) becomes
-
-\begin{equation}
+\end{equation}
+
+Since $X_1$ is normal with mean $\mu_1$ and variance $\sigma_1^2$, the expectation in (\ref{eq:kre3}) becomes
+
+\begin{equation}
+\label{eq:kre4}
 \mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = \int_{-\infty}^\infty x_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (x_1-\mu_1)} \frac{1}{\sqrt{2\pi}\sigma_1} e^{-\frac{1}{2\sigma_1^2}(x_1-\mu_1)^2}.
-(\#eq:kre4)
 \end{equation}
@@
 \begin{equation}
+\label{eq:kre5}
 \mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = e^{\frac{1}{2}\sigma_2^2\rho^2} \int_{-\infty}^\infty (u+\mu_1)  \frac{1}{\sqrt{2\pi}\sigma_1} e^{-\frac{1}{2\sigma_1^2}(u-\rho\sigma_1\sigma_2)^2}.
-(\#eq:kre5)
-\end{equation}
-
-Since the integral in (\ref{kre5}) is the expectation $\mathbb E(X+\mu_1)$, where $X$ is normal with mean $\rho\sigma_1\sigma_2$ and variance $\sigma_1^2$, we conclude that
-
-\begin{equation}
+\end{equation}
+
+Since the integral in (\ref{eq:kre5}) is the expectation $\mathbb E(X+\mu_1)$, where $X$ is normal with mean $\rho\sigma_1\sigma_2$ and variance $\sigma_1^2$, we conclude that
+
+\begin{equation}
+\label{eq:kre6}
 \mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = e^{\frac{1}{2}\sigma_2^2\rho^2} (\rho\sigma_1\sigma_2 +\mu_1),
-(\#eq:kre6)
-\end{equation}
-
-which, in view of (\ref{kre3}), leads to
-
-\begin{equation}
+\end{equation}
+
+which, in view of (\ref{eq:kre3}), leads to
+
+\begin{equation}
+\label{eq:kre7}
 \mathbb E (Y_1Y_2) = e^{\mu_2 + \frac{1}{2}\sigma_2^2} (\rho\sigma_1\sigma_2 +\mu_1).
-(\#eq:kre7)
-\end{equation}
-
-Finally, (\ref{covy1y2}), along with the expressions for the means of $Y_1$ and $Y_2$, produce the covariance of $Y_1$ and $Y_2$:
-
-\begin{equation}
+\end{equation}
+
+Finally, (\ref{eq:covy1y2}), along with the expressions for the means of $Y_1$ and $Y_2$, produce the covariance of $Y_1$ and $Y_2$:
+
+\begin{equation}
+\label{eq:kre8}
 \mbox{Cov}(Y_1, Y_2) = e^{\mu_2 + \frac{1}{2}\sigma_2^2} (\rho\sigma_1\sigma_2 +\mu_1) - \mu_1 e^{\mu_2 + \frac{1}{2}\sigma_2^2} = \rho\sigma_1\sigma_2 e^{\mu_2 + \frac{1}{2}\sigma_2^2}.
-(\#eq:kre8)
 \end{equation}
@@
 \begin{equation}
+\label{eq:kram8}
 \rho_{Y_1, Y_2} = \rho \frac{\sigma_2}{\sqrt{e^{\sigma_2^2} -1}},
-(\#eq:kram8)
 \end{equation}
@@
 \(Y_1\) and \(Y_2\) in the above example. It can be shown that the
-factor on the far right in (\ref{kram8}) is a monotonically decreasing
-function of \(\sigma_2\) on \((0,\infty)\), with the limits of 1 and 0
-at zero and infinity, respectively. Thus, in principle, the range of
-correlation of \(Y_1\) and \(Y_2\) is the same as that of \(X_1\) and
-\(X_2\), as the factor on the far right in (\ref{kram8}) can be made
-arbitrarily close to 1. However, by changing this factor we may affect
-the marginal distributions of \(Y_1\) and \(Y_2\). It can be shown that
-if the marginal distributions of \(Y_1\) and \(Y_2\) are fixed, then the
-relation (\ref{kram8}) becomes
-
-\begin{equation}
+factor on the far right in (\ref{eq:kram8}) is a monotonically
+decreasing function of \(\sigma_2\) on \((0,\infty)\), with the limits
+of 1 and 0 at zero and infinity, respectively. Thus, in principle, the
+range of correlation of \(Y_1\) and \(Y_2\) is the same as that of
+\(X_1\) and \(X_2\), as the factor on the far right in (\ref{eq:kram8})
+can be made arbitrarily close to 1. However, by changing this factor we
+may affect the marginal distributions of \(Y_1\) and \(Y_2\). It can be
+shown that if the marginal distributions of \(Y_1\) and \(Y_2\) are
+fixed, then the relation (\ref{eq:kram8}) becomes
+
+\begin{equation}
+\label{eq:kram9}
 \rho_{Y_1, Y_2} = \rho \sqrt{\frac{\log(1+c_2^2)}{c_2^2}},
-(\#eq:kram9)
 \end{equation}
@@
 \begin{equation}
+\label{eq:kram10}
 - \sqrt{\frac{\log(1+c_2^2)}{c_2^2}} \leq \rho_{Y_1, Y_2} \leq \sqrt{\frac{\log(1+c_2^2)}{c_2^2}}. 
-(\#eq:kram10)
 \end{equation}
@@
 \begin{equation}
+\label{eq:kram11}
 \frac{\hat{\sigma}_2}{\sqrt{e^{\hat{\sigma}_2^2} -1}} = 0.881.
-(\#eq:kram11)
 \end{equation}

### 2021-05-06 010-background.Rmd
Drafting comparison table in background v2

@@
 
-```{r tab:compare-table}
-knitr::kable(
-           head(mtcars[, 1:8], 10),  booktabs = TRUE, format = "latex",
-                 linesep = c("", "", "\\addlinespace"),
-           caption = 'A table of the first 10 rows of the mtcars data.'
-  )
-```
+To place `Bigsimr`'s implementation and algorithms in context with existing approaches, as well as make this work's contribution more evident, we make some qualitative comparisons to other `R` multivariate simulation packages.
+The `MASS` packages `mvnorm()` has long been available to produce random vectors from multivariate normal (and $t$) distributions.
+This familiar procedure cannot simulate from arbitrary margins and dependency measures, as well as fails to readily scale to high dimensions.
+Recently, the high-performance package `mvnfast` has been released that provides scaleable multivariate normal generation, but this doesn't completely satistify our goals of flexible margins and dependency modeling.
+For dependent discrete data, small-dimesional packages exist, such as `GenOrd`.
+The full-featured package `copula` provides many different copulas and a fairly general random generator `rCopula`.
+For low-dimensional modeling situtations, copula is suitable, yet the entent of exact matching of margins automatically is unclear.
+High-dimesional random vectors via `copula` takes an impractical amount of computing time [@Li2019gpu].
+The `nortaRA` R package matches Pearson coefficients near exactly and is reasonable for low-dimenional settings.
+In sum, we know of no other `R` package that can produce HD random vectors with flexible margins and dependence modeling.
+Table (\#tab:compare-table) summarizes the above discussion.
 
-<!-- 
-| Software | Designed for | Scales to Large P | Hetereogenous Margins | Flexible Dependence |
-|----------|--------------|-------------------|-----------------------|---------------------|
-|`copula`|              |                   |                       |                     |
-|`genord`|              |                   |                       |                     |
-|`mvnfast`|              |                   |                       |                     |
-
-your comment -->
+\begin{table}[]
+\centering
+\caption{Selected feature comparison of `bigsimr` to other existing `R` multivariate simulation packages.}
+\label{tab:compare-table}
+\begin{tabular}{@{}llll@{}}
+R Package & Scales to HD & Hetereogenous Margins & Flexible Dependence \\ \midrule
+MASS      & No           & No                    & No                  \\
+mvnfast   & Yes          & No                    & No                  \\
+GenOrd    & No           & Discrete margins      & Yes                 \\
+copula    & No           & Yes                   & No                  \\
+nortaRA   & No           & Yes                   & No                  \\
+bigsimr   & Yes          & Yes                   & Yes                
+\end{tabular}
+\end{table}

### 2021-05-06 040-monte-carlo-evaluation.Rmd
Drafting comparison table in background v2

@@
 Simulation Reps $B$ & Correlation Types & Identical-margin 2D distribution \\ \midrule
-$1000$ & Pearson ($\rho_P$) & ${\bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho_i ), i=1,\ldots,100$ \\
+$1,000$ & Pearson ($\rho_P$) & ${\bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho_i ), i=1,\ldots,100$ \\
 $10,000$ & Spearman ($\rho_S$) & ${\bf Y} \sim MVG( shape = 10, rate = 1, \rho_i ), i=1,\ldots,100$ \\
@@
 \setstretch{1.5}
-i. Estimate the correlation matrix from the "data" in the *Compute Correlation* step.
-1. Map the correlations to initialize the algorithm (Pearson to Pearson, Spearman to Pearson, or Kendall to Pearson) in the *Adjust Correlation* step.
+1. a) Estimate the correlation matrix from the "data" in the *Compute Correlation* step.
+1. b) Map the correlations to initialize the algorithm (Pearson to Pearson, Spearman to Pearson, or Kendall to Pearson) in the *Adjust Correlation* step.
 2. Check whether the mapping produces a valid correlation matrix and, if not, find the nearest PD correlation matrix in the *Check Admissibility* step.
@@
 
-```{r ch040-moderateDtab}
+```{r tab:ch040-moderateDtab}
 dat_long <- benchmark_dependences %>%
@@
     kable(booktabs = TRUE,
-          linesep = c("", "", "\\addlinespace"),
+          linesep = c("", "", "\\addlinespace"), format="latex",
           col.names = c("Dimension",

### 2021-05-06 article_bigsimr.tex
Drafting comparison table in background v2

@@
 
-\begin{CodeChunk}
-\begin{table}
-
-\caption{\label{tab:tab:compare-table}A table of the first 10 rows of the mtcars data.}
+To place \texttt{Bigsimr}'s implementation and algorithms in context
+with existing approaches, as well as make this work's contribution more
+evident, we make some qualitative comparisons to other \texttt{R}
+multivariate simulation packages. The \texttt{MASS} packages
+\texttt{mvnorm()} has long been available to produce random vectors from
+multivariate normal (and \(t\)) distributions. This familiar procedure
+cannot simulate from arbitrary margins and dependency measures, as well
+as fails to readily scale to high dimensions. Recently, the
+high-performance package \texttt{mvnfast} has been released that
+provides scaleable multivariate normal generation, but this doesn't
+completely satistify our goals of flexible margins and dependency
+modeling. For dependent discrete data, small-dimesional packages exist,
+such as \texttt{GenOrd}. The full-featured package \texttt{copula}
+provides many different copulas and a fairly general random generator
+\texttt{rCopula}. For low-dimensional modeling situtations, copula is
+suitable, yet the entent of exact matching of margins automatically is
+unclear. High-dimesional random vectors via \texttt{copula} takes an
+impractical amount of computing time \citep{Li2019gpu}. The
+\texttt{nortaRA} R package matches Pearson coefficients near exactly and
+is reasonable for low-dimenional settings. In sum, we know of no other
+\texttt{R} package that can produce HD random vectors with flexible
+margins and dependence modeling. Table (\#tab:compare-table) summarizes
+the above discussion.
+
+\begin{table}[]
 \centering
-\begin{tabular}[t]{lrrrrrrrr}
-\toprule
-  & mpg & cyl & disp & hp & drat & wt & qsec & vs\\
-\midrule
-Mazda RX4 & 21.0 & 6 & 160.0 & 110 & 3.90 & 2.620 & 16.46 & 0\\
-Mazda RX4 Wag & 21.0 & 6 & 160.0 & 110 & 3.90 & 2.875 & 17.02 & 0\\
-Datsun 710 & 22.8 & 4 & 108.0 & 93 & 3.85 & 2.320 & 18.61 & 1\\
-\addlinespace
-Hornet 4 Drive & 21.4 & 6 & 258.0 & 110 & 3.08 & 3.215 & 19.44 & 1\\
-Hornet Sportabout & 18.7 & 8 & 360.0 & 175 & 3.15 & 3.440 & 17.02 & 0\\
-Valiant & 18.1 & 6 & 225.0 & 105 & 2.76 & 3.460 & 20.22 & 1\\
-\addlinespace
-Duster 360 & 14.3 & 8 & 360.0 & 245 & 3.21 & 3.570 & 15.84 & 0\\
-Merc 240D & 24.4 & 4 & 146.7 & 62 & 3.69 & 3.190 & 20.00 & 1\\
-Merc 230 & 22.8 & 4 & 140.8 & 95 & 3.92 & 3.150 & 22.90 & 1\\
-\addlinespace
-Merc 280 & 19.2 & 6 & 167.6 & 123 & 3.92 & 3.440 & 18.30 & 1\\
-\bottomrule
+\caption{Selected feature comparison of `bigsimr` to other existing `R` multivariate simulation packages.}
+\label{tab:compare-table}
+\begin{tabular}{@{}llll@{}}
+R Package & Scales to HD & Hetereogenous Margins & Flexible Dependence \\ \midrule
+MASS      & No           & No                    & No                  \\
+mvnfast   & Yes          & No                    & No                  \\
+GenOrd    & No           & Discrete margins      & Yes                 \\
+copula    & No           & Yes                   & No                  \\
+nortaRA   & No           & Yes                   & No                  \\
+bigsimr   & Yes          & Yes                   & Yes                
 \end{tabular}
 \end{table}
-
-\end{CodeChunk}
 
@@
 \begin{enumerate}
-\def\labelenumi{\roman{enumi}.}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  \begin{enumerate}
+  \def\labelenumii{\alph{enumii})}
   \tightlist
   \item
-  Estimate the correlation matrix from the ``data'' in the \emph{Compute
-  Correlation} step.
+    Estimate the correlation matrix from the ``data'' in the
+    \emph{Compute Correlation} step.
   \end{enumerate}
-
+\item
   \begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
+  \def\labelenumii{\alph{enumii})}
+  \setcounter{enumii}{1}
   \tightlist
   \item
-  Map the correlations to initialize the algorithm (Pearson to Pearson,
-  Spearman to Pearson, or Kendall to Pearson) in the \emph{Adjust
-  Correlation} step.
+    Map the correlations to initialize the algorithm (Pearson to
+    Pearson, Spearman to Pearson, or Kendall to Pearson) in the
+    \emph{Adjust Correlation} step.
+  \end{enumerate}
 \item
@@
 \begin{CodeChunk}
-
-Table: Total time to produce 10,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.
-
-|Dimension |Correlation type | Total Time (Seconds)|
-|:---------|:----------------|--------------------:|
-|100       |Pearson          |                0.081|
-|100       |Spearman         |                0.027|
-|100       |Kendall          |                0.075|
-|250       |Pearson          |                0.378|
-|250       |Spearman         |                0.062|
-|250       |Kendall          |                0.375|
-|500       |Pearson          |                1.425|
-|500       |Spearman         |                0.135|
-|500       |Kendall          |                1.495|
+\begin{table}
+
+\caption{\label{tab:tab:ch040-moderateDtab}Total time to produce 10,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.}
+\centering
+\begin{tabular}[t]{llr}
+\toprule
+Dimension & Correlation type & Total Time (Seconds)\\
+\midrule
+100 & Pearson & 0.081\\
+100 & Spearman & 0.027\\
+100 & Kendall & 0.075\\
+\addlinespace
+250 & Pearson & 0.378\\
+250 & Spearman & 0.062\\
+250 & Kendall & 0.375\\
+\addlinespace
+500 & Pearson & 1.425\\
+500 & Spearman & 0.135\\
+500 & Kendall & 1.495\\
+\bottomrule
+\end{tabular}
+\end{table}
 
@@
 \hypertarget{examples}{%
-\section{RNA-seq data applications}\label{examples}}
+\section{RNA-seq data application}\label{examples}}
 

### 2021-05-25 000-introduction.Rmd
Revise for JSS submission up to Chapter 4

@@
 Massive high-dimensional (HD) data sets are now commonplace in many areas of scientific inquiry.
-As new methods are developed for data, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of analyzing such proposed methodology --- such as false positive rates, statistical power, interval coverage, and robustness --- often with comparison to existing methods.
+As new methods are developed for data analysis, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of proposed methodology --- such as false positive rates, statistical power, interval coverage, and robustness --- often to compare to methods.
 Further, efficient simulation empowers statistical computing strategies, such as the parametric bootstrap [@Chernick2008] to simulate from a hypothesized null model, providing inference in analytically challenging settings.
-Such Monte Carlo (MC) techniques become difficult for high-dimensional data using existing algorithms and tools.
-This is particularly true when simulating massive multivariate, non-normal distributions, arising naturally in many fields of study.
+Such Monte Carlo (MC) techniques become difficult for HD dependent data using existing algorithms and tools.
+This is particularly true when simulating massive multivariate, non-normal distributions, arising in many fields of study.
 
-As many have noted, it can be vexing to simulate dependent, non-normal/discrete data, even for low dimensional settings [@MB13; @XZ19].
-For continuous non-normal multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula approaches [@Nelsen2007] are well-studied, with flexible, robust software available [@Yan2007; @Chen2001].
-Yet these approaches do not scale in a timely fashion to high-dimensional problems [@Li2019gpu].
+As many have noted, it can be vexing to simulate dependent, non-normal/discrete data, even for low-dimensional (LD) settings [@MB13; @XZ19].
+For continuous non-normal LD multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula approaches [@Nelsen2007] are well-studied, with flexible, robust software available [@Yan2007; @Chen2001].
+Yet these approaches do not scale in a timely fashion to HD problems [@Li2019gpu].
 For discrete data, early simulation strategies had major flaws, such as failing to obtain the full range of possible correlations (e.g., admitting only positive correlations: see @Park1996).
-While more recent approaches [@MB13; @Xia17; @BF17] have largely remedied this issue for low-dimensional problems, the existing tools are not designed to scale to high dimensions.
+While more recent approaches [@MB13; @Xia17; @BF17] have largely remedied this issue for LD problems, the existing tools are not designed to scale to high dimensions for discrete marginal distributions.
 
-Another central issue lies in characterizing dependence between components in the high-dimensional random vector.
+Another central issue lies in characterizing dependence between components in the HD random vector.
 The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc).
@@
 Throughout, we focus on matching these nonparametric dependence measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology.
-We do, however, provide Pearson matching with some caveats.
+We do, however, implement Pearson matching, but several layers of approximation is required.
 
@@
 The crux of the method lies in the construction of a Gaussian copula in the spirit of the NORTA procedure.
-Further, we introduce the `bigsimr` R package that provides high-performance software implementing our algorithm.
-The algorithm design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed-form relationships between dependence measures for the multivariate normal (MVN) distribution.
+As we will describe in more detail, the algorithm's design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed-form relationships between dependence measures for the multivariate normal (MVN) distribution.
+For our method, we developed a high-performance implementation: the `bigsimr` R package, written in native Julia code.
 
@@
 Then we describe and justify our simulation methodology and related algorithms.
-Next, we detail an illustrative low-dimensional example of basic use of the `bigsimr` R package.
-Then we proceed with Monte Carlo studies under various bivariate distributional assumptions to assess accuracy.
-We conclude the Monte Carlo evaluations by summarizing the computation time for increasingly higher dimensional vectors.
+We proceed by providing an illustrative LD `bigsimr` workflow.
+Next we conduct MC studies under various bivariate distributional assumptions to evaluate performance and accuracy.
 After the MC evaluations, we simulate random vectors motivated by our RNA-seq example, evaluate the accuracy, and provide example statistical computing tasks, namely MC estimation of joint probabilities and evaluating HD correlation estimation efficiency.

### 2021-05-25 010-background.Rmd
Revise for JSS submission up to Chapter 4

@@
 
-The `bigsimr` `R` package presented here provides multiple algorithms that operate with high-dimensional multivariate data; however, all these algorithms were originally designed to support a single task: to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics.
-Specifically, our goal is to efficiently simulate a large number, $B$, of HD random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with *correlated* components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i$.
+The `bigsimr` `R` package presented here provides multiple algorithms that operate with HD multivariate data; however, all these algorithms were originally designed to support a single task: to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics.
+Specifically, our goal is to efficiently simulate a large number, $B$, of HD random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with *correlated* components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i, i=1,\ldots,d$.
 
@@
 * BP2: Flexible dependence, meaning that the number of bivariate marginals can be equal to the number of dependence parameters.
-* BP3: Flexible marginal modeling, generating heterogeneous data --- possibly from differing probability families.
+* BP3: Flexible marginal modeling, generating heterogeneous data --- including mixed continuous and discrete margins.
 
@@
 
-Simulating high-dimensional, non-normal, correlated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients.
-The RNA-seq data-generating process involves counting how often a particular form of human messenger RNA (mRNA) is expressed in a biological sample.
-RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data.
+Simulating HD, non-normal, correlated data motivates this work, in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients.
+The RNA-seq data-generating process involves counting how often a particular form of messenger RNA (mRNA) is expressed in a biological sample.
+RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in HD data.
 For human-derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted [@Schissler2019].
@@
 
-We illustrate our methodology using a well-studied Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments).
+We illustrate our methodology using the Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments).
 For ease of modeling and simplicity of exposition, we only consider high expressing genes.
@@
 
-```{r ch010-realDataTab}
+```{r tab:ch010-realDataTab}
 set.seed(2020-02-25)
@@
 
-```{r ch010-realDataFig, cache=FALSE, fig.width=8, fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes from Table 1.
+```{r ch010-realDataFig, cache=FALSE, fig.width=8, fig.cap="Pairwise scatterplots for three example genes (Table 1), with estimated Spearman's correlations and marginal density estimates displayed.
 The data possess outliers, heavy-right tails, are discrete, and have non-trivial intergene correlations.
@@
 In multivariate analysis, an analyst must select a metric to quantify dependency.
-The most widely-known is the Pearson (product-moment) correlation coefficient that describes the linear association between two random variables $X$ and $Y$, and, it is given by
+The most widely-known is the Pearson correlation coefficient that describes the linear association between two random variables $X$ and $Y$, and, it is given by
 
@@
 For non-normal marginals with monotone correlation patterns, however, $\rho_P$ suffers some drawbacks and may mislead or fail to capture important relationships [@MK01].
-Alternatively in these settings, analysts often prefer rank-based correlation measures to describe the degree of monotonic association.
+Alternatively, analysts often prefer rank-based correlation measures to describe the degree of monotonic association.
 
 Two nonparametric, rank-based measures common in practice are Spearman's correlation (denoted $\rho_S$) and Kendall's $\tau$.
-Spearman's $\rho_S$ has an appealing correspondence as the Pearson correlation coefficient on *ranks* of the values, thereby captures nonlinear yet monotone relationships.
-Kendall's $\tau$, on the other hand, is the difference in probabilities of concordant and discordant pairs of observations, $(X_i, Y_i)$ and $(X_j, Y_j)$, with concordance meaning that orderings have the same direction (e.g., if $X_i < X_j$, then $Y_i < Y_j$).
+Spearman's $\rho_S$ has an appealing correspondence as the Pearson correlation coefficient on *ranks* of the values, thereby capturing nonlinear yet monotone relationships.
+Kendall's $\tau$, on the other hand, is the difference in probabilities of concordant and discordant pairs of observations, $(X_i, Y_i)$ and $(X_j, Y_j)$ (concordance meaning that orderings have the same direction, e.g., if $X_i < X_j$, then $Y_i < Y_j$).
 Note that concordance is determined by the ranks of the values, not the values themselves.
@@
 Given two marginal distributions, $\rho_P$ is not free to vary over the entire range of possible correlations $[-1,1]$.
-The so-called *Frechet-Hoeffding bounds* are well-studied [@Nelsen2007; @BF17].
+The *Frechet-Hoeffding bounds* are well-studied [@Nelsen2007; @BF17].
 These constraints cannot be overcome through algorithm design.
@@
 
-There is a strong connection of our simulation strategy to Gaussian *copulas* (see @Nelsen2007 for a technical introduction).
+There is a connection of our simulation strategy to Gaussian *copulas* [@Nelsen2007].
 A copula is a distribution function on $[0,1]^d$ that describes a multivariate probability distribution with standard uniform marginals.
-This provides a powerful, natural way to characterize joint probability structures.
+This provides a powerful, natural way to characterize joint probability.
 Consequently, the study of copulas is an important and active area of statistical theory and practice.
@@
 
-A Gaussian copula has marginal CDFs that are all standard normal, $F_i = \Phi, \forall \, i$.
+A Gaussian copula has marginal CDFs that are all standard normal, $F_i = \Phi, \forall \, i$, where $\Phi(\cdot)$ is the standard normal CDF.
 This representation corresponds to a multivariate normal (MVN) distribution with standard normal marginal distributions and covariance matrix ${\bf R_P}$.
-As the marginals are standardized to have unit variance, however, ${\bf R_P}$ is a Pearson correlation matrix.
+Since the marginals are standardized to have unit variance, however, ${\bf R_P}$ is a Pearson correlation matrix.
 If $F_{{\bf R}}$ is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula $C_{{\bf R}}$ is defined through
@@
 
-where $\Phi(\cdot)$ is the standard normal CDF.
-Note that the copula $C_{{\bf R}}$ is the familiar multivariate normal CDF of the random vector $(\Phi(X_1), \ldots, \Phi(X_d))$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})$.
+Here the copula function $C_{{\bf R}}$ is the familiar multivariate normal CDF of the random vector $(\Phi(X_1), \ldots, \Phi(X_d))$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})$.
 
@@
 
-## Relevant Multivariate Simulation software Comparison
+## Other Multivariate Simulation R Packages
 
-To place `Bigsimr`'s implementation and algorithms in context with existing approaches, as well as make this work's contribution more evident, we make some qualitative comparisons to other `R` multivariate simulation packages.
+To place `bigsimr`'s implementation and algorithms in context with existing approaches, here we qualitatively compare to other `R` multivariate simulation packages.
 The `MASS` packages `mvnorm()` has long been available to produce random vectors from multivariate normal (and $t$) distributions.
 This familiar procedure cannot simulate from arbitrary margins and dependency measures, as well as fails to readily scale to high dimensions.
-Recently, the high-performance package `mvnfast` has been released that provides scaleable multivariate normal generation, but this doesn't completely satistify our goals of flexible margins and dependency modeling.
-For dependent discrete data, small-dimesional packages exist, such as `GenOrd`.
-The full-featured package `copula` provides many different copulas and a fairly general random generator `rCopula`.
-For low-dimensional modeling situtations, copula is suitable, yet the entent of exact matching of margins automatically is unclear.
+Recently, the high-performance package `mvnfast` [@Fasiolo2016] has been released that provides HD multivariate normal generation, but this lacks flexibility in marginal models and dependency modeling.
+For dependent discrete data, LD packages exist, such as `GenOrd`.
+For copula computation, the full-featured package `copula` provides many different copulas and a fairly general LD random generator `rCopula()`.
 High-dimesional random vectors via `copula` takes an impractical amount of computing time [@Li2019gpu].
-The `nortaRA` R package matches Pearson coefficients near exactly and is reasonable for low-dimenional settings.
-In sum, we know of no other `R` package that can produce HD random vectors with flexible margins and dependence modeling.
+The `nortaRA` R package matches Pearson coefficients near exactly and is reasonable for LD settings.
+Finally, we know of no other `R` package that can produce HD random vectors with flexible margins and dependence modeling.
 Table (\#tab:compare-table) summarizes the above discussion.
@@
 \centering
-\caption{Selected feature comparison of `bigsimr` to other existing `R` multivariate simulation packages.}
+\caption{Selected feature comparison of bigsimr to existing R packages for multivariate simulation.}
 \label{tab:compare-table}

### 2021-05-25 020-simulation-algorithm.Rmd
working template
Revise for JSS submission up to Chapter 4

@@
 To characterize dependency, every pair $(Y_i, Y_j)$ has a given Pearson correlation $\rho_P$, Spearman correlation $\rho_S$, and/or Kendall's $\tau$.
-The method can be described as a *high-performance Gaussian copula* (Equation \@ref(eq:gauss)) providing a high-dimensional NORTA-inspired algorithm.
+The method can be described as a *high-performance Gaussian copula* (Equation \@ref(eq:gauss)) providing a HD NORTA-inspired algorithm.
 
@@
 
-1. Simulate a random vector $\bf Z$ with $d$ *independent and identically distributed* (iid) standard normal components.
+1. Simulate a random vector $\bf Z$ with $d$ independent and identically distributed (iid) standard normal components.
 2. Determine the input matrix $\Sigma_{\bf Z}$ that corresponds with the specified output $\Sigma_{\bf Y}$.
@@
 
-With modern parallel computing, steps 1, 3, 4, 5 are readily implemented as high-performance, multi-core and/or graphical-processing-unit (GPU) accelerated algorithms --- providing fast scalability using readily-available hardware.
+With modern parallel computing, steps 1, 3, 4, 5 are readily implemented as high-performance, multi-core and/or graphical-processing-unit (GPU) accelerated algorithms --- providing fast scalability to HD.
 
 Matching specified Pearson correlation coefficients exactly (step 2 above), however, is computationally costly.
-In general, there is no closed-form correspondence between the components of the input $\Sigma_{\bf Z}$ and target $\Sigma_{\bf Y}$.
+In general, there is no convenient correspondence between the components of the input $\Sigma_{\bf Z}$ and target $\Sigma_{\bf Y}$.
 Matching the correlations involves evaluating or approximating $\binom{d}{2}$ integrals of the form
@@
 where $\phi(\cdot)$ is the joint probability distribution function of two correlated standard normal variables.
-For HD data, (nearly) exact evaluation may become too costly to enable practical simulation studies.
-For low-dimensional problems, however, methods and tools exist to match Pearson correlations precisely; see [@Xia17] and the publicly available `nortaRA` R package [@Chen2001].
-As described later, to enable HD Pearson matching we approximate these integrals.
+For HD simulation, exact evaluation may become too costly to enable practical simulation studies.
+Our remedy to enable HD Pearson matching in `bigsimr` is Hermite polynomial approximation of these $\choose{d}{2}$ double integrals, as suggested by @XZ19.
 
@@
 Yet the choice of copula-based simulation algorithm affects which joint distributions may be simulated.
-Even in low-dimensional spaces (e.g., $d=3$), there exist valid multivariate distributions with *feasible* Pearson correlation matrices that NORTA cannot match exactly [@LH75].
+Even in LD spaces (e.g., $d=3$), there exist valid multivariate distributions with *feasible* Pearson correlation matrices that NORTA cannot match exactly [@LH75].
 This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined the resultant matrix is indefinite.
@@
 There is hope, however, as @GH02 also showed that replacing an indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$.
-This provides evidence that our nearest positive definite (PD) augmented approach will maintain reasonable accuracy if our input matching scheme returns an indefinite matrix, at least for the rank-based matching scheme described above.
+This provides guidance that our nearest positive definite (PD) augmented approach has promise provide adequate accuracy if our input matching scheme returns an indefinite matrix.
 
-## Random vector generator {#rand-vec-gen}
+## bigsimr rvec: Random vector generator {#rand-vec-gen}
 
-We now describe our algorithm to generate random vectors, which mirrors the classical NORTA algorithm above with some modifications for rank-based dependency matching:
+We now describe our algorithm to generate random vectors, which extends NORTA to HD via nearest correlation matrix replacement and provides rank-based dependency matching:
 
@@
 
-1. Mapping step
-    * (i) Convert the target Spearman correlation matrix $R_S$ to the corresponding MVN Pearson correlation $R_X$. Alternatively,
-    * (i') Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$. Alternatively,
-    * (i'') Convert the target Pearson correlation matrix to $R_P$ to the corresponding, approximate MVN Pearson correlation $R_X$.
+1. Mapping step, either
+    * Convert the target Spearman correlation matrix $R_S$ to the corresponding MVN Pearson correlation $R_X$. Alternatively,
+    * Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$. Alternatively,
+    * Convert the target Pearson correlation matrix to $R_P$ to the corresponding, approximate MVN Pearson correlation $R_X$.
 2. Check admissibility and, if needed, compute the nearest correlation matrix.
-    * (i) Check that $R_X$ is a correlation matrix, a positive definite matrix with 1's along the diagonal.
-    * (ii) If $R_X$ is a correlation matrix, the input matrix is *admissible* in this scheme. Otherwise
-    * (ii') Replace $R_X$ with the nearest correlation matrix $\tilde{R}_X$, in the Frobenius norm.
+    * Check that $R_X$ is a correlation matrix, a positive definite matrix with 1's along the diagonal.
+    * If $R_X$ is a correlation matrix, the input matrix is *admissible* in this scheme. Otherwise,
+    * Replace $R_X$ with the nearest correlation matrix $\tilde{R}_X$, in the Frobenius norm.
 3. Gaussian copula
-    * (i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)$.
-    * (ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz. $U_i=\Phi(X_i)$, $i=1, \ldots, d$.
-    * (iii) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$.
+    * Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)$.
+    * Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz. $U_i=\Phi(X_i)$, $i=1, \ldots, d$.
+    * Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$.
 
-\setstretch{2.0}
-
-*Step 1*
-
-The first two descriptions of the *Mapping step* employ the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertKendall) and \@ref(eq:convertSpearman), respectively (implemented as `cor_covert`).
+*Step 1: Mapping step.* We first employ the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertKendall) and \@ref(eq:convertSpearman), respectively (implemented as `cor_covert()`).
 Initializing our algorithm to match the nonparametric correlations by computing these equations for all pairs is computationally trivial.
@@
 Briefly, the many double integrals of the form in \@ref(eq:pearsonIntegralRelation) are approximated by weighted sums of Hermite polynomials.
-Matching coefficients for pairs of continuous distributions is made tractable by this method, but for discrete distributions (especially discrete distributions with large support sets or infinite support), the problem is not so simple or efficient.
+Matching coefficients for pairs of continuous distributions is made tractable by this method, but for discrete distributions (especially discrete distributions with large support sets or infinite support), the approximation is computationally expensive.
+To remedy this, we further approximate discrete distributions by a continuous distribution using Generalized S-Distributions [@muino2006gs].
 
-Our solution is to approximate discrete distributions by a continuous distribution.
-The question then becomes that of which distribution to use.
-We found that Generalized S-Distributions [@muino2006gs] solve this problem by approximating a wide range of unimodal distributions, both continuous and discrete.
-Using the GS-Distribution approximation of discrete distributions in Pearson matching scheme above yields favorable results.
-
-*Step 2*
-
-Once $R_X$ has been determined, we check admissibility of the adjusted correlation via the steps described above.
+*Step 2: Admissibility check and nearest correlation matrix computation.* Once $R_X$ has been determined, we check admissibility of the adjusted correlation via the steps described above.
 If $R_X$ is not a valid correlation matrix, then we compute the nearest correlation matrix.
 Finding the nearest correlation matrix is a common statistical computing problem.
-The defacto default function in `R` is `Matrix::nearPD`, an alternating projection algorithm due to @higham2002computing.
+The defacto function in `R` is `Matrix::nearPD`, an alternating projection algorithm due to @higham2002computing.
 As implemented the function fails to scale to HD.
-Instead, we provide the quadratically convergent algorithm based on the theory of strongly semi-smooth matrix functions (@qi2006quadratically).
-The nearest correlation matrix problem can be written down as the following convex optimization problem:
+Instead, we provide the quadratically-convergent algorithm based on the theory of strongly semi-smooth matrix functions (@qi2006quadratically).
+The nearest correlation matrix problem can be written down as the following convex optimization problem:  $\mathrm{min} \quad \frac{1}{2} \Vert G - X \Vert^2, \quad \mathrm{s.t.} \quad X_{ii} = 1, \quad i = 1, \ldots , n, \quad X \in S_{+}^{n}$.
 
-\begin{align*}
-    \mathrm{min}\quad & \frac{1}{2} \Vert G - X \Vert^2 \\
-    \mathrm{s.t.}\quad & X_{ii} = 1, \quad i = 1, \ldots , n, \\
-    & X \in S_{+}^{n}
-\end{align*}
+For nonparametric correlation measures, our algorithm allows the generation of HD multivariate data with arbitrary marginal distributions with a broad class of admissible Spearman correlation matrices and Kendall $\tau$ matrices.
+The admissible classes consist of the matrices that map to a Pearson correlation matrix for a MVN.
+In particular, if we let $X$ be MVN with $d$ components and denote $\Omega_P = \{ R_P : R_P \textrm{ is a correlation matrix for } X \}, \quad \Omega_K = \{ R_K : R_K \textrm{ is a correlation matrix for } X \}, \quad \Omega_S = \{ R_S : R_S \textrm{ is a correlation matrix for } X \}$.
+There are 1-1 mappings between these sets.
+We conjecture that the sets of admissible $R_S$ and $R_K$ are not highly restrictive.
+In particular, $R_P$ is approximately $R_S$ for a MVN, suggesting that the admissible set $\Omega_S$ should be flexible as $R_P$ can be any PD matrix with 1's along the diagonal.
+We provide methods to check whether a target $R_S$ is an element of the *admissible set* $\Omega_S$.
 
-For nonparametric correlation measures, our algorithm allows the generation of high-dimensional multivariate data with arbitrary marginal distributions with a broad class of admissible Spearman correlation matrices and Kendall $\tau$ matrices.
-The admissible classes consist of the matrices that map to a Pearson correlation matrix for a MVN.
-In particular, if we let $X$ be MVN with $d$ components and set
-
-\begin{align*}
-\Omega_P &= \{ R_P : R_P \textrm{ is a correlation matrix for } X \} \\
-\Omega_K &= \{ R_K : R_K \textrm{ is a correlation matrix for } X \} \\
-\Omega_S &= \{ R_S : R_S \textrm{ is a correlation matrix for } X \} \\
-\end{align*}
-
-then there are 1-1 mappings between these sets.
-We conjecture informally that the sets of admissible $R_S$ and $R_K$ are not highly restrictive.
-In particular, $R_P$ is approximately $R_S$ for a MVN, suggesting that the admissible set $\Omega_S$ should be flexible as $R_P$ can be any PD matrix with 1's along the diagonal.
-We provide methods to check whether a target $R_S$ is an element of the *admissible set* $\Omega_S$ (and similarly for $R_K$).
-
-There is an increasing probability of encountering a non-admissible correlation matrix as dimension increases.
+There is an increasing probability of encountering a inadmissible correlation matrix as dimension increases.
 In our experience, the mapping step for large $d$ almost always produces a $R_X$ that is not a correlation matrix.
@@
 
-*Step 3*
-
-The final step implements a NORTA-inspired, Gaussian copula approach to produce the desired margins.
+*Step 3: Gaussian copula.* The final step implements a NORTA-inspired, Gaussian copula approach to produce the desired margins.
 Steps 1 and 2 determine the MVN Pearson correlation values that will eventually match the target correlation.
-Then all that is required is a fast MVN simulator, a standard normal CDF, and well-defined quantile functions for marginals.
-The MVN is transformed to a copula (distribution with standard uniform margins) by applying the normal CDF $\phi(\cdot)$.
-The quantile functions $F^{-1}$ are applied across the margins to return the desired random vector ${\bf Y}$.
+Step 3 requires a fast MVN simulator, a standard normal CDF, and well-defined quantile functions for marginals.
+The MVN is transformed to a copula (distribution with standard uniform margins) by applying the normal CDF $\Phi(\cdot)$.
+Finally, the quantile functions $F^{-1}$ are applied across the margins to return the desired random vector ${\bf Y}$.

### 2021-05-25 060-conclusion-discussion.Rmd
Revise for JSS submission up to Chapter 4

@@
 
-We have introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a user-friendly, high-performance `R` package [`bigsimr`].
+We introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a user-friendly, high-performance `R` package [`bigsimr`].
 The random vector generation method is inspired by NORTA [@Cario1997] and Gaussian copula-based approaches [@MB13, @BF17, @Xia17].
@@
 
+<!-- 
 It is customary to compare new tools and algorithms directly to existing competing methods and software.
@@
 For the bivariate simulations, existing packages such as [`nortaRA`](https://github.com/superdesolator/NORTARA) work well to match Pearson correlations exactly.
+-->
 

### 2021-05-25 article_bigsimr.Rmd
Revise for JSS submission up to Chapter 4

@@
   - name: Edward J. Bedrick
-    affiliation: University of Nevada
+    affiliation: University of Arizona
   - name: Alexander D. Knudson
@@
   - name: Walter W. Piegorsch
-    affiliation: University of Nevada
+    affiliation: University of Arizona
   - name: Duc Tran
@@
 title:
-  formatted: "Simulating High-Dimensional Multivariate Data using the \\pkg{Bigsimr} Package"
-  plain:     "Simulating High-Dimensional Multivariate Data using the Bigsimr Package"
-  short:     "\\pkg{Bigsimr}: Simulate High-Dimensional Data"
+  formatted: "Simulating High-Dimensional Multivariate Data using the \\pkg{bigsimr} Package"
+  plain:     "Simulating High-Dimensional Multivariate Data using the bigsimr Package"
+  short:     "\\pkg{bigsimr}: Simulate High-Dimensional Data"
 abstract: >
-  It is critical to realistically simulate data when employing Monte Carlo techniques and evaluating statistical methodology.
+  It is critical to accurately simulate data when employing Monte Carlo techniques and evaluating statistical methodology.
 Measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments.
 Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations.
-This article introduces the \\pkg{Bigsimr} R package.
-This high-level package provides a flexible and scalable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions and known Pearson, Spearman, or Kendall correlation matrix.
-\\pkg{Bigsimr} contains additional high-performance features, including multi-core algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix.
+To meet this gap, we developed the bigsimr R package for high-dimensional random vector simulation with arbitrary marginal distributions and known Pearson, Spearman, or Kendall correlation matrix.
+bigsimr contains additional high-performance features, including multi-core algorithms to estimate correlation and compute the nearest correlation matrix.
 Monte Carlo studies quantify the accuracy and scalability of our approach, up to $d=10,000$.
-Finally, we demonstrate example applications enabled via \\pkg{Bigsimr} by applying these methods to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples.
+Finally, we describe example workflows and apply to a motivating high-dimensional data set --- RNA-sequencing data obtained from breast cancer tumor samples.
 keywords:

### 2021-05-25 article_bigsimr.tex
Revise for JSS submission up to Chapter 4

@@
 Alfred G. Schissler\\University of Nevada \And Edward J.
-Bedrick\\University of Nevada \And Alexander D. Knudson\\University of
+Bedrick\\University of Arizona \And Alexander D. Knudson\\University of
 Nevada \AND Tomasz J. Kozubowski\\University of Nevada \And Tin
@@
 \AND Juli Petereit\\University of Nevada \And Walter W.
-Piegorsch\\University of Nevada \And Duc Tran\\University of Nevada
+Piegorsch\\University of Arizona \And Duc Tran\\University of Nevada
 }
 \title{Simulating High-Dimensional Multivariate Data using the
-\pkg{Bigsimr} Package}
+\pkg{bigsimr} Package}
 
@@
 \Plaintitle{Simulating High-Dimensional Multivariate Data using the
-Bigsimr Package}
-\Shorttitle{\pkg{Bigsimr}: Simulate High-Dimensional Data}
+bigsimr Package}
+\Shorttitle{\pkg{bigsimr}: Simulate High-Dimensional Data}
 
 \Abstract{
-It is critical to realistically simulate data when employing Monte Carlo
+It is critical to accurately simulate data when employing Monte Carlo
 techniques and evaluating statistical methodology. Measurements are
@@
 to simulation designs that posit independence or perform arbitrary data
-transformations. This article introduces the
-\textbackslash pkg\{Bigsimr\} R package. This high-level package
-provides a flexible and scalable procedure to simulate high-dimensional
-random vectors with arbitrary marginal distributions and known Pearson,
-Spearman, or Kendall correlation matrix. \textbackslash pkg\{Bigsimr\}
-contains additional high-performance features, including multi-core
-algorithms to simulate random vectors, estimate correlation, and compute
-the nearest correlation matrix. Monte Carlo studies quantify the
-accuracy and scalability of our approach, up to \(d=10,000\). Finally,
-we demonstrate example applications enabled via
-\textbackslash pkg\{Bigsimr\} by applying these methods to a motivating
-dataset --- RNA-sequencing data obtained from breast cancer tumor
-samples.
+transformations. To meet this gap, we developed the bigsimr R package
+for high-dimensional random vector simulation with arbitrary marginal
+distributions and known Pearson, Spearman, or Kendall correlation
+matrix. bigsimr contains additional high-performance features, including
+multi-core algorithms to estimate correlation and compute the nearest
+correlation matrix. Monte Carlo studies quantify the accuracy and
+scalability of our approach, up to \(d=10,000\). Finally, we describe
+example workflows and apply to a motivating high-dimensional data set
+--- RNA-sequencing data obtained from breast cancer tumor samples.
 }
@@
 Massive high-dimensional (HD) data sets are now commonplace in many
-areas of scientific inquiry. As new methods are developed for data, a
-fundamental challenge lies in designing and conducting simulation
-studies to assess the operating characteristics of analyzing such
-proposed methodology --- such as false positive rates, statistical
-power, interval coverage, and robustness --- often with comparison to
-existing methods. Further, efficient simulation empowers statistical
-computing strategies, such as the parametric bootstrap
-\citep{Chernick2008} to simulate from a hypothesized null model,
-providing inference in analytically challenging settings. Such Monte
-Carlo (MC) techniques become difficult for high-dimensional data using
-existing algorithms and tools. This is particularly true when simulating
-massive multivariate, non-normal distributions, arising naturally in
-many fields of study.
+areas of scientific inquiry. As new methods are developed for data
+analysis, a fundamental challenge lies in designing and conducting
+simulation studies to assess the operating characteristics of proposed
+methodology --- such as false positive rates, statistical power,
+interval coverage, and robustness --- often to compare to methods.
+Further, efficient simulation empowers statistical computing strategies,
+such as the parametric bootstrap \citep{Chernick2008} to simulate from a
+hypothesized null model, providing inference in analytically challenging
+settings. Such Monte Carlo (MC) techniques become difficult for HD
+dependent data using existing algorithms and tools. This is particularly
+true when simulating massive multivariate, non-normal distributions,
+arising in many fields of study.
 
 As many have noted, it can be vexing to simulate dependent,
-non-normal/discrete data, even for low dimensional settings
-\citep{MB13, XZ19}. For continuous non-normal multivariate data, the
+non-normal/discrete data, even for low-dimensional (LD) settings
+\citep{MB13, XZ19}. For continuous non-normal LD multivariate data, the
 well-known NORmal To Anything (NORTA) algorithm \citep{Cario1997} and
@@
 flexible, robust software available \citep{Yan2007, Chen2001}. Yet these
-approaches do not scale in a timely fashion to high-dimensional problems
+approaches do not scale in a timely fashion to HD problems
 \citep{Li2019gpu}. For discrete data, early simulation strategies had
@@
 \citet{Park1996}). While more recent approaches
-\citep{MB13, Xia17, BF17} have largely remedied this issue for
-low-dimensional problems, the existing tools are not designed to scale
-to high dimensions.
+\citep{MB13, Xia17, BF17} have largely remedied this issue for LD
+problems, the existing tools are not designed to scale to high
+dimensions for discrete marginal distributions.
 
 Another central issue lies in characterizing dependence between
-components in the high-dimensional random vector. The choice of
-correlation in practice usually relates to the eventual analytic goal
-and distributional assumptions of the data (e.g., non-normal, discrete,
+components in the HD random vector. The choice of correlation in
+practice usually relates to the eventual analytic goal and
+distributional assumptions of the data (e.g., non-normal, discrete,
 infinite support, etc). For normal data, the Pearson product-moment
@@
 rank-based measures possess invariance properties favorable in our
-proposed methodology. We do, however, provide Pearson matching with some
-caveats.
+proposed methodology. We do, however, implement Pearson matching, but
+several layers of approximation is required.
 
@@
 simulation algorithm. The crux of the method lies in the construction of
-a Gaussian copula in the spirit of the NORTA procedure. Further, we
-introduce the \texttt{bigsimr} R package that provides high-performance
-software implementing our algorithm. The algorithm design leverages
-useful properties of nonparametric correlation measures, namely
-invariance under monotone transformation and well-known closed-form
-relationships between dependence measures for the multivariate normal
-(MVN) distribution.
+a Gaussian copula in the spirit of the NORTA procedure. As we will
+describe in more detail, the algorithm's design leverages useful
+properties of nonparametric correlation measures, namely invariance
+under monotone transformation and well-known closed-form relationships
+between dependence measures for the multivariate normal (MVN)
+distribution. For our method, we developed a high-performance
+implementation: the \texttt{bigsimr} R package, written in native Julia
+code.
 
@@
 (RNA-seq) breast cancer data. Then we describe and justify our
-simulation methodology and related algorithms. Next, we detail an
-illustrative low-dimensional example of basic use of the
-\texttt{bigsimr} R package. Then we proceed with Monte Carlo studies
-under various bivariate distributional assumptions to assess accuracy.
-We conclude the Monte Carlo evaluations by summarizing the computation
-time for increasingly higher dimensional vectors. After the MC
-evaluations, we simulate random vectors motivated by our RNA-seq
-example, evaluate the accuracy, and provide example statistical
-computing tasks, namely MC estimation of joint probabilities and
-evaluating HD correlation estimation efficiency. Finally, we discuss the
-method's utility, limitations, and future directions.
+simulation methodology and related algorithms. We proceed by providing
+an illustrative LD \texttt{bigsimr} workflow. Next we conduct MC studies
+under various bivariate distributional assumptions to evaluate
+performance and accuracy. After the MC evaluations, we simulate random
+vectors motivated by our RNA-seq example, evaluate the accuracy, and
+provide example statistical computing tasks, namely MC estimation of
+joint probabilities and evaluating HD correlation estimation efficiency.
+Finally, we discuss the method's utility, limitations, and future
+directions.
 
@@
 The \texttt{bigsimr} \texttt{R} package presented here provides multiple
-algorithms that operate with high-dimensional multivariate data;
-however, all these algorithms were originally designed to support a
-single task: to generate random vectors drawn from multivariate
-probability distributions with given marginal distributions and
-dependency metrics. Specifically, our goal is to efficiently simulate a
-large number, \(B\), of HD random vectors
-\({\bf Y}=(Y_1, \ldots, Y_d)^\top\) with \emph{correlated} components
-and heterogeneous marginal distributions, described via cumulative
-distribution functions (CDFs) \(F_i\).
+algorithms that operate with HD multivariate data; however, all these
+algorithms were originally designed to support a single task: to
+generate random vectors drawn from multivariate probability
+distributions with given marginal distributions and dependency metrics.
+Specifically, our goal is to efficiently simulate a large number, \(B\),
+of HD random vectors \({\bf Y}=(Y_1, \ldots, Y_d)^\top\) with
+\emph{correlated} components and heterogeneous marginal distributions,
+described via cumulative distribution functions (CDFs)
+\(F_i, i=1,\ldots,d\).
 
@@
   BP3: Flexible marginal modeling, generating heterogeneous data ---
-  possibly from differing probability families.
+  including mixed continuous and discrete margins.
 \end{itemize}
@@
 
-Simulating high-dimensional, non-normal, correlated data motivates this
-work --- in pursuit of modeling RNA-sequencing (RNA-seq) data
+Simulating HD, non-normal, correlated data motivates this work, in
+pursuit of modeling RNA-sequencing (RNA-seq) data
 \citep{Wang2009b, Conesa2016b} derived from breast cancer patients. The
 RNA-seq data-generating process involves counting how often a particular
-form of human messenger RNA (mRNA) is expressed in a biological sample.
+form of messenger RNA (mRNA) is expressed in a biological sample.
 RNA-seq platforms typically quantify the entire transcriptome in one
-experimental run, resulting in high-dimensional data. For human-derived
-samples, this results in count data corresponding to over 20,000 genes
-(protein-coding genomic regions) or even over 77,000 isoforms when
-alternatively-spliced mRNA are counted \citep{Schissler2019}.
-Importantly, due to inherent biological processes, gene expression data
-exhibit correlation (co-expression) across genes
-\citep{BE07, Schissler2018}.
-
-We illustrate our methodology using a well-studied Breast Invasive
-Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see
-Acknowledgments). For ease of modeling and simplicity of exposition, we
-only consider high expressing genes. In turn, we begin by filtering to
-retain the top 1000 of the highest-expressing genes (in terms of median
-expression) of the over 20,000 gene measurements from \(N=878\)
-patients' tumor samples. This gives a great number of pairwise
-dependencies among the marginals (specifically,
-\(\ensuremath{4.995\times 10^{5}}\) correlation parameters). Table
-@ref(tab:ch010-realDataTab) displays RNA-seq counts for three selected
-high-expressing genes for the first five patients' breast tumor samples.
-To help visualize the bivariate relationships for these three selected
-genes across all patients, Figure @ref(fig:ch010-realDataFig) displays
-the marginal distributions and estimated Spearman's correlations.
+experimental run, resulting in HD data. For human-derived samples, this
+results in count data corresponding to over 20,000 genes (protein-coding
+genomic regions) or even over 77,000 isoforms when alternatively-spliced
+mRNA are counted \citep{Schissler2019}. Importantly, due to inherent
+biological processes, gene expression data exhibit correlation
+(co-expression) across genes \citep{BE07, Schissler2018}.
+
+We illustrate our methodology using the Breast Invasive Carcinoma (BRCA)
+data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments).
+For ease of modeling and simplicity of exposition, we only consider high
+expressing genes. In turn, we begin by filtering to retain the top 1000
+of the highest-expressing genes (in terms of median expression) of the
+over 20,000 gene measurements from \(N=878\) patients' tumor samples.
+This gives a great number of pairwise dependencies among the marginals
+(specifically, \(\ensuremath{4.995\times 10^{5}}\) correlation
+parameters). Table @ref(tab:ch010-realDataTab) displays RNA-seq counts
+for three selected high-expressing genes for the first five patients'
+breast tumor samples. To help visualize the bivariate relationships for
+these three selected genes across all patients, Figure
+@ref(fig:ch010-realDataFig) displays the marginal distributions and
+estimated Spearman's correlations.
 
@@
 
-\caption{\label{tab:ch010-realDataTab}mRNA expression for three selected high-expressing genes, STAU1, FKBP1A, NME2, for the first five patients in the TCGA BRCA data set.}
+\caption{\label{tab:tab:ch010-realDataTab}mRNA expression for three selected high-expressing genes, STAU1, FKBP1A, NME2, for the first five patients in the TCGA BRCA data set.}
 \centering
@@
 
-\caption[Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes from Table 1]{Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes from Table 1.
+\caption[Pairwise scatterplots for three example genes (Table 1), with estimated Spearman's correlations and marginal density estimates displayed]{Pairwise scatterplots for three example genes (Table 1), with estimated Spearman's correlations and marginal density estimates displayed.
 The data possess outliers, heavy-right tails, are discrete, and have non-trivial intergene correlations.
@@
 In multivariate analysis, an analyst must select a metric to quantify
-dependency. The most widely-known is the Pearson (product-moment)
-correlation coefficient that describes the linear association between
-two random variables \(X\) and \(Y\), and, it is given by
+dependency. The most widely-known is the Pearson correlation coefficient
+that describes the linear association between two random variables \(X\)
+and \(Y\), and, it is given by
 
@@
 drawbacks and may mislead or fail to capture important relationships
-\citep{MK01}. Alternatively in these settings, analysts often prefer
-rank-based correlation measures to describe the degree of monotonic
-association.
+\citep{MK01}. Alternatively, analysts often prefer rank-based
+correlation measures to describe the degree of monotonic association.
 
@@
 \(\rho_S\) has an appealing correspondence as the Pearson correlation
-coefficient on \emph{ranks} of the values, thereby captures nonlinear
+coefficient on \emph{ranks} of the values, thereby capturing nonlinear
 yet monotone relationships. Kendall's \(\tau\), on the other hand, is
 the difference in probabilities of concordant and discordant pairs of
-observations, \((X_i, Y_i)\) and \((X_j, Y_j)\), with concordance
-meaning that orderings have the same direction (e.g., if \(X_i < X_j\),
-then \(Y_i < Y_j\)). Note that concordance is determined by the ranks of
-the values, not the values themselves.
+observations, \((X_i, Y_i)\) and \((X_j, Y_j)\) (concordance meaning
+that orderings have the same direction, e.g., if \(X_i < X_j\), then
+\(Y_i < Y_j\)). Note that concordance is determined by the ranks of the
+values, not the values themselves.
 
@@
 Given two marginal distributions, \(\rho_P\) is not free to vary over
-the entire range of possible correlations \([-1,1]\). The so-called
+the entire range of possible correlations \([-1,1]\). The
 \emph{Frechet-Hoeffding bounds} are well-studied
@@
 
-There is a strong connection of our simulation strategy to Gaussian
-\emph{copulas} (see \citet{Nelsen2007} for a technical introduction). A
-copula is a distribution function on \([0,1]^d\) that describes a
-multivariate probability distribution with standard uniform marginals.
-This provides a powerful, natural way to characterize joint probability
-structures. Consequently, the study of copulas is an important and
-active area of statistical theory and practice.
+There is a connection of our simulation strategy to Gaussian
+\emph{copulas} \citep{Nelsen2007}. A copula is a distribution function
+on \([0,1]^d\) that describes a multivariate probability distribution
+with standard uniform marginals. This provides a powerful, natural way
+to characterize joint probability. Consequently, the study of copulas is
+an important and active area of statistical theory and practice.
 
@@
 A Gaussian copula has marginal CDFs that are all standard normal,
-\(F_i = \Phi, \forall \, i\). This representation corresponds to a
-multivariate normal (MVN) distribution with standard normal marginal
-distributions and covariance matrix \({\bf R_P}\). As the marginals are
-standardized to have unit variance, however, \({\bf R_P}\) is a Pearson
-correlation matrix. If \(F_{{\bf R}}\) is the CDF of such a multivariate
-normal distribution, then the corresponding Gaussian copula
-\(C_{{\bf R}}\) is defined through
+\(F_i = \Phi, \forall \, i\), where \(\Phi(\cdot)\) is the standard
+normal CDF. This representation corresponds to a multivariate normal
+(MVN) distribution with standard normal marginal distributions and
+covariance matrix \({\bf R_P}\). Since the marginals are standardized to
+have unit variance, however, \({\bf R_P}\) is a Pearson correlation
+matrix. If \(F_{{\bf R}}\) is the CDF of such a multivariate normal
+distribution, then the corresponding Gaussian copula \(C_{{\bf R}}\) is
+defined through
 
@@
 
-where \(\Phi(\cdot)\) is the standard normal CDF. Note that the copula
-\(C_{{\bf R}}\) is the familiar multivariate normal CDF of the random
-vector \((\Phi(X_1), \ldots, \Phi(X_d))\), where
-\((X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})\).
+Here the copula function \(C_{{\bf R}}\) is the familiar multivariate
+normal CDF of the random vector \((\Phi(X_1), \ldots, \Phi(X_d))\),
+where \((X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})\).
 
@@
 
-\hypertarget{relevant-multivariate-simulation-software-comparison}{%
-\subsection{Relevant Multivariate Simulation software
-Comparison}\label{relevant-multivariate-simulation-software-comparison}}
-
-To place \texttt{Bigsimr}'s implementation and algorithms in context
-with existing approaches, as well as make this work's contribution more
-evident, we make some qualitative comparisons to other \texttt{R}
-multivariate simulation packages. The \texttt{MASS} packages
+\hypertarget{other-multivariate-simulation-r-packages}{%
+\subsection{Other Multivariate Simulation R
+Packages}\label{other-multivariate-simulation-r-packages}}
+
+To place \texttt{bigsimr}'s implementation and algorithms in context
+with existing approaches, here we qualitatively compare to other
+\texttt{R} multivariate simulation packages. The \texttt{MASS} packages
 \texttt{mvnorm()} has long been available to produce random vectors from
@@
 as fails to readily scale to high dimensions. Recently, the
-high-performance package \texttt{mvnfast} has been released that
-provides scaleable multivariate normal generation, but this doesn't
-completely satistify our goals of flexible margins and dependency
-modeling. For dependent discrete data, small-dimesional packages exist,
-such as \texttt{GenOrd}. The full-featured package \texttt{copula}
-provides many different copulas and a fairly general random generator
-\texttt{rCopula}. For low-dimensional modeling situtations, copula is
-suitable, yet the entent of exact matching of margins automatically is
-unclear. High-dimesional random vectors via \texttt{copula} takes an
-impractical amount of computing time \citep{Li2019gpu}. The
+high-performance package \texttt{mvnfast} \citep{Fasiolo2016} has been
+released that provides HD multivariate normal generation, but this lacks
+flexibility in marginal models and dependency modeling. For dependent
+discrete data, LD packages exist, such as \texttt{GenOrd}. For copula
+computation, the full-featured package \texttt{copula} provides many
+different copulas and a fairly general LD random generator
+\texttt{rCopula()}. High-dimesional random vectors via \texttt{copula}
+takes an impractical amount of computing time \citep{Li2019gpu}. The
 \texttt{nortaRA} R package matches Pearson coefficients near exactly and
-is reasonable for low-dimenional settings. In sum, we know of no other
-\texttt{R} package that can produce HD random vectors with flexible
-margins and dependence modeling. Table (\#tab:compare-table) summarizes
-the above discussion.
+is reasonable for LD settings. Finally, we know of no other \texttt{R}
+package that can produce HD random vectors with flexible margins and
+dependence modeling. Table (\#tab:compare-table) summarizes the above
+discussion.
 
@@
 \emph{high-performance Gaussian copula} (Equation @ref(eq:gauss))
-providing a high-dimensional NORTA-inspired algorithm.
+providing a HD NORTA-inspired algorithm.
 
@@
 \item
-  Simulate a random vector \(\bf Z\) with \(d\) \emph{independent and
-  identically distributed} (iid) standard normal components.
+  Simulate a random vector \(\bf Z\) with \(d\) independent and
+  identically distributed (iid) standard normal components.
 \item
@@
 as high-performance, multi-core and/or graphical-processing-unit (GPU)
-accelerated algorithms --- providing fast scalability using
-readily-available hardware.
+accelerated algorithms --- providing fast scalability to HD.
 
@@
 above), however, is computationally costly. In general, there is no
-closed-form correspondence between the components of the input
+convenient correspondence between the components of the input
 \(\Sigma_{\bf Z}\) and target \(\Sigma_{\bf Y}\). Matching the
@@
 where \(\phi(\cdot)\) is the joint probability distribution function of
-two correlated standard normal variables. For HD data, (nearly) exact
+two correlated standard normal variables. For HD simulation, exact
 evaluation may become too costly to enable practical simulation studies.
-For low-dimensional problems, however, methods and tools exist to match
-Pearson correlations precisely; see \citep{Xia17} and the publicly
-available \texttt{nortaRA} R package \citep{Chen2001}. As described
-later, to enable HD Pearson matching we approximate these integrals.
+Our remedy to enable HD Pearson matching in \texttt{bigsimr} is Hermite
+polynomial approximation of these \(\choose{d}{2}\) double integrals, as
+suggested by \citet{XZ19}.
 
@@
 distributions through copulas. Yet the choice of copula-based simulation
-algorithm affects which joint distributions may be simulated. Even in
-low-dimensional spaces (e.g., \(d=3\)), there exist valid multivariate
-distributions with \emph{feasible} Pearson correlation matrices that
-NORTA cannot match exactly \citep{LH75}. This occurs when the bivariate
+algorithm affects which joint distributions may be simulated. Even in LD
+spaces (e.g., \(d=3\)), there exist valid multivariate distributions
+with \emph{feasible} Pearson correlation matrices that NORTA cannot
+match exactly \citep{LH75}. This occurs when the bivariate
 transformations are applied to find the input correlation matrix, yet
@@
 approximate matching to the target --- with adequate performance for
-moderate \(d\). This provides evidence that our nearest positive
-definite (PD) augmented approach will maintain reasonable accuracy if
-our input matching scheme returns an indefinite matrix, at least for the
-rank-based matching scheme described above.
+moderate \(d\). This provides guidance that our nearest positive
+definite (PD) augmented approach has promise provide adequate accuracy
+if our input matching scheme returns an indefinite matrix.
 
 \hypertarget{rand-vec-gen}{%
-\subsection{Random vector generator}\label{rand-vec-gen}}
-
-We now describe our algorithm to generate random vectors, which mirrors
-the classical NORTA algorithm above with some modifications for
+\subsection{bigsimr rvec: Random vector generator}\label{rand-vec-gen}}
+
+We now describe our algorithm to generate random vectors, which extends
+NORTA to HD via nearest correlation matrix replacement and provides
 rank-based dependency matching:
@@
 \item
-  Mapping step
+  Mapping step, either
 
   \begin{itemize}
-  \item
-    \begin{enumerate}
-    \def\labelenumii{(\roman{enumii})}
   \tightlist
@@
     corresponding MVN Pearson correlation \(R_X\). Alternatively,
-    \end{enumerate}
-  \item
-    (i') Convert the target Kendall \(\tau\) matrix \(R_K\) to the
+  \item
+    Convert the target Kendall \(\tau\) matrix \(R_K\) to the
     corresponding MVN Pearson correlation \(R_X\). Alternatively,
   \item
-    (i'\,') Convert the target Pearson correlation matrix to \(R_P\) to
-    the corresponding, approximate MVN Pearson correlation \(R_X\).
+    Convert the target Pearson correlation matrix to \(R_P\) to the
+    corresponding, approximate MVN Pearson correlation \(R_X\).
   \end{itemize}
@@
   \begin{itemize}
-  \item
-    \begin{enumerate}
-    \def\labelenumii{(\roman{enumii})}
   \tightlist
@@
     matrix with 1's along the diagonal.
-    \end{enumerate}
-  \item
-    \begin{enumerate}
-    \def\labelenumii{(\roman{enumii})}
-    \setcounter{enumii}{1}
+  \item
+    If \(R_X\) is a correlation matrix, the input matrix is
+    \emph{admissible} in this scheme. Otherwise,
+  \item
+    Replace \(R_X\) with the nearest correlation matrix \(\tilde{R}_X\),
+    in the Frobenius norm.
+  \end{itemize}
+\item
+  Gaussian copula
+
+  \begin{itemize}
   \tightlist
   \item
-      If \(R_X\) is a correlation matrix, the input matrix is
-      \emph{admissible} in this scheme. Otherwise
-    \end{enumerate}
-  \item
-    (ii') Replace \(R_X\) with the nearest correlation matrix
-    \(\tilde{R}_X\), in the Frobenius norm.
-  \end{itemize}
-\item
-  Gaussian copula
-
-  \begin{itemize}
-  \item
-    \begin{enumerate}
-    \def\labelenumii{(\roman{enumii})}
-    \tightlist
-    \item
     Generate \({\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)\).
-    \end{enumerate}
-  \item
-    \begin{enumerate}
-    \def\labelenumii{(\roman{enumii})}
-    \setcounter{enumii}{1}
-    \tightlist
   \item
@@
     viz.~\(U_i=\Phi(X_i)\), \(i=1, \ldots, d\).
-    \end{enumerate}
-  \item
-    \begin{enumerate}
-    \def\labelenumii{(\roman{enumii})}
-    \setcounter{enumii}{2}
-    \tightlist
   \item
@@
     \(Y_i=F_i^{-1}(U_i)\), \(i=1, \ldots, d\).
-    \end{enumerate}
   \end{itemize}
@@
 
-\setstretch{2.0}
-
-\emph{Step 1}
-
-The first two descriptions of the \emph{Mapping step} employ the
-closed-form relationships between \(\rho_S\) and \(\tau\) with
-\(\rho_P\) for bivariate normal random variables via Equations
-@ref(eq:convertKendall) and @ref(eq:convertSpearman), respectively
-(implemented as \texttt{cor\_covert}). Initializing our algorithm to
-match the nonparametric correlations by computing these equations for
-all pairs is computationally trivial.
+\emph{Step 1: Mapping step.} We first employ the closed-form
+relationships between \(\rho_S\) and \(\tau\) with \(\rho_P\) for
+bivariate normal random variables via Equations @ref(eq:convertKendall)
+and @ref(eq:convertSpearman), respectively (implemented as
+\texttt{cor\_covert()}). Initializing our algorithm to match the
+nonparametric correlations by computing these equations for all pairs is
+computationally trivial.
 
@@
 this method, but for discrete distributions (especially discrete
-distributions with large support sets or infinite support), the problem
-is not so simple or efficient.
-
-Our solution is to approximate discrete distributions by a continuous
-distribution. The question then becomes that of which distribution to
-use. We found that Generalized S-Distributions \citep{muino2006gs} solve
-this problem by approximating a wide range of unimodal distributions,
-both continuous and discrete. Using the GS-Distribution approximation of
-discrete distributions in Pearson matching scheme above yields favorable
-results.
-
-\emph{Step 2}
-
-Once \(R_X\) has been determined, we check admissibility of the adjusted
-correlation via the steps described above. If \(R_X\) is not a valid
-correlation matrix, then we compute the nearest correlation matrix.
-Finding the nearest correlation matrix is a common statistical computing
-problem. The defacto default function in \texttt{R} is
+distributions with large support sets or infinite support), the
+approximation is computationally expensive. To remedy this, we further
+approximate discrete distributions by a continuous distribution using
+Generalized S-Distributions \citep{muino2006gs}.
+
+\emph{Step 2: Admissibility check and nearest correlation matrix
+computation.} Once \(R_X\) has been determined, we check admissibility
+of the adjusted correlation via the steps described above. If \(R_X\) is
+not a valid correlation matrix, then we compute the nearest correlation
+matrix. Finding the nearest correlation matrix is a common statistical
+computing problem. The defacto function in \texttt{R} is
 \texttt{Matrix::nearPD}, an alternating projection algorithm due to
 \citet{higham2002computing}. As implemented the function fails to scale
-to HD. Instead, we provide the quadratically convergent algorithm based
+to HD. Instead, we provide the quadratically-convergent algorithm based
 on the theory of strongly semi-smooth matrix functions
@@
 can be written down as the following convex optimization problem:
-
-\begin{align*}
-    \mathrm{min}\quad & \frac{1}{2} \Vert G - X \Vert^2 \\
-    \mathrm{s.t.}\quad & X_{ii} = 1, \quad i = 1, \ldots , n, \\
-    & X \in S_{+}^{n}
-\end{align*}
+\(\mathrm{min} \quad \frac{1}{2} \Vert G - X \Vert^2, \quad \mathrm{s.t.} \quad X_{ii} = 1, \quad i = 1, \ldots , n, \quad X \in S_{+}^{n}\).
 
 For nonparametric correlation measures, our algorithm allows the
-generation of high-dimensional multivariate data with arbitrary marginal
-distributions with a broad class of admissible Spearman correlation
-matrices and Kendall \(\tau\) matrices. The admissible classes consist
-of the matrices that map to a Pearson correlation matrix for a MVN. In
-particular, if we let \(X\) be MVN with \(d\) components and set
-
-\begin{align*}
-\Omega_P &= \{ R_P : R_P \textrm{ is a correlation matrix for } X \} \\
-\Omega_K &= \{ R_K : R_K \textrm{ is a correlation matrix for } X \} \\
-\Omega_S &= \{ R_S : R_S \textrm{ is a correlation matrix for } X \} \\
-\end{align*}
-
-then there are 1-1 mappings between these sets. We conjecture informally
-that the sets of admissible \(R_S\) and \(R_K\) are not highly
-restrictive. In particular, \(R_P\) is approximately \(R_S\) for a MVN,
-suggesting that the admissible set \(\Omega_S\) should be flexible as
-\(R_P\) can be any PD matrix with 1's along the diagonal. We provide
-methods to check whether a target \(R_S\) is an element of the
-\emph{admissible set} \(\Omega_S\) (and similarly for \(R_K\)).
-
-There is an increasing probability of encountering a non-admissible
+generation of HD multivariate data with arbitrary marginal distributions
+with a broad class of admissible Spearman correlation matrices and
+Kendall \(\tau\) matrices. The admissible classes consist of the
+matrices that map to a Pearson correlation matrix for a MVN. In
+particular, if we let \(X\) be MVN with \(d\) components and denote
+\(\Omega_P = \{ R_P : R_P \textrm{ is a correlation matrix for } X \}, \quad \Omega_K = \{ R_K : R_K \textrm{ is a correlation matrix for } X \}, \quad \Omega_S = \{ R_S : R_S \textrm{ is a correlation matrix for } X \}\).
+There are 1-1 mappings between these sets. We conjecture that the sets
+of admissible \(R_S\) and \(R_K\) are not highly restrictive. In
+particular, \(R_P\) is approximately \(R_S\) for a MVN, suggesting that
+the admissible set \(\Omega_S\) should be flexible as \(R_P\) can be any
+PD matrix with 1's along the diagonal. We provide methods to check
+whether a target \(R_S\) is an element of the \emph{admissible set}
+\(\Omega_S\).
+
+There is an increasing probability of encountering a inadmissible
 correlation matrix as dimension increases. In our experience, the
@@
 
-\emph{Step 3}
-
-The final step implements a NORTA-inspired, Gaussian copula approach to
-produce the desired margins. Steps 1 and 2 determine the MVN Pearson
-correlation values that will eventually match the target correlation.
-Then all that is required is a fast MVN simulator, a standard normal
-CDF, and well-defined quantile functions for marginals. The MVN is
-transformed to a copula (distribution with standard uniform margins) by
-applying the normal CDF \(\phi(\cdot)\). The quantile functions
-\(F^{-1}\) are applied across the margins to return the desired random
-vector \({\bf Y}\).
+\emph{Step 3: Gaussian copula.} The final step implements a
+NORTA-inspired, Gaussian copula approach to produce the desired margins.
+Steps 1 and 2 determine the MVN Pearson correlation values that will
+eventually match the target correlation. Step 3 requires a fast MVN
+simulator, a standard normal CDF, and well-defined quantile functions
+for marginals. The MVN is transformed to a copula (distribution with
+standard uniform margins) by applying the normal CDF \(\Phi(\cdot)\).
+Finally, the quantile functions \(F^{-1}\) are applied across the
+margins to return the desired random vector \({\bf Y}\).
 
@@
         [,1]    [,2]
-[1,]  1.000 -0.878
-[2,] -0.878  1.000
+[1,]  1.0000 -0.8815
+[2,] -0.8815  1.0000
 \end{CodeOutput}
@@
        [,1]   [,2]
-[1,] 1.000 0.879
-[2,] 0.879 1.000
+[1,] 1.0000 0.8803
+[2,] 0.8803 1.0000
 \end{CodeOutput}
@@
 Julia Object of type NamedTuple{(:lower, :upper),Tuple{Float64,Float64}}.
-(lower = -0.8801373451421908, upper = 0.8801357032571077)
+(lower = -0.8818561157816974, upper = 0.8818503162627974)
 \end{CodeOutput}
@@
 
-We have introduced a general-purpose high-dimensional multivariate
-simulation algorithm and provide a user-friendly, high-performance
-\texttt{R} package {[}\texttt{bigsimr}{]}. The random vector generation
-method is inspired by NORTA \citep{Cario1997} and Gaussian copula-based
-approaches \citep[\citet{BF17}, \citet{Xia17}]{MB13}. The major
-contributions of this work are methods and software for flexible,
-scalable simulation of HD multivariate probability distributions with
-broad potential data analytic applications for modern, big-data
-statistical computing. For example, one could simulate high-resolution
-time series data, such as those consistent with an auto-regressive
-moving average model exhibiting a specified Spearman structure. Or our
-methods could be used to simulate sparsely correlated data, as many HD
-methods assume, via specifying a \emph{spiked correlation matrix}.
-
-It is customary to compare new tools and algorithms directly to existing
-competing methods and software. In this study, however, we only employ
-our proposed methodology, as our previous work has shown that existing
-\texttt{R} tools are simply not designed to meet our high-dimensional
-goal (see \citet{Li2019gpu} for evaluations of the \texttt{R}
-\texttt{copula} package and others). For the bivariate simulations,
-existing packages such as
-\href{https://github.com/superdesolator/NORTARA}{\texttt{nortaRA}} work
-well to match Pearson correlations exactly.
+We introduced a general-purpose high-dimensional multivariate simulation
+algorithm and provide a user-friendly, high-performance \texttt{R}
+package {[}\texttt{bigsimr}{]}. The random vector generation method is
+inspired by NORTA \citep{Cario1997} and Gaussian copula-based approaches
+\citep[\citet{BF17}, \citet{Xia17}]{MB13}. The major contributions of
+this work are methods and software for flexible, scalable simulation of
+HD multivariate probability distributions with broad potential data
+analytic applications for modern, big-data statistical computing. For
+example, one could simulate high-resolution time series data, such as
+those consistent with an auto-regressive moving average model exhibiting
+a specified Spearman structure. Or our methods could be used to simulate
+sparsely correlated data, as many HD methods assume, via specifying a
+\emph{spiked correlation matrix}.
 

### 2021-06-01 030-bigsimr-package.Rmd
idk
Revise for JSS submission pre cross ref fix

--- 2021-03-01 030-bigsimr-package.Rmd
@@
-# The `bigsimr` R package {#package}
+# The \pkg{bigsimr} R package {short-title="bigsimr-pkg" #package}
 
@@
 
-This section describes a low-dimensional (2D) random vector simulation workflow via the `bigsimr` `R` package (https://github.com/SchisslerGroup/r-bigsimr).
-The package `bigsimr` provides an interface to the native code written in Julia, registered as the `Bigsimr` Julia package (https://github.com/adknudson/Bigsimr.jl).
-In addition to the native Julia `Bigsimr` package and `R` interface `bigsimr`, we also provide a python interface `bigsimr` (https://github.com/SchisslerGroup/python-bigsimr/) that interfaces with the Julia `Bigsimr` package.
+This section describes a LD random vector simulation workflow via the `bigsimr` `R` package.
+Our `R` package provides an interface to the native code written in Julia (registered as the `Bigsimr` Julia package.
+In addition to the native Julia `Bigsimr` package and `R` interface `bigsimr`, we also provide a python interface `bigsimr` that interfaces with the Julia `Bigsimr` package.
 The Julia package provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms)).
@@
 
-## Bivariate example description
+## Bivariate example
 
-We illustrate the use of `bigsimr` motivated by the New York air quality data set (`airquality`) included in the R `datasets` package.
+We illustrate the use of `bigsimr` using the New York air quality data set (`airquality`) included in the R `datasets` package.
 First, we load the `bigsimr` library and a few other convenient data science packages, including the syntactically-elegant `tidyverse` suite of `R` packages.
@@
 
-For simplicity and to provide a minimal working example, we consider bivariate simulation of the two `airquality` variables `Temperature`, in degrees Fahrenheit, and `Ozone` level, in parts per billion.
+Below, we describe a minimal working example --- a bivariate simulation of two airquality variables: `Temperature`, in degrees Fahrenheit, and `Ozone` level, in parts per billion.
 
@@
 
-Figure \@ref(fig:ch030-aq-joint-dist) visualizes the bivariate relationship between Ozone and Temperature.
+Figure \@ref(fig:ch030-aq-joint-dist) visualizes the bivariate relationship between `Ozone` and `Temperature`.
 We aim to simulate random two-component vectors mimicking this structure.
-The margins are not normally distributed; particularly the Ozone level exhibits a strong positive skew.
+The margins are not normally distributed; `Ozone` level exhibits a strong positive skew.
 
-```{r ch030-aq-joint-dist, cache=FALSE, fig.width= 8, fig.cap="Bivariate scatterplot of Ozone vs.
-Temp with estimated marginal densities.
-We model the Ozone data as marginally log-normal and the Temperature data as normal."}
+```{r ch030-aq-joint-dist, cache=FALSE, fig.width= 8, fig.cap="Bivariate scatterplot of Ozone versus Temp with estimated marginal densities.
+The Ozone data are modeled marginally as log-normal and the Temperature data as normal."}
 p0 <- ggplot(df, aes(Temp, Ozone)) +
@@
 For this example, we avoid goodness-of-fit considerations to determine the marginal distributions.
-But it seems sensible without domain knowledge to estimate these quantities from the data, and `bigsimr` contains fast functions designed for this task.
+It is sensible without domain knowledge to estimate these quantities from the data, and `bigsimr` contains fast functions designed for this task.
 
@@
 Based on the estimated densities in Figure \@ref(fig:ch030-aq-joint-dist), we assume `Temp` is normally distributed and `Ozone` is log-normally distributed, as the latter values are positive and skewed.
-We use the well-known, unbiased estimators for the normal distribution's parameters and maximum likelihood estimators for the log-normal parameters:
+We use the classical unbiased estimators for the normal distribution's parameters and maximum likelihood estimators for the log-normal parameters:
 
@@
 
-Next, we configure the input marginals for later input into `rvec`. The marginal distributions are specified using `Julia`'s `Distributions` package and stored in a vector.
+Now, we configure the input marginals for later input into `rvec`. The marginal distributions are specified using `Julia`'s `Distributions` package and stored in a vector.
 
@@
 For non-normal data and for improved simulation accuracy/scalability in our scheme, we advocate the use of Spearman's $\rho$ correlation matrix $R_S$ and Kendall's $\tau$ correlation matrix $R_K$.
-We also support Pearson correlation coefficient matching, while cautioning the user to check the performance for their parametric multivariate model (see [Monte Carlo evaluations](#simulations) below for evaluation strategies and guidance).
-These estimation methods are classical approaches, not designed for high-dimensional correlation estimation (see the [Conclusion and Discussion]({#discussion) sections for more on this).
+We also support Pearson correlation coefficient matching, while cautioning the user to check the performance for the distribution at hand (see [Monte Carlo evaluations](#simulations) below for evaluation strategies and guidance).
+Note that hese estimation methods are classical approaches, not specifically designed for high-dimensional correlation estimation.
 
@@
 
-First we use `cor_bounds` to ensure that the pairwise target correlation values are valid prior to mapping step which constructs $R_X$ for the MVN input into NORTA (see Section \@ref(algorithms)).
-`cor_bounds` estimates the pairwise lower and upper correlation bounds using the Generate, Sort, and Correlate algorithm of @DH2011.
-
-```{r ch030-cor-bounds, echo=TRUE}
-bounds <- bs$cor_bounds(margins)
-bounds$lower
-bounds$upper
-```
-
-Since our single estimated Spearman correlation is within the theoretical bounds, the correlation is valid as input to `rvec`.
-But even if the 2-dimensional bounds are satisfied for each pair of margins, this does not guarantee the feasibility of a $d-$variate distribution [@BF17].
-
-To provide higher dimensional feasibility/admissibility checking, we begin by mapping using `cor_convert` and check admissibility.
+Once a target correlation matrix is specified, first convert the dependency type to Pearson to correctly specify the MVN inputs.
+Then, check that the converted matrix $R_X$ is a valid correlation matrix (PD and 1s on the diagonal).
+For this bivariate example, $R_X$ is a valid correlation matrix.
+Typically in HD the resultant MVN Pearson correlation matrix is indefinite and requires approximation (see the HD examples in subsequent sections).
 
@@
 (R_X <- bs$cor_convert(R_S, bs$Spearman, bs$Pearson))
+```
+
+```{r ch030-cor-check, echo=TRUE}
 # Step 2. Check admissibility
@@
 
-The bounds on the Pearson correlation coefficient $R_P$ between these margins are restricted as seen below in our MC estimated correlation bounds:
+Despite being a valid correlation matrix under this definition, marginal distributions induce Frechet bounds on the possible Pearson correlation values.
+To check that the converted correlation values fall within these bounds, use `bigsimr::cor_bounds` to estimate the pairwise lower and upper correlation bounds.
+`cor_bounds` uses the Generate, Sort, and Correlate algorithm of @DH2011.
+For the assumed marginals, the Pearson correlation coefficient is not free to vary in [-1,1], as seen below.
+Our MC estimate of the bounds slightly underestimates the theoretic bounds of $(-0.881, 0.881)$.
+(See an analytic derivation presented in the Appendix).
+Since our single Pearson correlation coefficient is within the theoretical bounds, the correlation matrix is valid for our simulation strategy.
 
@@
 
-Our MC estimate of the bounds slightly underestimates the theoretic bounds of $(-0.881, 0.881)$. An analytic derivation is presented as an Appendix.
-
 ## Simulating random vectors
 
-Finally, we arrive at the main function of `bigsimr`: `rvec`. We now simulate $B=10,000$ random vectors from the assumed joint distribution of Ozone levels and Temp.
+A single `rvec` execution simulates the desired $10,000$ random vectors from the assumed bivariate distribution of `Ozone` and `Temp`.
+Note that the target input in specified using the pre-computed MVN correlation matrix.
 

### 2021-06-01 040-monte-carlo-evaluation.Rmd
Revise for JSS submission pre cross ref fix

@@
 
-Before applying our methodology to real data simulation, we conduct several Monte Carlo studies to investigate method performance.
-Since marginal parameter matching in our scheme is essentially a sequence of univariate inverse probability transforms, the challenging aspects are the accuracy of dependency matching and computational efficiency at high dimensions.
+In this section, we conduct MC studies to investigate method performance.
+Marginal parameter matching is straightforward in our scheme as it is a sequence of univariate inverse probability transforms.
+This process is both fast and exact.
+On the other hand, accurate and computationally efficent dependency matching presents a challenge, especially in the HD setting.
 To evaluate our methods in those respects, we design the following numerical experiments to first assess accuracy of matching dependency parameters in bivariate simulations and then time the procedure in increasingly large dimension $d$.
@@
 
-We select bivariate simulation configurations to ultimately simulate our motivating discrete-valued RNA-seq example, and, so we proceed in increasing complexity, leading to the model in our motivating application in Section \@ref(examples).
-We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson's, Spearman's, and Kendall's --- in identical, bivariate marginal configurations.
-For each pair of identical margins, we vary the target correlation across $\Omega$, the set of possible admissible values for correlation type, to evaluate the simulation's ability to obtain the theoretic bounds.
-The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (mimicking RNA-seq counts).
+We select bivariate simulation configurations to ultimately simulate our motivating discrete-valued RNA-seq example.
+Thus, we proceed in increasing departure from normality, leading to the multivariate negative binomial (MVNB) model for our motivating data.
+We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson, Spearman, and Kendall --- in identical, bivariate marginal configurations.
+For each pair of identical margins, we vary the target correlation across $\Omega$, the set of possible admissible values for each correlation type, to evaluate the simulation's ability to simulate all possible correlations.
+The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (non-normal and discrete).
 
 Table \@ref(tab:sims) lists our identical-marginal, bivariate simulation configurations.
-We increase the simulate replicates $B$ to check that our results converge to the target correlations and gauge statistical efficiency.
+We increase the simulate replicates $B$ to visually ensure that our results converge to the target correlations and gauge efficiency.
 We select distributions beginning with a standard multivariate normal (MVN) as we expect the performance to be exact (up to MC error) for all correlation types.
-Then, we select a non-symmetric continuous distribution: a standard (rate =1) two-component multivariate gamma (MVG).
-Finally, we select distributions and marginal parameter values that are motivated by our RNA-seq data, namely values proximal to probabilities and sizes estimated from the data (see [Example applications](examples) for estimation details).
-Thus we arrive at a multivariate negative binomial (MVNB) $p_1 = p_2 = 3\times10^{-4}, r_1 = r_2 = 4, \rho \in \Omega$.
+Then, we select a non-symmetric continuous distribution: a standard (rate =1), two-component multivariate gamma (MVG).
+Finally, we select distributions and marginal parameter values that are motivated by our RNA-seq data, namely values proximal to probabilities and sizes estimated from the data (see [RNA-seq data application](examples) for estimation details).
+Specifically, we assume a MVNB with $p_1 = p_2 = 3\times10^{-4}, r_1 = r_2 = 4, \rho \in \Omega$.
 
-Table: (\#tab:sims) Identical margin, bivariate simulation configurations to evaluate correlation matching accuracy and efficiency.
-
+<!-- 
+Table: (\#tab:sims) Identical margin, bivariate simulation configurations to evaluate matching accuracy.
 \begin{table}[]
-\begin{tabular}{@{}lcr@{}}
 \toprule
@@
 \end{table}
+-->
 
 For each of the unique 9 simulation configurations described above, we estimate the correlation bounds and vary the correlations along a sequence of 100 points evenly placed within the bounds, aiming to explore $\Omega$.
-Specifically, we set correlations $\{ \rho_1 = ( \hat{l} + \epsilon), \rho_2 = (\hat{l} + \epsilon) + \delta, \ldots, \rho_{100} = (\hat{u} - \epsilon) \}$, with $\hat{l}$ and $\hat{u}$ being the estimated lower and upper bounds, respectively, and increment value $\delta$.
+Specifically, we set correlations $\{ \rho_1 = ( \hat{l} + \epsilon), \rho_2 = (\hat{l} + \epsilon) + \delta, \ldots, \rho_{100} = (\hat{u} - \epsilon) \}$, with $\hat{l}$ and $\hat{u}$ being the estimated lower and upper bounds, respectively, with increment value $\delta$.
 The adjustment factor, $\epsilon=0.01$, is introduced to handle numeric issues when the bound is specified exactly.
@@
 
-Overall, the studies show that our methodology is generally accurate across the entire range of possible correlation values all three dependency measures, at least in these limited simulation settings for the rank-based correlations.
+Overall, the studies show that our methodology is generally accurate across the entire range of possible correlation for all three dependency measures.
 Our Pearson matching performs nearly as well as Spearman or Kendall, except for a slight increase in error for negative binomial case.
-This is due the particularly large counts generated with our choice of parameters for $p$ and $r$.
 
@@
 
-With information of our method's accuracy from a low-dimensional perspective, we now turn to assessing whether `bigsimr` can scale to larger dimensional problems with practical computation times.
-Specifically, we ultimately generate $B=1,000$ random vectors for $d=\{100, 250, 500, 1000, 2500, 5000, 10000\}$ for each correlation type, $\{Pearson, Spearman, Kendall\}$ while timing the algorithm's major steps.
+With information of our method's accuracy from a low-dimensional perspective, we now assess whether `bigsimr` can scale to larger dimensional problems with practical computation times.
+We ultimately generate $B=1,000$ random vectors for $d=\{100, 250, 500, 1000, 2500, 5000, 10000\}$ for each correlation type, {Pearson, Spearman, Kendall} while timing the algorithm's major steps.
 
-To mimic the workflow, we first produce a synthetic "data set" by completing the following steps:
+We first produce a synthetic "data set" by completing the following steps:
 
@@
 2. Produce a random full-rank Pearson correlation matrix via `cor_randPD` of size $d \times d$.
-3. Simulate a "data set" of $1,000 \times d$ random vectors via `rvec` (without matching the Pearson exactly).
+3. Simulate a "data set" of $1,000 \times d$ random vectors via `rvec`.
 \setstretch{2.0}
 
-With the synthetic data set in hand, we complete and time the following 4 steps:
+With the synthetic data set in hand, we complete and time the following four steps involved in a typical workflow (including correlation estimation).
 
 \setstretch{1.5}
-1. a) Estimate the correlation matrix from the "data" in the *Compute Correlation* step.
-1. b) Map the correlations to initialize the algorithm (Pearson to Pearson, Spearman to Pearson, or Kendall to Pearson) in the *Adjust Correlation* step.
-2. Check whether the mapping produces a valid correlation matrix and, if not, find the nearest PD correlation matrix in the *Check Admissibility* step.
-3. Simulate $1,000$ vectors via the Inverse Transform method in the *Simulate Data* step.
+1. Estimate the correlation matrix from the "data" in the *Compute Correlation* step.
+2. Map the correlations to initialize the algorithm (Pearson to Pearson, Spearman to Pearson, or Kendall to Pearson) in the *Adjust Correlation* step.
+3. Check whether the mapping produces a valid correlation matrix and, if not, find the nearest PD correlation matrix in the *Check Admissibility* step.
+4. Simulate $1,000$ vectors in the *Simulate Data* step.
 \setstretch{2.0}
@@
 The experiments are conducted on a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, with all 16 threads employed during computation.
-The results for $d \leq 500$ are fast with all times (except Kendall) under 3 seconds.
-Table \@ref(tab:ch040-moderateDtab) displays the total computation time of the algorithms steps 1 through 3, including estimation step i.
+Table \@ref(tab:ch040-moderateDtab) displays the total computation time for moderate dimesions ($d \leq 500$).
+In every simullation setting, the 1,000 vectors are generated rapidly, executing in under 2 seconds.
 
@@
                         "Total Time (Seconds)"),
-          caption = 'Total time to produce 10,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.')
+          caption = 'Total time to produce 1,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.')
 ```
 
-The results with larger dimensional vectors show scalability to ultra-high dimensions for all three correlation types, although the total times do become much larger. 
+The results for $d > 500$ show scalability to ultra-high dimensions for all three correlation types, although the total times do become much larger. 
 Figure \@ref(fig:ch040-largeDfig) displays computation times for $d=\{1000, 2500, 5000, 10000\}$.
@@
 Interestingly, Pearson is actually faster than Kendall for $d=10,000$ due to bottlenecks in *Compute Correlation* and *Check Admissibility*.
-Uniformly, matching Spearman correlations is faster, with total times under 5 minutes for $d=10,000$, making Spearman correlations the most computationally-friendly dependency type.
+Uniformly, matching Spearman correlations is faster, with total times under 5 minutes for $d=10,000$, making Spearman the most computationally-friendly dependency type for our package.
 With this in mind, we scaled the simulation to $d=20,000$ for the Spearman type and obtained the 1,000 vectors in under an hour (data not shown).
 In principle, this would enable the simulation of an entire human-derived RNA-seq data set.
-We note that for a given target correlation matrix and margins, steps i, 1, and 2 only need to be computed once and the third step, *Simulate Data*, is fast under all schemes for all $d$ under consideration.
+We note that for a given target correlation matrix and margins, steps 1, 2, and 3 only need to be computed once and the fourth step, *Simulate Data*, is nearly instaneous for all settings considered.
 
-```{r ch040-largeDfig, fig.cap="Computation times as d increases."}
+```{r ch040-largeDfig, fig.cap="Computation times for HD multivariate gamma simulation."}
 # Set colors
@@
 
-In the bivariate studies, we chose arbitrary simulation parameters for three distributions, moving from the Gaussian to the discrete and non-normal, multivariate negative binomial.
+In the bivariate studies, we chose arbitrary simulation parameters for three distributions, moving from the Gaussian to discrete and non-normal MVNB.
 Under these conditions, the simulated random vectors sample the desired bivariate distribution across the entire range of pairwise correlations for the three dependency measures.
@@
 This could mitigate some of the current algorithmic/implementation shortcomings.
-Despite this, Kendall matching is still feasible for most high-dimensional data sets.
-Finally, we note that one could use our single-pass algorithm `cor_fastPD` to produce a 'close' (not nearest PD) to scale to even higher dimensions with greater loss of accuracy.
+Despite this, Kendall matching is still feasible for most HD data sets.
+Finally, we note that one could use our single-pass algorithm `cor_fastPD` to produce a 'close' (not nearest PD) to scale to even higher dimensions with some loss of accuracy.

### 2021-06-01 050-examples.Rmd
Drafting comparison table in background v2
Revise for JSS submission pre cross ref fix

@@
-# RNA-seq data applications  {#examples}
+# RNA-seq data application  {#examples}
 
@@
 
-This section demonstrates how to simulate multivariate data using `bigsimr`, aiming to replicate the structure of high-dimensional dependent count data.
+This section demonstrates how to simulate multivariate data using `bigsimr`, aiming to replicate the structure of HD dependent count data.
 In an illustration of our proposed methodology, we seek to simulate RNA-sequencing data by producing simulated random vectors mimicking the observed data and its generating process.
-Modeling RNA-seq using multivariate probability distributions is natural as inter-gene correlation is an inherent part of biological processes [@Wang2009b].
+Modeling RNA-seq using multivariate probability distributions is natural as inter-gene correlation occurs during biological processes [@Wang2009b].
 And yet, many models do not account for this, leading to major disruptions to the operating characteristics of statistical estimation, testing, and prediction.
 See @Efron2012 for a detailed discussion with related methods and @Wu2012b, @Schissler2018; @Schissler2019 for applied examples.
-The following subsections apply `bigsimr`'s methods to real RNA-seq data, including replicating an estimated parametric structure, MC probability estimation, and MC evaluation of correlation estimation efficiency.
+The following subsections apply `bigsimr`'s methods to real RNA-seq data, including replicating an estimated parametric structure, probability estimation, and evaluation of correlation estimation efficiency.
 
@@
 
-We begin by estimating the structure of the TCGA BRCA RNA-seq data set (see [Background](background)).
+We begin by estimating the structure of the TCGA BRCA RNA-seq data set.
 Ultimately, we will simulate $B=10,000$ random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with $d=`r d`$.
-We assume a multivariate negative binomial (MVNB) model as RNA-seq counts are often over-dispersed and correlated.
+We assume a MVNB model as RNA-seq counts are often over-dispersed and correlated.
 Since all $d$ selected genes exhibit over-dispersion (data not shown), we proceed to estimate the NB parameters $(r_i, p_i), i=1,\ldots,d$, to determine the target marginal PMFs $f_i$.
-To complete specification of the simulation algorithm inputs, we estimate the Spearman correlation matrix ${ \bf R}_{S}$ to characterize dependency.
+To complete specification of the simulation algorithm inputs, we estimate the Spearman correlation matrix $R_S$ to characterize dependency.
 
@@
 Next, we estimate the marginal parameters.
-We use the method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model.
-While, marginal distributions are from the same probability family (NB), they are heterogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.
+We use method of moments to estimate the marginal parameters for the multivariate negative binomial model.
+The marginal distributions are from the same probability family (NB), yet they are heterogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.
 The functions below support this estimation for later use in `rvec`.
@@
 
-We apply these estimators to all `r d` genes across the `r nrow(brca1000)` patients:
+We apply these estimators to the highest-expressing `r d` genes across the `r nrow(brca1000)` patients:
 
@@
 Notably, the estimated marginal NB probabilities $\{ \hat{p}_i \}$ are small --- ranging in the interval $[`r min(probs)` , `r max(probs)`]$.
-This gives rise to highly variable counts and, typically, less restriction on potential pairwise correlation pairs.
-Once the functions are defined/executed to complete marginal estimation, we specify targets and generate the desired random vectors using `rvec`.
-Now we check admissibility of the specified correlation matrix.
+This gives rise to highly variable counts and, thus, less restriction on potential pairwise correlation pairs.
+Given the marginals, we now specify targets and check admissibility of the specified correlation matrix.
 
@@
 
-While the exact $d \times d$ Spearman correlation matrix is not strictly admissible in our scheme (as seen by the non-positive definite result above), the approximation is close with a maximum absolute error of `r max( abs(targets - approximates))` and average absolute error of `r mean(abs(targets - approximates))` across the `r choose(d,2)` correlations.
+As seen above $R_X$ is not strictly admissible in our scheme (as seen by the non-positive definite result above).
+However, the approximation is close with a maximum absolute error of `r max( abs(targets - approximates))` and average absolute error of `r mean(abs(targets - approximates))` across the `r choose(d,2)` correlations.
+With the inputs configured, `rvec` is executed to produce a synthetic RNA-seq data set:
 
@@
 sim_nbinom <- bs$rvec(10000, R_X, nb_margins) 
-colnames(sim_nbinom) <- colnames(brca1000)
 ```
@@
 Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parametric form and with less extreme points then the observed data in Figure \@ref(fig:ch010-realDataFig).
-Figure \@ref(fig:ch050-figBRCA) displays the aggregated results of our simulation by comparing the specified target parameter (horizontal axes) with the corresponding quantities estimated from the simulated data (vertical axes).
-The evaluation shows that the simulated counts approximately match the target parameters and exhibit the full range of estimated correlation from the data.
+Figure \@ref(fig:ch050-figBRCA) compares the specified target parameter (horizontal axis) with the corresponding quantities estimated from the simulated data (vertical axis).
+The evaluation shows that the simulated counts approximately match the target parameters and exhibit the full range of estimated correlations from the data.
 
@@
 
-The results show overall good simulation performance for our choice of parameters settings.
+The results show overall aquedate simulation performance for our choice of parameters settings.
 Our settings were motivated by modeling high-expressing genes from the TCGA BRCA data set.
@@
 where $\chi_i$ is the sample space for the $i^{th}$ component of the random vector $\bf{Y}$.
-Compact representations with convenient computational forms are rare for high-dimensional constructions, especially with heterogeneous, correlated marginal distributions (or margins of mixed data types).
+Compact representations with convenient computational forms are rare for high-dimensional constructions, especially with heterogeneous marginal distributions.
 Given a large number of simulated vectors as produced above, estimated probabilities are readily given by counting the proportion of simulated vectors meeting the desired condition.
-In our motivating application, one may ask what is the probability that all genes expressed greater than a certain threshold value ${\bf y}_0$.
+In our motivating application, one could ask: what is the probability that all genes are expressed greater than a certain threshold value ${\bf y}_0$?
 This can be estimated as
@@
 
-MC methods are routinely used in many statistical inferential tasks including estimation, hypothesis testing, error rates, and empirical interval coverage rates.
-To conclude the example applications, we demonstrate how `bigsimr` can be used to evaluate estimation efficiency.
+To conclude the example applications, we demonstrate how `bigsimr` can be used to evaluate correlation estimation efficiency.
 In particular, we wish to assess the error in our correlation estimation above.
 We used a conventional method, based on classical statistical theory which was not designed for high-dimensional data.
-Indeed, high-dimensional covariance estimation (and precision matrices) is an active area of statistical science, .e.g., [@Won2013g; @VanWieringen2016].
+Indeed, high-dimensional covariance estimation is an active area of statistical science; see, for example, @Won2013g or @VanWieringen2016.
 
 In this small example, we simulate $m=30$ data sets with the number of simulated vectors matching the number of patients in the BRCA data set, $N=`r nrow(example_brca)`$.
-Since our simulation is much faster for the Pearson correlation type (see Figure \@ref(fig:ch040-largeDfig)), we only convert the Spearman correlation matrix once (and ensure it is positive definite).
 At each iteration, we estimate the quadratic loss (residual sum of squared errors) from the specified ${\bf R}_{S}$, producing a distribution of loss values.
@@
 }, simplify = FALSE)
-# Evaluate the residual sum of squared error
-sapply(simRho, function(R) sum((R - R_S)^2))
+frobenius_loss <- sapply(simRho, function(R) norm( R - R_S, type = "F")  )
 ```
 
-```{r ch050-lossResults}
-frobenius_loss <- sapply(simRho, function(R) sum((R - R_S)^2))
-```
-
-The `R` summary function supplies the mean-augmented five-number summary of the quadratic loss distribution computed above.
+The `R` summary function supplies the mean-augmented five-number summary of the Frobenius loss distribution computed above.
 
@@
 
-This distribution could be compared to high-dimensional designed covariance estimators for guidance on whether the additional complexity and computation time are warranted.
+This distribution could be compared to bespoke HD covariance estimators for guidance on whether the additional complexity and computation time are warranted.

### 2021-06-01 060-conclusion-discussion.Rmd
Revise for JSS submission pre cross ref fix

@@
 
-We introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a user-friendly, high-performance `R` package [`bigsimr`].
+We developed a general-purpose high-dimensional multivariate simulation algorithm and provide a user-friendly, high-performance `R` package `bigsimr`, Julia package `Bigsimr` and Python package `bigsimr`.
 The random vector generation method is inspired by NORTA [@Cario1997] and Gaussian copula-based approaches [@MB13, @BF17, @Xia17].
-The major contributions of this work are methods and software for flexible, scalable simulation of HD multivariate probability distributions with broad potential data analytic applications for modern, big-data statistical computing.
+The major contributions of this work are methods and software for flexible, scalable simulation of HD multivariate probability distributions with broad data analytic applications for modern, big-data statistical computing.
 For example, one could simulate high-resolution time series data, such as those consistent with an auto-regressive moving average model exhibiting a specified Spearman structure.

### 2021-06-01 article_bigsimr.tex
Revise for JSS submission pre cross ref fix

@@
 
-This section describes a low-dimensional (2D) random vector simulation
-workflow via the \texttt{bigsimr} \texttt{R} package
-(\url{https://github.com/SchisslerGroup/r-bigsimr}). The package
-\texttt{bigsimr} provides an interface to the native code written in
-Julia, registered as the \texttt{Bigsimr} Julia package
-(\url{https://github.com/adknudson/Bigsimr.jl}). In addition to the
-native Julia \texttt{Bigsimr} package and \texttt{R} interface
-\texttt{bigsimr}, we also provide a python interface \texttt{bigsimr}
-(\url{https://github.com/SchisslerGroup/python-bigsimr/}) that
-interfaces with the Julia \texttt{Bigsimr} package. The Julia package
-provides a high-performance implementation of our proposed random vector
-generation algorithm and associated functions (see Section
-@ref(algorithms)).
+This section describes a LD random vector simulation workflow via the
+\texttt{bigsimr} \texttt{R} package. Our \texttt{R} package provides an
+interface to the native code written in Julia (registered as the
+\texttt{Bigsimr} Julia package. In addition to the native Julia
+\texttt{Bigsimr} package and \texttt{R} interface \texttt{bigsimr}, we
+also provide a python interface \texttt{bigsimr} that interfaces with
+the Julia \texttt{Bigsimr} package. The Julia package provides a
+high-performance implementation of our proposed random vector generation
+algorithm and associated functions (see Section @ref(algorithms)).
 
@@
 
-\hypertarget{bivariate-example-description}{%
-\subsection{Bivariate example
-description}\label{bivariate-example-description}}
-
-We illustrate the use of \texttt{bigsimr} motivated by the New York air
-quality data set (\texttt{airquality}) included in the R
-\texttt{datasets} package. First, we load the \texttt{bigsimr} library
-and a few other convenient data science packages, including the
-syntactically-elegant \texttt{tidyverse} suite of \texttt{R} packages.
-The code chunk below prepares the computing environment:
+\hypertarget{bivariate-example}{%
+\subsection{Bivariate example}\label{bivariate-example}}
+
+We illustrate the use of \texttt{bigsimr} using the New York air quality
+data set (\texttt{airquality}) included in the R \texttt{datasets}
+package. First, we load the \texttt{bigsimr} library and a few other
+convenient data science packages, including the syntactically-elegant
+\texttt{tidyverse} suite of \texttt{R} packages. The code chunk below
+prepares the computing environment:
 
@@
 
-For simplicity and to provide a minimal working example, we consider
-bivariate simulation of the two \texttt{airquality} variables
-\texttt{Temperature}, in degrees Fahrenheit, and \texttt{Ozone} level,
-in parts per billion.
+Below, we describe a minimal working example --- a bivariate simulation
+of two airquality variables: \texttt{Temperature}, in degrees
+Fahrenheit, and \texttt{Ozone} level, in parts per billion.
 
@@
 Figure @ref(fig:ch030-aq-joint-dist) visualizes the bivariate
-relationship between Ozone and Temperature. We aim to simulate random
-two-component vectors mimicking this structure. The margins are not
-normally distributed; particularly the Ozone level exhibits a strong
-positive skew.
+relationship between \texttt{Ozone} and \texttt{Temperature}. We aim to
+simulate random two-component vectors mimicking this structure. The
+margins are not normally distributed; \texttt{Ozone} level exhibits a
+strong positive skew.
 
@@
 
-\caption[Bivariate scatterplot of Ozone vs]{Bivariate scatterplot of Ozone vs.
-Temp with estimated marginal densities.
-We model the Ozone data as marginally log-normal and the Temperature data as normal.}\label{fig:ch030-aq-joint-dist}
+\caption[Bivariate scatterplot of Ozone versus Temp with estimated marginal densities]{Bivariate scatterplot of Ozone versus Temp with estimated marginal densities.
+The Ozone data are modeled marginally as log-normal and the Temperature data as normal.}\label{fig:ch030-aq-joint-dist}
 \end{figure}
@@
 this example, we avoid goodness-of-fit considerations to determine the
-marginal distributions. But it seems sensible without domain knowledge
-to estimate these quantities from the data, and \texttt{bigsimr}
-contains fast functions designed for this task.
+marginal distributions. It is sensible without domain knowledge to
+estimate these quantities from the data, and \texttt{bigsimr} contains
+fast functions designed for this task.
 
@@
 distributed and \texttt{Ozone} is log-normally distributed, as the
-latter values are positive and skewed. We use the well-known, unbiased
+latter values are positive and skewed. We use the classical unbiased
 estimators for the normal distribution's parameters and maximum
@@
 
-Next, we configure the input marginals for later input into
+Now, we configure the input marginals for later input into
 \texttt{rvec}. The marginal distributions are specified using
@@
 coefficient matching, while cautioning the user to check the performance
-for their parametric multivariate model (see
-\protect\hyperlink{simulations}{Monte Carlo evaluations} below for
-evaluation strategies and guidance). These estimation methods are
-classical approaches, not designed for high-dimensional correlation
-estimation (see the \href{\%7B\#discussion}{Conclusion and Discussion}
-sections for more on this).
+for the distribution at hand (see \protect\hyperlink{simulations}{Monte
+Carlo evaluations} below for evaluation strategies and guidance). Note
+that hese estimation methods are classical approaches, not specifically
+designed for high-dimensional correlation estimation.
 
@@
 
-First we use \texttt{cor\_bounds} to ensure that the pairwise target
-correlation values are valid prior to the mapping step which constructs
-\(R_X\) for the MVN input into NORTA (see Section @ref(algorithms)).
-\texttt{cor\_bounds} estimates the pairwise lower and upper correlation
-bounds using the Generate, Sort, and Correlate algorithm of
-\citet{DH2011}.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> bounds <- bs$cor_bounds(margins)
-R> bounds$lower
-\end{CodeInput}
-\begin{CodeOutput}
-        [,1]    [,2]
-[1,]  1.0000 -0.8815
-[2,] -0.8815  1.0000
-\end{CodeOutput}
-\begin{CodeInput}
-R> bounds$upper
-\end{CodeInput}
-\begin{CodeOutput}
-       [,1]   [,2]
-[1,] 1.0000 0.8803
-[2,] 0.8803 1.0000
-\end{CodeOutput}
-\end{CodeChunk}
-
-Since our single estimated Spearman correlation is within the
-theoretical bounds, the correlation is valid as input to \texttt{rvec}.
-But even if the 2-dimensional bounds are satisfied for each pair of
-margins, this does not guarantee the feasibility of a \(d-\)variate
-distribution \citep{BF17}.
-
-To provide higher dimensional feasibility/admissibility checking, we
-begin by mapping using \texttt{cor\_convert} and check admissibility.
+Once a target correlation matrix is specified, first convert the
+dependency type to Pearson to correctly specify the MVN inputs. Then,
+check that the converted matrix \(R_X\) is a valid correlation matrix
+(PD and 1s on the diagonal). For this bivariate example, \(R_X\) is a
+valid correlation matrix. Typically in HD the resultant MVN Pearson
+correlation matrix is indefinite and requires approximation (see the HD
+examples in subsequent sections).
 
@@
 \end{CodeOutput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
 \begin{CodeInput}
@@
 
-The bounds on the Pearson correlation coefficient \(R_P\) between these
-margins are restricted as seen below in our MC estimated correlation
-bounds:
+Despite being a valid correlation matrix under this definition, marginal
+distributions induce Frechet bounds on the possible Pearson correlation
+values. To check that the converted correlation values fall within these
+bounds, use \texttt{bigsimr::cor\_bounds} to estimate the pairwise lower
+and upper correlation bounds. \texttt{cor\_bounds} uses the Generate,
+Sort, and Correlate algorithm of \citet{DH2011}. For the assumed
+marginals, the Pearson correlation coefficient is not free to vary in
+{[}-1,1{]}, as seen below. Our MC estimate of the bounds slightly
+underestimates the theoretic bounds of \((-0.881, 0.881)\). (See an
+analytic derivation presented in the Appendix). Since our single Pearson
+correlation coefficient is within the theoretical bounds, the
+correlation matrix is valid for our simulation strategy.
 
@@
 
-Our MC estimate of the bounds slightly overestimates the theoretic
-bounds of \((-0.881, 0.881)\). An analytic derivation is presented as an
-Appendix.
-
 \hypertarget{simulating-random-vectors}{%
@@
 
-Finally, we arrive at the main function of \texttt{bigsimr}:
-\texttt{rvec}. We now simulate \(B=10,000\) random vectors from the
-assumed joint distribution of Ozone levels and Temp.
+A single \texttt{rvec} execution simulates the desired \(10,000\) random
+vectors from the assumed bivariate distribution of \texttt{Ozone} and
+\texttt{Temp}. Note that the target input in specified using the
+pre-computed MVN correlation matrix.
 
@@
 
-Before applying our methodology to real data simulation, we conduct
-several Monte Carlo studies to investigate method performance. Since
-marginal parameter matching in our scheme is essentially a sequence of
-univariate inverse probability transforms, the challenging aspects are
-the accuracy of dependency matching and computational efficiency at high
-dimensions. To evaluate our methods in those respects, we design the
-following numerical experiments to first assess accuracy of matching
-dependency parameters in bivariate simulations and then time the
-procedure in increasingly large dimension \(d\).
+In this section, we conduct MC studies to investigate method
+performance. Marginal parameter matching is straightforward in our
+scheme as it is a sequence of univariate inverse probability transforms.
+This process is both fast and exact. On the other hand, accurate and
+computationally efficent dependency matching presents a challenge,
+especially in the HD setting. To evaluate our methods in those respects,
+we design the following numerical experiments to first assess accuracy
+of matching dependency parameters in bivariate simulations and then time
+the procedure in increasingly large dimension \(d\).
 
@@
 We select bivariate simulation configurations to ultimately simulate our
-motivating discrete-valued RNA-seq example, and, so we proceed in
-increasing complexity, leading to the model in our motivating
-application in Section @ref(examples). We begin with empirically
-evaluating the dependency matching across all three supported
-correlations --- Pearson's, Spearman's, and Kendall's --- in identical,
-bivariate marginal configurations. For each pair of identical margins,
-we vary the target correlation across \(\Omega\), the set of possible
-admissible values for correlation type, to evaluate the simulation's
-ability to obtain the theoretic bounds. The simulations progress from
-bivariate normal, to bivariate gamma (non-normal yet continuous), and
-bivariate negative binomial (mimicking RNA-seq counts).
+motivating discrete-valued RNA-seq example. Thus, we proceed in
+increasing departure from normality, leading to the multivariate
+negative binomial (MVNB) model for our motivating data. We begin with
+empirically evaluating the dependency matching across all three
+supported correlations --- Pearson, Spearman, and Kendall --- in
+identical, bivariate marginal configurations. For each pair of identical
+margins, we vary the target correlation across \(\Omega\), the set of
+possible admissible values for each correlation type, to evaluate the
+simulation's ability to simulate all possible correlations. The
+simulations progress from bivariate normal, to bivariate gamma
+(non-normal yet continuous), and bivariate negative binomial (non-normal
+and discrete).
 
 Table @ref(tab:sims) lists our identical-marginal, bivariate simulation
-configurations. We increase the simulate replicates \(B\) to check that
-our results converge to the target correlations and gauge statistical
+configurations. We increase the simulate replicates \(B\) to visually
+ensure that our results converge to the target correlations and gauge
 efficiency. We select distributions beginning with a standard
@@
 to MC error) for all correlation types. Then, we select a non-symmetric
-continuous distribution: a standard (rate =1) two-component multivariate
-gamma (MVG). Finally, we select distributions and marginal parameter
-values that are motivated by our RNA-seq data, namely values proximal to
-probabilities and sizes estimated from the data (see
-\href{examples}{Example applications} for estimation details). Thus we
-arrive at a multivariate negative binomial (MVNB)
+continuous distribution: a standard (rate =1), two-component
+multivariate gamma (MVG). Finally, we select distributions and marginal
+parameter values that are motivated by our RNA-seq data, namely values
+proximal to probabilities and sizes estimated from the data (see
+\href{examples}{RNA-seq data application} for estimation details).
+Specifically, we assume a MVNB with
 \(p_1 = p_2 = 3\times10^{-4}, r_1 = r_2 = 4, \rho \in \Omega\).
-
-Table: (\#tab:sims) Identical margin, bivariate simulation
-configurations to evaluate correlation matching accuracy and efficiency.
-
-\begin{table}[]
-\begin{tabular}{@{}lcr@{}}
-\toprule
-Simulation Reps $B$ & Correlation Types & Identical-margin 2D distribution \\ \midrule
-$1,000$ & Pearson ($\rho_P$) & ${\bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho_i ), i=1,\ldots,100$ \\
-$10,000$ & Spearman ($\rho_S$) & ${\bf Y} \sim MVG( shape = 10, rate = 1, \rho_i ), i=1,\ldots,100$ \\
-$100,000$ & Kendall ($\tau$) & ${\bf Y} \sim MVNB(p = 3\times10^{-4}, r = 4,\rho_i), i=1,\ldots,100$ \\ \bottomrule
-\end{tabular}
-\end{table}
 
@@
 with \(\hat{l}\) and \(\hat{u}\) being the estimated lower and upper
-bounds, respectively, and increment value \(\delta\). The adjustment
+bounds, respectively, with increment value \(\delta\). The adjustment
 factor, \(\epsilon=0.01\), is introduced to handle numeric issues when
@@
 Overall, the studies show that our methodology is generally accurate
-across the entire range of possible correlation values all three
-dependency measures, at least in these limited simulation settings for
-the rank-based correlations. Our Pearson matching performs nearly as
-well as Spearman or Kendall, except for a slight increase in error for
-negative binomial case. This is due the particularly large counts
-generated with our choice of parameters for \(p\) and \(r\).
+across the entire range of possible correlation for all three dependency
+measures. Our Pearson matching performs nearly as well as Spearman or
+Kendall, except for a slight increase in error for negative binomial
+case.
 
@@
 With information of our method's accuracy from a low-dimensional
-perspective, we now turn to assessing whether \texttt{bigsimr} can scale
-to larger dimensional problems with practical computation times.
-Specifically, we ultimately generate \(B=1,000\) random vectors for
+perspective, we now assess whether \texttt{bigsimr} can scale to larger
+dimensional problems with practical computation times. We ultimately
+generate \(B=1,000\) random vectors for
 \(d=\{100, 250, 500, 1000, 2500, 5000, 10000\}\) for each correlation
-type, \(\{Pearson, Spearman, Kendall\}\) while timing the algorithm's
-major steps.
-
-To mimic the workflow, we first produce a synthetic ``data set'' by
-completing the following steps:
+type, \{Pearson, Spearman, Kendall\} while timing the algorithm's major
+steps.
+
+We first produce a synthetic ``data set'' by completing the following
+steps:
 
@@
   Simulate a ``data set'' of \(1,000 \times d\) random vectors via
-  \texttt{rvec} (without matching the Pearson exactly). \setstretch{2.0}
+  \texttt{rvec}. \setstretch{2.0}
 \end{enumerate}
@@
 With the synthetic data set in hand, we complete and time the following
-4 steps:
+four steps involved in a typical workflow (including correlation
+estimation).
 
@@
 \def\labelenumi{\arabic{enumi}.}
-\item
-  \begin{enumerate}
-  \def\labelenumii{\alph{enumii})}
 \tightlist
 \item
-    Estimate the correlation matrix from the ``data'' in the
-    \emph{Compute Correlation} step.
-  \end{enumerate}
-\item
-  \begin{enumerate}
-  \def\labelenumii{\alph{enumii})}
-  \setcounter{enumii}{1}
-  \tightlist
-  \item
-    Map the correlations to initialize the algorithm (Pearson to
-    Pearson, Spearman to Pearson, or Kendall to Pearson) in the
-    \emph{Adjust Correlation} step.
-  \end{enumerate}
+  Estimate the correlation matrix from the ``data'' in the \emph{Compute
+  Correlation} step.
+\item
+  Map the correlations to initialize the algorithm (Pearson to Pearson,
+  Spearman to Pearson, or Kendall to Pearson) in the \emph{Adjust
+  Correlation} step.
 \item
@@
 \item
-  Simulate \(1,000\) vectors via the Inverse Transform method in the
-  \emph{Simulate Data} step. \setstretch{2.0}
+  Simulate \(1,000\) vectors in the \emph{Simulate Data} step.
+  \setstretch{2.0}
 \end{enumerate}
@@
 Intel Core i9 processor, with all 16 threads employed during
-computation. The results for \(d \leq 500\) are fast with all times
-(except Kendall) under 3 seconds. Table @ref(tab:ch040-moderateDtab)
-displays the total computation time of the algorithms steps 1 through 3,
-including estimation step i.
+computation. Table @ref(tab:ch040-moderateDtab) displays the total
+computation time for moderate dimesions (\(d \leq 500\)). In every
+simullation setting, the 1,000 vectors are generated rapidly, executing
+in under 2 seconds.
 
@@
 
-\caption{\label{tab:tab:ch040-moderateDtab}Total time to produce 10,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.}
+\caption{\label{tab:tab:ch040-moderateDtab}Total time to produce 1,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.}
 \centering
@@
 
-The results with larger dimensional vectors show scalability to
-ultra-high dimensions for all three correlation types, although the
-total times do become much larger. Figure @ref(fig:ch040-largeDfig)
-displays computation times for \(d=\{1000, 2500, 5000, 10000\}\). For
-\(d\) equal to 1000 and 2500, the total time is under a couple of
-minutes. At \(d\) of 5000 and 10,000, Pearson correlation matching in
-the \emph{Adjust Correlation} step becomes costly. Interestingly,
-Pearson is actually faster than Kendall for \(d=10,000\) due to
-bottlenecks in \emph{Compute Correlation} and \emph{Check
-Admissibility}. Uniformly, matching Spearman correlations is faster,
-with total times under 5 minutes for \(d=10,000\), making Spearman
-correlations the most computationally-friendly dependency type. With
-this in mind, we scaled the simulation to \(d=20,000\) for the Spearman
-type and obtained the 1,000 vectors in under an hour (data not shown).
-In principle, this would enable the simulation of an entire
+The results for \(d > 500\) show scalability to ultra-high dimensions
+for all three correlation types, although the total times do become much
+larger. Figure @ref(fig:ch040-largeDfig) displays computation times for
+\(d=\{1000, 2500, 5000, 10000\}\). For \(d\) equal to 1000 and 2500, the
+total time is under a couple of minutes. At \(d\) of 5000 and 10,000,
+Pearson correlation matching in the \emph{Adjust Correlation} step
+becomes costly. Interestingly, Pearson is actually faster than Kendall
+for \(d=10,000\) due to bottlenecks in \emph{Compute Correlation} and
+\emph{Check Admissibility}. Uniformly, matching Spearman correlations is
+faster, with total times under 5 minutes for \(d=10,000\), making
+Spearman the most computationally-friendly dependency type for our
+package. With this in mind, we scaled the simulation to \(d=20,000\) for
+the Spearman type and obtained the 1,000 vectors in under an hour (data
+not shown). In principle, this would enable the simulation of an entire
 human-derived RNA-seq data set. We note that for a given target
-correlation matrix and margins, steps i, 1, and 2 only need to be
-computed once and the third step, \emph{Simulate Data}, is fast under
-all schemes for all \(d\) under consideration.
+correlation matrix and margins, steps 1, 2, and 3 only need to be
+computed once and the fourth step, \emph{Simulate Data}, is nearly
+instaneous for all settings considered.
 
@@
 
-\caption[Computation times as d increases]{Computation times as d increases.}\label{fig:ch040-largeDfig}
+\caption[Computation times for HD multivariate gamma simulation]{Computation times for HD multivariate gamma simulation.}\label{fig:ch040-largeDfig}
 \end{figure}
@@
 In the bivariate studies, we chose arbitrary simulation parameters for
-three distributions, moving from the Gaussian to the discrete and
-non-normal, multivariate negative binomial. Under these conditions, the
-simulated random vectors sample the desired bivariate distribution
-across the entire range of pairwise correlations for the three
-dependency measures. The simulation results could differ for other
-choices of simulation settings. Specifying extreme correlations near the
-boundary or Frechet bounds could result in poor simulation performance.
-Fortunately, it is straightforward to evaluate simulation performance by
-using strategies similar to those completed above. We expect our random
-vector generation to perform well for the vast majority of
-NORTA-feasible correlation matrices, but advise to check the performance
-before making inferences/further analyses.
+three distributions, moving from the Gaussian to discrete and non-normal
+MVNB. Under these conditions, the simulated random vectors sample the
+desired bivariate distribution across the entire range of pairwise
+correlations for the three dependency measures. The simulation results
+could differ for other choices of simulation settings. Specifying
+extreme correlations near the boundary or Frechet bounds could result in
+poor simulation performance. Fortunately, it is straightforward to
+evaluate simulation performance by using strategies similar to those
+completed above. We expect our random vector generation to perform well
+for the vast majority of NORTA-feasible correlation matrices, but advise
+to check the performance before making inferences/further analyses.
 
@@
 some of the current algorithmic/implementation shortcomings. Despite
-this, Kendall matching is still feasible for most high-dimensional data
-sets. Finally, we note that one could use our single-pass algorithm
+this, Kendall matching is still feasible for most HD data sets. Finally,
+we note that one could use our single-pass algorithm
 \texttt{cor\_fastPD} to produce a `close' (not nearest PD) to scale to
-even higher dimensions with greater loss of accuracy.
+even higher dimensions with some loss of accuracy.
 
@@
 This section demonstrates how to simulate multivariate data using
-\texttt{bigsimr}, aiming to replicate the structure of high-dimensional
-dependent count data. In an illustration of our proposed methodology, we
-seek to simulate RNA-sequencing data by producing simulated random
-vectors mimicking the observed data and its generating process. Modeling
-RNA-seq using multivariate probability distributions is natural as
-inter-gene correlation is an inherent part of biological processes
-\citep{Wang2009b}. And yet, many models do not account for this, leading
-to major disruptions to the operating characteristics of statistical
-estimation, testing, and prediction. See \citet{Efron2012} for a
-detailed discussion with related methods and \citet{Wu2012b},
-\citet{Schissler2018}; \citet{Schissler2019} for applied examples. The
-following subsections apply \texttt{bigsimr}'s methods to real RNA-seq
-data, including replicating an estimated parametric structure, MC
-probability estimation, and MC evaluation of correlation estimation
-efficiency.
+\texttt{bigsimr}, aiming to replicate the structure of HD dependent
+count data. In an illustration of our proposed methodology, we seek to
+simulate RNA-sequencing data by producing simulated random vectors
+mimicking the observed data and its generating process. Modeling RNA-seq
+using multivariate probability distributions is natural as inter-gene
+correlation occurs during biological processes \citep{Wang2009b}. And
+yet, many models do not account for this, leading to major disruptions
+to the operating characteristics of statistical estimation, testing, and
+prediction. See \citet{Efron2012} for a detailed discussion with related
+methods and \citet{Wu2012b}, \citet{Schissler2018};
+\citet{Schissler2019} for applied examples. The following subsections
+apply \texttt{bigsimr}'s methods to real RNA-seq data, including
+replicating an estimated parametric structure, probability estimation,
+and evaluation of correlation estimation efficiency.
 
@@
 
-We begin by estimating the structure of the TCGA BRCA RNA-seq data set
-(see \href{background}{Background}). Ultimately, we will simulate
-\(B=10,000\) random vectors \({\bf Y}=(Y_1, \ldots, Y_d)^\top\) with
-\(d=1000\). We assume a multivariate negative binomial (MVNB) model as
-RNA-seq counts are often over-dispersed and correlated. Since all \(d\)
-selected genes exhibit over-dispersion (data not shown), we proceed to
-estimate the NB parameters \((r_i, p_i), i=1,\ldots,d\), to determine
-the target marginal PMFs \(f_i\). To complete specification of the
-simulation algorithm inputs, we estimate the Spearman correlation matrix
-\({ \bf R}_{S}\) to characterize dependency.
+We begin by estimating the structure of the TCGA BRCA RNA-seq data set.
+Ultimately, we will simulate \(B=10,000\) random vectors
+\({\bf Y}=(Y_1, \ldots, Y_d)^\top\) with \(d=1000\). We assume a MVNB
+model as RNA-seq counts are often over-dispersed and correlated. Since
+all \(d\) selected genes exhibit over-dispersion (data not shown), we
+proceed to estimate the NB parameters \((r_i, p_i), i=1,\ldots,d\), to
+determine the target marginal PMFs \(f_i\). To complete specification of
+the simulation algorithm inputs, we estimate the Spearman correlation
+matrix \(R_S\) to characterize dependency.
 
@@
 
-Next, we estimate the marginal parameters. We use the method of moments
-(MoM) to estimate the marginal parameters for the multivariate negative
-binomial model. While, marginal distributions are from the same
-probability family (NB), they are heterogeneous in terms of the
-parameters probability and size \((p_i, n_i)\) for \(i,\ldots,d\). The
-functions below support this estimation for later use in \texttt{rvec}.
+Next, we estimate the marginal parameters. We use method of moments to
+estimate the marginal parameters for the multivariate negative binomial
+model. The marginal distributions are from the same probability family
+(NB), yet they are heterogeneous in terms of the parameters probability
+and size \((p_i, n_i)\) for \(i,\ldots,d\). The functions below support
+this estimation for later use in \texttt{rvec}.
 
@@
 
-We apply these estimators to all 1000 genes across the 878 patients:
+We apply these estimators to the highest-expressing 1000 genes across
+the 878 patients:
 
@@
 \([\ensuremath{7.5305\times 10^{-7}} , \ensuremath{5.6592\times 10^{-4}}]\).
-This gives rise to highly variable counts and, typically, less
-restriction on potential pairwise correlation pairs. Once the functions
-are defined/executed to complete marginal estimation, we specify targets
-and generate the desired random vectors using \texttt{rvec}. Now we
-check admissibility of the specified correlation matrix.
+This gives rise to highly variable counts and, thus, less restriction on
+potential pairwise correlation pairs. Given the marginals, we now
+specify targets and check admissibility of the specified correlation
+matrix.
 
@@
 
-While the exact \(d \times d\) Spearman correlation matrix is not
-strictly admissible in our scheme (as seen by the non-positive definite
-result above), the approximation is close with a maximum absolute error
-of 0.0069 and average absolute error of
-\ensuremath{5.3366\times 10^{-4}} across the
-\ensuremath{4.995\times 10^{5}} correlations.
+As seen above \(R_X\) is not strictly admissible in our scheme (as seen
+by the non-positive definite result above). However, the approximation
+is close with a maximum absolute error of 0.0069 and average absolute
+error of \ensuremath{5.3366\times 10^{-4}} across the
+\ensuremath{4.995\times 10^{5}} correlations. With the inputs
+configured, \texttt{rvec} is executed to produce a synthetic RNA-seq
+data set:
 
@@
 R> sim_nbinom <- bs$rvec(10000, R_X, nb_margins) 
-R> colnames(sim_nbinom) <- colnames(brca1000)
 \end{CodeInput}
@@
 and with less extreme points then the observed data in Figure
-@ref(fig:ch010-realDataFig). Figure @ref(fig:ch050-figBRCA) displays the
-aggregated results of our simulation by comparing the specified target
-parameter (horizontal axes) with the corresponding quantities estimated
-from the simulated data (vertical axes). The evaluation shows that the
-simulated counts approximately match the target parameters and exhibit
-the full range of estimated correlation from the data.
+@ref(fig:ch010-realDataFig). Figure @ref(fig:ch050-figBRCA) compares the
+specified target parameter (horizontal axis) with the corresponding
+quantities estimated from the simulated data (vertical axis). The
+evaluation shows that the simulated counts approximately match the
+target parameters and exhibit the full range of estimated correlations
+from the data.
 
@@
 
-The results show overall good simulation performance for our choice of
-parameters settings. Our settings were motivated by modeling
+The results show overall aquedate simulation performance for our choice
+of parameters settings. Our settings were motivated by modeling
 high-expressing genes from the TCGA BRCA data set. In general, the
@@
 computational forms are rare for high-dimensional constructions,
-especially with heterogeneous, correlated marginal distributions (or
-margins of mixed data types). Given a large number of simulated vectors
-as produced above, estimated probabilities are readily given by counting
-the proportion of simulated vectors meeting the desired condition. In
-our motivating application, one may ask what is the probability that all
-genes expressed greater than a certain threshold value \({\bf y}_0\).
-This can be estimated as
+especially with heterogeneous marginal distributions. Given a large
+number of simulated vectors as produced above, estimated probabilities
+are readily given by counting the proportion of simulated vectors
+meeting the desired condition. In our motivating application, one could
+ask: what is the probability that all genes are expressed greater than a
+certain threshold value \({\bf y}_0\)? This can be estimated as
 
@@
 
-MC methods are routinely used in many statistical inferential tasks
-including estimation, hypothesis testing, error rates, and empirical
-interval coverage rates. To conclude the example applications, we
-demonstrate how \texttt{bigsimr} can be used to evaluate estimation
+To conclude the example applications, we demonstrate how
+\texttt{bigsimr} can be used to evaluate correlation estimation
 efficiency. In particular, we wish to assess the error in our
@@
 classical statistical theory which was not designed for high-dimensional
-data. Indeed, high-dimensional covariance estimation (and precision
-matrices) is an active area of statistical science, .e.g.,
-\citep{Won2013g, VanWieringen2016}.
+data. Indeed, high-dimensional covariance estimation is an active area
+of statistical science; see, for example, \citet{Won2013g} or
+\citet{VanWieringen2016}.
 
@@
 simulated vectors matching the number of patients in the BRCA data set,
-\(N=878\). Since our simulation is much faster for the Pearson
-correlation type (see Figure @ref(fig:ch040-largeDfig)), we only convert
-the Spearman correlation matrix once (and ensure it is positive
-definite). At each iteration, we estimate the quadratic loss (residual
+\(N=878\). At each iteration, we estimate the quadratic loss (residual
 sum of squared errors) from the specified \({\bf R}_{S}\), producing a
@@
 + }, simplify = FALSE)
-R> # Evaluate the residual sum of squared error
-R> sapply(simRho, function(R) sum((R - R_S)^2))
+R> frobenius_loss <- sapply(simRho, function(R) norm( R - R_S, type = "F")  )
 \end{CodeInput}
@@
 The \texttt{R} summary function supplies the mean-augmented five-number
-summary of the quadratic loss distribution computed above.
+summary of the Frobenius loss distribution computed above.
 
@@
 
-This distribution could be compared to high-dimensional designed
-covariance estimators for guidance on whether the additional complexity
-and computation time are warranted.
+This distribution could be compared to bespoke HD covariance estimators
+for guidance on whether the additional complexity and computation time
+are warranted.
 
@@
 
-We introduced a general-purpose high-dimensional multivariate simulation
+We developed a general-purpose high-dimensional multivariate simulation
 algorithm and provide a user-friendly, high-performance \texttt{R}
-package {[}\texttt{bigsimr}{]}. The random vector generation method is
+package \texttt{bigsimr}, Julia package \texttt{Bigsimr} and Python
+package \texttt{bigsimr}. The random vector generation method is
 inspired by NORTA \citep{Cario1997} and Gaussian copula-based approaches
@@
 this work are methods and software for flexible, scalable simulation of
-HD multivariate probability distributions with broad potential data
-analytic applications for modern, big-data statistical computing. For
-example, one could simulate high-resolution time series data, such as
-those consistent with an auto-regressive moving average model exhibiting
-a specified Spearman structure. Or our methods could be used to simulate
+HD multivariate probability distributions with broad data analytic
+applications for modern, big-data statistical computing. For example,
+one could simulate high-resolution time series data, such as those
+consistent with an auto-regressive moving average model exhibiting a
+specified Spearman structure. Or our methods could be used to simulate
 sparsely correlated data, as many HD methods assume, via specifying a
@@
 Thus, the possible range of correlation becomes \((-0.881, 0.881)\).
-This approximately agrees with the MC results provided.
+This agrees with the MC results provided.
 

### 2021-06-02 010-background.Rmd
get bookdown cross refs working

@@
 
-```{r tab:ch010-realDataTab}
+```{r ch010-realDataTab}
 set.seed(2020-02-25)
@@
 Finally, we know of no other `R` package that can produce HD random vectors with flexible margins and dependence modeling.
-Table (\#tab:compare-table) summarizes the above discussion.
+Table \@ref(tab:compare-table) summarizes the above discussion.
 
@@
 \caption{Selected feature comparison of bigsimr to existing R packages for multivariate simulation.}
-\label{tab:compare-table}
+\label{compare-table}
 \begin{tabular}{@{}llll@{}}

### 2021-06-02 040-monte-carlo-evaluation.Rmd
get bookdown cross refs working

@@
 
-<!-- 
-Table: (\#tab:sims) Identical margin, bivariate simulation configurations to evaluate matching accuracy.
 \begin{table}[]
-\toprule
-Simulation Reps $B$ & Correlation Types & Identical-margin 2D distribution \\ \midrule
-$1,000$ & Pearson ($\rho_P$) & ${\bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho_i ), i=1,\ldots,100$ \\
-$10,000$ & Spearman ($\rho_S$) & ${\bf Y} \sim MVG( shape = 10, rate = 1, \rho_i ), i=1,\ldots,100$ \\
-$100,000$ & Kendall ($\tau$) & ${\bf Y} \sim MVNB(p = 3\times10^{-4}, r = 4,\rho_i), i=1,\ldots,100$ \\ \bottomrule
+\centering
+\caption{Selected feature comparison of `bigsimr` to other existing `R` multivariate simulation packages.}
+\label{compare-table}
+\begin{tabular}{llll}
+R Package & Scales to HD & Hetereogenous Margins & Flexible Dependence \\ \hline
+MASS      & No           & No                    & No                  \\
+mvnfast   & Yes          & No                    & No                  \\
+GenOrd    & No           & Discrete margins      & Yes                 \\
+copula    & No           & Yes                   & No                  \\
+nortaRA   & No           & Yes                   & No                  \\
+bigsimr   & Yes          & Yes                   & Yes                
 \end{tabular}
 \end{table}
--->
 

### 2021-06-02 050-examples.Rmd
get bookdown cross refs working

@@
 ```{r ch050-densityEvaluationECHO, echo=TRUE}
-<<ch050-densityEvaluation>>
 pHat
@@
 }, simplify = FALSE)
-frobenius_loss <- sapply(simRho, function(R) norm( R - R_S, type = "F")  )
 ```
 
-The `R` summary function supplies the mean-augmented five-number summary of the Frobenius loss distribution computed above.
+The `R` summary function supplies the mean-augmented five-number summary of the Frobenius loss from the target.
+This distribution could be compared to bespoke HD covariance estimators for guidance on whether the additional complexity and computation time are warranted.
 
 ```{r ch050-loss-summary, echo=TRUE}
+frobenius_loss <- sapply(simRho, function(R) norm( R - R_S, type = "F")  )
 summary(frobenius_loss)
 ```
-
-This distribution could be compared to bespoke HD covariance estimators for guidance on whether the additional complexity and computation time are warranted.

### 2021-06-02 article_bigsimr.tex
get bookdown cross refs working

@@
 
-\caption{\label{tab:tab:ch010-realDataTab}mRNA expression for three selected high-expressing genes, STAU1, FKBP1A, NME2, for the first five patients in the TCGA BRCA data set.}
+\caption{\label{tab:ch010-realDataTab}mRNA expression for three selected high-expressing genes, STAU1, FKBP1A, NME2, for the first five patients in the TCGA BRCA data set.}
 \centering
@@
 Julia Object of type NamedTuple{(:lower, :upper),Tuple{Float64,Float64}}.
-(lower = -0.8818561157816974, upper = 0.8818503162627974)
+(lower = -0.8807818345178882, upper = 0.8817164736117079)
 \end{CodeOutput}
@@
 \begin{CodeInput}
-R> d <- ncol(sim_nbinom)
-R> B <- nrow(sim_nbinom)
-R> y0 <- 1
-R> pHat <- mean(apply(sim_nbinom, 1, function(Y) {
-+   all(Y >= y0)
-+ }))
 R> pHat
@@
 + }, simplify = FALSE)
+\end{CodeInput}
+\end{CodeChunk}
+
+The \texttt{R} summary function supplies the mean-augmented five-number
+summary of the Frobenius loss from the target. This distribution could
+be compared to bespoke HD covariance estimators for guidance on whether
+the additional complexity and computation time are warranted.
+
+\begin{CodeChunk}
+\begin{CodeInput}
 R> frobenius_loss <- sapply(simRho, function(R) norm( R - R_S, type = "F")  )
-\end{CodeInput}
-\end{CodeChunk}
-
-The \texttt{R} summary function supplies the mean-augmented five-number
-summary of the Frobenius loss distribution computed above.
-
-\begin{CodeChunk}
-\begin{CodeInput}
 R> summary(frobenius_loss)
@@
    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-    675     813     974     998    1080    1851 
+   26.0    28.5    31.2    31.4    32.9    43.0 
 \end{CodeOutput}
 \end{CodeChunk}
-
-This distribution could be compared to bespoke HD covariance estimators
-for guidance on whether the additional complexity and computation time
-are warranted.
 

### 2021-06-02 index.Rmd
get bookdown cross refs working

@@
 biblio-style: jss
-link-citations: yes
+output:
+  bookdown::pdf_book:
+    base_format: rticles::jss_article
+    includes:
+      in_header: preamble.tex
 ---
 
-```{r include=FALSE}
-# automatically create a bib database for R packages
+```{r, setup, include=FALSE}
 knitr::write_bib(c(
@@
         digits = 4)
+
+get_rmd <- function(str_no = "") list.files(path = ".", pattern = str_no)
+
+knitr::opts_chunk$set(
+  # Text Results
+  eval = TRUE,
+  echo = FALSE,
+  results = 'markup',
+  collapse = TRUE,
+  message = FALSE,
+  warning = FALSE,
+  error = FALSE,
+  # Code Decoration
+  comment = NA,
+  background = '#F7F7F7',
+  # Cache
+  cache = TRUE,
+  # Plots
+  fig.show = "hold",
+  fig.width = 6.5,
+  fig.asp = 0.618,  # 1 / phi
+  out.width = "85%",
+  fig.align = 'center'
+)
+
+# options(dplyr.print_min = 6, dplyr.print_max = 6)
+
+seed_number <- 2021-03-23
+set.seed(seed = seed_number)
 ```
 
-\newpage
+\clearpage

### 2021-06-03 000-introduction.Rmd
get bookdown cross refs working
Final push before pre-JSS submit internal review

@@
 
+\setstretch{2.0}
+
 Massive high-dimensional (HD) data sets are now commonplace in many areas of scientific inquiry.
-As new methods are developed for data analysis, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of proposed methodology --- such as false positive rates, statistical power, interval coverage, and robustness --- often to compare to methods.
+As new methods are developed for data analysis, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of proposed methodology --- such as false positive rates, statistical power, interval coverage, and robustness.
 Further, efficient simulation empowers statistical computing strategies, such as the parametric bootstrap [@Chernick2008] to simulate from a hypothesized null model, providing inference in analytically challenging settings.
@@
 As many have noted, it can be vexing to simulate dependent, non-normal/discrete data, even for low-dimensional (LD) settings [@MB13; @XZ19].
-For continuous non-normal LD multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula approaches [@Nelsen2007] are well-studied, with flexible, robust software available [@Yan2007; @Chen2001].
+For continuous non-normal LD multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula approaches [@Nelsen2007] are well-studied, with existing software available [@Yan2007; @Chen2001].
 Yet these approaches do not scale in a timely fashion to HD problems [@Li2019gpu].
 For discrete data, early simulation strategies had major flaws, such as failing to obtain the full range of possible correlations (e.g., admitting only positive correlations: see @Park1996).
-While more recent approaches [@MB13; @Xia17; @BF17] have largely remedied this issue for LD problems, the existing tools are not designed to scale to high dimensions for discrete marginal distributions.
+While more recent approaches [@MB13; @Xia17; @BF17] have remedied this issue for LD problems, the existing tools are not designed to scale to high dimensions.
 
@@
 Throughout, we focus on matching these nonparametric dependence measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology.
-We do, however, implement Pearson matching, but several layers of approximation is required.
+We do, however, implement Pearson matching, but several layers of approximation are required.
 
@@
 As we will describe in more detail, the algorithm's design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed-form relationships between dependence measures for the multivariate normal (MVN) distribution.
-For our method, we developed a high-performance implementation: the `bigsimr` R package, written in native Julia code.
+For our method, we developed a high-performance implementation: the `Bigsimr` Julia package, with R and Python interfaces `bigsimr`.
 
-Our study proceeds by providing background information, including a description of a motivating example application: RNA-sequencing (RNA-seq) breast cancer data.
+This article proceeds by providing background information, including a description of a motivating example application: RNA-sequencing (RNA-seq) breast cancer data.
 Then we describe and justify our simulation methodology and related algorithms.

### 2021-06-03 010-background.Rmd
Final push before pre-JSS submit internal review

@@
 
-The `bigsimr` `R` package presented here provides multiple algorithms that operate with HD multivariate data; however, all these algorithms were originally designed to support a single task: to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics.
+The `Bigsimr` package presented here provides multiple high-performance algorithms that operate with HD multivariate data.
+All these algorithms were originally designed to support a single task: to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics.
 Specifically, our goal is to efficiently simulate a large number, $B$, of HD random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with *correlated* components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i, i=1,\ldots,d$.
@@
 When designing this methodology, we developed the following properties to guide our effort.
-We divide the properties into two categories: (1) basic properties (BP) and "scalability" properties (SP).
+We divide the properties into two categories: (1) basic properties (BP) and 'scalability' properties (SP).
 The BPs are adapted from an existing criteria due to @Nik13a.
@@
 * BP1: A wide range of dependences, allowing both positive and negative values, and, ideally, admitting the full range of possible values.
-* BP2: Flexible dependence, meaning that the number of bivariate marginals can be equal to the number of dependence parameters.
-* BP3: Flexible marginal modeling, generating heterogeneous data --- including mixed continuous and discrete margins.
+* BP2: Flexible dependence, meaning that the number of dependence parameters can be equal to the number of bivariate pairs in the vector.
+* BP3: Flexible marginal modeling, generating heterogeneous data --- including mixed continuous and discrete marginal distributions.
 
@@
 Note that concordance is determined by the ranks of the values, not the values themselves.
-
 Both $\tau$ and $\rho_S$ are *invariant under monotone transformations* of the underlying random variates.
@@
 Consequently, the study of copulas is an important and active area of statistical theory and practice.
-
 For any random vector ${\bf X}=(X_1, \ldots, X_d)^\top$ with CDF $F$ and marginal CDFs $F_i$ there is a copula function $C(u_1, \ldots, u_d)$ satisfying
@@
 \begin{equation}
-F(x_1, \ldots, x_d) = {\mathbb P}(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), 
+F(x_1, \ldots, x_d) = {\mathbb P}(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), \quad x_i \in {\mathbb R}, i=1,\ldots,d.
 \label{eq:copula}
@@
 
-$x_i \in {\mathbb R}, i=1,\ldots,d.$
-
 A Gaussian copula has marginal CDFs that are all standard normal, $F_i = \Phi, \forall \, i$, where $\Phi(\cdot)$ is the standard normal CDF.
-This representation corresponds to a multivariate normal (MVN) distribution with standard normal marginal distributions and covariance matrix ${\bf R_P}$.
-Since the marginals are standardized to have unit variance, however, ${\bf R_P}$ is a Pearson correlation matrix.
-If $F_{{\bf R}}$ is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula $C_{{\bf R}}$ is defined through
+This representation corresponds to a multivariate normal (MVN) distribution with standard normal marginal distributions and covariance matrix $R_P$.
+Since the marginals are standardized to have unit variance, however, $R_P$ is a Pearson correlation matrix.
+If $F_R$ is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula $C_R$ is defined through
 
@@
 \label{eq:gauss}
-F_{{\bf R}}(x_1, \ldots, x_d) = C_{{\bf R}}(\Phi(x_1), \ldots, \Phi(x_d)),
+F_R(x_1, \ldots, x_d) = C_R(\Phi(x_1), \ldots, \Phi(x_d)),
 \end{equation}
 
-Here the copula function $C_{{\bf R}}$ is the familiar multivariate normal CDF of the random vector $(\Phi(X_1), \ldots, \Phi(X_d))$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})$.
+Here the copula function $C_R$ is the familiar multivariate normal CDF of the random vector $(\Phi(X_1), \ldots, \Phi(X_d))$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_P)$.
 
@@
 For example, using Gaussian copulas, we can construct a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)^\top$ with $Y_i \sim F_i$, viz.
-$Y_i = F_i^{-1}(\Phi(X_i)), i=1, \ldots, d$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})$.
+$Y_i = F_i^{-1}(\Phi(X_i)), i=1, \ldots, d$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_P)$.
 
-## Other Multivariate Simulation R Packages
+## Other Multivariate Simulation Packages
 
-To place `bigsimr`'s implementation and algorithms in context with existing approaches, here we qualitatively compare to other `R` multivariate simulation packages.
-The `MASS` packages `mvnorm()` has long been available to produce random vectors from multivariate normal (and $t$) distributions.
-This familiar procedure cannot simulate from arbitrary margins and dependency measures, as well as fails to readily scale to high dimensions.
-Recently, the high-performance package `mvnfast` [@Fasiolo2016] has been released that provides HD multivariate normal generation, but this lacks flexibility in marginal models and dependency modeling.
-For dependent discrete data, LD packages exist, such as `GenOrd`.
-For copula computation, the full-featured package `copula` provides many different copulas and a fairly general LD random generator `rCopula()`.
+To place `Bigsimr`'s implementation and algorithms in context with existing approaches, here we qualitatively compare to other multivariate simulation packages.
+In `R`, the `MASS` packages `mvnorm()` has long been available to produce random vectors from multivariate normal (and $t$) distributions.
+This procedure cannot simulate from arbitrary margins and dependency measures, as well as fails to readily scale to high dimensions.
+SAS PROC SIMNORMAL and Python's NumPy `random.multivariate_normal` both do similar tasks.
+Recently, the high-performance `R` package `mvnfast` [@Fasiolo2016] has been released that provides HD multivariate normal generation, but this lacks flexibility in marginal models and dependency modeling.
+For dependent discrete data, LD `R` packages exist, such as `GenOrd`.
+For copula computation, the full-featured `R` package `copula` provides many different copulas and a fairly general LD random generator `rCopula()`.
 High-dimesional random vectors via `copula` takes an impractical amount of computing time [@Li2019gpu].
@@
 
-\begin{table}[]
+\begin{table}[h]
 \centering
-\caption{Selected feature comparison of bigsimr to existing R packages for multivariate simulation.}
-\label{compare-table}
+\caption{\label{tab:compare-table}Comparison of selected features of \pkg{Bigsimr} to existing packages for multivariate simulation.}
 \begin{tabular}{@{}llll@{}}
-R Package & Scales to HD & Hetereogenous Margins & Flexible Dependence \\ \midrule
+Package & Scales to HD & Hetereogenous Margins & Flexible Dependence \\ \midrule
 MASS      & No           & No                    & No                  \\
@@
 \end{table}
+
+\clearpage

### 2021-06-03 020-simulation-algorithm.Rmd
Final push before pre-JSS submit internal review

@@
 For HD simulation, exact evaluation may become too costly to enable practical simulation studies.
-Our remedy to enable HD Pearson matching in `bigsimr` is Hermite polynomial approximation of these $\choose{d}{2}$ double integrals, as suggested by @XZ19.
+Our remedy to enable HD Pearson matching in `bigsimr` is Hermite polynomial approximation of these $\binom{d}{2}$ double integrals, as suggested by @XZ19.
 
@@
 Sklar's theorem provides a useful characterization of multivariate distributions through copulas.
-Yet the choice of copula-based simulation algorithm affects which joint distributions may be simulated.
+In practice, however, the particulars of the simulation algorithm affect which joint distributions may be simulated.
 Even in LD spaces (e.g., $d=3$), there exist valid multivariate distributions with *feasible* Pearson correlation matrices that NORTA cannot match exactly [@LH75].
@@
 
-@GH02 conducted a Monte Carlo study to estimate the probability of encountering NORTA defective matrices while increasing the dimension $d$.
+@GH02 conducted a study to estimate the probability of encountering NORTA defective matrices while increasing the dimension $d$.
 They found that for what is now considered low-to-moderate dimensions ($d \approx 20$), almost *all* feasible matrices are NORTA defective.
@@
 There is hope, however, as @GH02 also showed that replacing an indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$.
-This provides guidance that our nearest positive definite (PD) augmented approach has promise provide adequate accuracy if our input matching scheme returns an indefinite matrix.
+This provides insight that our nearest positive definite (PD) augmented approach has promise to provide adequate accuracy if our input matching scheme returns an indefinite Pearson correlation matrix.
 
-## bigsimr rvec: Random vector generator {#rand-vec-gen}
+## Random vector generator {#rand-vec-gen}
 
@@
 
-There is an increasing probability of encountering a inadmissible correlation matrix as dimension increases.
+There is an increasing probability of encountering an inadmissible correlation matrix as dimension increases.
 In our experience, the mapping step for large $d$ almost always produces a $R_X$ that is not a correlation matrix.
@@
 The MVN is transformed to a copula (distribution with standard uniform margins) by applying the normal CDF $\Phi(\cdot)$.
-Finally, the quantile functions $F^{-1}$ are applied across the margins to return the desired random vector ${\bf Y}$.
+Finally, the quantile functions $F_i^{-1}$ are applied across the margins to return the desired random vector ${\bf Y}$.

### 2021-06-03 030-bigsimr-package.Rmd
Final push before pre-JSS submit internal review

@@
-# The \pkg{bigsimr} R package {short-title="bigsimr-pkg" #package}
+# The \pkg{Bigsimr} package {short-title="Bigsimr-pkg" #package}
 
@@
 
-This section describes a LD random vector simulation workflow via the `bigsimr` `R` package.
-Our `R` package provides an interface to the native code written in Julia (registered as the `Bigsimr` Julia package.
-In addition to the native Julia `Bigsimr` package and `R` interface `bigsimr`, we also provide a python interface `bigsimr` that interfaces with the Julia `Bigsimr` package.
+This section describes a bivariate random vector simulation workflow via the `bigsimr` `R` package.
+This `R` package provides an interface to the native code written in Julia (registered as the `Bigsimr` Julia package).
+In addition to the native Julia `Bigsimr` package and `R` interface `bigsimr`, we also provide a Python interface `bigsimr` that interfaces with the Julia `Bigsimr` package.
 The Julia package provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms)).
 
-The subsections below describe the basic use of `bigsimr` by stepping through an example workflow using the data set `airquality` that contains daily air quality measurements in New York, May to September 1973 [@Chambers1983].
+The subsections below describe the basic use of `bigsimr` R package by stepping through an example workflow using the data set `airquality` that contains daily air quality measurements in New York, May to September 1973 [@Chambers1983].
 This workflow proceeds from setting up the computing environment, to data wrangling, estimation, simulation configuration, random vector generation, and, finally, result visualization.
@@
 
-Below, we describe a minimal working example --- a bivariate simulation of two airquality variables: `Temperature`, in degrees Fahrenheit, and `Ozone` level, in parts per billion.
+Here, we describe a minimal working example --- a bivariate simulation of two airquality variables: `Temperature`, in degrees Fahrenheit, and `Ozone` level, in parts per billion.
 
@@
 df <- airquality %>% select(Temp, Ozone) %>% drop_na()
-```
-
-```{r ch030-aq-glimpse}
 glimpse(df)
@@
 
-As mentioned, the user must decide how to describe correlation based on the particulars of the problem.
-For non-normal data and for improved simulation accuracy/scalability in our scheme, we advocate the use of Spearman's $\rho$ correlation matrix $R_S$ and Kendall's $\tau$ correlation matrix $R_K$.
+The user must decide how to describe correlation based on the particulars of the problem.
+For non-normal data in our scheme, we advocate the use of Spearman's $\rho$ correlation matrix $R_S$ or Kendall's $\tau$ correlation matrix $R_K$.
 We also support Pearson correlation coefficient matching, while cautioning the user to check the performance for the distribution at hand (see [Monte Carlo evaluations](#simulations) below for evaluation strategies and guidance).
-Note that hese estimation methods are classical approaches, not specifically designed for high-dimensional correlation estimation.
+Note that these estimation methods are classical approaches, not specifically designed for high-dimensional correlation estimation.
 
@@
 For this bivariate example, $R_X$ is a valid correlation matrix.
-Typically in HD the resultant MVN Pearson correlation matrix is indefinite and requires approximation (see the HD examples in subsequent sections).
+Note that typically in HD the resultant MVN Pearson correlation matrix is indefinite and requires approximation (see the HD examples in subsequent sections).
 
@@
 
-A single `rvec` execution simulates the desired $10,000$ random vectors from the assumed bivariate distribution of `Ozone` and `Temp`.
+Finally, we execute `rvec` to simulate the desired $10,000$ random vectors from the assumed bivariate distribution of `Ozone` and `Temp`.
 Note that the target input in specified using the pre-computed MVN correlation matrix.
+Figure \@ref(fig:ch030-plot-sim) plots the 10,000 simulated points.
 
@@
 ```
-
-Figure \@ref(fig:ch030-plot-sim) plots the 10,000 simulated points.
 
@@
 ```
+
+\clearpage

### 2021-06-03 040-monte-carlo-evaluation.Rmd
Final push before pre-JSS submit internal review

--- 2021-06-01 040-monte-carlo-evaluation.Rmd
@@
 We select bivariate simulation configurations to ultimately simulate our motivating discrete-valued RNA-seq example.
-Thus, we proceed in increasing departure from normality, leading to the multivariate negative binomial (MVNB) model for our motivating data.
+Thus, we proceed in increasing departure from normality, leading to a multivariate negative binomial (MVNB) model for our motivating data.
 We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson, Spearman, and Kendall --- in identical, bivariate marginal configurations.
@@
 
-<!-- 
-Table: (\#tab:sims) Identical margin, bivariate simulation configurations to evaluate matching accuracy.
 \begin{table}[]
+\centering
+\caption{ \label{tab:sims} Identical margin, bivariate simulation configurations to evaluate matching accuracy.}
+\begin{tabular}{@{}lcr@{}}
 \toprule
@@
 \end{table}
--->
 
@@
 Table \@ref(tab:ch040-BiError) contains the mean absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.
+Overall, the studies show that our methodology is generally accurate across the entire range of possible correlation for all three dependency measures.
+Our Pearson matching performs nearly as well as Spearman or Kendall, except for a slight increase in error for negative binomial case.
 
@@
 ```
-
-Overall, the studies show that our methodology is generally accurate across the entire range of possible correlation for all three dependency measures.
-Our Pearson matching performs nearly as well as Spearman or Kendall, except for a slight increase in error for negative binomial case.
 
@@
 We ultimately generate $B=1,000$ random vectors for $d=\{100, 250, 500, 1000, 2500, 5000, 10000\}$ for each correlation type, {Pearson, Spearman, Kendall} while timing the algorithm's major steps.
-
-We first produce a synthetic "data set" by completing the following steps:
+We begin by producing a synthetic "data set" by completing the following steps:
 
@@
 
-```{r tab:ch040-moderateDtab}
+```{r ch040-moderateDtab}
 dat_long <- benchmark_dependences %>%

### 2021-06-03 050-examples.Rmd
Final push before pre-JSS submit internal review

@@
 To complete specification of the simulation algorithm inputs, we estimate the Spearman correlation matrix $R_S$ to characterize dependency.
-
 With this goal in mind, we first estimate the desired correlation matrix using the fast implementation provided by `bigsimr`:
@@
 In this small example, we simulate $m=30$ data sets with the number of simulated vectors matching the number of patients in the BRCA data set, $N=`r nrow(example_brca)`$.
-At each iteration, we estimate the quadratic loss (residual sum of squared errors) from the specified ${\bf R}_{S}$, producing a distribution of loss values.
+At each iteration, we estimate the quadratic loss (residual sum of squared errors) from the specified $R_S$, producing a distribution of loss values.
 

### 2021-06-03 070-supplementary-materials.Rmd
Final push before pre-JSS submit internal review

@@
 We provide an open-source implementation of our methodology as the `Bigsimr.jl` Julia package, hosted on github at https://github.com/adknudson/Bigsimr.jl.
-Additionally we provide R and Python interfaces to `Bigsimr`, respectively at https://github.com/SchisslerGroup/r-bigsimr and https://github.com/SchisslerGroup/python-bigsimr.
+Additionally we provide R and Python interfaces to `Bigsimr`, respectively on https://github.com/SchisslerGroup/r-bigsimr and https://github.com/SchisslerGroup/python-bigsimr.
+A stable release of the R package `bigsimr` is also available on CRAN.
 `Bigsimr.jl` is an ongoing project, and feature requests or issues can be submitted to https://github.com/adknudson/Bigsimr.jl/issues.

### 2021-06-03 index.Rmd
Final push before pre-JSS submit internal review

@@
 title:
-  formatted: "Simulating High-Dimensional Multivariate Data using the \\pkg{bigsimr} Package"
-  plain:     "Simulating High-Dimensional Multivariate Data using the bigsimr Package"
-  short:     "\\pkg{bigsimr}: Simulate High-Dimensional Data"
+  formatted: "Simulating High-Dimensional Multivariate Data using the \\pkg{Bigsimr} Package"
+  plain:     "Simulating High-Dimensional Multivariate Data using the Bigsimr Package"
+  short:     "\\pkg{Bigsimr}: Simulate High-Dimensional Multivariate Data"
 abstract: >
@@
 Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations.
-To meet this gap, we developed the bigsimr R package for high-dimensional random vector simulation with arbitrary marginal distributions and known Pearson, Spearman, or Kendall correlation matrix.
-bigsimr contains additional high-performance features, including multi-core algorithms to estimate correlation and compute the nearest correlation matrix.
+To meet this gap, we developed the \pkg{Bigsimr} Julia package with R and Python interfaces.
+These packages empower high-dimensional random vector simulation with arbitrary marginal distributions and dependency via a Pearson, Spearman, or Kendall correlation matrix.
+\pkg{Bigsimr} contains high-performance features, including multi-core and graphical-processing-unit-accelerated algorithms to estimate correlation and compute the nearest correlation matrix.
 Monte Carlo studies quantify the accuracy and scalability of our approach, up to $d=10,000$.
-Finally, we describe example workflows and apply to a motivating high-dimensional data set --- RNA-sequencing data obtained from breast cancer tumor samples.
+We describe example workflows and apply to a high-dimensional data set --- RNA-sequencing data obtained from breast cancer tumor samples.
 keywords:

### 2021-06-21 020-simulation-algorithm.Rmd
removed duplicated ref
Revise for JOSS submission. Address Ed's comments.

@@
 For the computationally expensive process to match Pearson correlations, we approximate Equation \@ref(eq:pearsonIntegralRelation) for all pairs of margins.
-To this end, we implement the approximation scheme introduced by @xiao2019matching.
+To this end, we implement the approximation scheme introduced by @XZ19.
 Briefly, the many double integrals of the form in \@ref(eq:pearsonIntegralRelation) are approximated by weighted sums of Hermite polynomials.
@@
 Instead, we provide the quadratically-convergent algorithm based on the theory of strongly semi-smooth matrix functions (@qi2006quadratically).
-The nearest correlation matrix problem can be written down as the following convex optimization problem:  $\mathrm{min} \quad \frac{1}{2} \Vert G - X \Vert^2, \quad \mathrm{s.t.} \quad X_{ii} = 1, \quad i = 1, \ldots , n, \quad X \in S_{+}^{n}$.
+The nearest correlation matrix problem can be written down as the following convex optimization problem:  $\mathrm{min} \quad \frac{1}{2} \Vert R_X - X \Vert^2, \quad \mathrm{s.t.} \quad X_{ii} = 1, \quad i = 1, \ldots , n, \quad X \in S_{+}^{n}$.
 
@@
 The admissible classes consist of the matrices that map to a Pearson correlation matrix for a MVN.
-In particular, if we let $X$ be MVN with $d$ components and denote $\Omega_P = \{ R_P : R_P \textrm{ is a correlation matrix for } X \}, \quad \Omega_K = \{ R_K : R_K \textrm{ is a correlation matrix for } X \}, \quad \Omega_S = \{ R_S : R_S \textrm{ is a correlation matrix for } X \}$.
+In particular, if we let $X$ be MVN with $d$ components and denote $\Omega_P = \{ R_P : R_P \textrm{ is a Pearson correlation matrix for } X \}, \quad \Omega_K = \{ R_K : R_K \textrm{ is a Kendall correlation matrix for } X \}, \quad \Omega_S = \{ R_S : R_S \textrm{ is a Spearman correlation matrix for } X \}$.
 There are 1-1 mappings between these sets.

### 2021-06-21 030-bigsimr-package.Rmd
Revise for JOSS submission. Address Ed's comments.

@@
 
+<!-- i guess another point is that even if all the pairwise frechet bounds are satisfied - does that imply the proposed correlation matrix is valid or not - ie does there exist a multivariate distribution with all those pairwise correlations???
+some discussion here might help (help me at least)...
+maybe put another way - need to clarify how useful those bounds are  -->
+
 ```{r ch030-cor-bounds-pearson, echo=TRUE}
@@
 ```
+
+In this 2D example, it is possible to verify the existence of a bivariate distribution with the target Pearson correlation / margins (See the [Appendix]).
+For general $d-$variate distributions, is it not guaranteed that such a multivariate distribution exists with exact the specified correlations and margins [@BF17].
+Theoretical guidance is lacking in this regard, but, in practice, we find that using bivariate bounds form rough approximations to the bounds for the feasible region for the $d-$variate construction when aggregated into a Pearson matrix.
+See the [RNA-seq data application] section for an example discussing this, NORTA feasibility, and use of the nearest PD matrix algorithm.
 

### 2021-06-21 050-examples.Rmd
Revise for JOSS submission. Address Ed's comments.

@@
 
-As seen above $R_X$ is not strictly admissible in our scheme (as seen by the non-positive definite result above).
+As seen above $R_X$ is not strictly admissible in our scheme.
+This can be seen from the negative result from `bs$iscorrelation(R_X)`, which checks 1) positive-definiteness, 2) symmetry, and 3) 1s along the diagonal.
 However, the approximation is close with a maximum absolute error of `r max( abs(targets - approximates))` and average absolute error of `r mean(abs(targets - approximates))` across the `r choose(d,2)` correlations.
@@
 
+<!-- 
 ## Simulation-based joint probability calculations
@@
 
-```{r ch050-densityEvaluation, include=FALSE}
+```{r ch050-densityEvaluation, echo=T}
 d <- ncol(sim_nbinom)
@@
 ```
+
+ -->

### 2021-06-21 _book/_main.tex
Revise for JOSS submission. Address Ed's comments.

@@
 \author{
-\\
+Alfred G.
+Schissler\\University of Nevada \And Edward J.
+Bedrick\\University of Arizona \And Alexander D.
+Knudson\\University of Nevada \AND Tomasz J.
+Kozubowski\\University of Nevada \And Tin Nguyen\\University of Nevada \And Anna K.
+Panorska\\University of Nevada \AND Juli Petereit\\University of Nevada \And Walter W.
+Piegorsch\\University of Arizona \And Duc Tran\\University of Nevada
 }
@@
 
-\Plainauthor{}
+\Plainauthor{Alfred G. Schissler, Edward J. Bedrick, Alexander D. Knudson, Tomasz J. Kozubowski, Tin Nguyen, Anna K. Panorska, Juli Petereit, Walter W. Piegorsch, Duc Tran}
 \Plaintitle{Simulating High-Dimensional Multivariate Data using the Bigsimr Package}
-\Shorttitle{\pkg{Bigsimr}: Simulate High-Dimensional Data}
+\Shorttitle{\pkg{Bigsimr}: Simulate High-Dimensional Multivariate Data}
 
 \Abstract{
-It is critical to realistically simulate data when employing Monte Carlo techniques and evaluating statistical methodology.
+It is critical to accurately simulate data when employing Monte Carlo techniques and evaluating statistical methodology.
 Measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments.
 Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations.
-This article introduces the \textbackslash pkg\{Bigsimr\} R package.
-This high-level package provides a flexible and scalable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions and known Pearson, Spearman, or Kendall correlation matrix.
-\textbackslash pkg\{Bigsimr\} contains additional high-performance features, including multi-core algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix.
+To meet this gap, we developed the \pkg{Bigsimr} Julia package with R and Python interfaces.
+These packages empower high-dimensional random vector simulation with arbitrary marginal distributions and dependency via a Pearson, Spearman, or Kendall correlation matrix.
+\pkg{Bigsimr} contains high-performance features, including multi-core and graphical-processing-unit-accelerated algorithms to estimate correlation and compute the nearest correlation matrix.
 Monte Carlo studies quantify the accuracy and scalability of our approach, up to \(d=10,000\).
-Finally, we demonstrate example applications enabled via \textbackslash pkg\{Bigsimr\} by applying these methods to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples.
+We describe example workflows and apply to a high-dimensional data set --- RNA-sequencing data obtained from breast cancer tumor samples.
 }
@@
 \usepackage{amsthm}
-
 \newtheorem{theorem}{Theorem}
 \newtheorem{lemma}[theorem]{Lemma}
-
 \usepackage{booktabs}
@@
 \usepackage{ragged2e}
-
 \usepackage{setspace}
-\setstretch{2.0}
 
 \begin{document}
+
+\clearpage
 
@@
 
+\setstretch{2.0}
+
 Massive high-dimensional (HD) data sets are now commonplace in many areas of scientific inquiry.
-As new methods are developed for data, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of analyzing such proposed methodology --- such as false positive rates, statistical power, interval coverage, and robustness --- often with comparison to existing methods.
+As new methods are developed for data analysis, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of proposed methodology --- such as false positive rates, statistical power, interval coverage, and robustness.
 Further, efficient simulation empowers statistical computing strategies, such as the parametric bootstrap \citep{Chernick2008} to simulate from a hypothesized null model, providing inference in analytically challenging settings.
-Such Monte Carlo (MC) techniques become difficult for high-dimensional data using existing algorithms and tools.
-This is particularly true when simulating massive multivariate, non-normal distributions, arising naturally in many fields of study.
-
-As many have noted, it can be vexing to simulate dependent, non-normal/discrete data, even for low dimensional settings \citep{MB13, XZ19}.
-For continuous non-normal multivariate data, the well-known NORmal To Anything (NORTA) algorithm \citep{Cario1997} and other copula approaches \citep{Nelsen2007} are well-studied, with flexible, robust software available \citep{Yan2007, Chen2001}.
-Yet these approaches do not scale in a timely fashion to high-dimensional problems \citep{Li2019gpu}.
+Such Monte Carlo (MC) techniques become difficult for HD dependent data using existing algorithms and tools.
+This is particularly true when simulating massive multivariate, non-normal distributions, arising in many fields of study.
+
+As others have noted, it can be vexing to simulate dependent, non-normal/discrete data, even for low-dimensional (LD) settings \citep{MB13, XZ19}.
+For continuous non-normal LD multivariate data, the well-known NORmal To Anything (NORTA) algorithm \citep{Cario1997} and other copula approaches \citep{Nelsen2007} are well-studied, with existing software available \citep{Yan2007, Chen2001}.
+Yet these approaches do not scale in a timely fashion to HD problems \citep{Li2019gpu}.
 For discrete data, early simulation strategies had major flaws, such as failing to obtain the full range of possible correlations (e.g., admitting only positive correlations: see \citet{Park1996}).
-While more recent approaches \citep{MB13, Xia17, BF17} have largely remedied this issue for low-dimensional problems, the existing tools are not designed to scale to high dimensions.
-
-Another central issue lies in characterizing dependence between components in the high-dimensional random vector.
+While more recent approaches \citep{MB13, Xia17, BF17} have remedied this issue for LD problems, the existing tools are not designed to scale to high dimensions.
+
+Another central issue lies in characterizing dependence between components in the HD random vector.
 The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc).
@@
 Throughout, we focus on matching these nonparametric dependence measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology.
-We do, however, provide Pearson matching with some caveats.
+We do, however, implement Pearson matching, but several layers of approximation are required.
 
@@
 The crux of the method lies in the construction of a Gaussian copula in the spirit of the NORTA procedure.
-Further, we introduce the \texttt{bigsimr} R package that provides high-performance software implementing our algorithm.
-The algorithm design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed-form relationships between dependence measures for the multivariate normal (MVN) distribution.
-
-Our study proceeds by providing background information, including a description of a motivating example application: RNA-sequencing (RNA-seq) breast cancer data.
+As we will describe in more detail, the algorithm's design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed-form relationships between dependence measures for the multivariate normal (MVN) distribution.
+For our method, we developed a high-performance implementation: the \texttt{Bigsimr} Julia package, with R and Python interfaces \texttt{bigsimr}.
+
+This article proceeds by providing background information, including a description of a motivating example application: RNA-sequencing (RNA-seq) breast cancer data.
 Then we describe and justify our simulation methodology and related algorithms.
-Next, we detail an illustrative low-dimensional example of basic use of the \texttt{bigsimr} R package.
-Then we proceed with Monte Carlo studies under various bivariate distributional assumptions to assess accuracy.
-We conclude the Monte Carlo evaluations by summarizing the computation time for increasingly higher dimensional vectors.
+We proceed by providing an illustrative LD \texttt{bigsimr} workflow.
+Next we conduct MC studies under various bivariate distributional assumptions to evaluate performance and accuracy.
 After the MC evaluations, we simulate random vectors motivated by our RNA-seq example, evaluate the accuracy, and provide example statistical computing tasks, namely MC estimation of joint probabilities and evaluating HD correlation estimation efficiency.
@@
 
-The \texttt{bigsimr} \texttt{R} package presented here provides multiple algorithms that operate with high-dimensional multivariate data; however, all these algorithms were originally designed to support a single task: to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics.
-Specifically, our goal is to efficiently simulate a large number, \(B\), of HD random vectors \({\bf Y}=(Y_1, \ldots, Y_d)^\top\) with \emph{correlated} components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) \(F_i\).
+The \texttt{Bigsimr} package presented here provides multiple high-performance algorithms that operate with HD multivariate data.
+All these algorithms were originally designed to support a single task: to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics.
+Specifically, our goal is to efficiently simulate a large number, \(B\), of HD random vectors \({\bf Y}=(Y_1, \ldots, Y_d)^\top\) with \emph{correlated} components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) \(F_i, i=1,\ldots,d\).
 
 When designing this methodology, we developed the following properties to guide our effort.
-We divide the properties into two categories: (1) basic properties (BP) and ``scalability'' properties (SP).
+We divide the properties into two categories: (1) basic properties (BP) and `scalability' properties (SP).
 The BPs are adapted from an existing criteria due to \citet{Nik13a}.
@@
 \item
-  BP2: Flexible dependence, meaning that the number of bivariate marginals can be equal to the number of dependence parameters.
-\item
-  BP3: Flexible marginal modeling, generating heterogeneous data --- possibly from differing probability families.
+  BP2: Flexible dependence, meaning that the number of dependence parameters can be equal to the number of bivariate pairs in the vector.
+\item
+  BP3: Flexible marginal modeling, generating heterogeneous data --- including mixed continuous and discrete marginal distributions.
 \end{itemize}
@@
 
-Simulating high-dimensional, non-normal, correlated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data \citep{Wang2009b, Conesa2016b} derived from breast cancer patients.
-The RNA-seq data-generating process involves counting how often a particular form of human messenger RNA (mRNA) is expressed in a biological sample.
-RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data.
+Simulating HD, non-normal, correlated data motivates this work, in pursuit of modeling RNA-sequencing (RNA-seq) data \citep{Wang2009b, Conesa2016b} derived from breast cancer patients.
+The RNA-seq data-generating process involves counting how often a particular form of messenger RNA (mRNA) is expressed in a biological sample.
+RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in HD data.
 For human-derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted \citep{Schissler2019}.
@@
 \begin{CodeChunk}
-\begin{CodeInput}
-R> set.seed(2020-02-25)
-R> num_genes    <- 3
-R> num_patients <- 5
-R> gene_sample  <- sample(example_genes[1:1000], num_genes)
-R> 
-R> cap <- paste0("mRNA expression for three selected high-expressing genes, ",
-+               paste(gene_sample, collapse = ", "), 
-+               ", for the first five patients in the TCGA BRCA data set.")
-R> 
-R> small_brca <- example_brca %>%
-+   select(all_of(gene_sample)) %>%
-+   head(n = num_patients) %>%
-+   as_tibble(rownames = "Patient ID") %>%
-+   mutate(`Patient ID` = str_sub(`Patient ID`, end = 12))
-R> 
-R> small_brca %>%
-+   kable(format = "latex", booktabs = TRUE, caption = cap)
-\end{CodeInput}
 \begin{table}
@@
 \begin{CodeChunk}
-\begin{CodeInput}
-R> ggpairs(
-+   data = example_brca[, gene_sample],
-+   upper = list(continuous = wrap('cor', method = "spearman"))
-+ ) + 
-+   theme_bw() +
-+   theme(axis.text.x = element_text(angle = -90, vjust = 0.5))
-\end{CodeInput}
-\begin{CodeOutput}
-Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
-exact p-value with ties
-\end{CodeOutput}
-\begin{CodeOutput}
-Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
-exact p-value with ties
-\end{CodeOutput}
-\begin{CodeOutput}
-Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
-exact p-value with ties
-\end{CodeOutput}
 \begin{figure}
 
-{\centering \includegraphics{_main_files/figure-latex/ch010-realDataFig-1}
+{\centering \includegraphics[width=0.85\linewidth]{_main_files/figure-latex/ch010-realDataFig-1}
 
@@
 In multivariate analysis, an analyst must select a metric to quantify dependency.
-The most widely-known is the Pearson (product-moment) correlation coefficient that describes the linear association between two random variables \(X\) and \(Y\), and, it is given by
+The most widely-known is the Pearson correlation coefficient that describes the linear association between two random variables \(X\) and \(Y\), and, it is given by
 
@@
 For non-normal marginals with monotone correlation patterns, however, \(\rho_P\) suffers some drawbacks and may mislead or fail to capture important relationships \citep{MK01}.
-Alternatively in these settings, analysts often prefer rank-based correlation measures to describe the degree of monotonic association.
+Alternatively, analysts often prefer rank-based correlation measures to describe the degree of monotonic association.
 
 Two nonparametric, rank-based measures common in practice are Spearman's correlation (denoted \(\rho_S\)) and Kendall's \(\tau\).
-Spearman's \(\rho_S\) has an appealing correspondence as the Pearson correlation coefficient on \emph{ranks} of the values, thereby captures nonlinear yet monotone relationships.
-Kendall's \(\tau\), on the other hand, is the difference in probabilities of concordant and discordant pairs of observations, \((X_i, Y_i)\) and \((X_j, Y_j)\), with concordance meaning that orderings have the same direction (e.g., if \(X_i < X_j\), then \(Y_i < Y_j\)).
+Spearman's \(\rho_S\) has an appealing correspondence as the Pearson correlation coefficient on \emph{ranks} of the values, thereby capturing nonlinear yet monotone relationships.
+Kendall's \(\tau\), on the other hand, is the difference in probabilities of concordant and discordant pairs of observations, \((X_i, Y_i)\) and \((X_j, Y_j)\) (concordance meaning that orderings have the same direction, e.g., if \(X_i < X_j\), then \(Y_i < Y_j\)).
 Note that concordance is determined by the ranks of the values, not the values themselves.
-
 Both \(\tau\) and \(\rho_S\) are \emph{invariant under monotone transformations} of the underlying random variates.
@@
 Given two marginal distributions, \(\rho_P\) is not free to vary over the entire range of possible correlations \([-1,1]\).
-The so-called \emph{Frechet-Hoeffding bounds} are well-studied \citep{Nelsen2007, BF17}.
+The \emph{Frechet-Hoeffding bounds} are well-studied \citep{Nelsen2007, BF17}.
 These constraints cannot be overcome through algorithm design.
@@
 
-There is a strong connection of our simulation strategy to Gaussian \emph{copulas} (see \citet{Nelsen2007} for a technical introduction).
+There is a connection of our simulation strategy to Gaussian \emph{copulas} \citep{Nelsen2007}.
 A copula is a distribution function on \([0,1]^d\) that describes a multivariate probability distribution with standard uniform marginals.
-This provides a powerful, natural way to characterize joint probability structures.
+This provides a powerful, natural way to characterize joint probability.
 Consequently, the study of copulas is an important and active area of statistical theory and practice.
-
 For any random vector \({\bf X}=(X_1, \ldots, X_d)^\top\) with CDF \(F\) and marginal CDFs \(F_i\) there is a copula function \(C(u_1, \ldots, u_d)\) satisfying
@@
 
-\(x_i \in {\mathbb R}, i=1,\ldots,d.\)
-
-A Gaussian copula has marginal CDFs that are all standard normal, \(F_i = \Phi, \forall \, i\).
-This representation corresponds to a multivariate normal (MVN) distribution with standard normal marginal distributions and covariance matrix \({\bf R_P}\).
-As the marginals are standardized to have unit variance, however, \({\bf R_P}\) is a Pearson correlation matrix.
-If \(F_{{\bf R}}\) is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula \(C_{{\bf R}}\) is defined through
+A Gaussian copula has marginal CDFs that are all standard normal, \(F_i = \Phi, \forall \, i\), where \(\Phi(\cdot)\) is the standard normal CDF.
+This representation corresponds to a multivariate normal (MVN) distribution with standard normal marginal distributions and covariance matrix \(R_P\).
+Since the marginals are standardized to have unit variance, however, \(R_P\) is a Pearson correlation matrix.
+If \(F_R\) is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula \(C_R\) is defined through
 
@@
 \label{eq:gauss}
-F_{{\bf R}}(x_1, \ldots, x_d) = C_{{\bf R}}(\Phi(x_1), \ldots, \Phi(x_d)),
-\end{equation}
-
-where \(\Phi(\cdot)\) is the standard normal CDF.
-Note that the copula \(C_{{\bf R}}\) is the familiar multivariate normal CDF of the random vector \((\Phi(X_1), \ldots, \Phi(X_d))\), where \((X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})\).
+F_R(x_1, \ldots, x_d) = C_R(\Phi(x_1), \ldots, \Phi(x_d)),
+\end{equation}
+
+Here the copula function \(C_R\) is the familiar multivariate normal CDF of the random vector \((\Phi(X_1), \ldots, \Phi(X_d))\), where \((X_1, \ldots, X_d) \sim N_d({\bf 0}, R_P)\).
 
 Sklar's Theorem \citep{Sklar1959, Ubeda-Flores2017} guarantees that given inverse CDFs \(F_i^{-1}\)s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions.
-For example, using Gaussian copulas, we can construct a random vector \({\bf Y} = (Y_1, \ldots, Y_d)^\top\) with \(Y_i \sim F_i\), viz.~\(Y_i = F_i^{-1}(\Phi(X_i)), i=1, \ldots, d\), where \((X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})\).
+For example, using Gaussian copulas, we can construct a random vector \({\bf Y} = (Y_1, \ldots, Y_d)^\top\) with \(Y_i \sim F_i\), viz.
+\(Y_i = F_i^{-1}(\Phi(X_i)), i=1, \ldots, d\), where \((X_1, \ldots, X_d) \sim N_d({\bf 0}, R_P)\).
+
+\hypertarget{other-multivariate-simulation-packages}{%
+\subsection{Other Multivariate Simulation Packages}\label{other-multivariate-simulation-packages}}
+
+To place \texttt{Bigsimr}'s implementation and algorithms in context with existing approaches, here we qualitatively compare to other multivariate simulation packages.
+In \texttt{R}, the \texttt{MASS} package \texttt{mvnorm()} has long been available to produce random vectors from multivariate normal (and \(t\)) distributions.
+This procedure cannot simulate from arbitrary margins and dependency measures, as well as fails to readily scale to high dimensions.
+SAS PROC SIMNORMAL and Python's NumPy \texttt{random.multivariate\_normal} both do similar tasks.
+Recently, the high-performance \texttt{R} package \texttt{mvnfast} \citep{Fasiolo2016} has been released that provides HD multivariate normal generation, but this lacks flexibility in marginal models and dependency modeling.
+For dependent discrete data, LD \texttt{R} packages exist, such as \texttt{GenOrd}.
+For copula computation, the full-featured \texttt{R} package \texttt{copula} provides many different copulas and a fairly general LD random generator \texttt{rCopula()}.
+High-dimesional random vectors via \texttt{copula} takes an impractical amount of computing time \citep{Li2019gpu}.
+The \texttt{nortaRA} R package matches Pearson coefficients near exactly and is reasonable for LD settings.
+Finally, we know of no other \texttt{R} package that can produce HD random vectors with flexible margins and dependence modeling.
+Table \ref{tab:compare-table} summarizes the above discussion.
+
+\begin{table}[h]
+\centering
+\caption{\label{tab:compare-table}Comparison of selected features of \pkg{Bigsimr} to existing packages for multivariate simulation.}
+\begin{tabular}{@{}llll@{}}
+Package & Scales to HD & Hetereogenous Margins & Flexible Dependence \\ \midrule
+MASS      & No           & No                    & No                  \\
+mvnfast   & Yes          & No                    & No                  \\
+GenOrd    & No           & Discrete margins      & Yes                 \\
+copula    & No           & Yes                   & No                  \\
+nortaRA   & No           & Yes                   & No                  \\
+bigsimr   & Yes          & Yes                   & Yes                
+\end{tabular}
+\end{table}
+
+\clearpage
 
@@
 To characterize dependency, every pair \((Y_i, Y_j)\) has a given Pearson correlation \(\rho_P\), Spearman correlation \(\rho_S\), and/or Kendall's \(\tau\).
-The method can be described as a \emph{high-performance Gaussian copula} (Equation \eqref{eq:gauss}) providing a high-dimensional NORTA-inspired algorithm.
+The method can be described as a \emph{high-performance Gaussian copula} (Equation \eqref{eq:gauss}) providing a HD NORTA-inspired algorithm.
 
@@
 \item
-  Simulate a random vector \(\bf Z\) with \(d\) \emph{independent and identically distributed} (iid) standard normal components.
+  Simulate a random vector \(\bf Z\) with \(d\) independent and identically distributed (iid) standard normal components.
 \item
@@
 
-With modern parallel computing, steps 1, 3, 4, 5 are readily implemented as high-performance, multi-core and/or graphical-processing-unit (GPU) accelerated algorithms --- providing fast scalability using readily-available hardware.
+With modern parallel computing, steps 1, 3, 4, 5 are readily implemented as high-performance, multi-core and/or graphical-processing-unit (GPU) accelerated algorithms --- providing fast scalability to HD.
 
 Matching specified Pearson correlation coefficients exactly (step 2 above), however, is computationally costly.
-In general, there is no closed-form correspondence between the components of the input \(\Sigma_{\bf Z}\) and target \(\Sigma_{\bf Y}\).
+In general, there is no convenient correspondence between the components of the input \(\Sigma_{\bf Z}\) and target \(\Sigma_{\bf Y}\).
 Matching the correlations involves evaluating or approximating \(\binom{d}{2}\) integrals of the form
@@
 where \(\phi(\cdot)\) is the joint probability distribution function of two correlated standard normal variables.
-For HD data, (nearly) exact evaluation may become too costly to enable practical simulation studies.
-For low-dimensional problems, however, methods and tools exist to match Pearson correlations precisely; see \citep{Xia17} and the publicly available \texttt{nortaRA} R package \citep{Chen2001}.
-As described later, to enable HD Pearson matching we approximate these integrals.
+For HD simulation, exact evaluation may become too costly to enable practical simulation studies.
+Our remedy to enable HD Pearson matching in \texttt{bigsimr} is Hermite polynomial approximation of these \(\binom{d}{2}\) double integrals, as suggested by \citet{XZ19}.
 
@@
 Sklar's theorem provides a useful characterization of multivariate distributions through copulas.
-Yet the choice of copula-based simulation algorithm affects which joint distributions may be simulated.
-Even in low-dimensional spaces (e.g., \(d=3\)), there exist valid multivariate distributions with \emph{feasible} Pearson correlation matrices that NORTA cannot match exactly \citep{LH75}.
+In practice, however, the particulars of the simulation algorithm affect which joint distributions may be simulated.
+Even in LD spaces (e.g., \(d=3\)), there exist valid multivariate distributions with \emph{feasible} Pearson correlation matrices that NORTA cannot match exactly \citep{LH75}.
 This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined the resultant matrix is indefinite.
@@
 
-\citet{GH02} conducted a Monte Carlo study to estimate the probability of encountering NORTA defective matrices while increasing the dimension \(d\).
+\citet{GH02} conducted a study to estimate the probability of encountering NORTA defective matrices while increasing the dimension \(d\).
 They found that for what is now considered low-to-moderate dimensions (\(d \approx 20\)), almost \emph{all} feasible matrices are NORTA defective.
@@
 There is hope, however, as \citet{GH02} also showed that replacing an indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate \(d\).
-This provides evidence that our nearest positive definite (PD) augmented approach will maintain reasonable accuracy if our input matching scheme returns an indefinite matrix, at least for the rank-based matching scheme described above.
+This provides insight that our nearest positive definite (PD) augmented approach has promise to provide adequate accuracy if our input matching scheme returns an indefinite Pearson correlation matrix.
 
@@
 
-We now describe our algorithm to generate random vectors, which mirrors the classical NORTA algorithm above with some modifications for rank-based dependency matching:
+We now describe our algorithm to generate random vectors, which extends NORTA to HD via nearest correlation matrix replacement and provides rank-based dependency matching:
 
@@
     Convert the target Spearman correlation matrix \(R_S\) to the corresponding MVN Pearson correlation \(R_X\). Alternatively,
-    \end{enumerate}
-  \item
-    (i') Convert the target Kendall \(\tau\) matrix \(R_K\) to the corresponding MVN Pearson correlation \(R_X\). Alternatively,
-  \item
-    (i'\,') Convert the target Pearson correlation matrix to \(R_P\) to the corresponding, approximate MVN Pearson correlation \(R_X\).
+  \item
+    Convert the target Kendall \(\tau\) matrix \(R_K\) to the corresponding MVN Pearson correlation \(R_X\). Alternatively,
+  \item
+    Convert the target Pearson correlation matrix to \(R_P\) to the corresponding, approximate MVN Pearson correlation \(R_X\).
   \end{itemize}
@@
     Check that \(R_X\) is a correlation matrix, a positive definite matrix with 1's along the diagonal.
-    \end{enumerate}
-  \item
-    \begin{enumerate}
-    \def\labelenumii{(\roman{enumii})}
-    \setcounter{enumii}{1}
+  \item
+    If \(R_X\) is a correlation matrix, the input matrix is \emph{admissible} in this scheme. Otherwise,
+  \item
+    Replace \(R_X\) with the nearest correlation matrix \(\tilde{R}_X\), in the Frobenius norm.
+  \end{itemize}
+\item
+  Gaussian copula
+
+  \begin{itemize}
   \tightlist
   \item
-      If \(R_X\) is a correlation matrix, the input matrix is \emph{admissible} in this scheme. Otherwise
-    \end{enumerate}
-  \item
-    (ii') Replace \(R_X\) with the nearest correlation matrix \(\tilde{R}_X\), in the Frobenius norm.
-  \end{itemize}
-\item
-  Gaussian copula
-
-  \begin{itemize}
-  \item
-    \begin{enumerate}
-    \def\labelenumii{(\roman{enumii})}
-    \tightlist
-    \item
     Generate \({\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)\).
-    \end{enumerate}
-  \item
-    \begin{enumerate}
-    \def\labelenumii{(\roman{enumii})}
-    \setcounter{enumii}{1}
-    \tightlist
-    \item
-      Transform \({\bf X}\) to \({\bf U} = (U_1, \ldots, U_d)\) viz.~\(U_i=\Phi(X_i)\), \(i=1, \ldots, d\).
-    \end{enumerate}
-  \item
-    \begin{enumerate}
-    \def\labelenumii{(\roman{enumii})}
-    \setcounter{enumii}{2}
-    \tightlist
+  \item
+    Transform \({\bf X}\) to \({\bf U} = (U_1, \ldots, U_d)\) viz. \(U_i=\Phi(X_i)\), \(i=1, \ldots, d\).
   \item
     Return \({\bf Y} = (Y_1, \ldots, Y_d)\), where \(Y_i=F_i^{-1}(U_i)\), \(i=1, \ldots, d\).
-    \end{enumerate}
   \end{itemize}
@@
 
-\setstretch{2.0}
-
-\emph{Step 1}
-
-The first two descriptions of the \emph{Mapping step} employ the closed-form relationships between \(\rho_S\) and \(\tau\) with \(\rho_P\) for bivariate normal random variables via Equations \eqref{eq:convertKendall} and \eqref{eq:convertSpearman}, respectively (implemented as \texttt{cor\_covert}).
+\emph{Step 1: Mapping step.} We first employ the closed-form relationships between \(\rho_S\) and \(\tau\) with \(\rho_P\) for bivariate normal random variables via Equations \eqref{eq:convertKendall} and \eqref{eq:convertSpearman}, respectively (implemented as \texttt{cor\_covert()}).
 Initializing our algorithm to match the nonparametric correlations by computing these equations for all pairs is computationally trivial.
@@
 For the computationally expensive process to match Pearson correlations, we approximate Equation \eqref{eq:pearsonIntegralRelation} for all pairs of margins.
-To this end, we implement the approximation scheme introduced by \citet{xiao2019matching}.
+To this end, we implement the approximation scheme introduced by \citet{XZ19}.
 Briefly, the many double integrals of the form in \eqref{eq:pearsonIntegralRelation} are approximated by weighted sums of Hermite polynomials.
-Matching coefficients for pairs of continuous distributions is made tractable by this method, but for discrete distributions (especially discrete distributions with large support sets or infinite support), the problem is not so simple or efficient.
-
-Our solution is to approximate discrete distributions by a continuous distribution.
-The question then becomes that of which distribution to use.
-We found that Generalized S-Distributions \citep{muino2006gs} solve this problem by approximating a wide range of unimodal distributions, both continuous and discrete.
-Using the GS-Distribution approximation of discrete distributions in Pearson matching scheme above yields favorable results.
-
-\emph{Step 2}
-
-Once \(R_X\) has been determined, we check admissibility of the adjusted correlation via the steps described above.
+Matching coefficients for pairs of continuous distributions is made tractable by this method, but for discrete distributions (especially discrete distributions with large support sets or infinite support), the approximation is computationally expensive.
+To remedy this, we further approximate discrete distributions by a continuous distribution using Generalized S-Distributions \citep{muino2006gs}.
+
+\emph{Step 2: Admissibility check and nearest correlation matrix computation.} Once \(R_X\) has been determined, we check admissibility of the adjusted correlation via the steps described above.
 If \(R_X\) is not a valid correlation matrix, then we compute the nearest correlation matrix.
 Finding the nearest correlation matrix is a common statistical computing problem.
-The defacto default function in \texttt{R} is \texttt{Matrix::nearPD}, an alternating projection algorithm due to \citet{higham2002computing}.
+The defacto function in \texttt{R} is \texttt{Matrix::nearPD}, an alternating projection algorithm due to \citet{higham2002computing}.
 As implemented the function fails to scale to HD.
-Instead, we provide the quadratically convergent algorithm based on the theory of strongly semi-smooth matrix functions (\citet{qi2006quadratically}).
-The nearest correlation matrix problem can be written down as the following convex optimization problem:
-
-\begin{align*}
-    \mathrm{min}\quad & \frac{1}{2} \Vert G - X \Vert^2 \\
-    \mathrm{s.t.}\quad & X_{ii} = 1, \quad i = 1, \ldots , n, \\
-    & X \in S_{+}^{n}
-\end{align*}
-
-For nonparametric correlation measures, our algorithm allows the generation of high-dimensional multivariate data with arbitrary marginal distributions with a broad class of admissible Spearman correlation matrices and Kendall \(\tau\) matrices.
+Instead, we provide the quadratically-convergent algorithm based on the theory of strongly semi-smooth matrix functions (\citet{qi2006quadratically}).
+The nearest correlation matrix problem can be written down as the following convex optimization problem: \(\mathrm{min} \quad \frac{1}{2} \Vert R_X - X \Vert^2, \quad \mathrm{s.t.} \quad X_{ii} = 1, \quad i = 1, \ldots , n, \quad X \in S_{+}^{n}\).
+
+For nonparametric correlation measures, our algorithm allows the generation of HD multivariate data with arbitrary marginal distributions with a broad class of admissible Spearman correlation matrices and Kendall \(\tau\) matrices.
 The admissible classes consist of the matrices that map to a Pearson correlation matrix for a MVN.
-In particular, if we let \(X\) be MVN with \(d\) components and set
-
-\begin{align*}
-\Omega_P &= \{ R_P : R_P \textrm{ is a correlation matrix for } X \} \\
-\Omega_K &= \{ R_K : R_K \textrm{ is a correlation matrix for } X \} \\
-\Omega_S &= \{ R_S : R_S \textrm{ is a correlation matrix for } X \} \\
-\end{align*}
-
-then there are 1-1 mappings between these sets.
-We conjecture informally that the sets of admissible \(R_S\) and \(R_K\) are not highly restrictive.
+In particular, if we let \(X\) be MVN with \(d\) components and denote \(\Omega_P = \{ R_P : R_P \textrm{ is a Pearson correlation matrix for } X \}, \quad \Omega_K = \{ R_K : R_K \textrm{ is a Kendall correlation matrix for } X \}, \quad \Omega_S = \{ R_S : R_S \textrm{ is a Spearman correlation matrix for } X \}\).
+There are 1-1 mappings between these sets.
+We conjecture that the sets of admissible \(R_S\) and \(R_K\) are not highly restrictive.
 In particular, \(R_P\) is approximately \(R_S\) for a MVN, suggesting that the admissible set \(\Omega_S\) should be flexible as \(R_P\) can be any PD matrix with 1's along the diagonal.
-We provide methods to check whether a target \(R_S\) is an element of the \emph{admissible set} \(\Omega_S\) (and similarly for \(R_K\)).
-
-There is an increasing probability of encountering a non-admissible correlation matrix as dimension increases.
+We provide methods to check whether a target \(R_S\) is an element of the \emph{admissible set} \(\Omega_S\).
+
+There is an increasing probability of encountering an inadmissible correlation matrix as dimension increases.
 In our experience, the mapping step for large \(d\) almost always produces a \(R_X\) that is not a correlation matrix.
@@
 
-\emph{Step 3}
-
-The final step implements a NORTA-inspired, Gaussian copula approach to produce the desired margins.
+\emph{Step 3: Gaussian copula.} The final step implements a NORTA-inspired, Gaussian copula approach to produce the desired margins.
 Steps 1 and 2 determine the MVN Pearson correlation values that will eventually match the target correlation.
-Then all that is required is a fast MVN simulator, a standard normal CDF, and well-defined quantile functions for marginals.
-The MVN is transformed to a copula (distribution with standard uniform margins) by applying the normal CDF \(\phi(\cdot)\).
-The quantile functions \(F^{-1}\) are applied across the margins to return the desired random vector \({\bf Y}\).
-
-\section[bigsimr-pkg]{The \pkg{bigsimr} R package}\label{package}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> box::use(
-+   patchwork[...],
-+   dplyr[...],
-+   ggplot2[...],
-+   tidyr[drop_na],
-+   bigsimr[bigsimr_setup, distributions_setup],
-+   JuliaCall[julia_call]
-+ )
-R> 
-R> Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
-R> bs <- bigsimr_setup(pkg_check = FALSE)
-\end{CodeInput}
-\begin{CodeOutput}
-Julia version 1.5.4 at location /home/alex/julia-1.5.4/bin will be used.
-\end{CodeOutput}
-\begin{CodeOutput}
-Loading setup script for JuliaCall...
-\end{CodeOutput}
-\begin{CodeOutput}
-Finish loading setup script for JuliaCall.
-\end{CodeOutput}
-\begin{CodeInput}
-R> dist <- distributions_setup()
-\end{CodeInput}
-\end{CodeChunk}
-
-This section describes a low-dimensional (2D) random vector simulation workflow via the \texttt{bigsimr} \texttt{R} package (\url{https://github.com/SchisslerGroup/r-bigsimr}).
-The package \texttt{bigsimr} provides an interface to the native code written in Julia, registered as the \texttt{Bigsimr} Julia package (\url{https://github.com/adknudson/Bigsimr.jl}).
-In addition to the native Julia \texttt{Bigsimr} package and \texttt{R} interface \texttt{bigsimr}, we also provide a python interface \texttt{bigsimr} (\url{https://github.com/SchisslerGroup/python-bigsimr/}) that interfaces with the Julia \texttt{Bigsimr} package.
+Step 3 requires a fast MVN simulator, a standard normal CDF, and well-defined quantile functions for marginals.
+The MVN is transformed to a copula (distribution with standard uniform margins) by applying the normal CDF \(\Phi(\cdot)\).
+Finally, the quantile functions \(F_i^{-1}\) are applied across the margins to return the desired random vector \({\bf Y}\).
+
+\section[Bigsimr-pkg]{The \pkg{Bigsimr} package}\label{package}
+
+This section describes a bivariate random vector simulation workflow via the \texttt{bigsimr} \texttt{R} package.
+This \texttt{R} package provides an interface to the native code written in Julia (registered as the \texttt{Bigsimr} Julia package).
+In addition to the native Julia \texttt{Bigsimr} package and \texttt{R} interface \texttt{bigsimr}, we also provide a Python interface \texttt{bigsimr} that interfaces with the Julia \texttt{Bigsimr} package.
 The Julia package provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \ref{algorithms}).
 
-The subsections below describe the basic use of \texttt{bigsimr} by stepping through an example workflow using the data set \texttt{airquality} that contains daily air quality measurements in New York, May to September 1973 \citep{Chambers1983}.
+The subsections below describe the basic use of \texttt{bigsimr} R package by stepping through an example workflow using the data set \texttt{airquality} that contains daily air quality measurements in New York, May to September 1973 \citep{Chambers1983}.
 This workflow proceeds from setting up the computing environment, to data wrangling, estimation, simulation configuration, random vector generation, and, finally, result visualization.
 
-\hypertarget{bivariate-example-description}{%
-\subsection{Bivariate example description}\label{bivariate-example-description}}
-
-We illustrate the use of \texttt{bigsimr} motivated by the New York air quality data set (\texttt{airquality}) included in the R \texttt{datasets} package.
+\hypertarget{bivariate-example}{%
+\subsection{Bivariate example}\label{bivariate-example}}
+
+We illustrate the use of \texttt{bigsimr} using the New York air quality data set (\texttt{airquality}) included in the R \texttt{datasets} package.
 First, we load the \texttt{bigsimr} library and a few other convenient data science packages, including the syntactically-elegant \texttt{tidyverse} suite of \texttt{R} packages.
@@
 
-For simplicity and to provide a minimal working example, we consider bivariate simulation of the two \texttt{airquality} variables \texttt{Temperature}, in degrees Fahrenheit, and \texttt{Ozone} level, in parts per billion.
+Here, we describe a minimal working example --- a bivariate simulation of two airquality variables: \texttt{Temperature}, in degrees Fahrenheit, and \texttt{Ozone} level, in parts per billion.
 
@@
 R> df <- airquality %>% select(Temp, Ozone) %>% drop_na()
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{CodeInput}
 R> glimpse(df)
@@
 
-Figure \ref{fig:ch030-aq-joint-dist} visualizes the bivariate relationship between Ozone and Temperature.
+Figure \ref{fig:ch030-aq-joint-dist} visualizes the bivariate relationship between \texttt{Ozone} and \texttt{Temperature}.
 We aim to simulate random two-component vectors mimicking this structure.
-The margins are not normally distributed; particularly the Ozone level exhibits a strong positive skew.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> p0 <- ggplot(df, aes(Temp, Ozone)) +
-+   geom_point(size = 1) +
-+   theme(legend.position = "none") + 
-+   labs(x = "Temperature")
-R> 
-R> pTemp <- ggplot(df, aes(Temp)) + 
-+   geom_density() +
-+   theme(axis.title.x = element_blank(),
-+         axis.title.y = element_blank(),
-+         axis.text = element_blank(),
-+         axis.ticks = element_blank())
-R> 
-R> pOzone <- ggplot(df, aes(Ozone)) + 
-+     geom_density() +
-+   theme(axis.title.y = element_blank(),
-+         axis.title.x = element_blank(),
-+         axis.ticks = element_blank(),
-+         axis.text = element_blank()) +
-+   coord_flip()
-R> 
-R> pTemp + plot_spacer() + p0 + pOzone + 
-+   plot_layout(widths = c(3,1), heights = c(1, 3)) 
-\end{CodeInput}
+The margins are not normally distributed; \texttt{Ozone} level exhibits a strong positive skew.
+
+\begin{CodeChunk}
 \begin{figure}
 
-{\centering \includegraphics{_main_files/figure-latex/ch030-aq-joint-dist-1}
+{\centering \includegraphics[width=0.85\linewidth]{_main_files/figure-latex/ch030-aq-joint-dist-1}
 
@@
 For this example, we avoid goodness-of-fit considerations to determine the marginal distributions.
-But it seems sensible without domain knowledge to estimate these quantities from the data, and \texttt{bigsimr} contains fast functions designed for this task.
+It is sensible without domain knowledge to estimate these quantities from the data, and \texttt{bigsimr} contains fast functions designed for this task.
 
@@
 Based on the estimated densities in Figure \ref{fig:ch030-aq-joint-dist}, we assume \texttt{Temp} is normally distributed and \texttt{Ozone} is log-normally distributed, as the latter values are positive and skewed.
-We use the well-known, unbiased estimators for the normal distribution's parameters and maximum likelihood estimators for the log-normal parameters:
+We use the classical unbiased estimators for the normal distribution's parameters and maximum likelihood estimators for the log-normal parameters:
 
@@
 
-Next, we configure the input marginals for later input into \texttt{rvec}. The marginal distributions are specified using \texttt{Julia}'s \texttt{Distributions} package and stored in a vector.
+Now, we configure the input marginals for later input into \texttt{rvec}. The marginal distributions are specified using \texttt{Julia}'s \texttt{Distributions} package and stored in a vector.
 
@@
 
-As mentioned, the user must decide how to describe correlation based on the particulars of the problem.
-For non-normal data and for improved simulation accuracy/scalability in our scheme, we advocate the use of Spearman's \(\rho\) correlation matrix \(R_S\) and Kendall's \(\tau\) correlation matrix \(R_K\).
-We also support Pearson correlation coefficient matching, while cautioning the user to check the performance for their parametric multivariate model (see \protect\hyperlink{simulations}{Monte Carlo evaluations} below for evaluation strategies and guidance).
-These estimation methods are classical approaches, not designed for high-dimensional correlation estimation (see the \href{\%7B\#discussion}{Conclusion and Discussion} sections for more on this).
+The user must decide how to describe correlation based on the particulars of the problem.
+For non-normal data in our scheme, we advocate the use of Spearman's \(\rho\) correlation matrix \(R_S\) or Kendall's \(\tau\) correlation matrix \(R_K\).
+We also support Pearson correlation coefficient matching, while cautioning the user to check the performance for the distribution at hand (see \protect\hyperlink{simulations}{Monte Carlo evaluations} below for evaluation strategies and guidance).
+Note that these estimation methods are classical approaches, not specifically designed for high-dimensional correlation estimation.
 
@@
 
-First we use \texttt{cor\_bounds} to ensure that the pairwise target correlation values are valid prior to the mapping step which constructs \(R_X\) for the MVN input into NORTA (see Section \ref{algorithms}).
-\texttt{cor\_bounds} estimates the pairwise lower and upper correlation bounds using the Generate, Sort, and Correlate algorithm of \citet{DH2011}.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> bounds <- bs$cor_bounds(margins)
-R> bounds$lower
-\end{CodeInput}
-\begin{CodeOutput}
-        [,1]    [,2]
-[1,]  1.0000 -0.8821
-[2,] -0.8821  1.0000
-\end{CodeOutput}
-\begin{CodeInput}
-R> bounds$upper
-\end{CodeInput}
-\begin{CodeOutput}
-       [,1]   [,2]
-[1,] 1.0000 0.8812
-[2,] 0.8812 1.0000
-\end{CodeOutput}
-\end{CodeChunk}
-
-Since our single estimated Spearman correlation is within the theoretical bounds, the correlation is valid as input to \texttt{rvec}.
-But even if the 2-dimensional bounds are satisfied for each pair of margins, this does not guarantee the feasibility of a \(d-\)variate distribution \citep{BF17}.
-
-To provide higher dimensional feasibility/admissibility checking, we begin by mapping using \texttt{cor\_convert} and check admissibility.
+Once a target correlation matrix is specified, first convert the dependency type to Pearson to correctly specify the MVN inputs.
+Then, check that the converted matrix \(R_X\) is a valid correlation matrix (PD and 1s on the diagonal).
+For this bivariate example, \(R_X\) is a valid correlation matrix.
+Note that typically in HD the resultant MVN Pearson correlation matrix is indefinite and requires approximation (see the HD examples in subsequent sections).
 
@@
 
-The bounds on the Pearson correlation coefficient \(R_P\) between these margins are restricted as seen below in our MC estimated correlation bounds:
+Despite being a valid correlation matrix under this definition, marginal distributions induce Frechet bounds on the possible Pearson correlation values.
+To check that the converted correlation values fall within these bounds, use \texttt{bigsimr::cor\_bounds} to estimate the pairwise lower and upper correlation bounds.
+\texttt{cor\_bounds} uses the Generate, Sort, and Correlate algorithm of \citet{DH2011}.
+For the assumed marginals, the Pearson correlation coefficient is not free to vary in {[}-1,1{]}, as seen below.
+Our MC estimate of the bounds slightly underestimates the theoretic bounds of \((-0.881, 0.881)\).
+(See an analytic derivation presented in the Appendix).
+Since our single Pearson correlation coefficient is within the theoretical bounds, the correlation matrix is valid for our simulation strategy.
 
@@
 Julia Object of type NamedTuple{(:lower, :upper),Tuple{Float64,Float64}}.
-(lower = -0.8814915390483397, upper = 0.8817369743444273)
+(lower = -0.8823026550919685, upper = 0.8815987574226597)
 \end{CodeOutput}
@@
 
-Our MC estimate of the bounds slightly overestimates the theoretic bounds of \((-0.881, 0.881)\). An analytic derivation is presented as an Appendix.
+In this 2D example, it is possible to verify the existence of a bivariate distribution with the target Pearson correlation / margins (See the \protect\hyperlink{appendix}{Appendix}).
+For general \(d-\)variate distributions, is it not guaranteed that such a multivariate distribution exists with exact the specified correlations and margins \citep{BF17}.
+Theoretical guidance is lacking in this regard, but, in practice, we find that using bivariate bounds form rough approximations to the bounds for the feasible region for the \(d-\)variate construction when aggregated into a Pearson matrix.
+See the \protect\hyperlink{examples}{RNA-seq data application} section for an example discussing this, NORTA feasibility, and use of the nearest PD matrix algorithm.
 
@@
 
-Finally, we arrive at the main function of \texttt{bigsimr}: \texttt{rvec}. We now simulate \(B=10,000\) random vectors from the assumed joint distribution of Ozone levels and Temp.
+Finally, we execute \texttt{rvec} to simulate the desired \(10,000\) random vectors from the assumed bivariate distribution of \texttt{Ozone} and \texttt{Temp}.
+Note that the target input in specified using the pre-computed MVN correlation matrix.
+Figure \ref{fig:ch030-plot-sim} plots the 10,000 simulated points.
 
@@
 
-Figure \ref{fig:ch030-plot-sim} plots the 10,000 simulated points.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> p1 <- df_sim %>%
-+   ggplot(aes(Temp, Ozone)) +
-+   geom_density_2d_filled() +
-+   theme(legend.position = "none") +
-+   labs(x = "Simulated Temperature", y = "Simulated Ozone") +
-+   scale_y_continuous(limits = c(0, max(df$Ozone))) +
-+   scale_x_continuous(limits = range(df$Temp))
-R> 
-R> p1Temp <- ggplot(df_sim, aes(Temp)) + 
-+   geom_density(alpha = 0.5, fill = "lightseagreen") +
-+   theme(axis.title.x = element_blank(),
-+         axis.title.y = element_blank(),
-+         axis.text = element_blank(),
-+         axis.ticks = element_blank())
-R> 
-R> p1Ozone <- ggplot(df_sim, aes(Ozone)) + 
-+   geom_density(alpha = 0.5, fill = "lightseagreen") + 
-+   theme(axis.title.y = element_blank(),
-+         axis.title.x = element_blank(),
-+         axis.ticks = element_blank(),
-+         axis.text = element_blank()) +
-+   coord_flip() 
-R> 
-R> p1Temp + plot_spacer() + p1 + p1Ozone + 
-+   plot_layout(widths = c(3,1), heights = c(1, 3))
-\end{CodeInput}
-\begin{CodeOutput}
-Warning: Removed 392 rows containing non-finite values (stat_density2d_filled).
-\end{CodeOutput}
+\begin{CodeChunk}
 \begin{figure}
 
-{\centering \includegraphics{_main_files/figure-latex/ch030-plot-sim-1}
+{\centering \includegraphics[width=0.85\linewidth]{_main_files/figure-latex/ch030-plot-sim-1}
 
@@
 
+\clearpage
+
 \hypertarget{simulations}{%
@@
 
-Before applying our methodology to real data simulation, we conduct several Monte Carlo studies to investigate method performance.
-Since marginal parameter matching in our scheme is essentially a sequence of univariate inverse probability transforms, the challenging aspects are the accuracy of dependency matching and computational efficiency at high dimensions.
+In this section, we conduct MC studies to investigate method performance.
+Marginal parameter matching is straightforward in our scheme as it is a sequence of univariate inverse probability transforms.
+This process is both fast and exact.
+On the other hand, accurate and computationally efficent dependency matching presents a challenge, especially in the HD setting.
 To evaluate our methods in those respects, we design the following numerical experiments to first assess accuracy of matching dependency parameters in bivariate simulations and then time the procedure in increasingly large dimension \(d\).
@@
 
-We select bivariate simulation configurations to ultimately simulate our motivating discrete-valued RNA-seq example, and, so we proceed in increasing complexity, leading to the model in our motivating application in Section \ref{examples}.
-We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson's, Spearman's, and Kendall's --- in identical, bivariate marginal configurations.
-For each pair of identical margins, we vary the target correlation across \(\Omega\), the set of possible admissible values for correlation type, to evaluate the simulation's ability to obtain the theoretic bounds.
-The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (mimicking RNA-seq counts).
+We select bivariate simulation configurations to ultimately simulate our motivating discrete-valued RNA-seq example.
+Thus, we proceed in increasing departure from normality, leading to a multivariate negative binomial (MVNB) model for our motivating data.
+We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson, Spearman, and Kendall --- in identical, bivariate marginal configurations.
+For each pair of identical margins, we vary the target correlation across \(\Omega\), the set of possible admissible values for each correlation type, to evaluate the simulation's ability to simulate all possible correlations.
+The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (non-normal and discrete).
 
 Table \ref{tab:sims} lists our identical-marginal, bivariate simulation configurations.
-We increase the simulate replicates \(B\) to check that our results converge to the target correlations and gauge statistical efficiency.
+We increase the simulate replicates \(B\) to visually ensure that our results converge to the target correlations and gauge efficiency.
 We select distributions beginning with a standard multivariate normal (MVN) as we expect the performance to be exact (up to MC error) for all correlation types.
-Then, we select a non-symmetric continuous distribution: a standard (rate =1) two-component multivariate gamma (MVG).
-Finally, we select distributions and marginal parameter values that are motivated by our RNA-seq data, namely values proximal to probabilities and sizes estimated from the data (see \href{examples}{Example applications} for estimation details).
-Thus we arrive at a multivariate negative binomial (MVNB) \(p_1 = p_2 = 3\times10^{-4}, r_1 = r_2 = 4, \rho \in \Omega\).
-
-Table: \label{tab:sims} Identical margin, bivariate simulation configurations to evaluate correlation matching accuracy and efficiency.
+Then, we select a non-symmetric continuous distribution: a standard (rate =1), two-component multivariate gamma (MVG).
+Finally, we select distributions and marginal parameter values that are motivated by our RNA-seq data, namely values proximal to probabilities and sizes estimated from the data (see \href{examples}{RNA-seq data application} for estimation details).
+Specifically, we assume a MVNB with \(p_1 = p_2 = 3\times10^{-4}, r_1 = r_2 = 4, \rho \in \Omega\).
 
 \begin{table}[]
+\centering
+\caption{ \label{tab:sims} Identical margin, bivariate simulation configurations to evaluate matching accuracy.}
 \begin{tabular}{@{}lcr@{}}
@@
 For each of the unique 9 simulation configurations described above, we estimate the correlation bounds and vary the correlations along a sequence of 100 points evenly placed within the bounds, aiming to explore \(\Omega\).
-Specifically, we set correlations \(\{ \rho_1 = ( \hat{l} + \epsilon), \rho_2 = (\hat{l} + \epsilon) + \delta, \ldots, \rho_{100} = (\hat{u} - \epsilon) \}\), with \(\hat{l}\) and \(\hat{u}\) being the estimated lower and upper bounds, respectively, and increment value \(\delta\).
+Specifically, we set correlations \(\{ \rho_1 = ( \hat{l} + \epsilon), \rho_2 = (\hat{l} + \epsilon) + \delta, \ldots, \rho_{100} = (\hat{u} - \epsilon) \}\), with \(\hat{l}\) and \(\hat{u}\) being the estimated lower and upper bounds, respectively, with increment value \(\delta\).
 The adjustment factor, \(\epsilon=0.01\), is introduced to handle numeric issues when the bound is specified exactly.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> bivariate_normal_sims %>%
-+     ggplot(aes(rho, rho_hat, color = type)) +
-+     geom_point() +
-+     geom_abline(slope = 1) +
-+     facet_wrap(~ type + N) + 
-+     theme_bw()
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> bivariate_gamma_sims %>%
-+     ggplot(aes(rho, rho_hat, color = type)) +
-+     geom_point() +
-+     geom_abline(slope = 1) +
-+     facet_wrap(~ type + N) + 
-+     theme_bw()
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> bivariate_nbinom_sims %>%
-+     ggplot(aes(rho, rho_hat, color = type)) +
-+     geom_point() +
-+     geom_abline(slope = 1) +
-+     facet_wrap(~ type + N) + 
-+     theme_bw()
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> allDat <- bind_rows(
-+     select(bivariate_normal_sims, margins, type, N, rho, rho_hat),
-+     select(bivariate_gamma_sims, margins, type, N, rho, rho_hat),
-+     select(bivariate_nbinom_sims, margins, type, N, rho, rho_hat)
-+ ) %>%
-+     mutate(margins = factor(margins, levels=c("norm", "gamma", "nbinom")),
-+            type = factor(type, levels=c("Pearson", "Spearman", "Kendall")),
-+            N = factor(N))
-\end{CodeInput}
-\end{CodeChunk}
 
@@
 Table \ref{tab:ch040-BiError} contains the mean absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> tabMAE <- allDat %>%
-+     group_by(N, type, margins) %>%
-+     summarize(MAE = mean(abs(rho - rho_hat))) %>%
-+     ungroup()
-\end{CodeInput}
-\begin{CodeOutput}
-`summarise()` has grouped output by 'N', 'type'. You can override using the `.groups` argument.
-\end{CodeOutput}
-\begin{CodeInput}
-R> kable(tabMAE, booktabs = TRUE, format = "latex",
-+       linesep = c("", "", "\\addlinespace"),
-+       col.names = c("No. of random vectors",
-+                     "Correlation type",
-+                     "Distribution",
-+                     "Mean abs. error"),
-+       caption = "Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.")
-\end{CodeInput}
+Overall, the studies show that our methodology is generally accurate across the entire range of possible correlation for all three dependency measures.
+Our Pearson matching performs nearly as well as Spearman or Kendall, except for a slight increase in error for negative binomial case.
+
+\begin{CodeChunk}
 \begin{table}
@@
 
-Overall, the studies show that our methodology is generally accurate across the entire range of possible correlation values all three dependency measures, at least in these limited simulation settings for the rank-based correlations.
-Our Pearson matching performs nearly as well as Spearman or Kendall, except for a slight increase in error for negative binomial case.
-This is due the particularly large counts generated with our choice of parameters for \(p\) and \(r\).
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> # https://www.datanovia.com/en/blog/how-to-change-ggplot-facet-labels/
-R> # New facet label names
-R> repsLabs <- paste0("B=", c("1,000", "10,000", "100,000") )
-R> names(repsLabs) <- c(1000, 10000, 100000)
-R> typeLabs <- c("Pearson", "Spearman", "Kendall")
-R> names(typeLabs) <- c("Pearson", "Spearman", "Kendall")
-R> 
-R> # Set colors
-R> ## RColorBrewer::display.brewer.all()
-R> numColors <- 4
-R> numGroups <- length(levels(allDat$margins))
-R> myColors <- rev(RColorBrewer::brewer.pal(n = numColors, name = "Greys")[ ((numColors - numGroups) + 1):numColors])
-R> 
-R> allDat %>%
-+     ggplot(aes(x = rho, y = rho_hat, color = margins)) +
-+     geom_point(size = 2) +
-+     scale_color_manual(values = myColors ) +
-+     geom_abline(slope = 1, linetype = "dashed") +
-+     labs(x = "Specified Correlation", y = "Estimated Correlation") +
-+     facet_wrap(~ type + N, labeller = labeller(N = repsLabs, type = typeLabs)) +
-+     theme_bw() +
-+     theme(legend.position = "bottom", legend.direction = "horizontal")
-\end{CodeInput}
+\begin{CodeChunk}
 \begin{figure}
 
-{\centering \includegraphics{_main_files/figure-latex/ch040-bPlot-1}
+{\centering \includegraphics[width=0.85\linewidth]{_main_files/figure-latex/ch040-bPlot-1}
 
@@
 
-With information of our method's accuracy from a low-dimensional perspective, we now turn to assessing whether \texttt{bigsimr} can scale to larger dimensional problems with practical computation times.
-Specifically, we ultimately generate \(B=1,000\) random vectors for \(d=\{100, 250, 500, 1000, 2500, 5000, 10000\}\) for each correlation type, \(\{Pearson, Spearman, Kendall\}\) while timing the algorithm's major steps.
-
-To mimic the workflow, we first produce a synthetic ``data set'' by completing the following steps:
+With information of our method's accuracy from a low-dimensional perspective, we now assess whether \texttt{bigsimr} can scale to larger dimensional problems with practical computation times.
+We ultimately generate \(B=1,000\) random vectors for \(d=\{100, 250, 500, 1000, 2500, 5000, 10000\}\) for each correlation type, \{Pearson, Spearman, Kendall\} while timing the algorithm's major steps.
+We begin by producing a synthetic ``data set'' by completing the following steps:
 
@@
 \item
-  Simulate a ``data set'' of \(1,000 \times d\) random vectors via \texttt{rvec} (without matching the Pearson exactly).
+  Simulate a ``data set'' of \(1,000 \times d\) random vectors via \texttt{rvec}.
   \setstretch{2.0}
@@
 
-With the synthetic data set in hand, we complete and time the following 4 steps:
+With the synthetic data set in hand, we complete and time the following four steps involved in a typical workflow (including correlation estimation).
 
 \setstretch{1.5}
-
-\begin{enumerate}
-\def\labelenumi{\roman{enumi}.}
-\tightlist
-\item
-  Estimate the correlation matrix from the ``data'' in the \emph{Compute Correlation} step.
-\end{enumerate}
 
@@
 \item
+  Estimate the correlation matrix from the ``data'' in the \emph{Compute Correlation} step.
+\item
   Map the correlations to initialize the algorithm (Pearson to Pearson, Spearman to Pearson, or Kendall to Pearson) in the \emph{Adjust Correlation} step.
@@
 \item
-  Simulate \(1,000\) vectors via the Inverse Transform method in the \emph{Simulate Data} step.
+  Simulate \(1,000\) vectors in the \emph{Simulate Data} step.
   \setstretch{2.0}
@@
 The experiments are conducted on a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, with all 16 threads employed during computation.
-The results for \(d \leq 500\) are fast with all times (except Kendall) under 3 seconds.
-Table \ref{tab:ch040-moderateDtab} displays the total computation time of the algorithms steps 1 through 3, including estimation step i.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> dat_long <- benchmark_dependences %>%
-+   select(-total_time, -n_sim, -needed_near_pd) %>%
-+   pivot_longer(cols = c(corr_time:sim_time), names_to = "Step", values_to = "Time") %>%
-+   mutate(dim = factor(dim),
-+          Step = factor(Step,
-+                        levels = c("corr_time", "adj_time", "admiss_time", "sim_time"),
-+                        labels = c("Compute Correlation",
-+                                   "Adjust Correlation",
-+                                   "Check Admissibility",
-+                                   "Simulate Data"))) %>%
-+     rename(Correlation = corr_type, Dimensions = dim)
-R> 
-R> tabTimes  <- dat_long %>%
-+     filter(Dimensions %in% c(100, 250, 500)) %>%
-+     group_by(Dimensions, Correlation) %>%
-+     summarize('Total Time(Seconds)' = sum(Time)) %>%
-+     ungroup()
-\end{CodeInput}
-\begin{CodeOutput}
-`summarise()` has grouped output by 'Dimensions'. You can override using the `.groups` argument.
-\end{CodeOutput}
-\begin{CodeInput}
-R> tabTimes %>%
-+     kable(booktabs = TRUE,
-+           linesep = c("", "", "\\addlinespace"),
-+           col.names = c("Dimension",
-+                         "Correlation type",
-+                         "Total Time (Seconds)"),
-+           caption = 'Total time to produce 10,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.')
-\end{CodeInput}
+Table \ref{tab:ch040-moderateDtab} displays the total computation time for moderate dimesions (\(d \leq 500\)).
+In every simullation setting, the 1,000 vectors are generated rapidly, executing in under 2 seconds.
+
+\begin{CodeChunk}
 \begin{table}
 
-\caption{\label{tab:ch040-moderateDtab}Total time to produce 10,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.}
+\caption{\label{tab:ch040-moderateDtab}Total time to produce 1,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.}
 \centering
@@
 \midrule
-100 & Pearson & 0.080\\
-100 & Spearman & 0.036\\
-100 & Kendall & 0.116\\
+100 & Pearson & 0.081\\
+100 & Spearman & 0.027\\
+100 & Kendall & 0.075\\
 \addlinespace
-250 & Pearson & 0.286\\
-250 & Spearman & 0.371\\
-250 & Kendall & 0.366\\
+250 & Pearson & 0.378\\
+250 & Spearman & 0.062\\
+250 & Kendall & 0.375\\
 \addlinespace
-500 & Pearson & 2.932\\
-500 & Spearman & 0.176\\
-500 & Kendall & 2.183\\
+500 & Pearson & 1.425\\
+500 & Spearman & 0.135\\
+500 & Kendall & 1.495\\
 \bottomrule
@@
 
-The results with larger dimensional vectors show scalability to ultra-high dimensions for all three correlation types, although the total times do become much larger.
+The results for \(d > 500\) show scalability to ultra-high dimensions for all three correlation types, although the total times do become much larger.
 Figure \ref{fig:ch040-largeDfig} displays computation times for \(d=\{1000, 2500, 5000, 10000\}\).
@@
 Interestingly, Pearson is actually faster than Kendall for \(d=10,000\) due to bottlenecks in \emph{Compute Correlation} and \emph{Check Admissibility}.
-Uniformly, matching Spearman correlations is faster, with total times under 5 minutes for \(d=10,000\), making Spearman correlations the most computationally-friendly dependency type.
+Uniformly, matching Spearman correlations is faster, with total times under 5 minutes for \(d=10,000\), making Spearman the most computationally-friendly dependency type for our package.
 With this in mind, we scaled the simulation to \(d=20,000\) for the Spearman type and obtained the 1,000 vectors in under an hour (data not shown).
 In principle, this would enable the simulation of an entire human-derived RNA-seq data set.
-We note that for a given target correlation matrix and margins, steps i, 1, and 2 only need to be computed once and the third step, \emph{Simulate Data}, is fast under all schemes for all \(d\) under consideration.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> # Set colors
-R> numColors <- 5
-R> numGroups <- 4
-R> myColors <- rev(brewer.pal(n=numColors, name="Greys")[((numColors-numGroups)+1):numColors])
-R> 
-R> dat_long %>%
-+     filter(Dimensions %in% c(1000, 2500, 5000, 10000, 20000)) %>%
-+     mutate(`Time (minutes)` = Time / 60) %>%
-+     ggplot(aes(Correlation, `Time (minutes)`, fill=Step)) +
-+     geom_bar(position = "stack", stat = "identity") +
-+     scale_fill_manual(values = myColors ) +
-+     scale_y_continuous(breaks = seq(0, 80, 10),
-+                      minor_breaks = NULL) +
-+     facet_grid(. ~ Dimensions) +
-+     theme(axis.text.x = element_text(angle = -90))
-\end{CodeInput}
+We note that for a given target correlation matrix and margins, steps 1, 2, and 3 only need to be computed once and the fourth step, \emph{Simulate Data}, is nearly instaneous for all settings considered.
+
+\begin{CodeChunk}
 \begin{figure}
 
-{\centering \includegraphics{_main_files/figure-latex/ch040-largeDfig-1}
+{\centering \includegraphics[width=0.85\linewidth]{_main_files/figure-latex/ch040-largeDfig-1}
 
@@
 This could mitigate some of the current algorithmic/implementation shortcomings.
-Despite this, Kendall matching is still feasible for most high-dimensional data sets.
-Finally, we note that one could use our single-pass algorithm \texttt{cor\_fastPD} to produce a `close' (not nearest PD) to scale to even higher dimensions with greater loss of accuracy.
+Despite this, Kendall matching is still feasible for most HD data sets.
+Finally, we note that one could use our single-pass algorithm \texttt{cor\_fastPD} to produce a `close' (not nearest PD) to scale to even higher dimensions with some loss of accuracy.
 
 \hypertarget{examples}{%
-\section{RNA-seq data applications}\label{examples}}
-
-This section demonstrates how to simulate multivariate data using \texttt{bigsimr}, aiming to replicate the structure of high-dimensional dependent count data.
+\section{RNA-seq data application}\label{examples}}
+
+This section demonstrates how to simulate multivariate data using \texttt{bigsimr}, aiming to replicate the structure of HD dependent count data.
 In an illustration of our proposed methodology, we seek to simulate RNA-sequencing data by producing simulated random vectors mimicking the observed data and its generating process.
-Modeling RNA-seq using multivariate probability distributions is natural as inter-gene correlation is an inherent part of biological processes \citep{Wang2009b}.
+Modeling RNA-seq using multivariate probability distributions is natural as inter-gene correlation occurs during biological processes \citep{Wang2009b}.
 And yet, many models do not account for this, leading to major disruptions to the operating characteristics of statistical estimation, testing, and prediction.
 See \citet{Efron2012} for a detailed discussion with related methods and \citet{Wu2012b}, \citet{Schissler2018}; \citet{Schissler2019} for applied examples.
-The following subsections apply \texttt{bigsimr}'s methods to real RNA-seq data, including replicating an estimated parametric structure, MC probability estimation, and MC evaluation of correlation estimation efficiency.
+The following subsections apply \texttt{bigsimr}'s methods to real RNA-seq data, including replicating an estimated parametric structure, probability estimation, and evaluation of correlation estimation efficiency.
 
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> d <- 1000
-R> brca1000 <- example_brca %>%
-+   select(all_of(1:d)) %>%
-+   mutate(across(everything(), as.double))
-\end{CodeInput}
-\end{CodeChunk}
-
-We begin by estimating the structure of the TCGA BRCA RNA-seq data set (see \href{background}{Background}).
+We begin by estimating the structure of the TCGA BRCA RNA-seq data set.
 Ultimately, we will simulate \(B=10,000\) random vectors \({\bf Y}=(Y_1, \ldots, Y_d)^\top\) with \(d=1000\).
-We assume a multivariate negative binomial (MVNB) model as RNA-seq counts are often over-dispersed and correlated.
+We assume a MVNB model as RNA-seq counts are often over-dispersed and correlated.
 Since all \(d\) selected genes exhibit over-dispersion (data not shown), we proceed to estimate the NB parameters \((r_i, p_i), i=1,\ldots,d\), to determine the target marginal PMFs \(f_i\).
-To complete specification of the simulation algorithm inputs, we estimate the Spearman correlation matrix \({ \bf R}_{S}\) to characterize dependency.
-
+To complete specification of the simulation algorithm inputs, we estimate the Spearman correlation matrix \(R_S\) to characterize dependency.
 With this goal in mind, we first estimate the desired correlation matrix using the fast implementation provided by \texttt{bigsimr}:
@@
 Next, we estimate the marginal parameters.
-We use the method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model.
-While, marginal distributions are from the same probability family (NB), they are heterogeneous in terms of the parameters probability and size \((p_i, n_i)\) for \(i,\ldots,d\).
+We use method of moments to estimate the marginal parameters for the multivariate negative binomial model.
+The marginal distributions are from the same probability family (NB), yet they are heterogeneous in terms of the parameters probability and size \((p_i, n_i)\) for \(i,\ldots,d\).
 The functions below support this estimation for later use in \texttt{rvec}.
@@
 
-We apply these estimators to all 1000 genes across the 878 patients:
+We apply these estimators to the highest-expressing 1000 genes across the 878 patients:
 
@@
 Notably, the estimated marginal NB probabilities \(\{ \hat{p}_i \}\) are small --- ranging in the interval \([\ensuremath{7.5305\times 10^{-7}} , \ensuremath{5.6592\times 10^{-4}}]\).
-This gives rise to highly variable counts and, typically, less restriction on potential pairwise correlation pairs.
-Once the functions are defined/executed to complete marginal estimation, we specify targets and generate the desired random vectors using \texttt{rvec}.
-Now we check admissibility of the specified correlation matrix.
+This gives rise to highly variable counts and, thus, less restriction on potential pairwise correlation pairs.
+Given the marginals, we now specify targets and check admissibility of the specified correlation matrix.
 
@@
     Min.  1st Qu.   Median     Mean  3rd Qu.     Max. 
-0.000000 0.000201 0.000432 0.000534 0.000753 0.006931 
+0.000000 0.000193 0.000413 0.000506 0.000717 0.006922 
 \end{CodeOutput}
@@
 
-While the exact \(d \times d\) Spearman correlation matrix is not strictly admissible in our scheme (as seen by the non-positive definite result above), the approximation is close with a maximum absolute error of 0.0069 and average absolute error of \ensuremath{5.3366\times 10^{-4}} across the \ensuremath{4.995\times 10^{5}} correlations.
+As seen above \(R_X\) is not strictly admissible in our scheme.
+This can be seen from the negative result from \texttt{bs\$iscorrelation(R\_X)}, which checks 1) positive-definiteness, 2) symmetry, and 3) 1s along the diagonal.
+However, the approximation is close with a maximum absolute error of 0.0069 and average absolute error of \ensuremath{5.0641\times 10^{-4}} across the \ensuremath{4.995\times 10^{5}} correlations.
+With the inputs configured, \texttt{rvec} is executed to produce a synthetic RNA-seq data set:
 
@@
 Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parametric form and with less extreme points then the observed data in Figure \ref{fig:ch010-realDataFig}.
-Figure \ref{fig:ch050-figBRCA} displays the aggregated results of our simulation by comparing the specified target parameter (horizontal axes) with the corresponding quantities estimated from the simulated data (vertical axes).
-The evaluation shows that the simulated counts approximately match the target parameters and exhibit the full range of estimated correlation from the data.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> set.seed(2020-02-25)
-R> num_genes <- 3
-R> gene_sample <- sample(example_genes[1:1000], num_genes)
-R> 
-R> ggpairs(data = as.data.frame(sim_nbinom[, gene_sample]),
-+         upper = list(continuous = wrap('cor', method = "spearman"))) + 
-+   theme_bw()
-\end{CodeInput}
-\begin{CodeOutput}
-Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
-exact p-value with ties
-\end{CodeOutput}
-\begin{CodeOutput}
-Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
-exact p-value with ties
-\end{CodeOutput}
-\begin{CodeOutput}
-Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
-exact p-value with ties
-\end{CodeOutput}
+Figure \ref{fig:ch050-figBRCA} compares the specified target parameter (horizontal axis) with the corresponding quantities estimated from the simulated data (vertical axis).
+The evaluation shows that the simulated counts approximately match the target parameters and exhibit the full range of estimated correlations from the data.
+
+\begin{CodeChunk}
 \begin{figure}
 
-{\centering \includegraphics{_main_files/figure-latex/ch050-simDataFig-1}
+{\centering \includegraphics[width=0.85\linewidth]{_main_files/figure-latex/ch050-simDataFig-1}
 
@@
 
-\hypertarget{simulation-based-joint-probability-calculations}{%
-\subsection{Simulation-based joint probability calculations}\label{simulation-based-joint-probability-calculations}}
-
-Many statistical tasks require evaluation of a joint probability mass (or density) function:
-
-\[
-P( {\bf Y} = {\bf y} ), \: y_i \in \chi_i.
-\]
-
-where \(\chi_i\) is the sample space for the \(i^{th}\) component of the random vector \(\bf{Y}\).
-Compact representations with convenient computational forms are rare for high-dimensional constructions, especially with heterogeneous, correlated marginal distributions (or margins of mixed data types).
-Given a large number of simulated vectors as produced above, estimated probabilities are readily given by counting the proportion of simulated vectors meeting the desired condition.
-In our motivating application, one may ask what is the probability that all genes expressed greater than a certain threshold value \({\bf y}_0\).
-This can be estimated as
-
-\[
-\hat{P}( {\bf Y} \ge {\bf y_0 } ) = \frac{1}{B} \sum_{b=1}^B I( {\bf Y}^{ (b) } \ge {\bf y_0} ),
-\]
-
-where \({\bf Y}^{(b)}\) is the \(b^{th}\) simulated vector from a total of \(B\) simulation replicates and \(I(\cdot)\) is the indicator function.
-For example, we can estimate from our \(B=10,000\) simulated vectors that the probability of all genes expressing (i.e., \({\bf y}_i \geq 1, \forall \; i\)) is \(0.5979\).
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> d <- ncol(sim_nbinom)
-R> B <- nrow(sim_nbinom)
-R> y0 <- 1
-R> pHat <- mean(apply(sim_nbinom, 1, function(Y) {
-+   all(Y >= y0)
-+ }))
-R> pHat
-\end{CodeInput}
-\begin{CodeOutput}
-[1] 0.5979
-\end{CodeOutput}
-\end{CodeChunk}
-
-\hypertarget{evaluation-of-correlation-estimation-efficiency}{%
-\subsection{Evaluation of correlation estimation efficiency}\label{evaluation-of-correlation-estimation-efficiency}}
-
-MC methods are routinely used in many statistical inferential tasks including estimation, hypothesis testing, error rates, and empirical interval coverage rates.
-To conclude the example applications, we demonstrate how \texttt{bigsimr} can be used to evaluate estimation efficiency.
-In particular, we wish to assess the error in our correlation estimation above.
-We used a conventional method, based on classical statistical theory which was not designed for high-dimensional data.
-Indeed, high-dimensional covariance estimation (and precision matrices) is an active area of statistical science, .e.g., \citep{Won2013g, VanWieringen2016}.
-
-In this small example, we simulate \(m=30\) data sets with the number of simulated vectors matching the number of patients in the BRCA data set, \(N=878\).
-Since our simulation is much faster for the Pearson correlation type (see Figure \ref{fig:ch040-largeDfig}), we only convert the Spearman correlation matrix once (and ensure it is positive definite).
-At each iteration, we estimate the quadratic loss (residual sum of squared errors) from the specified \({\bf R}_{S}\), producing a distribution of loss values.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> # Simulate random vectors equal to the sample size
-R> N <- nrow(example_brca)
-R> # create m random vectors and estimate correlation
-R> simRho <- replicate(n = m, expr = {
-+   tmpSim <- bs$rvec(N , R_X, nb_margins)
-+   bs$cor(tmpSim, bs$Spearman)
-+ }, simplify = FALSE)
-R> # Evaluate the residual sum of squared error
-R> sapply(simRho, function(R) sum((R - R_S)^2))
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> frobenius_loss <- sapply(simRho, function(R) sum((R - R_S)^2))
-\end{CodeInput}
-\end{CodeChunk}
-
-The \texttt{R} summary function supplies the mean-augmented five-number summary of the quadratic loss distribution computed above.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> summary(frobenius_loss)
-\end{CodeInput}
-\begin{CodeOutput}
-   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-    691     829     890     925     986    1224 
-\end{CodeOutput}
-\end{CodeChunk}
-
-This distribution could be compared to high-dimensional designed covariance estimators for guidance on whether the additional complexity and computation time are warranted.
-
 \hypertarget{discussion}{%
@@
 
-We have introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a user-friendly, high-performance \texttt{R} package {[}\texttt{bigsimr}{]}.
+We developed a general-purpose high-dimensional multivariate simulation algorithm and provide a user-friendly, high-performance \texttt{R} package \texttt{bigsimr}, Julia package \texttt{Bigsimr} and Python package \texttt{bigsimr}.
 The random vector generation method is inspired by NORTA \citep{Cario1997} and Gaussian copula-based approaches \citep[\citet{BF17}, \citet{Xia17}]{MB13}.
-The major contributions of this work are methods and software for flexible, scalable simulation of HD multivariate probability distributions with broad potential data analytic applications for modern, big-data statistical computing.
+The major contributions of this work are methods and software for flexible, scalable simulation of HD multivariate probability distributions with broad data analytic applications for modern, big-data statistical computing.
 For example, one could simulate high-resolution time series data, such as those consistent with an auto-regressive moving average model exhibiting a specified Spearman structure.
 Or our methods could be used to simulate sparsely correlated data, as many HD methods assume, via specifying a \emph{spiked correlation matrix}.
-
-It is customary to compare new tools and algorithms directly to existing competing methods and software.
-In this study, however, we only employ our proposed methodology, as our previous work has shown that existing \texttt{R} tools are simply not designed to meet our high-dimensional goal (see \citet{Li2019gpu} for evaluations of the \texttt{R} \texttt{copula} package and others).
-For the bivariate simulations, existing packages such as \href{https://github.com/superdesolator/NORTARA}{\texttt{nortaRA}} work well to match Pearson correlations exactly.
 
@@
 We provide an open-source implementation of our methodology as the \texttt{Bigsimr.jl} Julia package, hosted on github at \url{https://github.com/adknudson/Bigsimr.jl}.
-Additionally we provide R and Python interfaces to \texttt{Bigsimr}, respectively at \url{https://github.com/SchisslerGroup/r-bigsimr} and \url{https://github.com/SchisslerGroup/python-bigsimr}.
+Additionally we provide R and Python interfaces to \texttt{Bigsimr}, respectively on \url{https://github.com/SchisslerGroup/r-bigsimr} and \url{https://github.com/SchisslerGroup/python-bigsimr}.
+A stable release of the R package \texttt{bigsimr} is also available on CRAN.
 \texttt{Bigsimr.jl} is an ongoing project, and feature requests or issues can be submitted to \url{https://github.com/adknudson/Bigsimr.jl/issues}.
@@
 \begin{CodeChunk}
-\begin{CodeInput}
-R> knitr::include_graphics('images/hex-bigsimr.png')
-\end{CodeInput}
 
@@
 \begin{equation}
+\label{eq:kram1}
 (Y_1, Y_2)^\top \stackrel{d}{=} \left(X_1, e^{X_2}\right)^\top,
-\label{eq:kram1}
 \end{equation}
@@
 \end{bmatrix}
-\label{eq:kram2}
 \end{equation}
@@
 \right].
-\label{eq:kram3}
 \end{equation}
@@
 \begin{equation}
+\label{eq:kram4}
 F_1^{-1}(u) = \mu_1+\sigma_1 \Phi^{-1}(u)\,\,\, \mbox{and} \,\,\, F_2^{-1}(u) = e^{\mu_2+\sigma_2 \Phi^{-1}(u)}, 
-\label{eq:kram4}
 \end{equation}
@@
 \begin{equation}
+\label{eq:kram5}
 F_1^{-1}(U_1) = \mu_1+\sigma_1 \Phi^{-1}(\Phi(Z_1)) =  \mu_1+\sigma_1 Z_1 \,\,\, \mbox{and} \,\,\, F_2^{-1}(U_2) = e^{\mu_2+\sigma_2 \Phi^{-1}(\Phi(Z_2))} = e^{\mu_2+\sigma_2 Z_2}, 
-\label{eq:kram5}
 \end{equation}
@@
 where \((X_1, X_2)^\top = (\mu_1+\sigma_1 Z_1, \mu_2+\sigma_2 Z_2)^\top \sim N_2(\boldsymbol \mu, \boldsymbol \Sigma)\).
-Consequently, the vector \((Y_1, Y_2)^\top\) obtained in Step 3 of the algorithm has the structure provided by (\ref{kram1}).
+Consequently, the vector \((Y_1, Y_2)^\top\) obtained in Step 3 of the algorithm has the structure provided by (\ref{eq:kram1}).
 
@@
 
-\noindent Next, we provide the exact covariance structure of the random vector \((Y_1, Y_2)^\top\) given by (\ref{kram1}), and relate the correlation of \(Y_1\) and \(Y_2\) to \(\rho\), which is the correlation of the normal variables \(Z_1\), \(Z_2\) (and also \(X_1\) and \(X_2\)).
+\noindent Next, we provide the exact covariance structure of the random vector \((Y_1, Y_2)^\top\) given by (\ref{eq:kram1}), and relate the correlation of \(Y_1\) and \(Y_2\) to \(\rho\), which is the correlation of the normal variables \(Z_1\), \(Z_2\) (and also \(X_1\) and \(X_2\)).
 A straightforward albeit somewhat tedious algebra produces the following result.
@@
 \begin{equation}
+\label{eq:kram6}
 \boldsymbol \mu_{{\bf Y}} = \left(\mu_1, e^{\mu_2+\sigma_2^2/2}\right)^\top
-\label{eq:kram6}
 \end{equation}
@@
 \end{bmatrix}
-\label{eq:kram7}
 \end{equation}
@@
 \begin{equation}
+\label{eq:covy1y2}
 \mbox{Cov}(Y_1, Y_2) = \mathbb E(Y_1 Y_2) - \mathbb E(Y_1)\mathbb E(Y_2), 
-\label{eq:covy1y2}
 \end{equation}
@@
 \begin{equation}
+\label{eq:kre1}
 \mathbb E(Y_1 Y_2)  =  \mathbb E\left(X_1 e^{X_2}\right)  =  \mathbb E \left\{ \mathbb E \left(X_1 e^{X_2} | X_1 \right) \right\} = \mathbb E \left\{ X_1 \mathbb E \left(e^{X_2} | X_1 \right) \right\}.
-\label{eq:kre1}
-\end{equation}
-
-Further, by using the fact that the conditional distribution of $X_2$ given $X_1=x_1$ is normal with mean $\mathbb E(X_2|X_1=x_1) = \mu_2+\rho\sigma_2(x_1-\mu_1)/\sigma_1$ and variance $\sigma_2^2(1-\rho^2)$, one can relate the inner expectation on the far right in (\ref{kre1}) to the moment generating function $M(t)$ of this conditional distribution evaluated at $t=1$, leading to
-
-\begin{equation}
+\end{equation}
+
+Further, by using the fact that the conditional distribution of $X_2$ given $X_1=x_1$ is normal with mean $\mathbb E(X_2|X_1=x_1) = \mu_2+\rho\sigma_2(x_1-\mu_1)/\sigma_1$ and variance $\sigma_2^2(1-\rho^2)$, one can relate the inner expectation on the far right in (\ref{eq:kre1}) to the moment generating function $M(t)$ of this conditional distribution evaluated at $t=1$, leading to
+
+\begin{equation}
+\label{eq:kre2}
 E \left(e^{X_2} | X_1 \right) = e^{ \mu_2+\rho\sigma_2(x_1-\mu_1)/\sigma_1 + \sigma_2^2(1-\rho^2)/2},
-\label{eq:kre2}
 \end{equation}
@@
 \begin{equation}
+\label{eq:kre3}
 \mathbb E(Y_1 Y_2)  =  e^{ \mu_2+ \frac{1}{2} \sigma_2^2(1-\rho^2)} \mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\}.
-\label{eq:kre3}
-\end{equation}
-
-Since $X_1$ is normal with mean $\mu_1$ and variance $\sigma_1^2$, the expectation in (\ref{kre3}) becomes
-
-\begin{equation}
+\end{equation}
+
+Since $X_1$ is normal with mean $\mu_1$ and variance $\sigma_1^2$, the expectation in (\ref{eq:kre3}) becomes
+
+\begin{equation}
+\label{eq:kre4}
 \mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = \int_{-\infty}^\infty x_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (x_1-\mu_1)} \frac{1}{\sqrt{2\pi}\sigma_1} e^{-\frac{1}{2\sigma_1^2}(x_1-\mu_1)^2}.
-\label{eq:kre4}
 \end{equation}
@@
 \begin{equation}
+\label{eq:kre5}
 \mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = e^{\frac{1}{2}\sigma_2^2\rho^2} \int_{-\infty}^\infty (u+\mu_1)  \frac{1}{\sqrt{2\pi}\sigma_1} e^{-\frac{1}{2\sigma_1^2}(u-\rho\sigma_1\sigma_2)^2}.
-\label{eq:kre5}
-\end{equation}
-
-Since the integral in (\ref{kre5}) is the expectation $\mathbb E(X+\mu_1)$, where $X$ is normal with mean $\rho\sigma_1\sigma_2$ and variance $\sigma_1^2$, we conclude that
-
-\begin{equation}
+\end{equation}
+
+Since the integral in (\ref{eq:kre5}) is the expectation $\mathbb E(X+\mu_1)$, where $X$ is normal with mean $\rho\sigma_1\sigma_2$ and variance $\sigma_1^2$, we conclude that
+
+\begin{equation}
+\label{eq:kre6}
 \mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = e^{\frac{1}{2}\sigma_2^2\rho^2} (\rho\sigma_1\sigma_2 +\mu_1),
-\label{eq:kre6}
-\end{equation}
-
-which, in view of (\ref{kre3}), leads to
-
-\begin{equation}
+\end{equation}
+
+which, in view of (\ref{eq:kre3}), leads to
+
+\begin{equation}
+\label{eq:kre7}
 \mathbb E (Y_1Y_2) = e^{\mu_2 + \frac{1}{2}\sigma_2^2} (\rho\sigma_1\sigma_2 +\mu_1).
-\label{eq:kre7}
-\end{equation}
-
-Finally, (\ref{covy1y2}), along with the expressions for the means of $Y_1$ and $Y_2$, produce the covariance of $Y_1$ and $Y_2$:
-
-\begin{equation}
+\end{equation}
+
+Finally, (\ref{eq:covy1y2}), along with the expressions for the means of $Y_1$ and $Y_2$, produce the covariance of $Y_1$ and $Y_2$:
+
+\begin{equation}
+\label{eq:kre8}
 \mbox{Cov}(Y_1, Y_2) = e^{\mu_2 + \frac{1}{2}\sigma_2^2} (\rho\sigma_1\sigma_2 +\mu_1) - \mu_1 e^{\mu_2 + \frac{1}{2}\sigma_2^2} = \rho\sigma_1\sigma_2 e^{\mu_2 + \frac{1}{2}\sigma_2^2}.
-\label{eq:kre8}
 \end{equation}
@@
 \begin{equation}
+\label{eq:kram8}
 \rho_{Y_1, Y_2} = \rho \frac{\sigma_2}{\sqrt{e^{\sigma_2^2} -1}},
-\label{eq:kram8}
 \end{equation}
@@
 The above result is useful when studying the possible range of correlation of \(Y_1\) and \(Y_2\) in the above example.
-It can be shown that the factor on the far right in (\ref{kram8}) is a monotonically decreasing function of \(\sigma_2\) on \((0,\infty)\), with the limits of 1 and 0 at zero and infinity, respectively.
-Thus, in principle, the range of correlation of \(Y_1\) and \(Y_2\) is the same as that of \(X_1\) and \(X_2\), as the factor on the far right in (\ref{kram8}) can be made arbitrarily close to 1.
+It can be shown that the factor on the far right in (\ref{eq:kram8}) is a monotonically decreasing function of \(\sigma_2\) on \((0,\infty)\), with the limits of 1 and 0 at zero and infinity, respectively.
+Thus, in principle, the range of correlation of \(Y_1\) and \(Y_2\) is the same as that of \(X_1\) and \(X_2\), as the factor on the far right in (\ref{eq:kram8}) can be made arbitrarily close to 1.
 However, by changing this factor we may affect the marginal distributions of \(Y_1\) and \(Y_2\).
-It can be shown that if the marginal distributions of \(Y_1\) and \(Y_2\) are fixed, then the relation (\ref{kram8}) becomes
-
-\begin{equation}
+It can be shown that if the marginal distributions of \(Y_1\) and \(Y_2\) are fixed, then the relation (\ref{eq:kram8}) becomes
+
+\begin{equation}
+\label{eq:kram9}
 \rho_{Y_1, Y_2} = \rho \sqrt{\frac{\log(1+c_2^2)}{c_2^2}},
-\label{eq:kram9}
 \end{equation}
@@
 \begin{equation}
+\label{eq:kram10}
 - \sqrt{\frac{\log(1+c_2^2)}{c_2^2}} \leq \rho_{Y_1, Y_2} \leq \sqrt{\frac{\log(1+c_2^2)}{c_2^2}}. 
-\label{eq:kram10}
 \end{equation}
@@
 \begin{equation}
+\label{eq:kram11}
 \frac{\hat{\sigma}_2}{\sqrt{e^{\hat{\sigma}_2^2} -1}} = 0.881.
-\label{eq:kram11}
-\end{equation}
-
-Thus, the possible range of correlation becomes \((-0.881, 0.881)\). This approximately agrees with the MC results provided.
-
-As an invited speaker at CMStatistics 2020, I was delighted to learn of the opportunity to have our work included in the first issue of CSDA Annals of Statistical Data Science.
-
-On behalf of my co-authors, I'm pleased to submit our manuscript entitled \emph{Simulating High-Dimensional Multivariate Data using the \texttt{Bigsimr} Package} for your consideration.
-We feel that the work is well suited to the indicated themes and hope that it will be considered a valuable contribution to this prestigious journal.
-
-The entire article can be reproduced from code and is housed at \url{https://github.com/SchisslerGroup/article_bigsimr}, which we are glad to make available upon request to the reviewers and, after acceptance, the general public.
-
-I hope that everything is in order. If not, or if you have any additional needs, please do not hesitate to contact me. Thank you and all those involved in this process.
+\end{equation}
+
+Thus, the possible range of correlation becomes \((-0.881, 0.881)\). This agrees with the MC results provided.
 

### 2021-06-22 010-background.Rmd
Revise for JOSS submission. Address Ed's comments.
Revise for JOSS submission. Address Walt's comments.

@@
 
-\noindent where $U$ is a uniform random variable on $(0,1)$ and $F^{-1}_1, F^{-1}_2$ are the inverse CDFs of $X_1$ and $X_2$, respectively, definitely by \@ref(eq:inverseCDF) when the variables are discrete.
+\noindent where $U$ is a uniform random variable on $(0,1)$ and $F^{-1}_1, F^{-1}_2$ are the inverse CDFs of $X_1$ and $X_2$, respectively, defined by \@ref(eq:inverseCDF) when the variables are discrete.
 
@@
 To place `Bigsimr`'s implementation and algorithms in context with existing approaches, here we qualitatively compare to other multivariate simulation packages.
-In `R`, the `MASS` packages `mvnorm()` has long been available to produce random vectors from multivariate normal (and $t$) distributions.
+In `R`, the `MASS` package's `mvnorm()` has long been available to produce random vectors from multivariate normal (and $t$) distributions.
 This procedure cannot simulate from arbitrary margins and dependency measures, as well as fails to readily scale to high dimensions.

### 2021-06-22 020-simulation-algorithm.Rmd
Revise for JOSS submission. Address Walt's comments.

@@
     * Convert the target Pearson correlation matrix to $R_P$ to the corresponding, approximate MVN Pearson correlation $R_X$.
-2. Check admissibility and, if needed, compute the nearest correlation matrix.
+2. Check correlation matrix admissibility and, if not, compute the nearest correlation matrix.
     * Check that $R_X$ is a correlation matrix, a positive definite matrix with 1's along the diagonal.
-    * If $R_X$ is a correlation matrix, the input matrix is *admissible* in this scheme. Otherwise,
-    * Replace $R_X$ with the nearest correlation matrix $\tilde{R}_X$, in the Frobenius norm.
+    * If the mapping produced an $R_X$ that is a correlation matrix, we describe the matrix as **admissible** in this scheme and retain $R_X$. Otherwise,
+    * Replace $R_X$ with the nearest correlation matrix $\tilde{R}_X$, in the Frobenius norm and the result will be approximate.
 3. Gaussian copula

### 2021-06-22 030-bigsimr-package.Rmd
Revise for JOSS submission. Address Walt's comments.

@@
 
-The subsections below describe the basic use of `bigsimr` R package by stepping through an example workflow using the data set `airquality` that contains daily air quality measurements in New York, May to September 1973 [@Chambers1983].
+The subsections below describe the basic use of the `bigsimr` R package by stepping through an example workflow using the data set `airquality` that contains daily air quality measurements in New York, May to September 1973 [@Chambers1983].
 This workflow proceeds from setting up the computing environment, to data wrangling, estimation, simulation configuration, random vector generation, and, finally, result visualization.
@@
 Finally, we execute `rvec` to simulate the desired $10,000$ random vectors from the assumed bivariate distribution of `Ozone` and `Temp`.
-Note that the target input in specified using the pre-computed MVN correlation matrix.
+Note that the target input is specified using the pre-computed MVN correlation matrix.
 Figure \@ref(fig:ch030-plot-sim) plots the 10,000 simulated points.

### 2021-06-22 050-examples.Rmd
Revise for JOSS submission. Address Walt's comments.

@@
 Next, we estimate the marginal parameters.
-We use method of moments to estimate the marginal parameters for the multivariate negative binomial model.
+We use the method of moments to estimate the marginal parameters for the multivariate negative binomial model.
 The marginal distributions are from the same probability family (NB), yet they are heterogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.
@@
 Figure \@ref(fig:ch050-simDataFig) displays the simulated counts and pairwise relationships for our example genes from Table \@ref(tab:ch010-realDataTab).
-Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parametric form and with less extreme points then the observed data in Figure \@ref(fig:ch010-realDataFig).
+Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parametric form and with less extreme points than the observed data in Figure \@ref(fig:ch010-realDataFig).
 Figure \@ref(fig:ch050-figBRCA) compares the specified target parameter (horizontal axis) with the corresponding quantities estimated from the simulated data (vertical axis).

### 2021-06-22 080-appendix.Rmd
Revise for JSS submission pre cross ref fix
Revise for JOSS submission. Address Walt's comments.

@@
 \noindent Consider a bivariate example where we have a correlated $(Y_1, Y_2)^\top$ with $Y_1\sim N(\mu_1, \sigma_1^2)$ (normal) and $Y_2\sim LN(\mu_2, \sigma_2^2)$ (lognormal).
-First, let us note that when we get such a vector from the `bigsimr` package then in fact it can be represented as
+First, let us note that when we generate such a vector from the `bigsimr` package then in fact it can be represented as
 
@@
 \begin{proof}
-The values of the means and the variances are obtained immediately from normal marginal distribution of $Y_1$ and lognormal marginal distribution of $Y_2$.
+The values of the means and the variances are obtained immediately from the normal marginal distribution of $Y_1$ and lognormal marginal distribution of $Y_2$.
 It remains to establish the covariance of $Y_1$ and $Y_2$,
@@
 
-Thus, the possible range of correlation becomes $(-0.881, 0.881)$. This approximately agrees with the MC results provided.
+Thus, the possible range of correlation becomes $(-0.881, 0.881)$. This agrees with the MC results provided.

### 2021-06-22 _book/_main.tex
Revise for JOSS submission. Address Walt's comments.

@@
 
-\noindent where \(U\) is a uniform random variable on \((0,1)\) and \(F^{-1}_1, F^{-1}_2\) are the inverse CDFs of \(X_1\) and \(X_2\), respectively, definitely by \eqref{eq:inverseCDF} when the variables are discrete.
+\noindent where \(U\) is a uniform random variable on \((0,1)\) and \(F^{-1}_1, F^{-1}_2\) are the inverse CDFs of \(X_1\) and \(X_2\), respectively, defined by \eqref{eq:inverseCDF} when the variables are discrete.
 
@@
 To place \texttt{Bigsimr}'s implementation and algorithms in context with existing approaches, here we qualitatively compare to other multivariate simulation packages.
-In \texttt{R}, the \texttt{MASS} package \texttt{mvnorm()} has long been available to produce random vectors from multivariate normal (and \(t\)) distributions.
+In \texttt{R}, the \texttt{MASS} package's \texttt{mvnorm()} has long been available to produce random vectors from multivariate normal (and \(t\)) distributions.
 This procedure cannot simulate from arbitrary margins and dependency measures, as well as fails to readily scale to high dimensions.
@@
 \item
-  Check admissibility and, if needed, compute the nearest correlation matrix.
+  Check correlation matrix admissibility and, if not, compute the nearest correlation matrix.
 
@@
   \item
-    If \(R_X\) is a correlation matrix, the input matrix is \emph{admissible} in this scheme. Otherwise,
-  \item
-    Replace \(R_X\) with the nearest correlation matrix \(\tilde{R}_X\), in the Frobenius norm.
+    If the mapping produced an \(R_X\) that is a correlation matrix, we describe the matrix as \textbf{admissible} in this scheme and retain \(R_X\). Otherwise,
+  \item
+    Replace \(R_X\) with the nearest correlation matrix \(\tilde{R}_X\), in the Frobenius norm and the result will be approximate.
   \end{itemize}

### 2021-06-22 000-introduction.Rmd
removed duplicated ref
Revise for JOSS submission. Address Tom's comments

@@
 
-As many have noted, it can be vexing to simulate dependent, non-normal/discrete data, even for low-dimensional (LD) settings [@MB13; @XZ19].
+As others have noted, it can be vexing to simulate dependent, non-normal/discrete data, even for low-dimensional (LD) settings [@MB13; @XZ19].
 For continuous non-normal LD multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula approaches [@Nelsen2007] are well-studied, with existing software available [@Yan2007; @Chen2001].
@@
 The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc).
-For normal data, the Pearson product-moment correlation describes the dependency perfectly.
-As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix is computationally intense [@Chen2001; @Xia17].
+For normal data, the Pearson product-moment correlation describes the dependence perfectly.
+However, simulating arbitrary random vectors that match a target Pearson correlation matrix is computationally intense [@Chen2001; @Xia17].
 On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependence, such as Spearman's $\rho$ and Kendall's $\tau$.

### 2021-06-22 010-background.Rmd
Revise for JOSS submission. Address Tom's comments

@@
 
-\noindent where $U$ is a uniform random variable on $(0,1)$ and $F^{-1}_1, F^{-1}_2$ are the inverse CDFs of $X_1$ and $X_2$, respectively, defined by \@ref(eq:inverseCDF) when the variables are discrete.
+\noindent where $U$ is a uniform random variable on $(0,1)$ and $F^{-1}_1, F^{-1}_2$ are the inverse CDFs of $X_1$ and $X_2$, respectively, defined as
 
 \begin{equation}
-F_{i}^{-1} = \inf\{y:F_{i}(y) \geq u \}.
+F_{i}^{-1} = \inf\{y:F_{i}(y) \geq u \}
 \label{eq:inverseCDF}
 \end{equation}
+
+when the variables are discrete.
 
@@
 
-A Gaussian copula has marginal CDFs that are all standard normal, $F_i = \Phi, \forall \, i$, where $\Phi(\cdot)$ is the standard normal CDF.
+A Gaussian copula has marginal CDFs that are all standard normal, $F_i = \Phi, \forall \, i$, .
 This representation corresponds to a multivariate normal (MVN) distribution with standard normal marginal distributions and covariance matrix $R_P$.
-Since the marginals are standardized to have unit variance, however, $R_P$ is a Pearson correlation matrix.
+Since the marginals are standardized to have unit variance, this $R_P$ is also Pearson correlation matrix.
 If $F_R$ is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula $C_R$ is defined through
@@
 
-Here the copula function $C_R$ is the familiar multivariate normal CDF of the random vector $(\Phi(X_1), \ldots, \Phi(X_d))$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_P)$.
+where $\Phi(\cdot)$ is the standard normal CDF.
 
-Sklar's Theorem [@Sklar1959; @Ubeda-Flores2017] guarantees that given inverse CDFs $F_i^{-1}$s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions.
+Here, the copula function $C_R$ is the CDF of the random vector $(\Phi(X_1), \ldots, \Phi(X_d))$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_P)$.
+
+Sklar's Theorem [@Ubeda-Flores2017] guarantees that given inverse CDFs $F_i^{-1}$s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions.
 For example, using Gaussian copulas, we can construct a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)^\top$ with $Y_i \sim F_i$, viz.
@@
 In `R`, the `MASS` package's `mvnorm()` has long been available to produce random vectors from multivariate normal (and $t$) distributions.
-This procedure cannot simulate from arbitrary margins and dependency measures, as well as fails to readily scale to high dimensions.
+This procedure cannot simulate from arbitrary margins and dependency measures and it fails to readily scale to high dimensions.
 SAS PROC SIMNORMAL and Python's NumPy `random.multivariate_normal` both do similar tasks.
-Recently, the high-performance `R` package `mvnfast` [@Fasiolo2016] has been released that provides HD multivariate normal generation, but this lacks flexibility in marginal models and dependency modeling.
+Recently, the high-performance `R` package `mvnfast` [@Fasiolo2016] has been released, which provides HD multivariate normal generation.
+However, lacks flexibility in marginal models and dependency modeling.
 For dependent discrete data, LD `R` packages exist, such as `GenOrd`.
 For copula computation, the full-featured `R` package `copula` provides many different copulas and a fairly general LD random generator `rCopula()`.
-High-dimesional random vectors via `copula` takes an impractical amount of computing time [@Li2019gpu].
+High-dimesional random vector generation via `copula` takes an impractical amount of computing time [@Li2019gpu].
 The `nortaRA` R package matches Pearson coefficients near exactly and is reasonable for LD settings.

### 2021-06-22 020-simulation-algorithm.Rmd
Revise for JOSS submission. Address Tom's comments

@@
 
-The well-known NORTA algorithm [@Cario1997] simulates a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$. Specifically, NORTA algorithm proceeds as:
+The well-known NORTA algorithm [@Cario1997] simulates a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$. Specifically, NORTA algorithm proceeds as follows:
 
@@
 4. Set $X$ by $X \gets MZ$.
-5. $\text{Return} \; Y \; \text{where} \; Y_i \gets F_i^{-1}[\Phi(X_i)], \; i=1,...,d$.
+5. $\text{Return} \; {\bf Y} \; \text{where} \; Y_i \gets F_i^{-1}[\Phi(X_i)], \; i=1,...,d$.
 
@@
 
-where $\phi(\cdot)$ is the joint probability distribution function of two correlated standard normal variables.
+where $\phi(\cdot)$ is the joint probability density function of two correlated standard normal variables.
 For HD simulation, exact evaluation may become too costly to enable practical simulation studies.
@@
 Even in LD spaces (e.g., $d=3$), there exist valid multivariate distributions with *feasible* Pearson correlation matrices that NORTA cannot match exactly [@LH75].
-This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined the resultant matrix is indefinite.
+This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined the resultant matrix becomes non-positive definiteness.
 These situations do occur, even using exact analytic calculations.
@@
 
-There is hope, however, as @GH02 also showed that replacing an indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$.
+There is hope, however, as @GH02 also showed that replacing an non-positive definite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$.
 This provides insight that our nearest positive definite (PD) augmented approach has promise to provide adequate accuracy if our input matching scheme returns an indefinite Pearson correlation matrix.
@@
 For the computationally expensive process to match Pearson correlations, we approximate Equation \@ref(eq:pearsonIntegralRelation) for all pairs of margins.
-To this end, we implement the approximation scheme introduced by @XZ19.
+To this end, we implement the approximation scheme introduced by [@XZ19].
 Briefly, the many double integrals of the form in \@ref(eq:pearsonIntegralRelation) are approximated by weighted sums of Hermite polynomials.
@@
 The defacto function in `R` is `Matrix::nearPD`, an alternating projection algorithm due to @higham2002computing.
-As implemented the function fails to scale to HD.
-Instead, we provide the quadratically-convergent algorithm based on the theory of strongly semi-smooth matrix functions (@qi2006quadratically).
+As implemented, the function fails to scale to HD.
+Instead, we provide the quadratically-convergent algorithm based on the theory of strongly semi-smooth matrix functions [@qi2006quadratically].
 The nearest correlation matrix problem can be written down as the following convex optimization problem:  $\mathrm{min} \quad \frac{1}{2} \Vert R_X - X \Vert^2, \quad \mathrm{s.t.} \quad X_{ii} = 1, \quad i = 1, \ldots , n, \quad X \in S_{+}^{n}$.

### 2021-06-22 040-monte-carlo-evaluation.Rmd
Revise for JOSS submission. Address Walt's comments.
Revise for JOSS submission. Address Tom's comments

@@
 Overall, the studies show that our methodology is generally accurate across the entire range of possible correlation for all three dependency measures.
-Our Pearson matching performs nearly as well as Spearman or Kendall, except for a slight increase in error for negative binomial case.
+Our Pearson matching performs nearly as well as Spearman or Kendall, except for a slight increase in error for the negative binomial case.
 
@@
 
-Somewhat surprising, Kendall estimation and nearest PD computation scale poorly compared to Spearman and, even, approximate Pearson matching.
+Somewhat surprisingly, Kendall estimation and nearest PD computation scale poorly compared to Spearman and, even, approximate Pearson matching.
 In our experience, Kendall computation times are sensitive to the number of cores, benefiting from multi-core parallelization.

### 2021-06-22 050-examples.Rmd
Revise for JOSS submission. Address Tom's comments

@@
 And yet, many models do not account for this, leading to major disruptions to the operating characteristics of statistical estimation, testing, and prediction.
-See @Efron2012 for a detailed discussion with related methods and @Wu2012b, @Schissler2018; @Schissler2019 for applied examples.
 The following subsections apply `bigsimr`'s methods to real RNA-seq data, including replicating an estimated parametric structure, probability estimation, and evaluation of correlation estimation efficiency.
@@
 
-As seen above $R_X$ is not strictly admissible in our scheme.
-This can be seen from the negative result from `bs$iscorrelation(R_X)`, which checks 1) positive-definiteness, 2) symmetry, and 3) 1s along the diagonal.
+As seen above, $R_X$ is not strictly admissible in our scheme.
+This can be seen from the negative result from `bs$iscorrelation(R_X)`, which checks if $R_X$ is positive-definite with 1's along the diagonal.
 However, the approximation is close with a maximum absolute error of `r max( abs(targets - approximates))` and average absolute error of `r mean(abs(targets - approximates))` across the `r choose(d,2)` correlations.

### 2021-06-22 060-conclusion-discussion.Rmd
Revise for JOSS submission. Address Tom's comments

@@
 For example, one could simulate high-resolution time series data, such as those consistent with an auto-regressive moving average model exhibiting a specified Spearman structure.
-Or our methods could be used to simulate sparsely correlated data, as many HD methods assume, via specifying a *spiked correlation matrix*.
+Our methods could also be used to simulate sparsely correlated data, as many HD methods assume, via specifying a *spiked correlation matrix*.
 
@@
 We could only investigate selected multivariate distributions in our Monte Carlo studies.
-There may be instances that the methods do not perform well.
+There may be instances where the methods do not perform well.
 Along those lines, we expect that correlation values close to the boundary of the feasible region could result in algorithm failure.

### 2021-06-22 _book/_main.tex
Revise for JOSS submission. Address Tom's comments

@@
 The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc).
-For normal data, the Pearson product-moment correlation describes the dependency perfectly.
-As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix is computationally intense \citep{Chen2001, Xia17}.
+For normal data, the Pearson product-moment correlation describes the dependence perfectly.
+However, simulating arbitrary random vectors that match a target Pearson correlation matrix is computationally intense \citep{Chen2001, Xia17}.
 On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependence, such as Spearman's \(\rho\) and Kendall's \(\tau\).
@@
 
-\noindent where \(U\) is a uniform random variable on \((0,1)\) and \(F^{-1}_1, F^{-1}_2\) are the inverse CDFs of \(X_1\) and \(X_2\), respectively, defined by \eqref{eq:inverseCDF} when the variables are discrete.
-
-\begin{equation}
-F_{i}^{-1} = \inf\{y:F_{i}(y) \geq u \}.
+\noindent where \(U\) is a uniform random variable on \((0,1)\) and \(F^{-1}_1, F^{-1}_2\) are the inverse CDFs of \(X_1\) and \(X_2\), respectively, defined as
+
+\begin{equation}
+F_{i}^{-1} = \inf\{y:F_{i}(y) \geq u \}
 \label{eq:inverseCDF}
 \end{equation}
+
+when the variables are discrete.
 
@@
 
-A Gaussian copula has marginal CDFs that are all standard normal, \(F_i = \Phi, \forall \, i\), where \(\Phi(\cdot)\) is the standard normal CDF.
+A Gaussian copula has marginal CDFs that are all standard normal, \(F_i = \Phi, \forall \, i\), .
 This representation corresponds to a multivariate normal (MVN) distribution with standard normal marginal distributions and covariance matrix \(R_P\).
-Since the marginals are standardized to have unit variance, however, \(R_P\) is a Pearson correlation matrix.
+Since the marginals are standardized to have unit variance, this \(R_P\) is also Pearson correlation matrix.
 If \(F_R\) is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula \(C_R\) is defined through
@@
 
-Here the copula function \(C_R\) is the familiar multivariate normal CDF of the random vector \((\Phi(X_1), \ldots, \Phi(X_d))\), where \((X_1, \ldots, X_d) \sim N_d({\bf 0}, R_P)\).
-
-Sklar's Theorem \citep{Sklar1959, Ubeda-Flores2017} guarantees that given inverse CDFs \(F_i^{-1}\)s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions.
+where \(\Phi(\cdot)\) is the standard normal CDF.
+
+Here, the copula function \(C_R\) is the CDF of the random vector \((\Phi(X_1), \ldots, \Phi(X_d))\), where \((X_1, \ldots, X_d) \sim N_d({\bf 0}, R_P)\).
+
+Sklar's Theorem \citep{Ubeda-Flores2017} guarantees that given inverse CDFs \(F_i^{-1}\)s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions.
 For example, using Gaussian copulas, we can construct a random vector \({\bf Y} = (Y_1, \ldots, Y_d)^\top\) with \(Y_i \sim F_i\), viz.
@@
 In \texttt{R}, the \texttt{MASS} package's \texttt{mvnorm()} has long been available to produce random vectors from multivariate normal (and \(t\)) distributions.
-This procedure cannot simulate from arbitrary margins and dependency measures, as well as fails to readily scale to high dimensions.
+This procedure cannot simulate from arbitrary margins and dependency measures and it fails to readily scale to high dimensions.
 SAS PROC SIMNORMAL and Python's NumPy \texttt{random.multivariate\_normal} both do similar tasks.
-Recently, the high-performance \texttt{R} package \texttt{mvnfast} \citep{Fasiolo2016} has been released that provides HD multivariate normal generation, but this lacks flexibility in marginal models and dependency modeling.
+Recently, the high-performance \texttt{R} package \texttt{mvnfast} \citep{Fasiolo2016} has been released, which provides HD multivariate normal generation.
+However, lacks flexibility in marginal models and dependency modeling.
 For dependent discrete data, LD \texttt{R} packages exist, such as \texttt{GenOrd}.
 For copula computation, the full-featured \texttt{R} package \texttt{copula} provides many different copulas and a fairly general LD random generator \texttt{rCopula()}.
-High-dimesional random vectors via \texttt{copula} takes an impractical amount of computing time \citep{Li2019gpu}.
+High-dimesional random vector generation via \texttt{copula} takes an impractical amount of computing time \citep{Li2019gpu}.
 The \texttt{nortaRA} R package matches Pearson coefficients near exactly and is reasonable for LD settings.
@@
 
-The well-known NORTA algorithm \citep{Cario1997} simulates a random vector \(\bf Y\) with variance-covariance matrix \(\Sigma_{\bf Y}\). Specifically, NORTA algorithm proceeds as:
+The well-known NORTA algorithm \citep{Cario1997} simulates a random vector \(\bf Y\) with variance-covariance matrix \(\Sigma_{\bf Y}\). Specifically, NORTA algorithm proceeds as follows:
 
@@
 \item
-  \(\text{Return} \; Y \; \text{where} \; Y_i \gets F_i^{-1}[\Phi(X_i)], \; i=1,...,d\).
+  \(\text{Return} \; {\bf Y} \; \text{where} \; Y_i \gets F_i^{-1}[\Phi(X_i)], \; i=1,...,d\).
 \end{enumerate}
@@
 
-where \(\phi(\cdot)\) is the joint probability distribution function of two correlated standard normal variables.
+where \(\phi(\cdot)\) is the joint probability density function of two correlated standard normal variables.
 For HD simulation, exact evaluation may become too costly to enable practical simulation studies.
@@
 Even in LD spaces (e.g., \(d=3\)), there exist valid multivariate distributions with \emph{feasible} Pearson correlation matrices that NORTA cannot match exactly \citep{LH75}.
-This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined the resultant matrix is indefinite.
+This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined the resultant matrix becomes non-positive definiteness.
 These situations do occur, even using exact analytic calculations.
@@
 
-There is hope, however, as \citet{GH02} also showed that replacing an indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate \(d\).
+There is hope, however, as \citet{GH02} also showed that replacing an non-positive definite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate \(d\).
 This provides insight that our nearest positive definite (PD) augmented approach has promise to provide adequate accuracy if our input matching scheme returns an indefinite Pearson correlation matrix.
@@
 For the computationally expensive process to match Pearson correlations, we approximate Equation \eqref{eq:pearsonIntegralRelation} for all pairs of margins.
-To this end, we implement the approximation scheme introduced by \citet{XZ19}.
+To this end, we implement the approximation scheme introduced by \citep{XZ19}.
 Briefly, the many double integrals of the form in \eqref{eq:pearsonIntegralRelation} are approximated by weighted sums of Hermite polynomials.
@@
 The defacto function in \texttt{R} is \texttt{Matrix::nearPD}, an alternating projection algorithm due to \citet{higham2002computing}.
-As implemented the function fails to scale to HD.
-Instead, we provide the quadratically-convergent algorithm based on the theory of strongly semi-smooth matrix functions (\citet{qi2006quadratically}).
+As implemented, the function fails to scale to HD.
+Instead, we provide the quadratically-convergent algorithm based on the theory of strongly semi-smooth matrix functions \citep{qi2006quadratically}.
 The nearest correlation matrix problem can be written down as the following convex optimization problem: \(\mathrm{min} \quad \frac{1}{2} \Vert R_X - X \Vert^2, \quad \mathrm{s.t.} \quad X_{ii} = 1, \quad i = 1, \ldots , n, \quad X \in S_{+}^{n}\).
@@
 
-The subsections below describe the basic use of \texttt{bigsimr} R package by stepping through an example workflow using the data set \texttt{airquality} that contains daily air quality measurements in New York, May to September 1973 \citep{Chambers1983}.
+The subsections below describe the basic use of the \texttt{bigsimr} R package by stepping through an example workflow using the data set \texttt{airquality} that contains daily air quality measurements in New York, May to September 1973 \citep{Chambers1983}.
 This workflow proceeds from setting up the computing environment, to data wrangling, estimation, simulation configuration, random vector generation, and, finally, result visualization.
@@
 Finally, we execute \texttt{rvec} to simulate the desired \(10,000\) random vectors from the assumed bivariate distribution of \texttt{Ozone} and \texttt{Temp}.
-Note that the target input in specified using the pre-computed MVN correlation matrix.
+Note that the target input is specified using the pre-computed MVN correlation matrix.
 Figure \ref{fig:ch030-plot-sim} plots the 10,000 simulated points.
@@
 And yet, many models do not account for this, leading to major disruptions to the operating characteristics of statistical estimation, testing, and prediction.
-See \citet{Efron2012} for a detailed discussion with related methods and \citet{Wu2012b}, \citet{Schissler2018}; \citet{Schissler2019} for applied examples.
 The following subsections apply \texttt{bigsimr}'s methods to real RNA-seq data, including replicating an estimated parametric structure, probability estimation, and evaluation of correlation estimation efficiency.
@@
 Next, we estimate the marginal parameters.
-We use method of moments to estimate the marginal parameters for the multivariate negative binomial model.
+We use the method of moments to estimate the marginal parameters for the multivariate negative binomial model.
 The marginal distributions are from the same probability family (NB), yet they are heterogeneous in terms of the parameters probability and size \((p_i, n_i)\) for \(i,\ldots,d\).
@@
 
-As seen above \(R_X\) is not strictly admissible in our scheme.
-This can be seen from the negative result from \texttt{bs\$iscorrelation(R\_X)}, which checks 1) positive-definiteness, 2) symmetry, and 3) 1s along the diagonal.
+As seen above, \(R_X\) is not strictly admissible in our scheme.
+This can be seen from the negative result from \texttt{bs\$iscorrelation(R\_X)}, which checks if \(R_X\) is positive-definite with 1's along the diagonal.
 However, the approximation is close with a maximum absolute error of 0.0069 and average absolute error of \ensuremath{5.0641\times 10^{-4}} across the \ensuremath{4.995\times 10^{5}} correlations.
@@
 Figure \ref{fig:ch050-simDataFig} displays the simulated counts and pairwise relationships for our example genes from Table \ref{tab:ch010-realDataTab}.
-Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parametric form and with less extreme points then the observed data in Figure \ref{fig:ch010-realDataFig}.
+Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parametric form and with less extreme points than the observed data in Figure \ref{fig:ch010-realDataFig}.
 Figure \ref{fig:ch050-figBRCA} compares the specified target parameter (horizontal axis) with the corresponding quantities estimated from the simulated data (vertical axis).
@@
 For example, one could simulate high-resolution time series data, such as those consistent with an auto-regressive moving average model exhibiting a specified Spearman structure.
-Or our methods could be used to simulate sparsely correlated data, as many HD methods assume, via specifying a \emph{spiked correlation matrix}.
+Our methods could also be used to simulate sparsely correlated data, as many HD methods assume, via specifying a \emph{spiked correlation matrix}.
 
@@
 \noindent Consider a bivariate example where we have a correlated \((Y_1, Y_2)^\top\) with \(Y_1\sim N(\mu_1, \sigma_1^2)\) (normal) and \(Y_2\sim LN(\mu_2, \sigma_2^2)\) (lognormal).
-First, let us note that when we get such a vector from the \texttt{bigsimr} package then in fact it can be represented as
+First, let us note that when we generate such a vector from the \texttt{bigsimr} package then in fact it can be represented as
 

### 2021-06-25 000-introduction.Rmd
Submitting to JSS, final pre-submission edits

@@
 
-Massive high-dimensional (HD) data sets are now commonplace in many areas of scientific inquiry.
+Massive high-dimensional (HD) data sets are now common in many areas of scientific inquiry.
 As new methods are developed for data analysis, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of proposed methodology --- such as false positive rates, statistical power, interval coverage, and robustness.
@@
 As others have noted, it can be vexing to simulate dependent, non-normal/discrete data, even for low-dimensional (LD) settings [@MB13; @XZ19].
-For continuous non-normal LD multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula approaches [@Nelsen2007] are well-studied, with existing software available [@Yan2007; @Chen2001].
+For continuous non-normal LD multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula approaches [@Nelsen2007] are well-studied and implemented in publicly-available software [@Yan2007; @Chen2001].
 Yet these approaches do not scale in a timely fashion to HD problems [@Li2019gpu].
-For discrete data, early simulation strategies had major flaws, such as failing to obtain the full range of possible correlations (e.g., admitting only positive correlations: see @Park1996).
+For discrete data, early simulation strategies had major flaws, such as failing to obtain the full range of possible correlations --- such as admitting only positive correlations [@Park1996].
 While more recent approaches [@MB13; @Xia17; @BF17] have remedied this issue for LD problems, the existing tools are not designed to scale to high dimensions.
@@
 Another central issue lies in characterizing dependence between components in the HD random vector.
-The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc).
+The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc.).
 For normal data, the Pearson product-moment correlation describes the dependence perfectly.
@@
 On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependence, such as Spearman's $\rho$ and Kendall's $\tau$.
-Throughout, we focus on matching these nonparametric dependence measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology.
+Throughout, we focus on matching these nonparametric dependence measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties enabling our proposed methodology.
 We do, however, implement Pearson matching, but several layers of approximation are required.

### 2021-06-25 010-background.Rmd
Submitting to JSS, final pre-submission edits

@@
 
-* SP1: Procedure must scale to high dimensions with practical compute times.
+* SP1: Procedure must scale to high dimensions with practical computation times.
 * SP2: Procedure must scale to high dimensions while maintaining accuracy.
@@
 
-To place `Bigsimr`'s implementation and algorithms in context with existing approaches, here we qualitatively compare to other multivariate simulation packages.
+To place `Bigsimr`'s implementation and algorithms in context of existing approaches, we qualitatively compare `Bbigsimr` to other multivariate simulation packages.
 In `R`, the `MASS` package's `mvnorm()` has long been available to produce random vectors from multivariate normal (and $t$) distributions.
@@
 The `nortaRA` R package matches Pearson coefficients near exactly and is reasonable for LD settings.
-Finally, we know of no other `R` package that can produce HD random vectors with flexible margins and dependence modeling.
+Finally, we know of no other `R` package that can produce HD random vectors with flexible margins and dependence structure.
 Table \@ref(tab:compare-table) summarizes the above discussion.

### 2021-06-25 020-simulation-algorithm.Rmd
Submitting to JSS, final pre-submission edits

@@
 
-This section describes our methods involved in simulating a random vector $\bf Y$ with $Y_i$ components for $i=1,\ldots,d$.
+This section describes our methods for simulating a random vector $\bf Y$ with $Y_i$ components for $i=1,\ldots,d$.
 Each $Y_i$ has a specified marginal CDF $F_i$ and its inverse $F^{-1}_i$.
@@
 Even in LD spaces (e.g., $d=3$), there exist valid multivariate distributions with *feasible* Pearson correlation matrices that NORTA cannot match exactly [@LH75].
-This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined the resultant matrix becomes non-positive definiteness.
+This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined, the resultant matrix becomes non-positive definite.
 These situations do occur, even using exact analytic calculations.
@@
 There is hope, however, as @GH02 also showed that replacing an non-positive definite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$.
-This provides insight that our nearest positive definite (PD) augmented approach has promise to provide adequate accuracy if our input matching scheme returns an indefinite Pearson correlation matrix.
+This provides evidence that our nearest positive definite (PD) augmented approach has promise to provide adequate accuracy if our input matching scheme returns an indefinite Pearson correlation matrix.
 
@@
 In our experience, the mapping step for large $d$ almost always produces a $R_X$ that is not a correlation matrix.
-In Section \@ref(package) we provide a basic description of how to quantify and control the approximation error.
+In Section \@ref(package), we provide a basic method of how to quantify and control the approximation error.
 Further, the RNA-seq example in Section \@ref(examples) provides an illustration of this in practice.

### 2021-06-25 040-monte-carlo-evaluation.Rmd
Submitting to JSS, final pre-submission edits

@@
 This process is both fast and exact.
-On the other hand, accurate and computationally efficent dependency matching presents a challenge, especially in the HD setting.
+On the other hand, accurate and computationally efficient dependency matching presents a challenge, especially in the HD setting.
 To evaluate our methods in those respects, we design the following numerical experiments to first assess accuracy of matching dependency parameters in bivariate simulations and then time the procedure in increasingly large dimension $d$.
@@
 We select bivariate simulation configurations to ultimately simulate our motivating discrete-valued RNA-seq example.
-Thus, we proceed in increasing departure from normality, leading to a multivariate negative binomial (MVNB) model for our motivating data.
+Thus, we proceed by increasing departure from normality, leading to a multivariate negative binomial (MVNB) model for our motivating data.
 We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson, Spearman, and Kendall --- in identical, bivariate marginal configurations.
-For each pair of identical margins, we vary the target correlation across $\Omega$, the set of possible admissible values for each correlation type, to evaluate the simulation's ability to simulate all possible correlations.
+For each pair of identical margins, we vary the target correlation across $\Omega$, the set of possible admissible values for each correlation type, to evaluate the simulation's ability to generate all possible correlations.
 The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (non-normal and discrete).
@@
 Table \@ref(tab:sims) lists our identical-marginal, bivariate simulation configurations.
-We increase the simulate replicates $B$ to visually ensure that our results converge to the target correlations and gauge efficiency.
+We increase the simulation replicates $B$ to visually ensure that our results converge to the target correlations and gauge efficiency.
 We select distributions beginning with a standard multivariate normal (MVN) as we expect the performance to be exact (up to MC error) for all correlation types.

### 2021-06-25 070-supplementary-materials.Rmd
Submitting to JSS, final pre-submission edits

@@
 
-We provide an open-source implementation of our methodology as the `Bigsimr.jl` Julia package, hosted on github at https://github.com/adknudson/Bigsimr.jl.
+We provide an open-source implementation of our methodology as the `Bigsimr.jl` Julia package, hosted on github at https://github.com/SchisslerGroup/Bigsimr.jl.
 Additionally we provide R and Python interfaces to `Bigsimr`, respectively on https://github.com/SchisslerGroup/r-bigsimr and https://github.com/SchisslerGroup/python-bigsimr.
-A stable release of the R package `bigsimr` is also available on CRAN.
+A stable release of the R package `bigsimr` is available on CRAN.
 `Bigsimr.jl` is an ongoing project, and feature requests or issues can be submitted to https://github.com/adknudson/Bigsimr.jl/issues.

### 2021-06-25 index.Rmd
Submitting to JSS, final pre-submission edits

@@
   It is critical to accurately simulate data when employing Monte Carlo techniques and evaluating statistical methodology.
-Measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments.
-Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations.
-To meet this gap, we developed the \pkg{Bigsimr} Julia package with R and Python interfaces.
+Measurements are often correlated and high dimensional in this era of big data, such as data obtained in high-throughput biomedical experiments.
+Due to the computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations.
+To close this gap, we developed the \pkg{Bigsimr} Julia package with R and Python interfaces.
 These packages empower high-dimensional random vector simulation with arbitrary marginal distributions and dependency via a Pearson, Spearman, or Kendall correlation matrix.

### 2021-06-28 _book/bigsimr_jss_submission_v3.tex
Submitting to JSS, final pre-submission edits v2

--- 2021-06-22 _book/_main.tex
@@
 It is critical to accurately simulate data when employing Monte Carlo techniques and evaluating statistical methodology.
-Measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments.
-Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations.
-To meet this gap, we developed the \pkg{Bigsimr} Julia package with R and Python interfaces.
+Measurements are often correlated and high dimensional in this era of big data, such as data obtained in high-throughput biomedical experiments.
+Due to the computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations.
+To close this gap, we developed the \pkg{Bigsimr} Julia package with R and Python interfaces.
 These packages empower high-dimensional random vector simulation with arbitrary marginal distributions and dependency via a Pearson, Spearman, or Kendall correlation matrix.
@@
 As others have noted, it can be vexing to simulate dependent, non-normal/discrete data, even for low-dimensional (LD) settings \citep{MB13, XZ19}.
-For continuous non-normal LD multivariate data, the well-known NORmal To Anything (NORTA) algorithm \citep{Cario1997} and other copula approaches \citep{Nelsen2007} are well-studied, with existing software available \citep{Yan2007, Chen2001}.
+For continuous non-normal LD multivariate data, the well-known NORmal To Anything (NORTA) algorithm \citep{Cario1997} and other copula approaches \citep{Nelsen2007} are well-studied and implemented in publicly-available software \citep{Yan2007, Chen2001}.
 Yet these approaches do not scale in a timely fashion to HD problems \citep{Li2019gpu}.
-For discrete data, early simulation strategies had major flaws, such as failing to obtain the full range of possible correlations (e.g., admitting only positive correlations: see \citet{Park1996}).
+For discrete data, early simulation strategies had major flaws, such as failing to obtain the full range of possible correlations --- such as admitting only positive correlations \citep{Park1996}.
 While more recent approaches \citep{MB13, Xia17, BF17} have remedied this issue for LD problems, the existing tools are not designed to scale to high dimensions.
@@
 On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependence, such as Spearman's \(\rho\) and Kendall's \(\tau\).
-Throughout, we focus on matching these nonparametric dependence measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology.
+Throughout, we focus on matching these nonparametric dependence measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties enabling our proposed methodology.
 We do, however, implement Pearson matching, but several layers of approximation are required.
@@
 \item
-  SP1: Procedure must scale to high dimensions with practical compute times.
+  SP1: Procedure must scale to high dimensions with practical computation times.
 \item
@@
 
-To place \texttt{Bigsimr}'s implementation and algorithms in context with existing approaches, here we qualitatively compare to other multivariate simulation packages.
+To place \texttt{Bigsimr}'s implementation and algorithms in context of existing approaches, we qualitatively compare \texttt{Bbigsimr} to other multivariate simulation packages.
 In \texttt{R}, the \texttt{MASS} package's \texttt{mvnorm()} has long been available to produce random vectors from multivariate normal (and \(t\)) distributions.
@@
 The \texttt{nortaRA} R package matches Pearson coefficients near exactly and is reasonable for LD settings.
-Finally, we know of no other \texttt{R} package that can produce HD random vectors with flexible margins and dependence modeling.
+Finally, we know of no other \texttt{R} package that can produce HD random vectors with flexible margins and dependence structure.
 Table \ref{tab:compare-table} summarizes the above discussion.
@@
 
-This section describes our methods involved in simulating a random vector \(\bf Y\) with \(Y_i\) components for \(i=1,\ldots,d\).
+This section describes our methods for simulating a random vector \(\bf Y\) with \(Y_i\) components for \(i=1,\ldots,d\).
 Each \(Y_i\) has a specified marginal CDF \(F_i\) and its inverse \(F^{-1}_i\).
@@
 Even in LD spaces (e.g., \(d=3\)), there exist valid multivariate distributions with \emph{feasible} Pearson correlation matrices that NORTA cannot match exactly \citep{LH75}.
-This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined the resultant matrix becomes non-positive definiteness.
+This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined, the resultant matrix becomes non-positive definite.
 These situations do occur, even using exact analytic calculations.
@@
 There is hope, however, as \citet{GH02} also showed that replacing an non-positive definite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate \(d\).
-This provides insight that our nearest positive definite (PD) augmented approach has promise to provide adequate accuracy if our input matching scheme returns an indefinite Pearson correlation matrix.
+This provides evidence that our nearest positive definite (PD) augmented approach has promise to provide adequate accuracy if our input matching scheme returns an indefinite Pearson correlation matrix.
 
@@
 In our experience, the mapping step for large \(d\) almost always produces a \(R_X\) that is not a correlation matrix.
-In Section \ref{package} we provide a basic description of how to quantify and control the approximation error.
+In Section \ref{package}, we provide a basic method of how to quantify and control the approximation error.
 Further, the RNA-seq example in Section \ref{examples} provides an illustration of this in practice.
@@
 For the assumed marginals, the Pearson correlation coefficient is not free to vary in {[}-1,1{]}, as seen below.
-Our MC estimate of the bounds slightly underestimates the theoretic bounds of \((-0.881, 0.881)\).
+Our MC estimate of the bounds slightly underestimates the theoretical bounds of \((-0.881, 0.881)\).
 (See an analytic derivation presented in the Appendix).
@@
 This process is both fast and exact.
-On the other hand, accurate and computationally efficent dependency matching presents a challenge, especially in the HD setting.
+On the other hand, accurate and computationally efficient dependency matching presents a challenge, especially in the HD setting.
 To evaluate our methods in those respects, we design the following numerical experiments to first assess accuracy of matching dependency parameters in bivariate simulations and then time the procedure in increasingly large dimension \(d\).
@@
 We select bivariate simulation configurations to ultimately simulate our motivating discrete-valued RNA-seq example.
-Thus, we proceed in increasing departure from normality, leading to a multivariate negative binomial (MVNB) model for our motivating data.
+Thus, we proceed by increasing departure from normality, leading to a multivariate negative binomial (MVNB) model for our motivating data.
 We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson, Spearman, and Kendall --- in identical, bivariate marginal configurations.
-For each pair of identical margins, we vary the target correlation across \(\Omega\), the set of possible admissible values for each correlation type, to evaluate the simulation's ability to simulate all possible correlations.
+For each pair of identical margins, we vary the target correlation across \(\Omega\), the set of possible admissible values for each correlation type, to evaluate the simulation's ability to generate all possible correlations.
 The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (non-normal and discrete).
@@
 Table \ref{tab:sims} lists our identical-marginal, bivariate simulation configurations.
-We increase the simulate replicates \(B\) to visually ensure that our results converge to the target correlations and gauge efficiency.
+We increase the simulation replicates \(B\) to visually ensure that our results converge to the target correlations and gauge efficiency.
 We select distributions beginning with a standard multivariate normal (MVN) as we expect the performance to be exact (up to MC error) for all correlation types.
@@
 
-We provide an open-source implementation of our methodology as the \texttt{Bigsimr.jl} Julia package, hosted on github at \url{https://github.com/adknudson/Bigsimr.jl}.
+We provide an open-source implementation of our methodology as the \texttt{Bigsimr.jl} Julia package, hosted on github at \url{https://github.com/SchisslerGroup/Bigsimr.jl}.
 Additionally we provide R and Python interfaces to \texttt{Bigsimr}, respectively on \url{https://github.com/SchisslerGroup/r-bigsimr} and \url{https://github.com/SchisslerGroup/python-bigsimr}.
-A stable release of the R package \texttt{bigsimr} is also available on CRAN.
+A stable release of the R package \texttt{bigsimr} is available on CRAN.
 \texttt{Bigsimr.jl} is an ongoing project, and feature requests or issues can be submitted to \url{https://github.com/adknudson/Bigsimr.jl/issues}.

### 2021-10-07 article_Bigsimr/article_Bigsimr.Rmd
rewriting as a single rmd to simplify pdf production

--- 2021-06-25 000-introduction.Rmd
@@
+---
+documentclass: jss
+author:
+  - name: Alfred G. Schissler
+    affiliation: University of Nevada
+    address: |
+      | 1664 N Virginia St.
+      | Reno, NV 89557
+    email: \email{aschissler@unr.edu}
+  - name: Edward J. Bedrick
+    affiliation: University of Arizona
+  - name: Alexander D. Knudson
+    affiliation: University of Nevada \AND
+  - name: Tomasz J. Kozubowski
+    affiliation: University of Nevada
+  - name: Tin Nguyen
+    affiliation: University of Nevada
+  - name: Anna K. Panorska
+    affiliation: University of Nevada \AND
+  - name: Juli Petereit
+    affiliation: University of Nevada
+  - name: Walter W. Piegorsch
+    affiliation: University of Arizona
+  - name: Duc Tran
+    affiliation: University of Nevada
+title:
+  formatted: "Simulating High-Dimensional Multivariate Data using the \\pkg{Bigsimr} Package"
+  plain:     "Simulating High-Dimensional Multivariate Data using the Bigsimr Package"
+  short:     "\\pkg{Bigsimr}: Simulate High-Dimensional Multivariate Data"
+abstract: >
+  It is critical to accurately simulate data when employing Monte Carlo techniques and evaluating statistical methodology.
+Measurements are often correlated and high dimensional in this era of big data, such as data obtained in high-throughput biomedical experiments.
+Due to the computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations.
+To close this gap, we developed the \pkg{Bigsimr} Julia package with R and Python interfaces.
+These packages empower high-dimensional random vector simulation with arbitrary marginal distributions and dependency via a Pearson, Spearman, or Kendall correlation matrix.
+\pkg{Bigsimr} contains high-performance features, including multi-core and graphical-processing-unit-accelerated algorithms to estimate correlation and compute the nearest correlation matrix.
+Monte Carlo studies quantify the accuracy and scalability of our approach, up to $d=10,000$.
+We describe example workflows and apply to a high-dimensional data set --- RNA-sequencing data obtained from breast cancer tumor samples.
+keywords:
+  formatted: [multivariate simulation, high-dimensional data, nonparametric correlation, Gaussian copula, RNA-sequencing data, breast cancer]
+  plain:     [multivariate simulation, high-dimensional data, nonparametric correlation, Gaussian copula, RNA-sequencing data, breast cancer]
+preamble: >
+  \usepackage{amsmath}
+output: rticles::jss_article
+bibliography: [bigsimr.bib, packages.bib, alex.bib]
+---
+
+```{r, setup, include=FALSE}
+options(prompt = 'R> ', continue = '+ ')
+```
+
+\clearpage
+
 # Introduction
-
-\setstretch{2.0}
 

### 2021-10-07 article_Bigsimr/article_Bigsimr.tex
rewriting as a single rmd to simplify pdf production

--- 2021-06-28 _book/bigsimr_jss_submission_v3.tex
@@
 \author{
-Alfred G.
-Schissler\\University of Nevada \And Edward J.
-Bedrick\\University of Arizona \And Alexander D.
-Knudson\\University of Nevada \AND Tomasz J.
-Kozubowski\\University of Nevada \And Tin Nguyen\\University of Nevada \And Anna K.
-Panorska\\University of Nevada \AND Juli Petereit\\University of Nevada \And Walter W.
+Alfred G. Schissler\\University of Nevada \And Edward J.
+Bedrick\\University of Arizona \And Alexander D. Knudson\\University of
+Nevada \AND Tomasz J. Kozubowski\\University of Nevada \And Tin
+Nguyen\\University of Nevada \And Anna K. Panorska\\University of Nevada
+\AND Juli Petereit\\University of Nevada \And Walter W.
 Piegorsch\\University of Arizona \And Duc Tran\\University of Nevada
 }
-\title{Simulating High-Dimensional Multivariate Data using the \pkg{Bigsimr} Package}
+\title{Simulating High-Dimensional Multivariate Data using the
+\pkg{Bigsimr} Package}
 
-\Plainauthor{Alfred G. Schissler, Edward J. Bedrick, Alexander D. Knudson, Tomasz J. Kozubowski, Tin Nguyen, Anna K. Panorska, Juli Petereit, Walter W. Piegorsch, Duc Tran}
-\Plaintitle{Simulating High-Dimensional Multivariate Data using the Bigsimr Package}
+\Plainauthor{Alfred G. Schissler, Edward J. Bedrick, Alexander D.
+Knudson, Tomasz J. Kozubowski, Tin Nguyen, Anna K. Panorska, Juli
+Petereit, Walter W. Piegorsch, Duc Tran}
+\Plaintitle{Simulating High-Dimensional Multivariate Data using the
+Bigsimr Package}
 \Shorttitle{\pkg{Bigsimr}: Simulate High-Dimensional Multivariate Data}
@@
 \Abstract{
-It is critical to accurately simulate data when employing Monte Carlo techniques and evaluating statistical methodology.
-Measurements are often correlated and high dimensional in this era of big data, such as data obtained in high-throughput biomedical experiments.
-Due to the computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations.
-To close this gap, we developed the \pkg{Bigsimr} Julia package with R and Python interfaces.
-These packages empower high-dimensional random vector simulation with arbitrary marginal distributions and dependency via a Pearson, Spearman, or Kendall correlation matrix.
-\pkg{Bigsimr} contains high-performance features, including multi-core and graphical-processing-unit-accelerated algorithms to estimate correlation and compute the nearest correlation matrix.
-Monte Carlo studies quantify the accuracy and scalability of our approach, up to \(d=10,000\).
-We describe example workflows and apply to a high-dimensional data set --- RNA-sequencing data obtained from breast cancer tumor samples.
+It is critical to accurately simulate data when employing Monte Carlo
+techniques and evaluating statistical methodology. Measurements are
+often correlated and high dimensional in this era of big data, such as
+data obtained in high-throughput biomedical experiments. Due to the
+computational complexity and a lack of user-friendly software available
+to simulate these massive multivariate constructions, researchers resort
+to simulation designs that posit independence or perform arbitrary data
+transformations. To close this gap, we developed the \pkg{Bigsimr} Julia
+package with R and Python interfaces. These packages empower
+high-dimensional random vector simulation with arbitrary marginal
+distributions and dependency via a Pearson, Spearman, or Kendall
+correlation matrix. \pkg{Bigsimr} contains high-performance features,
+including multi-core and graphical-processing-unit-accelerated
+algorithms to estimate correlation and compute the nearest correlation
+matrix. Monte Carlo studies quantify the accuracy and scalability of our
+approach, up to \(d=10,000\). We describe example workflows and apply to
+a high-dimensional data set --- RNA-sequencing data obtained from breast
+cancer tumor samples.
 }
 
-\Keywords{multivariate simulation, high-dimensional data, nonparametric correlation, Gaussian copula, RNA-sequencing data, breast cancer}
-\Plainkeywords{multivariate simulation, high-dimensional data, nonparametric correlation, Gaussian copula, RNA-sequencing data, breast cancer}
+\Keywords{multivariate simulation, high-dimensional data, nonparametric
+correlation, Gaussian copula, RNA-sequencing data, breast cancer}
+\Plainkeywords{multivariate simulation, high-dimensional
+data, nonparametric correlation, Gaussian copula, RNA-sequencing
+data, breast cancer}
 
@@
 % Pandoc header
+
 \usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{amsfonts}
-\usepackage{amsthm}
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}[theorem]{Lemma}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{ragged2e}
-\usepackage{setspace}
 
@@
 
-\setstretch{2.0}
+Massive high-dimensional (HD) data sets are now common in many areas of
+scientific inquiry. As new methods are developed for data analysis, a
+fundamental challenge lies in designing and conducting simulation
+studies to assess the operating characteristics of proposed methodology
+--- such as false positive rates, statistical power, interval coverage,
+and robustness. Further, efficient simulation empowers statistical
+computing strategies, such as the parametric bootstrap
+\citep{Chernick2008} to simulate from a hypothesized null model,
+providing inference in analytically challenging settings. Such Monte
+Carlo (MC) techniques become difficult for HD dependent data using
+existing algorithms and tools. This is particularly true when simulating
+massive multivariate, non-normal distributions, arising in many fields
+of study.
 
-Massive high-dimensional (HD) data sets are now common in many areas of scientific inquiry.
-As new methods are developed for data analysis, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of proposed methodology --- such as false positive rates, statistical power, interval coverage, and robustness.
-Further, efficient simulation empowers statistical computing strategies, such as the parametric bootstrap \citep{Chernick2008} to simulate from a hypothesized null model, providing inference in analytically challenging settings.
-Such Monte Carlo (MC) techniques become difficult for HD dependent data using existing algorithms and tools.
-This is particularly true when simulating massive multivariate, non-normal distributions, arising in many fields of study.
+As others have noted, it can be vexing to simulate dependent,
+non-normal/discrete data, even for low-dimensional (LD) settings
+\citep{MB13, XZ19}. For continuous non-normal LD multivariate data, the
+well-known NORmal To Anything (NORTA) algorithm \citep{Cario1997} and
+other copula approaches \citep{Nelsen2007} are well-studied and
+implemented in publicly-available software \citep{Yan2007, Chen2001}.
+Yet these approaches do not scale in a timely fashion to HD problems
+\citep{Li2019gpu}. For discrete data, early simulation strategies had
+major flaws, such as failing to obtain the full range of possible
+correlations --- such as admitting only positive correlations
+\citep{Park1996}. While more recent approaches \citep{MB13, Xia17, BF17}
+have remedied this issue for LD problems, the existing tools are not
+designed to scale to high dimensions.
 
-As others have noted, it can be vexing to simulate dependent, non-normal/discrete data, even for low-dimensional (LD) settings \citep{MB13, XZ19}.
-For continuous non-normal LD multivariate data, the well-known NORmal To Anything (NORTA) algorithm \citep{Cario1997} and other copula approaches \citep{Nelsen2007} are well-studied and implemented in publicly-available software \citep{Yan2007, Chen2001}.
-Yet these approaches do not scale in a timely fashion to HD problems \citep{Li2019gpu}.
-For discrete data, early simulation strategies had major flaws, such as failing to obtain the full range of possible correlations --- such as admitting only positive correlations \citep{Park1996}.
-While more recent approaches \citep{MB13, Xia17, BF17} have remedied this issue for LD problems, the existing tools are not designed to scale to high dimensions.
+Another central issue lies in characterizing dependence between
+components in the HD random vector. The choice of correlation in
+practice usually relates to the eventual analytic goal and
+distributional assumptions of the data (e.g., non-normal, discrete,
+infinite support, etc.). For normal data, the Pearson product-moment
+correlation describes the dependence perfectly. However, simulating
+arbitrary random vectors that match a target Pearson correlation matrix
+is computationally intense \citep{Chen2001, Xia17}. On the other hand,
+an analyst might consider use of nonparametric correlation measures to
+better characterize monotone, non-linear dependence, such as Spearman's
+\(\rho\) and Kendall's \(\tau\). Throughout, we focus on matching these
+nonparametric dependence measures, as our aim lies in modeling
+non-normal data and these rank-based measures possess invariance
+properties enabling our proposed methodology. We do, however, implement
+Pearson matching, but several layers of approximation are required.
 
-Another central issue lies in characterizing dependence between components in the HD random vector.
-The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc.).
-For normal data, the Pearson product-moment correlation describes the dependence perfectly.
-However, simulating arbitrary random vectors that match a target Pearson correlation matrix is computationally intense \citep{Chen2001, Xia17}.
-On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependence, such as Spearman's \(\rho\) and Kendall's \(\tau\).
-Throughout, we focus on matching these nonparametric dependence measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties enabling our proposed methodology.
-We do, however, implement Pearson matching, but several layers of approximation are required.
+With all this in mind, we present a scalable, flexible multivariate
+simulation algorithm. The crux of the method lies in the construction of
+a Gaussian copula in the spirit of the NORTA procedure. As we will
+describe in more detail, the algorithm's design leverages useful
+properties of nonparametric correlation measures, namely invariance
+under monotone transformation and well-known closed-form relationships
+between dependence measures for the multivariate normal (MVN)
+distribution. For our method, we developed a high-performance
+implementation: the \texttt{Bigsimr} Julia package, with R and Python
+interfaces \texttt{bigsimr}.
 
-With all this in mind, we present a scalable, flexible multivariate simulation algorithm.
-The crux of the method lies in the construction of a Gaussian copula in the spirit of the NORTA procedure.
-As we will describe in more detail, the algorithm's design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed-form relationships between dependence measures for the multivariate normal (MVN) distribution.
-For our method, we developed a high-performance implementation: the \texttt{Bigsimr} Julia package, with R and Python interfaces \texttt{bigsimr}.
-
-This article proceeds by providing background information, including a description of a motivating example application: RNA-sequencing (RNA-seq) breast cancer data.
-Then we describe and justify our simulation methodology and related algorithms.
-We proceed by providing an illustrative LD \texttt{bigsimr} workflow.
-Next we conduct MC studies under various bivariate distributional assumptions to evaluate performance and accuracy.
-After the MC evaluations, we simulate random vectors motivated by our RNA-seq example, evaluate the accuracy, and provide example statistical computing tasks, namely MC estimation of joint probabilities and evaluating HD correlation estimation efficiency.
-Finally, we discuss the method's utility, limitations, and future directions.
-
-\hypertarget{background}{%
-\section{Background}\label{background}}
-
-The \texttt{Bigsimr} package presented here provides multiple high-performance algorithms that operate with HD multivariate data.
-All these algorithms were originally designed to support a single task: to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics.
-Specifically, our goal is to efficiently simulate a large number, \(B\), of HD random vectors \({\bf Y}=(Y_1, \ldots, Y_d)^\top\) with \emph{correlated} components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) \(F_i, i=1,\ldots,d\).
-
-When designing this methodology, we developed the following properties to guide our effort.
-We divide the properties into two categories: (1) basic properties (BP) and `scalability' properties (SP).
-The BPs are adapted from an existing criteria due to \citet{Nik13a}.
-A suitable simulation strategy should possess the following properties:
-
-\setstretch{1.5}
-
-\begin{itemize}
-\tightlist
-\item
-  BP1: A wide range of dependences, allowing both positive and negative values, and, ideally, admitting the full range of possible values.
-\item
-  BP2: Flexible dependence, meaning that the number of dependence parameters can be equal to the number of bivariate pairs in the vector.
-\item
-  BP3: Flexible marginal modeling, generating heterogeneous data --- including mixed continuous and discrete marginal distributions.
-\end{itemize}
-
-\setstretch{2.0}
-
-Moreover, the simulation method must \emph{scale} to high dimensions:
-
-\setstretch{1.5}
-
-\begin{itemize}
-\tightlist
-\item
-  SP1: Procedure must scale to high dimensions with practical computation times.
-\item
-  SP2: Procedure must scale to high dimensions while maintaining accuracy.
-\end{itemize}
-
-\setstretch{2.0}
-
-\hypertarget{motivating-example-rna-seq-data}{%
-\subsection{Motivating example: RNA-seq data}\label{motivating-example-rna-seq-data}}
-
-Simulating HD, non-normal, correlated data motivates this work, in pursuit of modeling RNA-sequencing (RNA-seq) data \citep{Wang2009b, Conesa2016b} derived from breast cancer patients.
-The RNA-seq data-generating process involves counting how often a particular form of messenger RNA (mRNA) is expressed in a biological sample.
-RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in HD data.
-For human-derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted \citep{Schissler2019}.
-Importantly, due to inherent biological processes, gene expression data exhibit correlation (co-expression) across genes \citep{BE07, Schissler2018}.
-
-We illustrate our methodology using the Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments).
-For ease of modeling and simplicity of exposition, we only consider high expressing genes.
-In turn, we begin by filtering to retain the top 1000 of the highest-expressing genes (in terms of median expression) of the over 20,000 gene measurements from \(N=878\) patients' tumor samples.
-This gives a great number of pairwise dependencies among the marginals (specifically, \(\ensuremath{4.995\times 10^{5}}\) correlation parameters).
-Table \ref{tab:ch010-realDataTab} displays RNA-seq counts for three selected high-expressing genes for the first five patients' breast tumor samples.
-To help visualize the bivariate relationships for these three selected genes across all patients, Figure \ref{fig:ch010-realDataFig} displays the marginal distributions and estimated Spearman's correlations.
-
-\begin{CodeChunk}
-\begin{table}
-
-\caption{\label{tab:ch010-realDataTab}mRNA expression for three selected high-expressing genes, STAU1, FKBP1A, NME2, for the first five patients in the TCGA BRCA data set.}
-\centering
-\begin{tabular}[t]{lrrr}
-\toprule
-Patient ID & STAU1 & FKBP1A & NME2\\
-\midrule
-TCGA-A1-A0SB & 10440 & 11354 & 17655\\
-TCGA-A1-A0SD & 21523 & 20221 & 14653\\
-TCGA-A1-A0SE & 21733 & 22937 & 35251\\
-TCGA-A1-A0SF & 11866 & 19650 & 16551\\
-TCGA-A1-A0SG & 12486 & 12089 & 10434\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics[width=0.85\linewidth]{_main_files/figure-latex/ch010-realDataFig-1}
-
-}
-
-\caption[Pairwise scatterplots for three example genes (Table 1), with estimated Spearman's correlations and marginal density estimates displayed]{Pairwise scatterplots for three example genes (Table 1), with estimated Spearman's correlations and marginal density estimates displayed.
-The data possess outliers, heavy-right tails, are discrete, and have non-trivial intergene correlations.
-Modeling these data motivate our simulation methodology.}\label{fig:ch010-realDataFig}
-\end{figure}
-\end{CodeChunk}
-
-\hypertarget{measures-of-dependency}{%
-\subsection{Measures of dependency}\label{measures-of-dependency}}
-
-In multivariate analysis, an analyst must select a metric to quantify dependency.
-The most widely-known is the Pearson correlation coefficient that describes the linear association between two random variables \(X\) and \(Y\), and, it is given by
-
-\begin{equation}
-\rho_P(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ \mathrm{Var}(X)\mathrm{Var}(Y)\right]^{1/2}}.
-\label{eq:pearson}
-\end{equation}
-
-As \citet{MB13} and \citet{MK01} discuss, for a bivariate normal \((X,Y)\) random vector, the Pearson correlation completely describes the dependency between the components.
-For non-normal marginals with monotone correlation patterns, however, \(\rho_P\) suffers some drawbacks and may mislead or fail to capture important relationships \citep{MK01}.
-Alternatively, analysts often prefer rank-based correlation measures to describe the degree of monotonic association.
-
-Two nonparametric, rank-based measures common in practice are Spearman's correlation (denoted \(\rho_S\)) and Kendall's \(\tau\).
-Spearman's \(\rho_S\) has an appealing correspondence as the Pearson correlation coefficient on \emph{ranks} of the values, thereby capturing nonlinear yet monotone relationships.
-Kendall's \(\tau\), on the other hand, is the difference in probabilities of concordant and discordant pairs of observations, \((X_i, Y_i)\) and \((X_j, Y_j)\) (concordance meaning that orderings have the same direction, e.g., if \(X_i < X_j\), then \(Y_i < Y_j\)).
-Note that concordance is determined by the ranks of the values, not the values themselves.
-Both \(\tau\) and \(\rho_S\) are \emph{invariant under monotone transformations} of the underlying random variates.
-As we will describe more fully in the \protect\hyperlink{algorithms}{Algorithms} section, this property enables matching rank-based correlations with speed (SP1) and accuracy (SP2).
-
-\emph{Correspondence among Pearson, Spearman, and Kendall correlations}
-
-There is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficient, as the marginal distributions \(F_i\) are intrinsic in their calculation.
-For \emph{bivariate normal vectors}, however, the correspondence is well-known:
-
-\begin{equation}
-\label{eq:convertKendall}
-\rho_{P} = \sin \left( \tau \times \frac{\pi}{2} \right), 
-\end{equation}
-
-\noindent and similarly for Spearman's \(\rho\) \citep{K58},
-
-\begin{equation}
-\label{eq:convertSpearman}
-\rho_P = 2 \times \sin \left( \rho_S \times \frac{\pi}{6} \right).
-\end{equation}
-
-\emph{Marginal-dependent bivariate correlation bounds}
-
-Given two marginal distributions, \(\rho_P\) is not free to vary over the entire range of possible correlations \([-1,1]\).
-The \emph{Frechet-Hoeffding bounds} are well-studied \citep{Nelsen2007, BF17}.
-These constraints cannot be overcome through algorithm design.
-In general, the bounds are given by
-
-\begin{equation}
-\label{eq:frechet}
-\rho_P^{max} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (U) \right), \quad \rho_P^{min} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (1 - U) \right)
-\end{equation}
-
-\noindent where \(U\) is a uniform random variable on \((0,1)\) and \(F^{-1}_1, F^{-1}_2\) are the inverse CDFs of \(X_1\) and \(X_2\), respectively, defined as
-
-\begin{equation}
-F_{i}^{-1} = \inf\{y:F_{i}(y) \geq u \}
-\label{eq:inverseCDF}
-\end{equation}
-
-when the variables are discrete.
-
-\hypertarget{gaussian-copulas}{%
-\subsection{Gaussian copulas}\label{gaussian-copulas}}
-
-There is a connection of our simulation strategy to Gaussian \emph{copulas} \citep{Nelsen2007}.
-A copula is a distribution function on \([0,1]^d\) that describes a multivariate probability distribution with standard uniform marginals.
-This provides a powerful, natural way to characterize joint probability.
-Consequently, the study of copulas is an important and active area of statistical theory and practice.
-For any random vector \({\bf X}=(X_1, \ldots, X_d)^\top\) with CDF \(F\) and marginal CDFs \(F_i\) there is a copula function \(C(u_1, \ldots, u_d)\) satisfying
-
-\begin{equation}
-F(x_1, \ldots, x_d) = {\mathbb P}(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), \quad x_i \in {\mathbb R}, i=1,\ldots,d.
-\label{eq:copula}
-\end{equation}
-
-A Gaussian copula has marginal CDFs that are all standard normal, \(F_i = \Phi, \forall \, i\), .
-This representation corresponds to a multivariate normal (MVN) distribution with standard normal marginal distributions and covariance matrix \(R_P\).
-Since the marginals are standardized to have unit variance, this \(R_P\) is also Pearson correlation matrix.
-If \(F_R\) is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula \(C_R\) is defined through
-
-\begin{equation}
-\label{eq:gauss}
-F_R(x_1, \ldots, x_d) = C_R(\Phi(x_1), \ldots, \Phi(x_d)),
-\end{equation}
-
-where \(\Phi(\cdot)\) is the standard normal CDF.
-
-Here, the copula function \(C_R\) is the CDF of the random vector \((\Phi(X_1), \ldots, \Phi(X_d))\), where \((X_1, \ldots, X_d) \sim N_d({\bf 0}, R_P)\).
-
-Sklar's Theorem \citep{Ubeda-Flores2017} guarantees that given inverse CDFs \(F_i^{-1}\)s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions.
-For example, using Gaussian copulas, we can construct a random vector \({\bf Y} = (Y_1, \ldots, Y_d)^\top\) with \(Y_i \sim F_i\), viz.
-\(Y_i = F_i^{-1}(\Phi(X_i)), i=1, \ldots, d\), where \((X_1, \ldots, X_d) \sim N_d({\bf 0}, R_P)\).
-
-\hypertarget{other-multivariate-simulation-packages}{%
-\subsection{Other Multivariate Simulation Packages}\label{other-multivariate-simulation-packages}}
-
-To place \texttt{Bigsimr}'s implementation and algorithms in context of existing approaches, we qualitatively compare \texttt{Bbigsimr} to other multivariate simulation packages.
-In \texttt{R}, the \texttt{MASS} package's \texttt{mvnorm()} has long been available to produce random vectors from multivariate normal (and \(t\)) distributions.
-This procedure cannot simulate from arbitrary margins and dependency measures and it fails to readily scale to high dimensions.
-SAS PROC SIMNORMAL and Python's NumPy \texttt{random.multivariate\_normal} both do similar tasks.
-Recently, the high-performance \texttt{R} package \texttt{mvnfast} \citep{Fasiolo2016} has been released, which provides HD multivariate normal generation.
-However, lacks flexibility in marginal models and dependency modeling.
-For dependent discrete data, LD \texttt{R} packages exist, such as \texttt{GenOrd}.
-For copula computation, the full-featured \texttt{R} package \texttt{copula} provides many different copulas and a fairly general LD random generator \texttt{rCopula()}.
-High-dimesional random vector generation via \texttt{copula} takes an impractical amount of computing time \citep{Li2019gpu}.
-The \texttt{nortaRA} R package matches Pearson coefficients near exactly and is reasonable for LD settings.
-Finally, we know of no other \texttt{R} package that can produce HD random vectors with flexible margins and dependence structure.
-Table \ref{tab:compare-table} summarizes the above discussion.
-
-\begin{table}[h]
-\centering
-\caption{\label{tab:compare-table}Comparison of selected features of \pkg{Bigsimr} to existing packages for multivariate simulation.}
-\begin{tabular}{@{}llll@{}}
-Package & Scales to HD & Hetereogenous Margins & Flexible Dependence \\ \midrule
-MASS      & No           & No                    & No                  \\
-mvnfast   & Yes          & No                    & No                  \\
-GenOrd    & No           & Discrete margins      & Yes                 \\
-copula    & No           & Yes                   & No                  \\
-nortaRA   & No           & Yes                   & No                  \\
-bigsimr   & Yes          & Yes                   & Yes                
-\end{tabular}
-\end{table}
-
-\clearpage
-
-\hypertarget{algorithms}{%
-\section{Algorithms}\label{algorithms}}
-
-This section describes our methods for simulating a random vector \(\bf Y\) with \(Y_i\) components for \(i=1,\ldots,d\).
-Each \(Y_i\) has a specified marginal CDF \(F_i\) and its inverse \(F^{-1}_i\).
-To characterize dependency, every pair \((Y_i, Y_j)\) has a given Pearson correlation \(\rho_P\), Spearman correlation \(\rho_S\), and/or Kendall's \(\tau\).
-The method can be described as a \emph{high-performance Gaussian copula} (Equation \eqref{eq:gauss}) providing a HD NORTA-inspired algorithm.
-
-\hypertarget{normal-to-anything-norta}{%
-\subsection{NORmal To Anything (NORTA)}\label{normal-to-anything-norta}}
-
-The well-known NORTA algorithm \citep{Cario1997} simulates a random vector \(\bf Y\) with variance-covariance matrix \(\Sigma_{\bf Y}\). Specifically, NORTA algorithm proceeds as follows:
-
-\setstretch{1.5}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Simulate a random vector \(\bf Z\) with \(d\) independent and identically distributed (iid) standard normal components.
-\item
-  Determine the input matrix \(\Sigma_{\bf Z}\) that corresponds with the specified output \(\Sigma_{\bf Y}\).
-\item
-  Produce a Cholesky factor \(M\) of \(\Sigma_{\bf Z}\) such that \(M M^{\prime}=\Sigma_{\bf Z}\).
-\item
-  Set \(X\) by \(X \gets MZ\).
-\item
-  \(\text{Return} \; {\bf Y} \; \text{where} \; Y_i \gets F_i^{-1}[\Phi(X_i)], \; i=1,...,d\).
-\end{enumerate}
-
-\setstretch{2.0}
-
-With modern parallel computing, steps 1, 3, 4, 5 are readily implemented as high-performance, multi-core and/or graphical-processing-unit (GPU) accelerated algorithms --- providing fast scalability to HD.
-
-Matching specified Pearson correlation coefficients exactly (step 2 above), however, is computationally costly.
-In general, there is no convenient correspondence between the components of the input \(\Sigma_{\bf Z}\) and target \(\Sigma_{\bf Y}\).
-Matching the correlations involves evaluating or approximating \(\binom{d}{2}\) integrals of the form
-
-\begin{equation}
-    \mathrm{E}\left[Y_i Y_j\right] = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} F_i^{-1}\left[\Phi(z_i)\right] F_j^{-1}\left[\Phi(z_j)\right] \phi(z_i, z_j, \rho_z) dz_i dz_j,
-    \label{eq:pearsonIntegralRelation}
-\end{equation}
-
-where \(\phi(\cdot)\) is the joint probability density function of two correlated standard normal variables.
-For HD simulation, exact evaluation may become too costly to enable practical simulation studies.
-Our remedy to enable HD Pearson matching in \texttt{bigsimr} is Hermite polynomial approximation of these \(\binom{d}{2}\) double integrals, as suggested by \citet{XZ19}.
-
-\emph{NORTA in higher dimensions}
-
-Sklar's theorem provides a useful characterization of multivariate distributions through copulas.
-In practice, however, the particulars of the simulation algorithm affect which joint distributions may be simulated.
-Even in LD spaces (e.g., \(d=3\)), there exist valid multivariate distributions with \emph{feasible} Pearson correlation matrices that NORTA cannot match exactly \citep{LH75}.
-This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined, the resultant matrix becomes non-positive definite.
-These situations do occur, even using exact analytic calculations.
-Such problematic target correlation matrices are termed \emph{NORTA defective}.
-
-\citet{GH02} conducted a study to estimate the probability of encountering NORTA defective matrices while increasing the dimension \(d\).
-They found that for what is now considered low-to-moderate dimensions (\(d \approx 20\)), almost \emph{all} feasible matrices are NORTA defective.
-This stems from the concentration of measure near the boundary of the space of all possible correlation matrices as dimension increases.
-Unfortunately, it is precisely near this boundary that NORTA defective matrices reside.
-
-There is hope, however, as \citet{GH02} also showed that replacing an non-positive definite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate \(d\).
-This provides evidence that our nearest positive definite (PD) augmented approach has promise to provide adequate accuracy if our input matching scheme returns an indefinite Pearson correlation matrix.
-
-\hypertarget{rand-vec-gen}{%
-\subsection{Random vector generator}\label{rand-vec-gen}}
-
-We now describe our algorithm to generate random vectors, which extends NORTA to HD via nearest correlation matrix replacement and provides rank-based dependency matching:
-
-\setstretch{1.5}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Mapping step, either
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Convert the target Spearman correlation matrix \(R_S\) to the corresponding MVN Pearson correlation \(R_X\). Alternatively,
-  \item
-    Convert the target Kendall \(\tau\) matrix \(R_K\) to the corresponding MVN Pearson correlation \(R_X\). Alternatively,
-  \item
-    Convert the target Pearson correlation matrix to \(R_P\) to the corresponding, approximate MVN Pearson correlation \(R_X\).
-  \end{itemize}
-\item
-  Check correlation matrix admissibility and, if not, compute the nearest correlation matrix.
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Check that \(R_X\) is a correlation matrix, a positive definite matrix with 1's along the diagonal.
-  \item
-    If the mapping produced an \(R_X\) that is a correlation matrix, we describe the matrix as \textbf{admissible} in this scheme and retain \(R_X\). Otherwise,
-  \item
-    Replace \(R_X\) with the nearest correlation matrix \(\tilde{R}_X\), in the Frobenius norm and the result will be approximate.
-  \end{itemize}
-\item
-  Gaussian copula
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Generate \({\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)\).
-  \item
-    Transform \({\bf X}\) to \({\bf U} = (U_1, \ldots, U_d)\) viz. \(U_i=\Phi(X_i)\), \(i=1, \ldots, d\).
-  \item
-    Return \({\bf Y} = (Y_1, \ldots, Y_d)\), where \(Y_i=F_i^{-1}(U_i)\), \(i=1, \ldots, d\).
-  \end{itemize}
-\end{enumerate}
-
-\emph{Step 1: Mapping step.} We first employ the closed-form relationships between \(\rho_S\) and \(\tau\) with \(\rho_P\) for bivariate normal random variables via Equations \eqref{eq:convertKendall} and \eqref{eq:convertSpearman}, respectively (implemented as \texttt{cor\_covert()}).
-Initializing our algorithm to match the nonparametric correlations by computing these equations for all pairs is computationally trivial.
-
-For the computationally expensive process to match Pearson correlations, we approximate Equation \eqref{eq:pearsonIntegralRelation} for all pairs of margins.
-To this end, we implement the approximation scheme introduced by \citep{XZ19}.
-Briefly, the many double integrals of the form in \eqref{eq:pearsonIntegralRelation} are approximated by weighted sums of Hermite polynomials.
-Matching coefficients for pairs of continuous distributions is made tractable by this method, but for discrete distributions (especially discrete distributions with large support sets or infinite support), the approximation is computationally expensive.
-To remedy this, we further approximate discrete distributions by a continuous distribution using Generalized S-Distributions \citep{muino2006gs}.
-
-\emph{Step 2: Admissibility check and nearest correlation matrix computation.} Once \(R_X\) has been determined, we check admissibility of the adjusted correlation via the steps described above.
-If \(R_X\) is not a valid correlation matrix, then we compute the nearest correlation matrix.
-Finding the nearest correlation matrix is a common statistical computing problem.
-The defacto function in \texttt{R} is \texttt{Matrix::nearPD}, an alternating projection algorithm due to \citet{higham2002computing}.
-As implemented, the function fails to scale to HD.
-Instead, we provide the quadratically-convergent algorithm based on the theory of strongly semi-smooth matrix functions \citep{qi2006quadratically}.
-The nearest correlation matrix problem can be written down as the following convex optimization problem: \(\mathrm{min} \quad \frac{1}{2} \Vert R_X - X \Vert^2, \quad \mathrm{s.t.} \quad X_{ii} = 1, \quad i = 1, \ldots , n, \quad X \in S_{+}^{n}\).
-
-For nonparametric correlation measures, our algorithm allows the generation of HD multivariate data with arbitrary marginal distributions with a broad class of admissible Spearman correlation matrices and Kendall \(\tau\) matrices.
-The admissible classes consist of the matrices that map to a Pearson correlation matrix for a MVN.
-In particular, if we let \(X\) be MVN with \(d\) components and denote \(\Omega_P = \{ R_P : R_P \textrm{ is a Pearson correlation matrix for } X \}, \quad \Omega_K = \{ R_K : R_K \textrm{ is a Kendall correlation matrix for } X \}, \quad \Omega_S = \{ R_S : R_S \textrm{ is a Spearman correlation matrix for } X \}\).
-There are 1-1 mappings between these sets.
-We conjecture that the sets of admissible \(R_S\) and \(R_K\) are not highly restrictive.
-In particular, \(R_P\) is approximately \(R_S\) for a MVN, suggesting that the admissible set \(\Omega_S\) should be flexible as \(R_P\) can be any PD matrix with 1's along the diagonal.
-We provide methods to check whether a target \(R_S\) is an element of the \emph{admissible set} \(\Omega_S\).
-
-There is an increasing probability of encountering an inadmissible correlation matrix as dimension increases.
-In our experience, the mapping step for large \(d\) almost always produces a \(R_X\) that is not a correlation matrix.
-In Section \ref{package}, we provide a basic method of how to quantify and control the approximation error.
-Further, the RNA-seq example in Section \ref{examples} provides an illustration of this in practice.
-
-\emph{Step 3: Gaussian copula.} The final step implements a NORTA-inspired, Gaussian copula approach to produce the desired margins.
-Steps 1 and 2 determine the MVN Pearson correlation values that will eventually match the target correlation.
-Step 3 requires a fast MVN simulator, a standard normal CDF, and well-defined quantile functions for marginals.
-The MVN is transformed to a copula (distribution with standard uniform margins) by applying the normal CDF \(\Phi(\cdot)\).
-Finally, the quantile functions \(F_i^{-1}\) are applied across the margins to return the desired random vector \({\bf Y}\).
-
-\section[Bigsimr-pkg]{The \pkg{Bigsimr} package}\label{package}
-
-This section describes a bivariate random vector simulation workflow via the \texttt{bigsimr} \texttt{R} package.
-This \texttt{R} package provides an interface to the native code written in Julia (registered as the \texttt{Bigsimr} Julia package).
-In addition to the native Julia \texttt{Bigsimr} package and \texttt{R} interface \texttt{bigsimr}, we also provide a Python interface \texttt{bigsimr} that interfaces with the Julia \texttt{Bigsimr} package.
-The Julia package provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \ref{algorithms}).
-
-The subsections below describe the basic use of the \texttt{bigsimr} R package by stepping through an example workflow using the data set \texttt{airquality} that contains daily air quality measurements in New York, May to September 1973 \citep{Chambers1983}.
-This workflow proceeds from setting up the computing environment, to data wrangling, estimation, simulation configuration, random vector generation, and, finally, result visualization.
-
-\hypertarget{bivariate-example}{%
-\subsection{Bivariate example}\label{bivariate-example}}
-
-We illustrate the use of \texttt{bigsimr} using the New York air quality data set (\texttt{airquality}) included in the R \texttt{datasets} package.
-First, we load the \texttt{bigsimr} library and a few other convenient data science packages, including the syntactically-elegant \texttt{tidyverse} suite of \texttt{R} packages.
-The code chunk below prepares the computing environment:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> library("tidyverse")
-R> library("bigsimr")
-R> # Activate multithreading in Julia
-R> Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
-R> # Load the Bigsimr and Distributions Julia packages
-R> bs <- bigsimr_setup()
-R> dist <- distributions_setup()
-\end{CodeInput}
-\end{CodeChunk}
-
-Here, we describe a minimal working example --- a bivariate simulation of two airquality variables: \texttt{Temperature}, in degrees Fahrenheit, and \texttt{Ozone} level, in parts per billion.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> df <- airquality %>% select(Temp, Ozone) %>% drop_na()
-R> glimpse(df)
-\end{CodeInput}
-\begin{CodeOutput}
-Rows: 116
-Columns: 2
-$ Temp  <int> 67, 72, 74, 62, 66, 65, 59, 61, 74, 69, 66, 68, 58, 64, 66, 57, ~
-$ Ozone <int> 41, 36, 12, 18, 28, 23, 19, 8, 7, 16, 11, 14, 18, 14, 34, 6, 30,~
-\end{CodeOutput}
-\end{CodeChunk}
-
-Figure \ref{fig:ch030-aq-joint-dist} visualizes the bivariate relationship between \texttt{Ozone} and \texttt{Temperature}.
-We aim to simulate random two-component vectors mimicking this structure.
-The margins are not normally distributed; \texttt{Ozone} level exhibits a strong positive skew.
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics[width=0.85\linewidth]{_main_files/figure-latex/ch030-aq-joint-dist-1}
-
-}
-
-\caption[Bivariate scatterplot of Ozone versus Temp with estimated marginal densities]{Bivariate scatterplot of Ozone versus Temp with estimated marginal densities.
-The Ozone data are modeled marginally as log-normal and the Temperature data as normal.}\label{fig:ch030-aq-joint-dist}
-\end{figure}
-\end{CodeChunk}
-
-Next, we specify the marginal distributions and correlation coefficient (both type and magnitude).
-Here the analyst is free to be creative.
-For this example, we avoid goodness-of-fit considerations to determine the marginal distributions.
-It is sensible without domain knowledge to estimate these quantities from the data, and \texttt{bigsimr} contains fast functions designed for this task.
-
-\hypertarget{specifying-marginal-distributions}{%
-\subsection{Specifying marginal distributions}\label{specifying-marginal-distributions}}
-
-Based on the estimated densities in Figure \ref{fig:ch030-aq-joint-dist}, we assume \texttt{Temp} is normally distributed and \texttt{Ozone} is log-normally distributed, as the latter values are positive and skewed.
-We use the classical unbiased estimators for the normal distribution's parameters and maximum likelihood estimators for the log-normal parameters:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> df %>% select(Temp) %>% 
-+   summarise_all(.funs = c(mean = mean, sd = sd))
-\end{CodeInput}
-\begin{CodeOutput}
-   mean    sd
-1 77.87 9.485
-\end{CodeOutput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> mle_mean <- function(x) mean(log(x))
-R> mle_sd <- function(x) mean( sqrt( (log(x) - mean(log(x)))^2 ) )
-R> df %>% 
-+   select(Ozone) %>% 
-+   summarise_all(.funs = c(meanlog = mle_mean, sdlog = mle_sd))
-\end{CodeInput}
-\begin{CodeOutput}
-  meanlog  sdlog
-1   3.419 0.6967
-\end{CodeOutput}
-\end{CodeChunk}
-
-Now, we configure the input marginals for later input into \texttt{rvec}. The marginal distributions are specified using \texttt{Julia}'s \texttt{Distributions} package and stored in a vector.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> margins <- c(dist$Normal(mean(df$Temp), sd(df$Temp)),
-+              dist$LogNormal(mle_mean(df$Ozone), mle_sd(df$Ozone)))
-\end{CodeInput}
-\end{CodeChunk}
-
-\hypertarget{specifying-correlation}{%
-\subsection{Specifying correlation}\label{specifying-correlation}}
-
-The user must decide how to describe correlation based on the particulars of the problem.
-For non-normal data in our scheme, we advocate the use of Spearman's \(\rho\) correlation matrix \(R_S\) or Kendall's \(\tau\) correlation matrix \(R_K\).
-We also support Pearson correlation coefficient matching, while cautioning the user to check the performance for the distribution at hand (see \protect\hyperlink{simulations}{Monte Carlo evaluations} below for evaluation strategies and guidance).
-Note that these estimation methods are classical approaches, not specifically designed for high-dimensional correlation estimation.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> (R_S <- bs$cor(as.matrix(df), bs$Spearman))
-\end{CodeInput}
-\begin{CodeOutput}
-      [,1]  [,2]
-[1,] 1.000 0.774
-[2,] 0.774 1.000
-\end{CodeOutput}
-\end{CodeChunk}
-
-\hypertarget{checking-target-correlation-matrix-admissibility}{%
-\subsection{Checking target correlation matrix admissibility}\label{checking-target-correlation-matrix-admissibility}}
-
-Once a target correlation matrix is specified, first convert the dependency type to Pearson to correctly specify the MVN inputs.
-Then, check that the converted matrix \(R_X\) is a valid correlation matrix (PD and 1s on the diagonal).
-For this bivariate example, \(R_X\) is a valid correlation matrix.
-Note that typically in HD the resultant MVN Pearson correlation matrix is indefinite and requires approximation (see the HD examples in subsequent sections).
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> # Step 1. Mapping
-R> (R_X <- bs$cor_convert(R_S, bs$Spearman, bs$Pearson))
-\end{CodeInput}
-\begin{CodeOutput}
-       [,1]   [,2]
-[1,] 1.0000 0.7886
-[2,] 0.7886 1.0000
-\end{CodeOutput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> # Step 2. Check admissibility
-R> bs$iscorrelation(R_X)
-\end{CodeInput}
-\begin{CodeOutput}
-[1] TRUE
-\end{CodeOutput}
-\end{CodeChunk}
-
-Despite being a valid correlation matrix under this definition, marginal distributions induce Frechet bounds on the possible Pearson correlation values.
-To check that the converted correlation values fall within these bounds, use \texttt{bigsimr::cor\_bounds} to estimate the pairwise lower and upper correlation bounds.
-\texttt{cor\_bounds} uses the Generate, Sort, and Correlate algorithm of \citet{DH2011}.
-For the assumed marginals, the Pearson correlation coefficient is not free to vary in {[}-1,1{]}, as seen below.
-Our MC estimate of the bounds slightly underestimates the theoretical bounds of \((-0.881, 0.881)\).
-(See an analytic derivation presented in the Appendix).
-Since our single Pearson correlation coefficient is within the theoretical bounds, the correlation matrix is valid for our simulation strategy.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> bs$cor_bounds(margins[1], margins[2], bs$Pearson, n_samples = 1e6)
-\end{CodeInput}
-\begin{CodeOutput}
-Julia Object of type NamedTuple{(:lower, :upper),Tuple{Float64,Float64}}.
-(lower = -0.8823026550919685, upper = 0.8815987574226597)
-\end{CodeOutput}
-\end{CodeChunk}
-
-In this 2D example, it is possible to verify the existence of a bivariate distribution with the target Pearson correlation / margins (See the \protect\hyperlink{appendix}{Appendix}).
-For general \(d-\)variate distributions, is it not guaranteed that such a multivariate distribution exists with exact the specified correlations and margins \citep{BF17}.
-Theoretical guidance is lacking in this regard, but, in practice, we find that using bivariate bounds form rough approximations to the bounds for the feasible region for the \(d-\)variate construction when aggregated into a Pearson matrix.
-See the \protect\hyperlink{examples}{RNA-seq data application} section for an example discussing this, NORTA feasibility, and use of the nearest PD matrix algorithm.
-
-\hypertarget{simulating-random-vectors}{%
-\subsection{Simulating random vectors}\label{simulating-random-vectors}}
-
-Finally, we execute \texttt{rvec} to simulate the desired \(10,000\) random vectors from the assumed bivariate distribution of \texttt{Ozone} and \texttt{Temp}.
-Note that the target input is specified using the pre-computed MVN correlation matrix.
-Figure \ref{fig:ch030-plot-sim} plots the 10,000 simulated points.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> x <- bs$rvec(10000, R_X, margins)
-R> df_sim <- as.data.frame(x)
-R> colnames(df_sim) <- colnames(df)
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics[width=0.85\linewidth]{_main_files/figure-latex/ch030-plot-sim-1}
-
-}
-
-\caption[Contour plot and marginal densities for the simulated bivariate distribution of Air Quality Temperatures and Ozone levels]{Contour plot and marginal densities for the simulated bivariate distribution of Air Quality Temperatures and Ozone levels.
-The simulated points mimic the observed data with respect to both the marginal characteristics and bivariate association.}\label{fig:ch030-plot-sim}
-\end{figure}
-\end{CodeChunk}
-
-\clearpage
-
-\hypertarget{simulations}{%
-\section{Monte Carlo evaluations}\label{simulations}}
-
-In this section, we conduct MC studies to investigate method performance.
-Marginal parameter matching is straightforward in our scheme as it is a sequence of univariate inverse probability transforms.
-This process is both fast and exact.
-On the other hand, accurate and computationally efficient dependency matching presents a challenge, especially in the HD setting.
-To evaluate our methods in those respects, we design the following numerical experiments to first assess accuracy of matching dependency parameters in bivariate simulations and then time the procedure in increasingly large dimension \(d\).
-
-\hypertarget{bivariate-experiments}{%
-\subsection{Bivariate experiments}\label{bivariate-experiments}}
-
-We select bivariate simulation configurations to ultimately simulate our motivating discrete-valued RNA-seq example.
-Thus, we proceed by increasing departure from normality, leading to a multivariate negative binomial (MVNB) model for our motivating data.
-We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson, Spearman, and Kendall --- in identical, bivariate marginal configurations.
-For each pair of identical margins, we vary the target correlation across \(\Omega\), the set of possible admissible values for each correlation type, to evaluate the simulation's ability to generate all possible correlations.
-The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (non-normal and discrete).
-
-Table \ref{tab:sims} lists our identical-marginal, bivariate simulation configurations.
-We increase the simulation replicates \(B\) to visually ensure that our results converge to the target correlations and gauge efficiency.
-We select distributions beginning with a standard multivariate normal (MVN) as we expect the performance to be exact (up to MC error) for all correlation types.
-Then, we select a non-symmetric continuous distribution: a standard (rate =1), two-component multivariate gamma (MVG).
-Finally, we select distributions and marginal parameter values that are motivated by our RNA-seq data, namely values proximal to probabilities and sizes estimated from the data (see \href{examples}{RNA-seq data application} for estimation details).
-Specifically, we assume a MVNB with \(p_1 = p_2 = 3\times10^{-4}, r_1 = r_2 = 4, \rho \in \Omega\).
-
-\begin{table}[]
-\centering
-\caption{ \label{tab:sims} Identical margin, bivariate simulation configurations to evaluate matching accuracy.}
-\begin{tabular}{@{}lcr@{}}
-\toprule
-Simulation Reps $B$ & Correlation Types & Identical-margin 2D distribution \\ \midrule
-$1,000$ & Pearson ($\rho_P$) & ${\bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho_i ), i=1,\ldots,100$ \\
-$10,000$ & Spearman ($\rho_S$) & ${\bf Y} \sim MVG( shape = 10, rate = 1, \rho_i ), i=1,\ldots,100$ \\
-$100,000$ & Kendall ($\tau$) & ${\bf Y} \sim MVNB(p = 3\times10^{-4}, r = 4,\rho_i), i=1,\ldots,100$ \\ \bottomrule
-\end{tabular}
-\end{table}
-
-For each of the unique 9 simulation configurations described above, we estimate the correlation bounds and vary the correlations along a sequence of 100 points evenly placed within the bounds, aiming to explore \(\Omega\).
-Specifically, we set correlations \(\{ \rho_1 = ( \hat{l} + \epsilon), \rho_2 = (\hat{l} + \epsilon) + \delta, \ldots, \rho_{100} = (\hat{u} - \epsilon) \}\), with \(\hat{l}\) and \(\hat{u}\) being the estimated lower and upper bounds, respectively, with increment value \(\delta\).
-The adjustment factor, \(\epsilon=0.01\), is introduced to handle numeric issues when the bound is specified exactly.
-
-Figure \ref{fig:ch040-bPlot} displays the aggregated bivariate simulation results.
-Table \ref{tab:ch040-BiError} contains the mean absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.
-Overall, the studies show that our methodology is generally accurate across the entire range of possible correlation for all three dependency measures.
-Our Pearson matching performs nearly as well as Spearman or Kendall, except for a slight increase in error for the negative binomial case.
-
-\begin{CodeChunk}
-\begin{table}
-
-\caption{\label{tab:ch040-BiError}Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.}
-\centering
-\begin{tabular}[t]{lllr}
-\toprule
-No. of random vectors & Correlation type & Distribution & Mean abs. error\\
-\midrule
-1000 & Pearson & norm & 0.0156\\
-1000 & Pearson & gamma & 0.0159\\
-1000 & Pearson & nbinom & 0.0162\\
-\addlinespace
-1000 & Spearman & norm & 0.0175\\
-1000 & Spearman & gamma & 0.0182\\
-1000 & Spearman & nbinom & 0.0151\\
-\addlinespace
-1000 & Kendall & norm & 0.0130\\
-1000 & Kendall & gamma & 0.0115\\
-1000 & Kendall & nbinom & 0.0126\\
-\addlinespace
-10000 & Pearson & norm & 0.0060\\
-10000 & Pearson & gamma & 0.0058\\
-10000 & Pearson & nbinom & 0.0058\\
-\addlinespace
-10000 & Spearman & norm & 0.0062\\
-10000 & Spearman & gamma & 0.0056\\
-10000 & Spearman & nbinom & 0.0049\\
-\addlinespace
-10000 & Kendall & norm & 0.0033\\
-10000 & Kendall & gamma & 0.0038\\
-10000 & Kendall & nbinom & 0.0033\\
-\addlinespace
-1e+05 & Pearson & norm & 0.0018\\
-1e+05 & Pearson & gamma & 0.0017\\
-1e+05 & Pearson & nbinom & 0.0030\\
-\addlinespace
-1e+05 & Spearman & norm & 0.0017\\
-1e+05 & Spearman & gamma & 0.0016\\
-1e+05 & Spearman & nbinom & 0.0015\\
-\addlinespace
-1e+05 & Kendall & norm & 0.0010\\
-1e+05 & Kendall & gamma & 0.0011\\
-1e+05 & Kendall & nbinom & 0.0012\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics[width=0.85\linewidth]{_main_files/figure-latex/ch040-bPlot-1}
-
-}
-
-\caption[Bivariate simulations match target correlations across the entire range of feasible correlations]{Bivariate simulations match target correlations across the entire range of feasible correlations.
-The horizontal axis plots the specified target correlations for each bivariate margin.
-Normal margins are plotted in dark dark grey, gamma in medium grey, and negative binomial in light grey.
-As the number of simulated vectors $B$ increases from left to right, the variation in estimated correlations (vertical axis) decreases.
-The dashed line indicates equality between the specified and estimated correlations.}\label{fig:ch040-bPlot}
-\end{figure}
-\end{CodeChunk}
-
-\hypertarget{scale-up-to-high-dimensions}{%
-\subsection{Scale up to High Dimensions}\label{scale-up-to-high-dimensions}}
-
-With information of our method's accuracy from a low-dimensional perspective, we now assess whether \texttt{bigsimr} can scale to larger dimensional problems with practical computation times.
-We ultimately generate \(B=1,000\) random vectors for \(d=\{100, 250, 500, 1000, 2500, 5000, 10000\}\) for each correlation type, \{Pearson, Spearman, Kendall\} while timing the algorithm's major steps.
-We begin by producing a synthetic ``data set'' by completing the following steps:
-
-\setstretch{1.5}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Produce heterogeneous gamma marginals by randomly selecting the \(j^{th}\) gamma shape parameter from \(U_j \sim uniform(1,10), j=1,\ldots,d\) and the \(j^{th}\) rate parameter from \(V_j \sim exp(1/5), j=1,\ldots,d\), with the constant parameters determined arbitrarily.
-\item
-  Produce a random full-rank Pearson correlation matrix via \texttt{cor\_randPD} of size \(d \times d\).
-\item
-  Simulate a ``data set'' of \(1,000 \times d\) random vectors via \texttt{rvec}.
-  \setstretch{2.0}
-\end{enumerate}
-
-With the synthetic data set in hand, we complete and time the following four steps involved in a typical workflow (including correlation estimation).
-
-\setstretch{1.5}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Estimate the correlation matrix from the ``data'' in the \emph{Compute Correlation} step.
-\item
-  Map the correlations to initialize the algorithm (Pearson to Pearson, Spearman to Pearson, or Kendall to Pearson) in the \emph{Adjust Correlation} step.
-\item
-  Check whether the mapping produces a valid correlation matrix and, if not, find the nearest PD correlation matrix in the \emph{Check Admissibility} step.
-\item
-  Simulate \(1,000\) vectors in the \emph{Simulate Data} step.
-  \setstretch{2.0}
-\end{enumerate}
-
-The experiments are conducted on a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, with all 16 threads employed during computation.
-Table \ref{tab:ch040-moderateDtab} displays the total computation time for moderate dimesions (\(d \leq 500\)).
-In every simullation setting, the 1,000 vectors are generated rapidly, executing in under 2 seconds.
-
-\begin{CodeChunk}
-\begin{table}
-
-\caption{\label{tab:ch040-moderateDtab}Total time to produce 1,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.}
-\centering
-\begin{tabular}[t]{llr}
-\toprule
-Dimension & Correlation type & Total Time (Seconds)\\
-\midrule
-100 & Pearson & 0.081\\
-100 & Spearman & 0.027\\
-100 & Kendall & 0.075\\
-\addlinespace
-250 & Pearson & 0.378\\
-250 & Spearman & 0.062\\
-250 & Kendall & 0.375\\
-\addlinespace
-500 & Pearson & 1.425\\
-500 & Spearman & 0.135\\
-500 & Kendall & 1.495\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\end{CodeChunk}
-
-The results for \(d > 500\) show scalability to ultra-high dimensions for all three correlation types, although the total times do become much larger.
-Figure \ref{fig:ch040-largeDfig} displays computation times for \(d=\{1000, 2500, 5000, 10000\}\).
-For \(d\) equal to 1000 and 2500, the total time is under a couple of minutes.
-At \(d\) of 5000 and 10,000, Pearson correlation matching in the \emph{Adjust Correlation} step becomes costly.
-Interestingly, Pearson is actually faster than Kendall for \(d=10,000\) due to bottlenecks in \emph{Compute Correlation} and \emph{Check Admissibility}.
-Uniformly, matching Spearman correlations is faster, with total times under 5 minutes for \(d=10,000\), making Spearman the most computationally-friendly dependency type for our package.
-With this in mind, we scaled the simulation to \(d=20,000\) for the Spearman type and obtained the 1,000 vectors in under an hour (data not shown).
-In principle, this would enable the simulation of an entire human-derived RNA-seq data set.
-We note that for a given target correlation matrix and margins, steps 1, 2, and 3 only need to be computed once and the fourth step, \emph{Simulate Data}, is nearly instaneous for all settings considered.
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics[width=0.85\linewidth]{_main_files/figure-latex/ch040-largeDfig-1}
-
-}
-
-\caption[Computation times for HD multivariate gamma simulation]{Computation times for HD multivariate gamma simulation.}\label{fig:ch040-largeDfig}
-\end{figure}
-\end{CodeChunk}
-
-\emph{Limitations, conclusions, and recommendations}
-
-In the bivariate studies, we chose arbitrary simulation parameters for three distributions, moving from the Gaussian to discrete and non-normal MVNB.
-Under these conditions, the simulated random vectors sample the desired bivariate distribution across the entire range of pairwise correlations for the three dependency measures.
-The simulation results could differ for other choices of simulation settings.
-Specifying extreme correlations near the boundary or Frechet bounds could result in poor simulation performance.
-Fortunately, it is straightforward to evaluate simulation performance by using strategies similar to those completed above.
-We expect our random vector generation to perform well for the vast majority of NORTA-feasible correlation matrices, but advise to check the performance before making inferences/further analyses.
-
-Somewhat surprisingly, Kendall estimation and nearest PD computation scale poorly compared to Spearman and, even, approximate Pearson matching.
-In our experience, Kendall computation times are sensitive to the number of cores, benefiting from multi-core parallelization.
-This could mitigate some of the current algorithmic/implementation shortcomings.
-Despite this, Kendall matching is still feasible for most HD data sets.
-Finally, we note that one could use our single-pass algorithm \texttt{cor\_fastPD} to produce a `close' (not nearest PD) to scale to even higher dimensions with some loss of accuracy.
-
-\hypertarget{examples}{%
-\section{RNA-seq data application}\label{examples}}
-
-This section demonstrates how to simulate multivariate data using \texttt{bigsimr}, aiming to replicate the structure of HD dependent count data.
-In an illustration of our proposed methodology, we seek to simulate RNA-sequencing data by producing simulated random vectors mimicking the observed data and its generating process.
-Modeling RNA-seq using multivariate probability distributions is natural as inter-gene correlation occurs during biological processes \citep{Wang2009b}.
-And yet, many models do not account for this, leading to major disruptions to the operating characteristics of statistical estimation, testing, and prediction.
-The following subsections apply \texttt{bigsimr}'s methods to real RNA-seq data, including replicating an estimated parametric structure, probability estimation, and evaluation of correlation estimation efficiency.
-
-\hypertarget{simulating-high-dimensional-rna-seq-data}{%
-\subsection{Simulating High-Dimensional RNA-seq data}\label{simulating-high-dimensional-rna-seq-data}}
-
-We begin by estimating the structure of the TCGA BRCA RNA-seq data set.
-Ultimately, we will simulate \(B=10,000\) random vectors \({\bf Y}=(Y_1, \ldots, Y_d)^\top\) with \(d=1000\).
-We assume a MVNB model as RNA-seq counts are often over-dispersed and correlated.
-Since all \(d\) selected genes exhibit over-dispersion (data not shown), we proceed to estimate the NB parameters \((r_i, p_i), i=1,\ldots,d\), to determine the target marginal PMFs \(f_i\).
-To complete specification of the simulation algorithm inputs, we estimate the Spearman correlation matrix \(R_S\) to characterize dependency.
-With this goal in mind, we first estimate the desired correlation matrix using the fast implementation provided by \texttt{bigsimr}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> # Estimate Spearman's correlation on the count data
-R> R_S <- bs$cor(as.matrix(brca1000), bs$Spearman)
-\end{CodeInput}
-\end{CodeChunk}
-
-Next, we estimate the marginal parameters.
-We use the method of moments to estimate the marginal parameters for the multivariate negative binomial model.
-The marginal distributions are from the same probability family (NB), yet they are heterogeneous in terms of the parameters probability and size \((p_i, n_i)\) for \(i,\ldots,d\).
-The functions below support this estimation for later use in \texttt{rvec}.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> make_nbinom_margins <- function(sizes, probs) {
-+   margins <- lapply(1:length(sizes), function(i) {
-+     dist$NegativeBinomial(sizes[i], probs[i])
-+   })
-+   do.call(c, margins)
-+ }
-\end{CodeInput}
-\end{CodeChunk}
-
-We apply these estimators to the highest-expressing 1000 genes across the 878 patients:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> nbinom_fit <- apply(brca1000, 2, mom_nbinom)
-R> sizes <- nbinom_fit["size",]
-R> probs <- nbinom_fit["prob",]
-R> nb_margins <- make_nbinom_margins(sizes, probs)
-\end{CodeInput}
-\end{CodeChunk}
-
-Notably, the estimated marginal NB probabilities \(\{ \hat{p}_i \}\) are small --- ranging in the interval \([\ensuremath{7.5305\times 10^{-7}} , \ensuremath{5.6592\times 10^{-4}}]\).
-This gives rise to highly variable counts and, thus, less restriction on potential pairwise correlation pairs.
-Given the marginals, we now specify targets and check admissibility of the specified correlation matrix.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> # 1. Mapping step first
-R> R_X <- bs$cor_convert(R_S, bs$Spearman, bs$Pearson)
-R> # 2a. Check admissibility
-R> (is_valid_corr <- bs$iscorrelation(R_X))
-\end{CodeInput}
-\begin{CodeOutput}
-[1] FALSE
-\end{CodeOutput}
-\begin{CodeInput}
-R> # 2b. compute nearest correlation
-R> if (!is_valid_corr) {
-+   R_X_pd <- bs$cor_nearPD(R_X)
-+   ## Quantify the error
-+   targets      <- R_X[lower.tri(R_X, diag = FALSE)]
-+   approximates <- R_X_pd[lower.tri(R_X_pd, diag = FALSE)]
-+   R_X          <- R_X_pd
-+ }
-R> summary(abs(targets - approximates))
-\end{CodeInput}
-\begin{CodeOutput}
-    Min.  1st Qu.   Median     Mean  3rd Qu.     Max. 
-0.000000 0.000193 0.000413 0.000506 0.000717 0.006922 
-\end{CodeOutput}
-\end{CodeChunk}
-
-As seen above, \(R_X\) is not strictly admissible in our scheme.
-This can be seen from the negative result from \texttt{bs\$iscorrelation(R\_X)}, which checks if \(R_X\) is positive-definite with 1's along the diagonal.
-However, the approximation is close with a maximum absolute error of 0.0069 and average absolute error of \ensuremath{5.0641\times 10^{-4}} across the \ensuremath{4.995\times 10^{5}} correlations.
-With the inputs configured, \texttt{rvec} is executed to produce a synthetic RNA-seq data set:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> sim_nbinom <- bs$rvec(10000, R_X, nb_margins) 
-\end{CodeInput}
-\end{CodeChunk}
-
-Figure \ref{fig:ch050-simDataFig} displays the simulated counts and pairwise relationships for our example genes from Table \ref{tab:ch010-realDataTab}.
-Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parametric form and with less extreme points than the observed data in Figure \ref{fig:ch010-realDataFig}.
-Figure \ref{fig:ch050-figBRCA} compares the specified target parameter (horizontal axis) with the corresponding quantities estimated from the simulated data (vertical axis).
-The evaluation shows that the simulated counts approximately match the target parameters and exhibit the full range of estimated correlations from the data.
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics[width=0.85\linewidth]{_main_files/figure-latex/ch050-simDataFig-1}
-
-}
-
-\caption[Simulated data for three selected high-expressing genes generally replicates the estimated data structure]{Simulated data for three selected high-expressing genes generally replicates the estimated data structure.
-The data do not exhibit outlying points, but do possess the desired Spearman correlations, central tendencies, and discrete values.}\label{fig:ch050-simDataFig}
-\end{figure}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics[width=0.8\linewidth]{fig/ch050-figBRCA}
-
-}
-
-\caption[Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set]{Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set.
-The dashed red lines indicate equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes).}\label{fig:ch050-figBRCA}
-\end{figure}
-\end{CodeChunk}
-
-\emph{Limitations, conclusions, and recommendations}
-
-The results show overall aquedate simulation performance for our choice of parameters settings.
-Our settings were motivated by modeling high-expressing genes from the TCGA BRCA data set.
-In general, the ability to match marginal and dependence parameters depends on the particular joint probability model.
-We recommend to evaluate and tune your simulation until you can be assured of the accuracy.
-
-\hypertarget{discussion}{%
-\section{Conclusion and discussion}\label{discussion}}
-
-We developed a general-purpose high-dimensional multivariate simulation algorithm and provide a user-friendly, high-performance \texttt{R} package \texttt{bigsimr}, Julia package \texttt{Bigsimr} and Python package \texttt{bigsimr}.
-The random vector generation method is inspired by NORTA \citep{Cario1997} and Gaussian copula-based approaches \citep[\citet{BF17}, \citet{Xia17}]{MB13}.
-The major contributions of this work are methods and software for flexible, scalable simulation of HD multivariate probability distributions with broad data analytic applications for modern, big-data statistical computing.
-For example, one could simulate high-resolution time series data, such as those consistent with an auto-regressive moving average model exhibiting a specified Spearman structure.
-Our methods could also be used to simulate sparsely correlated data, as many HD methods assume, via specifying a \emph{spiked correlation matrix}.
-
-There are limitations to the methodology and implementation.
-We could only investigate selected multivariate distributions in our Monte Carlo studies.
-There may be instances where the methods do not perform well.
-Along those lines, we expect that correlation values close to the boundary of the feasible region could result in algorithm failure.
-Another issue is that for discrete distributions, we use continuous approximations when the support set is large.
-This could limit the scalability/accuracy for particular discrete/mixed multivariate distributions.
-Our method would also benefit computationally from faster Kendall estimation and more finely tuned nearest PD calculation for non-admissible Kendall matrices.
-
-\hypertarget{misc}{%
-\section*{Supplementary Materials}\label{misc}}
-\addcontentsline{toc}{section}{Supplementary Materials}
-
-We provide an open-source implementation of our methodology as the \texttt{Bigsimr.jl} Julia package, hosted on github at \url{https://github.com/SchisslerGroup/Bigsimr.jl}.
-Additionally we provide R and Python interfaces to \texttt{Bigsimr}, respectively on \url{https://github.com/SchisslerGroup/r-bigsimr} and \url{https://github.com/SchisslerGroup/python-bigsimr}.
-A stable release of the R package \texttt{bigsimr} is available on CRAN.
-\texttt{Bigsimr.jl} is an ongoing project, and feature requests or issues can be submitted to \url{https://github.com/adknudson/Bigsimr.jl/issues}.
-
-\begin{CodeChunk}
-
-\begin{center}\includegraphics[width=0.05\linewidth]{images/hex-bigsimr} \end{center}
-
-\end{CodeChunk}
-
-\hypertarget{acknowledgments}{%
-\section*{Acknowledgment(s)}\label{acknowledgments}}
-\addcontentsline{toc}{section}{Acknowledgment(s)}
-
-The authors gratefully acknowledge Heather Knudson's graphic design for the \texttt{bigsimr} R Package.
-The results published here are in whole or part based upon data generated by the TCGA Research Network: \url{https://www.cancer.gov/tcga}.
-
-\hypertarget{coi}{%
-\section*{Disclosure statement}\label{coi}}
-\addcontentsline{toc}{section}{Disclosure statement}
-
-The authors report no conflict of interest.
-
-\hypertarget{funding}{%
-\section*{Funding}\label{funding}}
-\addcontentsline{toc}{section}{Funding}
-
-Research reported in this publication was supported by grants from the National Institutes of General Medical Sciences (5 U54 GM104944, GM103440) from the National Institutes of Health.
-
-\hypertarget{appendix}{%
-\section*{Appendix}\label{appendix}}
-\addcontentsline{toc}{section}{Appendix}
-
-\noindent Consider a bivariate example where we have a correlated \((Y_1, Y_2)^\top\) with \(Y_1\sim N(\mu_1, \sigma_1^2)\) (normal) and \(Y_2\sim LN(\mu_2, \sigma_2^2)\) (lognormal).
-First, let us note that when we generate such a vector from the \texttt{bigsimr} package then in fact it can be represented as
-
-\begin{equation}
-\label{eq:kram1}
-(Y_1, Y_2)^\top \stackrel{d}{=} \left(X_1, e^{X_2}\right)^\top,
-\end{equation}
-
-where \((X_1, X_2)^\top \sim N_2(\boldsymbol \mu, \boldsymbol \Sigma)\), which is bivariate normal with mean vector \(\boldsymbol \mu = (\mu_1, \mu_2)^\top\) and variance-covariance matrix
-
-\begin{equation}
-\label{eq:kram2}
-\boldsymbol \Sigma = 
-\begin{bmatrix}
-\sigma_1^2 & \rho \sigma_1\sigma_2\\
-\rho \sigma_1\sigma_2 & \sigma_2^2
-\end{bmatrix}
-\end{equation}
-
-To see this, consider the three steps of the NORTA construction:
-
-\begin{enumerate}
-
-\item Generate $(Z_1, Z_2)^\top \sim N_2(\boldsymbol 0, \boldsymbol R)$, where
-
-\begin{equation}
-\label{eq:kram3}
-\boldsymbol R = 
-\left[
-\begin{array}{cc}
-1 & \rho \\
-\rho & 1
-\end{array}
-\right].
-\end{equation}
-
-\item Transform $(Z_1, Z_2)^\top$ to $(U_1, U_2)^\top$ viz. $U_i =\Phi(Z_i)$,  $i=1,2$, where $\Phi(\cdot)$ is the standard normal CDF.
-
-\item Return $(Y_1, Y_2)^\top$, where $Y_i=F_i^{-1}(U_i)$, $i=1,2$, and $F_i(\cdot)$ is the CDF of $Y_i$ and $F_i^{-1}(\cdot)$ is its inverse (the quantile function).
-
-\end{enumerate}
-
-In this example, \(F_1\) is the CDF of \(N(\mu_1, \sigma_1^2)\) while \(F_2\) is the CDF of \(LN(\mu_2, \sigma_2^2)\), so that the two required quantile functions are
-
-\begin{equation}
-\label{eq:kram4}
-F_1^{-1}(u) = \mu_1+\sigma_1 \Phi^{-1}(u)\,\,\, \mbox{and} \,\,\, F_2^{-1}(u) = e^{\mu_2+\sigma_2 \Phi^{-1}(u)}, 
-\end{equation}
-
-as can be seen by standard calculation. Thus, when we apply Step 3 of the NORTA algorithm, we get
-
-\begin{equation}
-\label{eq:kram5}
-F_1^{-1}(U_1) = \mu_1+\sigma_1 \Phi^{-1}(\Phi(Z_1)) =  \mu_1+\sigma_1 Z_1 \,\,\, \mbox{and} \,\,\, F_2^{-1}(U_2) = e^{\mu_2+\sigma_2 \Phi^{-1}(\Phi(Z_2))} = e^{\mu_2+\sigma_2 Z_2}, 
-\end{equation}
-
-where \((X_1, X_2)^\top = (\mu_1+\sigma_1 Z_1, \mu_2+\sigma_2 Z_2)^\top \sim N_2(\boldsymbol \mu, \boldsymbol \Sigma)\).
-Consequently, the vector \((Y_1, Y_2)^\top\) obtained in Step 3 of the algorithm has the structure provided by (\ref{eq:kram1}).
-
-\vspace{0.1in}
-
-\noindent Next, we provide the exact covariance structure of the random vector \((Y_1, Y_2)^\top\) given by (\ref{eq:kram1}), and relate the correlation of \(Y_1\) and \(Y_2\) to \(\rho\), which is the correlation of the normal variables \(Z_1\), \(Z_2\) (and also \(X_1\) and \(X_2\)).
-A straightforward albeit somewhat tedious algebra produces the following result.
-
-\begin{lemma}
-Let ${\bf Y} = (Y_1, Y_2)^\top$ admit the stochastic representation (\ref{eq:kram1}), where ${\bf X} = (X_1, X_2)^\top \sim N_2(\boldsymbol \mu, \boldsymbol \Sigma)$ with $\boldsymbol \mu$ and $\boldsymbol \Sigma$ as above.
-Then the mean vector and the variance-covariance matrix of ${\bf Y}$ are given by
-
-\begin{equation}
-\label{eq:kram6}
-\boldsymbol \mu_{{\bf Y}} = \left(\mu_1, e^{\mu_2+\sigma_2^2/2}\right)^\top
-\end{equation}
-
-and
-
-\begin{equation}
-\label{eq:kram7}
-\boldsymbol \Sigma = 
-\begin{bmatrix}
-\sigma_1^2 & \rho \sigma_1\sigma_2  e^{\mu_2+\sigma_2^2/2}  \\
-\rho \sigma_1\sigma_2 e^{\mu_2+\sigma_2^2/2} & \left[ e^{\mu_2+\sigma_2^2/2}\right]^2 \left[e^{\sigma_2^2} -1 \right],
-\end{bmatrix}
-\end{equation}
-
-respectively. 
-\end{lemma}
-
-\begin{proof}
-The values of the means and the variances are obtained immediately from the normal marginal distribution of $Y_1$ and lognormal marginal distribution of $Y_2$.
-It remains to establish the covariance of $Y_1$ and $Y_2$,
-
-\begin{equation}
-\label{eq:covy1y2}
-\mbox{Cov}(Y_1, Y_2) = \mathbb E(Y_1 Y_2) - \mathbb E(Y_1)\mathbb E(Y_2), 
-\end{equation}
-
-and in particular the first expectation on the right-hand-side above. By using the tower property of expectations, the latter expectation can be expressed as
-
-\begin{equation}
-\label{eq:kre1}
-\mathbb E(Y_1 Y_2)  =  \mathbb E\left(X_1 e^{X_2}\right)  =  \mathbb E \left\{ \mathbb E \left(X_1 e^{X_2} | X_1 \right) \right\} = \mathbb E \left\{ X_1 \mathbb E \left(e^{X_2} | X_1 \right) \right\}.
-\end{equation}
-
-Further, by using the fact that the conditional distribution of $X_2$ given $X_1=x_1$ is normal with mean $\mathbb E(X_2|X_1=x_1) = \mu_2+\rho\sigma_2(x_1-\mu_1)/\sigma_1$ and variance $\sigma_2^2(1-\rho^2)$, one can relate the inner expectation on the far right in (\ref{eq:kre1}) to the moment generating function $M(t)$ of this conditional distribution evaluated at $t=1$, leading to
-
-\begin{equation}
-\label{eq:kre2}
-E \left(e^{X_2} | X_1 \right) = e^{ \mu_2+\rho\sigma_2(x_1-\mu_1)/\sigma_1 + \sigma_2^2(1-\rho^2)/2},
-\end{equation}
-
-so that
-
-\begin{equation}
-\label{eq:kre3}
-\mathbb E(Y_1 Y_2)  =  e^{ \mu_2+ \frac{1}{2} \sigma_2^2(1-\rho^2)} \mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\}.
-\end{equation}
-
-Since $X_1$ is normal with mean $\mu_1$ and variance $\sigma_1^2$, the expectation in (\ref{eq:kre3}) becomes
-
-\begin{equation}
-\label{eq:kre4}
-\mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = \int_{-\infty}^\infty x_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (x_1-\mu_1)} \frac{1}{\sqrt{2\pi}\sigma_1} e^{-\frac{1}{2\sigma_1^2}(x_1-\mu_1)^2}.
-\end{equation}
-
-Upon substituting $u=x_1-\mu_1$, followed by some algebra, we arrive at
-
-\begin{equation}
-\label{eq:kre5}
-\mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = e^{\frac{1}{2}\sigma_2^2\rho^2} \int_{-\infty}^\infty (u+\mu_1)  \frac{1}{\sqrt{2\pi}\sigma_1} e^{-\frac{1}{2\sigma_1^2}(u-\rho\sigma_1\sigma_2)^2}.
-\end{equation}
-
-Since the integral in (\ref{eq:kre5}) is the expectation $\mathbb E(X+\mu_1)$, where $X$ is normal with mean $\rho\sigma_1\sigma_2$ and variance $\sigma_1^2$, we conclude that
-
-\begin{equation}
-\label{eq:kre6}
-\mathbb E \left\{ X_1 e^{ \rho\frac{\sigma_2}{\sigma_1} (X_1-\mu_1)} \right\} = e^{\frac{1}{2}\sigma_2^2\rho^2} (\rho\sigma_1\sigma_2 +\mu_1),
-\end{equation}
-
-which, in view of (\ref{eq:kre3}), leads to
-
-\begin{equation}
-\label{eq:kre7}
-\mathbb E (Y_1Y_2) = e^{\mu_2 + \frac{1}{2}\sigma_2^2} (\rho\sigma_1\sigma_2 +\mu_1).
-\end{equation}
-
-Finally, (\ref{eq:covy1y2}), along with the expressions for the means of $Y_1$ and $Y_2$, produce the covariance of $Y_1$ and $Y_2$:
-
-\begin{equation}
-\label{eq:kre8}
-\mbox{Cov}(Y_1, Y_2) = e^{\mu_2 + \frac{1}{2}\sigma_2^2} (\rho\sigma_1\sigma_2 +\mu_1) - \mu_1 e^{\mu_2 + \frac{1}{2}\sigma_2^2} = \rho\sigma_1\sigma_2 e^{\mu_2 + \frac{1}{2}\sigma_2^2}.
-\end{equation}
-
-This concludes the proof
-\end{proof}
-
-Using the above result, we can directly relate the correlation coefficient of \(Y_1\) and \(Y_2\) with that of \(X_1\) and \(X_2\),
-
-\begin{equation}
-\label{eq:kram8}
-\rho_{Y_1, Y_2} = \rho \frac{\sigma_2}{\sqrt{e^{\sigma_2^2} -1}},
-\end{equation}
-
-where \(\rho\) is the correlation of \(X_1\) and \(X_2\).
-The above result is useful when studying the possible range of correlation of \(Y_1\) and \(Y_2\) in the above example.
-It can be shown that the factor on the far right in (\ref{eq:kram8}) is a monotonically decreasing function of \(\sigma_2\) on \((0,\infty)\), with the limits of 1 and 0 at zero and infinity, respectively.
-Thus, in principle, the range of correlation of \(Y_1\) and \(Y_2\) is the same as that of \(X_1\) and \(X_2\), as the factor on the far right in (\ref{eq:kram8}) can be made arbitrarily close to 1.
-However, by changing this factor we may affect the marginal distributions of \(Y_1\) and \(Y_2\).
-It can be shown that if the marginal distributions of \(Y_1\) and \(Y_2\) are fixed, then the relation (\ref{eq:kram8}) becomes
-
-\begin{equation}
-\label{eq:kram9}
-\rho_{Y_1, Y_2} = \rho \sqrt{\frac{\log(1+c_2^2)}{c_2^2}},
-\end{equation}
-
-where \(c_2=\sigma_{Y_2}/\mu_{Y_2}\) is the \emph{coefficient of variation} (CV) of the variable \(Y_2\).
-Thus, the range of possible correlations in this model is not affected by the distribution of \(Y_1\), and is determined by the CV of \(Y_2\) as follows:
-
-\begin{equation}
-\label{eq:kram10}
-- \sqrt{\frac{\log(1+c_2^2)}{c_2^2}} \leq \rho_{Y_1, Y_2} \leq \sqrt{\frac{\log(1+c_2^2)}{c_2^2}}. 
-\end{equation}
-
-Plugging in the estimates of these quantities from Section \ref{package}, we see that
-
-\begin{equation}
-\label{eq:kram11}
-\frac{\hat{\sigma}_2}{\sqrt{e^{\hat{\sigma}_2^2} -1}} = 0.881.
-\end{equation}
-
-Thus, the possible range of correlation becomes \((-0.881, 0.881)\). This agrees with the MC results provided.
+This article proceeds by providing background information, including a
+description of a motivating example application: RNA-sequencing
+(RNA-seq) breast cancer data. Then we describe and justify our
+simulation methodology and related algorithms. We proceed by providing
+an illustrative LD \texttt{bigsimr} workflow. Next we conduct MC studies
+under various bivariate distributional assumptions to evaluate
+performance and accuracy. After the MC evaluations, we simulate random
+vectors motivated by our RNA-seq example, evaluate the accuracy, and
+provide example statistical computing tasks, namely MC estimation of
+joint probabilities and evaluating HD correlation estimation efficiency.
+Finally, we discuss the method's utility, limitations, and future
+directions.
 

### 2021-10-17 030-bigsimr-package.Rmd
Submitting to JSS, final pre-submission edits
moved jss re-submission to a fresh git repo in case editor looks

@@
 )
+Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
+## bs <- bigsimr_setup(pkg_check = FALSE)
+## dist <- distributions_setup()
+```
 
-Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
-bs <- bigsimr_setup(pkg_check = FALSE)
-dist <- distributions_setup()
+```{julia bigsimrSetup}
+using Bigsimr
+using Distributions
 ```
@@
 For the assumed marginals, the Pearson correlation coefficient is not free to vary in [-1,1], as seen below.
-Our MC estimate of the bounds slightly underestimates the theoretic bounds of $(-0.881, 0.881)$.
+Our MC estimate of the bounds slightly underestimates the theoretical bounds of $(-0.881, 0.881)$.
 (See an analytic derivation presented in the Appendix).

### 2021-10-19 arXiv_bigsimr/arXiv_bigsimr.Rmd
preparing arXiv release of bigsimr R focused article

---
title: Simulating High-Dimensional Multivariate Data using the Bigsimr R Package
authors:
  - name: David S. Hippocampus
    thanks: Use footnote for providing further information about author (webpage, alternative address)---*not* for acknowledging funding agencies. Optional.
    department: Department of Computer Science
    affiliation: Cranberry-Lemon University
    location: Pittsburgh, PA 15213
    email: hippo@cs.cranberry-lemon.edu
  - name: Elias D. Striatum
    department: Department of Electrical Engineering
    affiliation: Mount-Sheikh University
    location: Santa Narimana, Levand
    email: stariate@ee.mount-sheikh.edu
abstract: |
  It is critical to accurately simulate data when employing Monte Carlo techniques and evaluating statistical methodology. Measurements are often correlated and high dimensional in this era of big data, such as data obtained in high-throughput biomedical experiments. Due to the computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations. To close this gap, we developed the Bigsimr Julia package with R and Python interfaces. This paper focuses on the R interface. These packages empower high-dimensional random vector simulation with arbitrary marginal distributions and dependency via a Pearson, Spearman, or Kendall correlation matrix. bigsimr contains high-performance features, including multi-core and graphical-processing-unit-accelerated algorithms to estimate correlation and compute the nearest correlation matrix. Monte Carlo studies quantify the accuracy and scalability of our approach, up to $d=10,000$. We describe example workflows and apply to a high-dimensional data set --- RNA-sequencing data obtained from breast cancer tumor samples.
keywords:
  - high-dimensional data
  - multivariate simulation
  - nonparametric correlation
  - Gaussian copula
  - RNA-sequencing data
  - breast cancer
bibliography: references.bib
biblio-style: unsrt
output:
  rticles::arxiv_article:
    includes:
      in_header: preamble.tex
---

# Introduction

Massive high-dimensional (HD) data sets are now common in many areas of scientific inquiry. As new methods are developed for data analysis, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of proposed methodology --- such as false positive rates, statistical power, interval coverage, and robustness. Further, efficient simulation empowers statistical computing strategies, such as the parametric bootstrap [@Chernick2008] to simulate from a hypothesized null model, providing inference in analytically challenging settings. Such Monte Carlo (MC) techniques become difficult for HD dependent data using existing algorithms and tools. This is particularly true when simulating massive multivariate, non-normal distributions, arising in many fields of study.

As others have noted, it can be vexing to simulate dependent, non-normal/discrete data, even for low-dimensional (LD) settings [@MB13; @XZ19]. For continuous non-normal LD multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula approaches [@Nelsen2007] are well-studied and implemented in publicly-available software [@Yan2007; @Chen2001]. Yet these approaches do not scale in a timely fashion to HD problems [@Li2019gpu]. For discrete data, early simulation strategies had major flaws, such as failing to obtain the full range of possible correlations --- such as admitting only positive correlations [@Park1996]. While more recent approaches [@MB13; @Xia17; @BF17] have remedied this issue for LD problems, the existing tools are not designed to scale to high dimensions.

Another central issue lies in characterizing dependence between components in the HD random vector. The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc.). For normal data, the Pearson product-moment correlation describes the dependence perfectly. However, simulating arbitrary random vectors that match a target Pearson correlation matrix is computationally intense [@Chen2001; @Xia17]. On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependence, such as Spearman's $\rho$ and Kendall's $\tau$. Throughout, we focus on matching these nonparametric dependence measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties enabling our proposed methodology. We do, however, implement Pearson matching, but several layers of approximation are required.

With all this in mind, we present a scalable, flexible multivariate simulation algorithm. The crux of the method lies in the construction of a Gaussian copula in the spirit of the NORTA procedure. As we will describe in more detail, the algorithm's design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed-form relationships between dependence measures for the multivariate normal (MVN) distribution. For our method, we developed a high-performance implementation: the `Bigsimr` Julia package, with R and Python interfaces `bigsimr`.

This article proceeds by providing background information, including a description of a motivating example application: RNA-sequencing (RNA-seq) breast cancer data. Then we describe and justify our simulation methodology and related algorithms. We proceed by providing an illustrative LD `bigsimr` workflow. Next we conduct MC studies under various bivariate distributional assumptions to evaluate performance and accuracy. After the MC evaluations, we simulate random vectors motivated by our RNA-seq example, evaluate the accuracy, and provide example statistical computing tasks, namely MC estimation of joint probabilities and evaluating HD correlation estimation efficiency. Finally, we discuss the method's utility, limitations, and future directions.

# Background {#background}

```{r ch010-background, include=FALSE, cache=FALSE, eval=T}
library(dplyr)
library(ggplot2)
library(stringr)
library(GGally)
library(knitr)
load("../data/example_genes.rda")
load("../data/example_brca.rda")
d <- 1000
```

The `Bigsimr` package presented here provides multiple high-performance algorithms that operate with HD multivariate data. All these algorithms were originally designed to support a single task: to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics. Specifically, our goal is to efficiently simulate a large number, $B$, of HD random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with *correlated* components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i, i=1,\ldots,d$.

When designing this methodology, we developed the following properties to guide our effort. We divide the properties into two categories: (1) basic properties (BP) and 'scalability' properties (SP). The BPs are adapted from an existing criteria due to @Nik13a. A suitable simulation strategy should possess the following properties:

* BP1: A wide range of dependences, allowing both positive and negative values, and, ideally, admitting the full range of possible values.
* BP2: Flexible dependence, meaning that the number of dependence parameters can be equal to the number of bivariate pairs in the vector.
* BP3: Flexible marginal modeling, generating heterogeneous data --- including mixed continuous and discrete marginal distributions.

Moreover, the simulation method must *scale* to high dimensions:

* SP1: Procedure must scale to high dimensions with practical computation times.
* SP2: Procedure must scale to high dimensions while maintaining accuracy.

## Motivating example: RNA-seq data

Simulating HD, non-normal, correlated data motivates this work, in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. The RNA-seq data-generating process involves counting how often a particular form of messenger RNA (mRNA) is expressed in a biological sample. RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in HD data. For human-derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted [@Schissler2019]. Importantly, due to inherent biological processes, gene expression data exhibit correlation (co-expression) across genes [@BE07; @Schissler2018].

We illustrate our methodology using the Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments). For ease of modeling and simplicity of exposition, we only consider high expressing genes. In turn, we begin by filtering to retain the top r d of the highest-expressing genes (in terms of median expression) of the over 20,000 gene measurements from $N=r nrow(example_brca)$ patients' tumor samples. This gives a great number of pairwise dependencies among the marginals (specifically, $r choose(d,2)$ correlation parameters). Table \@ref(tab:ch010-realDataTab) displays RNA-seq counts for three selected high-expressing genes for the first five patients' breast tumor samples. To help visualize the bivariate relationships for these three selected genes across all patients, Figure \@ref(fig:ch010-realDataFig) displays the marginal distributions and estimated Spearman's correlations.

```{r ch010-realDataTab, eval=T}
set.seed(2020-02-25)
num_genes    <- 3
num_patients <- 5
gene_sample  <- sample(example_genes[1:1000], num_genes)

cap <- paste0("mRNA expression for three selected high-expressing genes, ",
              paste(gene_sample, collapse = ", "), 
              ", for the first five patients in the TCGA BRCA data set.")

small_brca <- example_brca %>%
  select(all_of(gene_sample)) %>%
  head(n = num_patients) %>%
  as_tibble(rownames = "Patient ID") %>%
  mutate(`Patient ID` = str_sub(`Patient ID`, end = 12))

small_brca %>%
  kable(format = "latex", booktabs = TRUE, caption = cap)
```

```{r ch010-realDataFig, cache=FALSE, fig.width=8, fig.cap="Pairwise scatterplots for three example genes (Table 1), with estimated Spearman's correlations and marginal density estimates displayed. The data possess outliers, heavy-right tails, are discrete, and have non-trivial intergene correlations. Modeling these data motivate our simulation methodology.", eval=T}
ggpairs(
  data = example_brca[, gene_sample],
  upper = list(continuous = wrap('cor', method = "spearman"))
) + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = -90, vjust = 0.5))
```

## Measures of dependency

In multivariate analysis, an analyst must select a metric to quantify dependency. The most widely-known is the Pearson correlation coefficient that describes the linear association between two random variables $X$ and $Y$, and, it is given by

\begin{equation}
\rho_P(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ \mathrm{Var}(X)\mathrm{Var}(Y)\right]^{1/2}}.
\label{eq:pearson}
\end{equation}

As @MB13 and @MK01 discuss, for a bivariate normal $(X,Y)$ random vector, the Pearson correlation completely describes the dependency between the components. For non-normal marginals with monotone correlation patterns, however, $\rho_P$ suffers some drawbacks and may mislead or fail to capture important relationships [@MK01]. Alternatively, analysts often prefer rank-based correlation measures to describe the degree of monotonic association.

Two nonparametric, rank-based measures common in practice are Spearman's correlation (denoted $\rho_S$) and Kendall's $\tau$. Spearman's $\rho_S$ has an appealing correspondence as the Pearson correlation coefficient on *ranks* of the values, thereby capturing nonlinear yet monotone relationships. Kendall's $\tau$, on the other hand, is the difference in probabilities of concordant and discordant pairs of observations, $(X_i, Y_i)$ and $(X_j, Y_j)$ (concordance meaning that orderings have the same direction, e.g., if $X_i < X_j$, then $Y_i < Y_j$). Note that concordance is determined by the ranks of the values, not the values themselves. Both $\tau$ and $\rho_S$ are *invariant under monotone transformations* of the underlying random variates. As we will describe more fully in the [Algorithms](#algorithms) section, this property enables matching rank-based correlations with speed (SP1) and accuracy (SP2).

*Correspondence among Pearson, Spearman, and Kendall correlations*

There is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficient, as the marginal distributions $F_i$ are intrinsic in their calculation. For *bivariate normal vectors*, however, the correspondence is well-known:

\begin{equation}
\label{eq:convertKendall}
\rho_{P} = \sin \left( \tau \times \frac{\pi}{2} \right), 
\end{equation}

\noindent and similarly for Spearman's $\rho$ [@K58],

\begin{equation}
\label{eq:convertSpearman}
\rho_P = 2 \times \sin \left( \rho_S \times \frac{\pi}{6} \right).
\end{equation}

*Marginal-dependent bivariate correlation bounds*

Given two marginal distributions, $\rho_P$ is not free to vary over the entire range of possible correlations $[-1,1]$. The *Frechet-Hoeffding bounds* are well-studied [@Nelsen2007; @BF17]. These constraints cannot be overcome through algorithm design. In general, the bounds are given by

\begin{equation}
\label{eq:frechet}
\rho_P^{max} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (U) \right), \quad \rho_P^{min} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (1 - U) \right)
\end{equation}

\noindent where $U$ is a uniform random variable on $(0,1)$ and $F^{-1}_1, F^{-1}_2$ are the inverse CDFs of $X_1$ and $X_2$, respectively, defined as

\begin{equation}
F_{i}^{-1} = \inf\{y:F_{i}(y) \geq u \}
\label{eq:inverseCDF}
\end{equation}

when the variables are discrete.

## Gaussian copulas

There is a connection of our simulation strategy to Gaussian *copulas* [@Nelsen2007]. A copula is a distribution function on $[0,1]^d$ that describes a multivariate probability distribution with standard uniform marginals. This provides a powerful, natural way to characterize joint probability. Consequently, the study of copulas is an important and active area of statistical theory and practice. For any random vector ${\bf X}=(X_1, \ldots, X_d)^\top$ with CDF $F$ and marginal CDFs $F_i$ there is a copula function $C(u_1, \ldots, u_d)$ satisfying

\begin{equation}
F(x_1, \ldots, x_d) = {\mathbb P}(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), \quad x_i \in {\mathbb R}, i=1,\ldots,d.
\label{eq:copula}
\end{equation}

A Gaussian copula has marginal CDFs that are all standard normal, $F_i = \Phi, \forall \, i$, . This representation corresponds to a multivariate normal (MVN) distribution with standard normal marginal distributions and covariance matrix $R_P$. Since the marginals are standardized to have unit variance, this $R_P$ is also Pearson correlation matrix. If $F_R$ is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula $C_R$ is defined through

\begin{equation}
\label{eq:gauss}
F_R(x_1, \ldots, x_d) = C_R(\Phi(x_1), \ldots, \Phi(x_d)),
\end{equation}

where $\Phi(\cdot)$ is the standard normal CDF.

Here, the copula function $C_R$ is the CDF of the random vector $(\Phi(X_1), \ldots, \Phi(X_d))$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_P)$.

Sklar's Theorem [@Ubeda-Flores2017] guarantees that given inverse CDFs $F_i^{-1}$s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions. For example, using Gaussian copulas, we can construct a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)^\top$ with $Y_i \sim F_i$, viz. $Y_i = F_i^{-1}(\Phi(X_i)), i=1, \ldots, d$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_P)$.

## Other Multivariate Simulation Packages

To place `Bigsimr`'s implementation and algorithms in context of existing approaches, we qualitatively compare `Bbigsimr` to other multivariate simulation packages.  In `R`, the `MASS` package's `mvnorm()` has long been available to produce random vectors from multivariate normal (and $t$) distributions. This procedure cannot simulate from arbitrary margins and dependency measures and it fails to readily scale to high dimensions. SAS PROC SIMNORMAL and Python's NumPy `random.multivariate_normal` both do similar tasks. Recently, the high-performance `R` package `mvnfast` [@Fasiolo2016] has been released, which provides HD multivariate normal generation. However, lacks flexibility in marginal models and dependency modeling. For dependent discrete data, LD `R` packages exist, such as `GenOrd`. For copula computation, the full-featured `R` package `copula` provides many different copulas and a fairly general LD random generator `rCopula()`. High-dimesional random vector generation via `copula` takes an impractical amount of computing time [@Li2019gpu]. The `nortaRA` R package matches Pearson coefficients near exactly and is reasonable for LD settings. Finally, we know of no other `R` package that can produce HD random vectors with flexible margins and dependence structure. Table \@ref(tab:compare-table) summarizes the above discussion.

\begin{table}[h]
\centering
\caption{\label{tab:compare-table}Comparison of selected features of bigsimr to existing R packages for multivariate simulation.}
\begin{tabular}{@{}llll@{}}
Package & Scales to HD & Hetereogenous Margins & Flexible Dependence \\ \midrule
MASS      & No           & No                    & No                  \\
mvnfast   & Yes          & No                    & No                  \\
GenOrd    & No           & Discrete margins      & Yes                 \\
copula    & No           & Yes                   & No                  \\
nortaRA   & No           & Yes                   & No                  \\
bigsimr   & Yes          & Yes                   & Yes                
\end{tabular}
\end{table}

# Algorithms {#algorithms}

This section describes our methods for simulating a random vector $\bf Y$ with $Y_i$ components for $i=1,\ldots,d$. Each $Y_i$ has a specified marginal CDF $F_i$ and its inverse $F^{-1}_i$. To characterize dependency, every pair $(Y_i, Y_j)$ has a given Pearson correlation $\rho_P$, Spearman correlation $\rho_S$, and/or Kendall's $\tau$. The method can be described as a *high-performance Gaussian copula* (Equation \@ref(eq:gauss)) providing a HD NORTA-inspired algorithm.

## NORmal To Anything (NORTA)

The well-known NORTA algorithm [@Cario1997] simulates a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$. Specifically, NORTA algorithm proceeds as follows:

1. Simulate a random vector $\bf Z$ with $d$ independent and identically distributed (iid) standard normal components.
2. Determine the input matrix $\Sigma_{\bf Z}$ that corresponds with the specified output $\Sigma_{\bf Y}$.
3. Produce a Cholesky factor $M$ of $\Sigma_{\bf Z}$ such that $M M^{\prime}=\Sigma_{\bf Z}$.
4. Set $X$ by $X \gets MZ$.
5. $\text{Return} \; {\bf Y} \; \text{where} \; Y_i \gets F_i^{-1}[\Phi(X_i)], \; i=1,...,d$.

With modern parallel computing, steps 1, 3, 4, 5 are readily implemented as high-performance, multi-core and/or graphical-processing-unit (GPU) accelerated algorithms --- providing fast scalability to HD.

Matching specified Pearson correlation coefficients exactly (step 2 above), however, is computationally costly. In general, there is no convenient correspondence between the components of the input $\Sigma_{\bf Z}$ and target $\Sigma_{\bf Y}$. Matching the correlations involves evaluating or approximating $\binom{d}{2}$ integrals of the form

\begin{equation}
    \mathrm{E}\left[Y_i Y_j\right] = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} F_i^{-1}\left[\Phi(z_i)\right] F_j^{-1}\left[\Phi(z_j)\right] \phi(z_i, z_j, \rho_z) dz_i dz_j,
    \label{eq:pearsonIntegralRelation}
\end{equation}

where $\phi(\cdot)$ is the joint probability density function of two correlated standard normal variables. For HD simulation, exact evaluation may become too costly to enable practical simulation studies. Our remedy to enable HD Pearson matching in `bigsimr` is Hermite polynomial approximation of these $\binom{d}{2}$ double integrals, as suggested by @XZ19.

*NORTA in higher dimensions*

Sklar's theorem provides a useful characterization of multivariate distributions through copulas. In practice, however, the particulars of the simulation algorithm affect which joint distributions may be simulated. Even in LD spaces (e.g., $d=3$), there exist valid multivariate distributions with *feasible* Pearson correlation matrices that NORTA cannot match exactly [@LH75]. This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined, the resultant matrix becomes non-positive definite. These situations do occur, even using exact analytic calculations. Such problematic target correlation matrices are termed *NORTA defective*.

@GH02 conducted a study to estimate the probability of encountering NORTA defective matrices while increasing the dimension $d$. They found that for what is now considered low-to-moderate dimensions ($d \approx 20$), almost *all* feasible matrices are NORTA defective. This stems from the concentration of measure near the boundary of the space of all possible correlation matrices as dimension increases. Unfortunately, it is precisely near this boundary that NORTA defective matrices reside.

There is hope, however, as @GH02 also showed that replacing an non-positive definite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$. This provides evidence that our nearest positive definite (PD) augmented approach has promise to provide adequate accuracy if our input matching scheme returns an indefinite Pearson correlation matrix.

## Random vector generator {#rand-vec-gen}

We now describe our algorithm to generate random vectors, which extends NORTA to HD via nearest correlation matrix replacement and provides rank-based dependency matching:

1. Mapping step, either
    * Convert the target Spearman correlation matrix $R_S$ to the corresponding MVN Pearson correlation $R_X$. Alternatively,
    * Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$. Alternatively,
    * Convert the target Pearson correlation matrix to $R_P$ to the corresponding, approximate MVN Pearson correlation $R_X$.
2. Check correlation matrix admissibility and, if not, compute the nearest correlation matrix.
    * Check that $R_X$ is a correlation matrix, a positive definite matrix with 1's along the diagonal.
    * If the mapping produced an $R_X$ that is a correlation matrix, we describe the matrix as **admissible** in this scheme and retain $R_X$. Otherwise,
    * Replace $R_X$ with the nearest correlation matrix $\tilde{R}_X$, in the Frobenius norm and the result will be approximate.
3. Gaussian copula
    * Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)$.
    * Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz. $U_i=\Phi(X_i)$, $i=1, \ldots, d$.
    * Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$.

*Step 1: Mapping step.* We first employ the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertKendall) and \@ref(eq:convertSpearman), respectively (implemented as `cor_covert()`). Initializing our algorithm to match the nonparametric correlations by computing these equations for all pairs is computationally trivial.

For the computationally expensive process to match Pearson correlations, we approximate Equation \@ref(eq:pearsonIntegralRelation) for all pairs of margins. To this end, we implement the approximation scheme introduced by [@XZ19]. Briefly, the many double integrals of the form in \@ref(eq:pearsonIntegralRelation) are approximated by weighted sums of Hermite polynomials. Matching coefficients for pairs of continuous distributions is made tractable by this method, but for discrete distributions (especially discrete distributions with large support sets or infinite support), the approximation is computationally expensive. To remedy this, we further approximate discrete distributions by a continuous distribution using Generalized S-Distributions [@muino2006gs].

*Step 2: Admissibility check and nearest correlation matrix computation.* Once $R_X$ has been determined, we check admissibility of the adjusted correlation via the steps described above. If $R_X$ is not a valid correlation matrix, then we compute the nearest correlation matrix. Finding the nearest correlation matrix is a common statistical computing problem. The defacto function in `R` is `Matrix::nearPD`, an alternating projection algorithm due to @higham2002computing. As implemented, the function fails to scale to HD. Instead, we provide the quadratically-convergent algorithm based on the theory of strongly semi-smooth matrix functions [@qi2006quadratically]. The nearest correlation matrix problem can be written down as the following convex optimization problem:  $\mathrm{min} \quad \frac{1}{2} \Vert R_X - X \Vert^2, \quad \mathrm{s.t.} \quad X_{ii} = 1, \quad i = 1, \ldots , n, \quad X \in S_{+}^{n}$.

For nonparametric correlation measures, our algorithm allows the generation of HD multivariate data with arbitrary marginal distributions with a broad class of admissible Spearman correlation matrices and Kendall $\tau$ matrices. The admissible classes consist of the matrices that map to a Pearson correlation matrix for a MVN. In particular, if we let $X$ be MVN with $d$ components and denote $\Omega_P = \{ R_P : R_P \textrm{ is a Pearson correlation matrix for } X \}, \quad \Omega_K = \{ R_K : R_K \textrm{ is a Kendall correlation matrix for } X \}, \quad \Omega_S = \{ R_S : R_S \textrm{ is a Spearman correlation matrix for } X \}$. There are 1-1 mappings between these sets. We conjecture that the sets of admissible $R_S$ and $R_K$ are not highly restrictive. In particular, $R_P$ is approximately $R_S$ for a MVN, suggesting that the admissible set $\Omega_S$ should be flexible as $R_P$ can be any PD matrix with 1's along the diagonal. We provide methods to check whether a target $R_S$ is an element of the *admissible set* $\Omega_S$.

There is an increasing probability of encountering an inadmissible correlation matrix as dimension increases. In our experience, the mapping step for large $d$ almost always produces a $R_X$ that is not a correlation matrix. In Section \@ref(package), we provide a basic method of how to quantify and control the approximation error. Further, the RNA-seq example in Section \@ref(examples) provides an illustration of this in practice.

*Step 3: Gaussian copula.* The final step implements a NORTA-inspired, Gaussian copula approach to produce the desired margins. Steps 1 and 2 determine the MVN Pearson correlation values that will eventually match the target correlation. Step 3 requires a fast MVN simulator, a standard normal CDF, and well-defined quantile functions for marginals. The MVN is transformed to a copula (distribution with standard uniform margins) by applying the normal CDF $\Phi(\cdot)$. Finally, the quantile functions $F_i^{-1}$ are applied across the margins to return the desired random vector ${\bf Y}$.

# The bigsimr R package

```{r ch030-basic-example-setup, cache=FALSE, eval=T, echo=F}
library(patchwork)
library(tidyr)
library(bigsimr)
Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores()-1)
bs <- bigsimr_setup(pkg_check = FALSE)
dist <- distributions_setup()
```

This section describes a bivariate random vector simulation workflow via the `bigsimr` `R` package. This `R` package provides an interface to the native code written in Julia (registered as the `Bigsimr` Julia package). In addition to the native Julia `Bigsimr` package and `R` interface `bigsimr`, we also provide a Python interface `bigsimr` that interfaces with the Julia `Bigsimr` package. The Julia package provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms)).

The subsections below describe the basic use of the `bigsimr` R package by stepping through an example workflow using the data set `airquality` that contains daily air quality measurements in New York, May to September 1973 [@Chambers1983]. This workflow proceeds from setting up the computing environment, to data wrangling, estimation, simulation configuration, random vector generation, and, finally, result visualization.

## Bivariate example

We illustrate the use of `bigsimr` using the New York air quality data set (`airquality`) included in the R `datasets` package. First, we load the `bigsimr` library and a few other convenient data science packages, including the syntactically-elegant `tidyverse` suite of `R` packages. The code chunk below prepares the computing environment:

```{r ch030-computeEnvironmentSetup, eval=FALSE, echo=TRUE}
library("tidyverse")
library("bigsimr")
# Activate multithreading in Julia
Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
# Load the Bigsimr and Distributions Julia packages
bs <- bigsimr_setup()
dist <- distributions_setup()
```

Here, we describe a minimal working example --- a bivariate simulation of two airquality variables: `Temperature`, in degrees Fahrenheit, and `Ozone` level, in parts per billion.

```{r ch030-air-quality-data, echo=TRUE, eval=T}
df <- airquality %>% select(Temp, Ozone) %>% drop_na()
glimpse(df)
```

Figure \@ref(fig:ch030-aq-joint-dist) visualizes the bivariate relationship between `Ozone` and `Temperature`. We aim to simulate random two-component vectors mimicking this structure. The margins are not normally distributed; `Ozone` level exhibits a strong positive skew.

```{r ch030-aq-joint-dist, cache=FALSE, fig.width= 8, fig.cap="Bivariate scatterplot of Ozone versus Temp with estimated marginal densities. The Ozone data are modeled marginally as log-normal and the Temperature data as normal.", eval=T}
p0 <- ggplot(df, aes(Temp, Ozone)) +
  geom_point(size = 1) +
  theme(legend.position = "none") + 
  labs(x = "Temperature")

pTemp <- ggplot(df, aes(Temp)) + 
  geom_density() +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())

pOzone <- ggplot(df, aes(Ozone)) + 
    geom_density() +
  theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank()) +
  coord_flip()

pTemp + plot_spacer() + p0 + pOzone + 
  plot_layout(widths = c(3,1), heights = c(1, 3)) 
```

Next, we specify the marginal distributions and correlation coefficient (both type and magnitude). Here the analyst is free to be creative. For this example, we avoid goodness-of-fit considerations to determine the marginal distributions. It is sensible without domain knowledge to estimate these quantities from the data, and `bigsimr` contains fast functions designed for this task.

## Specifying marginal distributions

Based on the estimated densities in Figure \@ref(fig:ch030-aq-joint-dist), we assume `Temp` is normally distributed and `Ozone` is log-normally distributed, as the latter values are positive and skewed. We use the classical unbiased estimators for the normal distribution's parameters and maximum likelihood estimators for the log-normal parameters:

```{r ch030-aq-temp-pars, echo=TRUE, eval=T}
df %>% select(Temp) %>% 
  summarise_all(.funs = c(mean = mean, sd = sd))
```

```{r ch030-aq-ozone-pars, echo=TRUE, eval=T}
mle_mean <- function(x) mean(log(x))
mle_sd <- function(x) mean( sqrt( (log(x) - mean(log(x)))^2 ) )
df %>% 
  select(Ozone) %>% 
  summarise_all(.funs = c(meanlog = mle_mean, sdlog = mle_sd))
```

Now, we configure the input marginals for later input into `rvec`. The marginal distributions are specified using `Julia`'s `Distributions` package and stored in a vector.

```{r ch030-margins-alist, echo=TRUE, eval=T}
margins <- c(dist$Normal(mean(df$Temp), sd(df$Temp)),
             dist$LogNormal(mle_mean(df$Ozone), mle_sd(df$Ozone)))
```

## Specifying correlation

The user must decide how to describe correlation based on the particulars of the problem. For non-normal data in our scheme, we advocate the use of Spearman's $\rho$ correlation matrix $R_S$ or Kendall's $\tau$ correlation matrix $R_K$. We also support Pearson correlation coefficient matching, while cautioning the user to check the performance for the distribution at hand (see [Monte Carlo evaluations](#simulations) below for evaluation strategies and guidance). Note that these estimation methods are classical approaches, not specifically designed for high-dimensional correlation estimation.

```{r ch030-aq-cor, echo=TRUE, eval=T}
(R_S <- bs$cor(as.matrix(df), bs$Spearman))
```

## Checking target correlation matrix admissibility

Once a target correlation matrix is specified, first convert the dependency type to Pearson to correctly specify the MVN inputs. Then, check that the converted matrix $R_X$ is a valid correlation matrix (PD and 1s on the diagonal). For this bivariate example, $R_X$ is a valid correlation matrix. Note that typically in HD the resultant MVN Pearson correlation matrix is indefinite and requires approximation (see the HD examples in subsequent sections).

```{r ch030-cor-convert, echo=TRUE, eval=T}
# Step 1. Mapping
(R_X <- bs$cor_convert(R_S, bs$Spearman, bs$Pearson))
```

```{r ch030-cor-check, echo=TRUE, eval=T}
# Step 2. Check admissibility
bs$iscorrelation(R_X)
```

Despite being a valid correlation matrix under this definition, marginal distributions induce Frechet bounds on the possible Pearson correlation values. To check that the converted correlation values fall within these bounds, use `bigsimr::cor_bounds` to estimate the pairwise lower and upper correlation bounds. `cor_bounds` uses the Generate, Sort, and Correlate algorithm of @DH2011. For the assumed marginals, the Pearson correlation coefficient is not free to vary in [-1,1], as seen below. Our MC estimate of the bounds slightly underestimates the theoretical bounds of $(-0.881, 0.881)$. (See an analytic derivation presented in the Appendix). Since our single Pearson correlation coefficient is within the theoretical bounds, the correlation matrix is valid for our simulation strategy.

<!-- i guess another point is that even if all the pairwise frechet bounds are satisfied - does that imply the proposed correlation matrix is valid or not - ie does there exist a multivariate distribution with all those pairwise correlations??? some discussion here might help (help me at least)... maybe put another way - need to clarify how useful those bounds are  -->

```{r ch030-cor-bounds-pearson, echo=TRUE, eval=T}
bs$cor_bounds(margins[1], margins[2], bs$Pearson, n_samples = 1e6)
```

In this 2D example, it is possible to verify the existence of a bivariate distribution with the target Pearson correlation / margins (See the [Appendix]).
For general $d-$variate distributions, is it not guaranteed that such a multivariate distribution exists with exact the specified correlations and margins [@BF17].
Theoretical guidance is lacking in this regard, but, in practice, we find that using bivariate bounds form rough approximations to the bounds for the feasible region for the $d-$variate construction when aggregated into a Pearson matrix.
See the [RNA-seq data application] section for an example discussing this, NORTA feasibility, and use of the nearest PD matrix algorithm.

## Simulating random vectors

Finally, we execute `rvec` to simulate the desired $10,000$ random vectors from the assumed bivariate distribution of `Ozone` and `Temp`. Note that the target input is specified using the pre-computed MVN correlation matrix. Figure \@ref(fig:ch030-plot-sim) plots the 10,000 simulated points.

```{r ch030-sim-margins, echo=TRUE, eval=T}
x <- bs$rvec(10000, R_X, margins)
df_sim <- as.data.frame(x)
colnames(df_sim) <- colnames(df)
```

```{r ch030-plot-sim, fig.cap="Contour plot and marginal densities for the simulated bivariate distribution of Air Quality Temperatures and Ozone levels. The simulated points mimic the observed data with respect to both the marginal characteristics and bivariate association.", eval=T}
p1 <- df_sim %>%
  ggplot(aes(Temp, Ozone)) +
  geom_density_2d_filled() +
  theme(legend.position = "none") +
  labs(x = "Simulated Temperature", y = "Simulated Ozone") +
  scale_y_continuous(limits = c(0, max(df$Ozone))) +
  scale_x_continuous(limits = range(df$Temp))

p1Temp <- ggplot(df_sim, aes(Temp)) + 
  geom_density(alpha = 0.5, fill = "lightseagreen") +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())

p1Ozone <- ggplot(df_sim, aes(Ozone)) + 
  geom_density(alpha = 0.5, fill = "lightseagreen") + 
  theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank()) +
  coord_flip()

p1Temp + plot_spacer() + p1 + p1Ozone + 
  plot_layout(widths = c(3,1), heights = c(1, 3))
```

# Monte Carlo evaluations {#simulations}

```{r ch040-LoadLib040, include=FALSE, cache=FALSE, eval=T}
library(RColorBrewer)
load("../data/bivariate_normal_sims.rda")
load("../data/bivariate_gamma_sims.rda")
load("../data/bivariate_nbinom_sims.rda")
load("../data/benchmark_dependences.rda")
prob <- 3e-04
```

In this section, we conduct MC studies to investigate method performance. Marginal parameter matching is straightforward in our scheme as it is a sequence of univariate inverse probability transforms. This process is both fast and exact. On the other hand, accurate and computationally efficient dependency matching presents a challenge, especially in the HD setting. To evaluate our methods in those respects, we design the following numerical experiments to first assess accuracy of matching dependency parameters in bivariate simulations and then time the procedure in increasingly large dimension $d$.

## Bivariate experiments

We select bivariate simulation configurations to ultimately simulate our motivating discrete-valued RNA-seq example. Thus, we proceed by increasing departure from normality, leading to a multivariate negative binomial (MVNB) model for our motivating data. We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson, Spearman, and Kendall --- in identical, bivariate marginal configurations. For each pair of identical margins, we vary the target correlation across $\Omega$, the set of possible admissible values for each correlation type, to evaluate the simulation's ability to generate all possible correlations. The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (non-normal and discrete).

Table \@ref(tab:sims) lists our identical-marginal, bivariate simulation configurations. We increase the simulation replicates $B$ to visually ensure that our results converge to the target correlations and gauge efficiency. We select distributions beginning with a standard multivariate normal (MVN) as we expect the performance to be exact (up to MC error) for all correlation types. Then, we select a non-symmetric continuous distribution: a standard (rate =1), two-component multivariate gamma (MVG). Finally, we select distributions and marginal parameter values that are motivated by our RNA-seq data, namely values proximal to probabilities and sizes estimated from the data (see [RNA-seq data application](examples) for estimation details). Specifically, we assume a MVNB with $p_1 = p_2 = 3\times10^{-4}, r_1 = r_2 = 4, \rho \in \Omega$.

\begin{table}[]
\centering
\caption{ \label{tab:sims} Identical margin, bivariate simulation configurations to evaluate matching accuracy.}
\begin{tabular}{@{}lcr@{}}
\toprule
Simulation Reps $B$ & Correlation Types & Identical-margin 2D distribution \\ \midrule
$1,000$ & Pearson ($\rho_P$) & ${\bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho_i ), i=1,\ldots,100$ \\
$10,000$ & Spearman ($\rho_S$) & ${\bf Y} \sim MVG( shape = 10, rate = 1, \rho_i ), i=1,\ldots,100$ \\
$100,000$ & Kendall ($\tau$) & ${\bf Y} \sim MVNB(p = 3\times10^{-4}, r = 4,\rho_i), i=1,\ldots,100$ \\ \bottomrule
\end{tabular}
\end{table}

For each of the unique 9 simulation configurations described above, we estimate the correlation bounds and vary the correlations along a sequence of 100 points evenly placed within the bounds, aiming to explore $\Omega$. Specifically, we set correlations $\{ \rho_1 = ( \hat{l} + \epsilon), \rho_2 = (\hat{l} + \epsilon) + \delta, \ldots, \rho_{100} = (\hat{u} - \epsilon) \}$, with $\hat{l}$ and $\hat{u}$ being the estimated lower and upper bounds, respectively, with increment value $\delta$. The adjustment factor, $\epsilon=0.01$, is introduced to handle numeric issues when the bound is specified exactly.

```{r ch040-biNormPlot, eval=T, fig.height=5, fig.width=8, fig.cap="`bigsimr` recovers the Pearson specified correlations for MVN."}
bivariate_normal_sims %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    geom_point() +
    geom_abline(slope = 1) +
    facet_wrap(~ type + N) + 
    theme_bw()
```

```{r ch040-biGammaPlot, eval=T, fig.height=5, fig.width=8, fig.cap="`bigsimr` recovers the Pearson specified correlations for Bivariate Gamma."}
bivariate_gamma_sims %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    geom_point() +
    geom_abline(slope = 1) +
    facet_wrap(~ type + N) + 
    theme_bw()
```

```{r ch040-biNegBinPlot, eval=T, fig.width=8, fig.height=5, fig.cap="`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
bivariate_nbinom_sims %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    geom_point() +
    geom_abline(slope = 1) +
    facet_wrap(~ type + N) + 
    theme_bw()
```

```{r ch040-combineBiSims, cache=FALSE, eval=T}
allDat <- bind_rows(
    select(bivariate_normal_sims, margins, type, N, rho, rho_hat),
    select(bivariate_gamma_sims, margins, type, N, rho, rho_hat),
    select(bivariate_nbinom_sims, margins, type, N, rho, rho_hat)
) %>%
    mutate(margins = factor(margins, levels=c("norm", "gamma", "nbinom")),
           type = factor(type, levels=c("Pearson", "Spearman", "Kendall")),
           N = factor(N))
```

Figure \@ref(fig:ch040-bPlot) displays the aggregated bivariate simulation results. Table \@ref(tab:ch040-BiError) contains the mean absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios. Overall, the studies show that our methodology is generally accurate across the entire range of possible correlation for all three dependency measures. Our Pearson matching performs nearly as well as Spearman or Kendall, except for a slight increase in error for the negative binomial case.

```{r ch040-BiError, eval=T}
tabMAE <- allDat %>%
    group_by(N, type, margins) %>%
    summarize(MAE = mean(abs(rho - rho_hat))) %>%
    ungroup()

kable(tabMAE, booktabs = TRUE, format = "latex",
      linesep = c("", "", "\\addlinespace"),
      col.names = c("No. of random vectors",
                    "Correlation type",
                    "Distribution",
                    "Mean abs. error"),
      caption = "Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.")
```

```{r ch040-bPlot, cache=FALSE, fig.asp=1.30, fig.weight=4, fig.cap="Bivariate simulations match target correlations across the entire range of feasible correlations. The horizontal axis plots the specified target correlations for each bivariate margin. Normal margins are plotted in dark dark grey, gamma in medium grey, and negative binomial in light grey. As the number of simulated vectors $B$ increases from left to right, the variation in estimated correlations (vertical axis) decreases. The dashed line indicates equality between the specified and estimated correlations.", eval=T}
# https://www.datanovia.com/en/blog/how-to-change-ggplot-facet-labels/
# New facet label names
repsLabs <- paste0("B=", c("1,000", "10,000", "100,000") )
names(repsLabs) <- c(1000, 10000, 100000)
typeLabs <- c("Pearson", "Spearman", "Kendall")
names(typeLabs) <- c("Pearson", "Spearman", "Kendall")

# Set colors
## RColorBrewer::display.brewer.all()
numColors <- 4
numGroups <- length(levels(allDat$margins))
myColors <- rev(RColorBrewer::brewer.pal(n = numColors, name = "Greys")[ ((numColors - numGroups) + 1):numColors])

allDat %>%
    ggplot(aes(x = rho, y = rho_hat, color = margins)) +
    geom_point(size = 2) +
    scale_color_manual(values = myColors ) +
    geom_abline(slope = 1, linetype = "dashed") +
    labs(x = "Specified Correlation", y = "Estimated Correlation") +
    facet_wrap(~ type + N, labeller = labeller(N = repsLabs, type = typeLabs)) +
    theme_bw() +
    theme(legend.position = "bottom", legend.direction = "horizontal")
```

## Scale up to High Dimensions

With information of our method's accuracy from a low-dimensional perspective, we now assess whether `bigsimr` can scale to larger dimensional problems with practical computation times. We ultimately generate $B=1,000$ random vectors for $d=\{100, 250, 500, 1000, 2500, 5000, 10000\}$ for each correlation type, {Pearson, Spearman, Kendall} while timing the algorithm's major steps. We begin by producing a synthetic "data set" by completing the following steps:

1. Produce heterogeneous gamma marginals by randomly selecting the $j^{th}$ gamma shape parameter from $U_j \sim uniform(1,10), j=1,\ldots,d$ and the $j^{th}$ rate parameter from $V_j \sim exp(1/5), j=1,\ldots,d$, with the constant parameters determined arbitrarily.
2. Produce a random full-rank Pearson correlation matrix via `cor_randPD` of size $d \times d$.
3. Simulate a "data set" of $1,000 \times d$ random vectors via `rvec`.

With the synthetic data set in hand, we complete and time the following four steps involved in a typical workflow (including correlation estimation).

1. Estimate the correlation matrix from the "data" in the *Compute Correlation* step.
2. Map the correlations to initialize the algorithm (Pearson to Pearson, Spearman to Pearson, or Kendall to Pearson) in the *Adjust Correlation* step.
3. Check whether the mapping produces a valid correlation matrix and, if not, find the nearest PD correlation matrix in the *Check Admissibility* step.
4. Simulate $1,000$ vectors in the *Simulate Data* step.

The experiments are conducted on a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, with all 16 threads employed during computation. Table \@ref(tab:ch040-moderateDtab) displays the total computation time for moderate dimesions ($d \leq 500$). In every simullation setting, the 1,000 vectors are generated rapidly, executing in under 2 seconds.

```{r ch040-moderateDtab, eval=T}
dat_long <- benchmark_dependences %>%
  select(-total_time, -n_sim, -needed_near_pd) %>%
  pivot_longer(cols = c(corr_time:sim_time), names_to = "Step", values_to = "Time") %>%
  mutate(dim = factor(dim),
         Step = factor(Step,
                       levels = c("corr_time", "adj_time", "admiss_time", "sim_time"),
                       labels = c("Compute Correlation",
                                  "Adjust Correlation",
                                  "Check Admissibility",
                                  "Simulate Data"))) %>%
    rename(Correlation = corr_type, Dimensions = dim)

tabTimes  <- dat_long %>%
    filter(Dimensions %in% c(100, 250, 500)) %>%
    group_by(Dimensions, Correlation) %>%
    summarize('Total Time(Seconds)' = sum(Time)) %>%
    ungroup()

tabTimes %>%
    kable(booktabs = TRUE,
          linesep = c("", "", "\\addlinespace"), format="latex",
          col.names = c("Dimension",
                        "Correlation type",
                        "Total Time (Seconds)"),
          caption = 'Total time to produce 1,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.')
```

The results for $d > 500$ show scalability to ultra-high dimensions for all three correlation types, although the total times do become much larger. 
Figure \@ref(fig:ch040-largeDfig) displays computation times for $d=\{1000, 2500, 5000, 10000\}$. For $d$ equal to 1000 and 2500, the total time is under a couple of minutes. At $d$ of 5000 and 10,000, Pearson correlation matching in the *Adjust Correlation* step becomes costly. Interestingly, Pearson is actually faster than Kendall for $d=10,000$ due to bottlenecks in *Compute Correlation* and *Check Admissibility*. Uniformly, matching Spearman correlations is faster, with total times under 5 minutes for $d=10,000$, making Spearman the most computationally-friendly dependency type for our package. With this in mind, we scaled the simulation to $d=20,000$ for the Spearman type and obtained the 1,000 vectors in under an hour (data not shown). In principle, this would enable the simulation of an entire human-derived RNA-seq data set. We note that for a given target correlation matrix and margins, steps 1, 2, and 3 only need to be computed once and the fourth step, *Simulate Data*, is nearly instaneous for all settings considered.

```{r ch040-largeDfig, fig.cap="Computation times for HD multivariate gamma simulation.", eval=T}
# Set colors
numColors <- 5
numGroups <- 4
myColors <- rev(brewer.pal(n=numColors, name="Greys")[((numColors-numGroups)+1):numColors])

dat_long %>%
    filter(Dimensions %in% c(1000, 2500, 5000, 10000, 20000)) %>%
    mutate(`Time (minutes)` = Time / 60) %>%
    ggplot(aes(Correlation, `Time (minutes)`, fill=Step)) +
    geom_bar(position = "stack", stat = "identity") +
    scale_fill_manual(values = myColors ) +
    scale_y_continuous(breaks = seq(0, 80, 10),
                     minor_breaks = NULL) +
    facet_grid(. ~ Dimensions) +
    theme(axis.text.x = element_text(angle = -90))
```

*Limitations, conclusions, and recommendations*

In the bivariate studies, we chose arbitrary simulation parameters for three distributions, moving from the Gaussian to discrete and non-normal MVNB. Under these conditions, the simulated random vectors sample the desired bivariate distribution across the entire range of pairwise correlations for the three dependency measures. The simulation results could differ for other choices of simulation settings. Specifying extreme correlations near the boundary or Frechet bounds could result in poor simulation performance. Fortunately, it is straightforward to evaluate simulation performance by using strategies similar to those completed above. We expect our random vector generation to perform well for the vast majority of NORTA-feasible correlation matrices, but advise to check the performance before making inferences/further analyses.

Somewhat surprisingly, Kendall estimation and nearest PD computation scale poorly compared to Spearman and, even, approximate Pearson matching. In our experience, Kendall computation times are sensitive to the number of cores, benefiting from multi-core parallelization. This could mitigate some of the current algorithmic/implementation shortcomings. Despite this, Kendall matching is still feasible for most HD data sets. Finally, we note that one could use our single-pass algorithm `cor_fastPD` to produce a 'close' (not nearest PD) to scale to even higher dimensions with some loss of accuracy.

# RNA-seq data application  {#examples}

```{r ch050-LoadLib, include=FALSE, cache=FALSE, eval=T}
load("../data/example_brca.rda")
load("../data/example_genes.rda")
sim_nbinom <- readRDS("../results/brca_1000_sim.rds")
simRho <- readRDS("../results/brca_1000_frobenius.rds")
```

This section demonstrates how to simulate multivariate data using `bigsimr`, aiming to replicate the structure of HD dependent count data. In an illustration of our proposed methodology, we seek to simulate RNA-sequencing data by producing simulated random vectors mimicking the observed data and its generating process. Modeling RNA-seq using multivariate probability distributions is natural as inter-gene correlation occurs during biological processes [@Wang2009b]. And yet, many models do not account for this, leading to major disruptions to the operating characteristics of statistical estimation, testing, and prediction. The following subsections apply `bigsimr`'s methods to real RNA-seq data, including replicating an estimated parametric structure, probability estimation, and evaluation of correlation estimation efficiency.

## Simulating High-Dimensional RNA-seq data

```{r ch050-readBRCA, cache=FALSE, eval=T}
d <- 1000
brca1000 <- example_brca %>%
  select(all_of(1:d)) %>%
  mutate(across(everything(), as.double))
```

We begin by estimating the structure of the TCGA BRCA RNA-seq data set. Ultimately, we will simulate $B=10,000$ random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with $d=`r d`$. We assume a MVNB model as RNA-seq counts are often over-dispersed and correlated. Since all $d$ selected genes exhibit over-dispersion (data not shown), we proceed to estimate the NB parameters $(r_i, p_i), i=1,\ldots,d$, to determine the target marginal PMFs $f_i$. To complete specification of the simulation algorithm inputs, we estimate the Spearman correlation matrix $R_S$ to characterize dependency. With this goal in mind, we first estimate the desired correlation matrix using the fast implementation provided by `bigsimr`:

```{r ch050-estRhoBRCA, echo=TRUE, cache=FALSE, eval=T}
# Estimate Spearman's correlation on the count data
R_S <- bs$cor(as.matrix(brca1000), bs$Spearman)
```

Next, we estimate the marginal parameters. We use the method of moments to estimate the marginal parameters for the multivariate negative binomial model. The marginal distributions are from the same probability family (NB), yet they are heterogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$. The functions below support this estimation for later use in `rvec`.

```{r ch050-nbHelpers, echo=TRUE, eval=T}
make_nbinom_margins <- function(sizes, probs) {
  margins <- lapply(1:length(sizes), function(i) {
    dist$NegativeBinomial(sizes[i], probs[i])
  })
  do.call(c, margins)
}
```

We apply these estimators to the highest-expressing r d genes across the r nrow(brca1000) patients:

```{r ch050-estMargins, echo=TRUE, eval=T}
mom_nbinom <- function(x) {
  m <- mean(x)
  s <- sd(x)
  c(size = m^2 / (s^2 - m), prob = m / s^2)
}
nbinom_fit <- apply(brca1000, 2, mom_nbinom)
sizes <- nbinom_fit["size",]
probs <- nbinom_fit["prob",]
nb_margins <- make_nbinom_margins(sizes, probs)
```

Notably, the estimated marginal NB probabilities $\{ \hat{p}_i \}$ are small --- ranging in the interval $[r min(probs) , r max(probs)]$. This gives rise to highly variable counts and, thus, less restriction on potential pairwise correlation pairs. Given the marginals, we now specify targets and check admissibility of the specified correlation matrix.

```{r ch050-cor-check, echo=TRUE, eval=T}
# 1. Mapping step first
R_X <- bs$cor_convert(R_S, bs$Spearman, bs$Pearson)
# 2a. Check admissibility
(is_valid_corr <- bs$iscorrelation(R_X))
# 2b. compute nearest correlation
if (!is_valid_corr) {
  R_X_pd <- bs$cor_nearPD(R_X)
  ## Quantify the error
  targets      <- R_X[lower.tri(R_X, diag = FALSE)]
  approximates <- R_X_pd[lower.tri(R_X_pd, diag = FALSE)]
  R_X          <- R_X_pd
}
summary(abs(targets - approximates))
```

As seen above, $R_X$ is not strictly admissible in our scheme. This can be seen from the negative result from bs$iscorrelation(R_X), which checks if $R_X$ is positive-definite with 1's along the diagonal. However, the approximation is close with a maximum absolute error of r max( abs(targets - approximates)) and average absolute error of r mean(abs(targets - approximates)) across the r choose(d,2) correlations. With the inputs configured, `rvec` is executed to produce a synthetic RNA-seq data set:

```{r ch050-runBRCA-echo, echo=TRUE, eval=FALSE}
sim_nbinom <- bs$rvec(10000, R_X, nb_margins) 
```

Figure \@ref(fig:ch050-simDataFig) displays the simulated counts and pairwise relationships for our example genes from Table \@ref(tab:ch010-realDataTab).
Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parametric form and with less extreme points than the observed data in Figure \@ref(fig:ch010-realDataFig). Figure \@ref(fig:ch050-figBRCA) compares the specified target parameter (horizontal axis) with the corresponding quantities estimated from the simulated data (vertical axis). The evaluation shows that the simulated counts approximately match the target parameters and exhibit the full range of estimated correlations from the data.

```{r ch050-simDataFig, cache=FALSE, fig.cap="Simulated data for three selected high-expressing genes generally replicates the estimated data structure. The data do not exhibit outlying points, but do possess the desired Spearman correlations, central tendencies, and discrete values.", eval=T}
set.seed(2020-02-25)
num_genes <- 3
gene_sample <- sample(example_genes[1:1000], num_genes)

ggpairs(data = as.data.frame(sim_nbinom[, gene_sample]),
        upper = list(continuous = wrap('cor', method = "spearman"))) + 
  theme_bw()
```

```{r ch050-figBRCA, out.width='80%', fig.cap="Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set. The dashed red lines indicate equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes).", eval=T}
include_graphics('../fig/ch050-figBRCA.png')
```

*Limitations, conclusions, and recommendations*

The results show overall aquedate simulation performance for our choice of parameters settings. Our settings were motivated by modeling high-expressing genes from the TCGA BRCA data set. In general, the ability to match marginal and dependence parameters depends on the particular joint probability model. We recommend to evaluate and tune your simulation until you can be assured of the accuracy.

# Conclusion and discussion {#discussion}

We developed a general-purpose high-dimensional multivariate simulation algorithm and provide a user-friendly, high-performance `R` package `bigsimr`, Julia package `Bigsimr` and Python package `bigsimr`. The random vector generation method is inspired by NORTA [@Cario1997] and Gaussian copula-based approaches [@MB13, @BF17, @Xia17]. The major contributions of this work are methods and software for flexible, scalable simulation of HD multivariate probability distributions with broad data analytic applications for modern, big-data statistical computing. For example, one could simulate high-resolution time series data, such as those consistent with an auto-regressive moving average model exhibiting a specified Spearman structure. Our methods could also be used to simulate sparsely correlated data, as many HD methods assume, via specifying a *spiked correlation matrix*.

There are limitations to the methodology and implementation. We could only investigate selected multivariate distributions in our Monte Carlo studies. There may be instances where the methods do not perform well. Along those lines, we expect that correlation values close to the boundary of the feasible region could result in algorithm failure. Another issue is that for discrete distributions, we use continuous approximations when the support set is large. This could limit the scalability/accuracy for particular discrete/mixed multivariate distributions. Our method would also benefit computationally from faster Kendall estimation and more finely tuned nearest PD calculation for non-admissible Kendall matrices.

### 2021-10-19 arXiv_bigsimr/arXiv_bigsimr.tex
preparing arXiv release of bigsimr R focused article

\documentclass{article}

\usepackage{arxiv}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{lmodern}        % https://github.com/rstudio/rticles/issues/343
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{graphicx}

\title{Simulating High-Dimensional Multivariate Data using the Bigsimr R
Package}

\author{
    David S. Hippocampus
    \thanks{Use footnote for providing further information about author
(webpage, alternative address)---\emph{not} for acknowledging funding
agencies. Optional.}
   \\
    Department of Computer Science \\
    Cranberry-Lemon University \\
  Pittsburgh, PA 15213 \\
  \texttt{\href{mailto:hippo@cs.cranberry-lemon.edu}{\nolinkurl{hippo@cs.cranberry-lemon.edu}}} \\
   \And
    Elias D. Striatum
   \\
    Department of Electrical Engineering \\
    Mount-Sheikh University \\
  Santa Narimana, Levand \\
  \texttt{\href{mailto:stariate@ee.mount-sheikh.edu}{\nolinkurl{stariate@ee.mount-sheikh.edu}}} \\
  }

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

% Pandoc citation processing
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc} % for calculating minipage widths
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{setspace}

\begin{document}
\maketitle

\def\tightlist{}

\begin{abstract}
It is critical to accurately simulate data when employing Monte Carlo
techniques and evaluating statistical methodology. Measurements are
often correlated and high dimensional in this era of big data, such as
data obtained in high-throughput biomedical experiments. Due to the
computational complexity and a lack of user-friendly software available
to simulate these massive multivariate constructions, researchers resort
to simulation designs that posit independence or perform arbitrary data
transformations. To close this gap, we developed the Bigsimr Julia
package with R and Python interfaces. This paper focuses on the R
interface. These packages empower high-dimensional random vector
simulation with arbitrary marginal distributions and dependency via a
Pearson, Spearman, or Kendall correlation matrix. bigsimr contains
high-performance features, including multi-core and
graphical-processing-unit-accelerated algorithms to estimate correlation
and compute the nearest correlation matrix. Monte Carlo studies quantify
the accuracy and scalability of our approach, up to \(d=10,000\). We
describe example workflows and apply to a high-dimensional data set ---
RNA-sequencing data obtained from breast cancer tumor samples.
\end{abstract}

\keywords{
    high-dimensional data
   \and
    multivariate simulation
   \and
    nonparametric correlation
   \and
    Gaussian copula
   \and
    RNA-sequencing data
   \and
    breast cancer
  }

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Massive high-dimensional (HD) data sets are now common in many areas of
scientific inquiry. As new methods are developed for data analysis, a
fundamental challenge lies in designing and conducting simulation
studies to assess the operating characteristics of proposed methodology
--- such as false positive rates, statistical power, interval coverage,
and robustness. Further, efficient simulation empowers statistical
computing strategies, such as the parametric bootstrap (Chernick 2008)
to simulate from a hypothesized null model, providing inference in
analytically challenging settings. Such Monte Carlo (MC) techniques
become difficult for HD dependent data using existing algorithms and
tools. This is particularly true when simulating massive multivariate,
non-normal distributions, arising in many fields of study.

As others have noted, it can be vexing to simulate dependent,
non-normal/discrete data, even for low-dimensional (LD) settings (Madsen
and Birkes 2013; Xiao and Zhou 2019). For continuous non-normal LD
multivariate data, the well-known NORmal To Anything (NORTA) algorithm
(Cario and Nelson 1997) and other copula approaches (Nelsen 2007) are
well-studied and implemented in publicly-available software (Yan 2007;
Chen 2001). Yet these approaches do not scale in a timely fashion to HD
problems (X. Li et al. 2019). For discrete data, early simulation
strategies had major flaws, such as failing to obtain the full range of
possible correlations --- such as admitting only positive correlations
(Park, Park, and Shin 1996). While more recent approaches (Madsen and
Birkes 2013; Xiao 2017; Barbiero and Ferrari 2017) have remedied this
issue for LD problems, the existing tools are not designed to scale to
high dimensions.

Another central issue lies in characterizing dependence between
components in the HD random vector. The choice of correlation in
practice usually relates to the eventual analytic goal and
distributional assumptions of the data (e.g., non-normal, discrete,
infinite support, etc.). For normal data, the Pearson product-moment
correlation describes the dependence perfectly. However, simulating
arbitrary random vectors that match a target Pearson correlation matrix
is computationally intense (Chen 2001; Xiao 2017). On the other hand, an
analyst might consider use of nonparametric correlation measures to
better characterize monotone, non-linear dependence, such as Spearman's
\(\rho\) and Kendall's \(\tau\). Throughout, we focus on matching these
nonparametric dependence measures, as our aim lies in modeling
non-normal data and these rank-based measures possess invariance
properties enabling our proposed methodology. We do, however, implement
Pearson matching, but several layers of approximation are required.

With all this in mind, we present a scalable, flexible multivariate
simulation algorithm. The crux of the method lies in the construction of
a Gaussian copula in the spirit of the NORTA procedure. As we will
describe in more detail, the algorithm's design leverages useful
properties of nonparametric correlation measures, namely invariance
under monotone transformation and well-known closed-form relationships
between dependence measures for the multivariate normal (MVN)
distribution. For our method, we developed a high-performance
implementation: the \texttt{Bigsimr} Julia package, with R and Python
interfaces \texttt{bigsimr}.

This article proceeds by providing background information, including a
description of a motivating example application: RNA-sequencing
(RNA-seq) breast cancer data. Then we describe and justify our
simulation methodology and related algorithms. We proceed by providing
an illustrative LD \texttt{bigsimr} workflow. Next we conduct MC studies
under various bivariate distributional assumptions to evaluate
performance and accuracy. After the MC evaluations, we simulate random
vectors motivated by our RNA-seq example, evaluate the accuracy, and
provide example statistical computing tasks, namely MC estimation of
joint probabilities and evaluating HD correlation estimation efficiency.
Finally, we discuss the method's utility, limitations, and future
directions.

\hypertarget{background}{%
\section{Background}\label{background}}

The \texttt{Bigsimr} package presented here provides multiple
high-performance algorithms that operate with HD multivariate data. All
these algorithms were originally designed to support a single task: to
generate random vectors drawn from multivariate probability
distributions with given marginal distributions and dependency metrics.
Specifically, our goal is to efficiently simulate a large number, \(B\),
of HD random vectors \({\bf Y}=(Y_1, \ldots, Y_d)^\top\) with
\emph{correlated} components and heterogeneous marginal distributions,
described via cumulative distribution functions (CDFs)
\(F_i, i=1,\ldots,d\).

When designing this methodology, we developed the following properties
to guide our effort. We divide the properties into two categories: (1)
basic properties (BP) and `scalability' properties (SP). The BPs are
adapted from an existing criteria due to Nikoloulopoulos (2013). A
suitable simulation strategy should possess the following properties:

\begin{itemize}
\tightlist
\item
  BP1: A wide range of dependences, allowing both positive and negative
  values, and, ideally, admitting the full range of possible values.
\item
  BP2: Flexible dependence, meaning that the number of dependence
  parameters can be equal to the number of bivariate pairs in the
  vector.
\item
  BP3: Flexible marginal modeling, generating heterogeneous data ---
  including mixed continuous and discrete marginal distributions.
\end{itemize}

Moreover, the simulation method must \emph{scale} to high dimensions:

\begin{itemize}
\tightlist
\item
  SP1: Procedure must scale to high dimensions with practical
  computation times.
\item
  SP2: Procedure must scale to high dimensions while maintaining
  accuracy.
\end{itemize}

\hypertarget{motivating-example-rna-seq-data}{%
\subsection{Motivating example: RNA-seq
data}\label{motivating-example-rna-seq-data}}

Simulating HD, non-normal, correlated data motivates this work, in
pursuit of modeling RNA-sequencing (RNA-seq) data (Wang, Gerstein, and
Snyder 2009; Conesa et al. 2016) derived from breast cancer patients.
The RNA-seq data-generating process involves counting how often a
particular form of messenger RNA (mRNA) is expressed in a biological
sample. RNA-seq platforms typically quantify the entire transcriptome in
one experimental run, resulting in HD data. For human-derived samples,
this results in count data corresponding to over 20,000 genes
(protein-coding genomic regions) or even over 77,000 isoforms when
alternatively-spliced mRNA are counted (Schissler et al. 2019).
Importantly, due to inherent biological processes, gene expression data
exhibit correlation (co-expression) across genes (Efron 2007; Schissler,
Piegorsch, and Lussier 2018).

We illustrate our methodology using the Breast Invasive Carcinoma (BRCA)
data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments).
For ease of modeling and simplicity of exposition, we only consider high
expressing genes. In turn, we begin by filtering to retain the top r d
of the highest-expressing genes (in terms of median expression) of the
over 20,000 gene measurements from \(N=r nrow(example_brca)\) patients'
tumor samples. This gives a great number of pairwise dependencies among
the marginals (specifically, \(r choose(d,2)\) correlation parameters).
Table @ref(tab:ch010-realDataTab) displays RNA-seq counts for three
selected high-expressing genes for the first five patients' breast tumor
samples. To help visualize the bivariate relationships for these three
selected genes across all patients, Figure @ref(fig:ch010-realDataFig)
displays the marginal distributions and estimated Spearman's
correlations.

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{set.seed}\NormalTok{(}\DecValTok{2020{-}02{-}25}\NormalTok{)}
\NormalTok{num\_genes    }\OtherTok{\textless{}{-}} \DecValTok{3}
\NormalTok{num\_patients }\OtherTok{\textless{}{-}} \DecValTok{5}
\NormalTok{gene\_sample  }\OtherTok{\textless{}{-}} \FunctionTok{sample}\NormalTok{(example\_genes[}\DecValTok{1}\SpecialCharTok{:}\DecValTok{1000}\NormalTok{], num\_genes)}

\NormalTok{cap }\OtherTok{\textless{}{-}} \FunctionTok{paste0}\NormalTok{(}\StringTok{"mRNA expression for three selected high{-}expressing genes, "}\NormalTok{,}
              \FunctionTok{paste}\NormalTok{(gene\_sample, }\AttributeTok{collapse =} \StringTok{", "}\NormalTok{), }
              \StringTok{", for the first five patients in the TCGA BRCA data set."}\NormalTok{)}

\NormalTok{small\_brca }\OtherTok{\textless{}{-}}\NormalTok{ example\_brca }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{select}\NormalTok{(}\FunctionTok{all\_of}\NormalTok{(gene\_sample)) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{head}\NormalTok{(}\AttributeTok{n =}\NormalTok{ num\_patients) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{as\_tibble}\NormalTok{(}\AttributeTok{rownames =} \StringTok{"Patient ID"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Patient ID}\StringTok{\textasciigrave{}} \OtherTok{=} \FunctionTok{str\_sub}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Patient ID}\StringTok{\textasciigrave{}}\NormalTok{, }\AttributeTok{end =} \DecValTok{12}\NormalTok{))}

\NormalTok{small\_brca }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{kable}\NormalTok{(}\AttributeTok{format =} \StringTok{"latex"}\NormalTok{, }\AttributeTok{booktabs =} \ConstantTok{TRUE}\NormalTok{, }\AttributeTok{caption =}\NormalTok{ cap)}
\end{Highlighting}
\end{Shaded}

\begin{table}

\caption{\label{tab:ch010-realDataTab}mRNA expression for three selected high-expressing genes, STAU1, FKBP1A, NME2, for the first five patients in the TCGA BRCA data set.}
\centering
\begin{tabular}[t]{lrrr}
\toprule
Patient ID & STAU1 & FKBP1A & NME2\\
\midrule
TCGA-A1-A0SB & 10440 & 11354 & 17655\\
TCGA-A1-A0SD & 21523 & 20221 & 14653\\
TCGA-A1-A0SE & 21733 & 22937 & 35251\\
TCGA-A1-A0SF & 11866 & 19650 & 16551\\
TCGA-A1-A0SG & 12486 & 12089 & 10434\\
\bottomrule
\end{tabular}
\end{table}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{ggpairs}\NormalTok{(}
  \AttributeTok{data =}\NormalTok{ example\_brca[, gene\_sample],}
  \AttributeTok{upper =} \FunctionTok{list}\NormalTok{(}\AttributeTok{continuous =} \FunctionTok{wrap}\NormalTok{(}\StringTok{\textquotesingle{}cor\textquotesingle{}}\NormalTok{, }\AttributeTok{method =} \StringTok{"spearman"}\NormalTok{))}
\NormalTok{) }\SpecialCharTok{+} 
  \FunctionTok{theme\_bw}\NormalTok{() }\SpecialCharTok{+}
  \FunctionTok{theme}\NormalTok{(}\AttributeTok{axis.text.x =} \FunctionTok{element\_text}\NormalTok{(}\AttributeTok{angle =} \SpecialCharTok{{-}}\DecValTok{90}\NormalTok{, }\AttributeTok{vjust =} \FloatTok{0.5}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
## exact p-value with ties

## Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
## exact p-value with ties

## Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
## exact p-value with ties
\end{verbatim}

\begin{figure}
\centering
\includegraphics{arXiv_bigsimr_files/figure-latex/ch010-realDataFig-1.pdf}
\caption{Pairwise scatterplots for three example genes (Table 1), with
estimated Spearman's correlations and marginal density estimates
displayed. The data possess outliers, heavy-right tails, are discrete,
and have non-trivial intergene correlations. Modeling these data
motivate our simulation methodology.}
\end{figure}

\hypertarget{measures-of-dependency}{%
\subsection{Measures of dependency}\label{measures-of-dependency}}

In multivariate analysis, an analyst must select a metric to quantify
dependency. The most widely-known is the Pearson correlation coefficient
that describes the linear association between two random variables \(X\)
and \(Y\), and, it is given by

\begin{equation}
\rho_P(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ \mathrm{Var}(X)\mathrm{Var}(Y)\right]^{1/2}}.
\label{eq:pearson}
\end{equation}

As Madsen and Birkes (2013) and Mari and Kotz (2001) discuss, for a
bivariate normal \((X,Y)\) random vector, the Pearson correlation
completely describes the dependency between the components. For
non-normal marginals with monotone correlation patterns, however,
\(\rho_P\) suffers some drawbacks and may mislead or fail to capture
important relationships (Mari and Kotz 2001). Alternatively, analysts
often prefer rank-based correlation measures to describe the degree of
monotonic association.

Two nonparametric, rank-based measures common in practice are Spearman's
correlation (denoted \(\rho_S\)) and Kendall's \(\tau\). Spearman's
\(\rho_S\) has an appealing correspondence as the Pearson correlation
coefficient on \emph{ranks} of the values, thereby capturing nonlinear
yet monotone relationships. Kendall's \(\tau\), on the other hand, is
the difference in probabilities of concordant and discordant pairs of
observations, \((X_i, Y_i)\) and \((X_j, Y_j)\) (concordance meaning
that orderings have the same direction, e.g., if \(X_i < X_j\), then
\(Y_i < Y_j\)). Note that concordance is determined by the ranks of the
values, not the values themselves. Both \(\tau\) and \(\rho_S\) are
\emph{invariant under monotone transformations} of the underlying random
variates. As we will describe more fully in the
\protect\hyperlink{algorithms}{Algorithms} section, this property
enables matching rank-based correlations with speed (SP1) and accuracy
(SP2).

\emph{Correspondence among Pearson, Spearman, and Kendall correlations}

There is no closed form, general correspondence among the rank-based
measures and the Pearson correlation coefficient, as the marginal
distributions \(F_i\) are intrinsic in their calculation. For
\emph{bivariate normal vectors}, however, the correspondence is
well-known:

\begin{equation}
\label{eq:convertKendall}
\rho_{P} = \sin \left( \tau \times \frac{\pi}{2} \right), 
\end{equation}

\noindent and similarly for Spearman's \(\rho\) (Kruskal 1958),

\begin{equation}
\label{eq:convertSpearman}
\rho_P = 2 \times \sin \left( \rho_S \times \frac{\pi}{6} \right).
\end{equation}

\emph{Marginal-dependent bivariate correlation bounds}

Given two marginal distributions, \(\rho_P\) is not free to vary over
the entire range of possible correlations \([-1,1]\). The
\emph{Frechet-Hoeffding bounds} are well-studied (Nelsen 2007; Barbiero
and Ferrari 2017). These constraints cannot be overcome through
algorithm design. In general, the bounds are given by

\begin{equation}
\label{eq:frechet}
\rho_P^{max} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (U) \right), \quad \rho_P^{min} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (1 - U) \right)
\end{equation}

\noindent where \(U\) is a uniform random variable on \((0,1)\) and
\(F^{-1}_1, F^{-1}_2\) are the inverse CDFs of \(X_1\) and \(X_2\),
respectively, defined as

\begin{equation}
F_{i}^{-1} = \inf\{y:F_{i}(y) \geq u \}
\label{eq:inverseCDF}
\end{equation}

when the variables are discrete.

\hypertarget{gaussian-copulas}{%
\subsection{Gaussian copulas}\label{gaussian-copulas}}

There is a connection of our simulation strategy to Gaussian
\emph{copulas} (Nelsen 2007). A copula is a distribution function on
\([0,1]^d\) that describes a multivariate probability distribution with
standard uniform marginals. This provides a powerful, natural way to
characterize joint probability. Consequently, the study of copulas is an
important and active area of statistical theory and practice. For any
random vector \({\bf X}=(X_1, \ldots, X_d)^\top\) with CDF \(F\) and
marginal CDFs \(F_i\) there is a copula function \(C(u_1, \ldots, u_d)\)
satisfying

\begin{equation}
F(x_1, \ldots, x_d) = {\mathbb P}(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), \quad x_i \in {\mathbb R}, i=1,\ldots,d.
\label{eq:copula}
\end{equation}

A Gaussian copula has marginal CDFs that are all standard normal,
\(F_i = \Phi, \forall \, i\), . This representation corresponds to a
multivariate normal (MVN) distribution with standard normal marginal
distributions and covariance matrix \(R_P\). Since the marginals are
standardized to have unit variance, this \(R_P\) is also Pearson
correlation matrix. If \(F_R\) is the CDF of such a multivariate normal
distribution, then the corresponding Gaussian copula \(C_R\) is defined
through

\begin{equation}
\label{eq:gauss}
F_R(x_1, \ldots, x_d) = C_R(\Phi(x_1), \ldots, \Phi(x_d)),
\end{equation}

where \(\Phi(\cdot)\) is the standard normal CDF.

Here, the copula function \(C_R\) is the CDF of the random vector
\((\Phi(X_1), \ldots, \Phi(X_d))\), where
\((X_1, \ldots, X_d) \sim N_d({\bf 0}, R_P)\).

Sklar's Theorem (Úbeda-Flores and Fernández-Sánchez 2017) guarantees
that given inverse CDFs \(F_i^{-1}\)s and a valid correlation matrix
(within the Frechet bounds) a random vector can be obtained via
transformations involving copula functions. For example, using Gaussian
copulas, we can construct a random vector
\({\bf Y} = (Y_1, \ldots, Y_d)^\top\) with \(Y_i \sim F_i\),
viz.~\(Y_i = F_i^{-1}(\Phi(X_i)), i=1, \ldots, d\), where
\((X_1, \ldots, X_d) \sim N_d({\bf 0}, R_P)\).

\hypertarget{other-multivariate-simulation-packages}{%
\subsection{Other Multivariate Simulation
Packages}\label{other-multivariate-simulation-packages}}

To place \texttt{Bigsimr}'s implementation and algorithms in context of
existing approaches, we qualitatively compare \texttt{Bbigsimr} to other
multivariate simulation packages. In \texttt{R}, the \texttt{MASS}
package's \texttt{mvnorm()} has long been available to produce random
vectors from multivariate normal (and \(t\)) distributions. This
procedure cannot simulate from arbitrary margins and dependency measures
and it fails to readily scale to high dimensions. SAS PROC SIMNORMAL and
Python's NumPy \texttt{random.multivariate\_normal} both do similar
tasks. Recently, the high-performance \texttt{R} package
\texttt{mvnfast} (Fasiolo 2016) has been released, which provides HD
multivariate normal generation. However, lacks flexibility in marginal
models and dependency modeling. For dependent discrete data, LD
\texttt{R} packages exist, such as \texttt{GenOrd}. For copula
computation, the full-featured \texttt{R} package \texttt{copula}
provides many different copulas and a fairly general LD random generator
\texttt{rCopula()}. High-dimesional random vector generation via
\texttt{copula} takes an impractical amount of computing time (X. Li et
al. 2019). The \texttt{nortaRA} R package matches Pearson coefficients
near exactly and is reasonable for LD settings. Finally, we know of no
other \texttt{R} package that can produce HD random vectors with
flexible margins and dependence structure. Table @ref(tab:compare-table)
summarizes the above discussion.

\begin{table}[h]
\centering
\caption{\label{tab:compare-table}Comparison of selected features of bigsimr to existing R packages for multivariate simulation.}
\begin{tabular}{@{}llll@{}}
Package & Scales to HD & Hetereogenous Margins & Flexible Dependence \\ \midrule
MASS      & No           & No                    & No                  \\
mvnfast   & Yes          & No                    & No                  \\
GenOrd    & No           & Discrete margins      & Yes                 \\
copula    & No           & Yes                   & No                  \\
nortaRA   & No           & Yes                   & No                  \\
bigsimr   & Yes          & Yes                   & Yes                
\end{tabular}
\end{table}

\hypertarget{algorithms}{%
\section{Algorithms}\label{algorithms}}

This section describes our methods for simulating a random vector
\(\bf Y\) with \(Y_i\) components for \(i=1,\ldots,d\). Each \(Y_i\) has
a specified marginal CDF \(F_i\) and its inverse \(F^{-1}_i\). To
characterize dependency, every pair \((Y_i, Y_j)\) has a given Pearson
correlation \(\rho_P\), Spearman correlation \(\rho_S\), and/or
Kendall's \(\tau\). The method can be described as a
\emph{high-performance Gaussian copula} (Equation @ref(eq:gauss))
providing a HD NORTA-inspired algorithm.

\hypertarget{normal-to-anything-norta}{%
\subsection{NORmal To Anything (NORTA)}\label{normal-to-anything-norta}}

The well-known NORTA algorithm (Cario and Nelson 1997) simulates a
random vector \(\bf Y\) with variance-covariance matrix
\(\Sigma_{\bf Y}\). Specifically, NORTA algorithm proceeds as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Simulate a random vector \(\bf Z\) with \(d\) independent and
  identically distributed (iid) standard normal components.
\item
  Determine the input matrix \(\Sigma_{\bf Z}\) that corresponds with
  the specified output \(\Sigma_{\bf Y}\).
\item
  Produce a Cholesky factor \(M\) of \(\Sigma_{\bf Z}\) such that
  \(M M^{\prime}=\Sigma_{\bf Z}\).
\item
  Set \(X\) by \(X \gets MZ\).
\item
  \(\text{Return} \; {\bf Y} \; \text{where} \; Y_i \gets F_i^{-1}[\Phi(X_i)], \; i=1,...,d\).
\end{enumerate}

With modern parallel computing, steps 1, 3, 4, 5 are readily implemented
as high-performance, multi-core and/or graphical-processing-unit (GPU)
accelerated algorithms --- providing fast scalability to HD.

Matching specified Pearson correlation coefficients exactly (step 2
above), however, is computationally costly. In general, there is no
convenient correspondence between the components of the input
\(\Sigma_{\bf Z}\) and target \(\Sigma_{\bf Y}\). Matching the
correlations involves evaluating or approximating \(\binom{d}{2}\)
integrals of the form

\begin{equation}
    \mathrm{E}\left[Y_i Y_j\right] = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} F_i^{-1}\left[\Phi(z_i)\right] F_j^{-1}\left[\Phi(z_j)\right] \phi(z_i, z_j, \rho_z) dz_i dz_j,
    \label{eq:pearsonIntegralRelation}
\end{equation}

where \(\phi(\cdot)\) is the joint probability density function of two
correlated standard normal variables. For HD simulation, exact
evaluation may become too costly to enable practical simulation studies.
Our remedy to enable HD Pearson matching in \texttt{bigsimr} is Hermite
polynomial approximation of these \(\binom{d}{2}\) double integrals, as
suggested by Xiao and Zhou (2019).

\emph{NORTA in higher dimensions}

Sklar's theorem provides a useful characterization of multivariate
distributions through copulas. In practice, however, the particulars of
the simulation algorithm affect which joint distributions may be
simulated. Even in LD spaces (e.g., \(d=3\)), there exist valid
multivariate distributions with \emph{feasible} Pearson correlation
matrices that NORTA cannot match exactly (S. T. Li and Hammond 1975).
This occurs when the bivariate transformations are applied to find the
input correlation matrix, yet when combined, the resultant matrix
becomes non-positive definite. These situations do occur, even using
exact analytic calculations. Such problematic target correlation
matrices are termed \emph{NORTA defective}.

Ghosh and Henderson (2002) conducted a study to estimate the probability
of encountering NORTA defective matrices while increasing the dimension
\(d\). They found that for what is now considered low-to-moderate
dimensions (\(d \approx 20\)), almost \emph{all} feasible matrices are
NORTA defective. This stems from the concentration of measure near the
boundary of the space of all possible correlation matrices as dimension
increases. Unfortunately, it is precisely near this boundary that NORTA
defective matrices reside.

There is hope, however, as Ghosh and Henderson (2002) also showed that
replacing an non-positive definite input correlation matrix with a close
proxy will give approximate matching to the target --- with adequate
performance for moderate \(d\). This provides evidence that our nearest
positive definite (PD) augmented approach has promise to provide
adequate accuracy if our input matching scheme returns an indefinite
Pearson correlation matrix.

\hypertarget{rand-vec-gen}{%
\subsection{Random vector generator}\label{rand-vec-gen}}

We now describe our algorithm to generate random vectors, which extends
NORTA to HD via nearest correlation matrix replacement and provides
rank-based dependency matching:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Mapping step, either

  \begin{itemize}
  \tightlist
  \item
    Convert the target Spearman correlation matrix \(R_S\) to the
    corresponding MVN Pearson correlation \(R_X\). Alternatively,
  \item
    Convert the target Kendall \(\tau\) matrix \(R_K\) to the
    corresponding MVN Pearson correlation \(R_X\). Alternatively,
  \item
    Convert the target Pearson correlation matrix to \(R_P\) to the
    corresponding, approximate MVN Pearson correlation \(R_X\).
  \end{itemize}
\item
  Check correlation matrix admissibility and, if not, compute the
  nearest correlation matrix.

  \begin{itemize}
  \tightlist
  \item
    Check that \(R_X\) is a correlation matrix, a positive definite
    matrix with 1's along the diagonal.
  \item
    If the mapping produced an \(R_X\) that is a correlation matrix, we
    describe the matrix as \textbf{admissible} in this scheme and retain
    \(R_X\). Otherwise,
  \item
    Replace \(R_X\) with the nearest correlation matrix \(\tilde{R}_X\),
    in the Frobenius norm and the result will be approximate.
  \end{itemize}
\item
  Gaussian copula

  \begin{itemize}
  \tightlist
  \item
    Generate \({\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)\).
  \item
    Transform \({\bf X}\) to \({\bf U} = (U_1, \ldots, U_d)\)
    viz.~\(U_i=\Phi(X_i)\), \(i=1, \ldots, d\).
  \item
    Return \({\bf Y} = (Y_1, \ldots, Y_d)\), where
    \(Y_i=F_i^{-1}(U_i)\), \(i=1, \ldots, d\).
  \end{itemize}
\end{enumerate}

\emph{Step 1: Mapping step.} We first employ the closed-form
relationships between \(\rho_S\) and \(\tau\) with \(\rho_P\) for
bivariate normal random variables via Equations @ref(eq:convertKendall)
and @ref(eq:convertSpearman), respectively (implemented as
\texttt{cor\_covert()}). Initializing our algorithm to match the
nonparametric correlations by computing these equations for all pairs is
computationally trivial.

For the computationally expensive process to match Pearson correlations,
we approximate Equation @ref(eq:pearsonIntegralRelation) for all pairs
of margins. To this end, we implement the approximation scheme
introduced by (Xiao and Zhou 2019). Briefly, the many double integrals
of the form in @ref(eq:pearsonIntegralRelation) are approximated by
weighted sums of Hermite polynomials. Matching coefficients for pairs of
continuous distributions is made tractable by this method, but for
discrete distributions (especially discrete distributions with large
support sets or infinite support), the approximation is computationally
expensive. To remedy this, we further approximate discrete distributions
by a continuous distribution using Generalized S-Distributions (Muino,
Voit, and Sorribas 2006).

\emph{Step 2: Admissibility check and nearest correlation matrix
computation.} Once \(R_X\) has been determined, we check admissibility
of the adjusted correlation via the steps described above. If \(R_X\) is
not a valid correlation matrix, then we compute the nearest correlation
matrix. Finding the nearest correlation matrix is a common statistical
computing problem. The defacto function in \texttt{R} is
\texttt{Matrix::nearPD}, an alternating projection algorithm due to
Higham (2002). As implemented, the function fails to scale to HD.
Instead, we provide the quadratically-convergent algorithm based on the
theory of strongly semi-smooth matrix functions (Qi and Sun 2006). The
nearest correlation matrix problem can be written down as the following
convex optimization problem:
\(\mathrm{min} \quad \frac{1}{2} \Vert R_X - X \Vert^2, \quad \mathrm{s.t.} \quad X_{ii} = 1, \quad i = 1, \ldots , n, \quad X \in S_{+}^{n}\).

For nonparametric correlation measures, our algorithm allows the
generation of HD multivariate data with arbitrary marginal distributions
with a broad class of admissible Spearman correlation matrices and
Kendall \(\tau\) matrices. The admissible classes consist of the
matrices that map to a Pearson correlation matrix for a MVN. In
particular, if we let \(X\) be MVN with \(d\) components and denote
\(\Omega_P = \{ R_P : R_P \textrm{ is a Pearson correlation matrix for } X \}, \quad \Omega_K = \{ R_K : R_K \textrm{ is a Kendall correlation matrix for } X \}, \quad \Omega_S = \{ R_S : R_S \textrm{ is a Spearman correlation matrix for } X \}\).
There are 1-1 mappings between these sets. We conjecture that the sets
of admissible \(R_S\) and \(R_K\) are not highly restrictive. In
particular, \(R_P\) is approximately \(R_S\) for a MVN, suggesting that
the admissible set \(\Omega_S\) should be flexible as \(R_P\) can be any
PD matrix with 1's along the diagonal. We provide methods to check
whether a target \(R_S\) is an element of the \emph{admissible set}
\(\Omega_S\).

There is an increasing probability of encountering an inadmissible
correlation matrix as dimension increases. In our experience, the
mapping step for large \(d\) almost always produces a \(R_X\) that is
not a correlation matrix. In Section @ref(package), we provide a basic
method of how to quantify and control the approximation error. Further,
the RNA-seq example in Section @ref(examples) provides an illustration
of this in practice.

\emph{Step 3: Gaussian copula.} The final step implements a
NORTA-inspired, Gaussian copula approach to produce the desired margins.
Steps 1 and 2 determine the MVN Pearson correlation values that will
eventually match the target correlation. Step 3 requires a fast MVN
simulator, a standard normal CDF, and well-defined quantile functions
for marginals. The MVN is transformed to a copula (distribution with
standard uniform margins) by applying the normal CDF \(\Phi(\cdot)\).
Finally, the quantile functions \(F_i^{-1}\) are applied across the
margins to return the desired random vector \({\bf Y}\).

\hypertarget{the-bigsimr-r-package}{%
\section{The bigsimr R package}\label{the-bigsimr-r-package}}

\begin{verbatim}
## Julia version 1.5.3 at location /Applications/Julia-1.5.app/Contents/Resources/julia/bin will be used.
\end{verbatim}

\begin{verbatim}
## Loading setup script for JuliaCall...
\end{verbatim}

\begin{verbatim}
## Finish loading setup script for JuliaCall.
\end{verbatim}

This section describes a bivariate random vector simulation workflow via
the \texttt{bigsimr} \texttt{R} package. This \texttt{R} package
provides an interface to the native code written in Julia (registered as
the \texttt{Bigsimr} Julia package). In addition to the native Julia
\texttt{Bigsimr} package and \texttt{R} interface \texttt{bigsimr}, we
also provide a Python interface \texttt{bigsimr} that interfaces with
the Julia \texttt{Bigsimr} package. The Julia package provides a
high-performance implementation of our proposed random vector generation
algorithm and associated functions (see Section @ref(algorithms)).

The subsections below describe the basic use of the \texttt{bigsimr} R
package by stepping through an example workflow using the data set
\texttt{airquality} that contains daily air quality measurements in New
York, May to September 1973 (Chambers et al. 1983). This workflow
proceeds from setting up the computing environment, to data wrangling,
estimation, simulation configuration, random vector generation, and,
finally, result visualization.

\hypertarget{bivariate-example}{%
\subsection{Bivariate example}\label{bivariate-example}}

We illustrate the use of \texttt{bigsimr} using the New York air quality
data set (\texttt{airquality}) included in the R \texttt{datasets}
package. First, we load the \texttt{bigsimr} library and a few other
convenient data science packages, including the syntactically-elegant
\texttt{tidyverse} suite of \texttt{R} packages. The code chunk below
prepares the computing environment:

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(}\StringTok{"tidyverse"}\NormalTok{)}
\FunctionTok{library}\NormalTok{(}\StringTok{"bigsimr"}\NormalTok{)}
\CommentTok{\# Activate multithreading in Julia}
\FunctionTok{Sys.setenv}\NormalTok{(}\AttributeTok{JULIA\_NUM\_THREADS =}\NormalTok{ parallel}\SpecialCharTok{::}\FunctionTok{detectCores}\NormalTok{())}
\CommentTok{\# Load the Bigsimr and Distributions Julia packages}
\NormalTok{bs }\OtherTok{\textless{}{-}} \FunctionTok{bigsimr\_setup}\NormalTok{()}
\NormalTok{dist }\OtherTok{\textless{}{-}} \FunctionTok{distributions\_setup}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

Here, we describe a minimal working example --- a bivariate simulation
of two airquality variables: \texttt{Temperature}, in degrees
Fahrenheit, and \texttt{Ozone} level, in parts per billion.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{df }\OtherTok{\textless{}{-}}\NormalTok{ airquality }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{select}\NormalTok{(Temp, Ozone) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{drop\_na}\NormalTok{()}
\FunctionTok{glimpse}\NormalTok{(df)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Rows: 116
## Columns: 2
## $ Temp  <int> 67, 72, 74, 62, 66, 65, 59, 61, 74, 69, 66, 68, 58, 64, 66, 5...
## $ Ozone <int> 41, 36, 12, 18, 28, 23, 19, 8, 7, 16, 11, 14, 18, 14, 34, 6, ...
\end{verbatim}

Figure @ref(fig:ch030-aq-joint-dist) visualizes the bivariate
relationship between \texttt{Ozone} and \texttt{Temperature}. We aim to
simulate random two-component vectors mimicking this structure. The
margins are not normally distributed; \texttt{Ozone} level exhibits a
strong positive skew.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{p0 }\OtherTok{\textless{}{-}} \FunctionTok{ggplot}\NormalTok{(df, }\FunctionTok{aes}\NormalTok{(Temp, Ozone)) }\SpecialCharTok{+}
  \FunctionTok{geom\_point}\NormalTok{(}\AttributeTok{size =} \DecValTok{1}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{theme}\NormalTok{(}\AttributeTok{legend.position =} \StringTok{"none"}\NormalTok{) }\SpecialCharTok{+} 
  \FunctionTok{labs}\NormalTok{(}\AttributeTok{x =} \StringTok{"Temperature"}\NormalTok{)}

\NormalTok{pTemp }\OtherTok{\textless{}{-}} \FunctionTok{ggplot}\NormalTok{(df, }\FunctionTok{aes}\NormalTok{(Temp)) }\SpecialCharTok{+} 
  \FunctionTok{geom\_density}\NormalTok{() }\SpecialCharTok{+}
  \FunctionTok{theme}\NormalTok{(}\AttributeTok{axis.title.x =} \FunctionTok{element\_blank}\NormalTok{(),}
        \AttributeTok{axis.title.y =} \FunctionTok{element\_blank}\NormalTok{(),}
        \AttributeTok{axis.text =} \FunctionTok{element\_blank}\NormalTok{(),}
        \AttributeTok{axis.ticks =} \FunctionTok{element\_blank}\NormalTok{())}

\NormalTok{pOzone }\OtherTok{\textless{}{-}} \FunctionTok{ggplot}\NormalTok{(df, }\FunctionTok{aes}\NormalTok{(Ozone)) }\SpecialCharTok{+} 
    \FunctionTok{geom\_density}\NormalTok{() }\SpecialCharTok{+}
  \FunctionTok{theme}\NormalTok{(}\AttributeTok{axis.title.y =} \FunctionTok{element\_blank}\NormalTok{(),}
        \AttributeTok{axis.title.x =} \FunctionTok{element\_blank}\NormalTok{(),}
        \AttributeTok{axis.ticks =} \FunctionTok{element\_blank}\NormalTok{(),}
        \AttributeTok{axis.text =} \FunctionTok{element\_blank}\NormalTok{()) }\SpecialCharTok{+}
  \FunctionTok{coord\_flip}\NormalTok{()}

\NormalTok{pTemp }\SpecialCharTok{+} \FunctionTok{plot\_spacer}\NormalTok{() }\SpecialCharTok{+}\NormalTok{ p0 }\SpecialCharTok{+}\NormalTok{ pOzone }\SpecialCharTok{+} 
  \FunctionTok{plot\_layout}\NormalTok{(}\AttributeTok{widths =} \FunctionTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{,}\DecValTok{1}\NormalTok{), }\AttributeTok{heights =} \FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{3}\NormalTok{)) }
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{arXiv_bigsimr_files/figure-latex/ch030-aq-joint-dist-1.pdf}
\caption{Bivariate scatterplot of Ozone versus Temp with estimated
marginal densities. The Ozone data are modeled marginally as log-normal
and the Temperature data as normal.}
\end{figure}

Next, we specify the marginal distributions and correlation coefficient
(both type and magnitude). Here the analyst is free to be creative. For
this example, we avoid goodness-of-fit considerations to determine the
marginal distributions. It is sensible without domain knowledge to
estimate these quantities from the data, and \texttt{bigsimr} contains
fast functions designed for this task.

\hypertarget{specifying-marginal-distributions}{%
\subsection{Specifying marginal
distributions}\label{specifying-marginal-distributions}}

Based on the estimated densities in Figure
@ref(fig:ch030-aq-joint-dist), we assume \texttt{Temp} is normally
distributed and \texttt{Ozone} is log-normally distributed, as the
latter values are positive and skewed. We use the classical unbiased
estimators for the normal distribution's parameters and maximum
likelihood estimators for the log-normal parameters:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{df }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{select}\NormalTok{(Temp) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{summarise\_all}\NormalTok{(}\AttributeTok{.funs =} \FunctionTok{c}\NormalTok{(}\AttributeTok{mean =}\NormalTok{ mean, }\AttributeTok{sd =}\NormalTok{ sd))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##       mean       sd
## 1 77.87069 9.485486
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{mle\_mean }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(x) }\FunctionTok{mean}\NormalTok{(}\FunctionTok{log}\NormalTok{(x))}
\NormalTok{mle\_sd }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(x) }\FunctionTok{mean}\NormalTok{( }\FunctionTok{sqrt}\NormalTok{( (}\FunctionTok{log}\NormalTok{(x) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{log}\NormalTok{(x)))}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{ ) )}
\NormalTok{df }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{select}\NormalTok{(Ozone) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{summarise\_all}\NormalTok{(}\AttributeTok{.funs =} \FunctionTok{c}\NormalTok{(}\AttributeTok{meanlog =}\NormalTok{ mle\_mean, }\AttributeTok{sdlog =}\NormalTok{ mle\_sd))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##    meanlog     sdlog
## 1 3.418515 0.6966689
\end{verbatim}

Now, we configure the input marginals for later input into
\texttt{rvec}. The marginal distributions are specified using
\texttt{Julia}'s \texttt{Distributions} package and stored in a vector.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{margins }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(dist}\SpecialCharTok{$}\FunctionTok{Normal}\NormalTok{(}\FunctionTok{mean}\NormalTok{(df}\SpecialCharTok{$}\NormalTok{Temp), }\FunctionTok{sd}\NormalTok{(df}\SpecialCharTok{$}\NormalTok{Temp)),}
\NormalTok{             dist}\SpecialCharTok{$}\FunctionTok{LogNormal}\NormalTok{(}\FunctionTok{mle\_mean}\NormalTok{(df}\SpecialCharTok{$}\NormalTok{Ozone), }\FunctionTok{mle\_sd}\NormalTok{(df}\SpecialCharTok{$}\NormalTok{Ozone)))}
\end{Highlighting}
\end{Shaded}

\hypertarget{specifying-correlation}{%
\subsection{Specifying correlation}\label{specifying-correlation}}

The user must decide how to describe correlation based on the
particulars of the problem. For non-normal data in our scheme, we
advocate the use of Spearman's \(\rho\) correlation matrix \(R_S\) or
Kendall's \(\tau\) correlation matrix \(R_K\). We also support Pearson
correlation coefficient matching, while cautioning the user to check the
performance for the distribution at hand (see
\protect\hyperlink{simulations}{Monte Carlo evaluations} below for
evaluation strategies and guidance). Note that these estimation methods
are classical approaches, not specifically designed for high-dimensional
correlation estimation.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{(R\_S }\OtherTok{\textless{}{-}}\NormalTok{ bs}\SpecialCharTok{$}\FunctionTok{cor}\NormalTok{(}\FunctionTok{as.matrix}\NormalTok{(df), bs}\SpecialCharTok{$}\NormalTok{Spearman))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##          [,1]     [,2]
## [1,] 1.000000 0.774043
## [2,] 0.774043 1.000000
\end{verbatim}

\hypertarget{checking-target-correlation-matrix-admissibility}{%
\subsection{Checking target correlation matrix
admissibility}\label{checking-target-correlation-matrix-admissibility}}

Once a target correlation matrix is specified, first convert the
dependency type to Pearson to correctly specify the MVN inputs. Then,
check that the converted matrix \(R_X\) is a valid correlation matrix
(PD and 1s on the diagonal). For this bivariate example, \(R_X\) is a
valid correlation matrix. Note that typically in HD the resultant MVN
Pearson correlation matrix is indefinite and requires approximation (see
the HD examples in subsequent sections).

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Step 1. Mapping}
\NormalTok{(R\_X }\OtherTok{\textless{}{-}}\NormalTok{ bs}\SpecialCharTok{$}\FunctionTok{cor\_convert}\NormalTok{(R\_S, bs}\SpecialCharTok{$}\NormalTok{Spearman, bs}\SpecialCharTok{$}\NormalTok{Pearson))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##           [,1]      [,2]
## [1,] 1.0000000 0.7885668
## [2,] 0.7885668 1.0000000
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Step 2. Check admissibility}
\NormalTok{bs}\SpecialCharTok{$}\FunctionTok{iscorrelation}\NormalTok{(R\_X)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] TRUE
\end{verbatim}

Despite being a valid correlation matrix under this definition, marginal
distributions induce Frechet bounds on the possible Pearson correlation
values. To check that the converted correlation values fall within these
bounds, use \texttt{bigsimr::cor\_bounds} to estimate the pairwise lower
and upper correlation bounds. \texttt{cor\_bounds} uses the Generate,
Sort, and Correlate algorithm of Demirtas and Hedeker (2011). For the
assumed marginals, the Pearson correlation coefficient is not free to
vary in {[}-1,1{]}, as seen below. Our MC estimate of the bounds
slightly underestimates the theoretical bounds of \((-0.881, 0.881)\).
(See an analytic derivation presented in the Appendix). Since our single
Pearson correlation coefficient is within the theoretical bounds, the
correlation matrix is valid for our simulation strategy.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{bs}\SpecialCharTok{$}\FunctionTok{cor\_bounds}\NormalTok{(margins[}\DecValTok{1}\NormalTok{], margins[}\DecValTok{2}\NormalTok{], bs}\SpecialCharTok{$}\NormalTok{Pearson, }\AttributeTok{n\_samples =} \FloatTok{1e6}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Julia Object of type NamedTuple{(:lower, :upper),Tuple{Float64,Float64}}.
## (lower = -0.8806384281235534, upper = 0.880733034303322)
\end{verbatim}

In this 2D example, it is possible to verify the existence of a
bivariate distribution with the target Pearson correlation / margins
(See the {[}Appendix{]}). For general \(d-\)variate distributions, is it
not guaranteed that such a multivariate distribution exists with exact
the specified correlations and margins (Barbiero and Ferrari 2017).
Theoretical guidance is lacking in this regard, but, in practice, we
find that using bivariate bounds form rough approximations to the bounds
for the feasible region for the \(d-\)variate construction when
aggregated into a Pearson matrix. See the
\protect\hyperlink{examples}{RNA-seq data application} section for an
example discussing this, NORTA feasibility, and use of the nearest PD
matrix algorithm.

\hypertarget{simulating-random-vectors}{%
\subsection{Simulating random vectors}\label{simulating-random-vectors}}

Finally, we execute \texttt{rvec} to simulate the desired \(10,000\)
random vectors from the assumed bivariate distribution of \texttt{Ozone}
and \texttt{Temp}. Note that the target input is specified using the
pre-computed MVN correlation matrix. Figure @ref(fig:ch030-plot-sim)
plots the 10,000 simulated points.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{x }\OtherTok{\textless{}{-}}\NormalTok{ bs}\SpecialCharTok{$}\FunctionTok{rvec}\NormalTok{(}\DecValTok{10000}\NormalTok{, R\_X, margins)}
\NormalTok{df\_sim }\OtherTok{\textless{}{-}} \FunctionTok{as.data.frame}\NormalTok{(x)}
\FunctionTok{colnames}\NormalTok{(df\_sim) }\OtherTok{\textless{}{-}} \FunctionTok{colnames}\NormalTok{(df)}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{p1 }\OtherTok{\textless{}{-}}\NormalTok{ df\_sim }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(Temp, Ozone)) }\SpecialCharTok{+}
  \FunctionTok{geom\_density\_2d\_filled}\NormalTok{() }\SpecialCharTok{+}
  \FunctionTok{theme}\NormalTok{(}\AttributeTok{legend.position =} \StringTok{"none"}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{labs}\NormalTok{(}\AttributeTok{x =} \StringTok{"Simulated Temperature"}\NormalTok{, }\AttributeTok{y =} \StringTok{"Simulated Ozone"}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{scale\_y\_continuous}\NormalTok{(}\AttributeTok{limits =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\FunctionTok{max}\NormalTok{(df}\SpecialCharTok{$}\NormalTok{Ozone))) }\SpecialCharTok{+}
  \FunctionTok{scale\_x\_continuous}\NormalTok{(}\AttributeTok{limits =} \FunctionTok{range}\NormalTok{(df}\SpecialCharTok{$}\NormalTok{Temp))}

\NormalTok{p1Temp }\OtherTok{\textless{}{-}} \FunctionTok{ggplot}\NormalTok{(df\_sim, }\FunctionTok{aes}\NormalTok{(Temp)) }\SpecialCharTok{+} 
  \FunctionTok{geom\_density}\NormalTok{(}\AttributeTok{alpha =} \FloatTok{0.5}\NormalTok{, }\AttributeTok{fill =} \StringTok{"lightseagreen"}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{theme}\NormalTok{(}\AttributeTok{axis.title.x =} \FunctionTok{element\_blank}\NormalTok{(),}
        \AttributeTok{axis.title.y =} \FunctionTok{element\_blank}\NormalTok{(),}
        \AttributeTok{axis.text =} \FunctionTok{element\_blank}\NormalTok{(),}
        \AttributeTok{axis.ticks =} \FunctionTok{element\_blank}\NormalTok{())}

\NormalTok{p1Ozone }\OtherTok{\textless{}{-}} \FunctionTok{ggplot}\NormalTok{(df\_sim, }\FunctionTok{aes}\NormalTok{(Ozone)) }\SpecialCharTok{+} 
  \FunctionTok{geom\_density}\NormalTok{(}\AttributeTok{alpha =} \FloatTok{0.5}\NormalTok{, }\AttributeTok{fill =} \StringTok{"lightseagreen"}\NormalTok{) }\SpecialCharTok{+} 
  \FunctionTok{theme}\NormalTok{(}\AttributeTok{axis.title.y =} \FunctionTok{element\_blank}\NormalTok{(),}
        \AttributeTok{axis.title.x =} \FunctionTok{element\_blank}\NormalTok{(),}
        \AttributeTok{axis.ticks =} \FunctionTok{element\_blank}\NormalTok{(),}
        \AttributeTok{axis.text =} \FunctionTok{element\_blank}\NormalTok{()) }\SpecialCharTok{+}
  \FunctionTok{coord\_flip}\NormalTok{() }

\NormalTok{p1Temp }\SpecialCharTok{+} \FunctionTok{plot\_spacer}\NormalTok{() }\SpecialCharTok{+}\NormalTok{ p1 }\SpecialCharTok{+}\NormalTok{ p1Ozone }\SpecialCharTok{+} 
  \FunctionTok{plot\_layout}\NormalTok{(}\AttributeTok{widths =} \FunctionTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{,}\DecValTok{1}\NormalTok{), }\AttributeTok{heights =} \FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{3}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: Removed 388 rows containing non-finite values (stat_density2d_filled).
\end{verbatim}

\begin{figure}
\centering
\includegraphics{arXiv_bigsimr_files/figure-latex/ch030-plot-sim-1.pdf}
\caption{Contour plot and marginal densities for the simulated bivariate
distribution of Air Quality Temperatures and Ozone levels. The simulated
points mimic the observed data with respect to both the marginal
characteristics and bivariate association.}
\end{figure}

\hypertarget{simulations}{%
\section{Monte Carlo evaluations}\label{simulations}}

In this section, we conduct MC studies to investigate method
performance. Marginal parameter matching is straightforward in our
scheme as it is a sequence of univariate inverse probability transforms.
This process is both fast and exact. On the other hand, accurate and
computationally efficient dependency matching presents a challenge,
especially in the HD setting. To evaluate our methods in those respects,
we design the following numerical experiments to first assess accuracy
of matching dependency parameters in bivariate simulations and then time
the procedure in increasingly large dimension \(d\).

\hypertarget{bivariate-experiments}{%
\subsection{Bivariate experiments}\label{bivariate-experiments}}

We select bivariate simulation configurations to ultimately simulate our
motivating discrete-valued RNA-seq example. Thus, we proceed by
increasing departure from normality, leading to a multivariate negative
binomial (MVNB) model for our motivating data. We begin with empirically
evaluating the dependency matching across all three supported
correlations --- Pearson, Spearman, and Kendall --- in identical,
bivariate marginal configurations. For each pair of identical margins,
we vary the target correlation across \(\Omega\), the set of possible
admissible values for each correlation type, to evaluate the
simulation's ability to generate all possible correlations. The
simulations progress from bivariate normal, to bivariate gamma
(non-normal yet continuous), and bivariate negative binomial (non-normal
and discrete).

Table @ref(tab:sims) lists our identical-marginal, bivariate simulation
configurations. We increase the simulation replicates \(B\) to visually
ensure that our results converge to the target correlations and gauge
efficiency. We select distributions beginning with a standard
multivariate normal (MVN) as we expect the performance to be exact (up
to MC error) for all correlation types. Then, we select a non-symmetric
continuous distribution: a standard (rate =1), two-component
multivariate gamma (MVG). Finally, we select distributions and marginal
parameter values that are motivated by our RNA-seq data, namely values
proximal to probabilities and sizes estimated from the data (see
\href{examples}{RNA-seq data application} for estimation details).
Specifically, we assume a MVNB with
\(p_1 = p_2 = 3\times10^{-4}, r_1 = r_2 = 4, \rho \in \Omega\).

\begin{table}[]
\centering
\caption{ \label{tab:sims} Identical margin, bivariate simulation configurations to evaluate matching accuracy.}
\begin{tabular}{@{}lcr@{}}
\toprule
Simulation Reps $B$ & Correlation Types & Identical-margin 2D distribution \\ \midrule
$1,000$ & Pearson ($\rho_P$) & ${\bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho_i ), i=1,\ldots,100$ \\
$10,000$ & Spearman ($\rho_S$) & ${\bf Y} \sim MVG( shape = 10, rate = 1, \rho_i ), i=1,\ldots,100$ \\
$100,000$ & Kendall ($\tau$) & ${\bf Y} \sim MVNB(p = 3\times10^{-4}, r = 4,\rho_i), i=1,\ldots,100$ \\ \bottomrule
\end{tabular}
\end{table}

For each of the unique 9 simulation configurations described above, we
estimate the correlation bounds and vary the correlations along a
sequence of 100 points evenly placed within the bounds, aiming to
explore \(\Omega\). Specifically, we set correlations
\(\{ \rho_1 = ( \hat{l} + \epsilon), \rho_2 = (\hat{l} + \epsilon) + \delta, \ldots, \rho_{100} = (\hat{u} - \epsilon) \}\),
with \(\hat{l}\) and \(\hat{u}\) being the estimated lower and upper
bounds, respectively, with increment value \(\delta\). The adjustment
factor, \(\epsilon=0.01\), is introduced to handle numeric issues when
the bound is specified exactly.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{bivariate\_normal\_sims }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(rho, rho\_hat, }\AttributeTok{color =}\NormalTok{ type)) }\SpecialCharTok{+}
    \FunctionTok{geom\_point}\NormalTok{() }\SpecialCharTok{+}
    \FunctionTok{geom\_abline}\NormalTok{(}\AttributeTok{slope =} \DecValTok{1}\NormalTok{) }\SpecialCharTok{+}
    \FunctionTok{facet\_wrap}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{ type }\SpecialCharTok{+}\NormalTok{ N) }\SpecialCharTok{+} 
    \FunctionTok{theme\_bw}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{arXiv_bigsimr_files/figure-latex/ch040-biNormPlot-1.pdf}
\caption{\texttt{bigsimr} recovers the Pearson specified correlations
for MVN.}
\end{figure}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{bivariate\_gamma\_sims }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(rho, rho\_hat, }\AttributeTok{color =}\NormalTok{ type)) }\SpecialCharTok{+}
    \FunctionTok{geom\_point}\NormalTok{() }\SpecialCharTok{+}
    \FunctionTok{geom\_abline}\NormalTok{(}\AttributeTok{slope =} \DecValTok{1}\NormalTok{) }\SpecialCharTok{+}
    \FunctionTok{facet\_wrap}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{ type }\SpecialCharTok{+}\NormalTok{ N) }\SpecialCharTok{+} 
    \FunctionTok{theme\_bw}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{arXiv_bigsimr_files/figure-latex/ch040-biGammaPlot-1.pdf}
\caption{\texttt{bigsimr} recovers the Pearson specified correlations
for Bivariate Gamma.}
\end{figure}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{bivariate\_nbinom\_sims }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(rho, rho\_hat, }\AttributeTok{color =}\NormalTok{ type)) }\SpecialCharTok{+}
    \FunctionTok{geom\_point}\NormalTok{() }\SpecialCharTok{+}
    \FunctionTok{geom\_abline}\NormalTok{(}\AttributeTok{slope =} \DecValTok{1}\NormalTok{) }\SpecialCharTok{+}
    \FunctionTok{facet\_wrap}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{ type }\SpecialCharTok{+}\NormalTok{ N) }\SpecialCharTok{+} 
    \FunctionTok{theme\_bw}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{arXiv_bigsimr_files/figure-latex/ch040-biNegBinPlot-1.pdf}
\caption{\texttt{bigsimr} recovers the correlations for bivariate
negative binomial only approximately for Pearson but (nearly) exactly
for the rank-based correlations.}
\end{figure}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{allDat }\OtherTok{\textless{}{-}} \FunctionTok{bind\_rows}\NormalTok{(}
    \FunctionTok{select}\NormalTok{(bivariate\_normal\_sims, margins, type, N, rho, rho\_hat),}
    \FunctionTok{select}\NormalTok{(bivariate\_gamma\_sims, margins, type, N, rho, rho\_hat),}
    \FunctionTok{select}\NormalTok{(bivariate\_nbinom\_sims, margins, type, N, rho, rho\_hat)}
\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{margins =} \FunctionTok{factor}\NormalTok{(margins, }\AttributeTok{levels=}\FunctionTok{c}\NormalTok{(}\StringTok{"norm"}\NormalTok{, }\StringTok{"gamma"}\NormalTok{, }\StringTok{"nbinom"}\NormalTok{)),}
           \AttributeTok{type =} \FunctionTok{factor}\NormalTok{(type, }\AttributeTok{levels=}\FunctionTok{c}\NormalTok{(}\StringTok{"Pearson"}\NormalTok{, }\StringTok{"Spearman"}\NormalTok{, }\StringTok{"Kendall"}\NormalTok{)),}
           \AttributeTok{N =} \FunctionTok{factor}\NormalTok{(N))}
\end{Highlighting}
\end{Shaded}

Figure @ref(fig:ch040-bPlot) displays the aggregated bivariate
simulation results. Table @ref(tab:ch040-BiError) contains the mean
absolute error (MAE) in reproducing the desired dependency measures for
the three bivariate scenarios. Overall, the studies show that our
methodology is generally accurate across the entire range of possible
correlation for all three dependency measures. Our Pearson matching
performs nearly as well as Spearman or Kendall, except for a slight
increase in error for the negative binomial case.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{tabMAE }\OtherTok{\textless{}{-}}\NormalTok{ allDat }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{group\_by}\NormalTok{(N, type, margins) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{summarize}\NormalTok{(}\AttributeTok{MAE =} \FunctionTok{mean}\NormalTok{(}\FunctionTok{abs}\NormalTok{(rho }\SpecialCharTok{{-}}\NormalTok{ rho\_hat))) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{ungroup}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## `summarise()` regrouping output by 'N', 'type' (override with `.groups` argument)
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{kable}\NormalTok{(tabMAE, }\AttributeTok{booktabs =} \ConstantTok{TRUE}\NormalTok{, }\AttributeTok{format =} \StringTok{"latex"}\NormalTok{,}
      \AttributeTok{linesep =} \FunctionTok{c}\NormalTok{(}\StringTok{""}\NormalTok{, }\StringTok{""}\NormalTok{, }\StringTok{"}\SpecialCharTok{\textbackslash{}\textbackslash{}}\StringTok{addlinespace"}\NormalTok{),}
      \AttributeTok{col.names =} \FunctionTok{c}\NormalTok{(}\StringTok{"No. of random vectors"}\NormalTok{,}
                    \StringTok{"Correlation type"}\NormalTok{,}
                    \StringTok{"Distribution"}\NormalTok{,}
                    \StringTok{"Mean abs. error"}\NormalTok{),}
      \AttributeTok{caption =} \StringTok{"Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal."}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{table}

\caption{\label{tab:ch040-BiError}Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.}
\centering
\begin{tabular}[t]{lllr}
\toprule
No. of random vectors & Correlation type & Distribution & Mean abs. error\\
\midrule
1000 & Pearson & norm & 0.0156184\\
1000 & Pearson & gamma & 0.0159287\\
1000 & Pearson & nbinom & 0.0162465\\
\addlinespace
1000 & Spearman & norm & 0.0175440\\
1000 & Spearman & gamma & 0.0182176\\
1000 & Spearman & nbinom & 0.0151124\\
\addlinespace
1000 & Kendall & norm & 0.0130284\\
1000 & Kendall & gamma & 0.0114740\\
1000 & Kendall & nbinom & 0.0126014\\
\addlinespace
10000 & Pearson & norm & 0.0060226\\
10000 & Pearson & gamma & 0.0057667\\
10000 & Pearson & nbinom & 0.0058180\\
\addlinespace
10000 & Spearman & norm & 0.0062388\\
10000 & Spearman & gamma & 0.0056132\\
10000 & Spearman & nbinom & 0.0049256\\
\addlinespace
10000 & Kendall & norm & 0.0032618\\
10000 & Kendall & gamma & 0.0038067\\
10000 & Kendall & nbinom & 0.0033203\\
\addlinespace
1e+05 & Pearson & norm & 0.0017570\\
1e+05 & Pearson & gamma & 0.0016962\\
1e+05 & Pearson & nbinom & 0.0029582\\
\addlinespace
1e+05 & Spearman & norm & 0.0016607\\
1e+05 & Spearman & gamma & 0.0016408\\
1e+05 & Spearman & nbinom & 0.0015269\\
\addlinespace
1e+05 & Kendall & norm & 0.0010441\\
1e+05 & Kendall & gamma & 0.0011077\\
1e+05 & Kendall & nbinom & 0.0011976\\
\bottomrule
\end{tabular}
\end{table}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# https://www.datanovia.com/en/blog/how{-}to{-}change{-}ggplot{-}facet{-}labels/}
\CommentTok{\# New facet label names}
\NormalTok{repsLabs }\OtherTok{\textless{}{-}} \FunctionTok{paste0}\NormalTok{(}\StringTok{"B="}\NormalTok{, }\FunctionTok{c}\NormalTok{(}\StringTok{"1,000"}\NormalTok{, }\StringTok{"10,000"}\NormalTok{, }\StringTok{"100,000"}\NormalTok{) )}
\FunctionTok{names}\NormalTok{(repsLabs) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\DecValTok{1000}\NormalTok{, }\DecValTok{10000}\NormalTok{, }\DecValTok{100000}\NormalTok{)}
\NormalTok{typeLabs }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"Pearson"}\NormalTok{, }\StringTok{"Spearman"}\NormalTok{, }\StringTok{"Kendall"}\NormalTok{)}
\FunctionTok{names}\NormalTok{(typeLabs) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"Pearson"}\NormalTok{, }\StringTok{"Spearman"}\NormalTok{, }\StringTok{"Kendall"}\NormalTok{)}

\CommentTok{\# Set colors}
\DocumentationTok{\#\# RColorBrewer::display.brewer.all()}
\NormalTok{numColors }\OtherTok{\textless{}{-}} \DecValTok{4}
\NormalTok{numGroups }\OtherTok{\textless{}{-}} \FunctionTok{length}\NormalTok{(}\FunctionTok{levels}\NormalTok{(allDat}\SpecialCharTok{$}\NormalTok{margins))}
\NormalTok{myColors }\OtherTok{\textless{}{-}} \FunctionTok{rev}\NormalTok{(RColorBrewer}\SpecialCharTok{::}\FunctionTok{brewer.pal}\NormalTok{(}\AttributeTok{n =}\NormalTok{ numColors, }\AttributeTok{name =} \StringTok{"Greys"}\NormalTok{)[ ((numColors }\SpecialCharTok{{-}}\NormalTok{ numGroups) }\SpecialCharTok{+} \DecValTok{1}\NormalTok{)}\SpecialCharTok{:}\NormalTok{numColors])}

\NormalTok{allDat }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ rho, }\AttributeTok{y =}\NormalTok{ rho\_hat, }\AttributeTok{color =}\NormalTok{ margins)) }\SpecialCharTok{+}
    \FunctionTok{geom\_point}\NormalTok{(}\AttributeTok{size =} \DecValTok{2}\NormalTok{) }\SpecialCharTok{+}
    \FunctionTok{scale\_color\_manual}\NormalTok{(}\AttributeTok{values =}\NormalTok{ myColors ) }\SpecialCharTok{+}
    \FunctionTok{geom\_abline}\NormalTok{(}\AttributeTok{slope =} \DecValTok{1}\NormalTok{, }\AttributeTok{linetype =} \StringTok{"dashed"}\NormalTok{) }\SpecialCharTok{+}
    \FunctionTok{labs}\NormalTok{(}\AttributeTok{x =} \StringTok{"Specified Correlation"}\NormalTok{, }\AttributeTok{y =} \StringTok{"Estimated Correlation"}\NormalTok{) }\SpecialCharTok{+}
    \FunctionTok{facet\_wrap}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{ type }\SpecialCharTok{+}\NormalTok{ N, }\AttributeTok{labeller =} \FunctionTok{labeller}\NormalTok{(}\AttributeTok{N =}\NormalTok{ repsLabs, }\AttributeTok{type =}\NormalTok{ typeLabs)) }\SpecialCharTok{+}
    \FunctionTok{theme\_bw}\NormalTok{() }\SpecialCharTok{+}
    \FunctionTok{theme}\NormalTok{(}\AttributeTok{legend.position =} \StringTok{"bottom"}\NormalTok{, }\AttributeTok{legend.direction =} \StringTok{"horizontal"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{arXiv_bigsimr_files/figure-latex/ch040-bPlot-1.pdf}
\caption{Bivariate simulations match target correlations across the
entire range of feasible correlations. The horizontal axis plots the
specified target correlations for each bivariate margin. Normal margins
are plotted in dark dark grey, gamma in medium grey, and negative
binomial in light grey. As the number of simulated vectors \(B\)
increases from left to right, the variation in estimated correlations
(vertical axis) decreases. The dashed line indicates equality between
the specified and estimated correlations.}
\end{figure}

\hypertarget{scale-up-to-high-dimensions}{%
\subsection{Scale up to High
Dimensions}\label{scale-up-to-high-dimensions}}

With information of our method's accuracy from a low-dimensional
perspective, we now assess whether \texttt{bigsimr} can scale to larger
dimensional problems with practical computation times. We ultimately
generate \(B=1,000\) random vectors for
\(d=\{100, 250, 500, 1000, 2500, 5000, 10000\}\) for each correlation
type, \{Pearson, Spearman, Kendall\} while timing the algorithm's major
steps. We begin by producing a synthetic ``data set'' by completing the
following steps:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Produce heterogeneous gamma marginals by randomly selecting the
  \(j^{th}\) gamma shape parameter from
  \(U_j \sim uniform(1,10), j=1,\ldots,d\) and the \(j^{th}\) rate
  parameter from \(V_j \sim exp(1/5), j=1,\ldots,d\), with the constant
  parameters determined arbitrarily.
\item
  Produce a random full-rank Pearson correlation matrix via
  \texttt{cor\_randPD} of size \(d \times d\).
\item
  Simulate a ``data set'' of \(1,000 \times d\) random vectors via
  \texttt{rvec}.
\end{enumerate}

With the synthetic data set in hand, we complete and time the following
four steps involved in a typical workflow (including correlation
estimation).

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Estimate the correlation matrix from the ``data'' in the \emph{Compute
  Correlation} step.
\item
  Map the correlations to initialize the algorithm (Pearson to Pearson,
  Spearman to Pearson, or Kendall to Pearson) in the \emph{Adjust
  Correlation} step.
\item
  Check whether the mapping produces a valid correlation matrix and, if
  not, find the nearest PD correlation matrix in the \emph{Check
  Admissibility} step.
\item
  Simulate \(1,000\) vectors in the \emph{Simulate Data} step.
\end{enumerate}

The experiments are conducted on a MacBook Pro carrying a 2.4 GHz 8-Core
Intel Core i9 processor, with all 16 threads employed during
computation. Table @ref(tab:ch040-moderateDtab) displays the total
computation time for moderate dimesions (\(d \leq 500\)). In every
simullation setting, the 1,000 vectors are generated rapidly, executing
in under 2 seconds.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{dat\_long }\OtherTok{\textless{}{-}}\NormalTok{ benchmark\_dependences }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{select}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{total\_time, }\SpecialCharTok{{-}}\NormalTok{n\_sim, }\SpecialCharTok{{-}}\NormalTok{needed\_near\_pd) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{pivot\_longer}\NormalTok{(}\AttributeTok{cols =} \FunctionTok{c}\NormalTok{(corr\_time}\SpecialCharTok{:}\NormalTok{sim\_time), }\AttributeTok{names\_to =} \StringTok{"Step"}\NormalTok{, }\AttributeTok{values\_to =} \StringTok{"Time"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{dim =} \FunctionTok{factor}\NormalTok{(dim),}
         \AttributeTok{Step =} \FunctionTok{factor}\NormalTok{(Step,}
                       \AttributeTok{levels =} \FunctionTok{c}\NormalTok{(}\StringTok{"corr\_time"}\NormalTok{, }\StringTok{"adj\_time"}\NormalTok{, }\StringTok{"admiss\_time"}\NormalTok{, }\StringTok{"sim\_time"}\NormalTok{),}
                       \AttributeTok{labels =} \FunctionTok{c}\NormalTok{(}\StringTok{"Compute Correlation"}\NormalTok{,}
                                  \StringTok{"Adjust Correlation"}\NormalTok{,}
                                  \StringTok{"Check Admissibility"}\NormalTok{,}
                                  \StringTok{"Simulate Data"}\NormalTok{))) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{rename}\NormalTok{(}\AttributeTok{Correlation =}\NormalTok{ corr\_type, }\AttributeTok{Dimensions =}\NormalTok{ dim)}

\NormalTok{tabTimes  }\OtherTok{\textless{}{-}}\NormalTok{ dat\_long }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{filter}\NormalTok{(Dimensions }\SpecialCharTok{\%in\%} \FunctionTok{c}\NormalTok{(}\DecValTok{100}\NormalTok{, }\DecValTok{250}\NormalTok{, }\DecValTok{500}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{group\_by}\NormalTok{(Dimensions, Correlation) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{summarize}\NormalTok{(}\StringTok{\textquotesingle{}Total Time(Seconds)\textquotesingle{}} \OtherTok{=} \FunctionTok{sum}\NormalTok{(Time)) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{ungroup}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## `summarise()` regrouping output by 'Dimensions' (override with `.groups` argument)
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{tabTimes }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{kable}\NormalTok{(}\AttributeTok{booktabs =} \ConstantTok{TRUE}\NormalTok{,}
          \AttributeTok{linesep =} \FunctionTok{c}\NormalTok{(}\StringTok{""}\NormalTok{, }\StringTok{""}\NormalTok{, }\StringTok{"}\SpecialCharTok{\textbackslash{}\textbackslash{}}\StringTok{addlinespace"}\NormalTok{), }\AttributeTok{format=}\StringTok{"latex"}\NormalTok{,}
          \AttributeTok{col.names =} \FunctionTok{c}\NormalTok{(}\StringTok{"Dimension"}\NormalTok{,}
                        \StringTok{"Correlation type"}\NormalTok{,}
                        \StringTok{"Total Time (Seconds)"}\NormalTok{),}
          \AttributeTok{caption =} \StringTok{\textquotesingle{}Total time to produce 1,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.\textquotesingle{}}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{table}

\caption{\label{tab:ch040-moderateDtab}Total time to produce 1,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.}
\centering
\begin{tabular}[t]{llr}
\toprule
Dimension & Correlation type & Total Time (Seconds)\\
\midrule
100 & Pearson & 0.081\\
100 & Spearman & 0.027\\
100 & Kendall & 0.075\\
\addlinespace
250 & Pearson & 0.378\\
250 & Spearman & 0.062\\
250 & Kendall & 0.375\\
\addlinespace
500 & Pearson & 1.425\\
500 & Spearman & 0.135\\
500 & Kendall & 1.495\\
\bottomrule
\end{tabular}
\end{table}

The results for \(d > 500\) show scalability to ultra-high dimensions
for all three correlation types, although the total times do become much
larger. Figure @ref(fig:ch040-largeDfig) displays computation times for
\(d=\{1000, 2500, 5000, 10000\}\). For \(d\) equal to 1000 and 2500, the
total time is under a couple of minutes. At \(d\) of 5000 and 10,000,
Pearson correlation matching in the \emph{Adjust Correlation} step
becomes costly. Interestingly, Pearson is actually faster than Kendall
for \(d=10,000\) due to bottlenecks in \emph{Compute Correlation} and
\emph{Check Admissibility}. Uniformly, matching Spearman correlations is
faster, with total times under 5 minutes for \(d=10,000\), making
Spearman the most computationally-friendly dependency type for our
package. With this in mind, we scaled the simulation to \(d=20,000\) for
the Spearman type and obtained the 1,000 vectors in under an hour (data
not shown). In principle, this would enable the simulation of an entire
human-derived RNA-seq data set. We note that for a given target
correlation matrix and margins, steps 1, 2, and 3 only need to be
computed once and the fourth step, \emph{Simulate Data}, is nearly
instaneous for all settings considered.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Set colors}
\NormalTok{numColors }\OtherTok{\textless{}{-}} \DecValTok{5}
\NormalTok{numGroups }\OtherTok{\textless{}{-}} \DecValTok{4}
\NormalTok{myColors }\OtherTok{\textless{}{-}} \FunctionTok{rev}\NormalTok{(}\FunctionTok{brewer.pal}\NormalTok{(}\AttributeTok{n=}\NormalTok{numColors, }\AttributeTok{name=}\StringTok{"Greys"}\NormalTok{)[((numColors}\SpecialCharTok{{-}}\NormalTok{numGroups)}\SpecialCharTok{+}\DecValTok{1}\NormalTok{)}\SpecialCharTok{:}\NormalTok{numColors])}

\NormalTok{dat\_long }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{filter}\NormalTok{(Dimensions }\SpecialCharTok{\%in\%} \FunctionTok{c}\NormalTok{(}\DecValTok{1000}\NormalTok{, }\DecValTok{2500}\NormalTok{, }\DecValTok{5000}\NormalTok{, }\DecValTok{10000}\NormalTok{, }\DecValTok{20000}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{mutate}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Time (minutes)}\StringTok{\textasciigrave{}} \OtherTok{=}\NormalTok{ Time }\SpecialCharTok{/} \DecValTok{60}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(Correlation, }\StringTok{\textasciigrave{}}\AttributeTok{Time (minutes)}\StringTok{\textasciigrave{}}\NormalTok{, }\AttributeTok{fill=}\NormalTok{Step)) }\SpecialCharTok{+}
    \FunctionTok{geom\_bar}\NormalTok{(}\AttributeTok{position =} \StringTok{"stack"}\NormalTok{, }\AttributeTok{stat =} \StringTok{"identity"}\NormalTok{) }\SpecialCharTok{+}
    \FunctionTok{scale\_fill\_manual}\NormalTok{(}\AttributeTok{values =}\NormalTok{ myColors ) }\SpecialCharTok{+}
    \FunctionTok{scale\_y\_continuous}\NormalTok{(}\AttributeTok{breaks =} \FunctionTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{80}\NormalTok{, }\DecValTok{10}\NormalTok{),}
                     \AttributeTok{minor\_breaks =} \ConstantTok{NULL}\NormalTok{) }\SpecialCharTok{+}
    \FunctionTok{facet\_grid}\NormalTok{(. }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Dimensions) }\SpecialCharTok{+}
    \FunctionTok{theme}\NormalTok{(}\AttributeTok{axis.text.x =} \FunctionTok{element\_text}\NormalTok{(}\AttributeTok{angle =} \SpecialCharTok{{-}}\DecValTok{90}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{arXiv_bigsimr_files/figure-latex/ch040-largeDfig-1.pdf}
\caption{Computation times for HD multivariate gamma simulation.}
\end{figure}

\emph{Limitations, conclusions, and recommendations}

In the bivariate studies, we chose arbitrary simulation parameters for
three distributions, moving from the Gaussian to discrete and non-normal
MVNB. Under these conditions, the simulated random vectors sample the
desired bivariate distribution across the entire range of pairwise
correlations for the three dependency measures. The simulation results
could differ for other choices of simulation settings. Specifying
extreme correlations near the boundary or Frechet bounds could result in
poor simulation performance. Fortunately, it is straightforward to
evaluate simulation performance by using strategies similar to those
completed above. We expect our random vector generation to perform well
for the vast majority of NORTA-feasible correlation matrices, but advise
to check the performance before making inferences/further analyses.

Somewhat surprisingly, Kendall estimation and nearest PD computation
scale poorly compared to Spearman and, even, approximate Pearson
matching. In our experience, Kendall computation times are sensitive to
the number of cores, benefiting from multi-core parallelization. This
could mitigate some of the current algorithmic/implementation
shortcomings. Despite this, Kendall matching is still feasible for most
HD data sets. Finally, we note that one could use our single-pass
algorithm \texttt{cor\_fastPD} to produce a `close' (not nearest PD) to
scale to even higher dimensions with some loss of accuracy.

\hypertarget{examples}{%
\section{RNA-seq data application}\label{examples}}

This section demonstrates how to simulate multivariate data using
\texttt{bigsimr}, aiming to replicate the structure of HD dependent
count data. In an illustration of our proposed methodology, we seek to
simulate RNA-sequencing data by producing simulated random vectors
mimicking the observed data and its generating process. Modeling RNA-seq
using multivariate probability distributions is natural as inter-gene
correlation occurs during biological processes (Wang, Gerstein, and
Snyder 2009). And yet, many models do not account for this, leading to
major disruptions to the operating characteristics of statistical
estimation, testing, and prediction. The following subsections apply
\texttt{bigsimr}'s methods to real RNA-seq data, including replicating
an estimated parametric structure, probability estimation, and
evaluation of correlation estimation efficiency.

\hypertarget{simulating-high-dimensional-rna-seq-data}{%
\subsection{Simulating High-Dimensional RNA-seq
data}\label{simulating-high-dimensional-rna-seq-data}}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{d }\OtherTok{\textless{}{-}} \DecValTok{1000}
\NormalTok{brca1000 }\OtherTok{\textless{}{-}}\NormalTok{ example\_brca }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{select}\NormalTok{(}\FunctionTok{all\_of}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\NormalTok{d)) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}\FunctionTok{across}\NormalTok{(}\FunctionTok{everything}\NormalTok{(), as.double))}
\end{Highlighting}
\end{Shaded}

We begin by estimating the structure of the TCGA BRCA RNA-seq data set.
Ultimately, we will simulate \(B=10,000\) random vectors
\({\bf Y}=(Y_1, \ldots, Y_d)^\top\) with \(d=1000\). We assume a MVNB
model as RNA-seq counts are often over-dispersed and correlated. Since
all \(d\) selected genes exhibit over-dispersion (data not shown), we
proceed to estimate the NB parameters \((r_i, p_i), i=1,\ldots,d\), to
determine the target marginal PMFs \(f_i\). To complete specification of
the simulation algorithm inputs, we estimate the Spearman correlation
matrix \(R_S\) to characterize dependency. With this goal in mind, we
first estimate the desired correlation matrix using the fast
implementation provided by \texttt{bigsimr}:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Estimate Spearman\textquotesingle{}s correlation on the count data}
\NormalTok{R\_S }\OtherTok{\textless{}{-}}\NormalTok{ bs}\SpecialCharTok{$}\FunctionTok{cor}\NormalTok{(}\FunctionTok{as.matrix}\NormalTok{(brca1000), bs}\SpecialCharTok{$}\NormalTok{Spearman)}
\end{Highlighting}
\end{Shaded}

Next, we estimate the marginal parameters. We use the method of moments
to estimate the marginal parameters for the multivariate negative
binomial model. The marginal distributions are from the same probability
family (NB), yet they are heterogeneous in terms of the parameters
probability and size \((p_i, n_i)\) for \(i,\ldots,d\). The functions
below support this estimation for later use in \texttt{rvec}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{make\_nbinom\_margins }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(sizes, probs) \{}
\NormalTok{  margins }\OtherTok{\textless{}{-}} \FunctionTok{lapply}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(sizes), }\ControlFlowTok{function}\NormalTok{(i) \{}
\NormalTok{    dist}\SpecialCharTok{$}\FunctionTok{NegativeBinomial}\NormalTok{(sizes[i], probs[i])}
\NormalTok{  \})}
  \FunctionTok{do.call}\NormalTok{(c, margins)}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

We apply these estimators to the highest-expressing r d genes across the
r nrow(brca1000) patients:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{mom\_nbinom }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(x) \{}
\NormalTok{  m }\OtherTok{\textless{}{-}} \FunctionTok{mean}\NormalTok{(x)}
\NormalTok{  s }\OtherTok{\textless{}{-}} \FunctionTok{sd}\NormalTok{(x)}
  \FunctionTok{c}\NormalTok{(}\AttributeTok{size =}\NormalTok{ m}\SpecialCharTok{\^{}}\DecValTok{2} \SpecialCharTok{/}\NormalTok{ (s}\SpecialCharTok{\^{}}\DecValTok{2} \SpecialCharTok{{-}}\NormalTok{ m), }\AttributeTok{prob =}\NormalTok{ m }\SpecialCharTok{/}\NormalTok{ s}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}
\NormalTok{\}}
\NormalTok{nbinom\_fit }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(brca1000, }\DecValTok{2}\NormalTok{, mom\_nbinom)}
\NormalTok{sizes }\OtherTok{\textless{}{-}}\NormalTok{ nbinom\_fit[}\StringTok{"size"}\NormalTok{,]}
\NormalTok{probs }\OtherTok{\textless{}{-}}\NormalTok{ nbinom\_fit[}\StringTok{"prob"}\NormalTok{,]}
\NormalTok{nb\_margins }\OtherTok{\textless{}{-}} \FunctionTok{make\_nbinom\_margins}\NormalTok{(sizes, probs)}
\end{Highlighting}
\end{Shaded}

Notably, the estimated marginal NB probabilities \(\{ \hat{p}_i \}\) are
small --- ranging in the interval \([r min(probs) , r max(probs)]\).
This gives rise to highly variable counts and, thus, less restriction on
potential pairwise correlation pairs. Given the marginals, we now
specify targets and check admissibility of the specified correlation
matrix.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# 1. Mapping step first}
\NormalTok{R\_X }\OtherTok{\textless{}{-}}\NormalTok{ bs}\SpecialCharTok{$}\FunctionTok{cor\_convert}\NormalTok{(R\_S, bs}\SpecialCharTok{$}\NormalTok{Spearman, bs}\SpecialCharTok{$}\NormalTok{Pearson)}
\CommentTok{\# 2a. Check admissibility}
\NormalTok{(is\_valid\_corr }\OtherTok{\textless{}{-}}\NormalTok{ bs}\SpecialCharTok{$}\FunctionTok{iscorrelation}\NormalTok{(R\_X))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] FALSE
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# 2b. compute nearest correlation}
\ControlFlowTok{if}\NormalTok{ (}\SpecialCharTok{!}\NormalTok{is\_valid\_corr) \{}
\NormalTok{  R\_X\_pd }\OtherTok{\textless{}{-}}\NormalTok{ bs}\SpecialCharTok{$}\FunctionTok{cor\_nearPD}\NormalTok{(R\_X)}
  \DocumentationTok{\#\# Quantify the error}
\NormalTok{  targets      }\OtherTok{\textless{}{-}}\NormalTok{ R\_X[}\FunctionTok{lower.tri}\NormalTok{(R\_X, }\AttributeTok{diag =} \ConstantTok{FALSE}\NormalTok{)]}
\NormalTok{  approximates }\OtherTok{\textless{}{-}}\NormalTok{ R\_X\_pd[}\FunctionTok{lower.tri}\NormalTok{(R\_X\_pd, }\AttributeTok{diag =} \ConstantTok{FALSE}\NormalTok{)]}
\NormalTok{  R\_X          }\OtherTok{\textless{}{-}}\NormalTok{ R\_X\_pd}
\NormalTok{\}}
\FunctionTok{summary}\NormalTok{(}\FunctionTok{abs}\NormalTok{(targets }\SpecialCharTok{{-}}\NormalTok{ approximates))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##      Min.   1st Qu.    Median      Mean   3rd Qu.      Max. 
## 0.0000000 0.0001929 0.0004129 0.0005064 0.0007168 0.0069224
\end{verbatim}

As seen above, \(R_X\) is not strictly admissible in our scheme. This
can be seen from the negative result from bs\$iscorrelation(R\_X), which
checks if \(R_X\) is positive-definite with 1's along the diagonal.
However, the approximation is close with a maximum absolute error of r
max( abs(targets - approximates)) and average absolute error of r
mean(abs(targets - approximates)) across the r choose(d,2) correlations.
With the inputs configured, \texttt{rvec} is executed to produce a
synthetic RNA-seq data set:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{sim\_nbinom }\OtherTok{\textless{}{-}}\NormalTok{ bs}\SpecialCharTok{$}\FunctionTok{rvec}\NormalTok{(}\DecValTok{10000}\NormalTok{, R\_X, nb\_margins) }
\end{Highlighting}
\end{Shaded}

Figure @ref(fig:ch050-simDataFig) displays the simulated counts and
pairwise relationships for our example genes from Table
@ref(tab:ch010-realDataTab). Simulated counts roughly mimic the observed
data but with a smoother appearance due to the assumed parametric form
and with less extreme points than the observed data in Figure
@ref(fig:ch010-realDataFig). Figure @ref(fig:ch050-figBRCA) compares the
specified target parameter (horizontal axis) with the corresponding
quantities estimated from the simulated data (vertical axis). The
evaluation shows that the simulated counts approximately match the
target parameters and exhibit the full range of estimated correlations
from the data.

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{set.seed}\NormalTok{(}\DecValTok{2020{-}02{-}25}\NormalTok{)}
\NormalTok{num\_genes }\OtherTok{\textless{}{-}} \DecValTok{3}
\NormalTok{gene\_sample }\OtherTok{\textless{}{-}} \FunctionTok{sample}\NormalTok{(example\_genes[}\DecValTok{1}\SpecialCharTok{:}\DecValTok{1000}\NormalTok{], num\_genes)}

\FunctionTok{ggpairs}\NormalTok{(}\AttributeTok{data =} \FunctionTok{as.data.frame}\NormalTok{(sim\_nbinom[, gene\_sample]),}
        \AttributeTok{upper =} \FunctionTok{list}\NormalTok{(}\AttributeTok{continuous =} \FunctionTok{wrap}\NormalTok{(}\StringTok{\textquotesingle{}cor\textquotesingle{}}\NormalTok{, }\AttributeTok{method =} \StringTok{"spearman"}\NormalTok{))) }\SpecialCharTok{+} 
  \FunctionTok{theme\_bw}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
## exact p-value with ties

## Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
## exact p-value with ties

## Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
## exact p-value with ties
\end{verbatim}

\begin{figure}
\centering
\includegraphics{arXiv_bigsimr_files/figure-latex/ch050-simDataFig-1.pdf}
\caption{Simulated data for three selected high-expressing genes
generally replicates the estimated data structure. The data do not
exhibit outlying points, but do possess the desired Spearman
correlations, central tendencies, and discrete values.}
\end{figure}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{include\_graphics}\NormalTok{(}\StringTok{\textquotesingle{}../fig/ch050{-}figBRCA.png\textquotesingle{}}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\includegraphics[width=0.8\linewidth]{../fig/ch050-figBRCA} \caption{Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set. The dashed red lines indicate equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes).}\label{fig:ch050-figBRCA}
\end{figure}

\emph{Limitations, conclusions, and recommendations}

The results show overall aquedate simulation performance for our choice
of parameters settings. Our settings were motivated by modeling
high-expressing genes from the TCGA BRCA data set. In general, the
ability to match marginal and dependence parameters depends on the
particular joint probability model. We recommend to evaluate and tune
your simulation until you can be assured of the accuracy.

\hypertarget{discussion}{%
\section{Conclusion and discussion}\label{discussion}}

We developed a general-purpose high-dimensional multivariate simulation
algorithm and provide a user-friendly, high-performance \texttt{R}
package \texttt{bigsimr}, Julia package \texttt{Bigsimr} and Python
package \texttt{bigsimr}. The random vector generation method is
inspired by NORTA (Cario and Nelson 1997) and Gaussian copula-based
approaches Xiao (2017). The major contributions of this work are methods
and software for flexible, scalable simulation of HD multivariate
probability distributions with broad data analytic applications for
modern, big-data statistical computing. For example, one could simulate
high-resolution time series data, such as those consistent with an
auto-regressive moving average model exhibiting a specified Spearman
structure. Our methods could also be used to simulate sparsely
correlated data, as many HD methods assume, via specifying a
\emph{spiked correlation matrix}.

There are limitations to the methodology and implementation. We could
only investigate selected multivariate distributions in our Monte Carlo
studies. There may be instances where the methods do not perform well.
Along those lines, we expect that correlation values close to the
boundary of the feasible region could result in algorithm failure.
Another issue is that for discrete distributions, we use continuous
approximations when the support set is large. This could limit the
scalability/accuracy for particular discrete/mixed multivariate
distributions. Our method would also benefit computationally from faster
Kendall estimation and more finely tuned nearest PD calculation for
non-admissible Kendall matrices.

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-BF17}{}}%
Barbiero, Alessandro, and Pier Alda Ferrari. 2017. {``{An R package for
the simulation of correlated discrete variables}.''}
\emph{Communications in Statistics - Simulation and Computation} 46 (7):
5123--40. \url{https://doi.org/10.1080/03610918.2016.1146758}.

\leavevmode\vadjust pre{\hypertarget{ref-Cario1997}{}}%
Cario, Marne C., and Barry L. Nelson. 1997. {``{Modeling and generating
random vectors with arbitrary marginal distributions and correlation
matrix}.''} Technical Report.

\leavevmode\vadjust pre{\hypertarget{ref-Chambers1983}{}}%
Chambers, J. M., W. S. Cleveland, B. Kleiner, and P. A. Tukey. 1983.
\emph{{Graphical Methods for Data Analysis}}. Belmont, CA: Wadsworth
{\&} Brooks.

\leavevmode\vadjust pre{\hypertarget{ref-Chen2001}{}}%
Chen, Huifen. 2001. {``{Initialization for NORTA: Generation of random
vectors with specified marginals and correlations}.''} \emph{INFORMS
Journal on Computing} 13 (4): 312--31.

\leavevmode\vadjust pre{\hypertarget{ref-Chernick2008}{}}%
Chernick, Michael R. 2008. \emph{{Bootstrap Methods: A Guide for
Practitioners and Researchers}}. 2nd ed. Hoboken, NJ.

\leavevmode\vadjust pre{\hypertarget{ref-Conesa2016b}{}}%
Conesa, Ana, Pedro Madrigal, Sonia Tarazona, David Gomez-Cabrero,
Alejandra Cervera, Andrew McPherson, Michal Wojciech Szcześniak, et al.
2016. {``{A survey of best practices for RNA-seq data analysis}.''}
\url{https://doi.org/10.1186/s13059-016-0881-8}.

\leavevmode\vadjust pre{\hypertarget{ref-DH2011}{}}%
Demirtas, Hakan, and Donald Hedeker. 2011. {``{A practical way for
computing approximate lower and upper correlation bounds}.''}
\emph{American Statistician} 65 (2): 104--9.
\url{https://doi.org/10.1198/tast.2011.10090}.

\leavevmode\vadjust pre{\hypertarget{ref-BE07}{}}%
Efron, Bradley. 2007. {``{Correlation and large-scale simultaneous
significance testing}.''} \emph{Journal of the American Statistical
Association} 102 (477): 93--103.
\url{https://doi.org/10.1198/016214506000001211}.

\leavevmode\vadjust pre{\hypertarget{ref-Fasiolo2016}{}}%
Fasiolo, Matteo. 2016. \emph{{An introduction to mvnfast. R package
version 0.1.6.}} \url{https://cran.r-project.org/package=mvnfast}.

\leavevmode\vadjust pre{\hypertarget{ref-GH02}{}}%
Ghosh, Soumyadip, and Shane G Henderson. 2002. {``{Properties of the
Norta method in higher dimensions}.''} \emph{Proceedings of the 2002
Winter Simulation Conference}, 263--69.

\leavevmode\vadjust pre{\hypertarget{ref-higham2002computing}{}}%
Higham, Nicholas J. 2002. {``Computing the Nearest Correlation
Matrix---a Problem from Finance.''} \emph{IMA Journal of Numerical
Analysis} 22 (3): 329--43.

\leavevmode\vadjust pre{\hypertarget{ref-K58}{}}%
Kruskal, William H. 1958. {``{Ordinal measures of association}.''}
\emph{Journal of the American Statistical Association} 53 (284):
814--61. \url{https://doi.org/10.1080/01621459.1958.10501481}.

\leavevmode\vadjust pre{\hypertarget{ref-LH75}{}}%
Li, Shing Ted, and Joseph L. Hammond. 1975. {``{Generation of
pseudorandom numbers with specified univariate distributions and
correlation coefficients}.''} \emph{IEEE Transactions on Systems, Man
and Cybernetics} 5: 557--61.
\url{https://doi.org/10.1109/TSMC.1975.5408380}.

\leavevmode\vadjust pre{\hypertarget{ref-Li2019gpu}{}}%
Li, Xiang, A. Grant Schissler, Rui Wu, Lee Barford, Jr. Harris, Fredrick
C., and Frederick C. Harris. 2019. {``{A Graphical Processing Unit
accelerated NORmal to Anything algorithm for high dimensional
multivariate simulation}.''} \emph{Advances in Intelligent Systems and
Computing}, 339--45. \url{https://doi.org/10.1007/978-3-030-14070-0_46}.

\leavevmode\vadjust pre{\hypertarget{ref-MB13}{}}%
Madsen, L., and D. Birkes. 2013. {``{Simulating dependent discrete
data}.''} \emph{Journal of Statistical Computation and Simulation} 83
(4): 677--91. \url{https://doi.org/10.1080/00949655.2011.632774}.

\leavevmode\vadjust pre{\hypertarget{ref-MK01}{}}%
Mari, Dominique Drouet, and Samuel Kotz. 2001. \emph{{Correlation and
dependence}}. World Scientific.

\leavevmode\vadjust pre{\hypertarget{ref-muino2006gs}{}}%
Muino, JM, Eberhard O Voit, and Albert Sorribas. 2006.
{``GS-Distributions: A New Family of Distributions for Continuous
Unimodal Variables.''} \emph{Computational Statistics and Data Analysis}
50 (10): 2769--98.

\leavevmode\vadjust pre{\hypertarget{ref-Nelsen2007}{}}%
Nelsen, Roger B. 2007. \emph{{An Introduction to Copulas}}. 2nd ed. New
York: Springer Science {\&} Business Media.

\leavevmode\vadjust pre{\hypertarget{ref-Nik13a}{}}%
Nikoloulopoulos, Aristidis K. 2013. {``{Copula-based models for
multivariate discrete response data}.''} In \emph{Lecture Notes in
Statistics: Copulae in Mathematical and Quantitative Finance}, 213th
ed., 231--49. Heidelberg: Springer.

\leavevmode\vadjust pre{\hypertarget{ref-Park1996}{}}%
Park, Chul Gyu, Taesung Park, and Dong Wan Shin. 1996. {``{A simple
method for generating correlated binary variates}.''} \emph{American
Statistician} 50 (4): 306--10.
\url{https://doi.org/10.1080/00031305.1996.10473557}.

\leavevmode\vadjust pre{\hypertarget{ref-qi2006quadratically}{}}%
Qi, Houduo, and Defeng Sun. 2006. {``A Quadratically Convergent Newton
Method for Computing the Nearest Correlation Matrix.''} \emph{SIAM
Journal on Matrix Analysis and Applications} 28 (2): 360--85.

\leavevmode\vadjust pre{\hypertarget{ref-Schissler2019}{}}%
Schissler, Alfred Grant, Dillon Aberasturi, Colleen Kenost, and Yves A.
Lussier. 2019. {``{A Single-Subject Method to Detect Pathways Enriched
With Alternatively Spliced Genes}.''} \emph{Frontiers in Genetics} 10
(414). \url{https://doi.org/10.3389/fgene.2019.00414}.

\leavevmode\vadjust pre{\hypertarget{ref-Schissler2018}{}}%
Schissler, Alfred Grant, Walter W Piegorsch, and Yves A Lussier. 2018.
{``{Testing for differentially expressed genetic pathways with
single-subject N-of-1 data in the presence of inter-gene
correlation}.''} \emph{Statistical Methods in Medical Research} 27 (12):
3797--3813. \url{https://doi.org/10.1177/0962280217712271}.

\leavevmode\vadjust pre{\hypertarget{ref-Ubeda-Flores2017}{}}%
Úbeda-Flores, Manuel, and Juan Fernández-Sánchez. 2017. {``{Sklar's
theorem: The cornerstone of the Theory of Copulas}.''} In \emph{Copulas
and Dependence Models with Applications}.
\url{https://doi.org/10.1007/978-3-319-64221-5_15}.

\leavevmode\vadjust pre{\hypertarget{ref-Wang2009b}{}}%
Wang, Zhong, Mark Gerstein, and Michael Snyder. 2009. {``{RNA-Seq: A
revolutionary tool for transcriptomics}.''}
\url{https://doi.org/10.1038/nrg2484}.

\leavevmode\vadjust pre{\hypertarget{ref-Xia17}{}}%
Xiao, Qing. 2017. {``{Generating correlated random vector involving
discrete variables}.''} \emph{Communications in Statistics - Theory and
Methods} 46 (4): 1594--1605.
\url{https://doi.org/10.1080/03610926.2015.1024860}.

\leavevmode\vadjust pre{\hypertarget{ref-XZ19}{}}%
Xiao, Qing, and Shaowu Zhou. 2019. {``{Matching a correlation
coefficient by a Gaussian copula}.''} \emph{Communications in Statistics
- Theory and Methods} 48 (7): 1728--47.
\url{https://doi.org/10.1080/03610926.2018.1439962}.

\leavevmode\vadjust pre{\hypertarget{ref-Yan2007}{}}%
Yan, Jun. 2007. {``{Enjoy the joy of copulas: with a package copula}.''}
\emph{Journal of Statistical Software} 21 (4): 1--21.
\url{http://www.jstatsoft.org/v21/i04}.

\end{CSLReferences}

\bibliographystyle{unsrt}
\bibliography{references.bib}

\end{document}

### 2021-10-19 arXiv_bigsimr/arXiv_bigsimr.Rmd
arXiv echoing and warnings

@@
 authors:
-  - name: David S. Hippocampus
-    thanks: Use footnote for providing further information about author (webpage, alternative address)---*not* for acknowledging funding agencies. Optional.
-    department: Department of Computer Science
-    affiliation: Cranberry-Lemon University
-    location: Pittsburgh, PA 15213
-    email: hippo@cs.cranberry-lemon.edu
-  - name: Elias D. Striatum
-    department: Department of Electrical Engineering
-    affiliation: Mount-Sheikh University
-    location: Santa Narimana, Levand
-    email: stariate@ee.mount-sheikh.edu
+  - name: A. Grant Schissler
+    department: Department of Mathematics \& Statistics
+    affiliation: University of Nevada, Reno
+    location: Reno, NV 89557
+    email: grant.schissler@gmail.com
+  - name: Edward J. Bedrick
+    affiliation: University of Arizona
+  - name: Alexander D. Knudson
+    affiliation: University of Nevada, Reno \AND
+  - name: Tomasz J. Kozubowski
+    affiliation: University of Nevada, Reno
+  - name: Tin Nguyen
+    affiliation: University of Nevada, Reno
+  - name: Anna K. Panorska
+    affiliation: University of Nevada, Reno \AND
+  - name: Juli Petereit
+    affiliation: University of Nevada, Reno
+  - name: Walter W. Piegorsch
+    affiliation: University of Arizona
+  - name: Duc Tran
+    affiliation: University of Nevada, Reno
 abstract: |
@@
 ---
+
+\clearpage
 
@@
 
-```{r ch010-realDataTab, eval=T}
+```{r ch010-realDataTab, eval=T, echo=F}
 set.seed(2020-02-25)
@@
 
-```{r ch010-realDataFig, cache=FALSE, fig.width=8, fig.cap="Pairwise scatterplots for three example genes (Table 1), with estimated Spearman's correlations and marginal density estimates displayed.
+```{r ch010-realDataFig, cache=FALSE, fig.width=6, fig.cap="Pairwise scatterplots for three example genes (Table 1), with estimated Spearman's correlations and marginal density estimates displayed.
 The data possess outliers, heavy-right tails, are discrete, and have non-trivial intergene correlations.
-Modeling these data motivate our simulation methodology.", eval=T}
+Modeling these data motivate our simulation methodology.", eval=T, warning=F, echo=F}
 ggpairs(
@@
 
-```{r ch030-aq-joint-dist, cache=FALSE, fig.width= 8, fig.cap="Bivariate scatterplot of Ozone versus Temp with estimated marginal densities.
-The Ozone data are modeled marginally as log-normal and the Temperature data as normal.", eval=T}
+```{r ch030-aq-joint-dist, cache=FALSE, fig.width=6, fig.cap="Bivariate scatterplot of Ozone versus Temp with estimated marginal densities.
+The Ozone data are modeled marginally as log-normal and the Temperature data as normal.", eval=T, echo=F}
 p0 <- ggplot(df, aes(Temp, Ozone)) +
@@
 ```{r ch030-plot-sim, fig.cap="Contour plot and marginal densities for the simulated bivariate distribution of Air Quality Temperatures and Ozone levels.
-The simulated points mimic the observed data with respect to both the marginal characteristics and bivariate association.", eval=T}
+The simulated points mimic the observed data with respect to both the marginal characteristics and bivariate association.", eval=T, echo=F, warning=F}
 p1 <- df_sim %>%
@@
 
-```{r ch040-biNormPlot, eval=T, fig.height=5, fig.width=8, fig.cap="`bigsimr` recovers the Pearson specified correlations for MVN."}
+```{r ch040-biNormPlot, eval=T, fig.height=5, fig.width=6, fig.cap="`bigsimr` recovers the Pearson specified correlations for MVN.", echo=F}
 bivariate_normal_sims %>%
@@
 
-```{r ch040-biGammaPlot, eval=T, fig.height=5, fig.width=8, fig.cap="`bigsimr` recovers the Pearson specified correlations for Bivariate Gamma."}
+```{r ch040-biGammaPlot, eval=T, fig.height=5, fig.width=6, fig.cap="`bigsimr` recovers the Pearson specified correlations for Bivariate Gamma.", echo=F}
 bivariate_gamma_sims %>%
@@
 
-```{r ch040-biNegBinPlot, eval=T, fig.width=8, fig.height=5, fig.cap="`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
+```{r ch040-biNegBinPlot, eval=T, fig.width=6, fig.height=5, fig.cap="`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations.", echo=F}
 bivariate_nbinom_sims %>%
@@
 
-```{r ch040-combineBiSims, cache=FALSE, eval=T}
+```{r ch040-combineBiSims, cache=FALSE, eval=T, echo=F}
 allDat <- bind_rows(
@@
 
-```{r ch040-BiError, eval=T}
+```{r ch040-BiError, eval=T, echo=F, message=F}
 tabMAE <- allDat %>%
@@
 
-```{r ch040-bPlot, cache=FALSE, fig.asp=1.30, fig.weight=4, fig.cap="Bivariate simulations match target correlations across the entire range of feasible correlations.
+```{r ch040-bPlot, cache=FALSE, echo=F, fig.asp=1.30, fig.weight=4, fig.cap="Bivariate simulations match target correlations across the entire range of feasible correlations.
 The horizontal axis plots the specified target correlations for each bivariate margin.
@@
 As the number of simulated vectors $B$ increases from left to right, the variation in estimated correlations (vertical axis) decreases.
-The dashed line indicates equality between the specified and estimated correlations.", eval=T}
+The dashed line indicates equality between the specified and estimated correlations.", eval=T, echo=F}
 # https://www.datanovia.com/en/blog/how-to-change-ggplot-facet-labels/
@@
 
-```{r ch040-moderateDtab, eval=T}
+```{r ch040-moderateDtab, eval=T, echo=F, message=F}
 dat_long <- benchmark_dependences %>%
@@
 
-```{r ch040-largeDfig, fig.cap="Computation times for HD multivariate gamma simulation.", eval=T}
+```{r ch040-largeDfig, fig.cap="Computation times for HD multivariate gamma simulation.", eval=T, echo=F}
 # Set colors
@@
 ```{r ch050-simDataFig, cache=FALSE, fig.cap="Simulated data for three selected high-expressing genes generally replicates the estimated data structure.
-The data do not exhibit outlying points, but do possess the desired Spearman correlations, central tendencies, and discrete values.", eval=T}
+The data do not exhibit outlying points, but do possess the desired Spearman correlations, central tendencies, and discrete values.", eval=T, warning=F, echo=F}
 set.seed(2020-02-25)
@@
 ```{r ch050-figBRCA, out.width='80%', fig.cap="Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set.
-The dashed red lines indicate equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes).", eval=T}
+The dashed red lines indicate equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes).", eval=T, echo=F}
 include_graphics('../fig/ch050-figBRCA.png')

### 2021-10-19 arXiv_bigsimr/arXiv_bigsimr.tex
arXiv echoing and warnings

@@
 \author{
-    David S. Hippocampus
-    \thanks{Use footnote for providing further information about author
-(webpage, alternative address)---\emph{not} for acknowledging funding
-agencies. Optional.}
+    A. Grant Schissler
    \\
-    Department of Computer Science \\
-    Cranberry-Lemon University \\
-  Pittsburgh, PA 15213 \\
-  \texttt{\href{mailto:hippo@cs.cranberry-lemon.edu}{\nolinkurl{hippo@cs.cranberry-lemon.edu}}} \\
+    Department of Mathematics \& Statistics \\
+    University of Nevada, Reno \\
+  Reno, NV 89557 \\
+  \texttt{\href{mailto:grant.schissler@gmail.com}{\nolinkurl{grant.schissler@gmail.com}}} \\
    \And
-    Elias D. Striatum
+    Edward J. Bedrick
    \\
-    Department of Electrical Engineering \\
-    Mount-Sheikh University \\
-  Santa Narimana, Levand \\
-  \texttt{\href{mailto:stariate@ee.mount-sheikh.edu}{\nolinkurl{stariate@ee.mount-sheikh.edu}}} \\
+    University of Arizona \\
+   \\
+  \texttt{} \\
+   \And
+    Alexander D. Knudson
+   \\
+    University of Nevada, Reno \AND \\
+   \\
+  \texttt{} \\
+   \And
+    Tomasz J. Kozubowski
+   \\
+    University of Nevada, Reno \\
+   \\
+  \texttt{} \\
+   \And
+    Tin Nguyen
+   \\
+    University of Nevada, Reno \\
+   \\
+  \texttt{} \\
+   \And
+    Anna K. Panorska
+   \\
+    University of Nevada, Reno \AND \\
+   \\
+  \texttt{} \\
+   \And
+    Juli Petereit
+   \\
+    University of Nevada, Reno \\
+   \\
+  \texttt{} \\
+   \And
+    Walter W. Piegorsch
+   \\
+    University of Arizona \\
+   \\
+  \texttt{} \\
+   \And
+    Duc Tran
+   \\
+    University of Nevada, Reno \\
+   \\
+  \texttt{} \\
   }
@@
 
+\clearpage
+
 \hypertarget{introduction}{%
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{set.seed}\NormalTok{(}\DecValTok{2020{-}02{-}25}\NormalTok{)}
-\NormalTok{num\_genes    }\OtherTok{\textless{}{-}} \DecValTok{3}
-\NormalTok{num\_patients }\OtherTok{\textless{}{-}} \DecValTok{5}
-\NormalTok{gene\_sample  }\OtherTok{\textless{}{-}} \FunctionTok{sample}\NormalTok{(example\_genes[}\DecValTok{1}\SpecialCharTok{:}\DecValTok{1000}\NormalTok{], num\_genes)}
-
-\NormalTok{cap }\OtherTok{\textless{}{-}} \FunctionTok{paste0}\NormalTok{(}\StringTok{"mRNA expression for three selected high{-}expressing genes, "}\NormalTok{,}
-              \FunctionTok{paste}\NormalTok{(gene\_sample, }\AttributeTok{collapse =} \StringTok{", "}\NormalTok{), }
-              \StringTok{", for the first five patients in the TCGA BRCA data set."}\NormalTok{)}
-
-\NormalTok{small\_brca }\OtherTok{\textless{}{-}}\NormalTok{ example\_brca }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{select}\NormalTok{(}\FunctionTok{all\_of}\NormalTok{(gene\_sample)) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{head}\NormalTok{(}\AttributeTok{n =}\NormalTok{ num\_patients) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{as\_tibble}\NormalTok{(}\AttributeTok{rownames =} \StringTok{"Patient ID"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{mutate}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Patient ID}\StringTok{\textasciigrave{}} \OtherTok{=} \FunctionTok{str\_sub}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Patient ID}\StringTok{\textasciigrave{}}\NormalTok{, }\AttributeTok{end =} \DecValTok{12}\NormalTok{))}
-
-\NormalTok{small\_brca }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{kable}\NormalTok{(}\AttributeTok{format =} \StringTok{"latex"}\NormalTok{, }\AttributeTok{booktabs =} \ConstantTok{TRUE}\NormalTok{, }\AttributeTok{caption =}\NormalTok{ cap)}
-\end{Highlighting}
-\end{Shaded}
-
 \begin{table}
@@
 \end{table}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{ggpairs}\NormalTok{(}
-  \AttributeTok{data =}\NormalTok{ example\_brca[, gene\_sample],}
-  \AttributeTok{upper =} \FunctionTok{list}\NormalTok{(}\AttributeTok{continuous =} \FunctionTok{wrap}\NormalTok{(}\StringTok{\textquotesingle{}cor\textquotesingle{}}\NormalTok{, }\AttributeTok{method =} \StringTok{"spearman"}\NormalTok{))}
-\NormalTok{) }\SpecialCharTok{+} 
-  \FunctionTok{theme\_bw}\NormalTok{() }\SpecialCharTok{+}
-  \FunctionTok{theme}\NormalTok{(}\AttributeTok{axis.text.x =} \FunctionTok{element\_text}\NormalTok{(}\AttributeTok{angle =} \SpecialCharTok{{-}}\DecValTok{90}\NormalTok{, }\AttributeTok{vjust =} \FloatTok{0.5}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
-## exact p-value with ties
-
-## Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
-## exact p-value with ties
-
-## Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
-## exact p-value with ties
-\end{verbatim}
 
@@
 
-\begin{verbatim}
-## Julia version 1.5.3 at location /Applications/Julia-1.5.app/Contents/Resources/julia/bin will be used.
-\end{verbatim}
-
-\begin{verbatim}
-## Loading setup script for JuliaCall...
-\end{verbatim}
-
-\begin{verbatim}
-## Finish loading setup script for JuliaCall.
-\end{verbatim}
-
 This section describes a bivariate random vector simulation workflow via
@@
 strong positive skew.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{p0 }\OtherTok{\textless{}{-}} \FunctionTok{ggplot}\NormalTok{(df, }\FunctionTok{aes}\NormalTok{(Temp, Ozone)) }\SpecialCharTok{+}
-  \FunctionTok{geom\_point}\NormalTok{(}\AttributeTok{size =} \DecValTok{1}\NormalTok{) }\SpecialCharTok{+}
-  \FunctionTok{theme}\NormalTok{(}\AttributeTok{legend.position =} \StringTok{"none"}\NormalTok{) }\SpecialCharTok{+} 
-  \FunctionTok{labs}\NormalTok{(}\AttributeTok{x =} \StringTok{"Temperature"}\NormalTok{)}
-
-\NormalTok{pTemp }\OtherTok{\textless{}{-}} \FunctionTok{ggplot}\NormalTok{(df, }\FunctionTok{aes}\NormalTok{(Temp)) }\SpecialCharTok{+} 
-  \FunctionTok{geom\_density}\NormalTok{() }\SpecialCharTok{+}
-  \FunctionTok{theme}\NormalTok{(}\AttributeTok{axis.title.x =} \FunctionTok{element\_blank}\NormalTok{(),}
-        \AttributeTok{axis.title.y =} \FunctionTok{element\_blank}\NormalTok{(),}
-        \AttributeTok{axis.text =} \FunctionTok{element\_blank}\NormalTok{(),}
-        \AttributeTok{axis.ticks =} \FunctionTok{element\_blank}\NormalTok{())}
-
-\NormalTok{pOzone }\OtherTok{\textless{}{-}} \FunctionTok{ggplot}\NormalTok{(df, }\FunctionTok{aes}\NormalTok{(Ozone)) }\SpecialCharTok{+} 
-    \FunctionTok{geom\_density}\NormalTok{() }\SpecialCharTok{+}
-  \FunctionTok{theme}\NormalTok{(}\AttributeTok{axis.title.y =} \FunctionTok{element\_blank}\NormalTok{(),}
-        \AttributeTok{axis.title.x =} \FunctionTok{element\_blank}\NormalTok{(),}
-        \AttributeTok{axis.ticks =} \FunctionTok{element\_blank}\NormalTok{(),}
-        \AttributeTok{axis.text =} \FunctionTok{element\_blank}\NormalTok{()) }\SpecialCharTok{+}
-  \FunctionTok{coord\_flip}\NormalTok{()}
-
-\NormalTok{pTemp }\SpecialCharTok{+} \FunctionTok{plot\_spacer}\NormalTok{() }\SpecialCharTok{+}\NormalTok{ p0 }\SpecialCharTok{+}\NormalTok{ pOzone }\SpecialCharTok{+} 
-  \FunctionTok{plot\_layout}\NormalTok{(}\AttributeTok{widths =} \FunctionTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{,}\DecValTok{1}\NormalTok{), }\AttributeTok{heights =} \FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{3}\NormalTok{)) }
-\end{Highlighting}
-\end{Shaded}
 
@@
 ## Julia Object of type NamedTuple{(:lower, :upper),Tuple{Float64,Float64}}.
-## (lower = -0.8806384281235534, upper = 0.880733034303322)
+## (lower = -0.8816136274594583, upper = 0.8815214379668451)
 \end{verbatim}
@@
 \end{Shaded}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{p1 }\OtherTok{\textless{}{-}}\NormalTok{ df\_sim }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(Temp, Ozone)) }\SpecialCharTok{+}
-  \FunctionTok{geom\_density\_2d\_filled}\NormalTok{() }\SpecialCharTok{+}
-  \FunctionTok{theme}\NormalTok{(}\AttributeTok{legend.position =} \StringTok{"none"}\NormalTok{) }\SpecialCharTok{+}
-  \FunctionTok{labs}\NormalTok{(}\AttributeTok{x =} \StringTok{"Simulated Temperature"}\NormalTok{, }\AttributeTok{y =} \StringTok{"Simulated Ozone"}\NormalTok{) }\SpecialCharTok{+}
-  \FunctionTok{scale\_y\_continuous}\NormalTok{(}\AttributeTok{limits =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\FunctionTok{max}\NormalTok{(df}\SpecialCharTok{$}\NormalTok{Ozone))) }\SpecialCharTok{+}
-  \FunctionTok{scale\_x\_continuous}\NormalTok{(}\AttributeTok{limits =} \FunctionTok{range}\NormalTok{(df}\SpecialCharTok{$}\NormalTok{Temp))}
-
-\NormalTok{p1Temp }\OtherTok{\textless{}{-}} \FunctionTok{ggplot}\NormalTok{(df\_sim, }\FunctionTok{aes}\NormalTok{(Temp)) }\SpecialCharTok{+} 
-  \FunctionTok{geom\_density}\NormalTok{(}\AttributeTok{alpha =} \FloatTok{0.5}\NormalTok{, }\AttributeTok{fill =} \StringTok{"lightseagreen"}\NormalTok{) }\SpecialCharTok{+}
-  \FunctionTok{theme}\NormalTok{(}\AttributeTok{axis.title.x =} \FunctionTok{element\_blank}\NormalTok{(),}
-        \AttributeTok{axis.title.y =} \FunctionTok{element\_blank}\NormalTok{(),}
-        \AttributeTok{axis.text =} \FunctionTok{element\_blank}\NormalTok{(),}
-        \AttributeTok{axis.ticks =} \FunctionTok{element\_blank}\NormalTok{())}
-
-\NormalTok{p1Ozone }\OtherTok{\textless{}{-}} \FunctionTok{ggplot}\NormalTok{(df\_sim, }\FunctionTok{aes}\NormalTok{(Ozone)) }\SpecialCharTok{+} 
-  \FunctionTok{geom\_density}\NormalTok{(}\AttributeTok{alpha =} \FloatTok{0.5}\NormalTok{, }\AttributeTok{fill =} \StringTok{"lightseagreen"}\NormalTok{) }\SpecialCharTok{+} 
-  \FunctionTok{theme}\NormalTok{(}\AttributeTok{axis.title.y =} \FunctionTok{element\_blank}\NormalTok{(),}
-        \AttributeTok{axis.title.x =} \FunctionTok{element\_blank}\NormalTok{(),}
-        \AttributeTok{axis.ticks =} \FunctionTok{element\_blank}\NormalTok{(),}
-        \AttributeTok{axis.text =} \FunctionTok{element\_blank}\NormalTok{()) }\SpecialCharTok{+}
-  \FunctionTok{coord\_flip}\NormalTok{() }
-
-\NormalTok{p1Temp }\SpecialCharTok{+} \FunctionTok{plot\_spacer}\NormalTok{() }\SpecialCharTok{+}\NormalTok{ p1 }\SpecialCharTok{+}\NormalTok{ p1Ozone }\SpecialCharTok{+} 
-  \FunctionTok{plot\_layout}\NormalTok{(}\AttributeTok{widths =} \FunctionTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{,}\DecValTok{1}\NormalTok{), }\AttributeTok{heights =} \FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{3}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Warning: Removed 388 rows containing non-finite values (stat_density2d_filled).
-\end{verbatim}
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{bivariate\_normal\_sims }\SpecialCharTok{\%\textgreater{}\%}
-    \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(rho, rho\_hat, }\AttributeTok{color =}\NormalTok{ type)) }\SpecialCharTok{+}
-    \FunctionTok{geom\_point}\NormalTok{() }\SpecialCharTok{+}
-    \FunctionTok{geom\_abline}\NormalTok{(}\AttributeTok{slope =} \DecValTok{1}\NormalTok{) }\SpecialCharTok{+}
-    \FunctionTok{facet\_wrap}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{ type }\SpecialCharTok{+}\NormalTok{ N) }\SpecialCharTok{+} 
-    \FunctionTok{theme\_bw}\NormalTok{()}
-\end{Highlighting}
-\end{Shaded}
-
 \begin{figure}
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{bivariate\_gamma\_sims }\SpecialCharTok{\%\textgreater{}\%}
-    \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(rho, rho\_hat, }\AttributeTok{color =}\NormalTok{ type)) }\SpecialCharTok{+}
-    \FunctionTok{geom\_point}\NormalTok{() }\SpecialCharTok{+}
-    \FunctionTok{geom\_abline}\NormalTok{(}\AttributeTok{slope =} \DecValTok{1}\NormalTok{) }\SpecialCharTok{+}
-    \FunctionTok{facet\_wrap}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{ type }\SpecialCharTok{+}\NormalTok{ N) }\SpecialCharTok{+} 
-    \FunctionTok{theme\_bw}\NormalTok{()}
-\end{Highlighting}
-\end{Shaded}
-
 \begin{figure}
@@
 \end{figure}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{bivariate\_nbinom\_sims }\SpecialCharTok{\%\textgreater{}\%}
-    \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(rho, rho\_hat, }\AttributeTok{color =}\NormalTok{ type)) }\SpecialCharTok{+}
-    \FunctionTok{geom\_point}\NormalTok{() }\SpecialCharTok{+}
-    \FunctionTok{geom\_abline}\NormalTok{(}\AttributeTok{slope =} \DecValTok{1}\NormalTok{) }\SpecialCharTok{+}
-    \FunctionTok{facet\_wrap}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{ type }\SpecialCharTok{+}\NormalTok{ N) }\SpecialCharTok{+} 
-    \FunctionTok{theme\_bw}\NormalTok{()}
-\end{Highlighting}
-\end{Shaded}
 
@@
 \end{figure}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{allDat }\OtherTok{\textless{}{-}} \FunctionTok{bind\_rows}\NormalTok{(}
-    \FunctionTok{select}\NormalTok{(bivariate\_normal\_sims, margins, type, N, rho, rho\_hat),}
-    \FunctionTok{select}\NormalTok{(bivariate\_gamma\_sims, margins, type, N, rho, rho\_hat),}
-    \FunctionTok{select}\NormalTok{(bivariate\_nbinom\_sims, margins, type, N, rho, rho\_hat)}
-\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
-    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{margins =} \FunctionTok{factor}\NormalTok{(margins, }\AttributeTok{levels=}\FunctionTok{c}\NormalTok{(}\StringTok{"norm"}\NormalTok{, }\StringTok{"gamma"}\NormalTok{, }\StringTok{"nbinom"}\NormalTok{)),}
-           \AttributeTok{type =} \FunctionTok{factor}\NormalTok{(type, }\AttributeTok{levels=}\FunctionTok{c}\NormalTok{(}\StringTok{"Pearson"}\NormalTok{, }\StringTok{"Spearman"}\NormalTok{, }\StringTok{"Kendall"}\NormalTok{)),}
-           \AttributeTok{N =} \FunctionTok{factor}\NormalTok{(N))}
-\end{Highlighting}
-\end{Shaded}
 
@@
 increase in error for the negative binomial case.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{tabMAE }\OtherTok{\textless{}{-}}\NormalTok{ allDat }\SpecialCharTok{\%\textgreater{}\%}
-    \FunctionTok{group\_by}\NormalTok{(N, type, margins) }\SpecialCharTok{\%\textgreater{}\%}
-    \FunctionTok{summarize}\NormalTok{(}\AttributeTok{MAE =} \FunctionTok{mean}\NormalTok{(}\FunctionTok{abs}\NormalTok{(rho }\SpecialCharTok{{-}}\NormalTok{ rho\_hat))) }\SpecialCharTok{\%\textgreater{}\%}
-    \FunctionTok{ungroup}\NormalTok{()}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## `summarise()` regrouping output by 'N', 'type' (override with `.groups` argument)
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{kable}\NormalTok{(tabMAE, }\AttributeTok{booktabs =} \ConstantTok{TRUE}\NormalTok{, }\AttributeTok{format =} \StringTok{"latex"}\NormalTok{,}
-      \AttributeTok{linesep =} \FunctionTok{c}\NormalTok{(}\StringTok{""}\NormalTok{, }\StringTok{""}\NormalTok{, }\StringTok{"}\SpecialCharTok{\textbackslash{}\textbackslash{}}\StringTok{addlinespace"}\NormalTok{),}
-      \AttributeTok{col.names =} \FunctionTok{c}\NormalTok{(}\StringTok{"No. of random vectors"}\NormalTok{,}
-                    \StringTok{"Correlation type"}\NormalTok{,}
-                    \StringTok{"Distribution"}\NormalTok{,}
-                    \StringTok{"Mean abs. error"}\NormalTok{),}
-      \AttributeTok{caption =} \StringTok{"Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal."}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# https://www.datanovia.com/en/blog/how{-}to{-}change{-}ggplot{-}facet{-}labels/}
-\CommentTok{\# New facet label names}
-\NormalTok{repsLabs }\OtherTok{\textless{}{-}} \FunctionTok{paste0}\NormalTok{(}\StringTok{"B="}\NormalTok{, }\FunctionTok{c}\NormalTok{(}\StringTok{"1,000"}\NormalTok{, }\StringTok{"10,000"}\NormalTok{, }\StringTok{"100,000"}\NormalTok{) )}
-\FunctionTok{names}\NormalTok{(repsLabs) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\DecValTok{1000}\NormalTok{, }\DecValTok{10000}\NormalTok{, }\DecValTok{100000}\NormalTok{)}
-\NormalTok{typeLabs }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"Pearson"}\NormalTok{, }\StringTok{"Spearman"}\NormalTok{, }\StringTok{"Kendall"}\NormalTok{)}
-\FunctionTok{names}\NormalTok{(typeLabs) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"Pearson"}\NormalTok{, }\StringTok{"Spearman"}\NormalTok{, }\StringTok{"Kendall"}\NormalTok{)}
-
-\CommentTok{\# Set colors}
-\DocumentationTok{\#\# RColorBrewer::display.brewer.all()}
-\NormalTok{numColors }\OtherTok{\textless{}{-}} \DecValTok{4}
-\NormalTok{numGroups }\OtherTok{\textless{}{-}} \FunctionTok{length}\NormalTok{(}\FunctionTok{levels}\NormalTok{(allDat}\SpecialCharTok{$}\NormalTok{margins))}
-\NormalTok{myColors }\OtherTok{\textless{}{-}} \FunctionTok{rev}\NormalTok{(RColorBrewer}\SpecialCharTok{::}\FunctionTok{brewer.pal}\NormalTok{(}\AttributeTok{n =}\NormalTok{ numColors, }\AttributeTok{name =} \StringTok{"Greys"}\NormalTok{)[ ((numColors }\SpecialCharTok{{-}}\NormalTok{ numGroups) }\SpecialCharTok{+} \DecValTok{1}\NormalTok{)}\SpecialCharTok{:}\NormalTok{numColors])}
-
-\NormalTok{allDat }\SpecialCharTok{\%\textgreater{}\%}
-    \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ rho, }\AttributeTok{y =}\NormalTok{ rho\_hat, }\AttributeTok{color =}\NormalTok{ margins)) }\SpecialCharTok{+}
-    \FunctionTok{geom\_point}\NormalTok{(}\AttributeTok{size =} \DecValTok{2}\NormalTok{) }\SpecialCharTok{+}
-    \FunctionTok{scale\_color\_manual}\NormalTok{(}\AttributeTok{values =}\NormalTok{ myColors ) }\SpecialCharTok{+}
-    \FunctionTok{geom\_abline}\NormalTok{(}\AttributeTok{slope =} \DecValTok{1}\NormalTok{, }\AttributeTok{linetype =} \StringTok{"dashed"}\NormalTok{) }\SpecialCharTok{+}
-    \FunctionTok{labs}\NormalTok{(}\AttributeTok{x =} \StringTok{"Specified Correlation"}\NormalTok{, }\AttributeTok{y =} \StringTok{"Estimated Correlation"}\NormalTok{) }\SpecialCharTok{+}
-    \FunctionTok{facet\_wrap}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{ type }\SpecialCharTok{+}\NormalTok{ N, }\AttributeTok{labeller =} \FunctionTok{labeller}\NormalTok{(}\AttributeTok{N =}\NormalTok{ repsLabs, }\AttributeTok{type =}\NormalTok{ typeLabs)) }\SpecialCharTok{+}
-    \FunctionTok{theme\_bw}\NormalTok{() }\SpecialCharTok{+}
-    \FunctionTok{theme}\NormalTok{(}\AttributeTok{legend.position =} \StringTok{"bottom"}\NormalTok{, }\AttributeTok{legend.direction =} \StringTok{"horizontal"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
 \begin{figure}
@@
 in under 2 seconds.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{dat\_long }\OtherTok{\textless{}{-}}\NormalTok{ benchmark\_dependences }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{select}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{total\_time, }\SpecialCharTok{{-}}\NormalTok{n\_sim, }\SpecialCharTok{{-}}\NormalTok{needed\_near\_pd) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{pivot\_longer}\NormalTok{(}\AttributeTok{cols =} \FunctionTok{c}\NormalTok{(corr\_time}\SpecialCharTok{:}\NormalTok{sim\_time), }\AttributeTok{names\_to =} \StringTok{"Step"}\NormalTok{, }\AttributeTok{values\_to =} \StringTok{"Time"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{dim =} \FunctionTok{factor}\NormalTok{(dim),}
-         \AttributeTok{Step =} \FunctionTok{factor}\NormalTok{(Step,}
-                       \AttributeTok{levels =} \FunctionTok{c}\NormalTok{(}\StringTok{"corr\_time"}\NormalTok{, }\StringTok{"adj\_time"}\NormalTok{, }\StringTok{"admiss\_time"}\NormalTok{, }\StringTok{"sim\_time"}\NormalTok{),}
-                       \AttributeTok{labels =} \FunctionTok{c}\NormalTok{(}\StringTok{"Compute Correlation"}\NormalTok{,}
-                                  \StringTok{"Adjust Correlation"}\NormalTok{,}
-                                  \StringTok{"Check Admissibility"}\NormalTok{,}
-                                  \StringTok{"Simulate Data"}\NormalTok{))) }\SpecialCharTok{\%\textgreater{}\%}
-    \FunctionTok{rename}\NormalTok{(}\AttributeTok{Correlation =}\NormalTok{ corr\_type, }\AttributeTok{Dimensions =}\NormalTok{ dim)}
-
-\NormalTok{tabTimes  }\OtherTok{\textless{}{-}}\NormalTok{ dat\_long }\SpecialCharTok{\%\textgreater{}\%}
-    \FunctionTok{filter}\NormalTok{(Dimensions }\SpecialCharTok{\%in\%} \FunctionTok{c}\NormalTok{(}\DecValTok{100}\NormalTok{, }\DecValTok{250}\NormalTok{, }\DecValTok{500}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
-    \FunctionTok{group\_by}\NormalTok{(Dimensions, Correlation) }\SpecialCharTok{\%\textgreater{}\%}
-    \FunctionTok{summarize}\NormalTok{(}\StringTok{\textquotesingle{}Total Time(Seconds)\textquotesingle{}} \OtherTok{=} \FunctionTok{sum}\NormalTok{(Time)) }\SpecialCharTok{\%\textgreater{}\%}
-    \FunctionTok{ungroup}\NormalTok{()}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## `summarise()` regrouping output by 'Dimensions' (override with `.groups` argument)
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{tabTimes }\SpecialCharTok{\%\textgreater{}\%}
-    \FunctionTok{kable}\NormalTok{(}\AttributeTok{booktabs =} \ConstantTok{TRUE}\NormalTok{,}
-          \AttributeTok{linesep =} \FunctionTok{c}\NormalTok{(}\StringTok{""}\NormalTok{, }\StringTok{""}\NormalTok{, }\StringTok{"}\SpecialCharTok{\textbackslash{}\textbackslash{}}\StringTok{addlinespace"}\NormalTok{), }\AttributeTok{format=}\StringTok{"latex"}\NormalTok{,}
-          \AttributeTok{col.names =} \FunctionTok{c}\NormalTok{(}\StringTok{"Dimension"}\NormalTok{,}
-                        \StringTok{"Correlation type"}\NormalTok{,}
-                        \StringTok{"Total Time (Seconds)"}\NormalTok{),}
-          \AttributeTok{caption =} \StringTok{\textquotesingle{}Total time to produce 1,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.\textquotesingle{}}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# Set colors}
-\NormalTok{numColors }\OtherTok{\textless{}{-}} \DecValTok{5}
-\NormalTok{numGroups }\OtherTok{\textless{}{-}} \DecValTok{4}
-\NormalTok{myColors }\OtherTok{\textless{}{-}} \FunctionTok{rev}\NormalTok{(}\FunctionTok{brewer.pal}\NormalTok{(}\AttributeTok{n=}\NormalTok{numColors, }\AttributeTok{name=}\StringTok{"Greys"}\NormalTok{)[((numColors}\SpecialCharTok{{-}}\NormalTok{numGroups)}\SpecialCharTok{+}\DecValTok{1}\NormalTok{)}\SpecialCharTok{:}\NormalTok{numColors])}
-
-\NormalTok{dat\_long }\SpecialCharTok{\%\textgreater{}\%}
-    \FunctionTok{filter}\NormalTok{(Dimensions }\SpecialCharTok{\%in\%} \FunctionTok{c}\NormalTok{(}\DecValTok{1000}\NormalTok{, }\DecValTok{2500}\NormalTok{, }\DecValTok{5000}\NormalTok{, }\DecValTok{10000}\NormalTok{, }\DecValTok{20000}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
-    \FunctionTok{mutate}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Time (minutes)}\StringTok{\textasciigrave{}} \OtherTok{=}\NormalTok{ Time }\SpecialCharTok{/} \DecValTok{60}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
-    \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(Correlation, }\StringTok{\textasciigrave{}}\AttributeTok{Time (minutes)}\StringTok{\textasciigrave{}}\NormalTok{, }\AttributeTok{fill=}\NormalTok{Step)) }\SpecialCharTok{+}
-    \FunctionTok{geom\_bar}\NormalTok{(}\AttributeTok{position =} \StringTok{"stack"}\NormalTok{, }\AttributeTok{stat =} \StringTok{"identity"}\NormalTok{) }\SpecialCharTok{+}
-    \FunctionTok{scale\_fill\_manual}\NormalTok{(}\AttributeTok{values =}\NormalTok{ myColors ) }\SpecialCharTok{+}
-    \FunctionTok{scale\_y\_continuous}\NormalTok{(}\AttributeTok{breaks =} \FunctionTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{80}\NormalTok{, }\DecValTok{10}\NormalTok{),}
-                     \AttributeTok{minor\_breaks =} \ConstantTok{NULL}\NormalTok{) }\SpecialCharTok{+}
-    \FunctionTok{facet\_grid}\NormalTok{(. }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Dimensions) }\SpecialCharTok{+}
-    \FunctionTok{theme}\NormalTok{(}\AttributeTok{axis.text.x =} \FunctionTok{element\_text}\NormalTok{(}\AttributeTok{angle =} \SpecialCharTok{{-}}\DecValTok{90}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
 \begin{figure}
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{set.seed}\NormalTok{(}\DecValTok{2020{-}02{-}25}\NormalTok{)}
-\NormalTok{num\_genes }\OtherTok{\textless{}{-}} \DecValTok{3}
-\NormalTok{gene\_sample }\OtherTok{\textless{}{-}} \FunctionTok{sample}\NormalTok{(example\_genes[}\DecValTok{1}\SpecialCharTok{:}\DecValTok{1000}\NormalTok{], num\_genes)}
-
-\FunctionTok{ggpairs}\NormalTok{(}\AttributeTok{data =} \FunctionTok{as.data.frame}\NormalTok{(sim\_nbinom[, gene\_sample]),}
-        \AttributeTok{upper =} \FunctionTok{list}\NormalTok{(}\AttributeTok{continuous =} \FunctionTok{wrap}\NormalTok{(}\StringTok{\textquotesingle{}cor\textquotesingle{}}\NormalTok{, }\AttributeTok{method =} \StringTok{"spearman"}\NormalTok{))) }\SpecialCharTok{+}
-  \FunctionTok{theme\_bw}\NormalTok{()}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
-## exact p-value with ties
-
-## Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
-## exact p-value with ties
-
-## Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
-## exact p-value with ties
-\end{verbatim}
-
 \begin{figure}
@@
 \end{figure}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{include\_graphics}\NormalTok{(}\StringTok{\textquotesingle{}../fig/ch050{-}figBRCA.png\textquotesingle{}}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
 

### 2021-10-19 arXiv_bigsimr/arXiv_bigsimr.Rmd
Working on cross references arXiv

@@
 This gives a great number of pairwise dependencies among the marginals (specifically, $r choose(d,2)$ correlation parameters).
-Table \@ref(tab:ch010-realDataTab) displays RNA-seq counts for three selected high-expressing genes for the first five patients' breast tumor samples.
-To help visualize the bivariate relationships for these three selected genes across all patients, Figure \@ref(fig:ch010-realDataFig) displays the marginal distributions and estimated Spearman's correlations.
+Table \ref{tab:ch010-realDataTab} displays RNA-seq counts for three selected high-expressing genes for the first five patients' breast tumor samples.
+To help visualize the bivariate relationships for these three selected genes across all patients, Figure \ref{fig:ch010-realDataFig} displays the marginal distributions and estimated Spearman's correlations.
 
@@
 
-```{r ch010-realDataFig, cache=FALSE, fig.width=6, fig.cap="Pairwise scatterplots for three example genes (Table 1), with estimated Spearman's correlations and marginal density estimates displayed.
+```{r ch010-realDataFig, cache=FALSE, fig.width=6, fig.cap="\\label{fig:ch010-realDataFig}Pairwise scatterplots for three example genes (Table 1), with estimated Spearman's correlations and marginal density estimates displayed.
 The data possess outliers, heavy-right tails, are discrete, and have non-trivial intergene correlations.
@@
 Finally, we know of no other `R` package that can produce HD random vectors with flexible margins and dependence structure.
-Table \@ref(tab:compare-table) summarizes the above discussion.
+Table \ref{tab:compare-table} summarizes the above discussion.
 
@@
 To characterize dependency, every pair $(Y_i, Y_j)$ has a given Pearson correlation $\rho_P$, Spearman correlation $\rho_S$, and/or Kendall's $\tau$.
-The method can be described as a *high-performance Gaussian copula* (Equation \@ref(eq:gauss)) providing a HD NORTA-inspired algorithm.
+The method can be described as a *high-performance Gaussian copula* (Equation \ref{eq:gauss}) providing a HD NORTA-inspired algorithm.
 
@@
 
-*Step 1: Mapping step.* We first employ the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertKendall) and \@ref(eq:convertSpearman), respectively (implemented as `cor_covert()`).
+*Step 1: Mapping step.* We first employ the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \ref{eq:convertKendall} and \ref{eq:convertSpearman}, respectively (implemented as `cor_covert()`).
 Initializing our algorithm to match the nonparametric correlations by computing these equations for all pairs is computationally trivial.
 
-For the computationally expensive process to match Pearson correlations, we approximate Equation \@ref(eq:pearsonIntegralRelation) for all pairs of margins.
+For the computationally expensive process to match Pearson correlations, we approximate Equation \ref{eq:pearsonIntegralRelation} for all pairs of margins.
 To this end, we implement the approximation scheme introduced by [@XZ19].
-Briefly, the many double integrals of the form in \@ref(eq:pearsonIntegralRelation) are approximated by weighted sums of Hermite polynomials.
+Briefly, the many double integrals of the form in \ref{eq:pearsonIntegralRelation} are approximated by weighted sums of Hermite polynomials.
 Matching coefficients for pairs of continuous distributions is made tractable by this method, but for discrete distributions (especially discrete distributions with large support sets or infinite support), the approximation is computationally expensive.
@@
 In our experience, the mapping step for large $d$ almost always produces a $R_X$ that is not a correlation matrix.
-In Section \@ref(package), we provide a basic method of how to quantify and control the approximation error.
-Further, the RNA-seq example in Section \@ref(examples) provides an illustration of this in practice.
+In Section \ref{package}, we provide a basic method of how to quantify and control the approximation error.
+Further, the RNA-seq example in Section \ref{examples} provides an illustration of this in practice.
 
@@
 In addition to the native Julia `Bigsimr` package and `R` interface `bigsimr`, we also provide a Python interface `bigsimr` that interfaces with the Julia `Bigsimr` package.
-The Julia package provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms)).
+The Julia package provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \ref{algorithms}).
 
@@
 
-Figure \@ref(fig:ch030-aq-joint-dist) visualizes the bivariate relationship between `Ozone` and `Temperature`.
+Figure \ref{fig:ch030-aq-joint-dist} visualizes the bivariate relationship between `Ozone` and `Temperature`.
 We aim to simulate random two-component vectors mimicking this structure.
@@
 
-```{r ch030-aq-joint-dist, cache=FALSE, fig.width=6, fig.cap="Bivariate scatterplot of Ozone versus Temp with estimated marginal densities.
+```{r ch030-aq-joint-dist, cache=FALSE, fig.width=6, fig.cap="\\label{fig:ch030-aq-joint-dist}Bivariate scatterplot of Ozone versus Temp with estimated marginal densities.
 The Ozone data are modeled marginally as log-normal and the Temperature data as normal.", eval=T, echo=F}
@@
 
-Based on the estimated densities in Figure \@ref(fig:ch030-aq-joint-dist), we assume `Temp` is normally distributed and `Ozone` is log-normally distributed, as the latter values are positive and skewed.
+Based on the estimated densities in Figure \ref{fig:ch030-aq-joint-dist}, we assume `Temp` is normally distributed and `Ozone` is log-normally distributed, as the latter values are positive and skewed.
 We use the classical unbiased estimators for the normal distribution's parameters and maximum likelihood estimators for the log-normal parameters:
@@
 Note that the target input is specified using the pre-computed MVN correlation matrix.
-Figure \@ref(fig:ch030-plot-sim) plots the 10,000 simulated points.
+Figure \ref{fig:ch030-plot-sim} plots the 10,000 simulated points.
 
@@
 
-```{r ch030-plot-sim, fig.cap="Contour plot and marginal densities for the simulated bivariate distribution of Air Quality Temperatures and Ozone levels.
+```{r ch030-plot-sim, fig.cap="\\label{fig:ch030-plot-sim}Contour plot and marginal densities for the simulated bivariate distribution of Air Quality Temperatures and Ozone levels.
 The simulated points mimic the observed data with respect to both the marginal characteristics and bivariate association.", eval=T, echo=F, warning=F}
@@
 
-Table \@ref(tab:sims) lists our identical-marginal, bivariate simulation configurations.
+Table \ref{tab:sims} lists our identical-marginal, bivariate simulation configurations.
 We increase the simulation replicates $B$ to visually ensure that our results converge to the target correlations and gauge efficiency.
@@
 
-```{r ch040-biNormPlot, eval=T, fig.height=5, fig.width=6, fig.cap="`bigsimr` recovers the Pearson specified correlations for MVN.", echo=F}
+```{r ch040-biNormPlot, eval=T, fig.height=5, fig.width=6, fig.cap="\\label{fig:ch040-biNormPlot}`bigsimr` recovers the Pearson specified correlations for MVN.", echo=F}
 bivariate_normal_sims %>%
@@
 
-```{r ch040-biGammaPlot, eval=T, fig.height=5, fig.width=6, fig.cap="`bigsimr` recovers the Pearson specified correlations for Bivariate Gamma.", echo=F}
+```{r ch040-biGammaPlot, eval=T, fig.height=5, fig.width=6, fig.cap="\\label{fig:ch040-biGammaPlot}`bigsimr` recovers the Pearson specified correlations for Bivariate Gamma.", echo=F}
 bivariate_gamma_sims %>%
@@
 
-```{r ch040-biNegBinPlot, eval=T, fig.width=6, fig.height=5, fig.cap="`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations.", echo=F}
+```{r ch040-biNegBinPlot, eval=T, fig.width=6, fig.height=5, fig.cap="\\label{fig:ch040-biNegBinPlot}`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations.", echo=F}
 bivariate_nbinom_sims %>%
@@
 
-Figure \@ref(fig:ch040-bPlot) displays the aggregated bivariate simulation results.
-Table \@ref(tab:ch040-BiError) contains the mean absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.
+Figure \ref{fig:ch040-bPlot} displays the aggregated bivariate simulation results.
+Table \ref{tab:ch040-BiError} contains the mean absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.
 Overall, the studies show that our methodology is generally accurate across the entire range of possible correlation for all three dependency measures.
@@
 
-```{r ch040-bPlot, cache=FALSE, echo=F, fig.asp=1.30, fig.weight=4, fig.cap="Bivariate simulations match target correlations across the entire range of feasible correlations.
+```{r ch040-bPlot, cache=FALSE, echo=F, fig.asp=1.30, fig.weight=4, fig.cap="\\label{fig:ch040-bPlot}Bivariate simulations match target correlations across the entire range of feasible correlations.
 The horizontal axis plots the specified target correlations for each bivariate margin.
@@
 The experiments are conducted on a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, with all 16 threads employed during computation.
-Table \@ref(tab:ch040-moderateDtab) displays the total computation time for moderate dimesions ($d \leq 500$).
+Table \ref{tab:ch040-moderateDtab} displays the total computation time for moderate dimesions ($d \leq 500$).
 In every simullation setting, the 1,000 vectors are generated rapidly, executing in under 2 seconds.
@@
 The results for $d > 500$ show scalability to ultra-high dimensions for all three correlation types, although the total times do become much larger. 
-Figure \@ref(fig:ch040-largeDfig) displays computation times for $d=\{1000, 2500, 5000, 10000\}$.
+Figure \ref{fig:ch040-largeDfig} displays computation times for $d=\{1000, 2500, 5000, 10000\}$.
 For $d$ equal to 1000 and 2500, the total time is under a couple of minutes.
@@
 
-```{r ch040-largeDfig, fig.cap="Computation times for HD multivariate gamma simulation.", eval=T, echo=F}
+```{r ch040-largeDfig, fig.cap="\\label{fig:ch040-largeDfig}Computation times for HD multivariate gamma simulation.", eval=T, echo=F}
 # Set colors
@@
 
-Figure \@ref(fig:ch050-simDataFig) displays the simulated counts and pairwise relationships for our example genes from Table \@ref(tab:ch010-realDataTab).
-Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parametric form and with less extreme points than the observed data in Figure \@ref(fig:ch010-realDataFig).
-Figure \@ref(fig:ch050-figBRCA) compares the specified target parameter (horizontal axis) with the corresponding quantities estimated from the simulated data (vertical axis).
+Figure \ref{fig:ch050-simDataFig} displays the simulated counts and pairwise relationships for our example genes from Table \ref{tab:ch010-realDataTab}.
+Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parametric form and with less extreme points than the observed data in Figure \ref{fig:ch010-realDataFig}.
+Figure \ref{fig:ch050-figBRCA} compares the specified target parameter (horizontal axis) with the corresponding quantities estimated from the simulated data (vertical axis).
 The evaluation shows that the simulated counts approximately match the target parameters and exhibit the full range of estimated correlations from the data.
 
-```{r ch050-simDataFig, cache=FALSE, fig.cap="Simulated data for three selected high-expressing genes generally replicates the estimated data structure.
+```{r ch050-simDataFig, cache=FALSE, fig.cap="\\label{fig:ch050-simDataFig}Simulated data for three selected high-expressing genes generally replicates the estimated data structure.
 The data do not exhibit outlying points, but do possess the desired Spearman correlations, central tendencies, and discrete values.", eval=T, warning=F, echo=F}
@@
 
-```{r ch050-figBRCA, out.width='80%', fig.cap="Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set.
+```{r ch050-figBRCA, out.width='80%', fig.cap="\\label{fig:ch050-figBRCA}Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set.
 The dashed red lines indicate equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes).", eval=T, echo=F}

### 2021-10-19 arXiv_bigsimr/arXiv_bigsimr.tex
Working on cross references arXiv

@@
 the marginals (specifically, \(r choose(d,2)\) correlation parameters).
-Table @ref(tab:ch010-realDataTab) displays RNA-seq counts for three
+Table \ref{tab:ch010-realDataTab} displays RNA-seq counts for three
 selected high-expressing genes for the first five patients' breast tumor
 samples. To help visualize the bivariate relationships for these three
-selected genes across all patients, Figure @ref(fig:ch010-realDataFig)
+selected genes across all patients, Figure \ref{fig:ch010-realDataFig}
 displays the marginal distributions and estimated Spearman's
@@
 \includegraphics{arXiv_bigsimr_files/figure-latex/ch010-realDataFig-1.pdf}
-\caption{Pairwise scatterplots for three example genes (Table 1), with
-estimated Spearman's correlations and marginal density estimates
-displayed. The data possess outliers, heavy-right tails, are discrete,
-and have non-trivial intergene correlations. Modeling these data
-motivate our simulation methodology.}
+\caption{\label{fig:ch010-realDataFig}Pairwise scatterplots for three
+example genes (Table 1), with estimated Spearman's correlations and
+marginal density estimates displayed. The data possess outliers,
+heavy-right tails, are discrete, and have non-trivial intergene
+correlations. Modeling these data motivate our simulation methodology.}
 \end{figure}
@@
 other \texttt{R} package that can produce HD random vectors with
-flexible margins and dependence structure. Table @ref(tab:compare-table)
+flexible margins and dependence structure. Table \ref{tab:compare-table}
 summarizes the above discussion.
@@
 Kendall's \(\tau\). The method can be described as a
-\emph{high-performance Gaussian copula} (Equation @ref(eq:gauss))
+\emph{high-performance Gaussian copula} (Equation \ref{eq:gauss})
 providing a HD NORTA-inspired algorithm.
@@
 relationships between \(\rho_S\) and \(\tau\) with \(\rho_P\) for
-bivariate normal random variables via Equations @ref(eq:convertKendall)
-and @ref(eq:convertSpearman), respectively (implemented as
+bivariate normal random variables via Equations \ref{eq:convertKendall}
+and \ref{eq:convertSpearman}, respectively (implemented as
 \texttt{cor\_covert()}). Initializing our algorithm to match the
@@
 For the computationally expensive process to match Pearson correlations,
-we approximate Equation @ref(eq:pearsonIntegralRelation) for all pairs
+we approximate Equation \ref{eq:pearsonIntegralRelation} for all pairs
 of margins. To this end, we implement the approximation scheme
 introduced by (Xiao and Zhou 2019). Briefly, the many double integrals
-of the form in @ref(eq:pearsonIntegralRelation) are approximated by
+of the form in \ref{eq:pearsonIntegralRelation} are approximated by
 weighted sums of Hermite polynomials. Matching coefficients for pairs of
@@
 mapping step for large \(d\) almost always produces a \(R_X\) that is
-not a correlation matrix. In Section @ref(package), we provide a basic
+not a correlation matrix. In Section \ref{package}, we provide a basic
 method of how to quantify and control the approximation error. Further,
-the RNA-seq example in Section @ref(examples) provides an illustration
+the RNA-seq example in Section \ref{examples} provides an illustration
 of this in practice.
@@
 high-performance implementation of our proposed random vector generation
-algorithm and associated functions (see Section @ref(algorithms)).
+algorithm and associated functions (see Section \ref{algorithms}).
 
@@
 
-Figure @ref(fig:ch030-aq-joint-dist) visualizes the bivariate
+Figure \ref{fig:ch030-aq-joint-dist} visualizes the bivariate
 relationship between \texttt{Ozone} and \texttt{Temperature}. We aim to
@@
 \includegraphics{arXiv_bigsimr_files/figure-latex/ch030-aq-joint-dist-1.pdf}
-\caption{Bivariate scatterplot of Ozone versus Temp with estimated
-marginal densities. The Ozone data are modeled marginally as log-normal
-and the Temperature data as normal.}
+\caption{\label{fig:ch030-aq-joint-dist}Bivariate scatterplot of Ozone
+versus Temp with estimated marginal densities. The Ozone data are
+modeled marginally as log-normal and the Temperature data as normal.}
 \end{figure}
@@
 Based on the estimated densities in Figure
-@ref(fig:ch030-aq-joint-dist), we assume \texttt{Temp} is normally
+\ref{fig:ch030-aq-joint-dist}, we assume \texttt{Temp} is normally
 distributed and \texttt{Ozone} is log-normally distributed, as the
@@
 ## Julia Object of type NamedTuple{(:lower, :upper),Tuple{Float64,Float64}}.
-## (lower = -0.8816136274594583, upper = 0.8815214379668451)
+## (lower = -0.8807315594247829, upper = 0.880286004475508)
 \end{verbatim}
@@
 and \texttt{Temp}. Note that the target input is specified using the
-pre-computed MVN correlation matrix. Figure @ref(fig:ch030-plot-sim)
+pre-computed MVN correlation matrix. Figure \ref{fig:ch030-plot-sim}
 plots the 10,000 simulated points.
@@
 \includegraphics{arXiv_bigsimr_files/figure-latex/ch030-plot-sim-1.pdf}
-\caption{Contour plot and marginal densities for the simulated bivariate
-distribution of Air Quality Temperatures and Ozone levels. The simulated
-points mimic the observed data with respect to both the marginal
-characteristics and bivariate association.}
+\caption{\label{fig:ch030-plot-sim}Contour plot and marginal densities
+for the simulated bivariate distribution of Air Quality Temperatures and
+Ozone levels. The simulated points mimic the observed data with respect
+to both the marginal characteristics and bivariate association.}
 \end{figure}
@@
 
-Table @ref(tab:sims) lists our identical-marginal, bivariate simulation
+Table \ref{tab:sims} lists our identical-marginal, bivariate simulation
 configurations. We increase the simulation replicates \(B\) to visually
@@
 \includegraphics{arXiv_bigsimr_files/figure-latex/ch040-biNormPlot-1.pdf}
-\caption{\texttt{bigsimr} recovers the Pearson specified correlations
-for MVN.}
+\caption{\label{fig:ch040-biNormPlot}\texttt{bigsimr} recovers the
+Pearson specified correlations for MVN.}
 \end{figure}
@@
 \includegraphics{arXiv_bigsimr_files/figure-latex/ch040-biGammaPlot-1.pdf}
-\caption{\texttt{bigsimr} recovers the Pearson specified correlations
-for Bivariate Gamma.}
+\caption{\label{fig:ch040-biGammaPlot}\texttt{bigsimr} recovers the
+Pearson specified correlations for Bivariate Gamma.}
 \end{figure}
@@
 \includegraphics{arXiv_bigsimr_files/figure-latex/ch040-biNegBinPlot-1.pdf}
-\caption{\texttt{bigsimr} recovers the correlations for bivariate
-negative binomial only approximately for Pearson but (nearly) exactly
-for the rank-based correlations.}
+\caption{\label{fig:ch040-biNegBinPlot}\texttt{bigsimr} recovers the
+correlations for bivariate negative binomial only approximately for
+Pearson but (nearly) exactly for the rank-based correlations.}
 \end{figure}
 
-Figure @ref(fig:ch040-bPlot) displays the aggregated bivariate
-simulation results. Table @ref(tab:ch040-BiError) contains the mean
+Figure \ref{fig:ch040-bPlot} displays the aggregated bivariate
+simulation results. Table \ref{tab:ch040-BiError} contains the mean
 absolute error (MAE) in reproducing the desired dependency measures for
@@
 \includegraphics{arXiv_bigsimr_files/figure-latex/ch040-bPlot-1.pdf}
-\caption{Bivariate simulations match target correlations across the
-entire range of feasible correlations. The horizontal axis plots the
-specified target correlations for each bivariate margin. Normal margins
-are plotted in dark dark grey, gamma in medium grey, and negative
-binomial in light grey. As the number of simulated vectors \(B\)
-increases from left to right, the variation in estimated correlations
-(vertical axis) decreases. The dashed line indicates equality between
-the specified and estimated correlations.}
+\caption{\label{fig:ch040-bPlot}Bivariate simulations match target
+correlations across the entire range of feasible correlations. The
+horizontal axis plots the specified target correlations for each
+bivariate margin. Normal margins are plotted in dark dark grey, gamma in
+medium grey, and negative binomial in light grey. As the number of
+simulated vectors \(B\) increases from left to right, the variation in
+estimated correlations (vertical axis) decreases. The dashed line
+indicates equality between the specified and estimated correlations.}
 \end{figure}
@@
 Intel Core i9 processor, with all 16 threads employed during
-computation. Table @ref(tab:ch040-moderateDtab) displays the total
+computation. Table \ref{tab:ch040-moderateDtab} displays the total
 computation time for moderate dimesions (\(d \leq 500\)). In every
@@
 for all three correlation types, although the total times do become much
-larger. Figure @ref(fig:ch040-largeDfig) displays computation times for
+larger. Figure \ref{fig:ch040-largeDfig} displays computation times for
 \(d=\{1000, 2500, 5000, 10000\}\). For \(d\) equal to 1000 and 2500, the
@@
 \includegraphics{arXiv_bigsimr_files/figure-latex/ch040-largeDfig-1.pdf}
-\caption{Computation times for HD multivariate gamma simulation.}
+\caption{\label{fig:ch040-largeDfig}Computation times for HD
+multivariate gamma simulation.}
 \end{figure}
@@
 
-Figure @ref(fig:ch050-simDataFig) displays the simulated counts and
+Figure \ref{fig:ch050-simDataFig} displays the simulated counts and
 pairwise relationships for our example genes from Table
-@ref(tab:ch010-realDataTab). Simulated counts roughly mimic the observed
+\ref{tab:ch010-realDataTab}. Simulated counts roughly mimic the observed
 data but with a smoother appearance due to the assumed parametric form
 and with less extreme points than the observed data in Figure
-@ref(fig:ch010-realDataFig). Figure @ref(fig:ch050-figBRCA) compares the
+\ref{fig:ch010-realDataFig}. Figure \ref{fig:ch050-figBRCA} compares the
 specified target parameter (horizontal axis) with the corresponding
@@
 \includegraphics{arXiv_bigsimr_files/figure-latex/ch050-simDataFig-1.pdf}
-\caption{Simulated data for three selected high-expressing genes
-generally replicates the estimated data structure. The data do not
-exhibit outlying points, but do possess the desired Spearman
-correlations, central tendencies, and discrete values.}
+\caption{\label{fig:ch050-simDataFig}Simulated data for three selected
+high-expressing genes generally replicates the estimated data structure.
+The data do not exhibit outlying points, but do possess the desired
+Spearman correlations, central tendencies, and discrete values.}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=0.8\linewidth]{../fig/ch050-figBRCA} \caption{Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set.
+\includegraphics[width=0.8\linewidth]{../fig/ch050-figBRCA} \caption{\label{fig:ch050-figBRCA}Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set.
 The dashed red lines indicate equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes).}\label{fig:ch050-figBRCA}

### 2022-01-07 arXiv_bigsimr/arXiv_bigsimr.Rmd
repo arXiv docs

@@
     affiliation: University of Nevada, Reno
-    location: Reno, NV 89557
-    email: grant.schissler@gmail.com
   - name: Edward J. Bedrick
+    department: Epidemiology and Biostatistics Department
     affiliation: University of Arizona
   - name: Alexander D. Knudson
-    affiliation: University of Nevada, Reno \AND
+    department: Department of Mathematics \& Statistics
+    affiliation: University of Nevada, Reno
   - name: Tomasz J. Kozubowski
+    department: Department of Mathematics \& Statistics
     affiliation: University of Nevada, Reno
   - name: Tin Nguyen
+    department: Department of Computer Science \& Engineering
     affiliation: University of Nevada, Reno
   - name: Anna K. Panorska
-    affiliation: University of Nevada, Reno \AND
+    department: Department of Mathematics \& Statistics
+    affiliation: University of Nevada, Reno
   - name: Juli Petereit
+    department: Nevada Bioinformatics Center
     affiliation: University of Nevada, Reno
   - name: Walter W. Piegorsch
+    department: Mathematics Department
     affiliation: University of Arizona
   - name: Duc Tran
+    department: Department of Computer Science \& Engineering
     affiliation: University of Nevada, Reno

### 2022-01-07 arXiv_bigsimr/arXiv_bigsimr_submission.tex
repo arXiv docs

--- 2021-10-19 arXiv_bigsimr/arXiv_bigsimr.tex
@@
     University of Nevada, Reno \\
-  Reno, NV 89557 \\
-  \texttt{\href{mailto:grant.schissler@gmail.com}{\nolinkurl{grant.schissler@gmail.com}}} \\
+   \\
+  \texttt{} \\
    \And
@@
    \\
+    Epidemiology and Biostatistics Department \\
     University of Arizona \\
@@
    \\
-    University of Nevada, Reno \AND \\
+    Department of Mathematics \& Statistics \\
+    University of Nevada, Reno \\
    \\
@@
    \\
+    Department of Mathematics \& Statistics \\
     University of Nevada, Reno \\
@@
    \\
+    Department of Computer Science \& Engineering \\
     University of Nevada, Reno \\
@@
    \\
+    Nevada Bioinformatics Center \\
     University of Nevada, Reno \\
@@
    \\
+    Mathematics Department \\
     University of Arizona \\
@@
 
+© A.G.
+Schissler.
+This is an Open Access article distributed under the terms of the Creative Commons License Attribution-NonCommercial-ShareAlike 4.0 International (CC BY-NC-SA 4.0) (https://creativecommons.org/licenses/by-nc-sa/4.0/), which permits reusers to distribute, remix, adapt, and build upon the material in any medium or format for noncommercial purposes only, and only so long as attribution is given to the creator.
+If you remix, adapt, or build upon the material, you must license the modified material under identical terms.
+
+\clearpage
+
 \begin{abstract}
@@
   }
-
-\clearpage
 
@@
 ## Julia Object of type NamedTuple{(:lower, :upper),Tuple{Float64,Float64}}.
-## (lower = -0.8807315594247829, upper = 0.880286004475508)
+## (lower = -0.8829854580269612, upper = 0.882607249144551)
 \end{verbatim}

### 2022-01-07 arXiv_bigsimr/arXiv_bigsimr_submission/arXiv_bigsimr_submission.tex
repo arXiv docs

--- 2022-01-07 arXiv_bigsimr/arXiv_bigsimr_submission.tex
@@
 \centering
-\includegraphics{arXiv_bigsimr_files/figure-latex/ch010-realDataFig-1.pdf}
+\includegraphics{ch010-realDataFig-1.pdf}
 \caption{\label{fig:ch010-realDataFig}Pairwise scatterplots for three
@@
 \centering
-\includegraphics{arXiv_bigsimr_files/figure-latex/ch030-aq-joint-dist-1.pdf}
+\includegraphics{ch030-aq-joint-dist-1.pdf}
 \caption{\label{fig:ch030-aq-joint-dist}Bivariate scatterplot of Ozone
@@
 \centering
-\includegraphics{arXiv_bigsimr_files/figure-latex/ch030-plot-sim-1.pdf}
+\includegraphics{ch030-plot-sim-1.pdf}
 \caption{\label{fig:ch030-plot-sim}Contour plot and marginal densities
@@
 \centering
-\includegraphics{arXiv_bigsimr_files/figure-latex/ch040-biNormPlot-1.pdf}
+\includegraphics{ch040-biNormPlot-1.pdf}
 \caption{\label{fig:ch040-biNormPlot}\texttt{bigsimr} recovers the
@@
 \centering
-\includegraphics{arXiv_bigsimr_files/figure-latex/ch040-biGammaPlot-1.pdf}
+\includegraphics{ch040-biGammaPlot-1.pdf}
 \caption{\label{fig:ch040-biGammaPlot}\texttt{bigsimr} recovers the
@@
 \centering
-\includegraphics{arXiv_bigsimr_files/figure-latex/ch040-biNegBinPlot-1.pdf}
+\includegraphics{ch040-biNegBinPlot-1.pdf}
 \caption{\label{fig:ch040-biNegBinPlot}\texttt{bigsimr} recovers the
@@
 \centering
-\includegraphics{arXiv_bigsimr_files/figure-latex/ch040-bPlot-1.pdf}
+\includegraphics{ch040-bPlot-1.pdf}
 \caption{\label{fig:ch040-bPlot}Bivariate simulations match target
@@
 \centering
-\includegraphics{arXiv_bigsimr_files/figure-latex/ch040-largeDfig-1.pdf}
+\includegraphics{ch040-largeDfig-1.pdf}
 \caption{\label{fig:ch040-largeDfig}Computation times for HD
@@
 \centering
-\includegraphics{arXiv_bigsimr_files/figure-latex/ch050-simDataFig-1.pdf}
+\includegraphics{ch050-simDataFig-1.pdf}
 \caption{\label{fig:ch050-simDataFig}Simulated data for three selected
@@
 \begin{figure}
-\includegraphics[width=0.8\linewidth]{../fig/ch050-figBRCA} \caption{\label{fig:ch050-figBRCA}Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set.
+\includegraphics[width=0.8\linewidth]{ch050-figBRCA} \caption{\label{fig:ch050-figBRCA}Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set.
 The dashed red lines indicate equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes).}\label{fig:ch050-figBRCA}
